\begin{tabular}{lrrrrrl}
\toprule
 Outcome & $\hat{\beta}$ & SE & SD($Y$) & SDE & SE(SDE) & Classification \\
 \midrule
 \multicolumn{7}{l}{\textit{Panel A: Pooled (binary treatment)}} \\
Saber 11 Global (binary) & 0.525 & 3.111 & 20.438 & 0.026 & 0.152 & Small positive \\
Saber 11 Math (binary) & 0.312 & 0.707 & 4.877 & 0.064 & 0.145 & Moderate positive \\
Saber 11 Reading (binary) & 0.102 & 0.554 & 3.647 & 0.028 & 0.152 & Small positive \\
\midrule
 \multicolumn{7}{l}{\textit{Panel B: Heterogeneity by municipality size}} \\
Global, small munis & -5.723 & 5.104 & 20.438 & -0.280 & 0.250 & Large negative \\
Global, large munis & 4.152 & 3.845 & 20.438 & 0.203 & 0.188 & Large positive \\
\bottomrule
 \end{tabular}
 \begin{tablenotes}
 \item \textit{Notes:}  \textbf{Country:} Colombia.  \textbf{Research question:} Does electing a mayor whose 2019 campaign donors went on to receive municipal-government contracts under the SECOP II procurement system degrade local public-school quality, as measured by Saber 11 standardized test scores?  \textbf{Policy mechanism:} Colombian municipal mayors control roughly 70--80\% of local procurement, much of it via direct-award contracts; if elected mayors steer contracts toward people who financed their campaigns, the resulting allocation may shift resources away from public services and toward rent-seeking.  \textbf{Outcome definition:} Municipality-year mean of the ICFES Saber 11 \textit{punt\_global} score (0--500 scale), aggregated from individual student records weighted by enrollment, with parallel estimates for math and critical-reading subscores.  \textbf{Treatment:} Binary indicator equal to one if at least one cedula appearing in the elected mayor's 2019 Cuentas Claras donor list also appears as a contractor in SECOP II 2020--2022; continuous version uses the share of total donor value contributed by such individuals.  \textbf{Data:} Datos Abiertos Colombia Socrata APIs --- Cuentas Claras 2019 (jgra-rz2t, 12{,}138 individual donor records), Elected Mayors 2019 (h236-q58p, 1{,}100 mayors), SECOP II Contratos Electronicos 2020--2022 (jbjy-vk9h, 879k contracts), ICFES Saber 11 2013--2022 (kgxf-xxbe, aggregated to 8{,}162 muni-period cells); 948 municipalities, 9 cohort years, 6{,}138 muni-year observations.  \textbf{Method:} Two-way fixed-effects (municipality + cohort year) DiD with standard errors clustered at the municipality level; event-study pre-trends test passes (max p $>$ 0.10 in pre-period).  \textbf{Sample:} The 948 of 1{,}100 elected mayors whose campaign donors could be matched to the Cuentas Claras file via fuzzy name matching within municipality (Jaccard $\geq$ 0.4).  SDE $= \hat{\beta} / \text{SD}(Y)$ with SD($Y$) computed in the pre-treatment cross-section (2019).  Classification refers to magnitude, not statistical significance: Large ($|$SDE$| > 0.15$), Moderate ($0.05$--$0.15$), Small ($0.005$--$0.05$), Null ($< 0.005$).
 \end{tablenotes}
