\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Growth Illusion: Romania's Fifteen-Fold Micro-Enterprise Expansion Created No New Firms}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Romania expanded its micro-enterprise turnover tax threshold fifteen-fold---from EUR 65,000 to EUR 1,000,000---between 2013 and 2018, slashing effective tax rates for hundreds of thousands of firms. Using Eurostat Structural Business Statistics for eleven Central and Eastern European countries (2008--2020), I find no effect on firm creation: micro-enterprise counts evolved identically to peer countries. Average turnover per firm rose 6\%, but uniformly across all size classes, suggesting economic catch-up rather than a policy response. A placebo test confirms no differential pre-trend. Simplified tax regimes---even dramatic expansions---may be capitalized into the intensive margin rather than stimulating entry.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H25, H32, L11, O23 \\
\noindent\textbf{Keywords:} micro-enterprise taxation, firm creation, turnover tax, Romania, difference-in-differences

\newpage

\section{Introduction}

Governments worldwide maintain simplified tax regimes for small firms, betting that lower compliance costs and reduced tax burdens will encourage entrepreneurship. Romania made the most aggressive bet in the European Union: between 2013 and 2018, it raised the eligibility ceiling for its micro-enterprise turnover tax from EUR~65,000 to EUR~1,000,000---a fifteen-fold expansion that brought roughly 90\% of all Romanian firms under a regime taxing revenue at 1--3\% instead of profits at 16\%. The implicit subsidy was enormous. Did it work?

This paper exploits Romania's staggered threshold expansions in a cross-country difference-in-differences framework, comparing Romania to ten Central and Eastern European peer countries using sector-level data from Eurostat's Structural Business Statistics (2008--2020). I study two margins: the \textit{extensive margin} (did more firms form?) and the \textit{intensive margin} (did existing firms grow?).

The answer is striking in its asymmetry. On the extensive margin, the expansion had no detectable effect. Romania's micro-enterprise count---firms with 0--9 employees---evolved on the same trajectory as its CEE peers throughout the entire reform period. The difference-in-differences estimate is $-0.089$ log points (SE $= 0.069$), statistically indistinguishable from zero and negative in sign. The micro-enterprise \textit{share} of all firms was essentially unchanged ($-0.03$ percentage points, SE $= 0.64$ pp). A placebo test placing the treatment at 2013---four years before the major expansion---produces a coefficient of similar magnitude ($-0.187$), confirming that the main estimate captures no systematic break.

On the intensive margin, average turnover per firm in the 0--9 employee class rose by approximately 6\% relative to peers (SE $= 3.0$\%, $p = 0.068$). But this finding dissolves under scrutiny: the same increase appears across \textit{all} size classes, from the smallest to the largest firms (10\% for firms with 250+ employees, $p = 0.004$). Since the micro-enterprise regime does not apply to large firms, this pervasive pattern reflects Romania's general economic outperformance relative to peers during 2017--2020, not a policy-specific response to the threshold change.

These results contribute to the literature on small-firm tax incentives. A large body of work documents behavioral responses to firm-size-dependent taxation---bunching below thresholds \citep{saez2010bunching, kleven2013using, best2015production}, evasion and informality \citep{gorodnichenko2009myth, bachas2019size}, and misallocation \citep{guner2008macroeconomic, garicano2016firm}. Most of this literature takes the existence of responsive firms as given and estimates the elasticity of their response. By contrast, this paper asks whether expanding the set of eligible firms \textit{creates new ones}---and finds that it does not.

The null is informative because Romania's experiment was uniquely large. Prior work on firm-size thresholds typically studies variation of 10--50\% around a fixed cutoff. Romania's threshold moved by 1,400\%, from EUR~65,000 to EUR~1,000,000, in just five years. If a simplified tax regime were ever going to stimulate entry, this was the place to find it. That it did not suggests the binding constraints on firm creation lie elsewhere---in regulatory barriers, credit access, human capital, or market demand---rather than in the tax code.

This paper also relates to the literature on tax competition and Eastern European flat-tax experiments. Romania itself adopted a 16\% flat corporate tax in 2005, and its micro-enterprise regime represents a second layer of simplification targeted at the smallest firms \citep{keen2008flat}. The finding that even dramatic rate reductions fail to stimulate entry echoes \citet{djankov2010effect}, who find that effective corporate tax rates affect formal business density but with elasticities too small to explain cross-country variation. My results suggest that the relevant margin for tax-induced entry is even smaller than their estimates imply, at least in the Eastern European context.

A natural approach would be to estimate revenue bunching at each threshold using firm-level data from Romania's tax authority (ANAF). While ANAF publishes annual financial statement data, these microdata are not available through programmatic access, precluding the bunching design envisioned in the prior literature. Instead, I exploit the cross-country variation in Eurostat's Structural Business Statistics, which provides enterprise counts and turnover by employee size class for all EU member states. This aggregate approach sacrifices the precision of firm-level bunching estimation but gains a credible counterfactual through comparison with peer economies that did not undertake similar reforms.

The empirical strategy exploits Romania's threshold changes as treatment events in a panel of 11 countries $\times$ 9 sectors $\times$ 13 years. The identifying assumption is that Romania's micro-enterprise sector would have evolved like its CEE peers absent the reform. I test this assumption with event-study specifications, placebo treatments, alternative control groups (narrow CEE neighbors versus a broader set including Western Europe), and sector-level heterogeneity tests. No specification produces a significant positive effect on firm counts.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes Romania's micro-enterprise regime and the sequence of threshold changes. \Cref{sec:data} presents the Eurostat data and summary statistics. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} reports the main findings and robustness checks. \Cref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

Romania introduced its micro-enterprise turnover tax (\textit{impozit pe veniturile microîntreprinderilor}) in 2001 as an alternative to the standard 16\% corporate income tax (CIT). Under this regime, eligible firms pay tax on gross revenue rather than profit, at rates between 1\% and 3\% depending on the number of employees and the fiscal year. The regime is designed to reduce compliance costs for small firms by eliminating the need for profit accounting.

Eligibility is determined by a turnover ceiling that has changed dramatically over time. From 2001 to 2015, the threshold was set at relatively modest levels: EUR~100,000 in the early years, reduced to EUR~65,000 in 2013. Beginning in 2016, Romania embarked on a series of expansions that would transform the regime from a niche small-business provision into the dominant tax structure for the vast majority of Romanian firms.

The key reforms occurred in rapid succession. In 2016, the threshold was raised to EUR~100,000 and the tax rate was reduced to 1\% for firms with at least one employee (3\% for firms without employees). In 2017, the ceiling jumped to EUR~500,000. In 2018, it reached EUR~1,000,000---fifteen times the 2013 level. By 2020, the regime was made mandatory for all eligible firms, meaning that firms below the threshold could no longer opt into the standard CIT regime.

The scale of this expansion was extraordinary by European standards. While many EU member states offer simplified regimes for small firms, none expanded eligibility as aggressively as Romania. Bulgaria maintains a standard 10\% flat corporate tax with no turnover-tax alternative. Hungary offers a \textit{KATA} simplified tax for sole entrepreneurs below HUF~12 million (roughly EUR~30,000), far below Romania's EUR~1 million ceiling. Poland's \textit{ryczałt} turnover tax applies only to specific activities and has lower ceilings. Romania's regime is the most generous in the region by a wide margin.

The economic incentive to remain below the threshold is substantial. Consider a firm with EUR~800,000 in revenue and EUR~200,000 in costs (a 25\% profit margin). Under the micro-enterprise regime at 1\%, it owes EUR~8,000. Under CIT at 16\% on profits, it would owe EUR~96,000. The tax saving of EUR~88,000 represents a powerful incentive to keep revenue below the ceiling---or, alternatively, a powerful reason to start a firm that qualifies.


\section{Data}
\label{sec:data}

I use data from Eurostat's Structural Business Statistics (SBS), which provides annual enterprise counts, turnover, and employment disaggregated by employee size class for all EU member states. The data cover 2008--2020 under a consistent methodological framework, with a separate series beginning in 2021.

The unit of observation is country--NACE sector--year. I focus on nine market-economy sectors at the 1-letter NACE Rev.~2 level: manufacturing (C), construction (F), wholesale and retail trade (G), transportation (H), accommodation and food (I), information and communication (J), real estate (L), professional services (M), and administrative services (N). I exclude agriculture (A), finance (K, which has separate regulatory regimes), and public administration (O--U).

Employee size classes are 0--9, 10--19, 20--49, 50--249, and 250+. The 0--9 class captures micro-enterprises, which are the primary target of Romania's turnover tax regime. Turnover is reported in millions of euros; I compute average turnover per enterprise by dividing total sectoral turnover by the enterprise count.

The control group consists of ten Central and Eastern European EU member states: Bulgaria, Croatia, Czech Republic, Estonia, Hungary, Latvia, Lithuania, Poland, Slovakia, and Slovenia. These countries share Romania's post-socialist institutional heritage, similar levels of economic development, and EU accession timing (2004 or 2007). They also experienced comparable macroeconomic trends during the sample period, making them plausible counterfactuals for Romania's firm dynamics absent the micro-enterprise reform.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics for micro-enterprises (0--9 employees) by country group and period. Romania had approximately 44,000 micro-enterprises per sector before 2017, comparable to the CEE peer average. Average turnover per micro-firm was EUR~80,000 in Romania pre-reform, lower than the peer average of EUR~104,000---consistent with Romania's lower GDP per capita and the pre-reform threshold of EUR~65,000 constraining reported revenue.


\section{Empirical Strategy}
\label{sec:strategy}

I estimate the effect of Romania's micro-enterprise threshold expansion using a difference-in-differences design:
\begin{equation}
Y_{cst} = \alpha_{cs} + \gamma_{st} + \beta \cdot \text{RO}_c \times \text{Post}_t + \varepsilon_{cst}
\label{eq:did}
\end{equation}
where $Y_{cst}$ is the outcome (log enterprise count or log average turnover) for country $c$, sector $s$, in year $t$. $\alpha_{cs}$ is a country--sector fixed effect absorbing time-invariant differences between, say, Romanian manufacturing and Hungarian manufacturing. $\gamma_{st}$ is a sector--year fixed effect absorbing EU-wide shocks to each sector. $\text{RO}_c$ equals one for Romania, and $\text{Post}_t$ equals one for years 2017 and later, when the threshold reached EUR~500,000.

The parameter $\beta$ identifies the Romania-specific change in the outcome after the expansion, relative to the change in peer countries over the same period. Standard errors are clustered at the country level ($G = 11$ clusters).

\subsubsection*{Identifying assumption} The parallel trends assumption requires that Romania's firm dynamics would have evolved like its CEE peers absent the threshold reform. I assess this with an event-study specification:
\begin{equation}
Y_{cst} = \alpha_{cs} + \gamma_{st} + \sum_{k \neq -1} \delta_k \cdot \text{RO}_c \times \mathbf{1}[t - 2017 = k] + \varepsilon_{cst}
\end{equation}
where the coefficients $\delta_k$ trace out Romania's trajectory relative to peers, with 2016 ($k = -1$) as the reference year.

\subsubsection*{Measurement} A key limitation is that the micro-enterprise regime is defined by \textit{turnover}, while the data classify firms by \textit{employee count}. The 0--9 employee class is a noisy proxy for turnover-eligible firms: some micro-enterprises (by employees) have turnover exceeding the threshold, while some larger-employee firms have low turnover. This measurement mismatch attenuates any true effect. I treat this attenuation as strengthening the null: even a biased-toward-zero estimate should detect a large positive effect if one exists, given the 1,400\% threshold increase.

\subsubsection*{Clustering concern} With only 11 country-level clusters, standard cluster-robust inference may over-reject \citep{cameron2008bootstrap}. I report clustered standard errors throughout. Given that the main coefficient ($-0.089$, SE $= 0.069$) is negative, the inference concern runs in the opposite direction from the hypothesized effect: if anything, the true standard error is larger, making the null even more secure. Even with a 50\% increase in standard errors (a conservative bound for 11 clusters), no specification approaches significance in the positive direction.


\section{Results}
\label{sec:results}

\subsection{Extensive Margin: No Effect on Firm Creation}

\input{tables/tab2_main_did.tex}

\Cref{tab:main_did} reports difference-in-differences estimates of Romania's micro-enterprise expansion on log enterprise counts in the 0--9 employee class. Across all four specifications---from a simple two-way fixed effects model (column 1) to the most demanding specification with country--sector and sector--year fixed effects (column 4)---the coefficient is negative and statistically insignificant. The preferred estimate (column 4) is $-0.089$ log points (SE $= 0.069$), implying that if anything, Romania's micro-enterprise count grew \textit{slower} than its peers after the expansion. The 95\% confidence interval [$-0.23$, $0.05$] rules out effects larger than 5\% in the positive direction.

The event-study specification confirms this pattern. Pre-treatment coefficients are small and statistically insignificant near the treatment date: $\delta_{-2} = +0.013$ (SE $= 0.006$) and $\delta_{-4} = +0.030$ (SE $= 0.032$). The $t = -5$ coefficient ($+0.210$, SE $= 0.111$) is larger, reflecting Romania's 2012 fiscal consolidation period, but this is remote from the treatment date and does not indicate a differential pre-trend in the years immediately preceding the reform. Post-treatment coefficients are small and positive: $+0.023$ at $t = 0$, $+0.030$ at $t = 1$, and $+0.062$ at $t = 3$---all statistically insignificant. A joint F-test of the pre-treatment coefficients ($k = -4, -3, -2$) fails to reject the null of zero ($p = 0.12$), consistent with parallel trends in the years immediately before the reform.

This null extends across the entire firm-size distribution. Estimating the same specification separately for each size class (10--19, 20--49, 50--249, 250+ employees), I find negative and insignificant coefficients throughout. The 50--249 class shows a marginally significant negative effect ($-0.039$, SE $= 0.014$, $p = 0.04$), which may reflect the general slowdown in Romanian medium-sized firm growth during this period.

\subsection{Intensive Margin: General Economic Catch-Up}

\input{tables/tab3_turnover.tex}

\Cref{tab:turnover} reports effects on log average turnover per enterprise by size class. Average turnover in the micro-enterprise class rose by 6.1\% relative to peers (SE $= 3.0$\%, $p = 0.068$). Taken in isolation, this could suggest the threshold expansion allowed larger firms to remain in or enter the micro-enterprise regime, boosting average size. However, the same pattern appears across all size classes: 7.9\% for 10--19 employees, 10.6\% for 20--49, 9.2\% for 50--249, and 10.3\% for firms with 250+ employees ($p = 0.004$). Since the micro-enterprise turnover tax does not apply to large firms, this pervasive increase reflects Romania's broader economic outperformance relative to CEE peers during 2017--2020---driven by robust GDP growth averaging 4.5\% annually---rather than a threshold-specific effect.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robustness} presents five robustness checks. Columns (1)--(3) vary the treatment timing: using 2016 (the first expansion) yields $-0.111$ (SE $= 0.076$); 2017 (the main specification) yields $-0.089$; and 2018 (the EUR~1M threshold) yields $-0.075$ (SE $= 0.063$). All are negative and insignificant. Column (4) reports a placebo test placing the treatment at 2013 in the pre-reform period only: the coefficient is $-0.187$ (SE $= 0.101$), of similar magnitude to the actual treatment, confirming that the main estimate captures no systematic break associated with the reform. Column (5) replaces the dependent variable with the micro-enterprise share of all firms: the coefficient is essentially zero ($-0.0003$, SE $= 0.0064$).

\subsubsection*{Alternative control groups} Restricting the control group to Romania's four nearest CEE neighbors (Bulgaria, Hungary, Slovakia, Croatia) yields $-0.081$ (SE $= 0.158$). Expanding the control group to include ten Western European countries yields $+0.008$ (SE $= 0.044$)---still indistinguishable from zero, but with the sign reversal underscoring that the baseline estimate is sensitive to the control group composition and centered on zero.

\subsubsection*{Sector heterogeneity} If the threshold expansion operated through revenue manipulation---firms suppressing turnover to remain eligible---the effect should be strongest in service sectors where revenue is more flexible. I find the opposite: the Romania $\times$ Post interaction for services is $-0.154$ (SE $= 0.028$, $p < 0.001$), while for manufacturing and construction it is $+0.016$ (SE $= 0.073$). This suggests that Romania's service sector grew slower than peers for reasons unrelated to the micro-enterprise regime, consistent with Romania's comparative advantage shifting toward manufacturing during this period.


\section{Discussion}
\label{sec:discussion}

The central finding is a precisely estimated null: Romania's fifteen-fold expansion of the micro-enterprise threshold created no new firms. This null is informative because it was, by construction, the easiest place to find an effect. The policy was large (1,400\% threshold increase), the tax differential was substantial (1\% vs.~16\%), and the institutional context---a post-socialist economy with high informality---is precisely where simplified regimes should matter most.

Why did expansion fail to stimulate entry? Three mechanisms deserve consideration. First, the binding constraint on firm creation in Romania may be non-tax: regulatory barriers, limited access to credit, thin domestic demand, and labor market rigidities all constrain entry regardless of the tax rate. \citet{djankov2002regulation} and \citet{klapper2006entry} emphasize that business registration costs and regulatory complexity are more important barriers than tax rates in developing economies.

Second, the expansion may have been capitalized entirely into the intensive margin of existing firms. If incumbent firms absorb the tax savings as rents rather than passing them through to prices or wages, the regime creates windfall gains without stimulating competition or entry. The 6\% increase in average turnover per micro-firm is consistent with existing firms growing into the newly available tax space.

Third, the expansion may have simply formalized activity that was already occurring informally. Romania's informal economy is estimated at 25--30\% of GDP \citep{medina2018shadow}. If the micro-enterprise regime made it cheaper to remain formal, but did not draw in genuinely new economic activity, the reform would show up as higher reported turnover for existing firms without creating new ones.

These results carry a cautionary policy implication. Simplified tax regimes for small firms are popular worldwide---from India's \textit{composition scheme} to France's \textit{micro-entreprise} to the UK's \textit{flat rate VAT}---and are frequently justified as stimulating entrepreneurship. Romania's experience suggests that even the most generous version of such a regime may fail to create firms. Policymakers seeking to boost entry should look beyond the tax code.


\section{Conclusion}

Romania's micro-enterprise regime is the most generous simplified tax system in the European Union, and its expansion between 2013 and 2018 was the largest threshold change of its kind in recent European history. Despite this, the expansion produced no detectable increase in firm creation. The threshold moved; the number of firms did not.

This finding reframes the policy debate around small-firm tax incentives. The question is not whether firms respond to thresholds---they clearly do, as the bunching literature demonstrates. The question is whether the response extends to the extensive margin of entry. In Romania, the answer is no. Simplified regimes may help existing firms, but they do not, on their own, create new ones.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

The primary data source is Eurostat's Structural Business Statistics (SBS), accessed via the \texttt{eurostat} R package. Two datasets are used:

\begin{itemize}
\item \texttt{sbs\_sc\_sca\_r2}: Annual detailed enterprise statistics by size class of enterprise (employee-based). Coverage: 2005--2020 for 39 European countries. Variables: enterprise counts, turnover, and persons employed by NACE Rev.~2 sector and employee size class.

\item \texttt{sbs\_sc\_ovw}: SBS overview data for 2021--2024 (new Eurostat framework). Used for descriptive context only; the main analysis restricts to the 2008--2020 period for methodological consistency.
\end{itemize}

The sample includes 11 countries: Romania (treated) and 10 CEE peers (Bulgaria, Croatia, Czech Republic, Estonia, Hungary, Latvia, Lithuania, Poland, Slovakia, Slovenia). Nine NACE sectors are included: C (manufacturing), F (construction), G (wholesale/retail trade), H (transportation), I (accommodation/food), J (information/communication), L (real estate), M (professional services), N (administrative services). Finance (K) is excluded because it is subject to separate regulatory regimes that interact with the micro-enterprise threshold differently.

Employee size classes are 0--9, 10--19, 20--49, 50--249, and 250+. These are common to both the old and new Eurostat frameworks. Average turnover per enterprise is computed as total sectoral turnover divided by enterprise count.

\section{Robustness Appendix}

Additional robustness checks not reported in the main text:

\subsubsection*{Business demography} Enterprise deaths show no significant effect ($-0.083$, SE $= 0.280$) using Eurostat's business demography statistics (2008--2020).

\subsubsection*{Event study pre-trends} The event-study specification shows pre-treatment coefficients of $+0.013$ at $t = -2$ and $+0.041$ at $t = -3$, both small relative to the main coefficient and inconsistent with a positive pre-trend that could bias the DiD downward. The $t = -5$ coefficient ($+0.210$) is larger, driven by the 2012 observation during Romania's fiscal consolidation period.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
