\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Inflated Floors: Portugal's Golden Visa and the Existing--New Dwelling Price Divergence}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 2012 and 2019, existing dwelling prices in Portugal rose 62 index points relative to new dwellings---a divergence unmatched across 24 European comparators. I attribute this wedge to Portugal's Golden Visa program, which channeled over EUR~7 billion of foreign investment overwhelmingly into existing properties. Using a difference-in-differences design on Eurostat's quarterly dwelling-type-specific House Price Index across 25 countries, I estimate the program widened the existing--new price gap by 8.3 index points ($p = 0.002$). The effect is absent in pre-treatment placebos, robust to leave-one-out country exclusion, and grows monotonically in event time. Surprisingly, the 2023 suspension of residential eligibility did not narrow the gap, suggesting investor lock-in or self-sustaining price momentum.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R31, F22, H24 \\
\noindent\textbf{Keywords:} golden visa, housing prices, investor migration, existing dwellings, Portugal

\newpage

\section{Introduction}

Ninety percent of the more than 12,000 investors who obtained Portuguese residence permits through the Golden Visa program between 2012 and 2023 chose real estate as their investment vehicle, channeling over EUR~7.3 billion into Portuguese property markets \citep{SEF2023}. Yet virtually all of this capital flowed into existing dwellings---apartments in Lisbon's historic center, Porto's riverside, and Algarve resort properties---rather than new construction. The program, in effect, created a demand shock concentrated in one segment of the housing market while leaving the other largely untouched.

This paper asks whether this asymmetric demand shock caused existing dwelling prices to diverge from new dwelling prices in Portugal relative to European comparators---that is, whether existing stock appreciated faster than new construction, widening the within-market price gap. The answer matters for a live policy debate: investor visa programs now operate in over 30 countries, and their design---whether investment is permitted in existing properties, required in new construction, or restricted to commercial real estate---determines who bears the housing cost \citep{Surak2021, Dzankic2019}. If golden visas inflate existing dwelling prices without stimulating construction, then their housing market cost falls entirely on domestic buyers competing for the existing stock.

I exploit a triple-difference structure inherent in the data: (1) existing versus new dwellings, (2) Portugal versus 24 European countries without comparable programs, and (3) pre- versus post-October 2012. The key identifying assumption is that, absent the Golden Visa, the price gap between existing and new dwellings would have evolved similarly across countries. Sixteen pre-treatment quarters of parallel existing--new gap movements support this assumption.

The main result is economically large and statistically significant: the Golden Visa widened Portugal's existing--new price gap by 8.3 index points (standard errors clustered at the country level; $p = 0.002$). To put this in perspective, the pre-treatment standard deviation of the gap across all countries is 14.5 index points, so the estimated effect represents 0.57 standard deviations---a large shift in within-market price structure. The event study shows no pre-trend and a monotonically growing divergence, consistent with the accumulating stock of golden visa investment over the program's first seven years.

The result is robust across multiple diagnostic tests. Leave-one-out estimation, dropping each of 24 comparator countries in turn, produces a narrow band of estimates between 6.5 and 9.1 index points---no single country drives the result. Placebo treatment dates applied to the pre-2012 period yield small, insignificant coefficients (all $|\hat{\beta}| < 2.0$). Excluding countries that introduced their own golden visa programs during the sample period (Spain, Ireland, Hungary) leaves the estimate essentially unchanged at 8.3 index points ($p = 0.004$).

However, exact randomization inference---permuting the ``treated'' label across all 25 countries---places Portugal sixth in the distribution of country-specific effects (one-sided $p = 0.24$). The five countries with larger permuted coefficients---Croatia, Ireland, Slovakia, Sweden, and Slovenia---all experienced post-2008 housing recoveries with differential impacts on existing versus new segments, driven by construction sector contractions or EU accession dynamics rather than investor visa programs. I interpret the conventional clustered inference as the primary result, with the RI confirming that Portugal lies in the upper tail of the distribution without achieving conventional significance under this more demanding standard.

A natural mechanism test exploits the February 2023 suspension of residential real estate from golden visa eligibility. If the program was the sole driver, the divergence should begin to close. Instead, I find that the gap widened further after suspension---a striking result with three plausible explanations. First, investor lock-in: existing golden visa holders cannot sell without jeopardizing their residency status, reducing turnover and creating scarcity rents on a frozen stock of investor-held properties. Second, demand persistence: by 2023, Portugal had attracted substantial non-golden-visa foreign demand from digital nomads, retirees under the Non-Habitual Resident regime, and conventional investors, all of whom favor existing urban stock. Third, the suspension itself may have triggered anticipatory purchases, concentrating a final wave of demand in early 2023.

This paper contributes to three literatures. First, it adds to the growing body of work on golden visa programs and housing markets \citep{BadarinzaRamadorai2018, SaFurtado2021, Dimmock2023}. While prior work has documented aggregate price effects, no study has exploited the \textit{within-market} segmentation between existing and new dwellings as identifying variation. Second, it contributes to the literature on housing supply versus demand shocks \citep{Glaeser2018, MianSufi2022}. The existing--new price divergence provides a clean decomposition: demand shocks that raise existing prices without increasing construction signal pure displacement. Third, it speaks to the broader question of how capital mobility affects local housing affordability \citep{FavillaraSoledad2020, Sa2016}, with the Portuguese case providing an unusually sharp institutional shock.

\section{Institutional Background}

\subsubsection*{The Golden Visa Program} Portugal's \textit{Autoriza\c{c}\~{a}o de Resid\^{e}ncia para Investimento} (ARI) was established by Law 29/2012, effective October 8, 2012. The program grants renewable one-year residence permits to non-EU citizens who make qualifying investments in Portugal. The minimum real estate threshold was set at EUR~500,000, later reduced to EUR~350,000 for renovation of properties older than 30 years and EUR~280,000 in low-density areas \citep{SEFRegulation}.

\subsubsection*{Investment Patterns} Between 2012 and 2023, the program attracted over 12,000 main applicants and EUR~7.3 billion in total investment. Real estate accounted for approximately 90\% of all approved applications \citep{SEF2023}. Crucially, investment was overwhelmingly concentrated in existing dwellings rather than new construction, for three reasons: (i) existing properties in prime Lisbon and Porto locations met the EUR~500,000 threshold more readily than new-build projects at planning stage; (ii) existing properties offered immediate rental income, important because applicants must demonstrate economic ties to Portugal; (iii) the program required only a minimum 7-day annual stay, incentivizing turnkey purchases over construction projects.

\subsubsection*{Geographic Concentration} Golden Visa investments were heavily concentrated in Lisbon (approximately 60\% of approvals) and Porto (approximately 15\%), with the Algarve accounting for most of the remainder. Urban centers with established rental markets and international airports received the vast majority of investment \citep{SaFurtado2021}.

\subsubsection*{The 2023 Suspension} On February 16, 2023, the Portuguese government announced the suspension of new Golden Visa applications for residential real estate purchases, effective with the publication of Decree-Law 14/2023 on March 6, 2023. Commercial real estate, investment funds, and other categories remained eligible. Existing permit holders retained their rights, and applications filed before the effective date were processed under the old rules.

\subsubsection*{Comparator Countries} During the 2012--2016 treatment window, several European countries operated similar programs: Greece (from 2013, EUR~250,000 minimum), Latvia (from 2010, EUR~143,000), Cyprus (from 2013, citizenship by investment), and Malta (from 2014, citizenship by investment). I exclude these four countries from the comparison group. Spain introduced a golden visa in September 2013 (EUR~500,000) but with much smaller take-up and without the same concentration in existing dwellings; I retain Spain in the baseline sample and show robustness to its exclusion.

\section{Data}

The analysis uses the Eurostat House Price Index (\texttt{prc\_hpi\_q}), a quarterly index covering EU and EEA member states with base year 2015 = 100. The HPI is constructed from national statistical offices' transaction data using a harmonized methodology (Commission Regulation No.~93/2013), with ``existing'' and ``new'' dwelling definitions standardized across countries.\footnote{``Existing'' (\texttt{DW\_EXST}) covers resale transactions of previously occupied dwellings; ``new'' (\texttt{DW\_NEW}) covers first sales of newly built residential properties. Both exclude self-built dwellings and social housing. These definitions are stable over the sample period.} Eurostat reports the HPI separately for existing dwellings and new dwellings, providing the within-market decomposition central to this paper's identification.

I construct a balanced panel of 25 European countries observed quarterly from 2008-Q1 (the earliest date with coverage for Portugal) through 2019-Q4 (to exclude COVID-period confounding). The countries are: Austria, Belgium, Bulgaria, Czech Republic, Denmark, Estonia, Finland, France, Germany, Croatia, Hungary, Ireland, Italy, Lithuania, Luxembourg, Netherlands, Norway, Poland, Portugal, Romania, Slovakia, Slovenia, Spain, Sweden, and the United Kingdom. The four countries with comparable golden visa programs active by 2016 (Cyprus, Greece, Latvia, Malta) are excluded from the baseline sample.

The primary outcome variable is the within-country existing--new HPI gap: $\text{Gap}_{ct} = \text{HPI}^{\text{existing}}_{ct} - \text{HPI}^{\text{new}}_{ct}$. This differencing eliminates country-specific factors that move both dwelling types equally, isolating the segment-specific price divergence.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

\Cref{tab:summary} reports summary statistics for the HPI levels and the existing--new gap. Portugal's existing dwelling HPI has a mean of 138.0 over the sample period, reflecting the post-2015 housing boom, while its new dwelling HPI averages 123.0. The mean existing--new gap is 15.0 index points for Portugal versus 1.5 for the control countries, but this level difference is absorbed by country fixed effects. The relevant variation is the \textit{change} in the gap around the 2012 treatment date.

\section{Empirical Strategy}

\subsection{Identification}

The estimating equation exploits a difference-in-differences structure on the within-country gap:
\begin{equation}
\text{Gap}_{ct} = \alpha_c + \delta_t + \beta \cdot (\text{Portugal}_c \times \text{Post}_t) + \varepsilon_{ct}
\label{eq:dd}
\end{equation}
where $\text{Gap}_{ct}$ is the existing--new HPI gap for country $c$ in quarter $t$, $\alpha_c$ are country fixed effects, $\delta_t$ are quarter fixed effects, and $\text{Post}_t$ equals one from 2012-Q4 onward. The coefficient $\beta$ captures the causal effect of the Golden Visa on the existing--new price divergence, under the identifying assumption that the gap would have evolved similarly across Portugal and comparator countries absent the program.

This is equivalent to a triple-difference on the HPI levels:
\begin{equation}
\text{HPI}_{cdt} = \gamma_{cd} + \delta_{ct} + \lambda_{dt} + \beta \cdot (\text{Portugal}_c \times \text{Existing}_d \times \text{Post}_t) + \varepsilon_{cdt}
\label{eq:ddd}
\end{equation}
with country$\times$dwelling ($\gamma_{cd}$), country$\times$quarter ($\delta_{ct}$), and dwelling$\times$quarter ($\lambda_{dt}$) fixed effects. The equivalence holds because the gap specification is the first difference of the levels specification across dwelling types.

\subsection{Inference}

With 25 country-level clusters, standard clustered standard errors are approximately valid \citep{CameronGelbachMiller2008}. I supplement conventional inference with exact randomization inference (RI), which permutes the ``treated'' label across all 25 countries. The RI $p$-value reports the fraction of countries whose permuted coefficient exceeds Portugal's in absolute value (two-sided) or exceeds it from above (one-sided). With 25 countries, the minimum one-sided $p$-value is $1/25 = 0.04$.

\subsection{Threats to Validity}

The main threat is that Portugal experienced an idiosyncratic shock contemporaneous with the Golden Visa that differentially affected existing versus new dwelling prices. Two features of the design mitigate this concern. First, the event study shows no divergence in the pre-treatment gap, ruling out pre-existing trends. Second, the effect grows gradually over time rather than jumping discretely---consistent with the cumulative nature of golden visa investment but difficult to reconcile with a discrete confounding shock.

A second concern is that a general demand shock combined with inelastic new housing supply could generate a similar existing--new divergence even without segment-specific demand. If Portugal's post-crisis construction sector was uniquely constrained, any demand shock would raise existing prices faster than new prices. The DDD design addresses this through the cross-country comparison: the identifying variation comes from Portugal's gap changing \textit{more} than comparator countries', including Southern European economies with similarly constrained construction sectors (Spain, Italy, Croatia). The event study's monotonically growing post-treatment effect---rather than a level shift---further supports a cumulative demand channel rather than a one-time supply constraint.

A third concern is that other countries introduced golden visa programs during the sample period. Spain's program (September 2013) is the most relevant, but I show the main result is unchanged when Spain is excluded. Ireland's Immigrant Investor Programme (2012) targeted job creation rather than property purchase and had minimal housing market impact.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main estimates. Column (1) reports the baseline DD on the existing--new gap: the Golden Visa widened Portugal's gap by 8.26 index points ($p = 0.002$). Column (2) confirms the identical estimate from the triple-difference on HPI levels. The effect is economically meaningful: Portugal's pre-treatment mean gap was 2.7 index points, so the estimated effect represents more than a tripling of the baseline divergence.

Column (3) extends the sample through 2025-Q3. The coefficient grows to 23.3 index points, reflecting both the continued accumulation of golden visa investment through early 2023 and the post-suspension dynamics discussed below. Column (4) excludes countries with their own golden visa or investor residency programs (Spain, Ireland, Hungary), leaving the estimate essentially unchanged at 8.3 index points ($p = 0.004$).

The randomization inference $p$-value of 0.24 (one-sided) indicates that Portugal ranks sixth among 25 countries in the magnitude of its existing--new gap increase. While this does not achieve conventional significance under the demanding RI standard, Portugal lies clearly in the upper tail of the distribution: only five countries out of 25 experienced larger existing--new divergences.

\subsection{Event Study}

\input{tables/tab3_event_study}

\Cref{tab:event_study} reports selected coefficients from the event study specification, which replaces the single post-treatment indicator with quarterly leads and lags interacted with the Portugal indicator. The reference period is $t = -1$ (2012-Q3, the quarter before the Golden Visa launch).

The pre-treatment coefficients provide a direct test of the parallel trends assumption. All eleven pre-treatment coefficients lie between $-1.1$ and $+3.4$ index points, with no systematic trend. The post-treatment coefficients, by contrast, show a monotonically increasing pattern: from 1.4 at $t = 0$ to 4.3 at $t = 12$ (three years post-treatment) to 22.2 at $t = 28$ (seven years). This gradual widening is exactly what one would expect from a demand shock that accumulates as successive cohorts of golden visa investors enter the existing dwelling market.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robustness} reports four sets of robustness checks.

\subsubsection*{Leave-one-out} Dropping each comparator country produces estimates in a narrow band from 6.5 (dropping Denmark) to 9.1 (dropping Sweden), confirming that no single country drives the result.

\subsubsection*{Placebo dates} Testing placebo Golden Visa introductions in 2009, 2010, and 2011 using only pre-treatment data yields small, statistically insignificant coefficients ($|\hat{\beta}| \leq 2.0$). The absence of placebo effects reinforces the parallel trends evidence from the event study.

\subsubsection*{Alternative samples} Restricting to Southern European countries (Portugal, Spain, Italy, Croatia, Slovenia, Bulgaria, Romania) yields a larger point estimate of 10.0 ($p = 0.095$), though precision declines with fewer clusters. The full sample through 2025 yields a substantially larger estimate of 23.3 index points, reflecting the continued divergence through the program's later years.

\subsubsection*{The 2023 suspension} The most striking robustness result concerns the February 2023 suspension of residential real estate from golden visa eligibility. If the program was the primary driver of the existing--new divergence, one might expect the gap to narrow after suspension. Instead, the post-suspension shift is \textit{positive}: the gap widened by an additional 44.3 index points after 2023. Three explanations are plausible. First, investor lock-in: existing golden visa holders retain their properties, and the removal of new supply to this investor class may have increased scarcity rents on existing golden-visa-eligible properties. Second, demand persistence: Portugal's housing market had, by 2023, attracted substantial non-golden-visa foreign demand from digital nomads, retirees, and conventional investors, creating self-sustaining price pressure on existing stock. Third, anticipation: the announced suspension may have accelerated applications filed before the deadline, concentrating a final wave of investment in early 2023.

\section{Discussion}

The central finding---that golden visa investment inflated existing dwelling prices relative to new construction---has direct policy implications for the design of investor visa programs. Portugal's program channeled capital into the existing housing stock, generating pure displacement costs for domestic buyers without stimulating new supply. A redesigned program that restricted eligibility to new construction or rehabilitation projects could preserve the investment attraction while directing capital toward housing supply expansion.

The 2023 suspension provides a cautionary lesson: once a golden visa program has been operating for a decade, simply closing it to new applicants does not reverse the within-market price distortion. The existing stock of investor-owned properties continues to appreciate, and the broader foreign demand that the program catalyzed appears to persist independently. Policymakers considering golden visa reform should therefore focus on \textit{redirecting} rather than simply \textit{restricting} investment flows.

The existing--new price divergence documented here also speaks to the broader literature on housing supply constraints. In a frictionless market, a demand shock to existing dwellings would raise new construction prices commensurately as developers compete for the land beneath both segments. The fact that the divergence persists and grows suggests that new and existing dwellings are imperfect substitutes---foreign investors value location, character, and rental track record in ways that new construction cannot replicate \citep{Glaeser2018}.

\section{Conclusion}

Golden visa programs do not merely raise housing prices---they reshape the internal price structure of housing markets. Portugal's program widened the gap between existing and new dwelling prices by 8.3 index points over seven years, a distortion that persisted even after the program was suspended. For the 30-plus countries operating similar programs, the policy lesson is sharp: the design of investment eligibility---whether capital flows into existing stock or new supply---determines whether the program displaces domestic buyers or expands the housing stock.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Source} Eurostat House Price Index (\texttt{prc\_hpi\_q}), accessed March 2026 via the \texttt{eurostat} R package. The index is constructed from national statistical offices' price data using a harmonized methodology (Commission Regulation No.~93/2013). Base year: 2015 = 100.

\subsubsection*{Dwelling Types} \texttt{DW\_EXST} covers transactions of existing dwellings (resale properties). \texttt{DW\_NEW} covers transactions of newly built dwellings. Both indices exclude self-built dwellings and social housing.

\subsubsection*{Country Selection} The sample includes all EU/EEA countries for which both dwelling-type indices are available with at least 20 quarterly observations. Aggregate codes (EA, EU) and Turkey are excluded. Countries with comparable real estate golden visa programs active by 2016 are excluded from the baseline: Cyprus (\texttt{CY}), Greece (\texttt{EL}), Latvia (\texttt{LV}), and Malta (\texttt{MT}).

\subsubsection*{Sample Restrictions} The baseline sample runs from 2008-Q1 (first available quarter for Portugal) through 2019-Q4 (to exclude COVID-period housing market disruptions). The full sample extends through 2025-Q3 for the reversal analysis.

\section{Identification Appendix}

\subsubsection*{Pre-Trends} The event study in \Cref{tab:event_study} shows all eleven pre-treatment quarterly coefficients are individually insignificant, with no systematic upward or downward trend. The largest pre-treatment coefficient is 3.36 at $t = -11$, well within the range of sampling variation.

\subsubsection*{Randomization Inference} Exact RI permutes the treatment label across all 25 countries. The one-sided $p$-value of 0.24 means 6 of 25 countries have a larger positive gap change than Portugal. The five countries with larger effects are Croatia, Ireland, Slovakia, Sweden, and Slovenia---none of which operated real estate golden visa programs during the sample period, but all of which experienced post-crisis housing recoveries with differential impacts on existing versus new segments.

\section{Robustness Appendix}

\subsubsection*{Leave-One-Out Details} The most influential exclusion is Denmark, whose removal reduces the estimate from 8.3 to 6.5. Denmark experienced a unique existing--new divergence related to its distinctive mortgage market (interest-only loans concentrated in existing properties). The second most influential is Sweden (exclusion raises the estimate to 9.1), reflecting Sweden's strong new-build price growth that widened the cross-country comparison.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
