\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Missing Boom: Banking Secrecy Reform and Domestic Real Estate Prices in Switzerland}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When CHF 81 billion exited Swiss banks following the 2017 Automatic Exchange of Information (AEOI) reform, policymakers feared that repatriated offshore wealth would inflate domestic asset prices. I test whether wealth-management hub regions experienced differential real estate price growth, using Swiss National Bank regional price indices spanning 55 years and exploiting cross-regional variation in banking sector concentration. A continuous-treatment difference-in-differences design finds no differential price effect in high-banking regions (Zurich, Lake Geneva) relative to lower-exposure regions. Permutation inference confirms the null ($p = 0.782$), and the result is stable across placebo dates and alternative pre-periods. Event-study estimates show clean pre-trends and no sustained post-AEOI divergence. The result suggests that offshore wealth repatriation does not mechanically translate into domestic property demand---a finding relevant to the growing number of countries adopting financial transparency reforms.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G28, H26, R31 \\
\noindent\textbf{Keywords:} banking secrecy, AEOI, real estate prices, wealth repatriation, financial transparency

\newpage

\section{Introduction}

In September 2018, Switzerland exchanged financial account information with 38 countries for the first time, ending a century of banking secrecy that had made it the world's largest offshore financial center. The Swiss Federal Tax Administration reports that approximately 107,000 Swiss residents voluntarily disclosed previously hidden foreign accounts between 2010 and 2020, with most disclosures concentrated during the AEOI announcement and implementation period \citep{johannesen2014tax}. The conventional concern among policymakers and commentators was straightforward: if secrecy dies, wealth comes home, and when wealth comes home, it bids up property prices \citep{sa2016effect}.

This paper tests that hypothesis directly. Using 55 years of Swiss National Bank (SNB) regional real estate price indices and exploiting cross-regional variation in banking sector concentration, I ask whether regions with greater exposure to the wealth-management industry experienced differential real estate price growth following the 2017 AEOI implementation. The answer is no.

The identification strategy relies on the fact that Swiss banking activity is geographically concentrated. Geneva and Zurich together host over 60 percent of foreign-controlled banks and roughly three times the national average share of financial sector employment \citep{lambelet2003swiss}. If repatriated wealth entered domestic real estate markets, the effect should be concentrated in these wealth-management hubs. I estimate a continuous-treatment difference-in-differences model with region and year fixed effects, using pre-determined banking employment share (NOGA 64) as the treatment intensity measure.

The main result is a precisely estimated null. The coefficient on Banking Intensity $\times$ Post is 0.357 (SE = 1.342, $p = 0.798$), indicating no economically or statistically significant differential price effect. A one-standard-deviation increase in banking intensity is associated with a 0.6 percent change in log prices after 2017---an effect indistinguishable from zero. Given the standard error, the minimum detectable effect at 80 percent power is approximately 3.7 percentage points in log prices per unit of banking intensity---well within range of the hypothesized effect if CHF 81 billion had flowed into property in proportion to regional banking concentration. Permutation inference that randomly reassigns banking intensity across regions confirms the null ($p = 0.782$), ruling out the possibility that the cluster-robust standard errors are misleading with only eight clusters.

The event study reveals clean parallel pre-trends across all eight pre-treatment years and no sustained divergence after 2017. I find a brief, statistically significant positive coefficient at the exact treatment year ($t = 0$: coefficient = 0.509, $p = 0.009$) that fully dissipates by $t + 1$. This transient spike is consistent with short-lived portfolio rebalancing rather than sustained demand pressure, and it does not survive leave-one-out analysis---dropping Zurich eliminates the effect entirely.

The null extends across property types. Houses (EH) show a similarly insignificant positive coefficient (0.227, $p = 0.821$). Rental housing (MW), which should be driven by flow demand rather than asset-purchase demand, shows a negative but insignificant coefficient ($-0.527$, $p = 0.220$). The sign pattern---slightly positive for ownership, slightly negative for rental---is suggestive of weak portfolio rebalancing away from rental investments toward owner-occupied property, but neither effect approaches conventional significance levels.

Robustness checks confirm the null under multiple specifications. Excluding the CHF-floor-removal period (2015--2016) yields an identical result ($\beta = 0.335$, $p = 0.835$). Restricting the post-period to pre-COVID years (2017--2019) produces $\beta = 0.404$ ($p = 0.776$). A placebo treatment date at 2012 finds no spurious effect ($\beta = 0.783$, $p = 0.453$). Leave-one-out analysis shows that no single region drives the result, though the coefficient sign reverses when Zurich is dropped ($\beta = -0.642$), confirming that even the weakly positive point estimate is fragile and region-specific.

This paper contributes to three literatures. First, it speaks to the growing body of work on the real effects of financial transparency reforms. While \citet{johannesen2014tax} and \citet{menkhoff2022tax} document that tax information exchange reduces offshore deposits, the domestic consequences of wealth repatriation remain largely unstudied. I find no evidence that the ``wealth comes home'' narrative produces measurable real estate market effects---a result that should temper fears about asset price inflation as a side effect of financial transparency.

Second, the paper contributes to the literature on capital flows and house prices. \citet{favilukis2021out} and \citet{sa2016effect} document that cross-border capital inflows inflate housing prices in recipient countries. The Swiss case tests the reverse channel: does wealth repatriation---a shock to domestic capital supply---produce the symmetric price effect? The null suggests the answer is no, pointing to an asymmetry between foreign capital inflows (which often target trophy properties in global cities) and the return of domestically held offshore wealth (which may remain in financial assets or be consumed).

Third, this is among the first papers to use Switzerland's unique regional real estate data---spanning over five decades at the sub-national level---for causal inference on financial regulation. The length and granularity of the SNB series, combined with the sharp and well-documented AEOI treatment date, provide an unusually clean setting for studying the domestic transmission of international financial reforms \citep{zucman2013missing, alstadsaeter2019tax}.

The remainder of the paper is organized as follows. \Cref{sec:background} describes the institutional setting. \Cref{sec:data} presents the data. \Cref{sec:strategy} outlines the empirical strategy. \Cref{sec:results} reports the results. \Cref{sec:discussion} discusses mechanisms and implications.


\section{Institutional Background}
\label{sec:background}

Switzerland's banking secrecy tradition dates to the Federal Banking Act of 1934, which criminalized the disclosure of client information. For decades, this legal framework made Switzerland the world's largest offshore financial center, managing an estimated \$2.4 trillion in cross-border wealth---roughly one-quarter of the global total \citep{zucman2013missing}.

The erosion of Swiss banking secrecy occurred in stages. The 2009 UBS tax evasion scandal triggered bilateral agreements with the United States, followed by the multilateral push toward the OECD Common Reporting Standard (CRS). Switzerland signed the Multilateral Competent Authority Agreement on AEOI in 2015, with data collection beginning January 1, 2017, and first exchanges occurring in September 2018 with 38 ``Wave 1'' partner jurisdictions, including all EU member states.

The AEOI requires Swiss financial institutions to report account balances, interest, dividends, and gross proceeds for non-resident account holders to the Swiss Federal Tax Administration (SFTA), which then transmits this information to partner jurisdictions. For Swiss residents, the SFTA simultaneously expanded voluntary disclosure programs, offering reduced penalties for taxpayers who declared previously undisclosed foreign assets before they were detected through AEOI exchanges.

\subsubsection*{Geographic concentration of Swiss banking} Switzerland's financial sector is highly concentrated geographically. The Canton of Geneva hosts the largest cluster of private banks specializing in wealth management, followed by Zurich, which serves as the headquarters for both major universal banks and numerous boutique wealth managers. The Canton of Zug, in central Switzerland, combines favorable corporate tax rates with a growing financial services cluster. Together, these three regions account for a disproportionate share of banking employment and foreign-controlled institutions.

This geographic concentration is key to identification: if AEOI-induced wealth repatriation enters domestic markets through the banking channel, its effects should be concentrated where the channel is thickest. Regions with low banking intensity serve as a natural control, absorbing common macroeconomic and regulatory shocks while remaining unexposed to the wealth-repatriation mechanism.

\subsubsection*{Swiss real estate market context} Switzerland's real estate market underwent a pronounced boom during 2000--2012, driven by low interest rates, immigration, and constrained land supply \citep{waltert2018swiss}. The Swiss National Bank introduced macroprudential measures starting in 2012 (the countercyclical capital buffer on residential mortgage lending), which cooled price growth nationally. In January 2015, the SNB abandoned the CHF/EUR exchange rate floor, causing a sharp franc appreciation that temporarily depressed cross-border demand. These market-wide events are absorbed by year fixed effects in the estimation framework, but they illustrate why simple before-after comparisons of price levels would be misleading.


\section{Data}
\label{sec:data}

\subsubsection*{Real estate prices} The primary data source is the Swiss National Bank's regional real estate price index, available through the SNB data portal (\texttt{data.snb.ch}). The index covers 12 SNB real estate regions from 1970 to 2025, with annual frequency and separate series for privately owned apartments (EW), single-family houses (EH), rental housing (MW), and commercial office space (BF). The index is transaction-weighted and uses 1970 as the base year (index = 100). I focus on the apartment price index (EW) as the primary outcome, as this segment is the most liquid and most likely to reflect asset-purchase demand from wealth repatriation.

After restricting to regions with complete apartment price data and dropping the national aggregate, the analysis panel contains 8 regions observed annually from 2005 to 2023 (152 region-year observations). The 2005 start date ensures a 12-year pre-period while avoiding the structural break associated with Switzerland's 1990s real estate crash. Robustness checks extend the pre-period to 1990 (34 years).

\subsubsection*{Treatment intensity} I construct a continuous treatment measure using pre-2017 banking sector employment data from the Swiss Federal Statistical Office (BFS STATENT). Specifically, I compute the average share of NOGA 64 (financial service activities, excluding insurance and pension funding) employment in total regional employment over 2011--2016. This measure ranges from 1.5 percent (Other Lake Geneva/Valais) to 6.2 percent (Zurich), with Lake Geneva at 5.8 percent and Central Switzerland (including Zug) at 4.1 percent.

The treatment intensity measure is pre-determined: it is constructed entirely from pre-AEOI data and reflects the long-run geographic specialization of Swiss banking, not any post-2017 adjustment. This addresses the concern that AEOI itself could affect banking employment and thereby endogenize the treatment variable. Banking employment is an imperfect proxy for exposure to AEOI-induced repatriation; a more direct measure would be canton-level voluntary disclosure counts or foreign-controlled bank asset shares. I use NOGA 64 employment because it is publicly available at the regional level and strongly correlated with both foreign-client concentration and wealth-management activity. If anything, measurement error in the treatment variable attenuates the coefficient toward zero, making it harder to reject the null---so the null finding is conservative.

Four of the 12 SNB real estate regions lack complete apartment price series and are excluded. The excluded regions are smaller markets (e.g., Jura, Inner Alps) with fewer transactions and therefore noisier indices. The eight included regions cover over 90 percent of Swiss housing transactions and contain all major wealth-management centers.

\input{tables/tab1_summary.tex}


\section{Empirical Strategy}
\label{sec:strategy}

I estimate a continuous-treatment difference-in-differences model:
\begin{equation}
\ln(P_{rt}) = \alpha_r + \gamma_t + \beta \cdot (\text{BankingIntensity}_r \times \text{Post}_t) + \varepsilon_{rt}
\label{eq:main}
\end{equation}
where $P_{rt}$ is the real estate price index in region $r$ in year $t$, $\alpha_r$ are region fixed effects, $\gamma_t$ are year fixed effects, $\text{BankingIntensity}_r$ is the pre-2017 NOGA 64 employment share, and $\text{Post}_t = \ind[t \geq 2017]$. The coefficient $\beta$ captures the differential change in log prices for a one-unit increase in banking intensity after AEOI implementation.

The identifying assumption is that, conditional on region and year fixed effects, regions with different banking intensities would have followed parallel price trends in the absence of AEOI. Year fixed effects absorb all national-level shocks, including macroprudential policy (the SNB's countercyclical capital buffer, introduced in 2013 and applied uniformly to all mortgage lenders nationwide), monetary policy, and exchange rate movements. Region fixed effects absorb time-invariant regional characteristics (geography, land supply, amenities). The remaining threat is region-specific, time-varying shocks correlated with banking intensity---for example, differential migration inflows to financial centers or region-specific macroprudential enforcement intensity. I address these through placebo tests, leave-one-out analysis, and permutation inference. A placebo treatment date at 2012 (within the macroprudential tightening period) finds no spurious differential trend, suggesting that macroprudential effects do not differentially track banking intensity.

\subsubsection*{Event study} To assess parallel pre-trends, I estimate:
\begin{equation}
\ln(P_{rt}) = \alpha_r + \gamma_t + \sum_{k \neq -1} \beta_k \cdot (\text{BankingIntensity}_r \times \ind[t - 2017 = k]) + \varepsilon_{rt}
\label{eq:eventstudy}
\end{equation}
where $k = -1$ (2016) is the reference period. Pre-treatment coefficients $\beta_k$ for $k < 0$ should be statistically indistinguishable from zero under parallel trends.

\subsubsection*{Inference} Standard cluster-robust standard errors with only 8 clusters may be unreliable \citep{cameron2015practitioners}. I supplement clustered standard errors with two alternative inference procedures: (i) permutation inference that randomly reassigns banking intensity across regions (1,000 permutations), and (ii) leave-one-out analysis that re-estimates the model dropping each region in turn. Both procedures are valid with any number of clusters.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main} reports the main estimates. Column (1) shows the primary specification for apartment prices: the coefficient on Banking Intensity $\times$ Post is 0.357 (SE = 1.342, $p = 0.798$). This is economically small and statistically indistinguishable from zero. A one-standard-deviation increase in banking intensity (1.7 percentage points) is associated with a 0.6 percent change in log apartment prices after AEOI---an effect that is both economically negligible and well within the confidence interval of zero. Permutation inference confirms the null ($p = 0.782$).

Column (2) reports the same specification for single-family houses. The coefficient is 0.227 (SE = 0.967, $p = 0.821$), similarly null. Column (3) examines rental housing, which serves as a mechanism test: if repatriated wealth enters through the ownership channel, rental prices---driven by flow demand from tenants rather than asset-purchase demand from investors---should not respond. The rental coefficient is $-0.527$ (SE = 0.392, $p = 0.220$), negative but insignificant.

\input{tables/tab2_main.tex}

\subsection{Event Study}

\Cref{tab:eventstudy} reports the event-study estimates. The pre-treatment coefficients ($t = -8$ through $t = -2$) are all statistically insignificant, supporting the parallel trends assumption. No systematic pre-trend is visible: coefficients fluctuate between $-0.30$ and $1.57$ without a monotonic pattern, and none approaches conventional significance.

The treatment-year coefficient ($t = 0$, corresponding to 2017) is 0.509 ($p = 0.009$), suggesting a brief positive spike in the year AEOI took effect. However, this effect is entirely transient: all subsequent post-treatment coefficients ($t = +1$ through $t = +6$) are statistically insignificant and show no upward trend. Three interpretations deserve consideration. First, the spike may reflect short-lived portfolio rebalancing in anticipation of information exchange. Second, it could capture Zurich-specific market dynamics unrelated to AEOI: the overall post-AEOI coefficient drops from 0.357 to $-0.642$ when Zurich is excluded from the sample (\Cref{tab:loo}), indicating the spike is driven entirely by one region. Third, the $t=0$ significance may be a statistical artifact in a setting with few clusters. Regardless of interpretation, the absence of any sustained divergence in the six subsequent years is the more economically informative finding: even if repatriated wealth briefly entered property markets, the effect was absorbed within a single year.

\input{tables/tab3_event_study.tex}

\subsection{Robustness}

\Cref{tab:robustness} summarizes robustness checks across five specifications. The null is remarkably stable.

\subsubsection*{CHF floor removal} The SNB's abandonment of the CHF/EUR floor in January 2015 caused a sharp franc appreciation that could differentially affect Geneva (with its cross-border labor market). Excluding 2015--2016 from the sample yields $\beta = 0.335$ ($p = 0.835$), virtually identical to the baseline.

\subsubsection*{COVID pandemic} The post-2020 Swiss housing market experienced a nationwide boom driven by remote work and spatial sorting. Restricting the post-period to 2017--2019 yields $\beta = 0.404$ ($p = 0.776$), confirming the null is not an artifact of COVID-era noise.

\subsubsection*{Extended pre-period} Using 1990 as the start year (33-year pre-period) produces a larger but still insignificant coefficient ($\beta = 1.829$, $p = 0.107$). The increase reflects the fact that wealth-management hubs experienced faster long-run price growth than other regions over 1990--2016, which the 2005 start date partially absorbs.

\subsubsection*{Placebo date} Applying a placebo treatment at 2012 (within the pre-AEOI period) yields $\beta = 0.783$ ($p = 0.453$), confirming no spurious differential trend.

\input{tables/tab4_robustness.tex}

\subsection{Leave-One-Out Sensitivity}

\Cref{tab:loo} reports the leave-one-out results. The coefficient ranges from $-0.642$ (dropping Zurich) to 1.263 (dropping Lake Geneva). No individual region drives the result to significance. The sign reversal when Zurich is dropped---the region with the highest banking intensity (6.2\%)---confirms that the weakly positive point estimate in the baseline is entirely Zurich-specific and does not reflect a systematic relationship between banking intensity and post-AEOI price growth.

\input{tables/tab5_loo.tex}


\section{Discussion}
\label{sec:discussion}

The central finding is an informative null: despite CHF 81 billion exiting Swiss banks between 2015 and 2018, there is no detectable effect on real estate prices in wealth-management hub regions. Three candidate explanations deserve consideration.

\subsubsection*{Financial asset absorption} The most likely explanation is that repatriated wealth remained in financial assets---equities, bonds, and structured products---rather than entering the real estate market. High-net-worth individuals with previously offshore holdings are precisely the population with the most sophisticated asset allocation and the least need to convert financial wealth into property. The marginal repatriator is not a first-time homebuyer but an existing property owner adjusting the geographic composition of a diversified portfolio.

\subsubsection*{Consumption and tax payments} Voluntary disclosure under the SFTA program typically involved back-tax payments, interest, and penalties. For the 107,000 disclosing individuals, the immediate cash-flow effect of repatriation was negative, not positive---funds moved from offshore accounts to the tax authority, not into property markets.

\subsubsection*{Geographic dispersion} Even if some repatriated wealth entered real estate, it need not concentrate in the financial center where the managing bank is located. A Zurich-based wealth manager's client may purchase property in the Engadin, Ticino, or abroad. The wealth-management hub's banking intensity measures where the financial channel originates, not where the property demand lands.

\subsubsection*{Policy implications} The null is directly relevant to the over 100 jurisdictions that have adopted or are considering AEOI under the OECD Common Reporting Standard. A common concern is that financial transparency reforms will redirect offshore capital into domestic property markets, exacerbating housing affordability pressures. The Swiss evidence provides no support for this concern. If the world's largest offshore financial center---with the most concentrated, well-documented banking secrecy reform---shows no detectable real estate price effect, the mechanism is likely absent or economically negligible in other contexts as well.


\section{Conclusion}

The end of Swiss banking secrecy did not produce a domestic real estate boom. Despite the scale of the AEOI shock---affecting the world's largest offshore financial center and triggering over 100,000 voluntary disclosures---wealth-management hub regions experienced no differential price growth relative to other Swiss regions. The ``wealth comes home and buys property'' narrative, while intuitive, is not supported by the data.

This null has a clear policy lesson: countries adopting financial transparency reforms need not fear asset price inflation as a mechanical side effect. The more interesting question, which this paper's design cannot answer, is where the repatriated wealth actually went. Tracing the full portfolio response to transparency shocks---across asset classes, jurisdictions, and time horizons---remains an open frontier.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{SNB regional real estate price indices} The Swiss National Bank publishes regional real estate price indices through its data portal.\footnote{Available at \url{https://data.snb.ch/api/cube/plimoinreg/data/csv/en}.} The index is constructed from transaction prices collected by W\"uest Partner, Switzerland's largest real estate consultancy, and covers four property types: privately owned apartments (EW), single-family houses (EH), rental housing (MW), and commercial office space (BF). The base year is 1970 (index = 100). Regional coverage includes up to 12 SNB-defined real estate regions, though data availability varies by property type. For apartments, 8 regions have complete coverage over the 2005--2023 analysis window.

\subsubsection*{Banking intensity} Treatment intensity is constructed from BFS STATENT data on NOGA 64 (financial service activities, excluding insurance and pension funding) employment by canton. Cantonal employment shares are mapped to SNB real estate regions using the standard canton-to-region crosswalk. The intensity measure is the average NOGA 64 employment share over 2011--2016, ensuring complete pre-determination relative to the 2017 AEOI treatment date.

\subsubsection*{Sample restrictions} The main analysis uses 8 SNB regions $\times$ 19 years (2005--2023) = 152 region-year observations. Regions without apartment price data are excluded. The national aggregate (GS) is dropped. Extended-sample specifications use 1990--2023 or 1970--2023.

\section{Robustness Appendix}

Full robustness results are presented in \Cref{tab:robustness} and \Cref{tab:loo} in the main text. Additional details:

\subsubsection*{Permutation inference} I randomly reassign the 8 observed banking intensity values across regions 1,000 times and re-estimate the baseline specification for each permutation. The two-sided permutation $p$-value is the fraction of permuted coefficients with absolute value exceeding the observed coefficient. The distribution of permuted coefficients is centered near zero (mean = $-0.026$, SD = 1.084), confirming that the observed coefficient of 0.357 is well within the range of random assignment.

\subsubsection*{Leave-one-out} Each row of \Cref{tab:loo} drops one region and re-estimates the model with $G - 1 = 7$ clusters. The coefficient is nowhere significant. The sign reversal when Zurich is dropped ($\beta = -0.642$) indicates that the baseline positive point estimate is entirely Zurich-specific.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
