\begin{table}[!h]
\centering
\caption{Standardized Effect Sizes\label{tab:sde}}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{lcccccc}
\toprule
Outcome & $\hat{\beta}$ & SE & SD($Y$) & SDE & SE(SDE) & Classification\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Panel A: Pooled}}\\
\hspace{1em}Total completions & -32.84 & 17.83 & 244.85 & -0.1341 & 0.0728 & Moderate negative\\
\hspace{1em}Minority completions & -10.99 & 5.97 & 91.37 & -0.1203 & 0.0653 & Moderate negative\\
\hspace{1em}Black completions & -6.26 & 3.83 & 53.20 & -0.1177 & 0.0720 & Moderate negative\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Panel B: Heterogeneous}}\\
\hspace{1em}Total comp. (high minority share) & -4.07 & 5.99 & 132.31 & -0.0308 & 0.0728 & Small negative\\
\hspace{1em}Total comp. (low minority share) & -61.50 & 34.68 & 341.20 & -0.1802 & 0.0653 & Large negative\\
\bottomrule
\end{tabular}
\end{table}
\begin{tablenotes}[flushleft]\small
\item \textit{Notes:} \textbf{Country:} United States. \textbf{Research question:} Does public disclosure of program-level quality scores under the Gainful Employment rule permanently reduce credential completions at for-profit college programs, even after regulatory enforcement is rescinded? \textbf{Policy mechanism:} The Department of Education published Debt-to-Earnings rates for 8,637 for-profit programs in January 2017, publicly labeling approximately 740 programs as ``failing'' (D/E $> 12\%$); the Trump administration paused enforcement in June 2017 and rescinded the rule in July 2019, removing the regulatory threat while leaving the published scores in the public record. \textbf{Outcome definition:} Annual program-level certificate and associate degree completions from IPEDS Completions survey (C\_A table), summed across award levels within each 6-digit CIP code at each institution. \textbf{Treatment:} Binary; 1 if program's official GE status was ``FAIL'' (annual D/E rate exceeding 12\% threshold), 0 if ``PASS'' (annual D/E rate below 8\%). \textbf{Data:} Department of Education GE Final D/E Rates (DMYR 2015, published January 2017) merged with IPEDS Completions 2013--2021 via OPEID-unitid crosswalk; 24,598 program-year observations across 3,429 programs at 1,200 for-profit institutions. \textbf{Method:} Two-way fixed effects DiD with program and year fixed effects; two-stage decomposition separating post-publication (2017) from post-rollback (2018+) effects; standard errors clustered at the institution level. \textbf{Sample:} Restricted to for-profit (proprietary) institutions with GE-evaluated programs; programs in GE ``zone'' (D/E 8--12\%) excluded from main analysis. SDE $= \hat{\beta} / \text{SD}(Y)$ where SD($Y$) is the pre-treatment standard deviation of the outcome. $\hat{\beta}$ is the cumulative effect (post-publication + post-rollback coefficients). Classification refers to magnitude, not statistical significance: Large ($|$SDE$| > 0.15$), Moderate ($0.05$--$0.15$), Small ($0.005$--$0.05$), Null ($< 0.005$).
\end{tablenotes}

