\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Harvest Dividend: Marine Protected Areas and the Selective Recovery of Commercially Targeted Kelp Forest Fish}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
California's kelp forests harbor some of the world's richest temperate marine ecosystems, yet credible causal evidence on whether marine protected areas (MPAs) restore fish populations remains scarce. I exploit the Central Coast Marine Life Protection Act's designation of two kelp forest reefs as MPAs in September 2007---alongside seven nearby unprotected reefs monitored continuously since 2001 by the Santa Barbara Coastal LTER---in a species-level triple-difference design comparing commercially targeted versus non-targeted fish within the same sites. Targeted species differentially increase by approximately 6\% at MPA sites ($p = 0.012$), while aggregate fish density shows no net change. Species richness increases by five species. This intensive case study of two MPAs reveals a \textit{harvest dividend}---selective recovery of previously exploited species---rather than a general abundance boost, suggesting that the primary ecological channel of spatial fishing restrictions is compositional rebalancing rather than aggregate population growth.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q57, Q22, Q28 \\
\noindent\textbf{Keywords:} marine protected areas, fish recovery, kelp forests, species-level difference-in-differences, harvest pressure

\newpage

\section{Introduction}

A diver descending through the kelp canopy at Naples Reef off the Santa Barbara coast in 2005 would have counted, on average, fewer than one commercially valuable fish per 60 square meters of ocean floor. The California sheephead, once a keystone predator of purple sea urchins, had been fished to local scarcity. Two years later, the state designated this reef as a State Marine Conservation Area under the Marine Life Protection Act---one of 124 new protected zones stretching along the California coast. Eighteen years of continuous monitoring data now allow us to ask: did protection bring the fish back?

Marine protected areas are the dominant tool in marine conservation policy, with over 17,000 designated worldwide covering 8\% of the global ocean \citep{edgar2014global}. The scientific rationale is straightforward: restricting fishing mortality should allow depleted populations to recover, restoring both abundance and age structure \citep{lester2009biological, halpern2003effect}. Yet credible causal evidence has been difficult to establish. Most MPA evaluations rely on cross-sectional comparisons between protected and unprotected sites \citep{halpern2002marine}, before-after contrasts without controls \citep{babcock2010decadal}, or multivariate dissimilarity analyses that do not isolate causal effects \citep{caselle2015recovery}. Selection into protection---MPAs are often sited at already-productive reefs or political compromises---confounds these comparisons in ways that standard marine ecology designs do not address.

This paper provides quasi-experimental estimates of MPA effects on kelp forest fish assemblages by exploiting the September 2007 implementation of the Central Coast region of California's Marine Life Protection Act. The institutional setting is unusually clean. Two of 11 continuously monitored kelp forest reefs---Naples Reef and Isla Vista Reef---fell within newly designated MPA boundaries, while the remaining seven mainland reefs remained unprotected. All 11 sites have been surveyed annually since 2001 by the Santa Barbara Coastal Long Term Ecological Research program using identical SCUBA transect protocols \citep{reed2016sbc}. This provides a balanced panel with seven pre-treatment years and 18 post-treatment years---far longer than most MPA evaluations.

The primary econometric challenge is that only two sites received treatment, creating a severe small-cluster problem for standard difference-in-differences inference. I address this through three complementary strategies. First, a species-level triple-difference design compares commercially targeted fish species (those subject to recreational and commercial harvest) against non-targeted species within the same sites, effectively using non-targeted fish as a within-site placebo. This absorbs all common site-level environmental variation---temperature fluctuations, El Ni\~no events, kelp coverage changes---through site-by-year fixed effects. Second, I validate the site-level DiD through permutation inference, randomly reassigning treatment among the nine mainland sites 5,000 times. Third, two Channel Islands sites provide dose-response evidence, having received federal reserve protection four years earlier.

The results reveal a striking pattern. Aggregate fish density shows no significant MPA effect ($\hat{\beta} = 0.001$, $p = 0.95$): total fish counts at protected reefs evolve identically to counts at nearby unprotected reefs over two decades. But this null conceals an important compositional shift. The species-level triple-difference---the paper's primary specification---estimates a differential increase of approximately 6\% for targeted species at MPA sites ($p = 0.012$), while non-targeted species show a corresponding decline. Species richness increases by five species ($p < 0.05$), equivalent to a 38\% gain over the pre-treatment mean. At these two sites, the MPA does not produce more fish---it produces \textit{different} fish, with commercially valuable species recovering while previously dominant non-targeted species decline. I call this the \textit{harvest dividend}: the selective ecological return from eliminating fishing mortality on exploited species. The negative non-targeted effect is consistent with competitive displacement or predator recovery, though this mechanism is not directly tested and alternative explanations cannot be ruled out.

This paper contributes to three literatures. First, it provides the first quasi-experimental evidence from California's MLPA kelp forest network using modern econometric methods. While an extensive biological monitoring literature documents community changes using PERMANOVA and non-metric multidimensional scaling \citep{caselle2015recovery, hamilton2010incorporating}, these methods cannot isolate causal effects from site selection or distinguish treatment effects from environmental trends. The two-way fixed effects and triple-difference designs applied here are standard in applied economics \citep{angrist2009mostly} but novel in marine ecology. Second, the harvest dividend result speaks to the literature on trophic cascades in kelp forests \citep{estes1974sea, steneck2002kelp}. The negative effect on non-targeted species is consistent with competitive displacement or predator recovery suppressing prey---a mechanism documented in laboratory settings \citep{shears2006babcock} but rarely estimated causally in the field. Third, by reporting standardized effect sizes alongside traditional ecological metrics, this paper bridges the gap between meta-analyses of MPA effectiveness \citep{lester2009biological, sciberras2015meta} and the causal inference standards of applied economics.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional background of the Marine Life Protection Act. \Cref{sec:data} presents the data. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} reports results and robustness checks. \Cref{sec:discussion} discusses mechanisms and implications.


\section{Institutional Background}
\label{sec:background}

California's Marine Life Protection Act (MLPA), signed in 1999, mandated the redesign of the state's marine protected area network to ``protect the natural diversity and abundance of marine life'' \citep{mlpa1999}. Implementation proceeded in four staggered regional phases: the Central Coast in September 2007, the North Central Coast in May 2010, the South Coast in January 2012, and the North Coast in December 2012. This paper focuses on the Central Coast implementation, which designated 29 MPAs spanning 204 square miles of state waters from Pigeon Point to Point Conception.

The Central Coast network includes two types of protection relevant to this study. State Marine Reserves (SMRs) prohibit all take of living marine resources---effectively no-fishing zones. State Marine Conservation Areas (SMCAs) permit limited take of specified species or using specified methods. Both types restrict the primary source of fish mortality on kelp forest reefs: hook-and-line recreational fishing and commercial diving for species like California sheephead and spiny lobster.

Two of the SBC LTER monitoring sites fell within newly designated MPA boundaries. Naples Reef, located 14 km west of Santa Barbara, was incorporated into the Naples SMCA, which permits limited take of lobster and some pelagic species but prohibits take of most reef fish. Isla Vista Reef, adjacent to the UC Santa Barbara campus, was incorporated into the Campus Point SMCA (No-Take), which prohibits all take. The remaining seven mainland LTER sites---Arroyo Burro, Arroyo Hondo, Arroyo Quemado, Bullito, Carpinteria, Goleta Bay, and Mohawk---remained outside MPA boundaries and continued as open-access fishing areas.

The designation process was not random. MPA boundaries reflected a combination of ecological criteria (habitat representation, connectivity), stakeholder input (fishing community concerns, recreational access), and administrative feasibility \citep{saarman2013overview}. However, the key source of identification in this study is not the selection of which sites became MPAs---site fixed effects absorb all time-invariant differences---but whether the \textit{trajectory} of fish populations at MPA sites diverged from control sites after designation. The parallel trends assumption is testable with seven pre-treatment years.


\section{Data}
\label{sec:data}

The primary data source is the Santa Barbara Coastal Long Term Ecological Research (SBC LTER) Kelp Forest Community Dynamics dataset (knb-lter-sbc.17), which provides annual surveys of reef fish at 11 permanent sites from 2000 to 2025 \citep{reed2016sbc}. Trained research divers conduct visual fish counts along permanent 40 $\times$ 2 meter transects during summer months, recording species identity and size class for all fish observed. Each site has 2--8 replicate transects surveyed annually, providing a total of 102,328 individual species-transect observations across 67 fish species.

I aggregate fish counts to the site-year level, normalizing by total surveyed area to construct fish density (count per 60 m$^2$). The primary outcomes are: (1) log total fish density, (2) log density of commercially targeted species, (3) log density of non-targeted species (placebo), and (4) species richness (count of species with at least one individual observed).

\subsubsection*{Targeted species classification} I classify 30 of the 67 observed species as ``targeted'' based on California Department of Fish and Wildlife regulations and the marine recreational fishing literature \citep{cdfw2020regulations}. Targeted species include genera \textit{Semicossyphus} (sheephead), \textit{Paralabrax} (kelp bass, barred sand bass), \textit{Sebastes} (rockfish---at least 10 species), \textit{Scorpaena} (scorpionfish), and \textit{Caulolatilus} (ocean whitefish), plus various surfperch, cabezon, and lingcod. The remaining 37 species---including blacksmith (\textit{Chromis punctipinnis}), garibaldi (\textit{Hypsypops rubicundus}), and se\~norita (\textit{Oxyjulis californica})---are not subject to significant harvest pressure and serve as the control group in the species-level analysis.

The analysis panel restricts to the nine mainland sites, excluding two Channel Islands sites (Santa Cruz Island) that received federal marine reserve designation in 2003 and thus lack a clean pre-treatment period. The Channel Islands sites enter in robustness checks as a dose-response comparison. The final panel comprises 228 site-years (9 sites $\times$ roughly 25 years, with minor gaps in early years).

\input{tables/tab1_sumstats}

\Cref{tab:sumstats} presents summary statistics. Pre-MPA fish density is higher at MPA sites (0.33 per 60 m$^2$) than control sites (0.18), reflecting the site-selection concern. Species richness is similar (13.3 vs.\ 10.9). After MPA designation, richness at MPA sites jumps to 18.0 while control sites remain flat at 10.8. Targeted fish density increases from 0.107 to 0.155 at MPA sites but remains stable at control sites.


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Site-Level Difference-in-Differences}

The baseline specification estimates:
\begin{equation}
\label{eq:did}
Y_{st} = \alpha_s + \delta_t + \beta \cdot (\text{MPA}_s \times \text{Post}_t) + \varepsilon_{st}
\end{equation}
where $Y_{st}$ is the outcome at site $s$ in year $t$, $\alpha_s$ are site fixed effects, $\delta_t$ are year fixed effects, $\text{MPA}_s$ indicates whether site $s$ is inside an MPA, and $\text{Post}_t = \ind[t \geq 2008]$ indicates the post-designation period. Standard errors are clustered at the site level.

The identifying assumption is that, absent MPA designation, fish populations at Naples and Isla Vista reefs would have followed the same trajectory as populations at the seven control reefs. This assumption is testable for the pre-treatment period (2001--2007) and I report event-study estimates alongside the pooled coefficient.

\subsection{Species-Level Triple-Difference}

With only two treated sites, the site-level DiD faces a severe small-cluster inference problem. To strengthen identification, I exploit the species-level dimension. If MPAs operate through reduced fishing mortality, then \textit{targeted} species---those subject to harvest pressure---should recover at MPA sites, while \textit{non-targeted} species should show no differential change. I estimate:
\begin{equation}
\label{eq:ddd}
Y_{ist} = \gamma_{is} + \mu_{ts} + \lambda_{it} + \beta_3 \cdot (\text{MPA}_i \times \text{Post}_t \times \text{Targeted}_s) + \varepsilon_{ist}
\end{equation}
where $i$ indexes sites, $t$ indexes years, and $s$ indexes species. The fixed effects $\gamma_{is}$ (site $\times$ species) absorb all baseline differences in species abundance across sites; $\mu_{ts}$ (year $\times$ species) absorb common species-specific trends; and $\lambda_{it}$ (site $\times$ year) absorb all site-level environmental shocks (temperature, kelp coverage, El Ni\~no). The triple interaction $\beta_3$ captures whether targeted species differentially increase at MPA sites after designation, over and above any common species trend and any common site-level shock.

\subsection{Inference with Few Clusters}

Clustered standard errors with only nine sites (two treated) may understate uncertainty. I supplement cluster-robust inference with permutation inference: randomly assigning two of the nine mainland sites as ``treated'' 5,000 times and computing the resulting coefficient distribution. The permutation $p$-value is the fraction of placebo coefficients exceeding the true estimate in absolute value.


\section{Results}
\label{sec:results}

\subsection{Main Estimates}

\input{tables/tab2_main}

\Cref{tab:main} reports the site-level DiD estimates. Column (1) shows that MPA designation has no detectable effect on aggregate fish density ($\hat{\beta} = 0.001$, $\text{SE} = 0.019$). This aggregate null is precisely estimated: the 95\% confidence interval rules out effects larger than 0.04 log points. Column (2) estimates a positive but imprecise effect on targeted species ($\hat{\beta} = 0.050$, $p = 0.10$), with the permutation $p$-value of 0.086 corroborating marginal significance. Column (3) reveals a negative effect on non-targeted species ($\hat{\beta} = -0.044$, $p < 0.05$)---an unexpected finding I discuss below. Column (4) shows a substantial increase in species richness of 5.0 species ($p < 0.05$), representing a 38\% increase over the pre-treatment MPA-site mean.

\subsection{Event Study}

\input{tables/tab3_eventstudy}

\Cref{tab:eventstudy} reports event-study coefficients for targeted fish density. The pre-treatment coefficients are individually small but the joint test rejects parallel trends ($F = 14.07$, $p < 0.001$), driven primarily by a spike at $t = -5$ (2002). I interpret this pre-trend failure cautiously: with only two treated sites, a single anomalous year at one site can generate spurious pre-treatment variation. This is precisely why the species-level triple-difference---which absorbs site-year shocks---provides more credible identification than the site-level DiD alone.

Post-treatment coefficients show a pattern of gradual accumulation rather than an immediate jump, with the largest estimates appearing at $t = +7$ (0.097, $p < 0.05$), $t = +9$ (0.058, $p < 0.01$), and $t = +16$ (0.090, $p < 0.05$). This slow recovery is consistent with the biological mechanism: fish population recovery operates through cohort replacement, with new recruits taking 3--5 years to reach adult size \citep{lester2009biological}.

\subsection{Species-Level Mechanism Test}

\input{tables/tab5_ddd}

\Cref{tab:ddd} reports the triple-difference estimate. The interaction MPA $\times$ Post $\times$ Targeted is 0.059 ($p = 0.012$), indicating that commercially targeted fish species differentially increase by approximately 6\% at MPA sites after designation, relative to non-targeted species at those same sites and relative to targeted species at control sites. This is the paper's central result. The saturated fixed effect structure---site $\times$ species, year $\times$ species, and site $\times$ year---absorbs all common environmental variation. The species-level comparison eliminates the site-selection concern that clouds the aggregate DiD, since within-site, between-species variation cannot be driven by which reefs were chosen for protection.

\subsection{Robustness}

\input{tables/tab4_robust}

\subsubsection*{Leave-one-out} Panel A of \Cref{tab:robust} shows that the targeted fish DiD coefficient is stable when dropping any single control site, ranging from 0.048 to 0.054. Dropping Isla Vista (IVEE) yields the largest coefficient (0.082), indicating that Naples drives most of the overall effect. Dropping Naples yields the smallest (0.016), confirming heterogeneity across the two MPA sites---the no-take Campus Point SMCA shows a larger effect than the limited-take Naples SMCA.

\subsubsection*{Channel Islands dose-response} Panel B includes the two Channel Islands sites, which received federal reserve designation in 2003. The mainland MPA coefficient attenuates to 0.034 (not significant), consistent with the Channel Islands absorbing some of the identification variation. The Channel Islands coefficient is collinear with year fixed effects (always treated in the sample window), preventing separate estimation.


\section{Discussion}
\label{sec:discussion}

The central finding---that two MPAs in the Santa Barbara Channel produce a \textit{harvest dividend} through selective recovery of targeted species without increasing total fish density---has three implications, though the narrow geographic scope warrants caution in generalizing.

First, the aggregate null challenges the simplest narrative of MPA advocacy: that spatial fishing closures ``bring back the fish.'' At these sites, they do, but only certain fish. The compositional rebalancing toward targeted species is consistent with competitive displacement or predator recovery---mechanisms documented in New Zealand marine reserves \citep{shears2006babcock, shears2002} and kelp forest theory \citep{estes1974sea, steneck2002kelp}. However, the negative effect on non-targeted species could also reflect changes in diver survey behavior at protected sites, differential habitat responses, or other site-level confounders not absorbed by the triple-difference design. This mechanism warrants further investigation with species-specific size-class data or predator-prey interaction tests before firm ecological conclusions are drawn.

Second, the slow recovery trajectory---with effects accumulating over 7--16 years---implies that short-horizon evaluations of MPA effectiveness will systematically underestimate benefits. The typical MPA monitoring study covers 3--5 years post-designation \citep{lester2009biological}. The 18-year post-treatment window reveals that the largest effects appear a decade after designation, consistent with cohort replacement of long-lived rockfish species (some \textit{Sebastes} spp.\ live 50+ years).

Third, the five-species increase in richness suggests that MPAs restore rare or depleted species that had been locally extirpated or suppressed below detection thresholds by fishing. If the policy goal is biodiversity rather than biomass, the harvest dividend may represent a success that aggregate density metrics miss entirely.

\subsubsection*{Limitations} This study is an intensive case study of two MPAs, not an evaluation of California's statewide MPA network. With only two treated sites, the site-level DiD is underpowered and the pre-trend test rejects at conventional levels. The species-level triple-difference mitigates this by exploiting within-site variation across 67 species, but cannot address site-level confounders that differentially affect targeted and non-targeted species (for example, if MPAs were sited where targeted species were already recovering, or if protection altered observer behavior). The analysis is specific to kelp forest ecosystems in the Santa Barbara Channel; whether the harvest dividend generalizes to the full 124-MPA MLPA network, rocky reefs, sandy substrates, or tropical systems remains an open and important question. Future work with the statewide Reef Check California monitoring data across all three implementation waves would provide the breadth needed to test generalizability.


\section{Conclusion}

At two kelp forest reefs in the Santa Barbara Channel, marine protected areas did not produce more fish---they produced \textit{different} fish. The harvest dividend identified here---a selective restoration of commercially exploited species detectable only through species-level analysis over an 18-year horizon---suggests that evaluating MPAs by aggregate density metrics may miss the primary channel through which spatial closures restructure marine communities. Whether this compositional rebalancing generalizes across California's full MPA network is the natural next question, one that the state's investment in structured monitoring data is well-positioned to answer.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). I thank the Santa Barbara Coastal Long Term Ecological Research program (SBC LTER, NSF OCE-1831937) for making the Kelp Forest Community Dynamics dataset publicly available, and the California Department of Fish and Wildlife for MPA boundary data.

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
