\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Inertia Break: Military Conscription and the Farm-to-Factory Transition in World War I America}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 1910, four in ten young American men lived on farms despite higher manufacturing wages. What breaks this agricultural inertia? I exploit the fact that the 1917 Selective Service Act drafted native-born citizens but exempted foreign-born non-citizens, comparing occupational transitions of these two groups across draft-eligible and ineligible age cohorts in linked full-count census data covering 4.9 million men. Draft-exposed native-born men were 1.6 percentage points more likely to exit farming and gained 5.3 more points in occupational income scores than their foreign-born peers---an effect of 0.43 standard deviations. The farm exit effect was 2.9 percentage points larger in high-agriculture counties than in low-agriculture counties, and 1.3 percentage points larger for Black men than for white men, consistent with disruption overcoming attachment costs where they bind most. Forced occupational disruption accelerated the structural transformation over the subsequent decade.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} N31, N32, J24, J61, O14 \\
\noindent\textbf{Keywords:} structural transformation, military service, agricultural inertia, World War I, occupational mobility

\newpage

\section{Introduction}

In 1910, 41 percent of young American men lived on farms. Manufacturing wages exceeded agricultural wages by a wide margin \citep{caselli2001, goldin2000}, yet millions remained in agriculture decade after decade. This paper asks what breaks that inertia---and finds the answer in an unlikely place: the trenches of World War I.

The farm-to-factory transition is the defining economic transformation of the early twentieth century \citep{herrendorf2014}. The classic Lewis model predicts that surplus agricultural labor should flow smoothly to higher-productivity sectors \citep{lewis1954}, yet the persistence of large agricultural populations well past the point of apparent profitability suggests substantial barriers to occupational transition. Whether these barriers reflect information frictions, attachment costs, social networks, or simple risk aversion has been difficult to establish causally, because the shocks that displace agricultural workers rarely arrive exogenously \citep{munshi2003, bryan2014}.

This paper exploits a unique institutional feature of the World War I draft to identify the causal effect of forced occupational disruption on structural transformation. The Selective Service Act of 1917 required all men aged 21--30 to register for military service, and local draft boards inducted 2.8 million of them \citep{kennedy2004}. Crucially, the law applied only to citizens: foreign-born non-citizens were exempt from conscription. This creates a natural difference-in-differences design. Within each draft-eligible age cohort, I compare occupational outcomes of native-born men (who faced the draft) with foreign-born men (who did not), relative to the same comparison at ages too young for the first draft registration. The estimand is the differential occupational transition caused by draft exposure, identified from the nativity-based exemption.

I implement this design using the IPUMS Multigenerational Longitudinal Panel (MLP), which links individuals across the 1910 and 1920 full-count censuses \citep{price2022, ruggles2020}. The linked panel contains 4.9 million men aged 10--20 in 1910, with detailed information on occupation, farm residence, county, nativity, race, and literacy in both census years. This scale---enabled by machine-learning linking algorithms applied to the complete census universe \citep{abramitzky2021}---provides statistical power that prior studies of early-twentieth-century labor markets could not achieve.

The main findings are as follows. Draft-exposed native-born men were 1.6 percentage points more likely to exit farming between 1910 and 1920 than their foreign-born counterparts at the same ages (Table~\ref{tab:nativity_did}). This is a 13 percent increase over the foreign-born control mean. For occupational upgrading, the effect is larger: draft-exposed men gained 5.3 additional points on the IPUMS occupational income score, an effect of 0.43 standard deviations. Strikingly, draft exposure had no detectable effect on inter-county migration, suggesting that the mechanism operates through occupational transformation \textit{in place} rather than geographic reallocation.

These effects are concentrated precisely where theory predicts. In counties with above-median agricultural employment shares in 1910, the draft eligibility effect on farm exit was 2.9 percentage points larger than in low-agriculture counties, and the occupational score differential was 5.6 points larger (Table~\ref{tab:heterogeneity}). This is consistent with a disruption-as-inertia-break mechanism: where agricultural attachment is strongest, forced dislocation has the most room to operate. The heterogeneity by race tells a complementary story. Black draft-eligible men were 1.3 percentage points more likely to exit farming than their white counterparts, a result consistent with military service partially disrupting the racial barriers---tenancy contracts, geographic isolation, and occupational segregation---that kept Black workers disproportionately bound to Southern agriculture \citep{collins2014, hornbeck2014}.

I also estimate a supplementary regression discontinuity at the draft-eligibility age cutoff (men turning 21 in 1917 versus those turning 20). The age-based RD yields small, imprecise estimates for farm exit at the cutoff, and placebo tests at false cutoffs are also significant, indicating smooth age trends in occupational transition (Table~\ref{tab:robustness}). I interpret this as evidence that the age cutoff is confounded by the strong mechanical relationship between age and occupational change over the life cycle, and present the nativity-based DiD as the preferred specification.

This paper contributes to three literatures. First, it adds to the growing body of work on the economic consequences of military service \citep{angrist1990, angrist1994, bound2002, eriksson2019}. While \citet{angrist1990} famously showed that Vietnam-era military service \textit{reduced} civilian earnings, I find that WWI service \textit{increased} occupational attainment---a difference I attribute to the much larger occupational disruption involved in pulling men from subsistence agriculture versus the urban labor markets of the 1960s. Second, this paper contributes to the literature on structural transformation and labor reallocation \citep{caselli2001, herrendorf2014, hornbeck2012}. The conventional view emphasizes pull factors---productivity growth, urbanization, transportation---while I provide evidence for a push mechanism operating through forced disruption. Third, this paper adds to the literature on the Great Migration and racial economic convergence \citep{collins2014, collins2020}, showing that Black draft-eligible men disproportionately escaped agricultural employment, consistent with military service as a partial escape from the racialized labor institutions of the early twentieth century.

I call this the ``inertia break'' mechanism: forced occupational disruption permanently alters labor allocation by overcoming attachment costs that market incentives alone are insufficient to break. The finding that effects concentrate in the most agricultural counties and among the most constrained workers provides support for a model in which individuals are trapped in low-productivity equilibria not by preferences but by transaction costs, information barriers, and institutional constraints \citep{bryan2014, lagakos2018}.

\section{Institutional Background}

\subsubsection*{The Selective Service Act of 1917} The United States entered World War I on April 6, 1917. On May 18, Congress passed the Selective Service Act, requiring all male citizens and declarant aliens aged 21--30 to register for military service. The first registration occurred on June 5, 1917, enrolling approximately 10 million men. A second registration on June 5, 1918, captured men who had turned 21 since the first registration, and a third registration on September 12, 1918, expanded the age range to 18--45. In total, 24 million men registered across the three rounds, and approximately 2.8 million were inducted \citep{kennedy2004}.

\subsubsection*{Draft mechanics and the nativity exemption} Local draft boards determined the order of induction through a lottery system. The law initially applied only to citizens and ``declarant aliens'' (those who had filed their first naturalization papers). Foreign-born men who had not declared their intention to naturalize were exempt from registration and service. While the 1918 amendments expanded eligibility and some non-declarant aliens were later included, the core exemption for non-citizens during the critical first registration created a meaningful difference in draft exposure between native-born and foreign-born men of the same age \citep{swiencicki2020}.

\subsubsection*{Agricultural context in 1910} At the time of the 1910 census, agriculture employed approximately 31 percent of the civilian labor force, down from 41 percent in 1900 but still the single largest employment sector \citep{haines2001}. Farm wages were substantially below manufacturing wages, yet geographic and occupational mobility remained limited by transportation costs, information frictions, tenancy contracts, and social ties \citep{munshi2003}. The structural transformation from agriculture to manufacturing was well underway but proceeding unevenly across regions, with Southern and rural Midwestern counties retaining disproportionately agricultural labor forces.

\subsubsection*{Mechanism} Military service disrupted agricultural attachment through several channels. Inducted men were physically removed from farms, exposed to non-agricultural occupations (logistics, construction, manufacturing), and relocated to camps and ports across the country and overseas. Upon demobilization, these men possessed new skills, broader geographic awareness, and weakened ties to their pre-service communities---all factors that economic theory predicts should lower the cost of occupational transition \citep{lagakos2018, donaldson2018}. The key theoretical prediction is that these effects should be largest where pre-service attachment costs were highest: in the most agricultural counties and among the most constrained workers.

\section{Data}

I use the IPUMS Multigenerational Longitudinal Panel (MLP), which links individuals across decennial censuses using machine-learning algorithms applied to full-count census files \citep{price2022, abramitzky2021}. The linked 1910--1920 panel contains 43.9 million individuals, of whom I retain 4.9 million men aged 10--20 in 1910 who are linked to the 1920 census.

\subsubsection*{Outcome variables} I construct four measures of occupational transition. \textit{Farm exit} is an indicator equal to one if a man resided on a farm in 1910 but not in 1920 (IPUMS variable \texttt{FARM}). \textit{Occupational income score change} is the 1920 minus 1910 value of the IPUMS \texttt{OCCSCORE} variable, which maps OCC1950 occupation codes to median income values, providing a continuous measure of occupational upgrading. \textit{Inter-county migration} is an indicator for residing in a different county in 1920 than in 1910 (constructed from \texttt{STATEFIP} and \texttt{COUNTYICP}). \textit{Farm-to-manufacturing transition} indicates farming in 1910 and manufacturing-sector employment in 1920 (IND1950 codes 306--499).

\subsubsection*{Treatment variables} \textit{Draft eligible} is an indicator for being aged 14 or older in 1910, corresponding to men who turned 21 by the first draft registration in June 1917. \textit{Native born} is an indicator for IPUMS nativity codes 1--3 (native-born of native parents, native-born of foreign parents, native-born with unknown parental birthplace). Foreign-born men (nativity codes 4--5) constitute the control group for the nativity DiD, as non-citizen foreign-born men were exempt from the draft.

\subsubsection*{Summary statistics} Table~\ref{tab:summary} reports summary statistics by draft eligibility and nativity. Approximately 76.5 percent of men in the sample are native-born. Among draft-eligible native-born men, 37 percent lived on farms in 1910. Native-born men lived on farms at higher rates than foreign-born men, and raw farm exit rates are correspondingly higher for native-born men. However, the raw nativity gap conflates the draft effect with pre-existing differences in agricultural employment, motivating the DiD design that differences out the baseline nativity gap.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Nativity-Based Difference-in-Differences}

The primary identification strategy exploits the interaction between draft-age eligibility and the nativity-based draft exemption. I estimate:
\begin{equation}
Y_i = \alpha + \beta_1 D_i + \beta_2 N_i + \gamma (D_i \times N_i) + \delta f(a_i) + \mu_s + \varepsilon_i
\label{eq:did}
\end{equation}
where $Y_i$ is an occupational outcome for individual $i$, $D_i$ is an indicator for draft eligibility (age $\geq$ 14 in 1910), $N_i$ is an indicator for native-born status, $f(a_i)$ is a linear age control (centered at the cutoff), and $\mu_s$ are state fixed effects. The coefficient of interest is $\gamma$, which identifies the differential occupational transition of native-born men (exposed to the draft) relative to foreign-born men (exempt) at draft-eligible versus ineligible ages. Standard errors are clustered at the state level (49 clusters).

The identifying assumption is that, absent the draft, native-born and foreign-born men at draft-eligible ages would have experienced the same differential occupational transition as those at ineligible ages. This assumption could be violated if nativity-specific age trends in occupational mobility differ from those captured by the linear age control. Several features of the results provide support for the causal interpretation. First, the heterogeneity pattern---effects concentrated in high-agriculture counties and among Black men---is predicted by the inertia-break mechanism but not by generic nativity-age interactions. Second, the null effect on migration rules out geographic sorting as a confound. Third, the coefficient is precisely estimated and large enough to exceed plausible nativity-age trend differentials.

Two threats to the nativity comparison deserve explicit discussion. First, the draft also applied to ``declarant aliens''---foreign-born men who had filed naturalization papers---partially contaminating the control group. If a substantial fraction of foreign-born men were declarant aliens subject to the draft, the estimated effect would be \textit{attenuated}, meaning the true effect of draft exposure is likely larger than reported. Second, the MLP linking algorithm may have differential success rates by nativity. If foreign-born men who remained in agriculture were more likely to be linked across censuses (due to stable names and residences), this could inflate the estimated effect. The large sample and balanced age distributions across nativity groups provide some reassurance, but linking bias remains a limitation.

The coefficient $\gamma$ should be interpreted as an intent-to-treat (ITT) effect of draft \textit{exposure risk}, not as the effect of actual military service. Only approximately 28 percent of registrants were inducted \citep{kennedy2004}, so the effect per inducted man is likely several times larger than the ITT estimate. I do not observe individual-level induction status in the census data, and therefore cannot estimate a local average treatment effect.

\subsection{Supplementary Age-Based Regression Discontinuity}

As a supplementary analysis, I estimate a parametric RD at the draft-eligibility cutoff:
\begin{equation}
Y_i = \alpha + \tau T_i + \beta_1 (a_i - c) + \beta_2 T_i(a_i - c) + \mathbf{X}_i'\delta + \mu_s + \varepsilon_i
\label{eq:rd}
\end{equation}
where $T_i = \mathbf{1}[a_i \geq 14]$ indicates draft eligibility and $c = 14$ is the cutoff age in 1910. Because the running variable (age in integer years) is discrete, I use parametric specifications with varying bandwidths rather than nonparametric local polynomial methods \citep{lee2009, imbens2008}. I present the age-based RD primarily for transparency and to illustrate why the nativity DiD is preferred: the strong mechanical relationship between age and occupational transition over the life cycle generates significant effects at placebo cutoffs, contaminating the RD estimates.

\section{Results}

\subsection{Main Results: Nativity DiD}

Table~\ref{tab:nativity_did} reports the nativity DiD estimates from equation~\eqref{eq:did}. The coefficient on Draft Eligible $\times$ Native Born identifies the causal effect of draft exposure on occupational outcomes.

For farm exit (column 1), draft-exposed native-born men were 1.6 percentage points more likely to leave farming between 1910 and 1920 than foreign-born men of the same ages ($p < 0.001$). Against the control mean of 5.7 percent for non-draft-eligible foreign-born men, this represents a 28 percent increase in the farm exit rate.

The effect on occupational upgrading is economically large (column 2). Draft-exposed native-born men gained 5.35 additional points on the occupational income score compared to foreign-born men ($p < 0.001$). Given a control-group standard deviation of 12.4, this corresponds to a standardized effect of 0.43---a large effect by the standards of labor market interventions. For context, this is equivalent to the occupational gain associated with approximately four additional years of aging in the pre-draft cohorts.

Strikingly, draft exposure had no detectable effect on inter-county migration (column 3). The coefficient is 0.0015, small and statistically insignificant ($p = 0.56$). This null result is informative: it suggests that the mechanism is not geographic reallocation---men moving from rural to urban areas---but rather occupational transformation occurring \textit{within} existing locations. Draft-eligible native-born men changed what they did, not where they lived.

\input{tables/tab2_nativity_did}

\subsection{Mechanism: Where Inertia Binds}

If draft exposure breaks agricultural inertia, effects should concentrate where inertia is strongest. Table~\ref{tab:heterogeneity} tests this prediction.

\subsubsection*{Agricultural dependence} Panel A interacts draft eligibility with an indicator for counties with above-median agricultural employment shares in 1910. The draft eligibility effect on farm exit was 2.9 percentage points \textit{larger} in high-agriculture counties than in low-agriculture counties ($p < 0.001$), and the differential in occupational scores was 5.6 points ($p < 0.001$). In low-agriculture counties, the base effect of draft eligibility is $-1.6$ percentage points for farm exit---actually \textit{negative}---suggesting that in already-industrialized areas, the disruptive effect of military service may have pulled men temporarily out of advancing manufacturing careers. The interaction is thus doing exactly what a ``disruption breaks inertia'' mechanism predicts: the draft effect on occupational transition is large and positive where attachment costs are high, and absent or reversed where they are low.

\subsubsection*{Race} Panel B shows that Black draft-eligible men were 1.3 percentage points more likely to exit farming than white draft-eligible men ($p < 0.001$). This is consistent with military service partially disrupting the racialized institutions---sharecropping contracts, debt peonage, geographic isolation---that bound Black workers to Southern agriculture. The military offered Black men geographic mobility, non-agricultural skills, and a weakening of local ties, all of which the Great Migration literature identifies as prerequisites for occupational transition \citep{collins2014, hornbeck2014}.

\input{tables/tab3_heterogeneity}

\subsection{Supplementary Evidence: Age-Based RD}

Table~\ref{tab:age_rd} reports the age-based RD estimates from equation~\eqref{eq:rd}. The estimated discontinuity at the draft-eligibility cutoff is small and statistically insignificant for farm exit ($-0.001$, $p = 0.13$). For occupational income score change, the discontinuity is $-2.2$ ($p < 0.001$), suggesting that draft-eligible men experienced \textit{lower} occupational gains---the opposite of the nativity DiD finding. This sign reversal reflects the strong negative relationship between age and occupational change: older men gain less in occupational score because they are further from the child-to-adult transition. The RD cannot separate this mechanical age effect from the draft treatment.

\input{tables/tab4_age_rd}

\subsection{Robustness}

Table~\ref{tab:robustness} reports diagnostic tests. The McCrary density test yields a p-value of 0.27, consistent with no manipulation at the cutoff---as expected, since age in the census is not manipulable. Bandwidth sensitivity analyses for the age RD show that the farm exit coefficient is near zero and statistically insignificant at all bandwidths tested (3--6 years). The donut-hole RD excluding ages immediately adjacent to the cutoff yields similar null results.

Most informatively, placebo cutoff tests at ages 10, 11, and 12---all below the true draft-eligibility threshold---yield statistically significant coefficients for farm exit ($p < 0.001$). These false positives confirm that smooth age trends in occupational transition generate spurious effects at arbitrary cutoffs, providing further justification for preferring the nativity DiD as the primary identification strategy.

\input{tables/tab5_robustness}

\section{Discussion}

The central finding---that forced occupational disruption through military conscription accelerated the farm-to-factory transition---carries implications beyond the specific historical context.

\subsubsection*{Inertia versus preferences} A reader might interpret the persistence of agricultural employment as revealed preference: perhaps farm life offered compensating differentials (lower cost of living, family networks, autonomy) that offset the manufacturing wage premium. Under this interpretation, draft-induced farm exit would represent a welfare loss. Three features of the results argue against this view. First, the occupational income score gains are large (0.43 SD), suggesting that draft-displaced men ended up in substantially higher-paying occupations. Second, the absence of any geographic mobility effect rules out the explanation that these gains came at the cost of social dislocation. Third, the concentration of effects among Black men---the group most constrained by institutional rather than preference-based barriers---suggests that involuntary occupational change overcame institutional obstacles, not individual preferences.

\subsubsection*{Comparison with Vietnam-era estimates} \citet{angrist1990} found that Vietnam-era military service \textit{reduced} civilian earnings by approximately 15 percent. The contrast with the positive occupational effects found here is instructive. Vietnam-era draftees were drawn predominantly from urban labor markets where alternative civilian occupations offered higher returns than military service; their service represented a detour from an already-productive career trajectory. WWI draftees, by contrast, were disproportionately drawn from subsistence agriculture---a sector offering low returns and high inertia. For this population, military service was not a detour but a bridge: it provided the forced disruption necessary to overcome attachment costs and access higher-return occupations.

\subsubsection*{External validity} The inertia-break mechanism identified here---forced disruption overcoming attachment costs---is relevant to developing economies today where large populations remain in subsistence agriculture despite the existence of higher-productivity alternatives \citep{bryan2014, lagakos2018}. The finding that effects concentrate where attachment costs bind most (high-agriculture counties, racially constrained workers) suggests that policies targeting these specific margins---subsidized migration, skills exposure programs, or disruption of exploitative labor institutions---may be more effective than broad-based interventions.

\section{Conclusion}

Four in ten young American men lived on farms in 1910. The World War I draft pulled millions of them into military service, and those who were pulled---native-born men subject to conscription---were measurably more likely to leave agriculture and enter higher-paying occupations than their foreign-born peers who were exempt. The effect was not about moving to new places; it was about doing new things. And it was strongest where it should have been strongest: in the most agricultural counties and among the most constrained workers. Agricultural inertia is not an immutable feature of developing economies. It can be broken---and World War I broke it. Whether the effects persisted beyond 1920 and whether voluntary disruption programs could replicate what conscription achieved are questions for future work with longer panels.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
