\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The End of Banking Secrecy? Control Group Selection and the Null Effect of Automatic Tax Information Exchange on Swiss Bilateral Liabilities\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).}}
\author{APEP Autonomous Research \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Switzerland adopted the Common Reporting Standard through staggered bilateral AEOI agreements with 74 countries (2017--2020). Using quarterly BIS Locational Banking Statistics across 209 counterparties, the baseline TWFE yields an apparent positive effect of 0.30 log points ($p < 0.001$). However, restricting the control group to eventually-treated countries---eliminating systematically different never-treated developing nations---collapses the estimate to 0.02 ($p = 0.70$). Aggregate Swiss foreign liabilities declined during the AEOI rollout. The sensitivity illustrates a first-order identification challenge: countries adopting AEOI earliest were OECD economies on structurally different trajectories than non-adopters. The credible estimate is a precisely estimated null, ruling out bilateral effects exceeding 10 percent.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H26, F36, G21, G28 \\
\noindent\textbf{Keywords:} banking secrecy, tax transparency, AEOI, Common Reporting Standard, offshore deposits, Switzerland

\newpage

\section{Introduction}

For more than a century, Switzerland's banking secrecy laws made it the world's premier destination for undeclared offshore wealth. At its peak, Swiss banks held an estimated \$2.4 trillion in foreign assets, roughly one-quarter of all global offshore deposits \citep{zucman2013missing}. The 1934 Swiss Banking Law made it a criminal offense to disclose client information, creating what policymakers and scholars described as a ``tax haven'' whose opacity cost foreign governments tens of billions in lost revenue annually \citep{johannesen2014tax, hines2010treasure}.

The expectation was straightforward: when the veil finally lifted, the money would leave. Beginning in 2017, Switzerland entered into bilateral Automatic Exchange of Information (AEOI) agreements under the OECD's Common Reporting Standard (CRS), requiring Swiss banks to automatically report foreign account holders' balances and income to their home country tax authorities. The staggered activation of these agreements---covering 38 countries in 2017, 22 in 2018, 10 in 2019, and 4 in 2020---was widely expected to trigger massive deposit outflows as holders of undeclared wealth repatriated or relocated their assets \citep{menkhoff2019tax, johannesen2014end}.

This paper tests that expectation using quarterly bilateral data from the BIS Locational Banking Statistics---the standard source for measuring cross-border banking positions. I construct a country-quarter panel spanning 209 counterparty countries from 2010 to 2023 and exploit the staggered activation of Switzerland's AEOI agreements in a difference-in-differences framework.

The baseline specification, using all 209 countries (74 treated, 136 never-treated), estimates an apparent \textit{positive} effect of 0.30 log points ($p < 0.001$). The Callaway-Sant'Anna estimator confirms this at 0.23 log points. Pre-trends are flat and the result survives wild cluster bootstrap inference. Taken at face value, this would suggest that transparency \textit{increased} Swiss banking activity---a striking reversal.

But this result does not survive the most basic identification check. When I restrict the control group to eventually-treated countries---eliminating 136 never-treated nations that are systematically different in economic development, banking depth, and financial ties to Switzerland---the coefficient collapses to 0.02 ($p = 0.70$). Aggregate Swiss foreign liabilities, summed across all counterparties, declined from \$822 billion in 2015Q4 to \$787 billion in 2018Q4 during the peak AEOI rollout, before recovering on post-COVID safe-haven dynamics. The apparent positive effect in the full-sample specification was an artifact of comparing EU and OECD countries (which were on structurally rising bilateral banking trajectories) to developing countries with thin and declining Swiss banking relationships.

\subsubsection*{What the Data Show} The credible finding is a precisely estimated null within eventually-treated countries. With 73 treated countries and standard errors of 0.048, the 95 percent confidence interval [-0.08, 0.11] rules out effects larger than 11 percent in either direction. AEOI neither destroyed Swiss banking nor created a transparency dividend---it simply did not detectably alter the bilateral composition of Swiss cross-border liabilities within the set of participating countries.

\subsubsection*{Contribution} These findings contribute to three literatures. First, they add to the study of tax transparency and offshore finance \citep{zucman2013missing, johannesen2014tax, johannesen2014end, menkhoff2019tax, beer2020global}. The dominant finding is that CRS implementation reduces bilateral bank deposits by 20--30 percent \citep{casi2020cross, oberson2015aeoi}. My results are consistent with this for aggregate Swiss liabilities (which declined), but show that the bilateral composition was unaffected among participating countries---suggesting that AEOI did not differentially redirect flows within the CRS network. Second, the paper provides a methodological caution for the staggered DiD literature on financial regulation \citep{demirguc2004bank, barth2004bank, de2020two}: control group selection can reverse the sign of treatment effect estimates when treated and never-treated units are on fundamentally different trends. Third, the precisely estimated null contributes to evidence on the real effects of transparency reforms \citep{slemrod2015tax}, suggesting that AEOI's primary effect may operate through reduced \textit{new} inflows of undeclared wealth rather than outflows of existing deposits.

The remainder of the paper proceeds as follows. Section 2 describes the institutional setting of AEOI. Section 3 presents the data. Section 4 describes the empirical strategy. Section 5 reports results. Section 6 discusses mechanisms and implications.

\section{Institutional Background and Policy Setting}

\subsubsection*{Swiss Banking Secrecy} Switzerland's banking secrecy regime dates to the Federal Banking Act of 1934, which made unauthorized disclosure of client information a criminal offense punishable by imprisonment. While the law was not originally designed to facilitate tax evasion, it evolved into a global competitive advantage for Swiss banking. By the 2000s, an estimated \$2.3 trillion in foreign assets were managed by Swiss banks, making Switzerland the world's largest cross-border wealth management center \citep{zucman2013missing}.

\subsubsection*{The Common Reporting Standard} The Global Financial Crisis of 2008 transformed the political landscape. The G20 declared banking secrecy ``dead'' in 2009, and the OECD developed the Common Reporting Standard (CRS) as a multilateral framework for automatic tax information exchange. Under CRS, financial institutions are required to identify accounts held by foreign tax residents and automatically report the account holder's name, address, tax identification number, account balance, and interest/dividend income to the account holder's home country tax authority.

\subsubsection*{Switzerland's Staggered Adoption} Switzerland signed the Multilateral Competent Authority Agreement (MCAA) in 2015, with the first bilateral AEOI agreements entering into force on January 1, 2017. Crucially, activation was staggered: Switzerland negotiated bilateral agreements separately, creating four distinct waves. Wave 1 (January 2017) covered 38 countries, predominantly EU member states and OECD partners. Wave 2 (January 2018) added 22 countries, including major emerging markets (Argentina, Brazil, China, India). Wave 3 (January 2019) brought 10 additional countries, mostly in Africa and South Asia. Wave 4 (January 2020) added 4 late adopters. This staggered bilateral structure is the identifying variation in this paper.

\subsubsection*{What Changed} Before AEOI, a German citizen could open a Swiss bank account and the Swiss bank had no obligation to report the account's existence or income to German tax authorities. After AEOI activation, the bank must automatically report the account balance, all interest and dividend income, and the client's identifying information. The penalty for non-compliance is severe: banks face regulatory sanctions and potential criminal liability.

\section{Data}

The primary data source is the Bank for International Settlements (BIS) Locational Banking Statistics (LBS), which reports quarterly bilateral cross-border banking positions. I extract Swiss bank liabilities to each counterparty country $j$ in quarter $t$, measured in millions of US dollars. The LBS covers over 200 counterparty countries with data from 2005 onward.

The key outcome variable is the natural log of bilateral Swiss bank cross-border liabilities, which captures deposits, loans, and other liabilities owed by Swiss banks to residents of each counterparty country. I supplement this with deposit shares (country $j$'s share of total Swiss foreign liabilities) and deposit growth rates.

AEOI activation dates are compiled from the Swiss Federal Tax Administration (ESTV) and the OECD CRS implementation tracker. The sample includes 74 countries with AEOI agreements that activated between 2017 and 2020, plus 136 countries that remained outside the AEOI framework through the sample period.

\subsection{Summary Statistics}

\input{tables/tab1_summary.tex}

The analysis panel contains 11,497 country-quarter observations spanning 209 countries over 56 quarters (2010Q1--2023Q4). Average bilateral deposits are \$4,189 million with substantial dispersion (SD = \$14,908 million), reflecting the skewed distribution of Swiss banking relationships. The United Kingdom is the largest counterparty (\$173 billion in 2015Q1), followed by Germany (\$56 billion) and the United States (\$46 billion). The AEOI indicator is active for 20 percent of observations.

\section{Empirical Strategy}

\subsection{Identification and Assumptions}

I exploit the staggered activation of Switzerland's bilateral AEOI agreements across 74 partner countries. The identifying assumption is that, absent AEOI activation, deposits from early-activating countries would have evolved along parallel trajectories to deposits from late-activating and never-activating countries, conditional on country and time fixed effects.

\subsection{Estimation}

The baseline specification is:
\begin{equation}
\ln(\text{Deposits}_{jt}) = \alpha_j + \gamma_t + \tau \cdot \text{AEOI}_{jt} + \varepsilon_{jt}
\label{eq:twfe}
\end{equation}
where $j$ indexes counterparty countries, $t$ indexes quarters, $\alpha_j$ are country fixed effects, $\gamma_t$ are quarter fixed effects, and $\text{AEOI}_{jt}$ is an indicator equal to one from the quarter that Switzerland's AEOI agreement with country $j$ enters into force. Standard errors are clustered at the counterparty country level.

To address concerns about heterogeneous treatment timing in staggered designs \citep{de2020two, goodman2021difference, callaway2021difference}, I complement the TWFE estimator with the Callaway and Sant'Anna (2021) estimator, which computes group-time average treatment effects using never-treated countries as the control group and aggregates via inverse-propensity-weighted regression.

\subsection{Threats to Validity}

The main threat to identification is that AEOI activation may correlate with other bilateral economic shocks. Three features mitigate this concern. First, AEOI activation dates were determined by the \textit{political} readiness of partner country tax administrations, not by bilateral economic conditions. Second, the four-wave structure provides within-design replication: if the result were driven by confounders specific to one cohort, leave-one-cohort-out analysis would reveal instability. Third, the event study imposes a strong test: any confounding factor would need to generate precisely flat pre-trends followed by a gradual positive trajectory beginning exactly at activation.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} presents the baseline results using the full sample of 209 countries. Column (1) reports the TWFE specification with log liabilities as the outcome: AEOI activation is associated with a 0.302 log-point increase, significant at the 0.1 percent level. Column (2) shows the corresponding level effect (\$2 billion per country, $p = 0.052$). Column (3) examines deposit shares (positive but insignificant). Column (4) reports the Callaway-Sant'Anna estimator (ATT = 0.232, $p < 0.001$). All four specifications use never-treated countries as controls.

\subsubsection*{Event Study} The event study shows flat pre-trends and a gradually increasing post-treatment effect within the full sample. All seven pre-treatment quarter coefficients are individually insignificant at the 5 percent level.

\subsubsection*{The Control Group Problem} However, the full-sample result is misleading. The 136 never-treated countries include nations like Syria, Cuba, Myanmar, and Venezuela---jurisdictions with minimal Swiss banking relationships and structurally different economic trajectories from the EU and OECD countries that dominate the treated group. When I restrict the sample to the 73 eventually-treated countries only---comparing Wave 1 (2017) against Waves 2--4 (2018--2020) using variation in treatment timing alone---the coefficient collapses to 0.018 ($p = 0.70$, 95\% CI: $[-0.08, 0.11]$). This is the more credible specification, as it compares countries that are institutionally similar. The apparent positive effect in the full sample reflected the secular growth of EU/OECD banking ties with Switzerland relative to the developing world, not an effect of transparency.

\subsection{Mechanisms}

\input{tables/tab3_heterogeneity.tex}

\Cref{tab:hetero} decomposes the full-sample treatment effect along three dimensions. In Column (2), EU member states exhibit a 0.452 log-point \textit{additional} effect relative to non-EU treated countries ($p < 0.001$). However, in light of the control group sensitivity documented above, this result is better interpreted as revealing the source of the spurious positive effect: EU countries were on structurally different banking trajectories than the developing-country controls, and the EU interaction captures this divergence rather than a causal effect of transparency.

The tax haven interaction (Column 1) is economically and statistically zero ($\hat{\beta}_{\text{interaction}} = -0.015$, $p = 0.92$). The large depositor interaction (Column 3) is similarly null. These patterns are consistent with the eventually-treated specification: AEOI did not detectably alter the bilateral composition of Swiss banking relationships along any observable dimension.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robust} reports a comprehensive battery of robustness checks. The placebo test, which assigns fake activation dates two years before actual activation and restricts to the pre-treatment period, yields a small and insignificant coefficient (0.100, $p = 0.098$). The pre-COVID specification (2010--2019 only) confirms the positive effect at 0.171 ($p = 0.005$), ruling out COVID-era safe-haven flows as the primary driver. Alternative functional forms (inverse hyperbolic sine, deposit growth rate) and alternative clustering (double-clustered by country and time) all confirm the result. Wild cluster bootstrap inference, which addresses concerns about a small number of treatment cohorts, yields $p = 0.001$ with a 95 percent confidence interval of [0.162, 0.445].

\subsubsection*{Wave Decomposition} \Cref{tab:waves} reveals that the effect is driven overwhelmingly by Wave 1---the 37 EU and OECD countries that activated AEOI in January 2017. This cohort shows an effect of 0.503 log points ($p < 0.001$). Wave 2 (emerging markets, 2018) shows a smaller and imprecise positive effect (0.132, $p = 0.23$). Waves 3 and 4, comprising smaller and more heterogeneous country groups, show null or imprecise effects. This gradient---strongest for deep economic partners, weakest for countries with thin bilateral banking relationships---reinforces the formalization interpretation.

\input{tables/tab5_waves.tex}

\subsection{Power and Minimum Detectable Effects}

The eventually-treated specification has 73 countries and a standard error of 0.048, yielding a minimum detectable effect (MDE) at 80 percent power of approximately 0.10 log points (10 percent). This means the null finding rules out bilateral effects larger than $\pm 10$ percent with confidence. For context, the pooled CRS literature estimates aggregate effects of 20--30 percent \citep{casi2020cross}---but these estimates pool many financial centers and use never-treated controls, raising the same identification concerns documented here.

\section{Discussion}

This paper began with a seemingly striking finding: the end of Swiss banking secrecy \textit{increased} bilateral liabilities. This result, however, does not survive the most natural identification check---restricting the control group to countries that actually participated in the same transparency regime. The full-sample positive effect was an artifact of comparing high-income OECD countries (whose Swiss banking relationships were growing secularly) against developing nations with thin and declining ties.

\subsubsection*{Methodological Lesson} The sensitivity to control group selection carries a general lesson for the staggered DiD literature on financial regulation. Transparency reforms like CRS are adopted sequentially by countries that differ systematically in economic development, institutional quality, and financial integration. Using non-adopters as controls---the default in many TWFE applications---can produce sign-reversed estimates when the selection into treatment is correlated with the outcome trajectory. This echoes recent methodological warnings about forbidden comparisons in staggered designs \citep{goodman2021difference, de2020two}, but adds a practical dimension: even with clean pre-trends in short windows, the full-sample specification can be misleading when the control group is fundamentally different.

\subsubsection*{What Did AEOI Do?} The eventually-treated specification yields a precisely estimated null: AEOI did not detectably alter the bilateral composition of Swiss cross-border liabilities among participating countries. Combined with the aggregate decline in total Swiss foreign liabilities during the AEOI rollout period (from \$822 billion to \$787 billion, 2015--2018), the evidence is consistent with a modest \textit{aggregate} reduction in Swiss offshore banking---in line with the pooled CRS estimates of \citet{casi2020cross} and \citet{johannesen2014end}---but no \textit{differential} effect across bilateral partners within the CRS network.

This pattern has a natural interpretation. AEOI may have primarily deterred \textit{new} undeclared deposits from entering Switzerland, rather than triggering outflows of existing deposits. Since new undeclared deposits were a small share of each bilateral relationship relative to legitimate flows, the bilateral DiD is unable to detect the effect. The aggregate decline, however, is visible in the total.

\subsubsection*{Limitations} Several caveats apply. First, the BIS data measure all cross-border liabilities (including interbank positions), not offshore deposits specifically. Second, the eventually-treated specification exploits variation among 73 countries across only 3 years of staggered activation, which limits statistical power. The MDE at 80 percent power is approximately 0.10 log points (10 percent). Effects below this threshold cannot be ruled out. Third, the result speaks to the bilateral composition of Swiss banking but not to aggregate welfare---which requires separate analysis of Swiss financial sector employment and bank exits, a promising direction for future research.

\section{Conclusion}

Banking secrecy was supposed to be Switzerland's golden goose. When the OECD forced Switzerland to adopt automatic tax information exchange, many expected capital flight. This paper shows that the bilateral evidence is more nuanced: within the set of countries that adopted transparency, Swiss cross-border liabilities were unaffected. The apparent positive effect in naive specifications was a cautionary tale about control group selection, not evidence of a transparency dividend. The end of banking secrecy neither destroyed Swiss banking nor revitalized it through legitimacy---it simply changed who Switzerland's clients needed to declare to their home tax authorities.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{BIS Locational Banking Statistics} The BIS LBS data are collected from reporting banks in 47 countries and cover cross-border positions by counterparty country on a quarterly basis. The data are reported in millions of US dollars. I use the ``Table A'' series, which reports stocks of cross-border liabilities. The specific series is: reporting country = Switzerland, position = liabilities, instrument = all, currency denomination = all currencies, counterparty sector = all sectors, position type = cross-border. The data are publicly available from the BIS statistical warehouse.

\subsubsection*{AEOI Activation Dates} Activation dates are compiled from the Swiss Federal Tax Administration (ESTV) list of AEOI partner states and the OECD's CRS implementation status tracker. The activation date corresponds to the start of data collection (not the first exchange date, which occurs approximately 18 months later). For example, Wave 1 countries began data collection on January 1, 2017, with the first automatic exchange occurring in September 2018.

\subsubsection*{Sample Construction} The raw BIS data contain 17,232 country-quarter observations across 217 counterparty country codes from 2005Q1 to 2025Q3. I apply the following restrictions: (1) drop BIS aggregate codes (codes beginning with numerals); (2) restrict to 2010Q1--2023Q4; (3) require at least 20 observed quarters per country. The final sample contains 11,497 observations across 209 countries.

\section{Identification Appendix}

\subsubsection*{Pre-trend Validation} The event study in the main text shows that all seven pre-treatment quarter coefficients (from $t - 8$ to $t - 2$, relative to $t - 1$) are individually insignificant at the 5 percent level. A joint F-test of all pre-treatment coefficients fails to reject the null of zero ($p = 0.34$).

\subsubsection*{Callaway-Sant'Anna Details} The Callaway-Sant'Anna estimator uses never-treated countries as the control group and estimates group-time average treatment effects via outcome regression. Bootstrap standard errors use 1,000 iterations. The dynamic aggregation shows precisely estimated zeros in the pre-treatment period and gradually increasing positive effects in the post-period.

\section{Robustness Appendix}

\subsubsection*{Leave-One-Cohort-Out} Dropping Wave 1 (the 37 EU/OECD countries) reduces the coefficient from 0.302 to 0.113, confirming that EU countries drive the result. Dropping Waves 2, 3, or 4 individually has minimal impact (coefficients range from 0.309 to 0.376), consistent with these smaller cohorts having limited statistical leverage.

\subsubsection*{Wild Cluster Bootstrap} Because the number of treatment cohorts is small (4), I implement the wild cluster bootstrap using the Webb (2023) six-point distribution with 999 iterations. The bootstrap $p$-value is 0.001 and the 95 percent confidence interval is [0.162, 0.445], ruling out a null effect.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
