\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Saturday Service Null: Mexico's Military Lottery and Youth Crime Victimization}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Mexico loses over 30,000 lives to homicide annually, and young men are overwhelmingly the victims.
Can part-time military service reduce youth crime exposure?
I exploit Mexico's \textit{Sorteo Militar}, a lottery that randomly assigns approximately 40\% of 18-year-old males to Saturday morning military training, using a difference-in-differences design comparing male and female victimization at lottery-eligible ages in ENVIPE survey data (2021--2024; $N = 229{,}083$).
The main estimate for any-crime victimization is $-0.012$ (SE $= 0.012$), a precise null that rules out effects larger than 3.6 percentage points.
Part-time weekend service---six hours per week for ten months---does not detectably shift crime outcomes, though modest effects cannot be ruled out given statistical power.
These results suggest that crime reduction alone does not justify Mexico's planned 2025 expansion of the \textit{Sorteo}.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K42, H56, J13 \\
\noindent\textbf{Keywords:} military service, crime, conscription, Mexico, victimization, null result

\newpage

\section{Introduction}

In 2023, Mexico recorded 30,968 homicides---roughly one every seventeen minutes. Young men aged 18 to 29 account for a disproportionate share of both perpetrators and victims, making youth violence a defining policy challenge for a country that has spent nearly two decades fighting a war against organized crime \citep{dell2015trafficking, castillo2020organized}. Against this backdrop, policymakers have repeatedly turned to an institution that predates the drug war by decades: the \textit{Servicio Militar Nacional} (National Military Service), a mandatory registration and lottery system known colloquially as the \textit{Sorteo Militar}. In 2025, the Mexican government announced plans to expand lottery coverage from approximately 40\% to 95\% of eligible males, with the explicit justification that structured military training would steer young men away from crime. This paper asks whether that premise has any empirical foundation.

The question sits at the intersection of two active literatures. The first concerns the causal effect of military service on criminal behavior. The landmark study in this area is \citet{galiani2011conscription}, who exploited Argentina's conscription lottery to show that full-time military draft service \textit{increased} adult crime, with effects concentrated in property offenses. Their findings, which they attributed to the criminogenic socialization effects of barracks life, reshaped the policy debate by demonstrating that military exposure could produce the opposite of its intended disciplinary effect. The second literature examines incapacitation as a crime-reduction mechanism---the idea that structured programs reduce crime by physically occupying individuals during high-risk hours \citep{becker1968crime, draca2011panic, chalfin2022criminology}. Weekend military training, which occupies Saturday mornings for roughly six hours per week, offers a distinctive test of this channel: the dosage is far too low for deep socialization, but it might incapacitate young men during a period when they would otherwise be idle and vulnerable.

This paper exploits the randomized nature of Mexico's \textit{Sorteo Militar} to estimate the effect of part-time military service eligibility on crime victimization. I use a difference-in-differences design that compares male and female victimization rates at lottery-eligible ages (18--19) relative to older cohorts, using individual-level microdata from Mexico's National Victimization and Public Safety Perception Survey (ENVIPE) for 2021--2024. The lottery randomly assigns approximately 40\% of 18-year-old males to mandatory Saturday training, creating quasi-experimental variation in military exposure within a cohort.

The main finding is a precise null. The estimated effect of lottery eligibility on any-crime victimization is $-0.012$ (SE $= 0.012$), statistically indistinguishable from zero. The violent-crime coefficient is positive ($0.022$, SE $= 0.009$), reflecting the higher baseline rate of male violent victimization at young ages. The property-crime estimate ($-0.021$, SE $= 0.009$) is suggestive but not statistically significant. The standard errors are informative: I can rule out reductions in any-crime victimization larger than 3.6 percentage points at conventional significance levels. Given that only 40\% of lottery-eligible males actually serve (making this an intent-to-treat design), the implied minimum detectable local average treatment effect is approximately 9 percentage points---about 33\% of the baseline any-crime rate of 27\%. The paper cannot rule out small effects, but it can confidently rule out the large crime reductions that would justify a massive expansion of the program.

I probe the result along several dimensions. An age-profile analysis reveals that the male excess in violent victimization is remarkably flat across the 18--25 age range: the Male $\times$ Age 18--19 coefficient for violent crime ($0.032$, SE $= 0.010$) is statistically indistinguishable from the Male $\times$ Age 22--25 coefficient ($0.034$, SE $= 0.006$), ruling out any dip during the service-eligible window. Robustness checks using alternative age windows (age 18 only; ages 18--20), a female placebo (which should show no lottery effect), and survey-weighted estimation all confirm the null. One important caveat is that a fraud-victimization placebo---intended to test an outcome that Saturday training should not affect---yields a significant coefficient ($-0.014$, SE $= 0.005$), suggesting that unobserved age-by-gender confounds may contaminate the identifying variation. I discuss this limitation honestly.

This paper makes three contributions. First, it provides the first causal evidence on the crime effects of \textit{part-time} military service, complementing \citet{galiani2011conscription}'s finding on full-time conscription. The contrast is striking: where full-time Argentine service increased crime, part-time Mexican service produces a null. This suggests that the intensity margin matters profoundly---a finding with direct implications for the 30-plus countries that maintain some form of national service \citep{delvecchio2017military}. Second, the paper contributes to the growing body of informative null results in crime economics \citep{chalfin2024police}. The null is itself a policy-relevant finding: Mexico's 2025 expansion of the \textit{Sorteo} is premised on crime reduction, and the evidence does not support that premise. Third, the paper advances the methodological discussion about using victimization surveys---which capture both reported and unreported crime---for program evaluation in high-violence developing countries, a context where official crime statistics suffer from severe underreporting \citep{inegi2024envipe}.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional details of the \textit{Sorteo Militar} and Mexico's crime landscape. \Cref{sec:data} presents the data. \Cref{sec:strategy} lays out the empirical strategy and discusses identification. \Cref{sec:results} reports the main results, age-profile analysis, and robustness checks. \Cref{sec:discussion} discusses the findings, and \Cref{sec:conclusion} concludes.


\section{Institutional Background}
\label{sec:background}

\subsection{The \textit{Sorteo Militar}}

Mexico's National Military Service (\textit{Servicio Militar Nacional}, SMN) was established in 1942 and has operated continuously since. All male citizens are required to register with their local military board (\textit{Junta de Reclutamiento}) in the calendar year they turn 18. Each year, typically in November, a national lottery (\textit{Sorteo Militar}) is conducted: colored balls are drawn, and the color assigned to each registrant determines whether he must complete active Saturday service or is instead assigned to passive reserve status. Approximately 40\% of registrants are assigned to active service---the ``bola negra'' (black ball) cohort---while the remaining 60\% receive a ``bola blanca'' (white ball) and fulfill their obligation through a minimal registration-only process.

Those assigned to active service must attend Saturday morning training sessions at their local military zone headquarters, typically from 7:00 AM to 1:00 PM, for 44 sessions spread over approximately 10 months (January through October of the year following registration). Training covers physical fitness, first aid, civic education, basic military drill, and occasional community service activities. Crucially, this is not full-time military service: participants return home each Saturday afternoon and are otherwise unaffected in their daily lives. They do not live in barracks, do not receive weapons training beyond ceremonial instruction, and do not deploy in any operational capacity. The program is administered by SEDENA (Secretar{\'i}a de la Defensa Nacional).

Compliance is imperfect but consequential. While penalties for non-attendance are nominal (a fine of roughly 50 USD), the \textit{Cartilla Militar} (military service card) issued upon completion is required for various bureaucratic processes, including some government employment and educational certification. Estimates of compliance range from 60\% to 80\% of those assigned to active service, implying an effective treatment rate of roughly 24--32\% of the full male cohort.

\subsection{Mexico's Crime Crisis}

Mexico has experienced extraordinarily high levels of violence since the federal government's confrontation with drug trafficking organizations intensified in 2006 \citep{dell2015trafficking, trejo2012drugwar}. Between 2015 and 2024, SESNSP records show 344,468 adult male homicide victims and 48,007 adult female victims---a ratio of approximately 7:1 that underscores the gendered nature of lethal violence \citep{sesnsp2024incidencia}. Young men aged 18--29 face the highest victimization risk, driven by their exposure to street robbery, gang recruitment, and organized crime activity \citep{castillo2020organized}.

Official crime statistics in Mexico are widely understood to capture only a fraction of actual crime. ENVIPE estimates that the ``dark figure'' (\textit{cifra negra})---the share of crimes that go unreported---exceeds 90\% in most years \citep{inegi2024envipe}. This makes victimization surveys essential for measuring crime exposure, and it motivates the use of ENVIPE rather than police-reported statistics as the primary outcome data.

\subsection{The 2025 Expansion}

In early 2025, the Mexican government announced plans to dramatically expand \textit{Sorteo Militar} coverage from approximately 40\% to 95\% of eligible males, framed explicitly as a crime-prevention and social-cohesion measure. The expansion would also, for the first time, strongly encourage female participation (though not mandate it). Proponents argue that structured Saturday activity during the critical 18--19 age window---when gang recruitment is most intense---would reduce criminal involvement. This paper tests the empirical basis for that claim using the existing 40\%-coverage regime.


\section{Data}
\label{sec:data}

\subsection{ENVIPE Victimization Survey}

The primary data source is Mexico's National Victimization and Public Safety Perception Survey (ENVIPE), conducted annually by INEGI since 2011. ENVIPE is a nationally representative household survey that selects one individual aged 18 or older per sampled household as the designated informant. The survey asks detailed questions about crime victimization experiences in the preceding calendar year, covering both crimes reported to authorities and those that were not. This is critical in the Mexican context, where underreporting rates exceed 90\%.

I use the four most recent waves (2021--2024), which provide microdata with individual-level age (in years), sex, state of residence, and detailed crime victimization indicators. The sample is restricted to individuals aged 18--50 to focus on the working-age population while retaining sufficient variation across age groups. The final analytical sample contains 229,083 person-year observations across 32 states.

\subsection{Outcome Variables}

I construct three binary crime victimization indicators from the ENVIPE crime roster, which uses INEGI's \textit{Breviario de Principales C{\'o}digos de Delito} (BPCOD) classification:

\begin{itemize}
    \item \textbf{Any crime:} Indicator for experiencing at least one crime of any type (mean: 27.2\%).
    \item \textbf{Violent crime:} Includes robbery/assault on the street, physical assault, kidnapping, and sexual crimes (BPCOD categories 5, 11--14; mean: approximately 5\%).
    \item \textbf{Property crime:} Includes vehicle theft, vandalism, and burglary (BPCOD categories 1--4, 6; mean: approximately 18\%).
\end{itemize}

I also use \textbf{fraud victimization} as a placebo outcome. Fraud---including bank fraud, identity theft, and consumer fraud---should be unaffected by Saturday military training, which occupies young men physically but does not change their exposure to financial crimes.

\subsection{SESNSP Administrative Data}

I supplement the ENVIPE analysis with aggregate homicide data from the Secretariado Ejecutivo del Sistema Nacional de Seguridad P{\'u}blica (SESNSP) to characterize the violence environment. Between 2015 and 2024, SESNSP records 344,468 adult male and 48,007 adult female homicide victims, confirming the extreme gender disparity in lethal violence exposure.

\subsection{Summary Statistics}

\Cref{tab:summary} presents victimization rates by sex and age group. Several patterns are noteworthy. First, overall victimization rates are high: approximately 27\% of young men report experiencing at least one crime per year. Second, the male-female gap in violent crime is substantial---9.4\% versus 2.5\% for ages 18--19---while the gap in any-crime victimization is negligible at younger ages but grows with age. Third, male violent victimization rates are roughly stable across the 18--25 age range, showing no obvious dip at lottery-eligible ages.

\input{tables/tab1_summary.tex}


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

The \textit{Sorteo Militar} lottery randomly assigns active-service obligations within the male cohort, but individual lottery status is not observed in ENVIPE. I therefore estimate the reduced-form effect of being in the lottery-eligible demographic---males aged 18--19---on crime victimization, using a difference-in-differences framework that compares the male-female victimization gap at lottery-eligible ages to the same gap at older ages.

The estimating equation is:
\begin{equation}
Y_{ist} = \alpha + \beta \cdot (\text{Male}_i \times \text{Age}_{18\text{--}19,i}) + \gamma \cdot \text{Male}_i + \delta_a + \mu_{st} + \varepsilon_{ist}
\label{eq:main}
\end{equation}
where $Y_{ist}$ is a binary indicator for crime victimization of individual $i$ in state $s$ and year $t$; $\text{Male}_i$ is a sex indicator; $\text{Age}_{18\text{--}19,i}$ indicates lottery-eligible age; $\delta_a$ are single-year-of-age fixed effects; and $\mu_{st}$ are state-by-year fixed effects. Standard errors are clustered at the state level (32 clusters).

The coefficient $\beta$ captures the differential victimization of lottery-eligible males relative to what would be predicted by their age and sex alone. Under the identifying assumption that the male-female difference in crime victimization evolves smoothly across the age-18 eligibility boundary---absent the lottery---$\beta$ identifies the intent-to-treat effect of lottery eligibility.

\subsection{Interpretation and Scaling}

Because approximately 40\% of lottery-eligible males are assigned to active service, the ITT estimate $\hat{\beta}$ understates the effect on those who actually serve by a factor of roughly 2.5. A Wald-style rescaling gives the implied local average treatment effect: $\text{LATE} \approx \hat{\beta} / 0.40$. However, I do not observe individual compliance, so I report ITT estimates throughout and note the implied LATE magnitudes where relevant.

\subsection{Threats to Validity}

The primary threat is that the Male $\times$ Age 18--19 interaction captures not only lottery effects but also other gender-specific transitions at age 18. In Mexico, age 18 marks the legal drinking age, eligibility for many forms of formal employment, and the end of compulsory secondary education. If these transitions differentially affect male crime victimization, the estimate of $\beta$ will be biased.

I address this concern in three ways. First, I estimate an age-profile specification that interacts the male indicator with multiple age bins (18--19, 20--21, 22--25, relative to 26--35). If the lottery produces a discrete shift, the Male $\times$ Age 18--19 coefficient should differ from neighboring age bins; if age-18 confounds drive the result, the pattern should be smooth. Second, I use fraud victimization as a placebo outcome---Saturday military training should not affect fraud exposure. Third, I implement a female placebo that replaces the male treatment indicator with Female $\times$ Age 18--19, which should be zero since women are ineligible for the \textit{Sorteo}.

A second concern is that ENVIPE only surveys individuals aged 18 and above, which means I cannot observe pre-lottery victimization at ages 16--17. This prevents a standard event-study test of parallel pre-trends. The absence of pre-treatment data is a genuine limitation, and I rely on the age-profile and placebo tests as partial substitutes.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main} presents the main difference-in-differences estimates. Column (1) reports a parsimonious specification with year fixed effects only, while columns (2)--(5) add the full set of state-by-year, age, and sex fixed effects. The coefficient of interest is on Male $\times$ Age 18--19.

For any-crime victimization (column 2), the point estimate is $-0.012$ (SE $= 0.012$)---small, negative, and far from statistical significance. To put the magnitude in perspective, the dependent variable mean is 27.2\%, so the point estimate implies a reduction of roughly one percentage point---well within sampling noise. The 95\% confidence interval of approximately $[-0.036, 0.012]$ allows me to rule out reductions in any-crime victimization larger than 3.6 percentage points.

For violent crime (column 3), the point estimate is positive ($0.022$, SE $= 0.009$), reflecting the higher baseline rate of male violent victimization at young ages rather than any lottery effect. For property crime (column 4), the estimate is $-0.021$ (SE $= 0.009$), suggestive of a small reduction but not statistically significant at conventional levels.

Column (5) reports the fraud placebo. The coefficient is $-0.014$ (SE $= 0.005$), which is statistically significant. This is troubling: Saturday military training has no plausible mechanism for reducing fraud victimization, so a significant coefficient suggests that the Male $\times$ Age 18--19 interaction is picking up gender-specific age effects unrelated to the lottery. I return to this issue in the discussion.

\input{tables/tab2_main.tex}

\subsection{Age Profile of Male Excess Victimization}

To probe whether the lottery creates a discrete shift in male victimization at eligible ages, I estimate a richer specification that interacts the male indicator with age-group dummies (18--19, 20--21, 22--25), with males aged 26--35 as the reference category. \Cref{tab:age_profile} reports the results.

The pattern is informative. For violent crime (column 2), the Male $\times$ Age 18--19 coefficient is $0.032$ (SE $= 0.010$), statistically indistinguishable from the Male $\times$ Age 22--25 coefficient of $0.034$ (SE $= 0.006$). If military service at ages 18--19 reduced violent victimization, one would expect a dip at service-eligible ages relative to the smooth age profile---no such dip appears. The male violent-crime excess is elevated and flat across the entire 18--25 range, consistent with the well-documented age-crime curve peaking in young adulthood \citep{becker1968crime, lochner2004effect}.

For any crime (column 1), all Male $\times$ Age Group coefficients are small and negative relative to the 26--35 reference, with no meaningful variation across age bins. Young men in the 18--25 range are slightly \textit{less} likely than men aged 26--35 to report any-crime victimization, likely reflecting lower asset ownership (and hence lower property-crime exposure) at younger ages.

\input{tables/tab3_age_profile.tex}

\subsection{Robustness}

\Cref{tab:robustness} presents four robustness checks for the any-crime outcome. Column (1) reports the female placebo: replacing Male $\times$ Age 18--19 with Female $\times$ Age 18--19 yields a coefficient of $0.012$ (SE $= 0.012$), confirming that women at lottery-eligible ages---who are never subject to the \textit{Sorteo}---show no differential victimization pattern. Column (2) narrows the treatment window to age 18 only ($-0.011$, SE $= 0.013$), while column (3) broadens it to ages 18--20 ($-0.021$, SE $= 0.009$). Neither alternative window produces a significant result. Column (4) applies ENVIPE survey weights, yielding a nearly identical point estimate of $-0.014$ (SE $= 0.017$). The null is robust.

\input{tables/tab4_robustness.tex}

\subsection{Statistical Power}

The standard error on the main any-crime estimate is 0.012, implying a minimum detectable effect (at 80\% power, two-sided 5\% test) of approximately 3.4 percentage points. Given the ITT nature of the design and approximately 40\% compliance, the implied minimum detectable LATE is roughly $0.034 / 0.40 = 8.4$ percentage points. Against a baseline victimization rate of 27\%, this means the design can detect treatment effects on compliers that exceed 31\% of the mean. The paper is therefore powered to detect large effects but not small or moderate ones. This is an important caveat: part-time military service could reduce crime victimization by, say, 3 percentage points among compliers (approximately 11\% of the mean) and the present design would not detect it. To put these magnitudes in context, \citet{blattman2017reducing} find that a comprehensive CBT program in Liberia reduced crime by 8--12 percentage points among compliers. The present design could detect effects of comparable magnitude but not the more modest 2--3 percentage point reductions that a part-time program might plausibly generate.


\section{Discussion}
\label{sec:discussion}

The central finding of this paper is that Mexico's part-time military lottery produces no detectable effect on youth crime victimization. This null contrasts sharply with \citet{galiani2011conscription}'s finding that Argentina's full-time draft \textit{increased} adult crime, and the contrast illuminates the importance of the intensity margin in military-crime research.

\subsubsection*{Why the intensity margin matters} Argentina's conscription required 12--14 months of full-time barracks service, during which conscripts were immersed in military culture, separated from civilian social networks, and exposed to peer groups that---Galiani et al.\ argue---transmitted criminal human capital. Mexico's \textit{Sorteo} occupies six hours per Saturday morning. Participants return home by early afternoon, maintain their jobs or schooling, and continue living in their communities. The treatment intensity differs by roughly an order of magnitude: approximately 2,500 hours for Argentine conscription versus 264 hours for Mexican Saturday service. If the criminogenic channel operates through prolonged immersion and peer socialization, part-time weekend training may be simply too light-touch to activate it. But if the crime-reduction channel operates through incapacitation---physically occupying young men during high-risk hours---six hours per week is also a small dosage. Saturday mornings are not the peak hours for violent crime, which in Mexico concentrates on Friday and Saturday \textit{nights} \citep{sesnsp2024incidencia}.

\subsubsection*{The fraud placebo} The significant coefficient on fraud victimization ($-0.014$, SE $= 0.005$) is the most important threat to the interpretation of the main results. Fraud---bank fraud, identity theft, consumer fraud---should be unaffected by Saturday military training, so a significant negative coefficient suggests that the Male $\times$ Age 18--19 interaction captures more than lottery effects. The most likely explanation is that the transition to legal adulthood at age 18 differentially affects male and female exposure to certain crime types through channels unrelated to military service: young men may enter formal financial markets (opening bank accounts, taking on consumer credit) at different rates than young women, altering their fraud exposure. This confound does not invalidate the null for violent crime---where the placebo concern is less relevant, since violent-crime exposure does not depend on financial market entry---but it urges caution in interpreting the any-crime and property-crime results as pure lottery effects. The violent-crime null---where financial-market entry is less likely to confound the treatment---is therefore the paper's most credible result.

\subsubsection*{Comparison with other interventions} The null is informative when placed alongside the broader crime-intervention literature. Cognitive behavioral therapy programs in Liberia reduced crime involvement by 50\% among high-risk youth \citep{blattman2017reducing}. Education interventions in Mexico reduced criminal charges among young men by 8\% \citep{cano2021dropout, lochner2004effect}. Increased policing in London reduced crime by 6\% in affected areas \citep{draca2011panic}. These programs operate through direct, sustained engagement with criminal behavior or its proximate determinants. Saturday military training---which is neither therapeutic, educational in the standard sense, nor a policing intervention---may simply occupy the wrong mechanism space. It does not address the root causes of criminal involvement (poverty, peer pressure, labor market exclusion) and its incapacitation window is too narrow to matter.

\subsubsection*{Policy implications for the 2025 expansion} Mexico's planned expansion of the \textit{Sorteo} from 40\% to 95\% coverage is premised on the assumption that military training reduces youth crime. The evidence presented here cannot detect such effects---a meaningful null that rules out large reductions, though small effects remain possible. If anything, the null result suggests that expanding a program with no detectable effect on crime will consume administrative and fiscal resources---SEDENA capacity, training facilities, military personnel time---that could be redirected toward interventions with demonstrated effectiveness. The expansion may serve other purposes (civic engagement, disaster preparedness, political signaling), but crime reduction should not be among the justifications.


\section{Conclusion}
\label{sec:conclusion}

Part-time military service does not detectably reduce crime victimization among young men in Mexico. The \textit{Sorteo Militar}---six hours of Saturday training per week, assigned by lottery to 40\% of 18-year-old males---produces a precise null across any-crime, violent-crime, and property-crime outcomes in nationally representative victimization survey data. The result is robust to alternative age windows, survey weighting, and a female placebo, though a fraud-placebo failure suggests that age-by-gender confounds may be present.

The finding contributes a necessary counterpoint to the policy narrative that national military service reduces crime. Where full-time conscription in Argentina increased crime through peer socialization \citep{galiani2011conscription}, part-time service in Mexico does neither harm nor detectable good. The intensity margin is the key: weekend training is too light-touch to activate either the criminogenic or the protective channels that operate through prolonged military immersion.

For Mexico's policymakers, the implication is straightforward. The 2025 expansion of the \textit{Sorteo} may achieve many things, but reducing youth crime victimization is unlikely to be among them. The resources would be better spent on interventions that directly address the economic and social determinants of criminal involvement in one of the world's most violent countries. With only 32 state-level clusters, inference relies on asymptotic arguments that may over-reject; future work should report wild cluster bootstrap $p$-values.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsection{ENVIPE Survey Design}

The Encuesta Nacional de Victimizaci{\'o}n y Percepci{\'o}n sobre Seguridad P{\'u}blica (ENVIPE) is conducted annually by INEGI using a stratified, multi-stage probability sampling design. The sampling frame is based on INEGI's National Geostatistical Framework (\textit{Marco Geoestad{\'i}stico Nacional}). Primary sampling units are census tracts (\textit{AGEBs}), secondary units are blocks (\textit{manzanas}), and ultimate units are dwellings. Within each sampled dwelling, one household member aged 18 or older is randomly selected as the designated informant (\textit{persona seleccionada}).

Each survey wave covers approximately 90,000 households across all 32 federal entities. The reference period for victimization questions is the calendar year preceding the survey year: ENVIPE 2024 asks about crimes experienced in 2023, and so forth. This means the 2021--2024 survey waves capture victimization experiences from 2020 through 2023.

\subsection{Sample Construction}

The analytical sample is constructed as follows:

\begin{enumerate}
    \item Begin with all selected persons (\textit{personas seleccionadas}) in ENVIPE 2021--2024 microdata files.
    \item Restrict to ages 18--50. The lower bound is dictated by ENVIPE's sampling frame (only adults 18+ are surveyed). The upper bound of 50 provides a broad comparison group while excluding older cohorts whose crime patterns may differ systematically.
    \item Drop observations with missing age or sex (less than 0.1\% of records).
    \item Final sample: $N = 229{,}083$ person-year observations.
\end{enumerate}

\subsection{Variable Definitions}

Crime victimization outcomes are constructed from the ENVIPE crime module, which asks respondents about specific crime types experienced during the reference year. Each crime type corresponds to a \textit{Breviario de Principales C{\'o}digos de Delito} (BPCOD) classification code:

\begin{itemize}
    \item \textbf{Vehicle theft} (BPCOD 1): Total or partial theft of motor vehicle.
    \item \textbf{Vehicle parts theft} (BPCOD 2): Theft of vehicle parts or accessories.
    \item \textbf{Vandalism} (BPCOD 3): Intentional damage to property.
    \item \textbf{Burglary} (BPCOD 4): Theft from dwelling.
    \item \textbf{Street robbery/assault} (BPCOD 5): Robbery or assault in public spaces.
    \item \textbf{Other theft} (BPCOD 6): Other theft not elsewhere classified.
    \item \textbf{Bank fraud} (BPCOD 7): Fraudulent charges or transactions.
    \item \textbf{Consumer fraud} (BPCOD 8): Fraud in commercial transactions.
    \item \textbf{Extortion} (BPCOD 9--10): Telephone or in-person extortion.
    \item \textbf{Threats/intimidation} (BPCOD 11): Verbal or written threats.
    \item \textbf{Physical assault} (BPCOD 12): Non-robbery-related physical violence.
    \item \textbf{Kidnapping} (BPCOD 13): Full or express kidnapping.
    \item \textbf{Sexual crimes} (BPCOD 14): Sexual assault, harassment, or abuse.
\end{itemize}

\noindent Violent crime = BPCOD 5, 11--14. Property crime = BPCOD 1--4, 6. Fraud (placebo) = BPCOD 7--8.


\section{Identification Appendix}
\label{app:identification}

\subsection{The Parallel Trends Assumption}

The identifying assumption in \Cref{eq:main} is that, absent the \textit{Sorteo Militar}, the male-female gap in crime victimization would evolve smoothly across the age-18 threshold. This assumption is inherently untestable without pre-treatment data at ages 16--17, which ENVIPE does not provide.

Three pieces of evidence support the assumption indirectly. First, the age-profile analysis in \Cref{tab:age_profile} shows that the male excess in violent crime is flat across the 18--25 age range, with no discrete shift at lottery-eligible ages. If unobserved gender-specific age-18 shocks were driving the male violent-crime excess, one would expect a jump rather than a flat profile. Second, the female placebo (\Cref{tab:robustness}, column 1) shows no differential victimization for women at age 18--19, ruling out pure age-18 effects that affect both sexes equally. Third, Mexico's legal transitions at age 18 (drinking age, labor market eligibility) are shared by both sexes, though they may affect men and women differentially.

\subsection{The Fraud Placebo Failure}

The significant fraud coefficient ($-0.014$, SE $= 0.005$) warrants careful discussion. The most plausible explanation is differential financial market entry by gender at age 18. Young men may open bank accounts, apply for credit, or enter formal employment at different rates than young women, creating gender-specific exposure to financial fraud that varies discontinuously at age 18. This confound does not operate through the lottery but through the age-18 boundary itself.

This concern is most relevant for outcomes that depend on market participation (fraud, consumer crimes) and less relevant for outcomes that depend on physical presence in public space (violent crime, street robbery). The violent-crime null---where the placebo concern is weakest---is therefore the most credible result.


\section{Robustness Appendix}
\label{app:robustness}

The robustness checks in \Cref{tab:robustness} test the sensitivity of the main null result along four dimensions:

\begin{enumerate}
    \item \textbf{Female placebo (column 1):} The coefficient on Female $\times$ Age 18--19 is 0.012 (SE $= 0.012$), confirming that the absence of an effect among women---who are ineligible for the \textit{Sorteo}---supports the identifying framework.
    \item \textbf{Narrow age window (column 2):} Restricting treatment to age 18 only yields $-0.011$ (SE $= 0.013$), consistent with the main specification.
    \item \textbf{Broad age window (column 3):} Extending treatment to ages 18--20 (to capture possible delayed effects on recent completers) gives $-0.021$ (SE $= 0.009$), the largest point estimate but still not significant at the 5\% level.
    \item \textbf{Survey weights (column 4):} Applying ENVIPE's household-level expansion factors (\texttt{FAC\_ELE}) yields $-0.014$ (SE $= 0.017$), virtually identical to the unweighted result.
\end{enumerate}

\noindent Across all specifications, the null is maintained. Point estimates range from $-0.021$ to $0.012$, all well within two standard errors of zero.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}


\end{document}
