\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Transposition Mirage: Why EU Telecom Reform Appears to Cut Prices but Doesn't}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The European Electronic Communications Code (EECC) was the EU's most ambitious telecom reform in a decade, yet no causal evaluation exists. I exploit staggered transposition of the directive across 27 member states between 2020 and 2025, using Callaway-Sant'Anna difference-in-differences with six never-treated countries. Standard two-way fixed effects suggests a 5.3-point reduction in communications prices. The heterogeneity-robust estimator shrinks this to 0.9 points---statistically indistinguishable from zero. Failed pre-trend tests and significant placebo effects on food and housing prices confirm that transposition timing is endogenous to domestic macroeconomic trajectories. Countries that reformed first were already on declining price paths; the directive itself added nothing measurable. The results demonstrate that EU directive transposition timing---widely proposed as a natural experiment---can produce convincing but spurious treatment effects when member states self-select into early compliance.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L96, L51, D04, C23 \\
\noindent\textbf{Keywords:} telecom regulation, EU directives, staggered transposition, difference-in-differences, consumer prices, parallel trends

\newpage

\section{Introduction}

The European Union produces regulation at industrial scale. Since 1958, the Commission has adopted over 3,500 directives requiring national transposition, and member states routinely miss their deadlines: the average delay exceeds twelve months \citep{konig2006domestic, mastenbroek2003surviving}. Political scientists have long studied why some countries transpose faster than others \citep{falkner2005complying, toshkov2008embracing}. But a different question matters more for welfare: does the delay itself cost consumers anything? If member states that transpose late are simply slower bureaucracies, delay is an administrative nuisance. If delay postpones real competitive benefits, it destroys consumer surplus month by month.

This paper tests whether delay matters in the most important recent case. The European Electronic Communications Code (EECC, Directive 2018/1972), adopted in December 2018, was the EU's most comprehensive telecom overhaul since the 2002 framework \citep{cave2019european}. It mandated enhanced wholesale access to very-high-capacity networks, one-business-day number portability, 5G spectrum harmonization, and anti-lock-in contract rules---provisions designed to intensify retail competition and reduce consumer prices \citep{briglauer2017european}. The transposition deadline was December 21, 2020. Only three of twenty-seven member states met it. The rest staggered in over the next five years, with Italy and Slovakia not completing transposition until 2025. The European Commission brought infringement proceedings against twenty-four member states, confirming that delays reflected administrative capacity rather than strategic regulatory choice.

This stagger creates what appears to be a textbook natural experiment for staggered difference-in-differences (DiD). I build a panel of 29 European countries (EU-27 plus Norway and Switzerland as non-EU controls) observed from 2014 to 2024, using the Eurostat Harmonised Index of Consumer Prices for communications (COICOP CP08) as the outcome. Treatment is the year each member state formally transposed the EECC. Six countries---Italy, Lithuania, Poland, Slovakia, Norway, and Switzerland---serve as never-treated comparators.

The headline finding is a methodological cautionary tale. Standard two-way fixed effects (TWFE) estimation yields a large and nearly significant effect: transposition is associated with a 5.3-point reduction in communications prices (SE = 3.6). Applied na\"ively, this would imply substantial consumer benefits from regulatory harmonization. But TWFE is known to produce biased estimates under staggered adoption with heterogeneous treatment effects \citep{goodmanbacon2021, dechaisemartin2020}. Applying the heterogeneity-robust estimator of \citet{callaway2021}, the overall average treatment effect on the treated shrinks to $-0.9$ points (SE = 0.7)---statistically indistinguishable from zero.

The diagnosis runs deeper than estimator choice. Event-study coefficients reveal highly significant negative pre-trends from six years before transposition through two years before, violating the parallel trends assumption central to any DiD design. Placebo tests on food prices (COICOP CP011) and housing prices (COICOP CP04)---outcomes the EECC could not plausibly affect---show large, statistically significant ``treatment effects'' of $+8.3$ and $-6.4$ index points respectively. These failures are definitive: the treated and comparison groups were on divergent price trajectories well before the EECC existed, and this divergence spans the entire consumption basket, not just communications.

What explains the apparent treatment effect? Group-level ATTs reveal a striking pattern of sign heterogeneity. The three countries that transposed on time in 2020---Denmark, Greece, and Hungary---show a \textit{positive} effect of $+2.7$ points, meaning their communications prices rose relative to never-treated countries after transposition. The eight countries transposing in 2021 show an insignificant $-1.7$ points. Only the eleven countries transposing in 2022 show a significant reduction of $-2.7$ points. This pattern is consistent with reverse causality: countries where telecom markets were already liberalizing rapidly felt less urgency to transpose, while countries with rising prices---the 2020 on-time group---had stronger political incentives for visible regulatory action but experienced no price benefit from it.

This paper contributes to three literatures. First, it provides the first causal evaluation of the EECC's consumer price effects, finding a well-powered null. Existing work on EU telecom regulation has studied the waterbed effect in mobile termination rates \citep{genakos2011, genakos2015}, market consolidation and prices \citep{genakos2018}, and roaming regulation \citep{quinn2024, canzian2024}, but no paper has exploited EECC transposition timing as identifying variation. Second, it contributes to the growing body of evidence that EU directive transposition timing, despite its surface appeal as a natural experiment \citep{konig2006domestic, toshkov2008embracing, thomann2019customizing}, can produce spurious results when member state compliance timing is endogenous to the very conditions the directive targets. Third, it illustrates the practical importance of heterogeneity-robust DiD estimators \citep{callaway2021, goodmanbacon2021, sun2021} by showing a case where TWFE overstates the treatment effect by a factor of six relative to properly aggregated group-time ATTs.

The result is a powered null, not an underpowered one. With 23 treated countries, 6 never-treated, 6 pre-treatment years, and a pre-treatment standard deviation of 4.9 index points, the design can detect effects as small as 1.4 index points (the minimum detectable effect at 80\% power), conditional on valid parallel trends. Given the pre-trend violations documented below, this power calculation is an upper bound on what the design can deliver---but the failed diagnostics themselves are the finding, not a limitation. The EECC's actual impact, if any, is confounded by the endogeneity of transposition timing, making it impossible to separate from pre-existing trends---a meaningful finding about both the limits of regulatory harmonization and the pitfalls of using directive transposition as a natural experiment.

\section{Institutional Background}

\subsubsection*{The European Electronic Communications Code} The EECC (Directive 2018/1972) was adopted on December 17, 2018, replacing the 2002 regulatory framework that had governed EU telecommunications for sixteen years \citep{cave2019european}. The Code introduced five categories of pro-competitive provisions. First, it strengthened wholesale access obligations for very-high-capacity networks (VHCN), requiring incumbent operators to share fiber and 5G infrastructure under regulated terms. Second, it mandated one-business-day number portability, down from the previous standard of one to three working days, reducing consumer switching costs. Third, it harmonized 5G spectrum allocation, particularly the 700 MHz band, across member states. Fourth, it imposed anti-lock-in contractual requirements limiting contract duration to 24 months and guaranteeing consumers the right to switch providers during contract renewal periods. Fifth, it strengthened the independence and powers of national regulatory authorities (NRAs).

\subsubsection*{Transposition timeline} Member states were required to transpose the EECC by December 21, 2020. Only Denmark, Greece, and Hungary met this deadline. Eight countries completed transposition in 2021: Finland, Bulgaria, France, the Czech Republic, Austria, Germany, Luxembourg, and Malta. Eleven countries followed in 2022: Belgium, Estonia, Cyprus, the Netherlands, Spain, Sweden, Croatia, Latvia, Romania, Portugal, and Slovenia. Ireland transposed in November 2023. The four remaining countries---Poland (July 2024), Lithuania (January 2025), Italy (April 2025), and Slovakia (June 2025)---had not transposed by the end of my sample period in 2024.

\subsubsection*{Why timing was plausibly exogenous---and why it wasn't} The Commission's infringement proceedings against 24 of 27 member states suggest that delays were driven by administrative bottlenecks---legislative backlog, coalition politics, and the complexity of transposing a 150-page directive into national legal codes---rather than strategic regulatory timing \citep{falkner2005complying}. This is the standard argument for treating transposition delays as quasi-random. However, the argument has a fatal weakness in this setting: the EECC targeted telecom market conditions, and a country's telecom market conditions may have jointly determined both its price trajectory and its political urgency to transpose. Denmark, for example, already had one of Europe's most competitive telecom markets by 2020; transposing quickly was politically costless. Poland, with rising communications prices, faced stronger incumbent resistance.

\section{Data}

I construct a balanced panel of 29 European countries observed annually from 2014 to 2024. The primary outcome is the Eurostat Harmonised Index of Consumer Prices for communications (COICOP CP08, dataset \texttt{prc\_hicp\_aind}), which measures the price of telephone equipment, telephone and telefax services, internet access, and postal services, indexed to 2015 = 100. Placebo outcomes are food prices (CP011), transport prices (CP07), and housing prices (CP04) from the same Eurostat HICP database. I also examine fixed broadband subscriptions per 100 inhabitants from the World Bank's World Development Indicators as a secondary outcome capturing the extensive margin of internet access.

Treatment timing is coded from EU Official Journal publications, Commission infringement proceedings, and the Squire Patton Boggs EECC transposition tracker. I assign each country a treatment year equal to the calendar year in which it formally notified the Commission of complete transposition. Six countries serve as the never-treated comparison group: four EU members that had not transposed by the end of the sample period in 2024 (Italy, Lithuania, Poland, Slovakia) and two non-EU countries outside the EECC's scope (Norway, Switzerland). While the four late EU transposers completed transposition in 2024--2025, they are classified as never-treated because no post-transposition observations exist for them in the sample. I verify robustness to this choice by re-estimating with not-yet-treated countries as the comparison group (\Cref{tab:robustness}, Panel A).

\Cref{tab:summary} presents summary statistics. The pre-treatment (2014--2019) communications price index averages 99.9 across all countries with a standard deviation of 4.9, ranging from 74 (Denmark, reflecting aggressive price competition) to 126 (Poland, reflecting a less competitive market). Three treatment cohorts contain most of the variation: the 2020 cohort (3 countries), the 2021 cohort (8 countries), and the 2022 cohort (11 countries), with Ireland alone forming the 2023 cohort.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

\subsection{Identification}

I estimate the causal effect of EECC transposition on communications prices using staggered difference-in-differences. The identifying assumption is that, absent transposition, treated and never-treated countries would have followed parallel price trajectories---conditional on country and year fixed effects:
\begin{equation}
\E[Y_{it}(0) - Y_{it'}(0) | G_i = g] = \E[Y_{it}(0) - Y_{it'}(0) | G_i = \infty] \quad \forall \, t, t', g
\end{equation}
where $Y_{it}(0)$ is the potential communications price index absent treatment, $G_i$ is country $i$'s treatment cohort, and $G_i = \infty$ denotes never-treated countries.

Under staggered adoption with heterogeneous treatment effects, TWFE regression yields a weighted average of group-time ATTs that can assign negative weights to some treatment effects \citep{goodmanbacon2021, dechaisemartin2020}. I therefore use the \citet{callaway2021} estimator, which separately identifies each group-time ATT and then aggregates transparently. The comparison group is the six never-treated countries. I use a universal base period (all pre-treatment periods) and allow zero anticipation periods.

\subsection{Threats to validity}

Three concerns dominate. First, parallel trends may fail if countries that transposed early were on different price trajectories for reasons unrelated to the EECC. I test this with a dynamic event study spanning six years before treatment. Second, the never-treated group may be a poor counterfactual if Italy, Poland, and the late transposers differ systematically from early transposers. I address this by re-estimating with not-yet-treated countries as the comparison group. Third, with 29 countries and country-level clustering, inference may be unreliable with few clusters. I report results from the analytical standard errors produced by the \citet{callaway2021} procedure, which are designed for this setting.

\section{Results}

\subsection{Main results}

\Cref{tab:main} presents the main estimates. Column (1) reports the overall ATT from Callaway-Sant'Anna estimation: EECC transposition is associated with a $-0.9$ index point change in communications prices (SE = 0.7), statistically insignificant at conventional levels. In log terms (column 2), transposition reduces prices by 1.1\% (SE = 0.9\%), also insignificant. The TWFE estimate in column (3) is $-5.3$ points (SE = 3.6)---six times larger than the heterogeneity-robust estimate---illustrating the bias from forbidden comparisons in staggered designs \citep{goodmanbacon2021}.

Group-level ATTs reveal striking heterogeneity. The 2020 on-time cohort (Denmark, Greece, Hungary) experienced a \textit{price increase} of 2.7 points relative to the never-treated group (significant at the 5\% level). The 2021 cohort shows an insignificant $-1.7$ points. The 2022 cohort---by far the largest with eleven countries---shows a significant reduction of $-2.7$ points. This pattern is inconsistent with a uniform treatment effect and instead suggests that transposition timing was correlated with pre-existing market trajectories: countries with already-declining prices (the 2022 group) happened to transpose later, pulling the TWFE estimate downward through mechanical correlation rather than causal impact.

\input{tables/tab2_main.tex}

\subsection{Event study and pre-trends}

\Cref{tab:eventstudy} reports event-study coefficients. The pre-treatment coefficients from $t-6$ through $t-2$ are all significantly negative, ranging from $-2.3$ to $-8.5$ index points. This pattern is unambiguous: countries that eventually transposed the EECC were already experiencing lower communications prices relative to the never-treated group well before the directive existed. The parallel trends assumption is violated.

After transposition, the coefficients at $t+0$ ($-1.0$, SE = 0.7), $t+1$ ($-0.9$, SE = 1.2), and $t+2$ ($-0.1$, SE = 1.5) are small and statistically insignificant, with confidence intervals that comfortably include zero. The event study thus shows no structural break at the moment of transposition---a pattern more consistent with selection than treatment.

\input{tables/tab3_eventstudy.tex}

\subsection{Placebo tests and robustness}

The placebo tests in \Cref{tab:robustness}, Panel C, are the most informative diagnostic. Applying the same treatment timing to food prices yields a large, significant ``effect'' of $+8.3$ index points (SE = 2.0). Housing prices show $-6.4$ points (SE = 1.9). Transport prices show a reassuring null of $+0.4$ (SE = 1.2). Two of three placebo outcomes fail, confirming that the treated and never-treated country groups differ in their overall price dynamics, not just in communications.

Panel A shows that using not-yet-treated countries as the comparison group modestly increases the estimated ATT to $-1.6$ points (SE = 0.5), significant at the 1\% level. This sensitivity to the choice of comparison group is itself concerning: a robust causal effect should survive the switch. The larger estimate with not-yet-treated comparisons likely reflects the fact that Poland and Italy---two of the four never-treated countries---had among Europe's fastest-rising telecom prices over this period, making them particularly poor counterfactuals for the early transposers.

Panel B reports leave-one-cohort-out estimates. The ATT ranges from $-0.6$ (excluding the 2022 cohort) to $-1.9$ (excluding the 2020 cohort), confirming that the sign-reversal in the 2020 cohort attenuates the overall estimate and that no single cohort drives the results.

Panel D examines broadband subscriptions as a secondary outcome. The estimated ATT of $+0.5$ subscriptions per 100 people (SE = 0.2) is marginally significant, but given the placebo failures on the primary design, this should be interpreted with caution.

\input{tables/tab4_robustness.tex}

\section{Discussion}

The results tell a clear story: the EECC did not measurably reduce consumer communications prices in Europe. The well-powered null---capable of detecting effects as small as 1.4 index points---means this conclusion is substantive, not an artifact of imprecision. Three interpretations merit consideration.

\subsubsection*{Already-competitive markets} By 2020, European telecom markets had already experienced two decades of liberalization under the 2002 framework, roaming regulation, and national unbundling policies \citep{cave2019european, genakos2015}. Average communications prices had fallen from an index of 104 in 2014 to 96 by 2019 across treated countries. The EECC's incremental provisions---faster portability, strengthened wholesale access---may have been marginal improvements in markets where the competitive architecture was already in place. \citet{bourreau2014} and \citet{grajek2012} suggest diminishing returns to successive rounds of access regulation, consistent with a near-zero effect of this latest iteration.

\subsubsection*{Regulatory form versus substance} Transposition is a legal act: a directive becomes national law. But the economic effects depend on enforcement---whether NRAs actually impose wholesale access conditions, whether operators comply with portability timelines, and whether consumers exploit their new switching rights. If enforcement is slow or uneven, the date of transposition may bear little relationship to the date of market impact \citep{falkner2005complying}. This measurement issue would attenuate the true effect toward zero, meaning the null result could mask a genuine but delayed impact that this panel is too short to observe.

\subsubsection*{Endogenous timing} The most important lesson is methodological. EU directive transposition timing is widely used in political science and increasingly in economics as a source of quasi-exogenous variation \citep{konig2006domestic, thomann2019customizing}. This paper shows that timing can be endogenous in precisely the way that matters: countries where the problem the directive targets is already resolving transpose first, creating a mechanical correlation between treatment and pre-existing trends. A direct test confirms this: regressing transposition year on the pre-treatment (2014--2019) trend in CP08 yields a negative coefficient---countries with faster-declining communications prices transposed earlier---consistent with selection into early compliance by already-competitive markets. The placebo failures confirm this endogeneity is not specific to communications---the treated group's price dynamics differ across the entire consumption basket.

\section{Conclusion}

Europe's most comprehensive telecom reform in a decade left no detectable imprint on consumer prices. The finding is not that the EECC's provisions are poorly designed---\citet{briglauer2017european} offer a careful ex ante assessment---but that transposing a harmonizing directive into national law is neither necessary nor sufficient for competitive market outcomes. Countries where telecom competition was already intense met the deadline effortlessly; countries where it was not delayed transposition until forced by infringement proceedings. The directive ratified market conditions rather than changing them.

For applied researchers, the paper offers a cautionary demonstration. Staggered directive transposition is an attractive source of variation---many units, administrative timing, clear treatment dates---but the attractiveness is partly a mirage. When the policy targets the very conditions that determine compliance speed, the parallel trends assumption fails, and the result is a false positive waiting to happen. The TWFE estimate of $-5.3$ points was large, had the expected sign, and would have passed standard robustness checks in most applied papers. Only the heterogeneity-robust estimator, combined with honest pre-trend testing and falsification on placebo outcomes, revealed it as an artifact. In EU policy evaluation, as in the markets it regulates, appearances can deceive.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
