\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Convergence Trap: Price Regulation, Dispersion Collapse,\\ and Insurance Market Dynamics}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The UK's 2022 GIPP remedy banned loyalty penalties in retail insurance, collapsing motor quote dispersion from 37.7\% to 11.4\%---widely cited as competitive success. Using ONS price indices (2015--2025) and FCA data, I document that this collapse coincided with an 83\% surge in transport insurance prices, far exceeding general inflation (14\%), health insurance (15\%), and house insurance (18\%). Comparative analysis shows transport insurance grew 7.5 percentage points more than health insurance post-GIPP, though differential pre-trends preclude causal interpretation. FCA loss ratios remain stable (motor 56\%$\to$54\%), indicating cost pass-through rather than margin expansion. However, eliminating introductory discounts may have reduced demand elasticity, enabling fuller pass-through than a high-dispersion regime would permit. These patterns suggest a ``convergence trap'': price discrimination bans can compress dispersion while weakening competitive pressure on pricing.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G22, L11, L51, D43 \\
\noindent\textbf{Keywords:} price discrimination, insurance regulation, price dispersion, tacit collusion, GIPP

\newpage

\section{Introduction}

The UK motor insurance market writes roughly \pounds 21 billion in annual premiums, covering over 30 million policyholders. In January 2022, the Financial Conduct Authority (FCA) implemented one of the most ambitious pricing interventions in insurance history: the General Insurance Pricing Practices (GIPP) remedy, which banned the practice of offering low introductory premiums to new customers while progressively raising renewal prices for loyal policyholders \citep{fca2021ps}. The stated objective was to eliminate ``loyalty penalties'' that cost existing customers an estimated \pounds 1.2 billion annually. By the FCA's own evaluation, the reform succeeded spectacularly: the inter-quartile range of motor insurance quotes collapsed from 37.7\% to 11.4\% \citep{fca2025ep}.

This paper argues that the dispersion collapse may have been a Pyrrhic victory. Between December 2021 and December 2023, transport insurance prices surged 83\%, dwarfing the 13.8\% increase in the general Consumer Prices Index Including Housing Costs (CPIH), the 15.3\% increase in health insurance, and the 17.9\% increase in house contents insurance. By February 2026, transport insurance remained 64.8\% above its pre-GIPP level while house contents insurance had risen only 3.3\%. The conventional interpretation---that narrower price dispersion signals fiercer competition---inverts the mechanism that industrial organization theory predicts. When all customers receive the same price, firms lose the ability to use introductory discounts as competitive weapons, and the compressed price distribution makes it easier to monitor rivals and detect deviations from tacit coordination \citep{stigler1964theory}. I call this dynamic the ``convergence trap'': a regulatory compression of price dispersion that simultaneously eliminates the competitive tool (discriminatory discounting) and creates the informational architecture (price transparency) that sustains higher equilibrium prices.

\subsubsection*{Identification limits} I emphasize at the outset that the evidence in this paper is suggestive rather than causal. GIPP's implementation coincided with severe claims inflation driven by semiconductor shortages, rising repair costs, and a post-COVID rebound in driving. A placebo test placing the treatment at January 2020 yields a large, statistically significant coefficient ($-0.154$, $\text{SE} = 0.018$), indicating that transport and health insurance prices were already on divergent trajectories before GIPP. The difference-in-differences estimates therefore cannot be interpreted as the causal effect of the pricing remedy. What the evidence does establish is a robust \textit{pattern}: dispersion collapse coincided with a level shift in prices that was far larger for the product most affected by the reform, and FCA loss ratio data show that the price increase was not fully absorbed by rising claims. This pattern is consistent with the convergence trap mechanism, even if it cannot rule out concurrent cost shocks as the primary driver.

\subsubsection*{What this paper does} I construct a monthly panel of CPIH insurance price indices from ONS Table 37 spanning 2015 to 2025, covering transport insurance (COICOP 12.5.4, primarily motor), house contents insurance (12.5.2), health insurance (12.5.3), and overall CPIH. I implement a difference-in-differences design comparing transport insurance to health insurance---both subject to GIPP, but motor insurance had far higher pre-reform loyalty penalty prevalence and switching rates, making it the binding target. The baseline comparative estimate indicates that transport insurance grew 7.5 percentage points more than health insurance post-GIPP ($\text{SE} = 0.021$, $p < 0.001$). The alternative comparison against house contents insurance yields a 13.6 percentage point differential ($\text{SE} = 0.024$), and a year-on-year growth specification shows a 9.2 percentage point gap ($\text{SE} = 2.77$). I complement the price evidence with FCA General Insurance Value Measures data on firm-level loss ratios, which show that motor insurers' claims-to-premium ratios remained broadly stable at 54--56\% in 2023--2024 \citep{fca2024gvm}---indicating that the price surge primarily reflected cost pass-through, though the regulation may have facilitated fuller pass-through by reducing demand elasticity.

\subsubsection*{Contribution} This paper contributes to three literatures. First, it engages the theoretical debate on whether banning price discrimination raises or lowers welfare. \citet{schmalensee1981output} and \citet{corts1998third} establish that the welfare effects of third-degree price discrimination bans are ambiguous in oligopoly; \citet{stole2007price} and \citet{rhodes2014can} show conditions under which bans can reduce total surplus by eliminating surplus-creating transactions. I provide empirical evidence from a major regulated market where the ban coincided with exactly the pattern theory predicts under coordination. Second, the paper connects to the literature on insurance market competition \citep{dafny2010competition, honka2014quantifying}, where search frictions and switching costs create the conditions for loyalty penalties. \citet{ericson2014consumer} and \citet{handel2013adverse} demonstrate that inertia in insurance markets generates large rents for incumbents. Third, the paper contributes to the growing literature on unintended consequences of consumer protection regulation \citep{grubb2009selling, cabral2023demand}, showing how a reform designed to protect inattentive consumers may have removed the competitive tool---discriminatory introductory pricing---that disciplined market outcomes for attentive ones.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of the GIPP remedy. Section 3 describes the data. Section 4 presents the empirical strategy and its limitations. Section 5 reports results. Section 6 presents robustness checks. Section 7 discusses the evidence. Section 8 concludes.


\section{Institutional Background}

\subsubsection*{The loyalty penalty problem} UK general insurance markets have long exhibited a ``dual pricing'' structure in which new business premiums are substantially lower than renewal premiums for equivalent risk. The practice is widespread across motor, home, and other personal lines. The FCA's 2019 market study (MS18/1) documented that the average motor insurance customer paid 12\% more at first renewal than a new customer with identical risk characteristics, with the penalty compounding to over 50\% for long-tenured policyholders. The FCA estimated that loyalty penalties transferred \pounds 1.2 billion annually from renewing to new customers across motor and home insurance.

\subsubsection*{The GIPP remedy} In response, the FCA issued PS21/5 in September 2021, introducing rule ICOBS 6B. The core provision, effective January 1, 2022, requires that the premium offered at renewal must be no higher than the ``equivalent new business price''---defined as the price the firm would offer a new customer with the same risk profile through the same channel \citep{fca2021ps}. The rule applies to all retail general insurance products, though the FCA explicitly identified motor and home insurance as the primary targets given the scale of loyalty penalties in those markets.

\subsubsection*{Implementation} The rule is not a price cap in the conventional sense. Firms remain free to set any price they choose for new business. The constraint binds only on the \textit{differential} between new and renewal premiums. In practice, firms had two strategic responses: raise new business prices to meet renewal levels (``leveling up''), lower renewal prices to meet new business levels (``leveling down''), or some combination. The FCA's own evaluation found that ``most firms adjusted their pricing by a combination of increasing new business prices and reducing renewal prices'' \citep{fca2025ep}, with the net effect depending on the firm's starting position.

\subsubsection*{Dispersion collapse} The FCA's EP25/2 evaluation documented a dramatic compression of the price distribution. Before GIPP, the inter-quartile range of motor insurance quotes for a standardized risk profile was 37.7\%. After implementation, it fell to 11.4\%. The FCA interpreted this as evidence that the market was converging toward the competitive price---that the removal of discriminatory mark-ups on loyal customers was forcing prices toward cost. An alternative interpretation, and the one this paper explores, is that dispersion collapse reflects the elimination of the introductory discount \textit{below} cost that firms previously used to acquire customers, with the compressed distribution settling at a level that reflects tacit coordination rather than cost.

\subsubsection*{Concurrent factors} Motor insurance claims costs rose sharply from 2021 onward, driven by three factors: pandemic-related supply chain disruptions that raised vehicle repair and replacement costs; semiconductor shortages that inflated used car values; and a rebound in driving miles after COVID-19 lockdowns. These cost shocks affect the interpretation of post-GIPP price increases, and any empirical strategy must grapple with the difficulty of separating regulatory effects from cost pass-through. I return to this identification challenge in Section 4.


\section{Data}

The analysis draws on two data sources: ONS Consumer Price Inflation detailed reference tables for monthly insurance price indices, and FCA General Insurance Value Measures for firm-level loss ratios and premium volumes.

\subsubsection*{ONS CPIH insurance indices} The ONS publishes monthly price indices for detailed consumption categories within the CPIH framework, based on the Classification of Individual Consumption According to Purpose (COICOP). I use Table 37 of the detailed reference tables, extracting four series: transport insurance (12.5.4), which primarily captures motor insurance premiums; house contents insurance (12.5.2); health insurance (12.5.3); and the aggregate CPIH all-items index. All indices are normalized to 2015 = 100. The sample spans January 2015 to February 2026, providing 84 monthly observations before GIPP implementation and 48 after. Transport insurance is the primary ``treated'' series: motor insurance was the largest market affected by GIPP and the one with the most extensive pre-reform loyalty penalties. Health insurance serves as the primary control: it is subject to GIPP but had lower pre-reform dual pricing prevalence and much lower switching rates, making the remedy less binding.

\subsubsection*{FCA General Insurance Value Measures} The FCA publishes annual firm-level data on premiums written, claims paid, and loss ratios by product line for firms supervised under the general insurance regime. I use the 2024 release, which provides data for reporting years 2023 and 2024 \citep{fca2024gvm}. The key variable is the loss ratio (claims paid divided by gross written premiums), which measures the share of premium income returned to policyholders as claims. A falling loss ratio indicates that firms are retaining a larger share of premiums---consistent with either declining claims costs or rising margins. Product groups include Motor, Home, Pet, and Travel, enabling cross-product comparisons between GIPP-targeted (Motor, Home) and comparison products (Pet, Travel).

\subsubsection*{Limitations} The CPIH indices are aggregate measures that cannot distinguish between compositional changes (e.g., shifts toward higher-coverage policies) and pure price changes. The FCA data cover only 2023--2024, precluding pre-post comparison of loss ratios. I cannot observe firm-level pricing, entry/exit, or customer switching directly.

\Cref{tab:summary} presents summary statistics. The post-GIPP period saw transport insurance prices averaging 167.8 (index, 2015 = 100) compared to 114.6 in the pre-period---a gap far larger than for any other insurance line or the overall CPIH.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Difference-in-Differences}

The primary specification compares the evolution of transport insurance prices to health insurance prices around the GIPP implementation date:
\begin{equation}
\label{eq:did}
\log Y_{it} = \alpha + \beta \cdot (\text{Transport}_i \times \text{Post}_t) + \gamma \cdot \text{Transport}_i + \delta \cdot \text{Post}_t + \phi \cdot \log(\text{CPIH}_t) + \varepsilon_{it}
\end{equation}
where $Y_{it}$ is the price index for insurance type $i$ in month $t$, $\text{Transport}_i$ is an indicator for transport insurance (versus health insurance), $\text{Post}_t$ equals one from January 2022 onward, and $\log(\text{CPIH}_t)$ controls for aggregate inflation. The coefficient $\beta$ captures the differential price change in transport insurance relative to health insurance after GIPP, net of common inflation trends.

Given the small number of cross-sectional groups ($N=2$) and potential serial correlation in monthly data, we report Newey-West standard errors with 12 lags. This choice accommodates autocorrelation up to one year in the residuals, which is appropriate given the annual renewal cycle of most insurance contracts. Cluster-robust standard errors are infeasible with only two groups.

The identifying assumption is that, absent GIPP, transport and health insurance prices would have evolved on parallel paths in logs after conditioning on overall CPIH. This assumption is testable in the pre-period and, as I show below, it fails.

\subsection{Threats to Validity}

\subsubsection*{Pre-trend violation} The central threat to identification is that transport and health insurance were already on divergent price trajectories before January 2022. A placebo test placing the GIPP treatment at January 2020 and using only pre-GIPP data yields an estimate of $-0.154$ ($\text{SE} = 0.018$, $p < 0.001$), indicating large and statistically significant differential pre-trends (\Cref{tab:placebo}). Transport insurance prices were rising more slowly than health insurance prices during 2015--2019, then accelerating through the COVID period. This pre-trend violation means the estimate of $\beta = 0.075$ cannot be interpreted as the causal effect of GIPP. I present the results as descriptive evidence of the differential price trajectory, not as a causal estimate.

\subsubsection*{Concurrent cost shocks} The post-2021 period saw severe claims inflation in motor insurance driven by supply chain disruptions, semiconductor shortages, and rising repair costs. These cost shocks would have raised motor insurance prices regardless of GIPP. The absence of pre-GIPP loss ratio data prevents me from measuring how much of the price increase reflects cost pass-through versus margin expansion.

\subsubsection*{Control group validity} Both transport and health insurance are subject to GIPP, so the comparison measures the \textit{differential} effect rather than the total effect. If GIPP also raised health insurance prices (through the same convergence mechanism), the differential growth estimate understates the total effect. Conversely, if motor-specific cost shocks drove the entire price increase, the estimate overstates any GIPP contribution.


\section{Results}

\subsection{Price Dynamics}

\Cref{tab:summary} documents the central fact. Transport insurance prices rose 83\% from December 2021 to December 2023, compared to 17.9\% for house contents insurance, 15.3\% for health insurance, and 13.8\% for the aggregate CPIH. By February 2026, transport insurance remained 64.8\% above its pre-GIPP level, while house contents insurance had returned to near its December 2021 level (3.3\% above). The magnitude of the transport insurance increase is extraordinary in the context of UK consumer prices: a near-doubling in two years, with only modest reversion by early 2026.

\subsection{Comparative Trends}

\Cref{tab:main} reports the differential growth estimates across specifications. In the baseline comparison of transport versus health insurance over the full 2015--2025 window (column 1), the estimated differential is 0.075 in log points ($\text{SE} = 0.021$, $p < 0.001$), indicating that transport insurance grew approximately 7.5 percentage points more than health insurance after GIPP, conditional on aggregate inflation. Substituting house contents insurance as the comparison group (column 2) yields a larger differential of 0.136 ($\text{SE} = 0.024$), consistent with house insurance tracking general inflation more closely. A shorter estimation window of 2019--2025 (column 3) produces a larger estimate of 0.154 ($\text{SE} = 0.022$), reflecting the sharper divergence when the distant 2015--2018 pre-period is excluded. A year-on-year growth specification (column 4) shows a 9.2 percentage point gap in annual price growth rates ($\text{SE} = 2.77$).

\input{tables/tab2_main}

\subsection{FCA Loss Ratios}

\Cref{tab:fca} reports FCA loss ratio data for 2023--2024. Motor insurance loss ratios were 0.563 in 2023 and 0.544 in 2024, meaning claims consumed 54--56\% of premium income. The stability of these ratios indicates that the post-GIPP price surge was primarily driven by cost pass-through: premiums rose roughly in proportion to claims costs, rather than through margin expansion. Motor loss ratios are comparable to Pet insurance (0.573 and 0.520) and higher than Home insurance (0.430 and 0.426), with no evidence of differential margin expansion in GIPP-targeted products.

However, the stable loss ratios do not exonerate the regulation. The relevant question is not whether margins expanded, but whether the regulation enabled \textit{fuller cost pass-through} than would have occurred under the pre-GIPP pricing regime. Before GIPP, firms competed aggressively for price-sensitive switchers through introductory discounts, constraining their ability to pass through cost increases to these elastic consumers. By banning differential pricing, GIPP eliminated the channel through which price-sensitive customers exerted competitive discipline. The remaining customer base---now receiving uniform quotes---exhibits lower demand elasticity, precisely because the introductory discounts that previously attracted elastic switchers no longer exist. In this framework, stable loss ratios are consistent with the convergence trap: the regulation did not expand margins, but it reduced competitive pressure that would otherwise have forced firms to absorb a larger share of cost increases. The 83\% price surge may reflect not inflated profits, but the full pass-through of cost shocks that a more competitive (higher-dispersion) market would have partially absorbed.

\input{tables/tab3_fca}


\section{Robustness}

\subsubsection*{Placebo test} \Cref{tab:placebo} presents the placebo test. Assigning a false GIPP treatment at January 2020 and restricting the sample to pre-GIPP data (2015--2021) yields a coefficient of $-0.154$ ($\text{SE} = 0.018$), statistically significant at the 1\% level. This indicates that transport insurance prices were evolving differently from health insurance prices even before GIPP---specifically, transport insurance was relatively cheaper in the 2020--2021 period compared to the pre-2020 pattern. The failed placebo test is the single most important caveat for interpreting the main results: it means the parallel trends assumption underlying the DiD is violated, and the post-GIPP differential cannot be cleanly attributed to the regulation.

\input{tables/tab4_placebo}

\subsubsection*{Alternative windows and controls} The sensitivity of the differential growth estimate to the estimation window (\Cref{tab:main}, columns 1 vs.\ 3) is consistent with the pre-trend concern. The shorter 2019--2025 window produces a larger estimate (0.154 vs.\ 0.075), likely because it places less weight on the period when transport insurance was relatively stable. The alternative control group (house contents insurance, column 2) produces a larger differential (0.136), consistent with house insurance being a cleaner control---it was less affected by the motor-specific cost shocks that contaminate the transport--health comparison.

\subsubsection*{Year-on-year growth specification} Differencing the outcome into year-on-year growth rates (column 4) is a partial remedy for non-parallel levels, as it effectively removes slow-moving trends. The 9.2 percentage point gap remains statistically significant ($p < 0.001$), suggesting that the post-GIPP acceleration in transport insurance prices was large even relative to its own pre-period growth rate.


\section{Discussion}

The evidence assembled in this paper is consistent with the convergence trap hypothesis but insufficient to confirm it. Three facts coexist uneasily with the narrative that GIPP improved consumer outcomes: transport insurance prices nearly doubled in two years; the dispersion collapse that the FCA celebrated as competitive success coincided with this surge; and stable loss ratios indicate that the regulation facilitated full cost pass-through rather than the partial absorption that competitive discounting would have imposed. At the same time, three facts prevent a causal attribution to GIPP: a failed placebo test reveals pre-existing differential trends; severe motor-specific claims inflation provides an alternative explanation for the price increase; and the stable cross-product loss ratios show no differential margin expansion in GIPP-targeted lines.

\subsubsection*{The IO mechanism} The theoretical prediction is clear. In \citet{varian1980model}'s model of sales, price dispersion arises endogenously as firms randomize between high prices (to exploit inattentive consumers) and low prices (to attract informed ones). A ban on differential pricing eliminates the low end of the distribution---the introductory discounts that served as competitive weapons---while the high end persists or rises. \citet{stigler1964theory} predicts that the resulting price transparency makes tacit coordination easier to sustain: when all firms charge similar prices, deviations are immediately detectable, and punishment is swift. The convergence trap is the intersection of these two forces: the regulation simultaneously removes the competitive tool (discriminatory discounting) and creates the monitoring infrastructure (price homogeneity) that stabilizes higher prices.

\subsubsection*{Why it matters beyond insurance} The pattern documented here---dispersion collapse coinciding with level increases---may arise in any market where regulators ban price discrimination with the expectation that convergence toward a single price is inherently pro-competitive. Uniform pricing mandates in electricity retail markets, anti-price-gouging laws, and most-favored-customer clauses share the feature of compressing the price distribution. In each case, the welfare implications depend on whether the resulting uniform price converges toward the competitive level or the collusive level. This paper provides a cautionary data point, even if the identification is imperfect.

\subsubsection*{FCA evaluation versus this evidence} The FCA's own EP25/2 evaluation emphasizes the dispersion collapse and reports modest net price effects, concluding that GIPP achieved its objectives. The agency does not examine the absolute price trajectory in the way this paper does, nor does it benchmark motor insurance inflation against other insurance lines. The discrepancy between the FCA's positive assessment and the patterns documented here illustrates a broader point: dispersion reduction is a necessary but not sufficient condition for improved competition. A regulator measuring success by dispersion alone may miss the possibility that the entire distribution shifted upward.


\section{Conclusion}

Price convergence is not the same thing as competitive convergence. The UK's GIPP remedy compressed the distribution of motor insurance quotes from a 37.7\% inter-quartile range to 11.4\%---a textbook regulatory success by the metric of dispersion. Yet the level at which prices converged was 83\% above the pre-reform baseline within two years, a magnitude that dwarfs contemporaneous inflation in every other major consumer price category. The convergence trap---in which banning price discrimination eliminates competitive discounting while facilitating tacit coordination---offers a coherent theoretical account of this pattern, though the evidence here is insufficient to rule out claims inflation as the primary driver.

The policy lesson is conditional but important: regulators considering price discrimination bans should monitor not only whether the price distribution narrows but \textit{where} it settles. A dispersion collapse to a higher mean is not a victory for consumers. Quote-level dispersion data---such as the comparison-site microdata underlying Confused.com or Willis Towers Watson reports---would provide a more direct test of the convergence mechanism by tracking within-risk dispersion over time. Future work with such data, firm-level pricing records, consumer switching behavior, and longer post-GIPP time series may permit cleaner identification of the mechanisms at work. Until then, the massive premium increases documented here---and the \pounds 21 billion that UK motorists pay annually for the privilege of mandatory coverage---demand scrutiny that goes beyond celebrating the convergence of prices that consumers cannot afford.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{ONS CPIH detailed reference tables} Price indices are sourced from the ONS Consumer Price Inflation detailed reference tables, Table 37 (``CPI and CPIH detailed goods and services indices''). The specific series used are: CPIH transport insurance (COICOP 12.5.4, series identifier L5EE), CPIH house contents insurance (COICOP 12.5.2, series identifier L5EC), CPIH health insurance (COICOP 12.5.3, series identifier L5ED), and CPIH all items (series identifier L522). All indices use 2015 = 100 as the base year. Data were accessed in March 2026 from the ONS website (\url{https://www.ons.gov.uk/economy/inflationandpriceindices/datasets/consumerpriceinflation}).

The transport insurance index (12.5.4) covers insurance connected with transport, which is dominated by motor vehicle insurance given its compulsory nature and market size. House contents insurance (12.5.2) covers insurance of dwelling contents against fire, theft, and other risks. Health insurance (12.5.3) covers private medical and health insurance premiums.

\subsubsection*{FCA General Insurance Value Measures} Firm-level data on premiums and claims by product group are from the FCA's General Insurance Value Measures 2024 data release. The data cover all firms regulated by the FCA for general insurance activities and are published annually. Product groups used in this analysis are Motor (combining comprehensive and third-party), Home (buildings and contents), Pet, and Travel. The loss ratio is defined as gross claims incurred divided by gross written premiums. Data were accessed from the FCA website in March 2026.

\subsubsection*{Sample construction} The CPIH panel consists of 132 monthly observations (January 2015 to December 2025) for each of the four price indices. For the DiD analysis, the panel is reshaped to long format with two series (transport and health insurance), yielding 264 observations. The pre-GIPP period (January 2015 to December 2021) contains 84 observations per series; the post-GIPP period (January 2022 to December 2025) contains 48 observations per series.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
