\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Escape Valve Illusion: State Forfeiture Reform and the Federal Equitable Sharing Program}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 2014 and 2022, 36 US states reformed their civil asset forfeiture laws. A persistent concern is that these reforms are undermined by an ``escape valve'': law enforcement agencies routing seizures through the federal equitable sharing program, which operates under federal law regardless of state restrictions. Using the universe of DOJ equitable sharing certifications---51,868 agency-year observations covering 6,562 agencies across 50 states from FY2009 to FY2024---I find no evidence of circumvention. TWFE and Callaway-Sant'Anna estimates show that post-reform agencies did not increase their equitable sharing receipts (TWFE: $-0.031$, SE $= 0.114$; CS-DiD: $-0.470$, SE $= 0.899$). Stronger reforms (conviction requirements and abolition) show a marginally significant \emph{decline} in equitable sharing ($-0.235$, SE $= 0.141$). The data can rule out large-scale substitution into federal channels, suggesting that state reform retains bite despite overlapping federal jurisdiction.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K42, H77, H11 \\
\noindent\textbf{Keywords:} civil asset forfeiture, equitable sharing, regulatory leakage, federalism, law enforcement, policing

\newpage

\section{Introduction}

In 2015, the Institute for Justice reported that the federal government returned \$6.8 billion in forfeited assets to state and local law enforcement agencies over the preceding decade through ``equitable sharing''---a program that allows agencies to bypass state forfeiture restrictions by partnering with federal authorities \citep{CarpenterIJ2020}. Reform advocates warned that no matter what states did to restrict civil forfeiture, this federal escape valve would render their efforts futile \citep{Holcomb2018, Pimentel2017}. By 2022, 36 states had enacted some form of forfeiture reform. The question remained: did the escape valve open?

This paper provides the first causal evidence on whether state forfeiture reform triggers circumvention through the federal equitable sharing program. The answer, using the universe of DOJ equitable sharing certifications, is no. Agencies in reformed states did not increase their equitable sharing receipts relative to agencies in states that did not reform.

The concern about regulatory leakage through equitable sharing is intuitive and well-documented descriptively. Under the DOJ's Asset Forfeiture Program, state and local agencies can ``adopt'' a seizure---transferring it to federal jurisdiction in exchange for up to 80\% of the proceeds \citep{Worrall2004, Baicker2004}. Because federal civil forfeiture law requires only a preponderance of evidence and imposes no conviction requirement, adoptive seizures effectively nullify state reforms that raise these thresholds. Several legal scholars have argued that this jurisdictional arbitrage makes state reform ``largely symbolic'' \citep{Carpenter2019, Nelson2016, Holcomb2018}.

Yet the empirical evidence on circumvention has been limited to cross-sectional correlations and case studies. \citet{CarpenterIJ2020} documented a positive correlation between state reforms and equitable sharing activity, but without controlling for secular trends or the staggered timing of reform adoption. \citet{Holcomb2018} examined individual states but could not distinguish circumvention from broader trends in federal forfeiture policy.

I exploit the staggered adoption of forfeiture reforms across 36 states between 2014 and 2022 in a difference-in-differences framework. The key outcome is agency-level equitable sharing funds received, drawn from a FOIA release of the DOJ Equitable Sharing Agreement and Certification (ESAC) database. This dataset covers every state and local law enforcement agency that participates in equitable sharing, yielding 51,868 agency-year observations across 6,562 unique agencies.

The main finding is a null: TWFE estimates show a coefficient of $-0.031$ (SE $= 0.114$) on log equitable sharing funds, and the heterogeneity-robust Callaway-Sant'Anna estimator yields an ATT of $-0.470$ (SE $= 0.899$). At the extensive margin (participation), the effect is similarly null ($-0.003$, SE $= 0.011$). The TWFE minimum detectable effect at 80\% power is 0.32 log points---equivalent to a 38\% increase in equitable sharing funds---meaning the data can rule out large-scale substitution into federal channels, though more modest circumvention remains possible.

Heterogeneity analysis provides suggestive evidence of a \emph{compliance spillover}: agencies in states with the strongest reforms (abolition of civil forfeiture or criminal conviction requirements) show a marginally significant \emph{decline} in equitable sharing of $-0.235$ log points (SE $= 0.141$, $p = 0.10$), while agencies in states with weaker reforms (burden-of-proof changes or reporting requirements) show no effect ($-0.015$, SE $= 0.142$). This pattern is inconsistent with circumvention---which predicts that stronger reforms would trigger \emph{more} leakage---and instead suggests that reform may shift institutional norms beyond its statutory scope.

This paper contributes to several literatures. First, it addresses the growing body of work on civil asset forfeiture and policing incentives \citep{Baicker2004, Worrall2004, Holcomb2018, Kelly2022}. Existing work has documented that equitable sharing revenues respond to fiscal incentives \citep{Baicker2004} and that forfeiture powers may distort policing priorities \citep{Worrall2004}, but the causal effect of reform on circumvention has not been estimated. Second, it speaks to the broader question of regulatory leakage in federal systems \citep{Sigman2014, Auffhammer2011}. The finding that sub-national regulation \emph{does not} trigger systematic jurisdictional arbitrage contrasts with settings like environmental regulation, where cross-border leakage is well-documented \citep{FowlieLevinson2014}. Third, it provides the first systematic analysis of the DOJ ESAC database as a research resource, demonstrating its utility for studying law enforcement behavior at the agency level.

The paper proceeds as follows. \Cref{sec:background} describes the institutional background of civil asset forfeiture and equitable sharing. \Cref{sec:data} presents the data. \Cref{sec:method} outlines the empirical strategy. \Cref{sec:results} presents results. \Cref{sec:discussion} discusses mechanisms and implications.

\section{Institutional Background}\label{sec:background}

\subsubsection*{Civil asset forfeiture} Civil asset forfeiture allows law enforcement agencies to seize property suspected of being connected to criminal activity without charging the owner with a crime. The legal proceeding is \emph{in rem}---against the property itself---and proceeds under civil rather than criminal standards of evidence. Historically, most states required only a preponderance of evidence, and the burden of proof often fell on the property owner to demonstrate innocence \citep{Worrall2004}.

\subsubsection*{The reform wave} Beginning in 2014, a bipartisan reform movement swept across US states. By 2022, 36 states had enacted some form of reform, varying in stringency: four states (New Mexico, Nebraska, North Carolina, and Maine) abolished civil forfeiture entirely, requiring criminal conviction and routing all proceeds through the general fund; eighteen states enacted conviction requirements; and twelve states raised the burden of proof to ``clear and convincing evidence'' or higher without requiring conviction. Four states enacted transparency and reporting requirements without changing substantive standards.

\subsubsection*{The equitable sharing program} The DOJ's equitable sharing program, established under the Comprehensive Crime Control Act of 1984, allows state and local agencies to share in the proceeds of federal forfeitures. There are two pathways. In \emph{joint investigations}, agencies collaborate with federal partners and share proceeds by agreement. In \emph{adoptive seizures}, agencies make seizures under state law and then transfer them to a federal agency for processing under federal law, receiving up to 80\% of the net proceeds. Adoptive seizures are the primary mechanism of concern, because they allow agencies to circumvent state restrictions: federal law governs the evidentiary standard, not state law.

\subsubsection*{The anti-circumvention response} Recognizing this vulnerability, three states---New Mexico (2015), Nebraska (2016), and Colorado (2017)---enacted explicit anti-circumvention statutes prohibiting agencies from transferring seizures to federal jurisdiction solely to avoid state restrictions. These laws provide a natural test: if circumvention is the primary channel, anti-circumvention laws should attenuate any post-reform increase in equitable sharing.

\section{Data}\label{sec:data}

The primary dataset is the DOJ Equitable Sharing Agreement and Certification (ESAC) database, obtained through FOIA. Every state and local law enforcement agency that participates in equitable sharing must file an annual ESAC certification reporting income received and expenditures made from the equitable sharing fund. The FOIA release covers the universe of certifications from FY2009 through FY2024.

\subsubsection*{Key variables} The main outcome variable is equitable sharing funds received (ESAC income type 1: ``Equitable Sharing Funds Received''), which captures direct federal disbursements to the agency. I construct three outcome measures: (1) log of equitable sharing funds received plus one, to accommodate zeros; (2) an extensive margin indicator for whether the agency received any equitable sharing funds; and (3) the level of funds in dollars.

\subsubsection*{Sample construction} I match agency records to states using the NCIC state code field and restrict to US states and DC. I drop North Carolina, which abolished civil forfeiture before the panel begins (2000). I keep agencies with at least three years of certification data. The final sample contains 51,868 agency-year observations across 6,562 unique agencies in 50 states, with 36 states classified as reformed and 14 as never-reformed.

\subsubsection*{Treatment coding} I code the reform year for each state based on the effective date of the relevant statute, drawn from Institute for Justice reports and NCSL legislative databases. Reform types are classified as: abolition (2 states), conviction required (21 states), burden of proof raised (12 states), or reporting only (4 states). Anti-circumvention statutes are coded for NM, NE, and CO. \Cref{tab:summary} presents summary statistics.

\input{tables/tab1_summary}

\section{Empirical Strategy}\label{sec:method}

The baseline specification is a two-way fixed effects (TWFE) regression:
\begin{equation}
Y_{a,t} = \alpha_a + \gamma_t + \beta \cdot \text{PostReform}_{s(a),t} + \varepsilon_{a,t}
\label{eq:twfe}
\end{equation}
where $Y_{a,t}$ is the outcome for agency $a$ in fiscal year $t$, $\alpha_a$ are agency fixed effects, $\gamma_t$ are fiscal year fixed effects, and $\text{PostReform}_{s(a),t}$ equals one if agency $a$'s state $s$ has enacted forfeiture reform by year $t$. Standard errors are clustered at the state level, yielding 50 clusters. The coefficient $\beta$ captures the average change in equitable sharing activity for agencies in reformed states relative to agencies in never-reformed states, after accounting for agency-specific levels and common year shocks.

\subsubsection*{Addressing staggered adoption bias} With treatment rollout spanning nine years (2014--2022), the TWFE estimator may be biased by heterogeneous treatment effects across cohorts \citep{GoodmanBacon2021, deChaisemartin2020}. I implement the \citet{CallawayS2021} estimator, which constructs group-time average treatment effects using only never-treated units as the comparison group and a varying base period. I aggregate to an overall ATT and to event-time coefficients for event study plots.

\subsubsection*{Identification} The identifying assumption is parallel trends: absent reform, agencies in states that eventually reformed would have followed the same trajectory as agencies in states that did not. I assess this using event-study estimates of pre-treatment coefficients from the Callaway-Sant'Anna framework. Year fixed effects absorb federal-level policy changes, including Attorney General Holder's 2015 restrictions on adoptive seizures and Attorney General Sessions' 2017 reversal, as these applied uniformly to all states.

\section{Results}\label{sec:results}

\subsubsection*{Main estimates} \Cref{tab:main} reports the main results. Across all specifications and both estimators, the effect of state forfeiture reform on equitable sharing is economically small and statistically insignificant. The TWFE estimate on log equitable sharing funds is $-0.031$ (SE $= 0.114$), essentially zero. The Callaway-Sant'Anna overall ATT is $-0.470$ (SE $= 0.899$), negative but imprecise. Neither the extensive margin (participation) nor the level specification yields significant effects.

\input{tables/tab2_main}

\subsubsection*{Power and minimum detectable effects} The TWFE standard error of 0.114 implies a minimum detectable effect (at 80\% power, 5\% size) of 0.32 log points, corresponding to a 38\% change in equitable sharing funds. The data can therefore rule out circumvention of economically meaningful magnitude. A 38\% increase would represent approximately \$24,000 per agency per year, or \$144 million in additional equitable sharing disbursements to agencies in reformed states---well below the \$6.8 billion cumulative total reported by the Institute for Justice but large enough to represent a meaningful policy response.

\subsubsection*{Pre-trends} The Callaway-Sant'Anna event study shows pre-treatment coefficients centered on zero: the estimates for event times $-5$ through $-1$ are 0.346, $-0.208$, 0.371, $-0.783$, and 0.960, with standard errors between 0.65 and 0.94. While individually noisy, they show no systematic upward or downward trend. Post-reform estimates decline monotonically from $-0.165$ at event time 0 to $-0.520$ at event time 5, consistent with a gradual compliance spillover rather than circumvention.

\subsubsection*{Robustness} \Cref{tab:robust} presents robustness checks. The inverse hyperbolic sine transformation yields nearly identical results ($-0.033$, SE $= 0.121$). The placebo test---randomly assigning reform years to agencies in never-reformed states---produces a null coefficient (0.066, SE $= 0.073$). Leave-one-state-out analysis shows the TWFE coefficient ranges from $-0.090$ to $+0.023$ across 50 jackknife iterations, confirming that no single state drives the result.

At the state level, I find a positive but statistically insignificant effect on total equitable sharing funds ($+0.760$, SE $= 0.477$, $p = 0.12$) and on the number of agencies filing certifications ($+12.3$, SE $= 13.6$). These point estimates suggest some extensive-margin entry---additional agencies filing ESAC certifications post-reform---but the effects are imprecise.

\input{tables/tab3_robust}

\subsubsection*{Heterogeneity} \Cref{tab:heterogeneity} explores heterogeneity by reform stringency and anti-circumvention legislation. Agencies in states with strong reforms (abolition or conviction requirements, 20 treated states) show a marginally significant decline of $-0.235$ log points (SE $= 0.141$, $p = 0.10$), while agencies in states with weak reforms (burden of proof or reporting, 16 treated states) show essentially no effect ($-0.015$, SE $= 0.142$). This gradient is inconsistent with circumvention, which predicts that stronger reforms---imposing greater restrictions on state-level seizures---would trigger \emph{more} substitution toward federal channels. Instead, the pattern suggests that strong reforms shift institutional norms, reducing equitable sharing alongside state-level forfeiture.

The anti-circumvention subsample is too small (3 treated states) for precise estimation but shows a positive point estimate ($+0.395$, SE $= 0.278$), consistent with these states being selected for anti-circumvention legislation precisely because their agencies were more actively using equitable sharing.

\input{tables/tab4_heterogeneity}

\section{Discussion}\label{sec:discussion}

The central finding is that state forfeiture reforms do not produce large average increases in agencies' reported equitable sharing receipts. This result pushes back against the widely held belief that equitable sharing renders state reform ``largely symbolic'' \citep{Carpenter2019}, though the data cannot fully separate adoptive seizures (the mechanism of concern) from joint investigations in the ESAC totals. Subject to this caveat, the evidence suggests that state-level action retains more bite than commonly assumed.

\subsubsection*{Why doesn't the escape valve open?} Three mechanisms may explain this result. First, administrative friction: adoptive seizures require coordination with federal agencies, paperwork, and processing delays that may make them unattractive relative to state-level forfeiture, even when state standards are tightened \citep{Worrall2004}. Second, institutional signaling: state reform may shift the political environment surrounding forfeiture, making agencies cautious about \emph{all} forms of seizure activity, including federal channels. The stronger effect of stringent reforms is consistent with this channel. Third, attention effects: reform may draw scrutiny from the public and media, making agencies reluctant to engage in any activity that could be perceived as circumvention.

\subsubsection*{Limitations} Several caveats apply. The ESAC data captures only agencies that file certifications; agencies that participate in equitable sharing without filing---or that exit the program entirely---are not observed. The data does not distinguish adoptive seizures (the circumvention channel) from joint investigations (which are less subject to circumvention concerns). With 50 state clusters, inference relies on cluster-robust standard errors, and the Callaway-Sant'Anna estimates are imprecise. Finally, the reform coding treats each state's reform date as a single event, abstracting from implementation lags and enforcement variation.

\subsubsection*{Broader implications} The absence of regulatory leakage in this setting is noteworthy. In environmental regulation, cross-border leakage is a well-documented response to sub-national policy \citep{FowlieLevinson2014, Sigman2014}. In forfeiture, the federal ``escape valve'' was structurally designed to enable circumvention, yet agencies did not exploit it. This suggests that institutional frictions and cultural norms may be more important constraints on regulatory arbitrage than the formal legal framework. For policymakers, the implication is that state forfeiture reform is not rendered futile by federal jurisdiction---though federal reform of the equitable sharing program would further strengthen the policy.

\label{apep_main_text_end}

\newpage

\bibliography{references}

\newpage
\appendix
\section*{Appendix: Standardized Effect Sizes}
\input{tables/tabF1_sde}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
