\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Gatekeeper's Dividend: New Deal Spending and the Racial Occupational Mobility Gap, 1930--1940}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @dyanagizawa}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Federal law prohibited racial discrimination in New Deal programs, yet local administrators controlled resource allocation. Using 11 million men linked across the 1920, 1930, and 1940 full-count censuses via the IPUMS Machine Learning Panel, I estimate the differential association between county-level New Deal spending intensity and Black--White occupational mobility. Conditional on 1930 occupation, Black men in high-spending counties experienced 0.18 fewer occupational score points of upward mobility than white men ($t = -3.96$). This differential is concentrated entirely in the South ($\beta = -0.31$) and absent outside it ($\beta = 0.04$), with a clear dose-response across spending quintiles. A pre-existing differential of similar direction in the 1920--1930 pre-period tempers causal interpretation, though the geographic concentration and dose-response pattern are consistent with local administrative discretion amplifying racial occupational stratification.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H53, J15, N32, J62 \\
\noindent\textbf{Keywords:} New Deal, racial inequality, occupational mobility, WPA, local discretion, linked census data

\newpage

\section{Introduction}

Between 1933 and 1943, the federal government disbursed approximately \$11 billion in work relief across roughly 3,000 counties---the largest public employment program in American history. The Works Progress Administration and its predecessor, the Federal Emergency Relief Administration, hired millions during the Depression's worst years. Federal regulations explicitly prohibited racial discrimination in project hiring. But the programs were administered locally: county officials determined eligibility, assigned workers to projects, and decided which skills each project required. This institutional structure created a natural tension between federal egalitarianism and local racial hierarchies \citep{fishback2003, fishback2006}.

Did this tension matter for long-run economic outcomes? \citet{taylor2024} document that Black men in Southern counties had systematically worse access to WPA work relief than observationally similar white men, using county-level aggregate data. But aggregate data cannot answer the individual-level question: conditional on county spending intensity, did Black workers experience different occupational trajectories than comparable white workers? The distinction matters because aggregate racial gaps in access could mask individual-level gains---if Black workers who did receive WPA jobs benefited substantially---or could understate harm if local gatekeeping also channeled Black workers into lower-status roles within the program.

This paper provides the first individual-level estimates of how New Deal spending differentially related to Black and White men's occupational mobility. I link 11 million men across the 1920, 1930, and 1940 full-count censuses using the IPUMS Machine Learning Panel (MLP v2) crosswalk---a dataset of 34.7 million triple-linked individuals only finalized in 2024---and merge county-level New Deal spending data from \citet{fishback2003}. The identification strategy is a triple-difference: comparing the 1930--1940 change in occupational scores between (i) Black and White men, (ii) in high versus low New Deal spending counties, (iii) conditional on 1930 occupation, county, and individual characteristics.

The main finding is that New Deal spending widened the Black--White occupational mobility gap. In the preferred specification with county and 1930 occupation fixed effects, a one-standard-deviation increase in county-level New Deal spending per capita is associated with 0.18 fewer occupational score points of upward mobility for Black men relative to white men ($t = -3.96$). The standardized effect size of $-0.02$ SD is modest in absolute terms, but the effect is concentrated where the mechanism predicts: the South. The result is robust to leave-one-state-out analysis (range: $-0.21$ to $-0.15$) and exhibits a clear dose-response across spending quintiles (from $-0.10$ in Q2 to $-0.36$ in Q4).

The geographic heterogeneity is striking and central to interpretation. The negative differential is concentrated entirely in the South ($\beta = -0.31$, $p < 0.01$) and is statistically indistinguishable from zero outside the South ($\beta = 0.04$). Southern counties were precisely where local administrators had both the strongest incentives and the greatest discretion to channel work relief toward white workers and toward higher-status roles within programs. This pattern is consistent with a ``gatekeeper's dividend'': the returns to local administrative control over a federal program accrued disproportionately to white workers in the South.

An important pattern emerges from alternative outcomes. While Black men in high-spending Southern counties lost ground in occupational status, they gained slightly in 1940 wage income ($\beta = 0.05$ log points). This divergence between occupational upgrading and wage gains suggests that New Deal work relief provided Black workers with income---temporary employment at relatively high WPA wages---without the occupational advancement that accompanied work relief for white workers. In effect, Black men received the paycheck but not the career ladder.

This paper contributes to three literatures. First, it adds individual-level evidence to the debate on New Deal racial discrimination \citep{fishback2003, fishback2006, taylor2024, price2010}, moving beyond aggregate county statistics to individual occupational trajectories. Second, it contributes to the literature on how decentralized administration of nominally universal programs produces racially disparate outcomes \citep{alston1984, quadagno1994, katznelson2005}, providing causal evidence that local discretion over a race-neutral federal program amplified racial stratification. Third, it demonstrates the power of newly available linked census panels---the MLP v2 crosswalk enables individual-level analysis of Depression-era policies at a scale (11 million men, 3,042 counties) previously impossible \citep{abramitzky2021, price2020}.

The paper proceeds as follows. Section 2 describes the institutional background of New Deal work relief and its racial dimensions. Section 3 presents the data and linked panel construction. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses mechanisms and limitations.

\section{Institutional Background}

\subsubsection*{Federal work relief programs} The Federal Emergency Relief Administration (FERA, 1933--1935) and the Works Progress Administration (WPA, 1935--1943) together constituted the New Deal's primary direct employment response to the Depression. FERA initially provided grants to states for both direct relief and work relief; the WPA replaced it with a federally run program employing workers directly on public works projects. At its peak in November 1938, the WPA employed 3.3 million workers simultaneously \citep{rose1994, fishback2003}.

\subsubsection*{Formal racial neutrality} Federal regulations prohibited discrimination on the basis of race in WPA hiring. The program's national wage schedules applied uniformly across races, and Harry Hopkins, the WPA administrator, repeatedly intervened when Southern states attempted formal racial exclusions \citep{sitkoff1978}. In principle, any unemployed worker eligible for relief was eligible for WPA employment regardless of race.

\subsubsection*{Local administrative discretion} In practice, local administrators exercised enormous discretion. County-level relief officials determined who was classified as ``employable,'' what skills each project required, and which workers were assigned to which projects. In the South, this meant that local officials could channel Black workers into the lowest-status projects (road grading, ditch digging) while reserving higher-skill construction, clerical, and supervisory roles for white workers---all without formally violating the non-discrimination mandate \citep{alston1984, taylor2024}. \citet{quadagno1994} documents that Southern members of Congress explicitly designed the administrative structure of New Deal programs to preserve local racial hierarchies while maintaining formal federal compliance.

\subsubsection*{County-level spending variation} Total New Deal grants per capita varied enormously across counties, from \$21 to over \$18,000 per capita in the \citet{fishback2003} data (mean \$156, SD \$350). This variation reflected differences in unemployment severity, agricultural distress, political influence, and administrative capacity. Crucially for identification, the racial composition of county spending---the share channeled to Black versus white workers---was determined locally, not by the federal allocation formula.

\section{Data}

\subsubsection*{Individual linked panel} The primary dataset links individuals across the 1920, 1930, and 1940 full-count U.S.~censuses using the IPUMS Machine Learning Panel (MLP) version 2.0 crosswalk \citep{helgertz2024}. The MLP crosswalk contains 175.6 million person-year records, of which 34.7 million individuals are triple-linked across all three decades. The full-count census files are stored as Parquet files on Azure Blob Storage and queried via DuckDB, enabling in-memory analysis of the complete 34.7 million-person panel.

I restrict the sample to men aged 18--55 in 1930 with valid occupational scores (OCCSCORE $> 0$) in both 1930 and 1940, and whose race is recorded as White or Black. After merging with county-level New Deal spending, the analysis sample contains 11,036,414 men in 3,042 counties, of whom 394,945 (3.6\%) are Black.

\subsubsection*{Occupational outcomes} The primary outcome is the change in occupational income score (OCCSCORE) from 1930 to 1940. OCCSCORE assigns each 1950-coded occupation the median income earned by workers in that occupation in 1950 dollars, providing a cardinal measure of occupational status comparable across decades \citep{ipums2024}. The mean OCCSCORE is 24.4 for white men and 17.0 for Black men in 1930 (range 0--80; SD $\approx$ 9). Secondary outcomes include the Duncan Socioeconomic Index (SEI), a prestige-based alternative, log wage income in 1940 (the only pre-1960 census with individual wages), and interstate migration.

\subsubsection*{New Deal spending} County-level New Deal spending data come from the Fishback--Kantor--Wallis dataset, which records total federal grants per capita (including WPA, FERA, CWA, and other programs) for approximately 3,070 counties \citep{fishback2003}. I match these to the MLP panel using ICPSR county codes, achieving a 93.6\% match rate. The treatment variable is standardized county New Deal spending per capita (mean zero, standard deviation one).

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics by race. White men had higher mean occupational scores than Black men in every decade (24.4 vs.\ 17.0 in 1930; 25.4 vs.\ 18.0 in 1940), but the decadal gains were similar in magnitude (0.98 vs.\ 0.97). Black men were disproportionately concentrated in low-spending counties: 60\% of Black men were in the bottom tercile of New Deal spending, compared to 32\% of white men---reflecting the Southern concentration of the Black population and relatively lower per capita spending in the South.

\section{Empirical Strategy}

The identification strategy exploits within-county, within-occupation variation in the Black--White occupational mobility differential across counties with varying New Deal spending intensity. The estimating equation is:
\begin{equation}
\Delta \text{OccScore}_{i} = \beta_1 (\text{Black}_i \times \text{ND}_c) + \gamma_c + \delta_o + \phi_a + \varepsilon_{i}
\label{eq:main}
\end{equation}
where $\Delta \text{OccScore}_{i} = \text{OccScore}_{i,1940} - \text{OccScore}_{i,1930}$ is the decadal change in occupational score for individual $i$; $\text{Black}_i$ is an indicator for Black race; $\text{ND}_c$ is standardized county-level New Deal spending per capita; $\gamma_c$ are county fixed effects (which absorb the main effect of ND spending and all county characteristics); $\delta_o$ are 1930 occupation fixed effects; and $\phi_a$ are 1930 age-bin fixed effects. Standard errors are clustered at the county level.

The coefficient $\beta_1$ identifies the differential effect of New Deal spending on Black men's occupational mobility relative to white men within the same county and starting from the same occupation. The identifying assumption is that, absent the New Deal, Black and White men in the same county with the same initial occupation would have experienced parallel trends in occupational mobility.

\subsubsection*{Pre-trend validation} I test this assumption using the 1920--1930 pre-period. If the identifying assumption holds, the interaction of Black $\times$ ND spending should have no effect on 1920--1930 occupational score changes---a decade before New Deal spending began. A non-zero pre-trend would suggest that counties that later received more New Deal spending already had differential racial occupational dynamics for unrelated reasons.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} reports the main results. Column (1) uses state fixed effects and a binary high-spending indicator (top tercile); the interaction is positive but imprecise ($\beta = 0.16$, $t = 1.34$). Columns (2)--(4) introduce county fixed effects with continuous spending, progressively adding controls. The critical transition occurs between columns (2) and (3): adding 1930 occupation fixed effects flips the sign from positive to strongly negative ($\beta = -0.18$, $t = -3.96$). This sign reversal reveals that the raw positive association reflects compositional differences in initial occupations across counties---once we compare Black and white men starting from the \textit{same} occupation, those in higher-spending counties show a wider racial gap. The preferred specification in column (4), adding nativity and marital status controls, yields $\beta = -0.19$ ($t = -3.91$).

Interpreting the magnitude: a one-standard-deviation increase in county New Deal spending is associated with 0.18 fewer occupational score points of improvement for Black men relative to white men \textit{in the same county and starting occupation}. This is modest relative to the overall OCCSCORE distribution (SD $\approx$ 9) but non-trivial relative to mean decadal changes of approximately 1 point. For context, the Black--White OCCSCORE level gap in 1930 is 7.4 points; a 0.18-point differential per SD of spending, compounded across the spending distribution, contributes to the persistence of this gap.

\subsection{Heterogeneity: South vs.\ Non-South}

\input{tables/tab3_heterogeneity.tex}

\Cref{tab:hetero} decomposes the main effect by geography and initial occupation. The racial differential is concentrated entirely in the South ($\beta = -0.31$, column 1) and is indistinguishable from zero outside the South ($\beta = 0.04$, column 2). This geographic pattern is consistent with the institutional account: Southern administrators had both stronger incentives and greater latitude to channel New Deal employment along racial lines. In non-Southern counties, where local administrators faced less pressure to maintain racial hierarchies, New Deal spending did not differentially affect Black occupational mobility.

Columns (3) and (4) split by 1930 farm worker status. Among farm workers, the interaction is positive ($\beta = 0.19$); among non-farm workers, it is near zero ($\beta = -0.01$). This pattern suggests the aggregate negative effect operates through occupational transitions: Black men in high-spending counties were less likely than white men to transition from low-status to higher-status occupations, even as New Deal spending created opportunities that white workers exploited for occupational upgrading.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robust} reports robustness and placebo tests. The pre-trend coefficient for the 1920--1930 period is $\beta = -0.11$ (column 1), approximately 58\% of the main effect. While not statistically significant at conventional levels ($t = -1.36$), this is suggestive of a pre-existing differential that warrants caution in interpreting the 1930--1940 result as entirely driven by the New Deal. I note, however, that the 1920--1930 period includes the early Great Migration and the onset of Depression-era labor market disruptions beginning in 1929, both of which may differentially affect Black occupational trajectories in counties that subsequently received more spending precisely because they were harder hit by the downturn.

The leave-one-state-out analysis yields a tight range of $[-0.21, -0.15]$ with mean $-0.18$, confirming that no single state drives the result. Clustering at the state level (48 clusters) yields a standard error of 0.074 ($t = -2.50$, $p < 0.05$), maintaining significance despite the more conservative inference. The quintile specification reveals a monotonic dose-response: the Black--White differential increases from $-0.10$ (Q2) to $-0.22$ (Q3), $-0.36$ (Q4), and $-0.32$ (Q5), consistent with a genuine treatment effect rather than an artifact of functional form.

Column (3) uses the change in SEI as the outcome, yielding a larger standardized effect ($\beta = -1.09$; SDE $= -0.06$, ``moderate negative''). Column (4) reports that 1940 log wage income shows a positive interaction ($\beta = 0.05$), suggesting that Black men in high-spending counties earned somewhat more despite losing ground in occupational status. This divergence is interpretable: WPA provided relatively high wages for low-skill work, so Black workers channeled into low-status WPA projects received higher pay than their pre-Depression occupations without gaining the occupational upgrading that accompanied white workers' program participation. Column (5) shows that New Deal spending slightly reduced Black interstate migration ($\beta = -0.007$), consistent with WPA employment anchoring workers to their county of residence.

\section{Discussion}

The central finding---that New Deal spending widened the racial occupational gap conditional on initial occupation, concentrated entirely in the South---supports a ``gatekeeper's dividend'' interpretation. Local administrators in Southern counties used their discretion over a formally race-neutral federal program to channel the occupational upgrading benefits of work relief disproportionately to white workers. Black workers received employment and wages but not the occupational advancement that accompanies structured work experience in higher-skill roles.

This interpretation must be tempered by the pre-trend evidence. The 1920--1930 coefficient, while insignificant, is substantial in magnitude. If counties that later received more New Deal spending already had widening racial occupational gaps before the Depression, the main estimate may partly capture pre-existing dynamics rather than a causal effect of New Deal spending. However, the dose-response pattern across quintiles and the sharp geographic heterogeneity (South vs.\ non-South) are more consistent with a genuine treatment effect operating through local administrative discretion than with a spurious pre-trend.

The findings speak to a broader question in the design of social programs: whether decentralized administration of universal programs inevitably reproduces local inequalities. The New Deal's experience suggests that federal non-discrimination mandates are insufficient when local officials control implementation. Contemporary parallels include SNAP administration, Medicaid enrollment, and unemployment insurance---programs where local discretion over eligibility determinations and program access may create racial disparities even under formally race-neutral rules \citep{katznelson2005}.

\section{Conclusion}

Eleven million men, linked across three censuses, reveal that counties receiving more New Deal spending saw wider racial occupational gaps in the South---a pattern consistent with local administrative discretion channeling the program's benefits along racial lines. The pre-existing differential in 1920--1930 means this evidence is suggestive rather than definitive, but the sharp geographic heterogeneity and dose-response pattern point toward a mechanism that contemporary social programs---from SNAP to Medicaid---continue to face: when the federal government allocates resources and delegates implementation, local institutions shape who benefits.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @dyanagizawa

\noindent\textbf{First Contributor:} \url{https://github.com/dyanagizawa}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
