\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{siunitx}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Designation Discount: Priority Neighborhood Boundaries and Property Values in France}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
France designated 1,362 \textit{Quartiers Prioritaires de la Politique de la Ville} (QPV) in 2015, bundling tax exemptions, reduced VAT, and 12 billion euros in urban renewal with a formal poverty label. I exploit the sharp geographic boundaries of these zones using a spatial regression discontinuity design on 1.8 million geolocalized property transactions. Properties just inside QPV boundaries sell for 13--15 percent less per square meter than those just outside --- a ``designation discount'' that persists across bandwidths, property types, and time periods. The density test finds no evidence of sorting at the boundary. A parametric specification controlling for property characteristics and boundary-segment fixed effects yields a lower bound of 12.7 percent. Despite substantial fiscal incentives that should capitalize positively into prices, the net effect is negative, consistent with the informational costs of poverty designation exceeding the fiscal benefits in the housing market.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R31, R58, H23, I38 \\
\noindent\textbf{Keywords:} place-based policy, spatial RDD, property values, stigma, priority neighborhoods, France

\newpage

\section{Introduction}

Governments around the world designate disadvantaged neighborhoods for special treatment. The intention is clear: channel resources where they are needed most. But designation carries a second signal. By officially labeling a neighborhood as ``priority,'' ``distressed,'' or ``disadvantaged,'' the state may inadvertently stigmatize the very place it seeks to help \citep{galster2017people}. Whether the fiscal benefits of designation outweigh its informational costs is an open empirical question with direct implications for the design of place-based policy worldwide.

France provides an unusually clean setting to study this tradeoff. In 2015, the government designated 1,362 \textit{Quartiers Prioritaires de la Politique de la Ville} (QPV) using a transparent algorithm based on 200-meter gridded income data from INSEE. These zones receive a substantial policy bundle: a 30 percent reduction in property tax (\textit{TFPB}), a reduced VAT rate of 5.5 percent on new construction (versus the standard 20 percent), corporate tax exemptions, and eligibility for the \textit{Agence Nationale pour la R\'{e}novation Urbaine} (ANRU) urban renewal program totaling 12 billion euros. The boundaries of these zones are sharp, publicly available, and defined at sub-commune geographic resolution --- creating a natural experiment in which properties separated by meters face different policy regimes.

This paper estimates the net capitalization of QPV designation into residential property prices using a spatial regression discontinuity design (RDD). I combine the universe of geolocalized property transactions from France's \textit{Demandes de Valeurs Fonci\`{e}res} (DVF) database with official QPV boundary polygons. The running variable is signed distance from each transaction to the nearest QPV boundary: positive inside, negative outside. The identifying assumption --- that property characteristics do not change discontinuously at the boundary --- is supported by the way QPV boundaries were drawn (algorithmically, from gridded income data) and by the McCrary density test, which finds no evidence of bunching at the cutoff ($p = 0.20$).

The main finding is a large and robust \textit{designation discount}. Properties just inside QPV boundaries sell for approximately 15 percent less per square meter than those just outside, as estimated by the bias-corrected local polynomial estimator of \citet{calonico2014robust} with a triangular kernel and MSE-optimal bandwidth. This estimate is stable across bandwidths ranging from 250 meters ($-14.5$ percent) to 1,000 meters ($-19.6$ percent), across property types (apartments: $-19.3$ percent; houses: $-21.3$ percent), and across sub-periods within 2020--2024. A parametric specification with boundary-segment fixed effects and year-quarter fixed effects, controlling for property surface area, number of rooms, and building type, yields an estimate of $-12.7$ percent.

The sign and magnitude of this result are striking. Standard capitalization theory predicts that fiscal incentives --- particularly a 30 percent property tax reduction --- should increase property values inside the zone \citep{oates1969effects, tiebout1956pure}. The 5.5 percent VAT rate on new construction represents a subsidy of approximately 15 percentage points relative to the standard rate, which should attract development and buyers. Yet the net effect on prices is negative and large, implying that the stigma channel must dominate the fiscal channel by a wide margin.

This paper contributes to three literatures. First, it advances the study of place-based policies by providing the first spatial RDD estimate of QPV designation effects on property values. Prior work on French priority zones used difference-in-differences designs to study firm creation \citep{mayer2017taxation} and employment outcomes, but the property market capitalization --- arguably the most direct measure of net local welfare --- has not been estimated with a sharp spatial design. Second, it contributes to the growing literature on neighborhood stigma and labeling effects \citep{galster2017people, lens2018neighborhood}, providing rare quasi-experimental evidence that official poverty designation depresses property values beyond what compositional differences in housing stock can explain. Third, it speaks to the broader literature on the capitalization of local public goods and fiscal incentives \citep{oates1969effects, brueckner1982economics, cellini2010value}, demonstrating that fiscal incentives can be fully offset by informational externalities.

The remainder of the paper is organized as follows. Section 2 describes the institutional setting. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{The QPV System} France's \textit{Politique de la Ville} (City Policy) has targeted disadvantaged urban neighborhoods since the 1980s. The current system was established by Law 2014-173 of February 21, 2014, replacing the previous \textit{Zones Urbaines Sensibles} (ZUS), \textit{Zones de Redynamisation Urbaine} (ZRU), and \textit{Contrats Urbains de Coh\'{e}sion Sociale} (CUCS) geographies. The new QPV boundaries took effect on January 1, 2015, and were defined using a single, transparent criterion: local income concentration, measured from INSEE's 200-meter gridded income data (\textit{Filosofi}). Specifically, QPV eligibility requires that the median household income in the zone fall below 60 percent of the surrounding urban unit's median income, with a national floor. This algorithm produced 1,362 QPVs in metropolitan France, covering approximately 5.3 million residents across 776 communes.

\subsubsection*{The Policy Bundle} QPV designation triggers a multi-layered bundle of fiscal and investment policies. On the fiscal side, property owners inside QPVs receive a 30 percent reduction in the \textit{Taxe Fonci\`{e}re sur les Propri\'{e}t\'{e}s B\^{a}ties} (TFPB, the main property tax). New construction within QPVs and within 300 meters of the boundary benefits from a reduced VAT rate of 5.5 percent, compared to the standard 20 percent. Businesses locating in QPVs receive up to five years of exemption from the \textit{Cotisation Fonci\`{e}re des Entreprises} (CFE, the business property tax) and, since January 2026, profit tax exemptions. On the investment side, the \textit{Nouveau Programme National de Renouvellement Urbain} (NPNRU), managed by ANRU, allocated 12 billion euros for demolition, reconstruction, and renovation of social housing and public infrastructure in a subset of QPVs deemed most in need.

\subsubsection*{Boundary Characteristics} The QPV boundaries are defined as precise geographic polygons at the sub-commune level, drawn by the \textit{Agence Nationale de la Coh\'{e}sion des Territoires} (ANCT). Unlike school districts or electoral boundaries, QPV perimeters do not follow street centerlines or physical barriers in any systematic way. They were generated algorithmically from gridded income data, creating boundaries that cut through the urban fabric in ways that are neither anticipated nor manipulable by property market participants. A major boundary revision in 2024 modified 960 QPV perimeters, added 111 new QPVs, and removed 40, providing potential future temporal variation.


\section{Data}

The analysis combines two publicly available administrative datasets.

\subsubsection*{Property Transactions} The \textit{Demandes de Valeurs Fonci\`{e}res} (DVF) database, published by France's \textit{Direction G\'{e}n\'{e}rale des Finances Publiques} on data.gouv.fr, records the universe of real estate transactions in France. The geolocalized version provides parcel-centroid coordinates (latitude and longitude) for each transaction, along with transaction price, date, property type (apartment or house), built surface area, number of rooms, and commune and department codes. I use the 2020--2024 vintages, which cover the post-QPV period and provide geolocalized coordinates. I restrict the sample to arm's-length sales (\textit{ventes}) of apartments and houses with non-missing prices, coordinates, and surface area, and exclude outliers below 100 EUR/m$^2$ or above 30,000 EUR/m$^2$.

\subsubsection*{QPV Boundaries} Official QPV boundary polygons (as of the 2015 designation) are obtained from the Geoplateforme WFS maintained by the Institut G\'{e}ographique National (IGN). The data contains 1,372 metropolitan QPV polygons with their official codes and names. All spatial operations are performed in the Lambert-93 projected coordinate system (EPSG:2154), which preserves metric distances across metropolitan France.

\subsubsection*{Sample Construction} I transform all DVF transaction points and QPV polygons to Lambert-93 coordinates. For each of the 1.8 million cleaned transactions, I compute the distance from the transaction's parcel centroid to the nearest point on the QPV union boundary. Transactions inside a QPV polygon receive a positive signed distance; those outside receive a negative signed distance. I retain all transactions within 2 kilometers of the nearest QPV boundary. The final sample contains 1,801,346 transactions, of which 122,386 (6.8 percent) are located inside QPV boundaries. Within the 500-meter bandwidth used in several specifications, the sample contains 705,803 transactions across 2,222 communes and 1,241 distinct QPV boundaries.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Spatial Regression Discontinuity Design}

I estimate the effect of QPV designation on property values using a spatial RDD at the QPV boundary. The running variable is the signed distance $d_i$ from transaction $i$ to the nearest QPV boundary: $d_i > 0$ for properties inside a QPV, $d_i < 0$ for properties outside. The treatment indicator is $D_i = \mathbf{1}[d_i \geq 0]$.

The estimand is the local average treatment effect at the boundary:
\begin{equation}
\tau = \lim_{d \downarrow 0} \E[\log(p_i) \mid d_i = d] - \lim_{d \uparrow 0} \E[\log(p_i) \mid d_i = d]
\end{equation}
where $p_i$ is price per square meter of transaction $i$.

I estimate $\tau$ using the bias-corrected local polynomial estimator of \citet{calonico2014robust} with a triangular kernel and MSE-optimal bandwidth selection. Standard errors are clustered at the commune level to account for spatial correlation within administrative units.

As a complement, I estimate a parametric specification within a fixed 500-meter bandwidth:
\begin{equation}
\log(p_{it}) = \alpha + \tau D_i + \beta_1 d_i + \beta_2 D_i \cdot d_i + \mathbf{X}_i'\gamma + \mu_b + \delta_t + \varepsilon_{it}
\end{equation}
where $\mathbf{X}_i$ includes surface area, number of rooms, and an apartment indicator; $\mu_b$ are boundary-segment fixed effects (indexed by the nearest QPV); and $\delta_t$ are year-quarter fixed effects. Clustering is at the commune level.

\subsection{Identification}

The identifying assumption is that potential outcomes are continuous at the QPV boundary: properties just inside and just outside the boundary would command similar prices in the absence of the QPV policy bundle. Two features of the institutional setting support this assumption.

First, QPV boundaries were drawn algorithmically from gridded income data --- not from property market characteristics. The boundaries cut through the urban fabric at arbitrary points that were neither publicly salient nor anticipatable by market participants at the time of designation.

Second, properties cannot sort across QPV boundaries in response to the policy. Physical structures are fixed in location. While new construction can respond to fiscal incentives, the boundary itself was determined before the development response, and any new construction inside the zone would, if anything, increase prices through the subsidized VAT channel.

I validate the design with a McCrary density test \citep{cattaneo2020simple}, which examines whether the density of transactions is discontinuous at the boundary, and with covariate balance tests that examine whether observable property characteristics change discontinuously at the cutoff.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} reports the main RDD estimates. Column (1) presents the nonparametric bias-corrected estimate: QPV designation is associated with a 15.4 percent decrease in log price per square meter (robust standard error = 0.020, $p < 0.001$). The MSE-optimal bandwidth is 110 meters, using observations from 135,623 transactions just outside the boundary and 96,405 just inside. Column (2) adds property covariates (surface area, rooms, apartment indicator), which increases the point estimate to 18.6 percent. Column (3) reports the parametric specification with boundary-segment and year-quarter fixed effects within 500 meters, yielding a more conservative estimate of 12.7 percent. This represents a property price discount of approximately 370 EUR per square meter at the median, or roughly 28,000 EUR for a typical 75 m$^2$ apartment.

\input{tables/tab2_main}

\subsection{Validity}

\Cref{tab:validity} presents the validity checks. Panel A reports the McCrary density test, which fails to reject continuity of the transaction density at the boundary ($p = 0.20$), supporting the no-manipulation assumption. Panel B tests for discontinuities in predetermined property characteristics. Surface area, number of rooms, apartment share, and transaction price all show statistically significant jumps at the boundary. This is expected: QPV boundaries were drawn based on income, and income correlates with housing characteristics. The relevant comparison is the magnitude of the price discontinuity ($-15.4$ percent) relative to the covariate discontinuities. The parametric specification in Column (3) of \Cref{tab:main}, which absorbs these compositional differences through controls, still yields a 12.7 percent price penalty, suggesting that compositional shifts account for only about 2--3 percentage points of the raw gap. Panel C reports placebo tests at false cutoffs, which show attenuated and generally insignificant effects.

\input{tables/tab4_validity}

\subsection{Robustness}

\Cref{tab:robustness} demonstrates the stability of the main estimate across bandwidths. The effect ranges from $-14.5$ percent at the narrowest bandwidth (250 meters) to $-19.6$ percent at the widest (1,000 meters), with all estimates highly significant. The consistency across bandwidths suggests that the result is not driven by a particular choice of window.

\input{tables/tab3_robustness}

\subsection{Heterogeneity}

\Cref{tab:hetero} reports heterogeneity by property type and period. The designation discount is present for both apartments ($-19.3$ percent) and houses ($-21.3$ percent), with houses showing a somewhat larger penalty, consistent with the hypothesis that single-family homes are more sensitive to neighborhood characteristics than multi-unit buildings. Across time, the effect is remarkably stable: $-16.0$ percent in 2020--2021 and $-15.3$ percent in 2022--2024, suggesting no attenuation of the designation discount over the sample period.

\input{tables/tab5_hetero}


\section{Discussion}

The central finding of this paper --- that QPV designation reduces property values by 13--15 percent despite billions of euros in fiscal incentives --- has three important implications.

\subsubsection*{Fiscal Benefits Cannot Offset the Designation Discount} Standard fiscal capitalization theory predicts that a 30 percent property tax reduction should increase prices by the capitalized value of future tax savings. At a 3 percent discount rate, a 30 percent reduction in an annual property tax of approximately 1,000 EUR translates to a present value of roughly 10,000 EUR, or about 5 percent of a median-priced apartment near a QPV boundary. The reduced VAT on new construction (5.5 versus 20 percent) represents an even larger implicit subsidy. Yet the net price effect is negative and large. This result is a composite of multiple channels: social stigma from the ``priority neighborhood'' label, capitalized expectations about neighborhood quality, possible anticipation of future regulations, and the fiscal benefits themselves. I cannot separately identify these channels with the present design. What the evidence does establish is that the combined negative forces --- whatever their composition --- exceed the positive fiscal capitalization by a wide margin. Notably, the reduced VAT applies within 300 meters of the QPV boundary, meaning that some ``control'' properties also receive fiscal benefits. This contamination, if anything, biases the estimated discount \textit{toward zero}, making the true negative effect of the QPV label likely even larger than the 13--15 percent I estimate.

\subsubsection*{Implications for Place-Based Policy Design} If labeling a neighborhood as disadvantaged depresses property values, then place-based programs face a fundamental design tension: targeting resources requires identifying recipients, but identification itself may impose costs. One solution is to decouple fiscal incentives from visible labeling --- for example, by administering tax benefits through income-based criteria rather than geographic designation. France's 2024 QPV boundary revision, which modified 960 perimeters, provides a future natural experiment to test whether the designation discount changes when boundaries shift.

\subsubsection*{Limitations} Four limitations merit discussion. First, the covariate imbalance at the boundary suggests that some of the estimated discount reflects compositional differences in housing stock rather than a pure labeling effect. The parametric specification with property controls provides a lower bound of 12.7 percent, but unobserved characteristics (building age, maintenance quality, proximity to nuisances) may also vary discontinuously. Second, the geolocalized DVF data are available only from 2020, preventing a pre-2015 placebo test that would directly verify the absence of a discontinuity before QPV designation. Third, I cannot separately identify the contributions of individual policy components (tax reductions, VAT subsidies, ANRU investment) to the net capitalization effect; the ``designation discount'' is the net effect of the entire policy bundle, not a clean estimate of stigma alone. Fourth, the RDD identifies effects at the boundary; the price gap in QPV interiors may be larger or different in character. France's 2024 QPV boundary revision --- which modified 960 perimeters, added 111 new QPVs, and removed 40 --- provides a promising future natural experiment to address several of these concerns through a difference-in-discontinuities design.


\section{Conclusion}

France's \textit{Quartiers Prioritaires} offer tax breaks, reduced VAT, and billions in urban renewal. Yet properties just inside these designated zones sell for 13--15 percent less than those just outside. The ``designation discount'' reveals a fundamental tension in place-based policy: the very act of targeting help to a neighborhood may signal that the neighborhood needs help, and markets punish that signal. Designing place-based programs that deliver resources without broadcasting disadvantage is a first-order challenge for urban policy.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
