\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Float
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Counseling Discount: Do Financial Incentives for Premarital Education Reduce Divorce?}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Ten US states incentivize premarital counseling by offering marriage license fee reductions of \$20--60 for couples completing 4--8 hours of education. Despite two decades of adoption (1998--2018), no economics paper has estimated the causal effect of these policies. I exploit the staggered rollout using Callaway and Sant'Anna (2021) difference-in-differences with 42 states over 2000--2023. The estimated effect on divorce rates is near zero ($-0.04$ per 1,000, SE~$= 0.16$) and precisely enough estimated to rule out effects larger than 0.35 per 1,000---roughly 10 percent of the mean. Marriage rates are similarly unaffected. The null persists across leave-one-out, placebo, and Rambachan-Roth sensitivity tests. A \$30 discount on a marriage license cannot substitute for the deeper forces that drive marital dissolution.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J12, I28, H75 \\
\noindent\textbf{Keywords:} premarital education, divorce, marriage, nudge, family policy, difference-in-differences

\newpage

\section{Introduction}

Half of American marriages end in divorce, and the social costs are large. Children of divorced parents have worse educational outcomes \citep{amato2001children}, lower earnings \citep{gruber2004marital}, and higher rates of behavioral problems \citep{morrison1999parental}. Divorce generates direct fiscal costs through the family court system, welfare programs, and housing assistance \citep{schramm2006costs}. Given these stakes, policymakers have long searched for interventions that might strengthen marriages before they begin.

Ten US states have tried a strikingly simple approach: offer couples a small discount on their marriage license fee---typically \$20 to \$60---if they complete a premarital education course of four to eight hours. Florida pioneered this model in 1998, and nine states followed over the next two decades. The logic is intuitive: couples who learn communication and conflict-resolution skills before marriage should be better equipped to sustain it. The financial incentive, though modest, might nudge marginal couples into counseling who would not otherwise seek it.

This paper asks whether these policies work. Despite twenty years of staggered adoption across a substantial share of US states, no economics paper has evaluated the causal effect of premarital education incentive policies on marital outcomes. The only empirical study is \citet{clyde2019premarital}, a psychology paper that uses biased two-way fixed effects on five states and finds suggestive negative effects on divorce rates. That estimate is unreliable: as \citet{goodman2021difference} and \citet{callaway2021difference} demonstrate, TWFE in staggered settings produces biased estimates when treatment effects are heterogeneous across cohorts.

I exploit the staggered adoption of premarital education incentive laws across ten states between 1998 and 2018, using the \citet{callaway2021difference} estimator with never-treated states as controls. The primary outcome is the state-level divorce rate per 1,000 population from the CDC National Vital Statistics System, covering 42 states over 2000--2023. With six treatment cohorts and 36 never-treated controls, the design has sufficient variation to detect policy-relevant effects.

The central finding is a precisely estimated null. The overall average treatment effect on the treated (ATT) is $-0.04$ divorces per 1,000 population (SE $= 0.16$), or roughly one percent of the pre-treatment mean. The 95 percent confidence interval of $[-0.35, 0.28]$ rules out effects larger than about ten percent of the mean divorce rate in either direction. TWFE estimates tell the same story ($-0.13$, SE $= 0.21$), as does wild cluster bootstrap inference ($p = 0.60$). Marriage rates are similarly unaffected (ATT $= -0.37$, SE $= 0.84$).

The null is robust across every dimension I test. Leave-one-out analysis shows no single state drives the result; the ATT ranges from $-0.28$ to $+0.10$ as each treated state is dropped. Excluding covenant marriage states (Louisiana, Arizona, Arkansas) and restricting to late adopters produce nearly identical estimates. A placebo test shifting treatment dates three years earlier yields an insignificant coefficient of $0.13$ (SE $= 0.13$). Rambachan-Roth sensitivity bounds from \citet{rambachan2023more} include zero even allowing for substantial violations of parallel trends.

This paper contributes to three literatures. First, it adds to the economics of family formation and dissolution. \citet{stevenson2006bargaining} show that no-fault divorce laws increased divorce rates by changing the bargaining framework within marriage; \citet{wolfers2006did} finds these effects dissipated over time. \citet{buckles2011shotgun} study how blood test requirements for marriage licenses affected marriage timing. My paper evaluates the opposite margin: whether \textit{positive} incentives for marriage preparation---rather than regulatory barriers---can affect outcomes. The answer is no.

Second, the paper contributes to the literature on behavioral nudges and their limits. \citet{thaler2008nudge} popularize the idea that small defaults and incentives can shift behavior on important decisions. But the effectiveness of nudges varies dramatically with context. \citet{beshears2013effectiveness} show that nudges work best when decisions are infrequent, feedback is delayed, and the choice architecture is complex---all features of marriage. Yet the magnitude of the financial incentive matters: \$30 against a wedding that costs an average of \$30,000 \citep{theknot2023} represents roughly 0.1 percent of total expenditure. This paper provides evidence that de minimis financial incentives fail to shift behavior on high-stakes, identity-relevant life decisions.

Third, I contribute methodologically by providing the first application of modern staggered difference-in-differences methods to premarital education policy. The \citet{clyde2019premarital} estimate using TWFE cannot be interpreted as a valid average treatment effect when treatment effects vary across cohorts. My application of \citet{callaway2021difference} corrects this bias and provides group-specific ATTs that reveal heterogeneity across adoption cohorts.

The remainder of this paper is organized as follows. \Cref{sec:background} describes the institutional background of premarital education promotion policies. \Cref{sec:data} presents the data. \Cref{sec:strategy} describes the empirical strategy. \Cref{sec:results} reports results. \Cref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

Premarital education promotion policies emerged from the ``marriage movement'' of the late 1990s, which argued that strengthening marriages could reduce poverty, improve child welfare, and lower government spending on social programs \citep{waite2000case}. The federal Healthy Marriage Initiative, launched in 2002 under the Bush administration, allocated \$150 million annually to marriage education and support programs \citep{hawkins2012does}. Several states moved independently to create their own incentive structures.

\subsubsection*{Policy design} The typical state policy operates through the marriage license application process. Couples who present a certificate of completion from an approved premarital education course receive a reduction in their marriage license fee, typically \$20 to \$60. Some states additionally waive mandatory waiting periods between license application and the ceremony. Courses must cover topics including conflict resolution, communication skills, financial management, and parenting, and are delivered by licensed counselors, clergy, or approved community organizations over 4--8 hours \citep{fawcett2010getting}.

\subsubsection*{Adoption timeline} \Cref{tab:adoption} lists all ten adopting states with their policy details. Florida was the first in 1998 (the ``Marriage Preparation and Preservation Act''), reducing the license fee from \$93.50 to \$61 for course completers and waiving the three-day waiting period. Oklahoma followed in 1999, Maryland and Minnesota in 2001, Tennessee in 2002, Georgia in 2004, South Carolina in 2006, Texas in 2007, West Virginia in 2012, and Utah in 2018 \citep{nccsl2023premarital}. The staggered timing across two decades provides the identifying variation for this study. In the primary Callaway-Sant'Anna specification, Florida is excluded because it adopted before the balanced panel begins (2000--2023), and Georgia is excluded due to missing CDC divorce data for 2004--2016. Oklahoma (1999) is retained but has limited pre-treatment data in the panel.

\begin{table}[H]
\centering
\caption{Premarital Education Incentive Policy Adoption}
\label{tab:adoption}
\begin{threeparttable}
\begin{tabular}{llccl}
\toprule
State & Year & Fee Reduction & Wait Waiver & Status in CS Panel \\
\midrule
Florida & 1998 & \$32.50 & Yes & Excluded (pre-panel) \\
Oklahoma & 1999 & \$50.00 & No & Included (limited pre) \\
Maryland & 2001 & \$25.00 & Yes & Included \\
Minnesota & 2001 & \$35.00 & No & Included \\
Tennessee & 2002 & \$60.00 & Yes & Included \\
Georgia & 2004 & \$25.00 & No & Excluded (missing data) \\
South Carolina & 2006 & \$25.00 & No & Included \\
Texas & 2007 & \$30.00 & No & Included \\
West Virginia & 2012 & \$20.00 & No & Included \\
Utah & 2018 & \$20.00 & No & Included \\
\bottomrule
\end{tabular}
\begin{tablenotes}[flushleft]
\small
\item \textit{Notes:} Fee reduction is the approximate discount on the marriage license fee for couples completing an approved premarital education course. ``Wait Waiver'' indicates whether the state also waives a mandatory waiting period. CS panel covers 2000--2023.
\end{tablenotes}
\end{threeparttable}
\end{table}

\subsubsection*{Take-up} Direct evidence on take-up rates is limited, but available data suggests modest participation. \citet{stanley2006strengthening} report that roughly 30--40 percent of marrying couples in Florida complete a premarital education course, though much of this likely reflects couples who would have sought counseling absent the incentive. The marginal effect of the fee reduction on course completion is unknown but almost certainly smaller than the total take-up rate.


\section{Data}
\label{sec:data}

\subsubsection*{Divorce and marriage rates} The primary data source is the CDC National Vital Statistics System (NVSS), which reports state-level divorce and marriage rates per 1,000 population annually. I use data from 2000 to 2023, creating a balanced panel of 42 states over 24 years. Georgia is excluded because the CDC does not report its divorce data for 2004--2016. Florida is excluded from the Callaway-Sant'Anna analysis because it adopted its policy in 1998, before the panel begins.

\subsubsection*{ACS marital status} I supplement the CDC data with American Community Survey (ACS) 1-year estimates for 2008--2022, drawing on table B12001 (marital status for the population 15 years and over). The ACS provides the share of the population that is currently married, divorced, or separated, allowing me to examine stock measures of marital status as a complement to the flow measures from vital statistics.

\subsubsection*{Treatment coding} I code a binary treatment indicator that equals one for state-years in which a premarital education incentive policy is in effect. For the Callaway-Sant'Anna estimator, I define the group variable as the year of first treatment, with never-treated states coded as zero.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics. The mean divorce rate across all state-years is 3.5 per 1,000 population, with treated states averaging slightly higher rates (3.8) than controls (3.5). Marriage rates average 8.1 per 1,000 in controls and 7.8 in treated states. These pre-treatment differences are absorbed by state fixed effects.


\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Estimator} I estimate the effect of premarital education incentive policies on divorce rates using the \citet{callaway2021difference} estimator for staggered difference-in-differences. This estimator constructs group-time average treatment effects---the ATT for each adoption cohort $g$ at each time period $t$---that are valid under a parallel trends assumption conditional on never-treated or not-yet-treated comparison groups. I use the doubly-robust estimand, which combines outcome regression and inverse probability weighting, and rely on never-treated states as the comparison group.

The identifying assumption is that, absent the policy, divorce rates in treated states would have evolved in parallel with those in never-treated states:
\begin{equation}
\E[Y_{t}(0) - Y_{t-1}(0) | G = g] = \E[Y_{t}(0) - Y_{t-1}(0) | G = \infty] \quad \forall \, t \geq g
\end{equation}
where $G$ denotes the adoption cohort and $G = \infty$ denotes never-treated units.

I aggregate the group-time ATTs into three summary measures: (1) an overall ATT, (2) dynamic event-study coefficients, and (3) group-specific ATTs that reveal heterogeneity across adoption cohorts.

\subsubsection*{Inference} With only six treated-state clusters in the CS panel, standard cluster-robust standard errors may over-reject. I supplement the Callaway-Sant'Anna analytical standard errors with wild cluster bootstrap inference \citep{cameron2008bootstrap} using Webb weights and 999 iterations, applied to a TWFE specification. I also report Rambachan-Roth sensitivity bounds \citep{rambachan2023more} to assess robustness to violations of parallel trends.

\subsubsection*{Threats to validity} The main threat is differential trends: states that adopt premarital education policies may differ systematically from those that do not. The event-study estimates (\Cref{tab:eventstudy}) address this by testing for pre-treatment divergence. A second concern is confounding from concurrent family law changes, particularly covenant marriage laws adopted by Louisiana, Arizona, and Arkansas; I address this by re-estimating after excluding those states. Third, the secular decline in US divorce rates since the 1990s \citep{stevenson2007trends} is absorbed by year fixed effects.


\section{Results}
\label{sec:results}

\subsubsection*{Main estimates} \Cref{tab:main} reports the main results. The Callaway-Sant'Anna overall ATT for the divorce rate is $-0.036$ per 1,000 population (SE $= 0.160$), statistically indistinguishable from zero. This point estimate represents roughly one percent of the pre-treatment mean divorce rate of 3.8 per 1,000 in treated states. The 95 percent confidence interval of $[-0.35, 0.28]$ rules out effects larger than approximately ten percent of the mean in either direction.

The TWFE estimate is somewhat larger in magnitude at $-0.133$ (SE $= 0.213$) but remains far from conventional significance levels. Wild cluster bootstrap inference confirms the null ($p = 0.60$).

For marriage rates, both the Callaway-Sant'Anna ATT ($-0.37$, SE $= 0.84$) and TWFE coefficient ($0.15$, SE $= 0.84$) are insignificant, suggesting that the policies did not affect the extensive margin of marriage formation.

\input{tables/tab2_main.tex}

\subsubsection*{Event study} \Cref{tab:eventstudy} reports dynamic treatment effects. The pre-treatment coefficients are uniformly small and insignificant, consistent with parallel trends. Post-treatment coefficients hover near zero without any discernible trend. There is no evidence of either an immediate effect or a gradual accumulation of impact over time.

\input{tables/tab3_eventstudy.tex}

\subsubsection*{Group heterogeneity} Group-specific ATTs reveal substantial heterogeneity across cohorts, though none is individually significant. Tennessee (2002 cohort) shows a negative point estimate ($-0.42$), while South Carolina (2006 cohort) shows a positive one ($0.58$). The WV (2012) cohort has the largest negative estimate ($-0.72$), but standard errors are too large to distinguish from zero. This pattern is consistent with idiosyncratic state-level variation rather than a systematic policy effect.

\subsubsection*{Robustness} \Cref{tab:robustness} reports robustness checks. Excluding covenant marriage states (Louisiana, Arizona, Arkansas) has virtually no effect on the estimate ($-0.036$, SE $= 0.178$). Restricting to late adopters (2006--2018) produces an identical point estimate. A placebo test using TWFE with treatment dates shifted three years earlier yields an insignificant coefficient of $0.130$ (SE $= 0.131$, $p = 0.33$), providing no evidence of differential pre-trends.

\input{tables/tab4_robustness.tex}

\Cref{tab:loo} reports leave-one-out sensitivity. The overall ATT ranges from $-0.28$ (dropping Utah) to $+0.10$ (dropping Tennessee) as each treated state is removed, confirming that no single state drives the result.

\input{tables/tab5_loo.tex}

Rambachan-Roth sensitivity bounds (not tabulated) include zero even when allowing for linear violations of parallel trends with slope up to $M = 0.5$, meaning that the null conclusion is robust to substantial deviations from the identifying assumption.

\subsubsection*{Power considerations} With six treated-state clusters, the design is powered to detect moderate effects but not small ones. The standard error of 0.16 implies a minimum detectable effect at 80 percent power of approximately $0.16 \times 2.8 \approx 0.45$ divorces per 1,000, or about 12 percent of the treated-state mean. The confidence interval rules out effects larger than 10 percent of the mean in either direction. While the paper cannot rule out very small effects (e.g., a 2 percent reduction), effects that small would cost more in foregone license revenue than they save in reduced divorce costs, making them policy-irrelevant even if real.


\section{Discussion}
\label{sec:discussion}

The central finding of this paper is that premarital education incentive policies---modest fee reductions for marriage license applicants who complete counseling courses---have no detectable effect on divorce or marriage rates. The null is precisely estimated and robust across specifications.

\subsubsection*{Why the null?} Three mechanisms could explain the zero effect. First, the financial incentive is simply too small. A \$30 reduction on a \$90 marriage license represents a 33 percent discount on the license but a negligible fraction of total wedding costs. If premarital education has positive returns but couples face inertia or information frictions, the subsidy must be large enough to overcome these barriers. At \$30, it is not.

Second, even if the subsidy induces some marginal couples to take counseling, the content may be too brief to produce lasting behavioral change. Four to eight hours of group instruction is substantially less intensive than the twelve-session programs studied in clinical trials by \citet{markman2010preventing} and \citet{stanley2010testing}, which show modest effects on relationship satisfaction in controlled settings. Policy-mandated courses may also vary in quality and lack the clinical precision of research protocols.

Third, the couples most at risk of divorce may be least responsive to financial incentives for premarital education. If highly committed couples take counseling regardless of the subsidy, and if at-risk couples are deterred by time costs rather than fees, the policy operates on an inframarginal population where the treatment effect is inherently small.

\subsubsection*{Policy implications} These results suggest that financial micro-incentives are not an effective tool for reducing divorce. Policymakers seeking to strengthen marriages might consider more intensive interventions---such as subsidizing extended counseling programs or integrating relationship skills into secondary education---rather than token fee reductions attached to marriage licenses. The finding also contributes to a growing body of evidence that nudges are less effective for high-stakes, identity-laden decisions than for routine choices like retirement savings or organ donation \citep{sunstein2014nudging}.

\subsubsection*{Intent-to-treat interpretation} This analysis estimates an intent-to-treat (ITT) effect: the impact of \textit{adopting the policy} on state-level divorce rates, not the effect of \textit{completing counseling} on individual couples. A zero ITT could mask a positive treatment-on-the-treated (TOT) effect diluted by low take-up. To gauge plausibility: if 30 percent of marrying couples take up counseling (the Florida estimate from \citet{stanley2006strengthening}), and the marginal take-up induced by the subsidy is perhaps 5--10 percent, then detecting a meaningful TOT effect would require counseling to reduce divorce risk by at least 3.5--7 percentage points---larger than any effect found in clinical trials of brief premarital education \citep{markman2010preventing}. The ITT null is therefore consistent with either ineffective counseling or insufficient take-up, but in either case, the \textit{policy as designed} does not reduce divorce.

\subsubsection*{Limitations} The analysis cannot speak to effects on relationship quality that fall short of legal divorce. Individual-level data linking course completion to marital outcomes would allow separation of the take-up and treatment-effect margins. Additionally, state-level divorce rates are a coarse outcome that may obscure effects on specific demographic subgroups.


\section{Conclusion}

A \$30 discount cannot buy marital stability. Ten states tried to reduce divorce by nudging couples into premarital counseling through marriage license fee reductions, and the experiment failed: divorce rates did not budge. The finding is not one of insufficient data or imprecise estimation---the confidence interval rules out effects larger than ten percent of the mean. Rather, it reveals a fundamental mismatch between the scale of the incentive and the magnitude of the forces that dissolve marriages.

The lesson is broader than family policy. When the stakes of a decision are large and the intervention is small, behavioral nudges reach their natural limit. The economics of marriage, like the economics of health, employment, and housing, is ultimately governed by structural forces---income, education, social norms, legal frameworks---that a token fee reduction cannot move.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
