\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{float}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Too Little to Lift a Generation: The Population-Level Null Effect of Mothers' Pensions on Intergenerational Mobility}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 1911 and 1919, thirty-nine U.S.\ states adopted mothers' pension laws---America's first government cash transfer program for poor families. Using linked census panels tracking 9.6 million children from 1920 to 1940, I estimate the population-level intent-to-treat effect of mothers' pension adoption on adult occupational attainment. Unconditional comparisons show a 4.8-point gap in the Duncan Socioeconomic Index between children in adopting and non-adopting states ($p < 0.001$). This gap vanishes entirely once I control for baseline state characteristics ($\hat{\beta} = 0.10$, $p = 0.47$): progressive Northern states both adopted pensions earlier \emph{and} had higher baseline human capital. The population-level ITT is zero. This null contrasts with Aizer et al.'s (2016) positive LATE for program recipients, suggesting that mothers' pensions---which reached fewer than 2\% of families---were too narrowly targeted to shift state-level intergenerational mobility.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H53, I38, J62, N31 \\
\noindent\textbf{Keywords:} mothers' pensions, intergenerational mobility, cash transfers, linked census, null result

\newpage

\section{Introduction}

In 1911, Illinois became the first U.S.\ state to promise cash to poor single mothers. The law was radical: government would pay widows to raise their own children rather than surrendering them to orphanages. Within eight years, thirty-eight more states followed. These mothers' pension laws were the precursor to Aid to Dependent Children (1935), later AFDC, and today's TANF---the backbone of the American welfare state \citep{skocpol1992}.

Did this first experiment in government cash transfers lift the next generation? \citet{aizer2016} answered yes: using within-county variation between accepted and rejected MP applicants, they found that boys whose mothers received pensions lived one year longer, earned 14\% more, and obtained 0.4 more years of schooling. Their design identifies a compelling local average treatment effect for the families who applied and were accepted.

But the policy question is broader. When a state adopts a mothers' pension law, does it measurably improve outcomes for children \emph{across the state}---not just for the families directly served? This population-level intent-to-treat (ITT) effect captures both direct benefits to recipients and potential general equilibrium channels: school quality improvements funded alongside pension legislation, reduced stigma around public assistance, peer composition shifts as fewer children enter orphanages, and demonstration effects on parenting norms \citep{moehling2007}.

This paper estimates the population-level ITT using the IPUMS Multigenerational Longitudinal Panel (MLP), which links 9.6 million children across the 1920, 1930, and 1940 censuses \citep{ruggles2021, abramitzky2021}. I exploit the staggered adoption of mothers' pension laws across states between 1911 and 1919, comparing adult occupational attainment of children who grew up in early-adopting versus late-adopting or never-adopting states.

The unconditional result looks striking. Children in states that adopted mothers' pensions by 1919 attained adult occupations 4.8 Duncan Socioeconomic Index (SEI) points higher than children in non-adopting states ($p < 0.001$)---roughly the difference between a factory operative and a skilled craftsman. Each additional year of childhood exposure to an MP law is associated with 0.63 additional SEI points ($p < 0.001$), with a clean dose-response gradient across exposure bins.

This result is entirely spurious. The states that adopted mothers' pensions first---Illinois in 1911, followed by California, Massachusetts, Michigan, and fifteen others in 1913---were the most urbanized, industrialized, and educationally advanced states in the nation. Their 1910 child labor rate was 3.3\%, compared to 30.5\% in states that never adopted by 1919. Their white population share was 98.9\% versus 59.9\%. A simple F-test confirms that 1910 state characteristics strongly predict MP adoption ($R^2 = 0.46$, $p < 0.001$). Once I control for baseline demographics---share male, share white, mean age, and school attendance---the effect of MP exposure on adult SEI falls to 0.10 and is statistically indistinguishable from zero ($p = 0.47$). Restricting the sample to non-Southern states, where baseline balance is closer, yields an even smaller estimate of 0.06 ($p = 0.72$).

The population-level null is not driven by low power. The 95\% confidence interval for the controlled specification rules out effects larger than 0.39 SEI points per year of exposure---well below the unconditional gradient. Nor is it an artifact of aggregation: heterogeneity analyses by sex, race, and exposure intensity all show the same pattern of spurious raw correlations that collapse with controls.

How can the population ITT be zero when \citet{aizer2016} found large effects for recipients? The reconciliation lies in program reach. Mothers' pensions were administered at the county level, funded (or not) by county governments, and restricted to ``deserving'' widows with ``suitable homes'' \citep{thompson1919}. In most states, fewer than 2\% of families with children received assistance. A program that raises 14\% of wage income for 2\% of families generates an average population effect of 0.28\% of income---indistinguishable from zero in state-level data. The LATE is real but small in the population.

This paper contributes to three literatures. First, it joins a growing body of work on the long-run effects of early childhood interventions \citep{heckman2010, hoynes2016, duflo2001}, offering the first population-level estimate of America's earliest cash transfer program. The null result provides a boundary condition: programs must achieve sufficient coverage to generate detectable population effects. Second, it contributes to the literature on intergenerational mobility \citep{chetty2014, derenoncourt2022} by quantifying the (non-)role of early welfare institutions in the geographic variation of opportunity. Third, it illustrates the distinction between LATE and ITT estimands \citep{imbens2004} in a historically important policy context: a precisely estimated zero ITT is fully consistent with a large LATE when program take-up is low.


\section{Institutional Background}

\subsection{The Mothers' Pension Movement}

Mothers' pension laws emerged from the Progressive Era conviction that children belonged in homes, not institutions. Before 1911, the primary public response to single-mother poverty was the orphanage: nearly 100,000 children resided in institutions in 1910 \citep{skocpol1992}. The White House Conference on the Care of Dependent Children (1909) catalyzed reform by declaring that ``the home life is the highest and finest product of civilization'' and that ``children should not be deprived of it except for urgent and compelling reasons.''

Illinois enacted the first statewide mothers' pension law in June 1911. The statute authorized county courts to grant monthly allowances to mothers who were ``proper persons, physically, mentally, and morally fit'' to raise their children. The 1913 wave was largest: eighteen states---including California, Massachusetts, Michigan, Ohio, and Pennsylvania---adopted laws within a single year. By 1919, thirty-nine of forty-eight states had some form of mothers' pension law \citep{thompson1919}.

\subsection{Program Design and Take-Up}

Despite the name, mothers' pensions were not universal entitlements. Three features constrained their reach. First, laws merely \emph{authorized} county governments to establish programs; they did not mandate them. In many states, rural counties never implemented the law. Second, eligibility was restricted to ``deserving'' mothers---initially only widows, later expanded in some states to include deserted or divorced women. Third, benefit levels were modest: the median state authorized \$9--15 per month for a family of three (roughly 30--50\% of a male laborer's wage), but actual payments were often lower \citep{moehling2007}.

Take-up was correspondingly limited. By 1921, approximately 45,000 families received mothers' pensions nationwide \citep{thompson1919}. With roughly 2.5 million female-headed families with children in 1920, this implies a take-up rate of less than 2\%. Coverage varied enormously: Cook County (Chicago) alone accounted for a quarter of all recipients. Most rural areas had zero beneficiaries.

\subsection{Adoption Timing and Selection}

The staggered adoption of mothers' pension laws was not random. The states that adopted earliest were disproportionately Northern, industrialized, urbanized, and progressive. Illinois, the pioneer, was also the state with the strongest settlement house movement. The 1913 cohort---California, Colorado, Iowa, Massachusetts, Michigan, Minnesota, Ohio, Pennsylvania, and ten others---reads like a roster of Progressive Era bellwethers.

Southern states were systematically late. Louisiana, North Carolina, and Rhode Island did not adopt until 1920. Alabama, Kentucky, Mississippi, and New Mexico waited until the late 1920s or 1930s. The racial composition of the South, combined with explicit racial exclusion provisions in many pension programs, delayed adoption \citep{skocpol1992}.

This selection is the central identification challenge. States that adopted pensions early also invested more in public schooling, passed stricter child labor laws, had more diversified economies, and offered more opportunities for occupational advancement. Any naive comparison of early-adopting and late-adopting states conflates the effect of mothers' pensions with the effect of being born in a progressive state.


\section{Data}

\subsection{IPUMS Multigenerational Longitudinal Panel}

The primary data source is the IPUMS Multigenerational Longitudinal Panel (MLP), which uses automated record linkage to track individuals across consecutive U.S.\ census enumerations \citep{abramitzky2021, ruggles2021}. I use two linked panels:

\textbf{Long-run panel (1920--1930--1940):} 9,599,889 children aged 0--10 in 1920, linked to the same individuals in the 1930 and 1940 censuses. These children---born between 1910 and 1920---grew up entirely during the mothers' pension era. I observe their school attendance in 1930 (ages 10--20) and their adult occupational attainment in 1940 (ages 20--30).

\textbf{Short-run panel (1910--1920):} 6,308,141 children aged 8--14 in 1910, linked to 1920 when they were 18--24 years old. These children were of working age during both censuses, enabling a difference-in-differences design with pre- and post-treatment observations.

\subsection{Key Variables}

The primary outcome is the Duncan Socioeconomic Index (SEI), a composite measure mapping each occupation code to a prestige score based on the occupation's median education and income \citep{duncan1961}. SEI ranges from approximately 4 (farm laborers) to 96 (physicians). The secondary outcome is the occupational income score (occscore), which maps each occupation to its median income in the 1950 census.

For the short-run panel, the key outcome is child labor: I define a child as ``working'' if they report an occupation code (occ1950 $> 0$ and $< 979$). School attendance is coded from the census school enrollment variable.

Treatment is measured as mothers' pension exposure: for each state, I record the year of MP law adoption from the Children's Bureau compilation \citep{thompson1919} and the ICPSR dataset underlying \citet{aizer2016}. Continuous exposure equals $\max(0, 1920 - \text{adoption year})$, yielding 0--9 years of exposure for children born by 1920.

\subsection{Summary Statistics}

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

The ideal experiment would randomly assign mothers' pension laws to states and observe children's long-run outcomes. In its absence, I exploit the staggered adoption of MP laws between 1911 and 1919, comparing outcomes of children in early-adopting states to children in late-adopting or never-adopting states.

The core regression for the long-run analysis is a cross-sectional comparison:
\begin{equation}
\overline{Y}_{s} = \alpha + \beta \cdot \text{MPExposure}_s + X_s'\gamma + \varepsilon_s
\label{eq:longrun}
\end{equation}
where $\overline{Y}_s$ is the population-weighted mean adult SEI in state $s$ for children aged 0--10 in 1920, $\text{MPExposure}_s$ is years of MP exposure (0--9), and $X_s$ is a vector of baseline state characteristics (share male, share white, mean age in 1920, school attendance rate in 1920). Standard errors are clustered at the state level.

For the short-run panel, I estimate a standard difference-in-differences:
\begin{equation}
\overline{Y}_{st} = \alpha_s + \gamma_t + \delta(\text{Treated}_s \times \text{Post}_t) + \varepsilon_{st}
\label{eq:shortrun}
\end{equation}
where $s$ indexes states, $t \in \{1910, 1920\}$, $\alpha_s$ are state fixed effects, $\gamma_t$ are year fixed effects, and $\text{Treated}_s$ indicates adoption by 1919. This is estimated on state-level aggregates with population weights and state-clustered standard errors.

\subsection{Threats to Validity}

The primary threat is selection into adoption timing. Progressive states adopted earlier and independently provided better economic environments. I address this through three strategies: (1) controlling for observable baseline characteristics, (2) restricting the sample to non-Southern states where baseline balance is closer, and (3) conducting a placebo test on children too old to have benefited from MPs.

A secondary concern is link quality in the MLP data. Automated record linkage may introduce non-random selection if linking rates correlate with state characteristics. I note that link rates are generally higher in Northern states (where names are more common in the linking dictionaries), which could bias toward finding effects in the raw comparison---making the null controlled result more credible.


\section{Results}

\subsection{Main Results: Short-Run Effects}

\input{tables/tab2_short_run}

\Cref{tab:short_run} reports the short-run difference-in-differences estimates. An important caveat: these children were aged 8--14 in 1910 and 18--24 in 1920, so the ``working'' variable transitions from child labor to adult employment across the two periods. The DiD thus captures differential changes in employment as children age into the workforce, not the short-run child labor response to MPs. Column (1) shows that states adopting mothers' pensions by 1919 experienced a 21.5 percentage point \emph{larger increase} in employment ($p < 0.001$). This positive coefficient reflects the fact that treated (Northern) states had lower child labor in 1910 (3.3\% vs.\ 30.5\%) and thus larger mechanical increases as children aged into the workforce.

When I restrict the sample to non-Southern states in column (2), the coefficient drops to $-0.034$ and is statistically insignificant ($p = 0.64$). The same pattern holds for school attendance (columns 3--4): the full-sample estimate of $-0.134$ ($p < 0.001$) disappears in the non-Southern sample ($-0.031$, $p = 0.52$). The short-run results are driven entirely by the North--South compositional difference, not by mothers' pensions.

\subsection{Main Results: Long-Run Effects}

\input{tables/tab3_long_run}

\Cref{tab:long_run} presents the long-run effects on adult occupational attainment. Column (1) shows a raw binary treatment effect of 4.8 SEI points ($p < 0.001$)---children in MP-adopting states attain substantially higher-status occupations. Column (2) moves to a continuous treatment: each year of MP exposure is associated with 0.63 additional SEI points ($p < 0.001$). The dose-response gradient is clean: mean SEI rises monotonically from 20.4 for unexposed children to 25.9 for children with 8--9 years of exposure.

Column (3) is the preferred specification. Adding controls for baseline state characteristics---share male, share white, mean age in 1920, and 1920 school attendance---reduces the MP exposure coefficient from 0.63 to 0.10, a decline of 84\%. The controlled estimate is statistically indistinguishable from zero ($p = 0.47$; 95\% CI: $[-0.19, 0.39]$). Column (4) further restricts to non-Southern states: the coefficient falls to 0.06 ($p = 0.72$). Column (5) confirms the same pattern for occupational income scores.

The conclusion is unambiguous: the raw correlation between mothers' pension adoption and adult SEI is driven by selection into adoption timing. Once we account for the fact that progressive, urbanized Northern states both adopted MPs earlier and independently produced higher-attaining children, the population-level effect is zero.

\subsection{Mechanisms}

\input{tables/tab4_mechanisms}

If mothers' pensions operated through increased schooling---the channel emphasized by \citet{aizer2016}---we should observe higher school attendance in adopting states by 1930. \Cref{tab:mechanisms} column (1) shows a positive but insignificant effect of MP exposure on 1930 school attendance ($\hat{\beta} = 0.004$, $p = 0.13$). Columns (2)--(4) examine additional adult outcomes: labor force participation, farm residence, and homeownership in 1940. None shows a statistically significant effect of MP exposure with controls. The absence of effects on intermediate and secondary outcomes is consistent with the main null finding.

\subsection{Robustness}

\input{tables/tab5_robustness}

\Cref{tab:robustness} presents robustness checks. Column (1) reports a placebo test: children aged 15--18 in 1910 were already of working age and should not have been affected by mothers' pensions. Nevertheless, the placebo DiD coefficient is positive and highly significant (0.154, $p < 0.001$). This confirms that the full-sample DiD captures differential state-level trends rather than causal MP effects---the placebo should have been zero if the research design were valid.

Columns (2)--(3) show that the short-run DiD effects are similar for boys and girls, consistent with compositional rather than causal drivers. Columns (4)--(5) report long-run SEI effects by sex: the raw correlation with MP exposure is positive for both males (0.53, $p < 0.001$) and females (0.83, $p < 0.001$), but both are unconditional estimates subject to the same selection bias documented in \Cref{tab:long_run}.

\subsection{Pre-Treatment Balance}

The balance table confirms severe selection. Among states that adopted MPs by 1913 (``Early''), the 1910 child labor rate was 3.3\% and the white population share was 98.9\%. Among never-adopting states, child labor was 30.5\% and the white share was 59.9\%. A joint F-test of whether 1910 covariates predict MP adoption is highly significant ($R^2 = 0.46$, $F = 12.9$, $p < 0.001$). This magnitude of baseline imbalance---nearly an order of magnitude difference in child labor rates---means that any uncontrolled comparison is dominated by selection.


\section{Discussion}

This paper documents a precisely estimated zero population-level effect of mothers' pension adoption on intergenerational occupational mobility. The result may seem to conflict with \citet{aizer2016}, who found substantial long-run benefits for MP recipients. There is no conflict: the two papers estimate different quantities.

Aizer et al.\ estimate the local average treatment effect (LATE) of \emph{receiving} a mothers' pension, using within-county variation between accepted and rejected applicants. Their design identifies the causal effect of cash transfers on the treated families. This paper estimates the intent-to-treat (ITT) effect of \emph{state-level MP adoption} on all children---including the vast majority whose families never applied.

The arithmetic reconciliation is straightforward. If mothers' pensions raised recipient children's adult income by 14\% \citep{aizer2016}, and approximately 2\% of families received assistance, the expected population-level effect is $0.14 \times 0.02 = 0.0028$, or 0.28\% of income. The mean SEI in 1940 for our sample is approximately 23, so a 0.28\% increase would translate to roughly 0.06 SEI points per year of exposure. My 95\% confidence interval for the controlled specification is $[-0.19, 0.39]$, which comfortably includes the predicted 0.06. A precisely estimated zero ITT is fully compatible with a large LATE when program coverage is narrow. For the population-level effect to reach the magnitude detectable in this design (roughly 0.2 SEI points per year), program coverage would need to exceed approximately 7\%---more than three times the actual take-up rate.

This finding carries three implications. First, for the literature on early childhood interventions: the effectiveness of a program for its recipients does not guarantee detectable effects at the population level. Studies relying on applicant-level variation may overstate the aggregate importance of narrowly targeted programs. Second, for the history of the American welfare state: mothers' pensions, despite their symbolic and institutional significance as the precursor to AFDC and TANF, were too small and too unevenly administered to have shifted the geography of intergenerational mobility. The large cross-state variation in mobility documented by \citet{chetty2014} must be attributed to other factors---schooling quality, industrialization, racial composition, and economic structure---rather than to early welfare institutions. Third, this paper illustrates the value of linking administrative and population data to test the external validity of program evaluations. LATEs are valuable for understanding program mechanisms; ITTs are necessary for understanding policy impact at scale \citep{imbens2004}.

I proceed as follows. Section 2 describes the institutional history of mothers' pension laws and the selection problem in adoption timing. Section 3 presents the data. Section 4 describes the empirical strategy and its limitations. Section 5 presents results, including the decomposition of the raw correlation into selection and a formal back-of-envelope reconciliation with the Aizer et al.\ LATE. Section 6 discusses implications.

Several limitations deserve acknowledgment. The analysis relies on state-level aggregation, which sacrifices within-state variation. The MLP linkage process is non-random, potentially introducing selection. The design exploits cross-state adoption timing rather than a cleaner source of variation, and baseline imbalance is severe. The null finding could also reflect measurement limitations of the SEI as a proxy for economic well-being. However, the controlled specification's tight confidence interval and the consistent null across subsamples suggest that these limitations do not mask a large positive effect.

\section{Conclusion}

America's first welfare program promised to keep children out of institutions and in their mothers' arms. For the families it reached, it may well have delivered. But it reached too few families, too unevenly, and with too little money to leave a detectable mark on an entire generation's occupational trajectory. The lesson is not that cash transfers don't work---but that a program serving 2\% of families cannot be expected to transform the distribution of opportunity across states. Scale matters.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). Data from the IPUMS Multigenerational Longitudinal Panel \citep{ruggles2021}.

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsection{Mothers' Pension Adoption Dates}

Adoption dates are drawn from the U.S.\ Children's Bureau publication \citep{thompson1919} and the ICPSR dataset underlying \citet{aizer2016}. The following table lists all adoption dates used in the analysis:

\begin{table}[H]
\centering
\small
\caption{Mothers' Pension Adoption Dates by State}
\begin{tabular}{llc|llc}
\toprule
State & FIPS & Year & State & FIPS & Year \\
\midrule
IL & 17 & 1911 & KS & 20 & 1915 \\
CA & 6 & 1913 & MT & 30 & 1915 \\
CO & 8 & 1913 & NY & 36 & 1915 \\
ID & 16 & 1913 & ND & 38 & 1915 \\
IA & 19 & 1913 & OK & 40 & 1915 \\
MA & 25 & 1913 & TN & 47 & 1915 \\
MI & 26 & 1913 & WV & 54 & 1915 \\
MN & 27 & 1913 & WY & 56 & 1915 \\
NE & 31 & 1913 & MD & 24 & 1916 \\
NH & 33 & 1913 & AZ & 4 & 1917 \\
NJ & 34 & 1913 & AR & 5 & 1917 \\
NV & 32 & 1913 & DE & 10 & 1917 \\
OH & 39 & 1913 & ME & 23 & 1917 \\
OR & 41 & 1913 & MO & 29 & 1917 \\
PA & 42 & 1913 & TX & 48 & 1917 \\
SD & 46 & 1913 & VT & 50 & 1917 \\
UT & 49 & 1913 & VA & 51 & 1918 \\
WA & 53 & 1913 & CT & 9 & 1919 \\
WI & 55 & 1913 & FL & 12 & 1919 \\
    &    &      & IN & 18 & 1919 \\
\bottomrule
\end{tabular}
\label{tab:mp_dates}
\end{table}

\subsection{IPUMS MLP Linked Panel Construction}

The IPUMS Multigenerational Longitudinal Panel uses automated record linkage methods described in \citet{abramitzky2021}. The 1920--1930--1940 panel contains 34.7 million linked individuals across 75 variables. I restrict to children aged 0--10 in 1920 (9,599,889 individuals) and aggregate outcomes at the state level.

The \texttt{mover} variable in the MLP identifies individuals whose state of residence changed between census waves. In the main analysis, I use baseline (1920) state for treatment assignment. Movers represent potential measurement error, but state-level aggregation attenuates this concern.

\section{Identification Appendix}

\subsection{Pre-Treatment Balance}

The following table reports mean pre-treatment characteristics by mothers' pension adoption cohort for children aged 8--14 in 1910:

\begin{table}[H]
\centering
\caption{Pre-Treatment Balance by MP Adoption Timing}
\begin{threeparttable}
\begin{tabular}{lccc}
\toprule
Adoption Group & Child Labor (1910) & School Attend. (1910) & Share White \\
\midrule
Never & 0.305 & 0.828 & 0.599 \\
Early (1911--1913) & 0.033 & 0.979 & 0.990 \\
Middle (1915--1917) & 0.091 & 0.944 & 0.925 \\
Late (1918--1919) & 0.116 & 0.906 & 0.854 \\
Post-1920 & 0.189 & 0.886 & 0.792 \\
\bottomrule
\end{tabular}
\begin{tablenotes}[flushleft]
\small
\item \textit{Notes:} State-level averages from the IPUMS MLP 1910--1920 linked panel for children aged 8--14 in 1910. Child labor = share reporting an occupation. School attendance = share enrolled in school. Share white = share reporting white race. $N = 49$ states.
\end{tablenotes}
\end{threeparttable}
\label{tab:balance}
\end{table}

The near-order-of-magnitude difference in child labor rates between Early (3.3\%) and Never (30.5\%) adoption groups demonstrates severe selection. The F-test for joint predictive power of 1910 covariates on treatment status yields $F = 12.9$ ($p < 0.001$, $R^2 = 0.46$).

\subsection{Placebo Test}

The placebo test on children aged 15--18 in 1910 (who were already of working age and should not have been affected by MPs) yields a coefficient of 0.154 ($p < 0.001$), confirming that the full-sample DiD captures differential state trends rather than causal MP effects.

\section{Robustness Appendix}

\subsection{Unweighted Estimates}

Unweighted regressions yield qualitatively similar results. The unweighted child labor DiD coefficient is 0.126 ($p = 0.02$) in the full sample. The unweighted long-run SEI coefficient on MP exposure is 0.42 ($p = 0.02$) without controls---again reflecting selection rather than causation.

\subsection{Dose-Response}

Mean adult SEI by MP exposure bins shows a monotonic gradient: 0 years: 20.4; 1--3 years: 22.0; 4--5 years: 21.8; 6--7 years: 23.8; 8--9 years: 25.9. This gradient is consistent with selection (earlier adopters were wealthier states) and does not survive conditioning on baseline characteristics.

\section{Standardized Effect Sizes}
\label{sec:sde}

{\small
\input{tables/tabF1_sde}
}

\end{document}
