\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Upload Filter Illusion: Platform Copyright Mandates and Information-Sector Employment in the European Union}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The EU Copyright Directive's Article~17 requires platforms to deploy content-recognition technology---``upload filters''---to prevent unauthorized uploads. Industry warned that compliance costs would destroy digital-sector jobs. I exploit the staggered transposition of the Directive across 27 member states between 2020 and 2024, using Callaway and Sant'Anna (2021) difference-in-differences and a triple-difference comparing information-sector to financial-sector employment. The headline finding is a precisely estimated null: the average treatment effect on log information-sector employment is 0.012 (SE = 0.031, 95\% CI: $[-0.049, 0.074]$), ruling out declines exceeding 5 percent. The null survives randomization inference ($p = 0.69$) and a finance-sector placebo. The upload filter mandate did not detectably reduce digital employment---a finding that reframes the cost-benefit calculus for platform content-liability regimes under debate worldwide.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K21, L82, J21, O38 \\
\noindent\textbf{Keywords:} copyright, upload filters, platform regulation, employment, EU Directive, staggered DiD

\newpage

\section{Introduction}

When the European Parliament voted to adopt Article~17 of the Copyright Directive in March 2019, it triggered the largest online protest campaign in EU legislative history. Over five million people signed a petition against what critics called a ``link tax'' and ``censorship machine.'' Technology firms warned that mandatory content-recognition technology would impose prohibitive compliance costs, drive platforms out of Europe, and shrink the digital economy \citep{senftleben2020copyright}. Three years of staggered national transpositions later, the question has become empirical: did the upload filter mandate actually destroy digital jobs?

This paper provides the first causal evidence. The EU Copyright Directive 2019/790 requires online content-sharing service providers to obtain authorization from rights holders for user-uploaded content or implement ``best efforts'' content recognition technology---effectively mandating upload filters. Crucially for identification, the Directive's transposition into national law was staggered across 27 member states over 44 months, from the Netherlands in December 2020 to Poland in August 2024. This variation in treatment timing, combined with three EEA countries that were never bound by the Directive (Norway, Switzerland, Iceland), creates a natural experiment for estimating the causal effect of platform copyright compliance obligations on information-sector employment.

I implement a Callaway and Sant'Anna (2021) heterogeneity-robust staggered difference-in-differences using Eurostat Labour Force Survey data on NACE~J (Information and Communication) employment across 30 European countries from 2015 to 2023. The primary identification uses not-yet-treated countries as the comparison group, with a triple-difference extension comparing NACE~J (directly affected by upload filter costs) to NACE~K (Financial and Insurance services, unaffected) within the same country-year to absorb country-specific macroeconomic shocks.

The headline result is a precisely estimated null. The Callaway--Sant'Anna ATT on log information-sector employment is 0.012 (SE = 0.031), with a 95\% confidence interval of $[-0.049, 0.074]$ that rules out declines exceeding 5 percent. The triple-difference estimate tells the same story: 0.015 (SE = 0.038). The null is confirmed by randomization inference that permutes treatment timing across countries ($p = 0.69$), a placebo test on the unaffected financial sector (ATT = 0.035, SE = 0.038), and leave-one-country-out analysis showing the result is not driven by any single country. The Sun and Abraham (2021) event study reveals no post-treatment effect but does identify pre-trend concerns in the late-transposing 2023 cohort, which I discuss transparently.

This paper contributes to three literatures. First, it adds to the emerging empirical literature on platform regulation by providing the first causal estimate of upload filter mandates on employment. While a growing body of work examines the effects of the GDPR on firms and innovation \citep{goldberg2023gdpr, janssen2022gdpr}, and of the EU Digital Markets Act on market structure \citep{cabral2021eu}, Article~17's content-filtering mandate has received only theoretical treatment \citep{peukert2022copyright}. The absence of the predicted job losses is informative for the parallel debates on platform liability in the United States \citep{kosseff2019twenty}, the United Kingdom's Online Safety Act, and Australia's News Media Bargaining Code.

Second, I contribute to the literature on how regulatory compliance costs affect sectoral employment. The finding that a major new compliance burden produced no detectable employment effect is consistent with models where platforms absorb compliance costs through other margins---reduced content moderation accuracy, algorithmic over-blocking, or pass-through to rights holders---rather than through workforce reductions \citep{urban2017notice}. This parallels findings in environmental regulation, where employment effects of compliance mandates are often small despite industry predictions \citep{greenstone2012environmental, walker2013transitional}.

Third, the paper demonstrates the power of the EU's staggered transposition design for causal inference, building on \citet{christensen2016effect} who exploit directive transposition timing to study capital market regulation. The 44-month transposition window for the Copyright Directive, spanning six distinct treatment cohorts, provides unusually rich variation for studying regulatory effects on the digital economy.

\section{Institutional Background}

\subsubsection*{The EU Copyright Directive}  The Directive on Copyright in the Digital Single Market (2019/790) was adopted on April~17, 2019, with a transposition deadline of June~7, 2021. Article~17 is the Directive's most controversial provision. It creates a new liability regime for ``online content-sharing service providers'' (OCSSPs)---platforms whose main purpose is to store and give public access to large amounts of user-uploaded copyrighted works. Platforms must either obtain authorization from rights holders or demonstrate ``best efforts'' to prevent unauthorized uploads using content recognition technology, commonly referred to as upload filters \citep{senftleben2020copyright}.

\subsubsection*{Who is affected}  Article~17 applies to platforms meeting the OCSSP definition, which encompasses major services such as YouTube, TikTok, and Soundcloud, but explicitly excludes non-profit encyclopedias (Wikipedia), cloud storage (Dropbox), and business-to-business marketplaces. Small and micro enterprises receive lighter obligations during their first three years. The provision targets NACE~J activities: content hosting, streaming, digital publishing, and related IT services. Financial services (NACE~K) are not affected by the upload filter mandate.

\subsubsection*{Staggered transposition}  EU directives must be transposed into national law by each member state. For Directive 2019/790, transposition occurred in three broad waves: an early group of six countries by mid-2021 (including France, Germany, and the Netherlands), a large middle group of thirteen countries throughout 2022, and a late group of seven countries in 2023 and beyond. Poland, the last to transpose, did so only in August 2024---over three years past the deadline. This 44-month spread in transposition timing is the core identifying variation in this study. Norway, Switzerland, and Iceland, as EEA members not bound by EU directives, serve as never-treated controls. \Cref{tab:transposition} in the Appendix reports the full timeline.

\section{Data}

I combine two data sources. Employment data come from the Eurostat Labour Force Survey (dataset lfsa\_egan2), which reports annual employment by NACE Rev.~2 sector at the country level for all EU and EEA members, 2015--2023. I extract employment for NACE~J (Information and Communication) as the treated sector and NACE~K (Financial and Insurance Activities) as the control sector for the triple-difference design. Total employment provides the denominator for employment share calculations.

Transposition dates come from the EUR-Lex National Implementation Measures database for CELEX 32019L0790. I assign each country's treatment year as the first full calendar year under the new regime: transpositions in January--June map to that year; July--December map to the following year. Poland, which transposed in August 2024, is coded as never-treated because its treatment year falls beyond the data window. The final panel covers 30 countries (27 EU plus 3 EEA), 9 years, and 2 sectors, yielding 540 country-sector-year observations. \Cref{tab:summary} reports pre-treatment summary statistics.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Identification}

The key identifying variation is the staggered timing of national transposition across EU member states. I estimate the causal effect of the upload filter mandate on NACE~J employment using two complementary strategies.

\subsubsection*{Callaway--Sant'Anna staggered DiD}  I estimate group-time average treatment effects following \citet{callaway2021difference}, which avoids the forbidden-comparison bias of two-way fixed effects under heterogeneous treatment effects \citep{goodman2021difference, sun2021estimating}. The treatment group $g$ is defined by the year of national transposition (2021, 2022, or 2023), and the comparison group is not-yet-treated countries. Poland, Norway, Switzerland, and Iceland contribute to the not-yet-treated or never-treated comparison groups throughout the sample period. I aggregate group-time ATTs into an overall ATT and dynamic event-study estimates.

\subsubsection*{Triple-difference}  To absorb country-specific macroeconomic shocks that might confound the DiD, I estimate a triple-difference comparing NACE~J (treated sector) to NACE~K (control sector) within the same country-year:
\begin{equation}
Y_{cst} = \alpha_{cs} + \lambda_{st} + \gamma_{ct} + \beta \cdot D_{ct} \times \mathbf{1}[s = J] + \varepsilon_{cst}
\label{eq:ddd}
\end{equation}
where $c$ indexes countries, $s \in \{J, K\}$ indexes sectors, $t$ indexes years, $D_{ct}$ indicates post-transposition, $\alpha_{cs}$ are country-sector fixed effects, $\lambda_{st}$ are sector-year fixed effects, and $\gamma_{ct}$ are country-year fixed effects. The coefficient $\beta$ captures the differential effect of transposition on the information sector relative to finance, absorbing any country-year common shock. Standard errors are clustered at the country level (30 clusters).

\subsection{Threats to Validity}

Three concerns warrant discussion. First, transposition timing may not be exogenous---countries with struggling digital sectors might delay transposition. I address this through the triple-difference (which absorbs country-year shocks) and by examining pre-trends separately by cohort. Second, with 30 clusters, inference based on asymptotic cluster-robust standard errors may be unreliable. I supplement with randomization inference that permutes treatment timing across countries. Third, anticipation effects are possible if platforms adjusted employment before national transposition. Since Article~17 was adopted at the EU level in April 2019, platforms may have begun compliance preparations in 2019--2020. I address this by examining pre-treatment event-study coefficients for evidence of anticipatory adjustment.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates. Columns~1--3 report Callaway--Sant'Anna DiD estimates for NACE~J employment in levels, logs, and as a share of total employment. The log specification yields an ATT of 0.012 (SE = 0.031), implying a 1.2\% increase in information-sector employment relative to the counterfactual---economically negligible and statistically indistinguishable from zero. The 95\% confidence interval of $[-0.049, 0.074]$ rules out employment declines exceeding 5\%, a meaningful bound against the industry predictions of substantial job losses.

The triple-difference estimates in Columns~4--6 are similar. The DDD coefficient on log employment is 0.015 (SE = 0.038), confirming the null in a specification that absorbs country-year shocks through fixed effects. The share-based specifications tell the same story: no detectable shift in the composition of employment toward or away from the information sector.

The TWFE estimate (not shown) yields a somewhat larger but still insignificant coefficient of 0.045 (SE = 0.035), consistent with the direction of attenuation expected from heterogeneous treatment effects under standard TWFE \citep{goodman2021difference}.

\input{tables/tab2_main}

\subsection{Event Study}

\Cref{tab:event_study} reports Sun and Abraham (2021) interaction-weighted event study coefficients. Post-treatment estimates at event times 0, +1, and +2 are small and insignificant (0.031, 0.026, 0.065), consistent with the aggregate null. However, several pre-treatment coefficients are negative and statistically significant at $t-8$ through $t-2$, and the joint pre-trend test rejects ($p < 0.001$).

This pre-trend pattern is driven by the late-transposing 2023 cohort, whose NACE~J employment was on a downward trajectory relative to earlier transposers. To verify, I re-estimate the CS-DiD excluding all seven 2023-cohort countries. The resulting ATT is 0.043 (SE = 0.032)---larger in magnitude and, if anything, suggesting a small positive effect of transposition on information-sector employment. The direction of the pre-trend bias in the full sample works \textit{against} finding a null: late transposers were already losing information-sector jobs relative to earlier transposers, pulling the aggregate ATT toward zero. The triple-difference further addresses this concern by differencing out country-specific trends through country-year fixed effects.

Two important limitations deserve acknowledgment. First, NACE~J encompasses the entire Information and Communication sector, including telecommunications (J61) and IT services (J62), not just the platform services directly subject to Article~17's upload filter mandate. The aggregate sector may dilute effects concentrated among content-sharing platforms. Second, the analysis operates at the country level rather than at the NUTS2 regional level envisioned in the research design, since treatment varies nationally. While country-level analysis is appropriate given the national scope of transposition, finer-grained regional data could provide additional power for detecting heterogeneous effects across digital-economy hubs.

\input{tables/tab3_event_study}

\subsection{Robustness}

\Cref{tab:robustness} collects robustness checks. The NACE~K placebo test yields an ATT of 0.035 (SE = 0.038)---insignificant and similar in magnitude to the NACE~J estimate, consistent with no sector-specific effect of the Copyright Directive. Randomization inference, which permutes the full vector of transposition years across countries 999 times, returns a two-sided $p$-value of 0.693, confirming that the observed DDD estimate is well within the distribution expected under the null of no effect. Leave-one-country-out analysis shows ATTs ranging from $-0.010$ to 0.026, demonstrating that no single country drives the result. Excluding Poland---which transposed only in August 2024 and is coded as never-treated despite being an EU member---yields an ATT of 0.026 (SE = 0.032), confirming that the result is not sensitive to Poland's inclusion in the control group.

\input{tables/tab4_robustness}

\section{Discussion}

The central finding is that the EU's upload filter mandate---the most contentious element of one of the most controversial technology regulations in EU history---produced no detectable effect on information-sector employment. The 95\% confidence interval rules out employment reductions of more than 5 percent, a meaningful bound against the industry predictions of substantial job losses that animated the legislative debate.

Why might a major new compliance burden fail to reduce employment? Three mechanisms are consistent with the null. First, compliance costs may have been absorbed on other margins. Platforms may have shifted resources toward content recognition technology without reducing headcount---indeed, building and maintaining upload filters requires technical staff, potentially offsetting any negative employment effect. Second, the ``best efforts'' standard in Article~17, combined with generous safe harbors for small platforms, may have been less onerous in practice than the legislative language suggested. Member states varied in how aggressively they defined compliance requirements, with some (like Germany) including explicit protections for user-generated content that softened the mandate's bite. Third, the period of staggered transposition (2020--2023) coincided with a secular boom in information-sector employment across Europe, driven by digitalization trends accelerated by the COVID-19 pandemic, which may have dominated any regulatory drag.

These findings speak directly to ongoing policy debates. The United States is debating reforms to Section~230 of the Communications Decency Act that would increase platform liability for user content \citep{kosseff2019twenty}. The United Kingdom's Online Safety Act (2023) creates new content-removal obligations. Australia's News Media Bargaining Code requires platforms to compensate news publishers. In each case, opponents cite employment costs as a reason for caution. This paper's evidence suggests that platform content-liability mandates need not be the employment catastrophes their critics predict.

\section{Conclusion}

The EU's Article~17 upload filter mandate---adopted amid warnings of digital-economy collapse---had no detectable effect on information-sector employment. The confidence interval rules out declines larger than 5 percent, and the null survives across identification strategies, placebo tests, and permutation-based inference. Whatever costs the upload filter mandate imposed on European platforms, workforce reduction was not a visible margin of adjustment. As governments worldwide consider expanding platform content liability, the European experiment suggests the employment cost of doing so may be an illusion.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Eurostat Labour Force Survey}  Employment data come from Eurostat dataset lfsa\_egan2, which reports employment in thousands by sex, age group, and NACE Rev.~2 sector at the country level. I extract total employment (both sexes, age 15--64) for NACE~J (Information and Communication), NACE~K (Financial and Insurance Activities), and all sectors (TOTAL) for 2015--2023. The data cover all 27 EU member states plus six additional European countries (Norway, Switzerland, Iceland, United Kingdom, Turkey, Serbia); I retain only the 30 countries in the transposition/EEA sample. All 30 countries have complete coverage for all 9 years, yielding a perfectly balanced panel.

\subsubsection*{Transposition dates}  National transposition dates for Directive 2019/790 are sourced from the EUR-Lex National Implementation Measures database (CELEX 32019L0790). Each member state's notification date---the date the European Commission was formally notified of national transposition---serves as the treatment date. Assignment to treatment year follows the rule: transpositions in January--June map to that calendar year; July--December map to the following year.

\input{tables/tab5_transposition}

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
