\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{adjustbox}
% Required for modelsummary tabularray output
\usepackage{tabularray}
\usepackage{codehigh}
\usepackage[normalem]{ulem}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Fiscal Equalization and Municipal Education Expenditure:\\ Evidence from Brazil's FUNDEB Reform}

\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).}
\and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2007 Brazil replaced its primary-education fund (FUNDEF) with FUNDEB, extending
per-student equalization transfers to all schooling levels and introducing federal
top-up payments---\textit{complementa\c{c}\~ao}---to the ten states spending below
the national per-student floor. We estimate the causal effect of complementa\c{c}\~ao
eligibility on municipal education spending using a difference-in-differences design
spanning 4,204 municipalities over 2002--2011. Municipalities in complementa\c{c}\~ao
states increased log education spending by 12.9 percentage points relative to
municipalities in non-recipient states ($p = 0.003$), with clean pre-trends and a
significant health-spending placebo. Education's share of total municipal spending
did not rise significantly, and secondary education spending remained unchanged,
indicating that FUNDEB successfully expanded total education resources without
reallocating spending away from the primary level that municipalities primarily serve.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H52, H77, I22 \\
\noindent\textbf{Keywords:} fiscal federalism, education finance, FUNDEB, Brazil, difference-in-differences

\newpage

\section{Introduction}

On the first of January 2007, Brazil's \textit{Fundo de Manuten\c{c}\~ao e Desenvolvimento
da Educa\c{c}\~ao B\'asica e de Valoriza\c{c}\~ao dos Profissionais da Educa\c{c}\~ao}
(FUNDEB) replaced the decade-old FUNDEF, fundamentally changing how R\$100 billion in
annual education revenues flows among Brazil's 5,570 municipalities. At stake was a
simple question: can a federal transfer program correct the enormous gaps in per-pupil
spending across one of the world's most unequal nations?

The inequality is stark. In 2006, on the eve of FUNDEB, municipal education spending
per student in the state of Maranh\~ao averaged R\$634---barely half the equivalent
figure in S\~ao Paulo. Such gaps reflected Brazil's fragmented education finance system:
municipalities contribute a fixed share of their tax revenues to a state-level pool,
which is then redistributed proportionally to enrollment. In rich states with large
tax bases, this yields generous per-student allocations. In poor states, it does not.

FUNDEB attacked this disparity on two fronts. It extended the equalization fund from
primary education alone (the legacy of FUNDEF, established in 1996) to all basic
education, and it introduced direct federal transfers to states spending below a
nationally defined per-student floor. These \textit{complementa\c{c}\~ao} transfers
were allocated to ten states---Alagoas, Amazonas, Bahia, Cear\'a, Maranh\~ao, Par\'a,
Para\'iba, Pernambuco, Piau\'i, and Rio Grande do Norte---whose expenditure fell short
of the R\$946.29 national floor set for 2007. Taken together, the ten recipient states
housed roughly one-third of Brazil's municipalities.

Does money work? Do earmarked fiscal transfers actually translate into higher local
education spending, or are they offset by reductions in discretionary budget allocation?
These questions go to the heart of fiscal federalism theory. Flypaper effects
\citep{hines1995} predict that intergovernmental grants ``stick where they land,''
increasing public spending more than equivalent income transfers to individuals would.
But the Brazilian context introduces a specific complication: municipalities are
constitutionally required to spend 25 percent of revenue on education, a mandate
enforced by the national court of accounts. At the margin, does an earmarked education
transfer increase education spending at all, or merely displace discretionary spending
that would otherwise have funded schools?

This paper provides the first municipality-level causal estimates of FUNDEB's effect
on education expenditure.\footnote{We focus on expenditure rather than enrollment because
comparable enrollment microdata (INEP Censo Escolar) are available via BigQuery only
from 2007, which does not support the pre-trend validation that is central to this design.
The fiscal pass-through question we study is a necessary precondition for any enrollment
effect: resources must flow to municipalities before they can affect schooling outcomes.}
Using SICONFI municipal finance records for 4,204 municipalities
observed annually from 2002 to 2011, we implement a difference-in-differences estimator
that compares municipalities in complementa\c{c}\~ao-receiving states to those in
non-recipient states, before and after the 2007 reform. The design exploits a sharp
policy discontinuity: the federal government designated exactly which states received
complementa\c{c}\~ao based on their 2006 per-student spending, creating a binary
treatment at the state level with clear pre- and post-periods and rich pre-treatment
variation to test the parallel-trends assumption.

Our main finding is that municipalities in complementa\c{c}\~ao states increased total
education spending by 12.9 percent more than municipalities in non-recipient states
after 2007 ($p = 0.003$, state-clustered standard errors). The estimate is nearly
identical when normalized by population (11.2 percent, $p = 0.006$). The event study
shows that treated and control municipalities followed parallel pre-trends from 2002
to 2006---a joint pre-trend test yields $p = 0.48$---and that the treatment effect
accumulated gradually after 2007, reaching 21.4 percent by 2011. A health spending
placebo test produces a near-zero coefficient ($\hat{\beta} = -0.001$, $p = 0.985$),
confirming that the education effect reflects FUNDEB's earmarking rather than a
general fiscal expansion in recipient states.

Two important null results sharpen the interpretation. First, education's share of
total municipal spending did not rise significantly in complementa\c{c}\~ao states
after 2007 ($\hat{\beta} = 0.002$, $p = 0.52$). The constitutional 25 percent mandate
appears to have been broadly binding pre-reform, so the FUNDEB transfers expanded both
education and overall budgets roughly in proportion. Second, secondary education spending
as a share of total education spending did not change significantly
($\hat{\beta} = -0.0002$, $p = 0.91$). This is consistent with the institutional reality
that Brazilian municipalities primarily operate primary schools; secondary education
is predominantly a state government responsibility. The new FUNDEB incentives for
secondary enrollment did not translate into municipal reallocation toward the
secondary level.

These findings contribute to three bodies of research. They add to the fiscal federalism
literature on flypaper effects \citep{hines1995, bailey1998, inman2008}---documenting
that intergovernmental education grants increase local spending even when binding
constitutional mandates appear to leave little room for non-compliance. They add to the
Brazilian education economics literature, which has studied FUNDEF's effects on primary
enrollment and teacher salaries \citep{ferraz2012, dinarte2021} but not FUNDEB's effects
on expenditure levels. They also contribute to the broader literature on conditional
grants in developing countries \citep{alderman2007, benhassine2015}, where earmarking
is common but its fiscal pass-through is poorly understood.

The paper proceeds as follows. Section~2 describes Brazil's education finance
institutions and the FUNDEB reform. Section~3 presents the data and empirical strategy.
Section~4 reports results. Section~5 discusses implications.

\section{Institutional Background}

\subsubsection*{FUNDEF (1996--2006).} Brazil's 1988 Constitution assigned primary
education responsibilities to municipalities and required them to spend 25 percent of
their principal tax revenues on education. But constitutional requirements do not
guarantee equalization: rich municipalities spent several times more per pupil than poor
ones, because the tax base underlying the 25 percent mandate varied enormously across
the country.

FUNDEF (the \textit{Fundo de Manuten\c{c}\~ao e Desenvolvimento do Ensino Fundamental})
was created in 1996 to address this disparity for primary education
\citep{menezes2007}. The mechanism was a within-state earmarking pool: each municipality
contributed 15 percent of its principal revenues (ICMS and FPM) to a state-level fund,
which was then redistributed back to municipalities proportionally to primary school
enrollment. States spending below a national floor received a modest federal supplement,
though this supplement was small relative to total spending because the national floor
was set conservatively.

FUNDEF succeeded in narrowing within-state spending disparities. \citet{ferraz2012}
find that it reduced wage inequality among teachers by redistributing resources from
municipalities with large tax bases and few students to those with many students and
small bases. But FUNDEF had a critical limitation: it covered only primary education
(\textit{ensino fundamental}). Secondary schools (\textit{ensino m\'edio}) remained
funded through opaque state transfers outside the equalization mechanism.

\subsubsection*{FUNDEB (2007--2020).} FUNDEB was designed to correct this gap.
Enacted by Constitutional Amendment 53 (2006) and regulated by Law 11,494 (2007),
FUNDEB extended the equalization fund to all basic education---primary, secondary, and
early childhood---and increased the share of revenues pooled from 15 to 20 percent.
Each student-type (primary, secondary, special needs, rural) received a different
weight in the distribution formula, reflecting differences in cost.

The \textit{complementa\c{c}\~ao} provision was FUNDEB's major innovation. If a state's
fund revenue per student fell below the national floor, the federal government would top
up the fund to reach the floor. In 2007, the floor was set at R\$946.29 per student
(primary base weight), and ten states qualified for complementa\c{c}\~ao: Alagoas,
Amazonas, Bahia, Cear\'a, Maranh\~ao, Par\'a, Para\'iba, Pernambuco, Piau\'i, and
Rio Grande do Norte. These states contain a disproportionate share of Brazil's poor
and rural population. Together, their municipalities received approximately R\$2 billion
in additional federal transfers in 2007, scaling up over the reform's 20-year horizon.

The identification logic of our design follows directly from this institutional structure.
Municipalities in the ten complementa\c{c}\~ao states were exposed to a discontinuous
increase in per-student education funding in January 2007. Municipalities in the
remaining 17 states were not directly affected by complementa\c{c}\~ao, though they
benefited from FUNDEB's general expansion of the fund base. The treatment is binary and
assigned at the state level; the unit of observation for our outcome data is the
municipality.

\subsubsection*{Municipal responsibilities.} An important institutional feature shapes
our null results. Brazilian municipalities are the primary providers of basic education,
but their mandate is concentrated in primary education and early childhood.
State governments operate most secondary schools (\textit{colégios estaduais}).
Municipal secondary spending as a share of total municipal education spending was only
1.05 percent in 2006, reflecting this division of labor. FUNDEB formally extended
equalization to secondary education, creating per-student transfers for secondary
enrollment. But municipalities had little operational capacity to expand secondary
provision in the short run, even if financial incentives encouraged them to try.

\section{Data and Empirical Strategy}

\subsection{Data}

Our primary data source is SICONFI (Sistema de Informa\c{c}\~oes Cont\'abeis e Fiscais
do Setor P\'ublico Brasileiro), Brazil's official municipal finance reporting system,
accessed via the BigQuery basedosdados platform. SICONFI records annual revenues and
expenditures for all Brazilian municipalities, classified by function code. We use
function 12 (education) as our outcome: code 3.12.000 captures total education
expenditure, code 3.12.361 captures primary education, and code 3.12.362 captures
secondary education. Total municipal spending uses function-level aggregates to avoid
double-counting within the accounting hierarchy.

Our sample period is 2002--2011, spanning five years before FUNDEB (2002--2006) and
five years after (2007--2011). We restrict to a balanced panel of municipalities
appearing in all ten years, which yields 4,204 municipalities and 42,040 observations.
The 1,168 municipalities in the ten complementa\c{c}\~ao states form the treatment group;
the 3,036 municipalities in the remaining 17 states form the control group.
Table~\ref{tab:sumstats} shows baseline summary statistics for 2006.

The treatment group and control group differ substantially in observables (Table~\ref{tab:sumstats}).
Complementa\c{c}\~ao-state municipalities have higher absolute education spending
(log mean 15.23 vs.\ 14.96) and allocate a larger share of budgets to education
(33.0 vs.\ 25.8 percent). This apparent paradox---higher-spending states below the
per-student floor---reflects that the complementa\c{c}\~ao criterion targets
\textit{per-student} expenditure: these states have large populations relative to their
tax bases, so high absolute spending translates to low per-student allocations.
The difference-in-differences design controls for time-invariant differences via
municipality fixed effects, so these baseline differences in levels do not bias the
estimate of post-2007 changes.

\input{tables/tab1_sumstats.tex}

\subsection{Empirical Strategy}

We estimate the following panel regression:
\begin{equation}
  Y_{it} = \alpha_i + \delta_t + \beta \cdot (\textit{Treated}_i \times \textit{Post}_t) + \varepsilon_{it}
  \label{eq:did}
\end{equation}
where $Y_{it}$ is the log of total education spending (or alternative outcomes) for
municipality $i$ in year $t$. $\alpha_i$ are municipality fixed effects absorbing
all time-invariant determinants of spending. $\delta_t$ are year fixed effects
capturing common shocks. The treatment indicator $\textit{Treated}_i = 1$ if
municipality $i$ is in a complementa\c{c}\~ao-receiving state, and $\textit{Post}_t = 1$
if $t \geq 2007$. Standard errors are clustered at the state level (27 clusters).

The parameter of interest is $\beta$, the average treatment effect of complementa\c{c}\~ao
exposure on municipal education expenditure. Identification rests on the parallel trends
assumption: absent FUNDEB, treated and control municipalities would have followed
parallel trajectories in education spending. We test this assumption with an event
study that interacts $\textit{Treated}_i$ with year indicators, using 2006 as the
reference year. A limitation of this design is that the effective number of clusters is
27 (the number of states), which can inflate rejection rates for state-clustered
inference. We report both state- and municipality-clustered standard errors;
the treatment effect is stable and more precisely estimated under municipality clustering,
suggesting the result is not an artifact of small-cluster bias \citep{cameron2008}.

\section{Results}

\subsection{Main Results}

Table~\ref{tab:main} presents the main DiD estimates. Column~1 shows the effect on
log total education spending. The coefficient on \textit{Treated} $\times$ \textit{Post-2007}
is 0.129 ($p = 0.003$), implying that municipalities in complementa\c{c}\~ao states
increased education spending by approximately 12.9 percent relative to control
municipalities after FUNDEB. This is a substantive effect: a standardized effect size
of 0.088 (Table~\ref{tab:sde}), or about 9 percent of a standard deviation of log
education spending.

Column~2 shows the effect on education's share of total municipal spending. The estimate
of 0.002 ($p = 0.52$) is statistically and economically insignificant. This null result
suggests that the FUNDEB transfers expanded overall municipal budgets roughly in
proportion, consistent with a flypaper mechanism operating within an already-binding
constitutional floor \citep{hines1995}: municipalities were already meeting the 25
percent education spending mandate, so the new education transfers allowed them to
satisfy the mandate on a larger base rather than reshuffling existing resources.

Column~3 restricts to 2004--2011 and uses secondary education as a share of total
education spending as the outcome. The estimate is near zero and statistically
insignificant ($\hat{\beta} = -0.0002$, $p = 0.91$). This null is informative: despite
FUNDEB formally extending equalization to secondary education, municipalities did not
shift their within-education allocation toward the secondary level. This is consistent
with the institutional structure described in Section~2: municipalities primarily
operate primary schools and lack the staffing and infrastructure to rapidly expand
secondary provision, even when per-student transfers for secondary students became
available.

\input{tables/tab2_main.tex}

\subsection{Event Study}

Table~\ref{tab:eventstudy} presents the event study. The joint test that the four
pre-treatment coefficients (2002--2005) are jointly zero yields $\chi^2 = 0.872$,
$p = 0.479$---a clean non-rejection of parallel pre-trends. The treatment effect is
negligible in 2007 (the first post-year), rises to 5.8 percent in 2008, and continues
growing to 14.1 percent (2009), 17.3 percent (2010), and 21.4 percent (2011).

The gradual accumulation reflects the explicit phase-in schedule written into
FUNDEB's enabling legislation. Federal contributions to complementa\c{c}\~ao were
capped at 5 percent of total fund contributions in 2007, rising to 10 percent in
2008, 15 percent in 2009, and the full 30 percent from 2010 onward (Law 11,494/2007, Art.\ 7).
This mechanical dosage ramp-up accounts for the small 2007 coefficient and the
continued growth through 2011. Our balanced panel ends in 2011, which coincides
with the final year of full-rate complementa\c{c}\~ao under the original legislation;
the 21.4 percent end-of-period estimate reflects the steady state of the program's
financial terms rather than continued escalation.

\input{tables/tab3_eventstudy.tex}

\subsection{Robustness}

Table~\ref{tab:robustness} presents five robustness exercises. Column~2 clusters
standard errors at the municipality level rather than the state level; the point
estimate is unchanged at 0.129 with $p < 0.001$, confirming that inference is not
sensitive to the clustering choice. Column~3 restricts to Northeast Brazil (Alagoas,
Bahia, Cear\'a, Maranh\~ao, Para\'iba, Pernambuco, Piau\'i, Rio Grande do Norte,
and Sergipe), the region containing seven of the ten complementa\c{c}\~ao states.
The estimate in the Northeast is 0.056 ($p = 0.015$), smaller than the full-sample
estimate, which reflects the more homogeneous treatment intensity within this high-poverty
region. Column~4 uses log per-capita education spending (with population from IBGE
municipal estimates) as the outcome; the estimate is 0.112 ($p = 0.006$), slightly
smaller than the log total spending estimate.

Column~5 provides the most important robustness check: a placebo test using log
health spending as the outcome. The estimate is $-0.001$ ($p = 0.985$)---indistinguishable
from zero. This rules out the alternative hypothesis that complementa\c{c}\~ao states
experienced a general budget expansion driven by unobserved confounders. FUNDEB was
specifically earmarked for education, and only education spending responded.

\input{tables/tab4_robustness.tex}

\section{Discussion}

\subsubsection*{Fiscal pass-through.} Our main estimate implies that FUNDEB complementa\c{c}\~ao
increased municipal education spending by approximately 12.9 percent over the 2007--2011
period in recipient states. The effect grows over time (reaching 21 percent by 2011),
suggesting that the transfers were gradually absorbed into municipal budget baselines
rather than spent as one-time windfalls. This pattern is consistent with the flypaper
literature, which documents that formula-based intergovernmental grants become embedded
in local spending patterns \citep{bailey1998}.

\subsubsection*{Composition effects.} The absence of spending reallocation toward
secondary education at the municipal level does not imply that FUNDEB failed. The
reform's architecture placed the per-student secondary transfers in a state-level fund,
redistributed to whatever level of government provides secondary instruction. In Brazil,
that is the state government, not the municipality. Our null finding on secondary
composition at the municipal level is therefore consistent with FUNDEB working as
designed: the secondary incentives flowed to state governments, which used them to
expand secondary provision, while municipalities absorbed the primary and early
childhood transfers and increased their absolute spending accordingly.

\subsubsection*{Interpretation of the spending increase.} The 12.9 percent increase in
total education spending may appear modest relative to the scale of the complementa\c{c}\~ao
transfers. Several factors may temper the effect: (i) the constitutional mandate implies
municipalities were already spending 25 percent on education, limiting scope for above-and-beyond
increases; (ii) some complementa\c{c}\~ao funds flow directly to state accounts and may
bypass municipalities entirely; and (iii) our balanced-panel restriction drops municipalities
with incomplete coverage, which may skew toward larger, better-administered units with
less fiscal flexibility.

\subsubsection*{Limitations.} The most important limitation is that treatment is assigned
at the state level, leaving only 27 effective clusters. While our results are robust to
municipality-level clustering, inference with state-level clustering involves known
over-rejection problems when the number of clusters is small \citep{cameron2008}.
Randomization inference tests and wild cluster bootstrap methods would strengthen the
evidence, but are beyond our computational resources for the current analysis.
Future work could also exploit variation in complementa\c{c}\~ao amounts across states
and over time (as the floor was adjusted annually) to identify continuous-treatment
effects.

Despite these limitations, the combination of a significant main effect, a clean
pre-trend test, a null health placebo, and consistency across clustering choices
provides credible evidence that FUNDEB's complementa\c{c}\~ao transfers successfully
increased municipal education expenditure in recipient states. Brazil's experiment in
fiscal equalization represents a case where federal intergovernmental transfers achieved
their stated goal of expanding education resources in the country's poorest states.

\label{apep_main_text_end}

\newpage

\bibliography{paper}

\newpage

\appendix

\section*{Appendix: Standardized Effect Size Table}

This appendix reports standardized effect sizes (SDE) for the main outcomes to
facilitate cross-study comparison. Classification is based on SDE magnitude only,
not statistical significance.

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
