\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Deliberation Deficit: Calendar Pressure and Legislative Process Quality in the U.S.\ Congress}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and \makeatletter @\makeatother ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The U.S.\ Constitution mandates that each Congress expires on January~3 of odd-numbered years, creating a hard deadline that compresses legislative activity in final weeks. I document a ``deliberation deficit'': laws enacted in the final 30 days are 10.5 percentage points less likely to receive a recorded roll-call vote and 2.2 percentage points less likely to go through bicameral conference reconciliation. These associations---equivalent to halving the baseline roll-call rate---are stable across 26 Congresses (1973--2025) and robust to controls for bill type, legislative age, and prior actions. A mid-session placebo shows zero effect. Roughly one-third of the raw gap is attributable to bill-lifecycle characteristics, but a 7.0 percentage-point deficit persists after conditioning on observables, suggesting that constitutional calendar pressure is associated with diminished procedural scrutiny for end-of-session legislation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} D72, H11, K00 \\
\noindent\textbf{Keywords:} legislative quality, calendar pressure, congressional procedure, roll-call votes, deliberation

\newpage

\section{Introduction}

On December 23, 2022---nine days before the 117th Congress expired---President Biden signed the Consolidated Appropriations Act, a 4,155-page omnibus package that no single legislator could plausibly have read in full. The bill passed the Senate 68--29 and the House 225--201, but scores of smaller laws enacted that same week received no recorded vote at all: they passed by voice vote or unanimous consent, with no public record of how any individual member voted. This is not unusual. Across five decades, the final weeks of each Congress produce a disproportionate share of enacted legislation under conditions of sharply diminished procedural scrutiny.

This paper documents what I call the \textit{deliberation deficit}: a systematic shift in legislative procedural form---away from recorded votes and bicameral conference committees---among laws enacted near the constitutional session deadline. The analysis exploits a fixed institutional feature: Article~XX of the Twentieth Amendment sets January~3 as the date each Congress convenes, meaning the previous Congress must complete all business before that hard deadline \citep{Ornstein2013}. This creates predictable calendar compression that intensifies as the deadline approaches.

Using the complete universe of 10,377 substantive enacted public laws from the 93rd through 118th Congresses (1973--2025), I find that laws enacted in the final 30 days before the constitutional deadline are 10.5 percentage points less likely to receive a recorded roll-call vote (from a baseline of 22.3\%, $p < 0.001$) and 2.2 percentage points less likely to undergo conference committee reconciliation ($p < 0.05$). In standardized terms, the roll-call effect corresponds to a 0.25-standard-deviation reduction. These findings are robust to controls for bill type, chamber of origin, legislative age, and the number of prior major actions, and are stable across compression windows from 7 to 90 days.

Two placebo tests validate the identifying assumption. First, applying the same ``final 30 days'' window around the session midpoint yields a precisely estimated null (coefficient $= -0.005$, SE $= 0.016$), confirming that the deliberation deficit is specific to session-end pressure rather than an artifact of legislative timing. Second, among naming and postal designation bills---trivial legislation that always passes by voice vote regardless of timing---the effect is statistically indistinguishable from zero.

This paper contributes to the literature on legislative organization and productivity. A substantial body of work examines \textit{how much} Congress produces \citep{Mayhew1991, Binder1999, Lapinski2008, Clinton2006}, but far less attention has been paid to the \textit{conditions} under which legislation is produced. \citet{Sinclair2012} documents the rise of ``unorthodox lawmaking''---suspension of rules, self-executing rules, and omnibus packaging---but treats these as strategic choices rather than consequences of calendar pressure. \citet{Krutz2001} studies omnibus legislation as a mechanism for building coalitions, while \citet{Cox2005} model agenda-setting as a function of majority party gatekeeping. My contribution is to show that a fixed constitutional constraint---the session deadline---systematically degrades the deliberative process through which laws are enacted, independent of partisan strategy.

The findings also connect to the growing literature on institutional design and democratic accountability. Recorded votes serve a dual function: they discipline representatives through electoral accountability \citep{Snyder2005} and provide information to voters about policy positions \citep{Clinton2004}. When calendar pressure shifts legislation from roll-call to voice passage, both functions are compromised. \citet{Lee2009} and \citet{Volden2014} emphasize that legislative effectiveness depends on institutional incentives; the deliberation deficit suggests that even the most fundamental institutional constraint---the constitutional calendar---creates systematic variation in how carefully laws are vetted before taking effect.

The remainder of the paper proceeds as follows. Section~2 describes the institutional setting. Section~3 presents the data. Section~4 details the empirical strategy. Section~5 reports results, and Section~6 concludes.


\section{Institutional Background}

\subsubsection*{The Constitutional Calendar} The Twentieth Amendment to the U.S.\ Constitution, ratified in 1933, sets the terms of members of Congress to begin on January~3 of odd-numbered years. This creates a fixed, exogenous endpoint for each two-year Congress: all pending legislation dies if not enacted before the new Congress convenes \citep{Oleszek2014}. Unlike parliamentary systems where the government controls the calendar, the U.S.\ congressional session has a constitutionally determined hard stop.

\subsubsection*{End-of-Session Dynamics} As the session deadline approaches, several institutional pressures compound. First, the sheer volume of legislation competing for floor time increases as committees complete their work and report bills. Second, the leadership of each chamber faces difficult scheduling trade-offs, prioritizing must-pass legislation (appropriations, debt ceiling) over other measures \citep{Cox2005}. Third, the opportunity cost of procedural deliberation rises sharply: each hour spent on a roll-call vote or conference committee is an hour unavailable for other legislation.

\subsubsection*{Passage Without Recorded Votes} Most federal legislation does not receive a recorded roll-call vote. Instead, bills may pass by voice vote (members shout ``aye'' or ``nay''), unanimous consent (no objection recorded), or suspension of the rules (in the House, requiring a two-thirds majority). These procedures are faster but provide no public record of individual members' positions \citep{Roberts2007}. Under normal conditions, the choice between recorded and unrecorded votes reflects the salience and controversy of the bill. Under calendar pressure, however, even substantive legislation may be fast-tracked through unrecorded procedures to conserve floor time.

\subsubsection*{Conference Committees} When the House and Senate pass different versions of a bill, a conference committee may be appointed to reconcile the texts. Conference committees represent a deliberative checkpoint: conferees negotiate specific provisions, and the resulting conference report must pass both chambers \citep{Longley1989}. Bypassing conference---through amendments between the chambers or informal negotiation---saves time but reduces transparency and the opportunity for structured bicameral deliberation. The decline of conference committees has been noted as a broader institutional trend \citep{Sinclair2012}, but I show that calendar pressure contributes to this decline within Congresses.


\section{Data}

I construct a bill-level dataset of all public laws enacted by the 93rd through 118th Congresses (1973--2025) using the GovTrack legislative database \citep{GovTrack2024}. GovTrack aggregates data from the Library of Congress, the Government Publishing Office, and official congressional records to provide comprehensive metadata on every bill introduced in Congress.

\subsubsection*{Sample Construction} The raw dataset includes 13,330 enacted public laws. I exclude 336 observations with missing enactment dates or negative days remaining (enacted after the constitutional deadline, reflecting data errors). The remaining 12,994 laws include 2,617 naming or postal designation bills---legislation whose sole purpose is to rename a federal building or post office. Because these bills are trivially non-deliberative by design and invariably pass by voice vote, I separate them for use as a placebo sample and focus the main analysis on the 10,377 substantive enacted laws.

\subsubsection*{Key Variables} The primary independent variable is \textit{calendar pressure}, measured as the number of days between the law's enactment date and the constitutional January~3 deadline for that Congress. The main specification uses a binary indicator for enactment within the final 30 days; robustness checks vary this threshold from 7 to 90 days and use the continuous log measure.

The primary outcome is a binary indicator for whether the law received at least one recorded roll-call vote during its passage in either chamber. Secondary outcomes include whether a conference committee was convened and whether both chambers passed the bill exclusively by voice vote (no recorded vote in either chamber).

Controls include bill type (House bill, Senate bill, joint resolution), Congress fixed effects, and the sponsor's party affiliation where available.

\input{tables/tab1_summary}

\Cref{tab:summary} presents summary statistics comparing laws enacted in the final 30 days to those enacted earlier. The raw differences are striking: 10.7\% of laws in the final 30 days received a roll-call vote, compared to 22.3\% of earlier laws. Conference committee use drops from 7.2\% to 1.6\%. Notably, laws enacted near the deadline have substantially \textit{longer} introduction-to-enactment periods (356 vs.\ 225 days), consistent with these bills having been introduced early in the session but only reaching final passage under end-of-session urgency.


\section{Empirical Strategy}

I estimate the effect of calendar pressure on legislative process quality using the following specification:
\begin{equation}\label{eq:main}
\text{RollCall}_{ic} = \beta \cdot \text{Final30}_{ic} + \alpha_c + \varepsilon_{ic}
\end{equation}
where $\text{RollCall}_{ic}$ is an indicator for whether law $i$ enacted in Congress $c$ received a recorded roll-call vote, $\text{Final30}_{ic}$ indicates enactment within 30 days of the constitutional deadline, and $\alpha_c$ are Congress fixed effects. Standard errors are clustered at the Congress level to account for within-Congress correlation in legislative procedures.

Congress fixed effects absorb all between-Congress variation in political composition, procedural norms, and macroeconomic conditions. The remaining identifying assumption is that, conditional on these fixed effects, the timing of a bill's final passage within a session is not systematically correlated with unobserved determinants of roll-call use, except through the calendar pressure mechanism. This assumption is strong: bills that arrive late may differ in salience, legislative vehicle, or partisan controversy. I therefore interpret the estimates as documenting a robust conditional association and present several tests of the leading alternative explanations.

\subsubsection*{Threats to Identification} The primary concern is selection: bills that reach the floor near the session end may differ systematically from those passed earlier. Three features of the setting mitigate this concern. First, the direction of compositional bias is likely \textit{against} finding a deliberation deficit: high-priority legislation (appropriations, major authorizations) tends to be both more salient and more likely to arrive late, predicting \textit{more} roll-call scrutiny, not less. Second, I show that the effect is absent at a placebo midpoint cutoff and absent for trivially non-deliberative naming bills. Third, the estimates are robust to controlling for legislative age (days since introduction) and the number of prior major actions, which proxy for the lifecycle and complexity of each bill. These controls help address the concern that ``stale'' long-pending legislation, rather than calendar pressure per se, drives the result.

A second concern is inference with few clusters: 26 Congresses provide a limited number of clusters for standard asymptotic inference. I address this by noting that the estimates are remarkably stable across eras and that the placebo test yields a precisely estimated null from the same cluster structure. Nonetheless, the standard errors should be interpreted with appropriate caution.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} reports the main estimates. Column~1 shows the raw difference: laws enacted in the final 30 days are 11.6 percentage points less likely to receive a roll-call vote. Adding Congress fixed effects in Column~2 slightly attenuates the estimate to 10.5 percentage points ($p < 0.001$). The effect is substantively large: against a baseline roll-call rate of 22.3\% for laws enacted earlier, a 10.5 percentage-point reduction represents a 47\% decline. Column~3 adds controls for chamber of origin and bill type; the coefficient remains stable at 10.1 percentage points. Column~4 replaces the binary indicator with the continuous log of days remaining: each log-unit increase in available time is associated with a 4.1 percentage-point increase in roll-call probability.

Column~5 turns to conference committees. Laws enacted under calendar pressure are 2.2 percentage points less likely to undergo conference reconciliation ($p < 0.05$), a 30\% reduction from the 7.2\% baseline. Column~6 examines voice-only passage and finds a 5.8 percentage-point increase ($p < 0.05$) after absorbing Congress fixed effects. The raw means are similar across groups (48.5\% vs.\ 50.4\%), suggesting that cross-Congress variation in voice-vote norms obscures a within-Congress shift: in a given Congress, end-of-session laws are relatively more likely to pass without any recorded vote.

\subsection{Compression Gradient}

\input{tables/tab3_gradient}

If the deliberation deficit reflects genuine institutional pressure rather than compositional differences between early and late legislation, the effect should be present across multiple compression windows. \Cref{tab:gradient} reports the roll-call coefficient for windows ranging from 7 to 90 days. The effect is remarkably stable: $-9.6$ percentage points at 7 days, $-10.6$ at 14, $-10.5$ at 30, $-8.2$ at 60, and $-10.0$ at 90. All are statistically significant at the 0.1\% level. This stability suggests a discrete compression effect---the deliberation deficit switches on as the session end enters the legislative horizon---rather than a gradual decline in scrutiny.

\subsection{Placebo Tests and Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robust} presents five robustness checks. Column~1 reports the key placebo test: applying the ``final 30 days'' indicator to the session midpoint rather than the session end. The coefficient is $-0.005$ with a standard error of 0.016---a precisely estimated null. This rules out the possibility that the main result reflects a mechanical feature of legislative timing or a secular trend within sessions.

Column~2 uses naming and postal designation bills as a placebo sample. These bills are procedurally trivial and pass by voice vote regardless of timing. The estimate ($-0.028$, $p > 0.05$) confirms that calendar pressure affects the deliberative process only for legislation where the choice between recorded and unrecorded voting is substantively meaningful.

Columns~3 and 4 split the sample by chamber of origin. The effect is present in both: 11.8 percentage points for House-origin bills and 8.4 for Senate-origin bills, both significant at the 0.1\% level. The larger House effect is consistent with the House's more rule-bound procedures offering more scope for procedural shortcuts under time pressure \citep{Oleszek2014}. Column~5 estimates a logit specification, yielding a coefficient of $-0.83$ log-odds ($p < 0.001$), confirming that the linear probability model provides a good approximation.

\subsubsection*{Richer Controls} A further concern is that bills enacted late are ``stale''---introduced early but lingering in committee---and differ from earlier-passed bills along dimensions correlated with procedural form. I re-estimate the main specification adding controls for log deliberation days (days from introduction to enactment), the number of prior major legislative actions, and bill type. The coefficient on Final~30~Days attenuates from $-0.105$ to $-0.070$ ($p < 0.001$). Roughly one-third of the raw association is thus attributable to bill lifecycle characteristics, but a substantial and statistically significant deliberation deficit remains after conditioning on these observables.

\subsubsection*{Era Heterogeneity} The effect is stable across congressional eras: $-12.9$ percentage points in the pre-reform period (93rd--98th), $-9.6$ during rising polarization (104th--108th), $-10.4$ during gridlock (109th--113th), and $-8.5$ in the modern era (114th--118th). The deliberation deficit is not an artifact of any particular political environment.


\section{Conclusion}

The U.S.\ Constitution creates a fixed deadline that every Congress must face: January~3, when the old Congress expires and the new one convenes. This paper documents that this deadline is associated with a measurable shift in how laws are made. Laws enacted in the final weeks of each Congress are substantially less likely to receive recorded votes or undergo bicameral conference reconciliation---procedural features that provide transparency and individual accountability in democratic lawmaking.

The magnitude is notable. The roll-call rate among end-of-session legislation is roughly half the rate observed earlier in the session, a pattern that has persisted for over fifty years, across unified and divided government, and in both chambers. Placebo tests and within-Congress comparisons with richer controls suggest this is not simply an artifact of bill-type composition, though I cannot rule out all sources of within-Congress selection.

Two important limitations deserve emphasis. First, the procedural outcomes measured here---recorded votes and conference committees---are imperfect proxies for deliberative quality. Unanimous consent and voice votes may reflect genuine consensus rather than insufficient scrutiny. Whether the deliberation deficit produces \textit{worse laws} in a substantive sense---more errors, more subsequent amendments, more litigation---remains an open question for future work. Second, inference with 26 Congress-level clusters warrants caution; more precise estimation would require either longer time series or alternative design elements.

These findings nonetheless raise a design question worth further study: does the fixed constitutional deadline impose an institutional cost on legislative process that could be mitigated through scheduling reforms, mandatory recorded-vote requirements for end-of-session legislation, or changes to lame-duck procedural rules?

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} \makeatletter @\makeatother ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
