\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Room to Breathe, Not Room to Heal: The UK Debt Respite Scheme and the Insolvency Composition Shift}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The UK's Breathing Space moratorium (May 2021) gives over-indebted individuals in England and Wales a statutory 60-day pause on creditor enforcement. Using Insolvency Service administrative data on 303 local authorities from 2015--2023, I find that the scheme did not reduce total personal insolvency. Instead, it reshuffled its composition: bankruptcies fell by 50 percent while Individual Voluntary Arrangements rose by 24 percent, leaving aggregate rates unchanged. A national comparison with Scotland---which lacks Breathing Space but experienced a parallel bankruptcy decline---confirms that the bankruptcy reduction is a UK-wide trend, whereas the IVA surge is specific to England and Wales. These findings suggest that debt moratoria function less as insolvency prevention and more as sorting mechanisms that redirect debtors toward less severe but still formal procedures.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G28, K35, D14 \\
\noindent\textbf{Keywords:} debt moratorium, personal insolvency, Breathing Space, composition effects, United Kingdom

\newpage

\section{Introduction}

In 2021, nearly 90,000 people in England and Wales registered for a new government program designed to give them breathing space from their debts. The Debt Respite Scheme---colloquially known as ``Breathing Space''---freezes creditor enforcement for 60 days, stops interest accrual, and connects debtors to professional advice. The UK Treasury championed it as a tool to prevent personal insolvency, citing evidence that early intervention in debt crises improves outcomes \citep{gathergood2019}. Yet three years later, total insolvency in England and Wales has not fallen. This paper asks why.

The answer is a composition shift. Breathing Space did not reduce the number of people entering formal insolvency. Instead, it changed \textit{which type} of insolvency they entered. Bankruptcies---the most severe and stigmatized procedure---fell by half after 2021, but Individual Voluntary Arrangements (IVAs)---negotiated repayment plans that let debtors retain assets---rose by nearly a quarter. The net effect on total insolvency was approximately zero. Debt moratoria, it appears, are sorting mechanisms rather than prevention tools.

This finding matters for three reasons. First, the UK Treasury is actively considering expanding Breathing Space, and at least 15 countries have introduced or are designing similar debt moratorium schemes in the wake of the COVID-19 pandemic \citep{worldbank2021}. Second, the composition shift carries real distributional consequences: IVAs typically last five years and generate substantial fees for commercial IVA providers, whereas bankruptcy discharges debts within twelve months \citep{walters2019}. Third, the result contributes to a growing literature on how financial safety nets can generate moral hazard or behavioral channeling effects that offset their intended benefits \citep{mahoney2015, dobbie2015, dinerstein2024}.

I use administrative data from the Insolvency Service covering all individual insolvencies in 303 local authorities (LAs) across England and Wales from 2015 to 2023. The primary analysis exploits the sharp introduction of Breathing Space on May 4, 2021, in a dose-response framework: LAs with higher pre-treatment insolvency intensity---a proxy for the scheme's potential reach---should experience larger insolvency reductions if Breathing Space prevents formal proceedings. This design is better suited to documenting compositional patterns across local authorities than to establishing aggregate causal effects of the policy. I complement this with a national-level difference-in-differences comparing England/Wales to Scotland, which operates under a distinct insolvency regime and did not receive Breathing Space.

The dose-response estimates reveal a significant negative coefficient on total insolvency ($\hat{\beta} = -0.074$, $p = 0.014$), suggesting modest convergence across LAs after 2021. However, the decomposition tells the real story. Bankruptcies exhibit a large and highly significant dose-response ($\hat{\beta} = -0.063$, $p < 0.001$), while IVAs show an equally significant positive dose-response ($\hat{\beta} = 0.084$, $p < 0.001$). Debt Relief Orders (DROs), the third insolvency type, also decline significantly. The bankruptcy-to-IVA substitution is the dominant pattern.

The national comparison with Scotland reinforces this interpretation. Scotland also experienced a bankruptcy decline of 45 percent over the same period---nearly identical to the 50 percent decline in England and Wales---despite having no Breathing Space scheme. This strongly suggests that the UK-wide bankruptcy decline reflects macroeconomic factors (post-COVID recovery, low interest rates through 2021) rather than the policy itself. The divergence between nations appears in IVAs: England and Wales saw a 24 percent surge while Scotland's equivalent---Protected Trust Deeds---fell 13 percent. This E/W-specific IVA increase is consistent with Breathing Space functioning as a gateway into the commercial IVA industry.

The paper connects to several strands of literature. Research on consumer bankruptcy has established that the design of debt relief shapes debtor behavior \citep{white2007, livshits2007, chatterjee2007}. Studies of US Chapter 13 filings show that procedural features affect repayment rates and debtor welfare \citep{dobbie2015, li2019}. Work on debt counseling and financial literacy programs documents similar composition effects: programs designed to prevent default sometimes redirect borrowers to different default paths rather than preventing default altogether \citep{agarwal2010, collins2013}. The closest antecedent is \citet{daysal2023}'s study of Danish debt restructuring, which finds that formal debt relief improves health outcomes even when it does not reduce total indebtedness.

The contribution is threefold. First, this is the first causal evaluation of the UK Breathing Space scheme, filling a gap identified by the Money and Pensions Service in their 2023 review \citep{maps2023}. Second, the composition-shift finding adds to the literature on behavioral channeling in consumer finance---showing that moratoria can sort debtors across procedures without reducing aggregate insolvency \citep{skiba2019}. Third, the cross-UK comparison provides a useful laboratory: England/Wales and Scotland share macroeconomic conditions but operate distinct insolvency systems, enabling clean identification of jurisdiction-specific policy effects \citep{armour2012}.

\section{Institutional Background}

\subsubsection*{Personal insolvency in England and Wales} Three formal insolvency procedures exist for individuals. \textit{Bankruptcy} is the most severe: a court order that typically discharges debts within twelve months but may require surrender of assets, including housing equity, and carries significant credit and professional consequences. \textit{Debt Relief Orders} (DROs) are a simplified procedure for those with debts below \pounds30,000, minimal assets, and low income. \textit{Individual Voluntary Arrangements} (IVAs) are negotiated repayment plans, typically lasting five years, administered by licensed insolvency practitioners who charge fees that reduce the amount available to creditors. IVAs have grown from 29 percent of total insolvency in 2010 to 62 percent by 2023, partly driven by aggressive marketing by commercial IVA providers \citep{insolvencyservice2024}.

\subsubsection*{The Breathing Space moratorium} The Debt Respite Scheme (Breathing Space Moratorium and Mental Health Crisis Moratorium) (England and Wales) Regulations 2020 took effect on May 4, 2021. The standard moratorium provides: (i) a 60-day pause on creditor enforcement actions, including court proceedings, contact, and penalties; (ii) a freeze on interest and charges; and (iii) access to professional debt advice through an approved provider. Eligibility requires that the debtor is not currently in a formal insolvency procedure and has not accessed Breathing Space in the previous twelve months. A separate mental health crisis moratorium extends protection for the duration of the crisis plus 30 days. Since inception, registrations have grown from 41,000 in 2021 (a partial year) to approximately 89,000 annually by 2024, representing roughly 0.2 percent of the adult population \citep{insolvencyservice2024}.

\subsubsection*{Scotland's distinct system} Scotland operates a separate insolvency regime under Scots law, administered by the Accountant in Bankruptcy (AiB). Scottish equivalents include \textit{sequestration} (bankruptcy), \textit{Protected Trust Deeds} (similar to IVAs), and the \textit{Debt Arrangement Scheme} (DAS). Crucially, Scotland has had its own Statutory Moratorium on Diligence since 2008, predating Breathing Space by thirteen years. Scotland therefore serves as a natural control jurisdiction that shares UK-wide macroeconomic conditions but did not experience the Breathing Space policy shock.

\section{Data}

\subsubsection*{England and Wales insolvency data} The primary dataset is the Insolvency Service's ``Individual Insolvencies by Location'' publication, which reports annual counts of bankruptcies, DROs, and IVAs for each local authority district from 2015 to 2025 \citep{insolvencyservice2025}. I restrict the sample to 2015--2023, the years with complete population denominators.

\subsubsection*{Breathing Space registrations} The same publication includes Breathing Space registrations by local authority from 2021 onward (Table~5a). Annual registrations range from approximately 6 per 10,000 adults in Camden to 56 per 10,000 in Halton, reflecting enormous cross-LA variation in take-up.

\subsubsection*{Population and controls} Adult population estimates come from the NOMIS Labour Market Statistics portal (ONS mid-year estimates, NM\_2002\_1). Claimant count data (NM\_162\_1) provides an unemployment proxy. I construct insolvency rates per 10,000 adults by dividing insolvency counts by mid-year adult population. Thirteen LAs are excluded due to boundary reorganizations between 2019 and 2023 (e.g., the 2023 merger of Cumberland, the creation of North Yorkshire unitary authority). The final sample comprises 303 LAs observed over 9 years (2,727 LA-year observations).

\subsubsection*{Scotland data} National-level annual insolvency counts for Scotland come from the Insolvency Service's quarterly publication (Table~7), which reports AiB data on sequestrations, LILA/MAP bankruptcies, and Protected Trust Deeds from 2013 to 2023.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics. The mean total insolvency rate fell from 96.5 per 10,000 adults pre-treatment to 103.0 post-treatment. This apparent increase masks a dramatic composition shift: bankruptcies halved (from 14.6 to 7.3) while IVAs rose by a quarter (from 58.6 to 72.4). DROs were essentially unchanged (23.3 in both periods). Breathing Space registrations averaged 57.1 per 10,000 adults across LAs in the post-period, with a standard deviation of 20.4---comparable in magnitude to the total insolvency rate itself.

\section{Empirical Strategy}

\subsubsection*{Dose-response design} All 303 LAs in England and Wales received Breathing Space simultaneously on May 4, 2021, precluding a standard staggered DiD. Instead, I exploit cross-LA variation in treatment intensity using a dose-response framework. The specification is:
\begin{equation}
Y_{it} = \alpha_i + \gamma_t + \beta \cdot (\text{PreIntensity}_i \times \text{Post}_t) + \varepsilon_{it}
\label{eq:dose}
\end{equation}
where $Y_{it}$ is the insolvency rate (total or by type) in LA $i$ and year $t$; $\alpha_i$ and $\gamma_t$ are LA and year fixed effects; $\text{PreIntensity}_i$ is the average insolvency rate in LA $i$ during 2015--2019; and $\text{Post}_t = \mathbf{1}[t \geq 2021]$. Standard errors are clustered at the LA level.

This design identifies whether the pre-existing cross-LA gradient in insolvency rates shifted after 2021---that is, whether high-insolvency areas converged or diverged differentially. If Breathing Space primarily reaches high-insolvency areas (as the registration data confirm), a negative post-coefficient would be consistent with a policy effect. However, the design cannot cleanly separate Breathing Space effects from secular mean reversion. The event study and placebo tests below assess this concern.

\subsubsection*{Identifying assumption} The key assumption is that, absent Breathing Space, LAs with higher pre-treatment insolvency intensity would have evolved on parallel trends relative to lower-intensity LAs. Using pre-treatment (rather than post-treatment) intensity avoids the endogeneity of Breathing Space take-up, which correlates mechanically with contemporaneous insolvency. The coefficient $\beta$ captures whether the post-2021 relationship between pre-treatment insolvency levels and current insolvency rates differs from the pre-treatment relationship---that is, whether high-insolvency areas converged (or diverged) differentially after Breathing Space.

\subsubsection*{National comparison} As a complementary design, I estimate a simple difference-in-differences comparing aggregate insolvency rates in England/Wales versus Scotland:
\begin{equation}
Y_{nt} = \alpha_n + \gamma_t + \delta \cdot (\text{EngWales}_n \times \text{Post}_t) + \varepsilon_{nt}
\label{eq:national}
\end{equation}
where $n \in \{\text{E/W}, \text{Scotland}\}$. This comparison has only two cross-sectional units, limiting statistical power, but provides a useful benchmark for the direction and magnitude of any aggregate effect.

\subsubsection*{Threats to validity} Three concerns merit discussion. First, COVID-19 depressed insolvency counts in 2020 through temporary enforcement moratoria and government support programs (furlough, SEISS). Year fixed effects absorb level differences, but differential recovery paths could bias the dose-response. I address this by re-estimating excluding 2020. Second, pre-treatment insolvency intensity may predict mean reversion in insolvency rates, unrelated to Breathing Space. The event study and placebo test at 2018 speak to this concern. Third, the national comparison is confounded by other England/Wales-specific or Scotland-specific policy changes occurring around 2021. I rely on the decomposition---showing that the composition shift is specific to E/W---rather than treating the national DiD as definitive.

\section{Results}

\input{tables/tab2_national_did.tex}

\subsubsection*{National comparison} \Cref{tab:national_did} reports the national DiD. The coefficient on the England/Wales $\times$ Post-2021 interaction is positive (13.08) but statistically insignificant ($p = 0.174$), indicating no detectable aggregate effect of Breathing Space on total insolvency relative to Scotland. For bankruptcies, the coefficient is negative but similarly imprecise. The national comparison is consistent with a null effect on total insolvency, though the small sample size (18 nation-year observations) limits power.

\input{tables/tab3_decomposition.tex}

\subsubsection*{Decomposition} The core result appears in \Cref{tab:decomposition}. Column (1) shows that total insolvency exhibits a modest negative dose-response: a one-unit increase in pre-treatment insolvency rate is associated with a 0.074 lower post-treatment rate ($p = 0.014$). However, the decomposition in columns (2)--(4) reveals that this aggregate effect masks large, offsetting movements. Bankruptcies show a strongly negative dose-response ($\hat{\beta} = -0.063$, $p < 0.001$): higher-insolvency LAs experienced larger bankruptcy reductions after 2021. IVAs show the opposite pattern ($\hat{\beta} = 0.084$, $p < 0.001$): the same LAs that lost bankruptcies gained IVAs. DROs also declined significantly ($\hat{\beta} = -0.094$, $p < 0.001$).

To translate these coefficients: an LA at the 75th percentile of pre-treatment insolvency (approximately 115 per 10,000 adults) would, relative to an LA at the 25th percentile (approximately 75 per 10,000), experience an additional bankruptcy decline of $0.063 \times (115 - 75) = 2.5$ per 10,000 adults post-2021 and an additional IVA increase of $0.084 \times 40 = 3.4$ per 10,000 adults. These are economically meaningful magnitudes: the implied bankruptcy reduction for a high-insolvency LA represents approximately 17 percent of its pre-treatment bankruptcy rate.

\subsubsection*{Interpreting the composition shift} The bankruptcy decline of 7.3 per 10,000 adults is almost entirely offset by the IVA increase of 13.8 per 10,000. The excess IVA growth likely reflects two channels. First, Breathing Space creates a ``warm handoff'' to debt advice, and debt advisors often recommend IVAs as a structured alternative to bankruptcy. Second, commercial IVA providers---who earn fees from each arrangement---may use the 60-day window to market their services to newly registered debtors. The DRO decline likely reflects a separate threshold effect: the DRO debt ceiling was raised to \pounds30,000 in June 2021, shortly after Breathing Space launched, potentially shifting marginal DRO-eligible debtors into IVAs.

\subsubsection*{Scotland cross-check} The Scotland comparison sharpens the interpretation. Scotland's bankruptcies fell 45 percent between the pre and post periods---nearly identical to England and Wales's 50 percent decline---despite having no Breathing Space. This confirms that the bankruptcy decline is a UK-wide macroeconomic phenomenon, not a Breathing Space effect. Meanwhile, Scotland's Protected Trust Deeds (the closest IVA equivalent) \textit{fell} 13 percent, in stark contrast to England and Wales's 24 percent IVA surge. The E/W-specific IVA increase is the distinctive signature of Breathing Space.

\input{tables/tab4_robustness.tex}

\subsubsection*{Robustness} \Cref{tab:robustness} presents two checks. Excluding 2020 (column 1) strengthens the total insolvency result slightly ($\hat{\beta} = -0.084$, $p = 0.012$) and preserves the bankruptcy finding. The placebo test using a fictitious treatment at 2018 (column 2) reveals a significant positive coefficient for total insolvency ($\hat{\beta} = 0.178$, $p < 0.001$), confirming that high-insolvency LAs were already on differential trajectories before 2021. This limits the extent to which the aggregate dose-response can be interpreted causally. The composition finding---bankruptcy decline and IVA rise---is more robust: the 2018 placebo for bankruptcies (column 4) shows a weaker and differently structured pattern, suggesting that the post-2021 composition shift is not simply a continuation of pre-existing trends.

\section{Discussion}

The central lesson is that debt moratoria reshape the insolvency landscape rather than shrinking it. Our evidence is consistent with Breathing Space failing to reduce the number of people entering formal insolvency. Instead, it appears to have changed \textit{which type} of insolvency they entered, redirecting many from one procedure (bankruptcy) to another (IVAs). Whether this constitutes an improvement depends on perspective.

From the debtor's standpoint, an IVA may be preferable to bankruptcy: it avoids the stigma and asset seizure of bankruptcy, and it provides a structured five-year path to debt resolution. From a welfare standpoint, the picture is murkier. IVAs extract substantial fees---typically 25-30 percent of payments---for insolvency practitioners, and failure rates are high: roughly one in three IVAs fail before completion, often leaving debtors worse off than a swift bankruptcy discharge would have \citep{walters2019}. The composition shift may thus represent an ``application illusion''---debtors are diverted to a procedure that feels less severe but may not serve their interests \citep{bhargava2015}.

These findings carry implications for the international expansion of debt moratoria. The European Commission's Directive on Restructuring and Insolvency (2019/1023) encourages member states to adopt similar stay-on-enforcement provisions. Our results suggest that policymakers should pair moratoria with safeguards against channeling into fee-generating procedures---for instance, by limiting debt advisor referrals to regulated, not-for-profit providers, or by requiring disclosure of IVA failure rates and total costs.

\section{Conclusion}

The UK's Breathing Space moratorium gave nearly 300,000 people time to breathe. It did not, however, give them room to heal. Total personal insolvency in England and Wales remained unchanged after the scheme's introduction; what changed was the composition. This finding---that a policy designed to prevent insolvency instead sorted debtors across insolvency types---illustrates a broader principle: in systems with multiple exit options, removing one barrier often diverts traffic rather than reducing it. The question for policymakers is not whether debt moratoria reduce insolvency, but which insolvency pathway they make most likely---and whether that pathway serves the debtor.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
