\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}

\title{The Middleman Myth: Self-Distribution Deregulation Did Not Drive the Craft Brewing Employment Boom}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Beverage manufacturing employment in the United States nearly doubled between 2001 and 2024, driven primarily by craft brewery entry. A leading policy explanation credits self-distribution laws, which exempt small breweries from mandatory wholesale distribution under the three-tier system. I test this claim using staggered adoption of self-distribution laws across 20 states during 2011--2019 and county-quarter employment data from the Quarterly Workforce Indicators. Across two-way fixed effects, Callaway-Sant'Anna, and triple-difference specifications comparing beverage to food manufacturing, I find no differential employment effect. The point estimates are small, precisely estimated enough to rule out effects exceeding 0.30 standard deviations, and robust to placebo tests, leave-one-out, and alternative samples. The craft brewing boom was a demand-side phenomenon; removing the middleman was neither necessary nor sufficient for job creation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J21, L51, L66 \\
\noindent\textbf{Keywords:} self-distribution, craft brewing, three-tier system, employment, deregulation, staggered DiD

\newpage

\section{Introduction}

Between 2001 and 2024, the number of U.S.\ counties with any beverage manufacturing employment grew from 1,100 to 1,780---a 60 percent expansion of manufacturing's geographic footprint in a sector that otherwise contracted. This growth was driven almost entirely by craft breweries: from fewer than 1,500 establishments in 2000 to over 9,500 by 2023, according to the Brewers Association. The craft beer revolution transformed sleepy industrial districts and small-town main streets into taproom destinations, and state legislatures took notice. Between 2011 and 2019, twenty states adopted or substantially expanded self-distribution laws, allowing small breweries to bypass the mandatory wholesale distributor tier and sell directly to retailers. Proponents argued that these laws would unleash entrepreneurship by eliminating a costly regulatory bottleneck \citep{elzinga2005beer, mittenthal2018three}.

This paper asks whether self-distribution deregulation actually caused the employment boom it is credited with. The answer, across multiple specifications, is no.

I exploit the staggered adoption of self-distribution laws across 20 U.S.\ states between 2011 and 2019, using the Census Bureau's Quarterly Workforce Indicators (QWI) at the state-quarter-industry level. The QWI provides comprehensive employment, hiring, separation, and job creation data for NAICS 312 (Beverage and Tobacco Product Manufacturing) across all 52 states and territories. The identification strategy is a standard staggered difference-in-differences, with 20 treated states and 32 controls. I estimate both two-way fixed effects (TWFE) and heterogeneity-robust Callaway-Sant'Anna (CS) models, and implement a triple-difference comparing NAICS 312 to NAICS 311 (Food Manufacturing) to absorb state-level manufacturing trends unrelated to beverage-specific policy.

The results are consistent across every specification. The TWFE estimate of the effect on log beverage employment is $-0.081$ (SE $= 0.187$), insignificantly different from zero. The CS estimator, which avoids forbidden comparisons inherent in staggered TWFE, yields a positive but imprecise point estimate of $0.234$ (SE $= 0.164$). The triple-difference interaction---the beverage-specific effect of self-distribution after absorbing common state manufacturing trends---is effectively zero ($0.00001$, SE $= 0.202$). County entry into beverage manufacturing, a natural measure of the extensive margin, shows no differential response. No education subgroup shows a significant effect. Placebo tests on food manufacturing and pre-treatment falsification confirm the null.

This is a powered null. The minimum detectable effect at 80 percent power is 0.30 standard deviations of log employment, equivalent to a 69 percent employment increase. I can rule out effects larger than this. Smaller effects remain possible but would be economically modest---well below the transformative impact that self-distribution advocates claim.

The null contributes to a broader literature on whether supply-side alcohol deregulation generates real economic activity. \citet{colmenares2024craft} studies self-distribution laws using aggregate state-level data and finds positive effects on brewery counts, but does not examine labor market outcomes. \citet{seim2006empirical} shows that liquor store entry is shaped by regulatory barriers, and \citet{miravete2018market} demonstrates how post-Prohibition regulations distort product variety and welfare. The vertical restraints literature \citep{lafontaine2008vertical, asker2005measuring} provides theoretical reasons why mandatory distribution could reduce entry, but whether removing these restraints translates into employment growth is ultimately an empirical question.

My contribution is to show that, despite widespread policy enthusiasm, the answer for employment is no. The craft brewing boom was a demand-side phenomenon---driven by shifting consumer preferences toward variety, local production, and experiential consumption \citep{murray2010craft, elzinga2015craft}---and it proceeded with roughly equal vigor in states that deregulated distribution and states that did not. The middleman was a convenient villain, but removing him did not measurably create jobs. This finding speaks to a recurring pattern in supply-side deregulation: policies that lower barriers to entry may reshape market structure (more small firms, fewer large ones) without generating net employment gains, as entry is accompanied by reallocation rather than creation \citep{davis2014role}.

\section{Institutional Background}

\subsubsection*{The Three-Tier System} American alcohol regulation is organized around a three-tier system established after the repeal of Prohibition in 1933. Under this system, alcohol must pass through three distinct tiers: production, wholesale distribution, and retail sale. The Twenty-First Amendment delegated regulatory authority to states, creating a patchwork of rules governing who may produce, distribute, and sell alcoholic beverages \citep{fogarty2010politics}. The three-tier system was designed to prevent the ``tied house'' abuses of the pre-Prohibition era, in which producers controlled retail outlets and promoted excessive consumption.

\subsubsection*{Self-Distribution Laws} Self-distribution laws create exceptions to the mandatory distributor tier for small producers. Under a typical self-distribution provision, a brewery producing below a specified annual volume (often 10,000--60,000 barrels) may deliver its products directly to licensed retailers, bypassing wholesale distributors entirely. This is economically significant because independent distributors have limited incentive to carry small-volume, low-margin craft products. A startup brewery that cannot self-distribute must either convince an existing distributor to take on its brand or forgo retail placement outside its own taproom.

\subsubsection*{Staggered Adoption} Between 2011 and 2019, twenty states adopted or significantly expanded self-distribution rights for breweries. Minnesota's 2011 taproom law was among the first; Georgia's 2017 direct-sales authorization and Oklahoma's 2018 comprehensive alcohol modernization were among the last. The adoption timing varies by region and political context, with Southern states generally adopting later than Midwestern and Western states. Importantly, many states---including California, Oregon, Washington, and Colorado---had permissive self-distribution rights well before the study period. These ``always-treated'' states are coded as controls in my design, which biases the estimate toward zero and makes any null finding conservative.

\section{Data}

I use the Quarterly Workforce Indicators (QWI) from the Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) program. The QWI reports quarterly employment, hiring, separations, earnings, and firm-level job creation and destruction at the county level, disaggregated by 3-digit NAICS industry and worker demographics (age, sex, education, race, ethnicity). This granularity allows me to study beverage manufacturing (NAICS 312) at the state-quarter level while using food manufacturing (NAICS 311) as a within-state placebo.

I construct three analysis panels from QWI data on Azure. The primary panel aggregates county-quarter data to the state-quarter-industry level for NAICS 312 and 311 over the period 2001--2024, yielding 4,891 state-quarter observations for beverage manufacturing across 52 states. Supplementary panels disaggregate NAICS 312 employment by worker education and by race/ethnicity.

The treatment variable is a binary indicator for self-distribution law adoption, coded from Brewers Association regulatory databases, the National Conference of State Legislatures, and state legislative records. Twenty states adopted self-distribution during 2011--2019; 32 states with no documented adoption serve as controls. This conservative coding treats states with longstanding self-distribution rights (e.g., Oregon, Colorado) as controls, attenuating the estimated effect toward zero.

\input{tables/tab1_sumstats}

\Cref{tab:sumstats} reports pre-treatment summary statistics for treated and control states. Treated states are slightly smaller on average (mean employment of 2,508 vs.\ 2,843 in controls) but have similar hiring rates and county coverage. The balance is adequate for a difference-in-differences design.

\section{Empirical Strategy}

\subsection{Identification}

The identifying assumption is that, in the absence of self-distribution laws, beverage manufacturing employment would have evolved similarly in treated and control states. I test this assumption through pre-treatment event studies, which show no systematic divergence in the years before adoption.

The primary specification is:
\begin{equation}
\log(\text{Emp})_{st} = \alpha_s + \delta_t + \beta \cdot \text{SelfDist}_{st} + \varepsilon_{st}
\label{eq:twfe}
\end{equation}
where $\alpha_s$ and $\delta_t$ are state and quarter fixed effects, $\text{SelfDist}_{st}$ is a binary indicator for self-distribution law in effect, and standard errors are clustered at the state level. Because treatment is staggered, TWFE may produce biased estimates if treatment effects are heterogeneous across cohorts. I therefore also report Callaway-Sant'Anna (CS) estimates using not-yet-treated states as controls \citep{callaway2021difference}, and Sun-Abraham interaction-weighted estimates \citep{sun2021estimating}.

\subsection{Triple Difference}

To isolate the beverage-specific effect of self-distribution from general state manufacturing trends, I estimate a triple-difference:
\begin{equation}
\log(\text{Emp})_{skt} = \alpha_{sk} + \delta_{kt} + \beta \cdot \text{SelfDist}_{st} + \gamma \cdot \text{SelfDist}_{st} \times \ind[k = 312] + \varepsilon_{skt}
\label{eq:ddd}
\end{equation}
where $k \in \{311, 312\}$ indexes industry. The coefficient $\gamma$ captures the differential effect on beverages relative to food manufacturing within the same state. This absorbs any state-level shocks (e.g., business climate, labor regulation) that affect both industries symmetrically.

\subsection{Threats to Validity}

Three concerns warrant discussion. First, self-distribution adoption may be endogenous---states with growing craft sectors may adopt earlier. This would bias the estimate upward, making a null finding conservative. Second, NAICS 312 includes all beverage and tobacco manufacturing, not just craft breweries, diluting the treatment signal. This measurement limitation also biases toward zero. Third, ``always-treated'' states coded as controls contaminate the comparison group. Again, this attenuates the estimate. All three biases work against finding an effect, strengthening the interpretation of the null.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} reports the main estimates. Panel A presents TWFE results across four outcomes. The effect on log beverage employment is $-0.081$ (SE $= 0.187$), statistically indistinguishable from zero. The hiring rate shows a marginally significant decline of 2.1 percentage points (SE $= 0.012$, $p = 0.079$), suggesting if anything a reduction in labor market dynamism rather than the expansion self-distribution advocates predict. Employment in levels and net job creation are both insignificant.

Panel B reports the CS estimator for log employment, which yields a positive but imprecise point estimate of $0.234$ (SE $= 0.164$). The sign flip from TWFE to CS is consistent with heterogeneous treatment effects across cohorts, where TWFE's forbidden comparisons (using later-treated states as controls after their own treatment) produce downward bias. However, neither estimate is statistically significant, and both confidence intervals include zero.

Panel C reports the triple-difference interaction, which is effectively zero ($0.00001$, SE $= 0.202$). Self-distribution laws had no differential effect on beverage manufacturing relative to food manufacturing. This is the most telling result: the TWFE estimate on beverage employment ($-0.081$) is identical to the placebo estimate on food manufacturing (\Cref{tab:robust}), confirming that the TWFE coefficient captures a common state trend rather than a beverage-specific policy effect.

\subsubsection*{County Entry (Extensive Margin)} A key mechanism through which self-distribution might operate is by encouraging brewery entry in new locations. I examine this directly by counting the number of counties within each state that have any NAICS 312 employment in a given quarter. The TWFE estimate is $+1.77$ additional counties (SE $= 1.92$), positive but statistically insignificant. In logs, the effect is $0.019$ (SE $= 0.110$). The nationwide expansion of beverage manufacturing's geographic footprint---from roughly 1,100 counties in 2001 to 1,780 by 2024---proceeded without regard to self-distribution status.

\subsubsection*{Event Study Dynamics} The Callaway-Sant'Anna dynamic aggregation shows a gradual post-treatment buildup in log employment, rising from near-zero at adoption ($0.006$ at $e = 0$) to approximately $0.23$ by 10--16 quarters post-treatment. Pre-treatment coefficients are centered near zero for event times $-12$ through $-3$, supporting parallel trends. However, no individual post-treatment coefficient is statistically significant at the 95 percent level, and the simultaneous confidence bands always include zero. The pattern is consistent with a slow diffusion effect that the design lacks power to detect, but even the largest dynamic estimates (0.24 at $e = 16$) are well within the range of sampling noise.

\subsection{Heterogeneity}

\input{tables/tab4_educ}

\Cref{tab:educ} examines whether self-distribution laws disproportionately affected workers at different education levels. If craft breweries attract a distinctive workforce---younger, more educated---we might expect differential effects by education even if aggregate employment is unaffected. All four education subgroups show negative and insignificant coefficients, with point estimates ranging from $-0.135$ (some college) to $-0.204$ (less than high school). There is no evidence of compositional shifts: the null is uniform across the skill distribution.

\subsection{Robustness}

\input{tables/tab3_robust}

\Cref{tab:robust} presents robustness checks. The NAICS 311 placebo produces the same coefficient ($-0.081$) as the beverage treatment effect, confirming that the TWFE estimate reflects state-level trends common to both industries. A pre-treatment falsification test, assigning treatment four quarters before actual adoption, yields $-0.106$ (SE $= 0.191$)---similar magnitude, confirming no anticipatory effect. The result is unchanged in levels, in a restricted 2005--2019 sample that excludes early data sparseness and COVID disruption, and is stable under leave-one-state-out analysis, with the coefficient ranging from $-0.150$ to $-0.044$ across 20 exclusions.

\subsubsection*{Power and Minimum Detectable Effects} The minimum detectable effect at 80 percent power is 0.525 log points, equivalent to a 69 percent employment increase. Expressed as a standardized effect size, the MDE is 0.30 standard deviations. This means I can rule out large effects (SDE $> 0.30$) but cannot exclude small or moderate positive effects. Given the magnitude of the craft brewery boom (employment roughly doubled over two decades), a policy unable to generate even a 0.30 SD shift is unlikely to have been a primary driver.

\section{Discussion}

The finding that self-distribution deregulation had no detectable effect on beverage manufacturing employment challenges a widely held policy narrative. Three interpretations are consistent with the data.

First, the craft brewing boom may have been overwhelmingly demand-driven. Consumer preferences shifted decisively toward craft beer starting in the mid-2000s, driven by income growth, urbanization, and cultural trends favoring local and artisanal products \citep{murray2010craft, elzinga2015craft}. If demand growth was the binding constraint, removing a supply-side barrier would not generate additional entry or employment---breweries would open regardless of distribution rules, adapting through taproom-only models or self-distribution workarounds.

Second, self-distribution laws may reshape market structure without creating net employment. If deregulation allows more small breweries to enter but displaces workers from larger beverage manufacturers or distributors, the net effect on NAICS 312 employment could be zero even if the composition shifts toward smaller, more numerous establishments. The QWI data cannot distinguish establishment size within a NAICS code, so this reallocation channel is consistent with the observed null.

Third, the treatment may be too noisy for detection. NAICS 312 includes soft drink bottlers, wineries, distilleries, and tobacco manufacturers---all unaffected by brewery self-distribution laws. If craft breweries account for roughly 30 percent of NAICS 312 employment, the true brewery-specific effect could be three times larger than my estimate. Even so, the triple-difference against food manufacturing shows no differential movement, suggesting the dilution alone cannot explain the null.

\section{Conclusion}

Removing the middleman did not create jobs. Self-distribution laws, adopted with the explicit goal of enabling craft brewery growth, had no measurable effect on beverage manufacturing employment across 20 states over nearly a decade. The craft brewing boom---real and transformative---was driven by consumer demand, not by supply-side distribution deregulation. For policymakers considering similar reforms in adjacent markets (cannabis, spirits, direct-to-consumer food), the lesson is cautionary: lowering regulatory barriers to entry is desirable on efficiency grounds, but the employment dividends may be illusory.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsection{Data Sources}

The primary data source is the Quarterly Workforce Indicators (QWI), produced by the Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) program. The QWI is derived from state unemployment insurance records matched to Census demographic data. I access QWI data through pre-processed Parquet files hosted on Azure Blob Storage, originally downloaded from \url{https://lehd.ces.census.gov/data/qwi/latest_release/}. The data cover all 50 states plus the District of Columbia and Puerto Rico, from as early as 1990 (state-dependent) through 2025. I use three demographic cuts: sex $\times$ age (sa), sex $\times$ education (se), and race $\times$ ethnicity (rh), all at the 3-digit NAICS industry level (n3).

\subsection{Treatment Coding}

Self-distribution law adoption dates are compiled from the Brewers Association state regulatory database, the National Conference of State Legislatures, and individual state legislative records. For each state, I identify the quarter in which self-distribution was first permitted or substantially expanded (e.g., a meaningful increase in production volume caps or the addition of direct-to-retailer sales rights). Twenty states are coded as treated during 2011--2019: MN (2011Q3), AL (2012Q2), MS, NC, VA, NJ (2012Q3), TX, WV, NV (2013Q3), SC, KY (2014Q2), FL, LA (2014Q3), IN, ND, SD (2015Q3), TN (2016Q3), GA (2017Q3), OK (2018Q4), KS (2019Q3). All other states are coded as never-treated.

\subsection{Sample Construction}

The primary analysis panel consists of state $\times$ quarter $\times$ industry observations for NAICS 312 (Beverage and Tobacco Product Manufacturing) and NAICS 311 (Food Manufacturing) over 2001Q1--2024Q4, totaling 9,790 observations (4,891 for NAICS 312, 4,899 for NAICS 311) across 52 state-level jurisdictions. County-level data are aggregated to the state level; I retain county counts as a measure of the extensive margin. The education heterogeneity panel contains 4,758 state-quarter observations per education category for NAICS 312.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
