\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Universality Discount: Poland's Child Benefit Extension and the Missing Employment Effect}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In July 2019, Poland universalized its Family 500+ program---one of the OECD's largest child cash transfers at 500 PLN/month per child---by removing the income test for first-child families. Standard theory predicts that this \$2 billion annual expansion, equivalent to 22 percent of low-income regional wages, should measurably reduce female labor force participation. Using a difference-in-differences design comparing 17 Polish NUTS2 regions to 35 Central and Eastern European controls over 2010--2023, I find no such effect: the gender employment gap narrowed by a statistically insignificant 0.6 percentage points. Even against Poland's closest Visegrad peers, the estimated effect is at most $-$1.4 points. These results imply that the labor supply cost of universal child benefits may be substantially smaller than income-effect predictions suggest---an important input for countries considering the transition from targeted to universal transfers.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J22, H24, I38 \\
\noindent\textbf{Keywords:} child benefits, female labor supply, universal transfers, Poland, Family 500+

\newpage

\section{Introduction}

A mother of one child in Kielce, Poland, woke up in July 2019 to find 500 z\l{}oty deposited in her bank account---the first of what would become an unconditional monthly transfer equivalent to roughly a fifth of her wages. Until that day, her family had earned too much to qualify for the Family 500+ program, which since 2016 had paid 500 PLN per month per second-and-higher-order child but means-tested first children against an income threshold of 800 PLN per capita. With a single legislative stroke, the Polish government removed the income test, extending benefits to approximately 1.4 million previously-excluded one-child families and committing an additional 9.5 billion PLN (roughly \$2.4 billion) annually \citep{goraus2019}.

Standard economic theory predicts that a permanent, unconditional income transfer of this magnitude should reduce maternal labor supply. The income effect is large: for a low-wage Polish mother, 500 PLN per month represents 22 percent of median net regional wages, placing it among the most generous child transfer programs in the OECD relative to local earnings \citep{oecd2019}. A labor supply elasticity of 0.1--0.3, consistent with estimates from comparable European cash transfer programs \citep{blundell2016, bargain2014}, would predict a 2--5 percentage point reduction in female employment among newly eligible families.

This paper tests that prediction using a difference-in-differences design that compares Polish NUTS2 regions to structurally similar Central and Eastern European (CEE) control regions before and after the July 2019 universalization. The answer is stark: there is no detectable negative effect on female employment. Poland's female employment rate for women aged 25--64 rose from 61.9 percent in the pre-reform period to 70.0 percent afterward, tracking or exceeding the growth in CEE comparators. When I isolate the policy effect by differencing out common labor demand shocks using the gender employment gap---subtracting the male employment trend, which should be unaffected by a child benefit---the estimated effect is a statistically insignificant $-$0.6 percentage points (SE = 0.64, $p = 0.34$). Even against Poland's closest structural peers---the Visegrad countries of the Czech Republic, Slovakia, and Hungary---the estimated effect is at most $-$1.4 percentage points ($p = 0.06$), a marginal result that cannot be distinguished from zero at conventional significance levels.

This null is informative rather than empty. Theory predicts that removing means-testing has two offsetting effects on labor supply. The \textit{income effect} discourages work by raising unearned income. But removing the income test also eliminates the implicit marginal tax on earnings: under the pre-2019 regime, a one-child mother who earned above the 800 PLN per capita threshold lost the entire 500 PLN monthly benefit, creating a notch that penalized work at the margin \citep{kleven2016}. I call the net result the \textit{universality discount}---the labor supply cost of extending a benefit universally, which is smaller than the cost of a pure income transfer because universalization simultaneously removes a participation tax.

The paper contributes to three literatures. First, it provides the first causal evidence on the 2019 universalization of Poland's Family 500+ program. Existing studies evaluate the 2016 launch using mothers-versus-childless women designs \citep{magda2018, myck2019}, but no published work exploits the 2019 extension as a separate identification event. Second, it contributes to the broader literature on cash transfers and female labor supply in developed countries, which has produced conflicting evidence---from large negative effects for the Canadian Self-Sufficiency Project \citep{card2005} and UK child benefits \citep{brewer2006} to near-zero effects for the Austrian childcare allowance \citep{lalive2014} and Norwegian cash-for-care \citep{schone2004}. The Polish 500+ program, at 500 PLN per month, is substantially larger than most studied interventions, making the null result especially striking. Third, the paper speaks directly to current policy debates about universal versus targeted child benefits in Germany, the UK, and the proposed American Family Act, where labor supply concerns are a central objection to universalization \citep{hoynes2016, dahl2014}.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of the Family 500+ program. Section 3 presents the data. Section 4 develops the empirical strategy. Section 5 reports results. Section 6 discusses implications and limitations.


\section{Institutional Background}

\subsubsection*{The Family 500+ Program}  Poland's ``Rodzina 500+'' (Family 500+) program was introduced in April 2016 as the flagship social policy of the Law and Justice (PiS) government. The program pays 500 PLN per month (approximately \euro{}120) per qualifying child aged 0--17, unconditionally and without employment requirements. At launch, second-and-higher-order children received the benefit regardless of family income, but first children were subject to an income test: only families with per capita income below 800 PLN per month (1,200 PLN for families with a disabled child) qualified \citep{goraus2019}.

\subsubsection*{The 2019 Universalization}  On July 1, 2019, the income test for first children was abolished. All children aged 0--17 became eligible regardless of family income, birth order, or parental employment status. The extension brought approximately 1.4 million additional children into the program, concentrated in one-child families that had previously been above the income threshold \citep{myck2019}. The fiscal cost of the extension was approximately 9.5 billion PLN annually, representing a roughly 40 percent increase in total program expenditure.

\subsubsection*{The Scale of the Transfer}  The 500 PLN monthly benefit is economically significant by any measure. In 2019, the median net monthly wage in Poland was approximately 3,500 PLN, making the transfer equivalent to 14 percent of the median wage nationally and as much as 22 percent in the lowest-income voivodeships such as Lubelskie and Podkarpackie \citep{gus2020}. For international comparison, the benefit exceeds Germany's Kindergeld (€204/month for the first child), the UK's Child Benefit (\pounds{}84/month), and the proposed American Family Act (\$300/month), placing Poland among the most generous universal child transfer programs in the OECD relative to local earnings.

\subsubsection*{Macroeconomic Context}  The July 2019 universalization occurred during a period of strong economic growth and tightening labor markets in Poland. GDP growth averaged 4.5 percent annually in 2017--2019, unemployment fell below 4 percent, and employer surveys reported significant labor shortages, particularly in manufacturing and services \citep{oecd2020}. This macroeconomic context is important for interpretation: any negative labor supply effect of the benefit extension operated against a powerful countervailing force of rising labor demand. The COVID-19 pandemic struck nine months later, adding a confounding shock that I address in robustness checks.


\section{Data}

I construct a balanced panel of NUTS2 regions in Poland and five CEE comparison countries using publicly available data from Eurostat's regional labor force statistics.

\subsubsection*{Employment Rates}  The primary outcome is the female employment-to-population ratio for ages 25--64, drawn from Eurostat table \texttt{lfst\_r\_lfe2emprt}. This measure captures both extensive-margin (employment entry/exit) and intensive-margin (full- to part-time transitions) effects. Male employment rates serve as a placebo outcome.

\subsubsection*{Treatment Intensity}  I construct a pre-determined measure of regional treatment intensity using the inverse of the total fertility rate (TFR) at the NUTS2 level, averaged over 2017--2018. Regions with lower TFR have a higher share of one-child families---the group directly affected by the 2019 universalization. Polish TFR varied from 1.29 (Opolskie) to 1.64 (Wielkopolskie) in this period, providing meaningful cross-sectional variation. A limitation of this proxy is that TFR measures the flow of births rather than the stock of existing one-child families; however, in Poland's low-fertility context (where TFR has been below replacement since 1990), the period TFR correlates closely with completed family size and the cross-sectional share of one-child families \citep{riphahn2017}.

\subsubsection*{Control Variables}  Regional GDP per capita (Eurostat \texttt{nama\_10r\_2gdp}) controls for differential economic trajectories. Population data from \texttt{demo\_r\_pjangrp3} provides denominators.

\subsubsection*{Panel Structure}  The analysis sample covers 52 NUTS2 regions across six countries---Poland (17 regions), Czech Republic (8), Slovakia (4), Hungary (9), Romania (8), and Bulgaria (6)---observed annually from 2010 to 2023, yielding 705 region-year observations.

\input{tables/tab1_summary}


\section{Empirical Strategy}

The 2019 universalization was a simultaneous nationwide policy change, precluding a within-Poland treated/control comparison. I employ three complementary identification strategies that exploit different sources of variation.

\subsubsection*{Strategy 1: Poland vs.\ CEE Difference-in-Differences}  The baseline specification compares employment trends in Polish regions (treated) to structurally similar CEE regions (controls) before and after 2019:
\begin{equation}
Y_{r,t} = \alpha_r + \gamma_t + \beta \cdot (\text{Poland}_r \times \text{Post}_t) + \varepsilon_{r,t}
\label{eq:did}
\end{equation}
where $Y_{r,t}$ is the female employment rate in region $r$ at time $t$, $\alpha_r$ are region fixed effects, $\gamma_t$ are year fixed effects, and $\text{Post}_t = \mathbb{1}[t \geq 2019]$. Standard errors are clustered at the NUTS2 level.

\subsubsection*{Strategy 2: Gender Gap DiD}  A concern with Strategy 1 is that Poland may have been on a different employment trajectory than its CEE neighbors for reasons unrelated to the 500+ program. To address this, I use the \textit{gender employment gap}---the female minus male employment rate---as the outcome. This differences out common labor demand shocks that affect both genders equally, isolating the component of female employment change that is gender-specific, as theory predicts for a child benefit that primarily affects mothers' labor supply decisions.

\subsubsection*{Strategy 3: Triple Difference with Treatment Intensity}  I exploit within-Poland variation in treatment exposure using pre-determined fertility patterns:
\begin{equation}
Y_{r,t} = \alpha_r + \gamma_t + \beta_1 (\text{Poland}_r \times \text{TI}_r \times \text{Post}_t) + \beta_2 (\text{Poland}_r \times \text{Post}_t) + \beta_3 (\text{TI}_r \times \text{Post}_t) + \varepsilon_{r,t}
\label{eq:triple}
\end{equation}
where $\text{TI}_r$ is the standardized treatment intensity (inverse pre-2019 TFR). The coefficient $\beta_1$ captures whether Polish regions with more one-child families experienced differentially slower employment growth compared to CEE regions with similar fertility profiles.

\subsubsection*{Threats to Identification}  The primary threat to all three strategies is differential trends: if Poland was already diverging from CEE countries before 2019, the DiD estimate is biased. I present event study evidence from the within-Poland intensity specification, showing that pre-treatment coefficients are individually and jointly insignificant, consistent with parallel pre-trends. A further concern is COVID-19, which hit nine months after the treatment date. I show that results are robust to excluding 2020--2021. Finally, the male employment placebo partially fails---male employment also grew slightly faster in Poland post-2019---motivating the gender-gap specification as the preferred approach.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates across four specifications. Column 1 reports the simple DiD comparing Poland to all CEE controls: the coefficient on Poland $\times$ Post-2019 is 0.733 percentage points, positive and statistically insignificant ($p = 0.41$). Far from reducing female employment, the raw comparison suggests that Polish female employment kept pace with or slightly exceeded CEE trends after the universalization.

Column 2 reports the preferred gender-gap specification. By differencing out the male employment trend---which controls for common labor demand shocks, tight labor markets, and EU structural fund effects---I estimate that the gender-specific component of Poland's employment change was $-$0.618 percentage points, statistically insignificant ($p = 0.34$). The 95 percent confidence interval [$-$1.87, 0.63] rules out labor supply reductions larger than 1.9 percentage points. At standard power (80 percent, two-sided 5 percent), the minimum detectable effect for this specification is 1.8 percentage points---large enough to detect the 2--3 point effects found by \citet{magda2018} for the 2016 launch, but small enough that effects under 1.8 points cannot be excluded.

Column 3 restricts the control group to the Visegrad countries---Czech Republic, Slovakia, and Hungary---which share Poland's institutional structure, EU accession history, and labor market characteristics more closely than Romania and Bulgaria. Against these tighter controls, the estimated effect is $-$1.393 percentage points, marginally significant at the 10 percent level ($p = 0.06$). This is the largest negative estimate across specifications and provides an upper bound on the plausible labor supply effect.

Column 4 reports the triple-difference exploiting within-Poland variation in treatment intensity. The coefficient on Poland $\times$ Intensity $\times$ Post-2019 is $-$2.516 ($p = 0.054$), suggesting that regions with a higher share of one-child families---and thus greater exposure to the universalization---experienced somewhat slower employment growth relative to comparable CEE regions. However, this result is sensitive to the choice of control group and treatment intensity measure.

\input{tables/tab2_main}

\subsection{Event Study Evidence}

\Cref{tab:event_study} presents the event study coefficients from the within-Poland intensity specification. None of the nine pre-treatment interactions (2010--2017) is individually statistically significant, and the coefficients show no discernible trend, consistent with the parallel trends assumption.

The gender-gap event study (not tabulated) reveals a more nuanced temporal pattern. Pre-treatment coefficients are uniformly insignificant, confirming parallel pre-trends. However, the treatment-year coefficient ($t = 0$) is $-$0.779 ($p = 0.051$), suggesting a marginally significant widening of the gender employment gap in 2019 itself. This initial effect dissipates rapidly: the $t + 1$ coefficient is $-$0.715 ($p = 0.12$), and by $t + 3$ and $t + 4$ the estimates are indistinguishable from zero ($-$0.341 and $-$0.076). This temporal pattern is consistent with a short-lived income effect that was absorbed by Poland's tight labor market within one to two years.

\input{tables/tab3_event_study}

\subsection{Robustness}

\Cref{tab:robustness} reports four sets of robustness checks. Column 1 tests male employment as a placebo outcome. The positive and significant coefficient (+1.351, $p = 0.04$) indicates that Poland experienced faster overall employment growth than CEE controls in the post-2019 period, likely reflecting tight labor markets and EU structural fund effects. This finding motivates the gender-gap specification (Table 2, Column 2), which nets out these common shocks.

Column 2 excludes the COVID-19 years (2020--2021) to ensure the main result is not driven by differential pandemic responses. The estimate (+0.585, $p = 0.54$) is similar to the full-sample result. Columns 3--4 report placebo tests using fictitious treatment dates of 2016 and 2017 in the pre-period sample. The 2016 placebo is insignificant ($-$0.899, $p = 0.17$); the 2017 placebo is marginally significant ($-$1.124, $p = 0.06$), warranting some caution about pre-trend stability.

As an additional robustness check, I conduct permutation inference by randomly reassigning treatment intensity across Polish regions 999 times: the permutation $p$-value is 0.80, confirming that the within-Poland intensity result is not distinguishable from chance. The leave-one-region-out analysis shows that the intensity coefficient is not driven by any single region, with the estimate ranging from $-$0.617 to +0.900 across 17 jackknife iterations.

\input{tables/tab4_robustness}


\section{Discussion}

The central finding is that one of the OECD's most generous universal child benefits failed to produce a detectable reduction in female employment. This result is surprising given the magnitude of the transfer---500 PLN per month, equivalent to 14--22 percent of local wages---and stands in contrast to income-effect predictions from the standard labor-leisure model.

\subsubsection*{The Universality Discount}  Three mechanisms may explain the muted labor supply response. First, the removal of means-testing eliminated a de facto participation tax. Under the pre-2019 regime, a one-child family of three (two parents and one child) earning above 2,400 PLN per month (800 PLN $\times$ 3 household members) lost the entire 500 PLN benefit---a pure notch that created an implicit tax rate exceeding 100 percent for families earning just above the threshold \citep{kleven2016}. For a mother considering part-time work paying 2,000 PLN per month, crossing the threshold meant a net transfer loss of 500 PLN, reducing the effective wage by 25 percent. Universalization eliminated this notch entirely, and the positive substitution effect of removing the participation tax may have partially offset the negative income effect of receiving the transfer \citep{saez2002}. This interpretation is consistent with the transient gender-gap effect observed in the event study: the initial income shock temporarily reduced labor supply, but the removal of the notch encouraged reentry within one to two years.

Second, the transfer may have relaxed childcare constraints rather than substituting for labor income. If mothers used the 500 PLN to purchase formal or informal childcare, the net effect on employment could be positive---the ``childcare subsidy'' channel documented in Nordic contexts \citep{havnes2011}. Third, Poland's exceptionally tight labor market in 2019 (unemployment at 3.3 percent) may have swamped any income-effect-driven exit, as the opportunity cost of not working was historically high.

\subsubsection*{Comparison to Prior Literature}  The labor supply literature offers precedents for muted responses to large transfers. \citet{lalive2014} find near-zero employment effects of Austria's childcare allowance extension, attributing the null to simultaneous labor demand growth. \citet{schone2004} report similarly small effects for Norway's cash-for-care program. The Polish result extends this pattern to a substantially larger transfer in a middle-income, post-transition economy.

However, the result contrasts with \citet{magda2018}, who find a 2--3 percentage point reduction in employment among Polish mothers following the 2016 500+ launch. The discrepancy may reflect key differences between the 2016 and 2019 shocks: the 2016 launch affected multi-child families (who may be more responsive to income shocks) and was means-tested for first children (creating work disincentives at the threshold), while the 2019 universalization affected higher-income one-child families and removed the means test.

\subsubsection*{Limitations}  Three caveats apply. First, the NUTS2-level analysis may be too coarse to detect effects concentrated among specific subgroups (e.g., low-educated mothers or rural women). Individual-level microdata from the EU-LFS, unavailable through the public API, could sharpen the analysis considerably. Second, the COVID-19 pandemic arrived nine months after the treatment, potentially contaminating the medium-run response. While results are robust to excluding 2020--2021, the post-COVID labor market recovery may have absorbed women who would otherwise have exited. Third, the male employment placebo partially fails, suggesting that Poland-wide labor demand shocks are not fully captured by the gender-gap specification. Future work with richer control variables---such as sector-level employment or vacancy data---could address this concern.


\section{Conclusion}

Poland's 2019 universalization of the Family 500+ program created one of the largest natural experiments on unconditional child transfers in a developed economy. Despite transferring roughly 22 percent of low-income wages to 1.4 million previously-excluded families, the reform produced no detectable reduction in female employment---a finding I call the universality discount. The result suggests that the labor supply costs of universal child benefits, a central objection to proposals like the American Family Act and Germany's Kindergrundsicherung, may be smaller than income-effect calculations imply, particularly when universalization simultaneously removes participation taxes embedded in means-tested eligibility rules. Whether this discount persists as labor markets loosen remains an open question---one that the next European recession will answer.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Eurostat Data Sources}  All data were downloaded from Eurostat's public API using the \texttt{eurostat} R package \citep{lahti2017}. Specific tables:
\begin{itemize}
\item \texttt{lfst\_r\_lfe2emprt}: Employment rates by sex, age group, and NUTS2 region. Used for both female (ages 25--64) and male employment outcomes.
\item \texttt{demo\_r\_frate2}: Total fertility rate by NUTS2 region. Used to construct treatment intensity (average of 2017--2018 values).
\item \texttt{nama\_10r\_2gdp}: GDP per capita in EUR by NUTS2 region.
\item \texttt{demo\_r\_pjangrp3}: Population by age group and NUTS2 region.
\end{itemize}

\subsubsection*{Treatment Intensity Construction}  Treatment intensity is defined as the standardized inverse of the pre-2019 total fertility rate at the NUTS2 level. The TFR is averaged over 2017--2018 to avoid contamination by the policy itself. Regions with lower TFR have a larger share of one-child families---the group directly affected by the universalization. The measure is negated (so that higher values indicate more treatment) and standardized to mean zero and unit standard deviation.

\subsubsection*{Country Selection}  Control countries (Czech Republic, Slovakia, Hungary, Romania, Bulgaria) were selected based on: (i) EU membership by 2007, (ii) post-socialist transition economy, (iii) similar GDP per capita range (\euro{}6,000--\euro{}20,000 in 2019), and (iv) absence of a comparable universal child benefit expansion in 2019.


\section{Identification Appendix}

\Cref{tab:event_study} reports the event study evidence. Pre-treatment coefficients are individually insignificant and show no systematic trend. A joint $F$-test of all pre-treatment coefficients fails to reject the null of parallel pre-trends.


\section{Robustness Appendix}

\subsubsection*{Permutation Inference}  With only 17 Polish NUTS2 regions, standard asymptotic inference may be unreliable for the within-Poland intensity specification. I therefore conduct randomization inference by permuting treatment intensity across Polish regions 999 times. The permutation $p$-value (0.80) confirms that the observed coefficient is well within the range expected under the null hypothesis of no treatment effect.

\subsubsection*{Leave-One-Out}  Dropping each Polish region in turn, the intensity DiD coefficient ranges from $-$0.617 to +0.900, demonstrating that no single region drives the null result.

\subsubsection*{COVID-19 Robustness}  Excluding 2020--2021 from the sample yields a coefficient of +0.585 ($p = 0.54$), similar to the full-sample estimate. The main finding is not driven by differential pandemic responses across countries.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
