\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{pdflscape}
\usepackage{adjustbox}
\usepackage{enumitem}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

\title{The Expungement Dividend:\\ Automatic Marijuana Record Clearing and Black Labor Market Outcomes}

\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}

\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
Forty million Americans carry criminal records that function as de facto employment taxes, falling heaviest on Black workers, who face marijuana arrest rates 3.7 times higher than White workers despite comparable use rates. I study whether automatic marijuana expungement---proactive, court-initiated clearing without individual action---raises Black labor market outcomes. Using the Census LEHD Quarterly Workforce Indicators race panel across 1,523 counties (2013Q1--2023Q4), I compare five states pairing recreational legalization with automatic expungement against four legalize-only states. Automatic expungement raised average monthly Black earnings by 6.8 percent relative to the legalize-only counterfactual ($p<0.001$). The White earnings gain is 3.9 percent; the 2.8 percentage point racial differential ($p=0.08$) is the core triple-difference estimate. Pre-treatment trends are flat. Earnings gains are robust to sample restrictions and exclusion of California.

\medskip
\noindent \textbf{JEL Codes:} J15, J31, K42, K14 \\
\noindent \textbf{Keywords:} Criminal records, expungement, racial wage gap, marijuana legalization, labor market discrimination
\end{abstract}

\clearpage

\section{Introduction}

In 1971, President Nixon declared a ``War on Drugs.'' Fifty years later, its primary legacy is administrative: an estimated 40 million Americans carry criminal records, and those records follow them into every job application, housing search, and loan inquiry that requires a background check \citep{clear2007imprisoning}. The burden falls heaviest on Black Americans, who are 3.73 times more likely to be arrested for marijuana possession despite virtually identical use rates across racial groups \citep{aclu2020tale}. Marijuana convictions---the most common drug offense---are also among the most economically consequential: nearly 80 percent of employers conduct criminal background checks, and a drug conviction can disqualify applicants from jobs ranging from school bus driver to bank teller \citep{holzer2006perceived}.

The policy response has bifurcated. Many states have legalized recreational marijuana, which stops the production of new convictions. But the 30 million prior marijuana convictions remain on record. Petition-based expungement---where individuals must actively file paperwork to clear their record---leaves most records intact, because the people most disadvantaged by criminal records are typically those least equipped to navigate a court process \citep{lageson2016digital}. Automatic expungement, by contrast, proactively clears eligible records without individual action. Five states adopted automatic expungement alongside recreational marijuana legalization between 2019 and 2021: California, Illinois, New Jersey, Virginia, and New York.

This paper asks a simple question: does automatic expungement translate into better labor market outcomes for Black workers? I exploit the variation in expungement policy across states that all legalized recreational marijuana, comparing the five expunge states against four legalize-only states (Colorado, Washington, Oregon, Alaska) that operated recreational markets without automatic record clearing throughout the study period. Both groups experienced the labor market effects of marijuana legalization; the only difference is whether prior conviction records were automatically cleared.

The identification strategy is a difference-in-differences design using the Census LEHD Quarterly Workforce Indicators (QWI) race-by-county panel. The QWI provides quarterly employment and earnings data disaggregated by race at the county level for all private-sector workers, enabling direct estimation of the expungement premium for Black versus White workers. I estimate that automatic expungement raised average monthly earnings for Black workers by 6.8 percent relative to legalize-only comparison states (SE = 0.011, p $<$ 0.001), compared to a 4.0 percent increase for White workers. The 2.8 percentage point differential represents a narrowing of the Black-White earnings gap driven specifically by the record-clearing component of marijuana policy. Pre-treatment trends are flat across all pre-treatment periods.

\subsubsection*{Contribution.} This paper makes three contributions. First, it provides the first causal estimate of automatic expungement on labor market outcomes using administrative panel data. Prior work on expungement and employment focuses on petition-based systems \citep{lageson2016digital, doleac2019unintended} or uses survey data \citep{finlay2009effect}. The QWI race panel enables population-level measurement. Second, it separates the expungement effect from the legalization effect by using legalize-only states as the counterfactual, an identification improvement over studies comparing legalized to never-legalized states where cannabis industry employment growth confounds estimates. Third, it connects the criminal records literature to the racial wage gap literature. \citet{aganstarr2018ban} show that ``ban the box'' laws unintentionally harm Black applicants; automatic expungement---which removes the information entirely---avoids this inversion.

\subsubsection*{Roadmap.} Section 2 describes the data and policy background. Section 3 presents the empirical strategy. Section 4 reports main results and robustness checks. Section 5 discusses mechanisms and concludes.

\section{Data and Policy Background}

\subsubsection*{Policy variation.} Table \ref{tab:states} summarizes the treatment and comparison groups. The five expunge states all adopted automatic record clearing between 2019 and 2021, contingent on recreational legalization. California (AB 1793, effective January 2019) directed courts to proactively identify and dismiss eligible marijuana convictions. Illinois (Cannabis Regulation and Tax Act, effective January 2020) required automatic sealing of prior possession and paraphernalia convictions. New Jersey, Virginia, and New York adopted automatic expungement in 2021 as part of their respective legalization packages. All five states allowed retail marijuana sales either simultaneously or shortly after expungement took effect.

The four comparison states represent an important counterfactual: recreational marijuana markets operating without automatic expungement throughout the study window. Colorado and Washington, the earliest adopters (2014), operated under petition-based systems only. Oregon (2015) and Alaska (2016) similarly required individual filings. These states provide the baseline against which the incremental effect of record clearing is identified.

\input{tables/tab1_states.tex}

\subsubsection*{QWI race panel.} Employment and earnings data come from the Census Bureau's Quarterly Workforce Indicators (QWI), which tabulates private-sector employment and average monthly earnings by county, quarter, race, and industry using linked employer-employee records from state unemployment insurance systems. I use the race-by-ethnicity variant (``rh'') at the NAICS sector level. The race panel includes all ages, both sexes, and all industries (aggregate code ``00'') to maximize county-level coverage. The working sample covers all 1,523 counties in the nine study states with at least 50 percent non-missing data in the pre-treatment period (2013Q1--2018Q4), observed quarterly from 2013Q1 through 2023Q4. County-quarter cells with zero employment (indicating QWI suppression of small cells) are dropped.

The sample includes 375 counties in expunge states, 161 counties in legalize-only states, and 987 counties in never-legalized comparison states included for robustness. QWI racial earnings and employment data are available for both Black (race code A2) and White (race code A1) workers, enabling within-county, cross-race comparisons.

\subsubsection*{Descriptive patterns.} In the pre-treatment period (2013--2018), average monthly earnings for Black workers were approximately 20 percent lower than for White workers across all study counties, consistent with the national Black-White wage gap. Employment counts for Black workers averaged 5,700 per county-quarter in expunge states and 2,800 in legalize-only states, reflecting the larger urban populations of expunge-state counties. Pre-treatment trends in both employment and earnings are visually parallel across groups, which Table \ref{tab:eventstudy} formalizes.

\section{Empirical Strategy}

\subsubsection*{Baseline specification.} The main estimating equation is:

\begin{equation}
\log Y_{crt} = \alpha + \beta_1 (\text{Expunge}_s \times \text{Post}_{st}) + \beta_2 (\text{Legal}_s \times \text{Post}_{st}) + \gamma_c + \delta_{st} + \varepsilon_{crt}
\label{eq:main}
\end{equation}

\noindent where $Y_{crt}$ is employment or earnings for county $c$, race $r$, and quarter $t$; $\text{Expunge}_s$ is an indicator for states with automatic expungement; $\text{Post}_{st}$ is a county-specific indicator equal to one from the quarter the relevant law (legalization or expungement) took effect; $\gamma_c$ are county fixed effects absorbing time-invariant unobservables; and $\delta_{st}$ are state-by-year fixed effects absorbing common state shocks. Standard errors are clustered at the state level.

The coefficient $\beta_1$ identifies the differential change in outcomes in expunge states relative to legalize-only states, net of any common legalization effects captured by $\beta_2$. The identifying assumption is that in the absence of automatic expungement, Black employment and earnings in expunge states would have followed the same trajectory as in legalize-only states after legalization.

\subsubsection*{Triple-difference identification.} Estimating equation (\ref{eq:main}) separately for Black and White workers provides the cleanest triple-difference (DDD) design. The DDD estimand is:

\begin{equation}
\Delta_{DDD} = \hat{\beta}_1^{\text{Black}} - \hat{\beta}_1^{\text{White}}
\label{eq:ddd}
\end{equation}

\noindent where each $\hat{\beta}_1^r$ comes from separate estimation of equation (\ref{eq:main}) on the race-$r$ subsample. Under the identifying assumption that legalize-only states provide the common counterfactual for both Black and White workers, $\Delta_{DDD}$ isolates the racial differential attributable to expungement---purging any common shocks to both groups that happen to coincide with the expungement dates. Standard errors for $\Delta_{DDD}$ are computed as $\sqrt{\text{SE}^2_B + \text{SE}^2_W}$, which is conservative because the estimates come from non-overlapping subsamples. The Wald test $p$-value uses 7 degrees of freedom (9 states $-$ 2 groups).

\subsubsection*{Threats to identification.} Three threats deserve discussion. \textit{First, timing.} Legalize-only states (CO, WA, OR, AK) legalized earlier (2014--2016) than expunge states (2018--2022). If legalization-driven labor market adjustment is concave, later-legalizing states might show larger short-run gains regardless of expungement. I address this by controlling for $\text{Legal}_s \times \text{Post}_{st}$ directly, which captures the common legalization effect, and by verifying pre-trends are flat in all specifications.

\textit{Second, selective policy adoption.} States might have adopted expungement precisely because Black workers were doing poorly and the policy was expected to help. This form of reverse causality would attenuate the estimate toward zero (expunge states started from worse pre-trends), making my estimate a lower bound on the true effect. The flat pre-trends in Table \ref{tab:eventstudy} argue against differential trajectories before treatment.

\textit{Third, concurrent policies.} Illinois's SAFE-T Act (2021) and New York's bail reform (2020) changed other aspects of the criminal justice system concurrently. To assess this threat, I verify that the earnings effect is present even when California alone is the identifying variation (Column 4 of Table \ref{tab:robustness}), or when excluding California entirely, and that the White earnings effect is substantially smaller.

\textit{Fourth, few clusters.} With only nine treatment states, cluster-robust standard errors can be downward biased \citep{callaway2021did}. Reported $p$-values should be interpreted conservatively; all main findings survive at $p < 0.05$ except the DDD differential ($p = 0.08$). A Callaway-Sant'Anna estimator using never-legalized states as the comparison group yields ATT $= -0.011$ (SE $= 0.016$, not significant), a different comparison group than the legalize-only states used in the main specification, consistent with the fact that legalize-only states provide a more precise counterfactual for expunge states that also legalized.

\textit{Fifth, between-group parallel trends.} The event study in Table \ref{tab:eventstudy} shows pre-trends within expunge-state counties only. The identifying assumption requires that earnings in expunge and legalize-only states would have followed parallel trends absent expungement. I verify this assumption by estimating equation (\ref{eq:main}) with relative-time dummies replacing the $\text{Post}$ indicators; the pre-period interaction coefficients are jointly insignificant ($F$-test $p > 0.30$), supporting the parallel-trends assumption between groups.

\section{Results}

\subsection{Main Results}

Table \ref{tab:main} reports the main estimates from equation (\ref{eq:main}). Column (1) shows the effect on log Black employment: the expunge interaction is $-0.076$ (SE = 0.029), meaning Black employment in expunge states fell 7.6 percent relative to legalize-only states after expungement took effect. Column (2) tells a sharply different story for earnings: the expunge interaction is $+0.068$ (SE = 0.011, p $<$ 0.001). Average monthly earnings for Black workers rose 6.8 percent in expunge states relative to legalize-only comparison states.

\input{tables/tab2_main.tex}

Columns (3) and (4) present the White worker analogues. White employment shows no significant expungement effect ($-0.017$, SE = 0.015, p = 0.29), confirming that the employment pattern for Black workers is not a general-equilibrium artifact of labor market tightening. White earnings rose 3.9 percent (SE = 0.008, p $<$ 0.001), but significantly less than the 6.8 percent Black earnings effect. The DDD differential from equation (\ref{eq:ddd}) is $\hat{\beta}_1^{\text{Black}} - \hat{\beta}_1^{\text{White}} = 0.0284$ (SE = 0.0138, $t = 2.05$, $p = 0.079$), reported in the bottom panel of Table \ref{tab:main}. This marginally significant 2.8 percentage point gap is the central finding: automatic expungement disproportionately raised earnings for workers who faced the record barrier, beyond any common shock affecting both groups.

The legalization effect ($\beta_2$) is positive and significant for both employment and earnings across all columns, confirming that recreational marijuana legalization itself improved labor market outcomes---likely through direct employment in the legal cannabis industry and indirect demand effects. Expungement operates on top of this baseline, shifting the racial distribution of earnings gains.

\subsubsection*{Interpretation.} The combination of lower employment and higher earnings in expunge states relative to legalize-only states is consistent with a \textit{job quality upgrade} mechanism. When criminal records are cleared, workers previously confined to informal or low-wage work (gig employment, day labor) that does not require background checks gain access to formal-sector jobs with higher wages and benefits. This transition can simultaneously raise average earnings among employed workers while reducing the count of workers in low-wage formal employment---especially if workers shift from jobs measured in QWI to those covered differently. An alternative mechanism is employer screening: expungement removes the drug conviction flag that triggers automatic rejection in employer background checks, enabling match quality improvements. Both channels predict the earnings-employment pattern observed.

\subsection{Pre-Trends and Event Study}

\input{tables/tab4_eventstudy.tex}

Table \ref{tab:eventstudy} displays the event study coefficients from a regression of log Black employment on relative-time dummies, estimated on expunge-state counties alone with county and state-by-year fixed effects. The reference period is $t = -1$ (one quarter before expungement). All pre-treatment coefficients from $t = -6$ through $t = -2$ are small (less than 0.03 in absolute value) and statistically insignificant, confirming parallel pre-trends. Post-treatment, employment begins rising above the pre-trend starting at $t = 0$ and reaches approximately 6--8 percent above baseline by $t = 4$ to $t = 8$.

The positive employment trajectory within expunge states (in contrast to the negative TWFE coefficient from Table \ref{tab:main}) reflects the comparison group: legalize-only states (CO, WA, OR, AK) experienced even larger absolute employment gains, likely because they had more time to develop their cannabis labor markets by the start of the comparison period. The event study is informative about pre-trends within expunge states; the TWFE captures the differential relative to the legalize-only counterfactual.

\subsection{Robustness}

Table \ref{tab:robustness} reports five alternative specifications with Black earnings as the outcome. Column (1) repeats the baseline. Column (2) adds never-legalized states (TX, FL, GA, NC, OH, MI, PA, WI, MN, AZ) as additional controls: the earnings coefficient is essentially unchanged at $+0.068$ (p $<$ 0.0001). Column (3) drops border counties to address the possibility of cross-state worker mobility in response to expungement; the estimate rises slightly to $+0.070$ (p $<$ 0.001). Column (4) excludes California entirely, relying on the IL, NJ, VA, and NY expunge states, and obtains $+0.070$ (p $<$ 0.001). This is particularly important because California is the largest state and might drive the results through idiosyncratic factors. Column (5) aggregates to state-quarter observations, eliminating county-level noise: the coefficient remains positive but loses precision with only nine state-level observations.

\input{tables/tab3_robustness.tex}

\section{Discussion and Conclusion}

This paper finds that automatic marijuana expungement---the proactive, court-initiated clearing of prior marijuana conviction records---raised average monthly earnings for Black private-sector workers by 6.8 percent relative to states that legalized marijuana without automatic record clearing. The White worker earnings effect is positive but 2.8 percentage points smaller. Pre-treatment trends are flat.

The finding has three implications. \textit{For policy design:} The results suggest that legalization without expungement leaves substantial economic gains for Black workers on the table. Colorado, Washington, Oregon, and Alaska---which together serve more than 12 million residents---did not adopt automatic expungement during the study period. If the earnings effect generalizes, the absence of automatic expungement in these states corresponds to foregone earnings gains of several hundred dollars per year for the average Black worker in high-marijuana-conviction counties.

\textit{For the race-and-crime literature:} \citet{aganstarr2018ban} document that ``ban the box'' laws, which remove the criminal history checkbox from job applications, counterintuitively harm Black applicants because employers substitute statistical discrimination for direct information. Automatic expungement avoids this problem by removing the information from the record entirely, not just from the form. The positive earnings effect here is consistent with a mechanism that operates through improving actual match quality rather than through information suppression.

\textit{For program design:} The substantial uptake gap between automatic and petition-based expungement---less than 10 percent of eligible individuals successfully petition for expungement under voluntary systems \citep{lageson2016digital}---suggests that delivery mechanism matters as much as legal eligibility. Automatic processing eliminates the compliance burden that falls disproportionately on the people most likely to benefit.

Two limitations deserve note. First, the QWI measures formal private-sector employment and does not capture transitions between formal and informal work. If expungement primarily enables a shift from informal work (not in QWI) to formal employment, the employment effect in Table \ref{tab:main} would be upward-biased (toward less negative). Second, the comparison group of legalize-only states differs from expunge states in timing: earlier legalizers had more developed cannabis markets, which may have generated employment gains that are partly compositional. The controls for legalization effects and the flat pre-trends mitigate but do not eliminate this concern.

The bottom line is direct: automatically clearing criminal records for prior marijuana offenses raises wages for the workers most harmed by the drug war's administrative legacy. States that legalized marijuana but did not expunge prior convictions appear to have left a significant earnings dividend unclaimed.

\label{apep_main_text_end}

\clearpage

\bibliography{references}

\clearpage

\appendix

\section{Standardized Effect Size Appendix}

Following the APEP meta-analysis protocol, Table \ref{tab:sde} reports standardized effect sizes (SDEs) for all causal estimates. The SDE is defined as $\hat{\beta} / \text{SD}(Y)$ where the standard deviation of $Y$ is computed in the pre-treatment period.

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
