\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{adjustbox}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Multiplier Mirage: Government Wage Shocks and Local Economic Activity in India}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
India's Sixth Central Pay Commission (2008) delivered a 20--40\% salary increase and Rs~18,060 crore in arrears to 4.7 million central government employees---the largest peacetime wage shock in history. A naive dose-response difference-in-differences using 610 Indian districts over 2003--2013 suggests that districts with higher government employment shares experienced substantially faster nightlight growth after the shock (SDE = 0.088). However, event-study diagnostics reveal strong differential pre-trends: government-heavy districts were already growing faster before 2008. After absorbing these trends, the estimated local spillover effect becomes small and statistically insignificant (SDE = $-$0.020). The apparent ``fiscal multiplier'' was a mirage created by the correlation between government employment concentration and underlying urbanization dynamics.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H30, O18, R11 \\
\noindent\textbf{Keywords:} fiscal multiplier, pay commission, nightlights, India, local economic activity, pre-trends

\newpage

\section{Introduction}

Between October and December 2008, the Government of India deposited Rs~7,224 crore---roughly \$1.5 billion---into the bank accounts of 4.7 million central government employees as the first installment of 6th Pay Commission arrears. By any standard, this was an extraordinary income shock: average salaries rose 20--40\%, the central wage bill jumped from 2\% to 3.5\% of GDP, and when state governments followed suit over the next five years, the total affected population reached 18--20 million workers \citep{Chakraborty2010}. If fiscal multipliers operate locally, as \citet{NakamuraSteinsson2014} demonstrate for U.S.\ military spending, this injection should have generated measurable spillovers in the districts where these employees live.

This paper tests that prediction by exploiting cross-district variation in pre-existing government employment intensity---measured two years before the pay commission was even constituted---in a dose-response difference-in-differences framework. Using DMSP-OLS calibrated nightlights as a proxy for local economic activity \citep{Henderson2012, Donaldson2016}, I construct a balanced panel of 610 Indian districts over 2003--2013. The identification strategy relies on the fact that some districts had government employment shares ten times higher than others in the 2005 Economic Census, generating substantial cross-sectional variation in exposure to the wage shock.

The naive estimate is large and highly significant: a one-standard-deviation increase in government employment share is associated with a 0.088 standard deviation increase in nightlights after 2008. This would imply a fiscal multiplier substantially above one, consistent with theoretical predictions for liquidity-constrained developing economies \citep{FarrierMendoza2014}.

But the event study tells a different story. All four pre-treatment coefficients (2003--2006, relative to 2007) are individually and jointly significant, revealing that government-heavy districts were already on steeper growth trajectories before the pay commission existed. This pattern is consistent with a simple but powerful confound: government employment is concentrated in state capitals and district headquarters---precisely the places undergoing rapid urbanization and service-sector growth during India's boom years \citep{Hnatkovska2012}. The differential trend is not a nuisance; it is the central finding.

After absorbing district-specific linear trends proportional to baseline government employment share, the pay commission effect becomes small and statistically insignificant ($\beta = -0.23$, $p > 0.10$). A one-standard-deviation increase in government employment share is associated with a 0.020 standard deviation \textit{decline} in nightlights after 2008, relative to the pre-existing trend. The ``fiscal multiplier'' estimated without trend controls was a statistical artifact.

This result speaks to a growing literature on fiscal multipliers in developing countries, where credible estimates remain scarce. \citet{Suarez2016} estimate local multipliers from U.S.\ stimulus spending; \citet{Chodorow2019} surveys the rich-country evidence. For developing countries, \citet{Muralidharan2023} study income effects from MGNREGA, and \citet{Egger2022} estimate multipliers from a Kenyan cash transfer experiment. My contribution is methodological rather than substantive: I show that the most natural identification strategy for studying pay commission spillovers---cross-sectional variation in government employment---is fatally confounded by urbanization trends. This is a cautionary tale for the macro-development literature, where dose-response designs using geographic concentration of fiscal exposure are common \citep{Clemens2012}.

The paper also contributes to the literature on nightlights as economic proxies. Following \citet{Henderson2012} and \citet{Asher2021}, I use SHRUG v2.1 district-level calibrated luminosity. The event-study failure here is not a limitation of the nightlights measure---it is a feature that reveals the endogeneity of government employment to local economic dynamism.

The remainder of the paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting. Section~\ref{sec:data} presents the data and summary statistics. Section~\ref{sec:strategy} outlines the empirical strategy and its key threat. Section~\ref{sec:results} presents the main results, event study, and robustness checks. Section~\ref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

India constitutes Central Pay Commissions roughly every ten years to revise compensation for central government employees. The 6th CPC was constituted in October 2006, submitted its report in March 2008, and was implemented with retrospective effect from January 1, 2006. Implementation involved two components: a permanent base pay revision (20--40\% increases, varying by grade) and lump-sum arrears covering January 2006 through September 2008 \citep{SixthCPC2008}.

\subsubsection*{Scale of the shock} The 6th CPC affected approximately 4.7 million central government employees and 2.8 million pensioners. The annual cost was Rs~7,975 crore (\$1.7 billion), and arrears totaled Rs~18,060 crore (\$3.8 billion). Forty percent of arrears were disbursed in fiscal year 2008--09, with the remaining 60\% in 2009--10. The central wage bill increased from approximately 2\% to 3.5\% of GDP \citep{Chakraborty2010}.

\subsubsection*{State adoption} State governments typically follow the central commission with their own pay revisions, but timing varies substantially. Most states constituted their own commissions between 2008 and 2010, with implementation dates spanning 2009--2013. This staggered adoption means that the total wage shock---central plus state---was distributed over roughly five years and affected an estimated 18--20 million public employees \citep{Rangarajan2011}. My identification strategy uses state $\times$ year fixed effects to absorb state-level fiscal policy variation, isolating within-state cross-district variation in exposure.

\subsubsection*{Geographic distribution of government employment} Central government employment is not uniformly distributed. It is concentrated in state capitals, district headquarters, and locations with military installations, railways (28\% of central employees), and postal services. The 2005 Economic Census records government employment shares ranging from under 5\% to over 50\% across districts, providing substantial cross-sectional variation.

\section{Data}
\label{sec:data}

I combine three SHRUG v2.1 \citep{Asher2021} datasets at the Census 2011 district level.

\subsubsection*{Treatment: Economic Census 2005} The treatment variable is the district-level share of government employees in total employment from the 5th Economic Census (2005), conducted two years before the 6th CPC was constituted. I use the SHRUG rural and urban SHRID key files to construct a concordance mapping Census 2001 district boundaries (used by the EC 2005) to Census 2011 boundaries, ensuring consistent geographic units across the panel. The resulting treatment variable, \textit{GovEmpShare}, has a mean of 0.157 (15.7\% of district employment is government) and a standard deviation of 0.101 (Table~\ref{tab:summary}).

\subsubsection*{Outcome: DMSP-OLS nightlights 2003--2013} I use calibrated DMSP-OLS annual nighttime luminosity at the district level as a continuous proxy for local economic activity, following \citet{Henderson2012}. For years with multiple satellite versions, I average across satellites. The analysis window spans 2003--2013, providing five pre-treatment years and six post-treatment years relative to the September 2008 implementation.

\subsubsection*{Controls: Census 2011} I draw district population from the Census 2011 Primary Census Abstract to control for urbanization trends.

The analysis panel consists of 6,710 observations (610 districts $\times$ 11 years).

\input{tables/tab1_summary}

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

I estimate a dose-response difference-in-differences:
\begin{equation}
\log(L_{dt} + 1) = \beta \cdot \text{GovEmpShare}_d \times \text{Post}_t + \alpha_d + \delta_{st} + \varepsilon_{dt}
\label{eq:main}
\end{equation}
where $L_{dt}$ is calibrated nightlights in district $d$ and year $t$, $\text{GovEmpShare}_d$ is the 2005 government employment share (continuous, time-invariant), $\text{Post}_t = \mathbf{1}[t \geq 2008]$, $\alpha_d$ are district fixed effects, and $\delta_{st}$ are state $\times$ year fixed effects. Standard errors are clustered at the state level.

The coefficient $\beta$ identifies the differential change in nightlights after 2008 between districts with higher versus lower pre-existing government employment concentration, within the same state-year. The identifying assumption is parallel trends: absent the pay commission, nightlight growth would not have varied systematically with baseline government employment share.

\subsection{Threats to validity}

The key threat is that government employment share correlates with local economic dynamism through channels unrelated to the pay commission. Government offices are located in administrative centers---state capitals and district headquarters---that are also hubs of urbanization, infrastructure investment, and service-sector growth. If these districts were growing faster for structural reasons, a dose-response DiD would spuriously attribute this growth to the pay commission.

I test this directly through an event study and through a de-trended specification:
\begin{equation}
\log(L_{dt} + 1) = \beta \cdot \text{GovEmpShare}_d \times \text{Post}_t + \gamma \cdot \text{GovEmpShare}_d \times (t - 2007) + \alpha_d + \delta_{st} + \varepsilon_{dt}
\label{eq:detrended}
\end{equation}
The additional term $\text{GovEmpShare}_d \times (t - 2007)$ absorbs any linear differential trend proportional to baseline government employment. If the pay commission generated real spillovers \textit{beyond} pre-existing trends, $\beta$ in Equation~\ref{eq:detrended} should remain positive.

\section{Results}
\label{sec:results}

\subsection{Main results}

Table~\ref{tab:main} presents the core findings. Column (1) estimates Equation~\ref{eq:main} with district and year fixed effects. The coefficient on $\text{GovEmpShare} \times \text{Post}$ is 1.377 ($p < 0.01$), implying substantial positive spillovers. Column (2) adds state $\times$ year fixed effects to absorb state-level fiscal variation; the coefficient falls to 0.978 but remains highly significant.

Columns (3)--(4) tell a different story. Adding $\text{GovEmpShare} \times \text{trend}$ to absorb differential pre-trends reverses the sign: the coefficient becomes $-0.417$ ($p < 0.10$). The trend term itself is large and highly significant (0.279, $p < 0.01$), confirming that government-heavy districts were growing faster throughout the sample period. Column (4) adds log population $\times$ year; the pay commission coefficient remains negative at $-0.417$.

The sign reversal from +0.978 to $-0.417$ is the central result. The apparent multiplier in columns (1)--(2) was driven entirely by differential trends, not by the causal effect of the pay commission.

\input{tables/tab2_main}

\subsection{Event study}

Table~\ref{tab:event_study} reports event-study coefficients from interactions of $\text{GovEmpShare}$ with year indicators, relative to the base year 2007. The pre-treatment coefficients are all negative and highly significant: $-0.52$ (2003), $-0.90$ (2004), $-1.30$ (2005), and $-0.88$ (2006). These coefficients are of similar magnitude and sign to each other, indicating that government-heavy districts had consistently lower nightlights relative to their 2007 level throughout the pre-period. This is not a gradual divergence (which might suggest a different confound) but rather a level pattern reflecting the underlying correlation between government employment and economic dynamism.

The post-treatment coefficients show a striking trajectory: $-0.92$ (2008), $-0.36$ (2009), then sharply positive from 2010 onward (+0.90, +0.61, +0.51, +0.50). The 2008--2009 dip is consistent with the Global Financial Crisis affecting Indian districts differentially \citep{RBI2010}, while the 2010--2013 rise continues the pre-existing growth trend. Crucially, the upward shift does not coincide with the pay commission implementation date (September 2008) but rather with the post-crisis recovery, further undermining a causal interpretation.

\input{tables/tab3_event_study}

\subsection{Robustness}

Table~\ref{tab:robustness} reports four robustness checks, all using the de-trended specification. Column (1) uses the inverse hyperbolic sine transformation of nightlights; the coefficient is virtually identical ($-0.417$). Column (2) drops the top 1\% of districts by maximum nightlight intensity to address DMSP saturation; results are unchanged ($-0.420$). Column (3) replaces the continuous treatment with a binary indicator for the top quartile of government employment; the coefficient is $-0.048$ ($p < 0.05$), confirming the null-to-negative result in a non-parametric specification. Column (4) assigns a placebo treatment date of 2005 using only pre-treatment data (2004--2007); the coefficient is positive but insignificant (0.174), consistent with the pre-existing trend being approximately linear.

\input{tables/tab4_robustness}

\subsection{Long differences in private sector activity}

Table~\ref{tab:mechanism} exploits the long difference between Economic Censuses (2005 vs.\ 2013) to examine whether government-heavy districts experienced faster private-sector growth. The coefficients are positive and significant: a 10 percentage point higher government employment share in 2005 predicts 5.6\% faster total employment growth, 6.1\% faster firm entry, and 4.7\% faster service-sector employment growth over 2005--2013.

These estimates are consistent with the nightlight trends: government-heavy districts were indeed growing faster. But the event study makes clear that this growth was not caused by the pay commission---it was already underway. The long differences cannot distinguish between pre-existing structural growth and pay-commission-induced multiplier effects because they span the full 2005--2013 period.

\input{tables/tab5_mechanism}

\section{Discussion}
\label{sec:discussion}

Using district-level government employment shares as exposure, I cannot detect positive spillovers from the 6th Pay Commission once differential trends are absorbed. This finding should be interpreted with several caveats.

First, the identification relies on a linear trend correction. If the true confound is non-linear or contains discrete shifts (e.g., infrastructure investments concentrated in administrative centers), the de-trended specification may over- or under-correct. The event-study coefficients suggest the pre-period pattern is roughly level rather than monotonically trending, which complicates the linear trend interpretation.

Second, the analysis uses district-level aggregates rather than the village-level variation envisioned in the original design. Indian districts average nearly two million residents; local multiplier effects concentrated around government office towns may be attenuated by averaging across vast rural hinterlands. Future work using SHRUG village-level data could detect effects that district aggregation obscures.

Third, the treatment period coincides with the 2008 Global Financial Crisis. While state $\times$ year fixed effects absorb state-level GFC impacts, within-state variation in crisis exposure (e.g., districts with higher export orientation or IT employment) could confound the pay commission effect. Controlling for baseline industry composition would strengthen the design.

Fourth, state governments adopted pay revisions at different times (2009--2013), but my specification uses a single post-2008 indicator for all districts. A staggered difference-in-differences exploiting state-level adoption timing \citep{Clemens2012} would provide a more credible design, though it requires data on exact state implementation dates.

Fifth, with approximately 30 state-level clusters, standard errors from conventional cluster-robust inference may be slightly downward-biased \citep{Cameron2008}. This concern is mitigated by the direction of the finding: any downward bias would make the naive positive estimate \textit{more} significant, reinforcing the conclusion that pre-trends drive the apparent multiplier.

The methodological lesson transcends the Indian context. Dose-response designs using geographic concentration of fiscal exposure are common in the multiplier literature \citep{NakamuraSteinsson2014, Suarez2016}. When fiscal exposure correlates with structural economic characteristics---as government employment inevitably does with administrative importance and urbanization---the identifying assumption of parallel trends is unlikely to hold without careful trend adjustment. Future work on pay commission effects should seek quasi-random variation in exposure, perhaps through the geographic concentration of railway employees (28\% of central government workers) or military installations, which are allocated for logistical rather than economic reasons.

\section{Conclusion}

India's 6th Pay Commission injected billions of rupees into local economies through wage increases and arrears payments. A naive analysis suggests large local multipliers, but the apparent effect is driven by pre-existing differential trends: government-heavy districts were already growing faster. Using district-level government employment shares as exposure and controlling for differential trends, I find no evidence that the pay commission generated positive local spillovers. The result highlights the difficulty of identifying fiscal multipliers from geographic variation in government employment, which is endogenous to local economic dynamism. Whether pay commissions generate multipliers through channels not captured by nightlights or at spatial scales finer than the district remains an open question.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). All data are from SHRUG v2.1 \citep{Asher2021}, which is publicly available at \url{https://www.devdatalab.org/shrug_download/}.

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
