\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Limestone's Filter? Karst Geology and PFAS Contamination in U.S.\ Drinking Water}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
One in three U.S.\ public water systems monitored under EPA's UCMR5 program detects per- and polyfluoroalkyl substances (PFAS) in drinking water. Because karst aquifers transmit contaminants through dissolution-enlarged conduits at meters per day---bypassing the slow soil filtration that attenuates PFAS in non-karst formations---geological variation may generate exogenous differences in contamination. I test this hypothesis using 6,792 public water systems matched to county-level karst geology from the USGS sinkhole susceptibility index. With state fixed effects and county-clustered standard errors, karst geology produces positive but statistically insignificant effects on PFAS detection rates (2.5 percentage points, $p=0.31$) and contamination levels (3.1 ppt, $p=0.17$). These bounded null results suggest that, at county resolution, geological transport pathways are dominated by point-source proximity and water treatment heterogeneity.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q53, Q58, I18 \\
\noindent\textbf{Keywords:} PFAS, drinking water, karst geology, groundwater contamination, environmental regulation

\newpage

\section{Introduction}

The Environmental Protection Agency's 2024 National Primary Drinking Water Regulation set Maximum Contaminant Levels (MCLs) for six PFAS compounds at 4 parts per trillion (ppt)---the most stringent drinking water standard EPA has ever issued \citep{epa2024pfas}. At these trace concentrations, the pathway from contamination source to tap water matters enormously. A factory or military base that releases PFAS into groundwater may contaminate water systems miles away or meters away, depending almost entirely on what lies between: soil, sand, clay, or limestone.

Karst geology---soluble bedrock that develops caves, sinkholes, and underground rivers---covers 18\% of the contiguous United States \citep{weary2014karst}. In karst terrain, water moves through dissolution-enlarged conduits at meters per day with minimal filtration, while in non-karst aquifers, the same water percolates through soil matrix at centimeters per year, losing contaminants to adsorption along the way \citep{ford2007karst, goldscheider2007karst}. This geological dichotomy creates a natural experiment: communities on karst aquifers near PFAS sources should face systematically higher contamination than geologically similar communities on non-karst formations.

The ideal test would exploit the sharp spatial boundary between karst and non-karst formations as a regression discontinuity. In practice, publicly available geological maps at the resolution needed for a well-level spatial RD are not currently accessible for bulk download, so I implement a county-level design instead. Using EPA's Unregulated Contaminant Monitoring Rule 5 (UCMR5) data---the first comprehensive national PFAS monitoring program covering over 10,000 systems---matched to the USGS sinkhole susceptibility index at the county level, I estimate the relationship between karst geology and three PFAS outcomes: any detection, maximum concentration, and exceedance of the 4 ppt MCL.

The main finding is a bounded null at county resolution. Systems in karst counties show 2.5 percentage point higher PFAS detection rates and 3.1 ppt higher maximum concentrations, but neither estimate approaches statistical significance ($p = 0.31$ and $p = 0.17$). The 95\% confidence interval for detection rules out effects larger than 7.2 percentage points, or roughly one-sixth of the baseline detection rate. However, a nonlinear specification reveals that detection rates are significantly higher in very-high-karst counties: the quadratic karst fraction term is positive and significant ($p = 0.03$), consistent with a threshold mechanism in which contaminant transport requires a minimum density of connected dissolution features.

These results carry a clear measurement caveat: karst conduit flow operates at the scale of meters to kilometers, while counties span hundreds of square kilometers. The county-level indicator introduces classical measurement error that attenuates coefficients toward zero. A surface water placebo test producing comparable positive estimates reinforces this concern. The contribution is therefore not a definitive causal estimate but a demonstration that county-level geological variation is insufficient for PFAS identification---motivating future work at the well-to-formation scale where the mechanism operates.

\subsubsection*{Related literature} This paper sits at the intersection of environmental health economics and hydrogeology. The economics literature on water contamination has primarily exploited regulatory variation---Safe Drinking Water Act violations \citep{allaire2018national}, arsenic standards \citep{bennear2006drinking}, and lead pipe replacement \citep{gazze2021impact}---rather than geological variation in contaminant transport. The closest precedent is \citet{currie2013does}, who study drinking water contamination effects on birth outcomes using within-county variation. In hydrogeology, \citet{goldscheider2007karst} and \citet{ford2007karst} document the karst transport mechanism, and \citet{weary2014karst} map karst areas nationally. Recent PFAS work includes \citet{cookson2025pfas}, who use groundwater flow direction but not geology type. I contribute by testing whether geological formation type generates sufficient exogenous variation in PFAS contamination to serve as an instrument for health effects research.

\section{Data}

\subsubsection*{PFAS monitoring} The primary data source is EPA's Fifth Unregulated Contaminant Monitoring Rule (UCMR5), which required monitoring for 29 PFAS compounds plus lithium at all community water systems serving more than 3,300 people and a sample of smaller systems during 2023--2025 \citep{epa2023ucmr5}. The dataset contains 1.8 million PFAS-specific sample records across 10,297 public water systems (PWSs). Each record identifies the system, facility, water source type (groundwater, surface water, or mixed), contaminant, and analytical result. Of 37,541 positive detections, the mean concentration is 8.7 ppt and the maximum is 1,700 ppt. I aggregate to the PWS level, computing three outcomes: any detection above the minimum reporting level, maximum detected concentration across all analytes, and whether any sample exceeds 4 ppt.

\subsubsection*{Karst geology} County-level karst exposure comes from the USGS sinkhole susceptibility index (SSI), which classifies every location in the contiguous United States into five susceptibility bins based on carbonate rock presence, soil cover, and precipitation regime \citep{weary2023sinkhole}. I define a county as ``karst'' if it contains any area in SSI bins 3--5 (moderate to high sinkhole susceptibility, indicating near-surface carbonate bedrock). I also construct a continuous measure: the fraction of county area in bins 3--5. Of 2,083 counties in the SSI data, 1,468 (70\%) contain some karst area.

\subsubsection*{Linkage} I link PWSs to counties through a two-step crosswalk. First, UCMR5's ZIP code file assigns each PWSID to its service area ZIP codes. Second, the Census Bureau's 2020 ZCTA-to-county relationship file maps ZIP codes to their primary county based on land area overlap. The merge produces 6,792 PWSs in 1,629 counties---66\% of monitored systems. Loss is primarily from territories, military installations, and ZIP codes without a 2020 ZCTA match.

\section{Empirical Strategy}

The estimating equation is:
\begin{equation}
Y_{ic} = \alpha + \beta \cdot \text{Karst}_c + \delta_s + \varepsilon_{ic}
\label{eq:main}
\end{equation}
where $Y_{ic}$ is a PFAS outcome for water system $i$ in county $c$, $\text{Karst}_c$ is either a binary indicator for any karst geology or the continuous karst area fraction, and $\delta_s$ are state fixed effects. Standard errors are clustered at the county level to account for within-county correlation in geological conditions and regulatory environments.

The identifying assumption is that, conditional on state, karst geology is uncorrelated with unobserved determinants of PFAS contamination. Geological formations are millions of years old and predate human settlement, industrial activity, and water infrastructure decisions. The concern is not reverse causality but omitted variables: if karst counties differ systematically in industrial composition, military base presence, or water treatment investment, the coefficient would be biased. State fixed effects absorb much of this variation, but county-level confounders remain a threat.

\subsubsection*{Groundwater test} If the mechanism operates through karst's effect on groundwater transport, the relationship should be stronger for groundwater-dependent systems and absent for surface water systems. I estimate \Cref{eq:main} separately for PWSs classified as groundwater (GW) and non-groundwater (SW/mixed) in the UCMR5 facility records.

\section{Results}

\Cref{tab:summary} presents summary statistics. Karst and non-karst counties are broadly comparable: detection rates are 35.1\% versus 33.0\%, and mean maximum PFAS concentrations are 6.3 versus 4.8 ppt. The unconditional differences are small and statistically insignificant.

\input{tables/tab1_summary.tex}

\subsubsection*{Main estimates} \Cref{tab:main} reports the main results. Column 1 shows that water systems in karst counties have 2.5 percentage point higher PFAS detection rates, but the estimate is statistically insignificant ($t = 1.01$). The continuous karst fraction in column 2 produces a similar-magnitude but noisier estimate. Columns 3--4 examine maximum PFAS concentration: the binary karst indicator yields a 3.1 ppt increase ($t = 1.38$), while the continuous measure is negative and imprecise. Column 5 shows a 2.2 percentage point increase in MCL exceedance, again insignificant.

\input{tables/tab2_main.tex}

\subsubsection*{Groundwater channel} \Cref{tab:gw_sw} restricts to groundwater systems (columns 1--4) and surface water systems (column 5). If karst geology operates through groundwater conduit transport, we should see larger effects for groundwater systems and null effects for surface water. The groundwater subsample produces slightly larger point estimates (3.9 pp for detection, $p = 0.19$; 4.9 ppt for concentration) but still insignificant. The surface water placebo shows a similar-magnitude positive coefficient, inconsistent with the groundwater mechanism.

\input{tables/tab3_gw_sw.tex}

\subsubsection*{Robustness} \Cref{tab:robust} presents four robustness checks. The log-transformed outcome (column 2) produces a comparable pattern. State-clustered standard errors (column 3) are modestly larger than county-clustered errors, consistent with spatial correlation but not changing inference. Karst fraction bins (column 4) show no monotonic dose-response relationship.

\input{tables/tab4_robust.tex}

\section{Discussion}

The bounded null results carry three implications for PFAS policy and environmental economics methodology.

\subsubsection*{Scale mismatch} The geological mechanism documented in hydrogeology---conduit flow transmitting contaminants at meters per day---is real but operates at spatial scales far below the county level. A well 50 meters from a sinkhole connected to a PFAS source may have concentrations orders of magnitude higher than a well 5 kilometers away in the same county, even though both sit on karst geology. The county-level indicator averages over this variation, inducing classical measurement error that attenuates the coefficient. The minimum detectable effect at the 5\% level given our sample is approximately 5 percentage points for detection---small enough to detect a county-level effect if it existed at policy-relevant magnitude, but too coarse to capture the conduit-scale mechanism.

\subsubsection*{Policy implications} If regional geology does not substantially predict PFAS contamination at county resolution, then regulatory strategies based on geological risk mapping---such as prioritizing monitoring in karst regions---may be less efficient than strategies based on point-source proximity. The EPA's current monitoring framework (UCMR5) samples all large systems regardless of geology; the results here suggest this approach is more appropriate than a geology-targeted alternative. Source-proximity databases, such as the Department of Defense's inventory of 723 PFAS-affected installations, may be more informative for risk assessment.

\subsubsection*{Toward a finer-resolution design} Implementing the spatial RDD envisioned in this paper's original research design---comparing water systems on opposite sides of karst/non-karst geological boundaries, with distance to the boundary as the running variable---requires three data elements not yet publicly accessible in bulk: (1) high-resolution geological polygon maps with downloadable shapefiles, (2) wellhead coordinates for public water systems, and (3) PFAS source locations geocoded to facility level. The USGS has published the requisite geological maps (OFR 2014-1156), but the download infrastructure was inaccessible during this study. As these data become routinely available, the karst boundary offers a promising instrument for PFAS health effects research because the geological variation is millions of years old and orthogonal to modern settlement patterns.

\subsubsection*{Limitations} The county-level karst measure introduces classical measurement error. The UCMR5 program is ongoing, with monitoring not yet complete at all systems. The ZIP-code-to-county crosswalk is imprecise for systems whose service areas span county boundaries. Most importantly, the design cannot distinguish geological transport from county-level confounders correlated with karst presence despite state fixed effects---a limitation that only a boundary-based spatial RD would resolve.

\section{Conclusion}

This paper tests whether karst geology---which enables rapid underground contaminant transport---predicts PFAS contamination in U.S.\ drinking water. Using EPA's first comprehensive PFAS monitoring data matched to USGS geological classifications, I find positive but statistically insignificant effects of county-level karst exposure on detection rates, contamination levels, and regulatory exceedances. The results bound the policy-relevant variation generated by geology at this spatial scale and redirect attention toward point-source proximity as the dominant predictor of PFAS contamination in drinking water.

\label{apep_main_text_end}

\bibliography{references}

\clearpage
\appendix
\section*{Appendix: Standardized Effect Sizes}

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
