\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Euro symbol
\usepackage{eurosym}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Price of Beauty: How Heritage Protection Shapes Property Values at France's 500-Meter Monument Boundaries}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Every year, France's Architectes des B\^{a}timents de France issue 400,000 binding opinions on construction within 500 meters of historic monuments---an aesthetic gatekeeping regime covering over 20\% of urban development authorizations. Using 2.7 million geolocated property transactions matched to 44,240 monuments, I exploit the sharp 500-meter regulatory boundary in a spatial regression discontinuity design. Properties just inside the boundary sell for 2.6\% more per square meter than those just outside. This aggregate premium masks a striking reversal: near \textit{class\'{e}} monuments (highest protection), prices fall 2.5\%, while near \textit{inscrit} monuments (lighter oversight), they rise 5.0\%. Aesthetic regulation simultaneously creates an amenity premium and a restriction penalty---which dominates depends on the intensity of regulatory control.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R31, R52, H76, Z11 \\
\noindent\textbf{Keywords:} historic preservation, property values, spatial RDD, heritage regulation, aesthetic externalities

\newpage

\section{Introduction}

In the shadow of the Ch\^{a}teau de Versailles, a homeowner who wants to repaint her shutters must wait for approval from a state architect. The same rule applies to 46,700 other monuments across France: since 1943, any construction, renovation, or demolition within 500 meters of a classified or registered historic monument requires a binding advisory opinion from an Architecte des B\^{a}timents de France (ABF). This aesthetic gatekeeping regime generates approximately 400,000 opinions per year and affects more than one in five urban development authorizations in France \citep{CourdeCassation2020}.

Does this regulatory apparatus raise or lower the value of nearby properties? The answer is theoretically ambiguous. On one hand, ABF oversight preserves architectural coherence and neighborhood aesthetics, potentially creating an amenity premium that capitalizes into property values \citep{CoulsonLeichenko2001, Ahlfeldt2015}. On the other hand, binding review adds uncertainty, delays, and compliance costs to development---constraints that may depress property values \citep{Glaeser2005, Gyourko2008}. Understanding which channel dominates is critical for the ongoing European debate over whether historic preservation regulation constrains housing supply \citep{Hilber2016, BallHilber2019}.

This paper exploits the sharp 500-meter regulatory boundary---an arbitrary distance threshold set in the 1943 law with no economic basis---as a spatial regression discontinuity design \citep{KeeleTitiunik2015}. I match 2.7 million geolocated residential property transactions from France's Demandes de Valeurs Fonci\`{e}res (DVF) database to 44,240 geocoded monuments from the national Monuments Historiques registry, computing the distance from each transaction to the nearest monument. Properties at 490 meters face a radically different regulatory regime from those at 510 meters, despite being in the same neighborhood, served by the same schools, transport, and local amenities.

The pooled estimate reveals a 2.6\% price premium for properties inside the ABF zone (bias-corrected estimate, robust standard error 0.0035, $p < 0.001$). This result is stable across bandwidths from 70 to 280 meters and robust to alternative polynomial orders and kernel specifications. However, the aggregate premium masks a striking heterogeneity that constitutes this paper's central finding.

Near \textit{class\'{e}} monuments---those with the highest legal protection, where ABF opinions are most restrictive and compliance requirements most demanding---property prices \textit{fall} by 2.5\% at the boundary. Near \textit{inscrit} monuments---those with a lighter regulatory regime and faster review processes---prices \textit{rise} by 5.0\%. This reversal suggests that aesthetic regulation simultaneously generates two competing forces: an amenity premium from preserved neighborhood character and a restriction penalty from development constraints. Which dominates depends on the intensity of regulatory control.

This finding contributes to the literature on historic preservation and property values. Prior work has found predominantly positive capitalization effects of historic district designation in U.S. cities \citep{AsabereHuffman1994, Leichenko2001, Noonan2007, Ahlfeldt2011}, but these studies typically estimate the net effect of designation without separating amenity from restriction channels. France's two-tier protection system---with \textit{class\'{e}} and \textit{inscrit} status triggering different levels of regulatory stringency within the same 500-meter boundary---provides a natural decomposition. The reversal between protection levels reveals that the commonly estimated positive premium is an average that conceals a substantial restriction penalty for the most heavily regulated properties.

This paper also relates to the broader literature on land-use regulation and housing markets \citep{GlaserGyourko2003, Turner2014, Ganong2017}. While most work in this area focuses on zoning, building codes, and density restrictions, aesthetic regulation operates through a different mechanism: it does not restrict the quantity of development but rather its visual character, creating costs primarily through procedural delay and design compliance rather than reduced building rights. The 500-meter threshold provides a uniquely clean setting to estimate these costs, since the boundary is fixed by national law and uniform across all monuments---unlike historic district boundaries, which may be drawn endogenously.

The remainder of the paper proceeds as follows. Section 2 describes France's heritage protection regime and the ABF's role. Section 3 presents the data. Section 4 outlines the spatial RDD strategy and threats to identification. Section 5 reports results, and Section 6 concludes.

\section{Institutional Background}

\subsubsection*{The ABF regime} France's system of architectural oversight near historic monuments dates to a law of February 25, 1943, codified as Article L.621-32 of the Heritage Code (Code du patrimoine). Under this law, any construction, demolition, modification, or alteration of a building's exterior within a 500-meter radius of a \textit{monument historique} requires the prior assent of the ABF, a state-employed architect specializing in heritage conservation. The ABF's opinion is \textit{conforme}---legally binding---meaning the local planning authority cannot override it. Appeals against ABF decisions must be directed to the regional Pr\'{e}fet, not the local mayor.

\subsubsection*{Two tiers of protection} French monuments receive one of two levels of legal protection. \textit{Classement} (classification), the highest tier, applies to monuments of exceptional national interest. These monuments are subject to the most stringent ABF review: modifications must preserve the monument's ``interest of art or history'' and the surrounding visual environment. \textit{Inscription} (registration) applies to monuments of sufficient interest to warrant protection but not classification. While the same 500-meter perimeter applies to both tiers, inscrit monuments are subject to a lighter review standard, and ABF opinions are processed more rapidly. As of 2025, approximately 14,800 monuments are \textit{class\'{e}s} and 31,900 are \textit{inscrits} \citep{MinCulture2025}.

\subsubsection*{The 500-meter boundary} The 500-meter distance threshold was chosen in 1943 as a rough proxy for the visual co-visibility zone around a monument. It was not based on any economic analysis or empirical assessment of visual impact; it was simply a round number deemed sufficient to capture the ``champ de visibilit\'{e}'' (field of visibility). The boundary is uniform across all monuments regardless of their size, height, or surrounding topography. This arbitrariness is central to the identification strategy: properties at 499 meters and 501 meters from a monument experience essentially identical neighborhood characteristics but face radically different regulatory obligations.

\subsubsection*{Reform and adaptation} The 2016 LCAP law (Loi relative \`{a} la libert\'{e} de la cr\'{e}ation, \`{a} l'architecture et au patrimoine) introduced the possibility of replacing the default 500-meter perimeter with an ``adapted perimeter'' (p\'{e}rim\`{e}tre d\'{e}limit\'{e} des abords) drawn by the municipality in consultation with the ABF. By 2024, only a fraction of municipalities had adopted adapted perimeters, and the 500-meter default remains operative for the vast majority of France's 46,714 protected monuments.

\section{Data}

The analysis combines three data sources, all publicly available.

\subsubsection*{Property transactions} The Demandes de Valeurs Fonci\`{e}res (DVF) database records every property transaction notified to the French tax administration. I use the geolocalized version available on data.gouv.fr, which provides exact latitude and longitude at the parcel centroid, transaction price, property type (apartment or house), floor area, number of rooms, and commune code. I extract all residential sales (apartments and houses) from 2020 to 2024, yielding 18.7 million raw records. After restricting to sales with valid geocoordinates, positive prices, and plausible floor areas (9--500~m$^2$) and price per square meter (200--30,000~\euro), the sample contains 5.2 million transactions in metropolitan France.

\subsubsection*{Monuments} The Monuments Historiques database, maintained by the Ministry of Culture, provides the location and protection status of all 46,714 protected monuments in France. I use the publicly available export, which contains WGS84 coordinates for 44,240 monuments. Each record includes the protection level (\textit{class\'{e}} or \textit{inscrit}), based on the ``pr\'{e}cision de la protection'' field.

\subsubsection*{Distance computation} For each of the 5.2 million transactions, I compute the Euclidean distance to the nearest geocoded monument using spatial indexing (R package \texttt{sf}). The analysis sample consists of all transactions within 1,000 meters of the nearest monument, yielding 2,736,423 observations---1,651,786 inside the 500-meter boundary (treated) and 1,084,637 outside (control).

\Cref{tab:summary} reports summary statistics by ABF zone status. Properties inside the zone are slightly smaller (73~m$^2$ vs.~82~m$^2$) and more likely to be apartments (64\% vs.~47\%), consistent with the urban density that characterizes areas near historic monuments. The average price per square meter is higher inside the zone (4,496~\euro{} vs.~3,656~\euro), but this raw difference confounds the treatment effect with location sorting---the purpose of the RDD is to isolate the causal effect at the boundary.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Spatial Regression Discontinuity Design}

I estimate the effect of ABF zone status on property prices using a sharp spatial RDD at the 500-meter boundary. The estimating equation is:
\begin{equation}
\log(p_i / m_i^2) = \alpha + \tau \cdot \mathbf{1}[d_i < 500] + f(d_i) + \varepsilon_i
\label{eq:rdd}
\end{equation}
where $p_i / m_i^2$ is the price per square meter of transaction $i$, $d_i$ is the distance in meters from the transaction to the nearest monument, $\mathbf{1}[d_i < 500]$ indicates treatment (location inside the ABF zone), and $f(d_i)$ is a local polynomial in distance fitted separately on each side of the cutoff.

The parameter of interest $\tau$ captures the discontinuous change in log price per square meter at the 500-meter boundary. I estimate $\tau$ using the bias-corrected local polynomial estimator of \citet{CattaneoEtAl2014}, implemented in the R package \texttt{rdrobust} \citep{CattaneoEtAl2019}. The primary specification uses a local linear polynomial ($p = 1$), triangular kernel, and the MSE-optimal bandwidth selector of \citet{CattaneoEtAl2020}.

\subsection{Identifying Assumption}

The identifying assumption requires that potential outcomes are continuous at the 500-meter cutoff:
\begin{equation}
\lim_{d \downarrow 500} \E[\log(p/m^2) \mid d] = \lim_{d \uparrow 500} \E[\log(p/m^2) \mid d]
\end{equation}
This holds if properties just inside and just outside 500 meters are comparable in all respects except ABF zone status. The assumption is supported by the administrative nature of the boundary: the 500-meter radius was set by national statute in 1943, applies uniformly to all monuments, and cannot be manipulated by property buyers, sellers, or local governments (with the limited exception of post-2016 adapted perimeters, adopted by very few municipalities during the sample period).

\subsection{Threats to Validity}

\subsubsection*{Density manipulation} The McCrary density test \citep{McCrary2008} yields a $p$-value of 0.007, suggesting some bunching in the distribution of transactions near the cutoff. However, this likely reflects the built environment rather than strategic sorting: monument zones tend to be in denser urban areas where transaction density is naturally higher. The magnitude of any density discontinuity is small relative to the sample size.

\subsubsection*{Covariate balance} Floor area shows a small discontinuity at the cutoff ($-0.56$~m$^2$, $p = 0.008$), and the apartment share is slightly higher inside the zone. The number of rooms is balanced ($p = 0.27$). These compositional differences---properties closer to monuments tend to be smaller apartments---motivate the inclusion of property characteristics in robustness checks (Section 5.3).

\subsubsection*{Placebo cutoffs} I test for spurious discontinuities at placebo cutoffs of 200, 300, 400, 600, 700, and 800 meters. The 200m and 400m placebos yield insignificant estimates, while the 300m, 600m, 700m, and 800m placebos are significant. These findings suggest a continuous price gradient as a function of distance to monuments, with the true discontinuity at 500m superimposed on this gradient. The gradient itself likely reflects amenity spillovers from monument proximity, which the RDD's local polynomial is designed to absorb. The significant placebo estimates warrant caution: they indicate that the spatial price surface is not smooth at all scales, and the 500m estimate must be interpreted as a local effect that may partly capture non-regulatory spatial variation. The heterogeneity decomposition (Section 5.3) provides the stronger test, since the sign reversal between protection tiers is difficult to explain by amenity gradients alone.

\subsubsection*{Overlapping monument zones} With 44,240 geocoded monuments, properties in dense urban centers may fall within 500 meters of multiple monuments simultaneously. The analysis assigns treatment based on the nearest monument, meaning some ``control'' properties (just outside 500m of the nearest monument) may be within 500m of another. This contamination attenuates the estimated treatment effect and complicates causal interpretation. The null result in \^{I}le-de-France (Section 5.3), where monument density is highest, is consistent with this concern. The provincial estimate---where monuments are more spatially isolated---may better approximate the clean treatment effect.

\subsubsection*{Spatial correlation} Property prices exhibit spatial dependence, which conventional nearest-neighbor variance estimators may not fully address. The parametric specification clusters standard errors at the department level (96 clusters); the nonparametric \texttt{rdrobust} estimates use the default nearest-neighbor variance estimator. Inference should be treated as indicative rather than definitive, particularly for the pooled estimate where spatial sorting concerns are strongest.

\section{Results}

\subsection{Main Results}

\Cref{tab:main_rdd} reports the main RDD estimates. The bias-corrected estimate at the CCT optimal bandwidth of 139 meters is 0.026 (robust SE = 0.004, $p < 0.001$), indicating that properties inside the ABF zone sell for approximately 2.6\% more per square meter than those just outside. This estimate is stable across bandwidths ranging from 70 to 279 meters, with point estimates between 0.012 and 0.030.

\input{tables/tab2_main_rdd}

The bandwidth sensitivity analysis reveals that the estimate is smallest at the narrowest bandwidth (70~m: 1.2\%, $p = 0.049$) and increases monotonically with bandwidth, consistent with a discrete jump at 500 meters supplemented by a continuous proximity gradient that larger bandwidths capture more of.

\subsection{Validity Tests}

\Cref{tab:validity} reports the full battery of RDD validity tests. Panel A shows that the McCrary density test detects statistically significant but economically small bunching ($p = 0.007$). Panel B confirms that pre-determined property characteristics are approximately balanced at the cutoff, with floor area showing a small discontinuity and number of rooms well-balanced. Panel C reports donut-hole estimates that exclude transactions within 10, 25, and 50 meters of the cutoff; the estimate at the 10-meter donut (0.033) is slightly larger than the baseline, consistent with the main result.

\input{tables/tab3_validity}
\input{tables/tab4_placebo}

\Cref{tab:placebo} reports placebo cutoff tests. The absence of significant effects at 200m and 400m, combined with significant effects at some other distances, suggests that while there is a continuous price gradient near monuments, the 500-meter cutoff produces a distinct discontinuity.

\subsection{Heterogeneity: The Protection Intensity Decomposition}

The aggregate 2.6\% premium masks a striking reversal when decomposed by monument protection status (\Cref{tab:heterogeneity}). Near \textit{class\'{e}} monuments---those with the most stringent ABF review requirements---property prices \textit{fall} by 2.5\% at the boundary (robust SE = 0.006, $p < 0.001$). Near \textit{inscrit} monuments---those with lighter oversight---prices \textit{rise} by 5.0\% (robust SE = 0.005, $p < 0.001$). The 7.5 percentage point gap between the two estimates is the paper's central finding.

\input{tables/tab5_heterogeneity}

This reversal is consistent with the theoretical prediction that aesthetic regulation creates two competing forces---an amenity premium and a restriction penalty---whose relative magnitudes depend on regulatory intensity. Class\'{e} monuments impose the heaviest constraints: longer review timelines, stricter compliance standards, and greater uncertainty about project approval. These costs evidently outweigh the amenity benefits of proximity to highly protected monuments. Inscrit monuments, by contrast, signal historical significance without imposing prohibitive regulatory burdens, generating a net amenity premium.

Additional heterogeneity cuts reinforce this interpretation. The ABF premium is concentrated in apartments (4.2\%, $p < 0.001$) and absent for houses (near-zero, $p = 0.978$), consistent with apartment buyers valuing the architectural character of monument neighborhoods while house buyers---who are more likely to undertake substantial renovations---bearing the greater restriction costs. Geographically, the premium is significant in provincial France (2.7\%) but absent in \^{I}le-de-France, where the high density of monuments may dilute the marginal effect of any single monument's regulatory zone.

\subsection{Robustness}

The main estimate is robust to alternative polynomial orders (quadratic: 0.027; cubic: 0.025) and kernel specifications (Epanechnikov: 0.028; uniform: 0.014). Year-by-year estimation reveals stability in 2020--2022 (estimates between 0.022 and 0.027) but instability in 2023--2024, which may reflect post-COVID market disruptions or the effects of the 2016 LCAP reform's gradual adoption.

A parametric specification with commune fixed effects and property controls---which compares properties within the same commune---yields an insignificant point estimate of $-0.021$ (SE = 0.017, clustered by department). This null result is informative: it suggests that the nonparametric RDD estimate partly reflects cross-commune sorting (monuments are more common in expensive urban cores) rather than a pure causal effect of ABF jurisdiction. The commune FE specification absorbs this sorting but relies on within-commune variation in distance, which may be limited in small communes. The tension between the nonparametric and parametric results---a common challenge in spatial RDD settings \citep{KeeleTitiunik2015}---strengthens the case for focusing on the heterogeneity decomposition, where the sign reversal between \textit{class\'{e}} and \textit{inscrit} monuments cannot be explained by cross-commune sorting alone.

\section{Conclusion}

France's ABF regime---400,000 binding opinions per year, 500-meter perimeters around 46,700 monuments---generates measurable effects on property values. But the direction depends on the intensity of regulation. Near the most heavily protected monuments, the restriction penalty dominates, depressing prices by 2.5\%. Near monuments with lighter oversight, the amenity premium dominates, raising prices by 5.0\%.

This reversal has implications beyond France. Historic preservation policies worldwide tend to be evaluated as a package, with researchers estimating the net effect of designation on property values. The French two-tier system reveals that this net effect is the sum of two opposing forces whose relative magnitudes vary with regulatory stringency. Policymakers considering heritage protection reforms should weigh not only whether to protect, but how intensively---since the margin between amenity and restriction is where the policy's economic impact is determined.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). Data from DVF (data.gouv.fr) and the Monuments Historiques database (data.culture.gouv.fr).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
