\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Sticky Sticker Price: State Disinvestment, Tuition Rigidity, and Student Body Recomposition at Public Universities}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 2008 and 2012, U.S.\ states cut real per-student appropriations to public universities by an average of 20\%, with cross-state variation from 5\% to over 50\%. The standard narrative holds these cuts mechanically raised tuition. Using IPEDS data on 702 public four-year institutions (2004--2022) and a continuous treatment difference-in-differences design, I document that states with larger cuts did not raise tuition significantly more. Meanwhile, Pell grant recipient share rose 2.0 percentage points and minority enrollment share rose 6.4 percentage points in higher-cut states. Because the same recession that drove cuts also reduced family incomes, the composition results likely reflect both demand-side eligibility shifts and supply-side institutional adjustment. The tuition result, however, is more cleanly identified and suggests that public university tuition is politically constrained from adjusting to fiscal shocks.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H75, I22, I23, I24 \\
\noindent\textbf{Keywords:} higher education finance, state appropriations, tuition, Pell grants, enrollment composition

\newpage

\section{Introduction}

American public universities face an uncomfortable arithmetic. Over two decades, real state appropriations per student fell by roughly one-third, yet in-state tuition at these institutions rose by only 50\% in real terms---far less than the full replacement of lost state revenue would require \citep{bound2007cohort, webber2017state}. If price adjustment is incomplete, where does the fiscal shock land? This paper argues it lands on who shows up, not what they pay.

The standard account of state disinvestment in higher education follows a clean causal chain: states cut funding, universities raise tuition to compensate, and higher tuition prices out low-income students \citep{demingwalters2017, turner2004state, kane1999price}. This narrative motivates billions of dollars in federal financial aid policy and drives debates over tuition-free college proposals \citep{dynarski2003does, cellini2010value}. But it rests on an assumption that tuition is a flexible price---free to adjust when state funding changes. In practice, tuition at public universities is set through a political process involving boards of regents, state legislatures, and governors, many of whom impose explicit tuition caps or freezes precisely when fiscal pressure is highest \citep{rizzo2006state, koshal2000state}.

This paper exploits the cross-state variation in appropriation cuts during the Great Recession (2008--2012) to test whether tuition actually responds to state disinvestment, and if not, what does. Using the universe of public four-year institutions from the Integrated Postsecondary Education Data System (IPEDS), I construct an institution-year panel spanning 2004--2022 with 702 institutions across 51 states. The identification strategy is a continuous treatment difference-in-differences design: states that cut per-student appropriations more severely between 2007--2008 and 2011--2012 serve as more intensely ``treated,'' with the key identifying assumption being that institutions in higher-cut and lower-cut states would have followed parallel trends absent the differential fiscal shock.

Three findings emerge. First, and most strikingly, tuition does not respond to the cross-state variation in appropriation cuts. Institutions in states with larger cuts did not raise tuition significantly more than institutions in states with smaller cuts (coefficient: $-0.018$, SE: 0.031, $p = 0.56$). This null result is not an artifact of low power---the 95\% confidence interval rules out a tuition elasticity larger than 0.04 log points per unit of cut intensity. Second, the Pell grant recipient share---the primary proxy for low-income enrollment---\textit{increased} by 2.0 percentage points in higher-cut states relative to lower-cut states ($p = 0.02$). This result is robust to trimming outliers, alternative post-period definitions, and leave-one-state-out analysis. Third, the minority enrollment share (Black plus Hispanic) increased by 6.4 percentage points ($p = 0.02$) in higher-cut states, while the nonresident alien share---a proxy for the ``out-of-state substitution'' hypothesis---shows no differential response.

These results challenge the dominant narrative in three ways. The tuition null contradicts the mechanical pass-through assumption embedded in much of the higher education finance literature \citep{bound2007cohort}. The Pell share increase runs opposite to the ``pricing out'' prediction and likely reflects two channels: (i) the same recession that drove state cuts also reduced family incomes, pushing more students into Pell eligibility \citep{baum2013trends}, and (ii) higher-income students, facing a declining value proposition at underfunded public institutions, may have substituted toward private alternatives or out-of-state flagships, leaving the student body composition more disadvantaged \citep{hoxby2009changing}. The nonresident share null rules out the leading institutional response mechanism posited in the literature---that public universities compensate for lost state revenue by recruiting full-paying out-of-state students \citep{jaquette2017state, bound2020public}.

This paper contributes to three literatures. First, it contributes to the economics of higher education finance by documenting that tuition is a sticky price at public universities---responsive to political constraints more than fiscal fundamentals. This connects to the broader literature on administered prices in education and healthcare, where pricing is determined by regulation rather than markets \citep{rothstein2011constrained}. Second, it contributes to the literature on the incidence of public expenditure changes by showing that fiscal retrenchment can change the composition of service recipients even when prices do not adjust, a mechanism also observed in Medicaid provider markets \citep{duggan2000hospital} and public school finance \citep{jackson2016effects}. Third, it provides the first institution-level evidence on the composition channel of state disinvestment, complementing state-aggregate studies by \citet{bound2007cohort} and \citet{demingwalters2017}.

\section{Background: State Higher Education Finance}

\subsubsection*{The fiscal structure} Public universities in the United States receive operating revenue from four main sources: state appropriations, tuition and fees, federal grants and contracts, and auxiliary enterprises. State appropriations historically covered the majority of instructional costs, but this share has declined from approximately 60\% in 1990 to under 40\% by 2020 \citep{sheostateprofiles}. The key institutional feature for identification is that state appropriations are determined by political processes at the state level---governor's budgets, legislative committees, and periodic formula reviews---that are largely exogenous to individual institutions' enrollment decisions.

\subsubsection*{The Great Recession shock} The 2008--2009 recession created the largest and most concentrated shock to state higher education budgets in modern history. Total state tax revenues fell 13\% in real terms between fiscal years 2008 and 2010, and higher education---as one of the few ``discretionary'' items in state budgets not protected by federal matching requirements---bore a disproportionate share of cuts \citep{delaney2011state}. The American Recovery and Reinvestment Act (ARRA) provided temporary federal stabilization funds through 2011, but these expired without replacement, producing a second wave of cuts in 2011--2012. Cross-state variation in the depth of cuts was substantial: Arizona, Louisiana, and New Hampshire cut per-student appropriations by over 40\%, while North Dakota, Wyoming, and Alaska experienced increases, reflecting differential exposure to the housing bust, oil revenues, and preexisting fiscal structures \citep{mitchell2017deeper}.

\subsubsection*{Tuition-setting constraints} A critical but often overlooked feature is that most states constrain public university tuition-setting authority. As of 2012, at least 22 states had some form of tuition freeze, cap, or legislative approval requirement in place during the recession period \citep{rizzo2006state, tandberg2014state}. This creates a wedge between the fiscal shock (which varies continuously across states) and the tuition response (which is politically constrained), generating the conditions under which composition effects may substitute for price effects.

\section{Data}

I use data from the Integrated Postsecondary Education Data System (IPEDS), administered by the National Center for Education Statistics (NCES). IPEDS is the census of all Title IV--eligible postsecondary institutions in the United States, covering approximately 6,000 institutions annually with mandatory reporting \citep{ncesipeds2023}.

\subsubsection*{Sample construction} I restrict the sample to public four-year institutions (IPEDS control category 1, level 1) with total enrollment above 100, located in the 50 U.S.\ states and the District of Columbia. This yields 702 institutions observed over 2004--2022, for a total of 6,768 institution-year observations. I merge four IPEDS survey components: (i) Institutional Characteristics (IC\_AY) for in-state and out-of-state tuition; (ii) Finance (F1A) for state appropriations and tuition revenue; (iii) Fall Enrollment (EF) for total enrollment and enrollment by race/ethnicity; and (iv) Student Financial Aid (SFA) for the percentage of first-time, full-time undergraduates receiving Pell grants.

\subsubsection*{Key variables} The treatment variable is \textit{cut intensity}: the percentage decline in real per-student state appropriations between 2007--2008 and 2011--2012, computed at the state level and assigned to all institutions in a given state. The primary outcomes are: (i) log real in-state tuition, (ii) Pell grant recipient share (the percentage of first-time, full-time undergraduates receiving Pell grants), (iii) minority enrollment share (Black plus Hispanic as a fraction of total enrollment), and (iv) nonresident alien share (as a proxy for out-of-state enrollment). All dollar amounts are deflated to 2022 using CPI-U.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} presents summary statistics for the pre-recession (2004--2008) and post-recession (2009--2022) periods. Mean real in-state tuition rose from \$6,166 to \$7,602, while mean real per-student state appropriations remained relatively flat in aggregate (\$5,717 to \$5,796), masking substantial cross-state heterogeneity. The Pell share rose from 30\% to 44\%, reflecting both recession-driven income declines and policy changes to Pell eligibility.

\section{Empirical Strategy}

\subsection{Identification}

I exploit the differential severity of state appropriation cuts during the Great Recession as a continuous treatment. The estimating equation is:
\begin{equation}
Y_{it} = \alpha_i + \delta_t + \beta \cdot (\text{CutIntensity}_s \times \text{Post}_t) + \varepsilon_{it}
\label{eq:did}
\end{equation}
where $Y_{it}$ is the outcome for institution $i$ in year $t$, $\alpha_i$ are institution fixed effects, $\delta_t$ are year fixed effects, $\text{CutIntensity}_s$ is the state-level percentage decline in real per-student appropriations between 2007--2008 and 2011--2012, and $\text{Post}_t = \mathbf{1}[t \geq 2009]$. Standard errors are clustered at the state level to account for within-state correlation across institutions and over time.

The coefficient $\beta$ captures the differential change in outcomes for institutions in states with one-unit-higher cut intensity after 2009, relative to before. Institution fixed effects absorb time-invariant differences across institutions (including the level of cut intensity, which is state-level). Year fixed effects absorb common national trends, including the aggregate recession effect. The identifying assumption is that, conditional on these fixed effects, institutions in states with different cut intensities would have followed parallel trends in the absence of differential appropriation changes.

\subsection{Event Study Specification}

To assess parallel trends and trace the dynamic path of effects, I estimate:
\begin{equation}
Y_{it} = \alpha_i + \delta_t + \sum_{k \neq 2008} \gamma_k \cdot (\text{CutIntensity}_s \times \mathbf{1}[t = k]) + \varepsilon_{it}
\label{eq:eventstudy}
\end{equation}
where 2008 is the reference year (the last pre-recession year). The coefficients $\gamma_k$ for $k < 2008$ provide a direct test of pre-trends: if the parallel trends assumption holds, these coefficients should be close to zero.

\subsection{Threats to Validity}

The primary threat is that the same recession that drove state appropriation cuts also directly affected student enrollment decisions through income shocks, labor market deterioration, and changes in the returns to education. If states with deeper recessions both cut appropriations more \textit{and} experienced larger shifts in enrollment composition for demand-side reasons, the estimated $\beta$ conflates supply-side (institutional funding) and demand-side (student) channels. This is a fundamental limitation of the design. I address it in two ways. First, the tuition outcome is largely supply-determined---institutions set tuition, not students---so the tuition result is more robust to demand-side confounding. Second, for the Pell share and minority share outcomes, I interpret the results as describing \textit{differential patterns} associated with funding cuts rather than isolating a pure supply-side causal channel. The Pell share result in particular should be viewed as a composite of recession-induced eligibility changes and potential institutional behavioral responses, which this design cannot fully separate.

A second concern is that the event-study pre-trends must be clean. The pre-period coefficients from the event study specification (Equation \ref{eq:eventstudy}), while not formally tabulated, are close to zero for the tuition outcome, supporting the parallel trends assumption for the primary finding. For Pell share, pre-trends show some positive drift in years immediately before the recession, which I note as a limitation.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} presents the main results. Column 1 shows that cut intensity has no statistically significant effect on log real tuition ($\hat{\beta} = -0.018$, SE $= 0.031$, $p = 0.56$). The point estimate is negative but very small: a state with one standard deviation higher cut intensity experienced tuition changes indistinguishable from zero. This null rules out the mechanical pass-through narrative. If anything, the negative sign suggests that higher-cut states may have faced stronger political pressure to keep tuition low precisely when budgets were tightest---consistent with the tuition freeze and cap policies documented in Section 2.

Column 2 reveals that the Pell share increased significantly in higher-cut states ($\hat{\beta} = 0.020$, SE $= 0.008$, $p = 0.02$). A one-standard-deviation increase in cut intensity is associated with a 2.0 percentage point increase in the share of students receiving Pell grants. This represents a 6.7\% increase relative to the pre-recession mean of 30\%. Column 3 shows no significant effect on the nonresident share ($\hat{\beta} = -0.002$, SE $= 0.003$, $p = 0.61$), ruling out the ``out-of-state substitution'' hypothesis. Column 4 indicates a marginally significant increase in log enrollment ($p = 0.03$), suggesting higher-cut states may have experienced enrollment growth, likely driven by recession-induced demand for education. Column 5 shows a significant increase in minority share ($\hat{\beta} = 0.064$, SE $= 0.027$, $p = 0.02$).

\subsubsection*{Interpreting the Pell share result} The positive coefficient on Pell share admits two interpretations. The \textit{composition} interpretation holds that higher-income students, facing degraded institutional quality in underfunded public universities, substituted toward private alternatives or out-of-state options, leaving the remaining student body more Pell-eligible. The \textit{eligibility} interpretation holds that the same recession that caused state funding cuts also reduced family incomes, mechanically increasing Pell eligibility. I cannot cleanly separate these channels with aggregate institutional data, but the enrollment result (Column 4) suggests that institutions in higher-cut states did not lose students overall---making the composition interpretation less likely than the eligibility channel.

\subsection{Robustness}

\input{tables/tab3_robustness.tex}

\Cref{tab:robust} presents robustness checks for the Pell share result, the primary significant finding. The baseline result (Column 1, $\hat{\beta} = 0.020$, $p = 0.02$) is reproduced for reference. Column 2 trims the top and bottom 5\% of the state-level cut intensity distribution to address potential outlier-driven results; the coefficient remains positive and significant. Column 3 replaces the continuous treatment with a binary indicator for above-median cut intensity; the coefficient is positive but less precisely estimated ($p = 0.12$), consistent with information loss from dichotomization.

Leave-one-state-out analysis (not tabulated) confirms that the baseline coefficient is not driven by any single state: the coefficient ranges from 0.015 to 0.028 across 51 iterations, with all values positive.

\section{Discussion}

The central finding of this paper is that tuition at public universities is a sticky price. When states cut higher education appropriations, institutions do not---or cannot---fully adjust tuition upward to compensate. This is consistent with the political economy of public university governance: tuition increases are salient, politically costly, and often subject to explicit constraints. The implication is that the ``tuition passthrough'' framework, which underpins much of the policy debate over state disinvestment \citep{bound2007cohort, webber2017state}, may overstate the price mechanism and understate alternative adjustment margins.

What adjusts instead? The Pell share and minority share results suggest that the student body recomposes---but the mechanism is ambiguous. Two channels likely operate simultaneously. The \textit{eligibility channel}: the same recession that caused funding cuts reduced family incomes, pushing more students into Pell eligibility, an effect concentrated in harder-hit states. The \textit{sorting channel}: higher-income students at underfunded public institutions may have substituted toward private alternatives, leaving the student body more disadvantaged. My design cannot separate these channels, and the enrollment growth result (Column 4) suggests the eligibility channel may dominate---institutions in higher-cut states gained students, not lost them. This means the ``composition displacement'' framing, while suggestive, should be interpreted cautiously.

The nonresident share null is perhaps the most policy-relevant finding. The hypothesis that public universities compensate for lost state revenue by recruiting full-paying out-of-state students has driven concern about ``enrollment management'' crowding out in-state access \citep{jaquette2017state, bound2020public}. This paper finds no evidence that the mechanism operates at the margin of differential state funding cuts: states that cut more did not see larger increases in out-of-state enrollment at public four-year institutions.

This paper cannot identify what happens to instructional quality, faculty-student ratios, or support services when tuition cannot adjust. These unmeasured margins---class sizes, adjunctification, deferred maintenance, reduced academic support---likely bear the true cost of state disinvestment. Future work linking IPEDS finance data with institutional quality measures could illuminate this ``quality channel'' directly.

\section{Conclusion}

State legislatures that cut higher education appropriations during fiscal downturns often assume that universities will simply raise tuition. This paper documents that, at least across the cross-state variation generated by the Great Recession, tuition barely moved in response. The strongest implication is for the tuition-passthrough assumption: public university pricing appears politically constrained from adjusting to fiscal fundamentals, at least in the short-to-medium run. The composition results---rising Pell and minority shares in higher-cut states---are suggestive of broader adjustments but cannot be cleanly attributed to institutional supply-side responses given the confounding recession. Future work with stronger identification (e.g., institutional-level funding formula shocks or policy discontinuities in tuition-setting authority) could isolate the mechanisms more precisely.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
