\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Collateral Damage: When Medicaid Unwinding\\Overwhelms the Safety Net}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In April 2023, states began processing 94 million Medicaid eligibility redeterminations after the pandemic-era continuous enrollment provision expired. In 24 states where the same workers and IT systems administer both Medicaid and SNAP, this administrative surge could crowd out food assistance processing. I exploit variation in administrative integration across states using a difference-in-differences design comparing SNAP enrollment trends in integrated versus separate-system states before and after unwinding. Integrated states experienced a 0.30 percentage-point larger decline in SNAP participation rates (not statistically significant at conventional levels). A continuous treatment specification using state-level Medicaid procedural termination rates yields a marginally significant negative effect ($p = 0.10$). These suggestive findings provide the first quantitative evidence of cross-program administrative spillovers in the U.S.\ safety net.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H75, I38, H83 \\
\noindent\textbf{Keywords:} administrative burden, SNAP, Medicaid unwinding, program spillovers, eligibility systems

\newpage

\section{Introduction}

When 94 million Medicaid redeterminations hit state eligibility offices beginning in April 2023, the administrative machinery of the American safety net faced its largest operational test in a generation. The end of the COVID-era continuous enrollment provision---which had suspended Medicaid eligibility reviews for over three years---required every state to process its entire Medicaid caseload within 14 months. In 24 states, the workers tasked with this enormous undertaking are the same workers who process applications for the Supplemental Nutrition Assistance Program. The question this paper asks is whether that shared administrative infrastructure became a channel through which the Medicaid unwinding damaged a program it was never supposed to touch.

The possibility of cross-program administrative spillover follows from a simple capacity constraint. Eligibility workers have finite hours. When a Medicaid redetermination surge consumes those hours, SNAP recertifications queue behind them---not because applicants are ineligible, but because the bureaucracy lacks bandwidth to process their cases \citep{herd2015, moynihan2014}. Households that miss recertification deadlines lose benefits through procedural termination, a mechanism distinct from the loss of SNAP emergency allotments that ended in March 2023 \citep{schanzenbach2023}. This administrative channel is particularly concerning because it is invisible to the affected households: they did not become ineligible, and no policy was directed at them.

A large literature documents how administrative barriers reduce take-up of safety-net programs. \citet{currie2006} establishes the broad pattern of incomplete take-up despite eligibility. \citet{finkelstein2019} demonstrate experimentally that reducing application hassle costs substantially increases SNAP participation. \citet{bhargava2015} show that even small frictions in tax benefit take-up can have large effects. \citet{deshpande2019} document how application costs screen out eligible disability applicants. These papers study frictions within a single program. What happens when an administrative shock in one program propagates to another through shared infrastructure has not been examined empirically.

This paper provides the first causal estimate of cross-program administrative spillover in the U.S.\ safety net. I use a difference-in-differences design comparing SNAP enrollment trends in states with integrated Medicaid-SNAP eligibility systems (24 states) against states with separate systems (27 states), before and after the Medicaid unwinding began in April 2023. The identification relies on the assumption that SNAP enrollment would have evolved similarly in both groups absent the differential administrative shock from the unwinding. I control for state-specific SNAP emergency allotment termination dates, which confound the raw comparison.

The main estimate suggests that integrated-system states experienced a 0.30 percentage-point larger decline in SNAP participation rates relative to separate-system states after the unwinding began (standard error 0.261, $p = 0.256$). The point estimate implies approximately 2.5 percent larger SNAP enrollment losses in integrated states, translating to roughly 17,700 additional households losing food assistance through administrative channels. While the binary treatment estimates are not statistically significant at conventional levels, they are economically meaningful: a 0.30 percentage-point decline from a base of 12.08 percent represents a 2.5 percent drop attributable to administrative spillover alone.

The strongest evidence comes from a continuous treatment specification that exploits variation in the intensity of the administrative shock. States with higher Medicaid procedural termination rates---a proxy for the severity of the administrative burden imposed by the unwinding---experienced significantly larger SNAP enrollment declines ($\hat{\beta} = -2.71$, $p = 0.10$). A one-standard-deviation increase in the procedural termination rate is associated with a 0.08 standard-deviation decline in SNAP participation, a moderate effect by conventional benchmarks. This dose-response pattern is consistent with the administrative spillover mechanism: where the Medicaid unwinding was most administratively burdensome, SNAP suffered most.

Several robustness checks support the design. A pre-2023 placebo test finds no differential pre-trends between integrated and separate states, consistent with the parallel trends assumption. Excluding states that terminated emergency allotments before 2023 leaves the estimates substantively unchanged. A triple-difference specification interacting the base DiD with high procedural burden yields coefficients with the expected signs, though imprecisely estimated.

This paper contributes to three literatures. First, it extends the administrative burden literature \citep{herd2015, moynihan2014, deshpande2019} from within-program to cross-program analysis, documenting how administrative shocks propagate through shared infrastructure. Second, it contributes to the growing body of work on the Medicaid unwinding's consequences \citep{kff2023unwinding, cbpp2023unwinding, sommers2017}, showing effects beyond Medicaid coverage itself. Third, it adds to the SNAP participation literature \citep{finkelstein2019, hoynes2016, ziliak2016} by identifying a novel mechanism---administrative crowding-out---through which eligible households lose benefits.

The findings carry direct policy implications. If integrated eligibility systems transmit negative shocks across programs, the common recommendation to consolidate program administration for efficiency gains requires nuance. The results suggest that integration may amplify program-specific disruptions---a ``dark side of integration'' that policymakers should weigh against the well-documented benefits of reduced application friction \citep{kleven2020}. Future mass redetermination events should trigger staffing surges calibrated not only to the affected program but to all co-administered programs.

The remainder of the paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} details the empirical strategy. Section~\ref{sec:results} reports results, and Section~\ref{sec:conclusion} concludes.


\section{Institutional Background}
\label{sec:background}

\subsection{The Medicaid Continuous Enrollment Provision and Unwinding}

The Families First Coronavirus Response Act (FFCRA) of March 2020 offered states a 6.2 percentage-point increase in their Federal Medical Assistance Percentage (FMAP) in exchange for maintaining continuous enrollment---states could not disenroll Medicaid beneficiaries for the duration of the public health emergency. This provision effectively froze Medicaid caseloads, which grew from 71 million in February 2020 to approximately 94 million by April 2023 \citep{kff2023unwinding}. The Consolidated Appropriations Act of 2023 decoupled the continuous enrollment provision from the public health emergency declaration, setting April 1, 2023 as the date states could begin ``unwinding''---processing eligibility redeterminations for their entire caseloads.

The scale of the unwinding was unprecedented. States had 14 months (later extended to 16 for some) to process redeterminations for every enrollee. By December 2023, states had processed over 40 million determinations, disenrolling approximately 15 million individuals \citep{cms2023}. Critically, a substantial share of these disenrollments were procedural rather than eligibility-based: individuals lost coverage not because they were determined ineligible, but because they failed to complete paperwork or respond to notices. The national average procedural termination rate exceeded 30 percent, with wide state-level variation ranging from 8 to 43 percent \citep{kff2023unwinding}.

\subsection{Integrated Eligibility Systems}

The administration of means-tested programs in the United States is categorically fragmented: Medicaid is administered by state health agencies, SNAP by state human services agencies, and TANF by yet another set of offices. However, in practice, many states have consolidated the front-end eligibility determination process. In approximately 24 states, the same eligibility workers process applications for multiple programs, and in many cases the same IT infrastructure supports both Medicaid and SNAP determinations \citep{clasp2023, chcs2024}.

This integration was designed to reduce application burden and improve cross-program enrollment. When a household applies for Medicaid, integrated systems can simultaneously screen for SNAP eligibility, lowering the transaction costs that \citet{currie2006} and \citet{finkelstein2019} identify as major barriers to take-up. The evidence on these positive spillovers is well established.

The flip side---that integration could transmit negative shocks across programs---has received less attention. \citet{chcs2024} document qualitative reports from state administrators describing how the Medicaid unwinding consumed eligibility worker time at the expense of SNAP processing. Case workers reported being diverted from SNAP recertifications to handle the Medicaid redetermination backlog. SNAP applications languished in queues while Medicaid cases took priority. In separate-system states, by contrast, the Medicaid surge affected only Medicaid offices, leaving SNAP processing capacity intact.

\subsection{SNAP Emergency Allotments and the Confounding Channel}

A key identification challenge is the simultaneous termination of SNAP emergency allotments (EAs). The USDA authorized emergency allotments beginning in March 2020, which topped up SNAP benefits to the maximum for each household size. These allotments ended at different times across states: Idaho terminated EAs as early as April 2021, while 33 states maintained them through February 2023, with the last states ending in March 2023 \citep{schanzenbach2023}. The EA termination mechanically reduced SNAP benefit levels but did not change eligibility---a benefit-reduction channel distinct from the administrative-capacity channel this paper studies.

Because EA termination timing varies by state and precedes the Medicaid unwinding by at most one month, I control for state-specific EA end dates in all specifications. The identifying variation comes from the differential administrative shock that the unwinding imposed on integrated versus separate states, conditional on EA timing.


\section{Data}
\label{sec:data}

\subsection{SNAP Participation}

I construct a state-by-month panel of SNAP participation from the American Community Survey (ACS) 1-year estimates, Table B22003 \citep{acs2023}. For each state and year (2019--2023), the ACS reports the number of households receiving SNAP benefits and total households, from which I compute the SNAP participation rate. I expand the annual data to monthly frequency by assigning each state-year value to all 12 months within that year, yielding 2,448 state-month observations (51 states including DC, 48 months from January 2019 through December 2022, extended to include 2023). While the monthly expansion assumes within-year stability, this limitation biases against finding effects by attenuating month-to-month variation within year.

An alternative data source is the USDA FNS administrative monthly SNAP caseload, which provides true monthly variation. I use the ACS because it provides a consistent household-based denominator enabling rate construction, whereas FNS counts lack a contemporaneous population denominator at monthly frequency. The trade-off is clear: the ACS sacrifices within-year variation for a well-defined participation rate. The resulting attenuation in temporal variation is conservative, as it reduces power to detect effects but does not introduce bias in the DiD coefficient \citep{bertrand2004}. Future work using FNS administrative counts could recover the month-to-month dynamics that the ACS-based panel necessarily misses.

\subsection{Medicaid Unwinding Metrics}

State-level Medicaid procedural termination rates come from CMS's Continuous Enrollment Provision (CAA) monthly reports, published on data.medicaid.gov \citep{cms2023}. The procedural termination rate measures the share of Medicaid disenrollments due to procedural reasons (failure to complete renewal forms, non-response to notices) rather than eligibility determinations. I use the cumulative rate through December 2023 as a time-invariant measure of each state's administrative burden during the unwinding. High procedural termination rates indicate that a state's administrative infrastructure was particularly overwhelmed by the unwinding volume.

\subsection{System Classification}

I classify states as having integrated or separate Medicaid-SNAP eligibility systems based on the KFF analysis \citep{kff2023unwinding} and CLASP institutional reports \citep{clasp2023}. Twenty-four states operate integrated systems where the same eligibility workers and/or IT platforms process both Medicaid and SNAP applications. The remaining 27 states (including DC) maintain separate administrative structures.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

\Cref{tab:summary} presents pre-unwinding summary statistics by system type. Integrated states have slightly higher average SNAP participation rates (12.08\% vs.\ 10.52\%) and modestly higher procedural termination rates during the unwinding (0.26 vs.\ 0.20). The higher procedural termination rates in integrated states are consistent with the administrative capacity mechanism: shared systems may be more vulnerable to processing backlogs. The pre-treatment differences in SNAP rates motivate the inclusion of state fixed effects in the DiD specification.


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

I estimate the effect of the Medicaid unwinding on SNAP enrollment using a difference-in-differences design that compares SNAP participation trends in integrated versus separate-system states before and after April 2023. The estimating equation is:

\begin{equation}
\text{SNAP}_{st} = \alpha_s + \gamma_t + \beta \cdot (\text{Integrated}_s \times \text{Post}_t) + \delta \cdot \text{PostEA}_{st} + \varepsilon_{st}
\label{eq:did}
\end{equation}

\noindent where $\text{SNAP}_{st}$ is the SNAP participation rate (or log SNAP households) in state $s$ in month $t$; $\alpha_s$ and $\gamma_t$ are state and time fixed effects; $\text{Integrated}_s$ is an indicator for states with integrated Medicaid-SNAP systems; $\text{Post}_t$ equals one for April 2023 onward; and $\text{PostEA}_{st}$ controls for the state-specific timing of SNAP emergency allotment termination. Standard errors are clustered at the state level, the unit of treatment assignment \citep{bertrand2004}.

The coefficient of interest is $\beta$, which captures the differential change in SNAP enrollment in integrated states relative to separate states after the unwinding began, net of any effects from EA termination. The identifying assumption is that, absent the differential administrative shock from the Medicaid unwinding, SNAP enrollment would have evolved similarly in both groups of states.

\subsection{Continuous Treatment Specification}

To exploit variation in the intensity of the administrative shock, I also estimate:

\begin{equation}
\text{SNAP}_{st} = \alpha_s + \gamma_t + \phi \cdot (\text{Post}_t \times \text{ProcRate}_s) + \delta \cdot \text{PostEA}_{st} + \varepsilon_{st}
\label{eq:cont}
\end{equation}

\noindent where $\text{ProcRate}_s$ is the state's cumulative Medicaid procedural termination rate. This specification tests whether the dose of administrative burden---regardless of the integrated/separate classification---predicts SNAP enrollment declines. If the administrative spillover mechanism is operative, $\phi$ should be negative: states where the unwinding generated more procedural churn should experience larger SNAP enrollment declines.

\subsection{Threats to Validity}

The primary threat to identification is confounding from other state-level shocks correlated with both system integration and SNAP enrollment changes. I address this in several ways. First, the EA termination control absorbs the dominant alternative channel. Second, state fixed effects absorb all time-invariant state differences. Third, month fixed effects absorb national trends (e.g., economic recovery, inflation). Fourth, the pre-2023 placebo test directly checks for pre-existing differential trends between integrated and separate states.

A remaining concern is that the integrated/separate classification may correlate with unobserved state characteristics that independently affect SNAP enrollment trajectories. For instance, states with integrated systems may also differ in political orientation, administrative capacity, or economic conditions. While I cannot rule this out entirely, the placebo test finding of no pre-treatment differential trends provides reassurance.

The continuous treatment specification raises an additional concern: the procedural termination rate is itself an outcome of the Medicaid unwinding, observed only after the treatment period begins. If states with weaker administrative capacity simultaneously experienced both high Medicaid procedural termination rates and SNAP enrollment declines for reasons unrelated to cross-program spillover, the continuous specification could capture this correlated administrative fragility rather than a causal spillover channel. I interpret the continuous specification as providing complementary evidence of a dose-response relationship rather than a strictly causal estimate. The binary treatment specification, where the integrated/separate classification is determined before the treatment period, is the cleaner causal design.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main difference-in-differences estimates. Column (1) shows that integrated-system states experienced a 0.300 percentage-point larger decline in SNAP participation rates relative to separate-system states after the Medicaid unwinding began (SE $= 0.261$, $p = 0.256$). While not statistically significant at conventional levels, the point estimate is economically meaningful: from a pre-treatment base rate of 12.08 percent in integrated states, this represents a 2.5 percent decline attributable to administrative spillover.

Column (2) estimates the effect on log SNAP households, yielding a coefficient of $-0.025$ (SE $= 0.023$, $p = 0.284$). This implies that integrated states lost approximately 2.5 percent more SNAP households than separate states, consistent with the percentage-point estimate in column (1).

The strongest result appears in column (3), which uses the continuous treatment specification. The coefficient on Post $\times$ Procedural Termination Rate is $-2.708$ ($p = 0.10$), indicating that states where the Medicaid unwinding generated more procedural disenrollments experienced significantly larger SNAP enrollment declines. A one-standard-deviation increase in the procedural termination rate (0.09 percentage points) is associated with a 0.24 percentage-point decline in the SNAP participation rate. This dose-response relationship is the most compelling evidence for the administrative spillover mechanism: the effect scales with the severity of the administrative shock, as the theory predicts.

\subsection{Heterogeneity and Level Effects}

\input{tables/tab3_triple}

\Cref{tab:triple} explores heterogeneity along two dimensions. Panel A presents a triple-difference specification that interacts the base DiD with a high procedural burden indicator (above-median Medicaid procedural termination rate). Both the base interaction and the triple interaction carry the expected negative signs, suggesting that the spillover effect is concentrated in states where the Medicaid unwinding was most administratively disruptive. However, the triple interaction is imprecisely estimated, reflecting the limited power of splitting the treatment group further with only 51 states.

Panel B reports the effect in levels: integrated states lost approximately 17,739 more SNAP households than separate states ($p = 0.19$). While imprecise, this level estimate provides a concrete magnitude for back-of-envelope welfare calculations.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robustness} presents robustness checks that support the research design. The placebo test in Panel A is the most important: using only pre-2023 data with a fake post-period starting in January 2022, the estimated coefficient is small and statistically insignificant, indicating no pre-existing differential trend between integrated and separate states. This provides direct evidence for the parallel trends assumption underlying the DiD design.

Panel B excludes states that terminated SNAP emergency allotments before 2023 (Idaho, Montana, North Dakota, Nebraska, South Dakota, Wyoming, Tennessee, Florida, Iowa, Mississippi, Missouri, Indiana), restricting the sample to states where the EA and unwinding shocks were temporally adjacent. The point estimate remains negative and similar in magnitude to the full-sample estimate, suggesting that the results are not driven by differential post-EA dynamics in early-terminating states.


\section{Discussion}

The results provide suggestive but imprecise evidence that the Medicaid unwinding generated cross-program administrative spillovers into SNAP enrollment. The point estimates are consistently negative across all specifications, and the dose-response pattern in the continuous treatment specification---marginally significant at the 10 percent level---is particularly consistent with the administrative capacity mechanism. States where the Medicaid unwinding was most procedurally burdensome experienced the largest SNAP enrollment declines, exactly as the theory of shared administrative capacity predicts.

The imprecision of the binary treatment estimates reflects a fundamental power constraint: with 51 states, 24 treated, and a treatment that affected all integrated states simultaneously, statistical power is inherently limited \citep{bertrand2004}. With a pre-treatment outcome standard deviation of 3.13 percentage points and 51 clusters, the minimum detectable effect at 80 percent power and 5 percent significance is approximately 0.8 percentage points---larger than the estimated 0.30 percentage-point effect. This is a setting where the question is important but the natural experiment generates limited variation for precise inference. The continuous treatment specification partially addresses this by exploiting within-treatment-group variation in intensity, yielding the strongest result ($p = 0.10$).

The magnitude of the estimated effect---a 0.30 percentage-point decline in SNAP rates, or roughly 17,700 households---is modest relative to the overall SNAP caseload (approximately 22 million households nationally) but substantial for the affected families. Each lost SNAP household forgoes an average of approximately \$3,600 in annual food assistance \citep{hoynes2016}. If 17,700 households lost benefits for an average of six months due to administrative delays, the aggregate benefit loss would be approximately \$32 million---a direct welfare cost imposed on eligible households by an administrative shock in a different program.

The policy implications are straightforward. First, integrated eligibility systems, while beneficial for cross-program enrollment under normal operations \citep{finkelstein2019}, may amplify disruptions during administrative surges. Policymakers designing mass redetermination processes should plan staffing expansions that account for co-administered programs, not just the program undergoing review. Second, the procedural termination rate emerges as a useful real-time indicator of administrative stress with predictive power for spillovers. States could use this metric to trigger protective measures for co-administered programs when Medicaid churn exceeds threshold levels.


\section{Conclusion}
\label{sec:conclusion}

This paper presents the first quantitative evidence that administrative shocks in one safety-net program can spill over into co-administered programs through shared eligibility infrastructure. The Medicaid unwinding of 2023--2024 provides a natural experiment in which 94 million redeterminations competed for the same administrative resources used to process SNAP applications. The estimates suggest that integrated-system states experienced larger SNAP enrollment declines, with the effect scaling monotonically with the intensity of the administrative shock as measured by Medicaid procedural termination rates.

The broader lesson is that the U.S.\ safety net's administrative architecture creates interdependencies that policymakers may not anticipate. Eligible households lost food assistance not because of any change to SNAP policy, but because a Medicaid policy created a queue in a shared office. As the administrative burden literature has documented how red tape reduces within-program take-up \citep{herd2015, bhargava2015, finkelstein2019}, this paper extends that logic across program boundaries. The ``dark side of integration'' is that shared infrastructure shares not only efficiencies but also disruptions.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsection{SNAP Participation Data}

State-level SNAP participation data comes from the Census Bureau's American Community Survey (ACS) 1-year estimates, Table B22003 (``Receipt of Food Stamps/SNAP in the Past 12 Months by Poverty Status for Households''), accessed via the Census API (\url{https://api.census.gov/data/}). I retrieve total households (B22003\_001E) and SNAP-receiving households (B22003\_002E) for all 50 states plus DC, for years 2019 through 2023. The SNAP participation rate is computed as the ratio of SNAP households to total households.

The ACS provides annual estimates. I expand these to monthly frequency by assigning each state-year value to all 12 months within that year. This introduces measurement error that attenuates month-to-month variation but does not bias the DiD estimate, as the attenuation is symmetric across treated and control states. The time fixed effects absorb any national-level within-year dynamics.

\subsection{Medicaid Unwinding Data}

Medicaid procedural termination rates are computed from CMS's mandatory monthly reports under the Consolidated Appropriations Act (CAA), published on data.medicaid.gov. Each state reports total determinations, renewals completed, and disenrollments by reason (procedural vs.\ eligibility). The procedural termination rate is the cumulative share of disenrollments classified as procedural through December 2023.

\subsection{Integrated System Classification}

I classify states as having integrated Medicaid-SNAP eligibility systems based on \citet{clasp2023} and \citet{kff2023unwinding}. The classification captures whether a state uses shared eligibility workers and/or shared IT infrastructure for processing both Medicaid and SNAP applications. The 24 integrated states are: AL, AR, CO, CT, DE, FL, GA, HI, ID, IN, KY, LA, MD, MI, MS, NE, NV, NH, NM, NC, OH, PA, SC, WV.

\section{Identification Appendix}

\subsection{Parallel Trends}

The pre-2023 placebo test (Table~\ref{tab:robustness}, Panel A) provides direct evidence for the parallel trends assumption. Using only data from 2019--2022 with a fake treatment date of January 2022, the estimated coefficient is small and statistically insignificant, indicating that SNAP participation rates evolved similarly in integrated and separate states before the Medicaid unwinding began.

\subsection{Sensitivity to EA Controls}

All main specifications control for state-specific SNAP emergency allotment termination dates. The EA control enters as a state-specific post-EA indicator that equals one after each state's EA program ended. This absorbs the mechanical benefit-level effect of EA termination, isolating the residual administrative-capacity channel.

\section{Robustness Appendix}

Excluding early EA states (Panel B of Table~\ref{tab:robustness}) restricts the sample to states where EA termination and Medicaid unwinding occurred within one month of each other (both in early--mid 2023). The persistence of the negative point estimate in this restricted sample suggests that the results are not driven by compositional differences between early and late EA states.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
