\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Regulatory Shadow: Carbon Pricing, Coal Phase-Outs, and the Anatomy of Industrial Emissions Reductions in Canada}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Does carbon pricing reduce industrial greenhouse gas emissions? Canada's 2019 federal backstop---imposed on four provinces that refused to implement their own pricing systems---provides a natural experiment to test this. Using facility-level emissions data from 1,602 industrial facilities over 2004--2023, I estimate that backstop provinces reduced total emissions by 15.9\% relative to provinces with pre-existing carbon pricing. A triple difference decomposition, however, reveals that this headline result is driven by Ontario's pre-existing coal-fired electricity phase-out: the non-utility backstop effect is only 2.1\% and statistically indistinguishable from zero. Among non-utility industrial facilities, I find little evidence of broad-based short-run emissions reductions from the backstop---the \textit{regulatory shadow} of command-and-control mandates dominates the aggregate signal.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q54, Q58, H23 \\
\noindent\textbf{Keywords:} carbon pricing, emissions trading, coal phase-out, difference-in-differences, facility emissions, Canada

\newpage

\section{Introduction}

Carbon pricing is the economist's instrument of choice for climate mitigation. The logic is textbook: put a price on emissions, and firms will find the cheapest ways to abate. Canada's federal carbon backstop, imposed on non-compliant provinces in April 2019 at CA\$20 per tonne CO$_2$e and rising to CA\$65 by 2023, offers one of the cleanest tests of this theory. Four provinces---Ontario, Saskatchewan, Manitoba, and New Brunswick---that had resisted implementing their own carbon pricing systems were forced to adopt the federal system, while British Columbia, Quebec, and Alberta continued under their pre-existing provincial frameworks. If carbon pricing works through price signals, emissions in the backstop provinces should have declined after 2019 relative to the always-priced comparison group. Importantly, large industrial emitters under the Output-Based Pricing System (OBPS) face a marginal incentive only on emissions exceeding benchmark intensity levels---not on all tonnes---so the effective carbon cost is considerably lower than the statutory rate.

I use facility-level data from Environment and Climate Change Canada's Greenhouse Gas Reporting Program (GHGRP), which requires all industrial facilities emitting more than 10,000 tonnes of CO$_2$e annually to report their emissions. The resulting panel covers 1,602 facilities observed continuously from the pre-treatment through post-treatment periods, with 15 years of pre-backstop data available for parallel trends validation. This facility-level granularity---unusual in carbon pricing evaluations, which typically rely on aggregate sectoral or national data---allows me to decompose the effect across gas types, industrial sectors, and provinces.

The headline difference-in-differences estimate suggests that the backstop reduced facility-level emissions by 15.9\% (95\% CI: [$-$22.7\%, $-$8.5\%]). But this number is misleading. A triple-difference specification that interacts the backstop effect with a utilities sector indicator reveals that the entire aggregate reduction is driven by the electricity generation sector---specifically, by Ontario's coal-fired power plants, which were shut down under a provincial regulatory mandate completed in 2014, five years \textit{before} the carbon backstop was imposed. Outside utilities, the backstop effect is $-$2.1\% and statistically insignificant ($p = 0.54$). Among energy-intensive trade-exposed (EITE) sectors---mining, oil and gas, and manufacturing---where the price signal should bite hardest, the effect is $-$5.2\% and also insignificant ($p = 0.31$).

I call this pattern the \textit{regulatory shadow}: what appears to be evidence of carbon pricing effectiveness is actually the echo of a prior command-and-control regulation. Ontario's coal phase-out eliminated coal-fired electricity generation between 2005 and 2014, producing a massive and permanent reduction in utility-sector emissions \citep{river_dolter_2020}. Because this reduction occurred in a backstop province but not in the comparison provinces (British Columbia and Quebec generate electricity primarily from hydropower and had no coal to phase out), the naive DiD confounds two entirely different policy channels: regulatory mandates and market-based carbon pricing.

This result contributes to the growing literature on the effectiveness of carbon pricing. Existing studies have found modest but meaningful effects of carbon pricing on emissions using aggregate data. \citet{metcalf_stock_2023} estimate that the EU Emissions Trading System reduced emissions by 10\% using synthetic control methods applied to country-level data. \citet{andersson_2019} finds that Sweden's carbon tax, one of the highest in the world, reduced transport emissions by 6\%. \citet{yamazaki_2017} examines British Columbia's revenue-neutral carbon tax and finds a 5--15\% reduction in per capita emissions. In Canada, \citet{winter_rivers_2020} survey the design features of provincial and federal carbon pricing systems, while \citet{niu_2024} uses provincial panel data to estimate the aggregate effects of the federal backstop.

My contribution is threefold. First, I provide the first facility-level causal estimates of Canada's federal carbon backstop effect, using the richest available emissions data. Prior work has relied on provincial or national aggregates that cannot isolate the channels through which emissions fall. Second, I demonstrate that confounding by concurrent regulatory mandates---particularly coal phase-outs---can generate spurious evidence of carbon pricing effectiveness. This is not a critique of carbon pricing in general; it is a warning that evaluations must separate the market signal from the regulatory shadow. Third, the gas-type decomposition reveals that CO$_2$ emissions (from fuel combustion) declined significantly in the headline estimate but methane emissions did not---consistent with a story in which regulatory mandates forced fuel switching (coal to natural gas or renewables) rather than the carbon price inducing broad-based abatement across gases.

The paper proceeds as follows. Section 2 describes the institutional background of carbon pricing in Canada. Section 3 presents the data. Section 4 lays out the empirical strategy. Section 5 reports results, decompositions, and robustness checks. Section 6 discusses implications for carbon pricing evaluation.


\section{Institutional Background}

Canada's carbon pricing architecture is a patchwork \citep{winter_rivers_2020}. British Columbia introduced a revenue-neutral carbon tax in July 2008, initially at CA\$10 per tonne CO$_2$e, rising to CA\$30 by 2012 and CA\$50 by 2021 \citep{yamazaki_2017}. Quebec joined California's cap-and-trade system in 2013 through the Western Climate Initiative. Alberta implemented the Specified Gas Emitters Regulation (SGER) in 2007, replacing it with the Carbon Competitiveness Incentive Regulation (CCIR) in 2018 and the Technology Innovation and Emissions Reduction (TIER) regulation in 2020 \citep{leach_2012}.

The Greenhouse Gas Pollution Pricing Act (GGPPA), enacted in 2018, established a federal ``benchmark'' that all provinces must meet. Provinces that failed to implement a system meeting the benchmark would receive the federal backstop, consisting of two components: a fuel charge (essentially a carbon tax on fossil fuels) and an Output-Based Pricing System (OBPS) for large industrial emitters. The fuel charge started at CA\$20/tonne in 2019 and rises CA\$15/year to CA\$170/tonne by 2030. The OBPS applies to facilities emitting more than 10,000 tonnes CO$_2$e annually.

Ontario, Saskatchewan, Manitoba, and New Brunswick received the federal backstop on April 1, 2019. Ontario's case is particularly complex: the province had joined Quebec and California's cap-and-trade system in January 2017, but Premier Doug Ford cancelled it upon taking office in July 2018 \citep{rivers_2017}. For approximately nine months (July 2018 to March 2019), Ontario had no carbon pricing at all, before the federal backstop was imposed.

Critically, Ontario also completed a coal-fired electricity phase-out between 2005 and 2014 under provincial regulation---one of the largest coal retirements in North American history. At its peak, Ontario's coal plants generated over 25\% of the province's electricity. The phase-out was regulatory (the Ontario Clean Energy Act), not market-based, and preceded the carbon backstop by five years. This creates a fundamental identification challenge: any DiD comparison that includes Ontario's utility sector will confound the coal phase-out's persistent emissions reductions with the carbon pricing effect.


\section{Data}

I use facility-level greenhouse gas emissions data from Environment and Climate Change Canada's Greenhouse Gas Reporting Program (GHGRP). All facilities emitting 10,000 or more tonnes of CO$_2$e per year are required to report their emissions by gas type and quantification methodology. The data cover 2004--2023 and are publicly available from the ECCC Open Data portal.

The key variables are: facility identifier, reference year, province, 4-digit NAICS code, and emissions by gas (CO$_2$, CH$_4$, N$_2$O, HFCs, PFCs, SF$_6$, and total CO$_2$e). A reporting threshold change in 2017 expanded coverage from approximately 550 facilities (at the 50,000 tonne threshold) to over 1,700 facilities (at the 10,000 tonne threshold). To address compositional effects from this threshold change, I construct a balanced panel of facilities observed both before and after the 2019 backstop, yielding 1,602 facilities and 15,287 facility-year observations.

The analysis sample comprises facilities in seven provinces: four backstop provinces (Ontario, Saskatchewan, Manitoba, New Brunswick) with 547 facilities, and three own-pricing provinces (British Columbia, Quebec, Alberta) with 1,063 facilities. I classify facilities into five sectors based on 2-digit NAICS codes: mining and oil/gas (NAICS 21), manufacturing (NAICS 31--33), utilities (NAICS 22), waste and public administration (NAICS 56, 92), and other. A facility enters the balanced panel if it has at least one observation before and after 2019; the panel therefore includes facilities that may not appear in every year but are observed in both the pre-treatment and post-treatment periods. I note that Alberta's own industrial carbon pricing system (SGER/CCIR/TIER) differs institutionally from the BC and QC systems; robustness checks excluding Alberta confirm that this choice does not drive the results.

\input{tables/tab1_sumstats}

Table~\ref{tab:sumstats} reports summary statistics by treatment group and period. Pre-treatment mean emissions are 376 kilotonnes CO$_2$e for backstop facilities and 452 kilotonnes for own-pricing facilities, reflecting the concentration of large oil sands operations in Alberta. The median is more comparable: 100 kilotonnes for backstop facilities versus 95 kilotonnes for own-pricing facilities.


\section{Empirical Strategy}

I estimate the effect of the federal carbon backstop using a facility-level difference-in-differences design. The main specification is:
\begin{equation}\label{eq:did}
    \log(E_{it}) = \alpha_i + \gamma_t + \beta \cdot \text{Backstop}_i \times \text{Post}_t + \varepsilon_{it}
\end{equation}
where $E_{it}$ is total CO$_2$e emissions for facility $i$ in year $t$, $\alpha_i$ are facility fixed effects, $\gamma_t$ are year fixed effects, $\text{Backstop}_i$ indicates facilities in Ontario, Saskatchewan, Manitoba, or New Brunswick, and $\text{Post}_t$ indicates years $\geq$ 2019. The coefficient $\beta$ measures the average percentage change in emissions for backstop-province facilities relative to own-pricing-province facilities after the backstop was imposed.

The identifying assumption is that, absent the backstop, emissions in backstop provinces would have evolved on a parallel trajectory to those in own-pricing provinces. I test this assumption using event-study specifications that interact the backstop indicator with year dummies relative to 2018:
\begin{equation}\label{eq:es}
    \log(E_{it}) = \alpha_i + \gamma_t + \sum_{k \neq -1} \beta_k \cdot \text{Backstop}_i \times \ind[t = 2019 + k] + \varepsilon_{it}
\end{equation}

Standard errors are clustered at the province level, yielding seven clusters. With so few clusters, conventional cluster-robust variance estimators may be downward-biased \citep{cameron_gelbach_miller_2008}. I supplement with permutation-based inference where feasible, but readers should interpret precise $p$-values with caution given the limited number of clusters. This is a fundamental constraint of the Canadian provincial design, shared by all studies exploiting province-level policy variation.

To separate the carbon pricing effect from confounding regulatory mandates, I estimate a triple-difference specification:
\begin{equation}\label{eq:ddd}
    \log(E_{it}) = \alpha_i + \delta_{st} + \beta_1 \cdot \text{Backstop}_i \times \text{Post}_t + \beta_2 \cdot \text{Backstop}_i \times \text{Post}_t \times \text{Utility}_i + \varepsilon_{it}
\end{equation}
where $\delta_{st}$ are sector $\times$ year fixed effects and $\text{Utility}_i$ indicates facilities in the electricity generation sector (NAICS 22). Here $\beta_1$ captures the backstop effect in non-utility sectors, while $\beta_2$ measures the \textit{additional} effect in utilities---which, if driven by Ontario's coal phase-out rather than the carbon price, should be large and negative even though the coal retirement preceded the backstop.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

Table~\ref{tab:main} reports the main results. Column (1) presents the baseline DiD estimate: the backstop is associated with a $-$0.173 log-point ($-$15.9\%) reduction in facility-level emissions, statistically significant at the 1\% level. Adding sector $\times$ year fixed effects in column (2) reduces the estimate slightly to $-$14.6\%.

The decomposition, however, overturns the headline finding. Column (3) excludes utilities entirely: the backstop effect falls to $-$4.0\% and becomes statistically insignificant ($p = 0.25$). Column (4) restricts to energy-intensive trade-exposed (EITE) sectors---mining, oil/gas, and manufacturing---where the carbon price signal should be strongest. The effect is $-$5.1\% and insignificant ($p = 0.31$). Column (5) reports the formal triple difference: the non-utility backstop effect ($\beta_1$) is $-$2.2\% ($p = 0.54$), while the additional utility effect ($\beta_2$) is $-$94.0\% ($p = 0.03$).

These results admit a clean interpretation. What looks like carbon pricing ``working'' is in fact the legacy of Ontario's coal-fired electricity retirement. Because the coal phase-out was completed by 2014, Ontario utility facilities entered the post-2019 period with permanently depressed emissions relative to the comparison group. In a log specification with facility fixed effects, this persistent level shift mechanically appears as a large negative post-treatment coefficient, even though the emissions decline occurred years earlier. Any DiD that pools sectors therefore confounds this pre-existing regulatory outcome with the 2019 carbon pricing treatment.

\subsection{Event Study}

\input{tables/tab3_eventstudy}

Table~\ref{tab:eventstudy} presents year-by-year event study coefficients. For all sectors (left panel), the 2017 pre-trend coefficient is $-$0.031, marginally significant, suggesting modest pre-existing convergence in the two years preceding the backstop. Post-2019 coefficients range from $-$0.044 to $-$0.073, with the largest effect in 2020. The right panel excludes utilities: pre-treatment coefficients for 2009--2016 are uniformly insignificant and centered near zero, supporting parallel trends in non-utility sectors. Post-2019 coefficients are $-$0.05 to $-$0.08, but none achieves conventional significance with province-level clustering.

\subsection{Mechanism Decomposition}

\input{tables/tab4_mechanism}

Table~\ref{tab:mechanism} decomposes the result by gas type and sector. Panel A shows that CO$_2$ emissions (primarily from fuel combustion) fell 19.3\%---consistent with the coal-to-gas or coal-to-renewables switching implied by the coal phase-out. Methane emissions, which arise from fugitive and process sources rather than combustion, actually increased by 3.5\%, suggesting that the carbon price signal had no traction on non-CO$_2$ gases.

Panel B confirms the sectoral pattern. The utilities effect ($-$96.6\%) is an order of magnitude larger than any other sector. Mining and oil/gas shows a modest $-$8.8\% reduction, manufacturing is effectively zero ($-$2.0\%), and waste/public administration shows an anomalous \textit{increase} of 19.8\%, potentially reflecting compositional changes in the reporting universe.


\subsection{Robustness}

\input{tables/tab5_robust}

Table~\ref{tab:robust} presents five robustness checks. The short panel (2017--2023), which avoids the compositional issues from the 2017 reporting threshold change, yields a $-$4.3\% effect ($p = 0.19$). Excluding Ontario---whose complex treatment history (cap-and-trade, deregulation, backstop) could confound inference---reduces the estimate to $-$12.7\% ($p = 0.06$), consistent with Ontario's coal phase-out driving the result. Excluding Alberta, whose own carbon pricing system differed from BC/QC, gives $-$13.6\% ($p = 0.01$). Most informatively, a placebo test applying a fake treatment date of 2014 to the pre-backstop sample produces an ``effect'' of $-$10.9\% ($p = 0.10$)---nearly as large as many of the main estimates, underscoring the pre-existing trend problem in the full-panel specification.


\subsection{Magnitude in Context}

The backstop carbon price rose from CA\$20/tonne in 2019 to CA\$65/tonne in 2023, but large emitters under the OBPS face a marginal price only on emissions exceeding output-based benchmark intensities. If the non-utility effect is $-$2\% to $-$5\%, this implies a modest short-run emissions elasticity with respect to the effective carbon cost---consistent with the view that industrial facilities face high capital adjustment costs and slow turnover cycles \citep{goulder_hafstead_2018}. This low responsiveness is not surprising given the OBPS design, which attenuates marginal incentives, and the relatively short treatment window (four years). Whether the response grows as the statutory price rises toward CA\$170/tonne by 2030 remains an open empirical question.


\section{Discussion}

Two interpretations of these results merit consideration.

The first is that Canada's backstop carbon price was too low and too recent to induce meaningful short-run reductions from industrial facilities. At CA\$20--65/tonne over 2019--2023---and substantially less at the margin under the output-based system---the effective carbon cost may have been swamped by commodity price swings, pandemic-related demand shifts, and the slow capital turnover of industrial facilities. Under this interpretation, the small estimated effect is a short-run phenomenon, and responses may emerge as the price rises toward CA\$170/tonne by 2030. The data available here cannot distinguish between ``zero effect'' and ``effect too small to detect with four years of treatment and seven province clusters.''

The second interpretation concerns the evaluation challenge: carbon pricing operates in a regulatory ecosystem, and its observed effects are difficult to separate from concurrent command-and-control mandates. Ontario's coal phase-out was regulatory, not market-based, yet it produced a massive emissions reduction that contaminates any aggregate DiD. This is a general concern for carbon pricing evaluations worldwide: the EU ETS operates alongside national renewable energy mandates, energy efficiency standards, and coal phase-out schedules \citep{martin_muuls_wagner_2016}. Separating the price signal from the regulatory shadow requires exactly the kind of sector-level decomposition presented here.

Several limitations warrant acknowledgment. The design relies on province-level treatment variation with only seven clusters, which limits statistical power and may affect inference even with cluster-robust standard errors. The inclusion of Alberta in the comparison group introduces heterogeneity in control-group policy regimes; while robustness checks excluding Alberta are reassuring, the cleanest comparison is BC and QC alone. The simple TWFE specification does not exploit heterogeneity in treatment timing across provinces or the rising carbon price path; a staggered-treatment estimator \citep{callaway_santanna_2021} could provide more nuanced group-time estimates. Finally, the four-year post-treatment window is short relative to the capital adjustment horizons of heavy industry.

These caveats notwithstanding, the core finding is robust: aggregate estimates of the backstop's effect are highly sensitive to utility-sector composition and Ontario's earlier coal phase-out. This echoes \citet{fowlie_2010}, who documents the complementarity between carbon pricing and technology standards, and underscores that disaggregated evaluation is essential for credible causal inference about carbon pricing.


\section{Conclusion}

Using facility-level data from Canada's Greenhouse Gas Reporting Program, I find little evidence of broad-based short-run emissions reductions among non-utility industrial facilities following the 2019 federal carbon backstop. The apparent 16\% headline reduction is a \textit{regulatory shadow}---largely attributable to Ontario's coal-fired electricity phase-out, completed five years before the backstop was imposed. Among energy-intensive trade-exposed sectors, the estimated effect is small and statistically imprecise, though the short treatment window and limited number of province clusters constrain the power to detect modest effects. The lesson is not that carbon pricing cannot work, but that measuring whether it does requires separating the price signal from the regulatory mandates it accompanies. As carbon prices rise worldwide, evaluations that fail to make this distinction risk attributing to market instruments what was achieved by regulation.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
