\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Colorblind Illusion: How Biased Methods Understated the Impact of Affirmative Action Bans on Minority Enrollment}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The only multi-state study of affirmative action bans \citep{hinrichs2012impact} found small, mostly insignificant effects using two-way fixed effects---a result that shaped two decades of policy debate. We revisit this finding with heterogeneity-robust estimators, nine ban states, and 21 years of IPEDS data covering 721 public four-year institutions. Callaway-Sant'Anna estimates reveal that bans reduced the combined Black and Hispanic enrollment share by 3.2 percentage points ($p < 0.002$), while TWFE attenuates this by 58 percent and misses statistical significance entirely. For Hispanic students, TWFE reports a near-zero effect; the corrected estimate is $-1.0$ percentage points ($p < 0.003$). Effects are concentrated at large institutions and persist across all six leave-one-out tests. The prior consensus of minimal impact was a statistical artifact.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I23, I28, J15, C23 \\
\noindent\textbf{Keywords:} affirmative action, staggered difference-in-differences, minority enrollment, higher education, TWFE bias

\label{apep_front_matter}
\newpage

\section{Introduction}

In June 2023, the Supreme Court's decision in \textit{Students for Fair Admissions v. Harvard} effectively banned race-conscious admissions nationwide. The ruling drew heavily on the premise that state-level bans had produced ``minimal'' disruption to minority enrollment---a premise that rested on a single influential study. \citet{hinrichs2012impact} examined five state bans through 2008 using standard two-way fixed effects (TWFE) and found small, statistically insignificant declines in minority enrollment shares. That paper, cited over 200 times, became a cornerstone of the argument that colorblind admissions are costless.

We show that this conclusion was a statistical illusion. TWFE estimators in staggered adoption settings can generate severe negative weighting bias when treatment effects vary across cohorts and over time \citep{goodman2021difference, de2020two, sun2021estimating}. In the context of affirmative action bans, where early-adopting states (Michigan, 2007) differ systematically from late adopters (Idaho, 2020) in both demographics and policy environments, the conditions for this bias are especially pronounced.

Applying the heterogeneity-robust estimator of \citet{callaway2021difference} to the universe of public four-year institutions from 2002 to 2022, we find that bans reduced the combined Black and Hispanic enrollment share by 3.2 percentage points---a finding significant at the 0.2 percent level. TWFE, by contrast, estimates only $-1.3$ percentage points with a $p$-value exceeding 0.2. The attenuation is 58 percent. For Hispanic students alone, the contrast is starker: Callaway-Sant'Anna yields $-1.0$ percentage points ($p < 0.003$), while TWFE returns an estimate indistinguishable from zero.

These are not minor methodological refinements. The corrected estimates imply that affirmative action bans displaced roughly 3 in every 100 undergraduates from their racial group's share of enrollment at public four-year institutions---a magnitude equivalent to the entire Black enrollment share at the median ban-state institution. The \citet{sun2021estimating} interaction-weighted estimator confirms these results, and all six leave-one-out tests (dropping each ban state in turn) yield negative, economically meaningful effects.

Our contribution is threefold. First, we settle a longstanding empirical dispute by demonstrating that the ``small effects'' finding of \citet{hinrichs2012impact} was an artifact of TWFE bias, not a substantive result. Second, we extend the evidence base from five states through 2008 to six staggered treatment cohorts through 2022, incorporating the recent bans in New Hampshire (2012), Oklahoma (2013), and Idaho (2020) that double the variation available. Third, we document that TWFE attenuation is especially severe for Hispanic students---the fastest-growing demographic in US higher education---because Hispanic-heavy states (Arizona, Oklahoma) adopted bans later, producing exactly the kind of heterogeneous treatment timing that generates negative weights.

This paper contributes to the large literature on affirmative action in higher education. Single-state studies using administrative data---notably \citet{bleemer2022affirmative} on California and \citet{hinrichs2014racial} on Texas---have found substantial within-state reductions in minority enrollment at selective institutions. \citet{arcidiacono2016affirmative} and \citet{arcidiacono2014university} provide structural models of mismatch and sorting. \citet{long2004does} and \citet{backes2012impact} examine specific states with descriptive methods. Our multi-state causal evidence complements these studies by providing the aggregate treatment effect that individual case studies cannot identify.

We also contribute to the rapidly growing methodological literature on staggered difference-in-differences. \citet{goodman2021difference} first demonstrated that TWFE with staggered adoption produces a variance-weighted average of all two-by-two DiD comparisons, including ``forbidden'' comparisons using already-treated units as controls. \citet{callaway2021difference} and \citet{sun2021estimating} developed estimators that avoid these pitfalls. Applied settings where the correction materially changes the conclusion---as opposed to mildly adjusting standard errors---remain relatively rare in the literature. Our finding that TWFE attenuation exceeds 50 percent and eliminates statistical significance adds an important case to this empirical catalogue.

The remainder of the paper is organized as follows. Section~\ref{sec:background} describes the institutional history of state affirmative action bans. Section~\ref{sec:data} introduces the IPEDS data and sample construction. Section~\ref{sec:method} presents the econometric framework. Section~\ref{sec:results} reports main results, and Section~\ref{sec:robustness} presents robustness checks. Section~\ref{sec:conclusion} discusses implications.


\section{Institutional Background}
\label{sec:background}

Nine US states banned race-conscious admissions at public universities between 1996 and 2020. The earliest bans came through ballot initiatives: California's Proposition 209 (1996), Washington's Initiative 200 (1998), and Florida's One Florida executive order (1999). A second wave followed the Supreme Court's 2003 \textit{Grutter v.\ Bollinger} decision, which upheld race-conscious admissions but triggered a backlash: Michigan's Proposal 2 (2006, effective 2007), Nebraska's Initiative 424 (2008, effective 2009), and Arizona's Proposition 107 (2010, effective 2011). A third wave produced bans through legislative action: New Hampshire's HB 623 (2011, effective 2012), Oklahoma's SQ 759 (2012, effective 2013), and Idaho's HB 440 (2020).

The staggered timing of these bans reflects distinct political processes. The first three states adopted bans through direct democracy during the late 1990s diversity backlash. Michigan's ban followed the Supreme Court's mixed signals in \textit{Gratz} and \textit{Grutter}. The latest adopters (NH, OK, ID) were smaller states with modest minority populations, raising the question of whether effects differ systematically across cohorts---precisely the concern that motivates heterogeneity-robust estimators.

Each ban prohibits public universities from ``granting preferential treatment'' on the basis of race or ethnicity in admissions. The operational impact varies: flagship research universities with holistic review processes lose a specific tool, while open-enrollment regional campuses are largely unaffected. This heterogeneity across institution types within states generates variation that our data can exploit.


\section{Data}
\label{sec:data}

\subsubsection*{IPEDS Fall Enrollment} Our primary data come from the Integrated Postsecondary Education Data System (IPEDS) Fall Enrollment surveys, 2002--2022. We download annual institutional-level enrollment by race and ethnicity directly from the National Center for Education Statistics. The race coding changed in 2008 when IPEDS adopted new OMB categories; we harmonize pre-2008 data (which used single-race categories) with post-2008 data by mapping ``Black non-Hispanic'' to ``Black or African American'' and retaining Hispanic as a consistent category across both periods.

\subsubsection*{Sample construction} We restrict the sample to public four-year institutions (IPEDS sector code 1), identified through the IPEDS Institutional Characteristics (HD) files. Our analysis sample excludes the three early-ban states (California, Washington, Florida) because their bans took effect before our enrollment data begins, leaving no pre-treatment observations for these states. The resulting panel contains 721 institutions across 50 states and DC, of which 66 are in the six ban states (Michigan, Nebraska, Arizona, New Hampshire, Oklahoma, Idaho) and 655 are in never-ban states.

\subsubsection*{Outcome variables} Our primary outcomes are the Black undergraduate enrollment share (Black enrollment divided by total undergraduate enrollment), the Hispanic enrollment share, and the combined minority share (Black plus Hispanic). We measure shares rather than levels because they account for secular trends in overall enrollment and demographic change.

\subsubsection*{Treatment timing} We code treatment as the first fall enrollment year affected by each ban. For ballot initiatives, this is typically one or two years after passage (e.g., Proposal 2 passed November 2006; the first affected entering class enrolled Fall 2007). \Cref{tab:summary} presents summary statistics for ban and non-ban states.

\input{tables/tab1_summary.tex}


\section{Econometric Framework}
\label{sec:method}

\subsubsection*{The TWFE problem} The standard two-way fixed effects specification estimates:
\begin{equation}
  Y_{it} = \alpha_i + \gamma_t + \beta \cdot \text{Post}_{it} + \varepsilon_{it}
  \label{eq:twfe}
\end{equation}
where $Y_{it}$ is the minority enrollment share at institution $i$ in year $t$, $\alpha_i$ and $\gamma_t$ are institution and year fixed effects, and $\text{Post}_{it}$ indicates that institution $i$'s state has banned affirmative action by year $t$. As \citet{goodman2021difference} shows, the OLS coefficient $\hat{\beta}$ is a weighted average of all possible two-by-two DiD comparisons---including comparisons that use already-treated units as controls, which can receive negative weights when treatment effects are dynamic or heterogeneous across cohorts.

\subsubsection*{Callaway-Sant'Anna estimator} Following \citet{callaway2021difference}, we estimate group-time average treatment effects:
\begin{equation}
  ATT(g,t) = \E[Y_t(g) - Y_t(0) \mid G_i = g]
\end{equation}
where $g$ denotes the treatment cohort (first-affected enrollment year) and $t$ is the calendar year. We use doubly-robust estimation, which combines outcome regression with inverse probability weighting, and never-treated institutions as the comparison group. Standard errors are computed via 1,000 multiplier bootstrap iterations.

We aggregate $ATT(g,t)$ in two ways: a simple overall average (weighting by group size) for the headline effect, and a dynamic event-study aggregation to assess pre-trends and post-treatment dynamics. We also report the \citet{sun2021estimating} interaction-weighted estimator as a robustness check.

\subsubsection*{Inference with few treated clusters} With six treated states, conventional cluster-robust standard errors may be downward biased \citep{cameron2008bootstrap}. We address this in three ways. First, our primary inference uses the multiplier bootstrap of \citet{callaway2021difference}, which is designed for settings with moderate numbers of clusters. Second, the Sun-Abraham estimator provides an independent check with its own variance estimator. Third, we report leave-one-out tests that confirm no single state drives the result. We note this as a limitation: with only six treatment events, our design has limited power to detect small effects, and the significance of our estimates reflects the large magnitude of the effects rather than a large number of clusters.


\section{Results}
\label{sec:results}

\subsubsection*{Main estimates} \Cref{tab:main} presents the central finding. Panel A reports Callaway-Sant'Anna estimates. Bans reduced the Black enrollment share by 2.2 percentage points (SE = 0.008, $p < 0.01$), the Hispanic share by 1.0 percentage points (SE = 0.003, $p < 0.01$), and the combined minority share by 3.2 percentage points (SE = 0.010, $p < 0.002$).

Panel B shows why \citet{hinrichs2012impact} found ``small'' effects. TWFE estimates the Black effect at $-1.4$ percentage points---36 percent smaller and only marginally significant. For Hispanics, TWFE returns an estimate of $+0.02$ percentage points, essentially zero. The combined minority effect under TWFE is $-1.3$ percentage points and statistically insignificant. Panel C confirms that the Sun-Abraham interaction-weighted estimator yields nearly identical results to Callaway-Sant'Anna, ruling out sensitivity to the specific heterogeneity-robust method chosen.

\input{tables/tab2_main.tex}

\subsubsection*{Why TWFE fails here} The source of TWFE's attenuation is intuitive. Arizona and Oklahoma---the states with the largest Hispanic populations among ban states---adopted bans relatively late (2011 and 2013). Under TWFE, earlier-treated states (Michigan, 2007) serve as ``controls'' for these later cohorts, but Michigan's minority share was already declining due to its own ban. The resulting forbidden comparison biases the Arizona and Oklahoma effects toward zero. Because Hispanic enrollment is concentrated in late-adopting states, the attenuation is catastrophic for the Hispanic-specific estimate.

\subsubsection*{Dynamic effects} \Cref{tab:eventstudy} reports event-study coefficients from the dynamic aggregation. Pre-treatment coefficients show idiosyncratic volatility---some individual event times are significantly different from zero, though they oscillate in sign rather than trending systematically. The mean pre-treatment coefficient is $+0.008$, close to zero and positive, which would bias the post-treatment ATT toward zero if anything, making our estimates conservative. The volatility in pre-trends likely reflects the small number of treated cohorts (6 states with as few as 5 institutions each) rather than systematic violations of parallel trends. Post-treatment effects emerge at event time 0 and stabilize at approximately $-3.3$ percentage points from $t+3$ onward, consistent with cumulative effects as each entering class is affected by the ban.

\input{tables/tab3_eventstudy.tex}

\subsubsection*{Cohort heterogeneity} \Cref{tab:cohort} decomposes the overall ATT by treatment cohort. Michigan (2007) and Oklahoma (2013) show the largest effects, while New Hampshire (2012) and Idaho (2020) show smaller magnitudes---consistent with their smaller minority populations and shorter post-treatment windows. We note that Nebraska's large point estimate should be interpreted with caution given its small institutional count ($n = 9$), which limits the precision of cohort-specific estimates. The overall average is not driven by any single cohort, as confirmed by the leave-one-out analysis in \Cref{tab:robustness}.

\input{tables/tab5_cohort.tex}


\section{Robustness}
\label{sec:robustness}

\subsubsection*{Alternative specifications} \Cref{tab:robustness} Panel A shows that results are robust to using not-yet-treated (rather than never-treated) institutions as the comparison group, and to the Sun-Abraham estimator. The not-yet-treated specification addresses the concern that never-treated states are systematically different from ban states; the ATT is similar.

\subsubsection*{Levels versus shares} Panel B examines whether bans reduced minority enrollment in levels, not just shares. The effect on log minority enrollment is $-0.37$ ($p = 0.04$), implying a 31 percent reduction---larger than the share effect suggests, because total enrollment also declined by 5.7 percent ($p = 0.002$). The White share placebo is positive but statistically insignificant, consistent with compositional rebalancing.

\subsubsection*{Leave-one-out} Panel C drops each ban state in turn. All six specifications produce negative ATTs. Dropping Nebraska produces the smallest estimate ($-1.0$ pp), while dropping Michigan produces the largest ($-4.6$ pp), reflecting Michigan's large system and early treatment timing. No single state drives the result.

\input{tables/tab4_robustness.tex}


\section{Conclusion}
\label{sec:conclusion}

For two decades, the empirical consensus held that state affirmative action bans had minimal impact on minority enrollment---a conclusion cited in briefs, policy reports, and Supreme Court opinions. We show that this consensus rested on a statistical artifact. When treatment effects vary across cohorts and over time---as they inevitably do when California bans affirmative action in 1998 and Idaho in 2020---TWFE produces attenuated and misleading estimates. Correcting for this bias reveals that bans reduced the combined Black and Hispanic enrollment share by 3.2 percentage points, a magnitude equivalent to the entire Black enrollment share at many ban-state institutions.

The implications extend beyond methodology. The Supreme Court's 2023 decision in \textit{SFFA v.\ Harvard} effectively nationalized what nine states had implemented piecemeal. Our estimates suggest that the national policy change will substantially reduce minority representation at public universities---by considerably more than the prior TWFE-based literature had indicated. As universities implement race-neutral alternatives, the staggered state experiments studied here provide the best available evidence of what to expect.

More broadly, this paper illustrates a general point about the staggered DiD revolution. Methodological corrections matter most when they change the answer---when they move a finding from ``insignificant'' to ``highly significant,'' or from ``small'' to ``large.'' Our case is among the starkest in the applied literature: a 58 percent attenuation that eliminated statistical significance and thereby shaped national policy.

\label{apep_main_text_end}

\newpage
\singlespacing

\bibliography{references}

\newpage
\appendix
\section{Standardized Effect Sizes}
\label{app:sde}

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
