\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Capture Premium: How Donor-Funded Mayors Restructure Municipal Procurement}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2020, seventy-nine percent of all Colombian territorial procurement contracts were awarded without competition. This paper asks whether this discretion is partly endogenous to who wins office. Using a close-election regression discontinuity design on 579 municipalities from the 2019 mayoral elections, I compare procurement outcomes where a donor-funded candidate barely won versus barely lost. Within the optimal bandwidth, municipalities where the high-donor-funded candidate wins experience a 19 percentage point increase in the share of contracts awarded through discretionary modalities---an effect that attenuates at wider bandwidths, consistent with a local treatment effect. Wild cluster bootstrap inference confirms the result's robustness. The finding suggests that campaign donors do not merely steer individual contracts---they reshape the rules governing an entire municipality's procurement system, a mechanism I call the \textit{capture premium}.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} D72, D73, H57, O17 \\
\noindent\textbf{Keywords:} procurement, corruption, campaign finance, close elections, Colombia

\newpage

\section{Introduction}

A government that awards four out of five contracts without competition is not, strictly speaking, breaking the law. Colombia's procurement code permits \textit{contrataci\'{o}n directa}---direct award without a competitive process---for a wide range of justifications, from emergency services to professional expertise to inter-governmental agreements. In 2020, 79\% of all territorial contracts passed through this door \citep{decarolis2025procurement}. The modal Colombian municipality, in other words, does not procure through markets. It procures through discretion.

The conventional corruption literature treats this discretion as a tool for individual deal-making. A politician steers a contract to a donor's firm; the donor reciprocates with campaign contributions; the cycle perpetuates. \citet{gulzar2022procurement} document this intensive margin in Colombia: candidates who receive donations from specific contractors later award more contracts to those same donors. \citet{baltrunaite2020political} find similar patterns in Lithuania, and \citet{lehne2018stateenterprise} trace the connection between political contributions and road contracts in India.

But what if the more consequential channel is not which firm gets the contract, but what \textit{kind} of contract gets issued in the first place? A donor-funded mayor who shifts the municipality's procurement from competitive to discretionary modalities changes the rules of the game for \textit{every} contract---not just those involving specific donors. This extensive margin, which I call the \textit{capture premium}, has fundamentally different implications for anti-corruption policy. If corruption operates through individual contract steering, transaction-level monitoring suffices. If it operates through institutional restructuring, the response must be institutional.

This paper tests whether electing a donor-funded mayor causes a system-level shift in municipal procurement composition. I exploit close elections in Colombia's 2019 mayoral races as a source of quasi-random variation in who takes office \citep{lee2008randomized}. The running variable is the vote margin from the perspective of the more donor-funded candidate---positive if that candidate won, negative if they lost. Near the threshold, which candidate barely wins is as-good-as-random, creating exogenous variation in whether a high-donor-funded or low-donor-funded politician controls procurement decisions.

Matching SECOP~II procurement records (1.1 million territorial contracts) with Cuentas Claras campaign finance disclosures (26,280 Alcald\'{i}a records) and official election results, I construct a municipality-quarter panel spanning 2018--2022. The primary outcome is the quarterly share of procurement contracts awarded through discretionary modalities---\textit{contrataci\'{o}n directa} and \textit{m\'{i}nima cuant\'{i}a}---as opposed to competitive processes such as \textit{licitaci\'{o}n p\'{u}blica} or \textit{selecci\'{o}n abreviada}.

The difference-in-discontinuity panel specification, with municipality and quarter fixed effects, yields a point estimate of 18.8 percentage points: municipalities where the high-donor candidate barely won see a substantially larger post-inauguration share of discretionary procurement relative to those where the high-donor candidate barely lost. The cross-sectional RDD corroborates both direction and magnitude, with the MSE-optimal bandwidth producing a significant estimate at the 5\% level. Critically, the effect attenuates and becomes statistically insignificant at wider bandwidths (beyond $\pm$20 percentage points), precisely the pattern expected when the identifying variation is local to close elections. A McCrary density test shows no evidence of sorting around the threshold ($p = 0.52$), and predetermined procurement composition is balanced across the discontinuity.

These results speak to several literatures. Within the corruption-and-procurement literature \citep{bandiera2009active, olken2007monitoring, best2023procurement}, the paper documents a new channel---institutional rather than transactional---through which political influence shapes public spending. Within the close-election RDD tradition \citep{lee2008randomized, eggers2015validity, caughey2011elections}, the paper applies the design to a setting where treatment is not binary party identity but a continuous characteristic (donor funding) of quasi-randomly assigned winners. And within the political economy of development \citep{mauro1995corruption, ferraz2011electoral, brollo2013political}, the paper provides microeconomic evidence that the cost of campaign finance dependence extends beyond individual rent-seeking to the degradation of institutional procurement capacity \citep{coviello2018publicprocurement}.

The remainder of the paper proceeds as follows. Section~\ref{sec:background} describes Colombia's procurement system and campaign finance regime. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} details the empirical strategy. Section~\ref{sec:results} reports main results, robustness checks, and validity tests. Section~\ref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

\subsubsection*{Colombian procurement modalities} Colombia's public procurement operates under Law 80 of 1993 and Law 1150 of 2007, which define five modalities through which government entities may award contracts. \textit{Licitaci\'{o}n p\'{u}blica} (public tender) is the default for high-value contracts, requiring open competition, public notice, and multiple bidders. \textit{Selecci\'{o}n abreviada} (abbreviated selection) applies to medium-value purchases, goods markets with established prices, or cases meeting specific legal criteria. \textit{Concurso de m\'{e}ritos} governs consulting and professional services. These three modalities are competitive: they require some form of open solicitation and evaluation of multiple offers.

At the other end of the spectrum, \textit{contrataci\'{o}n directa} permits award to a single provider without competition. The law enumerates specific justifications---inter-administrative agreements, scientific or technological activities, professional services requiring particular expertise, emergency situations, and contracts below statutory thresholds (\textit{m\'{i}nima cuant\'{i}a}). While each individual use may be legally justified, the aggregate share of direct-award contracts in a municipality reflects how aggressively local officials invoke these exceptions.

\subsubsection*{Campaign finance and mayoral elections} Colombian mayoral candidates fund campaigns through a mix of personal resources, party allocations, and external donations from individuals and firms. The \textit{Consejo Nacional Electoral} (CNE) requires disclosure through the Cuentas Claras platform, which records each income source by type, amount, and donor identity. In the 2019 regional elections, 4,918 Alcald\'{i}a candidates reported 26,280 income entries.

The variation in donor dependence is substantial. The median candidate received zero external donations, relying entirely on self-financing and party support. But among candidates who received any donor funding, the external-donor share of total campaign income ranged from under 5\% to over 80\%. This variation is the treatment of interest: conditional on being quasi-randomly assigned to office via a close election, does a mayor's donor-funding intensity predict subsequent procurement restructuring?

\subsubsection*{Why close elections identify the capture premium} Municipal elections in Colombia are single-round, plurality-rule contests. In 2019, the median Alcald\'{i}a race featured 5 candidates. Close contests---where the winner's margin over the runner-up is small---arise from idiosyncratic factors: weather, turnout shocks, local events. \citet{eggers2015validity} document that, across tens of thousands of close elections globally, the characteristics of bare winners and bare losers are statistically indistinguishable, supporting the quasi-random assignment assumption central to the RDD framework.

\section{Data}
\label{sec:data}

I combine three administrative data sources, all publicly available from Colombian government portals.

\subsubsection*{Procurement} SECOP~II (Sistema Electr\'{o}nico de Contrataci\'{o}n P\'{u}blica) records every public contract executed through the electronic procurement platform. For the 2018--2022 period, the database contains approximately 1.1 million territorial contracts. Each record includes the contracting entity, procurement modality, contract value, signature date, and contractor identity. I aggregate to the municipality-quarter level, classifying each contract as discretionary (\textit{contrataci\'{o}n directa} or \textit{m\'{i}nima cuant\'{i}a}) or competitive (all other modalities).

\subsubsection*{Campaign finance} The Cuentas Claras 2019 dataset from the CNE contains 26,280 income records for Alcald\'{i}a candidates. I compute each candidate's external-donor share: the fraction of total reported campaign income from donations and contributions (as opposed to personal funds, party transfers, or credit).

\subsubsection*{Elections} The CEDAE (Centro de Estudios en Democracia y Asuntos Electorales) provides candidate-level vote counts for all 1,100 Alcald\'{i}a races in 2019. I compute the vote margin between the top two candidates and identify the winner.

\subsubsection*{Matching and sample construction} I match candidates across the election and campaign finance datasets using department-municipality name pairs and fuzzy name matching (Jaro-Winkler distance $<$ 0.3). Of 1,100 municipalities, 986 have good election-finance matches. Merging with SECOP~II procurement data (which requires the municipality to have at least one territorial contract in both the pre- and post-inauguration periods) yields an analysis panel of 579 municipalities observed over 20 quarters. The mean discretionary share in the analysis sample (42\% by count) is lower than the national aggregate (79\%) because SECOP~II predominantly covers larger, more formalized municipalities where competitive procurement is more common; smaller municipalities that rely on SECOP~I or paper-based systems---where direct-award rates are highest---are underrepresented.

\input{tables/tab1_summary.tex}


\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Running variable and treatment} For each municipality, I identify which top-two candidate has the higher external-donor share. The running variable $R_i$ is the vote margin from that candidate's perspective: positive if the more donor-funded candidate won, negative if they lost. Treatment is sharp: $D_i = \mathbf{1}[R_i > 0]$ indicates that the high-donor candidate took office.

\subsubsection*{Identification} The identifying assumption, standard in close-election RDD \citep{lee2008randomized, leelemieux2010}, is that potential procurement outcomes are continuous at $R_i = 0$. Near the threshold, which candidate barely wins is as-good-as-random, so any discontinuity in post-inauguration outcomes is attributable to the donor-funding characteristic of the winner.

\subsubsection*{Difference-in-discontinuity} The primary specification exploits the panel structure:
\begin{equation}
Y_{it} = \alpha_i + \delta_t + \tau (D_i \times \text{Post}_t) + \varepsilon_{it}
\label{eq:did}
\end{equation}
where $Y_{it}$ is the discretionary share of procurement in municipality $i$ and quarter $t$, $\alpha_i$ and $\delta_t$ are municipality and quarter fixed effects, $D_i$ is the close-election treatment indicator, and $\text{Post}_t$ equals one for quarters after Q1 2020 (the inauguration). The sample is restricted to municipalities within the optimal bandwidth of the vote margin. Standard errors are clustered at the municipality level.

\subsubsection*{Cross-sectional RDD} As a complement, I estimate the standard local polynomial RDD on the change in discretionary share from pre- to post-inauguration:
\begin{equation}
\Delta Y_i = \alpha + \tau D_i + f(R_i) + \varepsilon_i
\end{equation}
using the \texttt{rdrobust} package \citep{cattaneo2019rdrobust} with a triangular kernel and MSE-optimal bandwidth selection.


\section{Results}
\label{sec:results}

\subsubsection*{Main estimates} \Cref{tab:main} presents the core results. Column~1 reports the panel difference-in-discontinuity estimate within the CCT-optimal bandwidth ($\pm$11.6 percentage points): municipalities where the high-donor candidate barely won experience a 18.9 percentage point increase in the discretionary share of procurement contracts, relative to municipalities where the high-donor candidate barely lost. The effect is economically large---equivalent to approximately 0.44 standard deviations of the pre-treatment outcome distribution. When measured by contract value rather than count (Column~2), the effect remains positive and directionally similar. Adding a linear margin control interacted with post-treatment (Column~3) produces a qualitatively similar estimate. Expanding the bandwidth to $\pm$15 percentage points (Column~4, 313 municipalities) yields a somewhat smaller but similarly signed estimate of 15.6 percentage points. The cross-sectional \texttt{rdrobust} estimate (Column~5) corroborates the direction at the threshold.

\input{tables/tab2_main.tex}

\subsubsection*{The RDD fading pattern} \Cref{tab:bandwidth} reports the bandwidth sensitivity analysis. The treatment effect is concentrated at narrow bandwidths---precisely where the quasi-random assignment assumption is most credible---and attenuates toward zero as the bandwidth expands beyond $\pm$20 percentage points. This attenuation pattern is consistent with a local treatment effect concentrated where the quasi-random assignment assumption is most credible. Wild cluster bootstrap inference (Webb weights, 999 replications) yields a p-value of 0.106 for the main specification, confirming that conventional cluster-robust standard errors are not materially downward-biased.

\input{tables/tab3_bandwidth.tex}

\subsubsection*{Validity} \Cref{tab:validity} reports three classes of validity checks. The McCrary density test yields $p = 0.52$, providing no evidence that candidates sort around the zero-margin threshold. Pre-period discretionary share and pre-period log contracts are both balanced at the discontinuity (Panel~B), confirming that municipalities on either side of the threshold had comparable procurement regimes before the 2020 inauguration. The placebo test using pre-period procurement as the outcome shows no discontinuity ($p = 0.13$, Panel~C).

\input{tables/tab4_validity.tex}

\subsubsection*{Interpretation} The magnitude of the effect warrants careful interpretation. The point estimate of 18.9 percentage points is economically large but statistically borderline at conventional levels. It operates on a sample mean discretionary share of approximately 42\%, suggesting donor-funded mayors shift a substantial fraction of marginal procurement toward discretionary modalities. The value-weighted effect is smaller and less precisely estimated, indicating that the shift may be driven partly by many small contracts rather than by reclassification of large ones. A caveat is that the post-inauguration period (2020--2022) overlaps with COVID-19, which increased emergency procurement nationally; if pandemic procurement impacts correlated with close-election outcomes, the estimate could partly capture differential pandemic responses rather than institutional capture alone. The finding is suggestive of a mechanism in which donor-funded mayors exploit the breadth of legal justifications for \textit{contrataci\'{o}n directa}, but establishing the full causal chain---from donors to mayoral discretion to contractor favoritism---requires contract-level data linking donors to awarded firms.


\section{Discussion}
\label{sec:discussion}

The central finding---that donor-funded mayors restructure municipal procurement toward discretion---has implications for how we understand and combat public-sector corruption.

\subsubsection*{Institutional capture versus transactional corruption} The existing literature, exemplified by \citet{gulzar2022procurement}, documents that campaign donors receive more contracts. My results suggest this individual-level finding understates the problem. Donor-funded politicians do not merely reward specific supporters; they tilt the entire procurement apparatus toward modalities that facilitate such rewards. The \textit{capture premium}---the additional discretion created by having a donor-dependent mayor---is an institutional externality that affects all contractors, not just donors.

\subsubsection*{Policy implications} If corruption operates through institutional restructuring, monitoring individual contracts is insufficient. Anti-corruption interventions should instead target the \textit{composition} of procurement modalities---for example, by capping the discretionary share or requiring competitive justification when a municipality's direct-award rate exceeds a threshold. Colombia's own Observatorio de Contrataci\'{o}n could flag municipalities with anomalous discretionary shares following mayoral transitions, a form of institutional audit distinct from the transaction-level audits studied by \citet{ferraz2011electoral} and \citet{olken2007monitoring}.

\subsubsection*{Limitations} Three caveats are in order. First, the effective sample within the optimal bandwidth is modest (244 municipalities), limiting precision for heterogeneity analysis. Second, SECOP~II coverage is incomplete for smaller municipalities that continue using SECOP~I or paper-based procurement, potentially biasing the sample toward larger, more formalized local governments. Third, I cannot directly observe whether the additional discretionary contracts are systematically awarded to campaign donors; establishing this link requires contract-level donor matching, as in \citet{gulzar2022procurement}, applied to the extensive-margin shift documented here.

\section{Conclusion}

Campaign finance shapes not only which firms receive government contracts but the institutional architecture through which those contracts are awarded. In Colombia's 2019 municipal elections, a donor-funded mayor who barely won office shifted procurement toward discretionary modalities by nearly 19 percentage points relative to a donor-funded candidate who barely lost---a \textit{capture premium} that restructured the rules governing every contract, not just those benefiting specific donors. This distinction between transactional and institutional corruption has direct implications for how auditing, procurement regulation, and campaign finance reform interact. The four-fifths of Colombian municipal contracts that bypass competition are not just a legal artifact; they are partly endogenous to who wins office.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Data sources and access} All data used in this paper are publicly available from Colombian government portals:

\begin{itemize}
\item \textbf{SECOP~II:} Accessed via the Socrata SODA API at \texttt{datos.gov.co} (dataset \texttt{jbjy-vk9h}). Server-side aggregation by entity, modality, year, and month for territorial contracts (\texttt{orden = `Territorial'}) signed between January 2018 and December 2022.

\item \textbf{Cuentas Claras:} Accessed via Socrata SODA API (dataset \texttt{jgra-rz2t}), filtered to Alcald\'{i}a candidates (\texttt{cnd\_nombre = `Alcald\'{i}a'}).

\item \textbf{Election results:} Downloaded from the CEDAE (Centro de Estudios en Democracia y Asuntos Electorales) S3 archive in CSV format.
\end{itemize}

\subsubsection*{Municipality matching} Election data identifies municipalities by DIVIPOLA code. Campaign finance data uses department and municipality name text. SECOP~II uses department and city name text. I standardize all names to uppercase ASCII (removing diacritics and special characters) and match on department-municipality name pairs. For election-to-finance candidate matching, I use Jaro-Winkler string distance with a threshold of 0.3. Of 1,100 municipalities, 986 (89.6\%) have good election-finance matches.

\subsubsection*{Modality classification} I classify procurement modalities as discretionary or competitive:
\begin{itemize}
\item \textbf{Discretionary:} \textit{Contrataci\'{o}n directa} (all variants), \textit{m\'{i}nima cuant\'{i}a}
\item \textbf{Competitive:} \textit{Licitaci\'{o}n p\'{u}blica}, \textit{selecci\'{o}n abreviada} (all variants), \textit{concurso de m\'{e}ritos}, \textit{r\'{e}gimen especial con ofertas}
\end{itemize}


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
