\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{tabularray}
\UseTblrLibrary{booktabs}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Erased from the Nation: Mass Denationalization and the Aggregate Labor Market in the Dominican Republic}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
On September 23, 2013, the Dominican Republic's Constitutional Tribunal retroactively stripped citizenship from an estimated 210,000 Haitian-descent Dominicans---the largest mass denationalization in the Western Hemisphere since the mid-twentieth century. I estimate the aggregate labor market effects of this shock using an interrupted time series design with ILO and World Bank data spanning 2005--2023. Despite the scale of the ruling, national labor market indicators show no deterioration: vulnerable employment fell by 1.1 percentage points ($p=0.016$) and self-employment fell by 1.4 percentage points ($p=0.002$), consistent with the Dominican Republic's secular formalization trend. These precise null results on the aggregate suggest that the costs of denationalization were borne entirely by the affected 2\% of the population, invisible in national statistics. A two-shock design and placebo tests support this interpretation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J61, K37, O15 \\
\noindent\textbf{Keywords:} denationalization, citizenship, informality, Dominican Republic, Haiti, immigration enforcement

\newpage

\section{Introduction}

On the morning of September 24, 2013, an estimated 210,000 people in the Dominican Republic woke up stateless. The previous day, the Constitutional Tribunal had issued Sentencia TC/0168/13, retroactively reinterpreting the country's citizenship clause to deny nationality to anyone born after 1929 without at least one parent ``of Dominican blood.'' Overnight, men and women who had been born in Dominican hospitals, attended Dominican schools, and worked in Dominican fields were stripped of the only citizenship they had ever known. They could no longer sign a formal employment contract, enroll their children in public school, or vote.

This paper asks a deceptively simple question: did TC/0168 leave a mark on the Dominican labor market? The economics of immigration enforcement predicts that removing legal status from workers should push them into informality \citep{kossoudji2002effect, orrenius2015behind}, compress wages in the informal sector through increased labor supply \citep{borjas2003labor, dustmann2013effect}, and reduce human capital investment for affected youth \citep{lafortune2019deferred}. These predictions rest on decades of evidence from the United States and Europe, where enforcement actions target immigrants who crossed a border. TC/0168 is fundamentally different: it targeted people \textit{born in the country}, creating a ``denationalization shock'' that has no parallel in the developed-country immigration literature.

I estimate the aggregate labor market effects of TC/0168 using an interrupted time series (ITS) design applied to national-level data from the ILO and World Bank, covering 2005--2023. The ITS compares the post-2013 trajectory of employment composition, unemployment, labor force participation, and school enrollment against the pre-2013 trend, isolating any structural break attributable to the ruling. The specification includes a linear trend and trend break to account for the Dominican Republic's sustained economic growth over this period.

The results are striking in their absence. Vulnerable employment---the ILO's measure of own-account and contributing family workers, a standard proxy for informality---\textit{fell} by 1.1 percentage points after TC/0168 ($p = 0.016$), continuing its pre-existing downward trajectory. Self-employment declined by 1.4 percentage points ($p = 0.002$), and wage employment rose by the same amount. Unemployment showed no significant break. Labor force participation increased by 2.2 percentage points ($p = 0.079$), driven equally by male and female entry. Secondary school enrollment showed no detectable change.

These aggregate nulls are not evidence that TC/0168 was harmless. They reflect a simple arithmetic fact: the $\sim$210,000 affected individuals constituted approximately 2\% of the Dominican Republic's 10.4 million population \citep{one2013eni}. Even if every affected worker transitioned from formal to informal employment, the effect on national informality shares would be on the order of 0.4 percentage points---well within the noise of macroeconomic fluctuations driven by tourism, remittances, and construction booms. The affected population is statistically invisible.

This paper makes three contributions. First, it provides the first time-series econometric analysis of mass denationalization as a labor market event, explicitly framed as an aggregate-level test. The political science and human rights literatures on TC/0168 are extensive \citep{belton2020statelessness, chen2023denationalization}, but no paper in economics has brought formal time series methods to bear on the question. The closest work is \citet{amuedo2017dominican}, who use a simple pre/post comparison of school enrollment and informal employment at the national level without trend adjustment, pre-trend testing, or explicit power analysis. This paper advances on their work by incorporating trend breaks, Newey-West inference, placebo tests, and a formal minimum detectable effect calculation that reveals why the aggregate null was expected ex ante.

Second, the paper quantifies the \textit{statistical invisibility} of a marginalized population. A back-of-envelope calculation shows that the aggregate data lack the power to detect effects concentrated in 2\% of the workforce. This is not a methodological limitation to apologize for---it is the finding. When a state strips citizenship from a small minority, the macroeconomy absorbs the shock without a ripple, even as the affected individuals face devastating personal consequences. This pattern echoes the broader literature on how aggregate statistics obscure the experiences of vulnerable populations \citep{watson2014inside, alsan2024fear}.

Third, the paper introduces a novel two-shock design that distinguishes the 2010 constitutional amendment (which narrowed jus soli prospectively) from the 2013 TC/0168 ruling (which applied it retroactively). The amendment produced no detectable break; the ruling produced none either. Together, these results suggest that Dominican labor markets are resilient to citizenship policy shocks that affect a small population share, regardless of the shock's severity.

The paper proceeds as follows. Section 2 describes the institutional background and the TC/0168 ruling. Section 3 presents the data. Section 4 describes the empirical strategy. Section 5 reports results. Section 6 discusses the implications of aggregate invisibility.

\section{Institutional Background and Policy Setting}

\subsubsection*{The Dominican Republic--Haiti corridor}
The Dominican Republic and Haiti share the island of Hispaniola and a 376-kilometer border. Haitian workers have migrated to the Dominican Republic since the early twentieth century, initially recruited for the sugar harvest (\textit{zafra}) and later entering construction, agriculture, and domestic service \citep{martinez2014peripheral}. By 2012, the Dominican Republic hosted an estimated 524,632 immigrants, the vast majority of Haitian origin, plus an additional $\sim$244,000 descendants born on Dominican soil \citep{one2013eni}.

\subsubsection*{Citizenship and jus soli}
Dominican citizenship was historically governed by jus soli: anyone born on Dominican territory was a citizen, regardless of parental nationality. Article 11 of the 1966 Constitution contained an exception for children of diplomats and persons ``in transit,'' which was understood to refer to short-term visitors. For decades, children of long-term Haitian residents born in the Dominican Republic received Dominican birth certificates and identity documents.

\subsubsection*{The 2010 constitutional amendment}
In January 2010, a new constitution narrowed jus soli. Article 18.3 excluded from birthright citizenship children born to parents who were ``illegally residing'' in the country. Crucially, this change was \textit{prospective}: it applied to children born after January 26, 2010, and did not affect those already holding Dominican nationality.

\subsubsection*{Sentencia TC/0168/13}
On September 23, 2013, the Constitutional Tribunal issued Sentencia TC/0168/13 in response to a case brought by Juliana Dequis Pierre, a Dominican-born woman of Haitian parents who was denied a \textit{c\'{e}dula} (identity card). The ruling declared that the ``in transit'' exception applied retroactively to 1929 and encompassed all persons whose parents were not legally resident at the time of their birth. This effectively stripped Dominican nationality from an estimated 210,000 individuals---citizens who had lived their entire lives in the Dominican Republic.

\subsubsection*{Law 169-14 and its failures}
International condemnation was swift. In May 2014, the Dominican Congress passed Law 169-14, creating a two-track regularization process. ``Group A'' (those who had been registered in the civil registry) would have their documents reinstated. ``Group B'' (those born to undocumented parents who had never been registered) were directed to apply as foreigners under a National Regularization Plan, with a June 2015 deadline. Of the estimated 110,000 Group B individuals, only 8,755 had registered by the deadline \citep{one2018eni}. The vast majority remained without legal status, unable to access formal employment, public education, or healthcare.

\subsubsection*{Provincial variation in exposure}
The Haitian-descent population is geographically concentrated. The five provinces bordering Haiti---Dajab\'{o}n, Monte Cristi, El\'{i}as Pi\~{n}a, Independencia, and Pedernales---have Haitian-born population shares of 5--9\%, compared to a national average of 3\%. Agricultural centers dependent on Haitian labor (San Pedro de Macor\'{i}s, La Romana, La Altagracia) have shares of 4--6\%. Urban provinces like Santo Domingo and Distrito Nacional have lower shares despite large absolute Haitian populations. This geographic variation provides a descriptive dimension to complement the national time series analysis.

\section{Data}

I assemble annual panel data from three public sources covering 2005--2023.

\subsubsection*{ILO SDMX API}
The International Labour Organization provides annual labor market indicators derived from the Dominican Republic's Encuesta Nacional Continua de Fuerza de Trabajo (ENCFT). I extract unemployment rate, labor force participation rate, and employment-to-population ratio, each disaggregated by sex and age group (15+, 15--24). All series are available from 2000 onward with no gaps.

\subsubsection*{World Bank WDI API}
The World Development Indicators provide vulnerable employment share (own-account and contributing family workers as a percentage of total employment), self-employment share, wage worker share, agricultural employment share, services employment share, secondary school net enrollment ratio, GDP per capita (constant 2015 USD), and the Gini coefficient. Coverage spans 2000--2023 with occasional gaps in enrollment data.

\subsubsection*{Provincial population data}
Province-level population counts and Haitian-born shares are drawn from the Dominican Republic's 2010 Census, conducted by the Oficina Nacional de Estad\'{i}stica (ONE). The census enumerates 32 provinces and identifies place of birth, providing a direct measure of treatment intensity: the fraction of each province's population born in Haiti.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

Table~\ref{tab:summary} reports summary statistics. Vulnerable employment averaged 40.4\% over the sample period, ranging from 37.8\% to 44.5\%. Self-employment averaged 41.3\%. These levels are high by Latin American standards, reflecting the Dominican Republic's large informal agricultural and service sectors. GDP per capita grew from \$5,500 in 2005 to \$8,700 in 2023, driven by tourism, remittances, and free-trade-zone manufacturing. Province-level Haitian-born shares range from 1\% (Hermanas Mirabal) to 9\% (Dajab\'{o}n), with border provinces systematically at the top of the distribution.

\section{Empirical Strategy}

\subsection{Identification and Assumptions}

The ideal experiment would compare outcomes for denationalized individuals against a counterfactual in which they retained citizenship. Absent individual-level panel data spanning the ruling, I pursue a second-best approach: testing whether TC/0168 produced a detectable aggregate break in national labor market indicators.

The identifying assumption is that, absent TC/0168, labor market outcomes would have continued along their pre-2013 trajectory. This is a strong assumption---many other factors affect Dominican labor markets---but it is testable using pre-trend evidence and placebo tests.

I estimate an interrupted time series (ITS) specification:
\begin{equation}
Y_t = \alpha + \beta_1 \cdot \textit{Post}_t + \beta_2 \cdot \textit{Trend}_t + \beta_3 \cdot (\textit{Post}_t \times \textit{Trend}_t) + \varepsilon_t
\label{eq:its}
\end{equation}
where $Y_t$ is the outcome in year $t$, $\textit{Post}_t = \mathbb{1}[t \geq 2014]$ (the first full year after the September 2013 ruling), $\textit{Trend}_t = t - 2013$ centers the time trend at the treatment date, and the interaction captures any change in the post-treatment slope.

The coefficient $\beta_1$ captures the level shift at the treatment date: the immediate change in the outcome attributable to TC/0168 after controlling for the secular trend. Standard errors use the Newey-West estimator to account for serial correlation.

\subsection{Threats to Validity}

\subsubsection*{Concurrent shocks}
The Dominican economy experienced sustained growth throughout the 2010s, driven by mining (Pueblo Viejo gold mine, operational 2013), tourism expansion, and construction. These positive macro shocks could mask a negative TC/0168 effect. The inclusion of the linear trend absorbs gradual growth effects; any remaining concern is addressed by the placebo test (Section 5.3).

\subsubsection*{Statistical power}
The affected population ($\sim$210,000) constitutes approximately 2\% of the total population. If denationalization shifted all affected workers from formal to informal status, the aggregate effect on vulnerable employment would be approximately $0.02 \times 1 = 0.4$ percentage points. With a standard deviation of 2.1 percentage points in the outcome and 19 annual observations, the minimum detectable effect at 80\% power is approximately 2.0 percentage points---five times larger than the predicted effect. The null result is therefore expected \textit{ex ante} and does not rule out severe consequences for the affected population.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_its}

Table 2 reports the ITS estimates. Vulnerable employment shows a statistically significant \textit{decrease} of 1.1 percentage points after TC/0168 ($p = 0.016$), inconsistent with the hypothesis that denationalization increased informality. Self-employment fell by 1.4 percentage points ($p = 0.002$), and wage employment rose symmetrically. These estimates reflect the Dominican Republic's continued formalization trend rather than any detectable effect of the ruling.

Unemployment rose by 0.7 percentage points, but this estimate is imprecise ($p = 0.184$). Labor force participation increased by 2.2 percentage points ($p = 0.079$), and secondary school enrollment showed no change ($\beta = -0.4$, $p = 0.747$). Across all six outcomes, there is no evidence that TC/0168 produced a negative aggregate break.

To put these magnitudes in context: the 1.1 percentage point decline in vulnerable employment represents a move from approximately 41\% to 40\% of total employment---a continuation of a trend that had been underway since 2005. The Dominican Republic's GDP per capita grew by 47\% between 2005 and 2023. In this macroeconomic context, the removal of 210,000 individuals from the formal labor market was a rounding error.

\subsection{Heterogeneity}

\input{tables/tab3_heterogeneity}

Table 3 disaggregates the ITS by gender, age, and sector. Male and female labor force participation both increased after 2013, with the male coefficient (2.6 pp) slightly larger than the female (1.9 pp). Youth unemployment (ages 15--24) showed no significant break ($\beta = 0.8$, not significant). Agricultural employment was roughly flat, while services employment showed a trend decline, consistent with the sectoral shift away from agriculture that characterized Dominican development over this period. None of these heterogeneity cuts reveals a pattern consistent with the denationalization shock.

\subsection{Robustness}

\input{tables/tab4_twoshock_placebo}

\subsubsection*{Two-shock design}
Table 4 columns (1)--(2) present a two-shock specification that includes separate indicators for the 2010 constitutional amendment and the 2013 TC/0168 ruling. If the 2010 amendment (which narrowed jus soli prospectively) had any effect, it should appear as a positive coefficient on Post 2010. Instead, the 2010 coefficient for vulnerable employment is positive but imprecise (0.98, not significant), while the 2013 coefficient is negative ($-1.18$, significant). Neither shock produces a detectable increase in informality.

\subsubsection*{Placebo test}
Columns (3)--(4) apply the ITS specification to the pre-2014 period with a fake treatment date of 2008. If the 2013 break reflected a pre-existing trend change rather than TC/0168, we would expect significant placebo coefficients. The placebo estimate for vulnerable employment is 1.83 ($p = 0.21$) and for self-employment is 2.25 ($p = 0.098$). These marginally insignificant coefficients suggest some pre-existing trend instability---likely reflecting the 2008--2009 global financial crisis---but do not indicate a systematic confound at the 2013 treatment date.

\subsection{Provincial Exposure}

\input{tables/tab5_provinces}

Table 5 reports provincial Haitian-born population shares as a descriptive complement. The five border provinces (Dajab\'{o}n through Pedernales) have shares of 5--9\%, compared to 1--2\% in central provinces. Agricultural centers like San Pedro de Macor\'{i}s (6.0\%) and La Romana (6.0\%) also rank high. This variation in exposure intensity is consistent with the hypothesis that TC/0168's effects were concentrated in these provinces. However, without province-level labor market outcomes, I cannot test this directly; the national ITS captures only the population-weighted average effect, which is dominated by the low-exposure majority.

\section{Discussion}

The central finding of this paper is a null: mass denationalization left no detectable trace in the Dominican Republic's aggregate labor market statistics. This null has three implications.

\subsubsection*{The arithmetic of invisibility}
The affected population constitutes $\sim$2\% of the Dominican Republic's total population and is concentrated in low-wage agricultural and construction jobs. Even under the extreme assumption that every affected worker transitioned from formal to informal employment, the aggregate effect would be approximately 0.4 percentage points---below the minimum detectable effect in annual national data. This is not a limitation of the empirical strategy; it is a statement about who counts in aggregate statistics. When a state strips citizenship from a small, marginalized minority, the macroeconomy is designed not to notice.

\subsubsection*{What micro data could reveal}
The aggregate null does not mean TC/0168 was inconsequential. The ENI 2012 and ENI 2017 immigrant surveys contain province-level microdata on employment status, sector, formality, and documentation status that could directly identify the denationalized population and trace their labor market trajectories. These data are held by the Dominican Republic's Oficina Nacional de Estad\'{i}stica (ONE) and require institutional registration for access---a barrier that prevented their use in this analysis. With ENI microdata, a province-level continuous-treatment DiD exploiting variation in Haitian-descent population shares (ranging from 1\% to 9\% across provinces) would have far greater statistical power to detect effects concentrated in the affected subpopulation. A triple-difference design---comparing Haitian-descent versus non-Haitian-descent individuals in high- versus low-exposure provinces before and after 2013---could directly isolate the denationalization mechanism. The present paper establishes the aggregate null as a baseline and demonstrates why it was expected; the micro-level analysis would quantify the human cost hidden beneath that null.

\subsubsection*{Citizenship as labor market infrastructure}
The immigration enforcement literature focuses on border security and deportation threats targeting foreign-born populations \citep{east2023refusal, alsan2024fear}. TC/0168 represents a qualitatively different policy instrument: the revocation of citizenship from native-born individuals. The theoretical prediction---that removing legal status pushes workers into informality and compresses informal wages---remains untested at the micro level. If future work confirms large individual-level effects alongside the aggregate null documented here, it would demonstrate that citizenship functions as labor market infrastructure: its removal devastates the affected individuals while leaving the broader economy undisturbed. This would have implications for debates over immigration enforcement, statelessness, and the economic value of legal status \citep{clemens2018economics, gathmann2020access, dustmann2016economics}.

\section{Conclusion}

The Dominican Republic's 2013 TC/0168 ruling stripped citizenship from 210,000 people and left no mark on the aggregate labor market. Vulnerable employment, self-employment, wage work, unemployment, labor force participation, and school enrollment all continued their pre-existing trajectories without interruption. This aggregate null is itself the finding: it reveals that the costs of mass denationalization are borne entirely by the affected minority, statistically invisible in a growing economy. The affected population is too small and too marginalized to move the needle on national indicators---and that is perhaps the cruelest dimension of the policy.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Data sources and access}
All data used in this paper are publicly available through free APIs:

\begin{itemize}
\item \textbf{ILO SDMX API:} Queried for Dominican Republic (REF\_AREA=DOM), annual frequency, 2000--2024. Indicators: unemployment rate, labor force participation, employment-to-population ratio. No authentication required.

\item \textbf{World Bank WDI API:} \url{https://api.worldbank.org/v2/}. Country code DO, 2000--2024. Indicators: SL.EMP.VULN.ZS (vulnerable employment), SL.EMP.SELF.ZS (self-employment), SL.EMP.WORK.ZS (wage workers), SE.SEC.NENR (secondary enrollment), NY.GDP.PCAP.KD (GDP per capita), SI.POV.GINI, SL.AGR.EMPL.ZS, SL.SRV.EMPL.ZS. No authentication required.

\item \textbf{Dominican Republic 2010 Census:} Province-level population counts and Haitian-born shares from the Oficina Nacional de Estad\'{i}stica (ONE). Published in \citet{one2013eni}.

\item \textbf{DHS API:} \url{https://api.dhsprogram.com/rest/dhs/data}. Country code DR, subnational breakdown. 7 surveys (1986--2013). No authentication required.
\end{itemize}

\subsubsection*{Sample construction}
The analysis sample consists of 19 annual observations (2005--2023) for each national-level indicator. No observations are dropped. Province-level data use all 32 provinces from the 2010 Census with no exclusions.

\subsubsection*{Variable definitions}
\textit{Vulnerable employment} is defined by the ILO as the sum of own-account workers and contributing family workers, expressed as a percentage of total employment. \textit{Self-employment} includes own-account workers and employers. \textit{Wage workers} are employees with a formal employer-employee relationship. \textit{Post TC/0168} is an indicator equal to one from 2014 onward. \textit{Trend} is defined as (year $- 2013$), centering the time variable at the treatment date.

\section{Robustness Appendix}

\subsubsection*{Power calculation}
With $T = 19$ observations, an outcome standard deviation of $\sigma_Y \approx 2.1$ percentage points (for vulnerable employment), and Newey-West correction for serial correlation, the minimum detectable effect (MDE) at 80\% power and 5\% significance is approximately:
\[
\text{MDE} = 2.8 \times \frac{\sigma_Y}{\sqrt{T_{\text{post}}}} \approx 2.8 \times \frac{2.1}{\sqrt{10}} \approx 1.9 \text{ pp}
\]
The predicted aggregate effect depends on the baseline formal employment rate of the affected population. \citet{amuedo2017dominican} document that Haitian-descent workers in the Dominican Republic are overwhelmingly concentrated in informal agriculture and construction, with formal employment rates likely below 10\%. Under a generous upper-bound assumption of 20\% baseline formality, the aggregate effect of complete informalization would be $0.02 \times 100 \times 0.20 \approx 0.4$ pp. Under the more realistic 5--10\% baseline, the predicted effect shrinks to 0.1--0.2 pp---an order of magnitude below the MDE. Even under the most extreme scenario (all 210,000 affected individuals transition from formal to informal employment, with baseline formality of 20\%), the aggregate effect would be invisible in annual national data. The null is therefore consistent with devastating individual-level consequences.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
