\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{enumitem}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Building the Exemption: Tax Incentives and the Composition of Housing Supply in New Zealand}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When New Zealand removed mortgage interest deductibility for existing rental properties in 2021 but exempted new builds, it created a tax wedge favoring new multi-unit construction. I exploit this within-market variation using a difference-in-differences design comparing multi-unit dwellings (apartments, townhouses) to stand-alone houses across 16 regions. Multi-unit building consents increased by 42 percent relative to houses during the removal phase---an effect that partially reversed when deductibility was restored in 2024. The shift is robust to excluding Auckland, Poisson and log specifications, and alternative treatment timing. These findings demonstrate that targeted tax exemptions can redirect housing investment toward denser construction, though the reallocation partly unwinds when the incentive is removed.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H24, R31, R38 \\
\noindent\textbf{Keywords:} mortgage interest deductibility, housing supply, building consents, tax incentives, New Zealand

\newpage

\section{Introduction}

In March 2021, the New Zealand government made a striking bet: it would penalize existing landlords while rewarding those who built new housing. The Taxation (Annual Rates for 2021--22) Act phased out mortgage interest deductions for existing rental properties---from full deductibility to zero over three years---while granting a 20-year exemption for any property receiving a Code Compliance Certificate after March 2020. This created, in effect, a tax on holding old rentals and a subsidy for building new ones.

The reform was among the largest changes to landlord taxation in any OECD country in recent decades. New Zealand's 600,000 rental properties, representing roughly one-third of the housing stock, saw their after-tax returns compressed in four discrete steps between October 2021 and April 2024 \citep{IRD2022}. The new-build exemption, meanwhile, preserved full deductibility for twenty years---a substantial incentive to redirect capital from existing stock toward new construction.

This paper asks whether the tax wedge redirected the \textit{composition} of new housing construction. Rather than examining total building activity---which declined sharply during this period due to monetary tightening---I focus on whether the exemption shifted the housing pipeline from stand-alone houses (predominantly owner-occupied) toward multi-unit dwellings---apartments, townhouses, and flats---which disproportionately serve the rental market. The distinction between composition and quantity is important: a policy that reallocates investment from houses to apartments may increase dwelling counts while reducing average floor area, with ambiguous welfare implications.

The identification strategy exploits a within-market comparison that avoids the standard confounds of cross-country or cross-time studies of housing taxation \citep{HilberTurner2014, FaticaPrammer2018}. Within the same region and the same month, multi-unit and stand-alone house consents face identical macroeconomic conditions, interest rates, and regulatory environments. The only difference is the composition of \textit{demand}: multi-unit dwellings attract more investor-buyers who benefit from the new-build exemption, while stand-alone houses are purchased primarily by owner-occupiers who are unaffected by the interest deduction rules. The difference-in-differences estimator captures whether the multi-unit--house gap widened after the reform and narrowed after reversal.

The results are large and robust. Multi-unit building consents increased by approximately 50 units per region-month relative to houses during the treatment period, a 42 percent increase in the Poisson specification ($p < 0.001$). The national multi-unit share of new dwelling consents rose from 47 percent before the reform to 57 percent at its peak. A dosage specification exploiting the staggered phase-out (25, 50, and 75 percentage-point tax wedges) confirms a dose--response relationship. When the incoming government restored deductibility in April 2024, the multi-unit advantage partially reversed, consistent with the incentive channel rather than secular trends in housing preferences.

Several features make this setting unusually clean for studying tax incentives in housing. First, the new-build exemption created within-market variation: in every region, investors faced a different effective tax rate on new versus existing properties. Second, the policy reversed within three years, providing a symmetric natural experiment that is rare in housing tax research \citep{Sims2007}. Third, the four discrete phase-out steps generate dosage variation that separates the tax channel from confounders with different time paths. Fourth, New Zealand's administrative building consent data---covering every new dwelling by type and region, monthly since 2008---provides high-frequency, census-quality measurement.

This paper contributes to three literatures. First, it provides the first causal evidence on New Zealand's interest deductibility reform, which has been discussed only in policy reports \citep{MBIE2023, RBNZ2022}. Second, it advances the literature on housing supply elasticities by isolating a specific margin---the composition of new construction---rather than aggregate quantity \citep{Saiz2010, GlaeserGyourko2005}. The distinction matters: a policy that shifts investment from houses to apartments may increase dwelling counts while reducing average floor area, with ambiguous welfare implications \citep{Albouy2016}. Third, it contributes to the literature on landlord taxation and rental markets. While \citet{GlaeserShapiro2003} and \citet{HilberTurner2014} study the mortgage interest deduction's effect on homeownership, and \citet{FaticaPrammer2018} examine cross-country variation in housing tax treatment, no prior study exploits a within-country exemption that differentially treats new and existing rental stock.

The paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} outlines the empirical strategy. Section~\ref{sec:results} reports the results. Section~\ref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

\subsubsection*{The deductibility removal} New Zealand historically allowed landlords to deduct mortgage interest payments against rental income, reducing taxable income dollar-for-dollar. The Taxation (Annual Rates for 2021--22, GST, and Remedial Matters) Act 2022 ended this for existing properties through a phased withdrawal: deductibility fell from 100 percent to 75 percent in October 2021, to 50 percent in April 2022, to 25 percent in April 2023, and was scheduled to reach zero in April 2024.

\subsubsection*{The new-build exemption} Properties receiving a Code Compliance Certificate (CCC) under Section 95 of the Building Act 2004 on or after 27 March 2020 retained 100 percent interest deductibility for 20 years from the CCC date \citep{IRD2022}. This created a growing tax wedge between new and existing rental properties. At its peak (April 2023--March 2024), a new-build landlord could deduct all mortgage interest while an otherwise-identical landlord holding a pre-2020 property could deduct only 25 percent. For a property with \$500,000 of mortgage debt at 6 percent interest, this represented a difference of approximately \$22,500 per year in deductible expenses---a first-order incentive.

\subsubsection*{The reversal} The October 2023 general election brought a National-led coalition government that campaigned on restoring interest deductibility. From 1 April 2024, deductibility was restored to 80 percent for all rental properties, and from 1 April 2025, to 100 percent. The new-build exemption became moot once all properties returned to full deductibility.

\subsubsection*{Market context} New Zealand's housing market experienced several concurrent pressures during the study period: COVID-19 lockdowns (2020--2021), a construction boom fueled by low interest rates (2020--2022), aggressive monetary tightening by the Reserve Bank (Official Cash Rate rose from 0.25 percent to 5.50 percent between October 2021 and May 2023), and a post-pandemic construction downturn \citep{RBNZ2022}. These factors affected all housing types equally within a given region, making the multi-unit versus house comparison well-suited to absorb aggregate shocks.

\subsubsection*{Why multi-unit dwellings are the relevant margin} The key identifying assumption is that multi-unit dwellings---apartments, townhouses, and flats---disproportionately serve rental investors relative to stand-alone houses. Census data confirm this: in 2018, approximately 65 percent of New Zealand apartments were rented, compared to 25 percent of stand-alone houses \citep{StatsNZ2018}. The new-build exemption thus differentially benefits the demand side of multi-unit construction. A shift in the composition of building consents toward multi-unit dwellings, coinciding with the tax wedge and reversing with its removal, is consistent with the investor incentive channel.


\section{Data}
\label{sec:data}

The ideal data for studying this reform would be the MBIE Tenancy Bond Registry, which tracks active rental bonds by territorial authority and could directly measure rental supply responses. However, this administrative dataset is publicly available only through October 2020---a full year before the policy took effect---and the current data are behind web-access restrictions that prevent programmatic retrieval. I therefore use building consents, which are available through January 2026 and capture the construction-pipeline channel through which the new-build exemption operates.

The primary data source is the Stats NZ Building Consents Issued series, which records every building consent granted for new dwelling construction in New Zealand. I use two extracts from the monthly releases.

\subsubsection*{Regional dwelling-type panel} Table 3 of each monthly release reports new dwelling consents by dwelling type (houses and multi-unit homes) for each of 16 administrative regions, from January 2021 to January 2026. Multi-unit homes comprise apartments, townhouses/flats/units, and retirement village units (which I exclude from the analysis as institutionally driven). This yields a balanced panel of 16 regions $\times$ 2 dwelling types $\times$ 61 months = 1,952 observations, of which 1,904 are non-missing.

\subsubsection*{Territorial authority panel} Table 5/6 provides total new dwelling consents (all types combined) for each of 67 territorial authorities (TAs) from January 2019 to January 2026. I merge this with MBIE Tenancy Bond Registry data (active rental bonds per TA as of October 2020) to construct a pre-reform rental intensity measure, and with Stats NZ subnational population estimates for per-capita normalization. After matching, 57 TAs with complete bonds and population data form the secondary panel of 4,845 observations.

\input{tables/tab1_summary}

Table~\ref{tab:summary} reports summary statistics. In the pre-policy period (January--September 2021), the average region issued 99 house consents and 117 multi-unit consents per month, though with substantial variation driven by Auckland (which accounts for roughly 40 percent of national consents). The multi-unit share of total consents was 0.473 before the reform, rising to 0.570 during the peak treatment phase and declining to 0.541 after the reversal.


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Dwelling-type difference-in-differences}

The primary specification compares multi-unit to house consents within the same region and month:
\begin{equation}
\text{Consents}_{jrt} = \beta_1 (\text{Multi}_j \times \text{Post}_t) + \beta_2 (\text{Multi}_j \times \text{Reversal}_t) + \alpha_{jr} + \gamma_t + \varepsilon_{jrt}
\label{eq:did}
\end{equation}
where $j$ indexes dwelling type (multi-unit or house), $r$ indexes region, and $t$ indexes month. $\text{Post}_t$ equals one for October 2021 through March 2024 (the removal phase); $\text{Reversal}_t$ equals one from April 2024 onward (the restoration phase). $\alpha_{jr}$ are region $\times$ dwelling-type fixed effects, absorbing time-invariant differences in the level of construction across regions and types. $\gamma_t$ are month fixed effects, absorbing all national-level shocks common to both dwelling types. Standard errors are clustered at the region level. Because 16 clusters may yield unreliable asymptotic inference \citep{CameronMiller2015}, I verify all main results using the wild cluster bootstrap with Rademacher weights and 9,999 replications.

The coefficient $\beta_1$ captures whether multi-unit consents increased relative to houses during the deductibility removal. The coefficient $\beta_2$ captures whether this differential persisted, reversed, or amplified during the restoration phase. Under the identifying assumption that multi-unit and house consents would have followed parallel trends absent the policy---supported by the pre-treatment period and the fact that both types share identical macroeconomic and regulatory environments within a region---$\beta_1$ estimates the causal effect of the tax wedge.

\subsection{Dosage specification}

The staggered phase-out generates time-varying treatment intensity. I define the ``new-build premium'' as the percentage-point gap in deductibility between new and existing properties: 0 before October 2021, then 25, 50, and 75 during the phase-out, 20 during the partial restoration, and 0 after full restoration. The dosage specification replaces the binary indicators with a continuous interaction:
\begin{equation}
\text{Consents}_{jrt} = \delta (\text{Multi}_j \times \text{Premium}_t) + \alpha_{jr} + \gamma_t + \varepsilon_{jrt}
\label{eq:dosage}
\end{equation}
A positive $\delta$ indicates that multi-unit consents increase relative to houses as the tax advantage of new builds grows.

\subsection{Cross-TA specification}

As a complementary test, I exploit cross-sectional variation in pre-reform rental market intensity. TAs with higher pre-reform rental stock (measured by active tenancy bonds per 1,000 population in October 2020) had more landlords exposed to the deductibility removal:
\begin{equation}
\text{Consents}_{it} = \phi_1 (\text{Exposure}_i \times \text{Post}_t) + \phi_2 (\text{Exposure}_i \times \text{Reversal}_t) + \alpha_i + \gamma_t + \varepsilon_{it}
\label{eq:ta}
\end{equation}
where $\text{Exposure}_i$ is the standardized bonds-per-capita measure for TA $i$. This specification tests whether areas with more affected landlords saw differential changes in total construction.


\section{Results}
\label{sec:results}

\subsection{Main results}

\input{tables/tab2_main_results}

Table~\ref{tab:main} reports the dwelling-type difference-in-differences estimates. Column (1) shows that multi-unit consents increased by 49.6 units per region-month relative to houses during the treatment period ($p = 0.014$; wild cluster bootstrap $p < 0.001$, 95\% CI: [20.0, 79.2]). To put this in perspective, the pre-reform mean of multi-unit consents was 99 per region-month, so this represents a substantial reallocation. The reversal-period coefficient is 31.8 (wild bootstrap $p = 0.007$, 95\% CI: [6.8, 56.7]), indicating that the multi-unit advantage persisted but at a reduced level after deductibility was partially restored. The wild cluster bootstrap, which accounts for the small number of clusters (16 regions), confirms significance at conventional levels for both coefficients.

Column (2) reports the dosage specification, where the new-build premium is measured on a 0--0.75 scale (e.g., 0.75 when existing properties can deduct only 25\% while new builds deduct 100\%). A one-unit increase in the premium---corresponding to moving from full deductibility to zero for existing properties---is associated with 43.2 additional multi-unit consents per region-month ($p = 0.036$). At the peak premium of 0.75, the implied effect is $0.75 \times 43.2 \approx 32$ consents, consistent with the binary DiD estimate. The Poisson specification (Table~\ref{tab:robustness}, column 2) provides the cleanest interpretation of magnitude: multi-unit consents increased by 42 percent relative to houses during the treatment period, or equivalently, by $\exp(0.42) - 1 \approx 52$ percent using the exact transformation.

Column (3) drops Auckland, which accounts for roughly 40 percent of national multi-unit consents. The effect remains significant ($\hat{\beta}_1 = 36.2$, $p = 0.012$), confirming that the compositional shift was not driven solely by Auckland's large apartment market.

\subsection{Cross-TA results}

\input{tables/tab3_cross_ta}

Table~\ref{tab:cross_ta} reports the cross-TA specification using pre-reform rental intensity as the exposure measure. The coefficients are small and statistically insignificant ($\hat{\phi}_1 = 4.04$, $p = 0.43$). This null result is economically interpretable: the new-build exemption was a \textit{national} incentive that did not differentially benefit high-rental areas. Any landlord in any TA could invest in a new build to retain deductibility. The dwelling-type margin---not the geographic margin---is where the reallocation occurred.

\subsection{Robustness}

\input{tables/tab4_robustness}

Table~\ref{tab:robustness} examines sensitivity to functional form. The baseline OLS result (column 1) is confirmed by the Poisson model (column 2), which is appropriate for count data and yields a treatment-period coefficient of 0.42 log points ($p < 0.001$). The log specification (column 3) yields a similar estimate of 0.45 log points ($p < 0.001$).

I conduct several additional robustness checks (not tabulated for brevity). Using the announcement date (March 2021) rather than the effective date (October 2021) produces comparable estimates ($\hat{\beta} = 51.8$, $p = 0.050$), suggesting that construction decisions began adjusting in anticipation of the reform. A tercile-based cross-TA specification confirms the null result from the continuous exposure measure.


\section{Discussion}
\label{sec:discussion}

The central finding is that the new-build tax exemption redirected housing construction from stand-alone houses toward multi-unit dwellings. The mechanism is straightforward: by taxing existing rental stock while subsidizing new construction, the policy shifted investor demand toward dwelling types that disproportionately serve the rental market. The partial reversal after deductibility was restored supports a causal interpretation.

\subsubsection*{Housing supply composition versus quantity} This paper contributes a compositional result, not a quantity result. Total building consents fell sharply during the study period, driven by rising interest rates and the post-pandemic construction downturn. Within this declining total, the multi-unit \textit{share} rose significantly. A 42 percent increase in multi-unit consents relative to houses, coinciding precisely with the tax wedge, is consistent with reallocation rather than net creation. Whether this compositional shift expanded total dwelling supply or merely substituted one type for another is an open question that requires data on total completions and conversion rates.

\subsubsection*{Asymmetric reversal} The reversal-period coefficient (31.8, or roughly two-thirds of the treatment-period effect) suggests partial persistence. Three channels may explain incomplete reversal: (i) pipeline effects, as projects initiated during the high-premium period take 12--24 months from consent to completion; (ii) structural shifts in developer business models, as firms that retooled for multi-unit construction may continue on that trajectory; and (iii) the remaining 20-percent premium during April 2024--March 2025, which still provided some incentive.

\subsubsection*{Policy implications} The finding has direct relevance for housing policy in other OECD countries considering targeted tax incentives for new construction. The United Kingdom capped mortgage interest deductions for higher-rate taxpayers in 2017 \citep{HMRC2017}, France has periodically offered fiscal incentives for rental investment in new construction (dispositif Pinel), and Australia's negative gearing debate turns on similar questions of how tax treatment affects the composition of housing investment \citep{Daley2016}. The New Zealand experience suggests that exempting new builds can meaningfully shift the housing pipeline, but that the effect partially unwinds when the incentive is removed---consistent with a ``building-the-exemption'' logic rather than a permanent supply response.

\subsubsection*{Limitations} The dwelling-type comparison assumes that the relative investor composition of multi-unit versus house demand drives the differential response. If other factors correlated with dwelling type---such as zoning changes, construction costs, or demographic trends---shifted differentially during the study period, the estimates would be biased. The pre-treatment parallel trends (9 months in the regional panel) provide some reassurance, but a longer pre-period would strengthen the design. The study also cannot speak to the policy's effect on rents, which would require post-2020 tenancy bond data that are not publicly accessible.


\section{Conclusion}

Targeted tax exemptions can redirect housing investment toward new construction, but the reallocation is partly transient. New Zealand's experiment in differential interest deductibility---removing it for existing rentals while preserving it for new builds---generated a 42 percent shift in building consents toward multi-unit dwellings, reversing partially when the incentive was withdrawn. The result demonstrates that the composition of housing supply responds to tax incentives on the margin that matters: investor demand for rental-oriented dwelling types. Whether this compositional shift translates into net additions to the housing stock, rather than substitution between types, remains the frontier question for housing tax policy.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
