\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Reshoring Myth: CHIPS Act Investments and the Absence of Local Housing Market Disruption}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The CHIPS and Science Act directed \$52.7 billion toward semiconductor manufacturing, concentrating billion-dollar fab investments in 21 U.S.\ counties. Media coverage and local housing advocates warned of surging rents and displacement. Using monthly Zillow data for 3,073 counties and a staggered difference-in-differences design exploiting the timing of funding announcements, I find no evidence that CHIPS Act investments increased local home values or rents. The Callaway--Sant'Anna ATT for home values is $-$0.24 percent (SE 0.60~pp), and the 95 percent confidence interval rules out effects larger than 1 percent. Rents, if anything, grew slightly slower in recipient counties. The null survives leave-one-out, randomization inference, and heterogeneity by award size. These results suggest that county-level housing markets absorbed the reshoring shock without measurable disruption---a finding relevant for the design of place-based industrial policies.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R31, L52, H25, O25 \\
\noindent\textbf{Keywords:} CHIPS Act, industrial policy, housing prices, semiconductor manufacturing, difference-in-differences

\newpage

\section{Introduction}

The United States is making the largest public investment in semiconductor manufacturing in its history. The CHIPS and Science Act of 2022 allocated \$52.7 billion to rebuild domestic chip production, with individual awards exceeding \$8 billion for facilities expected to employ thousands of construction and permanent workers. Within months of the first funding announcements in early 2024, media reports warned of ``CHIPS Act housing crunches'' in recipient communities, and local advocacy groups raised alarms about displacement of low-income renters. The fear was intuitive: inject billions into a county, attract thousands of workers, and watch housing costs surge.

This paper tests that fear directly. I exploit the staggered timing of CHIPS Act funding announcements across 21 U.S.\ counties to estimate the causal effect on local housing costs. Using monthly county-level data from the Zillow Home Value Index (ZHVI) and the Zillow Observed Rent Index (ZORI) for over 3,000 counties, I implement a Callaway--Sant'Anna \citeyearpar{callaway2021difference} staggered difference-in-differences design. Treatment is defined as the first public funding announcement for a semiconductor facility in a given county, with announcement dates spanning January 2024 through March 2025.

The main finding is a precisely estimated null. The overall ATT for home values is $-$0.24 percent (SE 0.60~pp), with a 95 percent confidence interval of [$-$1.4\%, 0.95\%]. This rules out home value increases larger than 1 percent---roughly two and a half months' national appreciation---at the 95 percent level. For rents, the evidence, if anything, points in the opposite direction: the Callaway--Sant'Anna ATT is $-$1.20 percent (SE 0.60~pp), and a static TWFE specification yields a statistically significant $-$3.46 percent effect. The event study shows clean pre-trends for both outcomes, with all pre-treatment coefficients within 0.2 percentage points of zero.

The null is robust across multiple dimensions. Leave-one-out analysis dropping each treated county in turn confirms that no single county drives the null result (see \Cref{tab:loo}). Randomization inference randomly reassigning treatment status across 500 permutations yields a $p$-value of 0.99, confirming that our null estimate is squarely within the placebo distribution. Splitting treated counties by award size (above and below the median of \$0.75 billion) produces null effects for both subgroups.

This paper contributes to three literatures. First, it adds to the nascent empirical literature on the CHIPS Act. \citet{erten2025chips} study the employment and wage effects of CHIPS investments, but housing market consequences remain unexamined. The housing dimension is the distributional flip side of job creation: if fab workers drive up local housing costs, the welfare gains from industrial policy are partly offset by losses to incumbent renters---the ``reshoring rent'' that policy advocates prefer not to discuss. My results suggest this cost may be smaller than feared.

Second, the paper contributes to the literature on local housing market responses to economic shocks. \citet{greenstone2010identifying} show that ``million dollar plant'' openings increase housing values by 1.1--1.7 percent, while \citet{diamond2019housing} and \citet{ganong2023housing} document substantial housing cost responses to local labor demand shocks. More recently, \citet{bartik2019should} and \citet{slattery2020evaluating} note that place-based subsidies can generate housing price capitalization that partially offsets their benefits. My null result contrasts with these findings, suggesting that either the CHIPS shock is too recent, too diffuse at the county level, or targeted at counties with sufficient housing supply elasticity.

Third, the finding speaks to the broader debate on housing supply elasticity and the distributional effects of place-based policies. \citet{saiz2010geographic} documents wide variation in housing supply elasticity across U.S.\ metro areas. CHIPS Act investments were disproportionately directed toward counties in the Sun Belt and exurban areas---precisely the geographies where \citet{glaeser2005housing} argue supply is most elastic. If housing supply absorbs labor demand shocks without price increases, the standard concern about capitalization is muted.

The paper proceeds as follows. \Cref{sec:background} describes the CHIPS Act and its implementation. \Cref{sec:data} presents the data. \Cref{sec:strategy} outlines the empirical strategy. \Cref{sec:results} reports results. \Cref{sec:discussion} discusses mechanisms and limitations.

\section{Institutional Background}
\label{sec:background}

The CHIPS and Science Act was signed into law on August 9, 2022, allocating \$52.7 billion for domestic semiconductor manufacturing incentives. The Department of Commerce's CHIPS Program Office administers the funds through competitive applications, evaluating proposals based on economic viability, workforce development, and national security considerations.

\subsubsection*{Staggered Funding Announcements} The Commerce Department announced awards in a staggered fashion beginning in January 2024. BAE Systems received the first award (\$35 million) on January 26, 2024, for its Nashua, New Hampshire facility. Major awards followed: GlobalFoundries (\$1.5 billion, February 2024), Intel (\$8.5 billion, March 2024), TSMC (\$6.6 billion, April 2024), Samsung (\$6.4 billion, April 2024), and Micron (\$6.1 billion, April 2024). Later awards extended through 2025, including Texas Instruments (\$1.6 billion, August 2024), Wolfspeed (\$750 million, November 2024), and SK Hynix (\$458 million, March 2025).

\subsubsection*{Geographic Distribution} Awards are concentrated in 21 counties across 15 states. Recipients span a wide range of existing economic conditions: Maricopa County, Arizona (TSMC, population 4.5 million) and Franklin County, Ohio (Intel, population 1.3 million) are large metropolitan counties, while Chatham County, North Carolina (Wolfspeed, population 76,000) and Tippecanoe County, Indiana (SK Hynix, population 195,000) are substantially smaller. The geographic distribution reflects the semiconductor industry's preference for sites with abundant water, reliable power, and available land---characteristics common in Sun Belt and exurban locations.

\subsubsection*{Expected Labor Demand} Each major fab is expected to employ 3,000--10,000 construction workers during peak building phases and 1,000--3,000 permanent workers upon completion. These employment figures, while large in absolute terms, represent a modest share of total county employment for the larger metropolitan recipients. In Maricopa County, for example, TSMC's projected 6,000 permanent jobs represent approximately 0.3 percent of the county's 2.1 million employed residents.

\section{Data}
\label{sec:data}

\subsubsection*{Housing Prices} I use the Zillow Home Value Index (ZHVI), a smoothed, seasonally adjusted measure of typical home values at the county-month level. ZHVI covers 3,073 counties from January 2000 through February 2026 (314 months). For rents, I use the Zillow Observed Rent Index (ZORI), available for 1,274 counties from January 2015 onward. Both measures are publicly available from Zillow Research.

\subsubsection*{CHIPS Act Announcements} I compile CHIPS Act funding announcements from Department of Commerce press releases, recording the county FIPS code, announcement date, award amount, and recipient company for each funded facility. When multiple awards target the same county (e.g., TSMC and Amkor Technology in Maricopa County), I use the earliest announcement date and sum the award amounts. This yields 21 unique treated counties with total awards of \$42.9 billion.

\subsubsection*{County Characteristics} Baseline county characteristics come from the 2021 American Community Survey 5-year estimates, including population, median household income, total housing units, median home value, and median gross rent.

\subsubsection*{Analysis Sample} I restrict the panel to January 2020 through February 2026 (74 months), providing at least 36 months of pre-treatment data for the earliest announcements.

\input{tables/tab1_summary}

\Cref{tab:summary} presents baseline characteristics as of January 2022. Treated counties are larger, wealthier, and have higher home values than the average U.S.\ county, reflecting the semiconductor industry's metropolitan orientation.

\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Identification} I exploit the staggered timing of CHIPS Act funding announcements to estimate the effect on housing costs using a difference-in-differences framework. The identifying assumption is that, absent the announcement, housing costs in treated counties would have evolved in parallel with never-treated counties. The key threat is that announcement timing is correlated with pre-existing county-level housing market trends. I assess this through the event study, which shows no evidence of differential pre-trends.

\subsubsection*{Estimation} The primary estimator is the group-time ATT of \citet{callaway2021difference}, which is robust to heterogeneous treatment effects across cohorts---a first-order concern given that award sizes range from \$24 million (Rocket Lab) to \$8.5 billion (Intel). I define the treatment group $g$ by the month of the first funding announcement in each county, with $g = 0$ for never-treated control counties. The group-time ATT is:
\begin{equation}
\text{ATT}(g,t) = \E\left[Y_{i,t}(g) - Y_{i,t}(0) \mid G_i = g\right]
\end{equation}
where $Y_{i,t}(g)$ is the potential outcome under treatment at time $g$, and $Y_{i,t}(0)$ is the potential outcome under no treatment. I aggregate group-time effects into an overall ATT and dynamic event-study coefficients. Standard errors are clustered at the county level.

I supplement the Callaway--Sant'Anna estimator with two TWFE specifications: (i) a static DiD interacting the treatment indicator with a post-announcement dummy, and (ii) a continuous dose--response specification interacting the award amount per housing unit with the post indicator. Both include county and month fixed effects with county-clustered standard errors.

\subsubsection*{Threats to Validity} The main concern is selection on pre-trends: counties receiving CHIPS awards may have been on different housing trajectories before the announcement. The event study addresses this directly---all 24 pre-treatment coefficients fall within $\pm$0.2 percentage points of zero, and a joint Wald test fails to reject the null of zero pre-treatment effects. A second concern is anticipation: housing markets may have priced in the expected investment before the formal announcement. I address this by testing for pre-trends extending 24 months before the announcement; no anticipatory movement is detected. A third concern is that county-level aggregation may mask localized effects near fab sites---a construction site employing 5,000 workers in a county with 500,000 housing units adds only 1 percent to housing demand. I discuss this limitation explicitly in \Cref{sec:discussion}. Fourth, with 21 treated clusters, asymptotic cluster-robust standard errors may be imprecise. The randomization inference (RI $p = 0.99$) provides a nonparametric check that does not depend on large-cluster asymptotics.

\section{Results}
\label{sec:results}

\subsubsection*{Main Estimates} \Cref{tab:main} reports the main results. The Callaway--Sant'Anna ATT for home values (column 1) is $-$0.24 percent with a standard error of 0.60 percentage points. The 95 percent confidence interval of [$-$1.42\%, 0.95\%] rules out economically meaningful positive effects: we can reject home value increases larger than 1 percent at the 95 percent level. For context, national ZHVI appreciation averaged approximately 0.4 percent per month over 2024--2025, so we can rule out an acceleration equivalent to roughly two and a half months' national trend.

The static TWFE estimate (column 2) is 0.71 percent (SE 1.59~pp), consistent with the Callaway--Sant'Anna null but less precise. The dose--response specification (column 3), which uses the dollar award per housing unit as a continuous treatment, is also indistinguishable from zero.

For rents, the results are more striking. The Callaway--Sant'Anna ATT (column 4) is $-$1.20 percent (SE 0.60~pp). The static TWFE estimate (column 5) shows a larger and statistically significant effect of $-$3.46 percent (SE 1.13~pp). Rather than the feared ``reshoring rent,'' CHIPS counties experienced rent growth that was, if anything, slower than comparable control counties.

\input{tables/tab2_main}

\subsubsection*{Event Study} The dynamic event-study estimates from the Callaway--Sant'Anna aggregation show clean pre-trends. All 24 pre-treatment coefficients for home values fall within $\pm$0.2 percentage points of zero, with simultaneous confidence bands comfortably covering zero throughout. Post-treatment coefficients drift slightly negative but remain small and statistically insignificant through 18 months post-announcement. The absence of pre-trends validates the parallel trends assumption. To translate the point estimate into dollar terms: the ATT of $-$0.24 percent applied to the treated-county average home value of \$339,000 corresponds to approximately \$810---an economically negligible magnitude that is indistinguishable from zero.

\subsubsection*{Robustness} \Cref{tab:robustness} summarizes the robustness analysis. The baseline Callaway--Sant'Anna ATT is insensitive to multiple perturbations. Leave-one-out analysis, reported in detail in \Cref{tab:loo}, confirms that no single treated county drives the null result. The narrow range across all 21 iterations demonstrates that the null is a feature of the entire treated sample, not the cancellation of large opposing effects from individual counties.

Randomization inference provides a nonparametric assessment of statistical significance. I randomly assign treatment status to 21 control counties 500 times, drawing treatment timing from the empirical distribution of actual announcement dates. The resulting RI $p$-value is 0.99: our observed ATT is unremarkably small relative to the placebo distribution, as expected under the null hypothesis.

\input{tables/tab3_robustness}

\subsubsection*{Heterogeneity} Splitting treated counties at the median award amount (\$0.75 billion) reveals null effects for both subgroups. Large-award counties (including Intel, TSMC, Samsung, and Micron, with individual awards exceeding \$6 billion) show an ATT of $-$0.18 percent (SE 1.23~pp). Small-award counties show an ATT of $+$0.13 percent (SE 0.73~pp). Even for the largest investments in U.S.\ industrial policy history, housing markets did not respond at the county level within the first one to two years.

\input{tables/tab4_loo}

\section{Discussion}
\label{sec:discussion}

The central finding---that \$42.9 billion in semiconductor investments produced no measurable county-level housing market response---admits several interpretations.

\subsubsection*{Housing Supply Elasticity} The most natural explanation is that CHIPS counties possess sufficient housing supply elasticity to absorb the labor demand shock. Many recipients are in Sun Belt metros (Maricopa County, Williamson County, TX) or mid-sized cities (Columbus, OH; Boise, ID) where land availability and permissive zoning allow construction to respond to demand increases. This aligns with \citet{saiz2010geographic} and \citet{glaeser2005housing}, who document that housing prices respond less to demand shocks in supply-elastic markets. If this interpretation is correct, the CHIPS Act's geographic targeting---whether intentionally or not---channeled investment toward locations where housing market disruption was least likely.

\subsubsection*{Aggregation and Scale} County-level measurement may be too coarse to detect localized effects near specific fab sites. A construction site employing 5,000 workers in a county with 500,000 housing units adds only 1 percent to housing demand---an effect easily absorbed at the county level even if rents in the immediate neighborhood increased substantially. Sub-county data (ZIP code or census tract) would be needed to detect such hyper-local capitalization.

\subsubsection*{Timing} Most CHIPS Act fabs remain under construction as of early 2026. The permanent workforce that would generate sustained housing demand has not yet arrived. The null may therefore reflect timing rather than a permanent absence of effect. However, the construction workforce alone is large---often 5,000--10,000 workers per site---and we find no price response even to this initial wave of labor demand.

\subsubsection*{The Muted Rent Effect} The slightly negative rent coefficient deserves comment. One possibility is that CHIPS announcements signaled future economic prosperity, encouraging new construction that expanded supply faster than demand grew. Another is that the large, metropolitan counties receiving CHIPS awards experienced rent growth moderation during 2024--2025 for reasons unrelated to the CHIPS Act---perhaps as part of a broader Sun Belt rental market correction documented by industry analysts.

\subsubsection*{Policy Implications} For policymakers designing place-based industrial subsidies, the short-run null is cautiously good news: the housing cost concern that dominates popular discussion of reshoring may be overstated, at least when investments target supply-elastic markets. But the null also challenges the mechanism through which place-based policies generate broader community benefits. If the housing market does not capitalize the income gains from new employment, then either the labor market adjustment is incomplete (workers are commuting from elsewhere rather than relocating) or the gains accrue narrowly to the directly employed. Both possibilities warrant further investigation as CHIPS Act fabs reach full operation.

\section{Conclusion}

The largest industrial policy investment in modern American history produced no detectable effect on county-level housing costs in its first two years. This finding challenges the ``reshoring rent'' narrative---the widely held fear that manufacturing investments in semiconductor fabs would drive up housing costs and displace incumbent residents. Our 95 percent confidence intervals rule out effects larger than 1 percent of home values, bounding the distributional cost of reshoring through local housing markets at a modest level.

The null is honest: it may reflect supply elasticity, geographic dilution, or simply early timing. As construction gives way to permanent operations at the 21 treated fab sites, the housing market response may eventually materialize. But the immediate evidence suggests that the counties selected for CHIPS investments---largely Sun Belt and exurban locations with available land and expanding housing stocks---were well positioned to absorb the initial shock. For the design of future place-based industrial policies, this is a useful fact: where you put the factory matters not just for labor markets but for whether the investment triggers the housing affordability spiral that critics fear.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
