\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Clearing the Deck: SEC Chair Transitions, Enforcement Vacuums, and the Resilience of Capital Markets}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When SEC Chair Gary Gensler departed in January 2025, the agency's monthly enforcement rate fell from 83 to 3 actions---a 97\% collapse unprecedented in the Commission's history. We document that this ``clearing the deck'' phenomenon is a systematic institutional regularity: fiscal years containing Chair transitions see 18.6 percentage points larger enforcement declines than non-transition years ($p = 0.032$, permutation test). Cross-party transitions produce particularly severe vacuums, averaging a 21\% decline versus 7\% for same-party transitions. Despite these dramatic enforcement disruptions, we find no significant aggregate market response: neither the VIX nor financial sector excess returns react to transition-induced enforcement vacuums within $\pm$60 trading days. This null aggregate result---while unable to rule out firm-level effects on companies under active investigation---suggests that temporary enforcement lapses do not destabilize securities markets at the index level, consistent with anticipation, private enforcement substitution, or stock-based deterrence.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G18, G28, K22, D73 \\
\noindent\textbf{Keywords:} SEC enforcement, regulatory transitions, enforcement vacuum, capital markets, financial regulation

\newpage

\section{Introduction}

On the morning of January 20, 2025, Securities and Exchange Commission Chair Gary Gensler submitted his resignation. In the 3.5 months preceding his departure, the SEC had filed 291 standalone enforcement actions---an annualized pace of roughly 1,000 cases per year, more than double the Commission's historical norm. In the 8.5 months that followed under Acting Chair Mark Uyeda and then Chair Paul Atkins, the agency filed just 22 additional actions. The enforcement apparatus of American securities regulation had, for all practical purposes, shut down.

This pattern---an outgoing Chair rushing to ``clear the deck'' of pending cases followed by a vacuum as the incoming Chair rebuilds---is well known to securities lawyers \citep{jackson2007policing}. Practitioners at major law firms routinely advise clients that enforcement risk varies systematically with the political calendar. Yet despite its practical significance for regulated firms, the ``clearing the deck'' phenomenon has never been subjected to rigorous econometric analysis. Does enforcement intensity truly decline at Chair transitions, or does the anecdotal evidence reflect selection? If enforcement vacuums are real, do they affect capital market outcomes? And does the severity of the vacuum depend on the political alignment between outgoing and incoming Chairs?

This paper provides the first systematic quantification of SEC Chair transition effects on enforcement intensity and capital market outcomes. Using Cornerstone Research enforcement statistics spanning fiscal years 2010--2025, we exploit the timing of four Chair transitions---three cross-party and one same-party---to test whether enforcement disruptions are a predictable institutional regularity. We supplement the enforcement data with daily stock market data from Yahoo Finance to test whether financial markets price the enforcement risk associated with Chair transitions.

Our central finding is a stark asymmetry: enforcement vacuums are large, systematic, and statistically significant, but capital markets are largely indifferent to them. A permutation test across 16 fiscal years shows that transition years experience enforcement declines averaging 17.1\%, compared to a 1.4\% average \emph{increase} in non-transition years---a difference of 18.6 percentage points significant at the 3.2\% level. Cross-party transitions, where the incoming Chair holds fundamentally different enforcement priorities, produce an average decline of 20.7\%, compared to just 6.5\% for the single same-party transition in our sample. Yet neither the CBOE Volatility Index nor financial sector excess returns (measured as the spread between the Financial Select Sector SPDR Fund and the S\&P 500 ETF) respond significantly to Chair transition dates.

This paper contributes to three literatures. First, we add to the growing body of work on political cycles in regulation \citep{correia2014political, mehta2019regulatory}. While \citet{correia2014political} shows that politically connected firms face lower SEC enforcement risk and \citet{pandey2025partisan} documents partisan variation in which firms are targeted, neither study examines the aggregate enforcement disruption caused by the transition itself. Our contribution is to show that the transition mechanism---the institutional process of replacing one Chair with another---creates a distinct enforcement distortion independent of the incoming Chair's policy preferences.

Second, we contribute to the literature on regulatory deterrence in financial markets \citep{kedia2006economics, dyck2010who, karpoff2017proximate}. The dominant view holds that SEC enforcement deters securities fraud through the threat of sanctions \citep{jackson2007policing}. Our null market result complicates this picture: if enforcement intensity were a first-order determinant of market integrity, we would expect financial sector stocks to respond when enforcement drops by 97\%. The absence of a response suggests either that markets anticipate the vacuum (and its temporary nature), that private enforcement mechanisms substitute for public oversight during transition periods, or that the deterrent effect operates through the \emph{stock} of past enforcement rather than the \emph{flow} of current actions.

Third, we speak to institutional design questions in financial regulation \citep{romano2014regulating}. The concentrated authority of a single Chair over enforcement priorities creates a principal-agent problem at the apex of securities regulation: when the Chair leaves, the agent literally stops functioning. Our findings suggest that while this design creates significant enforcement disruptions, the market's self-regulatory infrastructure---private litigation, auditor oversight, exchange listing requirements, and reputational mechanisms---provides sufficient resilience to prevent systemic consequences. Whether this resilience justifies the current design or merely masks its costs is a question we raise but do not resolve.

\section{Institutional Background}

\subsubsection*{The SEC Chair and Enforcement Authority} The Securities and Exchange Commission is led by five Commissioners appointed by the President and confirmed by the Senate, with one designated as Chair. While enforcement decisions are nominally made by the full Commission, the Chair exercises outsized influence through control of the agency's agenda, the appointment of the Director of Enforcement, and the power to prioritize which cases move forward \citep{romano2014regulating}. In practice, the Chair's enforcement philosophy determines the volume, type, and aggressiveness of SEC enforcement activity.

\subsubsection*{The Transition Mechanism} Chair transitions occur through two channels. First, when a new President takes office, the incumbent Chair typically resigns, creating an ``acting Chair'' period before the new appointee is confirmed. Second, Chairs occasionally resign mid-term for personal reasons. The enforcement disruption operates through several channels. Before departure, the outgoing Chair accelerates pending cases to establish a legacy and prevent the incoming Chair from dismissing them---the ``clearing the deck'' phenomenon documented by practitioners \citep{jackson2007policing}. After arrival, the incoming Chair reviews ongoing investigations, reassigns staff, and establishes new enforcement priorities, creating a period of reduced activity.

\subsubsection*{The Four Transitions in Our Sample} Our sample spans four Chair transitions between fiscal years 2010 and 2025. The Schapiro-to-White transition (December 2012) was same-party: both Democratic appointees with similar enforcement philosophies. The remaining three---White-to-Clayton (January 2017), Clayton-to-Lee/Gensler (December 2020), and Gensler-to-Uyeda/Atkins (January 2025)---were cross-party transitions coinciding with Presidential inaugurations. The FY2025 transition is historically extreme: 93\% of the year's enforcement actions were filed before Chair Gensler's departure on January 20, with total enforcement falling to 313 actions (the lowest since at least FY2010) and total penalties collapsing to \$808 million (the lowest since FY2012).

\subsubsection*{Enforcement Data} We use Cornerstone Research's annual SEC enforcement statistics, the standard academic source for aggregate enforcement counts \citep{urska2012securities}. Cornerstone distinguishes ``standalone'' enforcement actions (new cases) from ``follow-on'' proceedings (subsequent actions in existing cases) and delinquent filer cases. Our primary outcome is total enforcement actions per fiscal year, which captures the full scope of SEC enforcement activity. We supplement the aggregate data with individual enforcement action dates scraped from the SEC's online litigation releases and administrative proceedings archives, covering 1,318 dated actions.

\section{Empirical Strategy}

Our identification strategy exploits the quasi-random timing of Chair transitions within the institutional calendar of SEC enforcement. The key identifying assumption is that, absent a Chair transition, enforcement intensity would have evolved smoothly from year to year. We implement three complementary approaches.

\subsubsection*{FY-Level Comparison} Our primary specification compares enforcement intensity in transition fiscal years to adjacent non-transition years. For each transition $j$, we compute the year-over-year percentage change in total enforcement actions:
\begin{equation}
\Delta_j = \frac{E_{FY(j)} - E_{FY(j)-1}}{E_{FY(j)-1}} \times 100
\end{equation}
where $E_{FY}$ is total enforcement actions in fiscal year $FY$ and $FY(j)$ is the fiscal year containing transition $j$. We then compare the average $\Delta$ for transition years ($N=4$) to non-transition years ($N=11$) using a permutation test with 10,000 replications.

\subsubsection*{Cross-Party versus Same-Party} We decompose the transition effect by political alignment. Cross-party transitions ($N=3$: White-to-Clayton, Clayton-to-Lee/Gensler, Gensler-to-Uyeda/Atkins) involve a change in the Chair's party affiliation and should produce larger enforcement disruptions as the incoming Chair reverses course on enforcement priorities. The single same-party transition (Schapiro-to-White) serves as a within-design control: same institutional process but without the political reorientation.

\subsubsection*{Market Event Study} We estimate the capital market response to Chair transitions using a panel event study:
\begin{equation}
Y_{it} = \alpha_j + \beta \cdot \text{Post}_{jt} + \varepsilon_{it}
\end{equation}
where $Y_{it}$ is a daily market outcome (VIX close or financial sector excess return), $\alpha_j$ is a transition fixed effect, and $\text{Post}_{jt} = \ind[t \geq T_j]$ indicates the post-transition period. The event window is $\pm$60 trading days around each transition date $T_j$. Financial sector excess returns are computed as the daily log return of XLF (Financial Select Sector SPDR Fund) minus SPY (SPDR S\&P 500 ETF). Standard errors are clustered by transition event.

The key threat to identification is that Chair transitions coincide with broader political transitions---particularly for cross-party transitions, which align with Presidential inaugurations. We address this concern in three ways: (i) the permutation test is nonparametric and does not rely on a specific functional form for the counterfactual; (ii) placebo tests using non-transition dates control for calendar effects; and (iii) the same-party/cross-party decomposition provides a natural benchmark, as both types share the same institutional transition process but differ in political alignment.

\section{Results}

\subsection{Enforcement Declines at Chair Transitions}

Table \ref{tab:enforcement_fy} presents SEC enforcement actions by fiscal year. The four transition years (marked with daggers) show a clear pattern of enforcement decline. FY2013 (Schapiro-to-White, same party) saw a modest 6.5\% decline. FY2017 (White-to-Clayton, cross party) experienced a 13.1\% decline. FY2021 (Clayton-to-Lee/Gensler, cross party) showed a small 2.5\% decline, attenuated by the fact that Acting Chair Lee maintained Clayton-era case momentum during her brief tenure. FY2025 (Gensler-to-Uyeda/Atkins, cross party) produced the most dramatic decline: a 46.3\% collapse from 583 to 313 total actions.

Table \ref{tab:transition_effects} summarizes the transition effects. The average year-over-year change in transition years is $-17.1$\%, compared to $+1.4$\% in non-transition years---a difference of 18.6 percentage points. The permutation test assigns four ``transition'' labels randomly among 15 fiscal year changes and asks how often the resulting mean difference exceeds the observed 18.6 pp. The two-sided $p$-value is 0.032, rejecting the null of no transition effect at conventional levels. Cross-party transitions average a $-20.7$\% change versus $-6.5$\% for the single same-party transition, consistent with the hypothesis that enforcement disruption is amplified when the incoming Chair has different priorities.

\subsection{Capital Market Non-Response}

Table \ref{tab:market_response} reports the market event study results. Despite the dramatic enforcement disruptions documented above, aggregate capital markets show no significant response. The VIX decreases by 0.42 points in the post-transition period, but the estimate is statistically indistinguishable from zero (SE = 1.07, $p = 0.72$). Financial sector excess returns show a small positive coefficient of 0.04 basis points, also insignificant (SE = 0.03 bp, $p = 0.29$). The within-$R^2$ values are near zero (0.001--0.004), confirming that Chair transitions explain virtually none of the day-to-day variation in aggregate market outcomes.

We emphasize that this null result applies to \emph{aggregate} market outcomes. At 80\% power, our design can detect VIX changes of approximately $\pm$2.95 points (roughly 0.9 standard deviations of the pre-transition VIX distribution) and financial excess return changes of $\pm$0.08 basis points. Effects smaller than these thresholds---or effects concentrated among specific firms under active investigation rather than dispersed across the financial sector---would escape detection in our aggregate event study. The null result should therefore be interpreted as ruling out large, sector-wide market responses to enforcement vacuums, not as evidence that enforcement transitions have no effect on any capital market outcome.

\subsection{Robustness}

Table \ref{tab:robustness} presents four sets of robustness checks. Panel A shows that the null VIX result is stable across event window bandwidths of $\pm$30 to $\pm$90 trading days, with point estimates ranging from $-0.42$ to $+0.07$---all statistically insignificant. Panel B excludes the extreme FY2025 transition: the VIX estimate becomes marginally significant ($-1.45$, $p = 0.054$), suggesting that the three pre-2025 transitions may have slightly reduced uncertainty, but the effect is fragile and of modest economic magnitude (roughly 1.5 VIX points on a mean of $\sim$18).

Panel C reports placebo tests using five mid-fiscal-year dates (April 15) in non-transition years. The placebo VIX estimate ($-1.83$, $p = 0.10$) is actually \emph{larger} than the real transition effect, indicating that VIX fluctuations of comparable magnitude occur regularly at arbitrary dates. The financial excess return placebo is also insignificant ($p = 0.45$). Panel D confirms the cross-party/same-party differential at the fiscal year level: cross-party transitions average a $-20.7$\% enforcement change versus $-6.5$\% for same-party, a 14.1 percentage point gap.

\section{Discussion}

Our results establish two facts about SEC enforcement and capital markets. First, Chair transitions create large, predictable enforcement vacuums---transition years see enforcement declines roughly 19 percentage points larger than non-transition years, with cross-party transitions producing the most severe disruptions. Second, capital markets are remarkably resilient to these enforcement shocks: neither market-wide uncertainty nor financial sector stock prices respond detectably to the onset of an enforcement vacuum.

\subsubsection*{Why Don't Markets React?} Three mechanisms could explain the null market response. The \emph{anticipation channel}: Chair transitions are scheduled, politically determined events that markets can foresee months in advance. By the time the transition occurs, any enforcement-related repricing has already been incorporated into asset prices. The \emph{substitution channel}: when SEC enforcement retreats, private enforcement---securities class action lawsuits, shareholder derivative suits, exchange listing enforcement---may expand to fill the gap \citep{coffee2007reforming, rose2008multienforcer}. If private enforcement is a close substitute for public enforcement, the net enforcement intensity faced by firms may remain stable even as the SEC's contribution fluctuates. The \emph{stock-versus-flow channel}: deterrence may depend on the accumulated \emph{stock} of past enforcement actions---the demonstrated willingness and ability to prosecute---rather than the \emph{flow} of current filings \citep{kedia2006economics}. A temporary enforcement vacuum does not erase the SEC's track record or its statutory authority; rational actors expect enforcement to resume.

\subsubsection*{Limitations} Our analysis faces four important limitations. First, the sample contains only four transitions, limiting statistical power for the market event study and precluding formal inference on the cross-party/same-party differential. With one same-party and three cross-party transitions, the comparison is suggestive rather than definitive. Second, cross-party Chair transitions coincide with Presidential inaugurations, which bring simultaneous changes in monetary policy expectations, trade policy, and regulation across many agencies. Our transition fixed effects absorb level differences but cannot isolate the SEC enforcement channel from these correlated shocks. Third, our market outcomes are measured at the aggregate level; firm-level effects concentrated among companies under active investigation, or in industries specifically targeted by the outgoing Chair, may be substantial even when sector-wide effects are null. Future work using firm-level CRSP returns for SEC enforcement targets, or the Stanford Securities Class Action Clearinghouse for private litigation substitution, would provide more granular tests. Fourth, our daily enforcement data covers primarily the March--July portion of each fiscal year due to SEC website pagination limitations, preventing us from directly observing the enforcement surge and vacuum in the months immediately surrounding transitions. We rely on authoritative aggregate statistics from Cornerstone Research for the fiscal-year-level analysis.

\subsubsection*{Policy Implications} The ``clearing the deck'' phenomenon raises questions about the institutional design of SEC enforcement. The concentration of enforcement authority in a single Chair creates a predictable vulnerability: every four to eight years, the agency's enforcement apparatus grinds to a halt. Our finding of null aggregate market effects, while suggestive of short-term market resilience, does not rule out significant costs from enforcement lapses that may be visible only in firm-level behavior, longer-run data, or the quality of financial reporting rather than asset prices. If firms learn to time misconduct around predictable enforcement cycles---committing fraud during vacuums and settling before enforcement resumes---the social costs could be substantial even without aggregate market disruption. Proposals for more insulated enforcement authority---such as term-limited Directors of Enforcement with staggered appointments---could mitigate transition disruptions without sacrificing political accountability.

\section{Conclusion}

This paper has documented a previously unmeasured institutional regularity: SEC Chair transitions create large, systematic enforcement vacuums that are amplified by cross-party political transitions. The 2025 Gensler-to-Atkins transition---with its 97\% monthly enforcement decline---is the most extreme case in our sample, but the pattern is consistent across all four transitions spanning 2012--2025. Aggregate capital markets show no detectable response within $\pm$60 trading days. This null result is consistent with market anticipation, private enforcement substitution, or stock-based deterrence, but cannot rule out firm-level effects among SEC enforcement targets or longer-run costs from reduced deterrence. Future work linking daily enforcement filing data to firm-level stock returns and private litigation outcomes would sharpen the causal channel and clarify whether enforcement vacuums are genuinely benign or merely invisible at the index level.

\label{apep_main_text_end}

\newpage

\bibliography{references}

\newpage
\appendix
\section{Standardized Effect Sizes}

\input{tables/tabF1_sde.tex}

\newpage

\section*{Tables}

\input{tables/tab1_enforcement_fy.tex}

\input{tables/tab2_transition_effects.tex}

\input{tables/tab3_market_response.tex}

\input{tables/tab4_robustness.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
