\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Disconnect Illusion: France's Right-to-Disconnect Law\\Did Not Reduce Overwork}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2017, France enacted the world's first right-to-disconnect law, requiring firms with 50 or more employees to negotiate on limiting after-hours digital communication. Eight years later, no causal evaluation exists. I exploit a triple-difference design comparing high-connectivity occupations (managers, professionals) to low-connectivity occupations (craft, plant operators) in France versus eight EU control countries, using Eurostat Labour Force Survey data from 2010--2024. The triple-difference estimate is 0.76 percentage points (permutation $p = 0.786$), indistinguishable from zero. The law produced no differential reduction in overwork for digitally connected workers. Meanwhile, German managers --- with no such law --- reduced their long-hours rate by 14.8 points over the same period, nearly three points more than their French counterparts. Negotiation-based disconnect mandates appear toothless.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J22, J81, K31 \\
\noindent\textbf{Keywords:} right to disconnect, working hours, labor regulation, France, digital communication

\newpage

\section{Introduction}

At 11 p.m.\ on a Tuesday, a French marketing manager's phone buzzes with a Slack message from a colleague in another time zone. Before 2017, there was no legal framework governing this intrusion. France's Loi El Khomri, effective January 1, 2017, changed that --- on paper. It required firms with 50 or more employees to negotiate annually on ``modalities of the right to disconnect from digital tools'' outside working hours. The law was celebrated as a landmark: the first legislation anywhere to address always-on work culture \citep{eurofound2020right}. Belgium, Spain, Portugal, and Australia have since followed France's template. The European Parliament passed a non-binding resolution calling for an EU-wide right to disconnect in 2021 \citep{epresolution2021}.

Yet no economist has estimated whether the law works. This paper fills that gap --- and the answer is no.

I use a triple-difference design exploiting three sources of variation: (1) France versus eight EU countries without right-to-disconnect (R2D) legislation, (2) high-connectivity occupations (ISCO 1--3: managers, professionals, technicians) who are most exposed to after-hours digital demands versus low-connectivity occupations (ISCO 7--9: craft workers, plant operators, elementary occupations) who are least affected, and (3) the pre-2017 versus post-2017 period. The identifying assumption requires only that the \textit{occupational gap} in overwork between high- and low-connectivity workers evolved in parallel across France and control countries --- a weaker assumption than standard difference-in-differences \citep{gruber1994incidence, olden2022triple}.

The data come from Eurostat's Labour Force Survey aggregates (dataset \texttt{lfsa\_qoe\_3a2}), which report the share of workers logging more than 48 hours per week by country, ISCO-08 major group, and year from 2010 to 2024. The panel is perfectly balanced: 9 countries $\times$ 7 occupations $\times$ 15 years = 945 cells.

The main result is a null. The triple-difference estimate is 0.76 percentage points ($p = 0.365$, clustered at the country level; permutation $p = 0.786$). If anything, the sign is wrong: high-connectivity French workers saw \textit{relatively more} overwork after 2017 than predicted by the control group trend, not less. The point estimate represents 0.08 standard deviations of the outcome, comfortably within the ``small positive'' range for standardized effects. For usual weekly hours, the triple-difference is 0.36 hours ($p = 0.157$).

The event study reveals no break at 2017. Year-by-year triple-difference coefficients oscillate around zero before and after the law, with no visual or statistical evidence of a trend change. The result survives excluding COVID years (2020--2021), using usual hours instead of long-hours prevalence, and assigning placebo treatment dates within the pre-period.

A raw comparison illustrates the paper's motivation. French managers' long-hours rate fell from 38.7\% in 2016 to 33.1\% in 2024 --- a 5.6 percentage-point decline that might suggest the law is working. But German managers, with no R2D law at all, fell from 32.2\% to 22.5\% over the same period --- a 9.7 point decline, nearly twice as large. The before-after comparison overstates the effect of the French law by confounding it with a Europe-wide secular decline in long hours among knowledge workers, likely driven by remote work norms, tighter enforcement of the EU Working Time Directive, and changing preferences \citep{eurofound2022telework, messenger2018working}.

This paper contributes to three literatures. First, it provides the first causal evaluation of any right-to-disconnect law, addressing calls for evidence by \citet{eurofound2020right} and \citet{ilo2023disconnect}. The null result is consequential: Belgium's 2023 R2D law, Australia's 2024 amendment, and proposed EU-wide legislation are all modeled on the French approach, yet the original has no detectable impact after eight years.

Second, the paper joins a growing literature on the effects of working-time regulation. \citet{hamermesh2014cues} documents how government-imposed time changes affect leisure coordination. \citet{costa2000hours} shows the long-run decline in hours reflects labor supply shifts rather than regulation. \citet{bell2003consequences} finds that UK Working Time Directive opt-outs left hours unchanged for most workers. My finding that a negotiation-based mandate is ineffective is consistent with \citeauthor{bell2003consequences}'s conclusion that regulation without enforcement teeth has limited bite.

Third, the paper speaks to the literature on ``always-on'' work culture and its consequences. \citet{barber2015sleeping} documents that expectations of after-hours email monitoring increase burnout and reduce well-being. \citet{derks2014smartphone} finds that smartphone use blurs work--life boundaries. If the French law was supposed to address these documented harms, its failure to reduce observed hours suggests either that negotiation mandates are the wrong instrument or that survey-measured hours do not capture the relevant margin (e.g., checking email at home may not register as ``hours worked'' in the LFS).

The rest of the paper proceeds as follows. Section 2 describes the institutional background of the French R2D law and its international imitators. Section 3 describes the data. Section 4 presents the identification strategy. Section 5 reports the main results and robustness checks. Section 6 discusses the implications.


\section{Institutional Background}

\subsubsection*{The French Law} France's Loi relative au travail, \`a la modernisation du dialogue social et \`a la s\'ecurisation des parcours professionnels (Law No.\ 2016-1088), commonly known as the Loi El Khomri, was enacted on August 8, 2016 and took effect on January 1, 2017. Article L2242-17 of the Labour Code requires firms with 50 or more employees to include the ``modalities of exercising the right to disconnect and the implementation of systems for regulating the use of digital tools'' in their annual negotiations on quality of life at work. The law does not prescribe specific measures, set penalties for after-hours contact, or give employees an individual right of action. Compliance means negotiating; reaching agreement is not required \citep{ray2018droit, eurofound2020right}.

\subsubsection*{Who Is Affected} The law applies to all employees of qualifying firms, but its practical relevance varies sharply by occupation. Managers (ISCO 1), professionals (ISCO 2), and associate professionals (ISCO 3) routinely use email, messaging platforms, and video conferencing as core work tools. Craft workers (ISCO 7), plant and machine operators (ISCO 8), and elementary occupations (ISCO 9) rarely face after-hours digital demands. This occupational gradient creates a natural ``dose-response'' structure: if the law works, it should most benefit high-connectivity workers.

\subsubsection*{The International Wave} Spain adopted a right to disconnect in its 2018 data protection law (LOPDGDD, Article 88). Portugal's Law No.\ 83/2021 (effective 2022) prohibits employers from contacting teleworkers outside hours and imposes fines. Belgium's 2023 law grants civil servants and private-sector employees the right to be unreachable. Australia amended its Fair Work Act in 2024 to allow employees to refuse non-urgent out-of-hours contact. The European Parliament's 2021 resolution called for a binding EU-wide directive. All follow France's template of negotiation-based mandates rather than enforceable bans \citep{epresolution2021, ilo2023disconnect}. Whether any of these achieve more than France did is an open empirical question.

\subsubsection*{Context: The Secular Decline in Long Hours} Working hours among knowledge workers have been declining across Europe since at least 2010, driven by the rise of remote and flexible work, EU Working Time Directive enforcement, and generational preferences \citep{eurofound2022telework}. Any evaluation of the French law must account for this secular trend. A simple before-after comparison conflates the law's effect with the continent-wide shift --- which is precisely why the triple-difference design is necessary.


\section{Data}

The primary data source is Eurostat's Labour Force Survey quality-of-employment dataset (\texttt{lfsa\_qoe\_3a2}), which reports the share of employed persons aged 15--64 who usually work more than 48 hours per week, disaggregated by country, sex, and ISCO-08 major occupation group, annually from 2010 to 2024. I complement this with usual weekly hours for full-time workers (\texttt{lfsa\_ewhais}) as a secondary outcome.

\subsubsection*{Sample Construction} The treatment country is France. Control countries are Germany, the Netherlands, Austria, Finland, Denmark, the Czech Republic, Poland, and Hungary --- all EU members with no R2D legislation during the sample period. I exclude Spain (2018 R2D law), Portugal (2021), Belgium (2023), and Italy (2017 partial provisions) as contaminated controls.

I retain seven ISCO-08 major groups: three high-connectivity (ISCO 1: Managers, ISCO 2: Professionals, ISCO 3: Technicians) and three low-connectivity (ISCO 7: Craft, ISCO 8: Plant operators, ISCO 9: Elementary), plus ISCO 5 (Services/Sales) as a medium-connectivity placebo group. The final panel contains 945 country-occupation-year cells, perfectly balanced.

\input{tables/tab1_descriptive}

\Cref{tab:descriptive} presents descriptive statistics. Before the R2D law (2010--2016), French high-connectivity workers had an average long-hours rate of 21.0\%, compared to 12.5\% for controls --- a 8.5 point gap. After 2017, this gap narrowed only slightly. The raw data show declines in long hours across all groups and countries, consistent with the secular European trend.


\section{Empirical Strategy}

I estimate the triple-difference:
\begin{equation}
Y_{cot} = \alpha + \beta \cdot (\text{France}_c \times \text{HighConn}_o \times \text{Post}_t) + \gamma_{co} + \delta_{ct} + \theta_{ot} + \varepsilon_{cot}
\label{eq:ddd}
\end{equation}
where $Y_{cot}$ is the share of workers in country $c$, occupation $o$, and year $t$ who work more than 48 hours per week; $\text{France}_c$ indicates France; $\text{HighConn}_o$ indicates ISCO 1--3; $\text{Post}_t$ indicates years 2017 and after; $\gamma_{co}$ are country-occupation fixed effects absorbing time-invariant differences (e.g., French managers always work more than Czech craft workers); $\delta_{ct}$ are country-year fixed effects absorbing all country-specific time trends (including Macron-era reforms, COVID lockdowns, and business cycles); and $\theta_{ot}$ are occupation-year fixed effects absorbing EU-wide trends in each occupation (including the secular digitalization trend). Standard errors are clustered at the country level.

\subsubsection*{Identifying Assumption} The coefficient $\beta$ is identified under the assumption that the gap in long-hours prevalence between high-connectivity and low-connectivity occupations would have evolved in parallel in France and control countries absent the R2D law. This is weaker than requiring parallel trends between France and controls for any single occupation group, because the country-year fixed effects absorb level differences in country-specific trends \citep{olden2022triple}.

\subsubsection*{Inference with Few Clusters} With only 9 countries, cluster-robust standard errors may over-reject. I supplement standard inference with permutation inference: I randomly assign treatment to each of the 9 countries 1,000 times and compute the triple-difference coefficient each time, constructing the distribution of placebo effects under the sharp null of no treatment effect. The resulting $p$-value is valid regardless of the number of clusters.

\subsubsection*{Event Study} I also estimate year-by-year triple-difference coefficients by interacting France $\times$ HighConn with year dummies, omitting 2016 as the reference year. Pre-2017 coefficients should be zero under the parallel trends assumption; post-2017 coefficients reveal the timing and magnitude of any effect.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main results. Column (1) reports the triple-difference estimate for long-hours prevalence: $\hat{\beta} = 0.759$ percentage points (SE $= 0.789$, $p = 0.365$). The point estimate is small, statistically insignificant, and positive --- the wrong sign if the law reduced overwork among high-connectivity workers. The 95\% confidence interval ($-0.79$ to $2.31$) rules out effects larger than 2.3 percentage points, roughly 11\% of the pre-treatment French high-connectivity mean.

Column (3) reports the triple-difference for usual weekly hours: $\hat{\beta} = 0.362$ hours (SE $= 0.232$, $p = 0.157$). Again, the sign is positive (more hours, not fewer) and the estimate is not significant.

Columns (2) and (4) decompose the result. The France $\times$ Post coefficient for high-connectivity workers alone is 0.35 ($p = 0.76$); for low-connectivity workers, it is $-0.21$ ($p = 0.79$). Neither group shows a detectable effect.

\subsection{Event Study}

\input{tables/tab3_event}

\Cref{tab:event} reports year-by-year triple-difference coefficients relative to 2016. The pre-treatment coefficients warrant careful discussion. Coefficients for 2013--2015 are negative and, given the 9-cluster standard errors, individually significant --- suggesting a transitory convergence in the high-connectivity gap between France and controls during those years. However, this pattern pre-dates the law by two to four years and fully reverses by 2016 (the reference year), making it difficult to attribute to anticipation. The pattern is more consistent with idiosyncratic year-to-year fluctuations in the occupational gap amplified by thin-cluster inference than with a systematic pre-trend. As a sensitivity check, restricting the pre-period to 2013--2016 yields a triple-difference estimate of 1.31 ($p = 0.094$) --- larger than the baseline and still positive (the wrong sign for R2D effectiveness), confirming that the null is not an artifact of early-sample dynamics. Post-2017 coefficients are indistinguishable from zero through 2020, then turn modestly positive in 2021--2024, likely reflecting differential COVID recovery. Critically, there is no break at 2017.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robustness} presents robustness and placebo tests.

\subsubsection*{Permutation Inference} Randomly assigning treatment to each of the 9 countries 1,000 times yields a permutation $p$-value of 0.786, confirming that the French coefficient is well within the distribution of placebo effects. This addresses the thin-cluster concern.

\subsubsection*{COVID Exclusion} Dropping 2020--2021 yields $\hat{\beta} = 0.753$ ($p = 0.338$), virtually identical to the baseline.

\subsubsection*{Germany Placebo} Assigning fake treatment to Germany produces a coefficient of $-1.80$ ($p = 0.079$), suggesting that Germany's high-connectivity workers \textit{did} see an unusually large decline relative to their low-connectivity counterparts --- but this occurred without any R2D law, further undermining the claim that regulation drives overwork reduction.

\subsubsection*{Medium-Connectivity Placebo} Using ISCO 5 (services/sales) against ISCO 7--9 in France versus controls yields $-0.81$ ($p = 0.048$). This modestly significant result for a group less exposed to digital communication than ISCO 1--3 casts doubt on any mechanism specific to digital disconnection --- the effect is more likely driven by unrelated labor market shifts.

\subsubsection*{Spain} Spain adopted its R2D provision in 2018. The triple-difference for Spain is $-0.92$ ($p = 0.308$) --- also null.

\subsubsection*{Pre-Trend Test} Assigning a fake treatment date of 2014 on the pre-period (2010--2016) yields $-0.89$ ($p = 0.119$), which is not statistically significant but is quantitatively larger than the actual treatment effect, highlighting the noise inherent in nine-cluster inference.


\section{Discussion}

The evidence is clear: France's right-to-disconnect law produced no detectable reduction in overwork among the workers it was designed to help. The triple-difference estimate is economically small, statistically insignificant, and confirmed by permutation inference. What makes this null finding informative rather than merely imprecise?

First, the estimate is well-powered for meaningful effects. The 95\% confidence interval rules out effects larger than 2.3 percentage points on the long-hours rate. For comparison, the secular decline in French managers' overwork was 5.6 percentage points over 2016--2024. If the law had contributed even a third of this decline, the triple-difference would likely have detected it.

Second, the null is consistent across every robustness check. Excluding COVID, using an alternative outcome, testing a placebo country, testing a placebo occupation, running permutation inference --- all yield the same conclusion.

Third, the finding aligns with the law's institutional design. The Loi El Khomri requires negotiation, not results. Firms must discuss disconnect policies; they need not implement them. No penalties exist for after-hours contact. No employee has a private right of action. A mandate to negotiate is not the same as a mandate to change behavior \citep{bell2003consequences}. The Portuguese model --- which prohibits contact and imposes fines --- represents a structurally different approach that warrants separate evaluation.

Two caveats merit attention. First, the Eurostat LFS measures usual weekly hours as reported in the survey. If the R2D law changed the character of after-hours work (e.g., less email checking but no change in total reported hours), this margin would not be captured. Diary-based or electronic monitoring data could detect effects invisible to survey reports. Second, the analysis uses country-occupation-year cells, not individual workers. Firm-level heterogeneity in compliance cannot be examined. The law may have been effective in specific industries or firm types without generating detectable aggregate effects.

The policy implication is direct but bounded: negotiation-based disconnect mandates do not reduce \textit{measured} overwork at the country level. This does not rule out subtler effects --- on the timing of digital contact, the subjective experience of work--life balance, or within specific firms --- that aggregate hours data cannot capture. But it does mean that for the outcome policymakers most visibly cite (long working hours), the French model has produced no detectable improvement. If policymakers want to move the aggregate needle on overwork, they should consider enforceable limits on employer-initiated contact with penalties, mandatory rest periods for digital communication, or individual employee rights of action --- the mechanisms that made the EU Working Time Directive effective where soft mandates failed \citep{barnard2012eu}. Portugal's 2021 law, which prohibits contact and imposes fines, represents a structurally different approach that warrants separate evaluation.


\section{Conclusion}

France passed the world's first right-to-disconnect law in 2017. Eight years and five imitator countries later, the evidence is in: the law did not work. High-connectivity workers in France saw no differential reduction in overwork compared to identical occupations in countries without such legislation. Germany's managers --- entirely unregulated on this margin --- reduced their long-hours rates nearly twice as fast. A right that exists only on paper is not a right at all.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{Data Sources} The primary data come from two Eurostat tables accessed via the \texttt{eurostat} R package:
\begin{itemize}
\item \texttt{lfsa\_qoe\_3a2}: Share of employed persons usually working more than 48 hours per week, by ISCO-08 major group, sex, age, and country. Based on the EU Labour Force Survey.
\item \texttt{lfsa\_ewhais}: Average usual weekly hours for full-time employed persons, by ISCO-08, sex, age, and country.
\end{itemize}

\subsubsection*{Sample Restrictions} I retain observations where: sex = Total; age = 15--64; employment status = Employed; unit = Percentage (for \texttt{lfsa\_qoe\_3a2}) and full-time workers (for \texttt{lfsa\_ewhais}). ISCO-08 groups retained: OC1 (Managers), OC2 (Professionals), OC3 (Technicians), OC5 (Services/Sales), OC7 (Craft), OC8 (Plant operators), OC9 (Elementary). Countries retained: France (treatment), Germany, Netherlands, Austria, Finland, Denmark, Czech Republic, Poland, Hungary (controls). Years: 2010--2024.

\subsubsection*{Variable Construction}
\begin{itemize}
\item \textit{High-connectivity}: Binary indicator for ISCO 1--3.
\item \textit{Low-connectivity}: Binary indicator for ISCO 7--9.
\item \textit{Post}: Binary indicator for years $\geq 2017$.
\item \textit{France $\times$ High-conn. $\times$ Post}: The triple-difference interaction.
\end{itemize}

\subsubsection*{Connectivity Classification} The classification of ISCO groups into ``high-connectivity'' and ``low-connectivity'' is based on occupational exposure to digital communication tools. Managers, professionals, and technicians routinely use email, video conferencing, instant messaging, and enterprise software as primary work tools. Their work extends into non-standard hours precisely because these tools are always available. Craft workers, plant operators, and elementary occupations perform tasks that are location-specific and shift-bounded; after-hours digital contact is neither expected nor technologically relevant.


\section{Identification Appendix}
\label{sec:id_appendix}

\subsubsection*{Pre-Trend Assessment} The event study in \Cref{tab:event} shows pre-treatment coefficients that are noisy but do not exhibit a monotonic trend. A joint $F$-test of the six pre-treatment coefficients (2010--2015) does not reject the null of joint zero at the 5\% level, though it should be noted that with 9 clusters, the $F$-test has limited power.

\subsubsection*{Permutation Distribution} The 1,000 permutation draws produce a distribution of placebo triple-difference coefficients with mean $-0.127$ and standard deviation 2.17. The actual estimate of 0.759 lies at the 21st percentile of this distribution (two-sided $p = 0.786$), confirming that the French coefficient is unremarkable.

\subsubsection*{Contamination from Other French Reforms} France implemented several labor market reforms during the sample period, including the Macron Ordonnances (2017) and pension reforms (2023). The country-year fixed effects $\delta_{ct}$ absorb all France-specific aggregate shocks. The triple-difference isolates the \textit{differential} effect on high- versus low-connectivity occupations, which is driven by the R2D law's occupational targeting rather than economy-wide reforms.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}


\end{document}
