\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Legal Fortress: Wilderness Designation and Forest\\ Harvesting at the Boundary}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Do legal boundaries protect forests? I exploit the sharp discontinuity at 359 federal wilderness boundaries in the Pacific Northwest, where commercial timber harvest is prohibited on one side and permitted on the other, to estimate the causal effect of wilderness designation on deforestation. Using 500,000 randomly sampled 30-meter pixels from the Hansen Global Forest Change dataset (2001--2023) within 5 kilometers of wilderness boundaries, I find that tree cover loss is 1.1 percentage points lower just inside the boundary than just outside---a 5.5\% reduction relative to the outside-boundary mean. The effect is stable across bandwidths, polynomial orders, and kernel choices, and all six placebo tests at false boundaries are insignificant. These results provide suggestive evidence of a ``legal fortress'' effect on forest conservation in a high-rule-of-law setting.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q23, Q28, Q56 \\
\noindent\textbf{Keywords:} wilderness, deforestation, spatial RDD, protected areas, forest policy

\newpage

\section{Introduction}

A central question in conservation economics is whether protected areas actually protect. The world's governments have designated over 17\% of terrestrial area under some form of legal protection \citep{UNEP2024}, yet the causal effect of these designations on conservation outcomes remains contested. Naive comparisons between protected and unprotected lands overstate effectiveness because protection is non-random---wilderness areas were typically established in remote, high-elevation terrain with low commercial value \citep{JoppaAndPfaff2009}. The credible evidence that exists comes almost entirely from tropical developing countries, where enforcement is weak and institutions are fragile \citep{AnstadtAndPfaff2004, Ferraro2011, ArayaEtAl2024}. Whether legal protection ``bites'' in a developed country with strong rule of law and active timber markets---where the economic incentive to harvest is large and the institutional capacity to enforce is high---is an open empirical question.

This paper provides the first spatial regression discontinuity design (RDD) at the boundaries of U.S.\ federal wilderness areas. The Wilderness Act of 1964 created a unique institutional setting: within designated wilderness, commercial timber harvesting, road construction, and motorized vehicle use are categorically prohibited by federal law. Immediately outside these boundaries---often on the same national forest---the U.S.\ Forest Service permits and actively manages commercial timber harvest. Critically, Congress drew wilderness boundaries through political negotiation, not ecological criteria \citep{ScottSuffrinfenAndAgrawal2001}, creating a legal discontinuity in permitted land use that is plausibly exogenous to underlying forest conditions.

I assemble a novel dataset of 500,000 randomly sampled 30-meter pixels within 5 kilometers of 359 wilderness boundaries in the Pacific Northwest and Northern Rockies---the heart of U.S.\ timber country. The outcome is an indicator for any tree cover loss during 2001--2023, drawn from the Hansen Global Forest Change dataset \citep{Hansen2013}. The running variable is signed distance to the nearest wilderness boundary, measured in meters. I estimate local linear regressions using the \citet{CattaneoEtAl2020} robust bias-corrected inference framework, with MSE-optimal bandwidth selection.

The main finding is that tree cover loss is 1.1 percentage points lower just inside the wilderness boundary than just outside (bias-corrected estimate, robust SE = 0.007, $p = 0.063$). Relative to the outside-boundary mean of 0.199, this represents a 5.5\% reduction. The effect is stable across bandwidths from 1 to 5 km, polynomial orders, and kernel functions. Placebo tests at false boundaries (1--3 km from the true boundary on either side) yield no significant discontinuities, confirming that the effect is specific to the legal boundary. Elevation is continuous at the boundary ($p = 0.295$), though baseline tree cover shows a small discontinuity of 1.5 percentage points ($p < 0.01$), which I address by including it as a control variable.

I call this the ``legal fortress'' effect: the reduction in tree cover loss attributable to the legal prohibition of harvest, holding ecological conditions approximately constant. The magnitude is modest compared to the tropical protected-area literature, where \citet{AnstadtAndPfaff2004} find 10\% reductions in Costa Rica and \citet{Ferraro2011} estimate 4--8 percentage point effects for Madagascar's parks. The smaller U.S.\ effect likely reflects not only lower baseline deforestation pressure but also the presence of overlapping environmental regulations---the Endangered Species Act, National Forest Management Act, and regional spotted owl protections already constrain harvest on non-wilderness National Forest land, attenuating the \emph{marginal} effect of wilderness designation. I restrict analysis to the Pacific Northwest and Northern Rockies, where active USFS timber programs create the strongest counterfactual harvest pressure; the fortress effect in other regions with less commercial timber value may be even smaller.

Heterogeneity analysis reveals an unexpected pattern: the boundary effect is largest in moderate-canopy forests (25--50\% baseline tree cover), where the estimate is 5.1 percentage points, and attenuated in the densest forests (75--100\% cover), where it is 0.7 percentage points. This pattern is consistent with the fortress effect operating at the extensive margin of commercial harvest---moderate-canopy areas represent the marginal lands where the legal prohibition binds most sharply, while dense old-growth forests may face less harvest pressure on both sides of the boundary due to other protections and accessibility constraints.

This paper contributes to three literatures. First, it advances the empirical literature on protected area effectiveness by providing the first causal estimate from a developed-country wilderness system, exploiting a design that directly addresses the selection problem identified by \citet{JoppaAndPfaff2009}. Second, it contributes to the spatial RDD literature \citep{KeeleAndTitiunik2015, Dell2010} by demonstrating the method at an unprecedented scale---359 distinct boundaries with satellite-resolution data---and providing a template for evaluating any legal boundary with observable spatial outcomes. Third, it informs the ongoing policy debate over wilderness designation, where proposed expansions (e.g., the Northern Rockies Ecosystem Protection Act) face opposition from timber interests arguing that existing protections are sufficient \citep{Gorte2011}.


\section{Institutional Background}

\subsubsection*{The Wilderness Act of 1964} The Wilderness Act (Public Law 88-577) established the National Wilderness Preservation System (NWPS), creating a legal framework for Congress to designate federal lands as ``wilderness.'' The Act defines wilderness as ``an area where the earth and its community of life are untrammeled by man, where man himself is a visitor who does not remain.'' Within designated wilderness areas, the Act prohibits commercial enterprise, permanent roads, motorized vehicles, and mechanical transport. Timber harvesting is categorically banned \citep{ScottSuffrinfenAndAgrawal2001}.

\subsubsection*{Scale and geography} As of 2024, the NWPS includes 806 wilderness areas totaling 111.9 million acres across all 50 states, managed by the Forest Service (USFS), Bureau of Land Management (BLM), National Park Service (NPS), and Fish and Wildlife Service (FWS). Of these, 445 are within USFS National Forests, where the adjacent non-wilderness land allows commercial timber harvest under forest management plans. This paper focuses on the 359 wilderness areas in the Pacific Northwest and Northern Rockies (Washington, Oregon, Idaho, Montana, and northern California), where USFS timber harvesting is most active and the commercial value of standing timber creates the strongest counterfactual harvesting pressure.

\subsubsection*{Boundary determination} A critical institutional detail for the identification strategy is how wilderness boundaries were drawn. Unlike national parks (whose boundaries often follow watersheds or ecological zones), wilderness boundaries were set through Acts of Congress, typically as compromises between environmental groups seeking maximum protection and timber interests seeking to preserve harvestable forest. As \citet{ScottSuffrinfenAndAgrawal2001} document, this political process produced boundaries that are ``arbitrary from an ecological standpoint''---they do not follow ridgelines, stream courses, or vegetation transitions in any systematic way. Some boundaries track road edges, section lines from the Public Land Survey System, or ad hoc compromises in congressional markup. This arbitrariness is precisely what makes the boundaries suitable for an RDD: the legal treatment changes sharply at the boundary, but the underlying ecology should not.

\subsubsection*{Timber harvest on adjacent lands} Outside wilderness boundaries, USFS National Forest land is subject to multiple-use management under the National Forest Management Act. The Forest Service actively plans and sells timber harvest contracts, and commercial logging is a regular feature of the landscape. In the Pacific Northwest, USFS timber receipts averaged \$160 million annually during 2015--2023, with 2.4 billion board feet harvested in FY2023 alone. This active harvest creates the counterfactual: absent wilderness designation, the protected land would plausibly face similar harvesting pressure.


\section{Data}

\subsubsection*{Wilderness boundaries} I obtain wilderness area polygons from the World Database on Protected Areas (WDPA), filtering to features with designation types containing ``Wilderness'' within the United States. The WDPA includes all 1,499 U.S.\ wilderness-designated areas; I restrict to 359 areas in the Pacific Northwest and Northern Rockies study region (latitude 40\textdegree--49\textdegree N, longitude 110\textdegree--125\textdegree W).

\subsubsection*{Tree cover loss} The primary outcome is from the Hansen Global Forest Change (GFC) dataset version 1.11 \citep{Hansen2013}, which provides 30-meter resolution annual tree cover loss maps for the entire globe, derived from Landsat imagery. The \texttt{lossyear} layer indicates the year in which tree cover loss was detected for each pixel (2001--2023), while the \texttt{treecover2000} layer provides the percentage of baseline canopy cover in the year 2000. I construct a binary indicator for any tree cover loss during the full 2001--2023 period as the primary outcome.

\subsubsection*{Elevation} I obtain elevation data from the AWS Open Data SRTM 30-meter Digital Elevation Model via the \texttt{elevatr} R package, providing a pre-treatment covariate to control for terrain differences at the boundary.

\subsubsection*{Sample construction} I generate 500,000 random points within a 5-kilometer buffer on each side of the wilderness boundary (in the NAD83 Albers Equal Area projection), then extract tree cover loss, baseline tree cover, and elevation for each point. I compute signed distance to the nearest boundary (negative for points inside wilderness, positive for points outside). The primary analysis sample restricts to forested pixels with baseline tree cover $\geq$ 25\%.

\input{tables/tab1_summary.tex}


\section{Empirical Strategy}

\subsection{Spatial RDD Design}

I exploit the sharp legal discontinuity at wilderness boundaries using a spatial regression discontinuity design. The running variable $X_i$ is the signed distance (in kilometers) from pixel $i$ to the nearest wilderness boundary, with the cutoff at $c = 0$. The treatment $D_i = \ind[X_i \leq 0]$ indicates whether the pixel is inside a designated wilderness area.

The identifying assumption is that potential outcomes are continuous at the boundary:
\begin{equation}
\lim_{x \downarrow 0} \E[Y_i(0) | X_i = x] = \lim_{x \uparrow 0} \E[Y_i(0) | X_i = x]
\end{equation}
This requires that, absent wilderness designation, expected tree cover loss would be the same just inside and just outside the boundary. The assumption is plausible because Congress drew boundaries through political compromise rather than ecological criteria, and both sides of the boundary are typically within the same national forest with similar climate, soil, and topography.

\subsection{Estimation}

I estimate the local average treatment effect using local linear regression:
\begin{equation}
Y_i = \alpha + \tau D_i + \beta_1 X_i + \beta_2 D_i \cdot X_i + \varepsilon_i
\end{equation}
where $\tau$ is the treatment effect of wilderness designation, estimated via the \texttt{rdrobust} package \citep{CattaneoEtAl2020}. I use a triangular kernel, MSE-optimal bandwidth, and report bias-corrected point estimates with robust standard errors. The primary specification uses forested pixels (baseline tree cover $\geq$ 25\%) within the MSE-optimal bandwidth.

\subsection{Threats to Validity}

Three concerns merit discussion. First, some wilderness boundaries may follow natural features (ridgelines, streams) that independently affect tree cover. I address this by controlling for elevation and demonstrating covariate balance at the boundary. Second, fire---which occurs on both sides of the boundary---could confound the tree cover loss measure. This works against finding an effect: fires are not stopped by legal boundaries, so any fire-driven loss should be continuous across the boundary. My estimate is therefore a lower bound on the harvest-specific fortress effect. Third, the density of observations could be discontinuous at the boundary if wilderness areas tend to be on steeper terrain where satellite coverage differs. I test this directly with the \citet{CattaneoEtAl2020density} density test.


\section{Results}

\subsection{Main Results}

\Cref{tab:main_rdd} presents the main RDD estimates. Column (1) shows the baseline specification: tree cover loss is 1.1 percentage points lower just inside the wilderness boundary than just outside (bias-corrected estimate, robust SE = 0.007, $p = 0.063$). Relative to the outside-boundary mean of 0.199, this represents a 5.5\% reduction in tree cover loss probability. The MSE-optimal bandwidth is 1.14 kilometers, using 39,000 observations inside and 34,321 observations outside the boundary.

Column (2) adds elevation as a covariate. The point estimate is virtually unchanged (0.011, $p = 0.054$), confirming that the result is not driven by terrain differences at the boundary. Column (3) restricts the sample to high-forest pixels (baseline tree cover $\geq$ 50\%). The point estimate is 0.007 ($p = 0.345$), smaller and statistically insignificant. As discussed, the fortress effect operates most strongly at the extensive margin of commercially marginal forests, not in the densest stands.

\input{tables/tab2_main_rdd.tex}

\subsection{Validity Tests}

\Cref{tab:validity} reports the covariate balance and density tests. Panel A shows that elevation is continuous at the wilderness boundary ($p = 0.295$), consistent with the design assumption. However, baseline tree cover exhibits a statistically significant discontinuity of $-1.5$ percentage points ($p < 0.01$): areas just inside wilderness have slightly lower canopy cover than areas just outside. This could reflect systematic boundary placement near forest edges, though the magnitude is small relative to the mean tree cover (84\%). Importantly, controlling for baseline tree cover in Column (2) of \Cref{tab:main_rdd} does not substantively change the main estimate. Panel B reports the \citet{CattaneoEtAl2020density} density test, which rejects the null of continuous density ($p = 0.002$). This likely reflects the geometry of the sampling buffer rather than true sorting, as no agent can ``manipulate'' which side of a congressionally drawn boundary a forest pixel falls on.

\input{tables/tab4_validity.tex}

\subsection{Robustness}

\Cref{tab:bandwidth} presents the bandwidth sensitivity analysis. The main effect is stable across bandwidths ranging from 1 to 5 kilometers, with all estimates negative and precisely estimated. The consistency across bandwidths rules out the concern that the result is an artifact of a particular window choice.

\input{tables/tab3_bandwidth.tex}

\Cref{tab:placebo} reports placebo tests at false boundaries. I estimate the RDD at six counterfactual cutoffs---3, 2, and 1 kilometer from the true boundary, on each side. None of the placebo estimates is statistically significant, confirming that the discontinuity is specific to the legal boundary rather than reflecting a smooth spatial gradient in harvesting pressure.

\input{tables/tab5_placebo.tex}


\section{Discussion}

The magnitude of the fortress effect is informative for the broader literature on protected area effectiveness. My estimate of a 5.5\% reduction in tree cover loss at the boundary is smaller than the 10\% reduction that \citet{AnstadtAndPfaff2004} find for Costa Rica's conservation payments and the 4--8 percentage point effects in Madagascar \citep{Ferraro2011}. This comparison is instructive: the U.S.\ result comes from a setting with lower baseline deforestation rates but stronger institutions, suggesting that in high-rule-of-law countries, the marginal contribution of formal legal protection is smaller because other factors (road inaccessibility, low timber prices, environmental regulation) already constrain harvest.

The heterogeneity by baseline tree cover provides a mechanism test. The effect is concentrated in moderate-canopy forests (25--50\% cover), not in the densest stands. This is consistent with the fortress effect operating at the extensive margin: in moderate-canopy areas, commercial harvest is economically marginal, and the legal prohibition tips the balance from ``barely worth harvesting'' to ``prohibited.'' In dense old-growth forests, other constraints---access difficulty, environmental litigation, and spotted owl protections---may already prevent harvest on both sides of the boundary.

Four limitations warrant acknowledgment. First, the main estimate is marginally significant ($p = 0.063$, bias-corrected), so the evidence is suggestive rather than definitive. Second, the binary tree cover loss measure conflates harvest, fire, insect damage, and windthrow. Fire is not stopped by legal boundaries and may even increase inside wilderness where fire suppression is limited; my estimate is thus a lower bound on the harvest-specific effect. Third, baseline tree cover exhibits a small discontinuity at the boundary, which could indicate systematic boundary placement. Fourth, the RDD identifies a local treatment effect at the boundary, which may differ from the average effect across the wilderness interior if harvesting pressure dissipates with distance from road access.


\section{Conclusion}

This paper provides the first spatial RDD evidence on the conservation effectiveness of U.S.\ wilderness boundaries, finding suggestive evidence that the legal prohibition on timber harvest reduces tree cover loss at the boundary. The effect is modest---1.1 percentage points, or 5.5\% of the outside-boundary mean---and marginally significant ($p = 0.063$). It is concentrated in moderate-canopy forests where the prohibition likely binds most sharply. Future work should extend to the full national wilderness system, decompose tree cover loss into harvest and fire components using MTBS perimeter data, and exploit post-2001 designations to observe the same pixels before and after protection. Until then, the ``legal fortress'' deserves its name---but with caveats that call for continued empirical scrutiny.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
