\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Float placement
\usepackage{float}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Denominator Shuffle: How MSA Boundary Redefinitions Reveal CRA's Causal Effect on Mortgage Lending}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When the Office of Management and Budget redefines Metropolitan Statistical Area boundaries, the area median income used to classify census tracts as CRA-eligible changes mechanically---reclassifying tracts despite no change in their own income. Using HMDA microdata from 12 US states (2018--2024), I compare 205 reclassified tracts to 20,541 stable neighbors in a difference-in-differences design. Lending volume and approval rates are unchanged. However, tracts gaining CRA eligibility see significantly higher rate spreads---0.13 percentage points for gained-LMI tracts (p = 0.009)---consistent with banks serving marginal borrowers. Pre-trends are flat (F = 0.26, p = 0.93). CRA reclassification reshuffles loan pricing, not volume.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G21, G28, R21, H81 \\
\noindent\textbf{Keywords:} Community Reinvestment Act, mortgage lending, MSA boundaries, credit allocation, LMI tracts

\newpage

\section{Introduction}

Nearly half a century after its passage, the Community Reinvestment Act remains one of the most debated pillars of American financial regulation. Banks must demonstrate that they serve the credit needs of their assessment areas, including Low-and-Moderate Income (LMI) neighborhoods---yet whether this mandate actually changes where credit flows, or merely relabels lending that would have happened anyway, has proven remarkably difficult to establish. The fundamental challenge is selection: tracts classified as LMI differ systematically from non-LMI tracts in ways that confound any simple comparison of lending volumes.

This paper introduces a new identification strategy that isolates CRA's causal effect by exploiting what I call the \textit{denominator shuffle}. The Office of Management and Budget (OMB) periodically redefines Metropolitan Statistical Area (MSA) boundaries through its Bulletin process. When a tract moves from one MSA to another, or when the MSA it belongs to changes composition, the area median family income---the denominator used to classify tracts as LMI---shifts mechanically. A tract whose own income has not changed at all can cross the 80\% threshold that separates CRA-eligible from CRA-ineligible status, gaining or losing the regulatory incentive overnight. OMB Bulletin 23-01, effective January 1, 2024, triggered the latest such reshuffling.

The design exploits a sharp institutional rule: CRA eligibility depends on whether a tract's median family income falls below 80\% of the MSA/MD area median. When OMB redraws MSA boundaries, some tracts mechanically cross this threshold---not because anything changed in the neighborhood, but because the reference area changed. This is akin to a student's class rank shifting when classmates transfer in or out. The reclassification is orthogonal to local economic conditions, providing quasi-random variation in CRA incentives.

Using Home Mortgage Disclosure Act (HMDA) loan-level microdata from the Consumer Financial Protection Bureau, aggregated to the census-tract-by-year level across 12 US states for 2018--2024, I construct a balanced panel and estimate difference-in-differences models comparing reclassified tracts to their stable neighbors within the same MSA. The 2024 MSA redefinition provides sharp, single-date treatment timing with six years of pre-treatment data for parallel trend verification.

The main finding is a null that teaches: CRA reclassification does not meaningfully change lending volume or approval rates. The DiD coefficient on log originations is 0.001 (SE = 0.057), precisely estimated around zero. Approval rates are similarly unaffected ($\beta = -0.001$, SE = 0.005). If CRA were the binding constraint on neighborhood credit access that both proponents and critics assume, the denominator shuffle should produce detectable volume effects---it does not.

What does change is loan pricing. Tracts gaining CRA eligibility experience a 0.13 percentage point increase in rate spreads (p = 0.009), while tracts losing eligibility show no pricing change. This asymmetry suggests that banks respond to newly gained CRA eligibility by extending credit to marginally riskier borrowers---borrowers who receive loans at higher rates than the tract's pre-reclassification average. The volume of lending is unchanged because these marginal borrowers replace, rather than supplement, existing lending activity.

Pre-trends are flat across all outcomes (joint F-test p = 0.93 for originations), and a placebo test comparing tracts far from the 80\% threshold shows no effect. A donut-hole specification excluding tracts within 2 percentage points of the cutoff confirms the results.

This paper contributes to a long literature on CRA's effects. \citet{avery2003community} documented the institutional framework and early evidence. \citet{bhutta2011community} found modest CRA effects on mortgage lending using assessment-area comparisons. \citet{agarwal2012did} examined whether CRA induced risky lending during the subprime crisis, finding increased originations in CRA-eligible tracts around exam dates but no evidence of higher default rates. \citet{ding2020effects} exploited the 2013 MSA redefinition for small business lending---the closest antecedent to this paper---and found significant CRA effects on SBA-backed loans. \citet{gabriel2020community} and \citet{lee2012banks} assessed CRA's cumulative impact on neighborhood investment. My contribution is to bring the MSA-redefinition design to mortgage lending, using the richer post-2018 HMDA data with rate spread and borrower characteristics, and to exploit the more recent 2024 boundary change \citep{omb2023}. The finding that CRA reclassification affects pricing but not volume is new.

More broadly, this paper speaks to the debate over whether place-based credit incentives can effectively steer bank lending \citep{barr2005credit}. The denominator shuffle provides an unusually clean test because it holds constant both the tract's economic fundamentals and the bank's overall lending capacity---only the regulatory classification changes. The finding that banks respond to newly gained eligibility by shifting lending composition---accepting marginal borrowers at higher rates rather than expanding total credit---reframes the CRA debate. CRA does not appear to ration or expand credit at the neighborhood level; instead, it reshapes who receives credit and at what price. This has immediate policy relevance as federal regulators implement the 2023 CRA modernization rule \citep{harvey2024cra}.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of CRA and the MSA redefinition process. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses implications and concludes.

\section{Institutional Background}

\subsubsection*{The Community Reinvestment Act} Congress enacted the CRA in 1977 to combat redlining---the practice of denying credit to neighborhoods based on racial composition rather than creditworthiness. The law requires federal bank regulators (OCC, FDIC, Federal Reserve) to evaluate whether depository institutions meet the credit needs of their entire communities, including LMI areas. CRA performance ratings (Outstanding, Satisfactory, Needs to Improve, Substantial Noncompliance) affect banks' ability to pursue mergers, acquisitions, and branch openings \citep{barr2005credit}.

\subsubsection*{LMI Classification} A census tract is classified as LMI if its median family income is below 80\% of the area median income (AMI) of the MSA or Metropolitan Division (MD) in which it is located. The Federal Financial Institutions Examination Council (FFIEC) publishes annual census files mapping every tract to its MSA/MD and reporting the tract-to-area income ratio. This ratio---tract-to-MSA income percentage in HMDA parlance---is the key variable: tracts at or below 80 are LMI; those above are not.

\subsubsection*{MSA Boundary Redefinitions} OMB periodically updates MSA delineations based on Census Bureau data and commuting patterns. OMB Bulletin 23-01, issued in July 2023 and effective January 1, 2024, redefined boundaries for numerous MSAs nationwide. When a tract is reassigned to a different MSA, or when counties are added to or removed from its MSA, the area median income changes. A tract with median family income of \$40,000 that was in an MSA with AMI of \$55,000 (ratio = 72.7\%, LMI) might find itself in a redefined MSA with AMI of \$48,000 (ratio = 83.3\%, non-LMI)---or vice versa. The tract's own income is unchanged; only the denominator moved.

\subsubsection*{Why This Creates a Natural Experiment} The MSA redefinition process is driven by commuting-pattern data and population thresholds that are orthogonal to individual tract lending outcomes. The OMB does not consider CRA implications when drawing boundaries. Moreover, the redefinition affects all financial institutions simultaneously---no individual bank's behavior can influence the reclassification. The resulting variation is mechanical, predetermined, and publicly announced in advance, creating a clean setting for causal inference \citep{ding2020effects}.

\section{Data}

I use two data sources, both publicly available.

\subsubsection*{HMDA Loan-Level Data} The Home Mortgage Disclosure Act requires most mortgage lenders to report detailed information on each loan application, including the census tract, action taken (originated, denied, withdrawn), loan amount, borrower race and ethnicity, and rate spread. Since 2018, HMDA data includes expanded fields such as debt-to-income ratios, automated underwriting system recommendations, and property values \citep{Avery2005CRA}. I download loan-level data for 2018--2024 from the CFPB Data Browser API for 12 US states with substantial MSA coverage, then aggregate to the census-tract-by-year level. The key CRA-relevant field is tract-to-MSA income percentage, which directly measures the tract's income as a percentage of the MSA area median---the variable that determines LMI status.

\subsubsection*{Sample Construction} From the raw loan-level data, I construct a balanced panel of census tracts observed in all seven years (2018--2024). I restrict to tracts with at least 10 mortgage applications per year on average to reduce noise from thinly-traded tracts. The final balanced panel contains 20,746 tracts across 12 states observed in all seven years.

\subsubsection*{Treatment Identification} I identify reclassified tracts by comparing each tract's tract-to-MSA income percentage in 2023 (pre-redefinition) and 2024 (post-redefinition). Tracts whose LMI status changed---crossing the 80\% threshold in either direction---form the treatment group. Tracts in the same MSAs whose status was unchanged form the control group.

\input{tables/tab1_sumstats}

\Cref{tab:sumstats} presents pre-treatment summary statistics for control and reclassified tracts. The two groups are broadly comparable in lending volume and approval rates, though reclassified tracts are mechanically closer to the 80\% income threshold.

\section{Empirical Strategy}

\subsection{Difference-in-Differences}

The primary specification is:
\begin{equation}
Y_{it} = \alpha_i + \gamma_t + \beta \cdot (\text{Reclassified}_i \times \text{Post}_t) + \varepsilon_{it}
\label{eq:did}
\end{equation}
where $Y_{it}$ is the outcome (log originations, approval rate, minority share, log total amount, or rate spread) in tract $i$ in year $t$; $\alpha_i$ are tract fixed effects absorbing all time-invariant tract characteristics; $\gamma_t$ are year fixed effects absorbing aggregate trends; and $\text{Reclassified}_i \times \text{Post}_t$ is the treatment indicator, equal to one for reclassified tracts in 2024. Standard errors are clustered at the MSA/MD level to account for within-MSA correlation \citep{roth2023s}.

The identifying assumption is parallel trends: absent the MSA redefinition, reclassified and control tracts would have followed the same trajectory. I test this with an event-study specification:
\begin{equation}
Y_{it} = \alpha_i + \gamma_t + \sum_{k \neq -1} \delta_k \cdot (\text{Reclassified}_i \times \mathbf{1}[t - 2024 = k]) + \varepsilon_{it}
\label{eq:eventstudy}
\end{equation}
where $k = -1$ (year 2023) is the omitted reference period. The pre-treatment coefficients $\delta_{-6}, \ldots, \delta_{-2}$ should be close to zero and statistically insignificant if pre-trends are parallel.

I also separate the treatment into tracts that gained versus lost LMI status to test for asymmetric CRA effects---a key prediction of models where CRA operates through regulatory incentives rather than information channels.

\subsection{RDD at the 80\% Threshold}

As a complementary design, I estimate a sharp RDD using the 2024 cross-section:
\begin{equation}
Y_i = \alpha + \tau \cdot \mathbf{1}[\text{Income\%}_i \leq 80] + f(\text{Income\%}_i - 80) + \varepsilon_i
\end{equation}
where $f(\cdot)$ is a local linear polynomial and the bandwidth is chosen by the MSE-optimal procedure of \citet{calonico2014robust}. The RDD tests whether there is a discrete jump in lending at the CRA eligibility cutoff, conditional on the smooth relationship between tract income and lending.

\subsection{Threats to Validity}

\subsubsection*{Anticipation} If banks anticipated the 2024 MSA redefinitions (announced July 2023, effective January 2024), they might adjust lending before the formal reclassification. This would bias the DiD toward zero (lending adjusts in 2023, counted as pre-treatment) and would appear as a coefficient on $\delta_{-1}$ in the event study. I use 2023 as the reference period rather than omitting it.

\subsubsection*{Composition} The MSA redefinition might coincide with other changes (ACS updates, local economic shocks). Tract fixed effects absorb permanent differences, and year fixed effects absorb aggregate shocks. The key concern is tract-specific trends correlated with reclassification. The event study provides a direct test.

\subsubsection*{CRA Exam Timing} Banks are evaluated on CRA performance at irregular intervals (typically every 3--5 years). If exam timing correlates with MSA redefinitions, this could confound identification. However, the redefinition affects all banks in an MSA simultaneously, and exam-timing variation is at the bank level, not the tract level \citep{agarwal2012did}.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main_did}

\Cref{tab:main_did} reports the main DiD estimates from \Cref{eq:did}. Column 1 shows that reclassified tracts experienced essentially no change in log mortgage originations ($\beta = 0.001$, SE = 0.057). The point estimate is near zero and the standard error rules out effects larger than 11\% in either direction. Approval rates (column 2) and total loan amounts (column 4) are similarly unaffected. Minority lending share shows a small positive point estimate (0.004, p = 0.18), suggestive but not statistically significant. The exception is rate spread (column 5): reclassified tracts experience a 0.082 percentage point increase (p = 0.045), suggesting that while CRA reclassification does not change how much lending occurs, it does change lending terms.

\subsection{Asymmetric Effects}

\input{tables/tab3_asymmetric}

\Cref{tab:asymmetric} separates the treatment into tracts that gained versus lost LMI status. The volume null is symmetric: neither gained-LMI nor lost-LMI tracts show significant changes in originations. But the pricing effect is strikingly asymmetric. Tracts that gained CRA eligibility experience a 0.127 percentage point increase in rate spreads ($p = 0.009$), while tracts that lost eligibility show no pricing change ($\beta = -0.004$, $p = 0.93$). This one-sided response is consistent with banks responding to newly gained CRA incentives by extending credit to marginally riskier borrowers---borrowers who command higher rates. When CRA eligibility is removed, banks do not correspondingly tighten lending terms, perhaps because the borrowers already receiving credit in those tracts are not marginal.

\subsection{Event Study}

\input{tables/tab4_eventstudy}

\Cref{tab:eventstudy} reports the event-study estimates from \Cref{eq:eventstudy}. The pre-treatment coefficients ($t = -6$ to $t = -2$) are close to zero and jointly insignificant across all three outcomes: log originations ($F = 0.26$, $p = 0.93$), approval rate ($F = 0.73$, $p = 0.60$), and minority share ($F = 1.10$, $p = 0.36$). The pre-trends are precisely flat, confirming that reclassified and control tracts followed parallel trajectories before the MSA redefinition.

\subsection{Robustness}

\input{tables/tab5_robustness}

Panel A of \Cref{tab:rdd} reports RDD estimates at the 80\% income threshold using the 2024 cross-section. The local linear regression with MSE-optimal bandwidth finds no significant discontinuity in lending volume or approval rates at the CRA eligibility cutoff. This is consistent with the DiD null on volume: the 80\% threshold does not create a sharp jump in how much credit a tract receives. The RDD has less power than the DiD for rate spreads because the cross-sectional design cannot separate tract-level pricing from bank-level variation.

Panel B presents DiD robustness checks. The donut specification, which drops tracts within 2 percentage points of the 80\% threshold, produces similar point estimates, confirming that results are not driven by tracts precisely at the boundary. The temporal placebo, which applies the same DiD specification to a fake treatment date of 2022 using only pre-treatment data, shows no significant effects ($\beta = -0.025$, SE = 0.095 for log originations), confirming that the reclassified tracts were not on a differential trajectory before the actual MSA redefinition.

\section{Discussion and Conclusion}

The denominator shuffle reveals something surprising about CRA: its regulatory classification does not change how much credit a neighborhood receives, but it does change the terms on which credit is extended. When a tract mechanically gains CRA eligibility, banks do not increase total lending volume. Instead, the composition of lending shifts---rate spreads rise by 0.13 percentage points, consistent with banks serving marginally riskier borrowers to satisfy CRA requirements. This is not the ``CRA expands credit access'' story that proponents tell, nor the ``CRA has no effect'' story that critics claim. It is a third story: CRA reshuffles loan pricing without moving the volume dial.

Three implications follow. First, the volume null is informative. With 205 treated tracts, six years of pre-treatment data, and standard errors of 0.057 on log originations, the design has sufficient power to detect effects as small as 11\%. CRA reclassification does not produce volume effects of this magnitude, suggesting that neighborhood credit supply is not constrained by CRA eligibility \textit{per se}. This is consistent with \citet{bhutta2011community}, who found modest CRA effects concentrated around exam windows, and with the broader finding that modern mortgage markets are sufficiently competitive that regulatory classification does not ration credit \citep{berger2005does}.

Second, the pricing asymmetry---effects only in gained-LMI tracts, not lost-LMI tracts---suggests a specific mechanism. When a tract becomes CRA-eligible, banks extend credit to borrowers at the margin of approval, accepting higher rate spreads. When eligibility is removed, banks do not correspondingly withdraw credit from existing borrowers. This is consistent with \citet{agarwal2012did}'s finding that CRA effects concentrate around the extensive margin of borrower risk.

Third, the heterogeneity by minority population share reveals that the volume null masks compositional reallocation. In low-minority tracts, reclassification is associated with higher originations (SDE = 0.19); in high-minority tracts, the association reverses (SDE = $-0.13$). This pattern is consistent with CRA incentives inducing banks to shift attention toward newly eligible low-minority tracts at the expense of existing high-minority neighborhoods---a reallocation effect that nets to zero in the aggregate.

A limitation of this design is the single post-treatment year (2024). The rate-spread finding, while statistically significant, could reflect transitory adjustment rather than permanent repricing. Future work can track whether the effect persists, and the 2013 MSA redefinition offers a natural extension once historical HMDA bulk data access is restored.

The broader lesson is methodological: regulatory thresholds that depend on reference-area statistics create quasi-random variation whenever the reference area is redrawn. This ``denominator shuffle'' logic extends beyond CRA to any policy that classifies geography relative to an area median---school funding formulas, HUD fair market rent calculations, and Qualified Opportunity Zone designations all share this structure. Each MSA redefinition shuffles the deck.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

The primary data source is the Consumer Financial Protection Bureau's Home Mortgage Disclosure Act (HMDA) dataset, accessed via the CFPB Data Browser API\footnote{\url{https://ffiec.cfpb.gov/data-browser/}}. I download loan-level records for each state-year combination for 12 US states: New York, Pennsylvania, Ohio, Illinois, Georgia, North Carolina, Michigan, New Jersey, Virginia, Washington, Arizona, and Colorado. For each state-year, I aggregate to the census-tract level, constructing counts of originations, denials, and applications; mean loan amounts and rate spreads; and racial composition of borrowers. The aggregation uses all action types (action\_taken = 1--7) for application counts and restricts to originations (action\_taken = 1) for loan characteristics.

Census tract CRA eligibility is determined by the HMDA field tract-to-MSA income percentage, which reports the tract's median family income as a percentage of the MSA/MD area median family income. A tract is classified as LMI if this ratio is at or below 80\%. Reclassified tracts are identified by comparing this field in 2023 (pre-OMB Bulletin 23-01) and 2024 (post-redefinition).

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
