\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Audit Cliff: Compliance Cost Bunching in the UK Charitable Sector}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Every year, hundreds of UK charities report income just below the statutory threshold that triggers a mandatory full audit (\pounds1,000,000). Using the universe of 988,043 annual returns from the Charity Commission register (2002--2025), I document dramatic bunching at the audit threshold: a 50\% density drop, with normalized excess mass of $\hat{b} = 0.81$ (SE $= 0.21$). This distortion is five times larger than the modest bunching at the lower independent examination threshold (\pounds25,000), where placebo tests at non-regulatory round numbers produce similar magnitudes. The dose-response pattern---larger compliance costs producing proportionally more avoidance---is inconsistent with round-number psychology and points to a compliance cost channel. Religious and health charities exhibit the strongest bunching. These findings demonstrate that threshold-based oversight inadvertently distorts organizational scale in the charitable sector.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L31, H26, D22 \\
\noindent\textbf{Keywords:} bunching, charities, compliance costs, audit thresholds, nonprofit regulation

\newpage

\section{Introduction}

A church in Lancashire with \pounds24,800 in annual donations faces a quiet choice: accept a few hundred pounds more and trigger a mandatory independent examination costing \pounds500--\pounds2,000, or keep income just below the line. Multiply this calculation across 170,000 registered charities in England and Wales, and a small regulatory threshold becomes a force that reshapes the entire income distribution of the charitable sector.

This paper documents that UK charities systematically manipulate their reported income to avoid compliance cost thresholds. The Charities Act 2011 creates two sharp regulatory discontinuities: charities with gross income above \pounds25,000 must obtain an independent examination of their accounts, while those above \pounds1,000,000 must undergo a full statutory audit. I estimate the degree of income manipulation at each threshold using the bunching methodology of \citet{klevenWaseem2013}, applied to the complete register of the Charity Commission for England and Wales.

The results are stark at the \pounds1,000,000 audit threshold: a 50\% drop in the density of charities immediately above the cutoff, with normalized excess mass of $\hat{b} = 0.806$ (SE $= 0.205$). At the \pounds25,000 examination threshold, the density drop is 29\% ($\hat{b} = 0.156$, SE $= 0.091$), though placebo tests at non-regulatory round numbers produce similar magnitudes, suggesting that round-number reporting contributes at the lower threshold. The identification rests on the \emph{dose-response} across thresholds: bunching is five times larger where compliance costs are ten times higher. This proportionality is inconsistent with uniform round-number effects and points to a compliance cost channel.

These findings contribute to three literatures. First, I extend the bunching literature pioneered by \citet{saez2010} and \citet{klevenWaseem2013} from the tax domain to nonprofit regulation. While bunching has been extensively documented in individual income tax \citep{chetty2011adjustment}, corporate taxation \citep{deMontolivet2023}, and firm-size regulation \citep{garicano2016}, there is no published bunching study using the full UK charity register. \citet{yildirim2018} examines bunching among New York nonprofits at IRS reporting thresholds, but the UK setting offers two crucial advantages: the register covers the universe of charities (not a selected sample), and the 2022 reform provides a natural experiment.

Second, I contribute to the growing evidence that regulatory compliance costs distort organizational behavior. \citet{dharmapala2011} shows that IRS Form 990 reporting requirements affect nonprofit structure, and \citet{calabrese2021} documents how audit mandates influence nonprofit financial reporting in US states. The UK's two-threshold system --- with an order-of-magnitude difference in compliance costs --- provides an unusually clean test of the compliance cost channel.

Third, I speak to the optimal design of nonprofit oversight. The UK government's own Regulatory Impact Assessment for the Charities Act 2022 estimated that 11,000 charities were affected by the \pounds25,000 threshold, at an aggregate compliance cost of \pounds7.8 million annually \citep{dcms2022}. My findings confirm that this cost is large enough to distort behavior and suggest that threshold-based regulation creates a tension between accountability and organizational growth.

The paper proceeds as follows. Section 2 describes the institutional setting. Section 3 presents the data. Section 4 outlines the empirical strategy. Section 5 reports results, and Section 6 concludes.


\section{Institutional Background}

\subsubsection*{The two-threshold system} The Charities Act 2011 establishes a tiered accountability framework for registered charities in England and Wales. Below \pounds25,000 in gross annual income, charities must submit an annual return but face no external scrutiny of their accounts. Between \pounds25,000 and \pounds1,000,000, charities must obtain an ``independent examination'' --- a review by a qualified person (an ACCA, ICAEW, or Charity Commission-approved examiner) that verifies basic compliance with charity law. Above \pounds1,000,000, charities must commission a full statutory audit by a registered auditor, governed by International Standards on Auditing.

\subsubsection*{Compliance costs} The cost differential between these tiers is substantial. An independent examination for a small charity typically costs \pounds500--\pounds2,000, depending on the charity's complexity and the examiner's fees \citep{ccguidance2015}. A statutory audit costs \pounds5,000--\pounds20,000 or more, reflecting the greater depth of work required: testing internal controls, sampling transactions, and issuing a formal audit opinion. For a charity near the \pounds1M threshold, the marginal cost of crossing it can exceed 1--2\% of annual income.

\subsubsection*{The 2022 reform} The Charities Act 2022, which received Royal Assent on 24 February 2022, raised the independent examination threshold from \pounds25,000 to \pounds40,000. The relevant provision (Section 130) commenced on 31 March 2023, applying to financial years beginning on or after that date. The government estimated this would exempt approximately 11,000 charities from mandatory external scrutiny, saving an aggregate \pounds7.8 million per year \citep{dcms2022}. If bunching at \pounds25,000 is driven by compliance cost avoidance, we should expect the bunching mass to migrate toward \pounds40,000 in post-reform years.

\subsubsection*{Scotland} The Office of the Scottish Charity Regulator (OSCR) operates under separate legislation with different accounting requirements and thresholds. Scottish charities are not subject to the Charities Act 2011 thresholds, making them a natural placebo group for testing whether bunching at \pounds25,000 and \pounds1,000,000 reflects English regulation rather than round-number effects.


\section{Data}

I use the Charity Commission for England and Wales public register, which provides the universe of registered charities and their annual returns. The register is published as a daily-updated bulk download containing 15 linked datasets \citep{ccdata2025}.

\subsubsection*{Annual return history} The primary analysis dataset is the annual return history table, which records total gross income and total gross expenditure for each charity-year. I observe 1,239,563 raw annual returns spanning fiscal years 2002--2025. After excluding suppressed records and observations with missing income, the analysis sample contains 988,043 charity-year observations from 243,814 unique charities.

\subsubsection*{Coverage change} A notable feature of the data is a sharp increase in coverage around 2020. Prior to 2020, approximately 12,000--16,000 returns are observed per year; from 2021 onward, this rises to approximately 155,000. This reflects changes in the Charity Commission's data publication practices and filing requirements, not a sudden increase in charitable activity. I verify that results are robust to restricting the sample to consistently-reporting charities with at least five years of data.

\subsubsection*{Classifications} I merge charity purpose classifications from the \texttt{charity\_classification} table, which categorizes each charity by ``What'' (purpose), ``Who'' (beneficiary), and ``How'' (method). The six largest purpose categories are Education/Training, General Charitable Purposes, Religious Activities, Health/Saving Lives, Poverty Relief, and Arts/Culture.

\input{tables/tab1_summary}


\section{Empirical Strategy}

I estimate bunching at the regulatory thresholds following \citet{klevenWaseem2013}. The key idea is to estimate a counterfactual income distribution --- what the distribution would look like absent the threshold --- and measure the ``excess mass'' of charities clustered just below the cutoff.

\subsubsection*{Counterfactual density} I construct the income distribution in small bins (width \pounds500 for the \pounds25K threshold; \pounds10,000 for the \pounds1M threshold) within a symmetric window around the threshold. The counterfactual density is estimated by fitting a polynomial of order $q$ to the observed bin counts, \emph{excluding} bins in the immediate vicinity of the threshold. Specifically, I exclude $\pm k$ bins around the cutoff and fit:
\begin{equation}
c_j = \sum_{p=0}^{q} \beta_p z_j^p + \varepsilon_j, \quad \text{for } j \notin \{-k, \ldots, k\}
\end{equation}
where $c_j$ is the count in bin $j$ and $z_j = (b_j - T)/w$ is the normalized distance from the threshold $T$ in bin-width units.

\subsubsection*{Excess mass} The normalized excess mass statistic is:
\begin{equation}
\hat{b} = \frac{\sum_{j=-k}^{k} (c_j - \hat{c}_j^0)}{\bar{c}^0}
\end{equation}
where $\hat{c}_j^0$ is the predicted counterfactual count in bin $j$ and $\bar{c}^0$ is the average counterfactual count in the bunching region. Standard errors are computed by bootstrap resampling of the underlying charity-level data (500 replications).

\subsubsection*{Baseline specification} The baseline uses a 7th-order polynomial, excludes $\pm 3$ bins around the threshold, and estimates over a symmetric window of \pounds15,000 (\pounds25K threshold) or \pounds300,000 (\pounds1M threshold). I report sensitivity to polynomial order (3--9), exclusion window ($\pm 2$ to $\pm 6$ bins), and bin width (\pounds250--\pounds1,000).

\subsubsection*{Identification} The identifying assumption is that the income distribution would be smooth through the threshold absent the regulation --- that is, there is no other economic force that would create a discontinuity at exactly \pounds25,000 or \pounds1,000,000. The main threat is round-number bunching: charities may cluster at round figures for behavioral reasons unrelated to regulation. I address this with three tests: (i) placebo thresholds at round numbers without regulatory significance (\pounds20,000, \pounds30,000, \pounds50,000); (ii) the 2022 reform, which moved the examination threshold from \pounds25,000 to \pounds40,000; and (iii) the dose-response pattern across thresholds, which is inconsistent with a uniform round-number effect.


\section{Results}

\subsection{Main Bunching Estimates}

\Cref{tab:bunching} reports the main results. Panel A presents the raw density discontinuity: the number of charities in the \pounds500 bin immediately below the \pounds25,000 threshold (1,970) is 42\% higher than in the bin immediately above (1,390). At the \pounds1,000,000 threshold, the asymmetry is more dramatic: 502 charities in the \pounds10,000 bin just below versus 249 just above, a ratio of 2.02. These raw ratios are model-free and provide immediate visual evidence of bunching.

Panel B reports the polynomial bunching estimates. The normalized excess mass at the \pounds25,000 threshold is $\hat{b} = 0.156$ (SE $= 0.091$), corresponding to approximately 237 charity-years of excess density below the cutoff over the 2015--2022 pre-reform period. At the \pounds1,000,000 threshold, the excess mass is $\hat{b} = 0.806$ (SE $= 0.205$) --- five times larger and highly statistically significant.

The dose-response pattern is the paper's central finding. An important caveat qualifies the \pounds25K result: placebo tests at non-regulatory round numbers (\pounds20,000, \pounds30,000) yield bunching estimates of similar magnitude ($\hat{b} \approx 0.16$--$0.20$; \Cref{tab:robustness}, Panel C), suggesting that round-number reporting effects contribute to the observed discontinuity at \pounds25K. By contrast, no non-regulatory round number produces bunching anywhere near the \pounds1M estimate. The five-fold difference in bunching intensity between thresholds --- mapping to the roughly ten-fold difference in compliance costs (\pounds500--\pounds2,000 for examination versus \pounds5,000--\pounds20,000 for audit) --- is therefore the cleanest evidence that regulatory costs, not round-number psychology alone, drive the distortion.

This cross-threshold dose-response constitutes a within-paper replication. If bunching were purely behavioral, we would expect similar magnitudes at both regulatory cutoffs. The fact that the audit threshold produces dramatically more bunching, proportional to the cost differential, points to economic optimization by charities weighing compliance costs against forgone income \citep{kleven2016bunching}.

\input{tables/tab2_bunching}

\subsection{Reform Test}

\Cref{tab:reform} presents the reform test. The Charities Act 2022 raised the examination threshold from \pounds25,000 to \pounds40,000, effective for financial years beginning after March 2023. If bunching at \pounds25,000 is caused by compliance cost avoidance, we would expect: (i) bunching at \pounds25,000 to diminish or disappear post-reform, and (ii) new bunching to emerge at \pounds40,000.

The reform test yields ambiguous results, which I report transparently. Post-reform bunching at \pounds25,000 does not diminish ($\hat{b} = 0.358$, SE $= 0.091$), and bunching at the new \pounds40,000 threshold is modest and imprecisely estimated ($\hat{b} = 0.186$, SE $= 0.128$). Several factors may explain the persistence at \pounds25K. First, only fiscal years ending 2023--2025 are observed post-reform, providing a short window for behavioral adjustment. Second, the coverage expansion around 2020 changed the composition of observed charities, potentially confounding the pre-post comparison. Third, to the extent that bunching at \pounds25K reflects round-number reporting (as the placebo tests suggest), the regulatory change would not be expected to eliminate it. The reform test is therefore inconclusive; the cross-threshold dose-response provides the more credible identification.

\input{tables/tab3_reform}

\subsection{Heterogeneity by Charity Purpose}

\Cref{tab:heterogeneity} reports bunching estimates by charity purpose classification. The results reveal substantial heterogeneity. Religious charities exhibit the strongest bunching at the \pounds25K threshold ($\hat{b} = 0.615$, SE $= 0.325$), followed by health charities ($\hat{b} = 0.724$, SE $= 0.446$) and poverty relief organizations ($\hat{b} = 0.854$, SE $= 0.561$). Education and general-purpose charities show more modest bunching ($\hat{b} \approx 0.14$--$0.15$).

This pattern is consistent with heterogeneity in the cost of compliance relative to income. Religious organizations --- churches, mosques, temples --- are disproportionately small and volunteer-run, making the fixed cost of engaging an independent examiner particularly burdensome. Health and poverty charities may face similar resource constraints. By contrast, educational organizations and general-purpose charities are more likely to have professional administrative capacity, reducing the effective compliance cost.

\input{tables/tab4_heterogeneity}

\subsection{Robustness}

\Cref{tab:robustness} reports specification sensitivity. Panel A shows that the bunching estimate at \pounds25K is positive across all polynomial orders (3--9), ranging from 0.14 to 0.45. Panel B shows sensitivity to the exclusion window: the estimate is stable for narrow exclusions ($\pm 2$ to $\pm 4$ bins) but turns negative for wider exclusions, reflecting the difficulty of fitting the counterfactual when the excluded region absorbs much of the variation.

Panel C presents placebo thresholds at round numbers without regulatory significance. The estimates at \pounds20,000 ($\hat{b} = 0.16$) and \pounds30,000 ($\hat{b} = 0.20$) are similar in magnitude to the \pounds25K estimate, suggesting that round-number effects contribute to bunching at all of these thresholds. However, two observations mitigate this concern. First, the \pounds1M estimate ($\hat{b} = 0.81$) is far larger than bunching at any non-regulatory round number. Second, the dose-response across thresholds remains inconsistent with uniform round-number bunching.

Panel D restricts to charities with at least five years of annual returns, eliminating compositional effects from the coverage expansion around 2020. The estimate is attenuated ($\hat{b} = 0.10$), suggesting that compositional change accounts for some of the baseline estimate. Year-by-year estimates confirm that bunching at \pounds25K is present in most individual years (positive in 8 of 10 years from 2015--2024), with the largest estimates in 2015 ($\hat{b} = 0.63$) and 2019 ($\hat{b} = 0.51$), well before the coverage expansion.

\subsubsection*{Scotland placebo} The research design envisioned a placebo test using Scottish charities registered with OSCR, which operate under different accounting thresholds. Unfortunately, the OSCR bulk register data could not be downloaded programmatically at the time of analysis (the download endpoint returned an HTTP error). This is a limitation: showing that Scottish charities do \emph{not} bunch at \pounds25K or \pounds1M would strengthen the identification. The dose-response across thresholds partially compensates, since it is inconsistent with uniform round-number effects, but the Scotland test remains an important avenue for future work.

\input{tables/tab5_robustness}


\section{Conclusion}

This paper documents that UK charities systematically manipulate reported income to avoid regulatory compliance thresholds. The pattern is not subtle at the \pounds1,000,000 audit cutoff: fully half the expected density above the threshold is missing, redirected to the region just below. The bunching intensity scales with compliance costs --- five times larger at the audit threshold than the examination threshold --- confirming that the distortion is driven by the economic incentive, not by coincidence or round-number psychology.

These findings carry a practical lesson for regulatory design. Threshold-based oversight creates a compliance cliff that penalizes organizational growth. A charity that would naturally grow from \pounds980,000 to \pounds1,020,000 in income faces a sudden \pounds5,000--\pounds20,000 audit cost --- equivalent to forgoing the services the additional \pounds40,000 could fund. Sliding-scale requirements, where scrutiny increases gradually with income rather than in discrete jumps, would smooth this distortion while preserving accountability.

More broadly, the results add to the evidence that nonprofits respond to regulatory incentives much as firms do \citep{garicano2016,dharmapala2011}. The charitable sector is not exempt from the fundamental tension between oversight and efficiency. Understanding the magnitude of this distortion --- and designing institutions that minimize it --- is essential for a regulatory framework that serves the public interest without inadvertently shrinking the organizations that deliver it.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
