\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{enumitem}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Tax Flight Gradient: Income-Stratified Migration Elasticities from the SALT Deduction Cap}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Does taxing the rich drive them away? The 2017 SALT deduction cap sharply raised the effective cost of residing in high-tax states, but only for high-income itemizers. Using the universe of IRS tax-return migration data across seven income brackets, I estimate the full income gradient of tax-induced migration. A triple-difference design---comparing high- versus middle-income filers, in high- versus low-SALT states, before and after 2018---reveals that net outmigration of filers earning over \$200,000 from high-SALT states increased by 0.39 percentage points, three times the response of filers under \$10,000. This gradient is monotonically increasing in income and driven primarily by accelerated outflows rather than reduced inflows. The results survive excluding the top three SALT states and pandemic years. These findings bridge the near-zero elasticities in Young and Varner (2011) with the larger responses in Kleven, Landais, and Saez (2013): migration responsiveness to taxation is sharply income-graded.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H24, H73, R23 \\
\noindent\textbf{Keywords:} tax migration, SALT cap, income gradient, interstate mobility, fiscal federalism

\newpage

\section{Introduction}

A perennial fear in fiscal federalism is that progressive state taxation is self-defeating: raise rates on the rich and they simply leave. If this is true, the revenue base is a treadmill---states compete by cutting top rates, underfunding public goods in the process. If it is false, states have more fiscal space than they think. Which is it?

The empirical literature has produced sharply conflicting answers. \citet{young2011} and \citet{young2016} found near-zero millionaire migration elasticities in California and New Jersey, concluding that ``the rich don't flee.'' But \citet{kleven2013} estimated elasticities near 1.0 for top European footballers, and \citet{moretti2017} documented significant inventor responses to state tax differentials. The disconnect may reflect measurement differences, population scope, or the fact that these studies examined different slices of the income distribution. What no study has done is estimate the \textit{full income gradient} of tax-induced migration---how responsiveness varies continuously from the bottom to the top of the distribution---using a single identification strategy and comprehensive data.

This paper fills that gap by exploiting the 2017 Tax Cuts and Jobs Act (TCJA) cap on the federal deduction for state and local taxes (SALT). Before 2018, high-income filers in states like New York, New Jersey, and Connecticut could deduct their full state income and property taxes from federal taxable income. The \$10,000 SALT cap eliminated most of this benefit, sharply raising the \textit{effective} after-tax cost of living in high-tax states---but only for upper-income itemizers whose SALT deductions exceeded the cap. This differential exposure across the income distribution within the same state creates ideal variation for a triple-difference design.

I construct a panel of 51 jurisdictions $\times$ 7 AGI brackets $\times$ 11 years (2011--2021) from the universe of IRS Statistics of Income migration files---approximately 125 million matched tax returns per year. My main specification estimates:
\begin{equation}
Y_{sbt} = \beta(\text{HighSALT}_s \times \text{Post}_t \times \text{HighIncome}_b) + \gamma_{sb} + \delta_{bt} + \theta_{st} + \varepsilon_{sbt}
\end{equation}
where state$\times$bracket, year$\times$bracket, and state$\times$year fixed effects absorb time-varying state-level shocks and bracket-level trends. The coefficient $\beta$ identifies the differential migration response of high-income filers in high-SALT states relative to all three counterfactuals.

The results reveal a striking income gradient. The triple-difference estimate for net migration is $-0.0048$ (SE = 0.0011, $p < 0.001$), meaning high-income filers in high-SALT states experienced an additional 0.48 percentage-point decline in net migration after 2018 relative to their low-income counterparts in those same states and high-income filers in low-SALT states. Bracket-by-bracket estimates show the SALT cap effect rising monotonically from $-0.0012$ for filers under \$10,000 to $-0.0039$ for those above \$200,000---a threefold gradient. Decomposing into channels, approximately 70\% of the effect comes from increased outflows (filers leaving faster) and 30\% from reduced inflows (fewer arriving).

The key threat to identification is that other post-2018 changes differentially affected high-income residents of high-SALT states. Several robustness checks address this. First, excluding New York, New Jersey, and Connecticut---the most extreme SALT states---strengthens rather than weakens the estimate ($-0.0054$, $p < 0.001$). Second, dropping the COVID-affected years 2020--2021 yields a smaller but still significant estimate ($-0.0032$, $p < 0.001$), suggesting the pandemic amplified but did not create the pattern. Third, using continuous SALT exposure rather than a binary threshold produces consistent results.

These findings offer new evidence on the income gradient of tax-induced migration, providing a bridge between the near-zero elasticities of \citet{young2011} and the larger responses of \citet{kleven2013}: migration responsiveness to tax-related cost shocks is sharply increasing in income, even within the same national setting. Second, the paper contributes to the optimal state taxation literature \citep{gordon1986,epple1991,haughwout2004} by quantifying the intensive margin of the revenue-mobility tradeoff across the income distribution. Third, the decomposition into outflow and inflow channels has direct implications for place-based policy: the SALT cap primarily \textit{pushed} people out rather than discouraging entry, suggesting that destination amenities may buffer inflow responses even when tax costs rise \citep{diamond2016}.

The balance of this paper proceeds as follows. Section 2 describes the SALT cap and why it creates differential exposure. Section 3 details the IRS SOI migration data. Section 4 presents the empirical strategy. Section 5 reports results, including the income gradient, mechanism decomposition, and robustness. Section 6 discusses implications and concludes.


\section{The SALT Deduction Cap}

Before 2018, taxpayers who itemized deductions on their federal returns could deduct the full amount of state and local income, sales, and property taxes paid. This deduction effectively subsidized residence in high-tax states: a filer in the 39.6\% federal bracket living in New York (combined state+local rate $\approx$ 12\%) faced an effective state tax rate of only 7.3\% after accounting for the federal offset. The subsidy was worth thousands of dollars annually to high-income filers in states like New York (average SALT deduction: \$22,200), New Jersey (\$18,600), and Connecticut (\$18,100), while providing minimal benefit to filers in low-tax or no-income-tax states \citep{sammartino2018}.

The TCJA, signed December 22, 2017, capped the total SALT deduction at \$10,000 beginning January 1, 2018. This cap was binding for the vast majority of itemizers in high-SALT states: in 2017, 89\% of New York itemizers, 77\% of New Jersey itemizers, and 77\% of Connecticut itemizers claimed SALT deductions exceeding \$10,000 \citep{taxfoundation2018}. The cap raised the effective cost of living in these states overnight.

\subsubsection*{Differential Exposure by Income} The SALT cap's bite is sharply income-graded. Filers earning under \$50,000 rarely itemize and, when they do, typically claim SALT deductions under \$10,000. The cap is essentially non-binding for them. By contrast, filers earning over \$200,000 in high-tax states routinely claimed SALT deductions of \$30,000--\$60,000; the cap represented a tax increase of \$5,000--\$15,000 per year. This within-state, across-income variation is the identifying variation in this paper.

\subsubsection*{Timing and Anticipation} Although the TCJA was debated throughout 2017, many analysts did not expect the SALT cap to survive conference committee negotiations \citep{nyt2017}. Its inclusion in the final bill came as a partial surprise, limiting anticipatory migration. I define the post-treatment period as 2018 onward, with 2017 as the last pre-treatment year.


\section{Data}

The primary data source is the IRS Statistics of Income (SOI) State-to-State Migration Files for filing years 2011--2021. These files are constructed by matching individual tax returns across consecutive years using Social Security numbers, providing a nearly complete census of interstate migration among the filing population \citep{irssoi2023}. Each annual file reports, for each state-of-origin, the number of returns (filers), exemptions (individuals), and adjusted gross income (AGI) flowing to each destination state, disaggregated by seven AGI brackets: under \$10,000; \$10,000--\$25,000; \$25,000--\$50,000; \$50,000--\$75,000; \$75,000--\$100,000; \$100,000--\$200,000; and \$200,000+.

The unit of analysis is the state $\times$ AGI bracket $\times$ year. For each cell, I construct the \textit{net migration rate} as:
\begin{equation}
\text{NetMigRate}_{sbt} = \frac{\text{Inflow}_{sbt} - \text{Outflow}_{sbt}}{\text{TotalReturns}_{sbt}}
\end{equation}
where inflows and outflows are measured in returns (number of filers). I also construct outflow rates and inflow rates separately for the decomposition analysis, and AGI-weighted net flows for dollar-magnitude estimates.

SALT exposure is measured using 2017 (pre-TCJA) average SALT deductions by state, sourced from IRS SOI individual statistics. I classify states with average SALT deductions above \$13,000 as ``high-SALT'' (9 states: NY, NJ, CT, CA, MA, IL, MN, MD, DC), with continuous SALT exposure used in robustness specifications.

\Cref{tab:sumstats} presents summary statistics by AGI bracket. Net migration rates are highest for the top bracket (\$200K+), reflecting the greater geographic mobility of high-income filers. Outflow and inflow rates decline with income, consistent with lower migration frequencies at higher income levels but larger net flows.

\input{tables/tab1_sumstats}


\section{Empirical Strategy}

\subsection{Triple-Difference Design}

The identification strategy exploits three dimensions of variation created by the SALT cap. First, across income brackets: high-income filers (AGI $\geq$ \$100,000, brackets 6--7) were disproportionately affected because their SALT deductions typically exceeded the \$10,000 cap, while low- and middle-income filers were largely unaffected. Second, across states: filers in high-SALT states (average 2017 SALT deduction $\geq$ \$13,000) lost a larger subsidy than filers in low-SALT states. Third, over time: the cap took effect in 2018, creating a sharp before-after comparison.

The estimating equation is:
\begin{equation}
Y_{sbt} = \beta \left(\text{HighSALT}_s \times \text{Post}_t \times \text{HighIncome}_b \right) + \gamma_{sb} + \delta_{bt} + \theta_{st} + \varepsilon_{sbt} \label{eq:triple}
\end{equation}
where $Y_{sbt}$ is the net migration rate for state $s$, AGI bracket $b$, year $t$. The three sets of two-way fixed effects absorb: all time-invariant differences across state-bracket pairs ($\gamma_{sb}$); common year-specific shocks to each income bracket ($\delta_{bt}$); and all state-year shocks that affect all brackets equally ($\theta_{st}$). This last term is particularly important, as it absorbs state-level economic conditions, housing markets, and any policy change that does not differentially affect income brackets.

Standard errors are clustered at the state level (51 clusters). The coefficient $\beta$ identifies the \textit{differential} change in net migration for high-income filers in high-SALT states after 2018, relative to all three counterfactual groups. A negative $\beta$ indicates that the SALT cap induced net outmigration of high-income filers from high-SALT states.

\subsection{Threats to Validity}

The key identifying assumption is that, absent the SALT cap, the difference in net migration between high- and low-income filers would have evolved similarly in high- and low-SALT states. Several concerns merit discussion.

\subsubsection*{Concurrent Shocks} The TCJA included many provisions beyond the SALT cap (rate cuts, pass-through deduction, estate tax threshold increase). However, most provisions were national in scope and do not generate differential state$\times$bracket$\times$time variation. The state$\times$year fixed effects absorb any provision that affects all brackets within a state equally.

\subsubsection*{COVID-19} The 2020--2021 period saw large migration shifts driven by remote work, which disproportionately affected high-income workers in coastal cities that also happen to be high-SALT states. I address this by re-estimating the model excluding 2020--2021, relying on 2018--2019 for the post-treatment period.

\subsubsection*{State Tax Changes} Several states changed their top income tax rates during the sample period (California Prop 30 in 2012, Minnesota in 2013, Connecticut in 2015, New Jersey in 2020, Arizona in 2021). These could confound the SALT cap effect if they coincide with it. The state$\times$year fixed effects absorb the bracket-constant component; the remaining concern is that state tax changes differentially affected top brackets. I address this with leave-one-out robustness.


\section{Results}

\subsection{The Income Gradient}

\Cref{tab:gradient} reports the core finding: bracket-by-bracket estimates of the SALT cap's effect on net migration. Each row presents a separate regression of net migration rate on the interaction of standardized SALT exposure with a post-2018 indicator, with state and year fixed effects, clustered at the state level.

\input{tables/tab2_gradient}

The income gradient is monotonically increasing. The SALT cap effect on net migration rises from $-0.0012$ for filers under \$10,000 (marginally significant) to $-0.0039$ for filers above \$200,000 (significant at the 1\% level). This threefold ratio is the paper's central finding: the rich \textit{do} move in response to taxes, and their responsiveness is roughly proportional to their exposure.

To put these magnitudes in perspective, the \$200K+ estimate implies that a one-standard-deviation increase in state SALT exposure (approximately \$4,000 in average deductions) is associated with an additional 0.39 percentage-point decline in net migration for top-bracket filers. Given roughly 137,000 top-bracket filers per state, this corresponds to approximately 535 net departing filers per state per year---modest in headcount, but substantial in AGI terms.

\subsection{Triple-Difference Estimates}

\Cref{tab:triple} presents the main triple-difference results. Column (1) reports the binary specification: the interaction of HighSALT $\times$ Post $\times$ HighIncome yields $\hat{\beta} = -0.0048$ (SE = 0.0011), significant at the 0.1\% level. Column (2) uses continuous SALT exposure, yielding a consistent estimate of $-0.0018$ per standard deviation.

\input{tables/tab3_triple}

Columns (3) and (4) decompose the effect into outflow and inflow channels. The outflow coefficient is $+0.0034$ (accelerated departures), while the inflow coefficient is $-0.0014$ (reduced arrivals). Approximately 70\% of the net migration response operates through the outflow channel---high-income filers \textit{leaving} high-SALT states faster---and 30\% through reduced inflows. This decomposition suggests that, while the SALT cap made high-tax states less attractive to prospective movers, its primary effect was on incumbent residents' departure decisions.

\subsection{Robustness}

\Cref{tab:robustness} presents robustness checks on the triple-difference estimate. The baseline effect of $-0.0048$ is robust to alternative clustering at the state$\times$bracket level ($-0.0048$, $p < 0.001$) and to excluding the three highest-SALT states ($-0.0054$, $p < 0.001$). Critically, dropping the pandemic years 2020--2021 yields a smaller but still significant estimate ($-0.0032$, $p < 0.001$), confirming that the pattern predates COVID-era remote work. Given the plausible confound between pandemic migration and SALT-driven flight, the pre-COVID estimate of 0.32 percentage points may be the more conservative and credible benchmark.

\input{tables/tab4_robustness}

The placebo test for low-income brackets (AGI $< \$50$K) returns a positive and statistically significant coefficient ($+0.003$). This warrants candid discussion. Low-income filers should face minimal SALT exposure, so a true placebo would show zero. One interpretation is that as high-income residents departed high-SALT states, housing costs moderated, drawing in lower-income movers---a general-equilibrium response rather than a direct SALT effect. Alternatively, broader post-2018 trends (urban-to-Sunbelt migration, housing affordability shifts) may have differentially affected high-SALT states across the income distribution. While the monotonic gradient---with top-bracket effects three times larger than bottom-bracket effects---is consistent with the SALT mechanism, the non-zero low-income placebo means this paper estimates an upper bound on the purely tax-driven component of the gradient.


\section{Discussion and Conclusion}

This paper estimates the income gradient of migration responses to the SALT deduction cap---a large, income-graded shock to the effective cost of residing in high-tax states. The gradient is monotonically increasing: top-bracket filers respond roughly three times more than the lowest-bracket filers, and the effect operates primarily through accelerated departures rather than reduced inflows.

These findings provide new evidence that bridges the conflicting results in the tax-migration literature. \citet{young2011} and \citet{young2016} found near-zero millionaire elasticities; \citet{kleven2013} found large superstar responses. The results here suggest that both findings are consistent with a steeply graded elasticity that is modest for most of the income distribution but becomes detectable and economically relevant near the top. However, this paper measures reduced-form responses to a \textit{federal} change in the deductibility of state taxes, not direct state tax changes. The SALT cap changed the salience and effective cost of state taxes simultaneously; disentangling these channels requires further work.

For state fiscal policy, the income gradient implies that the revenue cost of progressive taxation is concentrated in a narrow band at the very top. A state raising its top rate faces meaningful base erosion among \$200K+ filers but minimal migration from middle-income taxpayers who constitute the majority of revenue. Whether the outflow of top filers is large enough to offset the mechanical revenue gain depends on the precise elasticity and the state's position on the tax-rate curve---a calculation this paper informs but does not resolve.

The outflow-inflow decomposition offers an additional insight. The finding that 70\% of the SALT effect operates through departures rather than reduced arrivals suggests that tax-driven migration is primarily a stock adjustment (incumbent residents re-optimizing) rather than a flow-diversion phenomenon (new residents choosing differently). This has implications for the dynamics of tax competition: a state tax increase may generate a transitory burst of outmigration that attenuates as the stock of dissatisfied marginal residents is exhausted.

This paper has important limitations. The IRS migration files measure only the filing population and cannot capture behavioral responses within brackets (income shifting, re-sourcing). The 2018 treatment coincides with other TCJA provisions that may differentially affect high-income taxpayers, though the triple-difference design addresses this concern. Finally, the post-2020 data comingles SALT effects with pandemic-induced remote work migration, though the pre-COVID estimates confirm the effect exists independently.

The central finding is that migration responsiveness to tax-related cost shocks is sharply income-graded. The gradient is steep enough that most of the income distribution shows modest responses, while the very top exhibits detectable and policy-relevant outmigration. For state fiscal policy, this means progressive taxation is not self-defeating in the aggregate---but neither is it costless. The fiscal exposure is concentrated precisely where the tax base is most valuable, and quantifying that exposure requires understanding the gradient, not just the average elasticity.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
