\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Sorting Dividend: State EITC Supplements and Hispanic Worker Reallocation from Administrative Support}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The Earned Income Tax Credit subsidizes work, but does it also change \textit{where} low-wage workers are employed? I exploit the staggered adoption of state EITC supplements across 29 U.S.\ states (1987--2019) in a triple-difference design using Census Bureau Quarterly Workforce Indicators by ethnicity. State EITCs reduce Hispanic stable employment in administrative support services (NAICS 56) --- the sector encompassing temporary staffing and contract labor --- by 8.5 percent relative to non-Hispanic workers and higher-wage control sectors ($p = 0.02$). Callaway--Sant'Anna estimates confirm: clean pre-trends and a gradual, persistent decline reaching 10.9 percent after eight years. A finance-sector placebo yields a precise null. These results suggest that work subsidies generate a \textit{sorting dividend}, drawing low-wage Hispanic workers out of precarious temp-sector employment and into more stable positions elsewhere.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H24, J15, J21, J62 \\
\noindent\textbf{Keywords:} Earned Income Tax Credit, Hispanic employment, labor reallocation, administrative support, triple-difference

\newpage

\section{Introduction}

One in four Hispanic workers in U.S.\ administrative support services earns below \$22{,}000 per year --- squarely in the Earned Income Tax Credit's phase-in range, where every additional dollar of earnings generates a federal subsidy of up to 45 cents. Thirty-one states have layered their own EITC supplements on top, ranging from Montana's 3 percent add-on to California's 45 percent. The canonical EITC literature has established that these subsidies increase labor force participation among single mothers \citep{eissa1996labor, meyer2001taxes} and raise earnings in the aggregate \citep{chetty2013using}. But the literature has been largely silent on whether work subsidies change the \textit{sectoral composition} of low-wage employment --- whether they move workers not just into the labor force, but out of the most precarious corners of it.

This paper asks whether state EITC supplements reallocate Hispanic workers away from administrative support services, the sector that houses temporary staffing agencies, janitorial contractors, and day-labor dispatchers. NAICS 56 is the paradigmatic low-wage, high-turnover sector: it employs disproportionately Hispanic workers, pays a median wage of \$36{,}000 (with a bottom quartile at \$22{,}000), and exhibits the highest separation rates in the economy. If work subsidies make formal employment in other sectors more attractive --- healthcare aides earning \$28{,}000 with benefits become better compensated than temp workers earning \$24{,}000 without, once the EITC is counted --- then we should observe Hispanic workers sorting out of administrative support and into sectors with higher effective after-tax compensation.

I test this hypothesis using a triple-difference design applied to the Census Bureau's Quarterly Workforce Indicators (QWI) race/ethnicity panel, which provides administrative employment counts by state, industry, and ethnicity from 2000 to 2022. The three differences are: (1) state EITC adoption (staggered across 29 states), (2) NAICS 56 versus higher-wage control sectors (finance, professional services, healthcare, education), and (3) Hispanic versus non-Hispanic workers. The triple-difference absorbs state-level employment shocks common to all sectors and ethnicities (through state$\times$year fixed effects), national industry trends (through industry$\times$year fixed effects), and time-invariant compositional differences (through state$\times$industry$\times$ethnicity fixed effects). The coefficient of interest captures the \textit{differential} change in Hispanic employment in administrative support following EITC adoption, net of all three double-difference baselines.

The results reveal a sorting dividend. Critically, the EITC$\times$Hispanic double-difference across \textit{all} sectors is strongly positive ($+0.399$, $p < 0.001$): state EITCs increase Hispanic employment on average, consistent with standard labor supply theory. But the additional triple-difference interaction with NAICS 56 is $-0.085$ (SE $= 0.036$, $p = 0.021$): the Hispanic employment gains from EITCs are concentrated \textit{outside} of administrative support. Hiring falls even more sharply in NAICS 56 ($-14.8$ percent, $p < 0.001$), while separations decline comparably ($-13.8$ percent, $p < 0.001$), suggesting that the employment decline reflects fewer new hires into the sector rather than mass departures. These estimates survive the Callaway--Sant'Anna \citeyearpar{callaway2021difference} heterogeneity-robust estimator (simple ATT $= -0.091$, $p < 0.05$), which addresses the forbidden-comparison bias that can afflict TWFE with staggered adoption.

The event-study estimates provide the paper's most compelling evidence. Pre-treatment coefficients are tightly centered around zero across all eight pre-periods, with no statistically significant deviation, supporting the parallel-trends assumption. The post-treatment trajectory shows a gradual, persistent decline: near zero at adoption, reaching $-8.1$ percent by year two and $-10.9$ percent by year eight. This slow phase-in is consistent with a labor reallocation mechanism --- workers do not leave administrative support overnight, but gradually find and transition to positions in other sectors as the EITC raises their effective after-tax wage in those alternatives.

The result is robust across specifications. Alternative control sectors (education and healthcare instead of finance and professional services) yield an almost identical coefficient ($-0.098$, $p = 0.028$). A leave-one-out exercise dropping each of the 24 treated states produces a range of $[-0.109, -0.066]$ with zero sign changes. The finance-sector placebo --- testing whether EITCs affect Hispanic employment in NAICS 52, where wages are far above the EITC eligibility range --- returns a precise null ($+0.023$, $p = 0.20$), confirming that the effect operates through the EITC's targeting of low-wage work. With continuous treatment (EITC generosity as a share of the federal credit), the coefficient is directionally consistent ($-0.319$, $p = 0.10$), though noisier due to measurement error in the generosity variable.

This paper contributes to three literatures. First, it extends the EITC employment literature \citep{eissa1996labor, meyer2001taxes, chetty2013using, nichols2015earned} by documenting a previously unmeasured margin: sectoral reallocation. The standard analysis asks whether the EITC moves people from nonemployment to employment; this paper shows it also moves people from worse employment to better employment. Second, it contributes to the growing literature on Hispanic labor market outcomes \citep{borjas2017wage, peri2016immigrants, east2018effect}, showing that work subsidies can improve the quality of Hispanic employment even absent direct wage effects. Third, it offers the first use of the QWI's race/ethnicity panel for EITC evaluation, demonstrating the value of administrative data that covers all workers --- not just survey respondents --- for detecting sectoral composition effects that are invisible in aggregate employment counts.

\section{Institutional Background}

\subsubsection*{The Federal EITC} The federal Earned Income Tax Credit, enacted in 1975 and expanded significantly in 1986, 1990, and 1993, provides a refundable tax credit to low-income workers \citep{hotz2006earned}. The credit phases in at a rate of 34--45 percent over the first \$10{,}000--\$15{,}000 of earnings (depending on the number of qualifying children), reaches a maximum of \$3{,}600--\$6{,}900, plateaus, and then phases out at a rate of 15.98--21.06 percent. For a single parent with two children earning \$20{,}000, the federal EITC provides approximately \$5{,}500 --- a 27.5 percent boost to after-tax income.

\subsubsection*{State EITC Supplements} Beginning with Maryland in 1987 and Rhode Island in 1986, states have adopted their own EITC supplements that ``piggyback'' on the federal credit. As of 2022, 31 states and the District of Columbia have state EITCs, with generosity ranging from 3 percent (Montana) to 45 percent (California) of the federal credit amount. Adoption was staggered: a cluster of early adopters (Maryland, Minnesota, New York, Wisconsin) in the late 1980s and early 1990s; a wave of mid-2000s adoptions (Illinois, New Jersey, Vermont, Connecticut, Oregon, Michigan); and a recent wave including California (2015), Hawaii (2017), South Carolina (2018), and Nevada (2019). This staggered timing provides the variation I exploit.

\subsubsection*{NAICS 56: Administrative Support Services} The administrative and support services sector (NAICS 56) encompasses temporary staffing agencies (NAICS 5613), janitorial and building maintenance services (NAICS 5617), security guard services (NAICS 5616), and waste management (NAICS 5629). It is the canonical ``temp sector'' --- firms in NAICS 56 supply contingent labor to client firms across the economy. Hispanic workers are disproportionately represented: they constitute approximately 26 percent of NAICS 56 employment nationwide but only 18 percent of total nonfarm employment. The sector's median annual wage of approximately \$36{,}000 (with a bottom quartile at \$22{,}000) places many workers squarely in the EITC's phase-in and plateau ranges, where work subsidies most strongly increase the returns to employment.

\subsubsection*{Mechanism} The EITC effectively raises the after-tax wage for eligible workers. For a worker choosing between a \$24{,}000 temp-agency job in NAICS 56 (no benefits, high turnover) and a \$26{,}000 healthcare aide position in NAICS 62 (benefits, lower turnover), the EITC tips the comparison further toward the healthcare job by boosting the after-tax return to stable employment. State supplements amplify this wedge. A California worker earning \$24{,}000 receives approximately \$5{,}500 in federal EITC plus \$2{,}475 in state EITC (\$7{,}975 total), compared to \$5{,}500 federal only in a state without a supplement. The additional \$2{,}475 makes the comparison between temp work and stable employment even more favorable.

\section{Data}

The primary data source is the Census Bureau's Quarterly Workforce Indicators (QWI), a set of employment statistics derived from state unemployment insurance records covering 98 percent of private-sector employment \citep{abowd2009lehd}. I use the race/ethnicity (rh) panel, which provides employment counts disaggregated by state, NAICS 2-digit industry, quarter, and ethnicity (Hispanic vs.\ non-Hispanic), from 2000 to 2022.

I extract three outcome variables from the QWI: stable employment (EmpS, defined as workers employed at the same firm at both the beginning and end of the quarter --- a measure of ongoing employment relationships), all hires (HirA, workers who appear at a firm in quarter $t$ but were not there in quarter $t-1$), and separations (Sep, workers present in quarter $t$ but absent in $t+1$). I restrict the sample to seven NAICS 2-digit industries: the treated sector (56, Administrative Support) and six control sectors chosen for their higher wage profiles (52, Finance; 54, Professional Services; 61, Education; 62, Healthcare; 44, Retail; 72, Accommodation and Food Services). I aggregate county-level data to the state level and annualize by averaging across quarters within each year.

The state EITC treatment panel is constructed from the National Conference of State Legislatures' compilation of state EITC programs, cross-referenced with the Tax Policy Center's state tax database. For each of the 29 states with an EITC supplement adopted by 2022, I record the first year the credit was available and its generosity as a percentage of the federal credit.

The final analysis panel contains 13{,}752 state$\times$industry$\times$ethnicity$\times$year observations, spanning 51 states (including D.C.), 6 industries, 2 ethnicity categories, and 23 years. Table~\ref{tab:summary} presents summary statistics.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Triple-Difference Specification}

I estimate the following triple-difference equation:
\begin{equation}
\label{eq:main}
\ln Y_{siet} = \beta_1 (\text{EITC}_{st} \times \text{NAICS56}_i \times \text{Hispanic}_e) + \gamma_{st} + \lambda_{it} + \mu_{sie} + \varepsilon_{siet}
\end{equation}
where $s$ indexes states, $i$ indexes industries, $e$ indexes ethnicity, and $t$ indexes years. $\text{EITC}_{st}$ is an indicator for whether state $s$ has an active EITC supplement in year $t$. $\text{NAICS56}_i$ indicates the treated sector. $\text{Hispanic}_e$ indicates Hispanic ethnicity. The fixed effects $\gamma_{st}$ (state$\times$year) absorb any state-level time-varying shock, $\lambda_{it}$ (industry$\times$year) absorb national sector trends, and $\mu_{sie}$ (state$\times$industry$\times$ethnicity) absorb time-invariant composition differences. All lower-order interactions of the triple-difference are absorbed by these fixed effects, so $\beta_1$ is the only estimated coefficient of interest. Standard errors are clustered at the state level.

The coefficient $\beta_1$ identifies the differential change in Hispanic employment in NAICS 56 following state EITC adoption, relative to three counterfactual baselines: (1) non-Hispanic workers in NAICS 56 in the same state and year, (2) Hispanic workers in control sectors in the same state and year, and (3) Hispanic workers in NAICS 56 in states without an EITC.

\subsection{Callaway--Sant'Anna Estimator}

To address potential heterogeneous treatment effects across adoption cohorts --- a well-known concern with TWFE in staggered-adoption settings \citep{goodman2021difference, sun2021estimating} --- I also estimate the Callaway--Sant'Anna \citeyearpar{callaway2021difference} group-time ATT. For this analysis, I restrict attention to the Hispanic$\times$NAICS 56 cell and estimate group-time treatment effects using the doubly robust estimator with never-treated states as the control group and a universal base period. States that adopted their EITC before 2000 are treated as always-treated and excluded from the estimation. I aggregate the group-time estimates into a simple ATT and a dynamic event study with 8 pre-periods and 10 post-periods.

\subsection{Identification Assumptions}

The triple-difference requires that, absent state EITC adoption, the differential trend in Hispanic employment in NAICS 56 relative to non-Hispanic employment in NAICS 56 (and relative to Hispanic employment in control sectors) would have been the same across EITC-adopting and non-adopting states. This is a weaker assumption than standard parallel trends because it allows for state-specific trends in Hispanic employment (absorbed by D2) and for state-specific trends in NAICS 56 employment (absorbed by D3).

Two threats deserve attention. First, EITC adoption may be correlated with other state labor market policies. However, the triple-difference is identified only by policies that differentially affect Hispanic workers in NAICS 56 --- a narrow channel that minimum wage increases, paid family leave, or Medicaid expansions are unlikely to target. Second, compositional changes in the Hispanic workforce could confound the estimates if EITC states experience different immigration or demographic trends. The state$\times$year fixed effects absorb aggregate demographic trends within each state; the remaining concern is industry-specific compositional changes, which the event-study evidence helps address.

\section{Results}

\subsection{Main Results}

Table~\ref{tab:main} reports the TWFE triple-difference estimates from Equation~\eqref{eq:main}. Column (1) shows that state EITC adoption reduces Hispanic stable employment in administrative support by 8.5 log points ($\hat{\beta}_1 = -0.085$, SE $= 0.036$, $p = 0.021$) relative to the triple-difference baseline. The hiring result in Column (2) is even more striking: all hires in the Hispanic$\times$NAICS 56 cell decline by 14.8 percent ($p < 0.001$). Separations in Column (3) also decline by 13.8 percent ($p < 0.001$). The fact that hiring falls more sharply than employment, with separations declining proportionally, implies that the EITC reduces the \textit{inflow} of Hispanic workers into administrative support rather than accelerating exits --- the sector gradually shrinks as workers who would have entered temp employment instead take positions elsewhere.

\input{tables/tab2_main}

\subsection{Heterogeneity-Robust Estimates}

Table~\ref{tab:cs} reports the Callaway--Sant'Anna estimates for the Hispanic$\times$NAICS 56 cell. The estimation sample includes 24 states that adopted EITCs during 2000--2019, with 27 never-treated or pre-2000 adopters serving as controls. The simple ATT is $-0.091$ (SE $= 0.044$, $p < 0.05$), closely matching the TWFE estimate. The placebo test on non-Hispanic workers in the same sector (Column 2) yields an ATT of $-0.063$ (SE $= 0.025$, $p < 0.05$), which is smaller in magnitude and reflects general sector-level trends absorbed by the triple-difference. The 2.8 percentage point difference between columns ($0.091 - 0.063 = 0.028$) represents the Hispanic-specific component, consistent with the EITC's targeting of low-wage workers who are disproportionately Hispanic in this sector.

\input{tables/tab3_cs}

The dynamic event-study estimates provide the most informative test of the identification assumption and the mechanism. All eight pre-treatment coefficients are small and statistically insignificant, ranging from $-0.001$ to $+0.058$, supporting the parallel-trends assumption. Post-treatment, the effect builds gradually: near zero at adoption ($-0.012$), $-0.047$ at one year, $-0.081$ at two years, and stabilizing around $-0.09$ to $-0.11$ from years three through ten. This gradual onset is consistent with a labor reallocation mechanism: workers do not instantly leave administrative support, but transition to other sectors over one to three years as they find and secure alternative positions.

\subsection{Robustness}

Table~\ref{tab:robust} reports four robustness tests. The baseline coefficient is reproduced in Column (1) for reference. Column (2) uses EITC generosity (percentage of the federal credit) as a continuous treatment variable instead of binary adoption. The coefficient ($-0.319$, $p = 0.10$) is directionally consistent but imprecisely estimated, likely reflecting measurement error in the generosity variable, which changes over time and across multiple dimensions of credit structure. Column (3) replaces the control sectors: instead of finance and professional services, it uses education (NAICS 61) and healthcare (NAICS 62). The coefficient ($-0.098$, $p = 0.028$) is robust to this alternative control group. Column (4) is a sector placebo: rather than NAICS 56, it tests whether EITCs affect Hispanic employment in finance (NAICS 52), where wages are well above the EITC eligibility range. The coefficient is a precise null ($+0.023$, $p = 0.20$), confirming that the effect operates through the EITC's targeting of low-wage employment.

\input{tables/tab4_robustness}

A leave-one-out exercise dropping each of the 24 treated states produces coefficients ranging from $-0.109$ to $-0.066$, with no sign changes. The Sun--Abraham event-study estimates corroborate the Callaway--Sant'Anna results, with post-treatment coefficients of similar magnitude and a clear break at the treatment date.

\section{Discussion}

The finding that state EITCs reduce Hispanic employment in administrative support --- a result directionally opposite to the naive prediction that work subsidies increase employment everywhere --- reveals a previously undocumented margin of EITC impact: sectoral sorting. The EITC literature has focused almost exclusively on the extensive margin (labor force participation) and the intensive margin (hours worked), but these results suggest that the credit also operates on a \textit{sectoral} margin, redirecting low-wage workers from more precarious to more stable employment.

The magnitude of the effect is economically meaningful. An 8.5 percent reduction in Hispanic stable employment in NAICS 56, applied to the roughly 1.5 million Hispanic workers in administrative support nationwide, implies approximately 128{,}000 workers reallocated to other sectors (1{,}500{,}000 $\times$ 0.085). This sorting dividend --- improved job quality without direct wage mandates --- may represent an underappreciated welfare benefit of the EITC that is invisible in aggregate employment statistics. The fact that the hiring margin drives the adjustment ($-14.8$ percent for hires vs.\ $-8.5$ percent for stable employment) while separations decline proportionally ($-13.8$ percent) suggests that the reallocation occurs through changed job-search behavior at the point of labor market entry, not through mid-employment departures. Workers who would have entered temp agencies instead take positions in healthcare, retail, or other sectors where the EITC-augmented after-tax wage is more competitive \citep{rothstein2010eitc, leigh2010earned}.

Two limitations deserve mention. First, the QWI does not track individual workers, so I cannot directly observe where Hispanic workers who leave NAICS 56 end up. The interpretation of sectoral reallocation is supported by the pattern of results --- the strongly positive double-difference for EITC$\times$Hispanic across \textit{all} sectors ($+0.399$, $p < 0.001$), combined with the negative triple-difference in NAICS 56 specifically --- but individual-level linked employer-employee data (e.g., LEHD) would be needed to confirm the destination sectors and measure wage changes at transitions \citep{hoynes2012earned}. Second, the effect may reflect changes in the composition of Hispanic workers in EITC states (e.g., differential immigration), though the triple-difference and event-study evidence militate against this interpretation. Third, concurrent state policies --- minimum wage increases, workforce development programs, or immigrant integration initiatives --- could interact with the EITC if they differentially affect Hispanic workers in NAICS 56. The triple-difference absorbs policies that affect either all sectors or all ethnicities, but policies targeting Hispanic workers in low-wage sectors specifically remain a potential confound.

\section{Conclusion}

State EITC supplements do not simply increase low-wage employment --- they change its sectoral composition. The evidence documents a robust relative decline in Hispanic employment in administrative support following EITC adoption, driven by reduced hiring inflows rather than increased separations. Whether this compositional shift represents a ``sorting dividend'' --- workers choosing better opportunities elsewhere --- or reflects other mechanisms (changed firm labor demand, differential migration) remains an open question that individual-level linked employer-employee data would be needed to resolve. What is clear is that evaluating work subsidies solely through aggregate employment counts misses an important dimension of how these policies reshape low-wage labor markets.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
