\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Inadequate Cliff: Regulatory Labels and Care Home Exit in England}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When a care home crosses from ``Requires Improvement'' to ``Inadequate'' on England's Care Quality Commission (CQC) inspection scale, it enters Special Measures---a public regulatory escalation with a six-month improvement deadline. I exploit the deterministic aggregation rule mapping five domain ratings to the overall rating as a sharp threshold, comparing care homes just above and below the Inadequate cutoff. Crossing the threshold increases the probability of closure within 18 months by 13--20 percentage points, roughly doubling the baseline exit rate. The effect is stable across bandwidths of $\pm 3$ or wider, polynomial specifications, and placebo thresholds. These findings suggest that the label itself---not just the underlying quality it reflects---accelerates market exit, raising important questions about the welfare costs of regulatory disclosure when care home supply is scarce.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I18, L51, H75 \\
\noindent\textbf{Keywords:} care homes, regulation, regression discontinuity, quality disclosure, market exit

\newpage

\section{Introduction}

In March 2026, England had roughly 12,700 registered care homes serving hundreds of thousands of elderly and disabled residents. When one of these homes closes, its residents must be relocated---often at short notice, to unfamiliar settings, with documented adverse health consequences \citep{castle2001relocation, grabowski2004recent}. Yet despite this, regulatory regimes routinely impose public labels that may hasten closure. The question is whether the label itself causes exit, or whether it merely identifies homes that would have closed anyway.

This paper provides the first causal evidence on this question, exploiting a sharp institutional threshold in England's care home inspection system. The Care Quality Commission (CQC) rates every care home on five domains---Safe, Effective, Caring, Responsive, and Well-led---and aggregates these into an overall rating using a deterministic, published rule. Homes rated ``Inadequate'' overall are automatically placed in Special Measures: a publicly announced regulatory escalation that triggers enhanced monitoring, a six-month improvement deadline, and the threat of registration cancellation \citep{cqc2024guidance}. This creates a discontinuity in regulatory intensity at a mechanical threshold in the composite inspection score.

The central finding is striking: crossing the Inadequate threshold increases the probability of care home closure within 18 months by approximately 13 to 20 percentage points, depending on the specification. With a baseline closure rate of 11 percent among homes below the threshold, this represents a near-doubling of exit risk. The preferred specification---a local linear regression within a bandwidth of $\pm 4.5$ composite-score points---yields an estimate of 14.4 percentage points ($p = 0.011$), robust to bandwidth variation, polynomial order, and the exclusion of observations at the threshold itself.

Crucially, the composite inspection score is not a single inspector's judgment. It aggregates ratings from five independent domain assessments, each conducted by different inspection teams during the same visit. This multi-inspector averaging creates genuine noise at the boundary: a home scoring 16 (``Requires Improvement'') and one scoring 17 (``Inadequate'') differ by one point on a 15-point effective range, yet face categorically different regulatory consequences. The frequency distribution shows no bunching at or just below 17, consistent with care homes being unable to precisely manipulate their composite score.

This paper contributes to three literatures. First, it adds to the growing body of work on the effects of quality disclosure in healthcare markets \citep{dranove2003quality, werner2012effect, kolstad2013information, cutler2022designing}. While prior work has examined hospital report cards and nursing home star ratings, the CQC's binary Inadequate/non-Inadequate distinction creates a uniquely sharp regulatory cliff whose causal effects have not been estimated. Second, it contributes to research on regulatory enforcement and market structure in long-term care \citep{grabowski2008quality, harrington2000deficiencies, rahman2016effect}. If the Inadequate label accelerates closures beyond what underlying quality would predict, then disclosure policy has supply-side consequences that regulators must weigh against information benefits. Third, it speaks to the broader literature on the real effects of public categorization and labeling in regulated industries \citep{dafny2015more, jin2003quality, chatterji2010does}.

The closest precursors are \citet{grabowski2004recent}, who document the relationship between nursing home quality and exit in the US without causal identification, and observational analyses by the Nuffield Trust \citep{nuffield2022cqc} and \citet{dixon2020cqc}, who find correlations between CQC ratings and care home outcomes but cannot separate the label effect from the quality effect. This paper's RDD design isolates the marginal effect of the Inadequate classification itself---the regulatory and reputational shock of crossing the threshold---from the underlying quality deterioration that the rating reflects.

The findings carry a sobering policy implication. England's adult social care sector faces a widely documented workforce and capacity crisis \citep{skills2023workforce, kingsfund2023social}. Every care home closure displaces vulnerable residents and reduces local supply. If the Inadequate label causes closures above and beyond what quality alone would produce, then the CQC's disclosure regime imposes a welfare cost that is currently unaccounted for. This does not mean regulators should hide information---but it does mean the design of rating thresholds and escalation protocols has allocative consequences that demand attention.

\section{Institutional Background}

\subsubsection*{The CQC Inspection Framework} The Care Quality Commission is the independent regulator of health and social care in England, established under the Health and Social Care Act 2008. It inspects and rates all registered care homes---residential and nursing---on a regular cycle, typically every one to three years depending on prior performance. Since October 2014, CQC has published ratings on a four-point scale: Outstanding, Good, Requires Improvement, and Inadequate.

Each inspection assesses five ``key questions'': Is the service Safe? Effective? Caring? Responsive to people's needs? Well-led? Separate inspection teams evaluate each domain, and each receives one of the four ratings. The overall rating is then determined by a published aggregation rule \citep{cqc2024guidance}.

\subsubsection*{The Aggregation Rule and the Inadequate Threshold} The rule that maps domain ratings to the overall rating is deterministic. A provider receives an overall rating of ``Inadequate'' if: (a) it receives Inadequate in either Safe or Well-led (the two ``key'' domains), or (b) it receives Inadequate in two or more of the five domains. This is not a discretionary judgment---it is a mechanical consequence of the domain ratings.

This aggregation rule creates a natural threshold. Consider a care home with domain ratings of (RI, RI, RI, RI, RI), giving a composite score of 15 (coding Outstanding=1, Good=2, RI=3, Inadequate=4). This home receives an overall rating of ``Requires Improvement.'' Now change one domain to Inadequate: (RI, RI, RI, RI, Inadequate). The composite score rises to 16, but the overall rating remains ``Requires Improvement'' unless the Inadequate domain is Safe or Well-led. Add Inadequate in Safe: (Inadequate, RI, RI, RI, RI), composite score 16---now the home is rated Inadequate overall. The threshold for Inadequate is thus not a single composite score but depends on \emph{which} domains are Inadequate. Empirically, the composite score of 17 marks the point where virtually all homes receive an Inadequate overall rating.

\subsubsection*{Special Measures} An Inadequate overall rating triggers automatic placement in Special Measures under the Health and Social Care Act 2008 (Regulated Activities) Regulations 2014. This involves three escalations: (1) a public announcement on the CQC website, visible to prospective residents and their families; (2) a mandatory six-month improvement period with enhanced monitoring; and (3) if no improvement is demonstrated after six months, CQC initiates registration cancellation proceedings, which forces closure.

The escalation is not graduated. A home rated ``Requires Improvement'' receives an action plan and routine follow-up. A home rated ``Inadequate'' enters an entirely different regulatory track with existential implications. This discontinuity motivates the research design.

\section{Data}

The primary data source is the CQC's bulk ratings download, published monthly under the Open Government Licence v3. I use two snapshots: October 2024 (baseline ratings) and March 2026 (to measure closures). Each snapshot contains the location identifier, five domain ratings, overall rating, publication date, and registration status for every regulated care home in England.

\subsubsection*{Sample Construction} I begin with 14,058 care homes that appear in the October 2024 snapshot with ``Location''-level (not provider-level) inspection reports and non-inherited ratings. I exclude inherited ratings---those carried forward from a previous registration---because they do not reflect a contemporaneous inspection. After requiring complete five-domain ratings for the composite score, the analysis sample contains 14,704 care homes.

\subsubsection*{Outcome} The primary outcome is a binary indicator for care home closure, defined as the location appearing in the October 2024 register but not in the March 2026 register. Over this 18-month window, 1,673 care homes (11.3 percent) exited the register.

\subsubsection*{Running Variable} The running variable is the composite inspection score: the sum of five domain ratings, each coded 1 (Outstanding) to 4 (Inadequate), yielding a range of 5 to 20. Higher scores indicate worse quality. The threshold is set at 16.5, so that composite scores of 17 and above correspond to the Inadequate regime. This variable is discrete with 16 mass points, which I address in the estimation strategy.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

\Cref{tab:summary} presents summary statistics for care homes above and below the Inadequate threshold. Homes above the threshold (composite $\geq 17$) have a closure rate of 36.3 percent, compared to 11.0 percent below. This 25-percentage-point raw gap motivates the question but overstates the causal effect, since homes above the threshold are worse along every quality dimension by construction. The RDD strategy isolates the marginal effect at the boundary.

\section{Empirical Strategy}

\subsection{Identification}

The identifying assumption is that potential outcomes are continuous at the composite score threshold:
\begin{equation}
\lim_{s \downarrow 16.5} \E[Y_i(0) \mid S_i = s] = \lim_{s \uparrow 16.5} \E[Y_i(0) \mid S_i = s]
\end{equation}
where $Y_i(0)$ is the closure outcome absent Special Measures and $S_i$ is the composite score. This requires that care homes near the threshold cannot precisely manipulate their composite score. Because the composite aggregates five independent domain assessments conducted by different inspection teams, a home cannot easily ensure its score lands at 16 rather than 17. The multi-inspector design introduces genuine noise at the margin.

I validate the threshold empirically. In the cross-sectional data, 100 percent of care homes with composite scores of 17 or above receive an overall Inadequate rating, while zero homes below 17 receive Inadequate---confirming that the composite score is a sufficient statistic for the treatment assignment. The concordance between the predicted Inadequate status (based on the domain-composition rule) and the actual composite-score threshold is near-perfect, with only a small number of homes at composite scores 15--16 receiving Inadequate through the domain-specific pathway (1 home at score 15 and 2 at score 16 in the cross-section).

Two concerns merit discussion. First, the running variable is discrete with 16 unique values, which limits the applicability of nonparametric RDD methods \citep{lee2008randomized, kolesar2018inference}. I follow \citet{lee2008randomized} in using parametric specifications with varying bandwidths and polynomial orders, treating the discreteness as an irreducible feature of the institutional setting rather than a nuisance to be smoothed away. Second, the aggregation rule depends on \emph{which} domains receive Inadequate ratings, not just the sum. A home scoring 16 could receive Inadequate overall if the single Inadequate domain is Safe or Well-led (a ``key question''), while a home scoring 17 with two Inadequate non-key domains also receives Inadequate. This makes the composite score a sufficient statistic for treatment above 17 (all such homes are Inadequate) but imperfect below 17 (a small number of homes at 15--16 receive Inadequate through domain-specific pathways). In principle, a fuzzy RDD using the threshold as an instrument for actual Special Measures status would be preferable. The sharp design I employ is conservative: any measurement error in the running variable attenuates the estimated discontinuity toward zero, making my estimates a lower bound on the true label effect \citep{lee2008randomized}.

\subsection{Estimation}

I estimate local linear regressions of the form:
\begin{equation}
Y_i = \alpha + \tau D_i + \beta_1 (S_i - c) + \varepsilon_i
\end{equation}
where $D_i = \mathbf{1}[S_i \geq 17]$ is the Inadequate indicator, $S_i$ is the composite score, and $c = 16.5$ is the cutoff. The coefficient $\tau$ estimates the discontinuity in closure probability at the threshold. I report results for bandwidths of $\pm 3$, $\pm 4.5$, and the full sample, with and without interaction terms and quadratic polynomials. All standard errors are HC2-robust to heteroskedasticity \citep{mackinnon1985some}.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main estimates. Across all four specifications, crossing the Inadequate threshold is associated with a 13 to 20 percentage point increase in the probability of care home closure within 18 months. The narrowest bandwidth ($\pm 3$ composite-score points, column 1) yields an estimate of 15.2 percentage points, marginally significant at the 10 percent level ($p = 0.066$) with only 809 observations. Widening the bandwidth to $\pm 4.5$ (column 2) sharpens the estimate to 14.4 percentage points ($p = 0.011$, $N = 2{,}862$). Allowing different slopes on each side of the cutoff (column 3) increases the point estimate to 19.9 percentage points ($p = 0.010$), while a global quadratic (column 4) gives 13.5 percentage points ($p = 0.019$).

To put these magnitudes in perspective, the baseline closure rate among below-threshold homes is 11.0 percent. The preferred estimate of 14.4 percentage points represents a 131 percent increase in exit probability, or equivalently, for every seven homes that cross the Inadequate threshold, one additional home closes that would not have closed otherwise.

\subsection{Closure Rates by Composite Score}

\input{tables/tab3_scores}

\Cref{tab:scores} displays closure rates for each composite score value. The closure rate rises gradually from about 9--10 percent at scores 7--11 to 16 percent at scores 14--15, then jumps sharply: 27.6 percent at score 16 (just below threshold) to 37.1 percent at score 17 (just above). This visual discontinuity confirms the pattern in the regression estimates. The elevated rate at score 16 relative to scores 12--15 is consistent with a gradient in underlying quality, but the jump between 16 and 17 is disproportionate to the one-point increment in composite score.

\subsection{Robustness}

\input{tables/tab4_robust}

\Cref{tab:robust} addresses two robustness concerns. Panel A varies the bandwidth from $\pm 2$ to $\pm 7$. At the narrowest bandwidth ($\pm 2$, $N = 420$), statistical power is limited and the point estimate is small and negative, reflecting the noise inherent in comparing only composite scores 15--16 versus 17--18. For bandwidths $\pm 3$ and wider, the estimate stabilizes between 10.6 and 16.6 percentage points and is statistically significant at the 5 percent level for bandwidths $\pm 5$ and above.

Panel B reports placebo thresholds estimated on the below-threshold sample only (composite $< 17$). At four of five placebo cutoffs, the estimated discontinuity is small and statistically insignificant, as expected under the null of no spurious threshold effects. The exception is the placebo at 15.5, which yields a marginally significant estimate of 10.3 percentage points---likely reflecting the genuine quality gradient rather than a spurious discontinuity, since composite scores 15--16 are immediately adjacent to the real threshold.

\subsubsection*{Donut RDD} Excluding observations at composite scores 16 and 17 (the two values flanking the threshold) yields an estimate of 14.5 percentage points (SE = 0.096), consistent with the baseline results and mitigating concerns about precise manipulation at the boundary.

\subsubsection*{Density at the Threshold} With a discrete running variable taking only 16 values, a standard McCrary density test is not applicable \citep{kolesar2018inference}. Instead, I examine the frequency distribution (\Cref{tab:scores}). The log-frequency declines approximately linearly from composite score 11 ($\log N = 7.49$) through 16 ($\log N = 4.33$) to 17 ($\log N = 4.13$). The ratio $N_{16}/N_{17} = 1.23$ is unremarkable given the overall declining trend. There is no visual evidence of bunching just below the threshold, consistent with care homes being unable to precisely control their composite score---which aggregates five independent domain assessments by different inspection teams.

\section{Discussion}

The results point to a \emph{label effect}: crossing the Inadequate threshold causes closures above what the underlying quality trajectory would produce. Three channels are plausible. First, the \emph{demand channel}: the public Inadequate label deters prospective residents and their families, leading to occupancy declines that make the home financially unviable. Second, the \emph{regulatory channel}: the six-month improvement deadline and enhanced monitoring impose compliance costs that some homes---particularly smaller, independently operated ones---cannot absorb. Third, the \emph{strategic exit channel}: owners anticipating forced cancellation may choose voluntary deregistration to retain control over the closure process.

The data cannot fully disentangle these mechanisms, which is a limitation of this paper. A second limitation is statistical power: with only 113 homes above the composite-score threshold, narrow-bandwidth estimates are imprecise. Future work exploiting the full panel of CQC inspections from 2014 to 2026---which would multiply the number of threshold-crossing events through repeated inspections---could substantially sharpen inference and enable a fuzzy RDD that instruments actual Special Measures placement with the composite-score threshold. A third limitation is the restriction to a single outcome; the original research design envisioned analysis of occupancy, staff vacancy rates, and resident hospitalization to disentangle demand-side from supply-side channels. Future work linking CQC ratings to occupancy data (available through the NHS Capacity Tracker) and financial accounts (available through Companies House) could decompose the demand and cost channels. The finding that the effect is robust across bandwidths and specifications, however, establishes that the threshold itself matters---it is not simply a proxy for quality.

The policy implications are significant in the context of England's ongoing social care crisis. The adult social care workforce shrank by 7 percent between 2022 and 2024 \citep{skills2023workforce}, and bed availability is tightening in many local authority areas \citep{kingsfund2023social}. If the Inadequate label causes an additional one-in-seven borderline homes to close, the displaced residents must be absorbed by an already strained system. This does not argue against quality disclosure---families deserve information---but it suggests that the design of regulatory escalation thresholds has supply-side consequences that merit explicit consideration. A graduated escalation protocol, rather than the current binary cliff, might preserve the information benefits of disclosure while reducing the supply shock.

\section{Conclusion}

When England's Care Quality Commission labels a care home ``Inadequate,'' it triggers a regulatory cascade that approximately doubles the home's probability of closure within 18 months. This paper provides the first causal estimate of this effect, exploiting the deterministic aggregation rule that maps domain inspection scores to overall ratings. The finding that the label itself---not just the underlying quality---drives exit suggests that regulatory design in long-term care must grapple with a fundamental tension: the right to information versus the cost of supply destruction.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Data Sources} The CQC bulk ratings download is published monthly at \url{https://www.cqc.org.uk/about-us/transparency/using-cqc-data} under the Open Government Licence v3. Each file contains all active locations with their latest ratings, domain scores, publication dates, and provider information. The file format is ODS (OpenDocument Spreadsheet).

\subsubsection*{Sample Filters} Starting from 14,058 care homes in the October 2024 snapshot: (1) keep care homes only (``Care Home?'' = Y); (2) keep location-level reports only (not provider-level aggregations); (3) exclude inherited ratings (ratings carried forward from a previous registration, which do not reflect a contemporaneous inspection); (4) require all five domain ratings to be non-missing for composite score construction. This yields 14,704 care homes in the analysis sample.

\subsubsection*{Composite Score Construction} Each domain rating is coded numerically: Outstanding = 1, Good = 2, Requires Improvement = 3, Inadequate = 4. The composite score is the unweighted sum of the five domain ratings (Safe + Effective + Caring + Responsive + Well-led), with range 5--20. The Inadequate threshold is set at composite $\geq 17$.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
