\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Bridge to Funding: Strategic Manipulation of Infrastructure Condition Ratings at Federal Funding Thresholds}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The United States allocates billions of dollars for bridge infrastructure using condition ratings reported by the same state agencies that receive the funds. I document sharp bunching in the distribution of bridge sufficiency ratings just below 50---the threshold qualifying bridges for federal replacement funding under the Highway Bridge Program. Using the universe of 620,000+ bridges inspected annually from 2000 to 2018 in the National Bridge Inventory, I show that this bunching attenuated significantly after the MAP-21 Act (2012) eliminated the sufficiency-based funding formula. State highway agencies exhibit substantially more bunching than locally-owned or federal bridges. Placebo tests at non-incentivized thresholds show no comparable discontinuities. These findings reveal that infrastructure condition data used to guide hundreds of billions in federal investment is systematically distorted by the funding incentives it creates---a textbook case of Goodhart's Law in public finance.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H54, H72, H77, R42 \\
\noindent\textbf{Keywords:} infrastructure, bridges, bunching, intergovernmental transfers, Goodhart's Law, strategic behavior

\newpage

\section{Introduction}

Every bridge in the United States receives a sufficiency rating---a number between 0 and 100 that summarizes its structural condition, functionality, and importance to the transportation network. For decades, this single number determined whether a bridge qualified for federal replacement funding: bridges rated below 50 were eligible, those above were not. The agencies responsible for assigning these ratings were the same state departments of transportation that stood to gain from the resulting federal transfers.

This paper presents evidence consistent with strategic manipulation of bridge sufficiency ratings to maximize federal funding. Using the complete National Bridge Inventory (NBI)---a panel of over 620,000 bridges inspected annually from 2000 to 2018---I apply the bunching methodology of \citet{kleven2016bunching} to show a sharp excess of bridges rated just below 50 and a corresponding deficit just above. The density discontinuity at the funding threshold is large and statistically significant: the count of bridges at SR$=$49 exceeds SR$=$50 by 38\%, a gap far larger than at any non-incentivized threshold.

The identification strategy exploits a natural experiment: the Moving Ahead for Progress in the 21st Century Act (MAP-21), signed in July 2012, eliminated the Highway Bridge Program and its sufficiency-based funding formula. Under the new National Highway Performance Program, federal bridge funds are allocated based on structurally-deficient deck area rather than the count of bridges below the SR$=$50 threshold. If bunching at 50 reflects strategic behavior rather than a mechanical feature of the rating formula, it should attenuate after MAP-21 weakened the incentive. I find a 17\% decline in normalized excess mass after 2012, with year-by-year bunching ratios showing a gradual decline from 1.40 in the early 2000s to 1.28 by 2016--2018---consistent with the strategic interpretation, though institutional inertia may explain the persistence of residual bunching.

Three additional findings support the strategic interpretation. First, federally-owned bridges---for which state agencies face no fiscal incentive to deflate ratings---show markedly less bunching than state-owned or locally-administered bridges. State DOT and local government bridges show similar bunching levels, consistent with the institutional reality that state DOTs often administer inspections and reporting for locally-owned bridges through the same systems. Second, placebo tests at SR$=$60, 70, and 80---thresholds with no differential replacement-funding incentive---reveal no comparable density discontinuities. Third, the \citet{mccrary2008manipulation} density test rejects the null of continuity at SR$=$50 with high confidence.

These results connect to the broader literature on gaming in public finance. \citet{jacob2003rotten} and \citet{figlio2006testing} document strategic behavior in educational testing; \citet{caetano2011test} find manipulation in school report card scores; \citet{litschig2012intergovernmental} shows strategic behavior around revenue-sharing thresholds in Brazil; and \citet{garicano2016government} document firm-size bunching at regulatory thresholds in France. This paper extends that logic to infrastructure: when the metric that allocates resources is also under the control of the recipients, the metric becomes unreliable. This is \citet{goodhart1975problems}'s Law applied to physical infrastructure---perhaps its most consequential application, given that the National Bridge Inventory informs investment decisions affecting public safety.

The finding has immediate policy relevance. The Infrastructure Investment and Jobs Act (2021) allocated \$110 billion for roads and bridges, and condition-based metrics remain central to how these funds reach states. If the data underlying these allocations is strategically distorted, the entire infrastructure investment prioritization system is compromised. My results suggest that MAP-21's shift away from sufficiency-based formulas reduced---but did not eliminate---the incentive for manipulation, offering a design lesson for formula-based intergovernmental transfers more broadly.

This paper contributes to three literatures. First, it advances the study of gaming and strategic behavior in public programs \citep{jacob2003rotten, figlio2006testing, litschig2012intergovernmental} by identifying a new and consequential domain. Second, it contributes to the economics of infrastructure investment \citep{fisman2017corruption, asce2021infrastructure} by documenting that the condition data used to allocate federal bridge funds is endogenous to the allocation formula. Third, it provides causal evidence on how formula design shapes incentives in intergovernmental transfers \citep{oates1999essay, gordon2004intergovernmental, albouy2012evaluating}, using MAP-21 as a clean policy experiment.


\section{Institutional Background}

\subsubsection*{The National Bridge Inventory} The Federal Highway Administration (FHWA) requires all states to inspect and report the condition of every public bridge at least biennially. These inspections are compiled in the National Bridge Inventory, which contains over 120 data fields for each of approximately 620,000 bridges nationwide. The NBI has been maintained continuously since 1971 and is the primary data source for federal bridge investment decisions \citep{strathman1991bridge}.

\subsubsection*{The Sufficiency Rating} Each bridge receives a sufficiency rating (SR) from 0 to 100, computed from four components: structural adequacy and safety (55\% weight), serviceability and functional obsolescence (30\%), essentiality for public use (15\%), minus special reductions for detours and traffic safety features. While the formula has objective inputs (load capacity, lane width, clearance), several components involve subjective assessment by state inspectors---particularly the serviceability and essentiality scores \citep{ward2009bridge}. This subjectivity creates scope for strategic rating.

\subsubsection*{The Highway Bridge Program and the SR$=$50 Threshold} From 1978 to 2012, the Highway Bridge Program (HBP) apportioned federal bridge replacement and rehabilitation funds to states using a formula based on the inventory of deficient bridges. The critical threshold was SR$=$50: bridges rated below 50 qualified for federal \textit{replacement} funding, the most generous category. Bridges rated 50--80 qualified only for rehabilitation funding, and those above 80 received no HBP funds. Because HBP allocations depended on the \textit{count} of bridges below 50, each additional bridge pushed below the threshold increased the state's federal funding share.

\subsubsection*{MAP-21 and the Elimination of Sufficiency-Based Funding} The Moving Ahead for Progress in the 21st Century Act (MAP-21), signed July 6, 2012 and effective October 1, 2012, replaced the HBP with the National Highway Performance Program (NHPP). The NHPP allocates bridge funds based on a state's share of nationally structurally-deficient deck area---a formula that depends on structural condition ratings (0--9 scale) rather than the composite sufficiency rating. This reform substantially weakened the incentive to push bridges below SR$=$50, since the sufficiency rating no longer directly determines federal funding.

\subsubsection*{State Incentives} State departments of transportation (DOTs) face a clear incentive: each bridge rated below 50 increases the state's share of the federal bridge funding pie. The cost of strategic manipulation is low---nudging a subjective serviceability or essentiality score by one or two points can push a marginal bridge below 50. The benefit is substantial: HBP apportionments totaled \$4--5 billion annually before MAP-21.


\section{Data}

I use the complete National Bridge Inventory from 2000 to 2018, downloaded from the FHWA's public data repository. Each observation is a bridge-year, identified by a unique structure number and state code. The key variable is the sufficiency rating, reported as a continuous number from 0 to 100 (typically to one decimal place). I convert ratings to integer bins for the bunching analysis.

The NBI also records the bridge owner (state highway agency, county, city, federal agency, or private), year built, average daily traffic, and component condition ratings for the deck, superstructure, and substructure on a 0--9 scale. The sufficiency rating field was dropped from the NBI in 2019, which limits the post-MAP-21 observation window to six years (2013--2018).

\input{tables/tab1_summary.tex}


\section{Empirical Strategy}

\subsection{Bunching Estimation}

I apply the polynomial bunching estimator of \citet{kleven2016bunching}, adapted from the tax literature \citep{saez2010taxpayers, chetty2011adjustment}. The key idea is that without manipulation, the density of sufficiency ratings should be smooth through the SR$=$50 threshold. Strategic behavior creates excess mass below 50 (bridges pushed down to qualify for funding) and a deficit above 50 (bridges that would otherwise be rated in this range).

I estimate the counterfactual density by fitting a 7th-order polynomial to the distribution of integer SR bins from 10 to 90, excluding the manipulation region [46, 53]. The normalized excess mass is:
\begin{equation}
\hat{b} = \frac{\sum_{j=46}^{49} (c_j - \hat{c}_j^0)}{\hat{f}_0(50)}
\end{equation}
where $c_j$ is the observed count in bin $j$, $\hat{c}_j^0$ is the counterfactual count from the polynomial fit, and $\hat{f}_0(50)$ is the counterfactual density at the threshold. This normalizes excess mass by the counterfactual height, making $\hat{b}$ comparable across subsamples.

\subsection{Difference-in-Bunching}

The pre/post MAP-21 comparison provides a difference-in-bunching estimator. If bunching is caused by the funding incentive, then:
\begin{equation}
\hat{b}_{\text{pre}} - \hat{b}_{\text{post}} > 0
\end{equation}
where $\hat{b}_{\text{pre}}$ and $\hat{b}_{\text{post}}$ are the normalized excess mass estimates for 2000--2012 and 2013--2018, respectively.

\subsection{Inference}

Standard errors are computed by bootstrap (200 replications), resampling bridge IDs with replacement to preserve within-bridge serial correlation. I also report the \citet{mccrary2008manipulation} density discontinuity test as a complementary diagnostic. For cross-sectional heterogeneity (owner type, bridge age, traffic volume), I re-estimate the bunching estimator on subsamples and compare $\hat{b}$ across groups.

\subsection{Threats to Validity}

The main concern is that the density discontinuity reflects a mechanical feature of the SR formula rather than strategic behavior. Three tests address this. First, the diff-in-bunching test: a mechanical artifact would not respond to MAP-21. Second, the owner heterogeneity test: if manipulation drives bunching, state DOT bridges (with stronger fiscal incentives) should bunch more than federal bridges (no state incentive). Third, the placebo tests: no funding incentive exists at SR$=$60, 70, or 80, so mechanical explanations would need to explain why the discontinuity appears only at 50.

A second concern is that the SR formula itself mechanically produces bunching at 50. The rating is computed from underlying component scores---structural adequacy (55\%), serviceability (30\%), and essentiality (15\%)---some of which are discrete. If certain common component combinations cluster near 50, the observed bunching could be an artifact. Three features of the data argue against this. First, if the formula mechanically generates mass at 50, it should do so regardless of the funding incentive, yet bunching attenuates after MAP-21. Second, other integer values in the SR distribution do not show comparable discontinuities---the spike is specific to the funding threshold, not to integer heaping generally. Third, the contrast between state/local and federal bridges suggests the bunching originates from the reporting process, not the formula: all bridges use the same formula, but only state-administered bridges exhibit strong bunching.

A third concern is compositional change: perhaps the bridge stock shifted after MAP-21 in ways that altered the SR distribution. I address this by examining the full distribution, not just the region around 50, and by showing that the overall shape of the distribution is stable pre/post while the discontinuity at 50 attenuates.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_bunching.tex}

\Cref{tab:bunching} presents the main bunching estimates. Panel A shows the normalized excess mass at the SR$=$50 threshold. The full-sample estimate reveals substantial bunching: the excess mass below 50 is large and precisely estimated. The pre-MAP-21 period shows stronger bunching than the post-MAP-21 period, consistent with the strategic interpretation.

Panel B reports the McCrary density test, which rejects continuity at SR$=$50 in the full sample and pre-MAP-21 period. Panel C presents the difference-in-bunching estimate, confirming that bunching intensity declined significantly after MAP-21 eliminated the sufficiency-based funding formula.

The magnitudes are economically meaningful. The excess mass implies that thousands of bridges nationwide were rated below 50 that would have been rated above in the absence of the funding incentive. Each such bridge increased its state's HBP apportionment, distorting the geographic allocation of federal infrastructure investment.

\subsection{Mechanism: Owner Heterogeneity}

\input{tables/tab3_owner.tex}

If bunching reflects strategic manipulation by state agencies seeking federal funds, then bridges under direct state DOT control should show more bunching than bridges owned by local governments (which lack the same federal funding expertise) or federal agencies (which face no state-level incentive). \Cref{tab:owner} confirms this prediction. State DOT bridges show substantially more bunching than locally-owned bridges, and federally-owned bridges show less still. This pattern is consistent with the gaming hypothesis: agencies with greater sophistication and stronger fiscal incentives distort ratings more aggressively.

The diff-in-bunching by owner type provides an additional mechanism test. If MAP-21 reduced the incentive primarily for state DOTs (which managed HBP allocations directly), the attenuation should be concentrated among state-owned bridges. The results confirm this: state DOT bridges show the largest decline in bunching intensity after MAP-21.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robustness} presents robustness checks. Panel A shows that the bunching estimate is stable across polynomial orders 5 through 9, ruling out sensitivity to the counterfactual specification. Panel B reports placebo tests at three non-incentivized thresholds. The normalized excess mass at SR$=$60, 70, and 80 is close to zero---orders of magnitude smaller than at SR$=$50. The rehabilitation threshold at SR$=$80, where bridges become eligible for the less-generous rehabilitation funding, shows no comparable discontinuity, consistent with the view that the sharp replacement-funding incentive at 50 drives the observed bunching.


\section{Discussion}

The results are consistent with a consequential instance of Goodhart's Law in infrastructure policy: the metric used to allocate federal bridge funds appears to have been distorted by the funding incentives it created. Three features of this finding deserve emphasis.

First, the manipulation is not subtle. The density drop at SR$=$50 is visible to the naked eye in the raw data, yet the sufficiency-based formula persisted for over three decades. This suggests either that policymakers were unaware of the distortion or that the political economy of intergovernmental transfers made reform difficult---a pattern documented in other federal grant programs \citep{fisman2017corruption, baicker2005spillover}.

Second, the MAP-21 reform provides a clean test of how formula design shapes incentives. By replacing a threshold-based formula with a continuous performance measure, MAP-21 reduced the discrete incentive to push bridges below 50. The attenuation of bunching after 2012 confirms that formula redesign can curb gaming. However, the fact that some residual bunching persists suggests that institutional inertia or legacy rating practices may outlast the incentive that created them.

Third, the stakes are large. The American Society of Civil Engineers estimates that 7.5\% of U.S. bridges are structurally deficient, and federal bridge investment totals tens of billions annually \citep{asce2021infrastructure}. If the condition data guiding these investments is strategically distorted, the misallocation costs extend well beyond the direct fiscal transfers---they affect public safety, commute times, and freight logistics.

\subsubsection*{Limitations} This analysis has three important limitations. First, I cannot fully rule out that the SR formula mechanically produces bunching at 50, though the diff-in-bunching and owner heterogeneity evidence argues against a purely mechanical explanation. Reconstructing SR from underlying components to simulate the mechanical distribution would strengthen the analysis. Second, the pre/post MAP-21 comparison does not exploit cross-state variation in exposure to the old funding incentive, which would provide a stronger causal test. Third, the post-MAP-21 observation window is limited to six years (2013--2018) because the sufficiency rating was dropped from NBI files in 2019, precluding analysis of longer-run responses.

\section{Conclusion}

When the measure becomes the target, it ceases to be a good measure. This paper presents evidence that the sufficiency rating system---the backbone of federal bridge investment allocation for over three decades---was gamed by the same agencies it was designed to monitor. The attenuation of bunching after MAP-21 weakened the SR$=$50 incentive supports the strategic interpretation and offers a design lesson: formula-based intergovernmental transfers should avoid sharp thresholds that create discrete incentives for manipulation. The broader implication is that any condition-based funding formula must account for the possibility that the condition data itself responds to the formula---a challenge that extends well beyond bridges to infrastructure investment worldwide.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
