\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{tabularray}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Does Financial Parity Follow Legal Parity?\\
Gender Quotas and Campaign Finance in Mexico}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Mexico's 2019 ``Parity in Everything'' constitutional mandate required political parties to nominate equal numbers of female and male candidates for all elected offices, including mayor. Did compliance extend to campaign resources? Using candidate-level fiscal disclosures from Mexico's National Electoral Institute for 20,226 mayoral candidates in 2018 and 2021, we estimate a triple-differences specification exploiting within-candidate variation across income types: party headquarters transfers (institutionally controlled) versus sympathizer donations (market-determined). Women received 47 cents per dollar in party transfers in 2018, rising to 60 cents by 2021. The triple-differences estimate of 0.125 log points (SE\,=\,0.275) is positive but indistinguishable from zero. The party funding gap narrowed in roughly half of Mexican states and widened in the other half. Legal parity opened the candidacy gate; the funding gate has only partially followed.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J16, D72, O54 \\
\noindent\textbf{Keywords:} gender quotas, campaign finance, political parity, Mexico, mayoral elections

\newpage

\section{Introduction}

In the 2018 Mexican local elections, a woman running for mayor received on average 51,789 pesos from her party's headquarters---less than half the 110,381 pesos directed to her male counterpart. Three years later, after Mexico's constitutional ``Parity in Everything'' reform required parties to nominate equal numbers of women and men for all elected positions, the female figure rose to 100,877 pesos---still only 60 percent of men's 167,448 pesos. Legal parity arrived. Financial parity did not.

This paper asks whether mandated nomination parity translates into financial parity in campaign resource allocation. The question is not merely about money. Campaign spending is among the most robust predictors of electoral success for challengers and first-time candidates \citep{fox2011}. Parties that comply with gender quota mandates at the nomination stage while systematically underfunding female nominees preserve a financial gatekeeping role---one that is invisible in standard representation statistics but consequential for electoral outcomes. The funding gate is the mechanism by which formal parity can coexist with informal exclusion.

We study Mexico's June 2019 constitutional reform, which amended Article 41 to require \textit{paridad en todo} (parity in everything): equal female and male nominations across all elected offices simultaneously. This followed the 2014 reform that had established parity only through alternating-gender proportional representation lists (``vertical parity''). The 2019 reform added ``horizontal parity''---equal female/male nominations for single-member mayoralty races---effective for the June 2021 local elections. Mexico's National Electoral Institute (INE) simultaneously required parties to allocate at least 40 percent of public campaign funds to women, though this mandate covered public matching funds only, not party-controlled transfers.

Our approach compares trends in two types of campaign income. Party headquarters transfers (\textit{transferencias de concentradoras}) are set by the party apparatus and represent institutional commitment. Sympathizer donations (\textit{aportaciones de simpatizantes}) reflect the candidate's personal networks and perceived market viability. We use the descriptive change in party transfers relative to sympathizer donations as an exploratory DDD comparison --- not a fully clean causal design. The limitation is that the mandate itself may have affected both channels, as we document. Nonetheless, the comparison is informative: if the mandate narrowed the party-transfer gap more than the sympathizer gap, it is suggestive (though not conclusive) that institutional pressure played a role.\footnote{An ideal design would exploit staggered state-level adoption of horizontal parity across elections from 2015 to 2021. This approach was infeasible because the INE's gender variable (\texttt{SEXO}) was not available in pre-2018 candidate records, limiting us to at most one pre-period --- below the threshold required for credible staggered DiD inference.} The triple-differences estimator compares the change in the gender gap for party transfers versus sympathizer donations, between 2018 and 2021:

\begin{equation}
\hat{\beta}_7 = \bigl[\Delta_{\text{party gap}} - \Delta_{\text{simpatiz gap}}\bigr]_{\text{post-2019}}
\label{eq:ddd_intuition}
\end{equation}

A positive $\hat{\beta}_7$ would indicate that parties redirected transfers toward women beyond any general trend in female candidates' fundraising ability.

Our main finding is a null. The triple-differences estimate $\hat{\beta}_7 = 0.125$ (SE\,=\,0.275, $p = 0.65$) is positive but statistically indistinguishable from zero. The party transfer gender gap narrowed from 0.757 to 0.507 log points---an improvement of 0.250---but the sympathizer gap also narrowed, from 1.006 to 0.851 log points (a change of 0.155). The differential improvement is 0.095 log points descriptively, and 0.125 in the regression controlling for party$\times$state fixed effects, but neither is significant. State-level analysis reveals sharp heterogeneity: the party gap narrowed substantially in Guerrero, Colima, and Nuevo Le\'on, while widening in Chiapas, Michoac\'an, and Quintana Roo. The variance across states explains the imprecision of the national estimate.

An auxiliary finding is more striking. In the sympathizer DiD---our placebo---the female$\times$post coefficient is $-0.372^{*}$ (SE\,=\,0.195), indicating that within party$\times$state cells, women's private donor support declined relative to men post-mandate. One interpretation is that horizontal parity pushed parties to nominate women in municipalities where they lacked pre-existing support networks, suppressing sympathizer donations even as institutional transfers increased modestly.

We contribute to three bodies of work. First, we advance the literature on gender quotas and political representation \citep{chattopadhyay2004, pande2003, besley2017, casas2017}. This literature asks whether mandated female representation produces policy shifts aligned with women's preferences \citep{clayton2023}. We study an upstream precondition: financial resources that enable the representation to function. \citet{folke2016} document a ``glass ceiling'' in political career advancement; we document a ``funding gate'' at the candidacy stage. Second, we connect to the emerging literature on campaign finance and gender \citep{buckley2023, sudulich2025, rydzewska2026}. Existing studies document funding gaps under voluntary and mandatory quotas in Ireland and Belgium; our contribution is a cross-year design exploiting a sharp mandatory quota covering all mayoral candidates nationally. Third, we extend the Mexico-specific parity literature \citep{piscopo2017, piscopo2023, baldez2004, htun2002, jones2009} from representation outcomes to financial behavior, using the INE's candidate-level disclosure data---first use of these data in published economics research.

A well-executed null result of this kind is genuinely informative. It tells us that parity mandates, at least in the short run, produce candidacy equity---they fill the ballot---without producing financial equity. Whether parties will converge to financial parity in subsequent election cycles, or whether the funding gate is a stable equilibrium, is a question for future research \citep{krook2009}.

The paper proceeds as follows. Section~\ref{sec:background} describes Mexico's parity reform and the INE fiscal disclosure system. Section~\ref{sec:data} presents the data and descriptive statistics. Section~\ref{sec:strategy} describes the triple-differences identification strategy. Section~\ref{sec:results} presents the regression results and state-level heterogeneity. Section~\ref{sec:conclusion} concludes.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{Mexico's Two-Stage Parity Reform.}
Mexico's constitutional gender parity architecture developed in two stages. The 2014 constitutional reform (Decreto DOF 10/02/2014, amending Articles 41 and 116) established gender parity through electoral list design: all political parties were required to ensure equal representation in their candidate nominations, implemented primarily through \emph{vertical parity}---alternating male and female names on proportional representation lists. This reform applied nationally and was enforced for the first time in the 2015 federal and local elections. Its practical impact on executive positions was limited: parties complied by populating gender-balanced lists while concentrating male nominees in single-member mayoralty (\textit{presidente municipal}) races, which are winner-take-all.

The second stage was the constitutional reform of June 6, 2019, which amended Article 41 to require \emph{paridad en todo}. The reform extended parity to all elected bodies and all elected positions, including municipal president. Parties were required to nominate roughly equal numbers of female and male candidates across their competed municipalities---\emph{horizontal parity}---with INE enforcement through candidate registration review. The 2021 local elections were the first cycle fully subject to this mandate.

This two-stage structure gives us a natural comparison: the 2018 local elections were conducted under vertical parity only (no horizontal mandate for alcalde), while the 2021 elections required horizontal parity. Our before/after comparison within mayoral candidates exploits this sharp institutional break.

\subsubsection*{The INE Fiscal Disclosure System.}
Mexico's National Electoral Institute administers a mandatory campaign finance disclosure system (\textit{fiscalizaci\'on}) requiring all candidates to file itemized income and expenditure reports. Disclosure covers the full campaign period and distinguishes income by source. The categories relevant to this paper are:

\begin{itemize}
  \item \textit{Transferencias de concentradoras}: funds transferred from the party's central treasury to the candidate. These are set by party leadership and constitute the party's institutional commitment to the candidate. In 2018, this category is captured as the sum of \textit{Transferencia de Recursos Federales} (federal resources) and \textit{Transferencia de Recursos Locales} (local party resources).
  \item \textit{Aportaciones de simpatizantes}: voluntary donations from individual supporters. These depend on the candidate's personal networks, local profile, and perceived electoral viability. They are market-determined rather than institutionally assigned.
  \item \textit{Financiamiento p\'ublico}: public matching funds provided by the INE based on party vote share. The 2021 regulations required parties to direct at least 40 percent of this category to female candidates.
\end{itemize}

The distinction between party transfers and sympathizer donations is central to our identification. Party transfers measure the party apparatus's willingness to financially back its female nominees; sympathizer donations measure community and donor confidence in those nominees. The mandate's institutional pressure bore directly on the former.

\subsubsection*{The 2021 Campaign Context.}
The June 2021 local elections coincided with the COVID-19 pandemic, which depressed overall campaign activity and shifted spending toward digital outreach. Total candidate income was lower in real terms in 2021 than 2018 for male candidates on average; for female candidates, party transfers rose substantially in nominal terms. The elections also took place in the context of MORENA's electoral dominance: having swept the 2018 presidential and 2021 midterm elections, MORENA expanded its municipal presence, which may have altered the composition of candidates across states. We absorb this compositional change through party$\times$state fixed effects.

\section{Data}
\label{sec:data}

\subsubsection*{INE Fiscalizaci\'on CSVs.}
Our data come from the publicly available CSV files on the INE fiscalization portal for local elections in 2018 and 2021. The 2018 file (\textit{CL2-Anexo-Ingreso-por-Rubro-Campa\~na-LOCAL}) covers 25 states and 15,148 candidate records. The 2021 file (\textit{PELO\_20-21\_CAMPA\~NA\_Y\_RUBRO}) covers 26 states and 18,800 records. After restricting to municipal president candidates (\textit{PRESIDENTE MUNICIPAL} or \textit{ALCALDE}) and dropping records missing the gender variable (\texttt{SEXO}), we obtain 20,226 mayoral candidates: 9,761 in 2018 and 10,465 in 2021.

The key gender variable records the candidate's registered sex. Female candidates (\texttt{SEXO}\,=\,M for Mujer) comprise 47.3 percent of the 2018 sample (4,629 of 9,761) and 49.4 percent of the 2021 sample (5,167 of 10,465). The near-50/50 split in 2021 reflects the mandate's effectiveness at the nomination stage.

\subsubsection*{Variable Construction.}
We construct three campaign income variables from the disclosure data: (i) party headquarters transfers, (ii) sympathizer donations, and (iii) candidate self-financing. Because many candidates receive zero in a given category, we use the $\ln(x + 1)$ transformation throughout. All variables are measured in nominal pesos (MXN). For the DDD estimation, we reshape the data to long format: each candidate appears once per income source, yielding 40,452 observations (20,226 candidates $\times$ 2 sources) in the main party-versus-sympathizer specification. We additionally define a post-2019 indicator (\texttt{post}\,=\,1 for 2021), a party-source indicator (\texttt{is\_party\_source}\,=\,1 for party transfers), and party$\times$state fixed effects (party abbreviation crossed with state name) as the unit of absorption.

Party headquarters transfers are zero for a substantial share of candidates: 50.2 percent of women and 41.1 percent of men received no party transfer in 2018. By 2021 these rates fell to 37.8 and 33.6 percent respectively, reflecting a general expansion in institutional support. The IHS robustness check confirms results are not sensitive to the log-plus-one transformation.

\subsection{Summary Statistics}

Table~\ref{tab:tab1_descriptive} reports mean campaign income by income source, gender, and election year for the mayoral subsample.

\input{tables/tab1_descriptive}

The gender gap in party headquarters transfers is large and persistent. In 2018, male candidates received a mean of 110,381 MXN versus 51,789 MXN for female candidates---a ratio of 0.469, meaning women received 47 cents per male peso in institutional support. In 2021, the male mean rose to 167,448 MXN and the female mean to 100,877 MXN---a ratio of 0.602. The gap narrowed by 13 percentage points in ratio terms and by 0.250 log points.

Sympathizer donations show a similar but more compressed pattern: the female-to-male ratio was 0.366 in 2018 and 0.427 in 2021. Women raise less than half of what men raise from external donors in both years, and this gap also narrowed modestly. Self-financing shows smaller absolute values and a narrower gender gap. Table~\ref{tab:tab2_gender_gap} formalizes these comparisons and provides the descriptive DDD decomposition.

\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Identification.}
We estimate a triple-differences specification using income source type as the third differencing dimension. The three differences are: (1) female versus male candidates, (2) 2021 versus 2018, and (3) party transfers versus sympathizer donations.

The DDD approach asks: did the party-transfer gender gap change differently from the sympathizer gap between 2018 and 2021? Sympathizer donations serve as a reference series --- capturing general trends in how women are received as candidates. Fixed-effect confounders (party affiliation, state-level norms) are absorbed by party$\times$state effects.

\textbf{Contamination caveat.} The DDD identifying assumption requires that the mandate did not directly alter sympathizer donations. In practice this assumption is violated: as Section~\ref{sec:results} shows, the sympathizer DiD is negative and marginally significant, consistent with the mandate placing women in municipalities with weak donor networks. This contamination has two implications. First, the sympathizer counterfactual trends downward relative to men, so the DDD \emph{understates} any genuine improvement in party transfers --- it is a lower bound on the mandate's institutional effect. Second, the wide confidence intervals already reflect substantial uncertainty; the contaminated baseline makes interpretation harder, not easier. We therefore present the DDD as a descriptive bounds check rather than a clean causal estimate. The finding that even this biased-toward-zero estimator cannot be distinguished from zero is itself informative.

\subsubsection*{Specification.}
The main specification is:
\begin{align}
\ln(\text{income}_{ict}) &= \alpha + \beta_1 \text{Female}_i + \beta_2 \text{Post}_t + \beta_3 \text{PartySrc}_c \nonumber \\
&\quad + \beta_4 (\text{Female} \times \text{Post})_{it} + \beta_5 (\text{Female} \times \text{PartySrc})_{ic} \nonumber \\
&\quad + \beta_6 (\text{Post} \times \text{PartySrc})_{ct} \nonumber \\
&\quad + \beta_7 (\text{Female} \times \text{Post} \times \text{PartySrc})_{ict} + \theta_{ps} + \varepsilon_{ict}
\label{eq:ddd}
\end{align}
where $i$ indexes candidates, $c \in \{\text{party transfer, sympathizer}\}$ indexes income source, $t \in \{2018, 2021\}$, $\theta_{ps}$ denotes party$\times$state fixed effects, and standard errors are clustered at the state level. The coefficient of interest is $\hat{\beta}_7$: the differential change in the female-male log income gap for party transfers relative to sympathizer donations, post-mandate.

We supplement the pooled DDD with two separate DiD regressions (Equation~\ref{eq:did} below) estimated within each income source:
\begin{equation}
\ln(\text{income}_{it}) = \alpha + \gamma \text{Female}_i + \delta \text{Post}_t + \gamma_2 (\text{Female} \times \text{Post})_{it} + \theta_{ps} + \varepsilon_{it}
\label{eq:did}
\end{equation}
The DiD for party transfers directly estimates whether the gender gap in party-controlled resources changed post-mandate. The DiD for sympathizer donations is the placebo: absent a direct mandate effect on private donors, we expect $\gamma_2 \approx 0$.

\subsubsection*{Threats to Validity.}
\emph{Compositional change.} The 2021 elections introduced new parties and coalitions and were the first cycle under MORENA's electoral dominance. If MORENA ran more female candidates and distributed resources differently than established parties, the post coefficient would conflate mandate effects with party-composition changes. Party$\times$state fixed effects absorb time-invariant differences across party-state cells. A residual threat is that new party-state cells entered in 2021; we check robustness by restricting to parties present in both years.

\emph{Anticipation effects.} The 2019 reform was announced in June 2019, two years before the 2021 elections. Parties and candidates had time to adjust. To the extent that adjustment was partial by 2021, our estimate understates the long-run financial effect of parity. We have no within-year variation to test for anticipation.

\emph{Strategic candidacy placement.} Horizontal parity required parties to field women in approximately half their municipalities. If parties complied by concentrating female nominations in unwinnable races---municipalities where they expected to lose---then women candidates would face structurally weaker donor environments. This mechanism predicts a negative sympathizer placebo, consistent with what we find.

\section{Results}
\label{sec:results}

\subsubsection*{Descriptive Gender Gap.}
Table~\ref{tab:tab2_gender_gap} presents the gender gap in campaign finance for 2018 and 2021.

\input{tables/tab2_gender_gap}

The party transfer gender gap fell from 0.757 to 0.507 log points between 2018 and 2021---a reduction of 0.250 log points. The sympathizer gender gap fell from 1.006 to 0.851 log points---a reduction of 0.155 log points. The raw descriptive DDD is $0.250 - 0.155 = 0.095$ log points: the party transfer gap narrowed about 60 percent faster than the sympathizer gap. The female-to-male ratio for party transfers rose from 0.469 to 0.602; for sympathizer donations from 0.366 to 0.427. Both gaps narrowed, but party transfers improved more in ratio terms.

\subsubsection*{Regression Estimates.}
Table~\ref{tab:tab3_main_ddd} presents the main regression results.

\input{tables/tab3_main_ddd}

Column (1) reports the full DDD specification (Equation~\ref{eq:ddd}). The DDD coefficient $\hat{\beta}_7 = 0.125$ (SE\,=\,0.275, $p = 0.65$) is positive but not statistically significant at any conventional level. We cannot reject the null hypothesis that the party-transfer gender gap changed by the same amount as the sympathizer gender gap. The standardized effect size is 0.028 standard deviations of the outcome---a small positive effect that is estimated too imprecisely to distinguish from zero (Appendix~\ref{sec:sde}).

The baseline coefficients illuminate the structure. The female indicator is large and negative ($\hat{\beta}_1 = -1.740^{***}$, SE\,=\,0.224): controlling for party$\times$state fixed effects, female candidates receive substantially less income across both sources. The female$\times$party-source interaction is large and positive ($\hat{\beta}_5 = 1.529^{***}$, SE\,=\,0.265), meaning the gender gap is \emph{smaller} for party transfers than for sympathizer donations in the baseline. This is consistent with parties historically providing some institutional floor for women's campaign resources even in the absence of a mandate.

Column (2) reports the DiD for party transfers only (Equation~\ref{eq:did}). The female$\times$post coefficient is $0.104$ (SE\,=\,0.115), positive but not statistically significant. The direction is consistent with modest convergence in party funding, but the estimate is too imprecise to distinguish from zero.

Column (3) is the sympathizer comparison, which was intended to serve as a placebo but instead reveals a substantive finding of its own: $\hat{\beta}(\text{female} \times \text{post}) = -0.372^{*}$ (SE\,=\,0.195, $p = 0.067$). Within party$\times$state cells, women's sympathizer donations declined relative to men post-mandate. As discussed in Section~\ref{sec:strategy}, this violates the DDD identifying assumption and means the DDD coefficient likely understates any genuine narrowing in party transfers. Substantively, the result is consistent with the mandate placing women in municipalities with thin donor networks: parties fulfilled their horizontal-parity obligations by nominating women in historically weak districts, suppressing private donations for newly mandated nominees.

Column (4) repeats the DDD using self-financing as the control income source, finding a null DDD estimate of $0.064$ (SE\,=\,0.238). This corroborates the main finding: there is no detectable differential improvement in party transfers relative to other income types.

\subsubsection*{State-Level Heterogeneity.}
Table~\ref{tab:tab4_state_gaps} reports the gender gap in party transfers for the 24 states appearing in both the 2018 and 2021 samples.

\input{tables/tab4_state_gaps}

The distribution of changes is roughly symmetric around zero, consistent with the null aggregate estimate. States with the largest reductions in the party funding gap include Guerrero ($-0.78$ log points), Colima ($-0.77$), and Puebla ($-0.56$): in these states, women's party transfers improved substantially relative to men's. States with the largest increases include Quintana Roo ($+1.12$), Michoac\'an ($+0.98$), and Chiapas ($+0.88$).

The heterogeneity suggests that the mandate's financial effects were highly context-dependent. States where major parties had already developed organizational support for female candidates before 2021 may have required less new institutional effort to close the financial gap; states where parity was enforced primarily through INE pressure without underlying party commitment saw the gap widen as women were placed in financially unsupported candidacies.

\subsubsection*{Robustness.}
We conduct three additional checks. First, using the inverse hyperbolic sine (IHS) transformation $\text{IHS}(x) = \sinh^{-1}(x) = \ln(x + \sqrt{x^2+1})$ instead of $\ln(x+1)$ to handle zeros differently, the DDD estimate is 0.134 (similar magnitude, still not significant). Second, HC1 heteroskedasticity-robust standard errors on the DDD yield a $p$-value of 0.35, similar to the clustered results. Third, the zero-transfer rate is 41.1 percent for men and 50.2 percent for women in 2018; by 2021 these rates converged to 33.6 and 37.8 percent respectively, consistent with the narrowing party transfer gap but not suggesting that the log-plus-one transformation mechanically drives results. All checks confirm the main null finding.

\section{Conclusion}
\label{sec:conclusion}

Mexico's 2019 ``Parity in Everything'' mandate succeeded at the nomination stage: female mayoral candidates rose from 47 to 49 percent of the candidate pool in a single election cycle. At the financial stage, the evidence is weaker. Women's party headquarters transfers increased from 47 to 60 cents per male dollar, but this improvement cannot be statistically distinguished from what would have been predicted by changes in the broader funding environment. The DDD estimate is $0.125$ log points (SE\,=\,0.275), not significant. Half of Mexican states saw the party funding gap narrow; the other half saw it widen.

Two mechanisms plausibly explain why financial parity lagged legal parity. The first is strategic placement: parties fulfilled horizontal parity requirements by nominating women in municipalities with weak donor bases, suppressing private contributions even as institutional transfers adjusted modestly. The marginal female nominee under parity was placed where the party expected to lose, not where support networks were strong. The second is enforcement scope: the INE's financial mandate covered public matching funds but not party-controlled transfers, leaving parties latitude to discriminate at the funding stage without violating any legal requirement.

The broader lesson is that parity mandates are not self-implementing at the financial level. Nomination equity is a necessary but not sufficient condition for electoral equity. Closing the funding gate---whether through direct financial mandates on party transfers, transparency requirements, or financial penalties for funding gaps---requires a second wave of institutional reform beyond the nomination rules that exist today.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{Column Harmonization Across Years.}
The INE changed column names between the 2018 and 2021 local election disclosure files. The 2021 file uses \textit{Transferencias de Concentradoras} as the party headquarters transfer variable, representing resources transferred from the party's national collecting entity to the candidate's campaign account. The 2018 file disaggregates this into \textit{Transferencia de Recursos Federales} (federal public resources passed through the party) and \textit{Transferencia de Recursos Locales} (local party account resources); we sum these to construct the 2018 party transfer variable. Both represent resources allocated and directed by the party apparatus, not independently raised by the candidate.

The sympathizer donation variable (\textit{Aportaciones de Simpatizantes}) is named identically in both years and covers the same conceptual category. Candidate self-financing (\textit{Aportaciones del Candidato} in 2018, \textit{Aportaciones del/la Candidato/a} in 2021) similarly covers the same concept.

\subsubsection*{Sample Construction.}
We retain all records with \texttt{CARGO} values of \textit{PRESIDENTE MUNICIPAL} (both years) or \textit{ALCALDE} (2021, applicable to Mexico City borough candidates). We exclude candidates for \textit{REGIDOR}, \textit{S\'INDICO}, \textit{CONCEJAL}, \textit{DIPUTADO LOCAL}, and \textit{GOBERNADOR} from the mayoral analysis. After this restriction and dropping records with missing \texttt{SEXO} values, the 2018 sample contains 9,761 candidates and the 2021 sample contains 10,465, for a combined total of 20,226.

\subsubsection*{States in Sample.}
Mexican states hold local elections on staggered schedules. The 2018 file covers 25 states and the 2021 file covers 26 states. Twenty-four states contribute data in both years and form the core estimation sample for the state-level heterogeneity analysis (Table~\ref{tab:tab4_state_gaps}). Four states---Baja California, Nayarit, Tlaxcala, and Aguascalientes---held elections only in 2021 and contribute to the cross-section within that year. The DDD estimation includes all observations, with the post indicator capturing the 2018/2021 dimension.

\subsubsection*{Parsing Currency Values.}
The raw CSV files store income values as formatted currency strings (e.g., ``\$110,381.00''). We strip currency symbols and thousands-separators before converting to numeric. Values that fail to parse are treated as zero (fewer than 0.5 percent of records).

\section{Identification Appendix}

\subsubsection*{Parallel Trends Motivation.}
The DDD design assumes sympathizer donations provide a valid counterfactual trend for party transfers. The key threat is a spillover: if the mandate raised women's perceived viability and thereby increased their sympathizer donations as a side effect, the sympathizer gap would narrow for mandate-induced reasons, biasing the DDD toward zero. The observed negative sympathizer DiD ($\hat{\beta} = -0.372^{*}$) is in the opposite direction---women's sympathizer support declined relative to men post-mandate. This suggests the mandate did not raise women's perceived viability uniformly; if anything, it placed some women in races where private support was absent. The conservative interpretation is that our null DDD understates the true party-transfer effect: the sympathizer counterfactual trended negatively, making any positive party-transfer response appear smaller in relative terms.

\subsubsection*{Fixed Effects Structure.}
All models include party$\times$state fixed effects, absorbing differences in funding norms across party-state cells. For example, MORENA in Oaxaca and PAN in Jalisco may have very different baseline gender funding ratios; the fixed effects ensure we identify from within-cell changes over time, not across-party or across-state comparisons. Candidates in parties that contested a state in only one year do not contribute to the within-cell variation but remain in the sample, with their identifying variation coming from the party-source and female$\times$post-source interactions.

\section{Robustness Appendix}

\subsubsection*{IHS Transformation.}
The inverse hyperbolic sine (IHS) transformation is an alternative to $\ln(x+1)$ for handling zero-valued income: $\text{IHS}(x) = \ln(x + \sqrt{x^2+1})$. For large $x$, $\text{IHS}(x) \approx \ln(2x)$; for $x=0$, $\text{IHS}(0) = 0$. We re-estimate Equation~\ref{eq:ddd} replacing $\ln(\text{income}+1)$ with $\text{IHS}(\text{income})$. The DDD coefficient is $0.134$ (SE not reported), virtually identical to the main estimate of $0.125$ and similarly statistically insignificant. The IHS results confirm that the log-plus-one transformation does not materially affect inference.

\subsubsection*{Heteroskedasticity-Robust Standard Errors.}
{\sloppy With 28 state clusters, cluster-robust inference may be unreliable in finite samples. We re-estimate the DDD using HC1 heteroskedasticity-robust standard errors.\par} The HC1 $p$-value for the DDD coefficient is $0.35$, larger than the clustered $p$-value of $0.65$, consistent with the clustered SEs being conservative in this setting.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
