\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Reimbursement Wage Floor: Medicaid Rate Increases and the Black-White Earnings Gap in Nursing Homes}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Medicaid simultaneously finances long-term care for disproportionately Black residents and sets the dominant price signal for a disproportionately Black workforce. I exploit 22 staggered state Medicaid nursing home reimbursement rate increases enacted between 2017 and 2023 to estimate the racial earnings incidence of these shocks. Using a triple-difference design comparing Black versus White workers in nursing homes versus ambulatory care before and after rate increases, I find that rate increases compress the Black-White quarterly earnings gap by 9.9 percent relative to the ambulatory care benchmark (SE = 2.9 pp). This compression is accompanied by a 15.7 percent decline in Black nursing home employment relative to White employment in the same comparison, consistent with compositional change in which facilities retain fewer but better-compensated Black workers. Because the outcome is quarterly earnings rather than hourly wages, the gap compression reflects both price and quantity margins.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I13, J15, J31, H75 \\
\noindent\textbf{Keywords:} Medicaid, nursing homes, racial earnings gap, reimbursement rates, wage incidence

\newpage

\section{Introduction}

In the United States, roughly 60 percent of nursing home residents are financed by Medicaid, making state-set reimbursement rates the dominant price signal for an industry that employs 1.5 million workers \citep{MACPAC2023}. In many Southern states, Black workers constitute a majority of the direct-care nursing home workforce, earning on average 87.5 cents for every dollar earned by their White counterparts. This dual role---Medicaid as both the primary insurer of predominantly Black long-term care residents and the implicit wage-setter of a predominantly Black workforce---creates a uniquely racialized transmission channel from public reimbursement policy to the labor market.

This paper asks whether Medicaid nursing home rate increases compress the Black-White earnings gap. The question is first-order for two reasons. First, nursing homes are among the most Medicaid-dependent sectors in the U.S. economy, with reimbursement rates directly constraining facility budgets and, through them, worker compensation \citep{Hackmann2019}. Second, the nursing home workforce exhibits extreme occupational segregation by race: Black workers are concentrated in lower-paying certified nursing assistant (CNA) positions while White workers are overrepresented in licensed practical nurse (LPN) and administrative roles \citep{Sloane2004, PHI2022}. If rate increases raise the floor of the earnings distribution where Black workers are concentrated, Medicaid reimbursement policy functions as a de facto racial wage policy.

I study 22 state-level Medicaid nursing home rate increases enacted between 2017 and 2023. These events were driven primarily by the COVID-19 staffing crisis, which created intense political pressure on state legislatures to raise rates---a source of variation that is plausibly exogenous to pre-existing racial earnings trends within the nursing home sector. I employ a triple-difference (DDD) design that compares (1) Black versus White workers, (2) in nursing homes (NAICS 623) versus ambulatory health care (NAICS 621), (3) before and after state rate increases. The comparison with ambulatory care absorbs race-specific earnings trends common to the broader health care sector, isolating the component attributable to Medicaid reimbursement.

The main result is that rate increases compress the Black-White earnings gap in nursing homes by 9.9 percent relative to ambulatory care (SE = 2.9 percentage points, $p = 0.001$). In dollar terms, the differential quarterly earnings gain for Black nursing home workers relative to what the DDD comparison group would predict is \$554 (SE = \$158). The result is stable across specifications: a saturated DDD with state$\times$year$\times$industry and state$\times$year$\times$race fixed effects yields an identical point estimate (0.099), and a within-nursing-home difference-in-differences produces a nearly identical coefficient (0.092, $p < 0.001$). Leave-one-out analysis shows the coefficient ranges from 0.091 to 0.105 when each treated state is dropped in turn, confirming that no single state drives the result.

However, the gap compression is accompanied by a reduction in relative Black employment. The employment DDD shows a $-$15.7 percent decline in Black nursing home employment relative to White employment and ambulatory care ($p = 0.001$). New hires follow a similar pattern ($-$9.5 percent, $p = 0.017$). This combination---rising relative earnings and falling relative employment for Black workers---is consistent with a compositional upgrading mechanism: rate increases allow facilities to offer higher wages, attracting or retaining more experienced workers while reducing reliance on the lowest-wage tier of the workforce, which is disproportionately Black.

This paper contributes to three literatures. First, I add to the literature on the incidence of public reimbursement in health care labor markets. \citet{Hackmann2019} shows that staffing mandates in nursing homes reduce profitability and increase wages; \citet{Grabowski2004} documents that Medicaid rate increases improve staffing levels; and \citet{Werner2012} links reimbursement to quality. None of these papers examines the racial incidence of rate changes within the nursing home workforce. Second, I contribute to the literature on the racial wage gap in low-wage service sectors \citep{Blau2017, Cajner2017}. Most work focuses on minimum wage policy \citep{Derenoncourt2021}; I show that sector-specific public reimbursement operates through a similar floor-raising mechanism but with distinct employment consequences. Third, I add to the growing literature on Medicaid's effects on the health care workforce \citep{Chen2016, Garthwaite2012, Duggan2000}, demonstrating that reimbursement affects not only the level of employment but its racial composition.

The rest of the paper proceeds as follows. Section 2 describes the institutional setting. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 presents results, including mechanism and robustness tests. Section 6 discusses implications and limitations.

\section{Institutional Background}

\subsubsection*{Medicaid and nursing home financing}  Medicaid is the single largest payer for nursing home care in the United States, financing approximately 60 percent of residents \citep{MACPAC2023}. Each state sets its own Medicaid nursing facility reimbursement rate, typically through a cost-based or prospective per-diem payment. Rates vary enormously across states: Texas pays approximately \$130 per resident-day while New York pays over \$200 \citep{KFF2023}. For nursing homes with high Medicaid census---the norm outside metropolitan areas---the state rate effectively determines the facility's revenue per bed, and through it, the wage structure for direct-care workers.

\subsubsection*{The nursing home workforce}  Nursing homes employ approximately 1.5 million workers, predominantly in direct-care roles: certified nursing assistants (CNAs), home health aides, and orderlies. This workforce is disproportionately Black and female. In several Southern states, Black workers constitute a majority of the nursing home labor force \citep{PHI2022}. Wages are low: median CNA earnings are roughly \$15 per hour nationally, with substantial geographic variation. Unionization is rare (12 percent coverage), and workers have limited outside options \citep{BLS2023}. In this setting, Medicaid reimbursement rates function as a binding wage floor for the lowest-paid workers because facilities cannot pay less than what the reimbursement revenue allows.

\subsubsection*{COVID-era rate increases}  The COVID-19 pandemic triggered an acute staffing crisis in nursing homes, with facilities competing for workers against hospitals, retail, and other sectors offering higher wages. In response, 22 state legislatures enacted major Medicaid nursing home rate increases between 2017 and 2023, with the majority concentrated in 2021--2022. These were not routine adjustments: Ohio's 2021 rebasing raised rates by approximately \$17 per day; Pennsylvania's 2022 Act 54 delivered a 17.5 percent increase; Michigan appropriated a 10 percent increase in its FY2023 budget. The staggering of these events across states and years provides the quasi-experimental variation for this paper. Critically, the political impetus was the staffing crisis itself---not pre-existing trends in the racial earnings gap---making treatment timing plausibly exogenous to the outcome of interest.

\section{Data}

\subsubsection*{Outcome data}  I use the Census Bureau's Quarterly Workforce Indicators (QWI), which provide employment and average quarterly earnings at the state$\times$quarter$\times$race$\times$industry level. I extract data for three NAICS 3-digit industries: 623 (Nursing and Residential Care Facilities), 621 (Ambulatory Health Care Services), and 721 (Accommodation). For race, I use non-Hispanic Black (QWI code A2) and non-Hispanic White (QWI code A1) workers, both sexes combined. The sample spans 2010--2024 across all 51 states (50 states plus DC), yielding 18,036 state-quarter-race-industry observations.

\subsubsection*{Treatment data}  I code a binary treatment indicator for 22 states that enacted major Medicaid nursing home rate increases ($\geq$5 percent above inflation or explicit rebasing events) between 2017 and 2023, with treatment assigned in the year of legislative enactment. The treatment events are documented from state budget bills, MACPAC annual reports, and National Conference of State Legislatures tracking. The 29 remaining states serve as the never-treated comparison group.

\subsubsection*{Summary statistics}  \Cref{tab:summary} presents the data. In nursing homes, Black workers earn \$2,610 per quarter on average compared to \$2,982 for White workers, a ratio of 87.5 percent. The gap is substantially wider in ambulatory care (64.4 percent ratio), reflecting sharper occupational segregation in physician-practice settings. This 23-percentage-point difference in Black-White earnings ratios across the two industries is the leverage point for the DDD design.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Triple-Difference Design}

The identifying variation comes from the interaction of three margins: race, industry Medicaid dependence, and the timing of state rate increases. I estimate:
\begin{multline}
\log Y_{sirj} = \beta_1 (\text{Post}_{s} \times \text{Black}_{r} \times \text{NH}_{j}) + \beta_2 (\text{Post}_{s} \times \text{NH}_{j}) \\
+ \beta_3 (\text{Post}_{s} \times \text{Black}_{r}) + \gamma_{st} + \delta_{rj} + \eta_{sj} + \mu_{sr} + \varepsilon_{sirj}
\label{eq:ddd}
\end{multline}
where $Y_{sirj}$ is average quarterly earnings in state $s$, year $i$, race $r$, and industry $j$; $\text{Post}_s$ is an indicator equal to one after state $s$ enacts a major rate increase; $\text{Black}_r$ indicates Black workers; and $\text{NH}_j$ indicates NAICS 623. The fixed effects $\gamma_{st}$ (state$\times$year) absorb all state-level shocks, $\delta_{rj}$ (industry$\times$race) absorb time-invariant differences, and $\eta_{sj}$ and $\mu_{sr}$ absorb state-specific industry and race effects. Standard errors are clustered at the state level.

The coefficient of interest $\beta_1$ identifies how rate increases differentially affect Black versus White earnings in nursing homes relative to ambulatory care. The key identifying assumption is that, absent the rate increase, the Black-White earnings gap would have evolved similarly in nursing homes and ambulatory care within the same state. This is more credible than a standard DD assumption because it allows for (1) state-specific earnings trends common to both industries, (2) industry-specific racial earnings dynamics, and (3) state-specific industry trends.

I also estimate a saturated version replacing the lower-order fixed effects with state-year-industry and state-year-race fixed effects, which absorbs all variation except the triple-interaction of interest.

\subsection{Callaway-Sant'Anna Staggered DiD}

To account for heterogeneous treatment effects across cohorts, I also estimate Callaway-Sant'Anna group-time ATT parameters separately for Black and White nursing home workers \citep{CallawayS2021}. The comparison group is not-yet-treated states, and I use a universal base period. I aggregate group-time effects to an event-study representation with five pre-treatment and four post-treatment years.

\subsection{Threats to Validity}

The primary threat is that rate increases coincided with other state-level labor market shocks that differentially affected Black workers in nursing homes. Three features of the design mitigate this concern. First, the DDD structure absorbs state$\times$year shocks (common to both industries) and race$\times$industry trends. Second, I use accommodation (NAICS 721, hotels) as a placebo industry: if the DDD captures state-level shocks rather than Medicaid-specific effects, hotels should show similar patterns. Third, the COVID-era timing is plausibly exogenous---the political pressure to raise rates came from the staffing crisis broadly, not from trends in the racial earnings gap specifically.

A second concern is compositional change. If rate increases alter who works in nursing homes rather than what incumbent workers earn, the earnings DDD may reflect selection rather than treatment. I test this directly by estimating the DDD on employment outcomes.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main results. Column (1) shows the basic DDD: rate increases compress the Black-White log earnings gap in nursing homes relative to ambulatory care by 9.9 percent ($\beta_1 = 0.099$, SE = 0.029, $p = 0.001$). Column (2) confirms this with saturated state$\times$year$\times$industry and state$\times$year$\times$race fixed effects, yielding an identical coefficient. Column (3) reports the levels specification: the differential quarterly earnings gain for Black workers is \$554 (SE = \$158, $p < 0.001$). To translate: in a typical nursing home with 100 employees, of whom 40 are Black, the rate increase raises Black workers' aggregate quarterly earnings by \$22,160 more than what the industry comparison would predict.

Column (4) restricts to nursing homes only and estimates a simple DD: the within-nursing-home Black earnings premium relative to White workers increases by 9.2 percent after rate increases ($p < 0.001$). Column (5) runs the DDD using hotels instead of nursing homes as the treatment industry (with ambulatory care still as comparison). The hotel placebo coefficient is 0.071 ($p = 0.027$), suggesting some state-level compression of the Black-White earnings gap around the time of rate increases that extends beyond nursing homes. However, the nursing home effect (0.099) exceeds the hotel effect by 40 percent, and the within-nursing-home DD (0.092) is unaffected by this concern, confirming a nursing-home-specific component.

\input{tables/tab2_main}

\subsection{Event Study}

\Cref{tab:eventstudy} reports Callaway-Sant'Anna event-study estimates separately for Black and White nursing home workers. Black workers experience a statistically significant earnings decline of \$68 at $t+1$ and \$202 at $t+3$, while White workers' earnings changes are smaller and less precisely estimated. The overall ATT for Black workers is $-$\$75 (SE = \$28, $p < 0.05$); for White workers it is $-$\$43 (SE = \$25, not significant). The differential is $-$\$32, with Black workers losing more---an apparent contradiction with the DDD result.

This divergence reflects two distinct estimands. The CS estimates measure the \textit{absolute} effect on earnings within nursing homes. During the COVID-disrupted 2020--2023 period, both racial groups experienced earnings declines---plausibly driven by reduced hours, staffing reorganization, and turnover---making the absolute ATT negative for both groups. The DDD asks a different question: did the racial gap change \textit{relative to the ambulatory care benchmark}? The answer is yes---the gap compressed because ambulatory care experienced even larger racial earnings divergence during this period, making nursing homes look relatively more equal after rate increases.

This distinction matters for interpretation. The finding is not that rate increases raised Black workers' absolute earnings, but that they compressed the racial gap within the Medicaid-dependent sector relative to the less-dependent comparison sector. Whether this reflects wage rate changes, hours reallocation, or compositional shifts cannot be determined from quarterly earnings data alone---a limitation I return to in the Discussion.

\input{tables/tab3_eventstudy}

\subsection{Mechanisms: Compositional Upgrading}

The employment DDD reveals a mechanism behind the earnings compression. \Cref{tab:robust} Panel A shows that rate increases reduce relative Black employment in nursing homes by 15.7 percent ($p = 0.001$) and relative Black new hires by 9.5 percent ($p = 0.017$). The combination of rising relative earnings and falling relative employment is consistent with compositional upgrading: when facilities receive higher reimbursement, they can afford to pay more and become more selective. If Black workers were previously concentrated at the lowest-wage tier, a floor increase mechanically raises average Black earnings while potentially displacing the lowest-productivity workers.

This pattern is important for policy interpretation. The 9.9 percent gap compression is not pure wage growth for incumbent Black workers; it partly reflects a shift in who works in nursing homes after the rate increase. The marginal Black worker who exits may earn less than the marginal White worker who exits, raising average Black earnings through selection.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robust} Panel B confirms stability. Leave-one-out analysis yields DDD coefficients ranging from 0.091 (dropping Kansas) to 0.105 (dropping Louisiana), with all 22 estimates significant at conventional levels. No single state drives the result.

The event-study DDD shows pre-treatment coefficients that trend from $-$0.079 at $t-5$ toward zero at $t-1$, then jump to +0.037 at $t+0$ and grow to +0.119 at $t+4$. The pre-trend suggests the racial gap in nursing homes relative to ambulatory care was already narrowing before rate increases, consistent with broader workforce dynamics. However, the discrete acceleration at $t+0$ and the magnitude of post-treatment effects (two to three times larger than the pre-trend pace) suggest a genuine treatment effect on top of any trend.

\section{Discussion}

The finding that Medicaid rate increases compress the Black-White earnings gap in nursing homes has a dual interpretation. On one hand, it demonstrates that public reimbursement policy can function as an effective tool for reducing racial earnings inequality in sectors where Black workers are concentrated. The 9.9 percent reduction in the gap is economically meaningful---comparable in magnitude to the effects estimated for minimum wage increases on racial wage gaps \citep{Derenoncourt2021}. On the other hand, the accompanying reduction in relative Black employment suggests this compression comes at a cost: it may price out the least-skilled Black workers, who have few outside options in the low-wage labor market.

This trade-off---what I call the \textit{reimbursement wage floor}---has not been recognized in the Medicaid literature. Prior work on nursing home reimbursement focuses on quality, staffing levels, and profitability \citep{Grabowski2004, Werner2012, Hackmann2019}. The racial dimension of rate pass-through is absent. Yet in a sector where 60 percent of financing is public and the workforce is racially stratified, every rate decision is implicitly a decision about the racial earnings distribution.

Several limitations warrant caution. First, the treatment coding relies on documented legislative events rather than continuous rate data, introducing measurement error that likely attenuates the estimates. A continuous dosage measure would allow testing whether larger rate increases produce larger gap compression, strengthening the causal chain.

Second, the QWI data provide average quarterly earnings, not hourly wage rates. The 9.9 percent gap compression may reflect changes in hours worked (e.g., remaining Black workers taking on more shifts after departures), compositional shifts in the workforce, or actual wage rate changes---most likely some combination of all three. The ``reimbursement wage floor'' mechanism is one explanation consistent with the data, but the earnings measure cannot isolate it from hours or composition channels.

Third, the hotel placebo coefficient of 0.071 ($p = 0.027$) is 72 percent as large as the main nursing home effect, suggesting that state-level factors correlated with rate increase timing---such as minimum wage increases or tight low-wage labor markets---contributed to Black-White earnings compression broadly. The nursing-home-specific component (the difference, approximately 0.028) is modest, though the within-nursing-home DD (0.092) provides reassurance that the finding is not entirely driven by the industry comparison.

Fourth, the concentration of treatment events in 2021--2022 raises concerns about COVID-specific confounds. Nursing homes experienced extreme staffing disruption, excess mortality, and federal emergency funding during this period, making it difficult to isolate the rate increase channel from other pandemic-related shocks. The DDD absorbs state$\times$year effects common to both industries, but if the pandemic affected nursing homes and ambulatory care differentially by race, the estimates may conflate the rate increase with COVID-specific compositional changes.

Fifth, the analysis cannot distinguish whether displaced Black workers find comparable or worse employment elsewhere, making the welfare implications ambiguous.

\section{Conclusion}

Medicaid nursing home reimbursement rates are not race-neutral. When states raise rates, the racial earnings gap in nursing homes compresses relative to other health care sectors---by 9.9 percent in the data studied here. Whether this compression reflects higher wage rates, hours reallocation, or workforce composition changes remains an open question that future work with hourly wage data could resolve. What is clear is that the compression coincides with reduced relative Black employment, suggesting a trade-off: policies that narrow the earnings gap within the sector may simultaneously reduce the sector's role as an employer of Black workers. Policymakers designing rate increases would do well to pair them with workforce retention incentives targeted at direct-care workers.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
