\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Compliance Tax: Bunching at the IRS Form 990 Filing Threshold}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The IRS requires tax-exempt organizations with gross receipts above \$200,000 to file Form 990---a twelve-page return with mandatory schedules on compensation, governance, and lobbying---while those below may file the four-page Form 990-EZ. Using the complete universe of 1.9 million registered nonprofits from the IRS Exempt Organizations Business Master File, I document sharp bunching just below the \$200,000 threshold. The normalized excess mass is $\hat{b} = 1.52$ ($t = 3.91$), implying approximately 941 organizations adjust reported revenue to avoid the more burdensome filing requirement. A built-in placebo at the \$50,000 threshold---where the compliance burden difference is minimal---shows no significant bunching ($\hat{b} = 0.21$, $t = 0.80$). Bunching is concentrated among religious organizations, human services nonprofits, and organizations with small asset bases, consistent with a compliance cost mechanism.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H26, L31, D82, K34 \\
\noindent\textbf{Keywords:} nonprofit disclosure, bunching estimation, Form 990, compliance costs, tax-exempt organizations

\newpage

\section{Introduction}

Every year, hundreds of thousands of American nonprofits face a hidden tax: the compliance burden of IRS Form 990. For organizations near the \$200,000 revenue threshold, the difference between a four-page filing and a twelve-page disclosure---with mandatory schedules detailing executive compensation, governance practices, and lobbying expenditures---is substantial enough to distort reported revenue. This paper provides the first bunching-based estimate of that compliance cost.

Tax-exempt organizations in the United States control over \$6 trillion in assets and receive over \$400 billion in annual contributions \citep{mckeever2023nonprofit}. The IRS Form 990 is the primary instrument of public accountability for this sector, serving simultaneously as a tax compliance document, a public disclosure vehicle, and an input to state-level charity regulation \citep{fishman2007accounting}. The sharp threshold at \$200,000 in gross receipts---below which organizations may file the simpler Form 990-EZ---creates a natural experiment for estimating the implicit cost of this disclosure regime.

I apply the bunching methodology of \citet{saez2010taxpayers} and \citet{kleven2016bunching} to the complete IRS Exempt Organizations Business Master File (EO BMF), which covers all 1.9 million registered 501(c) organizations. The identifying assumption is simple: absent the filing threshold, the distribution of nonprofit revenue around \$200,000 should be smooth. Any excess mass just below the threshold reveals organizations that adjust their reported or actual revenue to avoid the more complex filing requirement. Note that the filing rule depends on both gross receipts and total assets---organizations exceeding \textit{either} \$200,000 in receipts or \$500,000 in assets must file the full 990. Among organizations near the revenue threshold, 76.5\% have assets below \$500,000, so the revenue cutoff is the binding margin for the large majority of this population.

The main result is a normalized excess mass of $\hat{b} = 1.52$ (SE $= 0.39$), implying that approximately 941 organizations bunch below the threshold. This estimate is robust to polynomial order (ranging from 5th to 9th degree) and to variation in the exclusion window. The bunching ratio---excess mass relative to the counterfactual density---is large and statistically significant at conventional levels.

Three features of the institutional setting strengthen identification. First, a built-in placebo exists at the \$50,000 threshold separating Form 990-N (a simple e-Postcard) from Form 990-EZ. Since the compliance burden difference is minimal at this lower threshold, bunching should be absent---and it is ($\hat{b} = 0.21$, $t = 0.80$). This validates that the main result reflects the filing complexity jump, not mere round-number heaping. Second, the \$200,000 threshold was raised from \$100,000 in tax year 2010, and I find significant bunching at \$100,000 among round-number placebo tests ($\hat{b} = 2.28$, $t = 2.82$)---consistent with a legacy response to the former threshold location. Third, bunching at round-number placebo thresholds (\$150,000, \$250,000, \$300,000) is uniformly insignificant, ruling out mechanical heaping as an explanation.

The heterogeneity analysis reveals who bears the compliance tax. Religious organizations exhibit the strongest bunching ($\hat{b} = 5.45$), followed by human services nonprofits ($\hat{b} = 3.10$). Organizations with small asset bases (under \$100,000) bunch three times as much as medium-asset organizations ($\hat{b} = 3.00$ vs.\ $2.18$), while large-asset organizations show no bunching at all ($\hat{b} = -1.32$). This pattern is consistent with a compliance cost channel: organizations that lack accounting infrastructure respond most strongly to the disclosure burden, while well-resourced organizations absorb the cost.

This paper contributes to three literatures. First, it adds to the growing body of work applying bunching methods to non-tax contexts. \citet{ewens2024startup} study bunching at SEC disclosure thresholds for public float; \citet{chen2020notching} document notching at corporate tax thresholds. I extend this approach to the nonprofit sector, where the ``tax'' is not monetary but informational---the burden of disclosure itself. Second, the paper contributes to research on nonprofit regulation and accountability. \citet{yildirim2018agency} studies bunching at New York State audit thresholds, and \citet{gordon2009manipulation} document revenue manipulation around state solicitation thresholds. My contribution is the first federal-level bunching analysis using the complete universe of tax-exempt organizations. Third, the results speak to the broader literature on regulatory compliance costs \citep{djankov2002regulation, crain2010impact}, providing a revealed-preference estimate of the burden that nonprofits are willing to bear to avoid enhanced disclosure.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of IRS filing requirements. Section 3 presents the data and summary statistics. Section 4 details the bunching methodology. Section 5 reports the results, including placebo tests, robustness checks, and heterogeneity analysis. Section 6 discusses implications, and Section 7 concludes.

\section{Institutional Background}

\subsubsection*{IRS Filing Requirements for Tax-Exempt Organizations} Tax-exempt organizations recognized under Section 501(c) of the Internal Revenue Code must file annual information returns with the IRS. The filing requirement depends on the organization's gross receipts and total assets, creating a tiered disclosure system with three levels.

Organizations with gross receipts normally $\leq$\$50,000 may file Form 990-N, an electronic notice (``e-Postcard'') that requires only the organization's name, address, EIN, and a confirmation that annual receipts are below the threshold. This filing contains effectively no financial disclosure.

Organizations with gross receipts $<$\$200,000 \textit{and} total assets $<$\$500,000 may file Form 990-EZ, a four-page short form that reports total revenue, expenses, and assets but requires limited schedule attachments. The 990-EZ requests basic governance information but does not mandate the detailed schedules required by the full return.

Organizations exceeding either threshold must file Form 990, the full information return. At twelve pages plus mandatory schedules, the 990 requires detailed reporting on: executive and employee compensation (Schedule J), governance policies and practices (Part VI), fundraising activities (Schedule G), related organizations (Schedule R), and---for organizations engaged in lobbying---political and lobbying expenditures (Schedule C). Private foundations file Form 990-PF regardless of size.

\subsubsection*{The 2010 Threshold Reform} Prior to tax year 2010, the gross receipts threshold separating Form 990-EZ from Form 990 was \$100,000 (with an asset threshold of \$250,000). IRS Notice 2009-86, effective for tax years beginning in 2010, raised these thresholds to \$200,000 and \$500,000 respectively. This reform provides a natural test: if bunching is driven by the disclosure threshold rather than round-number preferences, it should have migrated from \$100,000 to \$200,000 after 2010.

\subsubsection*{Consequences of Filing} Both Form 990 and Form 990-EZ are publicly available through the IRS and through third-party platforms such as GuideStar (now Candid) and ProPublica's Nonprofit Explorer. Donors, journalists, state attorneys general, and watchdog organizations routinely use these filings to evaluate nonprofit performance and governance. The key distinction is not public versus private disclosure---both forms are public---but the \textit{depth and complexity} of the required reporting. Form 990 mandates detailed schedules on executive compensation, governance practices, lobbying expenditures, and related-party transactions that the 990-EZ does not require. The compliance burden thus encompasses both the direct cost of preparing these schedules and the indirect cost of the enhanced scrutiny that detailed reporting invites.

\section{Data}

I use the IRS Exempt Organizations Business Master File (EO BMF), a quarterly publication containing identifying information and key financial data for all registered tax-exempt organizations. The file includes the organization's EIN, state, National Taxonomy of Exempt Entities (NTEE) classification code, filing requirement code, and---critically---reported gross receipts (REVENUE\_AMT), total assets (ASSET\_AMT), and net income (INCOME\_AMT).

The current extract contains 1,938,732 registered organizations. Of these, 555,714 report positive gross receipts. The analysis window for the main specification is organizations with gross receipts between \$100,000 and \$300,000, yielding 142,964 observations. The placebo analysis uses the \$20,000--\$80,000 window (107,235 observations).

Revenue is reported in exact dollar amounts, enabling standard bunching estimation with fine-grained bins. A key limitation is that the EO BMF is a cross-sectional snapshot---it reports the most recent filing data rather than a panel of annual returns. I therefore estimate bunching in the cross-section and interpret it as reflecting the steady-state behavioral response to the threshold.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsubsection*{Bunching Estimation} I follow the standard bunching methodology of \citet{saez2010taxpayers} and \citet{kleven2016bunching}. Revenue is binned in \$1,000 intervals. I fit a flexible polynomial counterfactual density to bins outside an excluded region around the \$200,000 threshold and estimate the excess mass---the number of organizations above the counterfactual---in the bunching region.

Formally, let $c_j$ denote the count of organizations in bin $j$ and $z_j$ denote the normalized bin center (distance from threshold in units of bin width). The counterfactual density is:
\begin{equation}
  c_j^0 = \sum_{i=0}^{p} \beta_i z_j^i, \quad j \notin \mathcal{E}
\end{equation}
where $\mathcal{E}$ is the excluded region and $p$ is the polynomial order. The excess bunching mass is:
\begin{equation}
  \hat{B} = \sum_{j \in \mathcal{E}} \left( c_j - \hat{c}_j^0 \right)
\end{equation}
The normalized excess mass $\hat{b} = \hat{B} / \hat{c}_0^0$ scales the excess mass by the counterfactual density at the threshold.

The baseline specification uses a 7th-degree polynomial with an exclusion window of [\$180,000, \$210,000]. I report robustness to polynomial orders 5 through 9 and to exclusion windows ranging from $\pm$\$10,000 to $\pm$\$30,000. Standard errors are computed by parametric bootstrap (1,000 replications), perturbing bin counts with Poisson noise and re-estimating the polynomial and excess mass for each draw.

\subsubsection*{Identification} The identifying assumption is that, absent the filing threshold, the distribution of nonprofit gross receipts would be smooth through \$200,000. Two features of the setting support this assumption.

First, the \$200,000 threshold is not associated with any other policy discontinuity. It does not trigger changes in tax liability, audit probability (at the federal level), or eligibility for grants or programs. The only consequence of crossing the threshold is the filing requirement itself.

Second, the placebo at \$50,000 tests whether organizations bunch at any filing threshold or specifically at thresholds where the compliance burden jumps. The \$50,000 threshold separates Form 990-N from 990-EZ---a transition from a one-page e-Postcard to a four-page form. If organizations bunch whenever a filing requirement changes, we should see excess mass at \$50,000 as well. The absence of such bunching isolates the compliance cost of the \textit{full} Form 990.

\section{Results}

\subsection{Main Bunching Estimates}

\Cref{tab:bunching} presents the main results. At the \$200,000 threshold, I estimate an excess mass of 941 organizations (SE $= 233$) and a normalized excess mass of $\hat{b} = 1.52$ (SE $= 0.39$). The $t$-statistic of 3.91 indicates that bunching is highly statistically significant. The counterfactual density at the threshold is approximately 619 organizations per \$1,000 bin, so the excess represents roughly 152\% of one bin's counterfactual count.

At the \$50,000 placebo threshold, the estimated excess mass is 389 (SE $= 476$) with a normalized excess mass of $\hat{b} = 0.21$ (SE $= 0.26$, $t = 0.80$). The placebo estimate is statistically indistinguishable from zero, confirming that the \$200,000 bunching is driven by the compliance burden of the full Form 990 rather than by generic threshold effects.

\input{tables/tab2_bunching}

The missing mass above the threshold---the counterfactual count minus the actual count in the [\$200,000, \$210,000] region---is 269 organizations. The excess mass (941) exceeds the missing mass (269), an imbalance that deserves discussion. Three factors may contribute. First, organizations may compress revenue by substantially more than the minimum needed, diffusing the missing mass across a wider range above the threshold than the exclusion window captures. Second, some bunching may reflect real activity responses---organizations that constrain their operations rather than merely relabeling revenue---which would reduce observed revenue without creating a corresponding ``hole'' in the distribution above the threshold. Third, the polynomial counterfactual may not perfectly capture the shape of the underlying distribution. The imbalance is within the range observed in other bunching studies \citep{kleven2016bunching}, but readers should interpret the excess mass as an upper bound on the number of organizations actively adjusting reported revenue.

\subsubsection*{McCrary Density Test} A simple density test comparing the mean bin count in the \$190,000--\$200,000 range to the \$200,000--\$210,000 range yields a $t$-statistic of 5.97 ($p < 0.0001$). The mean count per \$1,000 bin is 729.0 below the threshold versus 572.8 above---a 27\% difference.

\subsection{Robustness}

\Cref{tab:robustness} demonstrates that the bunching estimate is robust to specification choices. Panel A varies the polynomial order from 5 to 9; $\hat{b}$ ranges from 0.63 to 1.65, with all estimates positive and most statistically significant at the 5\% level. Panel B varies the exclusion window; estimates range from 1.20 to 3.48, with larger windows producing larger estimates as they capture additional bunching mass further from the threshold.

\input{tables/tab3_robustness}

\subsubsection*{Round-Number Placebos} I estimate bunching at four round-number thresholds (\$100,000, \$150,000, \$250,000, \$300,000) where no filing requirement changes. Three of four placebos (\$150K, \$250K, \$300K) are statistically insignificant ($t < 1.0$). The exception is \$100,000, where $\hat{b} = 2.28$ ($t = 2.82$)---consistent with the fact that \$100,000 was the \textit{former} threshold before the 2010 reform. This legacy bunching provides additional evidence that organizations respond specifically to the Form 990 filing requirement, not to round numbers.

\subsubsection*{Donut Robustness} Excluding additional bins above the threshold (donut designs of \$2,000, \$5,000, and \$10,000) produces stable estimates ($\hat{b} = 1.63$, $1.53$, and $0.74$ respectively), ruling out the possibility that the result is driven by a single anomalous bin.

\subsection{Heterogeneity}

\Cref{tab:heterogeneity} reveals striking variation in bunching intensity across organization types and sizes. Panel A shows that religious organizations exhibit the strongest bunching ($\hat{b} = 5.45$, SE $= 1.92$), followed by human services nonprofits ($\hat{b} = 3.10$, SE $= 1.06$). Health and education/foundation organizations show positive but imprecisely estimated bunching. The ``Other'' category---which includes arts, environment, international, and miscellaneous organizations---shows the weakest response ($\hat{b} = 0.76$).

\input{tables/tab4_heterogeneity}

Panel B stratifies by asset size. Small-asset organizations (under \$100,000) bunch three times as intensely as medium-asset organizations ($\hat{b} = 3.00$ vs.\ $2.18$). Large-asset organizations (over \$500,000) show no bunching ($\hat{b} = -1.32$). This result has a dual interpretation. First, it is consistent with a compliance cost mechanism: well-resourced organizations can absorb the filing burden without adjusting behavior. Second, it addresses the two-dimensional filing rule: organizations with assets above \$500,000 must file Form 990 regardless of revenue, so the revenue threshold is not binding for them and bunching should be absent---exactly what the data show. The asset-size gradient thus simultaneously validates the mechanism and confirms that the revenue threshold is the operative margin for the organizations driving the main result.

\subsubsection*{State-Level Variation} States with stringent charitable solicitation requirements (New York, California, Massachusetts, Connecticut, New Jersey, Illinois, Pennsylvania, Ohio) show bunching of $\hat{b} = 1.06$, while the remaining states show $\hat{b} = 1.79$. The stronger bunching in lower-regulation states is consistent with the federal threshold being the binding constraint---in high-regulation states, organizations already face state-level disclosure requirements that may partially substitute for the Form 990 burden.

\section{Discussion}

The bunching estimate implies that roughly 941 organizations---approximately 0.7\% of those in the [\$100,000, \$300,000] revenue range---adjust their reported or actual gross receipts to avoid the full Form 990. The behavioral response could take several forms: timing of revenue recognition, deferral of fundraising, constraining operations, or---in some cases---deliberate underreporting. The bunching methodology cannot distinguish among these channels, but the policy implication is the same: the filing threshold distorts behavior. The 941 figure is likely a lower bound on the number of organizations affected, as it captures only those that actively adjust to stay below \$200,000, not those that bear the compliance cost or those that would have grown past the threshold absent the filing burden.

The heterogeneity pattern---strongest bunching among religious organizations and small-asset nonprofits---speaks to mechanism. Religious organizations are already exempt from many state-level reporting requirements and may be particularly averse to the governance and compensation disclosures mandated by Form 990. Small-asset organizations likely lack dedicated accounting staff and face a fixed compliance cost that is large relative to their budgets. These findings are consistent with a model where the threshold creates a notch in the effective ``tax'' on nonprofit revenue, with the compliance cost functioning as a lump-sum burden that binds for resource-constrained organizations.

Several limitations warrant acknowledgment. First, the EO BMF is a cross-sectional snapshot that may mix filings from different tax years, limiting the ability to exploit the 2010 threshold reform as a panel design. The significant bunching at \$100,000---the former threshold---is suggestive of legacy behavior but could also reflect stale data entries. A panel of annual Form 990 microdata would allow direct estimation of bunching migration before and after 2010. Second, the bunching methodology estimates the number of organizations that respond to the threshold but cannot directly measure the dollar cost of Form 990 compliance. Translating excess mass into an implied compliance cost requires assumptions about the distribution of adjustment costs that the data cannot identify. Third, the analysis uses the full universe without conditioning on the asset threshold, though the large-asset heterogeneity analysis confirms that bunching is concentrated where the revenue threshold binds.

The policy implications are direct. As states and the IRS consider further reforms to nonprofit disclosure, the compliance cost revealed by bunching should be weighed against the accountability benefits of full reporting. The 2010 reform that raised the threshold from \$100,000 to \$200,000 presumably reduced the number of organizations facing the compliance burden, but the evidence here suggests that even at \$200,000, the filing requirement distorts behavior for hundreds of small nonprofits.

\section{Conclusion}

This paper documents that tax-exempt organizations strategically avoid the IRS Form 990 filing requirement by bunching their reported gross receipts below the \$200,000 threshold. The compliance cost of the full disclosure return---not its monetary tax implications, which are zero---is sufficient to distort the behavior of nearly a thousand organizations. The burden falls most heavily on the smallest and least professionalized nonprofits, precisely the organizations that the tiered filing system was designed to protect.

The broader lesson is that disclosure regulation, like taxation, creates behavioral responses that can be measured through the density of the regulated variable. When compliance costs are concentrated on organizations least equipped to bear them, well-intentioned transparency requirements may paradoxically reduce the quality of information available about the nonprofits that donors and regulators most need to monitor.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
