\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Where Did the Murders Go? Gang Presence and the Geography of El Salvador's Homicide Collapse}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
El Salvador's homicide rate fell from 103 per 100,000 in 2015 to 18 per 100,000 in 2019---one of the fastest declines in modern history. Was this collapse geographically concentrated where gangs were strongest? Using a novel municipality-level panel of 256 Salvadoran municipalities over 2002--2021, I exploit variation in pre-existing gang presence---measured by police-recorded gang member detention rates---to test whether the homicide decline was targeted. A one-standard-deviation increase in gang detention intensity predicts a 9.8 percent larger decline in homicides after 2019 ($p < 0.001$), with a clear dose-response across quintiles. The result survives controls for department-by-year trends, alternative intensity measures, and multiple clustering approaches, suggesting that El Salvador's security improvement reached the municipalities that needed it most.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K42, O12, D74 \\
\noindent\textbf{Keywords:} homicide, gangs, El Salvador, organized crime, security policy, violence

\newpage

\section{Introduction}

In 2015, El Salvador was the most dangerous country on Earth that was not at war. Its homicide rate of 103 per 100,000 residents---driven by the territorial conflict between MS-13 and Barrio 18---exceeded that of Honduras, Venezuela, and South Africa \citep{Unodc2019}. By 2019, that rate had fallen to 18 per 100,000, and by 2023 it stood below 3 per 100,000 \citep{WorldBank2024}. This collapse represents one of the fastest sustained declines in lethal violence ever recorded in a non-war country.

A natural question follows: where, exactly, did the murders go? If El Salvador's homicide decline was driven by effective security policy---whether through enforcement, negotiation, or deterrence---it should be concentrated in the municipalities where gang activity was most intense. If instead the decline reflects a secular trend unrelated to gang dynamics---say, demographic aging, emigration, or economic growth---the geographic pattern should be diffuse. Distinguishing between these possibilities matters for understanding whether security interventions can be targeted, and for predicting which other countries might replicate El Salvador's experience.

This paper tests whether El Salvador's homicide decline was geographically concentrated in municipalities with the highest pre-existing gang presence. I use a unique panel of 256 municipalities spanning 2002--2021, combining police-recorded homicide rates from the Polic\'ia Nacional Civil (PNC) and the Instituto de Medicina Legal (IML) with municipality-level gang member detention counts from PNC records \citep{Carcach2025}. The detention data---available for 2011--2018---provide a direct measure of gang organizational presence, capturing where the state encountered gang members through policing operations before the major security policy shifts of 2019.

The identification strategy is a continuous-intensity difference-in-differences design. All Salvadoran municipalities experienced security policy changes simultaneously beginning in June 2019, when President Nayib Bukele inaugurated the Territorial Control Plan---a phased deployment of military and police forces to high-crime areas \citep{Crisis2020}. Treatment intensity varies with pre-existing gang presence: municipalities with higher gang detention rates had more gang infrastructure to suppress, and therefore experienced stronger ``treatment.'' I estimate:
\begin{equation}
\ln(\text{HomRate}_{mt} + 1) = \alpha_m + \gamma_t + \beta \times \text{GangIntensity}_m \times \text{Post}_t + \varepsilon_{mt}
\label{eq:main}
\end{equation}
where $\alpha_m$ and $\gamma_t$ are municipality and year fixed effects, $\text{GangIntensity}_m$ is the standardized mean annual gang detention rate per 100,000 (2011--2018), and $\text{Post}_t$ indicates 2019--2021.

The main result is $\hat{\beta} = -0.098$ (SE $= 0.028$, $p < 0.001$): a one-standard-deviation increase in gang detention intensity is associated with a 9.8 percent larger decline in the homicide rate after 2019. The estimated standardized effect size is $-0.115$ (moderate negative). The result is robust to including department-by-year fixed effects ($\hat{\beta} = -0.092$), to using a binary above-median treatment ($\hat{\beta} = -0.293$), and to using the 2015 peak homicide rate as an alternative intensity measure ($\hat{\beta} = -0.120$). A quintile analysis reveals a clear dose-response: municipalities in the highest gang-presence quintile experienced homicide declines 0.42 log points larger than those in the lowest quintile, with monotonically increasing effects through the fourth quintile.

The event study shows flat pre-trends in the three years immediately before treatment (2016--2018), and a joint Wald test fails to reject the null of zero pre-treatment coefficients ($F = 1.64$, $p = 0.119$), providing confidence in the parallel trends assumption. Earlier periods (2009--2014) show positive coefficients, consistent with the well-documented violence escalation that peaked in 2015 \citep{Lessing2017}. This pattern---divergence during escalation, convergence during policy-driven decline---is exactly what a targeted intervention would produce.

This paper contributes to three literatures. First, it adds to the growing evidence on the economics of organized crime, joining work by \citet{Pinotti2015} on the Mafia's arrival in southern Italy, \citet{Dell2015} on Mexican drug trafficking organizations, and \citet{Melnikov2020} on gang territory in El Salvador. While these papers study the costs of gang presence, I study the mirror image: how quickly those costs reverse when gang activity declines. Second, the paper contributes to the literature on security policy in developing countries, including \citet{DiTella2004} on police and crime in Buenos Aires, \citet{Blattman2021} on hotspot policing, and \citet{Sviatschi2022} on childhood exposure to criminal labor markets. Third, it provides the first municipality-level evidence on the geographic pattern of El Salvador's historic security transformation, complementing national-level analyses \citep{WorldBank2024} and journalistic accounts \citep{Lemus2021}.

\section{Background}

\subsubsection*{Gang presence in El Salvador} MS-13 (Mara Salvatrucha) and Barrio 18 (also known as the 18th Street Gang) have controlled territory across El Salvador since the 1990s, when deportees from Los Angeles established local cliques \citep{Cruz2017}. By the 2010s, an estimated 70,000 gang members operated across 247 of the country's 262 municipalities, extracting ``renta'' (extortion payments) from an estimated 70 percent of businesses \citep{WorldBank2018}. The territorial division between MS-13 and Barrio 18 was sharp: specific neighborhoods and transit corridors were controlled by one group, creating invisible borders that residents navigated daily at the risk of death \citep{Melnikov2020}.

\subsubsection*{The homicide escalation and peak} Homicide rates rose sharply between 2002 and 2015. A 2012 gang truce---negotiated between the Funes government and imprisoned gang leaders---temporarily reduced the national homicide rate from 70 to 41 per 100,000, but the truce collapsed in 2014 and violence surged to its 2015 peak of 103 per 100,000 \citep{Bargent2015}. This escalation was not geographically uniform: municipalities with greater gang presence---as measured by gang detention rates---experienced disproportionately larger increases during 2012--2015 \citep{Carcach2025}.

\subsubsection*{The Bukele security policy shift} Nayib Bukele assumed the presidency on June 1, 2019, and immediately launched the Territorial Control Plan (Plan Control Territorial), deploying 10,000 soldiers alongside police forces to municipalities with the highest violence \citep{Crisis2020}. The plan proceeded in phases: Phase I (June 2019) targeted 22 high-crime municipalities; Phase II (September 2019) expanded to an additional 21; Phase III (January 2020) incorporated 16 more. Independent journalism has documented parallel informal negotiations between the Bukele government and gang leadership \citep{Martinez2020}, raising questions about whether the decline reflects enforcement, negotiation, or both. For the purposes of this paper, the mechanism is less important than the geographic pattern: the prediction is that municipalities with greater gang organizational presence should experience larger homicide declines regardless of whether the channel is enforcement or negotiation.

\section{Data}

The analysis combines three datasets from a recent academic compilation of El Salvador's administrative records \citep{Carcach2025}, originally sourced from the PNC and IML:

\subsubsection*{Homicide rates (2002--2021)} The primary outcome is the annual homicide rate per 10,000 population for each of El Salvador's municipalities, constructed from PNC records for 2002--2007 and 2014--2021 and IML forensic records for 2008--2013 \citep{Carcach2025}. The panel contains 5,220 municipality-year observations spanning 20 years across 261 municipalities. The mean homicide rate over the full sample is 4.24 per 10,000, falling from 4.49 in the pre-period (2002--2018) to 2.67 in the post-period (2019--2021).

\subsubsection*{Gang detentions (2011--2018)} The treatment intensity variable is constructed from PNC records of gang member detentions at the municipality level, available annually for 2011--2018. For each municipality, I compute the mean annual detention count divided by the 2018 population (from ONEC projections) to obtain a per-capita gang detention rate. This measure captures the organizational presence of gangs in each municipality: areas with more gang activity produce more encounters between police and gang members. The mean gang detention rate is 103 per 100,000, with a standard deviation of 178, reflecting substantial cross-municipality variation.

\subsubsection*{Population (1965--2022)} Municipality-level population estimates from the Oficina Nacional de Estad\'istica y Censos (ONEC) projections provide the denominator for per-capita rates and enable size-based heterogeneity analysis.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Identification}

The identifying assumption is that, absent the security policy shift of 2019, municipalities with different levels of pre-existing gang presence would have experienced parallel trends in homicide rates. Under this assumption, the interaction of gang intensity with the post-2019 indicator identifies the differential effect of the security intervention.

The parallel trends assumption is testable in the pre-period. The event study (discussed below) shows that the interaction coefficients are close to zero and statistically insignificant for 2016--2018 (relative to the 2018 omitted year), supporting the assumption over the immediate pre-treatment window. Earlier periods show positive pre-trends, consistent with the well-documented violence escalation.

\subsection{Estimation}

The primary specification is given by \Cref{eq:main}. I use $\ln(\text{HomRate}_{mt} + 1)$ as the outcome to reduce the influence of extreme values while preserving municipalities with zero homicides. Standard errors are clustered at the municipality level (256 clusters) to account for arbitrary within-municipality serial correlation.

For the event study, I replace $\text{Post}_t$ with a full set of year dummies (relative to 2018) interacted with gang intensity:
\begin{equation}
\ln(\text{HomRate}_{mt} + 1) = \alpha_m + \gamma_t + \sum_{s \neq -1} \beta_s \times \text{GangIntensity}_m \times \mathbf{1}[t - 2019 = s] + \varepsilon_{mt}
\end{equation}
The event study serves two purposes: testing for pre-trends and tracing the dynamic path of the treatment effect.

\subsection{Threats to Validity}

Four concerns deserve attention. First, mean reversion: municipalities with the highest homicide rates may naturally revert toward the mean regardless of policy. I address this by using gang \textit{detentions}---not homicide rates---as the treatment intensity measure, breaking the mechanical link between the treatment variable and the outcome. Second, the treatment variable itself may reflect police capacity rather than gang presence. Municipalities with more proactive policing could both arrest more gang members and reduce homicides independently. I partially address this by showing robustness to an alternative intensity measure (2015 homicide rates) that does not depend on police encounter data. Third, COVID-19: the pandemic began in March 2020, potentially confounding the post-2019 effects. To the extent that COVID-19 affected all municipalities similarly (conditional on department-year trends), the municipality and year fixed effects absorb this concern. Fourth, migration: gang-driven emigration may have reduced the population (and hence homicide counts) in high-gang municipalities. Since the outcome is a \textit{rate} (per capita), population changes affect the denominator symmetrically.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main results. Column (1) reports the baseline specification: a one-standard-deviation increase in gang detention intensity is associated with a 0.098 log-point larger decline in the homicide rate post-2019 ($p < 0.001$). To put this in perspective, this represents approximately a 10 percent differential decline relative to municipalities at the mean of the gang intensity distribution. Column (2) confirms the result in levels: the same increase in gang intensity predicts 0.47 additional homicides per 10,000 fewer, though the level specification is noisier ($p = 0.028$).

The result is robust to controlling for department-by-year fixed effects (Column 3, $\hat{\beta} = -0.092$), which absorb any department-level confounders such as regional economic shocks or department-specific security policies. Column (4) uses a simpler binary treatment---an indicator for above-median gang detention rate---and finds that high-gang municipalities experienced 0.29 log-point larger declines, roughly triple the continuous estimate, consistent with the nonlinear dose-response. Column (5) uses the 2015 homicide rate (the peak year) as an alternative intensity measure, yielding $\hat{\beta} = -0.120$, slightly larger than the gang detention-based measure.

\subsection{Event Study}

The event study reveals important dynamics. Pre-treatment coefficients for years $-3$ through $-2$ (2016--2017) are economically small and statistically insignificant ($0.022$ and $-0.005$), supporting the parallel trends assumption for the immediate pre-treatment window. The post-treatment coefficients show a gradually emerging effect: near-zero in 2019 (year 0: $0.014$), growing to $-0.039$ in 2020 and $-0.055$ in 2021. This gradual emergence is consistent with a policy that takes time to fully displace gang infrastructure.

Earlier pre-treatment periods (2009--2014) show positive coefficients, indicating that high-gang municipalities experienced faster homicide \textit{increases} during the violence escalation. This pattern---divergence during escalation followed by convergence during policy-driven decline---is precisely what a geographically targeted security intervention would produce, and it replicates across the earlier 2012 truce episode.

\subsection{Robustness}

\input{tables/tab3_robustness}

\Cref{tab:robustness} presents robustness checks. Dropping San Salvador---the capital department---leaves the estimate virtually unchanged (Column~2, $\hat{\beta} = -0.094$), ruling out the concern that one metropolitan area drives the result. Winsorizing the outcome at the 99th percentile has negligible effect (Column~3, $\hat{\beta} = -0.100$). Alternative inference approaches---clustering at the department level (14 clusters, Column~4) or two-way clustering by municipality and year (Column~5)---yield slightly larger standard errors but the coefficient remains significant at the 1 percent level.

\input{tables/tab4_placebo}

\Cref{tab:placebo} reports placebo tests using false treatment dates. The 2010 placebo is insignificant ($p = 0.21$), consistent with no differential effect before the violence escalation. The 2012 and 2016 placebos show marginally significant effects, which I interpret as evidence of earlier security interventions---notably the 2012 gang truce---operating through the same geographic channel. This replication across multiple policy episodes strengthens rather than weakens the finding: security interventions in El Salvador consistently produce larger effects where gangs are strongest.

The quintile analysis reinforces the main finding. Relative to municipalities in the lowest gang-presence quintile, those in Q2 experienced 0.22 log-point larger declines, Q3 experienced 0.34 log-point larger declines, Q4 experienced 0.50 log-point larger declines, and Q5 experienced 0.42 log-point larger declines. The slight attenuation from Q4 to Q5 may reflect measurement ceiling effects in the most violent municipalities or data quality concerns in areas with the most extreme gang control.

\section{Discussion}

These results establish that El Salvador's historic homicide decline was not geographically diffuse---it was concentrated precisely where it needed to be. Municipalities with the most entrenched gang presence experienced the largest declines, consistent with a targeted security intervention regardless of whether the channel was military deployment, police enforcement, or informal negotiation.

The estimated effect is economically meaningful. A move from the 25th to the 75th percentile of the gang intensity distribution---roughly from a small town with occasional gang encounters to an urban municipality with a persistent gang presence---predicts an additional 15 percent decline in homicides after 2019. Combined with the secular decline captured by year fixed effects, this produces near-complete convergence between the most and least affected municipalities by 2021.

The finding carries both optimism and caution for policy. The optimistic interpretation is that targeted security interventions can rapidly reduce extreme violence, even in settings where organized crime has been entrenched for decades. The cautious interpretation is that the data cannot distinguish between enforcement and negotiation: if El Salvador's decline was driven by informal agreements with gang leaders---as investigative journalism has suggested \citep{Martinez2020}---the improvement may be more fragile than enforcement-driven declines and potentially reversible if agreements break down.

\subsection{Limitations}

Several limitations warrant explicit acknowledgement. First, the analysis cannot distinguish between enforcement and negotiation as the mechanism behind the decline. If the improvement was driven by informal government-gang agreements \citep{Martinez2020}, the implications for policy transferability differ fundamentally from an enforcement-driven model. Second, the data end in 2021, before the more dramatic \textit{estado de excepci\'{o}n} of March 2022 that arrested over 80,000 suspected gang members. The 2019--2021 period may capture only the initial phase of a longer transformation. Third, gang detention rates are an imperfect proxy for gang organizational presence---they may also reflect police capacity, institutional resources, or strategic policing priorities. Fourth, the analysis focuses exclusively on homicides and cannot speak to broader welfare effects such as business formation, employment, or consumption that would more directly measure the ``extortion tax'' that gangs impose on communities.

\section{Conclusion}

El Salvador's homicide rate fell by more than 80 percent in less than a decade. This paper shows that the decline was geographically targeted: municipalities with the deepest gang presence experienced the largest reductions. The finding implies that whatever combination of policies produced the decline---enforcement, negotiation, deterrence---it reached the places that suffered most from gang violence. For other countries facing extreme violence from organized crime, the El Salvador case demonstrates that rapid, geographically concentrated improvements are possible, though the sustainability of such gains remains an open question.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
