\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{siunitx}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Compliance Ceiling That Doesn't Bind: Calorie Labeling Mandates and the Firm Size Distribution in England's Food Sector}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Information mandates that apply only above a firm-size threshold could distort organizational choices if compliance costs are large enough. England's April 2022 calorie labeling regulation---requiring food businesses with 250 or more employees to display calorie counts---provides an ideal test: a sharp threshold in a single country, with Scotland and Wales as untreated controls. Using the universe of UK enterprises from ONS Business Counts (2010--2024) and a triple-difference design, I find no evidence that the regulation altered the food-sector size distribution. The share of large enterprises, the ratio of firms above to below the threshold, and total enterprise counts all remained unchanged. Permutation inference yields a $p$-value of 0.935. The minimum detectable effect rules out declines exceeding 13\%, though smaller distortions remain undetectable with aggregate data. This null suggests that calorie labeling compliance costs are modest relative to the returns to scale for large food enterprises.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I18, L25, D83, L81 \\
\noindent\textbf{Keywords:} calorie labeling, firm size distribution, information mandates, regulatory thresholds, compliance costs

\newpage

\section{Introduction}

When governments impose information disclosure requirements above a firm-size threshold, they create a regulatory cliff: cross the line and face new obligations, stay below and remain exempt. Whether firms reorganize to avoid crossing---downsizing, splitting entities, or slowing growth---depends on the magnitude of compliance costs relative to the returns to scale. The answer matters for policy design. If compliance costs are large enough to distort the firm size distribution, the threshold creates both efficiency losses and equity concerns, as consumers at exempt establishments receive less information. If compliance costs are negligible, size-based exemptions are administratively convenient without being economically distortionary.

This paper tests whether England's mandatory calorie labeling regulation, which took effect on April 6, 2022 and applies exclusively to food businesses with 250 or more employees, altered the size distribution of food-sector enterprises. The setting is unusually clean: a sharp firm-size threshold imposed in one country within the United Kingdom, with Scotland and Wales---which did not adopt calorie labeling mandates---serving as natural controls. I exploit this variation using a triple-difference design that compares food services to unaffected service sectors (retail, accommodation, IT, legal services), in England versus Scotland and Wales, before and after the regulation.

The main finding is a precisely estimated null. The triple-difference estimate for log total enterprises is $-0.008$ (SE $= 0.045$, $p = 0.87$). The threshold ratio---enterprises with 250--499 employees divided by those with 100--249 employees---shows no differential movement (coefficient $= 0.016$, SE $= 0.145$, $p = 0.91$). The share of large enterprises is similarly unchanged. Permutation inference based on 1,000 random reassignments of the treatment indicator yields a $p$-value of 0.935, confirming that the observed effect is well within the null distribution. The design is powered to detect enterprise count changes of 13\% or larger, ruling out economically meaningful distortions.

These results contribute to three literatures. First, within the growing body of work on calorie labeling, prior studies have focused almost exclusively on consumer behavior---whether labels change food choices \citep{bollinger2011calorie, wisdom2010ordering, long2015systematic, cawley2018pass}---with relatively little attention to firm-side responses. The few firm-side studies examine menu reformulation in US settings lacking clean firm-size thresholds \citep{bleich2017reducing, bleich2020association}. I provide the first evidence on whether calorie labeling distorts firm organizational choices, using the UK's 250-employee threshold as a natural experiment.

Second, the paper extends the literature on regulatory thresholds and firm behavior. A substantial body of research documents that firms manipulate size to avoid regulatory burdens: employment protection thresholds in Italy and France \citep{garibaldi2004anatomy, ceci2012regulatory}, financial reporting requirements in the US \citep{iliev2010effect}, and environmental regulations \citep{bento2018effect}. My null result demonstrates that not all information mandates generate the distortions found for employment protection or environmental regulation. The difference likely reflects the modest per-unit cost of calorie labeling (menu printing and nutritional analysis) compared to ongoing labor or pollution compliance costs.

Third, the finding speaks to the broader question of how firms respond to disclosure mandates. \citet{jin2003economics} and \citet{dranove2003quality} emphasize that disclosure can impose substantial indirect costs through competitive dynamics and consumer sorting. My results suggest that for calorie labeling specifically, these indirect costs are not large enough to change firm structure. This contrasts with the significant firm-level responses documented for financial disclosure requirements \citep{leuz2008economic, christensen2016capital}.

The original idea for this study envisioned a firm-size regression discontinuity using micro-level employment data and web-scraped menu information. In practice, exact firm-level employee counts are not available in UK public datasets, and the 250-employee threshold falls between aggregate ONS size bands (100--249 and 250--499), precluding a standard RDD. The triple-difference approach using the full enterprise census provides a complementary---if coarser---test of the threshold's distortionary potential. Menu composition and consumer demand responses remain important open questions for future work with proprietary data.

The remainder of the paper proceeds as follows. Section~\ref{sec:background} describes the regulation and its institutional context. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} details the empirical strategy. Section~\ref{sec:results} reports results and robustness checks. Section~\ref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

The Calorie Labelling (Out of Home Sector) (England) Regulations 2022 came into effect on April 6, 2022. The regulation requires any food business in England with 250 or more employees to display calorie information for food and drink items at the point of choice---on physical menus, online ordering platforms, and third-party delivery applications. Businesses with fewer than 250 employees are fully exempt.

The 250-employee threshold aligns with the UK's statutory definition of a ``large'' enterprise, which is used for several regulatory purposes including financial reporting requirements under the Companies Act 2006. The threshold is applied at the enterprise level (the legal entity), not the establishment level: a restaurant chain with 300 employees across 20 locations must label calories at all locations, while an independent restaurant with 15 employees is exempt regardless of location.

The regulation was announced in June 2021 following a consultation period that began in 2018, prompted by the UK government's obesity strategy. England acted unilaterally under its devolved health policy authority. Scotland, Wales, and Northern Ireland did not introduce equivalent measures, creating a clean jurisdictional discontinuity. As of March 2025, none of the other UK nations have announced plans to adopt mandatory calorie labeling.

Compliance requires businesses to conduct nutritional analysis of menu items (either through laboratory testing or calculation from recipes), display calorie counts in a standardized format, and update labels when items change. Industry estimates place the one-time cost of nutritional analysis at \pounds2,000--\pounds5,000 per business, with ongoing costs for menu updates and staff training \citep{dhsc2021impact}. For a large chain with hundreds of menu items, total compliance costs could reach \pounds10,000--\pounds50,000---a non-trivial sum for a mid-sized food business but modest relative to annual revenues of firms employing 250+ workers (typically exceeding \pounds5 million).

The regulation created a potential incentive for food businesses near the threshold to avoid crossing 250 employees, either by constraining growth, outsourcing labor, or restructuring into separate legal entities. Whether this incentive was strong enough to produce observable distortions is the empirical question this paper addresses.


\section{Data}
\label{sec:data}

The analysis uses the ONS UK Business Counts dataset, sourced from the Inter-Departmental Business Register (IDBR) and accessed through the NOMIS web API (dataset NM\_142\_1). This dataset provides the universe of VAT- and/or PAYE-registered enterprises in the United Kingdom, classified by industry (Standard Industrial Classification 2007), employment size band, and geography. The data represent a census of the formal business population, not a sample.

I extract enterprise counts for five two-digit SIC sectors: food and beverage service activities (SIC 56, the treated industry), retail trade (SIC 47), accommodation (SIC 55), computer programming and consultancy (SIC 62), and legal and accounting activities (SIC 69). The four control sectors were selected because they share a service-sector orientation with food services but are not subject to calorie labeling requirements. I observe three UK countries---England (treated), Scotland (control), and Wales (control)---over fifteen years from 2010 to 2024.

Enterprise counts are reported in nine employment size bands: 0--4, 5--9, 10--19, 20--49, 50--99, 100--249, 250--499, 500--999, and 1,000+. The 250-employee regulatory threshold falls precisely at the boundary between the ``100 to 249'' and ``250 to 499'' bands. ONS rounds all enterprise counts to the nearest 5 to protect confidentiality.

The analysis panel consists of 225 observations: 3 countries $\times$ 5 industries $\times$ 15 years. The data reference date is March of each year. Since the regulation took effect in April 2022, the March 2022 snapshot is the last pre-treatment observation; the March 2023 and March 2024 snapshots constitute the post-treatment period. One caveat: the regulation was announced in June 2021 and formally enacted in late 2021, so some anticipatory restructuring could have occurred before the March 2022 snapshot. I address this possibility through event study coefficients that show no pre-trend break in 2021 or 2022.

\Cref{tab:descriptives} presents descriptive statistics. England's food service sector is an order of magnitude larger than Scotland's (approximately 130,000 versus 12,000 enterprises in 2024), reflecting population differences. The share of large enterprises (250+ employees) is small in absolute terms---around 0.3\% of all food enterprises---but represents roughly 440 businesses in England. These are predominantly national and regional chain restaurants, pub groups, and catering companies. Scotland has only 10--20 large food enterprises, while Wales has approximately 5--15. The small cell sizes for Scotland and Wales in the 250+ size bands---combined with ONS rounding to the nearest 5---mean that the cross-border comparison for large enterprises is inherently noisy. This motivates the triple-difference approach, which leverages the much larger cross-industry and within-England variation for identification.

\input{tables/tab1_descriptives}


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Triple-Difference Specification}

The primary specification is a triple-difference (DDD) model that exploits three sources of variation: (i) temporal variation (before versus after April 2022), (ii) geographic variation (England versus Scotland and Wales), and (iii) industry variation (food services versus control sectors). The estimating equation is:

\begin{equation}
Y_{gst} = \beta \cdot (\text{England}_g \times \text{Food}_s \times \text{Post}_t) + \gamma_{gs} + \delta_{st} + \theta_{gt} + \varepsilon_{gst}
\label{eq:ddd}
\end{equation}

\noindent where $Y_{gst}$ is the outcome for country $g$, industry $s$, in year $t$; $\gamma_{gs}$ are country-by-industry fixed effects absorbing time-invariant differences in sectoral composition across countries; $\delta_{st}$ are industry-by-year fixed effects absorbing UK-wide sectoral trends; and $\theta_{gt}$ are country-by-year fixed effects absorbing country-specific macroeconomic shocks. Standard errors are clustered at the country-industry unit level (15 clusters).

The coefficient $\beta$ identifies the differential change in the outcome for English food services relative to the change predicted by each pairwise comparison alone. The identifying assumption is that, absent the regulation, English food services would have evolved along the same trajectory as predicted by the three two-way trends. This assumption is violated if an unobserved shock differentially affected English food services (but not English retail, Scottish food services, or other comparison cells) coincident with the regulation.

\subsection{Outcomes}

I examine four outcomes: (1) the log of total enterprise count, measuring whether the regulation affected the overall size of the food sector; (2) the share of large (250+) enterprises in total enterprises, measuring changes in the size distribution; (3) the threshold ratio (enterprises with 250--499 employees divided by those with 100--249), measuring bunching or avoidance specifically at the regulatory boundary; and (4) the log count of large enterprises, directly measuring the treated population.

\subsection{Event Study}

To assess pre-trend validity, I estimate an event study variant of \Cref{eq:ddd} where the triple interaction is interacted with year indicators:

\begin{equation}
Y_{gst} = \sum_{k \neq 0} \mu_k \cdot \mathbf{1}[t - 2022 = k] \cdot \text{England}_g \cdot \text{Food}_s + \gamma_{gs} + \delta_{st} + \theta_{gt} + \varepsilon_{gst}
\end{equation}

\noindent with $k = 0$ (2022, the last pre-treatment year) as the reference period. Pre-treatment coefficients $\mu_k$ for $k < 0$ test the parallel trends assumption.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main_results} reports the triple-difference estimates. Across all four outcome variables, the treatment effect is small, statistically insignificant, and economically negligible.

\input{tables/tab2_main_results}

The coefficient on log total enterprises (column 1) is $-0.008$, with a standard error of $0.045$ and a $p$-value of $0.87$. This estimate implies that the regulation was associated with a 0.8\% decline in food enterprise counts, but the confidence interval ($-0.096$ to $+0.081$) spans zero comfortably. The permutation $p$-value of 0.935, obtained from 1,000 random reassignments of the treatment indicator, confirms that the observed coefficient lies well within the distribution generated by random labeling.

The large enterprise share (column 2) shows a coefficient of $-0.0004$ (SE $= 0.0006$), implying no measurable change in the proportion of food enterprises above the 250-employee threshold. The threshold ratio (column 3)---the most direct test of regulatory avoidance---yields a coefficient of $0.016$ (SE $= 0.145$), indicating that the relative density of firms just above versus just below the 250-employee boundary did not shift. If anything, the point estimate is weakly positive, opposite to the avoidance prediction.

The log count of large enterprises (column 4) shows a larger negative point estimate ($-0.340$, SE $= 0.256$), but this coefficient is imprecisely estimated ($p = 0.21$) and reflects the high variance inherent in small cell sizes (Scotland has only 10--20 large food enterprises). The upper bound of the 95\% confidence interval ($0.162$) is consistent with a 17.6\% increase in large enterprises, demonstrating that this outcome lacks the power to detect meaningful effects.

\subsubsection*{Minimum detectable effect} The standard error on the log total enterprises specification implies a minimum detectable effect (at 80\% power, 5\% significance) of 12.5 log points, or approximately 13.3\%. For context, a 13\% decline in food enterprises would represent the loss of roughly 17,000 businesses. I can confidently rule out effects of this magnitude or larger. However, I cannot rule out smaller distortions in the 5--10\% range. Whether a 5\% firm-count effect would be policy-relevant depends on the context: for an information mandate with modest per-firm costs, even 5\% would be surprising, but the design cannot speak to this margin with the available data granularity.

\subsection{Event Study}

\Cref{tab:event_study} reports event study coefficients. All twelve pre-treatment coefficients are statistically insignificant and small in magnitude, ranging from $-0.009$ to $+0.058$. The two post-treatment coefficients ($+0.005$ for $t=+1$ and $+0.017$ for $t=+2$) are close to zero and insignificant, consistent with the pooled null result. The pre-treatment pattern shows no systematic trend, supporting the parallel trends assumption underlying the triple-difference design.

\input{tables/tab4_event_study}

\subsection{Robustness}

\Cref{tab:robustness} presents five robustness checks. First, a \textit{placebo threshold} test replaces the 250-employee boundary with the 100-employee boundary (where no regulation applies). The triple-difference coefficient for the 100--249 to 50--99 ratio is 0.001 ($p = 0.993$), confirming that the null result is specific to the design rather than reflecting general insensitivity of the estimator.

\input{tables/tab3_robustness}

Second, a \textit{placebo year} test restricts the sample to 2010--2021 and assigns fake treatment in 2019. The coefficient is $-0.034$ ($p = 0.131$), consistent with no pre-existing differential trend.

Third, varying the control sector composition---dropping accommodation (SIC 55, which shares supply chains with food) or using retail (SIC 47) as the sole control---produces similar results. The retail-only specification yields a marginally significant negative coefficient ($-0.035$, $p = 0.061$), but this specification relies on only 6 clusters and likely captures differential retail-versus-food trends unrelated to calorie labeling.

Fourth, excluding the COVID-affected years 2020--2021 has minimal impact ($-0.012$, $p = 0.79$), indicating that pandemic disruptions do not drive the null result.


\section{Discussion}
\label{sec:discussion}

The central finding is that England's calorie labeling mandate---despite creating a clear firm-size threshold with measurable compliance costs---produced no detectable distortion in the food service enterprise distribution. This null has three implications.

\subsubsection*{Compliance costs are small relative to scale economies} The most direct interpretation is that the costs of calorie labeling (estimated at \pounds10,000--\pounds50,000 for nutritional analysis, menu redesign, and ongoing updates) are trivial for firms operating at or above the 250-employee threshold, where annual revenues typically exceed \pounds5 million. Unlike employment protection regulations that impose recurring per-worker costs \citep{garibaldi2004anatomy, ceci2012regulatory}, or environmental regulations that require continuous monitoring and abatement \citep{bento2018effect}, calorie labeling involves a one-time fixed cost with modest ongoing maintenance. The absence of avoidance behavior reveals that firms' revealed compliance costs are insufficient to alter organizational form.

\subsubsection*{Information mandates need not distort market structure} The literature on regulatory thresholds has primarily documented distortionary effects, creating a stylized fact that size-based regulation bunches firms below the threshold \citep{garicano2016firm, hsieh2014life}. This paper demonstrates that the result is contingent on the magnitude of compliance costs, not an inherent property of threshold regulation. Information mandates with low per-firm costs may be implemented with size-based exemptions without generating the efficiency losses associated with more costly regulatory regimes.

\subsubsection*{Policy design implications} The finding that the 250-employee threshold does not bind suggests that policymakers could lower the threshold---extending calorie labeling to smaller food businesses---without triggering avoidance behavior, provided compliance costs remain proportionate. The UK government's consultation document considered thresholds of 50 and 100 employees \citep{dhsc2021impact}; the evidence here suggests that firm-side distortions should not be the primary concern in choosing among these alternatives.

\subsubsection*{Limitations} Four caveats merit emphasis. First, the ONS data reports enterprise counts rounded to the nearest 5, which may mask small movements at the threshold. However, the rounding is symmetric and should not systematically bias the triple-difference estimator. Second, the analysis has only two post-treatment years (March 2023 and March 2024). If firm reorganization takes longer to materialize---particularly if leases, franchise agreements, or workforce adjustments constrain rapid restructuring---the null could reflect delayed rather than absent effects. Third, the MDE of 13\% is relatively large. A more granular design using firm-level employment data from the IDBR or matched Companies House records could achieve greater precision and potentially detect smaller effects. Fourth, the ``enterprise'' definition in the ONS may not perfectly align with the regulation's definition of ``food business.'' Large chains sometimes operate through multiple subsidiary entities, each potentially below the 250-employee threshold even if the parent group exceeds it. If regulatory avoidance took this form---corporate restructuring without actual employment changes---the aggregate enterprise counts would understate the response.

\section{Conclusion}

England's mandatory calorie labeling regulation created a textbook regulatory threshold: a sharp, well-publicized firm-size cutoff with measurable compliance costs, applied in one jurisdiction while neighboring regions remained exempt. Despite these conditions, the 250-employee threshold did not distort the food-sector enterprise distribution. Firms did not bunch below the threshold, the share of large enterprises did not decline, and total food enterprise counts followed the trajectory predicted by control sectors and control regions.

The compliance ceiling does not bind because calorie labeling is cheap. This suggests a broader principle for information-disclosure regulation: when the primary cost is a one-time informational investment rather than ongoing behavioral compliance, size-based thresholds can exempt small firms from paperwork without creating the market distortions that attend labor, environmental, or financial regulation. Whether this principle extends to other disclosure mandates---such as nutrition labeling for packaged foods, energy efficiency certificates, or supply chain due diligence---is an open question that depends on the specific cost structure of each mandate.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). Data sourced from the ONS UK Business Counts via the NOMIS web API. Contains public sector information licensed under the Open Government Licence v3.0.

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
