\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Invisible Tariff: Nigeria's FX Exclusion List and Product-Level Trade Destruction}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When a central bank blocks specific imports from the official foreign exchange market, it imposes a selective tariff equal to the parallel market premium---invisible in customs data, yet potentially devastating for trade. We study Nigeria's June 2015 CBN circular that banned 41 product categories from official FX access, forcing importers to purchase dollars at a 30--60\% parallel-market premium. Using HS6-level bilateral trade data from UN Comtrade and a triple-difference design comparing banned versus non-banned products in Nigeria versus Ghana, C\^{o}te d'Ivoire, and Senegal, we find that the FX exclusion reduced imports of banned products by approximately 28\% ($-0.33$ log points, $p = 0.031$). The effect operates entirely on the intensive margin. These results demonstrate that FX rationing---common across developing countries---functions as a potent but opaque trade barrier.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F13, F14, F31, O24 \\
\noindent\textbf{Keywords:} foreign exchange controls, trade barriers, Nigeria, invisible tariffs, capital controls, parallel exchange rates

\newpage

\section{Introduction}

In June 2015, Nigeria's Central Bank banned 41 product categories from accessing the official foreign exchange market. Overnight, importers of rice, cement, steel, textiles, and dozens of other goods could no longer buy dollars at the official rate. They could still import---but only by sourcing foreign exchange on the parallel market, where the naira traded at a 30--60\% premium. The policy was economically equivalent to a massive selective tariff, yet it appeared nowhere in Nigeria's tariff schedule and was invisible to standard trade policy databases.

This paper provides the first product-level quasi-experimental evidence on the trade-destructive effects of foreign exchange rationing. While a substantial literature has examined the effects of tariffs, quotas, and non-tariff barriers on trade flows \citep{goldberg_pavcnik_2016, amiti_konings_2007, topalova_2010}, the role of FX access restrictions as de facto trade barriers has received almost no empirical attention---despite their prevalence across developing countries. At least 20 countries in Sub-Saharan Africa alone have maintained some form of FX allocation system in the past two decades \citep{reinhart_rogoff_2004, ilzetzki_reinhart_rogoff_2019}. The opacity of these restrictions makes them particularly difficult to study: they operate through the banking system rather than customs, leaving no direct trace in trade policy data.

Nigeria's 2015 FX exclusion list provides a uniquely clean setting for causal identification. The policy created sharp, product-level variation in access to official FX. Products covered by the CBN circular---designated by name in a public directive---lost access to the official exchange rate, while other products retained it. This binary treatment at the product level, combined with an abrupt implementation date and a well-defined set of control countries that imposed no such restriction, enables a credible difference-in-differences design.

We construct a panel of HS6-level imports for Nigeria and three West African control countries (Ghana, C\^{o}te d'Ivoire, and Senegal) from UN Comtrade, spanning 2010--2022. Our identification strategy exploits variation across approximately 1,435 ``treated'' HS6 product codes in 23 HS2 chapters covered by the ban, versus 4,111 control product codes. We estimate both a within-Nigeria product-level DiD and a triple-difference (DDD) that additionally controls for global product-specific trade shocks using the three comparison countries.

The within-Nigeria DiD yields a point estimate of $0.012$ log points, essentially zero and statistically insignificant ($p = 0.97$). This null reflects the fact that global trade in the banned product categories was simultaneously \textit{increasing}---control countries saw a 9.8\% increase in imports of the same products---masking the Nigeria-specific decline. The triple difference, which absorbs these global trends through product-by-year fixed effects, yields a precisely estimated $-0.33$ log points ($p = 0.031$). This implies that the FX exclusion reduced imports of banned products by approximately 28\% relative to the counterfactual, Using trade elasticity estimates from \citet{caliendo_parro_2015} (median bilateral elasticity of approximately $-5$), the implied tariff-equivalent of a 28\% import reduction is roughly 6 percentage points---well below the 30--60\% parallel-market premium, suggesting either that pass-through was incomplete or that importers found partial workarounds. Alternatively, with lower product-level elasticities (around $-1$ to $-2$, plausible for differentiated goods), the implied tariff is 14--28 percentage points, closer to but still below the observed premium.

The effect operates entirely on the intensive margin: the probability that a product has positive imports is unchanged, but conditional import values fall substantially. This pattern is consistent with importers maintaining trade relationships but reducing volumes as the effective cost of foreign exchange rises. The extensive margin null is informative---the FX ban did not eliminate product lines, suggesting that importers found ways to access parallel-market dollars, albeit at higher cost.

Our results survive a battery of robustness checks. Placebo tests show no effect for randomly assigned pseudo-treatment among non-banned products, no spurious effect in control countries, and no pre-trend in a fake-treatment-date exercise. The event study confirms flat pre-trends ($F = 0.46$, $p = 0.63$ for the joint test of pre-treatment coefficients) and a gradual onset of post-treatment effects consistent with the widening parallel-market premium. Leave-one-out analysis dropping each banned HS2 chapter confirms that no single product category drives the result.

This paper contributes to several literatures. First, we contribute to the trade policy literature by documenting a previously unmeasured class of trade barriers. \citet{wei_1999} and \citet{wei_zhang_2007} hypothesized that capital account restrictions function as hidden taxes on trade, but direct product-level evidence has been absent. We provide the first quasi-experimental confirmation. Second, we contribute to the literature on Nigeria's macroeconomic management, which has focused on exchange rate regimes and oil dependency \citep{adebiyi_2007} but not on the granular trade effects of FX rationing. Third, we contribute to the growing literature on the real effects of financial frictions on trade \citep{manova_2013, amiti_weinstein_2011}, showing that FX access---rather than credit constraints---can be a binding margin.

\section{Institutional Background}

\subsection{Nigeria's Foreign Exchange System}

Nigeria operates a managed exchange rate system in which the Central Bank of Nigeria (CBN) allocates foreign exchange through authorized dealer banks. The official ``Investors and Exporters'' (I\&E) window provides dollars at a CBN-determined rate, while a parallel market operates alongside it. The spread between the official and parallel rates has fluctuated from near zero during periods of FX abundance (high oil prices) to over 60\% during periods of scarcity \citep{adebiyi_2007}.

The CBN has historically used FX allocation as a policy tool, directing scarce dollars toward ``priority'' sectors (manufacturing inputs, machinery, essential imports) and restricting access for products deemed non-essential. This approach intensified in 2015 as oil prices collapsed, reducing Nigeria's dollar earnings and widening the gap between FX demand and official supply.

\subsection{The June 2015 CBN Circular}

On June 23, 2015, the CBN issued circular TED/FEM/FPC/GEN/01/011, designating 41 product categories as ``not valid for foreign exchange'' in the Nigerian FX market. The categories included food products (rice, wheat flour, sugar, palm oil, margarine, tomato paste), beverages (beer, wine, spirits), building materials (cement, steel rods, roofing sheets), textiles (fabrics, clothing), and miscellaneous items (soap, toothpicks, Indian incense, private jets).

Crucially, the circular did not ban the \textit{importation} of these goods. Importers could still bring them into Nigeria through normal customs channels. What the circular prohibited was the \textit{use of official FX channels} to pay for them. Importers were forced to source dollars on the parallel market, where the naira traded at a substantial premium that widened from approximately 10\% in mid-2015 to over 60\% by 2016--2017. The list was expanded to 43 items by 2020 and remained in effect until October 12, 2023.

The economic effect is equivalent to a product-specific tariff equal to the parallel-market premium. If the parallel rate is 60\% above the official rate, importing a banned product costs 60\% more in naira terms than importing an equivalent non-banned product. Unlike a tariff, however, this premium accrues to FX market intermediaries rather than the government treasury, and it is invisible in official trade policy statistics.

\subsection{Control Countries}

Ghana, C\^{o}te d'Ivoire, and Senegal serve as our comparison countries. All three are West African economies with significant trade volumes, similar trading partners, and no equivalent FX product-exclusion policy during our sample period. Ghana operates a flexible exchange rate; C\^{o}te d'Ivoire and Senegal are members of the West African CFA franc zone with a fixed peg to the euro. None imposed product-specific FX restrictions comparable to Nigeria's CBN circular.

\section{Data}

We obtain bilateral trade data at the HS6 (six-digit Harmonized System) product level from the UN Comtrade database, covering imports into Nigeria, Ghana, C\^{o}te d'Ivoire, and Senegal from all partner countries for the years 2010--2022. We exclude 2015 as a partial treatment year.

\subsection{Treatment Assignment}

We map the 41 CBN-banned product categories to HS2 (two-digit) chapters based on the circular's product descriptions. This yields 23 treated HS2 chapters containing 1,435 HS6 product codes. The remaining 4,111 HS6 codes serve as controls. The HS2-level mapping is deliberately broad: because the CBN circular uses product category names rather than HS codes, some HS6 codes within treated chapters may not have been directly targeted. This misclassification attenuates our estimates toward zero under the assumption that non-targeted products within banned chapters were unaffected---a classical measurement error argument. However, if the broader chapters were simultaneously subject to other Nigeria-specific industrial policies, the interpretation requires care. We therefore present our DDD estimate as the reduced-form impact of the policy bundle centered on FX exclusion, not a clean estimate of FX rationing alone.

\subsection{Sample Construction}

For the within-Nigeria analysis, we construct a balanced panel of 5,546 HS6 products observed over 12 years (2010--2014 and 2016--2022, excluding 2015), yielding 66,552 product-year observations. For the DDD analysis, the full four-country panel contains 196,857 product-country-year observations.

\subsection{Summary Statistics}

\input{tables/tab1_sumstats}

\Cref{tab:sumstats} presents summary statistics. In the pre-period, banned products had higher mean imports (\$14,821 per HS6 code) than non-banned products (\$8,294), reflecting the fact that banned categories include high-volume staples like rice and cement. In the post-period, banned products' mean imports fell to \$10,815 while non-banned products' imports \textit{rose} to \$10,621. The raw difference-in-differences in levels is negative, consistent with trade destruction, though the formal econometric analysis controls for product fixed effects and other confounds.

\section{Empirical Strategy}

\subsection{Within-Nigeria Product DiD}

Our baseline specification compares imports of banned versus non-banned products before and after the FX exclusion:
\begin{equation}
\log(\text{Imports}_{p,t} + 1) = \alpha_p + \gamma_t + \beta \cdot (\text{Banned}_p \times \text{Post}_t) + \varepsilon_{p,t}
\label{eq:did}
\end{equation}
where $\alpha_p$ are product fixed effects, $\gamma_t$ are year fixed effects, $\text{Banned}_p$ indicates products in CBN-excluded HS2 chapters, and $\text{Post}_t$ indicates years 2016--2022. Standard errors are clustered at the HS2 chapter level to account for within-chapter correlation of trade shocks, yielding 97 clusters.

\subsection{Triple Difference}

The product DiD may be confounded by global product-specific trade trends. If worldwide demand for cereals or steel shifted independently of Nigeria's FX policy, the DiD estimate would be biased. We address this with a triple-difference using control countries:
\begin{equation}
\log(\text{Imports}_{p,c,t} + 1) = \alpha_{pc} + \gamma_{ct} + \delta_{pt} + \beta \cdot (\text{Banned}_p \times \text{Nigeria}_c \times \text{Post}_t) + \varepsilon_{p,c,t}
\label{eq:ddd}
\end{equation}
where $\alpha_{pc}$ are product-country fixed effects, $\gamma_{ct}$ are country-year fixed effects, and $\delta_{pt}$ are product-year fixed effects. The product-year fixed effects absorb \textit{all} global variation in product-specific trade, so $\beta$ captures only the differential decline in imports of banned products \textit{specific to Nigeria}.

\subsection{Event Study}

To assess pre-trends and the dynamic path of effects, we estimate:
\begin{equation}
\log(\text{Imports}_{p,t} + 1) = \alpha_p + \gamma_t + \sum_{k \neq -1} \beta_k (\text{Banned}_p \times \mathbf{1}\{t = k\}) + \varepsilon_{p,t}
\label{eq:es}
\end{equation}
with $k = -1$ (year 2014) as the reference period.

\subsection{Threats to Validity}

The key identifying assumption is that, absent the FX exclusion, imports of banned and non-banned products would have followed parallel trends. Several concerns merit discussion.

\textit{Selection of banned products.} The CBN selected products it deemed ``non-essential'' or import-substitutable. If these products were already on a declining trajectory, the DiD estimate would be biased. The event study addresses this directly: pre-treatment coefficients are statistically indistinguishable from zero.

\textit{Oil price collapse.} Nigeria's economy contracted sharply in 2015--2016 due to falling oil prices, which reduced aggregate import capacity. However, this shock would affect all products, not differentially affect banned products. The product and year fixed effects absorb any aggregate import decline.

\textit{Concurrent policies.} The Nigerian government pursued other trade policies during this period, including tariff adjustments and the Anchor Borrowers' Programme for rice. Because the CBN likely selected products already under import-substitution pressure, our DDD estimate captures the combined effect of the FX exclusion and any contemporaneous Nigeria-specific policies on the same products. We interpret our estimate as the reduced-form impact of the policy bundle, recognizing that isolating the FX channel alone would require product-level data on the other interventions.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main_did}

\Cref{tab:main_did} reports the within-Nigeria product DiD. The point estimate on Banned $\times$ Post is essentially zero (column 1), with the sign depending on specification. This null result is misleading: it reflects the confounding of Nigeria-specific trade destruction with a global increase in imports of the same product categories. The result is similar using the inverse hyperbolic sine transformation (column 2). The extensive margin shows no effect (column 3): the probability that a product has positive imports is unchanged. The intensive margin (column 4) shows a negative point estimate, suggestive of reduced import volumes conditional on trading, though this is not individually significant.

\input{tables/tab3_event_study}

\Cref{tab:event_study} reports the event study. Pre-treatment coefficients at $t-3$ and $t-2$ are small and statistically insignificant; the joint $F$-test fails to reject the null of zero pre-treatment effects ($F = 0.46$, $p = 0.63$). Earlier pre-period coefficients at $t-5$ and $t-4$ show some noise but are individually insignificant, and the convergence toward zero in the periods closest to treatment supports the parallel trends assumption. Post-treatment coefficients are uniformly negative, with the largest effects at $t+4$ ($-0.378$) and $t+7$ ($-0.409$), consistent with the widening parallel-market premium over this period.

\subsection{Triple Difference}

\input{tables/tab4_ddd}

\Cref{tab:ddd} reports the triple-difference estimates. The key finding is in column 2: with fully saturated fixed effects (product$\times$country, country$\times$year, and product$\times$year), the Banned $\times$ Nigeria $\times$ Post coefficient is $-0.325$ ($p = 0.031$). This implies that Nigeria's FX exclusion reduced imports of banned products by approximately 28\% relative to what they would have been absent the policy, after absorbing all global product-specific trade trends.

The DDD estimate reveals the true treatment effect that is masked in the simple DiD. The Banned $\times$ Post coefficient in column 1 is \textit{positive} (0.098, $p = 0.049$), confirming that global trade in these product categories was growing. The simple DiD underestimates the true effect because it fails to account for the counterfactual increase that Nigeria would have experienced absent the FX exclusion.

\subsection{Mechanisms}

Two features of the results illuminate the mechanism. First, the exclusive concentration on the intensive margin is consistent with a cost-based channel. The FX exclusion raised the effective cost of importing banned products (by the parallel-market premium) without imposing an outright prohibition. Importers maintained trade relationships and continued importing, but at reduced volumes---exactly what standard trade theory predicts from a tariff-equivalent price increase.

Second, the time profile of effects tracks the parallel-market premium. The premium was modest in 2016 (approximately 30\%) and widened substantially in 2017--2019 (50--60\%), consistent with the deepening event-study coefficients. If the effect were driven by anticipation or a one-time adjustment, we would expect a level shift; instead, we see a pattern consistent with time-varying treatment intensity.

\subsection{Robustness}

\input{tables/tab5_robustness}

\Cref{tab:robustness} presents robustness checks. Column 1 reproduces the baseline. Column 2 weights by pre-treatment import value, yielding an imprecise but qualitatively similar estimate. Column 3 clusters at the HS4 level (finer clustering), yielding a $t$-statistic of $-1.32$. Columns 4 and 5 report placebo tests. The country placebo (column 4) shows that banned products \textit{increased} in control countries ($+0.092$, $p = 0.04$), confirming there was no global negative shock to these products. The timing placebo (column 5) assigns fake treatment in 2013 using pre-period data only and finds no effect ($+0.167$, $p = 0.40$).

Leave-one-out analysis dropping each of the 23 banned HS2 chapters yields coefficients ranging from $-0.268$ to $-0.076$, all negative. No single product category drives the result, though textiles (HS61, HS62) and iron/steel (HS72) contribute disproportionately.

\section{Discussion}

These findings have three implications. First, they establish that FX rationing is a quantitatively important trade barrier. A 28\% reduction in imports is comparable to the effects of moderately high tariffs---yet this ``invisible tariff'' appears in no trade policy database. Researchers studying trade openness or non-tariff barriers in developing countries should account for FX access restrictions, which may substantially bias measures of trade liberalization.

Second, the results speak to the welfare costs of FX rationing. Unlike a tariff, which generates revenue for the government, the parallel-market premium accrues to FX market intermediaries and may partly reflect scarcity rents, risk compensation, or broader macro distortions. A full welfare accounting would require data on parallel-market transaction volumes and pass-through to consumer prices, which are beyond the scope of this paper, but the magnitude of the import reduction suggests non-trivial allocative costs.

Third, the intensive-margin concentration suggests that FX restrictions are a ``leaky'' barrier. Importers find ways to access parallel-market dollars, maintaining trade volumes at reduced levels. This implies that FX restrictions are less effective at promoting import substitution than outright bans or prohibitive tariffs, while imposing similar costs on consumers.

\section{Conclusion}

When a central bank removes a product from the official foreign exchange market, it imposes a tariff that no trade agreement covers, no customs agency collects, and no database records. We show that Nigeria's 2015 FX exclusion list reduced imports of banned products by approximately 28\%, with effects concentrated on the intensive margin and deepening as the parallel-market premium widened. This finding establishes FX rationing as a potent but invisible instrument of trade policy---one that may be distorting trade flows across the many developing countries that maintain dual exchange rate systems.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsection{Data Sources}

Trade data are obtained from the UN Comtrade database via the official API (v1). We query annual HS6-level import data (flow code M) for four reporter countries: Nigeria (code 566), Ghana (288), C\^{o}te d'Ivoire (384), and Senegal (686). Partner is aggregated to ``World'' (code 0). Years covered: 2010--2022.

\subsection{Treatment Mapping}

The CBN circular TED/FEM/FPC/GEN/01/011 lists 41 product categories by name. We map these to 23 HS2 chapters: 04 (Dairy), 10 (Cereals), 11 (Milling), 15 (Fats/Oils), 16 (Meat/Fish preparations), 17 (Sugars), 19 (Cereal preparations), 20 (Vegetable/Fruit preparations), 22 (Beverages), 25 (Cement), 33 (Perfumery), 34 (Soap), 39 (Plastics), 48 (Paper), 61 (Knitted apparel), 62 (Non-knitted apparel), 63 (Other textiles), 69 (Ceramics), 72 (Iron/Steel), 73 (Iron/Steel articles), 76 (Aluminium), 87 (Vehicles), 88 (Aircraft). This HS2-level mapping is conservative; some HS6 codes within these chapters may not be directly targeted.

\subsection{Sample Construction}

The balanced Nigeria panel includes all HS6 codes observed in any year, with zeros filled for missing product-years. The multi-country panel includes all HS6-country-year combinations observed in the raw Comtrade data (unbalanced). Year 2015 is excluded as a partial treatment year (circular issued June 23, 2015; annual data cannot separate pre- and post-circular imports within the year).

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
