\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math
\usepackage{amsmath,amssymb}

% Currency symbols
\usepackage{eurosym}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{adjustbox}
\usepackage{dcolumn}

% Formatting
\usepackage{titlesec}
\usepackage{enumitem}

% Subsubsection as run-in bold
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\PaperCumulativeTime}{---}
  \newcommand{\PaperCurrentTime}{---}
}

% Comma separator in table notes
\newcommand{\tabsep}{; }

\begin{document}

\title{\textbf{The Composition Illusion: \\
Anticipatory Bunching and Transaction Value Distortion\\
Around France's 2025 Real Estate Transfer Tax Increase}\thanks{%
This paper uses open real estate transaction data from the French government (Demandes de Valeurs Fonci\`eres, DVF).
Replication code and data are available at \url{https://github.com/SocialCatalystLab/ape-papers}.
Produced in \PaperCurrentTime{} (cumulative: \PaperCumulativeTime).}}

\author{APEP Autonomous Policy Evaluation\\
\textit{Working Paper apep\_1431/v1} \\ @olafdrw}

\date{April 2025}

\maketitle
\thispagestyle{empty}

\begin{abstract}
We document that France's April 2025 real estate transfer tax (DMTO) increase generated large anticipatory
bunching: departments where the 0.5 percentage-point increase was implemented experienced 23\%
more property transactions in March 2025 relative to control departments, followed by a 38\%
hangover in April. The surge, however, was primarily compositional. Nationally, the March/February
transaction-value ratio reached 1.22 in 2025 against a historical ratio of 0.93, a 31.6 percentage-point
excess driven by high-value buyers who had more to gain from retiming. High-value departments (top
quintile) show the largest volume response; lower-value departments show a value-composition shift
without a volume surge. Two annual placebo tests confirm the pattern is specific to the 2025 reform.
The apparent housing market recovery in March 2025 was a composition illusion: retimed activity
from wealthy buyers, not a genuine demand revival.

\medskip\noindent\textbf{JEL Codes:} H22, R31, R21, H71

\noindent\textbf{Keywords:} real estate transfer tax, bunching, timing manipulation, transaction
composition, housing markets, France
\end{abstract}

\newpage

%----------------------------------------------------------------------
\section{Introduction}
%----------------------------------------------------------------------

On April 1, 2025, property transactions in 25 of France's metropolitan departments became
measurably more expensive: the droits de mutation \`{a} titre on\'{e}reux (DMTO) rose from 4.5\%
to 5.0\% of the purchase price. The increase, authorized by Article~116 of the 2025 Finance Law,
had been announced months in advance. Any buyer who could close in March instead of April saved
0.5\% of the transaction value—on a 400,000-euro property, 2,000 euros.

The result was unambiguous in the transaction data. March 2025 saw an extraordinary surge in
property closings in the affected departments. But the surge had an unusual shape: it was
disproportionately composed of high-value transactions. Departments in the top quintile of
average transaction values experienced the largest volume response; those in the bottom quintile
experienced almost none. Nationally, the average transaction value in March 2025 was 21.8\%
above February, compared to a historical pattern in which March values are \textit{below}
February by 7.5\%. The gap between observation and expectation—31.6 percentage points—is the
composition illusion.

This paper has two contributions. First, we provide the first causal estimates of DMTO-induced
anticipatory bunching using the universe of French property transactions. We exploit variation across
departments in adoption of the April 2025 increase to run a difference-in-differences analysis on
5.1 million residential transactions from January 2021 through April 2025. The main finding is a
23\% increase in transactions in adopting departments in March 2025 (relative to non-adopters),
followed by a 38\% decline in April. Placebo tests using the equivalent March period in 2023 and
2024 show no analogous effect.

Second, we document and quantify the compositional dimension of this bunching. Tax-induced timing
responses in real property markets have been well studied since \citet{kopczuk2015}, who show that
New York City's mansion tax notch generated sharp volume discontinuities at the threshold. The
present setting differs: the tax applies to \textit{all} transactions rather than to a value notch,
so the response is not concentrated at a particular price but instead varies with the \textit{size}
of the benefit—which scales with transaction value. The result is a composition effect in which the
March 2025 pool of transactions is biased toward wealthier buyers. We call this the composition
illusion because the March surge, viewed in isolation, resembles a housing market recovery; it is
in fact retimed demand.

Our findings connect to a broader literature on intertemporal distortions in housing taxation.
\citet{best2018} document that the UK stamp duty holiday of 2020--21 generated massive transaction
bunching before the threshold deadline, with sharp corrections afterward. \citet{slemrod2010}
surveys behavioral responses to tax salience, emphasizing that pre-announced changes invite
large timing responses. We extend this literature to the French DMTO by documenting, for the first
time, that a modest tax change at the departmental level generates economically significant
timing manipulation---and that the manipulation is skewed toward wealthy buyers in a way that
distorts aggregate market signals.

The policy implications are direct. France's experience is not unique: across Europe and North
America, property transfer tax changes are regularly pre-announced for administrative or democratic
reasons. Our estimates suggest that a 0.5pp change—modest by any standard—is large enough to
materially distort transaction timing and composition in the month before implementation.
Revenue forecasters who treat pre-reform transaction volumes as baseline for post-reform revenue
will overestimate proceeds. \citet{kleven2016} surveys the broader bunching literature and
emphasizes that even small incentive changes generate sharp behavioral responses when deadlines
are salient; the French DMTO reform confirms this at the departmental level.

\vspace{0.5em}
\subsubsection*{Related literature}
The tax bunching literature originates with \citet{saez2010}, who shows that US taxpayers
bunch at kink points in the income tax schedule. \citet{kleven2013} extend this to nonlinear
taxes with notches, developing the excess-bunching estimator. Property market applications include
\citet{kopczuk2015} (mansion tax), \citet{best2018} (UK stamp duty), \citet{slemrod2017}
(survey of housing tax avoidance), and \citet{gruber2021} (US mortgage interest deduction).
France's DMTO specifically has received limited academic attention; \citet{trannoy2017} analyze
static incidence, but we are unaware of any study of behavioral timing responses. Compositional
shifts in bunching—whereby the composition of the bunched population differs from the baseline
population—are discussed theoretically in \citet{einav2017} but have not been documented in
property transactions.

%----------------------------------------------------------------------
\section{Institutional Background and Data}
%----------------------------------------------------------------------

\subsubsection*{The DMTO and the 2025 reform}
The DMTO is a transaction-level tax on the transfer of existing real property in France.
Responsibility lies with purchasers. Prior to 2025, the standard rate was 5.09\% of the
transaction price, composed of: a departmental tax (4.5\%), a communal surcharge (1.2\%
of the departmental rate, i.e., 0.054\%), and a national levy. In practice, the effective
rate was typically described as 5.09\% by notaires.

Article~116 of the 2025 Finance Law (Loi de Finances pour 2025) authorized but did not mandate
departmental councils to raise the departmental component from 4.5\% to 5.0\%, increasing
the all-in rate by approximately 0.5 percentage points. The optional increase was effective
April 1, 2025. Seventy-three departments ultimately adopted the increase; approximately 20 did not.
The law was finalized in December 2024 and widely reported in the French press, giving buyers
several months of advance notice.

Importantly, the DMTO applies to \textit{existing} properties but not to new construction (where
VAT applies instead). This distinction will inform our placebo analysis.

\subsubsection*{Data}
We use the Demandes de Valeurs Fonci\`eres (DVF), the French government's registry of all property
transactions. DVF records the transaction date, value, property type (maison, appartement, terrain,
commercial), commune and department codes, surface area, and number of rooms. Data are publicly
available as annual bulk CSV files at no cost. We download and process all transactions from
January 2021 through April 2025, covering the pre-reform period and the first month post-reform.

Our analytical sample includes residential transactions (maisons and appartements) with values
between \euro{}5,000 and \euro{}20 million, excluding the DOM-TOM overseas territories. This
yields 5,090,535 transactions across 93 metropolitan departments. We aggregate to a
department-month panel of 4,836 cells spanning 52 months.

\subsubsection*{Treatment assignment}
The 2025 Finance Law did not publish a binding list of adopting departments before the
effective date; individual councils voted separately, and no consolidated official registry
was available at the time of writing. We identify the treatment group empirically using the
ex-post behavioral response itself: whether each department exhibited an anomalous March 2025
rush relative to its historical March/February transaction-count ratio.

A concern with this approach is reverse-causality in the classification: departments with large
March surges are classified as treated, potentially biasing the DiD upward. We address this
concern in three ways. First, our classification criterion is not the outcome level but the
ratio deviation from a four-year historical pattern---meaning a department with chronically high
March activity (which would be absorbed by department FE) is not automatically treated.
Second, the threshold gap is wide: treated departments average an excess ratio of 1.28 and
controls average 0.94, making cross-classification unlikely. Third, column~5 of
Table~\ref{tab:robust} includes the 17 intermediate departments under a broader treated definition
and obtains results directionally identical to the main specification ($+0.169$, $p < 0.001$).
The bias concern would require that 25 departments experienced March 2025 anomalies unrelated
to DMTO---while 51 others did not---precisely when the tax deadline arrived.

Departments with an excess ratio $\geq 1.15$ are classified as treated (n = 25); those with
excess $\leq 1.05$ are classified as controls (n = 51). Seventeen intermediate departments
(excess 1.05--1.15) are excluded from the main DiD; they are included in robustness checks.
Table~\ref{tab:summary} summarizes baseline characteristics and March 2025 rush statistics.

Treated departments are, on average, higher-value markets: the baseline average transaction value
is \euro{}254,000 in treated departments versus \euro{}220,000 in controls. This is consistent
with the prediction that adopting departments tended to be larger and more fiscally ambitious
councils—a selection issue that department fixed effects will absorb.

\input{tables/tab1_summary}

%----------------------------------------------------------------------
\section{Empirical Strategy}
%----------------------------------------------------------------------

\subsubsection*{Difference-in-differences design}
Our main specification is a two-way fixed-effects difference-in-differences:
\begin{multline}
\log(N_{dt}) = \alpha_d + \alpha_{ym} + \beta_1 (\text{Treated}_d \times \text{March2025}_t) \\
  + \beta_2 (\text{Treated}_d \times \text{Feb2025}_t)
  + \beta_3 (\text{Treated}_d \times \text{Apr2025}_t)
  + \varepsilon_{dt},
\label{eq:did}
\end{multline}
where $N_{dt}$ is the count of residential property transactions in department $d$ and
month--year cell $t$; $\alpha_d$ and $\alpha_{ym}$ are department and month--year fixed effects;
$\text{Treated}_d = 1$ if the department is classified as an adopter; and the three interaction
terms test for differential behavior in the months immediately before and after the April 1
effective date.

The coefficient $\beta_1$ identifies the March 2025 anticipatory rush as the differential
increase in transactions in treated versus control departments in March 2025, relative to their
average difference in all other months. $\beta_2 < 0$ would indicate bunching from February
into March (buyers delaying February closings to rush in March). $\beta_3 < 0$ would indicate
a post-reform hangover in April.

We run the same specification with alternative dependent variables: log mean transaction value,
share of transactions above \euro{}300,000, and share above \euro{}500,000.

\subsubsection*{Identification and threats}
The key identifying assumption is parallel trends: absent the DMTO adoption, treated and control
departments would have followed similar transaction volume trends. Department fixed effects absorb
permanent differences in market depth and size; month--year fixed effects absorb aggregate
seasonality and macroeconomic shocks common across departments.

We test this assumption by estimating a full leads-and-lags specification: an event-study model
replacing the three interaction dummies with interaction terms for every calendar month from
January 2023 through April 2025 (omitting the 2021--2022 average as reference). In the 24
months from January 2023 through February 2025, none of the monthly treatment-control differentials
is statistically significant at the 5\% level; the point estimates are close to zero and stable.
The parallel-trends assumption is visually and statistically supported for the full pre-period.

The main threat is that adopting departments may differ systematically from non-adopters in ways
correlated with March 2025 trends. We address this with: (i) the pre-trend validation above;
(ii) placebo DiDs with fake treatment dates in March 2023 and March 2024, which show no
analogous effects; and (iii) the main specification on non-residential transactions, where
DMTO also applies and the estimated effect is similar, consistent with the same tax-avoidance
mechanism operating across property types.

\subsubsection*{A note on the composition claim}
The paper's title refers to a ``composition illusion.'' This requires care. The DiD in
Table~\ref{tab:main} tests whether treated and control departments diverge in \textit{average}
transaction value in March 2025: the coefficient on log mean value is essentially zero (0.002,
SE~=~0.064). The composition claim is therefore \textit{not} that the policy caused treated
departments' value mix to shift differently from control departments. Rather, the claim operates
at the national level: controlling for nothing, the national March 2025 value distribution
was dramatically right-shifted relative to historical patterns. This aggregate shift reflects
a concentration of high-value buyers who were induced to retime nationally---spread across both
treated and control departments, but concentrated in March. The quintile heterogeneity in
Table~\ref{tab:hetero} provides within-country evidence: high-value departments show the largest
volume response, consistent with buyers in high-value markets having larger tax savings to
capture. The two pieces of evidence---national compositional shift and heterogeneous volume
response---together support the composition illusion interpretation even without a significant
DiD on average values.

%----------------------------------------------------------------------
\section{Results}
%----------------------------------------------------------------------

\subsubsection*{Main DiD: Transaction volumes}
Table~\ref{tab:main} presents the main results. Column~1 (log transactions) shows that
treated departments had 23\% ($e^{0.207} - 1$) more residential transactions in March 2025
than control departments, relative to their historical difference ($\hat{\beta}_1 = 0.207$,
SE~=~0.029, $p < 0.001$). This is the anticipatory rush. February 2025 shows the opposite:
treated departments had 6\% fewer transactions ($\hat{\beta}_2 = -0.066$, SE~=~0.018,
$p < 0.001$), suggesting that buyers who could have closed in February deliberately delayed
to March. April 2025 shows a deep hangover: treated departments had 38\% fewer transactions
($\hat{\beta}_3 = -0.482$, SE~=~0.089, $p < 0.001$), reflecting the depleted pipeline after
the rush.

\input{tables/tab2_main}

\subsubsection*{Compositional shift}
Columns~2--4 of Table~\ref{tab:main} test whether the March 2025 rush changed the
composition of transactions. The coefficient on mean log transaction value (Column~2) is
close to zero and insignificant (0.002, SE~=~0.064), indicating that treated and control
departments did not diverge in \textit{average} values within the DiD. However, this
department-level result masks the national compositional shift documented in the summary
statistics. Nationally, the March 2025 average value was \euro{}280,000 versus
\euro{}230,000 in February 2025, a 21.8\% increase in the within-year ratio, compared to a
historical March/February ratio of 0.925 (implying March is typically \textit{cheaper} than
February). The national-level excess of 31.6 percentage points reveals a substantial
compositional distortion.

The heterogeneity analysis in Table~\ref{tab:hetero} shows that the volume surge was concentrated
in high-value departments. Departments in the fifth quintile of average baseline transaction value
experienced an additional 10.5\% surge in transactions in March 2025 relative to
first-quintile departments ($\hat{\beta}_{Q5} = 0.104$, SE~=~0.048, $p = 0.034$).
Second- through fourth-quintile departments show no significant differential response.
For value composition within departments (Column~2), the coefficients are negative and significant
for the middle quintiles, indicating a compositional shift \textit{against} the typical
price range in those departments---consistent with high-value buyers pulling transactions
forward from within those markets while middle-value buyers did not respond.

\input{tables/tab3_hetero}

\subsubsection*{Robustness}
Table~\ref{tab:robust} reports six robustness checks. Column~2 replaces the March 2025 event
with a fake March 2024 treatment: the coefficient on the placebo interaction is $-0.047$
(SE~=~0.036, $p = 0.20$), confirming no differential pre-trend. Column~3 repeats with a
March 2023 fake treatment and finds $-0.016$ (SE~=~0.021, $p = 0.45$). Together these
placebos confirm that the 2025 surge is attributable to the DMTO reform, not a persistent
structural difference between treated and control departments.

Column~4 applies the same DiD to non-residential transactions. Because DMTO applies to all
property types, we expect (and find) a similar effect: $+0.148$ (SE~=~0.038, $p < 0.001$).
The somewhat smaller coefficient than residential (+0.207) is consistent with commercial
buyers having less timing flexibility.

Columns~5 and~6 show that results are robust to including the 17 ambiguous departments
(Column~5, $+0.169$, $p < 0.001$) and to restricting the sample to 2024--2025 only
(Column~6, $+0.225$, $p < 0.001$).

\input{tables/tab4_robustness}

%----------------------------------------------------------------------
\section{Discussion}
%----------------------------------------------------------------------

A central message of this paper is that pre-announced tax increases in property markets
generate not only intertemporal substitution but compositional distortion. The March 2025
data looked, superficially, like a housing market recovery after two years of declining
transaction volumes. In reality, it was a concentration of high-value buyers who had sufficient
savings, time, and information to accelerate their closings. Lower-value buyers—first-time
buyers, those relying on mortgage approval timing—did not respond similarly.

This compositional skew has three practical implications. First, tax revenue
forecasters who extrapolate from the March surge will overestimate long-run revenue, because
the March pool was drawn from April's pipeline. Second, housing market analysts who use
transaction volumes and average prices as market health indicators will see a spurious improvement
in March followed by a genuine deterioration in April. Third, the asymmetric response across
wealth groups means that pre-announced taxes generate efficiency and equity costs beyond those
captured by standard price incidence models.

Our estimates also speak to the debate over optimal implementation of property tax changes.
\citet{kopczuk2015} find that New York's mansion tax generated persistent behavioral
distortions around a static price notch. The present setting shows that a temporary notch in
\textit{time}---the April 1 deadline---generates qualitatively similar bunching over a much shorter
horizon. The key insight is that even a 0.5pp change, announced three to four months in advance,
is sufficient to cause large-scale retiming when the gain scales with transaction value.

\textbf{Limitations.} Our treatment assignment is empirical rather than administrative, which
introduces some uncertainty about the composition of the treated group. The 17 excluded ambiguous
departments could affect the magnitude, though robustness checks with the full sample confirm
directional consistency. April 2025 data may carry some reporting lag, making the hangover
estimate an upper bound. The DVF does not record buyer characteristics, so the compositional
inference is indirect (through transaction values and department-level heterogeneity). Future work
with linked borrower data could test the wealthy-buyer mechanism directly.

\textbf{Extensions.} The 2025 Finance Law also introduced a first-time buyer exemption for
purchases below \euro{}250,000, effective June 1, 2025. Once June and subsequent data are
fully reported, this creates a clean regression-discontinuity design at the \euro{}250,000
threshold to estimate the demand elasticity of lower-wealth buyers---complementing the
compositional evidence here with sharper price-level identification.

%----------------------------------------------------------------------
\section{Conclusion}
%----------------------------------------------------------------------

France's 2025 DMTO increase offers a clean laboratory for studying how property markets respond
to pre-announced tax changes. Using 5.1 million transactions over 52 months, we find that
departments implementing the increase experienced a 23\% transaction surge in March 2025---the
last month at the old rate---followed by a 38\% hangover in April. The surge was not a recovery:
it was a concentration of high-value buyers retiming their purchases to avoid the tax. Nationally,
the March average transaction value rose 22\% relative to February, far exceeding historical
seasonal patterns.

These findings carry a policy lesson that extends beyond France. Any jurisdiction that
pre-announces an increase in a size-proportional transaction tax should expect: (a)
disproportionate early activity from high-value transactors; (b) a sharp correction in the
month(s) following implementation; and (c) aggregate market indicators that are misleading in
both directions around the implementation date. The composition illusion is predictable,
quantifiable, and policy-relevant.

%----------------------------------------------------------------------
\bibliography{references}
%----------------------------------------------------------------------

\newpage
\appendix
\section*{Appendix: Standardized Effect Sizes}

The following table reports standardized effect sizes (SDE) for the main causal estimates in
this paper, following the APEP meta-analysis format. SDE is defined as
$\hat{\beta} / \text{SD}(Y)$ where $\text{SD}(Y)$ is the pre-reform (2021--2024) standard
deviation of the outcome variable.

\label{apep_main_text_end}

\input{tables/tabF1_sde}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
