\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Euro symbol
\usepackage{eurosym}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Kinks Without Bunching: Purchase-Tax Rate Jumps and Manufacturer CO\textsubscript{2} Manipulation}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Governments use graduated vehicle taxes to discourage CO\textsubscript{2} emissions, but whether manufacturers respond by manipulating type-approval emissions is unknown. I exploit four kinks in the Dutch BPM purchase tax where marginal rates jump by 1.6--39.5 times, using 1.25 million Dutch registrations and 11.5 million German registrations as a placebo. McCrary density tests find no significant discontinuity at any kink. Polynomial bunching estimates are unstable across specifications. The apparent concentration of vehicles near 79~g/km---entirely plug-in hybrids---is present equally in Germany, attributable to EU fleet-average regulation rather than national taxation. Purchase-tax kinks, unlike level notches, fail to distort manufacturer behavior.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H23, L62, Q58 \\
\noindent\textbf{Keywords:} vehicle taxation, bunching, CO\textsubscript{2} emissions, kink, purchase tax, manufacturer behavior

\newpage

\section{Introduction}

A Toyota Corolla registered in the Netherlands at 142~g/km of CO\textsubscript{2} faces a one-time purchase tax of \euro 9,767. The same car at 141~g/km pays \euro 9,483---a \euro 284 savings from a single gram. Across four kink points in the Dutch BPM (\textit{Belasting van Personenauto's en Motorrijwielen}), manufacturers could save buyers hundreds of euros per gram by calibrating type-approval emissions to land just below each threshold. Whether they do so is the question this paper answers. I find no detectable evidence that they do.

A large and influential literature uses bunching at tax thresholds to estimate behavioral elasticities in settings from income taxation \citep{saez2010, chetty2011} to housing markets \citep{best2014}. In vehicle markets, \citet{sallee2012} documented significant bunching at the US gas guzzler tax---a level notch that imposes a discrete tax increase above the threshold. \citet{ito2018} showed that Japanese fuel economy standards generate bunching at weight class boundaries. These findings underpin a growing policy consensus that vehicle CO\textsubscript{2} regulation creates strategic gaming incentives \citep{reynaert2021, grigolon2018}. But the evidence base conflates two distinct tax instruments: notches, where the tax level jumps discontinuously, and kinks, where only the marginal rate changes. This distinction is theoretically crucial \citep{kleven2016}: notches create a dominated region where no optimizing agent should locate, generating sharp bunching; kinks create only a change in slope, producing diffuse bunching that scales with the elasticity of adjustment.

This paper provides the first multi-cutoff test of manufacturer response to purchase-tax kinks in the CO\textsubscript{2} distribution. The Dutch BPM offers an unusually clean setting: four kinks at 79, 101, 141, and 157~g/km where the marginal tax rate increases by factors of 1.6 to 39.5. If manufacturers manipulate type-approval emissions to avoid these kinks, we should observe excess mass just below each threshold and missing mass above. The multi-cutoff structure enables a dose-response test: larger rate jumps should generate more bunching.

I use vehicle-level registration data from the European Environment Agency's CO\textsubscript{2} monitoring program, which records the WLTP CO\textsubscript{2} emissions, fuel type, manufacturer, and mass for every new car registered in the European Union. The analysis covers 1.25 million Dutch and 11.5 million German registrations from 2020--2022, the WLTP era of BPM taxation. Germany serves as a placebo country: it has no purchase tax with CO\textsubscript{2} kinks at the Dutch thresholds, allowing a difference-in-bunching design \citep{kleven2013}.

The results are consistently null for the Netherlands. McCrary density discontinuity tests \citep{mccrary2008} find no statistically significant break at any of the four BPM kinks in the Dutch distribution ($|t| < 1$ at all four thresholds). Polynomial bunching estimates in the tradition of \citet{chetty2011} are completely unstable: the sign of the estimated excess mass reverses across polynomial degrees and bandwidth choices at the 141 and 157~g/km kinks, and the 79 and 101~g/km kinks produce degenerate estimates due to the sharp transition between plug-in hybrid (PHEV) and conventional powertrains. Year-by-year estimates show no temporal stability.

The most striking data pattern---a concentration of vehicles near 79~g/km---turns out to be uninformative about BPM taxation. This cluster is composed entirely of PHEVs, and an identical pattern appears in German data, where no equivalent purchase-tax kink exists. The 79~g/km cluster reflects EU fleet-average CO\textsubscript{2} regulation under Regulation (EU) 2019/631 \citep{eu2019631}, which creates strong manufacturer incentives to calibrate PHEV type-approval emissions regardless of national tax policy. In the conventional (ICE) vehicle range above 86~g/km---where BPM kinks at 101, 141, and 157~g/km should bite---the CO\textsubscript{2} distribution is lumpy but shows no systematic response to tax thresholds.

This null result carries specific policy implications. Graduated purchase taxes with kink-based rate structures---used in the Netherlands, Norway, Denmark, and other EU member states---are less effective at distorting manufacturer behavior than the notch-based designs studied by \citet{sallee2012} and \citet{best2014}. The per-gram tax savings at a kink---\euro 79 to \euro 568---are likely modest relative to the engineering cost of recalibrating an ICE powertrain to achieve a specific type-approval CO\textsubscript{2} target \citep{reynaert2021}. If policymakers want to use vehicle taxation to shift the CO\textsubscript{2} distribution, the evidence suggests they need notches---not kinks.

The paper contributes to three literatures. First, it advances the bunching literature \citep{kleven2016} by documenting a setting where theory predicts diffuse bunching and the data confirm its absence---a rare multi-cutoff null in a high-stakes policy domain. Second, it informs the vehicle taxation literature \citep{anderson2012, rivers2017, allcott2014} by demonstrating that the instrument's design (kink vs.\ notch) dominates its level in determining manufacturer response. Third, it contributes to the debate over CO\textsubscript{2} manipulation in vehicle markets \citep{reynaert2021} by showing that PHEV CO\textsubscript{2} clustering is an EU-wide fleet regulation phenomenon, not a response to national fiscal incentives.

\section{Institutional Background}

\subsubsection*{The Dutch BPM} The BPM is a one-time vehicle purchase tax levied at the point of registration in the Netherlands. Since July 2020, the tax has been calculated from WLTP CO\textsubscript{2} emissions using a piecewise-linear schedule with five bands. \Cref{tab:bpm_schedule} presents the full schedule. The tax is \textit{continuous}---there is no level jump at any threshold---but the \textit{marginal rate} increases sharply at each kink. At the first kink (79~g/km), the per-gram rate rises from \euro 2 to \euro 79, a 39.5-fold increase. At subsequent kinks, the rate increases by factors of 2.2, 1.6, and 2.0. A vehicle at 158~g/km pays \euro 14,595 in BPM; the same vehicle at 79~g/km pays \euro 825.

The piecewise-linear design creates four kink points with varying incentive strengths. Unlike the US gas guzzler tax studied by \citet{sallee2012}---which imposes a discrete \$1,000--\$7,700 penalty above a fuel economy threshold---the BPM never creates a dominated region. A manufacturer moving a vehicle from 80 to 79~g/km saves the buyer \euro 79 (one gram at the higher rate); moving from 158 to 157~g/km saves \euro 568. These savings are meaningful but modest relative to vehicle prices of \euro 25,000--\euro 50,000.

\input{tables/tab1_bpm_schedule}

\subsubsection*{EU fleet-average regulation} Alongside national taxation, all European manufacturers face fleet-average CO\textsubscript{2} targets under Regulation (EU) 2019/631. The 2021--2024 target is 95~g/km, with manufacturer-specific adjustments based on fleet mass. Exceeding the target triggers fines of \euro 95 per g/km per vehicle sold. PHEVs, which typically achieve 30--80~g/km under WLTP, serve as compliance vehicles that pull down manufacturers' fleet averages \citep{reynaert2021}. This creates a strong incentive to calibrate PHEV type-approval emissions as low as possible---independent of any national purchase tax.

\subsubsection*{Germany as placebo} Germany levies an annual CO\textsubscript{2}-based vehicle tax (\textit{Kfz-Steuer}) at a flat rate of \euro 2/g above 95~g/km, with no kinks at the Dutch BPM thresholds. Any bunching at 79, 101, 141, or 157~g/km in the German distribution would indicate an EU-wide phenomenon unrelated to BPM taxation, providing a clean placebo test.

\section{Data}

I use vehicle-level registration data from the European Environment Agency's CO\textsubscript{2} monitoring program under Regulation (EU) 2019/631. Manufacturers report every new passenger car registration to the EEA with WLTP CO\textsubscript{2} emissions (g/km), fuel type, manufacturer, mass, engine capacity, and power. The data are publicly available via the EEA's SQL API.

The analysis sample covers new passenger car registrations in the Netherlands and Germany from 2020 to 2022---the period during which the WLTP-based BPM schedule has been in effect. I exclude battery-electric vehicles (0~g/km CO\textsubscript{2}), which are outside all BPM bands. The final sample comprises 1,253,380 Dutch and 11,478,416 German registrations.

I construct CO\textsubscript{2} frequency distributions by counting registrations at each integer CO\textsubscript{2} value from 1 to 250~g/km, pooled across years. This is the standard data structure for bunching estimation \citep{kleven2016}. I also decompose the distribution by fuel type (petrol, diesel, petrol/electric, diesel/electric) near each kink to identify powertrain composition effects.

\input{tables/tab2_summary}

\Cref{tab:summary} reports summary statistics. Dutch registrations have lower mean CO\textsubscript{2} (122.1 vs.\ 135.6~g/km in Germany), reflecting higher PHEV penetration driven by both BPM incentives and other Dutch EV subsidies. The Dutch distribution has substantial mass at both the PHEV range (30--85~g/km) and the ICE range (120--160~g/km).

\section{Empirical Strategy}

\subsection{Bunching Estimation}

I employ two complementary approaches to test for bunching at each BPM kink. The first is the standard polynomial bunching estimator in the tradition of \citet{chetty2011} and \citet{kleven2016}. For each kink point $k$, I fit a polynomial of degree $p$ to the CO\textsubscript{2} frequency distribution using bins outside an excluded region $[k - h, k + h]$, predict the counterfactual density inside the excluded region, and calculate excess mass $\hat{b}$ as the difference between observed and counterfactual counts below the kink, normalized by the counterfactual height at the kink point. Standard errors are computed via Poisson bootstrap with 500 replications.

The polynomial bunching estimator is the theoretically appropriate tool for kinks, where the prediction is diffuse excess mass below the threshold rather than a sharp density break. However, the lumpiness of the CO\textsubscript{2} distribution (driven by popular car models clustering at specific values) makes the polynomial fit sensitive to specification choices. I therefore complement the polynomial approach with a McCrary-style density discontinuity test \citep{mccrary2008}, which provides a specification-free check for any break in the density. While the McCrary test is designed primarily for notches (level jumps), a kink that generates substantial bunching should also produce a visible density break at the threshold. I estimate:
\begin{equation}
\log(n_{\text{co2}}) = \alpha + \beta \cdot \ind[\text{CO}_2 > k] + \gamma_1 ({\text{CO}_2 - k}) + \gamma_2 ({\text{CO}_2 - k})^2 + \gamma_3 \cdot \ind[\text{CO}_2 > k] \cdot ({\text{CO}_2 - k}) + \varepsilon
\end{equation}
where $n_{\text{co2}}$ is the count of vehicles at each integer CO\textsubscript{2} value within $\pm 15$~g/km of the kink. The coefficient $\beta$ captures the log-density discontinuity. Under the null of no manipulation, $\beta = 0$.

\subsection{Difference-in-Bunching}

The key identification concern is that the CO\textsubscript{2} distribution may be lumpy for reasons unrelated to taxation---popular car models cluster at specific CO\textsubscript{2} values, and EU-wide regulation affects the distribution in all member states. I address this through a difference-in-bunching design: I estimate bunching in both the Netherlands (where BPM kinks apply) and Germany (where no equivalent kinks exist), and test whether Dutch excess bunching exceeds German excess bunching at each threshold. The difference-in-bunching parameter is:
\begin{equation}
\Delta \hat{b}_k = \hat{b}^{\text{NL}}_k - \hat{b}^{\text{DE}}_k
\end{equation}
with standard error $\text{SE}(\Delta \hat{b}_k) = \sqrt{\text{SE}(\hat{b}^{\text{NL}}_k)^2 + \text{SE}(\hat{b}^{\text{DE}}_k)^2}$.

\subsection{What This Design Can and Cannot Identify}

The design identifies manufacturer response to BPM kinks under the assumption that, absent the BPM, the Dutch CO\textsubscript{2} distribution would be smooth (or at least as smooth as Germany's) at the kink points. The multi-cutoff structure with dose-response provides power: under the alternative hypothesis, bunching should be largest at the 79~g/km kink (39.5$\times$ rate jump) and smallest at the 141~g/km kink (1.6$\times$). The design primarily identifies manufacturer-side manipulation (engine calibration to specific CO\textsubscript{2} targets), which is the channel that creates sharp bunching at integer values. Consumer-side sorting across existing models would shift demand toward lower-emission vehicles but not generate excess mass at specific gram thresholds. The BPM is paid by buyers, so manufacturer manipulation requires the tax incentive to be transmitted through pricing or competitive pressure.

\section{Results}

\subsection{Main Results: No Density Discontinuity}

\Cref{tab:mccrary} reports McCrary density discontinuity tests at all four BPM kinks. None of the four Dutch kinks shows a statistically significant density break. The $t$-statistics range from $-0.78$ (kink at 79~g/km) to 0.27 (kink at 101~g/km). For comparison, Germany shows marginally significant negative discontinuities at 79~g/km ($t = -2.25$) and 141~g/km ($t = -2.00$), both in the direction of \textit{more} mass above the kink---the opposite of what BPM-driven bunching would predict. The difference-in-bunching estimates (NL $-$ DE) are small and statistically indistinguishable from zero at 141 and 157~g/km.

\input{tables/tab3_mccrary}

\subsubsection*{Statistical power} The null result is informative only if the design has sufficient power to detect economically meaningful bunching. At the 141~g/km kink, the average bin count is approximately 16,000 vehicles, with standard deviation across nearby bins of roughly 3,000. Excess mass of 5\% (800 vehicles) would produce a McCrary coefficient of $\log(1.05) \approx 0.049$ against a standard error of 0.13, yielding $t = 0.38$---undetectable. Excess mass of 20\% (3,200 vehicles) would produce $t = 1.5$---still below conventional thresholds. The design can reliably detect excess mass of approximately 35\% or more at a single kink. This implies that if manufacturers were shifting CO\textsubscript{2} by even 1--2 grams for a substantial fraction of vehicles, the effect would need to be concentrated within a narrow window to be detectable. The multi-cutoff structure partially compensates: if all four kinks showed consistent positive excess mass, even individually insignificant estimates would jointly constitute evidence. The data show no such pattern.

\subsubsection*{Raw density patterns} \Cref{tab:density} shows raw vehicle counts and normalized densities within $\pm 2$~g/km of each kink. At 79~g/km, the Dutch density drops sharply from 0.37 per 1,000 (at 79) to 0.08 per 1,000 (at 80)---but the German density shows a nearly identical pattern (0.20 to 0.02). At 141~g/km, the Dutch and German densities are comparably smooth across the threshold. At 157~g/km, the NL/DE density ratio is stable at approximately 0.5, showing no differential response.

\input{tables/tab4_density}

\subsection{Instability of Polynomial Bunching Estimates}

\Cref{tab:robustness} demonstrates that the polynomial bunching estimator produces fundamentally unstable results. At the 141~g/km kink (the cleanest ICE-range kink), the sign of the difference-in-bunching estimate reverses three times across four polynomial degrees ($p = 3, 5, 7, 9$) and twice across four bandwidth choices ($h = 3, 5, 7, 10$). This instability is the signature of noise, not signal: when a genuine density discontinuity exists, bunching estimates are robust to polynomial order \citep{kleven2016}. Here, the polynomial is fitting lumps from popular car models, not systematic bunching from tax optimization.

\input{tables/tab5_robustness}

\subsection{The PHEV Composition Channel}

The most visually striking feature of the Dutch CO\textsubscript{2} distribution is a cluster of vehicles near 79~g/km. Decomposition by fuel type reveals that this cluster is composed entirely of PHEVs. Below 86~g/km, 100\% of Dutch registrations are petrol/electric or diesel/electric hybrids. The first substantial mass of conventional petrol vehicles appears at 87~g/km (6,336 vehicles) and 92~g/km (11,688 vehicles), with no presence at or below the 79~g/km kink.

This PHEV cluster is not a BPM response. Germany, which has no BPM kinks, shows an identical pattern: PHEV mass at 70--85~g/km with a sharp transition to ICE at 87~g/km. The cluster reflects EU fleet-average regulation under Regulation (EU) 2019/631, which creates powerful incentives for manufacturers to offer PHEVs with type-approval CO\textsubscript{2} below the fleet target. PHEV CO\textsubscript{2} is highly calibratable---it depends on battery size, electric range assumptions, and utility factor weighting in the WLTP test cycle---making PHEVs the instrument of fleet compliance. This EU-wide calibration dominates any marginal incentive from national purchase taxation.

\subsection{Robustness}

\subsubsection*{Year-by-year estimates} Bunching estimates at the 141~g/km kink range from $+1.59$ (2020) to $-2.75$ (2022) for the Netherlands and from $+0.43$ to $-0.73$ for Germany. If manufacturers were systematically calibrating to the BPM schedule, we would expect stable or increasing bunching over time as the policy became established. Instead, the estimates fluctuate wildly, consistent with noise from model-specific CO\textsubscript{2} variation.

\subsubsection*{PHEV-specific bunching at 79~g/km} Restricting the sample to PHEVs only and estimating bunching within the PHEV CO\textsubscript{2} range (50--110~g/km), I find normalized bunching of $\hat{b} = 0.087$ (SE $= 0.146$, $t = 0.59$). Even among the vehicles most susceptible to CO\textsubscript{2} calibration, there is no evidence of BPM-driven bunching at the 79~g/km threshold.

\section{Discussion}

The absence of bunching at Dutch BPM kinks carries a specific mechanistic interpretation. The key distinction is between kinks and notches. At a notch, crossing the threshold imposes a discrete tax increase---a lump-sum cost. This creates a dominated region where purchasing is strictly worse than at the threshold, generating sharp and visible bunching \citep{kleven2013}. At a kink, the tax is continuous but its slope changes. There is no dominated region; the incentive to locate below the threshold is proportional to the marginal rate change times the distance from the kink. For small adjustments (1--2~g/km), the savings are modest---\euro 79--568 per gram---relative to the cost of powertrain re-engineering.

Engineering estimates suggest that reducing type-approval CO\textsubscript{2} by 1~g/km through powertrain recalibration costs manufacturers \euro 50--150 per vehicle for ICE engines \citep{reynaert2021}. At the 141~g/km kink, the marginal tax saving is \euro 111 per gram (the rate jump from \euro 173 to \euro 284). In a competitive market where tax savings are fully passed through, manufacturers would manipulate only if the tax saving exceeds the engineering cost for the \textit{marginal} gram. At the lower kinks (79 and 101), the per-gram savings (\euro 77 and \euro 94) may fall below the engineering cost, rationalizing the null as an economically predicted non-response rather than an absence of incentive transmission.

The contrast with \citet{sallee2012} is instructive. The US gas guzzler tax creates level jumps of \$1,000--\$7,700 at fuel economy thresholds---these are notches, not kinks---and generates clear bunching. The Dutch BPM, despite marginal rate changes of up to 39.5$\times$, produces no observable response because the \textit{level} of the tax is continuous. This suggests that the functional form of vehicle taxation---not its magnitude---determines whether manufacturers manipulate type-approval emissions.

For policymakers, the implication is that graduated purchase taxes with piecewise-linear rate schedules are poor instruments for influencing manufacturer CO\textsubscript{2} calibration. If the policy goal is to create incentives for lower-emission vehicles, a notch-based design (with discrete tax jumps at thresholds) would be more effective, as demonstrated by the UK Vehicle Excise Duty bands, the Norwegian registration tax, and the French bonus-malus system.

\section{Conclusion}

The Dutch BPM's four CO\textsubscript{2} kinks---the largest increasing the marginal tax rate by a factor of 39.5---generate no detectable bunching in the vehicle registration distribution. The apparent clustering near 79~g/km is entirely PHEVs, driven by EU fleet regulation rather than national taxation, and present identically in Germany. In this setting, purchase-tax kinks---even with dramatic rate jumps---do not generate detectable manufacturer manipulation, consistent with the absence of a dominated region that makes notches effective. Where the engineering cost of CO\textsubscript{2} adjustment exceeds the per-gram tax saving, the rational manufacturer response is non-response.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Data source} Vehicle-level CO\textsubscript{2} monitoring data from the European Environment Agency, collected under Regulation (EU) 2019/631. Available at \url{https://discodata.eea.europa.eu}. Accessed April 2026.

\subsubsection*{Sample construction} I query registration counts at each integer CO\textsubscript{2} value (1--250~g/km) for each country (NL, DE) and year (2020--2022). Battery-electric vehicles (CO\textsubscript{2} $= 0$) are excluded. The unit of observation is a CO\textsubscript{2} value $\times$ country $\times$ year bin. For fuel-type decomposition, I separately query counts for petrol, diesel, petrol/electric, and diesel/electric within $\pm 30$~g/km of each kink.

\subsubsection*{BPM rates} Tax rates from the \textit{Wet op de belasting van personenauto's en motorrijwielen} 1992, as amended by the 2020 Climate Agreement implementation. The rates in \Cref{tab:bpm_schedule} apply to the full 2020--2022 period without change.

\section{Robustness Appendix}

Robustness checks are reported in the main text (\Cref{tab:robustness}). Additional details: all polynomial bunching estimates use a window of $\pm 25$ CO\textsubscript{2} bins around each kink, with the excluded region defined symmetrically ($h$ below and $h$ above the kink). Poisson bootstrap uses 500 replications with seed 42 for replicability. McCrary tests use $\pm 15$ bins with a quadratic in the running variable and a slope interaction.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
