\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Inspection Lottery: How Regulatory Stringency\\ Crowds Out Nursing Home Staffing}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Nursing home quality depends on staffing, but does regulatory inspection improve it? I exploit quasi-random variation in state survey agency stringency---the ``inspection lottery''---to identify the causal effect of deficiency citations on nurse staffing. Using leave-one-out state mean severity as an instrument for 11,427 facilities in 2025, I find that stricter enforcement \textit{reduces} total nurse staffing by 0.73 hours per resident per day per severity unit (5.2\% of the mean for a one-standard-deviation stringency shift). The effect is confirmed within multi-state chains sharing identical management (--0.59 with chain fixed effects), concentrated in for-profit facilities, and accompanied by a 12.7 percentage point increase in nursing staff turnover. These results reveal a compliance crowding mechanism: regulatory pressure diverts resources from clinical care to administrative remediation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I11, I18, L51, J23 \\
\noindent\textbf{Keywords:} nursing homes, health inspection, regulatory enforcement, staffing, compliance costs

\newpage

\section{Introduction}

Every year, a state surveyor walks into each of America's 15,000 nursing homes, clipboard in hand, and decides how many deficiencies to cite and how severely to grade them. The consequences are real: citations trigger mandatory remediation plans, follow-up inspections, and potential financial penalties. But here is the paradox---identical conditions in identical chain-operated facilities receive wildly different regulatory treatment depending on which state's agency conducts the survey. The Government Accountability Office found that 70 percent of state surveys miss deficiencies identified by federal follow-up teams, and 15 percent miss serious ones \citep{GAO2008}. This variation is not merely statistical noise. It is a natural experiment.

The policy stakes are large. Nursing homes serve 1.3 million residents, consume over \$100 billion annually in Medicare and Medicaid spending, and have been the subject of persistent quality concerns \citep{Bowblis2011, Harrington2012}. The conventional wisdom holds that stricter regulation improves quality: more citations discipline facilities into investing in better care, primarily through higher staffing \citep{Mukamel2012, Castle2011}. If true, the dramatic cross-state variation in enforcement intensity represents a massive quality gap---and the policy response is straightforward standardization. But this paper tells a different story.

I use state survey agency stringency as an examiner-leniency instrument for deficiency severity, in the spirit of judge-IV designs \citep{Kling2006, Doyle2007}. The instrument is the leave-one-out mean scope-severity score assigned by the same state agency to \textit{other} facilities in the same year. The identifying assumption is that, conditional on facility characteristics, a nursing home's surveyor regime is determined by geography, not by the facility's own quality choices. Multi-state chains provide the sharpest test: a Genesis Healthcare facility in strict New Hampshire faces a different regulatory regime than a Genesis facility in lenient Louisiana, despite identical corporate governance and budget policies.

The first stage is powerful: a one-unit increase in state stringency raises the facility's mean severity score by 0.94 (robust $F$ = 4,764). But the second stage delivers a surprise. Instrumented deficiency severity \textit{reduces} total nurse staffing by 0.73 hours per resident per day (HPRD), significant at the 5 percent level ($p$ = 0.024). This is 15 times larger than the OLS estimate of --0.05, and it goes the wrong way for the enforcement-improves-quality hypothesis. Facilities facing stricter surveyors invest \textit{less} in clinical staffing, not more.

The within-chain specification confirms this finding. Among 8,192 chain-affiliated facilities spanning 604 chains, the IV estimate with chain fixed effects is --0.59 HPRD ($p$ = 0.003). Since chain fixed effects absorb all corporate-level policies, this estimate reflects pure within-chain variation in regulatory exposure---the closest feasible approximation to random surveyor assignment.

The mechanism I identify is \textit{compliance crowding}. Each deficiency citation triggers a cascade of administrative responses: plans of correction must be drafted, submitted, and implemented; follow-up surveys must be prepared for; staff time is redirected from patient care to documentation. I decompose the staffing reduction into its components: registered nurses lose 0.20 HPRD, licensed practical nurses 0.25, and certified nursing assistants 0.28. The reduction is spread across all staff types, with the lowest-paid caregivers bearing the largest share. Simultaneously, nursing staff turnover rises by 12.7 percentage points per unit of instrumented severity ($p$ = 0.002), consistent with the hypothesis that regulatory burden drives clinical workers away from cited facilities.

This paper makes three contributions. First, it provides the first examiner-leniency IV estimates of the causal effect of regulatory inspection stringency on nursing home staffing. While a large literature studies nursing home quality determinants \citep{Grabowski2004, Bowblis2011, Konetzka2014}, no prior work exploits the quasi-random assignment of surveyor regimes to identify the causal chain from citations to staffing investment. Second, it identifies the compliance crowding mechanism, adding to the literature on unintended consequences of regulation \citep{Bardach2000, ShortToffel2010, Parker2013, Johnson2020}. This mechanism is distinct from the well-studied compliance costs literature, which focuses on firms exiting the market; here, firms \textit{stay} but redirect resources internally. Third, the multi-state chain design provides a novel identification strategy that separates management quality from regulatory pressure, extending the examiner-leniency framework to a setting where the ``examiners'' are state agencies rather than individual judges.

The results have direct policy implications for the Centers for Medicare and Medicaid Services (CMS). If stringent enforcement crowds out the very staffing it is supposed to promote, then uniform citation standards---a frequent policy proposal---may be counterproductive without complementary reforms that reduce the administrative burden of compliance. The findings suggest that the optimal regulatory design must balance deterrence against the real resource costs that enforcement imposes on frontline care.

The paper proceeds as follows. Section 2 describes the institutional background of nursing home regulation. Section 3 presents the data. Section 4 develops the empirical strategy. Section 5 reports results. Section 6 discusses implications and limitations.

\section{Institutional Background}

\subsubsection*{The Federal-State Survey System} Medicare and Medicaid certification requires nursing homes to meet federal quality standards under 42 CFR Part 488. CMS delegates inspection authority to state survey agencies, which conduct unannounced standard health inspections approximately annually. Surveyors assess compliance with over 180 federal requirements covering areas from infection control to resident rights \citep{CMS2017}. When a violation is identified, it is classified on a scope-severity matrix with 12 cells, from A (isolated deficiency, no actual harm, potential for minimal harm) to L (widespread deficiency, immediate jeopardy to resident health or safety). The citation severity determines the required corrective action and potential penalties.

\subsubsection*{Cross-State Variation in Enforcement} The decentralized structure creates substantial variation in enforcement intensity. States differ in surveyor staffing ratios (from 2.5 to 12.5 surveyors per 1,000 nursing home residents), training protocols, citation norms, and organizational culture \citep{GAO2008, CMS2019}. Several studies have documented that identical conditions elicit different regulatory responses across states \citep{Harrington2004, Mukamel2012}. Federal comparative surveys, where CMS resurveys a sample of recently state-surveyed facilities, consistently find that state agencies undercount deficiencies relative to federal teams, with the degree of undercounting varying by state \citep{GAO2008}.

\subsubsection*{Consequences of Citations} Deficiency citations trigger a graduated enforcement response. At the lower end (A--F, no actual harm), facilities must submit plans of correction and may face follow-up surveys. At the moderate level (G--I, actual harm), facilities face directed plans of correction and potential civil monetary penalties (CMPs) of up to \$21,393 per day. At the highest level (J--L, immediate jeopardy), facilities face immediate CMPs, potential termination from Medicare/Medicaid, and state monitoring \citep{CMS2017}. The administrative burden of responding to citations is substantial: plans of correction require detailed documentation, implementation, and verification; resurveys consume staff time in preparation and during the visit.

\subsubsection*{Multi-State Chains} Approximately 69 percent of U.S. nursing homes are affiliated with chains. The largest---Ensign (14 states), Sabra Healthcare (43 states), and Genesis Healthcare (25 states)---operate facilities across diverse regulatory environments. Chain management typically sets corporate-wide policies on staffing ratios, budget allocation, and quality improvement, creating a setting where within-chain variation in regulatory exposure is driven by geography rather than management quality \citep{Grabowski2004, Bowblis2011}.

\section{Data}

\subsubsection*{Sources} I use four datasets from the CMS Provider Data Catalog, all publicly available without authentication. The Health Deficiencies dataset contains 418,972 individual deficiency citations with facility identifiers, survey dates, deficiency tags, and scope-severity codes. The Provider Information dataset provides facility characteristics for 14,703 Medicare/Medicaid-certified nursing homes, including certified bed count, ownership type, chain affiliation, and reported nurse staffing hours per resident per day from Payroll-Based Journal (PBJ) submissions. The Quality Measures dataset contains 249,951 MDS-derived quality indicators. I focus on the 2025 survey year (the latest complete year) for the cross-sectional analysis.

\subsubsection*{Variable Construction} I convert scope-severity codes to a numeric scale (A = 1, ..., L = 12) and aggregate to the facility level, computing mean severity, number of deficiencies, and the proportion classified as actual harm or immediate jeopardy (codes G--L). The leave-one-out (LOO) state stringency instrument for facility $i$ in state $s$ equals the mean severity score of all \textit{other} facilities in state $s$, excluding facility $i$. Staffing outcomes use PBJ-reported hours per resident per day (HPRD) for total nursing staff, registered nurses (RN), licensed practical nurses (LPN), and certified nursing assistants (CNA). Turnover is the annual percentage of nursing staff who separated from the facility.

\subsubsection*{Sample} The analysis sample includes 11,427 facilities with non-missing staffing data and valid instrument values. The chain subsample comprises 8,192 facilities across 604 chains.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

Table~\ref{tab:summary} reports summary statistics. The mean deficiency severity is 4.63 on the 1--12 scale, with 8 percent of citations at the actual harm or jeopardy level. Total nurse staffing averages 3.86 HPRD, with substantial variation (SD = 1.24). The LOO state stringency instrument ranges from 4.09 to 5.42 (cross-state SD = 0.29), providing meaningful variation for identification. Seventy-six percent of facilities are for-profit, and 72 percent are chain-affiliated.

\section{Empirical Strategy}

\subsection{Identification}

I estimate the effect of deficiency severity on staffing investment using two-stage least squares (2SLS):

\textit{First stage:}
\begin{equation}
\text{Severity}_i = \alpha + \gamma Z_{s(i)} + X_i'\beta + \varepsilon_i
\label{eq:first_stage}
\end{equation}

\textit{Second stage:}
\begin{equation}
Y_i = \mu + \delta \widehat{\text{Severity}}_i + X_i'\lambda + u_i
\label{eq:second_stage}
\end{equation}

\noindent where $Y_i$ is a staffing outcome for facility $i$, $\text{Severity}_i$ is the endogenous mean deficiency severity score, $Z_{s(i)}$ is the leave-one-out state stringency instrument, and $X_i$ includes log certified beds and ownership type indicators. Standard errors are clustered at the state level to account for within-state correlation in the instrument.

The instrument $Z_{s(i)}$ assigns to each facility the average severity score of all other facilities surveyed by the same state agency in the same year:
\begin{equation}
Z_{s(i)} = \frac{1}{N_s - 1} \sum_{j \in s, j \neq i} \overline{\text{Severity}}_j
\end{equation}

\subsubsection*{Identifying Assumptions} Instrument relevance requires that state agency stringency predicts facility-level deficiency severity. This is strongly satisfied: a one-unit increase in $Z_{s(i)}$ raises facility severity by 0.94 (first-stage $F$ = 4,764). The exclusion restriction requires that state stringency affects staffing outcomes only through the deficiency citations it generates. The key threat is that state-level factors (Medicaid reimbursement rates, nurse labor markets, regulatory culture) may independently affect both stringency and staffing. The within-chain specification addresses this by absorbing all chain-level confounders; the remaining variation is purely geographic. I also report balance tests and weak-IV-robust Anderson-Rubin inference.

\subsubsection*{Estimand} The 2SLS coefficient $\delta$ estimates the average causal response of staffing to deficiency severity among facilities whose measured severity is shifted by state stringency---the population of facilities that would receive different citations under a different surveyor regime. With a continuous instrument and continuous treatment, this is an average causal response along the margin moved by the instrument, analogous to the LATE interpretation in binary settings \citep{Angrist1996}.

\subsection{Threats to Validity}

\subsubsection*{State-level confounders} The primary concern is that state characteristics correlated with survey stringency may independently affect staffing. Balance tests (Section 5) show that log bed count and for-profit status are uncorrelated with the instrument. The urban indicator shows some imbalance ($\hat{\beta}$ = --0.30, SE = 0.12), suggesting that stricter states tend to have more rural facilities. I note this as a limitation; however, the within-chain specification, which compares facilities in different states within the same chain, substantially mitigates this concern.

\subsubsection*{Monotonicity} The instrument shifts all facilities in the same direction: stricter state agencies issue higher severity scores across the board. This is supported by the near-unity first-stage coefficient (0.94), indicating a roughly one-for-one relationship between state stringency and facility severity.

\section{Results}

\subsection{OLS and First Stage}

\input{tables/tab2_ols_fs}

Table~\ref{tab:ols_fs} presents OLS and first-stage results. Panel A shows that the na\"ive OLS relationship between deficiency severity and total HPRD is weakly negative (--0.048, SE = 0.026) and attenuated with controls (--0.046, SE = 0.027). This small OLS estimate is consistent with two offsetting forces: worse facilities have both lower staffing \textit{and} more citations (negative selection), while citation-induced compliance investment raises staffing (positive causal effect). The IV estimate, by isolating the regulatory pressure channel, separates these forces.

Panel B reports the first stage. The instrument is powerful across all specifications. Unconditionally, a one-unit increase in LOO state stringency raises facility severity by 0.94 ($F$ = 5,398). Adding controls barely changes the estimate (0.94, $F$ = 4,764). Even within chains, where the identifying variation comes from cross-state differences in surveyor regimes facing the same corporate management, the coefficient remains large (0.80, $F$ = 118). These $F$-statistics far exceed conventional weak-instrument thresholds.

\subsection{Main IV Results}

\input{tables/tab3_iv_main}

Table~\ref{tab:iv_main} presents the core finding. A one-unit increase in instrumented deficiency severity reduces total nurse staffing by 0.74 HPRD (column 1, SE = 0.31, $p$ = 0.018). Adding controls changes the estimate negligibly (--0.73, column 2). The within-chain specification (column 3) is the preferred estimate: among 8,192 chain-affiliated facilities, the IV estimate with chain fixed effects is --0.59 HPRD ($p$ = 0.003). Since chain fixed effects absorb all corporate-level budget and management policies, this estimate compares facilities within the same chain that face different state surveyors---the closest feasible approximation to random surveyor assignment.

To translate to policy-relevant magnitudes: a one-standard-deviation increase in state stringency (0.29 units) shifts facility severity by $0.29 \times 0.94 = 0.27$ points, reducing total HPRD by $0.27 \times 0.73 = 0.20$ hours per resident per day, or 5.2 percent of the mean (3.86). For a facility with 80 residents, this represents 16 fewer nursing hours per day---approximately two full-time nurse positions.

Columns 4--5 show that RN staffing declines by 0.20 HPRD in the full sample and 0.13 in the chain subsample, though the latter is imprecisely estimated.

\subsection{Mechanisms}

\input{tables/tab4_mechanisms}

Table~\ref{tab:mechanisms} decomposes the staffing reduction and examines turnover. Panel A reveals that all three staff categories decline: RN (--0.20 HPRD), LPN (--0.25), and CNA (--0.28). The three components sum exactly to the total (--0.73), confirming internal consistency. The largest absolute reduction falls on CNAs, the lowest-cost and most numerous caregiving staff---consistent with facilities reducing the cheapest-to-cut positions first.

The turnover results strengthen the compliance crowding interpretation. Total nursing staff turnover rises by 12.7 percentage points per unit of instrumented severity ($p$ = 0.002), and RN turnover rises by 6.9 points ($p$ = 0.027). Strict enforcement does not merely cause facilities to hire fewer nurses; it drives existing nurses away.

Panel B shows that the effect is concentrated in for-profit facilities (--0.73, SE = 0.28), with a smaller and imprecise effect among non-profits (--0.44, SE = 0.36). This heterogeneity is consistent with the compliance crowding mechanism: for-profit facilities, with tighter margin constraints, are more likely to redirect resources from staffing to administrative compliance when faced with regulatory pressure.

\subsection{Robustness}

\input{tables/tab5_robustness}

Table~\ref{tab:robustness} reports robustness checks. The main finding is stable across specifications (Panel B): using maximum severity rather than mean (--0.41), restricting to chain or independent facilities separately (--0.71 and --0.72), and including chain fixed effects (--0.59). The leave-one-state-out jackknife (Panel C) shows remarkable stability: the mean across 52 regressions is --0.73, with a range of [--0.88, --0.42]. No single state drives the result.

\subsubsection*{Balance tests} The instrument is uncorrelated with log bed count (0.025, SE = 0.14) and for-profit status (--0.009, SE = 0.10). The urban indicator shows some correlation (--0.30, SE = 0.12), a limitation I acknowledge. However, controlling for urban status does not meaningfully change the IV estimates, and the within-chain specification absorbs state-level urban-rural composition.

\subsubsection*{Alternative treatment measures} Using maximum severity rather than mean severity as the endogenous variable yields an IV coefficient of --0.41 (SE = 0.19), smaller in magnitude but the same sign. Using the \textit{count} of deficiencies produces a positive coefficient (+0.14, SE = 0.06), consistent with the interpretation that it is the \textit{severity} of regulatory action---not merely its volume---that drives the compliance crowding response.

\subsubsection*{Weak-IV inference} The Anderson-Rubin $F$-statistic is 5.12 ($p$ = 0.028), confirming that the reduced-form relationship is significant under weak-IV-robust inference. Given the extremely large first-stage $F$ (over 4,700), weak instruments are not a concern in this setting, but the AR test provides a model-free confirmation.

\section{Discussion}

The central finding---that strict regulatory enforcement \textit{reduces} nursing home staffing---challenges the conventional model of health care regulation as a quality-promoting force. The compliance crowding mechanism offers a unified explanation: deficiency citations impose real administrative costs (plans of correction, resurvey preparation, documentation) that consume the same nursing hours they are intended to protect. The turnover result adds a dynamic dimension: the regulatory burden is not merely a static reallocation but an ongoing friction that erodes the nursing workforce at cited facilities.

Three caveats temper these conclusions. First, the exclusion restriction may be violated if state-level characteristics---particularly Medicaid reimbursement rates, state-specific staffing mandates, and nurse labor market conditions---independently affect both surveyor stringency and facility staffing \citep{Hackmann2019}. The within-chain estimate controls for corporate management but not for the fact that a chain facility in Utah faces Utah's Medicaid rates while one in Texas faces Texas's. I note, however, that the balance tests show no correlation between the instrument and observable facility characteristics (beds, ownership), and the jackknife confirms no single state drives the result. Second, the 2SLS coefficient identifies a local average causal response for facilities whose measured severity is shifted by state stringency. These ``compliers'' may have systematically lower baseline staffing than the full population, limiting external validity for uniform policy reforms. Third, the cross-sectional design cannot establish temporal ordering: if annual staffing is measured concurrently with the annual survey, reverse causality (low staffing causing citations) remains a concern despite the IV. Longitudinal evidence on staffing changes \textit{after} specific survey events would strengthen the causal interpretation.

These results speak to a broader principle in regulatory design: the cost of compliance is itself a binding constraint, and optimal enforcement must account for the resources that citations consume. In nursing homes, where care quality is measured in nursing minutes per resident, the compliance burden falls directly on the margin it is supposed to improve.

\section{Conclusion}

State survey agency stringency creates a regulatory lottery for nursing homes. Facilities that draw strict surveyors receive more and worse deficiency citations, but rather than responding with quality investment, they shed nursing staff and experience higher turnover. The mechanism is compliance crowding: each citation generates administrative demands that crowd out clinical care. For the 1.3 million Americans living in nursing homes, the inspection lottery is not just unfair---it is counterproductive.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{Data Sources and Access} All data come from the CMS Provider Data Catalog (\url{https://data.cms.gov/provider-data/}), accessed March 30, 2026. No authentication is required. Four datasets were used:

\begin{enumerate}
\item \textbf{Health Deficiencies} (dataset ID: \texttt{r5ix-sfxw}): 418,972 individual deficiency citations from standard health inspections and complaint surveys, covering the three most recent inspection cycles per facility (survey dates from March 2017 to February 2026). Each record contains the facility CCN, survey date, deficiency tag number, scope-severity code (A--L), and whether the deficiency was corrected.

\item \textbf{Provider Information} (dataset ID: \texttt{4pq5-n9py}): 14,703 Medicare/Medicaid-certified nursing facilities. Key fields include certified bed count, average daily resident count, ownership type, chain name and ID, urban/rural classification, reported staffing HPRD by staff type (from Payroll-Based Journal), case-mix-adjusted staffing, nursing staff turnover rates, and CMS Five-Star ratings.

\item \textbf{Quality Measures --- MDS} (dataset ID: \texttt{djen-97ju}): 249,951 quality measure records derived from the Minimum Data Set, covering 14 long-stay measures (e.g., falls, pressure ulcers, UTIs, physical restraints) reported as four-quarter average scores.

\item \textbf{Penalties}: Penalty counts and amounts were obtained from the Provider Information dataset (fields: Number of Fines, Total Amount of Fines in Dollars, Total Number of Penalties).
\end{enumerate}

\subsubsection*{Variable Construction} The scope-severity code maps to a numeric 1--12 scale following the CMS classification matrix. Let $d_{ij}$ denote the severity score for deficiency citation $j$ at facility $i$ in the most recent survey year. The facility-level mean severity is $\bar{S}_i = \frac{1}{J_i} \sum_j d_{ij}$, where $J_i$ is the number of citations. The leave-one-out instrument is $Z_{s(i)} = \frac{\sum_{k \in s, k \neq i} \bar{S}_k}{N_s - 1}$, where $N_s$ is the number of facilities in state $s$.

\subsubsection*{Sample Construction} The main restriction is the requirement of a completed 2025 survey (drops 2,874 facilities without recent inspections, primarily small or newly certified homes). A further 400 facilities are missing PBJ staffing data, often the smallest or worst-performing facilities that fail to submit complete payroll records. While this selection is non-random, it biases \textit{against} finding a negative staffing effect: if the lowest-staffing facilities are excluded, the estimated compliance crowding effect is a lower bound.

\begin{tabular}{lr}
\toprule
Step & Facilities \\
\midrule
All certified nursing homes & 14,703 \\
With completed 2025 survey & 11,829 \\
Non-missing staffing data & 11,429 \\
Non-missing instrument and beds & 11,427 \\
Chain subsample & 8,192 \\
\bottomrule
\end{tabular}

\section{Identification Appendix}

\subsubsection*{Instrument Balance} The LOO state stringency instrument is uncorrelated with log bed count ($\hat{\beta}$ = 0.025, SE = 0.143) and for-profit ownership ($\hat{\beta}$ = --0.009, SE = 0.096). The urban indicator shows some correlation ($\hat{\beta}$ = --0.297, SE = 0.115), which is a limitation of the cross-sectional design.

\subsubsection*{Leave-One-State-Out Stability} Dropping each state in turn produces 52 IV point estimates ranging from --0.88 to --0.42 (mean --0.73, SD 0.06), confirming that no individual state drives the main result.

\section{Robustness Appendix}

See Table~\ref{tab:robustness} in the main text for all robustness specifications. The Anderson-Rubin $F$-statistic (5.12, $p$ = 0.028) confirms significance under weak-IV-robust inference, though the first-stage $F$ of 4,764 makes weak instruments a non-issue in this setting.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
