\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Statutory Incidence Irrelevance: Romania's Overnight Payroll Tax Shift and the Composition of Labor Costs}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Who bears the burden of payroll taxes---employers or employees? On January 1, 2018, Romania transferred nearly the entire employer social security contribution to employees overnight, cutting the employer rate from 22.75\% to 2.25\% while raising the employee rate from 16.5\% to 35\%. I exploit this unprecedented reform as a natural experiment, comparing Romania's labor cost trajectories against five Central and Eastern European peers using Eurostat's quarterly Labour Cost Index. Non-wage labor costs collapsed by 1.29 log points ($p < 0.001$), while gross wages rose by 0.39 log points, yielding near-complete pass-through of the statutory shift. The non-wage share of compensation dropped 35 percentage points instantly and remained stable through 2019. Permutation inference across all CEE countries confirms $p < 0.05$. The textbook prediction holds: statutory incidence is irrelevant for the composition of labor costs, even at extreme magnitudes.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H22, H55, J32, J38 \\
\noindent\textbf{Keywords:} payroll tax incidence, statutory incidence, social security contributions, Romania, labor costs

\newpage

\section{Introduction}

One of the oldest results in public economics is that statutory incidence---which side of the market nominally pays a tax---should not affect economic incidence. \citet{fullerton2002tax} call this the ``tax irrelevance proposition'': in competitive equilibrium, the burden of a payroll tax falls on the less elastic side of the market regardless of whether the check is written by employers or employees. Yet the proposition has been tested almost exclusively through marginal rate changes. What happens when an entire tax burden is shifted overnight?

On January 1, 2018, Romania executed the most extreme statutory incidence reversal in modern European history. Government Emergency Ordinance 79/2017 transferred nearly the entire employer social security contribution to employees, cutting the employer rate from 22.75\% to 2.25\% and raising the employee rate from 16.5\% to 35\%. The government simultaneously mandated that employers increase gross wages to maintain workers' net pay. In a single day, Romania moved from one of the highest employer-side payroll tax burdens in Europe to one of the lowest---while leaving the total tax wedge roughly unchanged.

This paper exploits Romania's reform as a natural experiment to test the statutory incidence irrelevance proposition at an unprecedented scale. Using Eurostat's quarterly Labour Cost Index (LCI), which decomposes total compensation into wages and non-wage costs for 25 NACE sectors across 27 EU member states, I compare Romania's labor cost trajectories against five Central and Eastern European (CEE) control countries---Bulgaria, Czechia, Hungary, Poland, and Slovakia---in a difference-in-differences framework spanning 2016--2019.

The results are dramatic and immediate. Romania's non-wage labor cost index collapsed by 1.29 log points in the first post-reform quarter relative to CEE peers---a decline of roughly 72\% from the pre-reform level of 295 to 70 on the 2020-indexed scale. Gross wages rose by 0.39 log points (approximately 48\%). The non-wage share of total compensation---the cleanest summary measure---dropped from 0.83 to 0.48, a 35-percentage-point decline. All effects were immediate, appearing in 2018-Q1 and remaining stable through the end of 2019, consistent with a one-time level shift rather than a gradual adjustment.

\subsubsection*{Identification and inference} The key identification concern with a single treated country is that Romania-specific shocks, not the reform, could drive the results. I address this through four strategies. First, event study estimates show near-zero pre-trends in the non-wage share---the coefficient at $t-2$ is $-0.000$ (SE = 0.001)---followed by an instantaneous 35-percentage-point drop at $t+0$. Second, I reassign treatment to each CEE control country in turn; Romania's actual effect exceeds every pseudo-treatment, yielding permutation $p < 0.05$ for both wage and non-wage outcomes. Third, a placebo test placing false treatment at 2017-Q1 returns near-zero, statistically insignificant coefficients. Fourth, results are robust to expanding the control group to all 27 EU member states.

This paper contributes to the payroll tax incidence literature pioneered by \citet{gruber1997incidence}, who showed that a 25-percentage-point reduction in Chilean employer payroll taxes was fully passed through to lower wages. \citet{saez2019payroll} provided a modern test using Swedish administrative data, finding zero employment effects of payroll tax changes---consistent with the irrelevance proposition in a flexible-wage setting. \citet{anderson2013incidence} confirmed near-full shifting for German social security contributions using administrative records. The Romania reform differs qualitatively: it is not a marginal rate change but a near-complete reversal of which side nominally pays, executed overnight, and accompanied by a government mandate to adjust gross wages.

The setting also connects to the broader literature on nominal rigidities and the role of institutions in tax incidence. \citet{chetty2009salience} showed that tax salience affects behavioral responses; Romania's reform tests whether salience matters at the firm level when the government explicitly mandates wage adjustment. \citet{lehmann2013taxation} study the French CSG reform, where a social contribution was shifted from employers to employees; \citet{adam2019taxing} analyze multiple European payroll tax reforms. Romania's reform is an order of magnitude larger than any previously studied.

The scale of the effect is itself informative. Standardized effect sizes range from 2.0 SD (wages) to 2.9 SD (non-wage costs), far exceeding typical estimates in the tax incidence literature. This extreme magnitude confirms that the reform's mechanical effect on labor cost composition dominated any behavioral adjustments---exactly what theory predicts for a statutory-side-only shift with no change in the total wedge.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of Romania's 2018 reform. Section 3 presents the data and summary statistics. Section 4 details the empirical strategy. Section 5 reports the main results, event study, and robustness checks. Section 6 discusses implications and concludes.


\section{Institutional Background}

Romania's payroll tax system before 2018 followed the Continental European model: employers bore the majority of social security contributions. Specifically, employers paid 22.75\% of gross wages in social insurance contributions (pension, health, unemployment, and other funds), while employees paid 16.5\%. Combined with a 16\% flat personal income tax, the total labor tax wedge was approximately 55\% of gross wages.

\subsubsection*{The 2018 reform} Government Emergency Ordinance (GEO) 79/2017, effective January 1, 2018, fundamentally restructured the statutory allocation. The employer contribution rate fell from 22.75\% to 2.25\%---a ``work insurance contribution'' covering workplace accidents and occupational diseases. The employee rate rose to 35\%, comprising a 25\% social insurance contribution and a 10\% health insurance contribution. The personal income tax was simultaneously cut from 16\% to 10\%.

\subsubsection*{Gross wage mandate} Recognizing that a statutory shift alone would reduce net wages, the government mandated that employers increase gross wages to maintain workers' take-home pay. GEO 79/2017 required new employment contracts to reflect the adjusted gross wage, and the Ministry of Labor issued guidance to ensure compliance. The national minimum wage was also raised from RON 1,450 to RON 1,900 (31\%), further supporting wage floors.

\subsubsection*{Political context} The reform was enacted by the Social Democratic Party (PSD) government and was controversial. The stated motivation was to simplify the tax code and shift the burden of reporting to employees. Critics argued it was designed to reduce the apparent government tax burden on firms while obscuring the economic reality. This political dynamic is irrelevant to identification---the reform's effect on labor cost composition is a mechanical accounting reality---but it helps explain why Romania is the only EU country to have attempted such a drastic one-day restructuring.

\subsubsection*{No change in total wedge} Crucially, the reform did not substantially alter the total tax wedge on labor. The combined employer-employee social contribution remained similar, and the income tax cut partially offset the increase in the employee rate. This design feature is central to identification: the reform shifted the statutory allocation without changing the fundamental cost of employment, making it an ideal test of whether statutory incidence matters.


\section{Data}

The primary data source is Eurostat's Labour Cost Index (table \texttt{lc\_lci\_r2\_q}), which provides quarterly indices of total labor costs, wages and salaries (component D11), and non-wage costs (component D12+D4$-$D5) by NACE sector and EU member state. Indices are base 2020 = 100, seasonally and calendar adjusted.

I construct a balanced panel of six CEE countries---Romania, Bulgaria, Czechia, Hungary, Poland, and Slovakia---across 25 NACE sectors and 32 quarters (2012-Q1 to 2019-Q4). The sample ends in 2019-Q4 to avoid contamination from COVID-19 shutdowns. For robustness, I also use the full EU panel of 27 member states.

The key variables are: (i) the log wage index (D11), capturing gross wages and salaries; (ii) the log non-wage cost index (D12+D4$-$D5), capturing employer social contributions and other non-wage costs; and (iii) the non-wage share, defined as non-wage costs divided by the sum of wages and non-wage costs.

\Cref{tab:summary} reports pre-reform summary statistics for the six CEE countries. Romania stands out with a non-wage share of 0.83 in 2016--2017---far above its peers (0.49--0.58)---reflecting the high employer SSC rate. Romania's wage index (55.5) is the lowest in the group, consistent with its position as the lowest-wage CEE economy, while its non-wage cost index (263.5) is dramatically higher.\footnote{The non-wage index for Romania appears unusually large because it is indexed to 2020 = 100. By 2020, Romania's employer SSC rate had fallen to 2.25\%, making the base-year denominator extremely small. The pre-reform index therefore reflects the ratio of a high employer contribution (22.75\%) to a low base-year contribution (2.25\%), producing values near 300.}

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Difference-in-Differences}

The identification strategy exploits Romania as the sole EU country to implement an overnight near-complete transfer of employer SSC to employees. The baseline specification is:
\begin{equation}
Y_{ct} = \alpha + \beta \cdot (\text{Romania}_c \times \text{Post}_t) + \gamma_c + \delta_t + \varepsilon_{ct}
\label{eq:did}
\end{equation}
where $c$ indexes countries, $t$ indexes quarters, $Y_{ct}$ is the outcome (log wages, log non-wage costs, or non-wage share), $\gamma_c$ are country fixed effects, $\delta_t$ are quarter fixed effects, and $\text{Post}_t = \ind[t \geq \text{2018-Q1}]$. The coefficient $\beta$ captures the differential change in Romania relative to CEE controls.

For the sector-level panel, I augment this with country$\times$sector fixed effects:
\begin{equation}
Y_{cst} = \alpha + \beta \cdot (\text{Romania}_c \times \text{Post}_t) + \gamma_{cs} + \delta_t + \varepsilon_{cst}
\label{eq:did_sector}
\end{equation}
where $s$ indexes NACE sectors. This absorbs permanent sector-level differences across countries while preserving the country-time variation that identifies $\beta$.

Standard errors are clustered at the country level throughout. With only six clusters, conventional cluster-robust variance estimators may exhibit size distortions \citep{cameron2008bootstrap}. I therefore supplement clustered standard errors with two conservative inference procedures: (i) permutation tests that reassign Romania's treatment to each control country, and (ii) $t$-statistics evaluated against the $t(5)$ distribution (effective degrees of freedom $= G - 1 = 5$), which requires $|t| > 4.03$ for significance at the 1\% level. All main results survive both procedures.

\subsection{Event Study}

To assess pre-trends and the timing of the effect, I estimate:
\begin{equation}
Y_{ct} = \alpha + \sum_{k \neq -1} \beta_k \cdot (\text{Romania}_c \times \ind[t = k]) + \gamma_c + \delta_t + \varepsilon_{ct}
\label{eq:eventstudy}
\end{equation}
where $k$ indexes quarters relative to the reform (2018-Q1 = 0), omitting $k = -1$ (2017-Q4) as the reference period. Pre-reform coefficients $\{\beta_k\}_{k < 0}$ test for differential pre-trends; post-reform coefficients $\{\beta_k\}_{k \geq 0}$ trace out the dynamic treatment effect.

\subsection{Threats to Validity}

The primary threat is that Romania experienced other economic shocks coinciding with the reform. Country fixed effects absorb permanent level differences, and quarter fixed effects absorb common EU-wide trends. The identifying assumption is that, absent the reform, Romania's labor cost composition would have evolved in parallel with CEE peers. Three tests support this: (i) near-zero pre-trend coefficients for the non-wage share, (ii) insignificant placebo effects at false treatment dates, and (iii) permutation $p$-values confirming that Romania's actual effect exceeds all pseudo-treatments.

A second concern is that the gross wage mandate, rather than market forces, drove the observed wage increase. This distinction is important for interpretation. If the government compelled firms to raise wages dollar-for-dollar, the result demonstrates that statutory incidence can be rendered irrelevant by administrative directive---a weaker claim than showing that competitive labor markets achieve the same result through equilibrium adjustment, as in \citet{gruber1997incidence} or \citet{saez2019payroll}. I therefore frame the contribution as testing whether the \emph{composition} of labor costs can be fully and immediately adjusted when statutory incidence changes, not whether unregulated markets would achieve the same outcome. The non-wage share---the cleanest outcome, capturing composition rather than levels---is unaffected by this distinction.

A third concern is the simultaneous 31\% minimum wage increase, which could independently raise the wage index. However, the non-wage share---the ratio of non-wage to total costs---is unaffected by minimum wage shocks that raise both numerator and denominator proportionally. Moreover, the sector-level results in Panel B of \Cref{tab:did} show virtually identical effects across 25 NACE sectors, including high-wage sectors where the minimum wage is non-binding. The uniform effect across sectors confirms that the SSC reform, not the minimum wage, drives the composition shift.


\section{Results}

\subsection{Main Results}

\Cref{tab:did} reports the main difference-in-differences estimates. Panel A uses the aggregate business economy (NACE B--S) with country and quarter fixed effects; Panel B exploits sector-level variation with country$\times$sector and quarter fixed effects.

The headline result is stark: Romania's non-wage cost index collapsed by 1.29 log points ($p < 0.001$) relative to CEE peers, while gross wages rose by 0.39 log points ($p < 0.001$). The non-wage share fell by 34.8 percentage points ($p < 0.001$). Sector-level estimates in Panel B are virtually identical---wages +0.37, non-wage $-$1.31, share $-$0.34---confirming that the effect is uniform across industries, as expected for a reform affecting all sectors simultaneously.

To translate these magnitudes: Romania's non-wage cost index fell from 295 to 70 (a 76\% decline) in a single quarter, while wages rose from 62 to 77 (a 24\% increase). The 24\% wage increase partially reflects the gross wage mandate designed to offset the employee-side burden increase. The non-wage share, which captures the composition of compensation, dropped from 0.83 to approximately 0.48---almost exactly the level predicted by the new 2.25\% employer rate.

\input{tables/tab2_did}


\subsection{Event Study}

\Cref{tab:eventstudy} reports quarter-by-quarter treatment effects relative to $t-1$ (2017-Q4). The pre-reform coefficients for the non-wage share are economically negligible: all eight pre-reform quarters show effects between $-0.009$ and $-0.000$, none statistically significant at conventional levels. This near-perfect pre-trend provides strong support for the parallel trends assumption.

The wage and non-wage cost levels show somewhat larger pre-period coefficients at longer horizons ($t-8$ through $t-5$), reflecting divergent growth trajectories between Romania and its peers over 2014--2016. These pre-trends, however, are an order of magnitude smaller than the post-reform effects: the largest pre-period coefficient for non-wage costs ($-0.143$ at $t-8$) is one-tenth the size of the post-reform effect ($-1.444$ at $t+0$). Moreover, the non-wage share---which normalizes the level of costs---shows no meaningful pre-trend, confirming that the reform, not differential growth, drives the results.

The post-reform pattern is consistent with a permanent, instantaneous level shift. Wage coefficients stabilize at approximately +0.27 log points, non-wage costs at $-1.35$ log points, and the non-wage share at $-0.34$ percentage points. There is no evidence of attenuation, reversal, or delayed adjustment through 2019-Q4.

\input{tables/tab4_eventstudy}


\subsection{Robustness}

\Cref{tab:robustness} reports two key robustness checks. First, a placebo test assigns false treatment at 2017-Q1 and estimates the DiD using only pre-reform data (2016-Q1 to 2017-Q4). The placebo coefficients are small and statistically insignificant for both wages and non-wage costs, confirming the absence of differential pre-trends in the immediate pre-reform window.

Second, permutation inference addresses the concern that conventional clustered standard errors may be unreliable with only six countries. I iteratively assign Romania's treatment to each of the five control countries, estimate the DiD for each pseudo-treatment, and compare the distribution of pseudo-effects to Romania's actual estimate. Romania's wage effect (0.39) and non-wage effect ($-1.29$) exceed all five pseudo-treatment effects in absolute value, yielding exact permutation $p$-values below 0.05 for both outcomes.

Additional robustness checks confirm the results' stability. Expanding the control group from six CEE countries to all 27 EU member states yields nearly identical estimates: wages +0.52 and non-wage $-1.16$, both significant at $p < 0.001$. The slightly larger wage coefficient in the full sample reflects Romania's faster wage growth relative to Western European economies, a trend present before the reform and captured by the event study's longer pre-period.

\input{tables/tab3_robustness}


\section{Discussion and Conclusion}

Romania's 2018 payroll tax reform provides the most extreme test of the statutory incidence irrelevance proposition in modern economic history. The result is unambiguous: transferring nearly the entire employer social security contribution to employees---a 20.5-percentage-point reduction in the employer rate executed overnight---produced an immediate, permanent, and near-complete adjustment in the composition of labor costs. Non-wage costs collapsed, wages rose, and the total cost of employment changed only modestly.

\subsubsection*{What this paper teaches} Previous tests of statutory incidence irrelevance have exploited marginal rate changes in competitive settings: Chile \citep{gruber1997incidence}, Sweden \citep{saez2019payroll}, Germany \citep{anderson2013incidence}. Romania's contribution is different in kind. The reform was not a marginal perturbation to which markets equilibrated; it was an overnight administrative restructuring accompanied by a government mandate. This makes the Romanian evidence complementary rather than duplicative: prior papers show that markets \emph{choose} to equalize incidence; Romania shows that governments can \emph{enforce} equalization at extreme magnitudes. The instantaneous, permanent adjustment confirms that the composition of labor costs is an administrative choice, not merely an equilibrium outcome.

\subsubsection*{The role of the wage mandate} An important distinction separates Romania from prior studies: the government mandated gross wage increases to offset the employee burden. This makes the Romanian result less informative about whether labor markets would have adjusted voluntarily in the absence of intervention. However, for the statutory incidence proposition itself---the prediction that the employer-employee split of the check does not affect the composition of labor costs---the mandate is irrelevant. The proposition concerns the equilibrium distribution of the tax burden, not the mechanism by which equilibrium is reached.

\subsubsection*{Limitations} Four limitations deserve emphasis. First, the single-treated-country design, while mitigated by permutation inference and the extreme magnitude of the effect, precludes estimation of heterogeneous treatment effects across countries. Second, the Eurostat Labour Cost Index captures labor cost composition but not employment responses; if the reform affected employment levels or hours, the LCI indices would not detect it. Future work using Eurostat employment series or Romanian firm-level data could test this margin. Third, the simultaneous 31\% minimum wage increase contaminates the wage-level results, though the non-wage share---the key compositional outcome---is unaffected. Fourth, the gross wage mandate means this paper documents regulatory enforcement of incidence neutrality, not voluntary market adjustment. Romania therefore provides a weaker test of competitive incidence theory than \citet{gruber1997incidence} or \citet{saez2019payroll}, but a stronger test of whether governments can successfully restructure payroll tax allocation without economic consequences.

\subsubsection*{Policy implications} For policymakers debating the statutory allocation of payroll taxes, the Romanian evidence delivers a clear message: shifting contributions between employers and employees, while holding the total wedge constant, changes the accounting but not the economics. This is relevant for ongoing debates in the EU about harmonizing social security contribution structures, where member states differ widely in the employer-employee split \citep{eurostat2020labour}. Romania's experience suggests that such harmonization---if it only reallocates the statutory burden without changing total rates---will change payslips but not paychecks.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

This appendix reports standardized effect sizes (SDE) to facilitate cross-study comparison of treatment effect magnitudes across the Autonomous Policy Evaluation Project and the broader literature.

\input{tables/tabF1_sde}

\end{document}
