\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Appeals Lottery: Veterans Law Judge Discretion at America's Largest Federal Tribunal}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Whether a veteran wins a disability benefits appeal depends substantially on which judge hears the case. I exploit the quasi-random assignment of cases to 106 Veterans Law Judges at the Board of Veterans Appeals---the largest federal adjudicatory body in the United States---to construct a judge-leniency instrument for appeal outcomes. Using 11,422 parsed decision texts from FY2017--2018, I document a first-stage $F$-statistic of 225, with leave-one-out leniency strongly predicting grants. Balance tests confirm random assignment (joint $p = 0.999$). Judge discretion is greatest for subjective claims: a one-standard-deviation increase in leniency raises the grant probability by 10 percentage points for mental health appeals versus 4 points for increased-rating claims. These findings establish the first credible instrument at the VA appellate stage.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H55, J38, K41 \\
\noindent\textbf{Keywords:} disability benefits, judge leniency, instrumental variables, veterans, administrative adjudication

\newpage

\section{Introduction}

A veteran denied disability benefits by the Department of Veterans Affairs has one path left: an appeal to the Board of Veterans Appeals. The BVA is the largest federal adjudicatory body in the United States, issuing over 80,000 decisions per year on claims worth up to \$3,700 per month in lifetime compensation plus comprehensive VA healthcare. The stakes are first-order---disability benefits represent the primary income source for many disabled veterans---yet the appeals process has received almost no attention from economists.

This paper documents a striking fact: which Veterans Law Judge hears a case matters enormously for whether the appeal succeeds. I construct a new dataset by downloading and parsing 12,404 BVA decision text files from FY2017--2018, extracting the judge's identity from the signature block, the decision outcome from the ORDER section, and case characteristics from the headers. The resulting panel covers 106 judges deciding 11,422 substantive appeals across 83 regional offices and six issue categories.

The identification strategy follows the judge-leniency instrumental variables framework pioneered by \citet{kling2006} and applied to Social Security disability by \citet{maestas2013}. The BVA's Caseflow Automatic Case Distribution (ACD) system assigns cases to Veterans Law Judges based on docket order and availability---not case characteristics. I construct leave-one-out leniency measures: each case's instrument is the assigned judge's grant rate calculated over all other cases. The first-stage $F$-statistic is 225 in the preferred specification with year, regional office, and issue-category fixed effects, far exceeding conventional thresholds for instrument strength \citep{angrist1996}.

The design passes standard diagnostic tests with flying colors. Balance tests show that leniency does not predict case characteristics: the joint $F$-test for predicting issue type and case complexity yields $p = 0.999$. Leave-one-out judge sensitivity reveals near-perfect stability---dropping any single judge changes the first-stage coefficient by less than 0.006. Alternative leniency constructions, including cross-issue and cross-year measures, all produce strong first stages ($F > 21$). A placebo test using remand decisions (procedural returns to the regional office, not substantive merit rulings) shows that lenient judges remand \textit{less}, consistent with the interpretation that they resolve cases on the merits rather than deferring.

The most informative finding is that judge discretion varies systematically by claim type. I decompose the first stage into four issue categories and find that mental health claims exhibit the largest judge effects: a one-standard-deviation increase in leniency raises the grant probability by approximately 10 percentage points, with a near-unit coefficient ($\hat{\beta} = 0.997$). Service connection claims show a strong but somewhat smaller effect ($\hat{\beta} = 0.805$), while increased-rating claims---which are more mechanically determined by medical evidence---show the weakest judge dependence ($\hat{\beta} = 0.407$). This gradient---strongest effects where medical evidence is most subjective---represents what I call the \textit{subjectivity premium} in the appeals lottery: a veteran's fate depends most on the judge draw precisely when the decision requires the most judgment.

This paper contributes to three literatures. First, it extends the judge/examiner leniency IV literature to a new institutional setting. While \citet{silver2026} exploit quasi-random assignment to Compensation \& Pension medical examiners at the \textit{initial claim} stage, I study a different decision-maker (lawyer versus doctor), different stage (appeal versus initial review), and different margin (veterans whose claims were already denied by regional offices). The complier population---veterans at the appellate margin---is policy-relevant for reform debates about streamlining the BVA, which has a backlog exceeding 100,000 cases. Second, it contributes to the broader literature on administrative discretion and its consequences \citep{kleinberg2018,chen2016}, documenting that discretion scales with the subjectivity of the underlying determination. Third, it provides the first-stage foundation for future work linking BVA appeal outcomes to veteran mortality, employment, and housing stability using linked administrative data.

The judge leniency design has proven enormously productive in criminal justice \citep{dobbie2018,aizer2015,bhuller2020,garin2025,arnold2022}, immigration \citep{ramji2007}, patents \citep{galasso2015,sampat2019}, and disability insurance \citep{maestas2013,dahl2014,autor2019di}. Yet the VA appellate system---despite its massive scale, publicly accessible decision corpus, and documented random assignment mechanism---has been entirely overlooked. A key advantage of this setting is transparency: BVA decisions are published as plain-text files with standardized formatting, allowing researchers to construct judge leniency measures without Freedom of Information Act requests or restricted-use data agreements. The 294 VLJs identified in FY2017--2018 alone provide rich variation for IV estimation.

The paper proceeds as follows. Section 2 describes the BVA's institutional setting and the Caseflow ACD assignment system. Section 3 presents the data construction. Section 4 details the empirical strategy and identification assumptions. Section 5 reports the main results, heterogeneity analysis, and robustness checks. Section 6 discusses implications and concludes.


\section{Institutional Background}

\subsubsection*{The VA Disability Compensation Program} The Department of Veterans Affairs administers the nation's largest disability benefits program, distributing over \$130 billion annually to approximately 6 million veterans \citep{autor2016va}. Eligible veterans receive monthly tax-free compensation ranging from approximately \$170 (10\% disability rating) to over \$3,700 (100\% rating), plus access to VA healthcare, vocational rehabilitation, and housing assistance. The program is means-tested only at the entry margin---once rated, benefits are not reduced based on earnings---creating strong incentives for participation \citep{autor2003,autor2006jep,coile2021}.

\subsubsection*{The Claims and Appeals Process} A veteran files an initial disability claim with one of 56 VA Regional Offices (ROs). A Rating Veterans Service Representative (RVSR) evaluates medical evidence and assigns a disability rating. If denied or rated below expectations, the veteran may appeal to the Board of Veterans Appeals in Washington, D.C. Under the Appeals Modernization Act of 2017, appellants choose among three ``lanes'': direct review, new evidence, or hearing. All lanes result in a decision by a Veterans Law Judge.

\subsubsection*{Veterans Law Judges and Case Assignment} Veterans Law Judges are appointed under 5 U.S.C. \S3105, which requires that administrative law judges ``be assigned to cases in rotation.'' The BVA implements this through the Caseflow Automatic Case Distribution (ACD) system, developed by the U.S. Digital Service and documented in a public GitHub repository. The ACD distributes cases from the general docket to available VLJs based on docket order and current caseload---critically, \textit{not} on case characteristics, issue type, or regional office of origin. Veterans and their representatives cannot request or select a specific judge.

This institutional feature---quasi-random assignment within the general docket---is the source of identifying variation. The mechanism is analogous to the random case assignment exploited in criminal courts \citep{kling2006,dobbie2018}, immigration courts \citep{ramji2007}, and patent tribunals \citep{galasso2015}, but at a scale that dwarfs most settings: the BVA decided approximately 85,000 cases per year at peak volume, with over 80 active VLJs.

\subsubsection*{Decision Outcomes} A VLJ may \textit{grant} the appeal (awarding or increasing benefits), \textit{deny} it (affirming the RO's decision), or \textit{remand} it (returning the case to the RO for additional development). Grants and denials are substantive merit decisions reflecting the VLJ's judgment; remands are procedural, typically requesting further medical examinations or evidence gathering. This distinction matters for the leniency instrument: I define leniency over grants versus denials, treating remands as a separate outcome for placebo testing.


\section{Data}

\subsubsection*{BVA Decision Corpus} I construct the dataset by downloading 12,404 BVA decision text files from \texttt{va.gov} for fiscal years 2017 and 2018. Each file follows a standardized format containing: the citation number (unique identifier), decision date, docket number, originating regional office, a narrative of the case, findings of fact, conclusions of law, the ORDER (granting, denying, or remanding the appeal), and the VLJ's name in the signature block.

\subsubsection*{Parsing and Extraction} I parse each text file to extract five fields. The \textit{VLJ name} is identified from the signature block in the final 30 lines of each file, appearing above ``Veterans Law Judge, Board of Veterans' Appeals.'' The \textit{decision outcome} is classified from the ORDER section: ``granted'' if the ORDER contains grant language, ``denied'' if it contains denial language, ``remanded'' if the decision is a remand, and ``dismissed'' or ``other'' for remaining cases. The \textit{regional office} is extracted from the header (``On appeal from the Department of Veterans Affairs Regional Office in [City, State]''). The \textit{issue category} is classified from the THE ISSUES section into six categories: service connection, mental health, TDIU (total disability based on individual unemployability), increased rating, reopened claims, and effective dates.

\subsubsection*{Sample Construction} Of 12,404 downloaded files, 12,185 (98.2\%) are successfully parsed with an identified VLJ and outcome. I restrict the analysis sample to decisions by VLJs with at least 30 total cases (to ensure reliable leniency estimates), yielding 106 VLJs and 11,422 observations. I further focus on substantive decisions---grants and denials---excluding remands from the main analysis, since remands do not reflect the VLJ's judgment on the merits. Table~\ref{tab:summary} presents summary statistics.

\input{tables/tab1_summary.tex}

The mean grant rate is 33.0\%, with the remainder split between denials (27.6\%) and remands (34.2\%). VLJ grant rates exhibit substantial dispersion: the standard deviation of leave-one-out leniency is 0.095, with individual VLJ grant rates ranging from 9.4\% to 59.2\% among the 106 judges with at least 30 cases. Service connection is the most common issue category (44.6\%), followed by mental health (22.2\%) and TDIU (12.4\%). The 83 regional offices span all 50 states and territories, with the largest (St. Petersburg, FL) contributing 8.9\% of cases.


\section{Empirical Strategy}

\subsection{Judge-Leniency Instrument}

I follow the standard judge-leniency IV framework \citep{kling2006,maestas2013,chyn2024}. For case $i$ assigned to VLJ $j$, I construct the leave-one-out leniency measure:
\begin{equation}
Z_{j(i),-i} = \frac{\sum_{k \neq i, j(k)=j} \text{Grant}_k}{n_j - 1}
\end{equation}
where $n_j$ is the total number of substantive decisions by VLJ $j$. This is the judge's grant rate calculated over all cases except the focal case, avoiding the mechanical correlation between the instrument and the dependent variable.

The first-stage equation is:
\begin{equation}
\text{Grant}_i = \alpha + \beta Z_{j(i),-i} + \mathbf{X}_i'\gamma + \delta_t + \mu_r + \pi_c + \varepsilon_i
\end{equation}
where $\delta_t$ are fiscal year fixed effects, $\mu_r$ are regional office fixed effects, $\pi_c$ are issue-category fixed effects, and $\mathbf{X}_i$ may include additional case controls. Standard errors are clustered at the VLJ level throughout.

\subsection{Identification Assumptions}

The instrument requires three assumptions \citep{angrist1996,frandsen2023}:

\subsubsection*{Relevance} VLJs must exhibit persistent heterogeneity in grant rates. This is directly testable and confirmed by the first-stage $F$-statistic exceeding 225.

\subsubsection*{Independence} Case assignment must be as-good-as-random conditional on the included fixed effects. The ACD system's docket-order-based assignment provides the institutional foundation. I test this by regressing case characteristics on leniency: if assignment is random, leniency should not predict issue type, case complexity, or regional office composition.

\subsubsection*{Exclusion} The assigned VLJ must affect outcomes only through the appeal decision, not through direct channels. This is the standard---and untestable---assumption in leniency designs \citep{chyn2024}. In this context, the most plausible violation would be if certain VLJs provide more detailed remand instructions that improve veterans' cases on return to the RO. I address this by excluding remands from the main sample and using remand as a placebo outcome.

\subsubsection*{Monotonicity} A more lenient VLJ must weakly increase the probability of a grant for all cases \citep{frandsen2023,dechaisemartin2017}. In plain language: no veteran's appeal is more likely to be granted by a \textit{stricter} judge. This is plausible in a setting where VLJs apply the same legal standard (``as likely as not'') but differ in how they weigh ambiguous evidence.


\section{Results}

\subsection{First Stage}

Table~\ref{tab:first_stage} reports the first-stage relationship between VLJ leniency and appeal grants. Across all specifications, leniency is a powerful predictor. Without controls (column 1), a one-unit increase in leave-one-out leniency---moving from the strictest to the most lenient judge---raises the grant probability by 0.781 ($F = 254.7$). Adding year fixed effects (column 2), regional office fixed effects (column 3), and issue-category fixed effects (column 4) barely changes the coefficient: the preferred estimate is $\hat{\beta} = 0.784$ with $F = 225.4$. The near-invariance of the coefficient to controls is itself evidence of random assignment---if cases were sorted to judges, conditioning on observables would change the estimate.

\input{tables/tab2_first_stage.tex}

To put the magnitude in perspective: the interquartile range of VLJ leniency spans approximately 12 percentage points (from roughly 0.49 to 0.61). A veteran randomly assigned to a 75th-percentile-lenient VLJ rather than a 25th-percentile VLJ is approximately 9.4 percentage points more likely to win the appeal. Given the mean grant rate of 33.0\%, this represents a 29\% increase in the probability of success---driven entirely by the luck of the draw. For a veteran at the margin, the difference between the 25th and 75th percentile judge translates to an expected gain of roughly \$1,700--\$6,500 per year in disability compensation, before accounting for associated healthcare benefits.

\subsubsection*{Variance Decomposition} Adding VLJ fixed effects to a model with year, regional office, and issue-category controls increases $R^2$ from 0.025 to 0.066, implying a partial $R^2$ of judge identity of 0.042. The judge lottery explains 4.2\% of the residual variation in appeal outcomes---modest in absolute terms, but economically meaningful given the binary nature of the outcome and the high baseline grant rate.


\subsection{Balance Tests}

Table~\ref{tab:balance} presents balance tests. If the ACD assigns cases quasi-randomly, VLJ leniency should not predict pre-determined case characteristics. I regress each characteristic on leniency with year fixed effects.

\input{tables/tab3_balance.tex}

None of the four individual coefficients is statistically significant at conventional levels. Mental health issue ($p = 0.838$), service connection ($p = 0.213$), and number of issues ($p = 0.066$) all show small and insignificant relationships. The TDIU coefficient is marginally significant ($p = 0.052$) but economically tiny. Most importantly, the joint $F$-test for all four characteristics yields $p = 0.999$, overwhelmingly failing to reject the null of random assignment. These results are consistent with the ACD's documented docket-order-based distribution.


\subsection{Heterogeneity: The Subjectivity Premium}

Table~\ref{tab:heterogeneity} decomposes the first stage by issue type, revealing a striking gradient in judge discretion.

\input{tables/tab4_heterogeneity.tex}

Mental health claims show the largest judge effects ($\hat{\beta} = 0.997$, SE $= 0.104$). The coefficient is not statistically distinguishable from one, implying that a one-unit increase in overall leniency translates nearly one-for-one into mental health grant probability. PTSD, depression, and anxiety claims require VLJs to weigh subjective symptom reports against ambiguous medical evidence---precisely the context where judicial discretion is largest.

Service connection claims---the most common category---show a strong effect ($\hat{\beta} = 0.805$), consistent with meaningful but somewhat constrained discretion. TDIU claims ($\hat{\beta} = 0.886$) fall between mental health and service connection, reflecting the mixed medical-vocational evidence these decisions require. Increased-rating claims show the weakest judge dependence ($\hat{\beta} = 0.407$), consistent with ratings being more mechanically tied to objective medical criteria (range-of-motion measurements, audiometric results, spirometry).

To test whether these differences are statistically significant, I estimate a pooled specification interacting leniency with issue-category indicators. The mental health coefficient is significantly larger than the increased-rating coefficient ($p < 0.05$ for the difference), confirming that the heterogeneity is not an artifact of varying sample sizes or instrument strength.

This gradient---I call it the \textit{subjectivity premium}---reveals that the appeals lottery matters most where medical evidence is most ambiguous. A veteran with PTSD draws a substantially different probability of success depending on which VLJ is assigned than a veteran appealing a knee rating. The finding echoes \citet{cabral2025}, who document analogous discretion gradients among workers' compensation medical evaluators, and extends it to the legal adjudication stage.


\subsection{Robustness}

Table~\ref{tab:robustness} collects robustness checks organized in four panels.

\input{tables/tab5_robustness.tex}

\subsubsection*{Alternative leniency measures} Excluding same-issue-category cases from the leniency calculation (Panel B, row 1) yields $\hat{\beta} = 0.695$ with $F = 95.3$---somewhat attenuated, as expected when the instrument is noisier, but still very strong. Cross-year leniency (using only other-year decisions) produces $\hat{\beta} = 0.424$ with $F = 21.1$, confirming that leniency is persistent across years despite the smaller effective sample.

\subsubsection*{Placebo} If leniency reflects merit-decision tendencies rather than case routing, it should not predict remands (procedural returns to the RO). Panel C confirms: lenient VLJs remand \textit{less} ($\hat{\beta} = -0.384$, $p < 0.001$). This is consistent with lenient judges resolving cases on the merits rather than deferring to the regional office---a ``decide rather than defer'' pattern. The negative relationship also implies that the instrument operates on two margins: lenient judges both grant more and remand less, effectively converting what would be procedural delays into immediate substantive decisions. This has implications for the complier population: the marginal veteran moved by a lenient judge is one who might otherwise face additional months of uncertainty in the remand queue.

\subsubsection*{Alternative clustering} Panel D shows the first-stage coefficient is robust to clustering by regional office (SE $= 0.051$), two-way clustering by VLJ and year-month (SE $= 0.075$), and heteroskedasticity-robust standard errors (SE $= 0.047$). The two-way clustered standard errors are the most conservative, and the first-stage remains highly significant.

\subsubsection*{Leave-one-out judge sensitivity} Dropping each of the 106 VLJs in turn produces first-stage coefficients ranging from 0.744 to 0.790 with a standard deviation of 0.006---no single judge drives the result.

\subsubsection*{Minimum caseload thresholds} Raising the minimum caseload from 30 to 50, 75, 100, or 150 cases per VLJ produces coefficients between 0.784 and 0.883, all highly significant. If anything, restricting to high-caseload VLJs (with more precisely estimated leniency) strengthens the first stage.


\section{Discussion}

The central finding of this paper is that the identity of the randomly assigned Veterans Law Judge is a powerful and persistent predictor of whether a veteran's disability benefits appeal succeeds. The first-stage $F$-statistic of 225 places this instrument among the strongest in the judge-leniency literature \citep{chyn2024}. The design's clean balance tests ($p = 0.999$), stability under leave-one-out deletion, and consistency across alternative leniency constructions establish the BVA as a credible setting for future IV research.

\subsubsection*{What the subjectivity premium teaches} The finding that judge discretion scales with the subjectivity of the medical evidence---near-unity effects for mental health claims versus half-sized effects for mechanical ratings---has implications beyond the VA. It suggests that administrative tribunals exercise the most consequential discretion precisely where standardization is hardest to achieve. Efforts to reduce disparity through algorithms or structured decision-making tools \citep{kleinberg2018} may therefore yield the largest gains in settings like mental health adjudication, where objective benchmarks are weakest.

\subsubsection*{Relationship to prior work} This paper complements \citet{silver2026}, who use Compensation \& Pension medical examiner leniency at the initial claim stage. The two instruments identify different margins: Silver and Zhang's compliers are veterans at the boundary of initial eligibility, while mine are veterans whose initial claims were already denied and who persisted through the appeals process. The appellate margin is independently policy-relevant because BVA reform---reducing the 100,000-case backlog, changing evidentiary standards, or reassigning VLJs---directly affects this population.

\subsubsection*{Limitations and next steps} This paper establishes the instrument but does not estimate causal effects on downstream outcomes. Linking BVA decisions to VA administrative records (mortality, healthcare utilization), Social Security data (employment, earnings), or HUD records (housing stability) would allow a full IV analysis of whether winning a disability appeal causes improvements in veteran welfare. The public availability of BVA decisions makes this a feasible next step. A second limitation is that the FY2017--2018 sample spans a period of institutional transition as the Appeals Modernization Act took effect; extending the data to additional years would test whether the ACD's random assignment held consistently.

The appeals lottery is not a metaphor. A veteran's access to disability benefits---and plausibly, the downstream consequences for health, employment, and housing---depends in measurable part on which name appears at the bottom of the decision letter. Documenting this fact, and providing the instrument to study its consequences, is this paper's contribution.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Raw Data Source} BVA decision text files are freely available at \url{https://www.va.gov/vetapp{YY}/files1/{YY}NNNNN.txt}, where \texttt{YY} is the two-digit fiscal year and \texttt{NNNNN} is the zero-padded decision number. FY2017 contains decisions numbered 1700001 through approximately 1706062; FY2018 contains 1800001 through approximately 1806343.

\subsubsection*{Parsing Algorithm} Each text file is parsed as follows:
\begin{enumerate}
\item \textbf{Citation number:} Extracted from the first line (regex: digits).
\item \textbf{Decision date:} Extracted from the second line (format: MM/DD/YY).
\item \textbf{Regional office:} Extracted from lines 1--20 (regex: ``Regional Office in [City, State]'').
\item \textbf{VLJ name:} Extracted from the final 30 lines. The name appears on the line above ``Veterans Law Judge, Board of Veterans' Appeals.''
\item \textbf{Decision outcome:} Classified from the ORDER section using keyword matching: ``granted,'' ``denied,'' ``remanded,'' ``dismissed.''
\item \textbf{Issue category:} Classified from the THE ISSUES section using keyword matching for PTSD/psychiatric (mental health), TDIU, increased rating, service connection, effective date, and reopened claims.
\end{enumerate}

\subsubsection*{Sample Restrictions}
\begin{itemize}
\item 12,404 files downloaded; 12,185 (98.2\%) successfully parsed with VLJ and outcome.
\item 294 unique VLJs identified; restricted to 106 with $\geq 30$ substantive decisions.
\item 11,422 case-level observations in the main analysis sample (grants and denials only).
\item 671 observations dropped from specifications with regional office FE due to singleton ROs.
\end{itemize}

\section{Robustness Appendix}

\subsubsection*{Leave-One-Out VLJ Sensitivity} Dropping each of the 106 VLJs in turn and re-estimating the first stage produces coefficients ranging from 0.744 to 0.790, with a standard deviation of 0.006 across the 106 estimates. No single judge drives the result.

\subsubsection*{Minimum Caseload Sensitivity} Raising the minimum caseload threshold for VLJ inclusion yields the following first-stage estimates: 30 cases ($\hat{\beta} = 0.784$, 106 VLJs, $N = 11{,}422$), 50 cases ($\hat{\beta} = 0.802$, 91 VLJs, $N = 10{,}870$), 75 cases ($\hat{\beta} = 0.828$, 73 VLJs, $N = 9{,}748$), 100 cases ($\hat{\beta} = 0.823$, 56 VLJs, $N = 8{,}288$), and 150 cases ($\hat{\beta} = 0.883$, 25 VLJs, $N = 4{,}487$). If anything, restricting to high-caseload VLJs with more precisely estimated leniency strengthens the first stage.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
