\begin{table}[!h]
\centering
\caption{\label{tab:sde}Standardized Distributional Effects}
\centering
\begin{tabular}[t]{lrrrrrl}
\toprule
Outcome & $\hat{\beta}$ & SE & SD($Y$) & SDE & SE(SDE) & Classification\\
\midrule
Convicted (Drug Trafficking) & 0.9701 & 0.0291 & 0.4405 & 2.2025 & 0.066 & Large\\
\bottomrule
\end{tabular}
\end{table}
\begin{flushleft}\small
\textbf{Country:} Brazil. \textbf{Research question:} Does legal indeterminacy amplify arbitrary punishment under random case assignment? \textbf{Policy mechanism:} Lei 11.343/2006 provides no objective quantity thresholds for distinguishing drug users from traffickers, granting judges broad discretion over a 5-year minimum sentence. \textbf{Outcome:} Binary conviction (\textit{Proced\^encia}) from judicial movement records. \textbf{Treatment:} Assignment to a more vs.\ less severe criminal vara via electronic lottery (\textit{sorteio}). \textbf{Data:} CNJ DataJud public API, TJSP first-instance cases (2015--2019), three offense types. \textbf{Method:} Cross-offense comparison of between-vara conviction rate variance; vara leniency IV; common severity factor decomposition. \textbf{Sample:} 31 criminal varas in S\~ao Paulo Central handling drug trafficking, robbery, and theft. \textbf{Key finding:} Drug trafficking conviction rates are uncorrelated ($r = 0.10$) with theft rates but robbery-theft correlation is $r = 0.67$, indicating trafficking operates on a separate judicial dimension. SDE $= \hat{\beta} / \text{SD}(Y)$. Classification: Large: $|\text{SDE}| > 0.15$; Moderate: 0.05--0.15; Small: 0.005--0.05; Null: $< 0.005$.
\end{flushleft}
