\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Relabeling Illusion: Procurement Set-Asides and Small Firm Market Access in the Dominican Republic}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Governments worldwide reserve procurement contracts for small firms, but whether these mandates expand market access or merely relabel incumbents is unknown. I study the Dominican Republic's 20\% MIPYME set-aside mandate, exploiting a sudden enforcement surge after the August 2020 presidential transition that shifted agency compliance from 3\% to 17\%. Using 655,000 contract awards across 256 agencies, I find that agencies with larger compliance increases saw \textit{fewer} unique suppliers, not more. The share of MIPYME-certified winners rose by 20 percentage points, but decomposition reveals that 35\% of post-period awards went to relabeled incumbents---existing suppliers who obtained MIPYME certification---versus only 17\% to genuinely new entrants. The effect concentrates in large agencies. Set-aside mandates may inadvertently create a certification premium that favors well-connected incumbents over the new firms they are designed to help.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H57, D44, L26, O17 \\
\noindent\textbf{Keywords:} procurement, set-asides, small firms, MIPYME, Dominican Republic, relabeling

\newpage

\section{Introduction}

Over 100 countries mandate that a share of government procurement go to small and medium enterprises \citep{WorldBank2017procurement, oecd2019sme}. The logic is straightforward: small firms face barriers to government contracting---complex paperwork, bonding requirements, minimum scale---and reserving a portion of contracts should help them overcome these hurdles. The United States reserves 23\% of federal contracts for small businesses; India reserves 25\% for micro and small enterprises; the EU encourages member states to divide contracts into SME-accessible lots. But do these mandates actually expand market access for new small firms, or do they merely redirect contracts toward incumbent suppliers who learn to carry the right certification?

This question is difficult to answer because procurement set-asides are typically permanent features of the regulatory environment, making credible causal identification elusive. Existing work on set-asides in developed countries---particularly the U.S.\ highway auction literature \citep{krasnokutskaya2011bid, marion2007bid, nakabayashi2013small, athey2013set}---examines whether bid preferences raise costs or distort allocation, but takes the supplier pool as given. Whether the mandates \textit{create} new suppliers or merely \textit{redistribute} contracts among existing ones remains an open empirical question.

This paper exploits a sharp quasi-experiment in the Dominican Republic to identify the causal effect of procurement set-aside enforcement on supplier market structure. Decree 543-12, enacted in 2012, requires every government agency to reserve 20\% of procurement processes for certified micro, small, and medium enterprises (MIPYMEs). For eight years, the mandate was a dead letter: compliance hovered around 3\%. The August 2020 presidential transition from Danilo Medina (PLD) to Luis Abinader (PRM) brought an aggressive anti-corruption agenda and a new procurement director who pressured agencies to comply. By 2025, the national MIPYME-directed share reached 17\%, with enormous cross-agency variation: the Ministry of Sports surged from 0\% to 71\%, while hospitals remained near zero.

I construct an agency-quarter panel from three administrative datasets published by the Direcci\'{o}n General de Contrataciones P\'{u}blicas (DGCP): 655,000 contract awards, 576,000 procurement processes, and 132,000 registered suppliers with firm creation dates. Using a generalized difference-in-differences design with continuous treatment intensity---the agency-level change in MIPYME-directed share from pre-period (2016Q1--2020Q2) to post-period (2020Q3--2025Q4)---I estimate effects on supplier diversity, market concentration, and firm entry.

The findings reveal two distinct effects. First, agencies with larger compliance increases experienced a significant \textit{decline} in the number of unique suppliers---approximately 2.4 fewer suppliers per agency-quarter (off a baseline of 36), driven by procurement consolidation: high-compliance agencies reduced their total number of processes by 9\% per 10-percentage-point increase in MIPYME direction. When procurement volume is held constant, the supplier effect disappears, suggesting that the mandate induced agencies to reorganize purchasing rather than exclude firms directly. Second, the share of MIPYME-certified suppliers among winners rose by 20 percentage points ($p < 0.001$)---but this primarily reflects relabeling, not entry.

A decomposition of post-period awards reveals the relabeling mechanism. Among the 420,000 contracts awarded after 2020, 35\% went to ``relabeled'' suppliers: established firms (median age 9.7 years) that won contracts before the enforcement surge and subsequently obtained MIPYME certification. Only 17\% went to genuinely new MIPYME entrants---young firms (median age 2.5 years, 77\% under five years old) with no prior government contracting history. The remaining contracts split between continuing non-MIPYME suppliers (35\%) and new uncertified entrants (13\%). For every new MIPYME entrant, there are two relabeled incumbents.

The supplier decline concentrates in large agencies. Agencies in the top half of pre-period procurement volume show a 9.3\% decline in unique suppliers ($p = 0.011$), while small agencies show no significant effect. This pattern is consistent with a certification premium: in large agencies with many contracts, the returns to obtaining MIPYME certification are highest, attracting incumbent suppliers to relabel rather than encouraging new entry.

Several robustness checks support the findings. Controlling for agency-specific pre-period trends leaves the estimate essentially unchanged ($-0.58$, $p = 0.024$). A placebo test using only pre-treatment data with a fake treatment date at 2018Q3 yields a null result ($p = 0.53$), and a joint F-test of pre-trend coefficients fails to reject parallel trends ($p = 0.54$). Leave-one-out analysis dropping each of the top-10 highest-shift agencies produces coefficients ranging from $-0.52$ to $-0.67$.

This paper contributes to the procurement literature by identifying a specific mechanism---the relabeling illusion---through which well-intentioned set-aside mandates can fail. Prior work has studied the cost effects of bid preferences \citep{krasnokutskaya2011bid, marion2007bid}, the role of discretion in procurement corruption \citep{szucs2023procurement, bandiera2009active, decarolis2020rules}, and the effectiveness of monitoring and transparency in reducing waste \citep{olken2007monitoring, ferraz2011electoral, coviello2018publicity}. I show that even when a mandate successfully redirects contracts toward certified small firms, the primary beneficiaries may be strategically relabeled incumbents---a finding with direct implications for the design of procurement set-asides in developing countries.

The paper also contributes to the literature on state capacity and policy implementation in developing countries \citep{best2023individuals, banerjee2020can, colonnelli2020corruption}. The Dominican Republic case illustrates how a dormant mandate can be activated by political will, but the resulting compliance may satisfy the letter of the law without achieving its spirit. The distinction between nominal compliance (directing processes to MIPYMEs) and substantive compliance (actually expanding small firm participation) is a general challenge for mandate-based policies.

\section{Institutional Background}

\subsubsection*{The MIPYME Set-Aside Mandate} The Dominican Republic's Decree 543-12, signed in September 2012 by President Danilo Medina, requires all government contracting agencies to direct at least 20\% of their procurement processes to certified micro, small, and medium enterprises. Firms qualify as MIPYMEs based on employee count and annual revenue thresholds set by the Ministry of Industry and Commerce (MICM), and must obtain official certification. The decree also created a sub-quota for women-led MIPYMEs (``MIPYME Mujer'').

\subsubsection*{The Enforcement Gap} For its first eight years, Decree 543-12 was largely unenforced. The national MIPYME-directed share was 1.5\% in 2015, 3.7\% in 2019---far below the 20\% target. No penalties existed for non-compliance, and agencies faced no systematic monitoring. The mandate existed on paper but had minimal operational impact.

\subsubsection*{The Abinader Transition} On August 16, 2020, Luis Abinader of the Partido Revolucionario Moderno (PRM) was inaugurated after defeating the incumbent PLD party. The transition brought an aggressive anti-corruption platform, including a new Director General of the DGCP with a mandate to increase transparency and compliance. The DGCP implemented a real-time data-driven monitoring system, published compliance dashboards, and applied direct political pressure on agency heads. By 2025, the national MIPYME-directed share reached 17.4\%, and the women-MIPYME share climbed from near zero to 3.4\%.

\subsubsection*{Cross-Agency Variation} The compliance response was highly heterogeneous. Among the 256 agencies in my analytical sample, the change in MIPYME-directed share from pre- to post-period ranged from $-12$ to $+71$ percentage points, with a mean of 10.7 and standard deviation of 14.7. This variation is central to my identification strategy.

\section{Data}

I use three administrative datasets from the DGCP, all publicly available as downloadable CSV files from the Dominican Republic's open government data portal.

\subsubsection*{Procurement Processes} The \textit{Procesos} dataset contains 576,000 procurement processes published between 2015 and 2025. Each record identifies the procuring agency, procurement modality (below-threshold, minor purchase, competitive bid, etc.), estimated amount, and critically, a binary flag indicating whether the process is directed to MIPYME-certified firms. I use this flag to construct the treatment variable.

\subsubsection*{Contract Awards} The \textit{Adjudicaciones} dataset contains 655,000 contract awards over the same period. Each record links a contract to a specific supplier via a unique supplier registration number (RPE), along with the award date and contracted value. I use this dataset to construct outcome variables: unique suppliers per agency-quarter, the Herfindahl-Hirschman Index of contract concentration, and the share of first-time winners.

\subsubsection*{Supplier Registry} The \textit{Proveedores} dataset lists 132,000 registered government suppliers with their MIPYME certification status, enterprise classification (micro, small, medium, large), firm creation date, and geographic location. The firm creation date is essential for distinguishing genuinely new entrants from relabeled incumbents.

\subsubsection*{Panel Construction} I merge processes and awards at the agency-quarter level using agency identifiers, constructing a panel of 256 agencies observed over 40 quarters (2016Q1--2025Q4), yielding 8,415 agency-quarter observations. I restrict the sample to agencies with at least four quarters of data in both the pre-period (2016Q1--2020Q2) and post-period (2020Q3--2025Q4).

\input{tables/tab1_summary.tex}

\Cref{tab:summary} presents pre-period summary statistics by treatment intensity tercile. Agencies in the top tercile ($\Delta$MIPYME $> 10.3$pp) had slightly higher procurement volume and MIPYME shares even before the transition, but similar HHI and first-time winner shares, suggesting that the cross-agency variation is not strongly correlated with pre-existing market structure.

\section{Empirical Strategy}

I estimate the effect of MIPYME set-aside enforcement on supplier outcomes using a generalized difference-in-differences design with continuous treatment intensity:

\begin{equation}
Y_{a,t} = \alpha_a + \gamma_t + \beta \cdot \Delta\text{MIPYME}_a \times \text{Post}_t + \varepsilon_{a,t}
\label{eq:main}
\end{equation}

\noindent where $Y_{a,t}$ is the outcome for agency $a$ in quarter $t$; $\alpha_a$ and $\gamma_t$ are agency and quarter fixed effects; $\Delta\text{MIPYME}_a$ is the change in agency $a$'s MIPYME-directed share from the pre-period to the post-period; and $\text{Post}_t = \ind[t \geq \text{2020Q3}]$. Standard errors are clustered at the agency level (256 clusters).

The identifying assumption is that, absent the enforcement surge, agencies with different $\Delta\text{MIPYME}$ values would have followed parallel trends in supplier outcomes. The key threat is that agencies that responded most aggressively to the new administration's pressure differed systematically in ways that independently affected supplier dynamics. I address this with pre-trend tests, agency-specific linear trend controls, a pre-period placebo, and leave-one-out analysis.

The treatment variable $\Delta\text{MIPYME}_a$ is the realized change in MIPYME-directed share, which is an endogenous choice by agency heads. My interpretation relies on the argument that this choice was driven by top-down political pressure from the new administration---pressure that varied across agencies due to visibility, political alignment, and administrative capacity, but was plausibly orthogonal to agency-level supplier market conditions.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} presents the core findings. Column 1 shows that agencies with larger compliance increases experienced a significant decline in the number of unique suppliers winning contracts: the coefficient on $\Delta\text{MIPYME} \times \text{Post}$ is $-0.630$ ($p = 0.029$). Given a mean $\Delta$MIPYME of 0.107, this implies approximately 2.4 fewer unique suppliers per agency-quarter---a 6.7\% decrease off a pre-period baseline of 36. Column 2 shows a positive but insignificant increase in HHI (market concentration). Columns 3 and 4 reveal no significant effect on the share of first-time winners or new firms. Column 5 confirms the first-stage: the share of MIPYME-certified suppliers among winners increased by 20.2 percentage points ($p < 0.001$).

Crucially, the supplier decline operates through procurement consolidation. Agencies with larger MIPYME shifts also significantly reduced their total number of procurement processes ($\hat{\beta} = -0.91$, $p = 0.009$). When log number of processes is included as a control, the coefficient on $\Delta\text{MIPYME} \times \text{Post}$ falls to $0.07$ ($p = 0.61$)---the mandate did not reduce suppliers \textit{per process}, but it induced agencies to reorganize their purchasing into fewer, more concentrated procurement rounds. This consolidation, combined with the relabeling of incumbent suppliers, is the channel through which set-aside enforcement contracted the supplier base.

\subsection{Mechanism: Supplier Decomposition}

\input{tables/tab4_decomp.tex}

\Cref{tab:decomp} decomposes the 420,000 post-period awards by supplier type. The largest category is ``Relabeled''---suppliers that won contracts before the enforcement surge and are now MIPYME-certified---accounting for 35.3\% of awards and 21.8\% of contract value. ``Continuing non-MIPYME'' suppliers represent another 35.2\% of awards. ``New MIPYME'' entrants---the intended beneficiaries of the mandate---account for only 16.7\% of awards and 13.0\% of value. The relabeling channel dominates: for every new MIPYME entrant, there are two relabeled incumbents.

Firm age data from the supplier registry sharpen the distinction. Relabeled suppliers are established firms with a median age of 9.7 years; only 15\% are younger than five years. New MIPYME entrants, by contrast, have a median age of 2.5 years, with 77\% under five years old. The certification premium thus accrues to mature, well-connected firms---not to the young micro-enterprises the mandate was designed to help. Relabeled incumbents capture DOP 285 billion in contracts, while genuinely new MIPYME entrants receive only DOP 170 billion.

\subsection{Event Study and Pre-Trends}

\input{tables/tab5_event.tex}

\Cref{tab:event} reports the event study, interacting year indicators with $\Delta$MIPYME (reference year: 2019). Pre-treatment coefficients for 2016--2018 are positive but statistically insignificant, and a joint F-test fails to reject the null of zero pre-trends ($p = 0.54$). The sharp negative shift appears in 2020---the year of the presidential transition---and persists through 2025, with point estimates ranging from $-0.37$ to $-0.61$. The timing aligns precisely with the enforcement surge rather than any pre-existing trend.

\subsection{Robustness}

\input{tables/tab3_robust.tex}

\Cref{tab:robust} presents robustness checks. Column 2 adds agency-specific linear time trends estimated from the pre-period; the coefficient attenuates slightly to $-0.580$ ($p = 0.024$) but remains significant. Column 3 runs a placebo test using only pre-treatment data with a fake treatment date at 2018Q3, yielding a null result ($p = 0.53$). Columns 4--5 reveal that the effect is concentrated in large agencies (above-median pre-period procurement volume): $-0.932$ ($p = 0.011$) versus $-0.255$ ($p = 0.53$) for small agencies. Leave-one-out analysis (not shown) produces coefficients ranging from $-0.52$ to $-0.67$ when dropping each of the ten highest-shift agencies, confirming that no single outlier drives the result.

\section{Discussion}

The finding that procurement set-asides \textit{reduced} supplier diversity while \textit{increasing} MIPYME-certified supplier share illuminates a specific failure mode of mandate-based industrial policy: the certification premium. When governments create a category (``MIPYME-certified'') and direct contracts toward it, the primary behavioral response comes from firms with the lowest cost of obtaining certification---which are precisely the established suppliers with existing government relationships, legal capacity, and administrative know-how. Genuinely new small firms, who lack these resources, face the same barriers to participation whether or not a set-aside exists.

This mechanism parallels findings in the broader literature on targeting and program capture. \citet{bandiera2009active} distinguish ``active'' waste (corruption) from ``passive'' waste (bureaucratic inefficiency) in Italian procurement; the relabeling illusion represents a third category---\textit{compliant} waste, where agencies technically satisfy the mandate while subverting its purpose. \citet{colonnelli2020corruption} show that political connections distort firm participation in Brazilian procurement; my findings suggest that even well-intentioned mandates can create analogous distortions through certification requirements.

An important caveat concerns the treatment variable. $\Delta$MIPYME is the realized change in MIPYME-directed procurement, which reflects agency choices rather than random assignment. While the enforcement pressure originated from the new presidential administration---plausibly exogenous to individual agency supplier markets---agencies that responded most aggressively may differ in unobservable ways. The parallel-trends evidence supports the design, but a shift-share instrument using pre-transition procurement composition to predict compliance intensity would strengthen the causal interpretation. The composition channel also matters: the supplier decline operates through procurement consolidation (fewer processes), suggesting that set-aside compliance imposes administrative costs that lead agencies to reorganize their purchasing. Whether this reorganization ultimately harms or helps small firms depends on whether consolidated processes remain accessible.

The concentration of the effect in large agencies is consistent with the certification premium mechanism. In agencies with hundreds of contracts, the returns to MIPYME certification are highest, attracting relabeling by incumbents. Small agencies, with fewer contracts at stake, offer weaker incentives for strategic certification.

These results have direct policy implications. Set-aside mandates may need to be paired with policies that reduce certification barriers for genuinely new firms---simplified registration, capacity-building programs, or direct matching services---rather than relying solely on mandated quotas. The Dominican Republic's experience also suggests that monitoring should track not just MIPYME-directed shares (nominal compliance) but the share of first-time winners and genuinely young firms among beneficiaries (substantive compliance).

\section{Conclusion}

Procurement set-asides for small firms are among the most popular industrial policies globally. Using a natural experiment from the Dominican Republic---a dormant mandate suddenly activated by a presidential transition---I find that set-aside enforcement contracted the supplier base rather than expanding it. The mechanism is relabeling: established suppliers obtained MIPYME certification to capture directed contracts, crowding out genuinely new entrants. The policy achieved nominal compliance while failing its substantive goal.

The relabeling illusion is likely not unique to the Dominican Republic. Wherever governments create certification categories and direct resources toward them, incumbents with lower certification costs will capture a disproportionate share of the benefits. The challenge for policymakers is to design mandates that reward genuine entry rather than administrative compliance---a challenge that the current generation of set-aside programs has not yet met.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). Data from the Direcci\'{o}n General de Contrataciones P\'{u}blicas (DGCP) of the Dominican Republic, publicly available at datos.gob.do.

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Data Sources} All three datasets are downloaded from the DGCP open data portal:
\begin{itemize}
\item \textbf{Adjudicaciones:} 655,476 contract awards (2015--2025). Fields: contract code, award date, contracted value, supplier RPE, supplier name.
\item \textbf{Procesos:} 575,869 procurement processes (2015--2025). Fields: process code, agency code, agency name, procurement modality, estimated amount, MIPYME-directed flag, women-MIPYME flag.
\item \textbf{Proveedores:} 132,190 registered suppliers. Fields: RPE, MIPYME certification status, enterprise classification, firm creation date, province, municipality.
\end{itemize}

\subsubsection*{Sample Restrictions} I restrict to 2016Q1--2025Q4 (dropping 2015 observations due to partial coverage) and require agencies to have at least 4 quarters in both the pre-period (2016Q1--2020Q2) and post-period (2020Q3--2025Q4). This yields 256 of 669 agencies (38\%) representing the vast majority of procurement volume.

\subsubsection*{Treatment Construction} $\Delta\text{MIPYME}_a$ is the difference between agency $a$'s mean MIPYME-directed share in the post-period and pre-period. This continuous measure captures the full dose-response gradient of the enforcement surge.

\subsubsection*{Supplier Matching} I link awards to the supplier registry using the RPE (Registro de Proveedor del Estado) identifier. Awards without a matching RPE in the registry are included in outcome calculations but excluded from the decomposition analysis.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
