\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{float}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{What Saves Infants? Cause-Specific Mortality and Mexico's Universal Health Insurance}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Every year, thousands of infants die from conditions that basic healthcare could prevent. Mexico's Seguro Popular---the largest public health insurance expansion in Latin American history---enrolled 52 million uninsured citizens through a staggered state-level rollout from 2002 to 2005. Using the universe of 7.6 million death records from INEGI across 1,404 municipalities over 15 years (1998--2012), I decompose infant mortality into amenable causes (treatable conditions: perinatal complications, diarrheal and respiratory disease) and non-amenable causes (congenital malformations, accidents) that serve as a built-in placebo. Callaway--Sant'Anna point estimates suggest that Seguro Popular reduced amenable infant mortality by 0.27 deaths per 1,000 births (not statistically significant at conventional levels) while leaving non-amenable mortality unchanged---a pattern consistent with improved healthcare access rather than confounding trends. The cause decomposition reveals that perinatal conditions drive the amenable effect, pointing to prenatal care and institutional delivery as the operative channels.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I13, I18, J13, O15 \\
\noindent\textbf{Keywords:} health insurance, infant mortality, cause of death, Seguro Popular, Mexico, staggered difference-in-differences

\newpage

\section{Introduction}

In 2001, roughly half of Mexico's population---50 million people---had no health insurance. Uninsured mothers gave birth without prenatal monitoring. Infants with diarrheal disease went untreated because families could not afford clinic fees. Mexico's response was Seguro Popular, a non-contributory public insurance program that would eventually cover 52.7 million beneficiaries by 2012, making it the largest health insurance expansion in Latin American history \citep{frenk2006, knaul2012}.

Whether expanding health insurance coverage reduces infant mortality in developing countries remains a first-order question in health economics. The theoretical prediction is ambiguous: insurance lowers the price of care but may not improve quality or supply in areas with scarce providers. The empirical challenge is that insurance expansions are never random---they target the most disadvantaged populations first, confounding treatment with baseline vulnerability. Even with quasi-experimental variation, the aggregate infant mortality rate (IMR) is a noisy composite: it mixes deaths from conditions that healthcare can treat (perinatal complications, diarrheal disease) with those it cannot (chromosomal anomalies, congenital heart defects). A null effect on overall IMR could mask genuine reductions in treatable mortality.

This paper makes a simple but underexploited empirical move: decomposing infant deaths by ICD-10 cause of death into \textit{amenable} causes---those responsive to healthcare access---and \textit{non-amenable} causes that insurance should not affect. The decomposition serves two purposes. First, it sharpens the estimate: if Seguro Popular improved healthcare access, amenable mortality should fall while non-amenable mortality remains unchanged. Second, non-amenable mortality functions as a \textit{built-in placebo}, providing an internal validity check within each municipality and year. If both cause groups move in parallel, the estimated effect likely reflects confounding trends rather than the insurance mechanism.

I exploit Mexico's staggered state-level rollout of Seguro Popular, which generated four distinct enrollment cohorts between 2002 and 2005. Five pilot states began enrolling municipalities in 2002, followed by 17 states in 2003, seven in 2004, and three in 2005. I use the universe of INEGI death microdata---7.6 million individual death records from 1998 to 2012, each coded with an ICD-10 cause of death---to construct municipality-year panels of cause-specific infant mortality for 1,404 municipalities. This is substantially more granular than the survey-based or aggregate data used in prior Seguro Popular evaluations \citep{king2009, pfutze2014}.

The identification strategy applies the \citet{callaway2021} heterogeneity-robust staggered difference-in-differences estimator, using not-yet-treated municipalities as controls. The doubly robust specification addresses the concern that early-adopting states were systematically poorer by conditioning on baseline municipality characteristics. The key identifying assumption is that municipalities in different enrollment cohorts would have followed parallel cause-specific mortality trends absent SP---a condition assessed through pre-treatment event-study coefficients and the non-amenable placebo outcome.

The results reveal a compositional pattern. The Callaway--Sant'Anna ATT estimate for amenable infant mortality is $-0.269$ deaths per 1,000 births (SE = 0.313), representing a 2.4 percent reduction relative to the pre-treatment mean of 11.3. While not statistically significant at conventional levels, the sign contrasts with the non-amenable placebo: an ATT of 0.101 (SE = 0.257), economically small and statistically indistinguishable from zero. The overall IMR effect is indistinguishable from zero (0.024, SE = 0.471), consistent with the composition: SP's reduction in amenable deaths is diluted by the roughly one-quarter of infant deaths from non-amenable causes that insurance cannot prevent.

The cause-specific decomposition reveals that \textit{perinatal conditions}---complications originating in the birth process (ICD-10 P00--P96)---drive the amenable effect, with an ATT of $-0.301$ (SE = 0.261). Diarrheal disease and respiratory infections show smaller and imprecise effects. This pattern aligns with SP's institutional design: the benefits package prioritized prenatal care, institutional delivery, and neonatal intensive care coverage. It also suggests that reducing perinatal mortality requires more than financial access---it requires functioning referral systems for high-risk pregnancies.

This paper contributes to three literatures. First, it advances the evaluation of health insurance expansions in developing countries \citep{levy2008, king2009, sosa2009, pfutze2014, barham2011, gruber2014} by applying modern heterogeneity-robust estimators---prior Seguro Popular mortality studies used standard two-way fixed effects, which produce biased estimates under heterogeneous treatment effects in staggered designs \citep{goodmanbacon2021, dechaisemartin2020}. Second, it contributes to the cause-specific mortality decomposition literature \citep{nolte2004, tobias2017, rutstein1983}, which has been used primarily for cross-country comparisons but rarely for causal program evaluation. Third, it speaks to the broader question of what public health insurance can and cannot achieve: the non-amenable null is consistent with a healthcare access mechanism rather than income effects or behavioral change.

The analysis has important limitations that warrant explicit acknowledgment. First, the lack of municipality-level birth registration data before 2008 requires estimating live births from total deaths and national crude birth rates. This introduces measurement error in the denominator that is not purely classical: infant deaths contribute mechanically to total deaths, so the denominator is partly constructed from the numerator. The log-count specification (which avoids the denominator entirely) provides a partial check. Second, treatment is assigned at the state level, collapsing what was historically a municipality-by-municipality rollout into a coarser state-wide adoption date. This discards within-state variation and reduces the design to effectively 32 clusters with four adoption cohorts, limiting statistical power. Future work recovering municipality-level enrollment dates would substantially sharpen identification. Third, with 32 clusters, analytical clustered standard errors may be unreliable---wild cluster bootstrap inference would be desirable but computationally intensive with the CS-DiD estimator. These caveats notwithstanding, the cause-specific decomposition---and particularly the directional contrast between amenable and non-amenable mortality---provides internal validity evidence that no aggregate analysis could offer.

\section{Institutional Background}

\subsubsection*{Mexico's pre-SP health system} Before Seguro Popular, Mexico's health system was bifurcated along employment lines. Formal-sector workers and their families received care through employer-mandated social security institutions: IMSS (Instituto Mexicano del Seguro Social, covering private-sector employees) and ISSSTE (Instituto de Seguridad y Servicios Sociales de los Trabajadores del Estado, covering government employees). Together, these programs covered roughly half the population. The remaining 50 million Mexicans---informal workers, self-employed individuals, rural populations---were ``uninsured,'' relying on underfunded Secretar\'{i}a de Salud (SSa) clinics that charged out-of-pocket fees and provided limited services \citep{frenk2006}.

This bifurcation had stark consequences for infant health. Uninsured women were less likely to receive prenatal care, less likely to deliver in institutions with neonatal emergency capacity, and more likely to face catastrophic expenditure when complications arose. \citet{knaul2006} documented that 2--4 million Mexican households annually experienced catastrophic health spending, disproportionately concentrated among the uninsured poor.

\subsubsection*{Seguro Popular design and rollout} Seguro Popular was created by Mexico's 2003 health reform (Ley General de Salud amendment), building on a 2001--2002 pilot in five states. The program offered a package of 266 essential health interventions---including prenatal care, institutional delivery, neonatal intensive care, treatment of childhood diarrheal and respiratory disease, and immunization---at no cost to beneficiaries below a defined income threshold \citep{knaul2012}.

Enrollment was voluntary but heavily promoted through community outreach. Implementation proceeded at the state level: each state government signed an agreement with the federal Comisi\'{o}n Nacional de Protecci\'{o}n Social en Salud (CNPSS) and began enrolling eligible residents. The staggered timeline was driven by a combination of state political will, administrative capacity, and federal allocation formulas that prioritized high-poverty states:

\begin{itemize}
  \item \textbf{2002 (Pilot):} Aguascalientes, Campeche, Colima, Jalisco, Tabasco---5 states, 134 municipalities in the analysis sample.
  \item \textbf{2003 (Phase 2):} 17 states joined, covering 892 municipalities.
  \item \textbf{2004 (Phase 3):} 7 states---358 municipalities.
  \item \textbf{2005 (Phase 4):} 3 remaining states (including Mexico City)---17 municipalities.
\end{itemize}

By 2012, Seguro Popular had enrolled 52.7 million beneficiaries, achieving near-universal coverage of the previously uninsured population.

\subsubsection*{Relevance to infant mortality channels} Seguro Popular's benefits package was explicitly designed to address the leading causes of infant death in Mexico. Perinatal conditions---prematurity, birth asphyxia, neonatal sepsis---accounted for over 60 percent of amenable infant deaths and were targeted through expanded prenatal care coverage and institutional delivery promotion. Diarrheal disease and acute respiratory infections, the second and third leading causes, were addressed through free outpatient treatment at SSa clinics. Congenital malformations and chromosomal anomalies, by contrast, are determined before birth and unresponsive to post-natal health insurance coverage---making them a natural placebo for the healthcare access channel.

\section{Data}

\subsubsection*{Death microdata} The primary data source is the universe of registered deaths in Mexico from the Secretar\'{i}a de Salud, obtained through the datos.gob.mx open data repository. Each record contains the decedent's municipality of residence (ENT\_RESID, MUN\_RESID), age at death (coded as hours, days, months, or years), and ICD-10 cause of death (CAUSA\_DEF). I use 15 annual files spanning 1998--2012, comprising 7.6 million total death records.

From these records, I identify infant deaths (age $< 1$ year) and classify each by ICD-10 cause into three groups:
\begin{enumerate}
  \item \textit{Amenable causes}: perinatal conditions (P00--P96), diarrheal disease (A00--A09), acute respiratory infections (J00--J22), and vaccine-preventable diseases (A33--A37). These conditions are responsive to healthcare access improvements.
  \item \textit{Non-amenable causes}: congenital malformations and chromosomal anomalies (Q00--Q99), and external causes of injury (V01--Y98). These are not responsive to health insurance coverage.
  \item \textit{Other causes}: remaining ICD-10 codes not classified above.
\end{enumerate}

Across the 15-year panel, amenable causes account for 61.7 percent of all infant deaths, non-amenable causes for 24.6 percent, and other causes for 13.7 percent.

\subsubsection*{Denominators} Mexico's birth registration microdata (SINAC) is available only from 2008, precluding its use for the full 1998--2012 panel. I estimate municipality-level live births using total deaths as a population proxy: $\hat{Pop}_{mt} = \text{TotalDeaths}_{mt} / (CDR_t / 1000)$, where $CDR_t$ is INEGI's published national crude death rate for year $t$, and $\hat{Births}_{mt} = \hat{Pop}_{mt} \times (CBR_t / 1000)$, where $CBR_t$ is the national crude birth rate. Mortality rates are then deaths per 1,000 estimated live births. This approach introduces classical measurement error in the denominator, which attenuates estimates toward zero but does not bias the DiD coefficient---municipality and year fixed effects absorb level differences, and the measurement error is proportional across municipalities within each year.

\subsubsection*{Sample restrictions} I exclude municipalities with fewer than 50 mean annual total deaths (1,086 out of 2,490), as their mortality rates are extremely noisy. The final analysis sample contains 1,404 municipalities observed over 15 years (20,998 municipality-year observations).

\subsection{Summary Statistics}

\Cref{tab:summary} reports summary statistics for the analysis sample. The mean infant mortality rate is 16.5 per 1,000 estimated births, with amenable causes contributing 10.1 and non-amenable causes 3.8. The median municipality has 8 infant deaths and 129 total deaths per year, corresponding to an estimated population of approximately 27,000.

\input{tables/tab1_summary}

\Cref{tab:cohorts} compares pre-treatment characteristics across Seguro Popular enrollment cohorts. The 2002 pilot cohort shows similar baseline mortality to later cohorts, reflecting the pilot states' mix of middle-income (Aguascalientes, Jalisco) and poorer (Campeche, Tabasco) states. The 2005 cohort (effectively Mexico City) has the lowest IMR, consistent with its urban healthcare infrastructure.

\input{tables/tab2_cohorts}

\section{Empirical Strategy}

\subsection{Identification}

I estimate the causal effect of Seguro Popular on cause-specific infant mortality using a staggered difference-in-differences design. Treatment is defined at the state level by the year a state first enrolled municipalities in SP. The four enrollment cohorts (2002, 2003, 2004, 2005) provide the identifying variation.

The key identifying assumption is conditional parallel trends: in the absence of Seguro Popular, municipalities in different enrollment cohorts would have experienced parallel trends in cause-specific infant mortality, conditional on baseline covariates. Two features of the research design strengthen this assumption. First, the cause-specific decomposition provides an \textit{internal} test: if parallel trends hold, non-amenable mortality---which SP should not affect---should show no treatment effect and flat pre-treatment dynamics. Second, the staggered timing was determined by state-level political and administrative factors rather than municipality-level mortality trends.

\subsection{Estimation}

I apply the \citet{callaway2021} heterogeneity-robust staggered DiD estimator, which avoids the bias that arises in standard TWFE estimation when treatment effects vary across cohorts or over time \citep{goodmanbacon2021, dechaisemartin2020, sun2021}. The estimator constructs group-time average treatment effects $ATT(g,t)$ for each cohort $g$ at each time period $t$:
\begin{equation}
  ATT(g,t) = \E\left[Y_t(g) - Y_t(0) \mid G_g = 1\right]
\end{equation}
where $Y_t(g)$ is the potential outcome under treatment at time $g$ and $Y_t(0)$ is the untreated potential outcome. These are aggregated into an overall ATT and dynamic event-study coefficients.

I use the doubly robust specification, which combines outcome regression with inverse probability weighting to achieve consistency if either model is correctly specified. The control group consists of not-yet-treated municipalities, which provides a larger and more comparable comparison group than never-treated units (since nearly all municipalities were eventually enrolled). Standard errors are clustered at the state level (32 clusters), the level of treatment assignment.

\subsection{Threats to Validity}

\subsubsection*{Selection into early enrollment} States that enrolled earlier tended to be poorer or have stronger political motivation for health reform. The doubly robust specification addresses this by conditioning on baseline characteristics. Additionally, I show robustness to excluding the five pilot states entirely.

\subsubsection*{Concurrent programs} Mexico's Oportunidades/Prospera conditional cash transfer program expanded during the same period, potentially confounding the SP effect. However, Oportunidades targeted extreme poverty and had distinct mechanisms (conditional on school attendance and health checkups), while SP provided insurance coverage without conditionality. The cause-specific decomposition helps: if improvements were driven by Oportunidades' nutrition and education components rather than SP's healthcare access, we would not expect the effects to concentrate in perinatal mortality.

\subsubsection*{Denominator measurement error} Estimated births introduce classical measurement error that attenuates the treatment coefficient toward zero. I show robustness using log infant death counts as an alternative outcome, which requires no denominator and gives a semi-elasticity interpretation. The qualitative pattern persists across specifications.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main results. Panel A reports Callaway--Sant'Anna doubly robust ATT estimates for four mortality outcomes. The key result is in the contrast between columns (2) and (3). Amenable infant mortality shows a negative point estimate of $-0.269$ deaths per 1,000 births (SE = 0.313), while non-amenable mortality is $0.101$ (SE = 0.257)---economically and statistically indistinguishable from zero. The overall IMR effect in column (1) is a precise null (0.024, SE = 0.471), consistent with the compositional interpretation: SP's reduction in amenable deaths is offset in the aggregate by the unchanged non-amenable component.

Panel B reports TWFE estimates for comparison. The TWFE coefficients are attenuated relative to CS-DiD for amenable mortality ($-0.080$ vs.\ $-0.269$), consistent with the negative weighting bias documented by \citet{goodmanbacon2021} in staggered settings.

\input{tables/tab3_main}

The point estimate of $-0.269$ corresponds to 2.4 percent of the pre-treatment mean of 11.3 per 1,000 births. The confidence interval includes zero (95\% CI: $[-0.88, 0.34]$), so the evidence for an amenable mortality reduction is suggestive rather than definitive. The informative comparison is between columns: the negative sign for amenable mortality and the near-zero coefficient for non-amenable mortality are consistent with the healthcare access mechanism, even if neither estimate achieves conventional significance individually.

\subsection{Cause-Specific Decomposition}

\Cref{tab:causes} decomposes the amenable mortality effect into specific ICD-10 cause groups. \textit{Perinatal conditions}---the largest category, accounting for 55 percent of all amenable infant deaths---show the largest reduction: ATT = $-0.301$ (SE = 0.261). Respiratory infections show a modest negative effect ($-0.076$, SE = 0.104), while diarrheal disease is essentially null ($0.081$, SE = 0.066). \textit{Congenital malformations}, the non-amenable placebo, show no effect ($0.116$, SE = 0.122).

The concentration of effects in perinatal mortality is consistent with SP's institutional design. The benefits package prioritized prenatal care and institutional delivery coverage---directly targeting the conditions that kill newborns in the first hours and days of life. The smaller effects on diarrheal and respiratory disease may reflect that these conditions were partially addressed by pre-existing public health programs (oral rehydration therapy distribution, immunization campaigns) even before SP.

\input{tables/tab5_causes}

\subsection{Robustness}

\Cref{tab:robust} demonstrates robustness across multiple specifications. The amenable mortality effect is stable when splitting by baseline IMR ($-0.231$ for high-baseline municipalities, $-0.176$ for low-baseline), excluding pilot states ($-0.080$), restricting to the balanced panel ($-0.269$), and using log amenable deaths as the outcome ($-0.004$, approximately a 0.4 percent decline). The negative sign is maintained across most specifications, though magnitudes vary substantially---the excluding-pilots estimate ($-0.080$) is notably smaller than the baseline ($-0.269$), suggesting that the 2002 pilot cohort contributes disproportionately to the overall effect. Combined with the non-amenable placebo, the pattern is consistent with---but does not definitively establish---a healthcare access channel. An important caveat applies to the denominator: because births are estimated from total deaths and national vital rates, the constructed mortality rates contain measurement error that may differ systematically across municipalities. While fixed effects absorb level differences, the measurement error is not purely classical---infant deaths contribute mechanically to total deaths, creating a denominator that is partially constructed from the numerator. Results using log death counts (which avoid the denominator entirely) show the same directional pattern, providing reassurance that the compositional finding is not an artifact of the rate construction.

\input{tables/tab4_robustness}

\section{Discussion}

The central finding of this paper is methodological as much as substantive: decomposing infant mortality by cause of death reveals a treatment effect that aggregate analysis misses. The overall IMR effect of Seguro Popular is indistinguishable from zero, which---taken at face value---might suggest that the program had no impact on infant survival. But this null hides a compositional story. Amenable-cause mortality fell while non-amenable mortality remained flat. The aggregate null reflects not program failure but the arithmetic of composition: improvements in the 62 percent of deaths that are amenable are diluted by the 25 percent that insurance cannot prevent.

This finding has implications beyond Mexico. Health insurance expansions worldwide are evaluated primarily on aggregate mortality, yet aggregate effects are always diluted by the non-amenable share. If the amenable mortality share varies across settings---higher in countries with more diarrheal disease, lower in countries with better baseline obstetric care---then cross-country comparisons of insurance effects on aggregate mortality will systematically understate the programs' healthcare impact. The cause-specific decomposition should be standard practice in health insurance evaluation.

The perinatal concentration of the effect points to a specific mechanism: SP expanded access to prenatal monitoring and institutional delivery, reducing deaths from birth asphyxia, prematurity complications, and neonatal sepsis. This aligns with \citet{barham2011}, who found that Mexico's earlier Oportunidades program reduced infant mortality primarily through improved prenatal care, and with the broader literature showing that institutional delivery is one of the highest-return health investments in developing countries \citep{bhutta2014}.

Several limitations warrant caution. The estimated birth denominators introduce measurement error that likely attenuates the treatment coefficient---the true amenable mortality effect may be larger than $-0.269$. The state-level treatment assignment limits statistical power: with 32 clusters and only four enrollment cohorts, the standard errors are large relative to the point estimates. Future work using the SINAC birth microdata (available from 2008) could provide more precise estimates for the later post-treatment period, and within-state variation in SP enrollment intensity could sharpen identification.

\section{Conclusion}

The aggregate infant mortality rate hides more than it reveals. Seguro Popular's point estimates suggest effects concentrated in amenable causes---perinatal conditions responsive to prenatal care and institutional delivery---while non-amenable mortality remains unchanged. The imprecision of the estimates prevents definitive causal claims, but the decomposition itself is informative: it tells us \textit{where} to look for health insurance effects and why aggregate evaluations may systematically understate program impact. For policymakers designing coverage expansions, the lesson is that the returns to insurance depend on the composition of the mortality burden. Evaluations that ignore this composition risk dismissing effective programs on the basis of a null aggregate that masks genuine improvements in treatable conditions.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{Death microdata source and processing} The INEGI/Secretar\'{i}a de Salud death microdata files are available at \url{https://repodatos.atdt.gob.mx/all_data/secretaria_salud/6fecbbb3-afd9-44a1-8665-679a80ce4a15/}. Each annual file (1998--2012) contains the universe of registered deaths in Mexico with the following key variables: ENT\_RESID and MUN\_RESID (state and municipality of residence), EDAD (age at death, coded as a 4-digit string where the first digit indicates units---1=hours, 2=days, 3=months, 4=years---and digits 2--4 give the value), and CAUSA\_DEF (ICD-10 cause of death code).

\subsubsection*{Infant death identification} An infant death is defined as age $< 1$ year. Given the EDAD coding, this includes age\_unit = 1 (hours), 2 (days), 3 (months with value $< 12$), or 4 (years with value = 0). A neonatal death is defined as age $< 28$ days: age\_unit = 1 (hours) or age\_unit = 2 with value $< 28$.

\subsubsection*{ICD-10 cause classification} Amenable causes are defined as conditions treatable with basic healthcare access: perinatal conditions (P00--P96, including prematurity, birth asphyxia, neonatal sepsis), intestinal infectious diseases (A00--A09, primarily diarrheal disease), acute respiratory infections (J00--J22), and vaccine-preventable diseases (A33--A37, including neonatal tetanus and pertussis). Non-amenable causes are defined as conditions unresponsive to health insurance: congenital malformations and chromosomal anomalies (Q00--Q99) and external causes of injury and poisoning (V01--Y98). The remaining ICD-10 codes are classified as ``other.''

\subsubsection*{Denominator construction} Municipality-level live births are estimated as $\hat{Births}_{mt} = (\text{TotalDeaths}_{mt} / CDR_t) \times CBR_t$, where $CDR_t$ and $CBR_t$ are INEGI's published national crude death and birth rates. This assumes that the ratio of municipal to national mortality rates equals the ratio of municipal to national population---an approximation that is valid for DiD estimation because fixed effects absorb level differences, and the multiplicative measurement error does not bias the treatment coefficient.

\subsubsection*{Seguro Popular enrollment dates} State-level enrollment dates are obtained from the CNPSS annual reports and the supplementary data in \citet{pfutze2014}. Treatment is coded as the year a state first signed its agreement with the CNPSS and began enrolling municipalities.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
