\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{enumitem}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Accommodation Gap: Permitless Carry Laws and Racial Disparities in Customer-Facing Employment}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 2003 and 2024, thirty U.S.\ states removed the permit requirement for concealed firearm carrying. While the crime effects of these ``permitless carry'' laws are extensively studied, their labor market consequences are unknown. Using Quarterly Workforce Indicators data on state-level employment by race and industry, I estimate a triple-difference comparing Black versus White workers in accommodation versus manufacturing before and after adoption. Black accommodation employment grew 7.2 log points less than predicted by the triple-difference baseline ($p = 0.011$), with no comparable shortfall for White accommodation or Black manufacturing. This ``accommodation gap'' survives excluding the 2021 adoption wave and appears in a pre-2020 sample. The pattern reflects relative sectoral reallocation---Black employment growth in adopting states bypassed accommodation---rather than absolute job loss, suggesting that gun-carrying liberalization reshapes the sectoral composition of minority employment.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J71, K14, J21 \\
\noindent\textbf{Keywords:} permitless carry, gun policy, racial employment gap, accommodation sector, triple-difference

\newpage

\section{Introduction}

In April 2021, Texas became the latest of an accelerating wave of states to pass ``constitutional carry'' legislation, allowing residents to carry concealed firearms in public without a government-issued permit. Within twelve months, five more states followed. By 2024, thirty states had adopted some form of permitless carry---covering more than half the U.S.\ population. The academic debate has focused almost exclusively on whether these laws affect crime \citep{donohue2019right, lott2020more}. Their consequences for the labor market remain entirely unstudied.

This paper asks whether permitless carry laws differentially reduce Black employment in customer-facing industries. The question rests on a simple observation: when legal barriers to public gun carrying are removed, the workers most exposed to the consequences are those who interact face-to-face with armed members of the public in the course of their jobs---hotel clerks, restaurant servers, bartenders. If racialized perceptions of threat alter the dynamics of these interactions \citep{eberhardt2004seeing, correll2007across}, or if Black workers differentially exit environments where armed encounters become more likely, then the employment effects of gun-carrying liberalization should vary by both race and industry.

I test this hypothesis using a triple-difference (DDD) design that exploits the staggered adoption of permitless carry across thirty states between 2003 and 2024. The key outcome is log quarterly employment from the Census Bureau's Quarterly Workforce Indicators (QWI), observed at the state-by-race-by-industry level. The three differences are: (1) before versus after adoption within each state, (2) Black versus White workers within the same state and time period, and (3) accommodation and food services (NAICS 72, a high-customer-contact sector) versus manufacturing (NAICS 31--33, a low-customer-contact sector). The triple interaction isolates the Black-accommodation-specific effect of permitless carry, netting out any general employment trends that affect Black workers, accommodation workers, or treated states.

The main finding is that permitless carry is associated with 7.2 log points lower Black accommodation employment relative to the triple-difference baseline (SE = 0.029, $p = 0.011$). Crucially, this reflects a relative sectoral reallocation, not absolute job loss: the direct Callaway-Sant'Anna ATT for Black accommodation is slightly positive (+0.050, SE = 0.039). The accommodation gap emerges because Black employment growth in adopting states concentrated in manufacturing rather than accommodation. This effect is driven entirely by the race-by-sector channel: White accommodation employment is essentially unaffected by adoption (Callaway-Sant'Anna ATT = $-0.003$, SE = 0.016), and Black manufacturing employment actually increases in adopting states (ATT = $+0.132$, SE = 0.052). The DDD design reveals that the broad Black employment gains in adopting states systematically bypass the accommodation sector.

Several robustness checks support this finding. Dropping the 2021 cohort---five states whose adoption coincided with post-COVID labor market turmoil---strengthens the estimate to $-0.077$ (SE = 0.030). The pre-2020 sample, which includes only the earlier adopters, yields a DDD coefficient of $-0.059$ (SE = 0.033, $p = 0.086$)---same sign, noisier with fewer cohorts. The standard two-way fixed effects (TWFE) estimator for Black accommodation alone gives a misleading positive coefficient, illustrating the classic heterogeneous-treatment-effect bias that the DDD and Callaway-Sant'Anna estimator correct for \citep{callaway2021difference, goodman2021difference}.

This paper makes three contributions. First, it introduces the labor market to the permitless carry debate, which has been confined to crime outcomes despite the large number of workers in customer-facing occupations \citep{rand2020effects}. Second, it provides the first evidence that gun policy has racially heterogeneous employment effects, linking the literatures on firearms regulation \citep{webster2020reducing} and racial labor market inequality \citep{lang2020race}. Third, the finding that gun-carrying liberalization has sector-specific costs for minority workers illustrates a broader class of policy externalities---regulatory changes designed for one domain (public safety) that spill over into labor markets through racialized channels. I call this the \emph{accommodation gap}: the divergence between Black employment gains in sectors insulated from public gun carrying and the dampened gains in sectors where armed interactions are routine.

The rest of the paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} outlines the empirical strategy. Section~\ref{sec:results} reports results. Section~\ref{sec:discussion} discusses implications. Section~\ref{sec:conclusion} concludes.

\section{Institutional Background}
\label{sec:background}

Prior to the permitless carry movement, all U.S.\ states required applicants for concealed carry permits to satisfy various eligibility criteria---typically including background checks, safety training, and residency requirements \citep{siegel2017easiness}. The permit system served as both a screening mechanism (filtering out prohibited persons and those unwilling to invest in training) and a deterrent to casual carrying (imposing time and monetary costs on prospective carriers).

Permitless carry---also called ``constitutional carry''---removes the permit requirement entirely, allowing any person not otherwise prohibited from firearm possession to carry a concealed handgun in public without government authorization. Alaska became the first state to adopt this regime in 2003. Adoption accelerated after 2010, with major waves in 2015--2017 (Kansas, Maine, Mississippi, West Virginia, Idaho, Missouri, New Hampshire, North Dakota) and 2019--2023 (Kentucky, South Dakota, Oklahoma, Montana, Iowa, Utah, Tennessee, Texas, Louisiana, Indiana, Ohio, Georgia, Alabama, Florida, Nebraska, North Carolina). By 2024, thirty states had adopted permitless carry.

The laws vary in scope: some apply only to state residents, some only to handguns, and some restrict carrying in specific locations such as courthouses and schools. However, all eliminate the core requirement of a government-issued permit for general concealed carrying in public spaces, including restaurants, hotels, and retail establishments---the workplaces of the accommodation and food services sector.

The accommodation sector (NAICS 72) is distinctive in its degree of customer-employee contact. Restaurant servers, hotel front-desk clerks, and bartenders interact with members of the public continuously throughout their shifts, often in enclosed spaces. Manufacturing workers, by contrast, operate in controlled environments with minimal public contact. This sectoral contrast provides the identifying variation for the triple-difference design: if permitless carry affects employment through the channel of armed customer-worker interactions, the effect should concentrate in accommodation.

\section{Data}
\label{sec:data}

The primary data source is the Quarterly Workforce Indicators (QWI), produced by the Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) program \citep{abowd2009lehd}. The QWI reports employment, hiring, separations, and earnings by state, industry, and demographic characteristics (including race) at quarterly frequency. I use the race/ethnicity tabulation at the NAICS sector level, which distinguishes White-alone and Black-alone workers.

The analysis sample covers 41 states (30 adopters plus 11 never-adopters: California, Connecticut, Delaware, District of Columbia, Hawaii, Illinois, Maryland, Massachusetts, New Jersey, New York, and Rhode Island) over the period 2005--2024. The unit of observation is the state-year-race-sector cell, where I collapse quarterly observations to annual averages to reduce noise. Two sectors are included: accommodation and food services (NAICS 72) and manufacturing (NAICS 31--33). The resulting panel contains 3,228 state-year-race-sector observations.

\input{tables/tab1_summary}

Table~\ref{tab:summary} reports summary statistics. Average state-level Black employment in accommodation is approximately 302,000, roughly one-fifth of White accommodation employment (1.39 million). Black manufacturing employment averages 234,000 per state-year. The substantial variation in employment levels across states and over time provides the identifying variation.

\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Triple-Difference Design} The primary specification is a triple-difference that identifies the effect of permitless carry on Black accommodation employment relative to three baselines: pre/post within state, Black versus White within the same state and year, and accommodation versus manufacturing within the same state, year, and race group. The estimating equation is:
\begin{equation}
\log(\text{Emp})_{s,t,r,j} = \alpha_{s,r,j} + \delta_{t,r,j} + \beta_1(\text{Post}_{s,t} \times \text{Black}_r \times \text{Accom}_j) + \mathbf{X}'\gamma + \varepsilon_{s,t,r,j}
\label{eq:ddd}
\end{equation}
where $s$ indexes states, $t$ years, $r$ race (Black or White), and $j$ sector (accommodation or manufacturing). $\alpha_{s,r,j}$ are state-by-race-by-sector fixed effects and $\delta_{t,r,j}$ are year-by-race-by-sector fixed effects. $\text{Post}_{s,t}$ equals one in years at or after the state's adoption of permitless carry (zero for never-adopters). The coefficient of interest is $\beta_1$, which captures the differential change in Black accommodation employment attributable to the policy, net of all additive race, sector, state, and time trends.

Standard errors are clustered at the state level (41 clusters), the level at which the policy varies. The specification also includes the two-way interactions $\text{Post} \times \text{Black}$ and $\text{Post} \times \text{Accom}$, which absorb any race-specific or sector-specific effects of adoption that are common across sectors or races.

\subsubsection*{Callaway-Sant'Anna DiD} To complement the DDD, I estimate standard two-group DiD models for each race-sector cell separately using the Callaway and Sant'Anna (2021) estimator \citep{callaway2021difference}. This estimator is robust to treatment-effect heterogeneity across adoption cohorts and time periods, a concern with thirty states adopting over two decades. I use the eleven never-adopted states as the control group and impose no anticipation. The estimator produces group-time average treatment effects $ATT(g,t)$ that are aggregated to an overall ATT.

\subsubsection*{Threats to Validity} The identifying assumption for the DDD is that, absent permitless carry, the Black-White employment gap in accommodation relative to manufacturing would have evolved similarly in adopting and non-adopting states. This is weaker than the standard DiD parallel trends assumption because it only requires parallel trends in the \emph{triple difference}---cross-race, cross-sector trends can be arbitrary. The main concern is that states adopting permitless carry are systematically different (more rural, more Republican, lower minority population). The DDD addresses this by comparing within the same state, differencing out any level differences across states.

COVID-19 is a potential confound for the 2021 adoption wave (five states). I address this by estimating the DDD on the pre-2020 sample and by dropping the 2021 cohort, showing that the result holds in both cases.

\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_main}

Table~\ref{tab:main} reports Callaway-Sant'Anna ATT estimates for each race-sector cell. The overall ATT for Black accommodation employment is $+0.045$ (SE = 0.042), positive but statistically insignificant. White accommodation employment shows no effect ($-0.003$, SE = 0.016). Black manufacturing employment increases substantially ($+0.132$, SE = 0.052). These results indicate that Black workers in adopting states experienced employment gains concentrated in manufacturing, not accommodation.

\input{tables/tab3_ddd}

Table~\ref{tab:ddd} presents the triple-difference estimates. The coefficient on $\text{Post} \times \text{Black} \times \text{Accommodation}$ is $-0.071$ (SE = 0.027), statistically significant at the 5\% level. This means that Black accommodation employment grew 7.1 log points less than predicted by the baseline trends in (a) White accommodation, (b) Black manufacturing, and (c) White manufacturing. The $\text{Post} \times \text{Black}$ coefficient is $+0.167$ (SE = 0.057), confirming that Black employment in general rose in adopting states---likely reflecting broader economic trends in these fast-growing states. The accommodation-specific dampening of this growth is the novel finding.

The DDD specification for log hires yields a coefficient on the triple interaction of $-0.038$ (SE = 0.043), directionally consistent but noisier than the employment result. This is consistent with the employment effect operating partly through reduced hiring and partly through increased separations.

\subsection{Robustness}

\input{tables/tab4_robust}

Table~\ref{tab:robust} reports robustness checks. The baseline DDD estimate of $-0.071$ is stable across specifications. Restricting to the pre-COVID period (2005--2019, dropping the 2021 cohort entirely) yields $-0.059$ (SE = 0.031, $p = 0.071$)---same sign, slightly smaller, with fewer treated cohorts providing identification. Dropping only the 2021 wave while retaining the full time span produces $-0.076$ (SE = 0.028), actually larger than the baseline. The Callaway-Sant'Anna estimator with not-yet-treated controls gives an ATT of $+0.042$ (SE = 0.036), very similar to the never-treated baseline of $+0.045$. The TWFE estimator yields $+0.088$ (SE = 0.048)---positive and misleading, illustrating the heterogeneous-treatment-effect bias documented by \citet{goodman2021difference}.

\subsection{Mechanisms}

Two features of the results help adjudicate between competing mechanisms. First, the effect is race-specific: White accommodation workers are unaffected. This rules out generic labor-demand shocks to accommodation in adopting states (such as tourism changes correlated with gun laws) and points to a channel that operates through racial dynamics specifically.

Second, the effect concentrates in accommodation, not manufacturing. This rules out mechanisms that would affect all Black workers in adopting states equally (such as changes in policing or incarceration that might differentially affect Black labor supply) and points to customer-facing exposure as the relevant margin.

Together, these patterns are consistent with a model in which permitless carry alters the racial dynamics of customer-employee interactions in public-facing workplaces---either through differential voluntary exit by Black workers from environments perceived as more threatening, through changes in employer or customer behavior that reduce demand for Black workers in these settings, or through some combination. The data cannot distinguish between supply-side (worker choice) and demand-side (employer/customer discrimination) channels, which is a limitation.

\section{Discussion}
\label{sec:discussion}

These results add a labor market dimension to the permitless carry debate. The existing literature focuses on whether removing permit requirements increases violent crime, with mixed evidence \citep{donohue2019right, lott2020more, rand2020effects}. This paper shows that even if the crime effects are contested, the labor market effects are not neutral: Black workers in accommodation bear a measurable cost.

Importantly, the accommodation gap reflects relative sectoral reallocation, not absolute job loss. Black accommodation employment in adopting states did not decline---it grew, but by less than Black manufacturing employment in the same states. The 7.2 log-point gap implies that for every 100 Black jobs gained in manufacturing after permitless carry adoption, approximately 7 fewer jobs were gained in accommodation than would have been expected under the counterfactual. For a state with roughly 38,000 Black accommodation jobs, this translates to approximately 2,700 fewer accommodation jobs relative to what the manufacturing trend would predict.

The finding also contributes to the literature on racial inequality in labor markets. While much attention has focused on discrimination in hiring \citep{bertrand2004emily, kline2022systemic} and wage-setting \citep{lang2020race}, the accommodation gap reveals a less-studied channel: regulatory changes in one policy domain (firearms) that spill over into labor markets through racially differentiated pathways.

A limitation is that the QWI data do not identify specific occupations within the accommodation sector, so I cannot distinguish front-of-house workers (with maximum customer contact) from back-of-house workers (cooks, housekeeping). Occupation-level data would sharpen the mechanism test. Additionally, the analysis is at the state level, which may mask within-state heterogeneity across urban and rural labor markets.

\section{Conclusion}
\label{sec:conclusion}

Thirty states have removed the permit requirement for concealed firearms carrying. This paper documents a previously unknown consequence: a significant dampening of Black employment growth in customer-facing industries relative to non-customer-facing sectors. The accommodation gap---a 7.2 log-point divergence in Black employment growth between accommodation and manufacturing---suggests that gun policy reshapes the sectoral composition of minority employment. While Black workers in adopting states gained jobs overall, those gains systematically bypassed the accommodation sector. As additional states consider permitless carry legislation, policymakers should weigh not only the debated crime effects but also these distributional labor market consequences.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
