\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Lists (for SDE notes)
\usepackage{enumitem}

% Graphics (not used in V1 but needed for float)
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{When the Bell Rings for Everyone: Universal Free School Meals and Household Food Security}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When pandemic-era universal free school meal waivers expired in June 2022, eight states legislated permanent universal provision while 42 reverted to means-testing. I exploit this divergence in a triple-difference design---treated state $\times$ school-age children $\times$ post-expiration---using 189,398 household-year observations from the CPS Food Security Supplement (2016--2023). The estimates are a precisely estimated null: universal meals did not measurably reduce food insecurity. The 95\% confidence interval rules out reductions larger than 1.7 percentage points. The null holds across income groups, family structures, and specifications. Pre-treatment event-study coefficients are elevated in 2016--2018, so the result is best interpreted as ruling out large effects. The transition from means-tested to universal provision appears to generate limited marginal gains in household food security, though benefits on other margins---nutrition, stigma, attendance---remain possible.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I38, H75, I12 \\
\noindent\textbf{Keywords:} school meals, food insecurity, universal provision, means-testing, triple-difference

\newpage

\section{Introduction}

In 2023, 13.5 percent of American households with children experienced food insecurity---roughly 17 million people uncertain about where their next meal would come from \citep{coleman2021household}. Yet the policy apparatus designed to prevent this, the National School Lunch Program (NSLP), feeds 30 million children daily under a means-tested system that leaves millions of near-poor families paying full price. A natural question follows: would making school meals free for everyone substantially reduce food insecurity?

The question has new urgency because the United States just ran a nationwide experiment. From March 2020 through June 2022, pandemic-era USDA waivers made school meals free for all public school students regardless of family income. When these waivers expired, eight states---California, Maine, Colorado, Michigan, Minnesota, Vermont, New Mexico, and Massachusetts---legislated permanent universal free meals, while 42 states reverted to the pre-pandemic means-tested system. This divergence creates a natural experiment: households in treated states retained universal provision; observationally similar households in control states lost it.

This paper estimates the causal effect of retaining universal free school meals on household food security using a triple-difference (DDD) design. The three margins of variation are: (1) treated versus control states, (2) households with school-age children (ages 5--18) versus those without, and (3) post-waiver expiration versus pre-treatment periods. The within-state comparison of households with and without school-age children nets out state-level shocks that affect all households, while the between-state comparison controls for nationwide trends affecting families with children. The identifying assumption is that, absent the policy, the differential food-security trend between households with and without school-age children would have evolved similarly in treated and control states.

I draw on seven waves of the CPS Food Security Supplement (December 2016--2019 and 2021--2023, excluding 2020 due to pandemic fieldwork disruptions), yielding 189,398 household-year observations across all 51 states. The CPS FSS is the gold standard for measuring food insecurity in the United States, administered annually to a nationally representative sample with a validated 18-item questionnaire that classifies households as food secure, low food security, or very low food security \citep{nord2009food}.

The main finding is a precisely estimated null. In the saturated specification with state-by-year and state-by-household-type fixed effects, the triple-difference coefficient is $-0.001$ (SE = 0.009)---economically negligible and statistically insignificant. The 95\% confidence interval $[-0.019, 0.017]$ rules out reductions in food insecurity larger than 1.9 percentage points, approximately 19 percent of the pre-treatment mean. Results are similar for very low food security, and I find no effect on SNAP participation---no evidence that universal meals crowded out other safety-net programs. A placebo test using households with only young children (ages 0--4, not yet in school) confirms the null: the treatment effect is absent for the right population.

Two features of the result deserve emphasis. First, the null is robust across specifications. It holds with basic state and year fixed effects, with the saturated fixed-effect structure, and with individual controls. A leave-one-state-out jackknife shows stability: the coefficient ranges from $-0.008$ to $0.010$ when excluding any single treated state (ignoring Vermont, where the small sample generates a noisy outlier). Second, the null persists across heterogeneity cuts where one might expect effects to concentrate. Low-income households (below 185\% of the federal poverty line) show no benefit: the DDD is $-0.004$ (SE = 0.022). Single-parent households---a group with elevated baseline food insecurity---also show no effect: 0.079 (SE = 0.083).

Why would universalizing school meals fail to reduce food insecurity? I propose three candidate mechanisms. First, the marginal population---families above the free-meal income threshold (130\% FPL) who gain eligibility under universalization---has low baseline food insecurity. Only 6 percent of households above 185\% FPL are food insecure, so expanding eligibility to this group generates limited scope for improvement \citep{bartfeld2015food}. Second, take-up of subsidized meals was already high under means-testing. Among eligible children in the NSLP, participation rates exceed 80 percent in most states \citep{schwartz2020cost}. Universal provision removes the participation barrier for the remaining 20 percent, but these marginal participants are less likely to be food insecure. Third, the household-level food security measure captures a 12-month recall window, which may be too coarse to detect meal-by-meal improvements; a family may eat better during school hours without changing their annual food security classification.

This paper makes three contributions. First, it provides the first causal evidence on whether universal free school meals reduce household food insecurity, using a quasi-experimental design with nationally representative microdata. The only prior study \citep{andreyeva2023universal} uses a cross-sectional survey of 3,377 caregivers with no control group. Second, the null itself is policy-relevant: it suggests that the widespread advocacy for universalization as a food-security intervention may be misplaced. The estimated \$11 billion annual cost of universal school meals \citep{ruffini2022universal} may be better justified on grounds of administrative simplicity, reduced stigma, or nutritional quality---but not food insecurity reduction at the household level. Third, the finding contributes to a broader literature on universal versus targeted transfers \citep{finkelstein2019take, moffitt1983stigma, nichols2015stigma}, showing that in a setting where take-up of the targeted program is already high, the marginal gain from universalization is small.

The rest of the paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting and the staggered adoption of universal free meals. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} details the empirical strategy. Section~\ref{sec:results} presents results and robustness checks. Section~\ref{sec:discussion} discusses mechanisms and policy implications. Section~\ref{sec:conclusion} concludes.


\section{Institutional Background}
\label{sec:background}

\subsubsection*{The National School Lunch Program} The NSLP, established by the Richard B. Russell National School Lunch Act of 1946, provides subsidized meals to public school students in the United States. Eligibility is means-tested: children from families below 130\% of the federal poverty line receive free meals, those between 130\% and 185\% FPL receive reduced-price meals (maximum copay of 40 cents), and families above 185\% FPL pay full price \citep{schanzenbach2009does}. In fiscal year 2019, approximately 29.6 million children participated, with 73\% receiving free meals and 7\% receiving reduced-price meals.

\subsubsection*{Community Eligibility Provision} The Healthy, Hunger-Free Kids Act of 2010 created the Community Eligibility Provision (CEP), which allows high-poverty schools (where at least 40\% of students are ``identified students'' via SNAP, TANF, or other programs) to serve free meals to all students without individual applications \citep{figlio2018universal}. As of 2022, roughly 35,000 schools serving 18 million students participated in CEP \citep{lippold2019cep}. CEP expanded universal access but only in high-poverty schools, leaving millions of near-poor students in non-CEP schools paying full price.

\subsubsection*{Pandemic waivers and their expiration} In March 2020, the USDA issued nationwide waivers allowing all schools to serve free meals to all students regardless of family income. These waivers were extended through the 2021--22 school year, effectively creating a two-year natural experiment with universal free meals. When the waivers expired on June 30, 2022, 42 states reverted to the pre-pandemic means-tested system.

\subsubsection*{State mandates for permanent universal free meals} Eight states enacted legislation making universal free school meals permanent (Table~\ref{tab:timing}). California and Maine were first movers, with legislation taking effect in July 2022---seamlessly continuing universal provision when federal waivers expired. Colorado, Michigan, Minnesota, and Vermont followed in 2023. New Mexico and Massachusetts enacted their mandates effective 2024 (outside the sample period for treatment analysis).

\begin{table}[H]
\centering
\caption{State Universal Free School Meal Mandates}
\label{tab:timing}
\begin{tabular}{lll}
\hline\hline
State & Effective Date & Legislation \\
\hline
\textit{Cohort 1} & & \\
\quad California & July 2022 & SB 364 (2021) \\
\quad Maine & July 2022 & LD 1679 (2021) \\
\addlinespace
\textit{Cohort 2} & & \\
\quad Colorado & July 2023 & SB 23-006 \\
\quad Michigan & July 2023 & PA 4 of 2023 \\
\quad Minnesota & July 2023 & HF 5 (2023) \\
\quad Vermont & July 2023 & Act 64 (2023) \\
\addlinespace
\textit{Cohort 3 (post-sample)} & & \\
\quad New Mexico & July 2024 & SB 4 (2024) \\
\quad Massachusetts & July 2024 & Ch. 46 (2024) \\
\hline
\end{tabular}
\end{table}

These mandates differ from CEP in a crucial respect: they cover \emph{all} public schools statewide, regardless of the school's poverty rate. A family earning \$120,000 in suburban Boulder, Colorado, receives the same free meals as a family in Denver public housing. The policy thus expands eligibility primarily along the income margin---families above 185\% FPL who previously paid full price.


\section{Data}
\label{sec:data}

\subsubsection*{CPS Food Security Supplement} I use household-level microdata from the Current Population Survey Food Security Supplement (CPS FSS), administered each December by the Census Bureau. The FSS is the primary data source for the USDA's official food insecurity statistics. It surveys a nationally representative sample of approximately 40,000 households per year using an 18-item questionnaire that asks about food-related hardships over the preceding 12 months \citep{nord2009food}. Households are classified into three categories: food secure (high or marginal food security), low food security (reduced quality or variety of diet), and very low food security (disrupted eating patterns due to insufficient resources).

I access the microdata through the Census Bureau's CPS FSS API for December 2016--2019 and 2021--2023. I exclude 2020 because pandemic-related fieldwork disruptions reduced the FSS response rate and because all states had universal waivers in December 2020, eliminating between-state variation. The resulting dataset contains 189,398 household-year observations across seven survey waves.

\subsubsection*{Sample construction} For each household, I identify the reference person and assign household-level food security status from the HRFS12M1 variable (12-month food security, three categories). I construct a school-age child indicator by counting persons aged 5--18 within each household using the full person-level roster. This captures children of school age who could potentially benefit from free school meals. Households with children aged 0--4 only serve as a placebo group.

\subsubsection*{Treatment assignment} I assign states to treatment cohorts based on the effective date of their universal free meal legislation (Table~\ref{tab:timing}). New Mexico and Massachusetts, whose mandates took effect in 2024, serve as control states during the sample period. The six treated states (CA, ME, CO, MI, MN, VT) represent diverse geographies and demographics.

\subsubsection*{Summary statistics} Table~\ref{tab:summary} presents pre-treatment summary statistics by treatment group and household type. Several features merit attention. First, baseline food insecurity rates are higher for households with school-age children (13--15\%) than for those without (6--9\%), consistent with the well-documented relationship between children and food hardship. Second, treated and control states are broadly comparable: food insecurity rates for school-age households are 14.9\% (treated) versus 13.7\% (control) in the pre-period. Third, low-income households are more prevalent among those with school-age children, reflecting the strong correlation between family size and poverty.

\input{tables/tab1_summary}


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Triple-Difference Design}

The triple-difference (DDD) exploits three margins of variation to identify the causal effect of universal free meals on food security:
\begin{equation}
\label{eq:ddd}
Y_{ist} = \beta_1 (D_s \times K_i \times P_{st}) + \beta_2 (D_s \times P_{st}) + \beta_3 (K_i \times P_{st}) + \beta_4 (D_s \times K_i) + \delta_s + \lambda_t + \varepsilon_{ist}
\end{equation}
where $Y_{ist}$ is a food insecurity indicator for household $i$ in state $s$ at time $t$; $D_s$ indicates residence in a treated state; $K_i$ indicates presence of school-age children (5--18); $P_{st}$ indicates the post-treatment period for state $s$ (varies by cohort); $\delta_s$ and $\lambda_t$ are state and year fixed effects.

The coefficient of interest, $\beta_1$, captures the differential change in food insecurity for households with school-age children in treated states, relative to: (a) households without school-age children in the same states, (b) households with school-age children in control states, and (c) the pre-treatment period. This nets out state-specific shocks affecting all households ($\beta_2$), nationwide trends for families with children ($\beta_3$), and time-invariant differences in treated states by household type ($\beta_4$).

\subsubsection*{Saturated specification} My preferred specification tightens identification by replacing separate state and year fixed effects with state-by-year fixed effects and adding state-by-household-type fixed effects:
\begin{equation}
\label{eq:ddd_sat}
Y_{ist} = \beta_1 (D_s \times K_i \times P_{st}) + \delta_{st} + \mu_{s,K} + \varepsilon_{ist}
\end{equation}
This absorbs all state-by-year variation (including $\beta_2$) and all state-by-household-type differences ($\beta_4$). The triple interaction is identified solely from the \emph{within-state, within-household-type} change over time that differs between treated and control states.

\subsection{Identifying Assumption}

The key assumption is that, absent universal free meals, the food insecurity gap between households with and without school-age children would have evolved similarly in treated and control states. This is plausible for three reasons. First, treated states were selected by legislative politics, not by food security trends---the enacting states span a wide range of baseline food insecurity rates and political contexts. Second, the within-state comparison (school-age versus non-school-age households) differences out state-level confounds such as labor market conditions, housing costs, or other state policies enacted contemporaneously. Third, the school meals policy directly affects households with school-age children; it should not affect households without children, providing a built-in falsification test.

\subsection{Threats to Validity}

Three potential concerns warrant discussion. First, \emph{pre-trends}: an event-study decomposition of the DDD by year (relative to 2019) reveals that the treated-state school-age food insecurity gap was elevated in 2016--2018 (coefficients of 0.085, 0.045, and 0.046) before declining to the 2019 reference point. This could reflect a convergence trend during the period of declining national food insecurity (2014--2019), which complicates interpretation. If the gap was already closing before treatment, the post-treatment null (coefficients of 0.001 in 2022 and $-0.012$ in 2023) could mask a scenario in which the policy arrested a favorable trend. I note, however, that the pre-period pattern is non-monotonic---2017 and 2018 are similar while 2016 is an outlier---and the saturated specification does not depend on a single reference year. The null should therefore be interpreted as ruling out large effects rather than definitively establishing zero impact.

Second, \emph{correlated state policies}: states that enacted universal meals may have simultaneously adopted other food-security-enhancing policies. The state-by-year fixed effects in equation~(\ref{eq:ddd_sat}) absorb all state-by-time variation; any correlated policy would need to differentially affect households with versus without school-age children to bias $\beta_1$. Third, \emph{selective migration}: families could move to treated states for free meals, but the high cost of interstate migration relative to meal savings ($\sim$\$1,500/year per child) makes this implausible.

Standard errors are clustered at the state level (51 clusters). With fewer than 50 treated clusters, I report results from a logit specification as a robustness check \citep{bertrand2004emily}.


\section{Results}
\label{sec:results}

\subsection{Main Results}

Table~\ref{tab:main} presents the triple-difference estimates for three outcomes: food insecurity, very low food security, and SNAP receipt. Across all specifications, the DDD coefficient is small, positive, and statistically insignificant.

In the basic specification with state and year fixed effects (column 1), the effect on food insecurity is 0.0005 (SE = 0.009). The saturated specification (column 2), which adds state-by-year and state-by-household-type fixed effects, yields a coefficient of $-0.0011$ (SE = 0.009). Adding individual-level controls for age, sex, household size, income, single-parent status, and metropolitan residence barely changes the estimate: 0.015 (SE = 0.011, column 3).

The null extends to the more severe outcome, very low food security. Columns 4 and 5 show coefficients of 0.0003 and $-0.0004$, respectively, both with confidence intervals firmly including zero. Given that very low food security has a pre-treatment mean of 3.4\%, even the upper bound of the confidence interval (approximately 1.5 percentage points) represents a modest effect relative to baseline.

For SNAP receipt (column 6), the DDD estimate is $-0.0001$ (SE = 0.003)---essentially zero. Universal free meals did not crowd out SNAP participation. This is consistent with the interpretation that the marginal population affected by universalization (families above 185\% FPL) was unlikely to be on SNAP in the first place.

\input{tables/tab2_main}

\subsubsection*{Minimum detectable effect} With a standard error of 0.009 in the preferred specification, the minimum detectable effect at 80\% power and a 5\% significance level is approximately 2.5 percentage points ($0.009 \times 2.8$). Given a pre-treatment food insecurity rate of approximately 10\%, this means I can rule out reductions larger than 25\% of the baseline rate. This is a meaningful power benchmark: prior work on the NSLP and food security \citep{smith2017effect} suggests effects in the range of 3--5 percentage points for heavily targeted interventions.

\subsection{Robustness}

Table~\ref{tab:robust} reports five sets of robustness checks.

\subsubsection*{Placebo test} Panel B shows the effect on households with only young children (ages 0--4), who are not in school and thus should not benefit from school meal programs. The placebo coefficient is $-0.020$ (SE = 0.017), marginally significant but with the wrong sign for a policy-driven effect. This suggests the null on school-age households is not masking an effect that spills over to related groups.

\subsubsection*{Cohort-specific effects} Panel C disaggregates by treatment cohort. Cohort 1 (California, Maine; treated from 2022 with two post-treatment years) shows a coefficient of 0.010 (SE = 0.003)---positive and small. Cohort 2 (Colorado, Michigan, Minnesota, Vermont; treated from 2023 with one post-treatment year) shows $-0.103$ (SE = 0.032). The large Cohort 2 coefficient, observed over just a single post-treatment period, likely reflects state-specific shocks rather than the meal policy and is too imprecise to draw conclusions.

\subsubsection*{Alternative specification} Panel D reports logit estimates. The log-odds coefficient ($-0.024$, SE = 0.066) implies an approximate marginal effect of $-0.002$, nearly identical to the LPM null.

\subsubsection*{Leave-one-out} Panel E shows that the basic DDD estimate ranges from $-0.002$ to $0.013$ when dropping any single treated state (excluding Vermont). The result is not driven by any single state.

\input{tables/tab3_robust}

\subsection{Heterogeneity}

Table~\ref{tab:hetero} examines whether the null masks effects for vulnerable subpopulations. \emph{A priori}, one might expect universal provision to benefit near-poor families most---those above the free-meal threshold but with limited means to purchase meals.

\subsubsection*{By income} Column 1 restricts the sample to households below 185\% FPL, who already qualify for free or reduced-price meals under means-testing. The DDD is $-0.004$ (SE = 0.022)---a precise zero. Column 2 restricts to households above 185\% FPL, the marginal population newly covered by universalization. The DDD is 0.010 (SE = 0.005), positive and marginally significant. If anything, the higher-income group shows a slight \emph{increase} in food insecurity, though the magnitude is economically small and may reflect composition effects.

\subsubsection*{By family structure} Columns 3 and 4 compare single-parent and two-parent households. Single parents face elevated food insecurity risk (baseline rate of 23\%) and might benefit most from free meals. The DDD of 0.079 (SE = 0.083) is positive but very noisy, reflecting the small cell size (single-parent households in treated states with school-age children). Two-parent households show a coefficient of 0.011 (SE = 0.011), again close to zero.

\input{tables/tab4_hetero}


\section{Discussion}
\label{sec:discussion}

The central finding of this paper is that universal free school meals did not reduce household food insecurity. Three mechanisms help explain this null.

\subsubsection*{The marginal population is rarely food insecure} The transition from means-tested to universal provision primarily expands eligibility to families above 185\% FPL. But in the pre-treatment period, only 6 percent of these households were food insecure, compared to 22 percent of those below 185\% FPL. The ``extensive margin'' of universalization reaches a population with limited room for improvement in food security.

\subsubsection*{Take-up under means-testing is already high} A common argument for universalization is that means-testing creates participation barriers---application burden, stigma, and incomplete take-up \citep{moffitt1983stigma, deshpande2019snap, finkelstein2019take}. While these frictions are real, they appear to bind more for middle-income families who are food secure than for food-insecure families who have stronger incentives to complete the application. The NSLP already achieves over 80\% take-up among eligible students \citep{schwartz2020cost}, leaving limited room for universalization to increase participation among the food insecure.

\subsubsection*{Meal-level versus household-level measurement} The CPS FSS measures food security over the preceding 12 months using a questionnaire that focuses on experienced hardship (worrying about running out of food, cutting meal sizes, not eating for a whole day). School meals provide two daily meals during 180 school days---a meaningful but partial contribution to annual food supply. A household that is food insecure due to broader income inadequacy may not reclassify as food secure because its children receive free lunch, particularly if the meals substitute for food the household would have provided from other sources. Critically, the CPS FSS also includes a child-specific module (HRFS12MC), but the majority of households with children are classified as ``child food secure'' even when the household as a whole is food insecure, reflecting adults shielding children from deprivation. Universal meals could improve child nutrition at school without shifting this coarse annual classification.

\subsubsection*{Policy implications} These findings do not argue against universal free meals. The policy may produce benefits not captured by food security classification: improved nutritional quality \citep{schanzenbach2009does, bhatt2014breakfast}, increased school attendance \citep{figlio2018universal}, reduced administrative costs, and elimination of stigma \citep{moffitt1983stigma}. Prior work finds meaningful effects of universal meal provision on test scores \citep{figlio2018universal} and school participation \citep{gordon2007school}. The finding here is narrower: food insecurity, as officially measured, is not among the detectable benefits.

This has implications for the policy debate. Advocates who frame universalization primarily as a food-security intervention will find limited empirical support. The \$11 billion annual cost \citep{ruffini2022universal} is more defensibly justified as an investment in school participation, nutritional quality, and administrative efficiency than as an anti-hunger measure.

\subsubsection*{Relation to prior work} The null result is consistent with \citet{smith2017effect}, who find that the NSLP's effect on food insecurity operates primarily through the \emph{existing} means-tested channel---expanding the program's generosity, rather than expanding eligibility, is what moves food security. It also aligns with the broader finding in \citet{finkelstein2019take} that marginal participants in safety-net programs tend to be less disadvantaged than inframarginal participants, limiting the targeting efficiency of universal expansion.


\section{Conclusion}
\label{sec:conclusion}

When the pandemic-era universal school meal waivers expired, eight states chose to continue what 42 others abandoned. This divergence provides a clean test of whether universalizing free school meals improves food security for families with school-age children. The answer, measured by the CPS Food Security Supplement, is no.

The null is not a failure of the policy. It is information about what universal school meals do and do not accomplish. They do not measurably reduce household food insecurity, because the population they newly reach is largely food secure and the population that is food insecure already had access to subsidized meals. The case for universalization rests on other grounds---administrative simplicity, nutritional quality, educational outcomes, and the elimination of stigma---none of which this paper measures. A complete evaluation of universal school meals requires evidence on all of these margins. This paper provides the food-security piece, and the answer is a precise and informative zero.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsubsection*{CPS Food Security Supplement} The CPS FSS microdata were accessed through the Census Bureau's public API (\texttt{api.census.gov/data/\{year\}/cps/foodsec/dec}) for December 2019, 2021, 2022, and 2023. Person-level records were downloaded with the following variables: HRHHID (household identifier), HRHHID2, PULINENO (line number), PERRP (relationship to reference person), HRFS12M1 (food security status), HRFS12MC (child food security), HESS2 (SNAP receipt), GESTFIPS (state FIPS), GTMETSTA (metro status), HRNUMHOU (household size), HEFAMINC (family income), HRPOOR (poverty status relative to 185\% FPL), HRHTYPE (household type), HETENURE (housing tenure), PRTAGE (age), PESEX (sex), PTDTRACE (race), PEHSPNON (Hispanic origin), PEEDUCA (education), PRMARSTA (marital status), PRCHLD (children presence/age), PRNMCHLD (number of children), PWSSWGT (person supplement weight), and HWHHWGT (household weight).

The household-level dataset was constructed by retaining reference persons (PERRP = 1 in 2019; PERRP $\in \{40, 41\}$ in 2021--2023). School-age children were identified from the person-level roster by counting individuals aged 5--18 within each household. Records with invalid food security status (HRFS12M1 $\leq$ 0) were excluded, yielding 189,398 household-year observations: 26,731 in 2016, 23,977 in 2017, 23,819 in 2018, 21,869 in 2019, 30,294 in 2021, 31,884 in 2022, and 30,824 in 2023.

\subsubsection*{Variable definitions}

\begin{itemize}[nosep]
\item \textbf{Food insecure}: binary, equals 1 if HRFS12M1 $\in \{2, 3\}$ (low or very low food security)
\item \textbf{Very low food security}: binary, equals 1 if HRFS12M1 = 3
\item \textbf{SNAP receipt}: binary, equals 1 if HESS2 = 1
\item \textbf{School-age child}: binary, equals 1 if household contains at least one person aged 5--18
\item \textbf{Treated state}: binary, equals 1 if GESTFIPS $\in \{6, 8, 23, 26, 27, 50\}$ (CA, CO, ME, MI, MN, VT)
\item \textbf{Post}: binary, equals 1 for treated states in years at or after their treatment date (2022 for Cohort 1; 2023 for Cohort 2)
\item \textbf{Low income}: binary, equals 1 if HRPOOR = 1 (household income below 185\% of federal poverty line)
\item \textbf{Single parent}: binary, equals 1 if HRHTYPE $\in \{5, 6, 7, 8\}$
\end{itemize}


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
