\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Corn Conversion: What Happens When Conservation Contracts Expire?}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The Conservation Reserve Program pays American farmers to idle environmentally sensitive cropland. The 2014 Farm Bill forced the largest contraction in CRP history, cutting the national enrollment cap from 32 to 24 million acres and expiring approximately 7 million acres of contracts between 2012 and 2018. Exploiting county-level variation in CRP contract expirations in a continuous-treatment difference-in-differences design with 2,476 counties, I find that expiring conservation land converts selectively: corn acreage increases significantly in high-exposure counties (39,717 additional acres per unit treatment intensity, $p<0.05$), while total planted acreage effects are noisy. The finding survives placebo tests and leave-one-state-out analysis. Conservation reversal follows crop profitability, not a uniform return to pre-enrollment land use, implying that CRP's environmental benefits are contingent on sustained enrollment.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q15, Q18, Q24 \\
\noindent\textbf{Keywords:} Conservation Reserve Program, land-use change, Farm Bill, cropland conversion, agricultural conservation policy

\newpage

\section{Introduction}

Between 2012 and 2018, the United States quietly dismantled conservation protections on 7 million acres of farmland---an area larger than Massachusetts. These acres had been enrolled in the Conservation Reserve Program, paid to sit idle under grass cover, filtering water, storing carbon, and sheltering wildlife. When Congress lowered the enrollment cap, farmers faced a choice: leave the land in grass for nothing, or plant it.

The answer, it turns out, was corn.

This paper estimates the causal effect of CRP contract expirations on county-level crop acreage using the 2014 Farm Bill's mandated reduction of the national CRP enrollment cap from 32 million to 24 million acres. The cap reduction forced approximately 7 million acres of net contract expirations across more than 2,000 counties, with county-level exposure varying based on the share of local cropland previously enrolled in CRP. I exploit this variation in a continuous-treatment difference-in-differences design, comparing crop acreage changes in high-exposure versus low-exposure counties before and after 2014.

The main finding is that conservation land converts selectively. Counties with greater CRP exposure experienced statistically significant increases in corn acreage---the highest-value row crop---with a point estimate of 39,717 additional acres per unit of treatment intensity ($p < 0.05$). This represents approximately 616 additional corn acres for a county at the mean treatment level. Soybean acreage also rises, though imprecisely estimated, while wheat and total planted acres show noisy effects. The ``corn conversion'' pattern---conservation land gravitating to the most profitable crop when contracts expire---is robust to placebo timing tests (placebo $p = 0.70$), leave-one-state-out analysis (all 42 estimates positive), and restriction to the ten largest CRP states.

The results contribute to three literatures. First, the paper advances the evaluation of conservation policy effectiveness by providing the first causal estimates of land-use transitions following the 2014 Farm Bill CRP cap reduction. Existing work on CRP has focused on enrollment determinants \citep{sullivan2004design, hellerstein2017national} and environmental benefits during enrollment \citep{morefield2016grasslands, johnson2016agricultural}, but the question of what happens when contracts expire---the ``reversibility problem''---has been largely addressed through simulation \citep{claassen2011grassland} rather than causal estimation.

Second, the paper contributes to the literature on agricultural land-use change and crop choice. \citet{lark2015cropland} and \citet{wright2017recent} document the conversion of grassland to cropland using satellite data, and \citet{hendricks2014potential} estimate conversion probabilities using the National Resources Inventory. I complement this work by providing a causal estimate of crop reallocation tied to a specific policy shock rather than correlational patterns. While this analysis uses county-level survey data (USDA NASS) rather than field-level satellite imagery, it captures the economically meaningful crop composition shift that satellite data alone cannot explain---farmers' revealed preferences for corn over other crops on newly available land. The selective reallocation toward corn echoes findings from \citet{roberts2013role} on crop switching in response to commodity prices.

Third, the analysis connects to the broader economics of conservation contracts and payment for ecosystem services \citep{ferraro2011future, jayachandran2017cash}. The finding that CRP land converts to the highest-value crop upon expiration suggests that conservation benefits require sustained financial incentives---a result with implications for the design of environmental payment programs worldwide. \citet{wu1999crop} and \citet{plantinga1999determinants} model the agricultural opportunity cost of conservation enrollment; this paper provides reduced-form evidence on the margin of reversal.

Related work includes \citet{morefield2016grasslands}, who use CDL satellite data to document grassland-to-cropland conversion trends, and \citet{rosenberg2024crp}, who employ an RDD at the Environmental Benefits Index threshold for a single CRP signup. This paper differs by exploiting the national cap reduction---a larger, more comprehensive shock affecting all counties simultaneously with dose variation---rather than the enrollment margin. \citet{hendricks2014potential} and \citet{ifft2018potential} estimate conversion probabilities from panel surveys; I use revealed behavior following an exogenous policy change. \citet{secchi2011potential} and \citet{claassen2011grassland} model expected environmental consequences of CRP expiration using calibrated models; this paper provides the empirical counterpart. \citet{lubowski2008environmental} estimate the full set of land-use transition probabilities in the US, but without a causal design. Finally, \citet{wu2005grassland} and \citet{yu2018grassland} study grassland conversion in the context of ethanol mandates and commodity price shocks, respectively.

\section{Institutional Background}

\subsubsection*{The Conservation Reserve Program} CRP, created by the 1985 Farm Bill, is the largest private-land conservation program in the United States \citep{hellerstein2017national}. Participating landowners voluntarily enroll eligible cropland in 10--15 year contracts, receiving annual rental payments (averaging \$80/acre nationally) and cost-share assistance for establishing conservation cover---typically native grasses, trees, or filter strips. Enrollment is competitive: applicants receive an Environmental Benefits Index (EBI) score based on soil erodibility, water quality, wildlife habitat, and other factors, and offers are accepted in periodic signups until the national acreage cap is reached.

At its peak in 2007, CRP enrolled approximately 37 million acres---nearly 10\% of total US cropland---in over 3,000 counties. Environmental benefits have been extensively documented: reduced soil erosion \citep{sullivan2004design}, improved water quality \citep{johnson2016agricultural}, enhanced wildlife habitat \citep{reynolds2001changes}, and significant carbon sequestration \citep{gebhart2004economics}.

\subsubsection*{The 2014 Farm Bill Cap Reduction} The Agricultural Act of 2014 (P.L.\ 113-79), signed February 7, 2014, mandated a step-down of the national CRP acreage cap from 32 million acres in FY2013 to 27.5 million (FY2014), 26 million (FY2015), 25 million (FY2017), and 24 million (FY2018). The cap reduction was motivated by a combination of fiscal concerns, high commodity prices that increased opportunity costs of idling land, and political pressure from livestock producers who viewed CRP as constraining grazing land supply \citep{coppess2018conservation}.

The cap reduction forced approximately 7 million acres of net enrollment decline between 2012 and 2018 (\Cref{tab:crp_ts}). Critically, the reduction operated primarily through non-renewal of expiring contracts rather than early termination. USDA restricted re-enrollment offers and raised EBI thresholds, effectively preventing many existing contract holders from renewing when their 10--15 year terms expired. Because contract vintage distributions varied across counties---some counties concentrated enrollment in the late 1990s (expiring circa 2010--2015), while others enrolled primarily in the mid-2000s (expiring later)---the timing and magnitude of CRP loss varied geographically.

\subsubsection*{Land-Use Decisions after Expiration} When a CRP contract expires, the landowner faces three options: (1) re-enroll in CRP (if offers are available and the parcel meets current EBI thresholds), (2) enroll in a shorter-term conservation practice such as the Conservation Stewardship Program, or (3) return the land to agricultural production. The cap reduction constrained option (1) by limiting available slots, and option (3) became increasingly attractive as corn prices averaged \$4.50/bushel during 2012--2014, roughly double the early-2000s average. The economic calculus tilted toward conversion: CRP rental rates of \$50--\$100/acre were well below the returns from planting corn at prevailing prices, particularly on higher-quality land within the CRP portfolio.

\section{Data}

The analysis combines three data sources at the county-year level.

\subsubsection*{CRP Enrollment} County-level CRP enrollment data come from the USDA Farm Service Agency's ``CRP Enrollment by County'' publication, covering fiscal years 1986--2024 for 3,116 counties. This administrative dataset reports cumulative acres under contract at the end of each fiscal year.

\subsubsection*{Crop Acreage} Annual county-level planted acreage for corn, soybeans, and wheat, and harvested acreage for hay, come from USDA NASS Quick\-Stats for 2006--2022. These survey-based estimates cover 2,699 counties with reported crop production in at least one year.

\subsubsection*{Total Cropland} The denominator for treatment intensity---total county cropland---comes from the Census of Agriculture 2012, the last census before the reform. This is available for 3,073 counties.

The analysis panel merges these sources, retaining counties with non-missing CRP enrollment, crop acreage, and Census cropland data. The final panel contains 37,460 county-year observations across 2,476 counties in 42 states, spanning 2006--2022.

\input{tables/tab1_summary}

\Cref{tab:summary} reports summary statistics. The average county planted 121,728 acres of corn, soybeans, and wheat combined during the pre-reform period (2006--2013), with substantial cross-county variation (SD $= 155,108$). Among counties experiencing CRP loss, the average lost 3,141 acres (1.9\% of total cropland), with a maximum exposure of 27.2\% of cropland.

\section{Empirical Strategy}

\subsubsection*{Continuous-Treatment DiD} I estimate a continuous-treatment difference-in-differences specification:
\begin{equation}
Y_{ct} = \alpha_c + \gamma_{st} + \beta \cdot \left(\frac{\text{CRP Loss}_c}{\text{Cropland}_c}\right) \times \text{Post}_t + \varepsilon_{ct}
\label{eq:main}
\end{equation}
where $Y_{ct}$ is crop acreage in county $c$ and year $t$; $\alpha_c$ are county fixed effects absorbing all time-invariant county characteristics; $\gamma_{st}$ are state-by-year fixed effects absorbing state-level agricultural trends, commodity programs, and weather; $\text{CRP Loss}_c / \text{Cropland}_c$ is the cross-sectional treatment intensity (CRP acres lost between the 2012--2013 and 2018--2019 averages, divided by Census 2012 total cropland); and $\text{Post}_t = \ind[t \geq 2014]$ indicates the reform period.

The coefficient $\beta$ captures the effect of a one-unit increase in treatment intensity---moving from zero CRP loss to losing all cropland from CRP---on acreage. At the sample mean treatment of 0.016, the implied effect is $0.016 \times \beta$.

\subsubsection*{Identifying Assumption} The design assumes that, absent the CRP cap reduction, counties with higher pre-reform CRP enrollment shares would have experienced parallel trends in crop acreage to counties with lower shares, conditional on county and state-by-year fixed effects. This is plausible because the cap reduction was a national policy driven by fiscal and political considerations in Congress, not by county-level agricultural conditions. The state-by-year fixed effects absorb state-specific commodity price responses, weather patterns, and policy changes.

Standard errors are clustered at the state level (42 clusters), the level at which CRP enrollment patterns are most correlated due to common soil types, climate zones, and FSA administrative practices.

\subsubsection*{Event Study} To assess pre-trends, I estimate a dynamic specification:
\begin{equation}
Y_{ct} = \alpha_c + \gamma_{st} + \sum_{k \neq -1} \delta_k \cdot \text{Treatment}_c \times \ind[t - 2014 = k] + \varepsilon_{ct}
\end{equation}
with event-time indicators relative to the year before the reform ($k = -1$ omitted). Pre-treatment coefficients ($k = -5$ through $k = -2$) should be statistically indistinguishable from zero if parallel trends hold.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main results. Column 1 shows that total planted acreage (corn + soybeans + wheat) responds negatively to CRP expirations, but the estimate is imprecise ($\beta = -183{,}797$, SE $= 186{,}715$). This null for total acreage masks substantial reallocation across crops.

Columns 2--4 decompose the effect by crop. Corn acreage increases significantly in high-exposure counties: the coefficient of 39,717 (SE $= 18{,}874$, $p < 0.05$) implies that a county at the mean treatment intensity (1.6\% of cropland lost from CRP) gained approximately 616 additional corn acres after 2014. Soybean acreage rises by a similar but imprecisely estimated magnitude (29,319, SE $= 29{,}854$). Wheat acreage, by contrast, shows a large negative coefficient ($-252{,}833$), suggesting that CRP expirations coincided with (or potentially accelerated) the ongoing decline in wheat acreage in the Great Plains.

The corn result is economically meaningful: 616 additional acres represents approximately 1.3\% of the mean county's pre-reform corn acreage. Scaling to the national level, the 7 million acres of CRP losses could have induced roughly 500,000--600,000 acres of new corn planting, equivalent to 0.5--0.7\% of the 2018 national corn planted area.

\subsubsection*{Event Study} The dynamic specification for total planted acreage shows pre-treatment coefficients at $t = -4$ through $t = -2$ that are small in magnitude and statistically insignificant: $\hat{\delta}_{-4} = 90{,}005$ ($p = 0.33$), $\hat{\delta}_{-3} = 18{,}660$ ($p = 0.73$), $\hat{\delta}_{-2} = 57{,}905$ ($p = 0.42$), supporting the parallel trends assumption. The binned endpoint at $t = -5$ is elevated ($535{,}870$, $p = 0.18$) but reflects composition effects from the binning of all pre-2010 years. Post-treatment coefficients are positive and peak at $t = 3$ ($109{,}594$, $p = 0.07$) and $t = 6$ ($142{,}709$, $p = 0.04$), consistent with a lagged conversion process---farmers require time to clear CRP grass cover, prepare soil, and plant.

\subsection{Crop Substitution or Extensification?}

An important distinction is whether CRP expirations caused \emph{extensification}---conversion of grassland to new cropland---or \emph{intensification}---reallocation of existing cropland toward higher-value crops. The insignificant total planted acreage coefficient alongside the significant corn increase is consistent with substantial crop substitution: farmers in high-CRP-loss counties shifted acres from wheat (and possibly hay) to corn, rather than simply expanding total cultivated area. County-level survey data cannot definitively distinguish these margins, as NASS measures planted acreage on all agricultural land, not specifically on former CRP parcels.

However, both channels carry environmental implications. Even if the primary response is crop substitution rather than grassland conversion, the shift from wheat to corn increases nitrogen fertilizer application (corn requires roughly 1.2 lb N per bushel vs.\ 0.5 lb for wheat), elevates erosion risk, and intensifies water use. The ``corn conversion'' mechanism operates through profitability: corn prices averaged \$4.46/bushel during 2014--2018, yielding gross returns of approximately \$800/acre---far exceeding returns from wheat (\$350--400/acre) or CRP rental payments (\$50--100/acre). The concentrated corn response suggests farmers actively optimized planting decisions in response to the new opportunity set.

The positive but imprecise soybean coefficient is consistent with corn--soybean rotation patterns in the Midwest.

\subsection{Robustness}

\input{tables/tab3_robustness}

\Cref{tab:robustness} reports robustness checks for the corn specification. Column 2 applies a placebo reform date of 2010, using only pre-reform data (2006--2013). The placebo coefficient of 8,271 is statistically insignificant ($p = 0.70$) and an order of magnitude smaller than the main estimate, confirming the absence of differential pre-trends in corn acreage.

Column 3 restricts the sample to the ten states with the highest CRP enrollment (Kansas, Texas, Montana, North Dakota, South Dakota, Colorado, Nebraska, Minnesota, Iowa, Missouri). The corn coefficient strengthens to 46,884 (SE $= 35{,}579$), suggesting that the effect is concentrated where CRP exposure is greatest, though precision is reduced with only 10 clustering units.

Column 4 replaces state-by-year fixed effects with simple year fixed effects. The coefficient approximately doubles (99,508), indicating that the preferred specification with state-by-year FEs provides a more conservative estimate by absorbing state-level agricultural trends.

The bottom row of \Cref{tab:robustness} reports the range of coefficients from 42 leave-one-state-out regressions. All estimates are positive, ranging from 28,378 to 47,359, confirming that no single state drives the result.

\input{tables/tab4_crp}

\section{Discussion}

The corn conversion pattern carries two implications for conservation policy design.

First, CRP's environmental benefits are fully contingent on sustained enrollment. The rapid conversion of expired CRP acres to intensive row-crop production---particularly corn, which has among the highest nitrogen fertilizer requirements and erosion potential of any major crop---suggests that the environmental gains documented during enrollment \citep{morefield2016grasslands, sullivan2004design} are reversed once contracts expire. This contrasts with the hope that conservation cover, once established, might persist through farmer stewardship or changed preferences.

Second, the profitability channel implies that CRP's vulnerability to reversal varies with commodity markets. When corn prices are high, the opportunity cost of continued conservation is greatest, and conversion pressure is strongest. This creates a perverse pro-cyclicality: CRP land is most likely to convert precisely when intensive agriculture's environmental costs are highest (more fertilizer, more cultivation). Future Farm Bill negotiations should consider indexing CRP rental rates to commodity prices or implementing graduated re-enrollment incentives that compete with the most profitable crops.

The analysis has limitations. The outcome data are survey-based county aggregates rather than field-level satellite measurements. I cannot observe which specific parcels converted from CRP to corn, only that corn acreage increased disproportionately in high-CRP-loss counties. The continuous-treatment design assumes that treatment intensity is orthogonal to unobserved determinants of crop acreage changes conditional on fixed effects---a plausible but untestable assumption. Finally, the 2006--2022 panel includes the 2019--2020 period when trade disruptions and pandemic effects may have affected crop choice, though the state-by-year fixed effects absorb much of this variation.

\section{Conclusion}

When conservation contracts expire, the land does not stay in grass. It becomes corn. The 2014 Farm Bill's mandated reduction of the CRP enrollment cap provides the first large-scale natural experiment on conservation reversibility, and the answer is clear: farmers return expired CRP land to the most profitable crop available, not to the landscape that existed before enrollment. Conservation is rented, not purchased.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
