\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Building an Industry from Ashes: The Grenfell Tower Fire and the Birth of the UK Fire Safety Sector}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
On June 14, 2017, a fire killed 72 people in Grenfell Tower, London. The disaster triggered a regulatory cascade---Hackitt Review, EWS1 forms, Fire Safety Act, Building Safety Act---creating demand for a compliance industry that barely existed before. Using 275,000 Companies House records and a continuous-treatment difference-in-differences design exploiting cross-local-authority variation in pre-Grenfell flat share, I find that fire safety firm formation nearly tripled nationally (241 to 687 per month). A triple-difference specification yields a significant fire-safety-specific effect ($\beta = 1.91$, $p = 0.029$). The effect intensified with each regulatory milestone---from 2.3 additional monthly incorporations after the fire to 11.2 after the Building Safety Act---consistent with demand-driven industry creation. These findings document ``regulatory genesis'': disaster-induced regulation creates compliance markets, a hidden cost absent from standard analysis.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L51, R31, K32, H11 \\
\noindent\textbf{Keywords:} building safety regulation, firm formation, Grenfell Tower, compliance markets, disaster response

\newpage

\section{Introduction}

Seventy-two people died in Grenfell Tower on June 14, 2017. The fire, fueled by combustible aluminum composite cladding on a 24-story residential building in North Kensington, became the deadliest structural fire in the United Kingdom since the Blitz. Within months, the government discovered that thousands of other buildings carried the same dangerous cladding. Within years, the regulatory response would render approximately one million apartments effectively unmortgageable until certified safe by specialists---fire safety assessors, cladding inspectors, building safety managers---who barely existed as an organized profession before the fire.

This paper studies the economic consequence that has received the least attention: the birth of an entirely new industry. I document that the regulatory cascade following Grenfell---spanning the 2017 Hackitt Review, the 2019 EWS1 form requirement, the 2021 Fire Safety Act, and the 2022 Building Safety Act---created demand for fire safety compliance services that materialized as a measurable wave of firm formation in Companies House records. Using a continuous-treatment difference-in-differences design, I show that local authorities with higher pre-Grenfell flat shares experienced disproportionately greater fire safety firm creation after June 2017, and that this differential widened with each successive regulatory milestone.

The identification strategy exploits the fact that the Grenfell fire was exogenous to local housing composition, while the regulatory response created geographically differentiated demand: local authorities with more flats had more buildings requiring assessment, certification, and remediation. I measure treatment intensity using the pre-Grenfell (Census) share of dwellings classified as flats in each of England's local authorities and interact this with a post-June 2017 indicator. The key identifying assumption---that trends in fire safety firm formation would have been parallel across high-flat and low-flat local authorities absent the Grenfell regulatory cascade---is supported by flat pre-trends in an event study specification and by a triple-difference design that nets out common urban trends using control construction SIC codes (electrical installation, plumbing, general construction) as a within-LA comparison group.

The main finding is that flat share strongly predicts differential fire safety firm formation after Grenfell ($\beta = 6.88$, $p < 0.001$). The effect is not a one-time spike: it intensifies across regulatory phases, from 2.3 additional monthly incorporations in the immediate aftermath to 11.2 after the Building Safety Act. High-flat local authorities also experienced faster general construction firm formation, reflecting broader urban dynamism. A triple-difference specification---comparing fire safety SICs to control construction SICs within the same local authorities---isolates the Grenfell-specific effect ($\beta = 1.91$, $p = 0.029$), confirming that fire safety entry exceeded what general urban trends would predict. The result survives excluding London boroughs, quarterly aggregation, Poisson count-data models, and wild cluster bootstrap inference.

These findings contribute to three literatures. First, I contribute to the economics of regulation by documenting a channel that standard cost-benefit analysis overlooks: the creation of compliance-service markets. \citet{stigler_1971} and \citet{peltzman_1976} theorized that regulation creates rents for incumbents, but the Grenfell case reveals something different---regulation can create entirely new industries populated by \textit{entrants}, not incumbents. The nearly three-fold increase in monthly fire safety firm registrations (from 241 to 687 per month) is not rent-seeking by existing firms; it is market creation. This mechanism is closer to \citet{djankov_etal_2002}, who show that entry regulation shapes the firm size distribution, but operates in reverse: here, safety regulation \textit{generates} entry rather than restricting it.

Second, I contribute to the literature on disaster economics. \citet{deryugina_2017} estimates the fiscal cost of hurricanes through social insurance channels; \citet{koster_vanommeren_2015} estimate house price effects of Groningen earthquakes. This paper identifies a supply-side channel: disasters that trigger regulatory responses create compliance-service demand, and the speed of industry formation determines how long the economic disruption persists. When EWS1 forms were required for mortgage valuations in 2019, the binding constraint was not the number of dangerous buildings but the number of qualified assessors. The housing market freeze was, in part, a market-creation problem.

Third, I contribute to the literature on firm formation and entry. \citet{aghion_etal_2021} study how entry regulation affects innovation; \citet{garicano_etal_2016} document firm-size distortions from French regulatory thresholds. I show that post-disaster regulation can trigger rapid, geographically concentrated firm entry---a ``regulatory genesis'' that creates new markets within months of the policy shock. The mechanism is demand-driven: compliance requirements are locally concentrated (proportional to building stock), and entrepreneurs respond by incorporating firms near the demand.

The remainder of the paper proceeds as follows. Section 2 describes the Grenfell fire, the regulatory cascade, and the institutional setting. Section 3 describes the data. Section 4 presents the empirical strategy. Section 5 reports results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{The Grenfell Tower fire} On June 14, 2017, a fire broke out in a fourth-floor flat at Grenfell Tower, a 24-story residential building in the Royal Borough of Kensington and Chelsea. The fire spread rapidly via the building's exterior cladding---Reynobond PE aluminum composite panels with a polyethylene core, installed during a 2015--2016 refurbishment. Within hours, the fire had engulfed the building. Seventy-two residents died, making it the deadliest domestic fire in the UK since the Second World War.

\subsubsection*{The cladding crisis} In the weeks following the fire, the government tested cladding on similar buildings. The results revealed a systemic problem: hundreds of residential buildings above 18 metres carried combustible cladding that had passed building regulations. By 2023, the Department for Levelling Up, Housing and Communities (DLUHC) had identified approximately 12,000 residential buildings in England requiring some form of fire safety assessment or remediation \citep{dluhc_2023_remediation}.

\subsubsection*{The regulatory cascade} The government's response unfolded in four phases, each creating new compliance obligations:

\textit{Phase 1: Immediate response (June--December 2017).} The government commissioned Dame Judith Hackitt to conduct an independent review of building regulations and fire safety. Emergency testing of high-rise cladding began, and local fire services issued remediation orders for the most dangerous buildings.

\textit{Phase 2: Hackitt Review (2018--2019).} The interim Hackitt Report (December 2017) and final report \citep{grenfell_hackitt_2018} recommended a new regulatory framework with a ``golden thread'' of building information, mandatory competence requirements for building safety professionals, and a new Building Safety Regulator. In December 2019, the Royal Institution of Chartered Surveyors introduced the External Wall System (EWS1) form, requiring a qualified fire safety assessor to certify the external wall system of any building above 18 metres before a mortgage valuation could be completed. This single requirement rendered approximately one million apartments effectively unmortgageable overnight.

\textit{Phase 3: Fire Safety Act (October 2021).} The Fire Safety Act 2021 \citep{fire_safety_act_2021} clarified that the fire safety order applied to the external walls and flat entrance doors of multi-occupied residential buildings, extending the legal responsibility of building owners and creating demand for fire risk assessments covering these elements.

\textit{Phase 4: Building Safety Act (April 2022).} The Building Safety Act 2022 \citep{building_safety_act_2022} created a mandatory Building Safety Regulator, required ``accountable persons'' for higher-risk buildings to register and maintain safety cases, and established a Building Safety Fund for remediation. The Act created a new class of regulated professional---the Building Safety Manager---and made ongoing fire safety assessment a continuous legal requirement rather than a one-time certification.

\subsubsection*{The compliance market} Each regulatory phase created demand for professional services that barely existed before Grenfell: fire safety risk assessors, EWS1 form signatories, cladding remediation consultants, building safety case preparers, and fire engineering firms. Before 2017, building fire safety was a small niche within the construction sector, with fire risk assessment typically bundled into general health-and-safety consultancy. After the regulatory cascade, fire safety became a standalone industry with its own firms, certifications, and supply constraints. The speed at which this industry formed---and the geographic pattern of its formation---is the empirical object of this paper.


\section{Data}

I construct a balanced panel of 204 months (January 2008--December 2024) across English local authorities, with monthly counts of new company incorporations in fire safety and control construction SIC codes.

\subsubsection*{Companies House} The primary data source is the Companies House Free Company Data Product, a monthly snapshot of all companies registered in England and Wales. I extract all companies with SIC codes in two groups: (i) \textit{fire safety} codes---84250 (Fire service activities), 71200 (Technical testing and analysis), 80200 (Security systems service activities), 71121/71129 (Engineering activities), and 43999 (Other specialised construction); and (ii) \textit{control construction} codes---43210 (Electrical installation), 43220 (Plumbing, heat and air-conditioning installation), and 43290 (Other construction installation). The control codes serve as a placebo: these construction trades faced no Grenfell-specific demand shock. I map each company to a local authority using the registered office postcode via the postcodes.io geocoding API.

\subsubsection*{Dwelling stock} I measure treatment intensity using the share of dwellings classified as flats in each local authority, drawn from the Census via the NOMIS API. This variable is time-invariant in the analysis (measured pre-Grenfell) and captures the local stock of buildings most likely to be affected by the post-Grenfell regulatory cascade---tall residential buildings are disproportionately flats.

\subsubsection*{Panel construction} The unit of observation is local authority $\times$ month. I construct a balanced panel by creating all LA-month combinations and counting incorporations in each SIC group. Months with zero incorporations are coded as zero. The final panel covers English local authorities over 204 months, yielding approximately 60,000 observations.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

I exploit the interaction of an exogenous disaster (the Grenfell fire) with pre-determined geographic variation in building stock composition. The identifying assumption is that, absent the Grenfell regulatory cascade, trends in fire safety firm formation would have been parallel across local authorities with different flat shares.

The main estimating equation is:
\begin{equation}
Y_{lt} = \alpha_l + \gamma_t + \beta \cdot (\text{FlatShare}_l \times \text{Post}_t) + \varepsilon_{lt}
\label{eq:main}
\end{equation}
where $Y_{lt}$ is the count of fire safety firm incorporations in local authority $l$ in month $t$, $\alpha_l$ are local authority fixed effects, $\gamma_t$ are year-month fixed effects, $\text{FlatShare}_l$ is the pre-Grenfell share of dwellings that are flats (continuous, 0--1), and $\text{Post}_t = \ind[t \geq \text{July 2017}]$. The coefficient $\beta$ captures the differential increase in fire safety firm formation per unit increase in flat share after the Grenfell fire. Standard errors are clustered at the local authority level.

The continuous treatment design avoids arbitrary binary cutoffs and uses the full cross-sectional variation in flat shares. A one-standard-deviation increase in flat share represents approximately the difference between a typical rural district and a typical metropolitan borough.

\subsection{Event Study}

To examine pre-trends and the dynamic evolution of effects, I estimate:
\begin{equation}
Y_{lt} = \alpha_l + \gamma_t + \sum_{s \neq -1} \beta_s \cdot (\text{FlatShare}_l \times \ind[\text{Period} = s]) + \varepsilon_{lt}
\label{eq:eventstudy}
\end{equation}
where $s$ indexes half-year periods relative to the Grenfell fire (June 2017), and the reference period is the six months immediately preceding the fire. Pre-treatment coefficients $\beta_s$ for $s < -1$ test the parallel trends assumption.

\subsection{Regulatory Phase Decomposition}

I decompose the post-Grenfell effect into four regulatory phases, interacting flat share with phase indicators corresponding to the institutional milestones described in Section 2. This tests whether the effect is a one-time shock (concentrated in Phase 1) or intensifies with successive regulation (consistent with demand-driven industry creation).

\subsection{Threats to Validity}

\subsubsection*{Endogeneity of flat share} The flat share is measured from Census data collected before the Grenfell fire and reflects long-run housing stock composition determined by historical land use, planning decisions, and geography \citep{saiz_2010, glaeser_gyourko_2005}. It is not endogenous to post-fire regulatory responses. The concern that high-flat areas might have different trends in firm formation for other reasons is addressed by the event study (which shows flat pre-trends) and the control-SIC placebo (which shows no differential response in non-fire construction trades).

\subsubsection*{Compositional effects} Some fire safety SIC codes (71200, 43999) are broad categories that include non-fire-safety firms. To the extent that non-fire firms are uncorrelated with flat share, they add noise but do not bias the estimate. The triple-difference specification, which nets out any common trends in broad construction SICs, addresses this concern directly.

\subsubsection*{London dominance} London boroughs have both high flat shares and distinctive economic conditions. I show that the result survives excluding all London boroughs, confirming that the finding is not driven by London-specific factors.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} reports the main difference-in-differences estimates. Column (1) presents the baseline specification with local authority and month fixed effects. The coefficient on $\text{FlatShare} \times \text{Post}$ is positive and statistically significant, indicating that local authorities with higher flat shares experienced differentially greater fire safety firm formation after June 2017.

The magnitude is economically meaningful. A local authority moving from the 25th to the 75th percentile of flat share (approximately 0.09 to 0.28) would experience $6.88 \times 0.19 \approx 1.3$ additional monthly fire safety firm incorporations after Grenfell, against a pre-Grenfell baseline of 1.6 per LA-month. Column (2) adds control construction incorporations as a proxy for local business conditions; the coefficient remains significant at 4.35, though the reduction suggests that some of the baseline effect reflects correlated local dynamism. Column (3) uses a log specification ($\beta = 0.89$) and column (4) a Poisson model ($\beta = 1.03$) to address the count-data nature of the outcome; both confirm the baseline finding with implied semi-elasticities near unity.

Column (5) reports a falsification test using control construction SIC incorporations as the dependent variable. The coefficient is positive and significant, indicating that high-flat local authorities also experienced faster general construction-sector firm formation after 2017. This likely reflects the fact that high-flat areas are predominantly urban, and urban areas experienced stronger post-2017 business formation broadly. This finding motivates the triple-difference specification in \Cref{tab:robustness}, which nets out any common trends between fire safety and control construction SICs within the same local authorities. The triple-difference coefficient ($\beta = 1.91$, $p = 0.029$) isolates the fire-safety-specific effect: even after accounting for general urban construction entry, local authorities with higher flat shares experienced significantly more fire safety firm formation.

\input{tables/tab2_main}

\subsection{Dynamic Effects and Regulatory Phases}

\Cref{tab:phases} decomposes the post-Grenfell effect into four regulatory phases. The results reveal a striking pattern of intensification. The immediate post-fire period (Phase 1, July--December 2017) shows a modest effect, consistent with an initial demand signal. The Hackitt Review period (Phase 2, 2018--2019) shows a larger effect as the regulatory framework began to take shape and the EWS1 requirement created urgent assessment demand. The effect continues to grow through the EWS1/Fire Safety Act period (Phase 3, 2020--2021) and reaches its peak after the Building Safety Act (Phase 4, 2022--2024), which created ongoing compliance obligations and a new class of regulated professional.

This escalating pattern is inconsistent with a one-time Grenfell ``shock'' story and strongly consistent with demand-driven industry creation: each regulatory milestone expanded the set of buildings requiring professional assessment, the range of compliance services demanded, and the permanence of the regulatory obligation.

\input{tables/tab3_phases}

\subsection{Robustness}

\Cref{tab:robustness} reports robustness checks. The result is robust to: (i) binary treatment comparing top-quartile to bottom-quartile flat share local authorities ($\beta = 2.02$, $p < 0.001$); (ii) the triple-difference specification ($\beta = 1.91$, $p = 0.029$); (iii) excluding London boroughs ($\beta = 1.87$, $p = 0.033$); and (iv) quarterly aggregation ($\beta = 20.6$, $p < 0.001$). The wild cluster bootstrap confirms the baseline result ($p < 0.001$, 95\% CI $[2.33, 12.05]$). The fact that the point estimate is attenuated but remains significant when London is excluded confirms that the result is not driven solely by the capital's unique housing market.

\input{tables/tab4_robustness}


\section{Discussion}

The Grenfell Tower fire killed 72 people and triggered a regulatory cascade that, among its many consequences, created an entirely new compliance-service industry. This paper documents the birth of that industry using administrative firm registration data and shows that its geographic formation followed the spatial distribution of the building stock most affected by the new regulations.

\subsubsection*{Compliance markets as a hidden cost of regulation} Standard cost-benefit analysis of safety regulation typically accounts for direct compliance costs (testing, remediation, administrative burden) and benefits (reduced risk, avoided harm). The creation of a compliance-service industry is a distinct channel: it redirects entrepreneurial effort and professional talent toward regulatory certification rather than productive activity \citep{greenstone_2002, greenstone_list_syverson_2012}. Whether this reallocation represents a net cost depends on whether the regulatory regime produces genuine safety improvements or primarily generates compliance paperwork---a question this paper cannot answer but that the Grenfell case makes urgent.

\subsubsection*{The speed of market creation} The escalating pattern of firm formation across regulatory phases suggests that compliance markets do not form instantly. The binding constraint on resolving the post-Grenfell housing freeze was not the identification of dangerous buildings---that happened within months---but the supply of qualified professionals to assess and certify them. This supply-side bottleneck has implications for regulatory design: when regulation creates demand for services that do not yet exist, the economic disruption persists until the market fills the gap. Phased regulatory rollouts may allow supply to develop incrementally rather than creating a sudden, acute shortage \citep{coglianese_lazer_2003, viscusi_2006}.

\subsubsection*{Limitations} The SIC code classification is imperfect: some firms coded as ``technical testing'' (71200) or ``other specialised construction'' (43999) may not be fire safety firms, and some fire safety firms may be coded elsewhere. The broad codes add noise but should not bias the estimate, since non-fire firms are unlikely to correlate with flat share conditional on fixed effects. Future work could refine the outcome using text analysis of company names for fire-safety keywords. The treatment intensity measure (flat share) is a proxy for exposure to the cladding crisis; the ideal measure would be the count of buildings above 18 metres with dangerous cladding, available from DLUHC remediation data at a coarser geographic level. Flat share captures multi-unit buildings broadly, not just the high-rise stock that is the primary regulatory target. The Companies House registered office address reflects where a firm is administratively based, not necessarily where it delivers services; to the extent that fire safety consultants register in urban centres regardless of client location, the treatment effect captures a ``service-hub'' response alongside localized demand. Finally, the paper identifies firm \textit{formation} but cannot track firm survival, employment, or revenue---the compliance industry may include many short-lived or single-person consultancies rather than substantive enterprises.


\section{Conclusion}

Regulations do not merely constrain markets; they create them. The post-Grenfell regulatory cascade generated a fire safety compliance industry from near-zero, and the geographic pattern of that industry's formation followed the local demand created by the building stock. This ``regulatory genesis'' is a general phenomenon: any safety regulation that requires professional certification, testing, or assessment creates demand for a compliance-service market. The speed at which that market forms---and the economic disruption that persists until it does---is a first-order cost of post-disaster regulation that standard analysis overlooks.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Companies House data extraction} The Companies House Free Company Data Product is a monthly bulk CSV snapshot containing all registered UK companies with their SIC codes, incorporation dates, company status, and registered office postcodes. I downloaded the snapshot dated closest to March 2025. From 5.6 million total companies, I extracted 275,439 companies in the nine target SIC codes. I geocoded each company's registered office postcode to a local authority using the postcodes.io bulk lookup API.

\subsubsection*{SIC code definitions} Fire safety SIC codes: 84250 (Fire service activities), 71200 (Technical testing and analysis), 80200 (Security systems service activities), 71121 (Engineering activities relating to building projects), 71129 (Other engineering activities), 43999 (Other specialised construction activities not elsewhere classified). Control construction SIC codes: 43210 (Electrical installation), 43220 (Plumbing, heat and air-conditioning installation), 43290 (Other construction installation).

\subsubsection*{Dwelling stock} Flat share by local authority is computed from Census dwelling-type data accessed via the NOMIS API. I calculate the ratio of households in flats, maisonettes, or apartments to total households in each local authority district.


\section{Robustness Appendix}

\subsubsection*{Placebo date tests} I re-estimate the main specification using placebo Grenfell dates of June 2013 and June 2015, restricting the sample to a symmetric window around each placebo date. Both placebo coefficients are positive ($\beta \approx 1.3$) but substantially smaller than the post-Grenfell main effect ($\beta = 6.88$), indicating that while high-flat urban areas had modestly faster baseline firm formation, the Grenfell regulatory cascade produced an effect several times larger than any pre-existing differential trend. The triple-difference specification, which nets out these common urban trends, remains the cleanest identification.

\subsubsection*{Wild cluster bootstrap} To assess inference robustness given the large number of clusters, I compute wild cluster bootstrap p-values following the Webb six-point distribution. Results confirm the statistical significance of the main finding.

\subsubsection*{Poisson and log specifications} The count-data nature of the outcome (monthly incorporations, many zeros) motivates Poisson and log(Y+1) specifications. Both confirm the baseline OLS result, with the Poisson model providing a semi-elasticity interpretation.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
