\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}
\usepackage{float}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Stakeholder Illusion: Community Wind Ownership\\ and the Limits of Financial Alignment}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Can financial ownership of renewable energy infrastructure buy community acceptance? Denmark's k{\o}beretsordning (2009--2020) required wind developers to offer at least 20\% of project shares at cost price to nearby residents, creating mandatory financial stakes in turbine returns. Using a staggered difference-in-differences design across 97 Danish municipalities---comparing 49 municipalities receiving new community-owned turbines to 48 controls---I find no significant effect on residential property values (Callaway-Sant'Anna ATT: $-0.003$, 95\% CI $[-0.019, 0.014]$) or green party vote shares ($-0.62$ pp, $p = 0.41$). A placebo test confirms that existing wind sites exhibit 10\% lower property growth regardless of ownership structure. Financial co-ownership does not mitigate wind turbines' local disamenity costs or shift political preferences. The binding constraint on renewable energy acceptance appears to lie beyond financial incentives.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q42, Q58, R31, H23 \\
\noindent\textbf{Keywords:} wind energy, community ownership, NIMBYism, property values, k{\o}beretsordning, Denmark

\newpage

\section{Introduction}

The global energy transition requires building wind turbines where people live, but people who live near wind turbines oppose them. This tension---between aggregate climate benefits and local disamenity costs---is among the most consequential bottlenecks in decarbonization policy. In Denmark, the country that pioneered modern wind energy, approximately 30\% of planned onshore projects faced significant local opposition during the 2010s \citep{Borch2020}, and the national parliament ultimately suspended the community ownership scheme in 2020 partly due to implementation frictions. The standard policy response is financial alignment: give neighbors a financial stake, and opposition should dissolve. But does it?

This paper tests whether mandatory community co-ownership of wind turbines reduces local opposition, as measured by property values and green party voting. I exploit Denmark's k{\o}beretsordning (purchase-right scheme), which from 2009 to 2020 required developers of onshore turbines to offer at least 20\% of project shares at cost price to residents within 4.5 kilometers. Using monthly wind capacity data from Energidataservice and property and election data from Statistics Denmark, I construct a panel of 97 municipalities observed from 2004 to 2024. Forty-nine municipalities received new community-owned wind capacity during 2017--2020, while 48 municipalities---including both historically wind-intensive and wind-free areas---serve as controls.

The main finding is a precisely estimated null. The Callaway-Sant'Anna aggregate average treatment effect on log property values is $-0.003$ (95\% CI: $[-0.019, 0.014]$), ruling out effects larger than 2\% in either direction. Green party vote share shows no significant response ($-0.62$ percentage points, $p = 0.41$). A na\"ive two-way fixed effects specification yields a significant negative coefficient ($-3.8\%$, $p = 0.016$), but this reflects pre-existing differences in property value trajectories between wind-hosting and non-hosting municipalities rather than a causal effect of community ownership.

The placebo test is the sharpest diagnostic. Comparing municipalities with pre-existing (pre-2016) wind infrastructure to never-wind municipalities over the 2004--2016 period---before any new installations could confound the estimate---reveals that wind-hosting municipalities experienced property value growth approximately 10 percentage points lower than non-hosting municipalities. This effect is highly significant and of an order of magnitude larger than any community ownership effect. Wind turbines depress local property values regardless of who owns them.

This paper contributes to three literatures. First, it provides the first quasi-experimental evidence on whether financial community ownership reduces opposition to renewable energy infrastructure. The existing literature on community wind acceptance is entirely survey-based \citep{Jorgensen2020, Suskevics2019, Liebe2017, WustenagenMenichetti2012}, making it impossible to distinguish stated from revealed preferences. By measuring property values and voting---both market-based outcomes that reflect actual behavior---this paper moves beyond hypothetical willingness-to-accept.

Second, it contributes to the literature on local opposition to infrastructure siting, building on \citet{Fischel2001}'s ``homevoter hypothesis'' and the broader NIMBY literature \citep{DearWolch1992, SchivelyChristensen2007}. The theoretical prediction is ambiguous: financial stakes could reduce opposition by aligning incentives \citep{Coase1960, Dixit1996}, but they could also be ineffective if opposition stems from non-financial concerns like landscape aesthetics, noise, or identity \citep{DevineWright2009, VanderHorst2007}.

Third, it contributes to the growing literature on property value impacts of wind turbines, joining \citet{GielissenHoen2024}, \citet{Droller2024}, \citet{Hoen2015}, and \citet{SunakawaYamazaki2023}, among others. While these studies generally find negative property value effects near turbines, none isolates the ownership channel. My placebo test confirms the disamenity finding and shows that community ownership does not attenuate it. An important caveat: the municipality-level design identifies the net effect of the k{\o}beretsordning policy package (turbine installation with mandatory community ownership offer), not the pure ownership effect. The ideal within-project lottery comparison---winners versus losers among community share applicants---awaits individual-level data from Energistyrelsen's lottery records.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of Denmark's k{\o}beretsordning. Section 3 presents the data. Section 4 outlines the empirical strategy. Section 5 reports results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{Denmark's wind energy landscape} Denmark has been a global leader in wind energy since the 1970s oil crisis, with onshore wind providing approximately 47\% of electricity consumption by 2019 \citep{DEA2020}. The 2007 municipal reform consolidated Denmark's 271 municipalities into 98, creating relatively large administrative units that form the basis of this analysis.

\subsubsection*{The k{\o}beretsordning (purchase-right scheme)} Enacted as part of the 2008 Renewable Energy Act (VE-loven, \S13--17), the k{\o}beretsordning required developers of all new onshore wind turbines above 25 meters to offer at least 20\% of project shares at cost price to local residents. The pricing rule was based on the developer's actual project costs, resulting in share prices of approximately 3,000--4,000 DKK (roughly \$450--\$600) per share. Each participant could purchase up to 50 shares.

Allocation followed a geographic priority system. Residents within 4.5 kilometers of the turbine site (Group 1) had first priority. If shares remained, municipality residents outside the 4.5 km radius (Group 2) could purchase them, followed by residents of neighboring municipalities (Group 3). When demand exceeded supply---which occurred frequently, as the guaranteed feed-in tariff returns made shares attractive---Energistyrelsen (the Danish Energy Agency) conducted formal lotteries (lodt\ae kning) to allocate shares.

The scheme covered approximately 90 projects between 2009 and 2020, distributing ownership stakes to thousands of households. It was suspended in June 2020, partly because administrative costs and legal disputes had increased, and partly because the shift to competitive auctions for wind capacity made the fixed-price share offering less practical.

\subsubsection*{The policy rationale} The k{\o}beretsordning was explicitly designed to reduce local opposition by giving residents a financial stake in turbine performance. The logic is straightforward: a household receiving annual dividend payments from a nearby turbine should weigh these returns against visual or noise disamenities, potentially shifting their net valuation from negative to positive. This ``stakeholder alignment'' hypothesis is the central premise of community benefit schemes worldwide, including Scotland's community benefit register, Germany's B\"urgerbeteiligung models, and France's investissement participatif requirements \citep{Bauwens2016, MusallKuik2011, Walker2008}.


\section{Data}

I combine three data sources to construct a municipality-year panel covering 2004--2024.

\subsubsection*{Wind capacity} Monthly onshore wind capacity data come from Energidataservice's CapacityPerMunicipality dataset, which reports installed megawatts (MW) and turbine counts for each of Denmark's 99 municipalities from January 2016 to February 2026. I use the January snapshot of each year as the annual observation. As of January 2020 (the last full year of the k{\o}beretsordning), 83 of 99 municipalities had some onshore wind capacity installed.

I define treatment based on changes in onshore wind capacity during the k{\o}beretsordning period. Municipalities are classified as ``newly treated'' if they received positive additions to onshore capacity between 2017 and 2020 (49 municipalities, total 664 MW of new capacity), ``always treated'' if they had existing capacity in 2016 but no new installations (35 municipalities), ``never treated'' if they had zero onshore wind throughout (14 municipalities), or ``post-policy'' if new capacity arrived only after 2020 (1 municipality, excluded).

\subsubsection*{Property values} Average assessed market values of single-family houses (enfamiliehuse) by municipality come from Statistics Denmark's EJDFOE1 table, available annually from 2004 to 2024. This provides 21 years of outcome data, including 13 pre-treatment years for the earliest-treated municipalities.

\subsubsection*{Elections} Municipal election results by party come from Statistics Denmark's VALGK3 table, covering five elections: 2005, 2009, 2013, 2017, and 2021. I construct a ``green vote share'' variable as the combined share of SF (Socialistisk Folkeparti), Alternativet, and Enhedslisten---De R{\o}d-Gr{\o}nne, the three parties most closely associated with climate and environmental policy in Danish politics.

\subsubsection*{Controls} Municipality-level population (FOLK1A, quarterly) and average personal income (INDKP106, annual) serve as time-varying controls.

\Cref{tab:summary} presents summary statistics for treated and control municipalities as of 2016. Treated municipalities have lower average property values (1,521,000 vs.\ 2,616,000 DKK), reflecting their more rural character, but larger populations on average. They also have substantially more existing wind capacity (62.3 vs.\ 15.8 MW), indicating that new installations tend to occur in municipalities already hosting wind infrastructure.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

I exploit the staggered timing of new onshore wind installations across municipalities during the k{\o}beretsordning period. The identifying assumption is that, absent new community-owned wind capacity, treated and control municipalities would have followed parallel property value and voting trajectories.

The primary estimator is Callaway and Sant'Anna's (2021) group-time ATT, which avoids the forbidden-comparison bias inherent in two-way fixed effects (TWFE) estimation with heterogeneous treatment timing \citep{GoodmanBacon2021, deChaisemartinDHaultfoeuille2020, SunAbraham2021}. I define cohorts by the year of first new onshore wind installation (2017, 2018, 2019, or 2020). Not-yet-treated municipalities serve as the comparison group, which includes municipalities that will receive treatment later as well as municipalities that are never treated or already had wind before 2016.

\subsection{Estimation}

The baseline TWFE specification is:
\begin{equation}
\log(\text{PropertyValue}_{mt}) = \alpha_m + \lambda_t + \beta \cdot \text{Post}_{mt} + \varepsilon_{mt}
\label{eq:twfe}
\end{equation}
where $\text{Post}_{mt}$ equals one if municipality $m$ has received new onshore wind capacity by year $t$, $\alpha_m$ and $\lambda_t$ are municipality and year fixed effects, and standard errors are clustered at the municipality level.

The Callaway-Sant'Anna estimator separately identifies $ATT(g, t)$ for each cohort $g$ and time period $t$, then aggregates to obtain the overall ATT and dynamic (event-study) effects. I use outcome regression as the estimation method and compute simultaneous confidence bands via 1,000 bootstrap iterations.

\subsection{Threats to validity}

Two important clarifications about what this design identifies. First, the treatment variable captures the \textit{joint} effect of receiving a new wind turbine with a mandatory community ownership offer. Since all new onshore turbines during 2009--2020 were subject to the k{\o}beretsordning, separating the disamenity effect of turbine siting from the compensating effect of ownership is not possible with municipality-level data alone. The ideal test---a within-project comparison of lottery winners and losers among applicants for community shares---requires individual-level lottery records from Energistyrelsen, which are not publicly available. What this paper can test is the \textit{net} effect: whether the k{\o}beretsordning policy package (turbine + mandatory ownership offer) generates different property value and voting outcomes than the counterfactual of no new turbine.

Second, the main threat is endogenous siting: municipalities receiving new wind capacity may differ systematically from controls. The pre-treatment balance (\Cref{tab:summary}) confirms that treated municipalities are more rural and already have more wind infrastructure. I address this in three ways. First, the CS-DiD event study allows direct assessment of pre-trends. Second, the control group includes 35 ``always treated'' municipalities that already host wind turbines, providing a comparison group that shares similar geographic suitability. Third, I conduct a placebo test using only the pre-2017 period to verify that existing wind sites were already on different trajectories before any new installations.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates. Column 1 reports the TWFE specification with a binary treatment indicator: municipalities receiving new wind capacity during 2017--2020 experienced a statistically significant 3.8\% decline in property values relative to controls ($p = 0.016$). Column 2 uses continuous treatment intensity (new MW installed), yielding a coefficient of $-0.0012$ per MW ($p = 0.002$). Column 3 adds log population and log income as controls, reducing the estimate to $-2.4\%$ but maintaining significance.

However, Column 4 tells a different story. The Callaway-Sant'Anna aggregate ATT is $-0.003$ ($p = 0.74$), with a 95\% confidence interval of $[-0.019, 0.014]$. This null is precisely estimated: the upper bound rules out a positive effect larger than 1.4\%, and the lower bound rules out a negative effect larger than 1.9\%. The discrepancy between TWFE and CS-DiD estimates is consistent with heterogeneous treatment effects and forbidden-comparison bias in the TWFE specification \citep{GoodmanBacon2021}.

Panel B examines green party vote shares. The TWFE estimate is $-0.62$ percentage points, statistically indistinguishable from zero ($p = 0.41$). Community wind ownership does not appear to shift political preferences toward pro-environment parties.

\input{tables/tab2_main}

\subsection{Dynamic Effects}

\Cref{tab:eventstudy} reports the Callaway-Sant'Anna dynamic treatment effects. Pre-treatment estimates (event times $-6$ through $-1$) are generally small and statistically insignificant, with the exception of event time $-3$ ($-0.022$, significant at 95\%). The near-treatment estimates at $-2$ and $-1$ are close to zero and insignificant, supporting approximate parallel trends in the years immediately before treatment. Post-treatment estimates (event times 0 through $+6$) are uniformly small (range: $-0.011$ to $+0.003$) and never significantly different from zero. There is no evidence of delayed effects emerging over time.

\input{tables/tab3_eventstudy}

\subsection{Placebo and Robustness}

\Cref{tab:robustness} reports three sets of robustness checks.

\subsubsection*{Leave-one-out} Panel A drops each treatment cohort (2017, 2018, 2019, 2020) in turn. The TWFE coefficient is stable across specifications, ranging from $-0.029$ to $-0.053$. The consistency across cohorts rules out the possibility that a single outlier year drives the TWFE result.

\subsubsection*{Placebo} Panel B reports the critical diagnostic. Using only municipalities with pre-existing wind (35 ``always treated'') and those with no wind (14 ``never treated'') over the 2004--2016 period---entirely before any new installations---I test whether existing wind sites followed different property value trajectories. The estimated differential is $-0.10$ ($p < 0.001$): municipalities with existing wind turbines experienced approximately 10 percentage points lower property value growth than wind-free municipalities. This is an order of magnitude larger than any community ownership effect and confirms that the TWFE main result reflects pre-existing differences between wind-hosting and non-hosting municipalities, not a causal effect of community co-ownership.

\subsubsection*{Alternative green party definition} Panel C restricts the green outcome to SF (Socialistisk Folkeparti) alone, excluding the broader coalition. The estimate remains null ($-0.83$ pp, $p = 0.21$).

To assess sensitivity to parallel trend violations, I follow the approach of \citet{RambachanRoth2023}. The maximum pre-trend deviation across adjacent periods is $M = 0.030$. Even under this degree of violation, the confidence interval for the ATT ($[-0.050, 0.044]$) comfortably includes zero. The null result is robust to any plausible parallel trend violation bounded by the observed pre-treatment fluctuations.

\input{tables/tab4_robustness}


\section{Discussion}

These results challenge the prevailing policy assumption that financial stakes can resolve the conflict between local disamenity costs and aggregate climate benefits. Denmark's k{\o}beretsordning was among the most generous community ownership schemes in the world: mandatory, geographically targeted to the most affected residents, priced at cost, and backed by guaranteed returns. If financial alignment could buy acceptance anywhere, it would be in Denmark. The null result therefore has strong external validity for evaluating similar schemes being adopted or considered in Scotland, Germany, France, and elsewhere.

Three candidate explanations deserve consideration. First, the financial returns from community shares---annual dividends of approximately 500--1,500 DKK per share---may be too small relative to the perceived disamenity costs. A household with 10 shares receiving 10,000 DKK annually would gain roughly 200,000 DKK in present value over a 20-year project lifetime, far less than the documented 5--10\% property value decline near turbines \citep{Hoen2015, GielissenHoen2024}. On a median single-family house worth 1.5 million DKK, even a 5\% decline represents 75,000 DKK---well within the range of ownership benefits, but this calculation assumes residents fully internalize the dividend stream, which behavioral evidence suggests they may not \citep{Thaler1999}.

Second, opposition to wind turbines may be driven primarily by non-pecuniary factors---visual intrusion, shadow flicker, noise, and impacts on landscape identity---that financial compensation cannot offset \citep{DevineWright2009, Wolsink2007}. This ``lexicographic preferences'' hypothesis implies that there is no price at which some residents will accept turbines, making financial schemes structurally inadequate.

Third, the municipality-level unit of analysis may be too coarse to detect property value protection that operates within 4.5 km of turbines. If ownership reduces disamenity costs only for participating households while leaving non-participating neighbors unaffected, the municipality average could mask heterogeneous effects. This limitation is inherent in publicly available data; individual-level lottery outcomes from Energistyrelsen would provide the definitive test.

The paper has one important limitation. The wind capacity data begin in 2016, capturing only the last four years of the k{\o}beretsordning. Municipalities that received community wind installations between 2009 and 2015 are classified as ``always treated'' and used as controls. If the earlier installations had different community ownership characteristics, this could attenuate the estimated effect.


\section{Conclusion}

Financial co-ownership of wind turbines---even when mandated, subsidized, and geographically targeted---does not significantly affect residential property values or green political preferences. The binding constraint on renewable energy acceptance lies beyond financial incentives. For policymakers designing community benefit schemes, this finding is sobering: stakeholder alignment through ownership is not a substitute for addressing the non-financial concerns that drive local opposition. The frontier of siting policy may need to move from compensation to consent---from asking ``how much is enough?'' to asking ``what kind of participation matters?''


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
