\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Sovereign Stigma Without Price Stigma: FATF Grey-Listing and the Cost of Sending Remittances}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Placement on the FATF grey list is widely believed to raise remittance costs by triggering de-risking of correspondent banking relationships. I test this claim using corridor-level data from the World Bank Remittance Prices Worldwide database, covering 378 corridors across 111 receiving countries from 2011 to 2025. Exploiting staggered grey-listing across 24 treatment cohorts in a Callaway-Sant'Anna difference-in-differences framework, I find no statistically significant effect on remittance costs (ATT = 0.16 percentage points, SE = 0.29). The null holds for banks and money transfer operators separately, for the exchange rate margin, and for provider counts on the extensive margin. These results challenge the prevailing narrative that FATF grey-listing imposes a regressive tax on diaspora families through higher transfer costs.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F24, G28, O16, F38 \\
\noindent\textbf{Keywords:} remittances, FATF, grey-listing, de-risking, financial regulation, anti-money laundering

\newpage

\section{Introduction}

Every year, migrants send over \$650 billion to families in low- and middle-income countries, often paying 5--10\% of the amount in fees and exchange rate markups \citep{worldbank2024}. For the one billion people in developing countries who receive these transfers, the cost of remittances functions as a tax on consumption, education, and healthcare. The UN Sustainable Development Goals target a maximum cost of 3\% by 2030, yet 56 quarters into the SDG era, the global average remains stuck near 6.4\%.

A common explanation for persistently high costs is ``de-risking'' --- the withdrawal of global banks from correspondent relationships with financial institutions in countries perceived as high-risk for money laundering. The FATF grey list, formally titled ``Jurisdictions under Increased Monitoring,'' publicly identifies countries with strategic deficiencies in their anti-money-laundering (AML) and counter-terrorism financing (CTF) frameworks. Since 2010, 114 countries have been placed on this list, and the claim that grey-listing raises remittance costs has been central to policy debates about FATF reform \citep{cgd2016, imf2021wps, fsb2018}.

This paper tests that claim directly. I use the World Bank's Remittance Prices Worldwide (RPW) database --- the most comprehensive public source of corridor-level remittance pricing --- to construct a panel of 378 sending-receiving country corridors observed quarterly from 2011 to 2025. I merge this with the complete history of FATF grey-list placements and removals, creating a dataset of 16,849 corridor-quarter observations. The staggered timing of grey-listing across 24 treatment cohorts provides rich variation for identification.

I estimate the causal effect of grey-listing using the \citet{callaway2021} staggered difference-in-differences estimator, which is robust to treatment-effect heterogeneity that can bias conventional two-way fixed effects (TWFE) estimates in staggered settings \citep{goodmanbacon2021, dechaisemartin2020}. My preferred specification compares corridors whose receiving countries enter the grey list to corridors whose receiving countries are never listed, using a doubly robust estimator with corridor and time fixed effects.

The main finding is a null: FATF grey-listing does not significantly raise remittance costs. The aggregate average treatment effect on the treated (ATT) is 0.16 percentage points with a standard error of 0.29, and the 95\% confidence interval of $[-0.40, 0.72]$ rules out economically large effects. An event study with eight pre-treatment quarters shows no evidence of differential pre-trends. The result is robust to using not-yet-treated corridors as controls (ATT = 0.20, SE = 0.29), conventional TWFE with source-country$\times$quarter fixed effects ($\hat{\beta} = -0.14$, SE = 0.13), and restricting the sample to bank or money-transfer-operator (MTO) channels separately.

Three additional tests strengthen the null. First, a sending-country placebo --- testing whether grey-listing the \emph{sending} country affects costs --- yields a precisely estimated zero (0.12, SE = 0.59), confirming that the identification strategy does not capture spurious sender-side shocks. Second, grey-listing has no detectable effect on the number of remittance providers in a corridor ($-0.01$, SE = 0.02 in logs), challenging the ``provider exit'' mechanism that is supposed to drive higher prices. Third, delisting --- removal from the grey list --- also has no effect on costs (0.004, SE = 0.18), consistent with the absence of an entry effect.

These results contribute to an active debate about whether and how AML regulation affects financial inclusion. \citet{cgd2016} used proprietary SWIFT data to document that countries experiencing de-risking saw declines in correspondent banking relationships, but could not observe consumer prices. \citet{imf2021wps} estimated aggregate capital flow effects of grey-listing using LASSO methods, but worked at the country level without corridor-specific pricing. \citet{fsb2018} surveyed banks about de-risking but acknowledged that the consumer price implications were unknown. \citet{erbenova2016} documented the withdrawal of correspondent banking relationships as a systemic risk but did not trace the retail price implications. My contribution is to bring the first corridor-level, publicly replicable test with modern heterogeneity-robust methods. The answer is that the \emph{retail price} channel --- the most politically salient argument for FATF reform --- does not operate as assumed. This does not rule out wholesale financial disruption: grey-listing may reduce correspondent banking claims even as consumer remittance prices remain stable, a distinction that matters for calibrating policy responses.

Why might grey-listing fail to raise remittance costs? I offer three candidate explanations. First, the remittance market is dominated by MTOs like Western Union and MoneyGram, which use their own proprietary networks rather than the correspondent banking infrastructure that FATF grey-listing most directly disrupts \citep{worldbank2024}. Second, mobile money platforms --- which have expanded rapidly since 2015 --- provide an alternative channel that is largely insulated from bank-level de-risking \citep{gsma2023}. Third, remittance pricing is highly competitive and transparent, with the World Bank database itself serving as a comparison-shopping tool, which may limit firms' ability to pass through compliance costs \citep{beck2018}.

The paper is organized as follows. Section 2 describes the FATF institutional setting. Section 3 presents the data. Section 4 lays out the empirical strategy. Section 5 reports the results, including mechanism tests and robustness checks. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{The FATF grey list} The Financial Action Task Force (FATF), an intergovernmental body headquartered at the OECD in Paris, sets global standards for anti-money-laundering and counter-terrorism financing. It evaluates member and non-member countries through a mutual evaluation process and, since 2010, has published a list of ``Jurisdictions under Increased Monitoring'' --- colloquially known as the grey list. Countries are added to the grey list when a mutual evaluation or follow-up review identifies strategic deficiencies in their AML/CTF framework that they have committed to addressing. Placement is decided at FATF plenaries held three times per year, typically in February, June, and October, creating sharp treatment timing.

\subsubsection*{De-risking and remittance costs} Grey-listing is not itself a legal prohibition on financial transactions. Rather, it signals to global banks that enhanced due diligence is warranted when transacting with institutions in listed countries. In practice, many international banks respond by terminating correspondent banking relationships altogether --- a practice known as ``de-risking'' \citep{fsb2018}. The hypothesized transmission to remittance costs runs through two channels: (i) reduced competition as banks exit corridors to grey-listed countries, enabling remaining providers to charge higher markups; and (ii) higher compliance costs for remaining providers, passed through to consumers \citep{cgd2016}.

\subsubsection*{Scale and scope} Between 2010 and 2025, 114 countries have been placed on the FATF grey list, with 86 subsequently removed. The median grey-listing episode lasts approximately 2.5 years. The geographical coverage is broad: listed countries span Sub-Saharan Africa (e.g., Nigeria, South Africa, Tanzania), South Asia (Pakistan, Bangladesh, Sri Lanka), Southeast Asia (Cambodia, Myanmar, Philippines), the Caribbean (Jamaica, Barbados, Trinidad and Tobago), and Eastern Europe (Albania, Croatia, Bosnia and Herzegovina). This diversity provides variation across income levels, remittance corridor characteristics, and financial market structures.


\section{Data}

\subsubsection*{Remittance prices} The primary outcome data come from the World Bank's Remittance Prices Worldwide (RPW) database, which collects quarterly price quotes from remittance service providers in major corridors worldwide \citep{worldbank2024}. The RPW records the total cost of sending a fixed amount (USD 200) through each provider in each corridor, decomposed into transfer fees and exchange rate margins. I aggregate the 247,490 firm-level observations to the corridor-quarter level, computing the average total cost across all providers in each corridor. The resulting panel covers 378 unique corridors (49 sending countries $\times$ 111 receiving countries) observed over 53 quarters from 2011Q1 to 2025Q1.

\subsubsection*{FATF grey-list status} I compile the complete history of grey-list placements and removals from FATF plenary statements, available at \texttt{fatf-gafi.org}. This yields 74 distinct grey-listing episodes across 63 countries. For each receiving country in each quarter, I construct a binary indicator equal to one if the country is on the grey list. Of the 16,849 corridor-quarter observations, 2,286 (13.6\%) have a grey-listed receiving country.

\subsubsection*{Remittance volumes} To assess remittance dependence, I use the World Development Indicators (WDI) series on personal remittances received as a share of GDP. This allows me to split the sample into high- and low-remittance-dependence destinations.

\subsection{Summary Statistics}

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

I exploit the staggered timing of FATF grey-list placements across receiving countries to identify the effect on remittance costs. The key identification assumption is that, conditional on corridor and time fixed effects, remittance cost trends in corridors to grey-listed countries would have paralleled those in corridors to non-listed countries absent the grey-listing. This assumption is supported by three features of the setting.

First, FATF grey-listing decisions are driven by the internal mutual evaluation cycle and reflect technical compliance assessments, not the economic conditions of the receiving country. The timing of listing is determined by when a country's evaluation is scheduled and when deficiencies are identified, creating plausibly exogenous variation in treatment timing \citep{imf2021wps}.

Second, the event study in \Cref{tab:eventstudy} shows no individually significant pre-treatment coefficients across eight pre-treatment quarters, consistent with parallel trends in the pre-period.

Third, a sending-country placebo confirms that the estimates are not driven by secular trends in corridors involving countries that are generally subject to international scrutiny. Grey-listing of the sending country has no effect on remittance costs (0.12, SE = 0.59; \Cref{tab:robust}, Panel A).

\subsection{Estimation}

My preferred estimator is the \citet{callaway2021} staggered difference-in-differences, which computes group-time average treatment effects $\text{ATT}(g,t)$ for each treatment cohort $g$ and calendar period $t$, then aggregates to an overall ATT. I use doubly robust estimation with the inverse probability weighting estimator of \citet{santanna2020}, which combines outcome regression and propensity score weighting to achieve consistency if either model is correctly specified. The control group consists of never-listed corridors. I also report results using not-yet-treated corridors as controls and conventional TWFE specifications as robustness checks.

For the dynamic specification, I aggregate to event time using:
\begin{equation}
\theta_\ell = \sum_{g} w_g \cdot \text{ATT}(g, g + \ell)
\end{equation}
where $\ell$ is event time (quarters since grey-listing), and $w_g$ are cohort-size weights.

Standard errors for the CS-DiD are computed analytically via the multiplier bootstrap at the corridor level. For TWFE regressions, I cluster standard errors at the receiving-country level, the unit at which treatment is assigned.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates. The Callaway-Sant'Anna ATT using never-treated controls (column 1) is 0.16 percentage points with a standard error of 0.29 --- statistically indistinguishable from zero. The 95\% confidence interval of $[-0.40, 0.72]$ rules out effects larger than 0.72 percentage points, which is economically meaningful: on a base cost of 7.35\%, even the upper bound of the confidence interval would represent a less than 10\% increase. Using not-yet-treated corridors as the control group (column 2) yields a similar estimate of 0.20 (SE = 0.29). The conventional TWFE specifications in columns 3--4 produce point estimates that are slightly negative ($-0.05$ and $-0.14$), further supporting the null.

\input{tables/tab2_main}

\subsubsection*{Event study} \Cref{tab:eventstudy} reports the dynamic treatment effects over eight pre-treatment and twelve post-treatment quarters. None of the eight pre-treatment coefficients are individually significant, supporting the parallel trends assumption. The post-treatment coefficients are uniformly small and statistically insignificant through three years after listing. There is a slight drift toward negative point estimates at longer event times, but these are imprecisely estimated and inconsistent with the hypothesis that grey-listing raises costs.

\input{tables/tab3_eventstudy}

\subsection{Mechanisms}

\subsubsection*{Channel heterogeneity} If de-risking operates through the correspondent banking system, the price effect should be concentrated among bank-originated remittances, with MTOs --- which operate proprietary payment networks --- relatively insulated. \Cref{tab:robust}, Panel B tests this by estimating the TWFE specification separately for bank and MTO corridor-quarters. Neither channel shows a significant effect: banks at $-0.42$ (SE = 0.31) and MTOs at $-0.17$ (SE = 0.16). The bank coefficient is directionally negative and the largest in absolute terms, but still statistically insignificant.

\subsubsection*{Extensive margin} The de-risking narrative predicts that grey-listing should reduce the number of providers in a corridor, particularly banks. Panel C of \Cref{tab:robust} shows that grey-listing has no detectable effect on the overall provider count ($-0.01$ log points, SE = 0.02), bank provider count ($-0.08$, SE = 0.10), or MTO provider count ($-0.33$, SE = 0.57). The absence of provider exit is consistent with the absence of a price effect: if providers are not leaving, there is no reduction in competition to drive up costs.

\subsection{Robustness}

\input{tables/tab4_robustness}

\subsubsection*{Sending-country placebo} If the empirical strategy were capturing general trends in corridors involving countries subject to international financial scrutiny, we would expect grey-listing of the sending country to also affect costs. The sending-country placebo in Panel A yields a point estimate of 0.12 (SE = 0.59, $p = 0.84$), confirming that the design isolates receiving-country shocks.

\subsubsection*{FX margin} The exchange rate margin --- which could be more directly affected by changes in foreign exchange access than transfer fees --- also shows no significant response to grey-listing ($-0.16$, SE = 0.12, $p = 0.19$). If anything, the point estimate is negative, further inconsistent with the hypothesis.

\subsubsection*{Entry vs. exit asymmetry} If grey-listing creates persistent damage to financial connectivity through sunk relationship costs (the ``sovereign stigma'' hypothesis), delisting should have no effect on costs even as formal status improves. Consistent with the absence of an entry effect, I find no delisting effect: the coefficient on post-exit is 0.004 (SE = 0.18) using 48 delisting events.


\section{Discussion}

The null result I document contradicts a widely held belief in the development finance community. Three candidate mechanisms may explain why FATF grey-listing does not translate into higher remittance costs.

First, the remittance market has become less dependent on correspondent banking. MTOs account for roughly 80\% of the firm-level price quotes in the RPW database and increasingly operate through proprietary settlement networks or partnerships that bypass the traditional correspondent banking infrastructure \citep{worldbank2024}. FATF grey-listing may disrupt bank-to-bank relationships while leaving the primary remittance channels untouched.

Second, the rapid expansion of mobile money --- particularly in Sub-Saharan Africa and South Asia, regions disproportionately affected by grey-listing --- has created alternative payment rails that are structurally insulated from correspondent banking de-risking \citep{gsma2023}. Between 2011 and 2025, mobile money transactions grew from near-zero to over \$1 trillion annually, fundamentally altering the competitive landscape.

Third, the RPW database itself may generate competitive pressure that limits price pass-through. By making corridor-level pricing transparent and easily comparable, the World Bank's initiative creates a form of ``yardstick competition'' that constrains providers' ability to exploit temporary reductions in competition \citep{beck2018}.

These results do not imply that FATF grey-listing is costless. The financial system effects of de-risking --- reduced correspondent banking relationships, higher compliance burdens, and increased transaction costs for banks --- are well-documented \citep{fsb2018, imf2021wps, erbenova2016}. My contribution is to show that these wholesale effects do not pass through to the retail remittance costs paid by migrant families. The distinction matters for policy calibration. If the goal is to protect diaspora families from higher transfer costs, the evidence suggests that remittance market structure --- dominated by MTOs and increasingly by mobile money --- provides a buffer against correspondent banking disruption. If the goal is to maintain correspondent banking infrastructure for broader financial inclusion (trade finance, access to dollar clearing, investment flows), the case for reform rests on different evidence that this paper does not address.

An important caveat applies: the RPW database covers only formal, priced remittance channels. If grey-listing diverts flows from formal to informal channels (e.g., hawala), the true cost to migrants may rise through reduced transparency, security, and regulatory protection --- even if the observed formal cost does not change \citep{freund2008}. The null result on formal pricing should therefore be interpreted as evidence about one specific margin of adjustment, not as a comprehensive welfare assessment.


\section{Conclusion}

FATF grey-listing does not raise formal remittance costs. Using 378 corridors observed quarterly over 14 years and a staggered difference-in-differences design with 24 treatment cohorts, I find a precisely estimated null effect that is robust across specifications, channels, and outcome measures. The confidence interval rules out effects larger than 0.72 percentage points on a base cost of 7.35\%. The finding challenges the prevailing view that de-risking imposes a ``regressive tax'' on diaspora families through higher transfer costs, and suggests that the remittance market's shift toward MTOs and mobile money has insulated consumer prices from the disruption of correspondent banking relationships. Whether grey-listing disrupts wholesale financial intermediation --- and whether those disruptions carry welfare costs through other channels --- remains an open question for future work.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}


\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsubsection*{World Bank Remittance Prices Worldwide (RPW)} The RPW database collects quarterly price quotes from remittance service providers operating in major corridors. Each observation records the total cost of sending \$200 (and separately \$500) through a specific provider in a specific corridor. Total cost comprises the transfer fee and the exchange rate margin. I use the \$200 benchmark, consistent with the SDG indicator 10.c.1. The raw data contain 247,490 firm-level observations across 378 corridors. I aggregate to corridor-quarter level by computing the unweighted mean cost across all providers, yielding 16,849 corridor-quarter observations.

\subsubsection*{FATF grey-list compilation} FATF plenary statements from 2010 to 2025 were used to construct the complete panel of grey-list entries and exits. Each plenary announcement identifies which countries are added to and removed from the ``Jurisdictions under Increased Monitoring'' list. I recorded 74 distinct episodes across 63 countries. Treatment timing is assigned to the quarter of the plenary announcement (February = Q1, June = Q2, October = Q4).

\subsubsection*{Cohort construction and sample restrictions} The FATF grey-list panel records 74 episodes across 63 countries since 2010. Of these, 24 distinct treatment cohorts enter the grey list during the RPW sample period (2011Q1--2025Q1). The remaining episodes either began before 2011 (and thus lack a clean entry event in the pricing data) or involve countries not covered by the RPW database. I exclude corridors to 18 receiving countries that were already on the grey list at the start of the sample, as the CS-DiD framework requires observing the transition from untreated to treated status. The remaining sample contains 14,469 corridor-quarter observations across 322 corridors, with 152 corridors to treated destinations and 170 corridors to never-treated destinations. Including the always-treated corridors in the TWFE specifications (\Cref{tab:main}, column 4) does not alter the null finding.

\subsubsection*{Scope limitation} The original research design envisioned a dual-outcome analysis combining remittance prices with BIS Locational Banking Statistics to measure effects on cross-border banking claims. The BIS analysis is deferred to future work due to the complexity of matching BIS counterparty codes to FATF listing status at quarterly frequency. The present paper focuses on the retail price channel, which is the most policy-relevant dimension for migrant welfare.

\section{Robustness Appendix}

Additional robustness checks are reported in the main text (\Cref{tab:robust}). The sending-country placebo, alternative control groups, channel heterogeneity, extensive margin, and entry-exit asymmetry tests all support the main null finding.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
