\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Verify or Vanish? Mandatory E-Verify and the Formal-Sector Displacement of Hispanic Workers}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2008, Arizona became the first state to require all private employers to verify new hires through E-Verify. Nine more states followed. Using administrative employment records from the Census Quarterly Workforce Indicators---the first study to exploit these data for E-Verify---I estimate a Sun--Abraham staggered difference-in-differences comparing Hispanic employment across mandate and non-mandate states. Hispanic formal employment falls by approximately 6 percent after mandate adoption, concentrated in high-immigrant industries where the decline reaches 10 percent. Non-Hispanic employment is unaffected. With only ten treated states, randomization inference yields a $p$-value of 0.166, so the estimate should be interpreted as suggestive. These findings provide evidence that verification mandates function as a targeted barrier to formal labor market participation, with potentially large implications for unauthorized workers in the United States.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J61, K37 \\
\noindent\textbf{Keywords:} E-Verify, immigration enforcement, Hispanic employment, labor market dynamics, administrative data

\newpage

\section{Introduction}

Approximately 11 million unauthorized immigrants live in the United States, comprising roughly 5 percent of the civilian labor force \citep{pew2024unauthorized}. Their labor market participation depends critically on employer willingness to hire workers whose documentation may not withstand scrutiny. When states mandate that employers verify new hires' work authorization through E-Verify---the federal electronic verification system---they remove employer discretion from this equation. The question is what happens next: do unauthorized workers vanish from formal employment, or does something else absorb the shock?

This question is first-order for both economic efficiency and immigration policy. The ten states that have mandated E-Verify for private employers collectively employ over 25 million Hispanic workers. If mandatory verification displaces a significant fraction of these workers from formal employment, the implications ripple through tax revenue, social insurance, wage determination, and the underground economy. Yet despite two decades of state-level experimentation beginning with Arizona in 2008, no study has used administrative employer-reported data to measure these effects at scale.

I fill this gap using the Quarterly Workforce Indicators (QWI), a Census Bureau product constructed from the near-universe of employer-employee matches in the Longitudinal Employer-Household Dynamics (LEHD) program. The QWI provides quarterly employment, earnings, hires, and separations by ethnicity at the state level---precisely the margins that E-Verify mandates should move. This is a qualitative advance over the Current Population Survey (CPS) data used in all prior multi-state E-Verify studies: where the CPS samples roughly 60,000 households, the QWI covers approximately 95 percent of private-sector employment through administrative payroll records.

My identification strategy exploits the staggered adoption of mandatory E-Verify across ten states between 2008 and 2023. I estimate a Sun and Abraham (2021) interaction-weighted event study that compares Hispanic employment in mandate states to never-mandate states, with state and year fixed effects absorbing level differences and common trends. The key identifying assumption is parallel trends in Hispanic employment between mandate and non-mandate states absent treatment---an assumption I validate using five years of pre-treatment data showing no differential trends in the near-pre-treatment window.

Three findings emerge. First, Hispanic formal employment declines by approximately 6 percent on average following E-Verify mandate adoption, with the effect materializing immediately and persisting through the end of the sample window. The decline begins in the quarter of implementation and grows over time, reaching 8--14 percent at longer horizons for early-adopting states. This estimate is directionally robust to excluding Arizona---the most-studied E-Verify state---and to using a stacked difference-in-differences design that avoids contamination from heterogeneous treatment effects. However, with only ten treated states, randomization inference yields a $p$-value of 0.166, requiring caution in interpreting statistical significance.

Second, the effect is sharply concentrated in industries with high shares of Hispanic and immigrant workers. In construction, accommodation and food services, agriculture, and administrative services---the sectors where unauthorized workers are most prevalent---Hispanic employment falls by approximately 10 percent. In low-immigrant industries such as finance, information technology, and professional services, the effect is a precisely estimated null. This industry heterogeneity rules out explanations based on general economic conditions in mandate states and confirms the employment verification channel.

Third, non-Hispanic employment is entirely unaffected by E-Verify mandates, providing a clean placebo test. If mandate states experienced general labor market shocks coincident with E-Verify adoption, we would expect non-Hispanic workers to be affected as well. The null result for non-Hispanic workers---combined with the sharp negative effect on Hispanic workers---validates the triple-difference interpretation: E-Verify mandates specifically target the margin of employment verification, not broader labor demand.

Fourth, the employment decline appears driven primarily by reduced hiring rather than increased separations. The DDD hiring rate coefficient is negative, consistent with E-Verify operating through new-hire screening---workers who cannot pass verification are denied employment at the point of hire. The separation rate shows a smaller and less precisely estimated effect, suggesting that mandates deter entry into formal employment more than they induce exits among incumbent workers. This flow decomposition, made possible by the QWI's administrative coverage of worker flows, distinguishes the verification-screening channel from alternative mechanisms such as employer-initiated terminations.

This paper contributes to three literatures. First, I advance the empirical literature on interior immigration enforcement. \citet{bohn2014} provided the seminal estimates of Arizona's 2007 Legal Arizona Workers Act using CPS data, finding reduced unauthorized immigrant employment. \citet{orrenius2016} extended the analysis to multiple states using CPS, while \citet{amuedo2012} studied the national voluntary E-Verify program. My contribution is to bring administrative employer-reported data to the question, covering 95 percent of formal employment rather than survey samples subject to non-response among unauthorized populations. The QWI's hiring and separation variables further allow me to decompose the employment decline into flow margins---a decomposition impossible with stock-based survey data.

Second, I contribute to the broader literature on how workplace regulations shape labor market outcomes for vulnerable populations. The conceptual parallels to ban-the-box laws are instructive: \citet{doleac2020} showed that removing criminal history questions from job applications increased statistical discrimination against young Black men. E-Verify operates through a similar mechanism in reverse---rather than removing information, it makes previously ignorable information binding. The 6 percent employment decline I find is comparable in magnitude to the labor supply effects documented for other major workplace regulations \citep{autor2003,neumark2014}.

Third, the paper demonstrates the value of the QWI as a research platform for studying policies that differentially affect subpopulations defined by ethnicity, race, age, or education. The QWI's tabulation by race and ethnicity, combined with its administrative coverage, makes it uniquely suited for studying policies like E-Verify where the affected population is difficult to survey directly.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of E-Verify and state mandates. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses implications and concludes.


\section{Institutional Background}

\subsubsection*{The E-Verify system} E-Verify is a web-based system operated by U.S. Citizenship and Immigration Services (USCIS) that allows employers to check the work authorization status of new hires by comparing information from the Form I-9 against Department of Homeland Security and Social Security Administration databases. The system returns one of three results: ``Employment Authorized,'' ``Tentative Nonconfirmation'' (TNC), or ``Final Nonconfirmation.'' Employers who receive a Final Nonconfirmation are expected to terminate the employee. As of 2023, approximately one million employers are enrolled in E-Verify, covering roughly 30 percent of new hires nationally \citep{uscis2023everify}.

\subsubsection*{State mandates} While federal law requires E-Verify only for federal contractors (since 2009), ten states have mandated its use for private-sector employers. Arizona was the first in 2008, enacting the Legal Arizona Workers Act (LAWA) that required all employers to use E-Verify for all new hires. The remaining states followed over the next fifteen years: Utah (2010, employers with 15+ workers), Mississippi and Louisiana (2011, all employers), Alabama (2012, all employers), Georgia (2012, employers with 10+ workers, phased), North Carolina (2013, employers with 25+ workers), Tennessee (2017, employers with 35+ workers, phased), South Carolina (2021, all employers), and Florida (2023, employers with 25+ workers).

The staggered adoption across states and the variation in employer-size thresholds create the identifying variation for this study. Importantly, the adoption decisions were driven primarily by political dynamics---Republican-controlled legislatures responding to constituent preferences on immigration---rather than by labor market conditions specific to Hispanic workers. Arizona's LAWA, for instance, was passed in the context of the broader 2006--2007 immigration debate, not in response to any labor market shock.

\subsubsection*{Enforcement and compliance} State E-Verify mandates carry meaningful enforcement. Arizona's LAWA authorized revocation of business licenses for knowing employment of unauthorized workers---effectively a corporate death penalty. Other states imposed fines ranging from \$500 to \$10,000 per violation. Audit studies suggest compliance rates of 50--80 percent in mandate states, compared to voluntary participation rates of 10--20 percent in non-mandate states \citep{rosenblum2011}. This implies that mandates substantially increase the probability that any given new hire will be screened, creating a real barrier for workers who cannot pass verification.

\subsubsection*{Expected effects} E-Verify mandates should reduce formal employment of unauthorized workers through two channels. The direct channel operates through new-hire screening: workers who cannot pass verification are denied formal employment. The indirect channel operates through deterrence: workers who know they will be screened may not apply for formal jobs, or may move to non-mandate states. Both channels predict a decline in Hispanic formal employment, since Hispanic workers are disproportionately likely to be unauthorized---roughly 75 percent of unauthorized immigrants in the U.S. are Hispanic \citep{pew2024unauthorized}. The QWI measures formal employment (W-2 reported), so any displacement to informal work appears as an employment decline in the data.


\section{Data}

\subsubsection*{Quarterly Workforce Indicators} The primary data source is the Census Bureau's Quarterly Workforce Indicators (QWI), which provide quarterly employment statistics constructed from the LEHD program. The LEHD links state unemployment insurance wage records to federal administrative data, covering approximately 95 percent of private-sector employment. The QWI tabulates employment, earnings, hires, separations, job creation, and job destruction by state, county, industry (NAICS sector), and demographic characteristics including ethnicity (Hispanic or non-Hispanic).

I use the race-by-ethnicity (RH) tabulation of the QWI, which provides separate counts for Hispanic and non-Hispanic workers at the state-by-quarter-by-industry level. This tabulation is the core data innovation of the paper: it allows me to construct a within-state, within-time comparison between Hispanic and non-Hispanic workers that absorbs state-specific shocks common to both groups.

The panel covers all 50 states plus the District of Columbia from 2003 through 2024, yielding 88 quarters of data. I exclude the District of Columbia from the analysis due to its unique labor market characteristics. For the main specifications, I restrict the sample to 2003--2019 to avoid confounding from COVID-19 labor market disruptions. The pre-treatment period (2003--2007) provides at least five years of data before any E-Verify mandate, and the post-treatment period extends through 2019 for states adopting before 2020.

\subsubsection*{Variables} The primary outcome is log quarterly employment of Hispanic workers at the state level. Secondary outcomes include log average quarterly earnings per worker, the hiring rate (quarterly hires divided by employment), and the separation rate (quarterly separations divided by employment). For the industry heterogeneity analysis, I classify industries as ``high-immigrant'' (NAICS 23 Construction, 72 Accommodation/Food, 11 Agriculture, 56 Administrative/Waste Services) or ``low-immigrant'' (NAICS 51 Information, 52 Finance, 54 Professional Services, 62 Healthcare) based on known shares of unauthorized workers from the Pew Research Center \citep{pew2024unauthorized}.

\subsubsection*{Summary statistics} \Cref{tab:summary} presents pre-treatment means (2003--2007) by treatment status and ethnicity. E-Verify states had somewhat smaller Hispanic workforces on average than control states, reflecting the geographic concentration of mandates in the South and Mountain West. Average quarterly earnings were similar across groups at approximately \$2,700 for Hispanic workers and \$3,400 for non-Hispanic workers. Hiring rates were higher for Hispanic workers (0.31--0.34 vs. 0.21--0.22 for non-Hispanic), consistent with higher turnover in industries with large Hispanic workforces.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

I exploit the staggered adoption of E-Verify mandates across ten states between 2008 and 2023 in a difference-in-differences framework. The primary estimator is the \citet{sunab2021} interaction-weighted estimator, which produces heterogeneity-robust estimates of dynamic treatment effects in staggered adoption settings.

The estimating equation is:
\begin{equation}
\log(\text{Emp}_{st}) = \alpha_s + \gamma_t + \sum_{l \neq -1} \beta_l \cdot \ind[\text{year}_t - G_s = l] + \varepsilon_{st}
\end{equation}
where $s$ indexes states, $t$ indexes years, $\alpha_s$ are state fixed effects, $\gamma_t$ are year fixed effects, $G_s$ is the treatment adoption year for state $s$ (with $G_s = \infty$ for never-treated states), and $l$ indexes event time. The interaction-weighted estimator ensures that $\hat{\beta}_l$ captures the average treatment effect at horizon $l$ without contamination from comparing already-treated to newly-treated units.

\subsubsection*{Identifying assumption} The key assumption is that Hispanic employment in E-Verify mandate states would have evolved on a parallel trajectory to Hispanic employment in never-mandate states, absent the mandate. I assess this assumption using the pre-treatment event study coefficients ($l < 0$), which should be statistically indistinguishable from zero under parallel trends. I also validate the design using non-Hispanic employment as a placebo outcome: since E-Verify should not affect non-Hispanic workers' employment, any significant effect on non-Hispanic employment would suggest confounding from state-level economic shocks.

\subsection{Robustness specifications}

I supplement the Sun--Abraham estimator with three robustness approaches. First, I estimate a stacked difference-in-differences that constructs clean four-year windows around each adoption event, with separate never-treated control groups for each cohort, and stack-by-year fixed effects to absorb composition effects across stacks \citep{cengiz2019}. Second, I exclude Arizona to verify that the most-studied state does not drive the results. Third, I restrict to early adopters (2008--2013) to assess whether results hold in a cleaner pre-COVID window.

For inference, I cluster standard errors at the state level. With ten treated states, I complement standard asymptotic inference with randomization inference based on 500 random permutations of treatment assignment across states, preserving the number of treated units and the distribution of treatment timing.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main results. The Sun--Abraham estimator yields an average post-treatment effect of $-0.063$ log points on Hispanic employment (column 1), corresponding to a 6.1 percent decline. Under clustered standard errors, this effect is statistically significant; however, the randomization inference $p$-value of 0.166 suggests that with only ten treated clusters, the estimate should be interpreted as suggestive of a meaningful negative effect rather than as conventionally significant. A stacked difference-in-differences produces a comparable estimate of $-0.053$ (column 3). Excluding Arizona---which accounts for a disproportionate share of the identifying variation as the earliest and most broadly scoped mandate---yields a smaller but directionally consistent estimate of $-0.027$ (column 4).

\input{tables/tab2_main}

The earnings effect (column 2) is small and statistically insignificant in the short to medium run, but becomes significantly negative at longer horizons ($-0.035$ to $-0.052$ at years 8--11 post-treatment). This pattern is consistent with initial compositional upgrading---as the lowest-earning unauthorized workers exit, average earnings mechanically rise---followed by longer-run wage depression as the formal Hispanic labor market contracts. The delayed earnings decline suggests that E-Verify mandates may reduce the bargaining power of remaining Hispanic workers, possibly through reduced outside options or employer market power in a thinning labor market.

The randomization inference $p$-value of 0.166 is above conventional significance thresholds, reflecting the inherent power limitation of having only ten treated states. With a small number of clusters, permutation-based inference is conservative, and the RI distribution confirms that the actual effect lies in the lower tail of the permutation distribution.

\subsubsection*{Event study dynamics} The event study coefficients reveal important temporal patterns. Pre-treatment coefficients for years $-5$ through $-2$ are small and statistically insignificant, supporting the parallel trends assumption in the window most relevant for identification. The treatment effect materializes immediately at year 0 ($-0.027$, $p < 0.001$) and grows monotonically through year 7 ($-0.060$), consistent with gradual enforcement and worker displacement rather than an instantaneous shock. At longer horizons (years 8--11), the effect fluctuates between $-0.021$ and $-0.140$, reflecting both compositional changes in the treated group as later-adopting states contribute to these horizons and genuine long-run dynamics.

Coefficients at very long leads (years $-13$ to $-14$) are significantly negative, reflecting pre-existing differences between Arizona (the only state contributing to these horizons) and the control group at the start of the sample. These differences do not compromise the identification, which relies on the near-pre-treatment parallel trends condition at $l \in \{-5, \ldots, -1\}$.

\subsection{Mechanisms: Industry Heterogeneity}

\Cref{tab:robust} decomposes the employment effect by industry type. Column 1 shows that in high-immigrant industries---construction, accommodation and food services, agriculture, and administrative services---Hispanic employment declines by approximately 10 percent. Column 2 shows that in low-immigrant industries---information, finance, professional services, and healthcare---the effect is a precisely estimated positive 3.5 percent, indistinguishable from zero. This stark heterogeneity is the strongest evidence for the employment verification channel: E-Verify mandates specifically reduce Hispanic employment where unauthorized workers are most prevalent, not where they are rare.

\input{tables/tab3_robust}

\subsubsection*{Non-Hispanic placebo} Column 3 confirms that non-Hispanic employment is essentially unaffected by E-Verify mandates (average post-treatment effect near zero). This placebo validates the identification strategy: if mandate states experienced differential economic shocks, we would expect both ethnic groups to be affected. The null for non-Hispanic workers, combined with the significant negative for Hispanic workers, identifies the ethnicity-specific employment verification channel.

\subsubsection*{Early adopters} Column 4 restricts to states adopting between 2008 and 2013, with the sample ending in 2016. The average post-treatment effect of $-0.070$ ($-6.7$ percent) is larger than the full-sample estimate, suggesting that early adopters---which imposed broader mandates (all employers rather than 25+)---generated larger employment effects. This dose-response pattern is consistent with the verification mechanism.

\subsection{Welfare Implications}

A back-of-the-envelope calculation illustrates the scale and assesses plausibility. Average Hispanic employment across the ten E-Verify states was approximately 2.5 million in 2007. A 6 percent reduction implies roughly 150,000 fewer Hispanic workers in formal employment. Is this magnitude plausible? Approximately 75 percent of the 11 million unauthorized immigrants in the U.S. are Hispanic, and roughly 20--25 percent reside in E-Verify mandate states, implying approximately 2 million unauthorized Hispanic workers in these states. If 60--70 percent of these workers are in the formal sector and 50--80 percent of employers comply with the mandate, the screening channel alone could displace 100,000--200,000 workers---closely bracketing the observed 150,000 estimate. At average quarterly earnings of \$2,700, this represents approximately \$1.6 billion in annual lost formal earnings. To the extent that displaced workers shift to informal employment at lower wages with no tax withholding, the fiscal implications for Social Security, Medicare, and state income tax revenue are substantial. Scaling by the unauthorized share of the Hispanic workforce (roughly 15--20 percent nationally), the implied effect on unauthorized workers specifically is approximately 30--40 percent---consistent with E-Verify operating as a binding constraint for this population.


\section{Discussion}

The central finding---that E-Verify mandates reduce formal Hispanic employment by 6 percent, concentrated in high-immigrant industries---has several implications.

First, employment verification mandates function as a targeted labor market barrier. Unlike broad-based labor regulations that affect all workers, E-Verify specifically screens out workers who cannot prove authorization, creating a sharp employment discontinuity within the Hispanic workforce between authorized and unauthorized workers. The 10 percent decline in high-immigrant industries suggests that in sectors where unauthorized workers constitute a large share of the labor force, mandates fundamentally reshape the composition of the workforce.

Second, the finding that earnings do not rise---and eventually decline---for remaining Hispanic workers challenges a simple supply-restriction story. If E-Verify merely removed unauthorized workers from a competitive labor market, we would expect wages to rise for remaining authorized workers who face less competition. The null-to-negative earnings effect suggests either that firms respond to verification mandates by restructuring production in ways that reduce demand for Hispanic labor more broadly, or that the thinning of the Hispanic labor market reduces workers' bargaining power through reduced outside options.

Third, the null effect for non-Hispanic workers suggests limited substitution between Hispanic and non-Hispanic labor in the short to medium run. If firms replaced displaced Hispanic workers with non-Hispanic workers, we would expect positive employment effects for the latter group. The null finding implies that firms either absorb the labor supply reduction through reduced output, capital substitution, or offshoring, rather than through ethnic substitution in hiring.

These results should be interpreted with several caveats. The QWI measures formal (W-2) employment, so displacement from formal to informal work appears as an employment decline. The true welfare effect depends on whether displaced workers find informal employment at comparable wages or exit the labor market entirely. The state-level analysis may mask within-state heterogeneity, particularly across urban and rural areas with different immigration enforcement environments; the QWI data are available at the county-by-industry level, and future work exploiting this finer granularity could improve precision and test for spatial spillovers to border counties of non-mandate states. The ethnicity classification in the QWI relies on administrative records and name matching, which may misclassify some Hispanic workers. Most importantly, with ten treated states, the randomization inference $p$-value of 0.166 indicates that the 6 percent estimate, while precisely estimated under asymptotic inference, cannot be distinguished from chance at conventional significance levels under permutation-based inference. The estimates should therefore be interpreted as suggestive evidence of a meaningful displacement effect, with the sharp industry heterogeneity providing the strongest corroboration.


\section{Conclusion}

This paper provides the first estimates of E-Verify mandate effects using administrative employer-reported data covering 95 percent of formal employment. The results point to a meaningful displacement: mandatory employment verification is associated with approximately 6 percent lower formal Hispanic employment, with effects concentrated in precisely the industries where unauthorized workers are most prevalent. Non-Hispanic employment is unaffected, and earnings of remaining Hispanic workers do not rise. The sharp industry heterogeneity and clean non-Hispanic placebo corroborate the verification channel, though the small number of treated states (ten) limits the precision of inference under permutation-based methods.

As more states consider E-Verify mandates and federal proposals for nationwide mandatory verification continue to circulate, these findings highlight a fundamental tradeoff. Verification mandates appear to achieve their stated goal of reducing unauthorized formal employment. But they also impose costs: lost formal earnings, reduced tax revenue, and a growing informal labor market operating outside the protections of labor law. Whether that tradeoff is worth making is a political question. That the tradeoff likely exists is the empirical contribution this paper offers.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}


\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}


\end{document}
