\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Examination Lottery: Measuring Regulatory Inconsistency\\ with Patent Continuation Twin Studies}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Patent examination is supposed to be consistent: the same invention should receive the same verdict regardless of which examiner reviews it. I test this assumption using 97,922 continuation-application twin pairs where the same invention family was reassigned to a different examiner within the same art unit at the USPTO. Reassignment increases outcome discordance---one application granted, the other abandoned---by 7.7 percentage points relative to same-examiner pairs (11.0 pp raw differential). A one-standard-deviation increase in examiner leniency raises the grant probability by 11.4 percentage points ($t = 33.8$). Small entities face 3.2 percentage points greater excess discordance than large entities. These results reveal a regulatory lottery in which examiner identity causally determines patent rights, with disproportionate consequences for resource-constrained inventors.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} O34, L51, K11 \\
\noindent\textbf{Keywords:} patent examination, regulatory inconsistency, examiner leniency, continuation applications, small entities

\newpage

\section{Introduction}

Nearly 30 percent of patent continuation applications receive a different verdict when reviewed by a different examiner in the same technology group. The same claims, the same prior art, the same legal standards---yet whether an invention becomes a patent depends substantially on which examiner happens to be available when the application arrives. This paper provides the first direct measurement of this examination lottery using a within-family twin study design that holds the invention family fixed.

The consistency of patent examination is a first-order policy question. Patents confer exclusive rights worth billions of dollars in aggregate, shaping innovation incentives, market structure, and technology diffusion \citep{jaffe2004patent}. If examination outcomes vary arbitrarily with examiner identity, then the patent system functions partly as a lottery---randomly allocating property rights in ways that distort resource allocation. Prior work has documented examiner heterogeneity using cross-sectional variation across different inventions \citep{lemley2001rational, cockburn2003all, lichtman2004rethinking}, but comparing different inventions across examiners confounds examiner effects with invention quality. The central challenge is isolating the causal contribution of examiner identity while holding the underlying invention constant.

I exploit the institutional structure of continuation applications at the United States Patent and Trademark Office (USPTO). When inventors file continuation or divisional applications, the new application may be assigned to a different examiner than the parent---even within the same art unit. This reassignment is driven by workload balancing and examiner availability, creating quasi-random variation in examiner identity conditional on art-unit-by-filing-year fixed effects. Crucially, because parent and child applications belong to the same invention family, comparing their outcomes holds invention quality approximately fixed, yielding a twin study design with 702,490 same-art-unit pairs.

The results are striking. Among 97,922 pairs where the child was reassigned to a different examiner, 29.8 percent produced discordant outcomes (one granted, the other abandoned), compared to 18.7 percent for same-examiner pairs---an excess discordance of 11.0 percentage points. In regressions with art-unit-by-filing-year fixed effects, reassignment increases discordance by 7.7 percentage points ($t = 21.1$). The examiner's leave-one-out grant rate is a powerful predictor of the child's outcome: a one-standard-deviation increase in leniency raises the grant probability by 11.4 percentage points, with a first-stage $F$-statistic of 1,158. These magnitudes imply that examiner identity explains a substantial share of outcome variation beyond what the invention itself warrants.

The burden of this regulatory lottery falls disproportionately on small entities. Among reassigned pairs, small-entity applications exhibit 38.4 percent discordance compared to 27.2 percent for large entities---a gap of 11.2 percentage points. In a triple-difference specification, the small-entity-by-reassignment interaction is 3.2 percentage points ($t = 4.3$). The leniency effect is also larger for small entities (0.692 versus 0.626), consistent with small inventors having fewer resources to appeal unfavorable decisions or navigate examiner-specific preferences.

This paper contributes to three literatures. First, it advances the measurement of examiner effects in patent examination. \citet{sampat2019examiner} show that examiner leniency affects downstream innovation outcomes, and \citet{farre2020patent} estimate that patents cause 36 percent higher employment for startups, using examiner leniency as an instrument. Both compare different inventions across examiners; I hold invention quality fixed by comparing applications within the same family. The 29.8 percent discordance rate among reassigned twins provides a direct measure of the examination lottery that these cross-sectional designs cannot isolate.

Second, this paper relates to the growing literature on regulatory discretion and its distributional consequences. \citet{kling2006incarceration} and \citet{dobbie2018effects} show that judge identity causally affects sentencing and downstream outcomes in criminal justice. \citet{maestas2013disability} demonstrate similar leniency effects in disability determination. My results extend the examiner-leniency paradigm to intellectual property, where the stakes---exclusive market rights rather than years of incarceration---are economic rather than punitive, and the small-entity penalty reveals how regulatory inconsistency compounds existing inequalities in the innovation system.

Third, I contribute to debates about patent quality and patent reform. \citet{frakes2017patent} show that examiner time pressure reduces examination thoroughness, and \citet{galasso2015patents} find that patent invalidation spurs follow-on innovation. My finding that 29.8 percent of reassigned continuations receive different verdicts suggests that the examination process itself introduces noise into the patent system at a scale comparable to the effects of intentional policy changes. The portability of this ``examination lottery'' concept extends beyond patents: any administrative adjudication system with examiner discretion---immigration, tax audits, disability reviews---faces the same consistency challenge.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional setting. \Cref{sec:data} presents the data. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} reports results. \Cref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

\subsubsection*{Patent examination at the USPTO} The USPTO receives over 600,000 utility patent applications annually. Each application is assigned to one of approximately 8,000 examiners organized into art units---groups of 10--20 examiners specializing in a narrow technology area. The examiner reviews the application against prior art and statutory requirements (novelty under 35 U.S.C. \S 102, non-obviousness under \S 103, written description under \S 112), issuing office actions that may accept or reject claims. Applications ultimately receive a ``disposal''---issuance as a patent (ISS) or abandonment (ABN)---typically within 3--5 years of filing \citep{frakes2017patent}.

\subsubsection*{Continuation and divisional applications} Inventors may file continuation applications that claim priority from a parent application, maintaining the same effective filing date while presenting different or refined claims. Divisional applications arise when the examiner determines that the parent contains multiple distinct inventions. Both continuation (CON) and divisional (DIV) applications enter the same art unit as the parent but may be assigned to a different examiner based on workload and availability. This reassignment is the source of identifying variation in this paper.

\subsubsection*{Examiner assignment} Within an art unit, applications are assigned to examiners through a combination of specialty matching and workload balancing. The assignment process does not condition on application quality or the identity of the inventor. While examiners may develop expertise in narrower subfields within an art unit, the decision of which examiner receives a particular continuation is primarily administrative---an ``as-good-as-random'' assignment conditional on art unit and filing year \citep{sampat2019examiner}.

\subsubsection*{Examiner heterogeneity} Prior work documents substantial variation in examiner grant rates even within art units. \citet{cockburn2003all} find that allowance rates range from 35 to 97 percent across examiners in the same technology center. \citet{lemley2001rational} argues that this heterogeneity reflects differences in effort, risk tolerance, and interpretation of legal standards rather than case mix. My data confirm this: among examiners with 50 or more decided applications, the within-art-unit standard deviation of grant rates is 0.172.


\section{Data}
\label{sec:data}

I construct the dataset from the USPTO Patent Examination Research Dataset (PatEx) \citep{marco2015patent}, accessed via Google BigQuery. PatEx contains the universe of 9.8 million patent applications with filing dates, examiner identifiers, art unit assignments, disposal types, and entity status. I link parent-child continuation pairs using the continuity data and merge office action records from the Office Action Research Dataset \citep{lu2017USPTO}.

\subsubsection*{Sample construction} I begin with all continuation (CON) and divisional (DIV) parent-child pairs filed between 1998 and 2015 where both applications have resolved outcomes (issued or abandoned) and non-missing examiner identifiers. I restrict to pairs where the child is in the same art unit as the parent, yielding 702,490 same-art-unit pairs. Of these, 97,922 (13.9 percent) were reassigned to a different examiner. The filing-year restriction ensures at least five years for disposition by the time of data collection.

\subsubsection*{Key variables} The outcome is a binary indicator for whether the child application was granted (disposal type ISS). The treatment variable in the twin comparison is an indicator for reassignment (different examiner than the parent). For the leniency analysis, I construct each examiner's leave-one-out (LOO) grant rate: the share of all other applications decided by that examiner that were granted. The LOO rate averages 0.749 with a standard deviation of 0.172, indicating substantial examiner-level variation in grant propensity.

\Cref{tab:summary} presents summary statistics. The overall grant rate for child applications is 78.1 percent, slightly below the parent grant rate of 84.0 percent---consistent with continuations facing somewhat tougher examination since the parent's claims have already been partially adjudicated. Small entities (those qualifying for reduced USPTO fees) account for 21.9 percent of the sample.

\input{tables/tab1_summary}


\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Twin study design} The core comparison exploits the within-family structure of continuation applications. For a given parent-child pair $(p, c)$ in art unit $a$ filed in year $t$, I estimate:
\begin{equation}
\text{Discordant}_{pc} = \alpha + \beta \cdot \text{Reassigned}_{pc} + \gamma \cdot \text{ParentGranted}_{pc} + \delta_{at} + \varepsilon_{pc}
\label{eq:discord}
\end{equation}
where $\text{Discordant}_{pc} = \ind[\text{Disposal}_c \neq \text{Disposal}_p]$, $\text{Reassigned}_{pc} = \ind[\text{Examiner}_c \neq \text{Examiner}_p]$, and $\delta_{at}$ are art-unit-by-filing-year fixed effects. The coefficient $\beta$ measures the causal effect of examiner reassignment on outcome inconsistency, identified under the assumption that reassignment is orthogonal to invention quality within art-unit-year cells.

\subsubsection*{Examiner leniency design} To measure the intensity of the examination lottery, I estimate:
\begin{equation}
\text{Granted}_c = \alpha + \beta \cdot \text{Leniency}_c^{-i} + \gamma \cdot \text{ParentGranted}_{pc} + \delta_{at} + \varepsilon_c
\label{eq:leniency}
\end{equation}
where $\text{Leniency}_c^{-i}$ is the leave-one-out grant rate of the child's examiner. This specification is estimated on the subsample of reassigned pairs, where the child examiner is plausibly as-good-as-random conditional on $\delta_{at}$. The coefficient $\beta$ captures the partial derivative of the grant probability with respect to examiner leniency, holding the invention family fixed through parent outcome and art-unit-year fixed effects.

\subsubsection*{Identification assumptions} The identifying assumption is that examiner assignment for continuation applications is independent of application quality conditional on art-unit-by-filing-year fixed effects. This follows from the administrative assignment mechanism described in \Cref{sec:background}: within an art unit and year, which examiner receives a continuation depends on workload and availability rather than case characteristics.

I probe this assumption with balance tests. \Cref{tab:balance} shows that conditional on art-unit-year fixed effects, the correlation between examiner leniency and small-entity status is $-0.009$ ($t = -5.5$), and the correlation with parent grant status is $0.007$ ($t = 4.7$). While both are statistically significant---as expected with nearly 100,000 observations---they are economically negligible, explaining less than 0.1 percent of leniency variation (within $R^2 < 0.001$). The identifying assumption requires conditional independence, not zero correlation in finite samples; these magnitudes are consistent with as-good-as-random assignment.

\subsubsection*{Scope of the twin comparison} An important caveat is that continuation and divisional applications are not identical to their parents. Children may present refined claims, different prosecution strategies, or narrower invention scope. The within-family design holds the invention \textit{family} fixed---not the precise claim set. The parent outcome control absorbs family-level quality differences, but residual within-family heterogeneity remains. The results should therefore be interpreted as measuring the contribution of examiner identity to outcome variation \textit{within families}, not as a pure test of whether identical applications receive different verdicts. Continuations (CON) provide the cleaner comparison because they pursue substantially similar claims; divisionals (DIV) may involve distinct inventions within the same specification. Robustness results in \Cref{tab:robust} confirm that the leniency effect is similar across both types.

\subsubsection*{Inference} Standard errors are clustered at the examiner level for specifications using examiner leniency and at the art-unit level for discordance specifications, reflecting the levels at which treatment varies. With 10,575 unique examiners and 671 art units, cluster counts are sufficiently large for reliable asymptotic inference.


\section{Results}
\label{sec:results}

\subsection{The Examination Lottery}

\Cref{tab:discordance} documents the core fact. Among same-art-unit pairs where the child was reassigned to a different examiner, 29.8 percent produced discordant outcomes---one application granted, the other abandoned. For same-examiner pairs, discordance is 18.7 percent. The raw excess discordance of 11.0 percentage points captures the additional inconsistency introduced by examiner reassignment.

The regression estimate in Panel B of \Cref{tab:discordance} confirms this pattern with art-unit-by-filing-year fixed effects: reassignment increases discordance by 7.7 percentage points ($\text{SE} = 0.004$). The difference from the raw gap reflects composition: reassigned pairs are somewhat more concentrated in technology areas with higher baseline discordance.

\input{tables/tab2_discordance}

\subsection{Balance}

\Cref{tab:balance} reports balance tests for the examiner leniency instrument. Conditional on art-unit-year fixed effects, examiner leniency is nearly orthogonal to observable characteristics of the application: the coefficient on small-entity status is $-0.009$ and on parent-grant status is $0.007$. Both are statistically significant but explain less than 0.1 percent of leniency variation, consistent with as-good-as-random assignment within art-unit-year cells.

\input{tables/tab3_balance}

\subsection{Examiner Leniency and Grant Outcomes}

\Cref{tab:main} presents the leniency results. Column (1) shows the preferred specification among reassigned pairs: a one-unit increase in examiner LOO grant rate raises the child's grant probability by 0.661 ($\text{SE} = 0.020$), controlling for parent outcome and art-unit-year fixed effects. At the mean leniency standard deviation of 0.172, this implies that a one-standard-deviation increase in examiner leniency raises the grant probability by 11.4 percentage points---a large effect relative to the 68.5 percent baseline grant rate for reassigned children. The $t$-statistic on leniency exceeds 33, confirming that examiner grant propensity is a powerful predictor of individual outcomes.

Column (3) extends to the full sample and interacts reassignment with leniency. The leniency effect for same-examiner pairs is 0.491, reflecting the mechanical correlation between an examiner's leniency and their own decisions. For reassigned pairs, the additional leniency effect is 0.112 ($t = 8.7$), consistent with reassigned children being more sensitive to the lottery of examiner assignment since they face a new decision-maker who has no memory of the parent case.

\input{tables/tab4_main}

\subsection{The Small-Entity Penalty}

The examination lottery falls disproportionately on small entities. In the raw data, reassigned small-entity pairs exhibit 38.4 percent discordance versus 27.2 percent for large entities---a gap of 11.2 percentage points (\Cref{tab:discordance}). The triple-difference estimate in the regression framework is 3.2 percentage points ($t = 4.3$): small entities face significantly greater excess discordance from reassignment even after absorbing art-unit-year fixed effects and controlling for parent outcome.

The leniency analysis confirms this pattern. The coefficient on examiner LOO rate is 0.692 for small entities versus 0.626 for large entities. Small inventors' grant outcomes are more sensitive to examiner identity, consistent with large entities having greater capacity to appeal adverse decisions, hire experienced patent attorneys, or restructure claims in response to examiner-specific feedback \citep{lanjouw2001characteristics}.

\subsection{Robustness}

\Cref{tab:robust} demonstrates that the leniency effect is stable across multiple dimensions. The coefficient is similar for continuations (0.639) and divisionals (0.694), for early filings (0.650, 1998--2006) and late filings (0.675, 2007--2015), and regardless of whether the parent was granted (0.647) or abandoned (0.666). Using within-art-unit-year standardized leniency yields a coefficient of 0.065 per standard deviation ($t = 31.0$), confirming that the result is not driven by cross-art-unit variation.

\input{tables/tab5_robustness}


\section{Discussion}
\label{sec:discussion}

These results reveal a regulatory lottery at the heart of the U.S. patent system. When the same invention family faces a different examiner, 29.8 percent of applications receive a different verdict. This is not a claim that examiners are incompetent or biased---the examination process involves genuine judgment under uncertainty, and reasonable experts may disagree. The point is that the magnitude of this disagreement is large enough to constitute a first-order distortion in the allocation of patent rights.

\subsubsection*{Economic magnitude} The examination lottery operates at considerable scale. Among 97,922 reassigned same-AU pairs, the 7.7-percentage-point excess discordance implies that roughly 7,500 applications received outcomes that would have been different under their parent's examiner. While this paper does not directly estimate downstream innovation effects, prior work suggests the stakes are material: \citet{farre2020patent} find that patent grants cause 36 percent higher employment for startups, implying that even a fraction of these discordant outcomes could meaningfully affect resource allocation in the innovation system.

\subsubsection*{The small-entity penalty} The disproportionate burden on small entities has implications for innovation policy. If the patent system functions as a lottery, resource-constrained inventors---who cannot afford repeated filings, experienced attorneys, or examiner-specific claim strategies---are most vulnerable to adverse draws. This compounds existing disparities documented by \citet{jensen2015gender} in patent prosecution success rates.

\subsubsection*{Limitations} The balance tests reveal small but statistically significant correlations between examiner leniency and application observables, suggesting that assignment is not perfectly random. While these correlations are economically negligible (within $R^2 < 0.001$), they warrant caution in interpreting the leniency estimates as strictly causal. The twin design addresses this concern by holding the invention family fixed within families, but some residual selection---for instance, if certain types of continuations are systematically routed to particular examiners---cannot be fully ruled out. Additionally, the LOO leniency measure uses an examiner's full career history rather than contemporaneous rates, potentially introducing measurement error that attenuates estimates.

\subsubsection*{Policy implications} Reducing examination inconsistency does not require eliminating examiner discretion. Targeted interventions---standardized examination guidelines with calibration exercises, blind second opinions on borderline cases, or statistical monitoring of examiner-level discordance rates---could reduce arbitrary variation while preserving the benefits of specialized human judgment. The twin study methodology developed here could serve as an ongoing audit tool for the patent office.


\section{Conclusion}

Patent examination operates partly as a lottery. Using 97,922 within-family twin comparisons at the USPTO, I show that examiner reassignment raises outcome discordance by 7.7 percentage points, with small entities bearing a disproportionate 3.2-percentage-point additional burden. The portable lesson is that any administrative adjudication system with examiner discretion faces a consistency challenge that can be measured---and potentially mitigated---using the twin study design introduced here. The examination lottery is not unique to patents; it is a structural feature of regulatory discretion that deserves systematic attention.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsubsection*{Data sources} The primary data come from the USPTO Patent Examination Research Dataset (PatEx), accessed via Google BigQuery (\texttt{patents-public-data.uspto\_oce\_pair}). This dataset contains the universe of 9.8 million utility patent applications filed at the USPTO, including examiner identifiers, art unit assignments, filing dates, disposal types, and entity status indicators. Continuation and divisional parent-child links come from the continuity tables within the same dataset. Office action data (rejection types and dates) come from the Office Action Research Dataset (\texttt{patents-public-data.uspto\_oce\_office\_actions}).

\subsubsection*{Sample construction} Starting from all continuation (CON) and divisional (DIV) parent-child pairs with child filing dates between January 1, 1998 and December 31, 2015:
\begin{enumerate}
\item Require both parent and child to have resolved outcomes (disposal type ISS or ABN): 1,160,090 pairs
\item Require non-missing examiner identifiers and art units for both: 1,160,090 pairs
\item Restrict to same-art-unit pairs (child art unit equals parent art unit): 702,490 pairs
\item Of these, 97,922 (13.9\%) have different examiners (reassigned)
\end{enumerate}

\subsubsection*{Variable definitions}
\begin{itemize}
\item \textbf{Granted:} Binary indicator equal to 1 if disposal type is ISS (issued).
\item \textbf{Discordant:} Binary indicator equal to 1 if child and parent disposal types differ.
\item \textbf{Reassigned:} Binary indicator equal to 1 if child examiner ID differs from parent examiner ID.
\item \textbf{Examiner leniency (LOO):} Leave-one-out grant rate: for child application $i$ assigned to examiner $j$, $\text{Leniency}_{j(-i)} = \frac{\sum_{k \neq i} \text{Granted}_k}{\sum_{k \neq i} 1}$ over all applications decided by examiner $j$.
\item \textbf{Small entity:} Binary indicator equal to 1 if the child application's entity status is ``small'' (qualifying for reduced USPTO fees under 37 CFR \S 1.27).
\item \textbf{Technology center:} First two digits of the art unit code, grouping related technology areas.
\end{itemize}


\section{Robustness Appendix}

The leniency effect is stable across all subsamples tested (\Cref{tab:robust}). Continuations (CON) and divisionals (DIV) yield similar coefficients (0.639 vs.\ 0.694), ruling out concerns that the results are driven by one application type. The temporal stability across early (1998--2006: 0.650) and late (2007--2015: 0.675) periods indicates that the examination lottery is not an artifact of a particular policy regime or examiner cohort. Conditioning on parent outcome (granted vs.\ abandoned) produces nearly identical estimates (0.647 vs.\ 0.666), confirming that the leniency effect does not reflect differential selection into continuation filing based on parent success.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}


\end{document}
