\begin{tabular}{lcccccc}
\toprule
Outcome & $\hat{\beta}$ & SE & SD($Y$) & SDE & SE(SDE) & Classification \\
\midrule
\multicolumn{7}{l}{\textit{Panel A: Pooled}} \\[3pt]
Log house price (all areas) & 0.0036 & 0.0016 & 0.4687 & 0.0078 & 0.0034 & Small positive \\
\midrule
\multicolumn{7}{l}{\textit{Panel B: Heterogeneous (by deprivation)}} \\[3pt]
Log house price (deprived areas) & -0.0108 & 0.0029 & 0.3725 & -0.0291 & 0.0078 & Small negative \\
Log house price (non-deprived areas) & 0.0097 & 0.0019 & 0.4787 & 0.0202 & 0.0040 & Small positive \\
\bottomrule
\end{tabular}
\par\vspace{0.3em}
{\footnotesize \textit{Notes:} \textbf{Country:} United Kingdom (England). \textbf{Research question:} Does receiving a negative Ofsted school inspection rating (Requires Improvement or Inadequate) cause residential property values in the school's postcode district to decline relative to areas near schools receiving positive ratings (Outstanding or Good)? \textbf{Policy mechanism:} Ofsted inspects all state-funded schools in England on a rolling cycle and publishes Overall Effectiveness ratings on a 1--4 scale; these ratings are publicly available and widely consulted by parents and homebuyers, creating an information channel through which school quality labels are capitalized into local house prices. \textbf{Outcome definition:} Log mean residential transaction price per postcode district per month, computed from HM Land Registry Price Paid Data standard transactions. \textbf{Treatment:} Binary indicator for receiving a rating of 3 (Requires Improvement) or 4 (Inadequate) versus 1 (Outstanding) or 2 (Good). \textbf{Data:} Ofsted Management Information (9,780 graded inspections, 2019--2024) linked to HM Land Registry Price Paid Data (8.7 million transactions, 2015--2024) via postcode district; school-month panel with 24 months pre/post inspection. \textbf{Method:} Difference-in-differences around inspection publication dates; school and calendar-month fixed effects; standard errors clustered at the postcode district level. \textbf{Sample:} State-funded schools in England with graded Ofsted inspections (2019--2024); restricted to schools with at least 12 months of house price data in their postcode district. SDE $= \hat{\beta} / \text{SD}(Y)$ where SD($Y$) is the pre-treatment standard deviation of log house prices. Classification refers to magnitude, not statistical significance: Large ($|$SDE$| > 0.15$), Moderate ($0.05$--$0.15$), Small ($0.005$--$0.05$), Null ($< 0.005$).}
