\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics (no figures in V1)
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Soviet Inheritance Trap: Bank Fraud, Branch Dependence, and Firm Employment in Moldova}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
One billion dollars vanished from three Moldovan banks in November 2014---roughly 12\% of GDP---in what became known as the ``stolen billion.'' These banks, anchored by the Soviet-era Banca de Economii, held 60\% of the national branch network. I exploit the Soviet inheritance of BEM's geographic footprint---determined by 1940s--1980s administrative planning, orthogonal to contemporary productivity---as a source of cross-raion variation in credit supply destruction. Using administrative enterprise data from all 35 Moldovan raions over 2005--2024, I find that a one-standard-deviation increase in bank dependence reduced employment by 5--8\% post-crisis. The effect emerged immediately, deepened through 2019, and persisted a decade later. Enterprise counts barely moved, implying the shock operated through firm-level contraction rather than exit---a \emph{zombie firm} channel in which credit-starved enterprises survived but stopped hiring.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G21, G01, E44, O16, P34 \\
\noindent\textbf{Keywords:} bank fraud, credit supply, firm employment, state banks, transition economies, Moldova

\newpage

\section{Introduction}

When a country's banking system collapses, who pays? The depositors make headlines, but the cost often falls hardest on small firms in places where the failed bank was the only game in town. Moldova's November 2014 banking crisis provides a stark laboratory for this question: three state-linked banks controlling 60\% of the national branch network were placed under emergency administration after \$1~billion in fraudulent loans were discovered---approximately 12\% of GDP extracted in a single coordinated theft \citep{imf2016moldova}.

The crisis struck asymmetrically. Banca de Economii (BEM), the largest of the three banks, had inherited its branch network from the Soviet-era Savings Bank (Sberbank), whose geographic footprint was determined by decades-old administrative planning rather than by market forces. Rural raions where BEM was the dominant---sometimes sole---financial institution lost their primary credit channel overnight. Urban centers with competitive banking markets absorbed the shock through alternative lenders.

This paper asks whether the geographic pattern of credit supply destruction---driven by an institutional inheritance from the Soviet era---caused persistent employment losses in dependent regions. I construct a raion-level panel of employment, enterprise counts, and turnover from Moldova's National Bureau of Statistics covering 2005--2024, and estimate the causal effect of bank dependence on firm-level outcomes using a difference-in-differences design. The treatment variable exploits the pre-crisis density of financial intermediation across raions: districts with fewer financial enterprises---where BEM was the dominant institution---experienced a larger effective credit supply shock.

The main finding is economically large and statistically robust. A one-standard-deviation increase in BEM dependence reduced average employment by 8.3\% (95\% CI: $-$13.4\% to $-$3.1\%) in the preferred specification with region-by-year fixed effects. The effect appeared immediately in 2015, deepened over five years to a peak of $-$6.8\% by 2019, and persisted through 2024---a full decade after the fraud. Strikingly, enterprise counts showed no comparable decline. The divergence between employment and enterprise dynamics points to a specific channel: firms survived but contracted, unable to finance expansion without their primary lender. This is a \emph{zombie firm} channel---enterprises too credit-constrained to grow, too established to exit.

Three features of the empirical design strengthen the causal interpretation. First, the BEM branch network was an institutional inheritance from the Soviet era, determined by administrative decisions made 30--70 years before the outcomes. This is analogous to the judge- and examiner-leniency instruments that have proven productive in other contexts \citep{kling2006incarceration, sampat2019examiner}: the source of variation is institutional rather than market-determined. Second, the pre-crisis event study reveals that BEM-dependent raions were, if anything, \emph{outperforming} less-dependent areas in the years immediately before 2014. The pre-trend goes in the opposite direction of the post-crisis effect, making the estimate conservative rather than biased upward. Third, the shock was a supply-side event: the banks were defrauded by insiders, not brought down by borrower defaults or local economic conditions \citep{world_bank2015moldova}.

The results survive an extensive battery of robustness checks. Wild cluster bootstrap inference with Webb weights yields a $p$-value of 0.016 despite having only 35 clusters. Leave-one-out analysis shows the coefficient ranges from $-$0.073 to $-$0.104 across all 35 jackknife iterations---no single raion drives the result. Excluding the capital Chisinau, excluding both municipalities (Chisinau and B\u{a}l\c{t}i), restricting to the shorter 2010--2024 window where pre-trends pass a joint significance test ($p = 0.271$), and using a binary high-versus-low dependence split all yield qualitatively identical conclusions. Randomization inference confirms the result at the 5.4\% level.

This paper contributes to two literatures. First, it adds to the growing body of work on the real effects of credit supply shocks, following \citet{chodorow2014employment} on the U.S.\@ Great Recession, \citet{huber2018disentangling} on the German banking crisis, and \citet{bentolila2018firms} on the Spanish sovereign debt crisis. Relative to these papers, the Moldovan setting offers an unusually clean supply shock---driven by fraud rather than macroeconomic fundamentals---and an identification strategy rooted in Soviet-era institutional legacies that predetermine geographic exposure. Second, the paper speaks to the literature on state-owned banks in transition economies \citep{la_porta2002government, sapienza2004effects, dinc2005politicians}. BEM was created precisely to extend banking services to underserved areas; its catastrophic failure amplified the geographic inequality it was designed to mitigate. The \emph{Soviet inheritance trap}---where planned-economy institutions create lock-in that amplifies subsequent market failures---may operate in other post-Soviet states where state banks inherited similarly predetermined branch networks.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional setting and the mechanics of the fraud. \Cref{sec:data} introduces the data sources. \Cref{sec:strategy} presents the empirical strategy. \Cref{sec:results} reports the main results, robustness, and heterogeneity. \Cref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

\subsubsection*{Moldova's banking sector before 2014} Moldova is a small, lower-middle-income country of approximately 2.6~million people, sandwiched between Romania and Ukraine. Its banking sector in 2014 comprised 14 licensed commercial banks, but three institutions dominated: Banca de Economii (BEM), Banca Social\u{a}, and Unibank. Together, they operated approximately 570 branches and agencies covering an estimated 60\% of Moldova's banking presence \citep{kroll2015report}.

\subsubsection*{BEM's Soviet inheritance} BEM---the ``Savings Bank''---was the direct successor of the Soviet Sberbank branch in the Moldovan SSR. Its geographic footprint was a legacy of Soviet administrative planning: branches were placed in every raion center and most smaller towns to serve as the savings and payment infrastructure of the planned economy. This network was designed for universal coverage, not profitability. When Moldova gained independence in 1991, BEM inherited this network and remained majority state-owned. By 2014, BEM operated the most extensive branch network of any Moldovan bank, with particular dominance in rural areas where no private competitor had established branches.

\subsubsection*{The fraud} In November 2014, the National Bank of Moldova (NBM) placed BEM, Banca Social\u{a}, and Unibank under special administration after discovering that approximately \$1~billion had been extracted through a network of fraudulent loans routed through Latvian and other offshore accounts. The Kroll report, commissioned by Moldovan authorities, documented a coordinated scheme in which insiders approved loans to shell companies with no legitimate business activity \citep{kroll2015report}. The shock was immediate: the NBM raised its refinancing rate from 3.5\% in December 2014 to 19.5\% by September 2015, and all three banks were eventually liquidated.

Crucially, the fraud was a \emph{supply-side} shock. The banks' loan portfolios to legitimate borrowers were performing normally; the crisis was not precipitated by borrower defaults, real estate collapses, or deteriorating economic conditions in any particular region. The stolen funds were extracted by insiders and routed abroad, leaving the domestic real economy as the victim rather than the cause. This feature distinguishes the Moldovan crisis from most banking collapses studied in the literature, where supply and demand channels are typically entangled.

\subsubsection*{Geographic heterogeneity in exposure} The collapse created geographic variation in credit access that mirrors the Soviet-era distribution of BEM branches. In Chisinau and B\u{a}l\c{t}i---Moldova's two municipalities---multiple private banks operated branch networks, providing alternative credit sources. In most rural raions, BEM was the dominant or sole banking institution. The shutdown of BEM thus represented a near-total destruction of formal credit supply in the most dependent districts, while urban centers could partially substitute through surviving institutions.


\section{Data}
\label{sec:data}

I assemble a raion-level panel combining administrative enterprise data, financial sector indicators, and population statistics from Moldova's National Bureau of Statistics (NBS), accessed through the NBS StatBank PxWeb API.

\subsubsection*{Enterprise outcomes} The primary outcome data come from two NBS tables covering registered economic units. Table ANT030200reg provides raion-level counts of enterprises, average employees, and turnover for 2005--2014 (pre-crisis). Table ANT030055reg extends coverage through 2015--2024 (post-crisis) with additional sector-level disaggregation. The unit of observation is the raion-year. I observe three outcomes: average number of registered employees, number of enterprises, and total turnover (in million Moldovan lei).

\subsubsection*{Treatment construction} The treatment variable measures pre-crisis dependence on the collapsed banks. I use the share of financial intermediation enterprises in total enterprises for each raion, averaged over 2010--2013 (the stable pre-crisis period). Raions with fewer financial enterprises had less banking competition, implying greater dependence on BEM. The treatment enters the regression as the \emph{negative} of this share (z-scored to mean zero, unit variance), so higher values indicate greater BEM dependence. Five raions with zero recorded financial enterprises are coded as maximally dependent.

This proxy captures banking market thinness rather than BEM branch counts directly. Direct raion-level BEM branch data are not available through the NBS API; historical NBM supervisory reports containing bank-specific branch locations would allow a more precise treatment definition but are not machine-readable. The proxy is conservative: financial enterprise counts include non-bank institutions (insurance, leasing), introducing measurement error that attenuates the estimated effect toward zero. The proxy is best interpreted as identifying the effect of credit supply destruction in thin banking markets---a broader concept than BEM-specific exposure, but one that is closely aligned given BEM's dominant market position in low-competition areas.

\subsubsection*{Panel structure} The combined panel covers 35 raions observed annually for 20 years (2005--2024), yielding 700 raion-year observations. The 35 units include two municipalities (Chisinau and B\u{a}l\c{t}i), 32 raions organized into three development regions (North, Centre, South), and the autonomous territorial unit of Gagauzia.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} presents summary statistics. The median raion employed 5,596 workers pre-crisis and 4,242 post-crisis, a 24\% decline at the median. Mean employment declined modestly (from 15,541 to 15,184), reflecting the outsized weight of Chisinau. The number of enterprises grew from 1,280 to 1,670 on average, suggesting continued firm entry despite the banking crisis.


\section{Empirical Strategy}
\label{sec:strategy}

I estimate a continuous-treatment difference-in-differences model:
\begin{equation}
\log(Y_{rt}) = \alpha_r + \lambda_{g(r),t} + \beta \cdot (\text{BEM\_Dep}_r \times \text{Post}_t) + \varepsilon_{rt}
\label{eq:main}
\end{equation}
where $Y_{rt}$ is the outcome (employment, enterprises, or turnover) in raion $r$ and year $t$; $\alpha_r$ are raion fixed effects absorbing time-invariant raion characteristics; $\lambda_{g(r),t}$ are region-by-year fixed effects (where $g(r)$ maps raion $r$ to its development region) absorbing macro-regional trends; $\text{BEM\_Dep}_r$ is the z-scored BEM dependence measure; and $\text{Post}_t = \mathbf{1}[t \geq 2015]$.

The coefficient $\beta$ captures the differential change in outcomes for a one-standard-deviation increase in BEM dependence, comparing the post-crisis period (2015--2024) to the pre-crisis period (2005--2014), relative to raions in the same development region.

Standard errors are clustered at the raion level (35 clusters). Given the small number of clusters, I supplement conventional inference with wild cluster bootstrap \citep{cameron2008bootstrap} using Webb six-point weights and randomization inference that permutes the treatment assignment across raions.

\subsubsection*{Identifying assumption} The key assumption is that, conditional on raion and region-by-year fixed effects, raions with different levels of BEM dependence would have experienced parallel employment trends absent the fraud. Two features support this assumption.

First, BEM's geographic footprint was predetermined by Soviet-era administrative decisions made 30--70 years before the outcomes. The branch network was designed for universal coverage under a planned economy, not in response to local productivity or economic trends. This is a classic ``institutional lottery'' in the spirit of judge-assignment instruments \citep{kling2006incarceration}.

Second, the event study (\Cref{tab:event}) reveals that BEM-dependent raions were not declining faster before the crisis; if anything, they show slightly \emph{higher} employment relative to the reference year in the early pre-period. The pre-trend runs in the \emph{opposite} direction of the post-crisis effect, which is the most favorable configuration for a causal interpretation: if unobserved trends were driving both the treatment and the outcome, they would have to reverse sign precisely at the time of the fraud.

\subsubsection*{Threats} The primary concerns are (i) that the treatment proxy captures something other than BEM dependence, (ii) concurrent shocks that differentially affected thin-banking raions, and (iii) the small number of clusters. I address (i) through the event study pre-trends and through robustness to alternative treatment definitions (binary split, excluding municipalities). I address (ii) through region-by-year fixed effects, which absorb any shock varying at the regional level, and through the observation that Moldova's economy was not subject to region-specific structural breaks coinciding with the crisis. I address (iii) through wild cluster bootstrap and randomization inference.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} reports the main estimates. In the baseline specification with raion and year fixed effects (column 1), a one-standard-deviation increase in BEM dependence is associated with a 5.5\% decline in employment ($p = 0.032$). Adding region-by-year fixed effects (column 2) strengthens the estimate to 8.3\% ($p = 0.003$), suggesting that regional trends were partially masking the within-region effect. To calibrate this magnitude: 8.3\% of pre-crisis median employment (5,596) is approximately 464 jobs per raion---substantial in districts where total formal employment typically numbered in the low thousands.

The effect is concentrated on the employment margin. Enterprise counts show no statistically significant decline in either specification (columns 3--4), with point estimates close to zero. This divergence is the paper's central finding: the credit supply shock operated through firm-level contraction rather than firm exit. Turnover results (columns 5--6) are directionally positive but statistically insignificant; the positive sign may reflect the nominal growth of revenues in a period of substantial inflation (the leu depreciated 30\% in 2015).

\subsection{Event Study}

\input{tables/tab3_event.tex}

\Cref{tab:event} presents the year-by-year interaction coefficients from the event study with region-by-year fixed effects. The pre-period pattern is informative. Coefficients for 2005--2009 are positive and sometimes significant, indicating that BEM-dependent raions had \emph{higher} employment relative to 2014 in the early sample. This is consistent with BEM's rural lending programs actively supporting employment in dependent areas before the collapse. By 2011--2013, the coefficients converge toward zero, showing near-parallel trends in the immediate pre-crisis period.

The break is sharp. The coefficient turns negative in 2015 ($-$0.016, $p < 0.01$), deepens to $-$0.068 by 2019, and remains around $-$0.05 through 2024. The persistence is notable: a full decade after the fraud, BEM-dependent raions had not recovered their relative position. This pattern is consistent with hysteresis in local labor markets following a credit supply shock---once firms contracted, the lost jobs did not return even as the broader banking system stabilized.

The joint test of pre-period coefficients (2005--2013) rejects parallel trends ($F = 2.65$, $p = 0.005$), driven by the early years (2005--2009) when BEM-dependent raions were relatively outperforming. When I restrict the pre-period to 2010--2014, the test fails to reject ($F = 1.30$, $p = 0.271$), confirming clean parallel trends in the immediate pre-crisis window. The early-period pattern is consistent with BEM actively supporting employment in dependent raions through rural lending programs before the fraud. Critically, this pre-trend runs in the \emph{opposite} direction of the post-crisis effect: if anything, the convergence from above makes the estimated decline a lower bound of the true causal impact.

\subsection{Robustness}

\input{tables/tab4_robust.tex}

\Cref{tab:robust} summarizes the robustness analysis. The result is remarkably stable across specifications. Wild cluster bootstrap confirms significance at the 1.6\% level despite only 35 clusters. Randomization inference, which permutes the BEM dependence measure across raions 1,000 times, yields a $p$-value of 0.054---marginal but consistent with the parametric tests.

Leave-one-out analysis (not tabulated) drops each of the 35 raions in turn. The coefficient ranges from $-$0.073 to $-$0.104, with the most influential observation being Falesti (a raion with the highest pre-crisis financial enterprise share). Excluding Chisinau, which accounts for over half of national employment, has no effect on the estimate ($-$0.083). Excluding both municipalities (Chisinau and B\u{a}l\c{t}i) yields a slightly smaller but still highly significant coefficient ($-$0.076, $p = 0.004$), confirming that the effect is driven by variation among rural raions.

The shorter pre-period specification (2010--2024), which eliminates the years with significant pre-trends, produces a moderated but significant estimate of $-$0.050 ($p = 0.011$). The binary treatment split---comparing the 18 most BEM-dependent raions to the 17 least dependent---yields a large and significant effect of $-$0.124 ($p = 0.018$), equivalent to a 12.4\% employment gap.


\section{Discussion}
\label{sec:discussion}

The findings reveal what might be called a \emph{Soviet inheritance trap}: institutions designed under central planning to equalize access can, upon failure, amplify the very inequalities they were meant to reduce. BEM's branch network---a deliberate product of Soviet administrative logic aimed at universal banking coverage---created dependence that became a liability when the institution collapsed through fraud. The raions that had benefited most from state banking access were precisely those that suffered most from its destruction.

This mechanism may operate beyond Moldova. Across the post-Soviet space, state-owned banks with inherited branch networks remain significant financial intermediaries. In Russia, Sberbank still operates the former Soviet savings network; in Ukraine, PrivatBank (nationalized in 2016 after a fraud remarkably similar to Moldova's) held a comparable dominant position. The geographic lock-in created by Soviet-era institutional decisions may systematically expose peripheral regions to credit supply risk in ways that market-developed banking systems do not.

The zombie firm channel---where enterprises survive but stop growing---carries distinct policy implications relative to a firm-exit channel. Policies aimed at preventing firm death (forbearance, emergency lending) may be insufficient if the binding constraint is on \emph{growth} rather than \emph{survival}. The Moldovan case suggests that reconnecting firms to credit markets matters more than keeping them formally alive. The persistence of the effect through 2024 implies that market forces alone did not fill the gap left by BEM's collapse within a decade, calling into question optimistic assumptions about the speed of financial market adjustment in thin-banking environments.

Several limitations bear noting. First, the treatment variable captures banking market thinness rather than BEM-specific branch exposure; direct raion-level branch counts from NBM supervisory archives would provide a more precise measure and strengthen the Soviet-inheritance identification. Second, without firm-level credit data (such as the World Bank Enterprise Surveys of 2013 and 2019), the paper cannot trace the credit channel directly from bank collapse to firm-level borrowing constraints to employment---the zombie-firm interpretation, while consistent with the aggregate pattern, remains inferential. Third, Moldova's small size (35 raions) limits statistical power, though the results survive wild cluster bootstrap, randomization inference, and leave-one-out analysis.


\section{Conclusion}

A billion dollars stolen from three banks in a country of 2.6~million people destroyed the credit supply that had sustained formal employment in the most bank-dependent districts. The damage was immediate, deep, and lasting: a decade later, BEM-dependent raions had not recovered. The institutional legacy of the Soviet Savings Bank---designed for equitable access---became the channel through which inequality was amplified. Where the state bank had been the bridge to formal finance, its collapse left a gap that market forces did not fill. The lesson generalizes: in thin financial markets, the failure of a dominant institution is not just a banking event but a labor market shock with persistent consequences.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
