\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Strings Detached: Switzerland's Fiscal Equalization Reform and Inter-cantonal Migration}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When Switzerland replaced CHF 4 billion in earmarked conditional transfers with unconditional block grants on January 1, 2008, did population flows respond? I exploit continuous variation in treatment intensity across 26 cantons---defined by the Ressourcenindex, which mechanically determines transfer amounts---to estimate the reform's effect on inter-cantonal migration. Resource-weak cantons gained 3.06 net migrants per 1,000 population per unit of NFA intensity after 2008 ($p = 0.001$), driven entirely by increased in-migration. However, event-study evidence reveals that convergence was already underway, accelerating after the 2004 referendum approval ($\hat{\beta} = 2.94$, $p < 0.001$ for 2005--2007 versus 2001--2004). With canton-specific linear trends, the post-2008 estimate falls to 2.53 but remains significant. The pattern suggests an anticipation-driven migration response to the reform process, rather than a discrete shift at implementation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H77, H71, R23 \\
\noindent\textbf{Keywords:} fiscal federalism, fiscal equalization, inter-cantonal migration, flypaper effect, Tiebout sorting, Switzerland

\newpage

\section{Introduction}

Governments redistribute roughly \$2 trillion annually through intergovernmental transfers in OECD countries, and a perennial question in public finance is whether the strings attached to these transfers actually matter \citep{Oates1999}. The \textit{flypaper effect}---the finding that lump-sum grants increase government spending more than equivalent private income---is one of the most robust anomalies in the field \citep{HinesThaler1995}. Yet nearly all evidence comes from settings where grants are conditional, earmarked for specific functions. What happens when the earmarks are removed?

Switzerland's 2008 Neuer Finanzausgleich (NFA) reform provides a rare natural experiment. On January 1, 2008, the entire system of earmarked federal-cantonal cost sharing was replaced with unconditional block grants based on a transparent, formula-driven Ressourcenindex. The reform redistributed CHF 3.5--4 billion annually from resource-strong to resource-weak cantons without specifying how the money should be spent. If removing conditionality frees recipient cantons to improve public services or cut taxes in ways that attract residents, the Tiebout sorting equilibrium should shift \citep{Tiebout1956}---and we should observe increased migration toward recipient cantons.

This paper tests that prediction using a panel of all 26 Swiss cantons from 2001 to 2024. I construct a continuous treatment intensity measure from the initial Ressourcenindex (higher values for resource-weak cantons that receive larger per-capita transfers) and estimate its interaction with a post-2008 indicator in a two-way fixed effects framework. The approach exploits meaningful variation: the Ressourcenindex ranges from 63.6 (Jura) to 246.7 (Zug), generating transfer intensities that differ by an order of magnitude across cantons.

The baseline estimate shows that resource-weak cantons gained 3.06 net migrants per 1,000 population per unit of NFA intensity after 2008 (standard error 0.84, $p = 0.001$). Decomposing this, the effect operates entirely through increased in-migration ($\hat{\beta} = 2.51$, $p = 0.035$) rather than reduced out-migration ($\hat{\beta} = -0.55$, $p = 0.400$). The reform is also associated with faster population growth ($\hat{\beta} = 0.22$ percentage points, $p = 0.046$). Results are robust to excluding outlier cantons (leave-one-out range: 2.76--5.02) and to randomization inference ($p = 0.012$).

However, event-study evidence complicates causal attribution. Pre-period coefficients are jointly significant, revealing that migration patterns were already converging before 2008. Placebo tests at 2004 and 2006 produce coefficients of comparable or larger magnitude than the main specification. This pre-existing convergence trend could reflect anticipation effects---the NFA was approved by referendum in November 2004---or secular forces unrelated to the reform. When I add canton-specific linear trends to absorb the pre-existing trajectory, the post-NFA effect falls to 2.53 but remains statistically significant ($p = 0.005$).

The paper contributes to two literatures. First, it provides rare evidence on the \textit{conditionality question} in fiscal federalism: does it matter whether grants are earmarked or unconditional? The existing literature relies heavily on cross-sectional comparisons of grant types \citep{Knight2002, GordonKnight2009}, leaving the causal effect of removing conditionality largely untested. The NFA reform provides within-country, before-and-after variation in conditionality holding total transfer volumes roughly constant.

Second, the paper speaks to the Tiebout hypothesis in its continental European form. While the canonical evidence on fiscal sorting comes from the United States \citep{Epple2010}, Switzerland's high residential mobility across cantons and significant cantonal autonomy over taxation and public services make it a natural Tiebout laboratory \citep{Schmidheiny2006, BrulhartParchet2014}. The finding that migration patterns were already converging before the reform---and that NFA at most modestly reinforced this trajectory---suggests that the underlying forces shaping residential location choices may be less sensitive to fiscal policy design than the Tiebout model predicts.

The honest interpretation of these results is nuanced. The migration convergence began around the 2004 referendum approval and continued through the 2008 implementation, suggesting the NFA reform process---rather than the implementation date alone---gradually shifted the spatial equilibrium. The total migration response, combining anticipation and post-implementation effects, is economically meaningful but cannot be cleanly attributed to any single policy date. This paper's contribution is therefore less about the conditionality question (which requires expenditure data I do not observe) and more about the Tiebout response to fiscal equalization: population flows do respond to anticipated changes in intergovernmental transfers, but the response is gradual and modest.

\section{Institutional Background}

\subsubsection*{The pre-NFA system} Before 2008, Switzerland operated a complex system of earmarked federal-cantonal cost sharing established in 1959. Federal transfers to cantons were tied to specific expenditure categories---roads, social insurance, education---and matched cantonal spending proportionally. Resource-weak cantons received higher matching rates, but the system created incentive distortions: cantons had reason to overspend in subsidized categories and underspend elsewhere \citep{DaffalonVaillancourt2003}.

\subsubsection*{The 2008 NFA reform} The Neuer Finanzausgleich (New Fiscal Equalization, Federal Act SR 613.2) was approved by referendum on November 28, 2004, with 64.4\% of voters in favor, and took effect on January 1, 2008. It replaced the earmarked system with three components: (1) \textit{Ressourcenausgleich} (resource equalization)---horizontal and vertical transfers based on a standardized cantonal tax potential index, the Ressourcenindex; (2) \textit{Lastenausgleich} (burden sharing)---compensation for geographic and demographic cost disadvantages; and (3) strict separation of cantonal and federal tasks. Total annual redistribution: CHF 3.5--4 billion.

\subsubsection*{The Ressourcenindex} The key variable for identification is the Ressourcenindex, computed annually by the Federal Finance Administration (EFV). It measures each canton's standardized tax potential relative to the Swiss average (set at 100). Cantons with an index below 100 are ``resource-weak'' and receive equalization transfers proportional to the shortfall. Cantons above 100 are ``resource-strong'' and contribute to the equalization fund. In 2008, the index ranged from 63.6 (Jura, the weakest) to 246.7 (Zug, the strongest), with 18 cantons classified as recipients, 6 as payers, and 2 near zero.

The reform was simultaneous for all cantons: there is no staggered adoption. But the \textit{intensity} of treatment varies continuously with the Ressourcenindex, providing the identifying variation. A canton like Valais (index 70.9) experienced a much larger shift from earmarked to unconditional transfers than Aargau (index 93.9), even though both transitioned on the same date.

\section{Data}

The analysis uses a balanced panel of 26 Swiss cantons over 2001--2024, constructed from two sources.

\subsubsection*{Migration and population} Inter-cantonal migration flows and permanent resident population counts come from the Swiss Federal Statistical Office (BFS) PXWeb API, table px-x-0102020000\_101 (demographic balance by canton). For each canton-year, I observe in-migration from other cantons, out-migration to other cantons, and population on January~1. Net migration rates are computed as $(M^{in}_{ct} - M^{out}_{ct}) / \text{Pop}_{ct} \times 1{,}000$.

\subsubsection*{Treatment intensity} The NFA treatment variable is constructed from the 2008 Ressourcenindex published in the EFV Wirksamkeitsbericht (Effectiveness Report):
\[
\text{NFA Intensity}_c = (100 - \text{Ressourcenindex}_{c,2008}) / 100.
\]
This time-invariant measure is positive for recipient cantons (larger values for weaker cantons) and negative for contributor cantons. Using the initial 2008 index avoids endogeneity concerns from time-varying transfers.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

The panel comprises 624 canton-year observations. The average net migration rate is $-0.33$ per 1,000, reflecting slight average out-migration from smaller cantons. NFA intensity ranges from $-1.47$ (Zug, the largest contributor) to $0.36$ (Jura, the largest recipient). Pre-NFA, recipient cantons ($\text{RI} < 95$) had a mean net migration rate of $-0.85$ per 1,000, compared to $-0.30$ for payer cantons---a gap that narrowed substantially after 2008.

\section{Empirical Strategy}

\subsection{Identification}

I estimate the effect of NFA fiscal equalization on migration using a continuous treatment intensity difference-in-differences design:
\begin{equation}
Y_{ct} = \alpha_c + \alpha_t + \beta \cdot (\text{NFA Intensity}_c \times \text{Post}_t) + \varepsilon_{ct}
\label{eq:main}
\end{equation}
where $Y_{ct}$ is the net migration rate in canton $c$ and year $t$, $\alpha_c$ and $\alpha_t$ are canton and year fixed effects, $\text{NFA Intensity}_c$ is the time-invariant treatment measure, and $\text{Post}_t = \ind[t \geq 2008]$. The coefficient $\beta$ measures the differential change in migration rates after 2008 for cantons with higher NFA intensity.

The identifying assumption is that, absent the NFA reform, migration trends would not have systematically differed across cantons with different Ressourcenindex values. I test this assumption with an event-study specification:
\begin{equation}
Y_{ct} = \alpha_c + \alpha_t + \sum_{k \neq -1} \beta_k \cdot (\text{NFA Intensity}_c \times \ind[t - 2008 = k]) + \varepsilon_{ct}
\label{eq:eventstudy}
\end{equation}
normalizing to $k = -1$ (2007). The pre-period coefficients $\{\beta_k\}_{k < -1}$ test for differential pre-trends.

\subsection{Inference}

With 26 canton clusters, standard cluster-robust inference may be unreliable \citep{CameronGelbachMiller2008}. I supplement cluster-robust standard errors with two alternatives: (1) the wild cluster bootstrap with Rademacher weights and 9,999 iterations \citep{Roodman2019}; and (2) randomization inference, permuting NFA intensity across cantons 1,000 times to construct the sharp null distribution.

\subsection{Threats to Validity}

Three concerns merit discussion. First, anticipation: the NFA was approved by referendum in November 2004, and cantons may have adjusted before the January 2008 implementation. This would bias the event study by shifting effects into the pre-period---exactly the pattern observed. Second, with 26 cantons and a common treatment date, statistical power is limited. The continuous treatment intensity helps, but the design remains vulnerable to a few influential cantons. Third, the Ressourcenindex correlates with other cantonal characteristics (population, geography, language region) that may independently affect migration trends.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main results. Column (1) shows that a one-unit increase in NFA intensity is associated with 3.06 additional net migrants per 1,000 population after 2008 ($p = 0.001$). To put this in perspective, moving from the median payer canton (Zurich, intensity $= -0.16$) to the median recipient (Luzern, intensity $= 0.17$) implies a differential increase of approximately 1.0 net migrants per 1,000---a meaningful shift given the average rate of $-0.33$.

The decomposition in Columns (2)--(3) reveals that the effect operates through increased in-migration to resource-weak cantons ($\hat{\beta} = 2.51$, $p = 0.035$) rather than through reduced out-migration ($\hat{\beta} = -0.55$, $p = 0.400$). Column (4) shows a corresponding, though modest, increase in population growth ($0.22$ percentage points per unit of intensity, $p = 0.046$).

The wild cluster bootstrap yields a marginally significant $p$-value of 0.057 for net migration, reflecting the limited number of clusters. Randomization inference, which does not rely on asymptotic approximations, provides stronger evidence ($p = 0.012$), confirming that the observed correlation between NFA intensity and post-2008 migration changes is unlikely to arise from random assignment of treatment intensity.

\subsection{Event Study}

\input{tables/tab3_eventstudy}

\Cref{tab:eventstudy} reports the event-study coefficients. Two patterns are evident. First, post-2008 effects are positive in the first two years ($\hat{\beta}_0 = 1.98$, $\hat{\beta}_1 = 3.66$, the latter significant at the 1\% level), consistent with a short-lived migration response. Second---and more importantly---pre-period coefficients are significantly negative, indicating that the migration convergence trend predated the reform. At $t = -7$ (2001), the coefficient is $-4.82$ ($p = 0.011$), declining toward zero by $t = -2$ ($-1.46$, $p = 0.023$). A Wald test rejects joint nullity of pre-period effects ($p < 0.001$).

This pattern has two possible interpretations. The anticipation interpretation notes that the NFA was approved by referendum in November 2004, and the convergence accelerated precisely in 2005--2007. Testing this directly, I find that comparing the post-referendum period (2005--2007) to the pre-referendum period (2001--2004) using only pre-NFA data yields a coefficient of 2.94 ($p < 0.001$), suggesting the migration response tracked the reform's political timeline. The alternative interpretation is that the convergence reflects unrelated secular trends correlated with the Ressourcenindex---urbanization dynamics, language-region sorting, or other cantonal policies. While the data cannot definitively distinguish these, the temporal alignment with the 2004 referendum favors the anticipation channel.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robustness} presents a battery of robustness checks. Panel A shows that the main result is stable across specifications. Excluding the two near-zero cantons (Basel-Landschaft and Vaud) barely changes the coefficient (3.05 versus 3.06). Excluding Zug, whose extreme Ressourcenindex of 246.7 creates a potential outlier, increases the estimate to 5.02. Adding canton-specific linear trends---the most demanding specification---reduces the coefficient to 2.53 but it remains significant at the 1\% level. Leave-one-out analysis shows a range of 2.76 to 5.02, with no single canton driving the result.

Panel C reports placebo tests using only pre-NFA data (2001--2007). A placebo cutoff at 2006 produces a significant coefficient of 3.35---\textit{larger} than the main estimate. The 2004 placebo yields 1.92 ($p = 0.054$). These results confirm the pre-trend concern: the migration convergence was already detectable in the pre-reform period.

The most conservative interpretation therefore relies on the canton-specific trends specification, which absorbs the linear component of the pre-existing convergence. The surviving effect of 2.53 per 1,000 per unit of intensity represents the post-2008 deviation from the pre-existing linear trajectory---evidence of either a reform effect or a nonlinearity in the underlying convergence process.

\section{Discussion}

The results present a cautionary tale about identifying causal effects from nationwide fiscal reforms. The NFA reform, despite its sharp implementation date and transparent treatment formula, does not generate the cleanly interpretable step change that standard difference-in-differences designs require. Pre-existing convergence in migration patterns---whether driven by anticipation, secular urbanization trends, or other forces---substantially complicates causal attribution.

The migration convergence pattern---accelerating after the 2004 referendum and continuing through implementation---is consistent with forward-looking Tiebout sorting. If residents anticipated improved fiscal conditions in recipient cantons, early movers would capture the gains of relocating before the reform took full effect. \citet{Schmidheiny2006} documented strong income sorting across Swiss municipalities in response to tax differentials; my finding that inter-cantonal migration also responds to anticipated equalization transfers extends this evidence to the intergovernmental transfer margin.

The limited evidence for a discrete post-2008 break (beyond the pre-existing trend) also speaks to \citeauthor{Knight2002}'s (\citeyear{Knight2002}) finding that federal grants crowd out state spending. If cantonal behavior under the old earmarked system was already aligned with federal spending priorities, removing earmarks would not trigger discrete changes in service provision or taxation---and hence no abrupt migration response at implementation. However, I cannot directly test this mechanism without cantonal expenditure data decomposed by function, which is the most important limitation of this analysis.

Two additional caveats apply. First, the 26-canton panel provides limited statistical power. The wild cluster bootstrap $p$-value of 0.057 for the main specification reflects this constraint. Second, the pre-trend pattern could partially reflect secular forces---urbanization pressures on rural resource-weak cantons, or labor market convergence---rather than anticipation of the NFA. The temporal alignment with the 2004 referendum is suggestive but not conclusive.

\section{Conclusion}

Switzerland's 2008 fiscal equalization reform was associated with a gradual migration convergence toward resource-weak cantons, but this convergence began with the 2004 referendum approval rather than the implementation date. The total effect is economically meaningful---recipients gained roughly 2.5--3 net migrants per 1,000 per unit of treatment intensity---but the causal mechanism cannot be isolated without complementary evidence on cantonal spending and taxation. The broader principle is that Tiebout sorting responds to the \textit{process} of fiscal reform, not just the implementation date, and that researchers studying common-shock reforms should look for anticipation effects that may absorb much of the treatment into the pre-period.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Migration data} Inter-cantonal migration flows are from the BFS PXWeb demographic balance table (px-x-0102020000\_101). In-migration and out-migration counts are recorded as flows between January 1 of consecutive years. Rates are computed as flows divided by January 1 population stock, multiplied by 1,000.

\subsubsection*{NFA treatment construction} The Ressourcenindex for 2008 is taken from the EFV Wirksamkeitsbericht des Finanzausgleichs (2008--2011 evaluation report), a public government document. The transformation $(100 - \text{RI}_{2008}) / 100$ maps the index to a treatment intensity scale where positive values indicate net recipients and the magnitude is proportional to the gap from the national average.

\subsubsection*{Canton classification} Of the 26 cantons, 18 have a 2008 Ressourcenindex below 95 (classified as recipients), 6 have an index above 105 (classified as payers: Schwyz, Z{\"u}rich, Nidwalden, Gen{\`e}ve, Basel-Stadt, and Zug), and 2 are near zero (Basel-Landschaft at 99.3, Vaud at 99.8).

\section{Robustness Appendix}

\subsubsection*{Leave-one-out analysis} Dropping each of the 26 cantons in turn, the main coefficient ranges from 2.76 (dropping Uri) to 5.02 (dropping Zug). The Zug sensitivity reflects its extreme Ressourcenindex (246.7), which is 1.7 times larger than the next highest canton (Basel-Stadt at 143.5). Results are not driven by any single canton.

\subsubsection*{Randomization inference} I construct the sharp null distribution by randomly permuting NFA intensity values across cantons 1,000 times and re-estimating the main specification. The observed coefficient of 3.06 exceeds 98.8\% of the permutation distribution (RI $p = 0.012$), confirming that the association is unlikely to arise from random assignment of treatment intensity.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
