\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Measurement Effect: How Minimum Energy Efficiency Standards Revealed England's Hidden Housing Problem}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
England's 2018 Minimum Energy Efficiency Standards prohibited letting properties rated below EPC band E. We exploit cross-Local Authority variation in pre-regulation rental intensity to test whether MEES reduced substandard housing or merely revealed its true extent. Contrary to the regulation's intent, Local Authorities with higher rental shares experienced a \textit{relative increase} in F+G-rated EPC shares post-MEES ($\beta = 0.045$, $p < 0.001$), driven by a 58 percent surge in rental EPC lodgements as landlords were forced to assess previously unrated properties. A placebo test on the unregulated D band confirms zero differential response. The regulation produced a ``measurement effect'': by mandating assessment, MEES first \textit{exposed} the scale of energy poverty in the rental sector before any improvements occurred.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q48, R31, K32 \\
\noindent\textbf{Keywords:} energy efficiency, building regulations, EPC, bunching, rental housing, MEES

\newpage

\section{Introduction}

In 2018, over one million privately rented homes in England failed to meet the government's minimum energy efficiency threshold. These ``F and G rated'' properties---with Energy Performance Certificate scores below 39 out of 100---cost their occupants an estimated \pounds 800 per year more in energy bills than comparable E-rated homes \citep{decc2015impact}. The regulation designed to eliminate this stock, the Minimum Energy Efficiency Standards (MEES), prohibited landlords from granting new tenancies for substandard properties starting April 2018, extending to all existing tenancies by April 2020. On the surface, the policy worked: England's aggregate F+G share among assessed properties fell from 6.6 percent in 2016--2017 to 1.3 percent by 2025.

But official statistics overstate the improvement. A regulation that mandates energy assessment as a precondition for rental activity does not merely \textit{change} energy efficiency---it \textit{measures} it. Before MEES, many landlords had no current EPC for their property. The regulation forced a compliance-driven wave of assessments that brought previously invisible properties into the statistical record. If this newly assessed stock was disproportionately old and energy-inefficient, the aggregate statistics would simultaneously reflect genuine improvements \textit{and} the first-time measurement of a problem that had always existed.

This paper identifies and quantifies what we term the ``measurement effect'' of MEES: the degree to which mandatory assessment revealed---rather than remedied---substandard energy performance in England's rental sector. We exploit cross-Local Authority variation in rental market intensity as a continuous treatment measure in a difference-in-differences framework. Local Authorities where rental properties constituted a larger pre-MEES share of EPC lodgements experienced greater exposure to the compliance-driven assessment surge. If MEES primarily drove genuine upgrades, high-rental LAs should show larger declines in their F+G share; if the measurement effect dominates, these LAs should show attenuated declines or relative increases as newly assessed properties enter the denominator.

Our main finding is striking. While the national F+G share declined everywhere (absorbed by quarter fixed effects), high-rental LAs experienced an \textit{attenuated} decline: a one-standard-deviation increase in pre-MEES rental intensity is associated with a 0.96 percentage point \textit{relative increase} in the F+G share ($\beta = 0.045$, SE = 0.007, $p < 0.001$). This result is robust to weighting by LA size ($\beta = 0.051$, $p < 0.001$), excluding London ($\beta = 0.041$, $p < 0.001$), and using binary rather than continuous treatment ($\beta = 0.010$, $p < 0.001$). A decomposition into three MEES phases reveals that the measurement effect concentrated in the ``new tenancies'' period (2018Q2--2020Q1, $\beta = 0.041$, $p < 0.001$) and persisted into the ``all tenancies'' period ($\beta = 0.012$, $p = 0.006$), consistent with a rolling wave of first-time assessments.

Three pieces of evidence support the measurement-effect interpretation. First, a placebo test on the D band---rated 55--68, well above the MEES threshold---shows zero differential response ($\beta = 0.004$, $p = 0.85$), confirming that the pattern is specific to the regulated margin. Second, the adjacent E band shows a significant \textit{negative} response ($\beta = -0.043$, $p < 0.01$), consistent with properties being upgraded from below-threshold to just above. Third, rental EPC lodgements surged 58 percent post-MEES (from 91,000 to 144,000 per quarter), while sale lodgements remained stable, directly documenting the compliance-driven assessment wave.

This paper contributes to three literatures. First, we add to the growing evidence on the ``energy efficiency gap'' \citep{allcott2014short, gerarden2017assessing} by showing that measurement distortions in administrative data may overstate the realized gains from building energy regulations. \citet{levinson2016energy} found that California building codes saved only one-third to one-half of their predicted energy savings; our finding suggests a complementary channel through which aggregate statistics overstate improvement. Second, we contribute to the literature on regulatory compliance and information revelation \citep{kleven2016bunching, houde2018consumer, costa2014energy} by identifying a setting where regulation simultaneously changes behavior \textit{and} the measurement process. \citet{collins2018bunching} documented bunching at Ireland's BER threshold; we show that the UK's MEES created a qualitatively different phenomenon---a compliance-driven census of the rental stock---that preceded behavioral responses. Third, we inform the policy design of building efficiency standards \citep{fowlie2018energy, christensen2024housing} by identifying a specific channel---mandatory assessment as a compliance mechanism---that generates information externalities beyond the regulation's direct effect on housing quality.

We note a key limitation: the individual-level EPC Register requires authenticated access not available for this study, preventing us from implementing the score-level bunching analysis that would directly test for assessor manipulation at the threshold \citep{kleven2013using, cattaneo2020manipulation}. Our aggregate band-level approach identifies the measurement effect but cannot distinguish between strategic scoring and genuine upgrading among the assessed stock. This remains an important avenue for future research with microdata access.

\section{Institutional Background}

\subsubsection*{Energy Performance Certificates} Since 2007, all domestic properties in England and Wales marketed for sale or rent must have a valid EPC. The certificate rates properties on a scale from 1 (worst) to 100 (best), mapped onto bands A (92--100) through G (1--20). EPCs are produced by accredited Domestic Energy Assessors using the government's Reduced Data Standard Assessment Procedure (RdSAP), which estimates energy use from dwelling characteristics including wall type, insulation, heating system, glazing, and floor area \citep{beis2021epc}. Assessors exercise judgment on observable characteristics, creating scope for measurement variation even for identical properties.

\subsubsection*{Minimum Energy Efficiency Standards} The Energy Efficiency (Private Rented Property) Regulations 2015 established MEES for England and Wales. From April 1, 2018, landlords could not grant new tenancies for properties rated F or G (score below 39). From April 1, 2020, the prohibition extended to all existing tenancies. Landlords of substandard properties could either invest in upgrades, apply for an exemption (if improvements up to a \pounds 3,500 cost cap would not raise the property to band E), or withdraw the property from the rental market entirely \citep{decc2015impact}.

\subsubsection*{The Compliance Mechanism} Crucially, MEES operates through the EPC system: compliance requires a valid EPC showing band E or above. Before MEES, many landlords---particularly those with long-term tenants or informal arrangements---had no current EPC. The regulation thus created a dual mandate: it prohibited substandard lettings \textit{and} required the information system that could detect non-compliance. This design feature is the source of the measurement effect we identify.

\subsubsection*{Three Policy Phases} We distinguish three periods: the \textit{anticipatory period} (2015--2018Q1), when regulations were announced but not enforced; the \textit{new tenancies period} (2018Q2--2020Q1), when only new and renewed tenancies required compliance; and the \textit{all tenancies period} (2020Q2 onward), when all existing tenancies were covered.

\section{Data}

We draw on two administrative datasets from GOV.UK's Energy Performance of Buildings Certificates statistics, updated through Q4 2025.

\subsubsection*{Live Table D1} Contains quarterly EPC lodgement counts by band (A through G) for each of 346 Local Authorities in England and Wales, from 2008Q4 through 2025Q4. We observe 23,084 LA-quarter cells containing 31.2 million rated lodgements.

\subsubsection*{Live Table D4B} Contains quarterly EPC lodgement counts by transaction type---new dwelling, marketed sale, non-marketed sale, rental, and other categories---for each LA from 2012Q4 through 2025Q4 (17,660 LA-quarter cells). This table identifies which lodgements were for rental versus sale purposes.

\subsubsection*{Panel Construction} We merge D1 band counts with D4B transaction types at the LA-quarter level. Our key outcome is the F+G share: the fraction of all rated EPC lodgements in a given LA-quarter that received an F or G rating. Our treatment intensity measure is the pre-MEES rental share: the fraction of lodgements classified as ``Rental'' in D4B during 2012--2017, computed at the LA level and held fixed as a baseline characteristic.

We restrict to England (324 LAs) and drop LA-quarters with fewer than 10 rated lodgements, yielding an analysis panel of 21,440 observations across 317 LAs and 69 quarters.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Identification}

MEES regulates rental properties but not sales. Local Authorities differ in their baseline rental intensity: London boroughs and university towns have rental shares exceeding 50 percent, while rural districts may be below 20 percent. We exploit this variation as a continuous treatment intensity measure.

The identifying assumption is that, absent MEES, the evolution of F+G shares would have been parallel across LAs with different rental intensities. While the \textit{level} of F+G may differ (absorbed by LA fixed effects) and the \textit{national trend} may shift (absorbed by quarter fixed effects), the differential evolution should be zero in the absence of MEES.

\subsection{Estimation}

Our main specification is:
\begin{equation}
\text{FG}_{it} = \alpha_i + \gamma_t + \beta \cdot (\text{PostMEES}_t \times \text{RentalIntensity}_i) + \varepsilon_{it}
\label{eq:main}
\end{equation}
where $\text{FG}_{it}$ is the F+G share of rated lodgements in LA $i$ in quarter $t$; $\alpha_i$ are LA fixed effects; $\gamma_t$ are quarter fixed effects; $\text{PostMEES}_t$ is an indicator for 2018Q2 onward; and $\text{RentalIntensity}_i$ is the pre-MEES (2012--2017) rental share of LA $i$'s lodgements. Standard errors are clustered at the LA level (317 clusters).

If MEES successfully reduces the F+G share of rental properties, $\beta < 0$: high-rental LAs should see larger declines. If instead the measurement effect dominates---MEES forces newly assessed, disproportionately substandard rental properties into the data---then $\beta > 0$.

We extend the baseline to decompose the post-period into three MEES phases:
\begin{equation}
\text{FG}_{it} = \alpha_i + \gamma_t + \sum_{p \in \{A, N, E\}} \beta_p \cdot (\text{Phase}_{pt} \times \text{RentalIntensity}_i) + \varepsilon_{it}
\end{equation}
where the phases are Anticipatory ($A$: 2015--2018Q1), New tenancies ($N$: 2018Q2--2020Q1), and all Existing tenancies ($E$: 2020Q2+).

\subsection{Threats to Validity}

The main threat is that LAs with high rental shares differ from low-rental LAs in ways that independently affect EPC quality trends. We address this through the D-band placebo (Section 5.2), London exclusion (Section 5.3), and by documenting the rental lodgement surge directly (Section 5.4).

A second concern is that rental intensity itself may respond to MEES---landlords may withdraw substandard properties, reducing the rental share. This would bias $\beta$ toward zero (attenuation), making our positive estimate a lower bound on the measurement effect.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main results. Column (1) shows the baseline specification: the coefficient on PostMEES $\times$ RentalIntensity is 0.045 ($p < 0.001$). At the mean pre-MEES rental share (0.288), this implies a 1.3 percentage point relative increase in the F+G share---roughly 20 percent of the pre-MEES mean F+G share (0.066). To be precise: all LAs experienced F+G declines (captured by the quarter fixed effects), but high-rental LAs experienced an \textit{attenuated} decline relative to low-rental LAs, consistent with the measurement effect partially offsetting genuine improvements in these areas.

Column (2) decomposes the effect by MEES phase. The anticipatory period shows a small, insignificant negative coefficient ($-0.006$, $p = 0.40$), consistent with no pre-trend. The new-tenancies phase generates the largest positive coefficient ($0.041$, $p < 0.001$), precisely when the compliance-driven assessment surge was most intense. The all-tenancies extension adds a further significant positive effect ($0.012$, $p = 0.006$), consistent with a second wave of first-time assessments reaching existing tenancies.

Column (3) uses tercile indicators: high-rental LAs saw a 1.06 percentage point relative increase compared to low-rental LAs ($p < 0.001$); the medium-rental effect is 0.40 percentage points ($p = 0.054$). Column (4) weights by total rated EPCs and confirms the baseline ($\beta = 0.051$, $p < 0.001$).

\input{tables/tab2_main}

\subsection{Placebo and Mechanism Tests}

\Cref{tab:robust} reports diagnostic tests. Column (1) shows the D-band placebo: the coefficient on PostMEES $\times$ RentalIntensity for D-band share (scores 55--68, well above the MEES threshold) is 0.004 with a standard error of 0.022---a precise zero ($p = 0.85$). If our main result were driven by compositional changes unrelated to the E/F threshold, the D band should respond similarly. It does not.

Column (2) shows that the adjacent E band responded in the opposite direction: $-0.043$ ($p < 0.01$). This is consistent with properties being upgraded from F/G to E---the intended policy effect---but appearing in the E-band statistics rather than disappearing from the F/G statistics, because the newly assessed F/G properties are simultaneously entering the data. The E-band result provides direct evidence that \textit{both} genuine improvement and measurement expansion occurred simultaneously.

Column (3) excludes London's 33 boroughs, which have atypically high rental shares and dense, older housing stock. The coefficient remains significant and similar in magnitude ($0.041$ vs.\ $0.045$). Column (4) uses binary treatment (above/below median rental share), confirming the pattern ($0.010$, $p < 0.001$).

\input{tables/tab3_robustness}

\subsection{The Compliance-Driven Assessment Surge}

\Cref{tab:surge} documents the direct evidence for the measurement effect. Mean quarterly rental EPC lodgements rose from 87,000 in 2012--2014 to 144,000 in 2018Q2--2020Q1---a 66 percent increase. Rental's share of all lodgements rose from 31 to 43 percent. This surge was not driven by a general increase in EPC activity: sale lodgements actually declined slightly. The lodgement surge is exactly the mechanism through which MEES expanded the measurement frontier: properties that would never have been assessed in the absence of MEES were now entering the EPC register, and many of these were substandard.

\input{tables/tab4_surge}

\subsection{Aggregate Time Series}

At the national level, England's F+G share declined continuously from 8.6 percent (2008Q4) to 1.3 percent (2025Q4). An interrupted time series with structural breaks at announcement and implementation reveals a pre-existing downward trend of $-0.12$ percentage points per quarter, a level shift of $+1.8$ percentage points at announcement ($p < 0.001$), and a further $-1.6$ percentage point shift at implementation ($p < 0.001$). The positive announcement coefficient is initially puzzling but reflects the same measurement dynamic: the 2015 announcement prompted early compliance assessments that temporarily increased the visible stock of substandard properties. By 2018, when enforcement began, the stock had already been substantially measured and the net effect turned negative.

\section{Discussion}

The measurement effect we document has three implications. First, it suggests that aggregate EPC statistics systematically overstate the realized energy efficiency gains from MEES. The 80 percent decline in England's F+G share between 2008 and 2025 reflects a combination of genuine upgrades, property withdrawal from the rental market, and the entry of newly assessed properties into an expanding denominator. A back-of-the-envelope calculation illustrates the scale: the rental lodgement surge brought roughly 53,000 additional rental EPCs per quarter into the system post-MEES. If even one-quarter of these were F- or G-rated properties that would not otherwise have been assessed, the measurement channel accounts for approximately 13,000 ``phantom'' F/G observations per quarter---equivalent to 14 percent of the total F+G lodgements observed nationally in the post-MEES period. Without individual-level panel data tracking the same properties over time, a precise decomposition remains infeasible, but the cross-LA evidence demonstrates that the measurement channel is quantitatively important.

Second, the finding connects to a broader pattern in regulatory evaluation: regulations that use administrative data as both a compliance mechanism and an outcome measure create an inherent tension between enforcement and measurement \citep{houde2018consumer}. Building codes, environmental standards, and financial regulations all share this feature. When compliance requires reporting, the act of compliance generates data that may undermine na\"ive before-after comparisons.

Third, the E-band result ($\beta = -0.043$) confirms that MEES did produce genuine upgrading---but this upgrading was concentrated at the threshold, consistent with the ``just passing'' behavior documented in the bunching literature \citep{kleven2016bunching}. Properties were upgraded from F/G to E, not from F/G to D or above. This minimal-compliance pattern raises questions about the long-term energy savings of MEES if the upcoming EPC C mandate (2030) requires a second round of threshold-targeted improvements.

\section{Conclusion}

England's Minimum Energy Efficiency Standards reduced substandard rental housing, but they also revealed how much of it had been invisible. By mandating energy assessment as a precondition for letting, MEES created a measurement effect that simultaneously exposed and addressed energy poverty in the rental sector. This dual nature of the regulation---as both information system and enforcement mechanism---means that aggregate statistics overstate the improvement and understate the baseline problem. For policymakers designing the next generation of building efficiency standards, the lesson is that mandatory assessment generates valuable information externalities, but the resulting data cannot be na\"ively compared to pre-regulation baselines.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
