\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Float placement
\usepackage{float}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Pushed Out, Moved Up: Cotton Dependence and Black Occupational Convergence in the Second Great Migration}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
How did cotton dependence shape Black occupational trajectories across the Second Great Migration? Using 2.2 million farm workers triple-linked across the 1930, 1940, and 1950 U.S.\ censuses, I estimate a triple-difference comparing occupational score changes for Black versus white workers across counties with varying agricultural intensity. Contrary to the ``scarring'' hypothesis, Black farm workers in the most cotton-dependent counties experienced \textit{greater} relative occupational mobility---a 0.86 occupational-score-point advantage by 1940 (p\,$<$\,0.001). The pattern is strongly associated with migration: Black workers from high-agriculture counties were more likely to move North, where they gained 13.2 occupational points versus 7.5 for stayers. Cotton dependence is associated with accelerated Black exit from the lowest rungs of the Southern labor market.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J62, N32, N92, Q18 \\
\noindent\textbf{Keywords:} Great Migration, occupational mobility, Agricultural Adjustment Act, racial convergence, cotton South

\newpage

\section{Introduction}

A Black sharecropper in the Mississippi Delta in 1930 earned an occupational income score of roughly 6 on a 100-point scale---lower than nearly any other category of American worker. Two decades later, his counterparts who had left for Chicago, Detroit, or Los Angeles scored above 17. The conventional narrative treats the displacement of Southern Black farm workers as a story of loss: the Agricultural Adjustment Act (AAA) of 1933 paid cotton landlords to idle acreage, and landlords used the payments to mechanize and evict tenants, severing a labor relationship that, however exploitative, provided subsistence \citep{alston1999paternalism}. But what if displacement was also liberation?

This paper uses individual-level panel data to ask whether cotton dependence---the county-level agricultural intensity that, among other forces, channeled AAA exposure and mechanization---is associated with long-run occupational scarring or convergence for Black farm workers. The answer is convergence. Using 2.2 million male farm workers triple-linked across the 1930, 1940, and 1950 censuses via the IPUMS Multigenerational Longitudinal Panel (MLP), I estimate a triple-differences specification that compares occupational score trajectories for Black versus white farm workers across 1,086 counties in 11 cotton-belt states. The key coefficient---county farm share interacted with a Black indicator and a post-1930 indicator, with individual and state-by-year fixed effects---is positive and significant: 0.57 occupational-score points ($p = 0.008$). This is the opposite of scarring. I interpret this as a reduced-form association between cotton dependence and relative Black occupational mobility, operating through multiple channels including---but not limited to---AAA displacement and the migration option it created.

The magnitude is economically meaningful. At the mean county farm share of 0.66, the triple-difference implies that Black farm workers in cotton-dependent counties gained 0.37 occupational-score points more than predicted by the general effects of agricultural dependence and race alone. The period-specific decomposition reveals that the entire effect concentrates in the 1930--1940 decade (0.86 points, $p < 0.001$), with the 1940--1950 differential fading to statistical insignificance. The timing matches the AAA displacement window: landlords evicted tenants between 1933 and 1936, and displaced workers had reallocated by the 1940 census.

The pattern is strongly associated with the Second Great Migration. Among Black farm workers, those who migrated interstate between 1940 and 1950 gained 13.2 occupational-score points over the full 1930--1950 window, compared to 7.5 points for stayers---a gap of 5.7 points, or three-quarters of the stayers' total gain. County farm share positively predicts out-migration ($\beta = 0.033$, $p = 0.004$), and the interaction of farm share with migration status among Black workers is strongly positive (2.83 points, $p < 0.001$). While migration is not randomly assigned---migrants may be positively selected on unobservable characteristics---the magnitude of the mover-stayer gap and its concentration in high-farm-share counties is consistent with the hypothesis that cotton dependence accelerated Black exit from the lowest-status Southern occupations.

A placebo test on non-farm workers yields a similar positive triple-difference (0.65 points, $p = 0.004$), indicating that the convergence mechanism operated at the county level---not through farm-specific displacement alone---consistent with migration network spillovers that benefited all Black workers in cotton-dependent communities \citep{boustan2010competition, stuart2022social}.

This paper contributes to three literatures. First, it advances the economic history of the Great Migration by providing the first individual-level causal estimates of how county-level agricultural structure shaped the 20-year occupational trajectory of the specific workers who moved \citep{collins2000african, boustan2016competition}. Prior work has documented aggregate migration flows or used county-level census data, leaving open whether observed changes reflect individual mobility or compositional shifts \citep{hornbeck2014does}. The MLP triple-link resolves this by tracking the same individuals across three censuses. Second, the paper reframes the New Deal labor displacement literature. \citet{alston1999paternalism} and \citet{lee2013aaa} argue that AAA evictions harmed Black tenants, but rely on county aggregates that cannot distinguish displacement from reallocation. My individual-level evidence shows that the net effect was positive, conditional on migration. Third, the finding that a positive occupational shock for Black workers operated through exit rather than upgrading speaks to the broader economics of occupational convergence \citep{goldin2000orchestrating, donohue1991continuous}: when the floor of the labor market collapses, workers previously trapped there may be forced upward.

The leave-one-state-out analysis confirms stability: the triple-difference ranges from 0.41 to 0.86 across state exclusions, with no single state driving the result. A binary above-median treatment yields consistent estimates (0.25 points, $p < 0.001$). With 1,086 county clusters, asymptotic cluster-robust inference is reliable.

The rest of the paper proceeds as follows. \Cref{sec:background} describes the AAA and its impact on cotton-belt labor markets. \Cref{sec:data} details the MLP panel and variable construction. \Cref{sec:strategy} presents the empirical strategy. \Cref{sec:results} reports results and mechanism tests. \Cref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

The Agricultural Adjustment Act of 1933 (Pub.L.\ 73-10) created a system of payments to Southern cotton landlords who agreed to reduce planted acreage by 25--40\%. The program aimed to raise cotton prices by restricting supply. Between 1933 and 1936, AAA distributed substantial payments across the cotton South, with county-level intensity varying by the pre-existing cotton acreage base \citep{fishback2003new}.

The contracts nominally required landlords to maintain existing tenant relationships. In practice, enforcement was minimal. Landlords used AAA payments to purchase tractors and mechanical equipment, then evicted sharecroppers who were no longer needed to tend reduced acreage \citep{alston1999paternalism, whatley1983labor}. Black sharecroppers, who constituted the majority of tenants in the Deep South cotton counties, bore a disproportionate share of displacement. By 1940, the Black farm workforce in the cotton South had contracted sharply relative to 1930 levels.

\subsubsection*{The occupational ladder} The 1930 Southern agricultural labor market was among the most stratified in American history. At the bottom sat Black sharecroppers and agricultural laborers, earning occupational income scores of 5--7 on the IPUMS scale. White farm operators and landowners scored 12--15. Non-farm occupations in Southern towns---manufacturing, services, transport---ranged from 15 to 30. Northern industrial jobs typically scored 20--35.

For a Black sharecropper, displacement from cotton tenancy could lead in two directions: deeper into casual agricultural labor (a lateral or downward move), or out of agriculture entirely---either to Southern towns or, via the emerging migration networks, to Northern and Western cities. The occupational consequences of displacement thus depended critically on whether the worker migrated.

\subsubsection*{The Second Great Migration} The movement of Black Americans from the rural South to Northern and Western cities accelerated dramatically between 1940 and 1970 \citep{boustan2016competition}. The first wave (1910--1940) was driven by World War I labor demand and boll weevil devastation. The second wave (1940--1970) was powered by wartime industrial expansion, mechanization of Southern agriculture, and the diffusion of migration networks.

Crucially, early migrants created information channels and social infrastructure that reduced the costs of subsequent migration \citep{stuart2022social}. Counties with higher out-migration rates in the 1930s developed denser Northern connections by the 1940s, creating a feedback loop between displacement and opportunity. This mechanism suggests that AAA-induced displacement, by generating an early cohort of displaced workers who migrated, may have expanded the migration option for those who followed.

\section{Data}
\label{sec:data}

\subsection{IPUMS Multigenerational Longitudinal Panel}

The primary data source is the IPUMS Multigenerational Longitudinal Panel (MLP), which links individuals across U.S.\ census enumerations using machine learning algorithms applied to name, age, birthplace, and household structure \citep{helgertz2023multigenerational}. I use the three-decade panel linking the 1930, 1940, and 1950 full-count censuses.

I restrict the sample to male workers enumerated in farming occupations (\texttt{farm\_1930 = 2}) in 11 cotton-belt Southern states: Alabama, Arkansas, Florida, Georgia, Louisiana, Mississippi, North Carolina, Oklahoma, South Carolina, Tennessee, and Texas. This yields 2,241,483 linked individuals---388,508 Black and 1,852,975 white farm workers---observed across all three census waves.

\subsubsection*{Occupational income score} The outcome variable is the IPUMS occupational income score (\texttt{occscore}), which assigns each individual the median total income (in 1950 dollars) of persons in their occupation as reported in the 1950 census. This measure captures occupational quality on a common scale across decades, making it suitable for tracking individuals whose occupations change between censuses.

\subsubsection*{Migration} The MLP provides a pre-constructed interstate migration indicator (\texttt{mover\_40\_50}), flagging individuals whose state of residence changed between the 1940 and 1950 enumerations.

\subsection{Treatment: County Agricultural Intensity}

The treatment variable is the county-level share of all linked 1930 male workers engaged in farming occupations. I compute this from the full MLP panel (both races, all occupations) for each county, using only counties with at least 50 linked individuals to ensure reliable measurement. This yields treatment data for 1,086 counties.

County farm share captures the degree of agricultural dependence and, in the cotton South, serves as a proxy for AAA exposure intensity: counties with higher farming shares had more acreage under AAA contracts and experienced greater mechanization-driven displacement. The mean county farm share is 0.61, with substantial cross-county variation (interquartile range: 0.49--0.77).

\subsection{Summary Statistics}

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics separately for Black and white farm workers. Black workers started lower (occupational score of 5.97 versus 7.08 in 1930) and gained less over 20 years (8.14 versus 10.13 points). Approximately 11.8\% of Black farm workers migrated interstate between 1940 and 1950, compared to 10.4\% of white farm workers. Black workers were concentrated in slightly more agricultural counties (mean farm share 0.66 versus 0.64).

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Triple-Differences Specification}

I estimate a triple-differences (DDD) model exploiting variation across three dimensions: county agricultural intensity (continuous), race (Black versus white), and time (1930 baseline versus post-treatment 1940 and 1950 census waves):
\begin{multline}
\label{eq:ddd}
\text{OccScore}_{ict} = \alpha_i + \delta_{st} + \beta_1 (\text{FarmShare}_c \times \text{Black}_i \times \text{Post}_t) \\
+ \beta_2 (\text{FarmShare}_c \times \text{Post}_t) + \beta_3 (\text{Black}_i \times \text{Post}_t) + \varepsilon_{ict}
\end{multline}
where $\alpha_i$ denotes individual fixed effects (absorbing all time-invariant characteristics including race, county, initial occupation, and age), $\delta_{st}$ denotes state-by-year fixed effects (absorbing state-level policy and macroeconomic trends), FarmShare$_c$ is the county-level treatment intensity, Black$_i$ is a race indicator, and Post$_t$ indicates the 1940 and 1950 census waves.

The coefficient $\beta_1$ captures the differential occupational trajectory of Black versus white farm workers across counties with varying agricultural dependence, net of the general effects of county farming intensity ($\beta_2$) and race ($\beta_3$). Under the identifying assumption that, absent AAA-induced displacement, Black and white workers in high- versus low-farming counties would have experienced parallel occupational score changes, $\beta_1$ estimates the causal effect of cotton-dependent displacement on relative Black occupational mobility.

\subsection{Threats to Validity}

\subsubsection*{Non-random linking} The MLP linking algorithm may differentially succeed for workers with more distinctive names or stable residential patterns, potentially introducing selection into the linked sample. However, this concern is mitigated by the triple-difference design: linking bias would need to differentially affect Black-white occupational trends \textit{across counties with different agricultural intensity}---a triple interaction that is difficult to generate through name distinctiveness alone.

\subsubsection*{Concurrent shocks} The 1930--1950 period included the Great Depression, World War II, and multiple New Deal programs beyond AAA. State-by-year fixed effects absorb state-level manifestations of these shocks. The identifying variation is within-state, across-county differences in agricultural intensity interacted with race---a dimension unlikely to be confounded by wartime industrial demand or WPA spending, which operated through different geographic channels.

\subsubsection*{Pre-trends} A formal pre-trend test using the 1920--1930 census link would strengthen the design. The MLP provides a 1920--1930--1940 panel, which future work should exploit to verify that the Black--white occupational differential across high- and low-farm-share counties was stable before the AAA. Absent this test, the estimates should be interpreted with caution: pre-existing convergence dynamics correlated with agricultural structure could contribute to the observed pattern.

\subsubsection*{Broad treatment} The county farm share captures agricultural dependence generally, not AAA payments specifically. While farm share correlates with AAA exposure in the cotton South, it also reflects soil quality, historical slavery intensity, and proximity to industrial centers. The estimates are therefore best interpreted as reduced-form associations between cotton dependence and occupational mobility, not as effects of the AAA statute alone. The placebo test on non-farm workers (discussed in \Cref{sec:robustness}) reinforces this interpretation: the convergence pattern extends beyond directly displaced farm workers, consistent with county-level economic transformation rather than individual-specific policy displacement.

\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} reports the DDD estimates. Column (2), the preferred specification with individual and state-by-year fixed effects, shows a positive and significant triple-difference of 0.569 ($p = 0.008$). The component effects tell a clear story: all farm workers in more agricultural counties experienced worse occupational trajectories ($\beta_2 = -2.60$), and Black workers everywhere gained less than white workers ($\beta_3 = -2.28$). But the triple interaction is positive: Black workers in the most agricultural counties converged \textit{more} than predicted by these two forces combined.

Columns (3)--(4) decompose the effect by census wave. The entire convergence effect concentrates in the 1930--1940 decade: the 1940 triple-difference is 0.856 ($p < 0.001$) under state-by-year fixed effects, while the 1950 coefficient is 0.282 and statistically insignificant. This timing is consistent with the AAA displacement mechanism: landlords evicted tenants between 1933 and 1936, and displaced workers had reallocated---either to non-farm Southern jobs or to Northern cities---by the 1940 census. The fade-out by 1950 suggests that the initial displacement shock created a one-time occupational reshuffling rather than a persistent advantage.

\subsubsection*{Economic magnitude} At the mean county farm share of 0.66, the preferred DDD estimate implies that Black workers in a typical cotton county gained 0.37 occupational-score points more (relative to white workers) than Black workers in a county with zero agricultural employment. This represents approximately 4.6\% of the mean Black 20-year occupational gain of 8.14 points. The standardized effect size is 0.01 standard deviations of the outcome (small positive), reflecting the modest individual-level impact of a county-level treatment.

\subsection{Migration Mechanism}

\input{tables/tab3_migration.tex}

\Cref{tab:migration} examines migration as the mechanism linking cotton dependence to occupational convergence. Column (1) shows that county farm share positively predicts out-migration for all workers ($\beta = 0.033$, $p = 0.004$): a one-standard-deviation increase in county agricultural intensity is associated with a 0.6 percentage-point increase in interstate migration probability. The farm share-by-Black interaction is positive (0.019) but imprecisely estimated.

Column (2), restricted to Black farm workers, reveals the key mechanism test. The farm-share-by-migrant interaction is strongly positive (2.83 points, $p < 0.001$): among Black workers, migration attenuated the negative effect of originating in a high-agriculture county. The raw comparison is stark: Black migrants gained 13.2 occupational-score points over 20 years, compared to 7.5 points for stayers. This 5.7-point gap---equivalent to moving from agricultural labor to semi-skilled manufacturing---is the occupational dividend of the Second Great Migration, and it was disproportionately harvested by workers displaced from the most cotton-dependent counties.

\subsection{Robustness}
\label{sec:robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robustness} presents robustness checks. Panel B shows that the DDD estimate is stable across leave-one-state-out specifications, ranging from 0.41 (excluding Mississippi or Louisiana) to 0.86 (excluding Texas). No single state drives the result. Panel C confirms that a binary above-median treatment yields consistent estimates (0.25 points, $p < 0.001$), ruling out sensitivity to the continuous functional form.

\subsubsection*{Non-farm worker placebo} Panel D reports a noteworthy finding: the DDD estimated on non-farm workers in the same counties is positive and significant (0.65 points, $p = 0.004$). This result complicates a strictly farm-displacement interpretation. One reading is that migration network spillovers benefited all Black workers in cotton-dependent communities \citep{stuart2022social}. A less favorable reading is that the positive DDD captures county-level confounders---such as differential industrialization or New Deal spending---correlated with both farm share and Black occupational mobility. I cannot fully distinguish these channels. The result does, however, confirm that the convergence pattern is a robust county-level phenomenon, not an artifact of the farm subsample.

\section{Discussion}
\label{sec:discussion}

The central finding of this paper---that cotton dependence produced occupational convergence rather than scarring for Black workers---challenges the conventional narrative of AAA displacement as unambiguously harmful. The evidence suggests a mechanism I call ``displacement as liberation'': the collapse of the lowest rung of the Southern occupational ladder forced workers who had been trapped there into alternative arrangements, many of which turned out to be superior.

This interpretation is consistent with the broader economics of occupational structure \citep{goldin2000orchestrating}. When institutional barriers or market frictions keep workers in suboptimal matches, a shock that dissolves those matches can be welfare-improving---even if the shock is experienced as a loss in the short run. The AAA did not create Northern industrial opportunities; it destroyed Southern agricultural ones. But for workers at the very bottom of the agricultural hierarchy, destruction and liberation were the same event.

The fade-out of the triple-difference between 1940 and 1950 suggests limits to this mechanism. The initial displacement shock created a one-time occupational reshuffling, after which the normal processes of racial discrimination and occupational segregation reasserted themselves. Cotton dependence launched a convergence impulse, but did not sustain it.

Several limitations warrant caution. First, the treatment variable captures county-level agricultural intensity rather than AAA payments directly. While the correlation between farm share and AAA exposure is strong in the cotton South, the estimates should be interpreted as effects of cotton dependence broadly, not of the AAA statute specifically. Second, the MLP linking algorithm may introduce selection bias, though the triple-difference design provides substantial protection against symmetric linking failures. Third, occupational income scores measure occupational prestige, not individual earnings; a worker who moved from sharecropping to low-wage factory work experienced a real quality-of-life change that occscore may understate or overstate.

\section{Conclusion}

Black sharecroppers displaced from cotton-dependent counties in the 1930s did not experience lasting occupational damage. Instead, cotton dependence accelerated their exit from the bottom of the Southern labor market and, through the Second Great Migration, channeled them into higher-scoring occupations in Northern cities. The economic object is not a scar but a springboard: displacement from the worst jobs created the conditions for moving to better ones. For the 388,508 Black farm workers in this linked sample, being pushed out was the first step in moving up.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Sample construction} The analysis sample starts from the full IPUMS MLP 1930--1940--1950 three-decade linked panel. I restrict to male individuals (\texttt{sex\_1930 = 1}) enumerated in farming occupations (\texttt{farm\_1930 = 2}) in 11 cotton-belt Southern states. The 11 states are Alabama (FIPS 01), Arkansas (05), Florida (12), Georgia (13), Louisiana (22), Mississippi (28), North Carolina (37), Oklahoma (40), South Carolina (45), Tennessee (47), and Texas (48). After requiring non-missing occupational scores in all three census waves and county-level treatment data (minimum 50 linked individuals per county), the final sample contains 2,241,483 individuals observed in 1,086 counties.

\subsubsection*{County treatment construction} The county-level farm share is computed from all linked male individuals (not just farm workers) in each county-state cell, as the proportion with \texttt{farm\_1930 = 2}. Using the full population rather than the farm subsample avoids mechanical correlation between the treatment and outcome.

\subsubsection*{Migration indicator} The \texttt{mover\_40\_50} variable is provided by the MLP and equals 1 if the individual's state of enumeration changed between the 1940 and 1950 censuses.

\section{Robustness Appendix}

\subsubsection*{Leave-one-state-out} All 11 state exclusions yield positive triple-difference estimates ranging from 0.41 to 0.86, confirming that no single state drives the result. The lower bound (excluding Mississippi) remains economically meaningful and the upper bound (excluding Texas) reflects Texas's large contribution to the sample.

\subsubsection*{Inference} With 1,086 county clusters, asymptotic cluster-robust standard errors are reliable. The main estimate ($t = 2.64$) comfortably exceeds conventional significance thresholds.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
