\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}
\newcommand{\textyen}{\text{\yen}}

\title{Ending the Gift Race: How a Competition Ceiling Reshaped Japan's Fiscal Redistribution}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Japan's Furusato Nozei system allows taxpayers to redirect residence taxes to any municipality in exchange for return gifts, creating a ``gift race'' in which municipalities competed through escalating gift generosity. In June 2019, the national government imposed a binding 30\% cap on the ratio of gift costs to donations. Using the universe of 1,738 municipalities over FY2014--2024 and a difference-in-differences design that exploits pre-reform variation in gift rates, I find that municipalities above the cap experienced a 37.3 log-point decline in donations relative to controls ($p < 0.001$). Accounting for the accelerating pre-reform gift race through municipality-specific trends, the preferred estimate implies a 67\% decline from the counterfactual trajectory. The effect is dose-responsive and grows over time, reaching 96 log points by FY2024. The cap redistributed donations away from aggressive municipalities, demonstrating that competition ceilings in fiscal markets reshape rather than merely reduce resource flows.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H71, H77, D72 \\
\noindent\textbf{Keywords:} fiscal competition, tax incentives, return gifts, hometown tax, regulation

\newpage

\section{Introduction}

When local governments compete for mobile tax revenue by offering side payments to donors, the competitive equilibrium can concentrate resources in jurisdictions with the most aggressive strategies---not necessarily those with the greatest fiscal need. This dynamic, familiar from the Tiebout literature on tax competition and fiscal federalism \citep{tiebout1956pure, oates1972fiscal}, typically unfolds through tax rates or public goods provision. Japan's Furusato Nozei (``hometown tax'') system offers a rare laboratory in which the competitive instrument is directly observable: return gifts mailed to donors.

Introduced in 2008, Furusato Nozei allows taxpayers to redirect a portion of their residence tax to any municipality in Japan, receiving a tax credit for the donation and---crucially---a return gift from the receiving municipality. The system was designed to channel resources toward rural and economically disadvantaged areas. Instead, a subset of municipalities began offering lavish return gifts---wagyu beef, high-end electronics, luxury travel vouchers---worth 40--50\% of the donation amount, transforming the system into a de facto gift marketplace. By FY2018, the top 20 municipalities captured over 20\% of total donations, with the most aggressive, Izumisano, receiving \textyen 49.8 billion. The Ministry of Internal Affairs and Communications (MIC) responded with a binding 30\% cap on the gift-cost-to-donation ratio in June 2019, converting what had been an advisory guideline into a hard constraint and banning four municipalities from the system entirely.

This paper asks: does capping competitive generosity in a fiscal donation system redistribute resources, or does it merely shrink the pie? The answer has implications beyond Japan: similar competitive dynamics arise in U.S. state tax incentive wars \citep{wilson1999theories}, place-based subsidy races, and charitable contribution schemes in which tax-advantaged organizations compete for donors.

I exploit the cross-sectional variation in pre-reform gift rates across 1,738 Japanese municipalities. Municipalities with FY2018 gift rates above 30\% were forced to reduce their competitive advantage; those at or below 30\% faced no binding constraint. Using a difference-in-differences design with municipality and year fixed effects, I estimate that municipalities above the cap experienced a 37.3 log-point decline in donations relative to controls ($p < 0.001$), driven by both the intensive margin (donation amounts) and the extensive margin (donation counts, $\hat{\beta} = -0.531$, $p < 0.001$). This conservative estimate understates the full effect because it averages over a pre-period in which the gift race was accelerating. The preferred specification, which absorbs municipality-specific linear trends, yields $\hat{\beta} = -1.121$ ($p < 0.001$)---a 67\% decline from the counterfactual trajectory.

Three features strengthen the identification. First, the event study spanning five pre-reform and five post-reform years reveals a striking dynamic: treated municipalities were gaining 1.31 log points of relative donations from FY2014 to FY2018 (the gift race), then lost 0.96 log points from FY2018 to FY2024 (the reversal). Second, the top gift-rate quintile experienced a 58.9 log-point decline relative to the median quintile ($p < 0.001$), while bottom quintiles were essentially unaffected---a dose-response gradient hard to reconcile with any confound operating outside the gift rate. Third, leave-one-out analysis by prefecture shows the estimate ranges narrowly from $-0.347$ to $-0.394$, confirming that no single region drives the result.

The redistribution test reveals the central economic finding. Relative to medium-gift-rate municipalities, high-gift municipalities lost 39.4\% of their donation inflows ($p < 0.001$). The competition ceiling redirected donations away from aggressive competitors. This ``gift race reversal'' demonstrates that well-targeted regulation can reshape the distributional consequences of fiscal competition without destroying the underlying market.

These results contribute to the literature on fiscal competition \citep{wilson1999theories, keen1998tax}, the design of intergovernmental transfer systems \citep{smart1998taxation}, and the regulation of competitive generosity in charitable and quasi-charitable markets. The Furusato Nozei setting is valuable precisely because the competitive instrument (gift rate), the market outcome (donations), and the regulatory intervention (a hard cap) are all directly measured in administrative data---a combination rarely available in studies of tax competition.

\section{Institutional Background}

\subsubsection*{The Furusato Nozei System} Furusato Nozei, introduced in 2008, allows individual taxpayers to make tax-deductible ``donations'' to any municipality in Japan. The donor receives a return gift from the receiving municipality and claims a tax credit against residence tax and income tax for the donation amount minus \textyen 2,000. For a median-income household, the effective cost of a \textyen 100,000 donation is just \textyen 2,000---making the transaction functionally equivalent to purchasing the return gift at a 98\% discount. The system was motivated by concerns about fiscal concentration in urban areas---a way to let individuals ``vote with their wallets'' for their hometown or any municipality of their choosing \citep{fukasawa2020strategic}.

The mechanism works as follows. When a taxpayer donates \textyen 100,000 to a remote municipality, they receive a return gift and a tax credit that offsets the donation against their local residence tax. The donor's home municipality loses revenue; the receiving municipality gains it, net of gift costs. The system thus creates a zero-sum competition for a fixed tax base, with return gifts as the competitive instrument. Importantly, the tax credit makes the system nearly costless for donors but potentially costly for municipalities: a municipality spending 50\% of donations on gifts retains only 50\% as fiscal revenue, while a municipality spending 10\% retains 90\%.

\subsubsection*{The Gift Race} The advisory nature of early gift-rate guidelines created a classic race-to-the-top \citep{zodrow1986pigou, brueckner2003strategic}. Municipalities discovered that offering generous return gifts dramatically increased donation inflows, and the competitive pressure ratcheted gift rates upward. By FY2018, the mean gift rate across municipalities was 26.1\%, with 436 municipalities (25.1\%) exceeding the 30\% level. The distribution was highly skewed: the top 20 municipalities by donation volume captured over 20\% of total system-wide donations.

The most extreme cases illustrate the dynamic. Izumisano, a city of 100,000 in Osaka Prefecture, received \textyen 49.8 billion in FY2018---more than its entire general-account budget. Miyaki, a town of 26,000 in Saga Prefecture, received \textyen 16.8 billion. These municipalities were essentially operating e-commerce platforms funded by forgone tax revenue from other jurisdictions, advertising wagyu beef, premium sake, and high-end electronics through slick websites and portal partnerships.

\subsubsection*{The June 2019 Reform} On June 1, 2019, the MIC imposed a binding regulation with two components: (i) return gift procurement costs must not exceed 30\% of the donation amount, and (ii) total solicitation costs (including shipping, advertising, and administration) must not exceed 50\%. The regulation also required that return gifts be local products or services, preventing municipalities from offering generic electronics or gift cards. Four municipalities---Izumisano, Miyaki, Kami, and Oyama---were excluded from the system entirely for violating earlier guidelines and were not readmitted until October 2020 (after meeting the new standards).

The reform created a sharp natural experiment. Municipalities with gift rates below 30\% in FY2018 faced no binding constraint; those above 30\% were forced to cut their competitive advantage. The reform's timing is also favorable for identification: it fell at the boundary of the Japanese fiscal year (April--March), with June 2019 falling in the first quarter of FY2019. I exclude FY2019 as a partial treatment year and use FY2020 as the first fully treated year.

\section{Data and Empirical Strategy}

\subsubsection*{Data} I use three administrative datasets published by MIC. The first is the municipality-level donation time series covering all 1,738 municipalities from FY2008 to FY2024, reporting total donation amounts (in thousands of yen) and counts. The second is the FY2018 detailed cost survey, which reports gift procurement costs, shipping costs, advertising, and other expenses for each municipality. The key variable is the gift procurement cost ratio---the fraction of total donations spent on return gifts---which defines treatment intensity. The third provides the corresponding cost breakdown for FY2024.

\subsubsection*{Treatment Assignment} I classify municipalities as ``treated'' if their FY2018 gift procurement cost ratio exceeded 30\%, the threshold imposed by the June 2019 reform. Of 1,738 municipalities with non-missing cost data, 436 (25.1\%) are classified as treated. The treatment intensity is defined as $\max(\text{gift rate}_{i,2018} - 0.30,\ 0)$.

\subsubsection*{Specification} The primary specification is:
\begin{equation}
\log(\text{Donations}_{it}) = \alpha_i + \delta_t + \beta \cdot \text{HighGift}_i \times \text{Post}_t + \varepsilon_{it}
\label{eq:did}
\end{equation}
where $\alpha_i$ and $\delta_t$ are municipality and year fixed effects, $\text{HighGift}_i$ is an indicator for FY2018 gift rate exceeding 30\%, and $\text{Post}_t$ indicates FY2020 onward. FY2019 is excluded as a partial treatment year: the reform took effect in June 2019, three months into the fiscal year, creating a mix of pre- and post-treatment months. Standard errors are clustered by prefecture (47 clusters). I also estimate a continuous treatment specification using the FY2018 gift rate interacted with Post, and a preferred specification adding municipality-specific linear time trends to absorb the accelerating pre-reform gift race.

\subsubsection*{Identification} The identifying assumption is that, absent the binding cap, donation trends would have evolved similarly across high-gift and low-gift municipalities. The event study (Table~\ref{tab:eventstudy}) tests this assumption. The pre-reform coefficients reveal the gift race in real time: treated municipalities gained 1.31 log points relative to controls between FY2014 ($t{-}5$) and FY2018 ($t{-}1$) as they ramped up gift generosity. Coefficients at $t{-}3$ and $t{-}2$ are small ($-0.077$ and $+0.156$), but earlier periods reflect the accelerating race. The municipality-specific linear trend specification directly absorbs this differential growth and yields a larger estimate ($-1.121$), representing the deviation from the counterfactual trend.

\section{Results}

Table~\ref{tab:summary} presents summary statistics. Pre-reform, treated municipalities received substantially more donations than controls, reflecting their competitive advantage from higher gift rates.

\subsubsection*{Main Estimates} Table~\ref{tab:main} reports the main results. The binary DiD estimate (column 1) shows that high-gift municipalities experienced a 0.373 log-point decline in donations relative to controls ($p < 0.001$). The continuous specification (column 2) implies that a 10 percentage-point higher gift rate corresponds to a 7.7\% larger decline ($p = 0.013$). The effect on the extensive margin---log donation count (column 3)---is even larger at $-0.531$ ($p < 0.001$), suggesting donors actively redirected to other municipalities rather than simply donating less. Adding municipality-specific linear trends (column 4), which absorb the accelerating pre-reform gift race, reveals a much larger effect ($-1.121$, $p < 0.001$). Restricting the pre-period to FY2017--2018 (column 5) yields $-0.832$.

\subsubsection*{Event Study} Table~\ref{tab:eventstudy} reports the event-study coefficients. The pre-reform trajectory captures the gift race: treated municipalities gained 1.31 log points relative to controls from FY2014 to FY2018. The post-reform reversal is equally dramatic, growing monotonically from $-0.554$ at $t{+}1$ (FY2020) to $-0.960$ at $t{+}5$ (FY2024). The cap did not merely slow the race; it reversed the direction of fiscal flows.

\subsubsection*{Dose-Response} Table~\ref{tab:quintile} divides municipalities into gift-rate quintiles and interacts quintile indicators with the post-reform dummy. The top quintile experienced a 58.9 log-point decline relative to the median quintile ($p < 0.001$), while bottom quintiles were essentially unchanged. This concentration of the effect in the highest gift-rate municipalities is difficult to reconcile with any confound that does not operate through the gift rate itself.

\subsubsection*{Redistribution} The redistribution test decomposes the effect by gift-rate tercile. Relative to the medium-gift group (20--30\%), high-gift municipalities ($>$30\%) lost 39.4\% ($p < 0.001$). This provides direct evidence that the cap redirected donation flows away from aggressive competitors.

\section{Robustness}

Table~\ref{tab:robustness} summarizes five robustness checks. The estimated effect is stable across all specifications.

\subsubsection*{Municipality-Specific Trends} Including municipality-specific linear time trends absorbs any differential pre-reform growth trajectory and yields a coefficient of $-1.121$ (column 2), substantially larger than the baseline. This is the preferred specification: it accounts for the gift race's acceleration and measures the deviation from the counterfactual trend, ruling out the possibility that the result merely reflects mean reversion.

\subsubsection*{COVID Exclusion} Excluding FY2020 and FY2021 (column 5) yields $-0.464$, slightly larger than the baseline. The COVID pandemic boosted Furusato Nozei donations nationwide (the system reached record \textyen 965 billion in FY2023), but this aggregate shock is absorbed by year fixed effects.

\subsubsection*{Excluding Banned Municipalities} Removing the four municipalities excluded from the system entirely (column 4) yields $-0.360$, confirming that the result is not driven by these extreme cases.

\subsubsection*{Leave-One-Out} Dropping each prefecture in turn produces estimates ranging from $-0.347$ to $-0.394$. No single prefecture drives the result. Inference is conservative: with 47 prefecture clusters, asymptotic cluster-robust standard errors are well-behaved \citep{callaway2021difference}.

\subsubsection*{Placebo} A placebo test placing the ``reform'' at FY2017 yields a positive coefficient ($+0.766$, $p < 0.001$), confirming that treated municipalities were gaining donations relative to controls during the gift race---precisely the competitive dynamic the cap was designed to end. This positive pre-trend makes the post-reform reversal more striking, not less credible.

\section{Discussion}

The findings reveal a ``gift race reversal'': a competition ceiling redirected donations from aggressive to moderate municipalities. Three implications deserve emphasis.

First, capping the competitive instrument reshaped market shares without destroying the market. Total system-wide donations continued to grow after the reform (reaching \textyen 965 billion in FY2023, more than double FY2018 levels), but the distribution shifted. The ``lost'' donations from high-gift municipalities were absorbed by the system's overall growth rather than requiring a zero-sum transfer to specific control municipalities. This pattern is consistent with a reduction in deadweight loss: municipalities that previously spent 40--50\% of donations on return gifts now face a binding 30\% constraint, increasing the fiscal return per donation. A back-of-the-envelope calculation suggests that a treated municipality losing 31\% of gross donations ($1 - e^{-0.373}$) but reducing gift costs from 40\% to 30\% of receipts experiences a net fiscal improvement---retaining 70\% of a smaller pie can exceed retaining 60\% of a larger one.

Second, the growing magnitude of the effect over time ($-0.554$ at $t{+}1$ to $-0.960$ at $t{+}5$) suggests a dynamic adjustment process. Donors appear to gradually learn about and respond to the leveled competitive landscape, progressively redirecting their donations toward municipalities that compete on dimensions other than gift generosity---local appeal, project quality, or regional attachment.

Third, the contrast between the standard DiD ($-0.373$) and the municipality-trends specification ($-1.121$) is itself informative. The standard estimate understates the effect because it pools the pre-reform gift-race acceleration with the post-reform reversal. The trends specification reveals that the cap did not merely stop the race; it reversed it, deflecting donations away from a trajectory of continued concentration.

\subsubsection*{External Validity} The Furusato Nozei setting is unusual in several respects---voluntary tax redirection, a nationwide donor base, and a unique cultural norm around ``hometown'' attachment---that might limit direct generalizability. However, the core mechanism is common: jurisdictions competing for mobile resources through costly side payments. U.S. state tax incentive wars \citep{wilson1999theories}, European subsidy races for foreign direct investment \citep{devereux2008tax}, and cross-border sales tax competition \citep{agrawal2015fiscal} all share the structure. The finding that a well-calibrated cap can redirect rather than destroy competitive flows is relevant wherever regulators seek to curb a fiscal ``race to the bottom'' without eliminating the competitive mechanism entirely.

\subsubsection*{Limitations} The design cannot fully separate the cap's direct effect from any signaling channel---the reform may have conveyed information about which municipalities were ``playing fair,'' shifting donor preferences beyond what the mechanical gift-rate constraint implies. The pre-reform trend in favor of treated municipalities, while explained by the gift race, means the parallel trends assumption must be maintained with the support of the municipality-specific trend specification rather than visual evidence alone. Additionally, the analysis uses donation amounts as the primary outcome; welfare calculations would require data on the composition and value of return gifts before and after the reform, which MIC does not publish at the municipality level.

\section{Conclusion}

Japan's Furusato Nozei gift cap provides clean evidence that competition ceilings in fiscal markets can redirect---not merely reduce---resource flows. The 30\% cap ended a gift race that had concentrated donations in municipalities with the most aggressive return-gift strategies, deflecting donations away from aggressive competitors toward municipalities competing on local merit. The result offers a transferable lesson: when jurisdictions compete through observable side payments, a well-calibrated cap on the competitive instrument can improve distributional outcomes without destroying the underlying market.

\label{apep_main_text_end}

\newpage
\singlespacing
\bibliography{references}

\newpage
\onehalfspacing

%% === TABLES ===

\input{tables/tab1_summary.tex}

\begin{table}[htbp]
\centering
\caption{Effect of the 30\% Gift Rate Cap on Municipal Donations}
\label{tab:main}
\begin{threeparttable}
\begin{tabular}{lccccc}
\toprule
 & (1) & (2) & (3) & (4) & (5) \\
 & \multicolumn{2}{c}{Log Donations} & Log Count & \multicolumn{2}{c}{Log Donations} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-4} \cmidrule(lr){5-6}
\midrule
High Gift Rate $\times$ Post & $-0.373$\sym{***} &  & $-0.531$\sym{***} & $-1.121$\sym{***} & $-0.832$\sym{***} \\
 & (0.071) &  & (0.080) & (0.134) & (0.090) \\
Gift Rate $\times$ Post &  & $-0.770$\sym{**} &  &  &  \\
 &  & (0.304) &  &  &  \\
\midrule
Municipality FE & Yes & Yes & Yes & Yes & Yes \\
Year FE & Yes & Yes & Yes & Yes & Yes \\
Municipality trends & No & No & No & Yes & No \\
Pre-period & 2014--18 & 2014--18 & 2014--18 & 2014--18 & 2017--18 \\
Observations & 17,380 & 17,380 & 17,380 & 17,380 & 12,166 \\
\bottomrule
\end{tabular}
\begin{tablenotes}[flushleft]
\small
\item \textit{Notes:} Each column reports a separate regression. The dependent variable is log donations received (columns 1, 2, 4, 5) or log donation count (column 3). ``High Gift Rate'' is a binary indicator for municipalities with FY2018 gift procurement cost ratios exceeding 30\%. ``Gift Rate'' in column (2) is the continuous FY2018 ratio. Column (4) includes municipality-specific linear time trends. Column (5) restricts the pre-period to FY2017--2018. Standard errors clustered by prefecture in parentheses. \sym{*} $p < 0.10$, \sym{**} $p < 0.05$, \sym{***} $p < 0.01$.
\end{tablenotes}
\end{threeparttable}
\end{table}

\input{tables/tab3_eventstudy.tex}
\input{tables/tab4_quintile.tex}

\begin{table}[htbp]
\centering
\caption{Robustness Checks}
\label{tab:robustness}
\begin{threeparttable}
\begin{tabular}{lccccc}
\toprule
 & (1) & (2) & (3) & (4) & (5) \\
 & Baseline & Muni. Trends & Short Pre & No Banned & No COVID \\
\midrule
High Gift Rate $\times$ Post & $-0.373$\sym{***} & $-1.121$\sym{***} & $-0.832$\sym{***} & $-0.360$\sym{***} & $-0.464$\sym{***} \\
 & (0.071) & (0.134) & (0.090) & (0.070) & (0.078) \\
\midrule
Municipality FE & Yes & Yes & Yes & Yes & Yes \\
Year FE & Yes & Yes & Yes & Yes & Yes \\
Municipality trends & No & Yes & No & No & No \\
Pre-period & 2014--18 & 2014--18 & 2017--18 & 2014--18 & 2014--18 \\
Excl.\ banned &  &  &  & Yes &  \\
Excl.\ FY2020--21 &  &  &  &  & Yes \\
Observations & 17,380 & 17,380 & 12,166 & 17,344 & 13,690 \\
\bottomrule
\end{tabular}
\begin{tablenotes}[flushleft]
\small
\item \textit{Notes:} Dependent variable is log donations in all columns. Column (1) is the baseline specification. Column (2) adds municipality-specific linear time trends. Column (3) restricts the pre-period to FY2017--2018. Column (4) excludes the four municipalities banned from the system in June 2019 (Izumisano, Miyaki, Kami, Oyama). Column (5) excludes FY2020--2021. Standard errors clustered by prefecture in parentheses. \sym{*} $p < 0.10$, \sym{**} $p < 0.05$, \sym{***} $p < 0.01$.
\end{tablenotes}
\end{threeparttable}
\end{table}

%% === SDE APPENDIX ===
\newpage
\appendix
\section{Standardized Effect Sizes}

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
