\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Walking Dead of the Nonprofit Sector: Mass Revocation and the Illusion of Creative Destruction}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2011, the IRS revoked the tax-exempt status of approximately 275,000 nonprofits that had failed to file annual returns for three consecutive years---removing roughly one-quarter of the sector's registered organizations in a single day. I exploit cross-county variation in revocation intensity to estimate effects on new nonprofit formation, sector employment, and charitable giving. Event study estimates reveal an initial burst of new entry in high-revocation counties, consistent with creative destruction. But this effect reverses sharply within two years: formations fall well below baseline, and nonprofit employment declines by 13 percent of a standard deviation. Rather than clearing deadwood, mass de-registration appears to damage the local ecosystem that sustains new entry. The collateral damage hypothesis dominates creative destruction in the medium run.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L31, H25, D22 \\
\noindent\textbf{Keywords:} nonprofits, tax-exempt status, creative destruction, organizational ecology, IRS enforcement

\newpage

\section{Introduction}

On June 8, 2011, the Internal Revenue Service published a list of 275,000 organizations whose tax-exempt status had been automatically revoked. These were not tax evaders or fraudulent charities. They were organizations---youth soccer leagues, volunteer fire companies, neighborhood associations---that had simply failed to file an annual return for three consecutive years. In a single administrative action, roughly one-quarter of America's registered nonprofit sector ceased to exist as tax-exempt entities.

The Pension Protection Act of 2006 mandated annual electronic filing for all tax-exempt organizations, including the smallest entities that had previously been exempt from any reporting requirement. Organizations that missed three consecutive deadlines faced automatic revocation---no warning, no appeal, no case-by-case review. The policy was designed to clean up the IRS Exempt Organizations registry, which had accumulated hundreds of thousands of ``zombie'' nonprofits: organizations that existed on paper but had long ceased meaningful activity \citep{irsdatareleases2011}.

Whether mass de-registration of dormant organizations helps or hurts the broader nonprofit sector is theoretically ambiguous. The \emph{creative destruction} hypothesis, drawing on the zombie firm literature in corporate finance \citep{caballero2008zombie}, predicts that removing dead organizations frees resources---donor attention, grant eligibility slots, volunteer energy, and administrative capacity at the IRS---for new, active entrants. Under this view, a cleaner registry lowers barriers to entry and improves the sector's overall vitality. The \emph{collateral damage} hypothesis predicts the opposite: many revoked organizations were merely filing-delinquent rather than truly inactive, and mass revocation damages the institutional infrastructure---networks, legitimacy, donor confidence---that sustains the local charitable ecosystem.

This paper tests these competing hypotheses using a continuous-treatment difference-in-differences design. I construct a county-year panel covering 3,060 U.S.\ counties over 2006--2020, measuring revocation intensity as the share of each county's pre-2010 registered nonprofits that were auto-revoked between 2010 and 2012. This creates substantial cross-county variation: in the median county, 47 percent of registered nonprofits lost their status, while the interquartile range spans 38 to 59 percent.

The main finding is a striking temporal pattern in new nonprofit formation. Event study estimates reveal that counties with higher revocation intensity experienced a significant \emph{increase} in new formations in the year of mass revocation (2011) and the year following---consistent with creative destruction. But this effect reverses sharply: by 2013--2014, the same counties show formations 1.8--2.1 standard deviations below baseline relative to low-revocation counties. The initial burst of entry was an illusion; in the medium run, mass revocation reduced, rather than stimulated, the formation of new organizations. The average effect across the full post-period is negative: a one-standard-deviation increase in revocation intensity reduces new formations per capita by 0.036 standard deviations ($\hat{\beta} = -0.192$, $p < 0.05$).

Nonprofit employment in the broader social-organization sector (NAICS 813) also declines significantly: the baseline specification yields $\hat{\beta} = -0.131$ ($p < 0.01$), corresponding to a small negative standardized effect. Per-worker earnings, however, show no meaningful change, suggesting that the employment margin adjusts through reduced headcount rather than wage compression. Cross-sectional evidence from IRS county income data indicates that higher-revocation counties have significantly lower charitable deductions per return even conditional on state and year effects, though the absence of pre-period giving data limits causal interpretation of this result.

These results speak to a broader question in organizational economics: when does removing low-quality incumbents improve the ecosystem, and when does it damage it? The zombie firm literature has focused almost exclusively on the for-profit sector, where creative destruction operates through market competition \citep{caballero2008zombie, mcgowan2018walking}. The nonprofit sector lacks this competitive mechanism---donors, volunteers, and community networks create complementarities that connect organizations, including dormant ones, to the ecosystem's vitality. Mass administrative culling may sever these connections in ways that have no analog in competitive markets.

The paper contributes to the nascent empirical literature on nonprofit organizational dynamics \citep{harrison2010does, calabrese2013running} and to the policy debate over IRS enforcement of tax-exempt organizations \citep{boris2010scope}. No prior study estimates the causal effects of mass revocation on local nonprofit ecosystems. This paper also contributes to the creative destruction literature by documenting a setting where the standard prediction fails: removing zombies does not help the survivors.

\section{Institutional Background}

\subsubsection*{The Pension Protection Act filing mandate}
Prior to 2006, tax-exempt organizations with annual gross receipts below \$25,000 were not required to file any annual return with the IRS. This exemption covered hundreds of thousands of small organizations---churches, PTAs, fraternal lodges, volunteer fire departments---that had obtained tax-exempt status decades earlier and never subsequently interacted with the IRS. The Pension Protection Act (PPA) of 2006, signed into law on August 17, 2006, changed this by requiring \emph{all} tax-exempt organizations to file an annual electronic notice (Form 990-N, or ``e-Postcard'') beginning with tax years starting after December 31, 2006.

\subsubsection*{Automatic revocation}
Section 6033(j) of the Internal Revenue Code, as amended by the PPA, mandates that any organization failing to file the required return or notice for three consecutive years shall have its tax-exempt status \emph{automatically revoked}. The statute provides no discretion: revocation is mechanical once the three-year window closes. Revoked organizations lose their listing in IRS Publication 78 (the database donors use to verify deductibility), cannot receive tax-deductible contributions, and must reapply for exemption.

\subsubsection*{The 2011 wave and subsequent enforcement}
The first three-year window closed in May 2010, and the IRS began posting automatic revocations in June 2010. The largest single wave occurred on June 8, 2011, when approximately 275,000 organizations were revoked simultaneously. Subsequent waves have continued annually, with cumulative revocations exceeding 800,000 by 2020. The geographic distribution of revocations reflects the historical pattern of nonprofit registration---denser in the South and Midwest, where small churches and fraternal organizations proliferated in the mid-20th century.

\subsubsection*{Reinstatement}
Revoked organizations can apply for reinstatement, but the process is costly: organizations must file a new application (Form 1023 or 1024) and pay the associated user fee. Reinstatement is retroactive only if the organization demonstrates reasonable cause for non-filing. As of 2020, fewer than 15 percent of revoked organizations had been reinstated, suggesting that the vast majority were either truly defunct or faced prohibitive transaction costs.

\section{Data}

I combine four administrative data sources to construct a county-year panel covering 2006--2020.

\subsubsection*{IRS Auto-Revocation List}
The IRS publishes a complete list of all organizations whose tax-exempt status has been automatically revoked, including EIN, organization name, ZIP code, revocation date, and exemption type. As of March 2026, the list contains 1,203,394 records. I restrict to revocations occurring between 2010 and 2020 (918,791 records) and geocode organizations to counties using a ZIP-to-county crosswalk from the 2010 Census ZCTA-County Relationship File. Approximately 90 percent of revocations match successfully.

\subsubsection*{IRS Exempt Organizations Business Master File}
The EO BMF is a monthly extract of all organizations recognized as tax-exempt by the IRS, containing 1,938,732 records. Each record includes a ruling date indicating when the IRS granted exempt status. I use ruling dates to construct county-year counts of new nonprofit formations and pre-2010 baseline counts as the denominator for the treatment variable.

\subsubsection*{Census Quarterly Workforce Indicators}
QWI data for NAICS 813 (Religious, Grantmaking, Civic, Professional, and Similar Organizations) provide county-quarter measures of employment and average earnings in the nonprofit sector. I aggregate to annual frequency using Q1 values for consistency and cover 50 states (excluding DC due to parsing issues with its single-county structure).

\subsubsection*{IRS Statistics of Income County Data}
SOI county income data provide annual measures of charitable deductions (total amount and number of returns claiming the deduction) for tax years 2011--2020. Pre-period data (2006--2010) are not available in the current release, limiting this outcome to cross-sectional analysis.

\subsubsection*{Treatment variable}
Revocation intensity for county $c$ is defined as:
\begin{equation}
\text{RevocationIntensity}_c = \frac{\text{Number of nonprofits revoked 2010--2012 in county } c}{\text{Number of registered nonprofits pre-2010 in county } c}
\end{equation}

\noindent I restrict the sample to counties with at least 10 pre-2010 registered nonprofits to avoid extreme ratios in sparsely populated areas. The resulting sample contains 3,060 counties. The median revocation intensity is 0.47, meaning that in the typical county, nearly half of all registered nonprofits lost their tax-exempt status.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

\subsubsection*{Identification}
The primary specification is a continuous-treatment difference-in-differences:
\begin{equation}
Y_{ct} = \alpha_c + \gamma_t + \beta \cdot \text{RevocationIntensity}_c \times \text{Post}_t + X_{ct}'\delta + \varepsilon_{ct}
\end{equation}

\noindent where $Y_{ct}$ is the outcome in county $c$ and year $t$; $\alpha_c$ and $\gamma_t$ are county and year fixed effects; $\text{Post}_t$ indicates years 2011 and later; and $X_{ct}$ includes log population. Standard errors are clustered at the county level.

The identifying assumption is that, absent the mass revocation, counties with different revocation intensities would have followed parallel trends in formation, employment, and giving. Revocation intensity is driven by the \emph{historical} composition of each county's nonprofit sector---specifically, the share of very small organizations that were exempt from filing prior to 2006 and subsequently failed to comply with the new mandate. This is plausibly orthogonal to contemporaneous trends in nonprofit formation conditional on county and year fixed effects.

\subsubsection*{Event study}
To test the parallel trends assumption and trace the dynamics of the treatment effect, I estimate an event study specification interacting revocation intensity with year dummies, omitting the year before the first mass revocation (2010) as the reference period:
\begin{equation}
Y_{ct} = \alpha_c + \gamma_t + \sum_{k \neq -1} \beta_k \cdot \text{RevocationIntensity}_c \times \mathbb{1}[t - 2011 = k] + \varepsilon_{ct}
\end{equation}

\subsubsection*{Threats}
The main concern is that counties with more zombie nonprofits may differ systematically. I address this through four checks: (1) placebo tests assigning a fake treatment date of 2009 to the pre-period only; (2) trimming the top and bottom deciles of revocation intensity; (3) binary treatment (above/below median); and (4) controlling for pre-period mean formation rates interacted with post.

\section{Results}

\subsubsection*{New nonprofit formations}
\Cref{tab:formations} presents the main DiD estimates for new formations per 10,000 county population. The baseline specification (column 1) yields $\hat{\beta} = -0.192$ ($p = 0.03$): a one-unit increase in revocation intensity---corresponding roughly to revoking all of a county's pre-2010 nonprofits---reduces new formations per 10,000 population by 0.19 per year. The estimate is stable when controlling for log population (column 2) and attenuates modestly with county-specific linear trends (column 3), though the trend specification substantially inflates standard errors.

\input{tables/tab2_formations.tex}

\subsubsection*{Event study dynamics}
\Cref{tab:event_study} reports event study coefficients, which reveal a rich dynamic pattern obscured by the average post-period effect. Pre-treatment coefficients for years $-5$ through $-3$ are small and statistically insignificant, supporting the parallel trends assumption. The coefficient at $t = -2$ is marginally significant and negative, which I discuss below.

The most striking feature is the temporal reversal. In the year of mass revocation ($t = 0$), the coefficient is large and positive (1.007, $p < 0.001$), indicating that counties with higher revocation intensity experienced a burst of new formations---consistent with creative destruction opening space for new entrants. This positive effect persists at $t = 1$ (0.560, $p < 0.001$). But by $t = 2$ and $t = 3$, the sign flips sharply: coefficients of $-2.083$ ($p < 0.001$) and $-1.805$ ($p < 0.001$) indicate that the initial burst was more than fully offset. By $t = 4$ through $t = 9$, effects return to near zero, suggesting a medium-run disruption that eventually dissipates.

\input{tables/tab4_event_study.tex}

This inverted-U pattern---an initial burst followed by a larger decline---is consistent with a ``false spring'' mechanism: mass revocation temporarily creates visible gaps in the organizational landscape, prompting new entry. But the same shock damages the institutional infrastructure (donor networks, grant pipelines, community awareness of the sector) that sustains new organizations, leading to a net negative effect once the initial gap-filling dissipates. An important concern is whether the $t = 0$ spike reflects genuine new organizations or merely reinstated entities counted as new formations. This is unlikely to explain the pattern: organizations reinstated after automatic revocation typically retain their original EIN and receive an effective date of reinstatement rather than a new ruling date in the BMF. The formation measure, based on ruling dates, should therefore capture genuinely new exempt organizations rather than administrative reinstatements.

The marginally significant $t = -2$ coefficient ($-0.369$, $p = 0.046$) warrants careful discussion. Three of the four pre-treatment coefficients ($t = -5, -4, -3$) are statistically and economically indistinguishable from zero, supporting parallel trends over the bulk of the pre-period. The $t = -2$ departure, while statistically marginal, could reflect anticipation: as the three-year clock wound down and the IRS publicized the coming enforcement wave in 2009--2010, some prospective founders may have delayed formation. Alternatively, the single significant lead out of four is consistent with a 5-percent false positive rate under the null. The controlled specification in \Cref{tab:robustness} (column 4), which interacts pre-period mean formation rates with post, strengthens the main effect to $-0.371$ ($p < 0.001$) while absorbing exactly this kind of differential pre-trend, reinforcing the causal interpretation.

\subsubsection*{Employment and earnings}
\Cref{tab:giving_employment} reports results for nonprofit sector employment and, separately, \emph{descriptive} associations between revocation intensity and charitable giving. Log nonprofit employment (NAICS 813) declines significantly in the baseline specification ($\hat{\beta} = -0.131$, $p = 0.009$). The effect attenuates to $-0.072$ ($p = 0.15$) when controlling for population, suggesting some of the employment decline reflects broader population dynamics in high-revocation counties. The charitable giving results (columns 1--3) are purely cross-sectional because IRS SOI county data are unavailable before 2011, precluding a DiD design for this outcome. Conditional on state and year fixed effects, post-period charitable deductions per return are negatively associated with revocation intensity ($\hat{\beta} = -0.374$, $p < 0.001$). This association is consistent with the collateral damage hypothesis but cannot be given a causal interpretation.

\input{tables/tab3_giving_employment.tex}

\subsubsection*{Robustness}
\Cref{tab:robustness} presents four robustness checks for the formation outcome. The placebo test applying a fake 2009 treatment date to the pre-period yields a small and insignificant coefficient ($-0.164$, $p = 0.13$), supporting the parallel trends assumption. Trimming the top and bottom deciles of revocation intensity attenuates the estimate and renders it insignificant ($-0.128$, $p = 0.36$), suggesting that extreme-intensity counties contribute disproportionately to the effect. Binary treatment (above/below median) confirms the direction ($-0.058$, $p = 0.049$). The controlled specification adding pre-period mean formations interacted with post \emph{strengthens} the main effect to $-0.371$ ($p < 0.001$), suggesting that failing to account for baseline formation rates biases the main estimate toward zero.

\input{tables/tab5_robustness.tex}

\section{Discussion}

These results offer a clear answer to the creative destruction question in the nonprofit context: removing zombie organizations does not help the survivors. The initial burst of new entry is a short-lived response to visible gaps in the organizational landscape. Once these gaps are filled, the deeper damage to the sector's institutional infrastructure dominates: donor networks are disrupted, community awareness of the sector declines, and the administrative burden of mass reinstatement falls disproportionately on the smallest, most resource-constrained organizations.

The finding that employment declines while per-worker earnings do not change suggests that the adjustment operates on the extensive margin: organizations shed positions or fail to grow, rather than reducing compensation. This is consistent with nonprofit labor markets where wages are rigid due to grant-funded salary structures and collective bargaining agreements in some sub-sectors.

Two caveats are important. First, the revocation intensity measure captures the \emph{share} of registered nonprofits revoked, which conflates organizational density with zombie prevalence. Counties with high revocation intensity may have had systematically different nonprofit sectors---older, more reliant on small volunteer organizations, less professionalized---rather than simply more deadwood. County fixed effects absorb time-invariant differences, but time-varying correlates cannot be ruled out. Second, the formations outcome measures IRS ruling dates, which reflect bureaucratic processing rather than actual organizational founding. If the IRS prioritized processing in low-revocation regions, the negative coefficient could reflect queue effects rather than real changes in entrepreneurial activity.

The policy implication is straightforward but counterintuitive: maintaining dormant organizations on the registry may be costless, while removing them is not. The ``clean registry'' objective that motivated automatic revocation assumed that zombie nonprofits impose costs through donor confusion and administrative burden. This paper finds no evidence of such costs and substantial evidence of unintended harm. A lighter-touch approach---suspending rather than revoking, or providing grace periods with outreach---might achieve the registry maintenance objective without disrupting local charitable ecosystems.

\section{Conclusion}

The IRS's mass revocation of tax-exempt status for non-filing nonprofits---affecting over 800,000 organizations since 2010---provides a rare natural experiment in the economics of organizational ecology. Exploiting cross-county variation in revocation intensity, I find that mass de-registration creates an initial illusion of creative destruction, with a burst of new formations in high-revocation counties. But this reverses within two years as the deeper damage to the sector's ecosystem dominates. The average post-period effect is negative for both formation and employment, with standardized effects in the ``small negative'' range. These findings extend the zombie firm literature to the nonprofit sector and suggest that administrative enforcement designed to ``clean up'' registries can impose real costs on the communities those organizations serve.

\label{apep_main_text_end}

\newpage
\singlespacing
\bibliography{references}

\newpage
\onehalfspacing

\appendix
\section{Standardized Effect Sizes}

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
