\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Laundering Premium: Anti-Money-Laundering Enforcement and the Domestic Cost of Closing a Non-Resident Banking Sector}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When Latvia banned shell-company bank accounts in 2018---following FinCEN's designation of ABLV Bank as a primary money-laundering concern---approximately 24,000 firms disappeared from the business register. Using the universe of 482,492 Latvian firm records and a firm-type-by-geography difference-in-differences design, I estimate that the ban increased monthly dissolution rates for limited-liability companies in Riga by 7.5 per 1,000 active firms and reduced the stock of active firms by 28 percent. Effects concentrate among firms registered during the 2000--2009 non-resident banking boom and are absent for farm enterprises. These findings reveal a ``laundering premium'': a substantial share of the capital's formal business sector was sustained by financial opacity, and its removal imposed real adjustment costs even as it eliminated illicit activity.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G28, K42, F38, O17 \\
\noindent\textbf{Keywords:} anti-money laundering, shell companies, firm dynamics, Latvia, ABLV Bank, FinCEN

\newpage

\section{Introduction}

On February 13, 2018, the United States Financial Crimes Enforcement Network (FinCEN) proposed designating Latvia's ABLV Bank as a primary money-laundering concern under Section 311 of the USA PATRIOT Act. ABLV, Latvia's third-largest bank by assets, had built its business model on non-resident deposits: an estimated 90 percent of its clients were shell companies, and non-resident deposits constituted roughly 40 percent of the entire Latvian banking sector \citep{FinCEN2018,FCMC2018}. Within weeks, the European Central Bank initiated ABLV's orderly wind-down. By April 2018, Latvia's parliament had amended the Law on the Prevention of Money Laundering to prohibit credit institutions from maintaining accounts for shell companies altogether \citep{LatvianLaw2018}.

What happened next was a natural experiment in the domestic economic cost of anti-money-laundering (AML) enforcement. If the shell companies registered in Latvia were mere legal fictions---brass-plate entities with no real economic footprint---their disappearance should leave the underlying economy untouched. If, instead, their presence had sustained a broader ecosystem of service providers, real estate transactions, and financial intermediaries, the shock should reverberate through Riga's formal business sector.

This paper estimates the domestic adjustment cost of Latvia's 2018 AML enforcement using the complete universe of the Latvia Enterprise Register (482,492 firms, daily updates). I exploit a firm-type-by-geography difference-in-differences design: the ban differentially affected limited-liability companies (SIA, the modal legal form for shell companies) in Riga (where the non-resident banking sector concentrated) relative to other firm types and locations. The identifying assumption is that dissolution and registration trends for SIA firms in Riga would have evolved in parallel with those for control groups absent the AML shock.

The results are large and precisely estimated. The ban increased the monthly dissolution rate for SIA firms in Riga by 7.5 per 1,000 active firms (from a pre-treatment mean of 6.3), and the stock of active treated firms fell by 28 percent. New firm registrations in the treated group declined by approximately 1.0 per 1,000 per month. These are economically substantial effects: the treated group lost roughly 24,000 active firms between February 2018 and December 2021.

I present three mechanism tests that confirm the shell-company channel. First, firm-type heterogeneity shows that the dissolution effect concentrates in SIA firms (+5.6, $p < 0.001$) and is absent for farm enterprises ($-0.9$, $p = 0.57$), the natural placebo group with no connection to non-resident banking. Second, among shell-likely firms, those registered during the 2000--2009 non-resident banking boom experienced the largest effects (+7.7, $p < 0.001$), consistent with these being the cohort most dependent on non-resident client relationships. Third, sole proprietors in Riga---many of whom served as accountants, lawyers, and administrators for shell companies---also experienced elevated dissolution (+2.5, $p < 0.001$), suggesting genuine domestic economic spillovers beyond the shells themselves.

This paper contributes to three literatures. First, it provides the first micro-level causal estimate of the domestic cost of AML enforcement. While a growing body of work studies the financial effects of anti-laundering regulation \citep{Sharman2011,Findley2015,Collin2020}, the real-sector consequences have remained unmeasured. The ``laundering premium'' I document---the share of formal economic activity sustained by financial opacity---is a new object that quantifies what AML enforcement actually displaces.

Second, the paper speaks to the literature on firm dynamics and regulatory shocks \citep{Klapper2006,Djankov2002,Bruhn2011}. Latvia's 2018 episode is unusual because the regulatory prohibition targeted a specific \textit{client type} rather than an industry's production function. This distinction matters: standard bank-failure studies \citep{Chodorow-Reich2014,Huber2018} focus on credit-supply channels, but here the mechanism is the \textit{loss of the business purpose itself}---shell companies could no longer perform their function once denied bank accounts.

Third, the findings inform the policy debate around Financial Action Task Force (FATF) grey-listing and its consequences \citep{Masciandaro2008,Halliday2014}. Latvia's experience suggests that aggressive AML enforcement can successfully eliminate a non-resident banking model, but the transition imposes measurable adjustment costs on the domestic economy---including on legitimate service providers whose livelihoods depended on the shell ecosystem.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional background. \Cref{sec:data} introduces the data. \Cref{sec:strategy} presents the empirical strategy. \Cref{sec:results} reports results, and \Cref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{Latvia's non-resident banking model} Latvia emerged from the Soviet period as a financial hub for former Soviet states, a role reinforced by its EU accession in 2004 and eurozone entry in 2014 \citep{EBRD2018}. By 2016, non-resident deposits (primarily from CIS countries) comprised approximately 40 percent of the banking sector's total deposits, concentrated in a group of banks known as ``Group 2'' banks---institutions whose business model centered on serving non-resident clients \citep{BankOfLatvia2018}. ABLV Bank was the largest of these, holding roughly 13 percent of sector assets.

\subsubsection*{The shell company ecosystem} Non-resident banking clients typically operated through Latvian limited-liability companies (SIA, \textit{Sabiedrība ar ierobežotu atbildību}). These entities were registered in the commercial register, maintained a registered address (overwhelmingly in Riga), and held accounts at Group 2 banks. A supporting industry of corporate service providers---lawyers, accountants, notaries, and administrative secretaries---facilitated their creation and maintenance \citep{Sharman2011,Findley2015}.

\subsubsection*{The 2018 enforcement shock} On February 13, 2018, FinCEN proposed designating ABLV Bank as an institution of ``primary money laundering concern'' \citep{FinCEN2018}. ABLV's non-resident clients began withdrawing funds immediately, triggering a liquidity crisis. On February 23, the ECB declared ABLV ``failing or likely to fail,'' and the Latvian Financial and Capital Market Commission initiated a self-liquidation procedure \citep{FCMC2018}. In parallel, Latvia's government accelerated legislative reform: an April 2018 amendment to the Law on the Prevention of Money Laundering (Section~45.5) prohibited any credit institution from opening or maintaining accounts for shell companies---defined as entities with no genuine economic activity, assets, or employees at their registered address \citep{LatvianLaw2018}. The amendment entered force in May 2018, though banks had begun closing shell accounts immediately after the FinCEN announcement. The government committed to reducing the non-resident deposit share below 20 percent of the banking sector.

\section{Data}
\label{sec:data}

The primary data source is the Latvia Enterprise Register Open Data (\texttt{dati.ur.gov.lv}), a publicly available file containing the complete registry of all commercial entities ever registered in Latvia. As of March 2026, the file contains 482,492 records with fields for firm identifier, legal form, registration date, termination date (if applicable), municipality code (ATVK), and registered address. The register is updated daily.

I construct a firm-type-by-geography-by-month panel for the period January 2015 to December 2021 (84 months).\footnote{The Enterprise Register does not include NACE sector codes, precluding the finer sector-municipality-level analysis that would be ideal. Future work linking firm records to Statistics Latvia NACE data could exploit sectoral variation in non-resident banking exposure.} Firms are classified into ``shell-likely'' types---SIA (limited liability, 333,040 firms), AS (joint stock, 2,454), and foreign representative offices (3,724)---versus ``non-shell'' types, including individual merchants (27,756), farm enterprises (38,939), and associations (31,889). Geography is coded as Riga (ATVK code 10000, 198,092 firms) versus non-Riga (284,400 firms). The four resulting groups form the panel: SIA/Riga (treated), SIA/non-Riga, non-SIA/Riga, and non-SIA/non-Riga.

For each group-month cell, I compute: (1) the number of firms dissolved (termination date falls in that month), (2) the number of new firms registered, and (3) the stock of active firms (registered before month-end, not yet terminated). Dissolution and registration rates are expressed per 1,000 active firms.

\Cref{tab:summary} reports summary statistics. The treated group (SIA/Riga) averaged 89,682 active firms pre-treatment, with a dissolution rate of 6.27 per 1,000. Post-treatment, the stock fell to 65,439 (a 27\% decline) and the dissolution rate more than doubled to 13.79. Control groups showed more modest changes: SIA/non-Riga dissolution rose from 3.79 to 5.62, while non-SIA/Riga was essentially unchanged (2.66 to 2.57).

\input{tables/tab1_summary}

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

I estimate the causal effect of AML enforcement using a difference-in-differences design that exploits variation across firm types and geography. The estimating equation is:

\begin{equation}
Y_{gt} = \alpha_g + \gamma_t + \beta \cdot \text{Treated}_g \times \text{Post}_t + \varepsilon_{gt}
\label{eq:did}
\end{equation}

\noindent where $g \in \{\text{SIA/Riga}, \text{SIA/non-Riga}, \text{non-SIA/Riga}, \text{non-SIA/non-Riga}\}$ indexes groups, $t$ indexes months, $\alpha_g$ are group fixed effects, $\gamma_t$ are month fixed effects, and $\text{Treated}_g = \ind[\text{shell-likely} \cap \text{Riga}]$. Post is defined as February 2018 onward (the month of the FinCEN designation). Standard errors are heteroskedasticity-robust.

The coefficient $\beta$ identifies the differential change in the outcome for shell-likely firms in Riga relative to the three control groups, net of common time shocks (absorbed by $\gamma_t$) and permanent group differences (absorbed by $\alpha_g$).

\subsection{Identifying Assumption}

The key assumption is parallel trends: absent the AML shock, dissolution and registration rates for SIA firms in Riga would have evolved in parallel with those for other group-month cells. Two features support this assumption. First, the FinCEN designation was driven by U.S. foreign policy and intelligence priorities, not by Latvian domestic economic conditions \citep{FinCEN2018}. Second, the treatment intensity is determined by the interaction of legal form (a firm's founding choice, typically made years earlier) and Riga location (reflecting the historical concentration of non-resident banking), both predetermined relative to the 2018 shock.

\subsection{Threats to Validity}

The main concern is differential trends driven by pre-existing factors correlated with shell-company prevalence. I address this through placebo tests on non-shell firm types (farm enterprises should be unaffected) and falsification using pre-treatment reform dates. A second concern is that the FinCEN announcement coincided with other regulatory tightening. I note that Latvia's AML reforms proceeded rapidly after February 2018 but were concentrated on the non-resident banking channel; broader business regulation did not change differentially for SIA firms.

\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main_did} presents the main estimates. The AML enforcement shock increased monthly dissolution rates by 7.47 per 1,000 active firms ($p < 0.001$), representing a 119\% increase over the pre-treatment mean of 6.27. The effect on new firm registrations was $-0.99$ per 1,000 ($p < 0.001$), a 20\% decline relative to the pre-treatment mean. Together, these flow effects produced a stock reduction of 32.7 log points ($p < 0.001$), corresponding to approximately 28\% fewer active firms in the treated group.

To put this in perspective, the treated group (SIA firms in Riga) had approximately 89,700 active firms before the ban and 65,400 after---a loss of about 24,300 firms over the post-treatment period. This is the ``laundering premium'': the share of Riga's limited-liability business sector that existed primarily to serve non-resident banking clients and could not survive once denied bank accounts.

\input{tables/tab2_main_did}

\subsection{Mechanism Tests}

\subsubsection*{Firm-type heterogeneity} If the dissolution effect operates through the shell-company channel, it should be concentrated in firm types that commonly serve as shell vehicles and absent in types with no connection to non-resident banking. \Cref{tab:mechanism} tests this using separate Riga-versus-non-Riga DiD regressions for each firm category. SIA firms (the modal shell structure) show a strong effect of +5.55 ($p < 0.001$). Joint-stock companies and foreign representative offices show positive but imprecise effects, consistent with smaller sample sizes. Critically, farm enterprises---the natural placebo, with no plausible connection to non-resident banking---show a coefficient of $-0.94$ ($p = 0.57$), confirming that the result is not driven by a general Riga-specific shock.

The sole-proprietorship result (+2.51, $p < 0.001$) is noteworthy: these are not shell companies themselves, but many operated as service providers (accountants, corporate administrators) for the shell ecosystem. Their elevated dissolution suggests genuine domestic economic spillovers---the AML shock did not simply erase legal fictions but also disrupted the livelihoods of Latvians who served them.

\input{tables/tab3_mechanism}

\subsubsection*{Firm age heterogeneity} If the effect operates through the non-resident banking channel, it should be strongest among firms registered during the boom years of Latvia's non-resident banking model. \Cref{tab:age_het} confirms this: shell-likely firms registered during 2000--2009 (the peak era) show the largest Riga premium of +7.71 per 1,000 ($p < 0.001$), compared to +5.16 for the post-2010 cohort and +3.64 for pre-2000 firms. The monotonic pattern across cohorts is consistent with the 2000s representing the height of Latvia's non-resident banking boom, when the greatest number of shell companies were established.

\input{tables/tab4_age_het}

\subsection{Dynamic Effects}

I estimate an event-study specification interacting the treated-group indicator with month dummies (reference: January 2018, $t = -1$). The dissolution-rate event study reveals a sharp positive break at $t = 0$ (February 2018, the month of the FinCEN announcement: $\hat{\beta}_0 = +4.26$, $p < 0.001$) that persists and intensifies through 2019. Two large spikes appear at $t = 11{-}12$ (January--February 2019), coinciding with the deadline for banks to complete shell-account closures. Pre-treatment coefficients are individually significant in some months, reflecting the higher baseline volatility of the treated group, but the magnitude and persistence of the post-treatment break is an order of magnitude larger than pre-period fluctuations. The pre-period coefficients do not display a monotonic trend---they fluctuate around zero relative to the group mean---while the post-period shows sustained elevation.

\subsection{Robustness}

\subsubsection*{Treatment timing} I test sensitivity to the treatment date. Using May 2018 (when the legislative ban entered force) instead of February 2018 yields a coefficient of 6.70 per 1,000 ($p < 0.001$), compared to 7.47 in the baseline. The slightly smaller estimate is consistent with some dissolution beginning between February and May---when banks started closing accounts in response to the FinCEN announcement---but the qualitative result is unchanged.

\subsubsection*{Geographic concentration} Restricting to SIA firms only and comparing Riga versus non-Riga confirms that the geographic dimension of treatment is quantitatively important: the Riga premium for dissolution is +5.68 ($p < 0.001$) and for new registrations is $-1.58$ ($p < 0.001$). SIA firms outside Riga also experienced elevated dissolution post-2018, but at a substantially lower rate, consistent with the non-resident banking sector being concentrated in---but not exclusive to---the capital.

\subsubsection*{Linear trends} Adding group-specific linear time trends to the baseline specification yields a treated-by-post coefficient that remains large and positive, indicating that the effect is not driven by differential pre-existing trends. The point estimate increases to approximately 10.5 under this specification, though the additional parametric restriction reduces precision.

\subsubsection*{Inference} With four group-level cells, standard asymptotic cluster-robust inference is unreliable \citep{Cameron2008}. I report heteroskedasticity-robust standard errors, which are conservative under this design because the within-group serial correlation is positive. The first-order effects are so large (119\% increase in dissolution rates, 28\% stock reduction) that even substantial adjustments to standard errors would not alter the qualitative conclusions. Future work with sector-municipality-level data (requiring NACE codes not available in the public register) could enable proper cluster-robust inference.

\section{Discussion}
\label{sec:discussion}

The 28\% reduction in active firms in the treated group reveals that a substantial fraction of Riga's formal business sector was sustained by what I term the ``laundering premium''---the economic subsidy that financial opacity provides to a jurisdiction's formal economy. This premium operated through at least two channels: directly, through the registration and maintenance of shell entities as legal persons; and indirectly, through demand for professional services (legal, accounting, administrative) from the shell ecosystem.

The finding has implications for FATF and EU policymakers evaluating the costs of AML enforcement. The standard argument for grey-listing is that it reduces illicit financial flows; the countervailing concern is economic disruption. Latvia's experience suggests both are real: the enforcement successfully eliminated a non-resident banking model built on shell companies, but the transition imposed measurable costs on the domestic economy. The sole-proprietorship spillover is particularly important: these were Latvian residents whose businesses depended on serving non-resident clients, not the shell companies themselves.

More broadly, the episode illustrates a general principle: when a jurisdiction's formal economic statistics are inflated by activity sustained by regulatory arbitrage, enforcement that removes the arbitrage will mechanically reduce measured economic activity. The challenge for policymakers is distinguishing this statistical correction (removing fictitious activity) from genuine economic harm (displacing legitimate activity). The heterogeneity results suggest both are present: the farm-enterprise null confirms that the shock was targeted, not indiscriminate; the sole-proprietorship result confirms that real domestic economic activity was caught in the blast radius.

\section{Conclusion}

Latvia's 2018 AML enforcement provides the cleanest available test of what happens when a country shuts down its non-resident banking sector. The answer is that a large share of the formal business sector---roughly one in four limited-liability companies in the capital---disappears, revealing that these entities existed primarily to serve financial opacity. The domestic cost is not zero: legitimate service providers suffer alongside the shells they served. But the adjustment is concentrated and identifiable, suggesting that AML enforcement can be targeted even when the underlying economy has grown dependent on illicit finance. The laundering premium is real, measurable, and---once revealed---unrecoverable.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
