\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Shell Games at the Municipal Border: Do Corporate Tax Cuts Attract Real Activity or Letterbox Companies?}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When a Swiss municipality cuts its corporate tax rate, do real firms follow---or just their letterheads? I exploit within-municipality variation in corporate tax multipliers across 246 municipalities in cantons Zurich and Basel-Landschaft (2011--2023), using the establishment-by-employment decomposition from Switzerland's STATENT census. A $\geq$5 percentage-point corporate Steuerfuss cut increases the number of establishments by 2.0\% but leaves employment unchanged, reducing employment per establishment by 2.1\%. The effect is three times larger in the tertiary sector ($-$6.6\%), where holding companies and letterbox entities cluster. Natural-person tax cuts---a placebo---show no effect. A dose-response pattern confirms the mechanism: larger cuts attract proportionally more employment-light entities. These results suggest that subnational corporate tax competition partly operates as a zero-sum relabeling of paper entities rather than a reallocation of real economic activity.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H25, H71, H73, R30 \\
\noindent\textbf{Keywords:} tax competition, corporate taxation, letterbox companies, Steuerfuss, Switzerland

\newpage

\section{Introduction}

In Zug, one of the world's most aggressive tax havens, each establishment employs an average of 6.1 workers. In Zurich, 20 kilometers away, the figure is 11.5. This gap---nearly double the employment intensity---exists despite Zug's spectacular success at attracting thousands of corporate registrations. It raises a question at the heart of the tax competition debate: when municipalities cut corporate taxes, are they attracting real economic activity, or just relocating letterheads?

The theoretical foundations of tax competition, laid by \citet{wilson1986theory} and \citet{zodrow1986pigou}, predict that jurisdictions competing for mobile capital will underprovide public goods. But the welfare consequences depend critically on what moves in response to tax differentials. If a corporate tax cut attracts a factory with 500 workers, the host municipality gains employment, income, and a broader tax base. If it attracts a holding company with two employees and a registered address, the host gains tax revenue siphoned from elsewhere without any real reallocation of production \citep{hines1994fiscal, desai2006demand}.

This distinction---real activity versus paper entities---has been central to the international profit-shifting literature. \citet{torslov2023missing} estimate that 36\% of multinational profits are shifted to tax havens, generating ``missing profits'' in high-tax countries. \citet{dharmapala2008what} documents the proliferation of shell structures in Caribbean tax havens. Yet almost all evidence on this margin comes from cross-country comparisons, where identification is confounded by differences in institutions, legal systems, and regulatory environments. Whether the same dynamic operates \textit{within} countries---where firms relocate across municipal borders under identical national law---remains an open question.

This paper provides the first subnational test of the real-activity-versus-letterbox distinction. I exploit the unique structure of Swiss fiscal federalism, where over 2,100 municipalities independently set corporate tax multipliers (Steuerfuss) that vary from 50\% to 150\% of the cantonal base rate \citep{brulhart2012agglomeration, brulhart2015taxes}. Crucially, I leverage Switzerland's STATENT census of establishments, which records \textit{both} the number of establishments and the number of employees at each establishment, at annual frequency and municipal granularity. This decomposition allows me to distinguish between the extensive margin (how many firms register) and the intensive margin (how many workers each firm brings).

My identification strategy uses within-municipality variation in the corporate Steuerfuss. I define treatment as a $\geq$5 percentage-point cut in the corporate tax multiplier and estimate two-way fixed effects regressions with municipality and year fixed effects. The panel covers 246 municipalities in cantons Zurich and Basel-Landschaft from 2011 to 2023, with 25 municipalities experiencing large tax cuts during this period.

The results are striking. A corporate tax cut increases the number of establishments by 2.0\% ($p < 0.10$) but has no detectable effect on total employment ($-$0.2\%, $p = 0.93$). The mechanical consequence: employment per establishment falls by 2.1\% ($p < 0.05$). The new entities are employment-light---consistent with letterbox structures rather than productive firms.

The mechanism test sharpens this finding. I decompose the effect by sector, exploiting the fact that letterbox companies---holding structures, IP vehicles, management shells---cluster overwhelmingly in the tertiary sector \citep{desai2006demand}. The effect on employment per establishment in the tertiary sector is $-$6.6\% ($p < 0.01$), three times the total effect. By contrast, the secondary sector (manufacturing)---where physical production precludes letterbox structures---shows no significant effect ($-$1.6\%, $p = 0.30$). The tertiary share of establishments rises by 0.5 percentage points ($p < 0.01$).

Three robustness checks support a causal interpretation. First, natural-person Steuerfuss cuts serve as a placebo: they should not affect corporate location decisions, and indeed show no significant effect ($-$1.3\%, $p = 0.38$). Second, a dose-response pattern emerges across thresholds: $\geq$3pp cuts reduce employment intensity by 1.5\%, $\geq$5pp cuts by 2.1\%, and $\geq$10pp cuts by 5.2\%. Third, the leave-one-out exercise shows the estimate is stable across all 25 treated municipalities, ranging from $-$1.4\% to $-$3.2\%.

This paper contributes to three literatures. First, it advances the corporate tax competition literature \citep{devereux2008business, slemrod2004tax} by providing the first subnational decomposition of tax-induced firm relocation into real activity and paper entities. \citet{krapf2021taxincidence} estimate a taxable income elasticity of 3.5 for Swiss municipalities but cannot decompose this into real versus accounting responses. My STATENT-based approach resolves this ambiguity. Second, it contributes to the literature on profit shifting and tax havens \citep{torslov2023missing, dharmapala2008what} by showing that the same letterbox dynamics documented internationally also operate within countries, under identical legal and institutional frameworks. Third, it informs the OECD Pillar Two debate \citep{oecd2021pillartwo} by demonstrating that minimum tax rules may need to address subnational competition alongside cross-border profit shifting.

\section{Institutional Background}

\subsubsection*{Swiss fiscal federalism} Switzerland's tax system is among the most decentralized in the world \citep{oates1972fiscal, agrawal2021fiscal}. The federal government, 26 cantons, and over 2,100 municipalities each levy separate taxes. For corporate income, the canton sets a base rate, and each municipality applies its own multiplier---the Steuerfuss---to determine the effective tax burden. This multiplier is set annually by the municipal legislature, creating frequent, discrete changes in corporate tax rates \citep{brulhart2012agglomeration}.

\subsubsection*{The Steuerfuss mechanism} In canton Zurich, the Steuerfuss for ``juristische Personen'' (legal entities) typically ranges from 80\% to 130\% of the cantonal base rate. A municipality with a Steuerfuss of 80\% charges its corporations 80\% of the cantonal rate; one with 120\% charges 120\%. When a municipality reduces its Steuerfuss, it immediately lowers the effective tax rate on all corporate profits earned within its borders. Importantly, this applies to the jurisdiction of \textit{registration}, not necessarily the jurisdiction of production.

\subsubsection*{Letterbox incentives} Swiss law permits corporations to register their legal domicile in any municipality, regardless of where their economic activity occurs. A holding company, IP management entity, or treasury center can register in a low-tax municipality while its operations---if it has any---take place elsewhere. The STATENT census records such entities as ``establishments'' (Arbeitsstätten) with minimal or zero ``employment'' (Beschäftigte). This creates a measurable wedge between the extensive margin (registrations) and the intensive margin (workers) that directly reveals the letterbox channel.

\subsubsection*{Variation in the panel} Between 2011 and 2023, 25 of the 246 municipalities in my sample experienced a $\geq$5 percentage-point corporate Steuerfuss cut. These cuts are spread across both cantons and occur at different times, providing staggered variation for a difference-in-differences design.

\section{Data}

I combine three data sources at the municipality-year level.

\subsubsection*{Establishment and employment data} The Swiss Federal Statistical Office's STATENT (Statistik der Unternehmensstruktur) provides an annual census of all establishments in Switzerland, disaggregated by municipality, economic sector, and observation unit. I extract two key variables: the number of establishments (Arbeitsstätten) and total employment (Beschäftigte) for each municipality-year cell. The ratio---employment per establishment---is my primary outcome. I also decompose by sector (secondary = manufacturing; tertiary = services) to isolate the letterbox channel. The data cover 2011--2023 (13 years) for all Swiss municipalities.

\subsubsection*{Tax multiplier data} I obtain corporate Steuerfuss data from the cantonal statistical offices of Zurich and Basel-Landschaft. The Zurich data (``JUR\_PERS'' variable) and Basel-Landschaft data (``Steuerfuss Ertrag juristische Personen'') provide annual corporate tax multipliers for each municipality. I harmonize across cantons using BFS municipality numbers as the join key.

\subsubsection*{Population data} Municipal population data from the BFS population statistics serve as a control variable. I use population at January~1 of each year.

\input{tables/tab1_summary}

\Cref{tab:summary} reports summary statistics. Treated municipalities (those experiencing a $\geq$5pp Steuerfuss cut) have higher baseline tax rates (118.6\% vs.\ 52.9\%), consistent with the mechanical fact that large cuts are only feasible from high initial levels. Employment per establishment is slightly higher in treated municipalities pre-treatment (5.9 vs.\ 5.2), and the tertiary share is similar across groups (80\% vs.\ 80\%).

\section{Empirical Strategy}

I estimate the effect of corporate tax cuts on establishment composition using a two-way fixed effects specification:
\begin{equation}
Y_{it} = \beta \cdot \text{PostCut}_{it} + \alpha_i + \gamma_t + \varepsilon_{it}
\label{eq:main}
\end{equation}
where $Y_{it}$ is the outcome (log employment per establishment, log establishments, or log employment) for municipality $i$ in year $t$. $\text{PostCut}_{it}$ is a binary indicator equal to one for all years at or after the first $\geq$5 percentage-point corporate Steuerfuss cut. $\alpha_i$ are municipality fixed effects absorbing time-invariant municipal characteristics. $\gamma_t$ are year fixed effects absorbing common macroeconomic trends. Standard errors are clustered at the canton level in the baseline specification. Because the sample contains only two cantons, I report municipality-clustered standard errors as an alternative and confirm that the key tertiary-sector result survives this more conservative inference.

A potential concern with TWFE in staggered settings is bias from heterogeneous treatment effects across cohorts \citep{callaway2021difference, sun2021estimating}. In this application, the concern is attenuated because the treatment is an absorbing state (once a municipality cuts, it remains treated) and the event study shows no evidence of dynamic heterogeneity in the pre-treatment period. Nevertheless, the estimates should be interpreted as weighted averages of cohort-specific effects.

\subsubsection*{Identification} The identifying assumption is parallel trends: absent the tax cut, treated and control municipalities would have followed the same trajectory in establishment composition. I provide event-study evidence supporting this assumption, showing no significant pre-trends in the 5 years before treatment.

\subsubsection*{Mechanism test} I decompose the main effect by sector. If tax cuts attract letterbox entities, the effect should concentrate in the tertiary sector (services, holding companies, financial intermediaries) where such structures are feasible, and should be absent in the secondary sector (manufacturing) where physical production anchors employment to location \citep{giroud2019state}.

\subsubsection*{Threats to validity} The main concern is reverse causality: municipalities may cut taxes precisely when they anticipate economic decline, which could independently reduce employment intensity. The event study addresses this by showing no differential pre-trends in the years before treatment. The placebo test using natural-person tax cuts provides additional evidence: if unobserved municipal shocks drove both tax cuts and compositional changes, personal income tax cuts---which often move in tandem with corporate cuts---should show similar effects, but they do not. A second concern is selection: municipalities that cut taxes may differ systematically from controls. The summary statistics show treated municipalities start with higher Steuerfuss (a mechanical precondition for large cuts) but similar tertiary establishment shares, and the municipality fixed effects absorb time-invariant differences. A third concern is that two cantons may not generalize. I discuss external validity in the conclusion.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} reports the main results. Column (1) shows that a corporate tax cut reduces log employment per establishment by 0.021 ($p < 0.05$). Given a pre-treatment mean of 5.91 employees per establishment, this represents a decline of approximately 0.12 workers per firm---modest in absolute terms but economically meaningful when aggregated across the hundreds of new registrations attracted by a tax cut. Column (2) reveals that the entire effect---and more---is driven by the tertiary sector, where employment per establishment falls by 6.6\% ($p < 0.01$). Column (3) shows that the tertiary share of establishments rises by 0.5 percentage points ($p < 0.01$), confirming the compositional shift toward service-sector entities.

The extensive and intensive margins move in opposite directions. Column (4) shows that total establishments increase by 2.0\% ($p < 0.10$)---municipalities successfully attract more registrations. But column (5) shows that total employment is statistically indistinguishable from zero ($-$0.2\%, $p = 0.93$). Tax cuts attract establishments without workers.

\subsubsection*{Event study} The event study for tertiary employment per establishment shows clean pre-trends: coefficients at $t = -5$ through $t = -2$ relative to the reference period ($t = -1$) are $+0.048$, $+0.027$, $-0.001$, and $+0.009$---all statistically insignificant and centered around zero (the largest $t$-statistic is 1.0). The absence of differential pre-trends supports the parallel-trends assumption. Post-treatment, the effect grows monotonically: $-4.3\%$ at $t = 1$, $-5.4\%$ at $t = 2$, $-9.6\%$ at $t = 3$, $-10.4\%$ at $t = 4$, and $-13.9\%$ at $t = 5$. This dynamic pattern---no pre-trend and a gradually widening post-treatment gap---is consistent with incremental relocation of paper entities rather than a compositional shock or mean reversion.

\subsection{Robustness}

\input{tables/tab3_robustness}

\Cref{tab:robustness} presents five robustness checks. Column (1) reproduces the baseline with canton-clustered standard errors. Column (2) clusters at the municipality level, which is more conservative given the small number of cantons; the tertiary-sector effect remains significant at the 1\% level. Column (3) controls for log population to address the concern that tax cuts attract residents who bring different firm types; the estimate is unchanged ($-$1.8\%, $p < 0.10$).

Column (4) reports the placebo test using natural-person Steuerfuss cuts. These personal income tax reductions should not directly affect corporate location decisions. The estimated effect is $-$1.3\% and statistically insignificant ($p = 0.38$), confirming that the main result is specific to corporate taxation.

Column (5) applies the same specification to the secondary sector (manufacturing), where letterbox structures are infeasible. The effect is small and insignificant ($-$1.6\%, $p = 0.30$), confirming the sector-specific mechanism.

\subsection{Dose-Response}

\input{tables/tab4_validity}

\Cref{tab:validity} varies the cut threshold. Larger cuts produce proportionally larger effects: a $\geq$3pp cut reduces employment intensity by 1.5\%, a $\geq$5pp cut by 2.1\%, and a $\geq$10pp cut by 5.2\%. This dose-response pattern is inconsistent with confounding stories and supports a causal interpretation where the magnitude of the tax incentive determines the volume of letterbox relocation.

\section{Discussion}

These results suggest that a substantial portion of subnational corporate tax competition operates through the relocation of paper entities---holding companies, IP vehicles, and management shells that register in low-tax municipalities without bringing meaningful employment. The tertiary sector concentration, the null effect on manufacturing, and the dose-response pattern all point to the same mechanism: tax cuts attract letterheads, not factories.

The economic magnitude is instructive. Consider a municipality with 400 establishments that cuts its corporate Steuerfuss by 5 percentage points. The estimated 2\% increase in establishments implies approximately 8 new registrations. But with zero net employment gain, these 8 entities bring roughly 0 additional workers. The municipality sacrifices revenue on its existing 400 firms---a 5-point cut on a base rate of, say, 8\% reduces the effective corporate rate by approximately 0.4 percentage points---while gaining tax revenue only from the profits these 8 new registrations report. Whether the municipality is a net fiscal winner depends on the profitability of these letterbox entities relative to the revenue forgone on existing firms. For the economy as a whole, the reallocation is zero-sum: each paper entity that registers in the low-tax municipality withdraws its registration---and its reported profits---from elsewhere \citep{fajgelbaum2019state}.

This finding has direct implications for the OECD Pillar Two framework \citep{oecd2021pillartwo}, which establishes a global minimum corporate tax of 15\%. While Pillar Two targets cross-border profit shifting, my results show that the same letterbox dynamics operate within countries. A jurisdiction need not be a Caribbean island to function as a corporate tax haven; it can be a Swiss municipality 20 kilometers from a major city. Policy responses may need to address subnational competition---for instance, through revenue-sharing formulas that account for where economic activity actually occurs, not just where it is registered.

\subsubsection*{Limitations} Two caveats are important. First, the sample is limited to cantons Zurich and Basel-Landschaft, which may not represent all of Switzerland. Zurich contains major financial and corporate centers, while Basel-Landschaft is a smaller, more suburban canton. A nationwide analysis---requiring Steuerfuss data from all 26 cantons---would strengthen external validity. Second, the 25 treated municipalities provide adequate power for detecting the main effect but limit precision for subgroup analyses. Future work with broader cantonal coverage could address both concerns.

\section{Conclusion}

When Swiss municipalities cut corporate taxes, they attract establishments but not employment. The new entities are concentrated in the tertiary sector, accumulate over time, and respond proportionally to the size of the tax cut---a pattern consistent with letterbox relocation rather than productive investment. While the evidence comes from two cantons and cannot rule out all forms of real activity associated with new registrations, the decomposition between extensive and intensive margins provides a sharper view of tax competition than aggregate outcomes alone.

The distinction between real and paper activity deserves greater attention in evaluations of tax competition. A municipality that ``wins'' the competition by attracting 100 new registrations may have gained less than headline counts suggest if those registrations bring minimal employment. The \citet{tiebout1956pure} vision of efficient sorting assumes that mobile agents bring their demand for public goods alongside their tax base. To the extent that corporate tax competition attracts registrations without workers, the welfare calculus is less favorable than simple revenue comparisons imply.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
