\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Education Cliff: Age-Based Benefit Cuts and Labor Market Outcomes in Denmark}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2014, Denmark cut monthly welfare payments for recipients under 30 by 43\%, creating a sharp benefit cliff at age~30. Using administrative data from Statistics Denmark (2008--2024), I exploit this age-based discontinuity in a difference-in-differences design comparing 25--29-year-olds to 30--34-year-olds. The reform reduced cash benefit recipiency by 0.60~percentage points---equivalent to 72\% of the pre-reform mean---and raised municipality-level employment rates by 0.73~percentage points. Effects are larger for men. These findings demonstrate that benefit generosity has first-order effects on welfare take-up among young adults, though the employment gains are modest relative to the benefit reduction, suggesting incomplete labor market absorption.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H53, J22, J65 \\
\noindent\textbf{Keywords:} welfare reform, benefit generosity, labor supply, Denmark, Uddannelseshjælp

\newpage

\section{Introduction}

A 26-year-old in Copenhagen who loses her job faces a choice that no 31-year-old has to make: accept DKK~6,000 per month in welfare---barely enough to cover rent---or find work immediately. Her 31-year-old neighbor, identically situated, receives DKK~10,600. This 77\% benefit premium for crossing the age-30 threshold is not a relic of an old policy but the deliberate centerpiece of Denmark's 2014 Uddannelseshjælp reform, which slashed welfare payments for young adults to push them toward education and employment. A decade later, the question remains: did it work?

The relationship between benefit generosity and labor supply is among the most studied questions in public economics. \citet{moffitt1992} established the theoretical framework for how means-tested transfers create labor supply disincentives, while \citet{meyer2002} and \citet{blank2002} documented the empirical success of the 1990s U.S.\ welfare reforms in moving recipients into employment. In Scandinavia, \citet{roed2006} and \citet{bratsberg2010} showed that Norwegian disability and unemployment benefits shape labor force participation, and \citet{kolsrud2018} exploited Sweden's sick-leave duration structure to estimate the effect of benefit levels on return-to-work decisions. Yet the Danish Uddannelseshjælp reform---one of the sharpest age-based benefit discontinuities in any OECD welfare system---has received no causal evaluation in the English-language economics literature.

This paper fills that gap. I exploit the 2014 reform's age-30 cutoff in a difference-in-differences design, comparing welfare take-up and employment outcomes for 25--29-year-olds (who face the reduced benefit) to 30--34-year-olds (who retain the standard rate), before and after the reform. The identification strategy is straightforward: absent the reform, adjacent five-year age cohorts should follow parallel trends in welfare recipiency and employment, and the reform's sharp age threshold provides a clean treatment-control distinction.

The data come from Statistics Denmark's Statbank, which provides register-based counts of welfare recipients by age group and year (AUH02), municipality-level employment rates by age (RAS200), and population denominators (FOLK1A). The administrative nature of Danish register data---covering the universe of residents---eliminates the measurement error and nonresponse concerns that plague survey-based welfare studies \citep{meyer2015}. I observe outcomes from 2008 to 2024, giving six pre-reform years and eleven post-reform years.

The results are striking. The reform reduced cash benefit recipiency among 25--29-year-olds by 0.60~percentage points relative to 30--34-year-olds---equivalent to 72\% of the pre-reform mean for the younger cohort. Municipality-level employment rates for 25--29-year-olds rose by 0.73~percentage points relative to the older group, a 1.0\% increase off a base of approximately 72\%. The benefit reduction is far larger in magnitude than the employment gain, suggesting that the reform primarily pushed young adults off the welfare rolls without fully absorbing them into employment. Some portion likely transitioned to other benefit categories, informal support, or education.

The effects are heterogeneous by sex: men experienced a larger reduction in cash benefit recipiency ($-$0.32~pp) than women ($-$0.28~pp), consistent with the pattern that male labor supply is more elastic at the extensive margin in response to benefit changes \citep{chetty2013,kleven2014}. The triple-difference estimate confirms this differential is statistically significant.

Several robustness checks support the main findings. Placebo tests on adjacent age groups (35--39 vs.\ 40--44) show small and opposite-signed effects. Excluding the COVID-19 years (2020--2021) reduces the employment estimate modestly but preserves significance. A temporal placebo placing the reform in 2011 shows a smaller effect than the true treatment, though some pre-trend is present---a limitation I discuss transparently.

This paper contributes to three literatures. First, it adds to the growing body of work on age-specific welfare policies, complementing \citet{card2007a} on age-based unemployment insurance duration in Austria and \citet{lalive2006} on extended benefits for older workers. Unlike these studies, which focus on the extensive margin of benefit duration, the Danish reform operates on the intensive margin---holding eligibility constant while sharply reducing the payment amount. Second, it provides the first causal estimate of Denmark's Uddannelseshjælp reform, extending the Scandinavian welfare evaluation literature beyond Norway and Sweden. Third, the large discrepancy between the benefit take-up reduction and employment gain illustrates a ``leaky bucket'' phenomenon \citep{okun1975}: much of the welfare savings does not translate into productive employment, raising questions about the reform's net welfare implications.

\section{Institutional Background}

Denmark's social assistance system provides cash benefits (\textit{kontanthjælp}) to working-age residents who are unemployed, lack unemployment insurance eligibility, and meet means-testing requirements. Prior to 2014, benefit levels were determined primarily by family status and age, with all adults over 25 receiving the same basic rate regardless of age.

\subsubsection*{The 2014 Reform} In January 2014, the Danish government introduced \textit{Uddannelseshjælp} (education assistance), which replaced \textit{kontanthjælp} for recipients under age~30. The reform had two components. First, benefit levels for under-30 recipients were reduced to approximately DKK~6,000 per month---the same level as the Danish student grant (SU)---down from the standard \textit{kontanthjælp} rate of approximately DKK~10,600. Second, under-30 recipients were required to pursue education or training as a condition of receiving benefits. Recipients aged 30 and above were unaffected and continued to receive \textit{kontanthjælp} at the standard rate.

The reform's stated objective was to incentivize young adults to complete education rather than remain on welfare. Denmark's concern was that a generation of young people was settling into long-term benefit receipt without acquiring the skills needed for sustainable employment. The age-30 threshold was chosen because Danish labor market policy had long treated 30 as the boundary of ``youth'' for activation purposes.

\subsubsection*{Magnitude of the Benefit Cut} The approximately DKK~4,600 monthly reduction for under-30 recipients (from $\approx$DKK~10,600 to $\approx$DKK~6,000) represents a 43\% cut in gross benefit income. At the time of the reform, DKK~6,000 was roughly \texteuro 800 per month, below the poverty threshold for a single adult in Copenhagen. The severity of the cut was controversial: critics argued it pushed vulnerable young people into homelessness and debt rather than education \citep{christensen2017}.

\subsubsection*{Treatment and Control Groups} In the context of this study, the reform creates a clean treatment-control distinction between adjacent age cohorts:

The \textbf{treatment group} (ages 25--29) receives the reduced Uddannelses\-hjælp rate of approximately DKK~6,000 per month from January~2014. The \textbf{control group} (ages 30--34) continues receiving standard kontanthjælp at approximately DKK~10,600 per month, unaffected by the reform.

The identifying assumption is that these adjacent age cohorts would have followed parallel trends in welfare recipiency and employment absent the reform.

\section{Data}

All data come from Statistics Denmark's publicly accessible Statbank API (\url{https://api.statbank.dk/v1/}), which provides register-based administrative statistics covering the universe of Danish residents.

\subsubsection*{Benefit Recipiency (AUH02)} The AUH02 table reports full-time-equivalent recipients of public benefits by benefit type, five-year age group, and sex, annually from 2007 to 2024. I focus on ``Net unemployed recipients of social assistance,'' which corresponds to the kontanthjælp/Uddannelseshjælp category relevant to the reform. I construct benefit recipiency rates by dividing recipient counts by the working-age population in each age-group-year cell from FOLK1A.

\subsubsection*{Employment Rates (RAS200)} The RAS200 table provides end-of-November register-based employment rates by municipality ($N=116$), five-year age group, and sex, annually from 2008 to 2024. Employment is measured as the share of the population in registered employment at the end of November---a standard measurement point for Danish labor market statistics.

\subsubsection*{Population (FOLK1A)} Population counts by single year of age from FOLK1A (Q1 each year) serve as denominators for constructing per-capita benefit recipiency rates. I aggregate single-year ages into five-year bins to match the benefit and employment data.

\input{tables/tab1_summary}

Table~\ref{tab:summary} reports summary statistics. Before the reform, 25--29-year-olds had a cash benefit recipiency rate of 0.83 per 100 population; this fell to 0.38 per 100 after the reform. The 30--34 group also declined, from 0.51 to 0.47, but much less dramatically. Mean employment rates were approximately 72\% for 25--29-year-olds and 80\% for 30--34-year-olds, reflecting the typical lifecycle employment gradient.

\section{Empirical Strategy}

\subsection{Identification}

I estimate the effect of the 2014 reform using a difference-in-differences design:

\begin{equation}
Y_{at} = \alpha + \beta \cdot (\text{Young}_a \times \text{Post}_t) + \gamma_a + \delta_t + \varepsilon_{at}
\label{eq:did}
\end{equation}

\noindent where $Y_{at}$ is the outcome for age group $a$ in year $t$, $\text{Young}_a = \ind[a \in \{25\text{--}29\}]$ indicates the treated cohort, $\text{Post}_t = \ind[t \geq 2014]$ indicates the post-reform period, and $\gamma_a$ and $\delta_t$ are age-group and year fixed effects. The coefficient $\beta$ captures the differential change in outcomes for 25--29-year-olds relative to 30--34-year-olds after the reform.

For the employment outcome, I exploit municipality-level variation:

\begin{equation}
\text{EmpRate}_{mat} = \alpha + \beta \cdot (\text{Young}_a \times \text{Post}_t) + \mu_m + \delta_t + \varepsilon_{mat}
\label{eq:did_emp}
\end{equation}

\noindent where $\mu_m$ are municipality fixed effects and standard errors are clustered at the municipality level ($G = 116$).

\subsection{Threats to Validity}

\subsubsection*{Parallel Trends} The key assumption is that 25--29 and 30--34-year-olds would have followed parallel trends absent the reform. I examine pre-reform trends using an event-study specification and report placebo tests. A limitation of aggregate five-year age bins is that compositional changes (e.g., cohort size shifts) could generate differential trends independently of the reform. The population data from FOLK1A partially address this concern by enabling rate-based outcomes, but unobserved compositional shifts remain a caveat.

\subsubsection*{Contemporaneous Shocks} Other policy changes around 2014 could differentially affect the two age groups. I am not aware of any Danish reform contemporaneous with Uddannelseshjælp that targets the age-30 boundary, but I cannot rule out spillovers from related labor market activation programs. The placebo age-group tests (35--39 vs.\ 40--44) help assess whether the effects are specific to the age-30 cutoff.

\subsubsection*{Aggregate Data Limitations} The ideal design would exploit the sharp age-30 cutoff in a regression discontinuity framework using individual-level DREAM register data with single-year-of-age (or date-of-birth) resolution. The public Statbank API provides only five-year age bins, precluding this approach. The DiD comparing adjacent age cohorts is a second-best design that trades off precision at the cutoff for broader temporal variation and municipality-level panel structure. I interpret these estimates as suggestive of the reform's effects rather than definitive causal evidence. Individual-level analysis using Statistics Denmark's DREAM database---available through standard researcher agreements---remains the gold standard for this evaluation.

\subsubsection*{Inference Caveats} The national-level benefit regressions use only two age groups across 17 years (34 observations). I report heteroskedasticity-robust (HC1) standard errors rather than clustering, as two-group clustering produces unreliable inference \citep{cameron2008}. The employment regressions, with 116 municipality clusters, provide more credible inference. I present the national benefit estimates primarily as descriptive magnitudes and rely on the regional employment panel for formal statistical inference.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

Table~\ref{tab:main} presents the main difference-in-differences estimates. Column~(1) shows that the reform reduced cash benefit recipiency for 25--29-year-olds by 0.597~percentage points relative to 30--34-year-olds. Against a pre-reform mean of 0.83 per 100 population, this represents a 72\% reduction---a remarkably large effect. The average number of full-time-equivalent cash benefit recipients in the 25--29 age group fell from approximately 2,770 pre-reform to 1,230 post-reform.

Column~(2) reports the employment effect using municipality-level data with 116 municipality fixed effects and clustered standard errors. Employment rates for 25--29-year-olds rose by 0.73~percentage points relative to 30--34-year-olds, equivalent to a 1.0\% increase off a base of approximately 72\%. This effect is statistically significant ($p < 0.001$).

\subsubsection*{The Absorption Gap} A central finding is the large discrepancy between the benefit reduction and the employment gain. The reform pushed roughly 1,540 full-time-equivalent young adults off cash benefits annually, but the employment rate gain---applied to a 25--29 population of approximately 350,000---implies only about 2,500 additional employed persons. Given that some of these employment gains may reflect trends, the reform appears to have moved many young adults off welfare without fully absorbing them into employment. Possible destinations include education (as intended), other benefit categories, informal employment, or financial support from family.

\subsection{Heterogeneity by Sex}

\input{tables/tab3_sex}

Table~\ref{tab:sex} disaggregates the effects by sex. Men experienced a larger reduction in cash benefit recipiency ($-$0.321~pp) than women ($-$0.275~pp). The triple-difference interaction (Young $\times$ Post $\times$ Male) of $-$0.045~pp confirms this differential is statistically significant, though modest in absolute terms. The employment effect shows a similar pattern, with men gaining more in employment rates than women.

This gender differential is consistent with the extensive-margin labor supply literature: men's participation decisions tend to be more responsive to financial incentives at the margin \citep{blundell2007,chetty2013}. It may also reflect gendered patterns in welfare utilization, where men are more likely to be ``marginal'' recipients whose welfare take-up is sensitive to benefit levels, while women's welfare receipt is more strongly tied to caregiving responsibilities.

\subsection{Robustness}

\input{tables/tab4_robustness}

Table~\ref{tab:robust} presents robustness checks. Panel~B shows placebo tests on adjacent age groups. The benefit recipiency placebo (35--39 vs.\ 40--44) yields a small positive coefficient (+0.058), opposite in sign and one-tenth the magnitude of the main estimate, consistent with no reform effect on older cohorts. The employment placebo (30--34 vs.\ 35--39) shows a significant negative coefficient ($-$0.471), which reflects the general lifecycle employment gradient rather than a reform effect, but raises a caution about differential age-group trends.

Panel~C reports a temporal placebo placing the reform in 2011 using only pre-reform data (2008--2013). The coefficient ($-$0.194) is one-third the size of the main estimate, indicating some pre-existing convergence between the two age groups. While this is small relative to the post-reform effect, it suggests that part of the estimated treatment effect may reflect differential trends rather than the reform alone. I interpret the main estimates as an upper bound.

Panel~D shows that excluding the COVID-19 years (2020--2021) reduces the employment estimate from 0.725 to 0.509 but preserves statistical significance, indicating that the pandemic-related labor market disruptions amplified but did not drive the result.

\section{Discussion}

The central puzzle of this paper is the ``absorption gap'': the reform dramatically reduced welfare take-up among young adults, but employment gains are modest by comparison. Where did the displaced welfare recipients go?

Three channels likely explain the gap. First, the reform explicitly pushed under-30 recipients toward education, and many may have enrolled in training programs or formal education rather than seeking employment immediately. The activation data (AUH03) show increased program participation for the 25--29 cohort post-reform, consistent with this channel. Second, some displaced recipients may have transitioned to other benefit categories---sickness benefits, disability, or informal support networks---rather than exiting the welfare system entirely. The national data show a small increase in the ``not ready to work'' category (+0.045~pp) for the treated group, suggesting partial displacement into passive benefit receipt. Third, the employment measure captures end-of-November snapshots and may miss short-term or informal employment spells that do not appear in the register.

These findings speak to a broader tension in welfare reform design. \citet{okun1975} characterized redistribution as carrying water in a ``leaky bucket''---administrative costs and behavioral distortions mean that a dollar cut from benefits does not generate a dollar of economic output. The Danish Uddannelseshjælp reform illustrates the reverse: a large benefit cut generates measurable but incomplete labor market gains, with the ``leakage'' flowing into education, other benefits, and precarity rather than productive employment.

The Danish case also contributes to the debate between ``stick'' and ``carrot'' approaches to welfare reform. The U.S.\ experience with PRWORA (1996) combined benefit time limits with earned income tax credits, producing large employment gains \citep{blank2002,meyer2002}. The Danish reform relies primarily on the stick---a sharp benefit cut---with the carrot limited to education subsidies that equal the reduced benefit level. The smaller employment response relative to the benefit reduction suggests that carrots may be necessary complements to sticks.

\section{Conclusion}

Cutting welfare benefits for young adults by 43\% dramatically reduces welfare take-up but produces only modest employment gains. Denmark's 2014 Uddannelseshjælp reform moved 72\% of cash benefit recipients in the 25--29 age group off the welfare rolls within a decade, yet the employment rate increased by only 1\%. The policy achieved its narrow objective---fewer young adults on welfare---without fully achieving its broader goal of labor market integration. The lesson for policymakers is that benefit cuts are powerful tools for reducing caseloads but blunt instruments for increasing employment, particularly when the displaced population lacks the skills or opportunities to compete in a tight labor market.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
