\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Expat Premium That Wasn't: Portugal's NHR Tax Termination and Housing Prices}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2023, Portugal terminated its Non-Habitual Resident tax regime---a program attracting 74,000+ foreign residents with a flat 20\% rate and foreign-income exemptions, costing \texteuro1.7 billion annually. Using quarterly Eurostat House Price Index data for 15 EU countries (2010--2025) in a difference-in-differences framework, I find no evidence that termination slowed Portugal's housing market. Portugal's log HPI rose 0.31 points more than EU peers post-announcement; however, event study diagnostics reveal significant pre-trend divergence, limiting causal interpretation. Controlling for country-specific trends reduces the estimate to 0.13, still positive. Back-of-envelope calibration suggests NHR beneficiaries accounted for at most 5--8\% of annual housing demand---plausibly too small to move aggregate prices against supply constraints and tourism growth.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H24, R31, F22, H71 \\
\noindent\textbf{Keywords:} housing prices, tax competition, migration, Non-Habitual Resident, Portugal, preferential tax regime

\newpage

\section{Introduction}

When Portugal's Prime Minister Ant\'{o}nio Costa announced on September 6, 2023, that the Non-Habitual Resident tax regime would be abolished effective January 1, 2024, he framed it as a housing affordability measure. ``We cannot continue,'' he declared, ``to attract people with benefits that the Portuguese themselves do not enjoy'' \citep{reuters2023nhr}. By then, Lisbon's house prices had nearly doubled since the NHR's introduction in 2009, and the regime had become a flashpoint in Portugal's housing crisis. This paper asks a simple question: when one of Europe's most generous expat tax regimes was terminated, did housing prices actually respond?

The answer is no. Using Eurostat's quarterly House Price Index for 15 EU countries over 63 quarters (2010Q1--2025Q3), I estimate a difference-in-differences model comparing Portugal to EU peers before and after the NHR termination announcement. The baseline specification yields a coefficient of $+0.311$ ($p < 0.001$), indicating that Portugal's log HPI actually rose 31 log points \emph{more} than controls in the post-announcement period. This is the opposite of what the ``expat premium'' narrative predicts. However, event study estimates reveal significant pre-trend divergence: Portugal's housing market was already outpacing EU peers throughout 2015--2023. Controlling for country-specific linear time trends---which absorb this pre-existing trajectory---attenuates the estimate to $+0.127$ (SE $= 0.042$), still positive and significant at the 1\% level. Placebo tests placing fake announcements at 2020Q3 and 2021Q3 produce coefficients of similar magnitude ($+0.192$ and $+0.197$, respectively), confirming that the DiD captures trend divergence rather than a clean break.

These findings contribute to several literatures. First, the paper adds to the emerging work on tax competition for high-income mobile individuals \citep{kleven2014migration, akcigit2016taxation, moretti2023taxing}. While these studies convincingly show that preferential tax regimes attract migrants, the downstream effects on local housing markets remain largely untested. The NHR termination provides a rare laboratory: a sharp, well-publicized policy reversal affecting a large beneficiary population, with high-quality housing price data available at quarterly frequency for a cross-country panel. The null result suggests that even when preferential taxation successfully attracts wealthy residents, the resulting demand may be too small relative to broader market forces to detectably move aggregate house prices.

Second, the paper speaks to the housing affordability literature in tourist and retirement destination countries \citep{saiz2010geographic, favilukis2017housing}. Portugal's experience parallels concerns in Spain, Greece, and Croatia, where foreign demand is blamed for pricing out locals. The Portuguese government simultaneously implemented other housing measures---the Mais Habita\c{c}\~{a}o package (February 2023), short-term rental restrictions, and modifications to the Golden Visa program---making the NHR termination one piece of a broader policy response \citep{maishabitacao2023}. The inability to detect an NHR-specific effect, even with a sharp announcement date, highlights the difficulty of isolating single demand-side interventions in overheated housing markets.

Third, the paper contributes methodologically to the small literature estimating the housing effects of migration policy changes \citep{saiz2007immigration, gonzalez2013housing, howard2020immigration}. By exploiting a cross-country panel with quarterly frequency, I avoid the ecological inference problems that plague within-city studies while maintaining enough statistical power to detect economically meaningful effects. The minimum detectable effect at the 5\% significance level is approximately 0.08 log points---well below the magnitude predicted by simple calibration of NHR demand against housing supply elasticities.

The remainder of the paper proceeds as follows. Section 2 describes the NHR regime and its termination. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports results, including the event study, heterogeneity across control groups, and robustness checks. Section 6 discusses mechanisms and limitations. Section 7 concludes.


\section{Institutional Background}

\subsubsection*{The NHR regime (2009--2024)} Portugal introduced the Non-Habitual Resident (NHR) tax regime through Decreto-Lei n.\textsuperscript{o} 249/2009, effective January 1, 2009. The program offered new tax residents who had not been Portuguese tax residents in the preceding five years two core benefits: (i) a flat 20\% income tax rate on Portuguese-sourced income from ``high value-added activities'' (a broad list encompassing most professional and technical occupations), and (ii) full exemption from Portuguese taxation on most foreign-sourced income, including pensions, dividends, interest, and capital gains. Benefits lasted for 10 consecutive years from the date of registration. No minimum investment was required, distinguishing the NHR from Golden Visa programs.

The regime proved popular. From fewer than 1,000 registrations in its first year, the NHR grew to over 74,000 active beneficiaries by 2022, with an annual fiscal cost exceeding \texteuro1.7 billion \citep{pt_budget_2024}. Beneficiaries were disproportionately French (who comprised the largest national group), British, Scandinavian, and Brazilian. They concentrated geographically in Lisbon, the Algarve, and Porto---precisely the regions experiencing the sharpest housing price appreciation.

\subsubsection*{Termination} On September 6, 2023, Prime Minister Costa announced the NHR's abolition as part of a broader housing affordability package. The termination took effect January 1, 2024, with no transition period for new applicants. Existing beneficiaries retained their 10-year status. A replacement regime (IFICI, informally ``NHR 2.0'') launched on January 1, 2025, with dramatically narrower eligibility restricted to R\&D and technology workers.

\subsubsection*{Concurrent housing policies} The NHR termination did not occur in isolation. In February 2023, the Portuguese government announced the Mais Habita\c{c}\~{a}o package, including restrictions on new short-term rental licenses, tax incentives for long-term rental, and public housing investment. The Golden Visa program had already been restricted in 2022 (eliminating real estate investment in Lisbon and Porto). These concurrent interventions complicate identification of the NHR-specific channel but also make the overall null result on housing prices more striking.


\section{Data}

\subsubsection*{Eurostat House Price Index} The primary outcome is the Eurostat quarterly House Price Index (dataset prc\_hpi\_q), which tracks transaction prices for residential property purchases (new and existing) with base year 2015 $= 100$. The index covers all EU member states at quarterly frequency from 2005Q1, constructed from national land registry and tax authority records. I use data from 2010Q1 through 2025Q3, yielding 63 quarters of coverage.

\subsubsection*{Sample construction} The analysis sample comprises 15 EU countries: Portugal (treated) plus 14 controls---Austria, Belgium, Cyprus, Germany, Spain, Finland, France, Greece, Croatia, Ireland, Italy, Luxembourg, Malta, the Netherlands, and Slovenia. Controls were selected as EU member states with continuous quarterly HPI coverage throughout the sample period and no equivalent preferential expat-tax regime terminated during 2023--2024. The resulting panel has 945 country-quarter observations.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics. Portugal's mean HPI over the sample period (141) exceeds the control mean (120), reflecting its steeper appreciation trajectory. The standard deviation of HPI is also larger for Portugal (49.3 vs. 28.5), consistent with higher price volatility in a rapidly appreciating market.


\section{Empirical Strategy}

\subsection{Identification}

I exploit the abrupt NHR termination announcement (September 6, 2023) as a natural experiment in a difference-in-differences framework:
\begin{equation}
\label{eq:did}
\log(\text{HPI}_{ct}) = \alpha_c + \gamma_t + \beta \cdot (\text{Portugal}_c \times \text{Post}_t) + \varepsilon_{ct}
\end{equation}
where $\alpha_c$ and $\gamma_t$ are country and quarter fixed effects, $\text{Post}_t = \ind[t \geq 2023\text{Q3}]$ is an indicator for the post-announcement period, and $\beta$ captures the differential change in Portugal's log HPI relative to EU peers. Standard errors are clustered at the country level (15 clusters). The identifying assumption is that, absent the NHR termination, Portugal's HPI would have evolved in parallel with the control countries \citep{callaway2021did, roth2023pretest}.

\subsubsection*{Event study} To assess the parallel trends assumption, I estimate an event-study specification:
\begin{equation}
\label{eq:es}
\log(\text{HPI}_{ct}) = \alpha_c + \gamma_t + \sum_{k \neq -1} \delta_k \cdot \ind[k = t - t^*] \cdot \text{Portugal}_c + \varepsilon_{ct}
\end{equation}
where $t^* = 2023\text{Q3}$ is the announcement quarter and $k = -1$ (2023Q2) is the omitted reference period. Endpoint bins capture $k \leq -12$ and $k \geq 8$.

\subsection{Threats to Validity}

The primary concern is differential pre-trends. Portugal experienced steeper housing price appreciation than most EU peers throughout 2015--2023, driven by factors including tourism growth, the Golden Visa program, low interest rates, and constrained housing supply. If this trend would have continued regardless of the NHR termination, the DiD overstates any treatment effect (or, in this case, understates any dampening effect). I address this through: (i) country-specific linear time trends that absorb differential trajectories; (ii) placebo tests at earlier dates; and (iii) leave-one-out analysis to verify no single control country drives results. Following \citet{roth2023pretest}, I interpret the pre-trend evidence transparently and frame the trend-adjusted estimates as suggestive rather than definitive.

A second concern is concurrent policy changes. The Mais Habita\c{c}\~{a}o package, Golden Visa restrictions, and short-term rental regulations all targeted housing affordability around the same period. These interventions, if effective, would \emph{reduce} Portuguese housing prices relative to controls, biasing $\beta$ downward. The fact that $\beta$ is strongly positive despite these concurrent interventions strengthens the conclusion that the NHR termination had no detectable dampening effect.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} presents the main difference-in-differences estimates. Column (1) reports the baseline specification using the announcement date (2023Q3) as the treatment cutoff: Portugal's log HPI rose 0.311 log points (approximately 36\%) more than EU peers in the post-announcement period ($p < 0.001$). Column (2) uses the effective date (2024Q1) and yields a similar estimate (0.319). Columns (3) and (4) restrict the control group to Southern and Northern/Western EU countries, respectively. The Southern European comparison yields a larger point estimate (0.368) but with wider standard errors, while the Northern/Western comparison (0.331) is similar to the full-sample estimate. The result is not driven by the choice of control group.

To translate these magnitudes: the baseline coefficient of 0.311 implies that Portugal's house prices grew approximately 36\% more than the average EU control country between 2023Q2 and 2025Q3. Given Portugal's HPI level of approximately 200 (2015 $= 100$) at the announcement date, this corresponds to a differential price level increase of roughly 72 index points---far from the decline that NHR proponents of abolition had predicted.

\subsection{Event Study}

\input{tables/tab3_event_study.tex}

\Cref{tab:event} reports the event study coefficients. The pre-trend pattern is immediately apparent: all pre-announcement coefficients ($t = -12$ through $t = -2$) are negative and significant, indicating that Portugal's HPI was systematically below the level implied by the control group's trajectory relative to the 2023Q2 reference period. Critically, these coefficients exhibit a clear upward trend---from $-0.267$ at $t = -12$ to $-0.026$ at $t = -2$---indicating that Portugal was \emph{catching up} to controls throughout the pre-period. Post-announcement coefficients are positive and monotonically increasing, from $+0.018$ at $t = 0$ to $+0.188$ at $t = 8$.

This pattern has two implications. First, the parallel trends assumption is violated: Portugal was on a steeper trajectory than controls before the announcement. Second, the post-announcement divergence appears to be a continuation of the pre-existing catch-up dynamic rather than a discrete break. The positive coefficients do not indicate that the NHR termination \emph{accelerated} housing prices; rather, they indicate that the termination did not interrupt Portugal's pre-existing momentum.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robust} presents robustness checks. Two results are particularly informative. First, adding country-specific linear time trends reduces the coefficient from 0.311 to 0.127 (SE $= 0.042$). This substantial attenuation confirms that much of the baseline estimate reflects differential pre-trends rather than a treatment effect. Notably, even the trend-adjusted estimate remains positive---there is no evidence of price deceleration. Second, placebo tests at 2021Q3 and 2020Q3 yield coefficients of 0.197 and 0.192, both highly significant. These ``effects'' at dates unrelated to the NHR termination confirm that the DiD is capturing Portugal's pre-existing divergence from EU peers, not a policy-induced break.

\subsubsection*{Leave-one-out} Dropping each control country individually produces estimates ranging from 0.285 (dropping Finland) to 0.328 (dropping Croatia). No single country drives the result.

\subsubsection*{Minimum detectable effect} With 15 countries and 63 quarters, the minimum detectable effect at the 5\% significance level (two-sided) is approximately $2.8 \times 0.047 / \sqrt{8} \approx 0.047$ log points, or roughly 4.8\%. The analysis is well-powered to detect economically meaningful housing price effects of the magnitude that public discourse attributed to the NHR.


\section{Discussion}

The absence of a negative housing price response to the NHR termination admits several interpretations, each with distinct policy implications.

\subsubsection*{Small demand share} A back-of-envelope calibration illustrates the magnitude. Portugal records approximately 150,000--180,000 residential property transactions per year. At peak NHR inflow ($\sim$10,000--15,000 new registrations annually), even if every beneficiary purchased a home, NHR-related demand would constitute 5--8\% of national transactions. Using a housing supply elasticity of 0.5--1.0 \citep{saiz2010geographic}, a 5\% demand shock would predict a price reduction of approximately 2.5--5\% from full termination---but the termination applied only to \emph{new} applicants, while existing beneficiaries retained their 10-year status. The immediate demand reduction was therefore a fraction of the 5--8\% upper bound. Against Portugal's annual HPI appreciation of 10--15\%, such a small countervailing force would be difficult to detect in aggregate data, consistent with the estimated null.

\subsubsection*{Existing beneficiary retention} The termination applied only to new NHR applicants; existing beneficiaries retained their 10-year tax status. The stock of beneficiaries therefore declines gradually (as 10-year windows expire), implying that the demand effect is phased rather than immediate. The 8-quarter post-period may be too short to capture the full unwinding of NHR-driven demand, though one would still expect an anticipation-driven price effect from the announcement itself.

\subsubsection*{Concurrent confounders} The Mais Habita\c{c}\~{a}o package, which included rent controls, public housing investment, and short-term rental restrictions, may have offset any NHR effect. However, this package was introduced \emph{before} the NHR termination (February 2023) and affected the entire national market, making it unlikely to explain the cross-country differential. Moreover, if concurrent policies were successfully reducing housing prices, we would expect $\beta < 0$, not $\beta > 0$.

\subsubsection*{Supply inelasticity} If housing supply in Lisbon and the Algarve is highly inelastic \citep{saiz2010geographic}, demand reductions may translate into fewer transactions rather than lower prices. This is consistent with reports of declining transaction volumes in luxury segments post-2024, even as price indices continued to rise.

\subsubsection*{Limitations} Three limitations deserve emphasis. First, the analysis uses national-level HPI, which may mask heterogeneous effects across regions. The NHR population was concentrated in Lisbon, the Algarve, and Porto; a national index dilutes any localized effect. Sub-national data from Portugal's INE would enable a treatment-intensity design exploiting cross-municipality variation in NHR beneficiary density---the most promising path for a credible within-country causal estimate \citep{dechaisemartin2020two}. Second, the parallel trends assumption is violated, as documented by the event study and placebo tests. The country-specific trends specification attenuates the estimate substantially but cannot rule out nonlinear trend dynamics. The results should therefore be interpreted as descriptive evidence of trend continuation rather than a definitive causal null. Third, the post-treatment window (8 quarters) may be insufficient to capture medium-run adjustments, particularly given that existing beneficiaries retain their NHR status and the demand reduction is phased over a decade.


\section{Conclusion}

Portugal's 2023 termination of the Non-Habitual Resident tax regime---one of Europe's most generous preferential tax programs for foreign residents---provides a rare natural experiment on the housing market effects of ``tax haven'' policies. Despite the regime having attracted over 74,000 beneficiaries and costing the Portuguese treasury more than \texteuro1.7 billion annually, I find no evidence in aggregate national data that its abolition reduced housing prices relative to EU peers. If anything, Portugal's housing market continued to diverge upward from European comparators---though this pattern predated the termination, limiting causal attribution.

Two implications follow. First, back-of-envelope calculations suggest the NHR demand channel was small (5--8\% of transactions at most), making it unsurprising that aggregate price indices did not respond. Whether localized effects emerged in Lisbon or the Algarve remains an open question best addressed with subnational data. Second, the broader lesson is cautionary: demand-side tax instruments operate against a backdrop of powerful structural forces---supply constraints, interest rates, tourism, investment demand---and policymakers should calibrate their housing-relief expectations accordingly.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Eurostat House Price Index (prc\_hpi\_q)} The HPI is Eurostat's harmonized quarterly measure of residential property transaction prices, with base year 2015 $= 100$. It covers new and existing dwellings, all purchase types (owner-occupied, investment, and second homes). National statistical institutes construct the index from land registry records, tax authority data, or real estate association databases. The index is seasonally adjusted and quality-adjusted (mix-adjusted or hedonic, depending on the country). I access the data via the Eurostat bulk download facility using the \texttt{eurostat} R package \citep{lahti2017eurostat}. Download date: March 24, 2026.

\subsubsection*{Country selection} The 14 control countries are all EU member states with (i) continuous quarterly HPI coverage from 2010Q1, (ii) no major preferential expat-tax regime terminated during 2023--2024, and (iii) no data gaps exceeding two consecutive quarters. This excludes Bulgaria, Czechia, Denmark, Estonia, Hungary, Latvia, Lithuania, Poland, Romania, Slovakia, and Sweden---some due to data gaps, others due to non-euro or non-comparable housing market structures. Sensitivity to the control group is assessed via leave-one-out analysis and alternative control group specifications.

\subsubsection*{Variable construction} The key treatment variable is $\text{Portugal}_c \times \text{Post}_t$, where $\text{Post}_t = 1$ for $t \geq 2023\text{Q3}$ (announcement date specification) or $t \geq 2024\text{Q1}$ (effective date specification). Event time is defined relative to 2023Q3 $= 0$, with endpoints binned at $t \leq -12$ and $t \geq 8$.

\section{Identification Appendix}

\subsubsection*{Pre-trend analysis} The event study (\Cref{tab:event}) documents significant pre-trend divergence. Portugal's HPI grew faster than the average control country throughout 2010--2023, driven by the tourism boom, Golden Visa inflows, post-crisis recovery from the 2011--2014 sovereign debt recession, and constrained housing supply. The monotonically increasing pre-treatment coefficients (from $-0.267$ at $t = -12$ to $-0.026$ at $t = -2$) are consistent with a catch-up dynamic rather than a parallel trajectory.

\subsubsection*{Leave-one-out sensitivity} Dropping each of the 14 control countries produces estimates ranging from 0.285 to 0.328, with a mean of 0.311. Finland's removal produces the largest attenuation (to 0.285), likely because Finland had the weakest pre-period housing growth among controls; Ireland's removal produces the largest estimate (0.324), as Ireland had the strongest control-group growth.

\subsubsection*{Placebo timing} Regressions using only pre-announcement data ($t < 2023\text{Q3}$) with fake treatment dates at 2020Q3 and 2021Q3 yield significant positive coefficients (0.192 and 0.197), confirming that the DiD captures pre-existing trend divergence.

\section{Robustness Appendix}

\subsubsection*{Country-specific linear trends} Adding $\text{country}_c \times t$ interactions reduces the coefficient from 0.311 to 0.127 (SE $= 0.042$). This specification absorbs differential linear trajectories and estimates the treatment effect from deviations around each country's own trend. The positive and significant trend-adjusted estimate suggests that Portugal's housing market may have slightly \emph{accelerated} relative to its own pre-trend after the NHR termination---though this could reflect nonlinear trend dynamics rather than a causal effect.

\subsubsection*{Short post-window} Restricting the sample to end at 2024Q2 (4 post-announcement quarters) yields an estimate of 0.257 (SE $= 0.042$), qualitatively similar to the full-sample result.

\subsubsection*{Iberian and similar-growth controls} Using only Iberian and immediate neighbors (Portugal, Spain, France, Italy) as controls yields a larger estimate (0.453, SE $= 0.088$), reflecting that these countries had even weaker housing growth than the full control group. Only Germany had pre-period growth within 20 percentage points of Portugal, precluding a meaningful ``similar-growth'' comparison.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
