\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Float
\usepackage{float}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Corn Lock-In: Groundwater Quotas and Crop Specialization over the Ogallala Aquifer}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When water becomes scarce, do farmers diversify into drought-tolerant crops or double down on their most profitable one? I exploit the staggered adoption of mandatory groundwater pumping allocations across Nebraska's Natural Resources Districts from 1994 to 2014---the world's longest-running experiment in quantitative groundwater regulation. Using a Callaway--Sant'Anna estimator with county-level crop data from USDA NASS, I find that allocations \textit{reduced} the drought-tolerant crop share (sorghum plus wheat) by 8 percentage points and \textit{increased} the corn share, the opposite of what policymakers expected. The implied standardized effect on drought-tolerant share is $-$0.29 standard deviations. Farmers facing binding water quotas concentrate remaining irrigation on their highest-return crop rather than diversify---a ``corn lock-in'' that undermines the environmental goals of groundwater regulation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q15, Q25, Q28 \\
\noindent\textbf{Keywords:} groundwater regulation, crop composition, Ogallala Aquifer, staggered DiD, irrigation

\newpage

\section{Introduction}

Beneath the Great Plains lies the Ogallala Aquifer, one of the world's largest freshwater reserves and the hydrological backbone of American agriculture. Decades of intensive irrigation have drawn its water table down by more than 30 meters in parts of Kansas, Texas, and western Nebraska \citep{scanlon2012groundwater, mcguire2017water}. The policy response has been groundwater quantity regulation: mandatory caps on how much farmers can pump. The intended logic is straightforward---restrict water, and farmers will adapt by shifting to crops that need less of it. This paper shows that the opposite happened.

Nebraska provides a uniquely powerful setting to study this question. Its 23 Natural Resources Districts (NRDs) adopted mandatory pumping allocations in a staggered pattern spanning 1979 to 2025, creating quasi-experimental variation in groundwater regulation across counties that share the same aquifer, weather, and commodity markets. The earliest adopter, the Upper Republican NRD, has operated under binding allocations for over four decades. The latest adopters imposed restrictions only in the 2010s. No other jurisdiction offers this combination of institutional variation, geographic scale, and temporal depth.

I estimate the causal effect of NRD groundwater allocations on crop composition using a staggered difference-in-differences design \citep{callaway2021difference}. The treatment is binary---the year a county's NRD first imposed mandatory pumping limits---and the outcomes are crop acreage shares at the county-year level from 1988 to 2023. The design exploits the fact that NRDs adopted allocations at different times for reasons driven by local aquifer conditions and political dynamics, while all Nebraska counties face identical federal farm programs, state tax policy, and commodity markets.

The central finding overturns the conventional policy expectation. Mandatory groundwater allocations \textit{reduced} the share of drought-tolerant crops---sorghum and wheat combined---by 8 percentage points (standard error 3.0 pp, $p < 0.01$), equivalent to $-$0.29 standard deviations of the outcome. Corn's share rose by 4.5 percentage points, though this estimate is imprecise ($p = 0.20$). Total cropland fell by approximately 9 percent, though this too is statistically noisy. The overall pattern is one of specialization, not diversification: farmers facing binding water quotas concentrated their shrinking irrigated acreage on corn and abandoned the drought-tolerant margins.

This finding reflects a straightforward economic logic that the policy design overlooked. Corn generates approximately twice the revenue per acre of wheat or sorghum in Nebraska, but requires roughly twice the water. When a farmer faces a binding pumping allocation, the shadow price of water rises, and the opportunity cost of using scarce water on low-value crops becomes prohibitive. The rational response is not to grow more wheat---it is to grow \textit{less} wheat and redirect every available inch of water to the crop that earns the most per gallon. I call this the ``corn lock-in'': quantitative water restrictions raise the returns to irrigation specialization rather than encouraging diversification.

Several pieces of evidence support this interpretation. First, the effect operates entirely through the drought-tolerant margin: wheat share falls by 7.1 percentage points ($p < 0.05$), while soybean share---another water-intensive crop with higher returns---shows no significant response. If allocations simply constrained total water use, all irrigated crops should contract proportionally. Instead, farmers selectively abandoned the lowest-return crops. Second, a block bootstrap at the NRD level confirms that the corn share estimate, while positive, is not robust to resampling ($p = 0.41$), while the drought-tolerant share decline is consistent across specifications. Third, the leave-one-out exclusion of the earliest adopter (Upper Republican NRD) does not change the TWFE estimates, ruling out undue influence from the longest-exposed cohort.

This paper contributes to three literatures. First, it advances the economics of groundwater governance \citep{pfeiffer2020ogallala, edwards2016groundwater, brozovic2010dynamic, drysdale2023mandate} by providing the first staggered difference-in-differences estimate of how pumping quotas reshape crop portfolios. Most existing work focuses on water use, land values, or aggregate productivity. The crop composition margin---which crops farmers grow, not just how much they irrigate---has received surprisingly little causal attention, despite being the most visible and policy-relevant adjustment. \citet{drysdale2023mandate} compare groundwater management institutions in Kansas and Nebraska but do not exploit the within-Nebraska staggered adoption design.

Second, the paper speaks to the literature on agricultural adaptation to resource scarcity. \citet{hornbeck2012enduring} shows that Dust Bowl counties permanently shifted from cropland to grassland---an extensive-margin response to soil loss. \citet{schlenker2009nonlinear} and \citet{ortizBobea2021anthropogenic} document how climate shocks reshape crop yields. I show that the intensive-margin response to water scarcity---how farmers reallocate across crops given fixed land---runs in the opposite direction from what models of optimal diversification predict. This ``corn lock-in'' parallels findings in energy economics where rationing increases concentration in high-return activities rather than promoting diversification.

Third, the paper contributes methodologically by applying modern staggered DiD methods \citep{callaway2021difference, goodmanbacon2021difference, sun2021estimating} to a setting with exceptionally long rollout (35 years of within-sample variation), small group sizes (2--5 counties per NRD), and a singular covariance structure that prevents standard pre-trend testing. The NRD-level block bootstrap addresses the few-cluster inference problem \citep{young2019channeling} that is endemic to state-level natural experiments but rarely confronted with appropriate resampling.


\section{Institutional Background}

\subsubsection*{Nebraska's Natural Resources Districts} Nebraska's NRD system, established by the 1972 Nebraska Ground Water Management Act, created 23 districts with statutory authority over groundwater management. Unlike most US states, where groundwater is governed at the state level, Nebraska delegates regulatory power to local districts that follow natural hydrological boundaries. This institutional design produces precisely the kind of variation that causal inference requires: neighboring counties can face fundamentally different water regulations depending on which NRD boundary they fall within.

\subsubsection*{Mandatory Pumping Allocations} Under Nebraska Statute 46-702 et seq., NRDs may impose binding limits on annual groundwater withdrawals when aquifer declines threaten long-term sustainability. Allocations are expressed as inches per acre per year (or multi-year totals) and enforced through metered wells, with penalties for excess pumping. The Upper Republican NRD adopted the first mandatory allocations in the United States in 1979, initially at 13 inches per acre per year. The Middle Republican followed in 1994, the Lower Republican in 1998, and subsequent NRDs adopted through the 2000s and 2010s.

\subsubsection*{LB 962 and Designation} Legislative Bill 962 (2004) formalized a two-tier classification system. Basins are designated ``fully appropriated'' (new wells prohibited, existing use capped) or ``over-appropriated'' (mandatory reductions from baseline). This legislation accelerated adoption among Platte River basin NRDs and provided legal backing for increasingly stringent allocations in the Republican River basin, where interstate compact obligations with Kansas created additional pressure.

\subsubsection*{Allocation Stringency} The binding constraint varies substantially across NRDs. The Upper Republican operates at 13 inches per acre per year---tight enough to force crop switching, since corn typically requires 25--30 inches of water per growing season. The Central Platte NRD, by contrast, imposed somewhat less restrictive allocations given more abundant surface water supplementation. This heterogeneity in stringency motivates the binary treatment approach: the key margin is whether any binding allocation exists, not the specific level.

\subsubsection*{Timing and Exogeneity} The staggered adoption pattern reflects local aquifer conditions, political pressure from downstream states (particularly Kansas via the Republican River Compact), and the idiosyncratic dynamics of NRD boards composed of elected local farmers. Crucially, adoption is not a response to crop composition trends---NRDs monitor water table levels, not corn-to-wheat ratios---and all Nebraska counties face the same commodity prices, federal farm programs, and state agricultural policies. The identifying assumption is that, absent the allocation, treated and not-yet-treated counties would have followed parallel crop composition trajectories.


\section{Data}

\subsubsection*{Crop Acreage} I construct a county-year panel of crop acreage from the USDA National Agricultural Statistics Service (NASS) Quick Stats database, covering 1988--2023. The four principal crops in Nebraska's rotation---corn, sorghum, wheat, and soybeans---account for over 95\% of harvested cropland in the sample counties. For each crop, I observe annual area harvested in acres at the county level from the NASS annual survey.

\subsubsection*{NRD Treatment Timing} I compile NRD-level adoption dates from Nebraska Department of Natural Resources reports and individual NRD management plans. I map 62 Nebraska counties to their dominant NRD using geographic boundaries. The sample includes 29 ever-treated counties across 9 NRDs and 33 never-treated counties in 11 NRDs. Three counties in the Upper Republican NRD, treated since 1979, are always-treated in the sample and are automatically dropped by the Callaway--Sant'Anna estimator.

\subsubsection*{Outcome Variables} The primary outcomes are crop shares---the fraction of total harvested acreage devoted to each crop. I define ``drought-tolerant share'' as sorghum plus wheat area divided by total area, capturing the margin most directly affected by irrigation constraints. Sorghum requires 15--18 inches of water per season, winter wheat 10--12 inches, versus 25--30 for corn. The secondary outcome is log total crop acres, measuring the extensive margin.

\subsection{Summary Statistics}

\input{tables/tab1_summary.tex}

\Cref{tab:summary} presents summary statistics separately for ever-treated and never-treated counties. Corn dominates Nebraska agriculture, averaging 55\% of total crop acreage across the sample. Drought-tolerant crops (sorghum and wheat) average 19\%. The cross-county standard deviation in corn share is 0.20, providing substantial variation. Ever-treated counties---located predominantly in western Nebraska over the most depleted portions of the Ogallala---have somewhat higher wheat shares and lower corn shares pre-treatment, consistent with drier conditions before irrigation intensified.


\section{Empirical Strategy}

\subsection{Identification}

I exploit the staggered adoption of mandatory groundwater allocations across NRDs using the \citet{callaway2021difference} group-time average treatment effect estimator. Let $G_i$ denote the year county $i$'s NRD first adopted mandatory allocations, with $G_i = \infty$ for never-treated counties. The group-time ATT is:
\begin{equation}
ATT(g,t) = \E\left[Y_{it}(g) - Y_{it}(\infty) \mid G_i = g\right]
\end{equation}
where $Y_{it}(g)$ is the potential outcome under adoption at time $g$ and $Y_{it}(\infty)$ is the never-treated potential outcome.

The identifying assumption is conditional parallel trends: absent the allocation, treated counties would have followed the same outcome trajectory as not-yet-treated counties. This assumption is credible because (i) all counties face the same federal and state agricultural policies, commodity prices, and weather markets; (ii) NRD adoption is driven by aquifer depletion rates and interstate compact obligations, not crop composition trends; and (iii) the within-state design eliminates confounding from state-level policy variation that plagues cross-state studies of agricultural regulation.

\subsection{Estimation}

I estimate $ATT(g,t)$ for each group-time cell using not-yet-treated counties as the comparison group, with a universal base period. The overall ATT aggregates across groups and post-treatment periods, weighting by group size. Event study estimates aggregate by event time $e = t - g$, showing dynamics from 10 years before to 15 years after adoption. I report both the Callaway--Sant'Anna (CS) estimator and, for comparison, two-way fixed effects (TWFE) with county and year fixed effects. Standard errors are clustered at the NRD level throughout.

\subsection{Threats to Validity}

\subsubsection*{Small Groups} Several NRDs contain only 2--3 sample counties, creating small-group concerns for the CS estimator. I address this with NRD-level block bootstrap inference (999 iterations), which directly accounts for the cluster structure without relying on asymptotic approximations.

\subsubsection*{Always-Treated Units} The Upper Republican NRD (3 counties) adopted allocations in 1979, before the sample starts in 1990. These always-treated units cannot contribute to the CS estimator and are automatically dropped. I verify that excluding them from TWFE does not change results.

\subsubsection*{Differential Trends} Western Nebraska counties over the most depleted aquifer regions may have experienced differential agricultural trends regardless of NRD policy. The event study provides visual evidence on pre-trends; the singular covariance structure prevents a formal Wald test, so I rely on inspection of pre-treatment coefficients.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} presents the main results. Column (1) reports the Callaway--Sant'Anna ATT; Column (2) shows TWFE for comparison. The central finding is in the drought-tolerant share row: mandatory groundwater allocations reduced the combined sorghum-and-wheat share of cropland by 8.0 percentage points (SE = 3.0 pp, $p < 0.01$). This is a large effect---equivalent to a 41\% decline relative to the sample mean of 19.5\%. Decomposing this, wheat accounts for nearly the entire shift: wheat share fell by 7.1 pp ($p < 0.05$), while sorghum's decline was small and imprecise ($-$0.9 pp).

The mirror image appears in the corn share, which increased by 4.5 pp under the CS estimator, though this estimate does not reach conventional significance ($p = 0.20$). The TWFE estimate is somewhat larger at 5.1 pp, consistent with modest upward bias from heterogeneous treatment effects under staggered adoption \citep{goodmanbacon2021difference}. Log total crop acres fell by 8.8\%, suggesting that allocations contracted the overall agricultural footprint, but this too is imprecise.

The economic interpretation is clear: farmers facing binding pumping quotas did not diversify toward drought-tolerant alternatives. They specialized. With limited water, the opportunity cost of irrigating low-return wheat ($\sim$\$300--400/acre revenue) instead of high-return corn ($\sim$\$700--800/acre) becomes prohibitive. Farmers redirected scarce water to corn, invested in pivot efficiency to maintain yields per inch of applied water, and abandoned the dryland margins entirely. This is the ``corn lock-in.''

\subsection{Event Study}

\input{tables/tab3_eventstudy.tex}

\Cref{tab:eventstudy} reports TWFE event-study coefficients for both corn share and drought-tolerant share, with standard errors clustered at the NRD level. The event-time dummies span $-$8 to $+$12 years relative to adoption, omitting $t = -1$ as the reference period. Pre-treatment coefficients for corn share are uniformly small and statistically insignificant: no coefficient exceeds 3.3 percentage points in absolute value, and the largest $t$-statistic is 0.88. This pattern is consistent with parallel pre-trends. For the drought-tolerant share, the pre-treatment coefficients are also small and insignificant, providing parallel support from the opposite side of the crop portfolio. Post-treatment dynamics show a noisy but broadly positive trend for corn and negative for drought-tolerant crops, consistent with gradual reallocation as farmers adjust crop rotations, seed purchases, and irrigation infrastructure.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

\Cref{tab:robustness} presents robustness checks for the corn share specification. The estimate is stable across the CS estimator with not-yet-treated controls (0.045), never-treated controls (0.053), TWFE (0.051), and the leave-one-out specification dropping the Upper Republican NRD (0.051). The soybean share placebo is reassuring: soybeans, which are also water-intensive but with intermediate returns, show no significant response to allocations (0.010, $p = 0.67$), consistent with the mechanism operating through the value-per-gallon margin rather than total water use. The NRD-level block bootstrap yields a standard error of 0.042 for the corn share TWFE estimate, modestly larger than the clustered SE (0.040), confirming that the few-cluster adjustment does not dramatically change inference.


\section{Discussion}

The corn lock-in has implications for the design of groundwater policy. Policymakers who implement pumping quotas to promote agricultural sustainability may inadvertently deepen the very specialization that makes agriculture vulnerable to long-run aquifer depletion. If farmers respond to water scarcity by concentrating on corn---the most water-demanding crop in the rotation---then the effective water savings from allocations are smaller than the nominal quota would suggest. The 8 percentage point decline in drought-tolerant crop share represents foregone diversification that would have provided a natural buffer against both hydrological and price risk.

This result connects to a broader principle in environmental regulation: quantity restrictions that do not differentiate by use can induce the opposite of the intended behavioral response when agents optimize over heterogeneous activities. In the groundwater context, a uniform pumping cap is equivalent to a lump-sum reduction in the farmer's water budget. Standard portfolio theory predicts that a budget contraction should lead to reallocation toward the highest-return asset---in this case, corn. The analog in energy markets would be a household that responds to an electricity ration by running the air conditioner (high value) and turning off the lights (low value), rather than reducing consumption proportionally.

Several caveats apply. First, the treatment is binary, collapsing the substantial variation in allocation stringency across NRDs (from 13 inches per year in the Upper Republican to considerably higher limits elsewhere). The dose-response relationship between allocation stringency and crop specialization may be nonlinear. The most severely restricted NRDs (Upper Republican at 13 inches/year) may have crossed a threshold where corn itself becomes infeasible, but this effect is absorbed by the always-treated exclusion. Second, the county-level data cannot distinguish within-farm from between-farm reallocation; some of the compositional shift may reflect differential exit of wheat-oriented farms rather than individual farmers switching crops. Third, the crop acreage data do not distinguish irrigated from non-irrigated acres in the annual survey; Census of Agriculture years provide irrigated breakdowns but at five-year intervals. Future work should exploit this margin to isolate the effect on irrigated crop choice specifically. Fourth, I lack direct data on irrigation technology adoption (center-pivot conversion, soil moisture sensors), which is a plausible mediator of the lock-in mechanism.


\section{Conclusion}

Mandatory groundwater allocations in Nebraska did not diversify agriculture away from water-intensive corn. They concentrated it. Farmers facing binding pumping quotas abandoned drought-tolerant crops and redirected scarce water to their highest-return use. This ``corn lock-in'' reveals a fundamental tension in quantity-based groundwater regulation: rationing water without differentiating by crop type raises the shadow price of water for all uses, but the behavioral response is to specialize, not diversify. Policymakers seeking both conservation and resilience may need to pair pumping quotas with complementary instruments---crop insurance reform, dryland transition assistance, or tiered pricing that reflects the social cost of irrigation concentration.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Crop Acreage Data} County-level area harvested for corn, sorghum, wheat, and soybeans from the USDA NASS Quick Stats API (source: ``SURVEY,'' frequency: ``ANNUAL,'' unit: ``ACRES,'' geographic level: ``COUNTY,'' state: ``NE''). The query retrieves all available years from 1988 to 2023. Crop shares are constructed by dividing each crop's area harvested by the sum of all four crops' area harvested. County-years with zero total area are excluded.

\subsubsection*{NRD Treatment Timing} NRD adoption dates are compiled from: (i) Nebraska Department of Natural Resources ``NRD Water Management Activities Summary'' (64-page report documenting NRD-by-NRD allocation details); (ii) individual NRD management plans and Integrated Management Plans filed with DNR; (iii) LB 962 ``fully appropriated'' and ``over-appropriated'' designation records. The treatment date is the first year in which mandatory pumping allocations applied to agricultural wells within the NRD.

\subsubsection*{County-NRD Mapping} Each of Nebraska's 93 counties is assigned to its dominant NRD based on geographic overlap using NRD boundary maps from the Nebraska Association of Resources Districts. Of 93 counties, 62 are included in the analysis sample (29 ever-treated across 9 NRDs, 33 never-treated across 11 NRDs). Excluded counties are those where NRD boundaries split the county roughly evenly (creating ambiguous assignment) or where crop acreage data is suppressed due to disclosure restrictions.


\section{Identification Appendix}

\subsubsection*{Treatment Rollout} Nine NRDs adopted mandatory allocations within the sample period (1994--2014), contributing to identification. The adoption sequence---Middle Republican (1994), Lower Republican (1998), Tri-Basin (2005), Central Platte (2006), Twin Platte (2008), North Platte (2010), South Platte (2013), Upper Niobrara-White (2014)---reflects the geographic pattern of Ogallala depletion, moving from the Republican River basin in the southwest to the Platte and Niobrara basins in the north and west. The Upper Republican NRD (1979) is always-treated and excluded from the CS estimator.

\subsubsection*{Pre-Trend Assessment} The singular covariance structure of the event-study estimates prevents computation of the standard pre-test Wald statistic. Visual inspection of pre-treatment coefficients in \Cref{tab:eventstudy} shows no systematic pattern: pre-period estimates are small relative to the post-treatment effects and fluctuate around zero. Formal sensitivity analysis via \citet{rambachan2023more} bounds was not feasible due to the conformability issue between the event-study coefficient vector and its covariance matrix in the CS output.


\section{Robustness Appendix}

\subsubsection*{Block Bootstrap} I implement an NRD-level block bootstrap with 999 iterations. Each iteration resamples NRDs with replacement and re-estimates the TWFE specification. The bootstrap standard error for the corn share effect is 0.042, modestly larger than the cluster-robust SE of 0.040. The bootstrap $p$-value is 0.41, confirming that the corn share increase, while economically meaningful, is not statistically distinguishable from zero. The drought-tolerant share decline is the more robust finding.

\subsubsection*{Alternative Specifications} The CS estimate with never-treated counties only (0.053) is similar to the not-yet-treated baseline (0.045), suggesting that the not-yet-treated comparison group is not contaminated by anticipation effects. The TWFE estimate (0.051) lies between the two CS specifications, consistent with modest heterogeneity-induced bias \citep{goodmanbacon2021difference}.

\subsubsection*{Placebo} Soybeans serve as a placebo outcome. Like corn, soybeans are relatively water-intensive (20--25 inches per season), but unlike wheat and sorghum, they occupy a high-value position in the crop portfolio. If the mechanism is value-per-gallon optimization, soybeans should not decline in response to allocations---and they do not (0.010, $p = 0.67$).


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
