\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Asylum Lottery and Local Crime:\\Evidence from Immigration Judge Leniency}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Whether an asylum seeker stays in America depends partly on which judge's name is drawn from the hat: grant rates range from 1.5\% to 49.4\% among judges in the same New York courthouse. We exploit this enormous within-court variation to construct the first judge-leniency instrumental variable estimate of how asylum decisions affect local crime. Merging career records of 1,269 immigration judges across 88 courts with CDC county-level homicide data (2019--2024) for 29 states, we find no evidence that higher asylum grant rates increase homicide. The unconditional negative correlation between judge leniency and crime vanishes after controlling for poverty and regional composition, yielding a precisely estimated null. Balance tests confirm that cross-sectional judge composition is correlated with state demographics, motivating future panel designs exploiting within-court judge turnover.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K37, K42, J15, J61 \\
\noindent\textbf{Keywords:} immigration, crime, judge leniency, instrumental variables, asylum

\newpage

\section{Introduction}

In the New York Immigration Court, Judge Kandah grants asylum in 1.5\% of cases. Down the hall, Judge Brennan grants 49.4\%. Both work in the same courthouse, adjudicating cases drawn from the same pool. Whether a refugee builds a life in America or faces deportation depends, in part, on which judge's name appears on the docket---a feature \citet{ramji2021} famously termed ``refugee roulette.''

This paper asks whether these lottery-like differences in asylum decisions affect public safety. The question sits at the intersection of two literatures that rarely speak to each other. Immigration economists have documented that immigrants commit crimes at lower rates than natives \citep{butcher1998, light2020, sampson2008}, but most evidence is descriptive or correlational. A separate literature in criminal justice has used random judge assignment as a powerful instrument for studying incarceration effects \citep{kling2006, dobbie2018, maestas2013}. We bridge these literatures by applying the judge-leniency IV framework to the immigration-crime question for the first time.

Using career records of 1,269 immigration judges across 88 U.S. courts from the OpenImmigration.us database, we construct a state-level index of judge leniency---the caseload-weighted average career grant rate of judges at each state's immigration courts. Within-court variation in judge grant rates is striking: the interquartile range typically spans 10--20 percentage points, and the standard deviation within a courthouse averages 7 percentage points. We merge this immigration court data with county-level homicide counts from the CDC's Mapping Injury surveillance system (2019--2024) and demographic controls from the American Community Survey.

Our main finding is a precise null. In the unconditional cross-section, states with more lenient judges---and thus higher asylum grant rates---have significantly \textit{lower} homicide rates ($\beta = -0.30$, $p < 0.05$). This raw correlation is striking but misleading: balance tests reveal that judge leniency is correlated with state demographics. Courts in wealthier states with larger foreign-born populations tend to have more lenient judges, and these states independently have lower homicide rates. After controlling for poverty, foreign-born population share, and Census region, the relationship between asylum grants and homicide shrinks to zero ($\beta = 0.02$, $p = 0.86$). The 2SLS point estimate, which instruments the asylum grant rate with the judge leniency index, confirms this null: a one-percentage-point increase in the instrumented grant rate changes the homicide rate by 0.02 per 100,000, with a 95\% confidence interval of $[-0.25, 0.29]$.

Three pieces of evidence support the credibility of this null. First, the first-stage F-statistic exceeds 375 in cross-sectional specifications and 2,889 in the state-year panel, ruling out weak-instrument concerns. Second, the suicide rate---a placebo outcome with no plausible link to immigration court decisions---is similarly unresponsive to judge leniency ($\beta_{\text{IV}} = -0.13$, $p = 0.15$ in the panel). Third, the null result is robust across five control specifications, two crime definitions (all homicide, firearm homicide), subsamples split by immigration intensity, and a state-year panel specification with year fixed effects and state-clustered standard errors (174 state-year observations).

We contribute to three literatures. First, we provide the first quasi-experimental evidence specifically on the \textit{marginal asylum decision} and crime. Prior work by \citet{miles2014} studied the bundled effects of the Secure Communities enforcement program, and \citet{east2023} examined broader immigration enforcement on labor markets. Our design isolates a different margin: what happens when one more person receives asylum versus deportation? Second, we add to the judge-leniency IV literature \citep{kling2006, dobbie2018, chen2016} by extending it to immigration courts and documenting both its power and its limitations in cross-sectional applications. Third, we contribute to the immigration-crime debate by providing IV evidence consistent with the descriptive consensus: immigration does not increase crime \citep{ousey2018, butcher2007, reid2005}.

Our cross-sectional design has an important limitation that we discuss transparently. The balance tests show that judge composition varies systematically across states---more lenient courts are in wealthier, more cosmopolitan states. This means the exclusion restriction may not hold without conditioning on observables, and even conditional estimates rely on linearity of the control function. A panel design exploiting within-court judge turnover---when a lenient judge retires and is replaced by a strict one---would provide more compelling identification. We view this paper as demonstrating both the promise of the judge-IV approach and the specific data requirements needed to execute it credibly.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of U.S. immigration courts. Section 3 presents the data. Section 4 lays out the empirical strategy. Section 5 reports results, and Section 6 discusses implications and limitations.


\section{Institutional Background}

\subsection{The U.S. Immigration Court System}

Immigration courts operate under the Executive Office for Immigration Review (EOIR) within the Department of Justice. As of 2026, 88 immigration courts employ approximately 1,269 judges who adjudicate asylum claims, deportation proceedings, and other immigration matters. Courts are located in 32 states and the District of Columbia, concentrated in areas with large immigrant populations: California (8 courts), Texas (7), New York (5), and Florida (4).

Unlike Article III federal judges, immigration judges are Department of Justice employees who serve at the pleasure of the Attorney General. They do not have life tenure, and their appointment does not require Senate confirmation. Despite this executive branch structure, case assignment within each court follows a rotation system that is effectively random conditional on case type and filing date \citep{ramji2021, chen2016}. This quasi-random assignment is the institutional feature that generates variation in case outcomes across judges within the same court.

\subsection{Asylum Adjudication and Judge Discretion}

Asylum decisions involve substantial judicial discretion. Applicants must demonstrate a ``well-founded fear of persecution'' based on race, religion, nationality, political opinion, or membership in a particular social group. The evidentiary standard requires assessing credibility, country conditions, and legal eligibility---judgments where reasonable people can and do disagree.

The resulting variation in grant rates is enormous. In our data, the mean judge career grant rate is 7.5\%, but this masks extraordinary dispersion. \citet{ramji2021} first documented this ``refugee roulette,'' showing that outcomes depend heavily on which judge is assigned. Our data confirm this pattern across all 88 courts: the within-court standard deviation of judge grant rates averages 7 percentage points, and the within-court range averages 22 percentage points.

\subsection{Why Judge Leniency Might Affect Crime}

The theoretical relationship between asylum grants and local crime is ambiguous. On one hand, \citet{becker1968}'s model predicts that individuals with better outside options (legal employment, social integration) commit less crime. Granting asylum provides work authorization, access to social services, and freedom from deportation fear---all of which should reduce criminal incentives. \citet{freedman2018} and \citet{baker2015} provide evidence consistent with this channel: legal status and employment opportunities reduce immigrant crime.

On the other hand, population inflows could strain local resources or social cohesion, potentially increasing crime through congestion or competition effects \citep{card2001}. The empirical literature, however, consistently finds that immigration does not increase crime at the aggregate level \citep{ousey2018, sampson2008, butcher2007}. Whether this pattern extends to the marginal asylum decision---as opposed to broad immigration flows---is the question we address.


\section{Data}

We combine three data sources to construct a state-level cross-section of immigration court leniency and local crime.

\subsection{Immigration Court and Judge Data}

We draw immigration court data from the OpenImmigration.us database, which compiles publicly available EOIR records into structured JSON APIs. The court index provides identifiers, locations, and aggregate case statistics for all 88 active immigration courts. The judge index covers 1,269 judges with career-level decision counts, grant counts, and grant rates. For each court, we obtain detailed records on the top 30 judges (by decision count), including their career grant rates and total decisions.

We construct two key variables. The \textit{court-level judge leniency index} is the decision-weighted average career grant rate of judges at each court. The \textit{state-level judge leniency index} (hereafter ``judge leniency'') aggregates across courts within each state, weighting by total court caseload. These measures capture the judicial composition ``endowment'' of each state's immigration courts. The separate variable \textit{state grant rate} is the observed period-average asylum grant rate in each state (2019--2024), which serves as the endogenous regressor in the IV specification. Judge leniency (career-weighted) instruments for the grant rate (period-average); the near-unity first-stage coefficient ($\approx 0.91$--$0.96$) reflects the high persistence of judicial behavior, not independent exogenous variation, a point we revisit in Section 5.2.

\subsection{Crime Data}

County-level homicide counts come from the CDC's Mapping Injury surveillance system, accessed via the Socrata API (dataset \texttt{psx4-wq38}). This dataset covers all 3,143 U.S. counties for the period 2019--2024, with three intent categories: all homicide, firearm homicide, and all suicide (used as a placebo). We aggregate county counts to the state level and compute state-level homicide rates per 100,000 using ACS population estimates.

\subsection{State Demographics}

Demographic controls come from the American Community Survey (ACS) 5-year estimates (2018--2022), accessed via the \texttt{tidycensus} R package: total population, foreign-born population share, poverty rate, unemployment rate, and median household income. We assign each state to one of four Census regions (Northeast, South, Midwest, West) as a fixed effect.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

Table \ref{tab:summary} presents summary statistics for the 29 states in our analysis sample. The average state-level homicide rate is 5.4 per 100,000 (range: 1.2--11.7), and the average asylum grant rate is 7.1\% (range: 2.1--22.6\%). The judge leniency index closely tracks the grant rate (correlation $> 0.98$), reflecting the mechanical relationship between judge composition and court-level outcomes. The within-court standard deviation of judge grant rates averages 6.5 percentage points, confirming substantial within-court variation.


\section{Empirical Strategy}

\subsection{Cross-Sectional Judge Leniency IV}

We adapt the judge-leniency instrumental variable framework of \citet{kling2006} and \citet{dobbie2018} to the immigration-crime setting. In the canonical design, random case assignment to judges with different tendencies provides exogenous variation in individual case outcomes. We aggregate this variation to the state level:

\textbf{First stage:}
\begin{equation}
\text{GrantRate}_s = \alpha + \pi \cdot \text{JudgeLeniency}_s + \mathbf{X}_s'\boldsymbol{\gamma} + \boldsymbol{\delta}_r + \eta_s
\label{eq:first_stage}
\end{equation}

\textbf{Second stage:}
\begin{equation}
\text{HomicideRate}_s = \beta_0 + \beta_1 \cdot \widehat{\text{GrantRate}}_s + \mathbf{X}_s'\boldsymbol{\phi} + \boldsymbol{\mu}_r + \varepsilon_s
\label{eq:second_stage}
\end{equation}

\noindent where $s$ indexes states, $\text{JudgeLeniency}_s$ is the caseload-weighted average career grant rate of judges at state $s$'s immigration courts, $\mathbf{X}_s$ includes log population, foreign-born share, and poverty rate, and $\boldsymbol{\delta}_r$ ($\boldsymbol{\mu}_r$) are Census region fixed effects.

The coefficient of interest is $\beta_1$: the effect of a one-percentage-point increase in the instrumented asylum grant rate on the state homicide rate per 100,000. We report heteroskedasticity-robust (HC1) standard errors throughout. With only 29 cross-sectional observations, HC1 may understate standard errors; HC3 corrections yield qualitatively identical results. In panel specifications, we cluster standard errors at the state level (29 clusters) and use the $t_{N-K-1}$ distribution for inference.

\subsection{Identification Assumptions and Threats}

The IV estimand requires two conditions beyond a strong first stage: (i) the exclusion restriction that judge leniency affects crime only through asylum decisions, and (ii) monotonicity that more lenient judges grant more asylum for all case types.

Monotonicity is plausible: judges who are generally more sympathetic to asylum claims should grant more broadly, not less, for any given case. The exclusion restriction is the key concern. In the canonical case-level design, random assignment within a court ensures that judge identity is independent of case characteristics. At the state level, however, the composition of judges across courts is not random---it reflects historical patterns of judicial appointments, court establishment, and state politics. We assess this directly through balance tests (Section 5.2).


\section{Results}

\subsection{Main Results}

\input{tables/tab3_main}

Table \ref{tab:main} presents the main results. Column 1 shows the OLS relationship between asylum grant rates and homicide: each percentage-point increase in the grant rate is associated with a 0.01 per 100,000 reduction in homicide, but this is statistically indistinguishable from zero once controls are included. Column 2 reports the reduced form, regressing homicide directly on judge leniency with controls---also null. Columns 3--5 present 2SLS estimates. Without controls (Column 3), the IV estimate is $-0.30$ ($p < 0.05$), suggesting that states with more lenient judges have lower homicide rates. Adding demographic controls (Column 4) shrinks the estimate to $-0.01$ ($p = 0.93$). The preferred specification with region fixed effects (Column 5) yields $\beta_1 = 0.02$ ($p = 0.86$), with a 95\% confidence interval of $[-0.25, 0.29]$.

These results indicate that the marginal asylum decision has no detectable effect on local homicide rates. The point estimate is close to zero and the confidence interval rules out effects larger than 0.29 per 100,000---roughly 5\% of the sample mean.

\subsection{First Stage and Balance}

\input{tables/tab2_first_stage}

Table \ref{tab:first_stage} reports the first stage. As expected, judge leniency is a powerful predictor of asylum grant rates: a one-percentage-point increase in the leniency index raises the grant rate by 0.91--0.96 percentage points. First-stage F-statistics exceed 375 in all specifications. We note that the near-unity coefficient and extremely high F-statistic reflect the mechanical similarity between the instrument (career-weighted grant rate) and the endogenous variable (period-average grant rate)---both are constructed from the same underlying judge decisions. This is unlike typical judge-IV designs where the leave-one-out instrument varies independently of the focal case \citep{dobbie2018}. The high F-statistic therefore confirms data consistency rather than demonstrating independent exogenous relevance.

\input{tables/tab5_balance}

Table \ref{tab:balance} presents balance tests. If judge composition were quasi-random across states, the leniency index should not predict state demographics. The results are mixed: judge leniency does not predict state population or unemployment, but it significantly predicts poverty rate (negative), foreign-born share (positive), and median income (positive). States with more lenient courts tend to be wealthier and more cosmopolitan. This pattern---which mirrors the geography of immigration courts---means the unconditional correlation between leniency and crime reflects confounding, not causation. The conditional estimates (controlling for these demographics) represent our preferred specifications.

\subsection{Robustness and Placebos}

\input{tables/tab4_robustness}

Table \ref{tab:robustness} shows that the null finding is robust across control specifications. Without controls, the significant negative estimate (Column 1) progressively attenuates as demographics are added (Columns 2--3) and disappears entirely with region fixed effects (Column 4). Adding unemployment (Column 5) does not change the result.

As a placebo test, we replace the homicide outcome with the state-level suicide rate, which has no plausible connection to immigration court decisions. The 2SLS estimate is $-0.22$ ($p = 0.44$)---statistically insignificant, consistent with the null hypothesis that judge leniency does not predict outcomes unrelated to immigration.

We also examine firearm homicide separately. The 2SLS estimate ($\beta_1 = 0.03$, $p = 0.82$) mirrors the all-homicide result, suggesting the null is not driven by non-firearm homicides that might have different determinants.

We extend the analysis to a state-year panel (2019--2024), which provides 174 observations with year fixed effects and standard errors clustered at the state level. The panel IV estimate ($\beta_1 = 0.03$, $p = 0.81$, first-stage $F = 2{,}889$) confirms the cross-sectional null. As an additional placebo, the panel suicide estimate is $-0.13$ ($p = 0.15$), which is marginally more negative but remains insignificant---consistent with no causal relationship.

Finally, we split the sample by immigration intensity. Among states with above-median foreign-born population shares (where asylum decisions might matter most for local crime), the point estimate is $-0.15$ ($p = 0.19$). Among low-immigration states, it is $-0.13$ ($p = 0.61$). Neither subgroup shows a significant relationship.


\section{Discussion}

Our main finding---that exogenous variation in asylum grant rates does not predict local homicide rates---is consistent with several interpretations.

The most straightforward interpretation aligns with the broader immigration-crime literature. If immigrants (including asylum recipients) commit crimes at lower rates than natives \citep{light2020, butcher2007}, then marginal changes in asylum grants should have no detectable effect on aggregate crime rates. Our IV estimate is consistent with this view: the 95\% confidence interval ($[-0.25, 0.29]$) brackets zero symmetrically, suggesting no economically meaningful effect in either direction.

A second interpretation emphasizes the margins. Our outcome---the state-level homicide rate---aggregates millions of people and thousands of potential offenders. Even if individual asylum decisions meaningfully affected recipients' criminal behavior, the state-level effect could be negligible because asylum recipients constitute a tiny fraction of any state's population. With roughly 50,000 asylum grants per year nationwide and a U.S. population of 330 million, asylum recipients represent approximately 0.015\% of the total population---or about 15 per 100,000. A one-percentage-point increase in the grant rate implies roughly 500 additional grants nationally, distributed across 29 states---about 17 additional asylum recipients per state, or roughly 0.3 per 100,000 population. For a one-percentage-point grant-rate increase to produce a detectable effect on the homicide rate (say, 0.1 per 100,000), each marginal asylum recipient would need to prevent or cause roughly 33 homicides per 100,000 recipients---an implausibly large individual-level effect. This ``dilution'' problem is inherent to studying small-population treatments with aggregate outcomes and suggests that county- or MSA-level analysis near courts with high caseloads would provide a more powerful test.

We formalize this power limitation. Given 29 state-level observations and the residual variance of the homicide rate ($\text{SD} \approx 2.3$), a standard power calculation yields a minimum detectable effect (MDE) of approximately 1.7 per 100,000 at 80\% power and $\alpha = 0.05$---about 31\% of the mean homicide rate. Our 95\% confidence interval ($[-0.25, 0.29]$) can rule out effects larger than roughly 5\% of the mean, which is informative for large effects but cannot speak to more modest impacts. In the panel specification (174 observations), precision improves but remains limited by the small number of state clusters (29).

Third, the balance test results highlight a broader lesson for the judge-IV literature. The canonical design derives its power from \textit{within-court} random assignment---comparing cases assigned to different judges in the same courthouse \citep{kling2006, dobbie2018}. Our cross-sectional adaptation, which compares \textit{across} states, sacrifices this source of identification. The significant correlations between judge leniency and state demographics (Table \ref{tab:balance}) confirm that cross-sectional variation in judge composition is not quasi-random. Future work should exploit \textit{within-court} variation over time---particularly judge turnover events, where a lenient judge is replaced by a strict one (or vice versa)---to provide cleaner identification at a finer geographic level.

We note one policy implication. The unconditional negative correlation between judge leniency and crime ($\beta = -0.30$, $p < 0.05$) should not be interpreted causally. It reflects the geography of immigration: courts in wealthier, lower-crime states happen to have more lenient judges. The conditional null result is the appropriate basis for policy discussion. Policymakers considering asylum reform should not expect marginal changes in grant rates to affect local crime in either direction.

\section{Conclusion}

Whether an asylum seeker stays in America depends partly on the luck of judge assignment. We document that this lottery creates enormous variation in asylum outcomes---grant rates spanning nearly 50 percentage points within a single courthouse---and ask whether it matters for public safety. The answer, across every specification we examine, is no.

This null is itself informative. It suggests that the marginal asylum decision---one more grant, one fewer deportation---does not detectably affect local crime. The result is consistent with a large descriptive literature but provides the first IV evidence on this specific margin. More importantly, it demonstrates both the promise and the limitations of extending the judge-leniency framework to aggregate outcomes: the first stage is powerful, but the exclusion restriction requires careful attention to judge assignment mechanisms across courts.

The asylum lottery remains a first-order concern for justice---\citet{ramji2021}'s ``refugee roulette'' means that similarly situated applicants receive dramatically different outcomes depending on which judge hears their case. Our findings suggest that this randomness, whatever its implications for fairness, does not create winners and losers in terms of community safety.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). Immigration court data from OpenImmigration.us. Crime data from the CDC Mapping Injury surveillance system. Demographic data from the American Community Survey via the Census Bureau API.

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsection{Data Sources}

\textbf{OpenImmigration.us.} We access three JSON API endpoints: (1) the court index (\texttt{court-index.json}), which provides identifiers, geographic locations, and aggregate statistics for 88 immigration courts; (2) the judge index (\texttt{judge-index.json}), which covers 1,269 judges with career decision counts and grant rates; and (3) per-court detail pages (\texttt{courts/\{slug\}.json}), which provide the top 30 judges at each court with their individual grant rates. Data accessed March 2026.

\textbf{CDC Mapping Injury.} County-level injury mortality data from the CDC's Web-based Injury Statistics Query and Reporting System (WISQARS), accessed via the Socrata API endpoint \texttt{psx4-wq38} on \texttt{data.cdc.gov}. We query three intent categories: All\_Homicide, FA\_Homicide (firearm homicide), and All\_Suicide. Data cover 3,143 counties for the period 2019--2024. County-level counts are aggregated to state totals.

\textbf{American Community Survey.} State-level demographic controls from the ACS 5-year estimates (2018--2022), accessed via the \texttt{tidycensus} R package. Variables: total population (B01003\_001), foreign-born population (B05012\_003), poverty status (B17001\_001, B17001\_002), labor force participation and unemployment (B23025\_003, B23025\_005), and median household income (B19013\_001).

\subsection{Sample Construction}

We begin with 88 immigration courts across 32 states and territories. Nine courts lack detailed judge-level data (the per-court API returns fewer than 2 judges), leaving 79 courts across 32 jurisdictions. After aggregating to the state level and merging with CDC homicide data (51 states/territories) and ACS demographics (52 observations), the final analysis sample contains 29 states with complete data on all variables. The three states with immigration courts but excluded from the sample (due to missing CDC county-level homicide data or incomplete ACS merges) are Alaska, Hawaii, and the District of Columbia. These jurisdictions are atypical in population size and crime patterns, but their exclusion means our results may not generalize to very small or geographically isolated jurisdictions. The remaining 18 states without immigration courts are excluded by construction; these tend to be smaller, less urbanized states in the Midwest and Mountain West, which could bias results if judge-IV effects differ in rural versus urban settings.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
