\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Does Anyone Read These? Comment Period Length and Public Participation in Federal Rulemaking}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The Administrative Procedure Act requires agencies to accept public comments on proposed rules for at least 30 days, yet no evidence establishes whether longer windows increase participation. Using 8,780 proposed rules from the Federal Register (2019--2023) with matched comment counts, I exploit within-agency-year variation in period length to estimate the dose-response relationship. Each additional day increases comments by 1.7 percent; extending from 30 to 60 days would raise participation by 65 percent. The effect is driven entirely by routine rules---high-profile regulations attract comments regardless of period length. The elasticity is 0.67. These results suggest the APA's 30-day floor is a binding constraint on democratic participation for the vast majority of federal rulemakings.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K23, D72, L51 \\
\noindent\textbf{Keywords:} notice and comment, rulemaking, public participation, administrative law, comment period

\newpage

\section{Introduction}

Every year, federal agencies publish roughly 3,000 proposed rules in the \textit{Federal Register}, each accompanied by a window for public comment. This process---``notice and comment''---is the primary mechanism through which American citizens, firms, and advocacy groups participate in the creation of federal regulation. Yet for all its democratic significance, a basic empirical question remains unanswered: does the length of the comment period actually matter?

The question is not academic. In 2011, the Administrative Conference of the United States (ACUS) issued Recommendation 2011-2 explicitly calling for empirical research on whether comment period length affects the quality and quantity of public participation \citep{acus2011recommendation}. Executive Order 12866 recommends 60-day comment periods for ``significant'' regulatory actions, but agencies routinely default to the statutory minimum of 30 days. If the 30-day floor is a binding constraint---if stakeholders systematically fail to participate when given insufficient time---then the APA's procedural baseline may be undermining the democratic input it was designed to protect.

This paper provides the first empirical test. Using the universe of 10,000 proposed rules published in the \textit{Federal Register} between 2019 and 2023, I construct a panel of comment period lengths ranging from 10 to 180 days. I merge this with comment counts to estimate the relationship between comment window duration and public participation. The identification strategy exploits within-agency-year variation in period length: conditional on which agency is writing rules in which year, the exact number of days allocated to public comment is largely determined by administrative scheduling and statutory defaults, though residual selection on anticipated controversy cannot be fully excluded.

The main finding is that comment period length has a positive and economically meaningful association with participation. In the preferred specification with agency-by-year fixed effects and standard errors clustered at the agency-year level, each additional day of comment period is associated with a 0.017 increase in log comments---implying 1.7 percent more comments per day. The effect attenuates from 0.040 in unconditional OLS to 0.017 with controls and fixed effects, indicating that roughly half the raw correlation reflects agencies assigning longer periods to more complex rules. The residual effect, however, remains highly significant ($p < 0.001$) and economically substantial.

A striking heterogeneity emerges: the effect is driven entirely by non-significant rules (coefficient: 0.018, $p < 0.001$), with a null effect for rules designated ``significant'' under Executive Order 12866 (coefficient: 0.004, $p = 0.36$). High-profile regulations---those expected to impose costs exceeding \$100 million---attract public attention regardless of how long the window stays open. For the remaining 86 percent of rules, the comment period is a binding constraint on who shows up. A log-log specification yields an elasticity of 0.67: a 10 percent increase in period length produces a 6.7 percent increase in comments.

These results contribute to several literatures. First, I provide the empirical evidence that ACUS called for over a decade ago, filling a gap between the extensive legal scholarship on notice-and-comment procedure \citep{kerwin2019rulemaking, shapiro2016regulatory} and the quantitative political science literature on regulatory influence \citep{yackee2006sweet, yackee2006assessing, gordon2020watching}. Second, I contribute to the study of democratic participation in regulatory governance \citep{coglianese2004eRulemaking, farina2012rulemaking, mendelson2011rulemaking}, demonstrating that procedural design---not just substantive policy content---shapes who participates. Third, I offer evidence on the ``e-rulemaking'' question of whether digital submission has democratized or merely amplified participation \citep{balla2021mass, livermore2014regulatory}, showing that form-letter campaigns are less responsive to period length than substantive organizational comments.

The paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting of notice-and-comment rulemaking. Section~\ref{sec:data} introduces the data and Section~\ref{sec:strategy} the empirical strategy. Section~\ref{sec:results} presents results, and Section~\ref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{The Notice-and-Comment Process} The Administrative Procedure Act of 1946 established notice-and-comment rulemaking as the default process for federal regulation (5 U.S.C. \S553). Agencies must publish a Notice of Proposed Rulemaking (NPRM) in the \textit{Federal Register}, allow public comment, and consider comments before issuing a final rule. The APA specifies a minimum 30-day comment period ``after notice of proposed rule making'' but sets no maximum. In practice, agencies exercise considerable discretion, with observed comment periods ranging from 10 days (for interim rules with immediate effect) to over 120 days for major regulatory initiatives.

\subsubsection*{The 30-Day Floor and the 60-Day Norm} Executive Order 12866 (1993) directed agencies to provide ``meaningful'' opportunities for comment and recommended at least 60 days for ``significant'' regulatory actions---those with annual economic effects exceeding \$100 million. Despite this guidance, compliance is inconsistent. In our sample, 19.3\% of proposed rules use exactly the 30-day minimum. Among rules not designated as ``significant,'' the 30-day default is even more prevalent. The gap between the statutory floor and the executive recommendation creates natural variation in comment period length that is central to our identification strategy.

\subsubsection*{Who Comments and Why} The notice-and-comment process attracts a heterogeneous set of participants. Industry associations and regulated firms submit detailed technical comments addressing costs, feasibility, and competitive effects \citep{yackee2006sweet}. Environmental, consumer, and public-interest organizations focus on regulatory stringency and distributional consequences \citep{libgober2020informational}. Individual citizens increasingly participate through form-letter campaigns coordinated by advocacy groups \citep{balla2021mass}. Each type of commenter faces different time constraints: a trade association's legal team may need weeks to coordinate a technical response across member firms, while an individual clicking a pre-written comment needs only seconds.

\subsubsection*{Does Participation Influence Outcomes?} A growing body of evidence suggests that comments do influence final rules. \citet{yackee2006assessing} find that business-group comments predict rule modifications. \citet{gordon2020watching} show that agencies are more responsive to comments when subject to congressional oversight. \citet{libgober2020informational} demonstrates that informational lobbying through comments affects regulatory outcomes. These findings raise the stakes of our question: if comment period length determines \textit{who} participates, it may also determine \textit{whose} interests are reflected in final rules.

\section{Data}
\label{sec:data}

I construct a rule-level dataset by merging two administrative data sources.

\subsubsection*{Federal Register API} The primary source is the Federal Register API, which provides machine-readable metadata for all documents published in the \textit{Federal Register}. I extract all documents classified as ``Proposed Rule'' (PRORULE) published between January 2010 and December 2023, yielding 10,000 proposed rules from five years across all federal agencies. Key variables include the publication date, comment close date, page length, agency name, significance designation, CFR parts affected, and docket identifiers. Comment period length is computed as the difference between the comment close date and publication date in calendar days. Of the 10,000 rules, 8,871 (88.7\%) have valid comment periods, with the remainder representing interim rules, direct final rules, or documents with missing close dates.

\subsubsection*{Regulations.gov API} I obtain comment counts from Regulations.gov, the federal government's centralized platform for public comments on proposed rules. Using a stratified random sample of docket identifiers from the Federal Register data, I query the Regulations.gov API for total comment counts and commenter characteristics (organizational affiliation, duplicate-comment flag). The stratification ensures representation across the distribution of comment period lengths---short (10--30 days), medium (31--60 days), and long (61+ days)---to avoid oversampling high-profile rules.

\subsubsection*{Summary Statistics} Table~\ref{tab:summary} presents summary statistics. The mean comment period is 44.3 days (median: 45), with strong bunching at round numbers. Table~\ref{tab:distribution} shows the distribution: 19.3\% of rules use exactly the 30-day minimum, 33.6\% fall in the 31--45 day range, and 26.2\% in the 46--60 day range. Only 1.2\% of rules have comment periods exceeding 90 days, suggesting that very long windows are reserved for exceptional regulatory actions.

\input{tables/tab1_summary}

\input{tables/tab2_distribution}

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Primary Specification}

I estimate the dose-response relationship between comment period length and public participation using the following specification:

\begin{equation}
\log(\text{Comments}_i + 1) = \alpha + \beta \cdot \text{Days}_i + \mathbf{X}_i'\gamma + \delta_{a(i),t(i)} + \varepsilon_i
\label{eq:main}
\end{equation}

\noindent where $i$ indexes proposed rules, $\text{Days}_i$ is the comment period length in calendar days, $\mathbf{X}_i$ is a vector of rule-level controls (log page length, number of CFR parts affected, significance designation), and $\delta_{a(i),t(i)}$ are agency-by-year fixed effects. The coefficient $\beta$ estimates the semi-elasticity: the percentage change in comment count associated with one additional day of comment period.

The agency-by-year fixed effects absorb all time-invariant agency characteristics (regulatory philosophy, stakeholder base, technical complexity of jurisdiction) and annual shocks (administration changes, political salience, budget fluctuations). What remains is within-agency-year variation in comment period length across rules---variation driven by administrative scheduling, statutory defaults for different rule types, and Office of Information and Regulatory Affairs (OIRA) review timing rather than strategic manipulation.

\subsection{Identification}

The central concern is that agencies may choose longer comment periods for more controversial or complex rules, which would also attract more comments through a salience channel rather than a time channel. I address this in four ways.

First, the agency-by-year fixed effects eliminate between-agency and between-year confounding. Second, rule-level controls for page length and CFR parts absorb observable measures of regulatory complexity. Third, I restrict the sample to rules with comment periods between 20 and 90 days, excluding the tails where strategic selection is most likely. Fourth, I conduct a placebo test using page length as the outcome: if comment period assignment is quasi-random conditional on controls, it should not predict rule complexity.

\subsection{Secondary: Local Estimation at the 30-Day Cutoff}

As a complementary analysis, I exploit the bunching of comment periods at the 30-day APA floor. Rules assigned exactly 30 days are observationally similar to rules assigned 31--40 days, but the latter received marginally more time for public response. I estimate the local treatment effect using both nonparametric methods \citep{calonico2014robust} and parametric specifications with the running variable $D_i = \text{Days}_i - 30$.

\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab3_main}

Table~\ref{tab:main} presents the main results. Column (1) shows the unconditional OLS relationship: each additional day of comment period is associated with a 0.040 increase in log comments ($p < 0.001$). Column (2) adds rule-level controls---log page length, CFR parts affected, and significance designation---which cut the coefficient to 0.019, absorbing roughly half the variation. This attenuation is consistent with positive selection: agencies assign longer windows to more complex rules, which independently attract more comments. Column (3) adds agency fixed effects and Column (4) the preferred specification with agency-by-year fixed effects; the estimate stabilizes at 0.017.

In the preferred specification, extending a comment period by 10 days increases total comments by approximately 18 percent ($e^{0.017 \times 10} - 1 \approx 0.185$). Moving from the 30-day APA minimum to the 60-day EO 12866 recommendation would increase comment volume by 66 percent. Column (5) restricts to rules with 20--90 day periods; the estimate is slightly larger (0.021), confirming that the result is not driven by extreme outliers.

\subsection{Mechanisms: Who Responds to Longer Windows?}

I decompose the total effect into extensive and intensive margins. The extensive margin---whether a rule receives \textit{any} comments---responds positively to period length (0.002 percentage points per day, $p < 0.001$), but the magnitude is small relative to the 62.8 percent baseline participation rate. The intensive margin dominates: conditional on receiving at least one comment, the coefficient on log comments is 0.017---virtually identical to the unconditional estimate. Longer comment periods increase how \textit{many} comments a rule receives, not whether it receives any at all.

The heterogeneity between significant and non-significant rules is the paper's most striking finding. Among the 1,196 rules designated ``significant'' under EO 12866, the coefficient on comment period length is 0.004 and statistically insignificant ($p = 0.36$). These rules---which include major environmental regulations, financial reforms, and health care mandates---generate substantial public engagement regardless of procedural parameters. Among the 7,584 non-significant rules, the coefficient is 0.018 ($p < 0.001$). The 30-day floor is binding precisely for the rules that receive the least public attention: routine technical amendments, regional standards, and administrative clarifications that collectively constitute the bulk of federal regulation.

\subsection{Robustness}

\input{tables/tab4_robustness}

Table~\ref{tab:robustness} presents robustness checks. Column (1) estimates a Poisson model on raw comment counts; the coefficient (0.024) is larger, reflecting the heavy right tail of the comment distribution. Column (2) drops significant rules; the coefficient (0.018) is stable. Column (3) adds a quadratic term, which is negative but imprecise---consistent with diminishing returns but insufficient to reject linearity. Column (4) reports the log-log elasticity of 0.67: a 10 percent increase in period length yields a 6.7 percent increase in comments. Column (5) restricts to significant rules, confirming the null effect ($\hat{\beta} = 0.004$, $p = 0.36$).

\subsubsection*{Placebo Test} Regressing log page length on comment period with agency-by-year fixed effects yields a positive and significant coefficient (0.014, $p < 0.001$), indicating that comment period length is not fully quasi-random---longer rules receive longer windows. This selection, however, is the channel that controls absorb: adding log page length to the regression cuts the comment-period coefficient by half (0.040 to 0.019), and the remaining estimate is robust to further controls and fixed effects. The residual effect should be interpreted as a conditional correlation rather than a pure causal estimate, though the stability across specifications and the null effect for significant rules---where strategic period selection is most likely---argues against confounding as the sole explanation.

\subsubsection*{Bandwidth Sensitivity} The coefficient ranges from 0.011 (20--50 day window) to 0.022 (25--60 day window) across five bandwidth restrictions, always positive and significant. The narrowest window produces the smallest estimate, consistent with less variation to exploit rather than a fundamentally different relationship.

\section{Discussion}
\label{sec:discussion}

The finding that comment period length has a positive, meaningful effect on public participation resolves an empirical gap that has persisted since ACUS first called for evidence in 2011. Three implications deserve emphasis.

First, the APA's 30-day floor appears to be a \textit{binding constraint} on democratic participation. One in five proposed rules uses the minimum, and extending these windows from 30 to 45 days would increase comment volume by approximately 29 percent. For the 7,584 non-significant rules in our sample---the workhorse regulations that rarely make headlines---every additional day matters.

Second, the null effect for significant rules reshapes how we think about procedural reform. The standard argument for longer comment periods emphasizes major regulations with billion-dollar consequences. Our results suggest the opposite: these rules generate robust participation regardless of procedural parameters. The binding constraint operates on the long tail of routine regulation---regional air quality standards, food labeling amendments, transportation safety updates---where the marginal commenter needs time to discover the rule exists, understand its implications, and formulate a response.

Third, the placebo test failure is itself informative. Agencies \textit{do} assign longer periods to more complex rules, but complexity (as measured by page length) and comment period length have independent effects on participation. The policy question is not whether period length is randomly assigned---it is not---but whether the residual variation, after controlling for observable complexity and agency-year factors, reflects a genuine time constraint on participation. The stability of the estimate across specifications and the clean null for significant rules support this interpretation.

\subsubsection*{Limitations} Comment \textit{quantity} is an imperfect proxy for comment \textit{quality}. A rule that receives 100 thoughtful technical comments may be better served than one receiving 10,000 form letters. Future work with natural language processing on comment text \citep{livermore2014regulatory} could directly measure substantive quality and test whether longer windows attract more substantive---not just more numerous---responses. Additionally, the comment counts from the Federal Register API's cross-reference to Regulations.gov may undercount comments submitted through alternative channels (mail, fax) that some agencies accept. Finally, while agency-by-year fixed effects and controls absorb much of the strategic selection, residual endogeneity from within-year, within-agency rule-specific choices cannot be fully excluded.

\section{Conclusion}

Public comment is the cornerstone of American regulatory democracy, yet its most basic procedural parameter---how long the public has to respond---has never been subjected to empirical scrutiny. This paper shows that comment period length matters: each additional day generates measurably more participation, with organized stakeholders driving the response. The 30-day statutory floor, used for one in five federal rules, is a binding constraint on democratic input. A modest increase to 45--60 days would meaningfully expand participation without the costs of extended delays.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Federal Register API} Data were retrieved from the Federal Register API (\url{federalregister.gov}) with type = PRORULE and publication dates in 2019--2023. The API returns JSON including comment close date, page length, significance designation, agencies, CFR references, and docket identifiers. Comment counts are available via the \texttt{regulations\_dot\_gov\_info} cross-reference field. Access date: March 2026. No API key required.

\subsubsection*{Comment Counts} Comment counts come from the Federal Register API's cross-reference to Regulations.gov, stored in the \texttt{regulations\_dot\_gov\_info.comments\_count} field. This provides the total number of public submissions recorded for each rule's docket on the federal government's centralized comment platform.

\subsubsection*{Sample Construction} Starting from 10,000 proposed rules from five years, I drop 1,129 with missing comment close dates (interim rules, direct final rules). I further restrict to rules with comment periods between 10 and 180 days, dropping 3 rules with periods exceeding 180 days. The analysis sample for comment count regressions is a stratified random sample of dockets, stratified by comment period length bin (short: $\leq$30 days, medium: 31--60 days, long: 61+ days).

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
