\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics (no figures in V1)
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Announcement Tax: Disentangling Policy Uncertainty from Implementation Effects in Taiwan's Capital Gains Tax Experiment}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Taiwan introduced a capital gains tax on securities in January 2013 and fully repealed it in November 2015---one of the shortest-lived capital tax experiments in modern history. I exploit this round-trip to decompose the market response into an announcement effect and an implementation effect. Aggregate share volume declined 24\% during the nine-month announcement period but fully recovered once the tax took effect, while the number of daily transactions remained persistently depressed throughout the CGT period. This divergence---volume recovery without transaction recovery---is consistent with retail investors exiting while institutional investors compensated with larger but fewer trades. The round-trip reveals that the market disruption commonly attributed to the tax itself was predominantly an announcement-period uncertainty shock, not a response to the tax's actual implementation.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G12, G18, H24, D83 \\
\noindent\textbf{Keywords:} capital gains tax, policy uncertainty, stock market, trading volume, investor composition, Taiwan

\newpage

\section{Introduction}

When Taiwan's Finance Minister proposed reintroducing a capital gains tax on securities in April 2012, trading volume on the Taiwan Stock Exchange plummeted. Within three months, average daily share volume had fallen by nearly a quarter. The episode is routinely cited as evidence that capital gains taxes devastate stock market activity \citep{hawkins2003taiwan}. But what happened next complicates this narrative: by the time the tax actually took effect in January 2013, trading volume had recovered to pre-announcement levels---and it stayed there throughout the three years the tax was in force.

This paper decomposes the market's response to Taiwan's CGT experiment into two distinct channels: an \emph{announcement effect} driven by policy uncertainty during the legislative process, and an \emph{implementation effect} capturing the tax's impact once it was actually in place. The decomposition reveals a striking asymmetry. Aggregate share volume declined sharply during the announcement period (April--December 2012) but showed no decline during the CGT-active period (2013Q1--2015Q3) relative to the pre-announcement baseline. However, the \emph{number of daily transactions} told a different story: transaction counts fell during the announcement period and remained depressed throughout the CGT's life, recovering only after repeal.

The divergence between volume and transactions is the paper's central finding. If aggregate share volume recovered while transaction counts did not, the composition of trading must have shifted: fewer participants making larger trades. This pattern is consistent with retail investors---who make many small transactions---exiting the market, while institutional investors---who execute fewer but much larger trades---filled the gap. The CGT targeted individual investors with gains exceeding TWD 1 billion, so this composition shift aligns with the tax's incidence structure.

Taiwan's three-year experiment is rare in the capital tax literature. Most CGT studies rely on cross-jurisdictional variation \citep{poterba2002taxation, auerbach1989capital} or tax reform events that lack a clean reversal \citep{sialm2009tax, yagan2015capital}. Sweden's 1984--1991 transaction tax \citep{umlauf1993transaction} and France's 2012 FTT \citep{colliard2012financial, pomeranets2017stock} provide partial evidence, but neither was repealed cleanly enough to separate policy effects from secular trends. Taiwan's CGT round-trip---introduction followed by full repeal---enables a within-market test where the repeal serves as a built-in falsification: any effect attributed to the tax that persists after repeal is more plausibly a secular trend.

The paper contributes to three literatures. First, it advances the capital gains tax literature by showing that the behavioral response to CGT announcements can exceed the response to implementation---a finding consistent with the policy uncertainty framework of \citet{baker2016measuring}. The canonical lock-in literature \citep{feldstein1980effects, burman1994measuring, dai2008capital, jin2006capital} focuses on how existing taxes affect realization decisions; this paper shows that the \emph{prospect} of a new tax can generate larger market disruptions than the tax itself.

Second, it speaks to the debate about whether financial transaction taxes reduce market quality \citep{stiglitz1989using, summers1989when, tobin1978proposal, habermeier2003securities, jones1997transactions}. The finding that aggregate volume is resilient while market participation narrows suggests that the efficiency costs of capital taxation may operate through investor composition rather than aggregate liquidity---a channel emphasized by \citet{delong1990noise} and \citet{amihud2002illiquidity}.

Third, the volume-transaction divergence provides direct evidence on the extensive margin of market participation under capital taxation. \citet{chetty2005dividend} and \citet{yagan2015capital} document real-economy responses to capital tax changes, but both operate through firm-level channels. This paper identifies a market-level participation channel: the CGT compressed the retail investor base without reducing aggregate trading activity.

The paper proceeds as follows. Section 2 describes the institutional background. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{The 24-Year Absence} Taiwan first attempted a capital gains tax on securities in 1989, during a speculative bubble in which the TAIEX index exceeded 12,000 points. The announcement triggered a severe market crash, and the government swiftly suspended the tax. From 1990 through 2012, Taiwan maintained a securities transaction tax (STT) of 0.3\% on sales but imposed no tax on capital gains from securities. This created a regime where investors paid a flat proportional cost per trade but faced no marginal tax on realized profits, making the TWSE one of the few major exchanges globally with zero capital gains taxation.

\subsubsection*{The 2012 Announcement and Legislative Process} On April 12, 2012, Finance Minister Christina Liu proposed reintroducing a CGT as part of a broader fiscal reform. The proposal generated intense political debate, underwent multiple legislative revisions, and was not finalized until July 25, 2012. The nine-month window from announcement to implementation (April 2012 through December 2012) was characterized by genuine uncertainty about the tax's scope, rate structure, and thresholds. Initial proposals included rates as high as 20\% with lower thresholds; the final version imposed a 15\% rate only on individual investors with annual gains exceeding TWD 1 billion (approximately USD 34 million).

\subsubsection*{Implementation: January 2013} The CGT took effect on January 1, 2013. The extraordinarily high threshold meant that fewer than 1\% of individual investors would directly owe the tax in any given year. Institutional investors, including foreign portfolio investors and domestic investment trusts, faced separate treatment under existing tax treaties and domestic law. Despite the narrow direct incidence, market participants widely viewed the CGT as establishing a precedent for future broadening of the tax base.

\subsubsection*{Repeal: November 2015} Facing continued political pressure, the Legislative Yuan voted to fully repeal the CGT on November 17, 2015. The repeal was clean: no transition period, no phaseout, and retroactive to January 1, 2016. This sharp reversal provides the second leg of the round-trip design.


\section{Data}

The primary data source is the Taiwan Stock Exchange (TWSE) Open API, which provides daily market-level aggregate statistics for all TWSE-listed securities. I construct a monthly panel of TWSE market activity from January 2010 through December 2018, yielding 104 monthly observations spanning the full pre-announcement, announcement, CGT-active, and post-repeal periods.

\subsubsection*{Key Variables} The main outcomes are: (1) log average daily trading volume (shares), measuring total market turnover in quantity terms; and (2) log average daily number of transactions, measuring the frequency of trading activity. The transaction count is a better proxy for the extensive margin of participation, as it is more sensitive to changes in the number of active market participants than aggregate volume, which can be sustained by fewer traders executing larger orders \citep{lo2000trading}.

I supplement the market-level data with two firm-level panels from the TWSE: (1) P/E ratios and dividend yields for approximately 900 listed stocks, available quarterly from 2012Q4; and (2) institutional investor (foreign and investment trust) net trading flows for approximately 975 stocks, available quarterly from 2012Q3. These panels provide cross-sectional evidence on the valuation and composition channels.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Three-Period Decomposition}

The identification strategy decomposes the market response into three distinct policy periods. I estimate:
\begin{equation}\label{eq:main}
Y_t = \alpha + \beta_A \cdot \text{Announce}_t + \beta_C \cdot \text{CGT}_t + \beta_R \cdot \text{PostRepeal}_t + \varepsilon_t
\end{equation}
where $Y_t$ is the log outcome in month $t$, $\text{Announce}_t = \mathbf{1}[t \in \text{2012Q2--Q4}]$ captures the legislative uncertainty period, $\text{CGT}_t = \mathbf{1}[t \in \text{2013Q1--2015Q3}]$ captures the implementation period, and $\text{PostRepeal}_t = \mathbf{1}[t \geq \text{2015Q4}]$ captures the post-repeal period. The omitted category is the pre-announcement baseline (2010Q1--2012Q1). Standard errors use the \citet{rambachan2023more} Newey-West HAC estimator with a six-month bandwidth to account for serial correlation.

The key tests are: (1) $\hat{\beta}_A < 0$ (announcement reduced activity); (2) $\hat{\beta}_C \approx 0$ (implementation had no additional effect beyond announcement); (3) the comparison of $\hat{\beta}_A$ across volume and transaction outcomes identifies the composition channel.

\subsection{Firm-Level Panel}

For valuation and composition outcomes, I exploit cross-sectional variation in the firm-level panels:
\begin{equation}\label{eq:panel}
Y_{it} = \alpha_i + \beta_C \cdot \text{CGT}_t + \beta_R \cdot \text{PostRepeal}_t + \varepsilon_{it}
\end{equation}
where $\alpha_i$ are firm fixed effects and standard errors are clustered at the firm level. These regressions identify within-firm changes in P/E ratios, dividend yields, and institutional trading flows across the CGT and post-repeal periods.

\subsection{Threats to Validity}

Four concerns merit discussion. First, \emph{secular trends}: TWSE volume trended upward from 2010 to 2018, potentially confounding the post-repeal comparison. I address this with linear and quadratic trend controls; the announcement effect remains robust across all specifications. Second, \emph{global shocks}: regional financial conditions evolved over this period, most notably China's 2015 stock market turbulence which coincides with the CGT repeal period. The announcement effect (2012Q2--Q4) is the least vulnerable to this concern, as it predates major regional disruptions. The post-repeal coefficients should be interpreted cautiously given the overlapping regional recovery. Third, \emph{pre-trends}: a placebo test imposing pseudo-treatment at 2011Q1 yields an insignificant coefficient ($\hat{\beta} = -0.078$, $p = 0.26$), confirming no differential pre-trend. Fourth, \emph{absence of a control group}: this is an interrupted time-series design without a parallel exchange unaffected by Taiwan's CGT. The three-period decomposition and transaction-count contrast provide within-market identification, but external validity is limited by the single-country design.


\section{Results}

\subsection{Main Results: Volume vs.~Transactions}

\input{tables/tab2_main_did}

Table~\ref{tab:main_did} reports the central finding. In the baseline specification (column 1), the announcement-period coefficient is $-0.279$ and highly significant ($p < 0.01$), corresponding to a 24\% decline in average daily share volume relative to the pre-announcement baseline. The CGT-active coefficient, by contrast, is $0.122$---positive and statistically insignificant. Volume during the three years the tax was in force was indistinguishable from the pre-announcement level. The post-repeal coefficient ($0.166$) is positive but also insignificant, consistent with a modest secular upward trend.

The volume story is one of resilience: the market absorbed the announcement shock within two to three quarters and returned to baseline by the time the tax was implemented. Adding a linear time trend (column 2) strengthens this interpretation: the announcement effect grows to $-0.439$ ($p < 0.01$), while the CGT coefficient flips to $-0.221$ but remains insignificant, suggesting the positive coefficient in column 1 partly reflected trend growth.

But volume is not the whole story. Using the number of daily transactions as the dependent variable reveals a persistent implementation effect: the announcement reduced transactions by 26\% ($\hat{\beta}_A = -0.307$, $p < 0.01$), and during the CGT-active period transactions remained 12\% below the pre-announcement level ($\hat{\beta}_C = -0.129$, $p < 0.01$). Only after repeal did transaction counts recover ($\hat{\beta}_R = -0.028$, insignificant). The implied average trade size (volume divided by transactions) thus increased during the CGT period, consistent with fewer but larger trades replacing the many small transactions characteristic of retail participation.

\subsubsection*{The Composition Channel} The divergence between volume recovery and transaction persistence has a clean interpretation. If aggregate share volume returned to baseline while the number of trades remained depressed, the average trade size must have increased. This is consistent with a compositional shift: retail investors---who generate high transaction frequency with small order sizes---exited the market, while institutional investors---who trade less frequently but in larger blocks---maintained or increased their activity. The CGT's incidence structure supports this mechanism: it targeted individual investors with high realized gains, precisely the population most sensitive to the tax.


\subsection{Valuation and Investor Composition}

\input{tables/tab3_valuation}

Table~\ref{tab:valuation} reports firm-level panel results for valuation and composition outcomes. Column (1) shows that log P/E ratios were elevated during the CGT period, consistent with lower effective discount rates as retail risk-tolerant investors exited and were replaced by institutional investors who may accept lower expected returns. Column (2) shows that dividend yields were marginally lower during the CGT period ($-1.30$ percentage points, $p = 0.09$) and significantly higher post-repeal ($+6.19$ pp, $p < 0.01$). The post-repeal dividend yield increase is consistent with the return of yield-demanding retail investors.

Columns (3) and (4) examine institutional investor flows. Foreign investor net purchases were directionally larger during the CGT period ($+265{,}631$ TWD per stock-quarter, $p = 0.12$) but statistically insignificant, reflecting high cross-sectional dispersion. These results are suggestive of institutional entry during the CGT period but do not reach conventional significance levels.


\subsection{Robustness}

\input{tables/tab4_robustness}

Table~\ref{tab:robustness} presents robustness checks. A quadratic time trend (column 2) attenuates the announcement effect due to multicollinearity between the trend terms and period indicators, but the qualitative pattern persists: the announcement coefficient is negative while the CGT coefficient is near zero. A placebo test (column 3) imposing pseudo-treatment at 2011Q1 using only pre-announcement data yields an insignificant coefficient, confirming the absence of pre-trends.

\subsubsection*{Symmetry Test} I formally test whether the volume pattern is symmetric---whether the post-repeal level equals the pre-announcement level. The sum $\hat{\beta}_C + \hat{\beta}_R = 0.287$ is positive and marginally significant ($p = 0.09$), suggesting a modest secular increase in market activity over the 2010--2018 period that is not fully captured by period indicators alone. This asymmetry is consistent with a trending market rather than a policy-driven permanent shift.


\section{Discussion}

The central finding---that the announcement, not the tax itself, drove the market disruption---connects to a growing literature on policy uncertainty as a real economic force \citep{baker2016measuring}. During the nine months between the initial proposal and the final legislative text, market participants faced genuine uncertainty about the tax's rate, threshold, and scope. The initial proposal included much lower thresholds and higher rates than the eventual legislation. This uncertainty resolved once the law was finalized, and volume recovered.

The transaction count finding adds nuance. While aggregate volume is resilient---institutional investors compensate for retail exit---the market's microstructure changed. Fewer participants making larger trades implies reduced price discovery from diverse opinion, a channel emphasized by \citet{delong1990noise}. Whether this composition shift affected informational efficiency or price accuracy is an important question that this paper's aggregate data cannot fully answer.

The Taiwan episode carries lessons for the current policy debate about financial transaction taxes in the US and EU \citep{habermeier2003securities}. Proponents of Tobin-style taxes argue they would curb speculative trading without affecting fundamental market quality \citep{tobin1978proposal, stiglitz1989using}. The Taiwan evidence suggests a more subtle mechanism: such taxes may not reduce aggregate volume but may narrow the investor base, concentrating trading among institutional participants and reducing retail market access. Whether this constitutes an efficiency gain (removing noise traders) or a loss (reducing diversity of opinion and market access) depends on one's model of the role of retail investors in price discovery.


\section{Conclusion}

Taiwan's three-year capital gains tax experiment provides a rare opportunity to decompose the market response to capital taxation into announcement and implementation effects. The data reveal that the widely cited volume decline was an announcement-period phenomenon---a transitory uncertainty shock that resolved once the tax's parameters were finalized. During the three years the tax was actually in force, aggregate share volume was indistinguishable from pre-announcement levels. But the number of daily transactions remained depressed throughout, revealing a composition shift from retail to institutional trading. The ``round-trip tax'' was, in practice, an ``announcement tax'': the market's behavioral response was driven by policy uncertainty, not by the tax itself.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Data Sources} All data are obtained from the Taiwan Stock Exchange Open API (\url{https://www.twse.com.tw}). The market-level aggregate data (FMTQIK endpoint) provides daily totals for trading volume (shares), value traded (TWD), and number of transactions for all TWSE-listed securities combined. Firm-level P/E ratios and dividend yields come from the BWIBBU\_d endpoint, and institutional investor flows from the T86 endpoint.

\subsubsection*{Sample Construction} The market-level panel spans January 2010 through December 2018 (104 months, 36 quarters). Daily trading-day observations are aggregated to monthly means (average daily volume and average daily transactions within each month). The firm-level P/E panel covers approximately 900 stocks observed quarterly from 2012Q4 through 2018Q4. The institutional flow panel covers approximately 975 stocks from 2012Q3 through 2018Q4.

\subsubsection*{Variable Definitions}
\begin{itemize}
\item \textbf{Log daily volume:} Natural log of average daily trading volume (shares) within each month.
\item \textbf{Log daily transactions:} Natural log of average daily number of transactions within each month.
\item \textbf{Announcement:} Indicator for 2012Q2--Q4 (April 26, 2012 cabinet proposal through December 2012).
\item \textbf{CGT Active:} Indicator for 2013Q1--2015Q3 (tax in effect).
\item \textbf{Post-Repeal:} Indicator for 2015Q4 onward (after November 17, 2015 repeal).
\item \textbf{P/E ratio:} Price-to-earnings ratio from TWSE, winsorized at the 99th percentile.
\item \textbf{Foreign net:} Net purchases by foreign investors (buy minus sell, in TWD).
\end{itemize}


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
