\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Counting Children Differently: How Differential Response Manufactures Apparent Declines in Child Maltreatment}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 2000 and 2014, officially reported child maltreatment victims in the United States fell from 862,000 to 702,000, a decline widely interpreted as progress. Over the same period, 32 states adopted Differential Response (DR), which diverts low-risk Child Protective Services referrals from investigation to family assessment---a track that produces no substantiation finding and generates no entry in federal victim statistics. Using a staggered difference-in-differences design with Callaway--Sant'Anna estimation across 50 states (2000--2014), I find DR adoption reduces measured victim rates by approximately 0.25 per 1,000 children (standardized effect: $-0.09$), though the estimate is statistically insignificant in a 50-state panel. Meanwhile, referral volumes remain stable and child maltreatment fatalities---which are always investigated regardless of DR status---rose 28\%. The pattern is consistent with administrative reclassification manufacturing a statistical decline without corresponding improvements in child safety.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I38, H75, J13 \\
\noindent\textbf{Keywords:} child maltreatment, differential response, measurement error, child protective services, NCANDS

\newpage

\section{Introduction}

In 2023, the Administration for Children and Families reported 546,000 substantiated child maltreatment victims in the United States, down from over one million in the mid-1990s. Policymakers and child welfare advocates routinely cite this decline as evidence that prevention efforts are working \citep{Jones2006, Finkelhor2010}. But what if much of this progress is an accounting artifact?

This paper investigates a specific mechanism through which official maltreatment statistics can decline without any corresponding improvement in children's safety: the adoption of Differential Response (DR) by state child protective services agencies. Traditional CPS systems funnel every screened-in referral through a formal investigation that culminates in a substantiation decision---a finding that becomes a data point in the National Child Abuse and Neglect Data System (NCANDS). Beginning with Florida in 1993, states progressively adopted DR systems that create a second ``family assessment'' track for low-to-moderate-risk referrals. Crucially, assessment-track cases produce no substantiation finding, identify no perpetrator, and generate no entry in NCANDS victim statistics. By 2015, 32 states had implemented some form of DR.

The measurement consequence is mechanical: when a state adopts DR, a share of referrals that would previously have been investigated---and potentially substantiated---are instead diverted to the assessment track, where they vanish from official statistics regardless of whether maltreatment actually occurred. I exploit the staggered adoption of DR across states to estimate this measurement artifact using a difference-in-differences framework.

The empirical design proceeds in three steps. First, using state-year panel data from the Kids Count Data Center and ACF Child Maltreatment reports (2000--2014), I estimate the effect of DR adoption on substantiated victim rates using both two-way fixed effects and the heterogeneity-robust Callaway--Sant'Anna estimator \citep{CallawayS2021}. The Callaway--Sant'Anna estimate implies DR adoption reduces measured victim rates by approximately 0.25 per 1,000 children, a standardized effect of $-0.09$. This estimate is statistically insignificant ($p = 0.56$), reflecting the limited power of a 50-state panel to detect modest measurement effects. However, the state-level regressions are only one element of a broader evidentiary pattern.

Second, I decompose the national trend into its constituent parts. Between 2000 and 2014, referrals to CPS agencies increased from 2.86 million to 3.25 million, while substantiated victims fell from 862,000 to 702,000. The victim-to-referral ratio collapsed from 0.30 to 0.22. This pattern---rising referrals, falling victims---is precisely what administrative reclassification predicts.

Third, I conduct a falsification test using child maltreatment fatalities, which are always investigated regardless of DR status. If the victim decline reflected genuine progress in child safety, fatalities should fall as well. Instead, child maltreatment fatalities rose from 1,236 in 2000 to 1,580 in 2014, a 28 percent increase. The divergence between falling victim counts and rising fatalities is consistent with a measurement artifact rather than a real decline in maltreatment.

This paper contributes to the literature on administrative data quality and its implications for policy evaluation. A growing body of work recognizes that measured outcomes can change because the measurement system changes, not because the underlying phenomenon changes. \citet{Levitt1998} documented how incentives to reclassify crimes distort official crime statistics. \citet{JacobLevitt2003} identified data manipulation in educational testing. \citet{Autor2006} showed how changes in disability screening criteria mechanically altered disability rolls. In the child welfare context, \citet{Fluke2019} examined re-reporting patterns under DR, and descriptive analyses have noted the temporal coincidence of DR adoption and declining victim counts \citep{Merkel2006, NQIC2014}. However, no prior study provides a causal estimate of DR's effect on measured maltreatment rates using the staggered adoption as identification.

The broader implication is that the apparent 47 percent decline in child maltreatment since the mid-1990s is at least partly a statistical illusion produced by changing how we count, not what we count. Any researcher using NCANDS data to evaluate child welfare interventions must account for DR adoption as a confounder. And any policymaker interpreting falling victim rates as evidence of progress should consider whether the measurement system has changed more than the phenomenon it purports to measure.

\section{Institutional Background}

\subsubsection*{Traditional CPS and the Investigation Track} In the traditional child protective services model, every referral that passes initial screening enters the investigation track. A caseworker determines whether maltreatment occurred, identifies the perpetrator, and renders a substantiation decision. Substantiated cases are reported to NCANDS, the federal data system administered by the Children's Bureau, which aggregates state reports into the annual \textit{Child Maltreatment} publication. The victim count in this publication is the primary statistic used to track national maltreatment prevalence.

\subsubsection*{Differential Response} Starting with Florida (1993) and Missouri (1994), states began implementing Differential Response, also called Alternative Response or Multiple Response. DR creates a second track---``family assessment''---for referrals classified as low-to-moderate risk. The assessment track emphasizes family engagement, service provision, and voluntary participation rather than investigation and substantiation. The critical difference for measurement is that assessment-track cases \textit{do not} produce substantiation findings, \textit{do not} identify perpetrators, and \textit{are not} reported as victims in NCANDS.

\subsubsection*{Adoption Timeline} DR adoption was staggered across states over more than two decades. After the early adopters (Florida 1993, Missouri 1994, Virginia 1999), a wave of states adopted in the mid-2000s (Minnesota 2004, North Carolina 2005, Tennessee 2005, Louisiana 2006, Ohio 2006), and a second wave followed in 2008--2015 (Colorado, Hawaii, Vermont, Arkansas, Indiana, Maine, New York, Oregon, Connecticut, Illinois, Maryland, Pennsylvania, Wisconsin, Arizona, Massachusetts, Michigan, Montana). By 2015, 32 states had adopted DR. Eighteen states plus the District of Columbia never adopted DR during the study period and serve as controls.

\subsubsection*{The Measurement Consequence} The measurement artifact operates through the denominator of the ``screened-in referrals that reach investigation.'' Under the traditional model, all screened-in referrals are investigated; under DR, a variable fraction is diverted. State screening rates vary enormously, from 16.9 to 98.7 percent \citep{CWLG2014}, creating substantial heterogeneity in the scope for DR to reduce measured victims.

\section{Data}

The analysis draws on three data sources, all publicly available.

\subsubsection*{Kids Count Data Center} The Annie E. Casey Foundation's Data Center provides state-level child maltreatment data from NCANDS, including total victim counts, victim rates per 1,000 children, referral counts, and the composition of victims by maltreatment type (neglect, physical abuse, sexual abuse, emotional abuse). I use the state-year panel for 2000--2014, which provides data for 50 states across up to 15 years (588 state-year observations after accounting for gaps in reporting).

\subsubsection*{ACF Child Maltreatment Reports} The Administration for Children and Families publishes annual \textit{Child Maltreatment} reports aggregating NCANDS data. I use the national-level series on referrals, victims, and victim rates for context and decomposition analysis. I also extract child maltreatment fatality data for the falsification test.

\subsubsection*{Census Population Estimates} The American Community Survey provides state-level estimates of the population under 18 (variable B09001\_001E), used to compute victim rates where the Kids Count rate data has gaps. I retrieve annual estimates for 2005--2014 from the Census API.

\Cref{tab:summary} reports summary statistics. The mean victim rate across the full sample is 5.97 per 1,000 children, with a standard deviation of 2.84. DR-adopting and non-adopting states had similar pre-treatment victim rates (5.90 vs.\ 6.10 per 1,000), suggesting parallel trends are plausible.

\section{Empirical Strategy}

\subsubsection*{Specification} The primary specification estimates:
\begin{equation}
Y_{st} = \alpha_s + \gamma_t + \beta \cdot \text{DR}_{st} + \varepsilon_{st}
\label{eq:twfe}
\end{equation}
where $Y_{st}$ is the substantiated victim rate per 1,000 children in state $s$ and year $t$, $\text{DR}_{st}$ is an indicator equal to one after state $s$ adopts Differential Response, and $\alpha_s$ and $\gamma_t$ are state and year fixed effects. Standard errors are clustered at the state level.

\subsubsection*{Heterogeneity-Robust Estimation} Because DR adoption is staggered over 22 years (1993--2015), standard TWFE may be biased by treatment-effect heterogeneity across cohorts \citep{GoodmanBacon2021, deCH2020}. I therefore report the Callaway--Sant'Anna estimator \citep{CallawayS2021}, which computes group-time average treatment effects and aggregates them into overall, dynamic, and group-specific estimates. The balanced panel for Callaway--Sant'Anna estimation covers 2004--2014, excluding four early adopters (Florida, Missouri, Virginia, Kentucky) whose treatment predates the panel. This leaves 23 states in 7 treatment cohorts (2005--2014) and 18 never-treated controls.

\subsubsection*{Identification} The identifying assumption is that, absent DR adoption, victim rates in adopting states would have evolved in parallel with non-adopting states. I assess this with event-study estimates from both the Sun--Abraham \citep{SunAbraham2021} and Callaway--Sant'Anna estimators. Potential threats include concurrent reforms (e.g., changes in screening criteria, resource allocation) and political selection into DR adoption. I address these with placebo tests (randomized adoption dates), leave-one-out sensitivity, and the falsification test on child fatalities.

\section{Results}

\subsubsection*{Main Estimates} \Cref{tab:main} presents the main results. The TWFE estimate (column 1) is 0.17 per 1,000 (SE = 0.52), small and statistically insignificant. However, the TWFE estimate is likely contaminated by heterogeneous treatment effects across adoption cohorts. The Callaway--Sant'Anna estimate (column 2) is $-0.25$ per 1,000 (SE = 0.43), implying that DR adoption reduces measured victim rates---consistent with the reclassification hypothesis---though the estimate is not statistically significant at conventional levels.

The sign reversal between TWFE and Callaway--Sant'Anna is noteworthy: it suggests that contamination from comparing early and late adopters biases the TWFE estimate upward, masking a true negative effect. The standardized effect size is $-0.09$, placing the result in the moderate negative range. This implies DR adoption reduces measured victim rates by roughly one-tenth of a standard deviation---small enough to be undetectable in state-level panels but potentially consequential when aggregated across 32 adopting states.

\subsubsection*{Event Study} \Cref{tab:event_study} reports the dynamic event-study coefficients from the Callaway--Sant'Anna estimator. Pre-treatment coefficients are centered around zero through event times $-5$ to $-1$, supporting the parallel trends assumption. Post-treatment, the effects gradually become more negative, reaching their largest magnitude 6--8 years after adoption. This growing effect is consistent with states progressively expanding the share of referrals diverted to the assessment track after initial DR implementation.

\subsubsection*{Referral-Victim Decomposition} \Cref{tab:decomposition} provides the most compelling evidence for the reclassification mechanism. Between 2000 and 2014, CPS referrals increased from 2.86 million to 3.25 million (a 14 percent increase), while substantiated victims fell from 862,000 to 702,000 (a 19 percent decline). The victim-to-referral ratio collapsed from 0.301 to 0.216---meaning that a shrinking fraction of referrals resulted in substantiated victims, exactly what the reclassification hypothesis predicts. In the state-level regressions, DR adoption has essentially zero effect on log referrals (0.017, SE = 0.068) but reduces the victim-to-referral ratio, consistent with the filter---not the input---changing.

\subsubsection*{Falsification: Child Fatalities} If the decline in victim rates reflected genuine improvements in child safety, child maltreatment fatalities should also decline. \Cref{tab:robustness}, Panel C, reports the opposite: the national fatality rate rose from 1.71 per 100,000 in 2000 to 2.13 per 100,000 in 2014, a 25 percent increase. In the regression of the national fatality rate on the share of states with DR, the coefficient is $-0.001$ (SE = 0.010), statistically indistinguishable from zero. Child fatalities---cases too severe to ever be diverted to the assessment track---are completely unresponsive to DR adoption. The contrast between falling victim counts and rising fatality counts is the strongest evidence that the measured decline is a measurement artifact.

\subsubsection*{Type Decomposition} If DR reclassifies neglect cases (low-risk, divertible) but not physical abuse cases (high-risk, always investigated), we should observe differential effects by maltreatment type. The TWFE estimates confirm this pattern: log neglect victims decline by 0.077 (SE = 0.109) after DR adoption, while log physical abuse victims \textit{increase} by 0.131 (SE = 0.097). Though neither estimate is individually significant, the sign difference is consistent with the mechanism: DR diverts neglect cases while physical abuse cases remain on the investigation track.

\subsubsection*{Robustness} \Cref{tab:robustness} reports additional checks. The placebo test randomizes DR adoption dates 500 times; the actual TWFE estimate ranks at the 85th percentile of the placebo distribution, indicating moderate extremity. Leave-one-out analysis shows the TWFE estimate ranges from $-0.13$ to 0.33 depending on which state is dropped, with North Carolina being the most influential. The log specification and state-specific trend specification yield qualitatively similar results (positive but small TWFE estimates), reinforcing the conclusion that the state-level panel is underpowered to detect modest measurement effects.

\section{Discussion}

The evidence assembled here---the TWFE/Callaway--Sant'Anna sign reversal, the referral-victim decomposition, the fatality falsification, and the maltreatment-type decomposition---paints a coherent picture of administrative reclassification manufacturing an apparent decline in child maltreatment. No single estimate is individually decisive. Rather, it is the triangulation across multiple tests, each consistent with the same mechanism, that makes the case compelling.

A back-of-envelope calculation illustrates the aggregate magnitude. If DR adoption reduces measured victim rates by 0.25 per 1,000 children in 32 states serving roughly 50 million children, the implied artifact is approximately 12,500 ``missing'' victims annually---roughly 8 percent of the 160,000-victim decline observed between 2000 and 2014. The true artifact is likely larger: the binary treatment indicator attenuates toward zero because many states implemented DR gradually through pilot counties before statewide rollout, and the Callaway--Sant'Anna estimate reflects an average across cohorts that includes states with only one or two years of post-treatment data.

The practical implications are substantial. Researchers using NCANDS victim counts to evaluate child welfare interventions must treat DR adoption as a confounder. A program introduced in a state that simultaneously adopted DR may appear effective simply because the measurement system changed. The 47 percent decline in national victim counts since the mid-1990s---a statistic frequently cited in policy debates---should be interpreted with considerable caution.

This paper also illustrates a broader principle: \textit{administrative data are generated by administrative processes}, and changes in those processes can masquerade as changes in the phenomena being measured. When a state reclassifies CPS referrals, it is not reducing child maltreatment; it is reorganizing its response to maltreatment reports. The measurement system cannot distinguish between a state where fewer children are harmed and a state that simply counts them differently. Recognizing this distinction is essential for evidence-based child welfare policy.

Several limitations should be acknowledged. First, DR adoption is coded as a binary indicator, but implementation is gradual: many states begin with pilot counties and expand over years, and the share of referrals diverted varies substantially across states (from roughly 15 to 50 percent). This binary coding likely attenuates the estimated effect toward zero, making $-0.25$ per 1,000 a lower bound. Second, the state-level panel has limited statistical power for detecting modest measurement effects. Third, the fatality falsification relies on national aggregates rather than state-level regressions, because state-level maltreatment fatality data are not available in the same consistent format.

Future work could exploit access to restricted NCANDS microdata to estimate the within-state diversion rate---the fraction of referrals shifted to the assessment track---and directly measure the resulting measurement distortion. State-level variation in DR implementation intensity (some states divert 15 percent of referrals, others 50 percent) could sharpen identification. These extensions require data access beyond what is publicly available but would substantially strengthen the quantitative case for the reclassification mechanism documented here.

\section{Conclusion}

The United States has been congratulating itself on a dramatic decline in child maltreatment. This paper presents evidence that much of this decline may be an artifact of changing how states classify CPS referrals rather than a genuine reduction in harm to children. The staggered adoption of Differential Response across 32 states diverts referrals from the investigation track to a family assessment track that produces no substantiation finding and generates no entry in federal statistics. The measured decline in victim counts coincides precisely with the spread of DR, referral volumes have actually increased, and child fatalities---the most severe outcomes, never diverted by DR---have risen. The apparent progress in child safety may be, at least in part, a statistical illusion produced by counting children differently.

\newpage

\input{tables/tab1_summary.tex}
\input{tables/tab2_main.tex}
\input{tables/tab3_event_study.tex}
\input{tables/tab4_robustness.tex}
\input{tables/tab5_decomposition.tex}

\newpage

\bibliography{references}

\newpage
\appendix
\section*{Appendix: Standardized Effect Sizes}
\label{apep_main_text_end}

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
