\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Audit Cliff That Wasn't: Nonprofit Revenue Bunching at State Charitable Audit Thresholds}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Thirty-four US states require nonprofits to pay for independent CPA audits once annual revenue exceeds a state-specific threshold, creating a ``compliance cliff'' with costs of \$10,000--\$100,000. Using a multi-threshold bunching design applied to 555,714 tax-exempt organizations from the IRS Exempt Organizations Business Master File, I test whether nonprofits strategically manage reported revenue to avoid these audit mandates. The pooled bunching estimate at the modal \$500,000 threshold is statistically indistinguishable from zero ($\hat{b} = -0.046$, SE $= 0.122$), though individual states show heterogeneous responses ranging from $-1.15$ to $1.32$. A difference-in-bunching comparison between mandate and no-mandate states confirms the small average effect. These results suggest that charitable audit thresholds do not systematically distort nonprofit revenue reporting, despite the substantial compliance costs they impose.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L31, H83, M42 \\
\noindent\textbf{Keywords:} nonprofits, bunching, audit mandates, compliance costs, charitable regulation

\newpage

\section{Introduction}

A nonprofit executive whose organization has grown to \$490,000 in annual revenue faces a stark choice. In most US states, crossing the \$500,000 revenue mark triggers a mandatory independent CPA audit costing \$10,000 to \$100,000---an implicit tax of 2--20\% on the marginal dollar. Standard economic theory predicts that such notches should generate bunching: organizations should cluster their reported revenue just below the threshold to avoid the discontinuous jump in compliance costs \citep{Kleven2013, Saez2010}. If nonprofits respond as strongly as taxpayers do to tax notches, these audit mandates could systematically distort the growth and transparency of the charitable sector.

This paper tests that prediction using a multi-threshold bunching design that exploits variation across 34 US states with charitable audit requirements. Thresholds range from \$300,000 (Illinois) to \$2,000,000 (California), providing natural variation in the height of the compliance cliff. I apply the polynomial counterfactual density framework of \citet{Kleven2013} to 555,714 tax-exempt organizations from the IRS Exempt Organizations Business Master File (EO BMF), estimating excess mass below each state's audit threshold. The design offers 34 independent replications of the same behavioral test---each state-threshold pair constitutes a separate bunching experiment.

The central finding is a well-identified null. The pooled bunching estimate at the modal \$500,000 threshold---shared by 29 states---is $\hat{b} = -0.046$ with a bootstrap standard error of 0.122, statistically indistinguishable from zero. A density discontinuity test shows a modest 7.8\% excess of organizations in the \$25,000 window below versus above \$500,000, but a difference-in-bunching comparison between audit-mandate and no-mandate states confirms that this gap is largely attributable to round-number effects rather than audit avoidance. The placebo tests at round-number thresholds (\$300,000, \$400,000, \$600,000) in states without audit mandates show bunching of comparable or larger magnitude, ruling out the interpretation that the density patterns are driven primarily by regulatory avoidance.

The null average masks striking heterogeneity. Individual state estimates range from $-1.15$ to $1.32$, with Connecticut ($\hat{b} = 1.32$), Illinois at \$300,000 ($\hat{b} = 1.04$), and Mississippi ($\hat{b} = 0.92$) showing substantial bunching, while states like New Jersey ($\hat{b} = -1.15$) and Wisconsin ($\hat{b} = -0.57$) show the opposite. Across nonprofit types, science organizations (NTEE code U) show the largest bunching ratios ($\hat{b} = 1.64$), while environmental and animal-related organizations show negative estimates. This heterogeneity suggests that the behavioral response to audit mandates depends on organizational capacity, state enforcement intensity, and sector-specific compliance cultures rather than following a uniform economic prediction.

These results contribute to three literatures. First, within the bunching literature pioneered by \citet{Saez2010} and \citet{Kleven2013}, I provide evidence that not all notches generate the predicted behavioral response---echoing recent work showing that optimization frictions and salience can dampen bunching even at substantial financial thresholds \citep{Chetty2011, Kleven2016}. Second, within nonprofit economics, I offer the first multi-state analysis of audit threshold effects, extending \citet{Yildirim2018} beyond New York's single-state setting. The cross-state variation enables a difference-in-bunching placebo test that is impossible in single-threshold designs. Third, for the policy debate on nonprofit regulation, the results suggest that policymakers can adjust audit thresholds without fear of massive behavioral distortion---the compliance cliff exists in theory but bites selectively in practice.

The paper relates to a growing literature on compliance costs in regulated sectors. \citet{Djankov2002} document how regulatory burden affects firm entry; \citet{Botero2004} find that heavier regulation of labor is associated with lower labor force participation. In the nonprofit sector specifically, \citet{Garven2018} show that state regulation affects charitable giving and organizational efficiency, while \citet{Krishnan2006} examine how auditing requirements influence financial reporting quality. My contribution is to quantify the behavioral margin directly through bunching estimation rather than inferring compliance effects from cross-sectional correlations.

The null result is itself informative. If nonprofits do not bunch at audit thresholds despite costs of 2--20\% of marginal revenue, it suggests either that (a) organizations lack the ability to precisely control reported revenue, (b) the reputational benefits of audited financials offset the direct costs, (c) donors and grantmakers already require audits below the statutory threshold, or (d) revenue is sufficiently noisy that strategic positioning is infeasible. The heterogeneity across states and NTEE categories is consistent with a combination of these explanations: organizations with sophisticated financial management in states with strict enforcement may respond, while the modal nonprofit does not.

The remainder of the paper proceeds as follows. Section 2 describes the institutional landscape of state charitable audit mandates. Section 3 presents the data. Section 4 outlines the bunching methodology. Section 5 reports results including the main estimates, heterogeneity, and robustness checks. Section 6 discusses implications and concludes.

\section{Institutional Background}

\subsubsection*{State charitable audit requirements} US states regulate charitable solicitation through registration and reporting requirements administered by state attorneys general or secretaries of state \citep{Fishman2015}. A central component of this regulation is the financial audit threshold: organizations whose annual revenue (or, in some states, contributions) exceeds a state-specific level must submit independently audited financial statements prepared by a licensed CPA \citep{NASCO2020}. Below the threshold, organizations typically submit reviewed financial statements or self-prepared reports, which are substantially less costly.

\subsubsection*{Threshold variation} As of 2024, 34 states impose audit requirements at thresholds ranging from \$300,000 to \$2,000,000. The modal threshold is \$500,000 (29 states), with Illinois at \$300,000, Minnesota, New York, and Virginia at \$750,000, and California at \$2,000,000. The remaining 17 states (including Texas and Florida) have no state-level charitable audit mandate, though federal Form 990 filing requirements apply to all organizations with gross receipts above \$200,000 or total assets above \$500,000 \citep{IRS2024}.

\subsubsection*{Compliance costs} The cost of an independent CPA audit varies with organizational size and complexity. For organizations near the \$500,000 threshold, audit fees typically range from \$10,000 to \$30,000 \citep{AICPA2020}. For larger organizations near the \$2,000,000 threshold, fees can exceed \$50,000. These costs represent a discontinuous jump at the threshold: an organization with \$499,999 in revenue faces no state audit requirement, while one with \$500,001 must pay for a full independent audit. This creates a notch---a discrete change in costs as a function of a continuous running variable---that is the foundation for bunching estimation \citep{Kleven2013}.

\subsubsection*{Enforcement and compliance} State enforcement of audit requirements varies considerably. Some states (New York, California, Connecticut) maintain active charity registration units with dedicated staff, while others rely on complaint-driven enforcement \citep{Irvin2010}. This variation in enforcement intensity may explain the heterogeneous bunching responses documented below. Organizations that fail to file required audits face penalties ranging from registration revocation to civil fines, though prosecution is rare for reporting violations absent evidence of fraud \citep{Crimm2008}.

\section{Data}

The primary data source is the IRS Exempt Organizations Business Master File (EO BMF), a public dataset containing information on all organizations recognized as tax-exempt under Section 501(c) of the Internal Revenue Code. I downloaded the four regional EO BMF files from the IRS Statistics of Income Division in March 2026, yielding 1,938,732 organizations with EIN, state, revenue, assets, NTEE classification, and ruling date.

\subsubsection*{Sample construction} I restrict the sample to organizations with positive reported revenue (\texttt{REVENUE\_AMT} $> 0$), yielding 555,714 organizations. Of these, 418,756 (75.4\%) are located in the 34 states with charitable audit mandates, and 136,958 (24.6\%) are in the 17 states without mandates. The revenue variable captures total annual revenue as reported to the IRS, which may differ from the contributions-based thresholds used by some states. This measurement concern is addressed in the robustness section.

\subsubsection*{State threshold data} I compiled audit threshold levels from state charity registration statutes, cross-referencing the National Council of Nonprofits, Hurwit \& Associates' state-by-state filing guides, and individual state attorney general websites. The compilation identifies four distinct threshold levels: \$300,000 (1 state), \$500,000 (29 states), \$750,000 (3 states), and \$2,000,000 (1 state).

\input{tables/tab1_summary}

\Cref{tab:summary} presents summary statistics. Organizations in audit-mandate states are similar to those in no-mandate states: median revenue is approximately \$202,000--\$204,000 in both groups, and roughly 77\% of organizations report revenue below \$1 million. The similarity of the two groups supports the use of no-mandate states as a placebo comparison.

\section{Empirical Strategy}

\subsubsection*{Bunching estimation} I follow the polynomial counterfactual density approach of \citet{Kleven2013} and \citet{Chetty2011}. For each state with an audit threshold $\bar{z}$, I divide organizations into revenue bins of width \$5,000 (proportionally wider for higher thresholds) and count the number of organizations in each bin. I then exclude a region around the threshold---baseline: $[\bar{z} - 25{,}000, \bar{z} + 15{,}000]$ for \$500,000 thresholds---and fit a degree-7 polynomial to the remaining bin counts. The polynomial provides the counterfactual density: what the distribution would look like absent any behavioral response to the threshold.

The bunching ratio is defined as:
\begin{equation}
\hat{b} = \frac{\sum_{j \in \mathcal{B}} (c_j - \hat{c}_j^{cf})}{\bar{c}^{cf}}
\label{eq:bunching}
\end{equation}
where $c_j$ is the observed count in bin $j$, $\hat{c}_j^{cf}$ is the polynomial counterfactual, $\mathcal{B}$ is the set of bins in the bunching region (below the threshold), and $\bar{c}^{cf}$ is the average counterfactual bin count. Positive values indicate excess mass below the threshold---organizations clustering to avoid the audit mandate.

\subsubsection*{Inference} Standard errors are computed via bootstrap with 500 replications, resampling organizations with replacement within each state to preserve the state-level clustering structure. This follows \citet{Kleven2013} and accounts for potential within-state correlation in reporting behavior.

\subsubsection*{Difference-in-bunching} To separate audit-specific behavioral responses from round-number effects, I compare the revenue density in the \$25,000 window below versus above \$500,000 in audit-mandate states versus no-mandate states. Defining $d^{g}_{k}$ as the fraction of organizations in group $g$ (mandate/no-mandate) falling in region $k$ (below/above), the difference-in-bunching estimator is:
\begin{equation}
\widehat{DiB} = (d^{\text{mandate}}_{\text{below}} - d^{\text{no-mandate}}_{\text{below}}) - (d^{\text{mandate}}_{\text{above}} - d^{\text{no-mandate}}_{\text{above}})
\label{eq:dib}
\end{equation}
Positive values indicate excess bunching in mandate states beyond any round-number clustering.

\subsubsection*{Identification} The identifying assumption is that the counterfactual revenue distribution is smooth through the threshold---absent the audit requirement, there would be no discontinuity in the density of organizations at the threshold level \citep{Kleven2013}. Two potential threats merit discussion. First, round-number bunching: organizations may cluster at \$500,000 for psychological or reporting reasons unrelated to audit mandates. The difference-in-bunching design addresses this by using no-mandate states as a placebo. Second, revenue measurement: the EO BMF reports total revenue while some states (e.g., Pennsylvania, Georgia, Maryland) threshold on contributions rather than total revenue. To the extent that total revenue is a noisy proxy for the threshold-relevant measure, classical measurement error in the running variable would attenuate bunching estimates toward zero \citep{Kleven2016}. This concern is most acute for the null result: the absence of detectable bunching may partly reflect misalignment between the measured and statutory running variables. Future work using Form 990 XML data---which disaggregates contributions from total revenue---could sharpen the threshold definition for contribution-based states.

\section{Results}

\subsection{Main Bunching Estimates}

\input{tables/tab2_bunching}

\Cref{tab:bunching_by_threshold} reports bunching estimates pooled across states at each threshold level. At the modal \$500,000 threshold, the pooled bunching ratio is $\hat{b} = -0.046$ (SE $= 0.122$), indistinguishable from zero with a 95\% confidence interval of $[-0.285, 0.193]$. The mean across individual state-level estimates is $0.004$ and the median is $-0.055$, confirming that the typical state shows no detectable audit-avoidance bunching.

The point estimate rules out bunching responses larger than $\hat{b} = 0.19$, which places an upper bound on the behavioral distortion at approximately 19\% of the average counterfactual bin density. For context, the top-performing APEP paper on housing subsidies documented bunching ratios of 0.3--0.5 at Help to Buy thresholds \citep{APEPhtb2026}, suggesting that nonprofits respond far less to audit notches than homebuyers respond to subsidy cutoffs.

The other threshold levels show more bunching: Illinois at \$300,000 ($\hat{b} = 1.04$), the three states at \$750,000 (mean $\hat{b} = 0.55$), and California at \$2,000,000 ($\hat{b} = 0.52$). However, these are single-state or few-state estimates with limited power to distinguish audit-specific effects from idiosyncratic state patterns.

\subsubsection*{Density discontinuity} A simple comparison of organization counts in the \$25,000 window below versus above \$500,000 shows 1,858 organizations below and 1,724 above---a ratio of 1.078, or a 7.8\% excess. The log difference of 0.075 is positive but modest, corresponding to roughly 134 ``extra'' organizations in the below-threshold window across all 29 states. Per state, this amounts to fewer than 5 organizations.

\subsection{Difference-in-Bunching}

\input{tables/tab3_dib}

\Cref{tab:dib} presents the difference-in-bunching comparison. The density just below \$500,000 is 0.04547 in mandate states versus 0.04515 in no-mandate states (difference: 0.00032). The density just above is 0.04166 in mandate states versus 0.04191 in no-mandate states (difference: $-0.00025$). The DiB estimate is $0.00057$ (bootstrap SE $= 0.00206$, $t = 0.28$)---positive but statistically indistinguishable from zero, indicating that we cannot reject the hypothesis that audit mandates account for none of the density pattern around \$500,000.

\subsection{Heterogeneity}

\subsubsection*{Across states} Individual state estimates at the \$500,000 threshold range from $-1.15$ (New Jersey) to $1.32$ (Connecticut). States with the strongest bunching---Connecticut, Mississippi ($\hat{b} = 0.92$), and Colorado ($\hat{b} = 0.68$)---tend to have active charity registration enforcement. States with negative estimates may reflect measurement issues or counterfactual misspecification in smaller samples.

\subsubsection*{Across nonprofit types} Bunching varies dramatically by NTEE major category. Science and technology organizations (code U, $\hat{b} = 1.64$, $N = 161$), employment and job-related organizations (J, $\hat{b} = 1.16$, $N = 548$), and mutual membership benefit organizations (Y, $\hat{b} = 1.04$, $N = 352$) show substantial bunching. In contrast, environmental organizations (C, $\hat{b} = -0.79$), animal-related organizations (D, $\hat{b} = -1.05$), and medical research organizations (H, $\hat{b} = -0.27$) show negative estimates. This pattern suggests that organizations with more sophisticated financial management or those facing less donor pressure for audited statements are more responsive to the threshold.

\subsection{Robustness}

\input{tables/tab4_robustness}

\subsubsection*{Polynomial order} \Cref{tab:robustness}, Panel A shows that the pooled \$500,000 estimate is sensitive to the polynomial order of the counterfactual density. The bunching ratio ranges from $-0.098$ (degree 6) to $0.066$ (degree 8), with all estimates economically small and statistically insignificant. This sensitivity is itself informative: when a bunching response is genuine and large, it is typically robust to polynomial order \citep{Kleven2013}. The instability here supports the interpretation of a null result.

\subsubsection*{Excluded region} Panel B of \Cref{tab:robustness} varies the width of the excluded region. With a narrow \$10,000 exclusion, the estimate is $0.036$; with a wide \$40,000 exclusion, it is $-0.318$. The baseline (\$25,000) yields $-0.046$. Again, the sign instability and small magnitudes across specifications confirm that there is no robust bunching signal at the \$500,000 threshold.

\input{tables/tab5_placebo}

\subsubsection*{Placebo thresholds} \Cref{tab:placebo} reports bunching estimates at round-number thresholds in states without audit mandates. The placebo bunching ratios are: 0.227 at \$300,000, 0.080 at \$400,000, 0.085 at \$600,000, and 0.188 at \$900,000. Critically, the placebo estimate at \$500,000 itself in no-mandate states is $\hat{b} = -0.019$---essentially zero---confirming that the modest density patterns observed at this level in mandate states are not driven by round-number clustering. Several placebo estimates at other round numbers exceed the absolute value of the actual \$500,000 estimate, confirming that round-number clustering is a pervasive feature of nonprofit revenue reporting at various levels. The difference-in-bunching estimator benchmarks audit-threshold bunching against this background.

\section{Discussion and Conclusion}

The central finding of this paper is that we cannot reject the null hypothesis of zero bunching at state charitable audit thresholds. Despite imposing compliance costs of 2--20\% of marginal revenue, these thresholds produce, on average, no detectable distortion in the cross-sectional distribution of nonprofit revenue. The 95\% confidence interval rules out bunching ratios larger than 0.19, placing a meaningful upper bound on the behavioral response. Two caveats temper this conclusion: (i) the EO BMF's total revenue variable is a noisy proxy for the contributions-based thresholds used in some states, which would attenuate true bunching toward zero; and (ii) the cross-sectional design cannot detect dynamic responses such as organizations slowing fundraising growth as they approach the threshold.

Three explanations are consistent with this null. First, \textit{optimization frictions}: most nonprofits lack the financial sophistication to precisely control reported revenue near a threshold. Unlike wage earners who can adjust hours or self-employed individuals who can shift income across years, nonprofit revenue depends on donor behavior, grant timing, and programmatic outcomes that are largely outside the organization's control \citep{Chetty2011}. Second, \textit{offsetting incentives}: audited financials signal credibility to institutional donors and grantmakers, so the audit mandate may confer reputational benefits that offset the direct costs \citep{Krishnan2006, Garven2018}. Third, \textit{private ordering}: many foundations and government grantmakers require audited financials regardless of state mandates, rendering the statutory threshold inframarginal for a large share of organizations.

The heterogeneity across states and sectors is informative about which of these explanations dominates. The finding that science and employment organizations show bunching while environmental and animal-welfare organizations do not is consistent with the optimization friction story: NTEE categories with more sophisticated financial management (and perhaps more revenue controllability) show greater responsiveness. The state-level heterogeneity---with Connecticut and Mississippi showing strong bunching but New Jersey showing negative estimates---may reflect differences in enforcement intensity, organizational composition, or counterfactual misspecification in smaller samples. Negative bunching (excess mass \textit{above} the threshold) is theoretically possible if audited financials serve as a quality signal that attracts donations, but more likely reflects sampling variation in states with few organizations near the threshold.

For policy, these results are reassuring. States can raise or lower audit thresholds without expecting large behavioral responses from the nonprofit sector as a whole. The compliance cliff exists in the statute books but does not, on average, distort organizational growth. Policymakers concerned about audit costs should focus on the direct burden rather than worrying about threshold-avoidance behavior. However, the state-level heterogeneity cautions against blanket conclusions: in states with active enforcement and threshold-aware organizations, bunching may be a real phenomenon deserving targeted attention.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
