\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{When Green Courts Create Brown Politics: The Dutch Nitrogen Ruling and the Rise of Agrarian Populism}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In May 2019, a Dutch court invalidated the national nitrogen permit system, freezing 18,000 construction permits and threatening thousands of livestock farms near Natura 2000 sites. Within four years, the BoerBurgerBeweging (BBB)---a party that did not exist before the ruling---won the largest vote share in all twelve provinces. I test whether this political realignment was driven by the ruling's economic disruption or by pre-existing agricultural identity. Using spatial variation in Natura 2000 exposure across 328 Dutch municipalities, I find no differential effect of the ruling on building permits. However, municipalities' pre-ruling agricultural employment share explains 40 percent of the cross-sectional variation in BBB vote share, even after controlling for population density and Natura 2000 proximity. The nitrogen crisis mobilized existing agricultural grievance rather than creating new economic losers.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} D72, Q58, R52 \\
\noindent\textbf{Keywords:} environmental regulation, political backlash, populism, nitrogen policy, Natura 2000

\newpage

\section{Introduction}

On May 29, 2019, the Dutch Council of State struck down the Programma Aanpak Stikstof (PAS)---the government's blanket nitrogen deposition permit system---ruling it incompatible with the EU Habitats Directive. Overnight, approximately 18,000 construction permits were frozen and 3,000 livestock farms faced uncertain futures \citep{vangrinsven2019}. The ``nitrogen crisis'' paralyzed Dutch infrastructure and agriculture, dominating public discourse for years. Then something remarkable happened: a party founded six months after the ruling, the BoerBurgerBeweging (BBB, ``Farmer-Citizen Movement''), swept the March 2023 Provincial Elections, becoming the largest party in all twelve provinces and winning 16 Senate seats \citep{delange2023}.

This paper asks a simple question with profound implications for the political economy of environmental regulation: did the nitrogen ruling \textit{cause} BBB's rise, or did BBB merely capture pre-existing agricultural grievances that the ruling made salient? The distinction matters. If environmental regulation directly creates political backlash through economic disruption---frozen permits, lost construction jobs, threatened farms---then the policy implication is to compensate losers. If instead the ruling activated a latent rural identity that was already diverging from urban political preferences, then compensation alone will not prevent backlash against future environmental policy.

I exploit spatial variation in municipalities' exposure to the nitrogen ruling, constructed from the intersection of Natura 2000 protected area boundaries with pre-existing agricultural and construction employment shares. Municipalities with more Natura 2000 land and higher agricultural dependence were mechanically more exposed to the ruling's permit freeze and farm restrictions. I use this variation in two complementary analyses.

First, I estimate a municipality-quarter panel difference-in-differences model for building permits, using 30 pre-treatment quarters (2012Q1--2019Q2) and 26 post-treatment quarters. If the ruling caused differential economic disruption, municipalities with higher nitrogen exposure should show a sharper decline in building permits after May 2019. I find no such effect: the coefficient on exposure interacted with the post-ruling indicator is small, positive, and statistically indistinguishable from zero ($ \hat{\beta} = 0.62$, $\text{SE} = 1.23$). Building permits declined nationally during this period, but the decline was not concentrated in high-exposure municipalities.

Second, I estimate cross-sectional regressions of BBB vote share on nitrogen exposure and agricultural characteristics. The key finding is a decomposition: in a bivariate regression, the composite exposure index significantly predicts BBB vote share ($ \hat{\beta} = 0.24$, $ p < 0.05$). But once I control for agricultural employment share, population density, and municipality size, the exposure coefficient collapses to near zero ($\hat{\beta} = -0.03$, $p = 0.74$). Agricultural employment share alone carries the explanatory power ($\hat{\beta} = 0.60$, $p < 0.001$), explaining roughly 40 percent of the cross-municipal variation in BBB support.

These results suggest that the nitrogen crisis was a catalyst, not a cause. BBB's support base is not the set of municipalities most directly harmed by frozen permits near Natura 2000 sites. It is the set of municipalities where agriculture is the economic backbone---places that saw the nitrogen ruling as an existential threat to their way of life, regardless of their proximity to protected areas. This is consistent with the ``identity threat'' mechanism documented in the political economy literature on cultural backlash \citep{norris2019, guriev2022}, but applied to a specific environmental policy context.

This paper contributes to three literatures. First, it adds to the growing body of work on the political consequences of environmental regulation \citep{walker2013, colmer2023, gazze2024}, providing the first evidence on how a specific court ruling reshaped an entire country's political landscape. The closest precedents are \citet{autor2020}, who trace trade shocks to political polarization, and \citet{fetzer2019}, who links austerity to Brexit---both finding that economic losses drive political realignment. My results complicate this narrative by showing that identity, not material loss, explains BBB support. Second, it contributes to the literature on agrarian populism in Europe \citep{mamonova2020, rodrik2018, funke2016}, showing that BBB's rise reflects agricultural identity rather than narrow economic losses. Third, it demonstrates that judicial decisions on environmental law can produce unintended political consequences extending far beyond the regulated parties \citep{guriev2022, dippel2022}.

The rest of the paper proceeds as follows. Section 2 describes the institutional background of the PAS ruling and BBB's emergence. Section 3 presents the data. Section 4 describes the empirical strategy. Section 5 presents results. Section 6 discusses implications and concludes.

\section{Institutional Background}

\subsubsection*{The Dutch nitrogen problem} The Netherlands has among the highest nitrogen deposition rates in Europe, driven by intensive livestock farming and industrial activity. Nitrogen oxides and ammonia settle on sensitive ecosystems, degrading biodiversity in Natura 2000 sites---162 protected areas covering roughly 14 percent of Dutch territory. Since 2004, the EU Habitats Directive has required member states to maintain ``favorable conservation status'' in these sites, creating tension with the Netherlands' export-oriented agricultural sector \citep{erisman2003}.

\subsubsection*{The PAS system (2015--2019)} In 2015, the Dutch government implemented the Programma Aanpak Stikstof (PAS), a permit system that allowed new economic activities near Natura 2000 sites if they could demonstrate that total nitrogen deposition would eventually decline. The PAS effectively front-loaded permits based on projected future emission reductions from planned policies. This ``borrow from the future'' approach enabled continued construction and agricultural expansion while promising ecological improvement \citep{backes2017}.

\subsubsection*{The ruling (May 29, 2019)} The Raad van State (Council of State) ruled that the PAS violated the Habitats Directive because the projected emission reductions were insufficiently certain. The ruling immediately invalidated the legal basis for all permits issued under PAS, freezing approximately 18,000 pending construction permits and placing thousands of livestock farms in legal uncertainty. Major infrastructure projects---highway expansions, housing developments, airport modifications---ground to a halt \citep{vangrinsven2019}. The economic impact was immediate: the construction sector lost an estimated \texteuro{}14 billion in planned investment, and the government convened an emergency advisory committee (the Remkes Commission) within weeks.

\subsubsection*{Differential exposure} The ruling's effects were geographically heterogeneous. Municipalities with large Natura 2000 areas within their borders faced the most acute permit disruptions, since any new construction or agricultural activity required nitrogen deposition calculations relative to nearby protected sites. Municipalities with high agricultural employment were doubly exposed: their farmers faced potential forced buyouts under emergency nitrogen reduction measures, while their construction sectors lost permits for new development. Rural municipalities in provinces like Gelderland, Overijssel, and Noord-Brabant---where intensive livestock farming overlaps with Natura 2000 sites---bore the brunt.

\subsubsection*{BBB's emergence} The BoerBurgerBeweging was founded on November 22, 2019---six months after the ruling---by Caroline van der Plas, a former agricultural journalist. The party's platform centered on defending farming communities against what it characterized as overreach by environmental regulators and urban elites. BBB's name (``Farmer-Citizen Movement'') deliberately linked agricultural interests with broader rural identity. In the March 2023 Provincial Elections, BBB received 1,488,373 votes (19.2\%), ranging from 5\% in Amsterdam to 59\% in Tubbergen (Overijssel). The party entered the national governing coalition in 2024.

\section{Data}

I combine four data sources covering Dutch municipalities from 2012 to 2025.

\subsubsection*{Building permits} CBS StatLine table 83671NED provides quarterly building permit counts by municipality from 2012Q1 to 2025Q4. I use the number of residential dwelling units approved per quarter as the primary economic outcome. The panel includes 328 municipalities observed over 56 quarters (18,104 municipality-quarter observations after matching).

\subsubsection*{Election results} Official results from the March 15, 2023 Provincial Elections come from the Kiesraad (Electoral Council) in Election Markup Language (EML) format, published on data.overheid.nl. I parse 342 municipality-level XML files to extract vote counts for all parties, computing BBB, FvD, and PVV vote shares as fractions of total valid votes.

\subsubsection*{Natura 2000 boundaries} I obtain spatial polygons for all 209 Natura 2000 sites from the PDOK WFS service (Rijksdienst voor Ondernemend Nederland). Municipality boundaries come from CBS Wijk- en Buurtkaart 2023. Both datasets are projected to Amersfoort/RD New (EPSG:28992) before computing the share of each municipality's area that lies within Natura 2000 boundaries.

\subsubsection*{Municipal characteristics} CBS table 70072NED provides annual municipality-level data on population, employment by sector (including agriculture, industry, and services), population density, average housing value (WOZ), and nitrogen excretion from livestock. I use 2018 values---the last pre-ruling year---to construct treatment intensity measures and control variables.

\input{tables/tab1_sumstats.tex}

\Cref{tab:tab:sumstats} presents pre-treatment summary statistics by exposure group. High-exposure municipalities (above-median Natura 2000 overlap $\times$ agricultural employment) are smaller, more rural, and have higher agricultural employment shares than low-exposure or unexposed municipalities.

\section{Empirical Strategy}

\subsection{Treatment Intensity}

I construct a continuous treatment intensity measure for each municipality $i$:
\begin{equation}
\text{Exposure}_i = \underbrace{\frac{\text{Natura 2000 area}_i}{\text{Municipal area}_i}}_{\text{N2K share}} \times \underbrace{\left(\frac{\text{Agri jobs}_i + \text{Construction jobs}_i}{\text{Total jobs}_i}\right)}_{\text{Ag+Construction share}} \label{eq:exposure}
\end{equation}

This composite captures both the spatial proximity to protected areas (which determines permit disruption) and the sectoral dependence on affected industries (which determines economic vulnerability). I use 2018 values for all components to avoid post-ruling endogeneity.

\subsection{Panel DiD for Building Permits}

For the economic disruption channel, I estimate:
\begin{equation}
\log(\text{Permits}_{it} + 1) = \alpha_i + \gamma_t + \beta \cdot \text{Exposure}_i \times \text{Post}_t + \varepsilon_{it} \label{eq:did}
\end{equation}
where $\alpha_i$ are municipality fixed effects, $\gamma_t$ are quarter fixed effects, and $\text{Post}_t$ equals one from 2019Q3 onwards (the first full quarter after the May 29 ruling). Standard errors are clustered at the municipality level. The parameter $\beta$ measures the differential change in log building permits for a one-unit increase in exposure after the ruling.

\subsection{Cross-Section for Political Outcome}

For the political realignment channel:
\begin{equation}
\text{BBB Share}_i = \alpha + \delta \cdot X_i + \mathbf{W}_i' \boldsymbol{\gamma} + \varepsilon_i \label{eq:cross}
\end{equation}
where $X_i$ alternates between the composite exposure index, agricultural employment share, nitrogen excretion, or Natura 2000 area share. Controls $\mathbf{W}_i$ include population density and log population. Heteroskedasticity-robust standard errors are reported throughout.

The cross-sectional design cannot establish causality---municipalities with high agricultural employment may differ in unobservable ways (historical voting patterns, religious composition, media markets) that independently predict BBB support. I do not claim that agriculture \textit{caused} BBB support. The contribution is in \textit{decomposing} the spatial pattern of BBB vote shares to ask which observable characteristics best explain the pattern: spatial proximity to Natura 2000 sites (the ``economic disruption'' channel) or pre-existing agricultural employment (the ``identity'' channel). This decomposition is informative about mechanism even without a causal design for the political outcome.

\section{Results}

\subsection{Building Permits: No Differential Effect}

\input{tables/tab2_did_permits.tex}

\Cref{tab:did_permits} reports difference-in-differences estimates for building permits. Across all five specifications---using Natura 2000 share, the composite exposure index, per-capita permits, excluding COVID quarters, and a binary high-exposure indicator---none shows a statistically significant differential effect of the nitrogen ruling on building permits. The point estimates are economically small and precisely estimated enough to rule out large effects: the 95\% confidence interval for the composite exposure specification (column 2) excludes effects larger than 3.0 in absolute value.

This null result does not mean the ruling had no effect on construction. National building permit issuance declined sharply after 2019. Rather, the decline was not \textit{differentially} concentrated in municipalities with higher Natura 2000 exposure. This is consistent with the ruling's administrative mechanism: the invalidation of PAS affected the entire national permit system, not only projects near specific Natura 2000 sites. Municipalities far from protected areas still faced uncertainty as the government scrambled to develop a replacement framework.

\subsection{BBB Vote Share: Agricultural Identity, Not Spatial Exposure}

\input{tables/tab3_bbb_cross.tex}

\Cref{tab:bbb_cross} decomposes the determinants of BBB vote share. Column 1 shows that Natura 2000 area share alone is a weak predictor of BBB support ($\hat{\beta} = 0.021$, $p = 0.23$). Column 2 reveals that agricultural employment share is strongly predictive ($\hat{\beta} = 0.60$, $p < 0.001$)---a one-standard-deviation increase in agricultural employment share (3.2 percentage points) is associated with a 1.9 percentage point increase in BBB vote share, or roughly one-fifth of the standard deviation of BBB support.

Column 4 includes both the composite exposure index and agricultural employment share as regressors. The exposure coefficient is essentially zero ($\hat{\beta} = -0.03$), while agriculture retains its strong predictive power ($\hat{\beta} = 0.60$). Population density enters negatively and significantly, confirming that BBB is a rural phenomenon. Together, agricultural employment share and population density explain 40 percent of the cross-municipal variation in BBB vote share (adjusted $R^2 = 0.39$).

Column 5 uses log nitrogen excretion from livestock as an alternative treatment variable. Nitrogen excretion captures actual pollution intensity rather than mere proximity to protected areas. It is a significant predictor ($p < 0.001$), consistent with the interpretation that municipalities whose economies depend on nitrogen-intensive activities---the activities most directly threatened by the ruling---are the strongest BBB supporters.

\subsection{Placebo: Other Populist Parties}

\input{tables/tab4_placebo_parties.tex}

If agricultural employment predicts BBB vote share, does it also predict support for other populist parties? \Cref{tab:placebo_parties} shows that it does not. Agricultural employment share is not significantly associated with either FvD or PVV vote shares (columns 2--3). This is consistent with BBB capturing a \textit{specifically agricultural} grievance rather than general populist sentiment. FvD and PVV draw from different constituencies---urban Eurosceptics and anti-immigration voters, respectively---that do not map onto agricultural geography.

\subsection{Robustness}

I conduct several robustness checks. First, placebo tests using 2016, 2017, and 2018 as artificial treatment dates for the building permits panel yield uniformly insignificant coefficients, confirming no differential pre-trends between high- and low-exposure municipalities. Second, adding province fixed effects to the BBB cross-section leaves the agriculture coefficient virtually unchanged ($\hat{\beta} = 0.62$, $p < 0.001$), ruling out concerns that the result is driven by regional political cultures rather than within-province variation in agricultural identity. Third, leave-one-province-out analysis shows the agriculture coefficient ranges from 0.52 to 0.68, confirming stability. Fourth, using a binary high-exposure indicator yields qualitatively identical results for both the permits DiD and the BBB cross-section.

\section{Discussion and Conclusion}

The Dutch nitrogen crisis offers a vivid case study of how environmental regulation can reshape political landscapes---but through a mechanism different from what simple models would predict. The standard narrative suggests a chain from regulation to economic harm to political backlash: the court ruling froze permits, construction workers lost jobs, farmers faced buyouts, and affected communities voted for the party that promised to reverse the regulation. My results challenge this narrative at its first link. The ruling's economic effects, at least as measured by building permits, were not differentially concentrated in the municipalities that subsequently voted most strongly for BBB.

Instead, BBB's support base maps onto agricultural identity. The municipalities where BBB won 40--59 percent of the vote are not necessarily those closest to Natura 2000 sites or those that lost the most building permits. They are municipalities where farming is the dominant employer---places like Tubbergen, Dinkelland, and Berkelland in eastern Netherlands, where agriculture is not just an industry but a way of life. The nitrogen ruling threatened this identity by framing farming as ecologically destructive, triggering a political response that expressed collective grievance rather than individual economic loss.

This finding has implications beyond the Netherlands. As countries implement increasingly stringent environmental regulations to meet climate and biodiversity targets, political backlash is inevitable. Understanding whether backlash is driven by material losses or identity threats determines the appropriate policy response. Compensation programs that reimburse economic losses may be insufficient if the underlying driver is cultural: the perception that environmental regulation delegitimizes an entire community's livelihood and values.

The paper has important limitations. The cross-sectional analysis of BBB vote share cannot definitively identify a causal relationship. BBB did not exist before the ruling, making pre-treatment political controls imperfect. And the null result on building permits may reflect measurement limitations rather than a true absence of differential effects---the ruling's economic consequences may operate through channels other than residential construction permits, such as agricultural investment, land values, or employment uncertainty.

Nevertheless, the decomposition result---that agricultural identity absorbs the apparent effect of nitrogen exposure on BBB support---is robust across specifications and consistent with qualitative accounts of the nitrogen crisis. When green courts create brown politics, the transmission mechanism runs through who people are, not just what they have lost.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
