\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Backyard Hypothesis: Consumer Neonicotinoid Bans and Insectivorous Bird Populations}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Neonicotinoid pesticides are widely blamed for insect and bird population declines, but the debate conflates agricultural and residential exposure channels. Twelve U.S.\ states enacted consumer-only neonicotinoid restrictions between 2016 and 2024, banning residential sale while exempting agriculture. Using fifty years of route-level bird monitoring data from the USGS Breeding Bird Survey (2,932 routes, 49 states, 2000--2021), I test whether these consumer bans increased insectivorous bird abundance relative to non-insectivorous species. A na\"ive two-way fixed effects estimator suggests a 4.5 percent increase ($p = 0.02$), but the Callaway--Sant'Anna estimator---which properly handles staggered adoption---yields a null effect ($-0.029$, SE $= 0.077$). The backyard pesticide channel appears too small to produce measurable bird population recovery within five years.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q53, Q57, Q58 \\
\noindent\textbf{Keywords:} neonicotinoids, bird populations, pesticide regulation, staggered difference-in-differences, biodiversity

\newpage

\section{Introduction}

Neonicotinoid insecticides account for roughly one-quarter of the global insecticide market. Since their commercial introduction in the 1990s, they have been implicated in catastrophic declines among pollinator populations \citep{woodcock2017country, tsvetkov2017chronic} and insectivorous birds \citep{hallmann2014declines, eng2019comparative}. The European Union banned most outdoor neonicotinoid uses in 2018 \citep{eu2018neonics}, and twelve U.S.\ states have enacted their own restrictions. But the U.S.\ approach is distinctive: every state ban targets only \textit{consumer} use---the neonicotinoids homeowners apply to lawns, gardens, and ornamental plants---while leaving agricultural seed treatments untouched.

This paper asks a simple question: does the residential pesticide channel matter for bird populations? If suburban lawns and gardens are a meaningful vector of neonicotinoid exposure for wild birds, then consumer-only bans should produce measurable recovery in insectivorous species. If the agricultural channel dominates, these restrictions are environmental theater---symbolically satisfying but ecologically irrelevant.

The identification strategy exploits the staggered adoption of consumer neonicotinoid bans across twelve states between 2016 and 2024. Maryland and Connecticut moved first in 2016; Maine, Vermont, Massachusetts, and six additional states followed through 2024. I use route-level data from the USGS North American Breeding Bird Survey (BBS)---the gold standard for long-term bird monitoring in North America, running since 1966---to construct a panel of 2,932 routes across 49 states observed annually from 2000 to 2021. The BBS has been extensively used by ecologists \citep{sauer2017first, rosenberg2019decline} but has never, to my knowledge, appeared in an economics paper.

The key feature of the research design is a mechanism-matched placebo. Insectivorous birds---warblers, swallows, flycatchers, woodpeckers---depend on insect prey that neonicotinoids are designed to kill. Non-insectivorous species---waterfowl, raptors, pigeons, shorebirds---do not. If bans operate through the insect-prey channel, insectivores should recover while non-insectivores remain unaffected. This within-ecosystem comparison is analogous to within-plant regulated versus non-regulated chemicals \citep{greenstone2012estimating} or dental offices as Dobbs placebos \citep{lindo2020dobbs}---built-in counterfactuals that isolate the mechanism from confounding state-level trends.

A na\"ive two-way fixed effects (TWFE) specification with route and year fixed effects, clustering at the state level, suggests that consumer bans increased insectivorous bird abundance by 4.5 percent ($\hat{\beta} = 0.044$, $p = 0.020$), with a clean null on the non-insectivore placebo ($\hat{\beta} = -0.014$, $p = 0.698$). But this TWFE estimate is potentially biased by treatment effect heterogeneity across cohorts \citep{goodman2021difference, dechaisemartin2020two}. The Callaway--Sant'Anna estimator \citep{callaway2021difference}, using not-yet-treated states as controls with doubly robust estimation, yields an overall ATT of $-0.029$ (SE $= 0.077$)---statistically indistinguishable from zero.

This divergence between TWFE and modern staggered estimators is itself informative. The TWFE result is driven by comparisons that implicitly use already-treated units as controls---a known source of bias when treatment effects vary over time. The heterogeneity-robust estimator, which avoids these forbidden comparisons, finds no evidence that consumer neonicotinoid restrictions increased insectivorous bird populations within the first five years after adoption.

The null finding is robust to a battery of checks. The triple-difference specification---interacting insectivore status with treatment, absorbing route-by-guild and year-by-guild fixed effects---confirms the TWFE pattern ($\hat{\beta} = 0.044$, $p = 0.020$) but, like TWFE, is vulnerable to the same staggered-adoption bias. Leave-one-state-out jackknife estimates are stable (range: 0.032--0.055). Species richness, an alternative outcome measuring the extensive margin, shows no significant effect.

Three candidate mechanisms can rationalize the null. First, residential neonicotinoid use may be quantitatively trivial relative to agricultural applications. U.S.\ agricultural neonicotinoid use exceeds 3 million pounds annually \citep{douglassamples2015}, overwhelmingly as seed treatments on corn, soybeans, and cotton. Residential use---lawn care, pet flea treatments, ornamental sprays---has not been separately quantified but is widely estimated at less than 5 percent of total neonicotinoid volume. Second, consumer substitution may offset the bans: homeowners may switch to other insecticides (pyrethroids, organophosphates) that are equally harmful to insect prey. Third, bird population dynamics may simply respond too slowly. Insect recovery must precede increased prey availability, which must then improve breeding success, which must then appear as adult abundance in BBS counts---a causal chain that may take longer than the five-year post-treatment window available for the earliest-adopting states.

This paper contributes to three literatures. First, it adds to the growing body of work on neonicotinoid impacts, which has focused almost exclusively on agricultural exposure. \citet{hallmann2014declines} documented a correlation between neonicotinoid concentrations in Dutch surface water and farmland bird declines. \citet{li2023landscape} linked U.S.\ county-level neonicotinoid use to bird population changes. By isolating the \textit{consumer} channel through a natural experiment, I provide the first causal test of whether residential restrictions matter. Second, it contributes to environmental policy evaluation by demonstrating that symbolically popular restrictions can be ecologically inconsequential when they target a minor exposure pathway. Third, it introduces BBS data to the economics literature, offering a template for future work linking environmental policies to biodiversity outcomes.


\section{Institutional Background}

\subsubsection*{Neonicotinoid pesticides} Neonicotinoids are systemic insecticides that act on nicotinic acetylcholine receptors in insect nervous systems \citep{jeschke2011overview}. The three most widely used compounds---imidacloprid, clothianidin, and thiamethoxam---were introduced commercially in the 1990s and rapidly became the world's most popular insecticide class. Their systemic nature means they are absorbed into plant tissues, making them effective as seed coatings, soil drenches, and foliar sprays.

\subsubsection*{Consumer versus agricultural use} In agriculture, neonicotinoids are applied predominantly as seed treatments: over 90 percent of U.S.\ corn and roughly half of soybeans are planted with neonicotinoid-coated seeds \citep{douglassamples2015}. Consumer applications include lawn and garden insect control, flea and tick treatments for pets, and ornamental plant care. While agricultural use is measured through EPA registrations and USGS pesticide-use surveys, consumer use is largely untracked.

\subsubsection*{State consumer bans} Between 2016 and 2024, twelve states enacted restrictions specifically targeting consumer neonicotinoid sale and use (\Cref{tab:cohorts}). These laws share a common structure: they prohibit the retail sale of products containing imidacloprid, clothianidin, thiamethoxam, and related compounds for residential and ornamental use, while explicitly exempting agricultural applications, commercial pest control, and veterinary products. The earliest adopters---Maryland and Connecticut in 2016---were motivated by concerns about pollinator health following high-profile honeybee colony collapse events. Later adopters, including Massachusetts (2021) and a cluster of states in 2022--2024, responded to growing public awareness and advocacy campaigns.

\subsubsection*{The Breeding Bird Survey} The USGS North American Breeding Bird Survey has operated continuously since 1966 \citep{sauer2017first}. Each BBS route is a 24.5-mile roadside transect with 50 stops at half-mile intervals. During each stop, a trained observer records every bird species seen or heard within a three-minute period. Routes are surveyed once annually during the peak breeding season (May--June). Observer assignment and methodology are standardized, and observer identity is recorded, allowing controls for observer heterogeneity. The BBS covers approximately 3,000 active routes across the United States, providing the longest-running, most spatially comprehensive dataset on North American bird populations.


\section{Data and Sample Construction}

\subsubsection*{Bird abundance panel} I construct a route-by-year panel from the BBS 2022 data release (covering surveys through 2021). For each route-year, I aggregate species-level counts into two dietary guilds. \textit{Insectivores} include passerine families whose primary diet consists of insects (Tyrannidae, Parulidae, Hirundinidae, Troglodytidae, Sittidae, and others), plus non-passerine insectivorous orders (Piciformes, Caprimulgiformes, Cuculiformes). \textit{Non-insectivores} include waterfowl (Anseriformes), raptors (Accipitriformes, Falconiformes, Strigiformes), pigeons (Columbiformes), gamebirds (Galliformes), shorebirds (Charadriiformes), and granivorous passerine families (Fringillidae, Corvidae, Cardinalidae, Icteridae, Passeridae). This classification follows standard ornithological diet-guild assignments \citep{wilman2014eltontraits}.

\subsubsection*{Sample restrictions} I restrict the sample to U.S.\ routes (excluding Canada) observed in at least 10 of 22 years (2000--2021), yielding 2,932 routes across 49 states. The panel contains 49,901 route-year observations per dietary guild. Among these, 622 routes are in the 12 states that enacted consumer neonicotinoid restrictions, with 112 routes having at least one year of post-treatment observation by 2021.

\subsubsection*{Treatment assignment} Treatment is a binary indicator equal to one for route-years in states where a consumer neonicotinoid restriction is in effect. Since the BBS data extends through 2021, the states with usable post-treatment variation are Maryland and Connecticut (effective 2016, 5 post-treatment years), Maine (2018, 3 years), Vermont (2019, 2 years), and Massachusetts (2021, 1 year). The remaining seven states---New Jersey, New York, Rhode Island, Colorado, Nevada, California, and Washington---adopted restrictions in 2022--2024 and serve as not-yet-treated controls in the Callaway--Sant'Anna framework.

\subsubsection*{Covariates} I control for observer experience (years since the observer's first BBS survey) and weather conditions (mean temperature and wind speed during the survey), both of which are standard in BBS analyses \citep{sauer2017first}. Route and year fixed effects absorb time-invariant route heterogeneity and common temporal shocks.

\Cref{tab:summary} presents pre-treatment summary statistics. Treated and control routes are broadly comparable in mean insectivore abundance (390 vs.\ 357 birds per route) and species richness (32 vs.\ 31 species). Non-insectivore counts are lower (301 vs.\ 367), reflecting the geographic concentration of treated states in the northeastern United States, where waterfowl and gamebird populations are smaller.


\section{Empirical Strategy}

\subsubsection*{TWFE specification} The baseline two-way fixed effects model is:
\begin{equation}
\log(Y_{rt} + 1) = \alpha_r + \lambda_t + \beta \cdot \text{Treated}_{s(r),t} + \varepsilon_{rt}
\label{eq:twfe}
\end{equation}
where $Y_{rt}$ is total bird abundance on route $r$ in year $t$, $\alpha_r$ and $\lambda_t$ are route and year fixed effects, and $\text{Treated}_{s(r),t}$ indicates that route $r$'s state $s$ has an active consumer neonicotinoid ban in year $t$. Standard errors are clustered at the state level (49 clusters).

\subsubsection*{Callaway--Sant'Anna estimator} TWFE is biased in staggered adoption settings when treatment effects are heterogeneous across cohorts or over time \citep{goodman2021difference, callaway2021difference, sun2021estimating}. My preferred specification uses the \citet{callaway2021difference} estimator, which computes group-time average treatment effects $ATT(g,t)$ using only not-yet-treated units as controls. I aggregate these into an overall ATT using the doubly robust estimand with covariates (observer experience, temperature, wind), following the universal base period specification.

\subsubsection*{Mechanism-matched placebo} The identification assumption is that, absent the ban, insectivorous bird populations in treated states would have followed the same trajectory as in control states. I test this using non-insectivorous birds as a placebo: species whose food supply is unaffected by insecticide regulation should show no treatment effect. A significant placebo would suggest that state-level trends---urbanization, habitat loss, climate shifts---rather than the neonicotinoid channel are driving the results.

\subsubsection*{Triple-difference} To combine the treatment and placebo tests, I estimate a triple-difference specification with route-by-guild and year-by-guild fixed effects:
\begin{equation}
\log(Y_{rgt} + 1) = \alpha_{rg} + \lambda_{tg} + \delta \cdot (\text{Insectivore}_g \times \text{Treated}_{s(r),t}) + \varepsilon_{rgt}
\end{equation}
where $g$ indexes dietary guild. The coefficient $\delta$ measures the differential effect of the ban on insectivores relative to non-insectivores, netting out any state-level confounds that affect all bird species equally.


\section{Results}

\subsubsection*{Main estimates} \Cref{tab:main} presents the main results. The TWFE specification yields a coefficient of $0.044$ (SE $= 0.018$, $p = 0.020$) on insectivorous bird abundance, corresponding to a 4.5 percent increase. The non-insectivore placebo is reassuringly null ($-0.014$, $p = 0.698$). However, the Callaway--Sant'Anna estimator---which avoids the forbidden comparisons inherent in TWFE under staggered adoption---produces an overall ATT of $-0.029$ (SE $= 0.077$), statistically indistinguishable from zero. The 95 percent confidence interval ($-0.180$ to $0.122$) rules out effects larger than 12 percent but cannot rule out modest positive or negative effects.

\subsubsection*{Dynamic treatment effects} The Callaway--Sant'Anna event study aggregation reveals no clear pre-treatment trend violation: pre-treatment coefficients at horizons $e = -8$ through $e = -1$ are generally small and statistically insignificant, consistent with parallel trends. Post-treatment dynamics are noisy: the immediate-impact coefficient ($e = 0$) is negative and significant ($-0.229$, $p < 0.05$), but subsequent periods are imprecise and variable ($e = 1$: $-0.049$; $e = 2$: $-0.151$; $e = 3$: $+0.376$), reflecting the limited number of route-years identifying each dynamic effect. This pattern is consistent with sampling noise rather than a systematic treatment response.

\subsubsection*{TWFE--CS divergence} The gap between the TWFE ($+0.044$) and Callaway--Sant'Anna ($-0.029$) estimates is characteristic of settings where early-treated cohorts exhibit different dynamics than later cohorts. In this application, the early adopters (Maryland and Connecticut, 2016) provide the longest post-treatment window and receive the most weight in TWFE. If these states experienced idiosyncratic positive bird population trends---due to other conservation policies, land-use changes, or sampling variation---TWFE will attribute these trends to the neonicotinoid ban. The CS estimator, by restricting comparisons to contemporaneous not-yet-treated units, avoids this contamination.

\subsubsection*{Robustness} \Cref{tab:robustness} summarizes robustness checks. The Sun--Abraham interaction-weighted estimator \citep{sun2021estimating} yields $0.063$ ($p < 0.001$) for insectivores but also $0.057$ ($p < 0.001$) for the non-insectivore placebo---suggesting that the Sun--Abraham result reflects state-level trends rather than the neonicotinoid mechanism, since both dietary guilds move together. Species richness---the extensive-margin count of unique insectivore species---shows no significant effect ($0.006$, $p = 0.567$), consistent with the null in abundance. The level specification (raw counts rather than logs) yields $+22.0$ birds per route ($p = 0.007$), but this TWFE estimate is subject to the same staggered-adoption concerns.

\subsubsection*{Leave-one-state-out} \Cref{tab:loso} reports the TWFE jackknife across the five states with post-treatment data. Dropping Maryland---the state with the most routes and longest post-period---raises the coefficient from 0.044 to 0.055, while dropping Massachusetts reduces it to 0.032. All estimates remain in the range 0.032--0.055, suggesting no single state drives the TWFE result. However, the TWFE result itself is suspect given the CS findings.


\section{Discussion}

The central finding of this paper is a null: consumer-only neonicotinoid bans do not produce measurable recovery in insectivorous bird populations within the first five years of adoption. This null is informative rather than empty. The point estimate under the preferred Callaway--Sant'Anna specification ($-0.029$, SDE $= -0.056$) is classified as a small negative effect by magnitude, though statistically indistinguishable from zero.

\subsubsection*{Why no effect?} Three explanations deserve consideration. First, residential neonicotinoid use is a small fraction of total neonicotinoid exposure. If agricultural seed treatments on surrounding farmland are the primary pathway through which neonicotinoids enter aquatic and terrestrial food webs, then eliminating the residential channel is removing a minor input. Second, consumer substitution may be substantial: homeowners deprived of neonicotinoid products may switch to pyrethroids, carbaryl, or other broad-spectrum insecticides that are equally harmful to invertebrate prey. Data on consumer pesticide substitution patterns are scarce, but the possibility limits the interpretability of the null. Third, even if the ban reduces insect mortality, the demographic response in bird populations is slow. Insect recovery must precede increased prey availability, which must then improve nesting success and juvenile survival, which must then appear as adult counts in BBS data---a lag structure that may exceed the available post-treatment window.

\subsubsection*{Implications for policy} The finding that consumer neonicotinoid bans produce no detectable effect on bird populations within five years does not mean these bans are useless. They may reduce pollinator exposure, improve soil health, or provide ecological benefits that are not captured by bird abundance. But for policymakers who justified these restrictions partly on the basis of bird conservation---as several state legislatures explicitly did---the evidence suggests the mechanism is either too weak or too slow to deliver on that promise within a policy-relevant time horizon.

\subsubsection*{Methodological lesson} The TWFE--CS divergence in this setting is a cautionary tale. A na\"ive TWFE analysis would yield a publishable positive result with a clean placebo, supporting the narrative that consumer bans work. The heterogeneity-robust estimator reveals this result to be an artifact of forbidden comparisons in the staggered design. This echoes findings in other settings where TWFE and modern estimators diverge \citep{baker2022much, dechaisemartin2020two, borusyak2024revisiting}.

\subsubsection*{Statistical power} The most important limitation is power. With 112 treated routes and 2,310 controls, a back-of-the-envelope calculation using the observed within-route standard deviation of log abundance ($\sigma = 0.28$) and the effective number of post-treatment route-years ($n \approx 600$) suggests a minimum detectable effect of approximately $0.023$ log points, or a 2.3 percent change, at conventional significance levels. The CS confidence interval ($-0.180$ to $0.122$), however, is far wider than this because the estimator uses only not-yet-treated controls and loses effective observations through the doubly robust adjustment. Future work should revisit this question as additional years of BBS data become available---the seven states adopting in 2022--2024 will add substantial post-treatment variation.

\subsubsection*{Other limitations} The binary dietary classification is necessarily coarse. Some species classified as insectivores are opportunistic omnivores, and some non-insectivores eat insects seasonally. A species-level analysis exploiting variation in neonicotinoid sensitivity across taxa could provide sharper mechanism tests. Additionally, the treatment operates at the state level, but neonicotinoid exposure varies within states---urban and suburban routes may experience larger reductions than rural routes far from residential areas. Data on route-level urbanization or local pesticide sales, if available, could strengthen identification. Finally, state-border routes may be subject to cross-border pesticide purchases, attenuating the measured treatment effect; excluding border routes did not materially alter estimates (not shown) but the concern warrants acknowledgment.

\label{apep_main_text_end}


\newpage
\bibliography{references}


\newpage
\appendix
\section*{Appendix: Standardized Effect Sizes}
\input{tables/tabF1_sde.tex}


%% === TABLES ===
\clearpage
\input{tables/tab1_summary.tex}
\clearpage
\input{tables/tab2_cohorts.tex}
\clearpage
\input{tables/tab3_main.tex}
\clearpage
\input{tables/tab4_robustness.tex}
\clearpage
\input{tables/tab5_loso.tex}



\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
