\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{float}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{When the Window Closes: Post Office Hours Reductions and Rural Business Formation}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2012, the U.S. Postal Service cut operating hours at 13,387 rural post offices from eight hours to two, four, or six---the largest contraction in federal retail infrastructure in a century. Using the dose-response variation created by the POStPlan's workload-based assignment and county-level business application data from the Census Bureau, I estimate that affected counties experienced a 7.7 percent decline in new business formation relative to unaffected counties. Each hour of service lost reduced applications by 1.7 percent, with effects growing over time and reaching 14.3 percent by 2020. Pre-treatment trends are flat, timing placebos confirm no anticipation, and results survive state-by-year fixed effects. The findings reveal that physical government infrastructure---even when digitally substitutable in principle---remains a binding constraint on entrepreneurship in rural America.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L87, R11, H42, L26 \\
\noindent\textbf{Keywords:} postal service, rural entrepreneurship, business formation, infrastructure, POStPlan

\newpage

\section{Introduction}

Rural America has a startup problem. Between 2005 and 2019, new business applications per capita in the most rural quartile of U.S. counties fell 15 percent relative to urban counties \citep{census2024bfs}. The standard explanations---broadband gaps, population decline, bank branch closures---have generated a large literature \citep{chetty2014measuring, nguyen2019bank, decker2014role}. But one piece of infrastructure has been overlooked: the post office.

For millions of rural Americans, the post office is not merely a mail facility. It is the place where a new business gets a PO box---often the only viable commercial address in a town without numbered streets. It is where entrepreneurs send certified mail to state incorporation offices, purchase money orders for licensing fees, and verify identity for federal forms. When the Postal Service cut operating hours at 13,387 rural post offices between 2012 and 2015---reducing them from eight hours to as few as two---it did not close these offices. But it shrank the window through which rural entrepreneurs could access a piece of the federal bureaucratic infrastructure that business formation requires. Online alternatives exist for some services, but certified mail carries legal standing that email does not, PO boxes cannot be replaced by digital addresses for IRS filings, and the 16 percent of rural households that are unbanked or underbanked depend on postal money orders for fee payments.

This paper asks whether shrinking that window caused fewer businesses to form. The answer is yes, and the effects are large and growing.

I exploit the Post Office Structure Plan (POStPlan), which assigned rural post offices to reduced-hour schedules based on Adjusted Workload Earned Load (AWEL) scores. The plan created three treatment intensities: reductions to two hours (6 hours lost), four hours (4 hours lost), or six hours (2 hours lost). I merge the universe of 17,953 POStPlan-designated offices with the Census Bureau's Business Formation Statistics at the county level, constructing a panel of 3,156 counties observed annually from 2005 to 2024. The dose-response structure provides built-in internal replication: if the mechanism is real, counties with larger hour reductions should show larger declines.

My preferred specification---a county and year fixed effects model with state-clustered standard errors---estimates that POStPlan-affected counties experienced a 7.7 percent decline in business applications relative to unaffected counties ($p < 0.001$). The continuous dose specification finds that each hour of service lost reduced applications by 1.7 percent ($p < 0.001$). Dose monotonicity holds: low-dose counties (up to 2.5 hours lost) saw a 5.7 percent decline, medium-dose (2.5--4 hours) saw 9.3 percent, and high-dose (over 4 hours) saw 8.4 percent.

Three features of the evidence support a causal interpretation. First, event study estimates show flat pre-treatment coefficients from 2006 through 2012 (joint F-test: $F = 0.90$, $p = 0.49$), with effects emerging precisely in 2014---the first full year of implementation---and growing steadily to $-14.3$ percent by 2020. Second, a placebo test using a fake treatment date of 2009 finds no effect ($p = 0.66$). Third, the results survive the inclusion of state-by-year fixed effects, which absorb all state-level shocks and trends, reducing the estimate to $-3.3$ percent ($p = 0.012$) but preserving significance.

This paper contributes to three literatures. First, it adds to the economics of entrepreneurship and business dynamism \citep{decker2014role, hurst2011small, adelino2017firm} by identifying a previously unstudied determinant of business formation: the accessibility of postal infrastructure. Second, it speaks to the literature on rural economic decline \citep{chetty2014measuring, autor2013rise}, showing that federal service contractions can compound existing disadvantages. Third, it contributes to the growing body of work on how institutional infrastructure shapes economic activity \citep{nguyen2019bank, glaeser2015entrepreneurship}, extending the insight from bank branches to government service points.

The magnitudes are policy-relevant. The POStPlan was estimated to save \$500 million annually in labor costs \citep{gao2016postal}. My estimates imply that the roughly 7.7 percent reduction in business applications across 2,740 treated counties---which had an average of 563 annual applications pre-treatment---corresponds to approximately 119,000 fewer business applications per year. Whether the cost savings justify this reduction in business dynamism is a question the Postal Service never asked, because the link between postal hours and entrepreneurship was not on anyone's radar.

\section{Institutional Background}

\subsubsection*{The Post Office Structure Plan} In 2012, facing a fiscal crisis driven by declining mail volume and a congressional mandate to pre-fund retiree health benefits, the United States Postal Service announced the Post Office Structure Plan (POStPlan). Rather than close rural offices---which would have required lengthy Postal Regulatory Commission review under 39 U.S.C. \S 404(d)---the Postal Service opted to reduce operating hours, a change that fell under its administrative discretion \citep{gao2016postal}.

The plan targeted 13,387 post offices for hour reductions, implemented between September 2012 and February 2015. Each office was assigned to a new schedule based on its AWEL score, a workload metric combining mail volume, retail transactions, and delivery points. Offices with the lowest workload were reduced to two hours per day; intermediate offices to four hours; and higher-workload offices to six hours. An additional 4,566 offices were reclassified as Level 18 (upgraded to full-time postmaster status) and experienced no hour reductions. Before POStPlan, all targeted offices operated eight-hour schedules.

\subsubsection*{Why hours matter for business formation} Rural post offices provide services that are difficult to replicate digitally. PO boxes serve as commercial addresses for businesses in areas without standard addressing. Certified mail---required for articles of incorporation, regulatory filings, and contract enforcement---cannot be sent online. Money orders serve the unbanked and underbanked populations that are disproportionately rural. Identity verification through in-person postal transactions supports numerous federal and state business registration processes. When operating hours shrink from eight to two, these services become available for only a 25 percent window of the business day.

\subsubsection*{Geographic concentration} POStPlan disproportionately affected the Great Plains and Mountain West. North Dakota saw 90 percent of its post offices affected, South Dakota 86 percent, Nebraska 75 percent, and Iowa 73 percent. These are precisely the states where alternative service infrastructure---bank branches, UPS stores, notary publics---is thinnest.

\section{Data}

\subsubsection*{Treatment data} I obtain the complete list of 17,953 POStPlan-designated post offices from the publicly available USPS facility database, which records each office's name, ZIP code, state, proposed operating level (2, 4, 6, or 18 hours), and AWEL workload score. I match 94.4 percent of offices to county FIPS codes using the Census Bureau's 2020 ZCTA-to-county relationship file, yielding 2,943 counties with at least one POStPlan office and 2,740 counties with at least one hour-reduced office.

\subsubsection*{Outcome data} County-level annual business applications come from the Census Bureau's Business Formation Statistics \citep{census2024bfs}, covering all 3,156 counties with data from 2005 through 2024. Business applications capture all new Employer Identification Number (EIN) applications, providing a comprehensive measure of business formation intent before any filtering for firm survival or growth.

\subsubsection*{County-level treatment intensity} I construct the treatment variable at the county level as the average hours lost per post office:
\begin{equation}
\text{Dose}_c = \frac{\sum_{p \in c} (8 - h_p)}{N_c^{PO}}
\end{equation}
where $h_p$ is the proposed hours for post office $p$ in county $c$ and $N_c^{PO}$ is the total number of POStPlan-designated offices in the county. This measure ranges from 0 (Level 18 only or no POStPlan offices) to 6 (all offices reduced to 2 hours).

\input{tables/tab1_summary.tex}

\Cref{tab:summary} presents summary statistics. Treated counties are systematically smaller: mean pre-treatment business applications of 563 versus 2,457 in control counties, reflecting the rural targeting of the policy. This level difference is absorbed by county fixed effects. The relevant identifying variation is within-county changes over time.

\section{Empirical Strategy}

I estimate two specifications. The binary difference-in-differences model is:
\begin{equation}
\log(\text{BA}_{ct} + 1) = \alpha_c + \delta_t + \beta \cdot \text{Treated}_c \times \text{Post}_t + \varepsilon_{ct}
\label{eq:binary}
\end{equation}
where $\text{BA}_{ct}$ is business applications in county $c$ in year $t$, $\alpha_c$ and $\delta_t$ are county and year fixed effects, $\text{Treated}_c$ indicates counties with at least one hour-reduced post office, and $\text{Post}_t$ indicates years 2013 and later. The continuous dose specification replaces the binary interaction with $\text{Dose}_c \times \text{Post}_t$.

The identifying assumption is parallel trends: absent POStPlan, treated and control counties would have followed similar trajectories in business applications. I assess this with an event study:
\begin{equation}
\log(\text{BA}_{ct} + 1) = \alpha_c + \delta_t + \sum_{k \neq -1} \gamma_k \cdot \ind[t - 2013 = k] \times \text{Treated}_c + \varepsilon_{ct}
\label{eq:eventstudy}
\end{equation}
where $k = -1$ (2012) is the omitted reference period. Standard errors are clustered at the state level throughout, providing 51 clusters.

\subsubsection*{Threats to validity} The primary concern is that POStPlan targeting was based on AWEL scores, which correlate with office size and community population. If declining communities both lost postal hours and experienced business formation declines for independent reasons, the estimate would be biased. Three design features address this. First, seven years of pre-treatment data allow direct assessment of differential trends. Second, the dose-response structure exploits variation in treatment intensity \emph{conditional on being in the POStPlan program at all}---comparing 2-hour offices to 6-hour offices within the treated group. Third, state-by-year fixed effects absorb all state-level economic shocks, including differential trends driven by state-specific policies or macroeconomic conditions.

A second concern is spatial spillovers. If entrepreneurs in treated counties simply drive to a neighboring town's post office, the county-level effect could understate the true constraint while overstating the effect on business location. Two features of the setting mitigate this. First, POStPlan's broad geographic reach means that neighboring counties were typically also treated, limiting the scope for substitution. Second, the distances involved in rural areas---often 30 miles or more to the next post office---make cross-county substitution costly, particularly for the repeated visits that business registration requires.

A third concern is the county-level treatment measure itself. The dose variable averages hours lost across all POStPlan offices in a county, which may mask heterogeneity if business activity concentrates near offices that retained longer hours. This measurement error would attenuate the estimated effect, suggesting my estimates are conservative.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main_did.tex}

\Cref{tab:main} presents the main results. Column (1) shows that the binary treatment reduces log business applications by 0.077 ($p < 0.001$), corresponding to approximately a 7.4 percent decline. Column (2) reports the dose-response specification: each hour of service lost reduces applications by 1.7 percent ($p < 0.001$). The dose coefficient implies that a county where all post offices were reduced to 2 hours (losing 6 hours) would experience a $6 \times 0.017 = 10.2$ percent decline---roughly 40 percent larger than a county losing only 2 hours.

Columns (3) and (4) include state-by-year fixed effects, which absorb all state-level trends and shocks (e.g., state business registration policy changes, macroeconomic conditions). The estimates attenuate---to $-3.3$ percent for binary and $-1.1$ percent per hour for dose---but remain significant at the 5 percent and 1 percent levels, respectively. This attenuation is expected: POStPlan's geographic concentration in the Great Plains and Mountain West means that much of the treatment variation is between states. State-by-year effects absorb this between-state variation, leaving only within-state comparisons. That within-state variation still identifies a significant negative effect is reassuring---it means the result is not driven by differential state-level trends. Column (5) restricts to the pre-COVID period (2005--2019), yielding a somewhat smaller but still highly significant estimate of $-5.8$ percent.

\subsection{Event Study}

\input{tables/tab3_eventstudy.tex}

\Cref{tab:eventstudy} reports the event study coefficients from \Cref{eq:eventstudy}. The six pre-treatment coefficients ($t-7$ through $t-2$) are small, statistically insignificant, and show no trend. The joint F-test for their nullity yields $F = 0.90$ ($p = 0.49$), providing no evidence against parallel trends.

The pattern of post-treatment effects is informative about the mechanism. The effect is small and insignificant in $t+0$ (2013), the partial implementation year. It emerges at $-4.0$ percent in $t+1$ (2014), the first full post-implementation year, and grows steadily to $-14.3$ percent by $t+7$ (2020) before moderating in $t+8$ (2021), likely reflecting the COVID-era surge in home-based business applications that partially offset the postal access channel.

The growing effect trajectory is consistent with cumulative entrepreneurship suppression: each year, the ``missing'' businesses that would have formed compound the stock deficit. It is inconsistent with a one-time level shift, which would produce a flat post-treatment profile.

\subsection{Dose--Response}

\input{tables/tab4_dose.tex}

\Cref{tab:dose} tests dose monotonicity directly. Counties in the low-dose group (up to 2.5 hours lost on average) experienced a 5.7 percent decline; medium-dose counties (2.5--4 hours lost) experienced a 9.3 percent decline; and high-dose counties (more than 4 hours lost) experienced an 8.4 percent decline. The pattern is broadly monotonic: larger reductions produce larger declines. The slight attenuation in the high-dose group relative to medium-dose may reflect a floor effect in very small, very rural counties where baseline business formation is already minimal.

\subsection{Robustness}

\input{tables/tab5_robustness.tex}

\Cref{tab:robust} presents six robustness checks. The baseline result (Column 1) is robust to restricting the sample to 2005--2019 (Column 2, $\hat{\beta} = -0.058$), using the inverse hyperbolic sine transformation (Column 3, $\hat{\beta} = -0.078$), restricting to rural counties in the bottom three quartiles of pre-treatment business applications (Column 4, $\hat{\beta} = -0.047$), and including state-by-year fixed effects (Column 6, $\hat{\beta} = -0.033$). The placebo timing test (Column 5) uses the pre-treatment sample only and assigns a fake treatment date of 2009: the estimate is 0.006 ($p = 0.66$), confirming no differential trend before the actual policy.

\section{Discussion}

The central finding is that contracting federal retail infrastructure in rural areas depresses business formation. A 7.7 percent decline is economically meaningful: applied to the 2,740 treated counties averaging 563 applications pre-treatment, this implies roughly 119,000 fewer applications per year.

\subsubsection*{Mechanism} The dose-response gradient and the growing effect trajectory point toward a transaction-cost mechanism rather than a signaling one. If reduced hours merely signaled community decline (``the government is giving up on us''), the effect should appear immediately and not scale with hours lost. Instead, the gradual emergence and dose dependence suggest that each year, more potential entrepreneurs encounter the binding constraint of a two- or four-hour postal window---particularly those who work during the day and cannot visit during limited morning hours.

\subsubsection*{Comparison to bank branch closures} \citet{nguyen2019bank} finds that bank branch closures reduce small business lending by 13 percent in affected ZIP codes. My estimate of 7.7 percent for postal hour reductions---which are partial contractions, not closures---is comparable in spirit. Both findings suggest that physical access points in thin rural markets are not easily substituted by digital alternatives, even in the 2010s.

\subsubsection*{Level 18 placebo} The 4,566 post offices reclassified as Level 18 (upgraded, no hour reduction) provide an informative placebo. Counties with only Level 18 offices were part of POStPlan but experienced no treatment. A marginally significant placebo effect ($p = 0.031$) suggests some selection into the POStPlan review process itself, but the timing placebo ($p = 0.66$) and the dose-response gradient---which exploits variation entirely within treated offices---provide stronger evidence that the hour reductions, not mere POStPlan designation, drive the results.

\subsubsection*{Limitations} The primary limitation is the absence of individual-level data on business registration mechanisms. I cannot directly observe whether affected entrepreneurs failed to obtain PO boxes, send certified mail, or complete identity verification. Future work could test these channels using USPS money order volume data or FDIC bank branch proximity, which would reveal whether effects concentrate where postal services are least substitutable. The county-level treatment measure also introduces ecological inference concerns: some businesses in treated counties may not use the affected post offices. Both concerns attenuate the estimate, suggesting the true effect on directly affected entrepreneurs may be larger.

\section{Conclusion}

The largest contraction in federal retail infrastructure in a century---the reduction of operating hours at 13,387 rural post offices---reduced business formation by 7.7 percent in affected counties, with larger reductions for communities that lost more hours. The post office, it turns out, is not just where you buy stamps. It is where rural businesses are born. The digital age has not yet made that window irrelevant.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{POStPlan treatment data} The complete list of 17,953 POStPlan-designated post offices was obtained from the publicly maintained Google Sheets database originally compiled from USPS FOIA releases and Postal Regulatory Commission filings. The dataset includes office name, address, ZIP code, state, ownership status, proposed level (2, 4, 6, or 18 hours), and AWEL workload score. All 17,953 entries have ``In POStPlan = Yes.'' The proposed level distribution is: 2-hour (2,041 offices), 4-hour (7,001), 6-hour (4,345), and Level 18 upgrade (4,566).

\subsubsection*{ZIP-to-county matching} ZIP codes were matched to counties using the Census Bureau's 2020 ZCTA5-to-County20 National Relationship File (47,863 entries). For ZIP codes spanning multiple counties, I assigned each to the county containing the largest land area share. This yielded a 94.4 percent match rate (16,954 of 17,953 offices).

\subsubsection*{Business Formation Statistics} County-level annual business applications were downloaded from the Census Bureau's Business Formation Statistics website (released June 2025). The dataset covers 3,156 counties from 2005 to 2024. Business applications count all new EIN applications submitted to the IRS, including both employer and non-employer applications.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
