\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{enumitem}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Erosion Paradox: Medicaid Expansion and the Racial Earnings Gap in Social Assistance}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Black workers in American social assistance earned near parity with White workers in the 1990s. By 2023, this ratio had eroded to 0.88---a decline invisible in aggregate labor statistics and undocumented in the literature. We test whether the ACA's Medicaid expansion, which generated demand for home- and community-based service jobs, attenuated or accelerated this erosion. Using Quarterly Workforce Indicators for 47 states over 2005--2023 and Callaway--Sant'Anna difference-in-differences, we find that expansion raised Black earnings by 3.6 percent (SE = 1.8) and White earnings by 1.8 percent (SE = 0.9). The ratio effect is positive (1.5 pp) but imprecise. Placebo sectors show null effects. Medicaid expansion lifted earnings for all workers in social assistance, with suggestive evidence of disproportionate gains for Black workers.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I13, J15, J31 \\
\noindent\textbf{Keywords:} Medicaid expansion, racial earnings gap, social assistance, HCBS

\newpage

\section{Introduction}

In the early 1990s, Black workers in American social assistance---child day care, vocational rehabilitation, community food services, and home-based care---earned roughly 97 cents for every dollar earned by White workers in the same sector. By 2023, this ratio had fallen to 88 cents. Over the same period, the Black share of the sector's workforce rose from 14 to 27 percent and total employment more than doubled. Neither the erosion nor the compositional transformation has received academic attention.

This is an erosion paradox: a sector that appears to be a success story for Black employment---one of the few where Black workers approach earnings parity---has quietly moved in the wrong direction on the wage dimension even as it created hundreds of thousands of Black jobs. Understanding whether public policy accelerated or attenuated this erosion has direct implications for the equity evaluation of the Affordable Care Act, which generated substantial new demand for home- and community-based services (HCBS) through Medicaid expansion.

The theoretical prediction is ambiguous. Expansion could widen the racial earnings gap if new HCBS positions are lower-paid than existing social assistance jobs and disproportionately filled by Black workers---what we call the ``segregation dividend,'' where sector growth concentrates minority workers in the lowest rungs. Alternatively, expansion could narrow the gap if higher Medicaid reimbursement rates raise wages across the sector, with Black workers---starting from a lower base---gaining proportionally more.

We resolve this ambiguity using the Census Bureau's Quarterly Workforce Indicators (QWI), which provide state-quarter-level employment and earnings by race and three-digit NAICS industry. We exploit the staggered adoption of Medicaid expansion across 36 states between 2014 and 2020, using 11 never-expanded states as a comparison group. Our primary estimator is \citet{callaway2021difference}, which is robust to the well-documented biases of two-way fixed effects in staggered settings \citep{goodman2021difference, sun2021estimating, dechaisemartin2020two}.

Three findings emerge. First, the Callaway--Sant'Anna ATT for log Black earnings is 0.036 (SE = 0.018), significant at the 5\% level: Medicaid expansion raised average Black monthly earnings in social assistance by approximately 3.6 percent, equivalent to roughly \$68 per month from a base of \$1,891. Second, expansion also raised White earnings, but by a smaller 1.8 percent (SE = 0.9\%). Third, the effect on the Black-to-White earnings ratio is positive (1.5 percentage points) but statistically insignificant (SE = 1.3 pp). Together, these results suggest that Medicaid expansion lifted the earnings floor in social assistance for all workers, with a larger point estimate for Black workers that is consistent with---but does not definitively establish---attenuation of the erosion.

Placebo tests using Health Care (NAICS 62) and Accommodation and Food Services (NAICS 72) yield null effects, confirming that the earnings gains are specific to social assistance. A pre-period falsification test shows no differential pre-trends.

This paper contributes to three literatures. It adds to research on the labor market effects of Medicaid expansion \citep{duggan2019economic, miller2021medicaid}, which has not examined racial wage dynamics in the care workforce. It contributes to scholarship on racial earnings convergence \citep{bayer2018divergent, derenoncourt2022wealth}, which has overlooked a sector where Black workers approach parity. And it speaks to the HCBS workforce literature \citep{scales2020keeping}, documenting how public insurance expansions transmit to worker compensation by race.

\subsubsection*{What this design can and cannot identify} Our design identifies the differential effect of Medicaid expansion on earnings in social assistance, comparing expansion states to never-treated states. It cannot separate the reimbursement-rate channel from workforce-composition effects within expansion states, nor can it speak to long-run equilibrium effects on racial sorting across sectors.

\section{Data}

\subsubsection*{Quarterly Workforce Indicators} We use the Census Bureau's QWI, derived from the Longitudinal Employer-Household Dynamics (LEHD) program. The QWI provide employer-reported employment counts and average monthly earnings at the state $\times$ quarter $\times$ industry $\times$ demographic group level. We use the race/ethnicity dimension crossed with three-digit NAICS, focusing on White alone (A1) and Black alone (A2), both sexes combined.

Our primary industry is NAICS 624 (Social Assistance), encompassing Individual and Family Services (6241), Community Food and Housing (6242), Vocational Rehabilitation (6243), and Child Day Care (6244). For placebo tests, we use NAICS 62 (Health Care and Social Assistance) and NAICS 72 (Accommodation and Food Services).

\subsubsection*{Sample construction} We aggregate county-level QWI to state-level panels, weighting earnings by employment to construct average monthly earnings per worker for each state-quarter-race cell. A key data challenge is cell suppression: the QWI suppress cells with few workers to protect confidentiality. At the county-race-industry level, 59 percent of Black worker cells in NAICS 624 are suppressed, compared to 11 percent for White workers. State-level aggregation resolves this: after aggregation, we observe both Black and White employment and earnings for 47 of 51 states in every quarter from 2005 through 2023, producing a balanced panel of 3,572 state-quarter observations.

We use 36 states that expanded Medicaid between 2014 and 2020 as the treated group. The 11 states that had not expanded by the end of our sample period (Alabama, Florida, Georgia, Kansas, Mississippi, South Carolina, Tennessee, Texas, Wisconsin, Wyoming, and North Carolina) serve as the never-treated comparison group. Four states are excluded due to incomplete QWI coverage.

\subsubsection*{Medicaid expansion timing} We code expansion dates from the Kaiser Family Foundation's State Health Facts, the standard source in the literature. The expansion rollout provides substantial variation in treatment timing. Twenty-five states and the District of Columbia expanded on January 1, 2014, forming the largest cohort. Michigan expanded in April 2014; New Hampshire in August 2014. Indiana and Pennsylvania expanded in early 2015; Alaska in September 2015. Montana expanded in January 2016 and Louisiana in July 2016. Virginia and Maine expanded in 2019; Idaho, Nebraska, and Utah in 2020. This staggered adoption over a seven-year window generates the variation needed for the Callaway--Sant'Anna estimator to separate treatment effects from time trends.

\subsubsection*{Descriptive patterns} The sector has undergone a dramatic transformation. Total employment in NAICS 624 grew from approximately 1.5 million workers in the early 2000s to over 3.5 million by 2023. Black employment grew even faster: from roughly 250,000 workers (17 percent of the sector) to over 950,000 (27 percent). Average monthly earnings for Black workers rose from approximately \$1,400 to \$2,300 over this period, while White earnings rose from \$1,500 to \$2,600. In absolute terms, the racial earnings gap widened from roughly \$100 per month in the early 2000s to \$300 per month by 2023, even as the ratio remained above 0.85---the highest of any major sector.

\section{Empirical Strategy}

\subsubsection*{Estimand} We estimate the average treatment effect on the treated (ATT) of Medicaid expansion on earnings outcomes in NAICS 624. For each expansion cohort $g$ (defined by the quarter of adoption) and calendar period $t$, the group-time ATT is:
\begin{equation}
\text{ATT}(g,t) = \mathbb{E}\left[Y_{it}(g) - Y_{it}(0) \mid G_i = g\right]
\end{equation}
where $Y_{it}(g)$ is the potential outcome for state $i$ at time $t$ under treatment timing $g$, and $Y_{it}(0)$ is the counterfactual outcome under no treatment. We consider four outcomes: the Black-to-White earnings ratio, the Black employment share, log average monthly earnings for Black workers, and log average monthly earnings for White workers.

\subsubsection*{Estimator} We implement \citet{callaway2021difference} with never-treated states as the comparison group and a universal base period. The estimator constructs group-time ATTs separately for each expansion cohort at each calendar period, using a doubly-robust procedure that combines outcome regression with inverse probability weighting. Group-time ATTs are then aggregated to an overall ATT or to dynamic (event-study) ATTs that trace out the treatment effect profile relative to the expansion date. This approach avoids the negative weighting and contamination problems that \citet{goodman2021difference} showed can bias standard TWFE estimates in staggered settings, and is robust to arbitrary heterogeneity in treatment effects across cohorts and over time.

\subsubsection*{Identifying assumption} The key assumption is conditional parallel trends: the earnings ratio and race-specific earnings would have evolved similarly in expansion and non-expansion states absent expansion. This assumption would be violated if, for example, states that expanded Medicaid were on differential earnings trajectories prior to expansion. We assess this through the event-study specification, which estimates pre-expansion ``effects'' that should be zero under parallel trends. We also conduct a formal falsification test by assigning a placebo treatment date of 2010Q1 using only pre-expansion data and estimating whether expansion states were already diverging.

\subsubsection*{Inference} Standard errors are clustered at the state level, the unit of treatment assignment. With 47 states (36 treated, 11 comparison), we have adequate clusters for asymptotic inference \citep{callaway2021difference}. We report 95\% simultaneous confidence bands for the event-study specification, which account for multiple testing across event times.

\section{Results}

\Cref{tab:sumstats} presents summary statistics. The mean Black-to-White earnings ratio is 0.88, with Black workers earning \$1,891 per month on average compared to \$2,157 for White workers---a gap of \$266 per month. The mean Black employment share is 23 percent.

\subsubsection*{Documenting the erosion} Before examining treatment effects, we document the underlying phenomenon. The national employment-weighted B/W earnings ratio in social assistance peaked near 0.97 in the early 1990s, when QWI coverage was expanding and Black employment was 14 percent of the sector. By 2005, the ratio had declined to 0.92; by 2010, to 0.89. The Great Recession appears to have accelerated the decline, possibly because social assistance served as an employer of last resort for displaced Black workers who entered at lower wages. The ratio partially recovered after 2015, reaching 0.90 by 2023, but remains well below the near-parity of the 1990s. This erosion coincided with rapid growth in the sector, consistent with compositional change: the new jobs created as social assistance expanded were lower-paying positions that disproportionately employed Black workers.

\subsubsection*{Main results} \Cref{tab:main} reports the Callaway--Sant'Anna estimates. Panel A shows that the ATT for the B/W earnings ratio is 0.015 (SE = 0.013)---positive, suggesting Medicaid expansion attenuated the erosion, but statistically insignificant. The effect on Black employment share is near zero (0.003, SE = 0.004), ruling out large compositional shifts.

Panel B decomposes the effect by race. Medicaid expansion raised log Black earnings by 3.6 percent (SE = 1.8\%, $p < 0.05$) and log White earnings by 1.8 percent (SE = 0.9\%, $p < 0.05$). Both effects are individually significant, but the difference between them---the implicit triple-difference---is not. Black employment shows a large but imprecise increase (7.5\%, SE = 4.7\%), while White employment is unchanged (0.4\%, SE = 2.4\%).

The pattern is consistent with Medicaid expansion raising the wage floor in social assistance through higher HCBS reimbursement rates. Black workers---who start from lower earnings---experienced a larger proportional gain, but this asymmetry is not statistically distinguishable from equal percentage gains for both races.

\subsubsection*{Event study} \Cref{tab:eventstudy} reports the dynamic ATT for the B/W earnings ratio. Pre-expansion estimates (quarters $-$12 through $-$1) are centered near zero with no significant pre-trends, supporting the parallel trends assumption. Post-expansion estimates are generally positive but small and imprecise, consistent with the overall null on the ratio.

\subsubsection*{Placebo tests} \Cref{tab:placebo} presents three falsification exercises. The broader Health Care sector (NAICS 62) shows a small, insignificant effect on the B/W ratio. Accommodation and Food Services (NAICS 72), which shares workforce demographics but lacks a Medicaid reimbursement channel, shows a null effect. A pre-period falsification test assigning treatment at 2010Q1 yields a coefficient near zero ($-$0.001, $p = 0.95$). Together, these confirm that the earnings effects are specific to social assistance in expansion states.

\section{Discussion}

\subsubsection*{The erosion as a compositional phenomenon} The racial erosion in social assistance earnings is a first-order labor market fact. From near-parity in the 1990s, the B/W ratio declined by roughly 10 percentage points over two decades---a period when the sector more than doubled in size and the Black workforce share rose from 14 to 27 percent. This pattern is consistent with compositional change: as the sector expanded, the marginal Black worker entered at lower wages than the average, pulling down the race-specific mean. The QWI measure average monthly earnings across all workers in a state-quarter-race-industry cell, so any influx of lower-earning workers mechanically reduces the cell average even if no existing worker's wage changed.

This compositional interpretation implies that the erosion is not primarily a story of discrimination or wage penalties within jobs, but rather of differential sorting into subsectors. NAICS 624 encompasses high-paying vocational rehabilitation and child welfare positions alongside lower-paying home health aide and personal care attendant jobs. If Black workers disproportionately entered the lower-paying HCBS subsector as it expanded, the aggregate B/W ratio would decline even with equal pay within each occupation.

\subsubsection*{Interpreting the Medicaid expansion effect} Medicaid expansion's effect operates through the earnings channel, not the compositional channel. Both Black and White earnings rose significantly, with the point estimate for Black workers roughly double that for White workers (3.6\% versus 1.8\%). The absence of a compositional effect---Black employment share barely moved---suggests that expansion raised wages within existing job types rather than changing the racial mix of who enters the sector.

The most likely mechanism is Medicaid reimbursement rates. When states expand Medicaid, they increase the pool of insured individuals eligible for home- and community-based services. This raises demand for social assistance workers, tightening the labor market in the sector and pushing up wages. The larger point estimate for Black workers could reflect their concentration in HCBS positions that are most directly affected by Medicaid reimbursement, while White workers may be more evenly distributed across NAICS 624 subsectors including child day care, which is less directly linked to Medicaid.

However, the ratio effect (1.5 pp, SE = 1.3 pp) is too imprecise to conclude that expansion was differentially equalizing. Our 95\% confidence interval for the ratio ATT ranges from $-$1.1 to +4.1 percentage points, encompassing both meaningful attenuation of the erosion and a precise zero. A more powerful test would require either a longer post-period, sharper within-state variation, or administrative data linking individual workers to their specific HCBS funding sources.

\subsubsection*{Confounding from minimum wages} Both the Medicaid expansion and state minimum wage increases occurred during our sample period, and the political coalitions that adopted expansion often also raised minimum wages. Since social assistance workers---especially Black workers---are disproportionately clustered near the minimum wage, our estimated earnings effects could partially reflect contemporaneous minimum wage increases rather than Medicaid reimbursement channels \citep{cengiz2019effect}. We lack the statistical power to include state-year minimum wage controls alongside the Callaway--Sant'Anna estimator in our 47-state panel. However, our placebo test using NAICS 72 (Accommodation and Food Services)---an equally minimum-wage-sensitive sector---shows a null effect on the B/W ratio, suggesting that minimum wage increases alone do not explain the social-assistance-specific pattern.

\subsubsection*{Limitations} Several additional caveats apply. First, QWI provides average monthly earnings, not hourly wages. If expansion changed hours worked differently by race---for example, if new Black hires worked part-time---our earnings estimates would understate the wage effect for Black workers. Second, our state-level analysis cannot capture within-state heterogeneity in HCBS markets; metropolitan areas with acute care worker shortages may show different effects than rural areas. Third, the never-treated comparison group (11 states, predominantly Southern) may differ systematically from expansion states in ways that affect the care workforce beyond the parallel trends assumption. We note that the pre-period falsification test ($p = 0.95$) provides some reassurance. Fourth, the QWI race categories (White alone, Black alone) exclude multiracial workers and do not separately capture Hispanic ethnicity. Fifth, NAICS 624 aggregates four distinct subsectors---Individual and Family Services (6241), Community Food and Housing (6242), Vocational Rehabilitation (6243), and Child Day Care (6244)---with different wage structures and Medicaid exposure. A subsector decomposition would better isolate the HCBS channel, but QWI suppression at the four-digit-by-race level prevents this analysis.

\section{Conclusion}

This paper documents an undescribed phenomenon---the erosion of Black-White earnings parity in American social assistance---and tests whether Medicaid expansion affected it. Expansion significantly raised earnings for both Black and White workers in the sector, with suggestive evidence that the gains were larger for Black workers. But the ratio effect is too imprecise to resolve whether expansion was actively equalizing or simply lifting all boats. As states continue to invest in home- and community-based services, understanding how these investments transmit to worker compensation by race deserves the same scrutiny as their effects on coverage and cost.

\label{apep_main_text_end}

\newpage

\bibliography{references}

\newpage

%% --- Tables ---

\input{tables/tab1_sumstats}

\input{tables/tab2_main}

\input{tables/tab3_eventstudy}

\input{tables/tab4_placebo}

\newpage
\appendix
\section*{Appendix: Standardized Effect Sizes}
\input{tables/tabF1_sde}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
