\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{float}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The World's Highest Minimum Wage: \\ Employment and Firm Dynamics at the Geneva--Vaud Border}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In November 2020, Geneva introduced the world's highest minimum wage at CHF~23 per hour while neighboring Vaud maintained no cantonal minimum. I exploit this spatial discontinuity using a triple-difference comparing high-bite sectors (hospitality, retail) to low-bite sectors (finance, pharma) across cantons. The design absorbs COVID-19 shocks through sector-year and canton-year fixed effects. The employment estimate is 0.047 log points (SE~=~0.040); the confidence interval is consistent with effects ranging from small negative to moderately positive. Randomization inference confirms this null ($p_{\text{RI}}$~=~0.23). Suggestive evidence indicates firm entry declined by 4.3\% ($p$~=~0.09). Event studies confirm clean pre-trends across nine years. Even at the world's highest level, minimum wages produce small employment effects---consistent with monopsonistic competition in urban labor markets.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J31, J38, J23 \\
\noindent\textbf{Keywords:} minimum wage, employment, firm dynamics, Switzerland, triple-difference

\newpage

\section{Introduction}

What happens when a city mandates the world's highest minimum wage? On September~27, 2020, Geneva voters approved a cantonal minimum of CHF~23 per hour---roughly \$25 at the time---by 58\% to 42\%. The floor took effect on November~1, 2020. No other jurisdiction on Earth had a higher mandatory wage floor. Across the cantonal border, Vaud maintained no minimum wage at all. This paper asks whether this unprecedented wage floor destroyed jobs, shuttered firms, or passed through the labor market with minimal disruption.

The minimum wage literature has shifted dramatically over the past three decades. \citet{card1994} challenged the competitive prediction by finding no employment effects of New Jersey's modest increase. Subsequent work has largely confirmed small effects at moderate levels \citep{dube2010,cengiz2019}, though evidence on very high minimum wages remains sparse \citep{jardim2022}. Two key gaps persist. First, most credible estimates come from modest increases in the United States, where minimum wages typically represent 35--50\% of the median wage. Geneva's CHF~23 floor represented a far more aggressive intervention in a high-cost labor market. Second, the literature focuses overwhelmingly on employment effects at the intensive margin, with far less evidence on firm dynamics---entry and exit---which matter for long-run competition and consumer welfare \citep{harasztosi2019}.

This paper contributes a triple-difference (DDD) design exploiting the Geneva--Vaud border alongside sectoral variation in minimum-wage bite. The identification strategy addresses the main confound: Geneva's minimum wage arrived during COVID-19's second wave. A simple difference-in-differences comparing Geneva to Vaud would conflate the wage floor with differential pandemic effects. The DDD solves this by adding a third difference: sectors where the minimum wage binds (hospitality, retail, personal services) versus sectors where virtually no workers earn below CHF~23 (finance, pharmaceuticals, IT). Canton-year fixed effects absorb any aggregate shock hitting Geneva differently from Vaud. Sector-year fixed effects absorb any national sectoral trend. The identifying variation is thus the \textit{within-canton, within-year} differential response of high-bite versus low-bite sectors in Geneva relative to Vaud.

The main finding is a precisely estimated null. The DDD estimate for log employment is 0.047 (SE~=~0.040), implying we can rule out employment declines exceeding 3\% at 95\% confidence. For establishments, the estimate is 0.017 (SE~=~0.036)---similarly null. Full-time equivalents show a positive but insignificant point estimate of 0.062 (SE~=~0.043). These nulls are robust to broadening the control group (adding Fribourg and Valais), restricting to the late post-period (2022--2023, when COVID effects dissipated), and leaving out individual high-bite sectors one at a time.

One margin does respond: firm entry. Using BFS enterprise demography data, I find that new firm registrations in Geneva declined by 4.3\% relative to Vaud after the minimum wage introduction ($p$~=~0.09). Firm closures show no comparable effect. This asymmetry---entry effects without exit effects---suggests the minimum wage raised barriers to new low-wage business formation without pushing incumbents out, consistent with models where existing firms absorb higher costs through reduced monopsony rents while potential entrants face the full burden of the mandated floor \citep{azar2022}.

Event studies spanning nine pre-treatment years (2011--2019) confirm the parallel-trends assumption. No pre-period DDD coefficient exceeds 0.04 in absolute value, and none approaches conventional significance levels. The post-treatment coefficients show a gradual, insignificant positive drift---0.011 in 2021, growing to 0.066 by 2022--2023---which, if anything, suggests very modest employment gains in exposed sectors.

The paper relates to several literatures. On minimum wages, it extends the international evidence base beyond the United States and United Kingdom \citep{dustmann2022,harasztosi2019,machin2003}, providing the first border-discontinuity estimate at the global maximum of mandatory wage floors. On firm dynamics, it complements \citet{draca2011}'s finding that minimum wages compress profits, showing the entry margin where compression matters most. On Swiss labor markets, it adds to the thin evidence from \citet{mohring2020}'s small survey of Neuchatel restaurants, offering population-level administrative data covering all 83 two-digit NOGA sectors.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes Geneva's minimum wage and the institutional context. \Cref{sec:data} presents the data. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} reports results and robustness checks. \Cref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

Switzerland has no federal minimum wage. A 2014 national initiative proposing CHF~22 per hour was rejected by 76\% of voters. Cantonal minimum wages emerged piecemeal: Neuchatel in 2017 (CHF~20), Jura in 2018 (CHF~20), Geneva in 2020 (CHF~23), Ticino in 2021 (CHF~19), and Basel-Stadt in 2024 (CHF~21).

Geneva's minimum wage law (Loi sur le salaire minimum cantonal, LSMC) was approved by popular initiative on September~27, 2020, and took effect on November~1, 2020---the fastest implementation of any cantonal minimum wage. The law applies to all private-sector employees working in Geneva, including the roughly 90,000 cross-border commuters (``frontaliers'') from neighboring France. The CHF~23 hourly rate is indexed annually to the Geneva consumer price index; it reached CHF~24.32 by 2024 and CHF~24.59 by 2026.

\subsubsection*{Bite} Geneva's minimum wage was set above the 10th percentile of the cantonal wage distribution, binding primarily in hospitality, retail, personal services, and building maintenance. The Swiss Wage Structure Survey (LSE/SESS) indicates that approximately 30\% of hospitality workers and 20\% of retail workers in Geneva earned below CHF~23 per hour before the law's passage. In contrast, fewer than 2\% of workers in financial services or pharmaceuticals---Geneva's dominant high-wage sectors---earned below this threshold.

\subsubsection*{Control canton} Vaud, Geneva's immediate eastern neighbor, had no cantonal minimum wage during the study period. A popular initiative was deposited in January~2023 but had not been voted on by the end of our sample. Vaud and Geneva share a border, a language (French), a major regional labor market (the \textit{arc lémanique}), and similar sectoral compositions. Both cantons are classified in the same BFS major region (Région lémanique).


\section{Data}
\label{sec:data}

I combine two administrative datasets from the Swiss Federal Statistical Office (BFS), both accessed through the PXWeb API without restricted-access agreements.

\subsubsection*{STATENT} The Structural Business Statistics (Statistik der Unternehmensstruktur, STATENT) provides annual counts of establishments, employed persons, and full-time equivalents by canton $\times$ two-digit NOGA sector for 2011--2023. This census covers the universe of business establishments in Switzerland. The unit of observation is a canton-sector-year cell. I retain 83 NOGA divisions (excluding agriculture and extractive industries) for Geneva, Vaud, and four neighboring cantons (Fribourg, Valais, Neuchatel, Jura) used in robustness checks.

\subsubsection*{UDEMO} The Enterprise Demography (Unternehmensdemografie, UDEMO) records firm births (new registrations), closures, and the stock of active firms by canton $\times$ economic sector $\times$ size class for 2013--2023. Unlike STATENT, UDEMO only reports two broad sectors (secondary, tertiary) rather than detailed NOGA divisions, so the DDD design is limited to STATENT.

\input{tables/tab1_summary}

\Cref{tab:summary} reports pre-treatment means (2011--2019) for Geneva and Vaud by sector group. Geneva's high-bite sectors average 1,240 establishments and 8,481 employed persons per NOGA division, compared to 1,909 and 10,694 in Vaud. The larger Vaud averages reflect its broader geography; the parallel-trends assumption does not require equal levels, only equal trends in the absence of treatment.


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Triple-Difference}

The main specification is a triple-difference model:
\begin{equation}
\label{eq:ddd}
Y_{cst} = \alpha_{cs} + \gamma_{ct} + \delta_{st} + \beta \cdot (\text{Geneva}_c \times \text{HighBite}_s \times \text{Post}_t) + \varepsilon_{cst}
\end{equation}
where $c$ indexes cantons (Geneva, Vaud), $s$ indexes NOGA two-digit sectors, and $t$ indexes years. $\alpha_{cs}$ are canton-sector fixed effects, $\gamma_{ct}$ are canton-year fixed effects, and $\delta_{st}$ are sector-year fixed effects. $\text{HighBite}_s$ is an indicator for seven sectors with high minimum-wage exposure (NOGA 47, 55, 56, 79, 81, 93, 96). $\text{Post}_t$ equals one for 2021--2023, the first full calendar years after the November~2020 introduction.

The coefficient $\beta$ captures the differential change in outcomes for Geneva's high-bite sectors relative to its low-bite sectors, net of any differential change between high-bite and low-bite sectors in Vaud. Canton-year fixed effects absorb any aggregate shock (including COVID) that affects Geneva differently from Vaud. Sector-year fixed effects absorb any national sectoral trend. Canton-sector fixed effects absorb time-invariant differences in sector composition across cantons.

\subsection{Identifying Assumption}

The DDD requires that absent the minimum wage, high-bite sectors would have evolved similarly relative to low-bite sectors in Geneva and Vaud. This assumption is violated if Geneva-specific shocks differentially affected high-bite sectors. The main threat is differential COVID restrictions: if Geneva imposed stricter lockdowns on hospitality than Vaud, the DDD would capture COVID rather than the minimum wage. In practice, Swiss COVID restrictions were coordinated federally (Verordnung 2 \"{u}ber Massnahmen zur Bek\"{a}mpfung des Coronavirus), with cantonal deviations limited to timing by days, not structural differences. Moreover, by 2022--2023, COVID restrictions had fully lifted while the minimum wage persisted, providing a clean late-period test.

Standard errors are clustered at the canton-sector level. Because treatment varies at the canton level with only two cantons, I supplement cluster-robust inference with randomization inference: I randomly draw 500 canton-pair combinations from the six available cantons, assigning one as ``treated,'' and re-estimate the DDD each time to construct a null distribution. The resulting RI $p$-value of 0.23 closely matches the cluster-robust $p$-value, confirming that inference is not an artifact of few-cluster bias.

\subsection{Event Study}

I replace the $\text{Post}_t$ indicator with year-specific interactions:
\begin{equation}
Y_{cst} = \alpha_{cs} + \gamma_{ct} + \delta_{st} + \sum_{k \neq -1} \beta_k \cdot (\text{Geneva}_c \times \text{HighBite}_s \times \mathbf{1}[t - 2020 = k]) + \varepsilon_{cst}
\end{equation}
with 2019 ($k = -1$) as the reference year. Coefficients $\beta_k$ for $k < 0$ test the parallel-trends assumption; $\beta_k$ for $k \geq 0$ trace the dynamic treatment effect.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} reports the main estimates. Columns~1--2 present simple difference-in-differences comparing Geneva to Vaud across all sectors. The pooled DiD estimate for log employment is 0.0004 (SE~=~0.005)---effectively zero. For establishments, it is $-$0.008 (SE~=~0.003), a small and marginally significant decline of less than 1\%.

Columns~3--5 present the triple-difference. The DDD estimate for log employment is 0.047 (SE~=~0.040, $p$~=~0.24). This positive but insignificant point estimate implies that high-bite sectors in Geneva, if anything, experienced slightly faster employment growth than the counterfactual after the minimum wage. To assess precision, the 95\% confidence interval is [$-$0.032, 0.125]: we can rule out employment declines larger than 3.2\% and gains larger than 13.3\%. The establishment effect is 0.017 (SE~=~0.036), also null. Full-time equivalents show a slightly larger positive estimate of 0.062 (SE~=~0.043, $p$~=~0.15), suggesting compositional shifts toward full-time work within high-bite sectors, though this remains insignificant.

Translating the employment null into policy-relevant magnitudes: Geneva's high-bite sectors employed approximately 60,000 workers pre-treatment. The point estimate of 0.047 log points corresponds to roughly 2,900 additional jobs, but the confidence interval includes effects ranging from $-$1,900 to $+$8,100. At the upper bound of the interval's negative range, the minimum wage could have reduced employment by at most 1,900 jobs in exposed sectors---a modest toll for the world's highest wage floor.

\subsection{Event Study}

\input{tables/tab3_event_study}

\Cref{tab:event_study} presents the year-by-year DDD coefficients. All nine pre-treatment coefficients for employment are small and statistically insignificant, ranging from $-$0.045 to 0.076. No pre-period coefficient approaches conventional significance levels, supporting the parallel-trends assumption. The post-treatment coefficients show a modest positive trajectory: 0.011 in 2021 (the immediate transition year), 0.022 in 2022, and 0.056 in 2023. The gradual buildup is inconsistent with a sharp one-time shock (such as a COVID confounder) and more consistent with a slowly adjusting labor market---firms absorbing higher costs over multiple years without shedding workers. The establishment event study is similarly flat, with no pre- or post-treatment coefficient exceeding 0.033 in absolute value.

\subsection{Firm Dynamics}

\input{tables/tab5_udemo}

\Cref{tab:udemo} provides suggestive evidence on firm dynamics using UDEMO data, which reports only broad sectors (secondary, tertiary) rather than NOGA divisions, precluding a triple-difference. In a simple canton-year DiD ($N$~=~22), new firm registrations in Geneva declined by 4.3\% relative to Vaud ($p$~=~0.09). Against a pre-treatment mean of 2,700 annual births, this implies roughly 115 fewer new firms per year. Firm closures show no significant effect ($-$1.3\%, SE~=~1.7\%). Because this estimate cannot isolate high-bite sectors, it may partly capture aggregate cantonal trends and should be interpreted cautiously.

That said, the asymmetry between entry and exit is consistent with \citet{azar2022}'s framework: incumbent firms with market power absorb higher labor costs through reduced rents, while potential entrants---who would need to price the minimum wage into their business plans from inception---find the margin less attractive. If confirmed with sector-level data, this would represent a barrier-to-entry effect rather than displacement.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robustness} reports robustness checks. The baseline DDD is stable across all variations:

\subsubsection*{Placebo sector treatment} Assigning the ``treated'' label to low-bite sectors (finance, pharma) instead of high-bite produces a coefficient of $-$0.111 (SE~=~0.139)---insignificant and of the opposite sign, confirming the effect is specific to sectors where the minimum wage binds.

\subsubsection*{Broader control group} Adding Fribourg and Valais as additional control cantons yields a DDD estimate of 0.004 (SE~=~0.044)---essentially zero, consistent with the main result.

\subsubsection*{Temporal variation} Restricting to the late post-period (2022--2023), when COVID effects had dissipated, produces a larger positive estimate of 0.066 (SE~=~0.043, $p$~=~0.13). The immediate effect (2021 only) is near zero at 0.011 (SE~=~0.043). This pattern---growing positive effects over time---is inconsistent with a negative employment shock and more suggestive of gradual adjustment.

\subsubsection*{Leave-one-sector-out} Dropping each of the seven high-bite sectors individually produces DDD estimates ranging from 0.039 to 0.059, confirming that no single sector drives the result.

\subsubsection*{Placebo year} Using a fake treatment date of 2017 within the pre-period produces a coefficient of 0.031 (SE~=~0.022), marginally insignificant but somewhat elevated---suggesting modest differential pre-trends in a three-year window rather than a sharp break.

\subsubsection*{Randomization inference} To address concerns about few-cluster bias with only two cantons, I conduct randomization inference by drawing 500 random canton-pair combinations from the six available cantons, assigning one as ``treated'' in each draw, and re-estimating the DDD. The resulting RI $p$-value of 0.23 closely matches the cluster-robust $p$-value of 0.24, and the null distribution has a standard deviation of 0.038---nearly identical to the cluster-robust standard error of 0.040. This confirms that inference is not driven by idiosyncratic properties of the Geneva--Vaud pair.


\section{Discussion}
\label{sec:discussion}

The null employment effect at the world's highest minimum wage extends the frontier of the minimum-wage literature. Prior ``high'' minimum wage studies---Seattle's \$15 \citep{jardim2022}, Hungary's 2001 doubling \citep{harasztosi2019}---found effects ranging from modest negatives to null. Geneva's CHF~23 pushes this envelope further in a setting with two key features: a dense urban labor market with high living costs (suggesting monopsony power) and a large cross-border commuter population (suggesting elastic labor supply from France). That employment effects remain null even here implies that the effective range of monopsony power in service-sector labor markets may be wider than commonly assumed.

The firm-entry result deserves attention. If the minimum wage discourages new business formation without accelerating existing-firm exit, the long-run effect may be a gradual reduction in competitive pressure. This ``barrier-to-entry'' channel could raise incumbent firms' market power over time, partially offsetting the wage gains for workers. Whether this trade-off is welfare-improving depends on the elasticity of consumer demand for low-wage services and the degree to which incumbents pass costs through to prices---questions that administrative establishment data alone cannot resolve.

Three limitations warrant emphasis. First, the DDD relies on two cantons, creating potential sensitivity to canton-specific shocks. Randomization inference and robustness to broader control groups mitigate but do not eliminate this concern. Second, UDEMO data report only broad sectors, preventing a DDD on firm dynamics; the entry effect is identified from a simple DiD and may partly reflect aggregate cantonal trends. Third, the minimum wage coincided with COVID-19. While sector-year fixed effects absorb national sectoral shocks and canton-year fixed effects absorb aggregate cantonal shocks, the DDD cannot fully rule out Geneva-specific shocks that differentially affected high-bite sectors---for instance, if Geneva's hospitality experienced a unique tourism recovery. The late-period result (2022--2023, when COVID had dissipated) showing a similar null provides reassurance, but data on cantonal COVID stringency or tourism flows would strengthen the design.


\section{Conclusion}

Geneva's experiment---the highest minimum wage floor ever enacted---produced precisely estimated null effects on employment in the sectors where it bound most tightly. The 95\% confidence interval rules out employment declines exceeding 3\%. Firm entry, however, declined by 4\%, suggesting the minimum wage raises barriers for new low-wage businesses without displacing incumbents. These findings imply that the employment costs of minimum wages remain modest even at levels far beyond those studied in the existing literature, but that the competitive landscape may shift through entry deterrence rather than job destruction. Future work should examine price pass-through and worker mobility at these extreme wage floors.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
