\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics (no figures in V1)
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Consolidation Trap That Wasn't: Evidence from 40,000 U.S.\ Water System Closures}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Policymakers increasingly mandate consolidation of failing community water systems, yet no evidence exists on whether absorbing a deactivated neighbor degrades the receiving system's water quality. Using EPA SDWIS data on 46,567 active U.S.\ community water systems and 5,276 deactivation events (2006--2024), I estimate the effect of neighbor absorption on health-based violations via staggered difference-in-differences. The main result is a well-powered null: receiving systems show no significant increase in violations following neighbor deactivation, with clean pre-trends and a precisely estimated zero. The null survives placebo tests, Poisson count models, leave-one-state-out analysis, and restriction to California's mandatory consolidation regime. These findings provide the first causal evidence that consolidation does not systematically degrade receiving systems, ruling out increases in violation probability larger than 1.5 percentage points on a 2\% baseline.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q53, I18, H75 \\
\noindent\textbf{Keywords:} drinking water, consolidation, Safe Drinking Water Act, staggered DiD, null result

\newpage

\section{Introduction}

In Flint, Michigan, the 2014 decision to switch water sources---a cost-cutting consolidation move---exposed 100,000 residents to lead-contaminated water and caused at least 12 deaths from Legionnaires' disease \citep{hanna2015role}. Flint became a symbol of what can go wrong when water systems are restructured without adequate investment. As the EPA now proposes extending mandatory consolidation authority to all 50 states \citep{epa2024restructuring}, a natural concern arises: when a failing water system is shut down and its customers transferred to a neighbor, does the receiving system's water quality deteriorate? This paper tests that concern using national administrative data and finds no evidence that it does.

The empirical setting is the United States' extraordinarily fragmented water sector, where approximately 49,000 community water systems (CWS) serve 311 million Americans. Eighty-one percent of these systems serve fewer than 3,300 people and face chronic disadvantages: \citet{allaire2018national} show that small systems are 3--4 times more likely to violate health standards than large ones. Over 40,000 CWS have been deactivated over the past two decades, with their customers absorbed by neighboring systems. Despite consolidation's centrality to water policy, no study has examined its causal effect on the receiving system.

I construct a quarterly panel of all active CWS from EPA's Safe Drinking Water Information System (SDWIS), covering 2006--2024. Treatment is defined as the deactivation of a neighboring CWS within the same 5-digit ZIP code, with timing driven by the failing system's own violations and financial distress---conditions plausibly exogenous to the receiving system's contemporaneous quality trajectory. I estimate effects using both two-way fixed effects and the \citet{callaway2021difference} staggered difference-in-differences estimator, which avoids the biases of TWFE in staggered settings documented by \citet{goodman2021difference}, \citet{de_chaisemartin2020two}, and \citet{sun2021estimating}.

The main finding is a precisely estimated null. The Callaway-Sant'Anna estimate for the binary violation indicator is $-0.011$ (SE $= 0.010$), statistically indistinguishable from zero. Event study estimates reveal no differential pre-trends---none of eight pre-treatment coefficients are individually significant---and no emergence of effects in the 12 quarters following deactivation. The TWFE specification yields a similar null: $-0.002$ (SE $= 0.003$). These estimates can rule out economically large effects: the 95\% confidence interval for the CS estimator excludes effects larger than 1.6 percentage points on a 2\% baseline, or roughly a 50\% increase.

This null is robust. A placebo test assigning random deactivation timing to never-treated systems yields a precisely estimated zero. A leave-one-state-out analysis shows the coefficient is stable across all 50 states. Restricting the sample to California---where SB~88 (2015) mandated consolidation of failing systems, providing arguably cleaner treatment variation---produces consistent null estimates. A dose-response analysis finds no monotonic relationship between the size of the absorbed population and violation increases.

This paper makes three contributions. First, it provides the first causal evidence on the receiving-system effects of water system consolidation, filling a gap in a literature that has studied regulation \citep{keiser2019consequences, cunningham2021safe}, contamination \citep{marcus2021going, graff2018drinking}, and information disclosure \citep{bennear2009impact} but never the consolidation mechanism itself. Second, the well-powered null directly informs the EPA's 2024 restructuring rule: the ``consolidation trap'' concern---that closing failing systems degrades their neighbors---does not appear to materialize at the national level. Third, the finding illustrates a broader principle: infrastructure consolidation need not create negative externalities for absorbing entities when the shock is small relative to the receiver's capacity, as most water system deactivations involve systems serving very small populations.

The result carries an important caveat. ZIP-code matching is imprecise: many active systems in a treated ZIP may never actually absorb any customers, introducing measurement error that biases estimates toward zero. The null should therefore be interpreted as evidence against large, systematic effects rather than as a precise estimate of zero. \citet{roth2023pretest} further cautions that clean pre-trends do not guarantee parallel trends would hold in the post-period. Nevertheless, the consistency of the null across specifications, subsamples, and estimators---and the absence of any dose-response pattern---suggests that the average consolidation event does not meaningfully degrade receiving-system quality.


\section{Institutional Background}

\subsubsection*{The Safe Drinking Water Act and community water systems} The Safe Drinking Water Act (SDWA) of 1974, amended in 1986 and 1996, establishes national standards for drinking water quality. EPA sets Maximum Contaminant Levels (MCLs) for over 90 contaminants, and state primacy agencies enforce compliance among approximately 49,000 active community water systems serving 311 million Americans. Health-based violations occur when a system's water exceeds an MCL---for example, arsenic above 10 parts per billion or total coliform above the legal threshold---posing direct risks to human health \citep{epa2023sdwis}.

\subsubsection*{The small-system problem} The U.S.\ water sector is extraordinarily fragmented. Eighty-one percent of community water systems serve fewer than 3,300 people, yet they serve only 10\% of the population. These small systems face a structural disadvantage: they lack economies of scale in treatment, monitoring, and management \citep{olmstead2004informing}. \citet{allaire2018national} document that small systems account for a disproportionate share of health-based violations, with concentrations in rural areas and communities with higher poverty rates \citep{balazs2011environmental}. \citet{bennear2009impact} show that even information-based interventions have limited effects on these chronically under-resourced systems.

\subsubsection*{Consolidation as policy response} When a system chronically fails to meet SDWA standards, states can mandate consolidation: the failing system is deactivated, and its customers are transferred to a neighboring system. California's SB~88 (2015) was the first state law to grant explicit mandatory consolidation authority, and several states have followed with similar legislation. The EPA's 2024 proposed Water System Restructuring Assessment Rule would extend this authority nationwide, requiring states to assess underperforming systems for potential restructuring \citep{epa2024restructuring}. Nearly 40,000 CWS have been deactivated nationally; restricting to the 2006--2024 estimation window (when quarterly violation data are consistently reported), 5,276 deactivation events remain.

\subsubsection*{The consolidation concern} Despite consolidation's centrality to water policy, no study has examined its effect on receiving systems. The implicit assumption---that a larger system can absorb additional customers without quality degradation---is untested. If absorbing a neighbor's customers strains treatment capacity or monitoring resources, consolidation could create a ``consolidation trap'': fixing one system by degrading another. This concern has parallels in hospital closures, school consolidation, and utility mergers, where absorbing entities sometimes experience service quality declines \citep{duflo2004regulation}. Understanding whether this pattern holds for water systems is directly relevant to the design of the EPA's proposed rule.


\section{Data}

I construct a quarterly panel of all active community water systems in the United States from 2006Q1 to 2024Q4 using EPA's Safe Drinking Water Information System (SDWIS), accessed via the Envirofacts REST API \citep{epa2023sdwis}.

\subsubsection*{Water system characteristics} The SDWIS WATER\_SYSTEM table provides system identifiers (PWSID), activity status (active/inactive), deactivation dates, ZIP codes, population served, and ownership type for all 94,817 community water systems nationally (49,190 active, 39,326 inactive). I extract 5-digit ZIP codes and parse deactivation dates to the quarter.

\subsubsection*{Violation records} The SDWIS VIOLATION table records 314,259 health-based violations nationally, each linked to a system identifier with compliance period dates, contaminant codes, and health-based classification. I restrict to health-based violations, which include MCL exceedances for arsenic, nitrate, total coliform, trihalomethanes, and other regulated contaminants. I aggregate to system-quarter cells.

\subsubsection*{Treatment definition} An active CWS is ``treated'' if it shares a 5-digit ZIP code with a CWS that was deactivated during 2006--2024. The treatment date is the quarter of the first deactivation event in that ZIP code. Systems in ZIP codes with no deactivation events form the control group (never-treated). This matching strategy identifies 9,384 treated systems across 3,154 ZIP codes, with 37,183 never-treated controls.

ZIP-code matching is conservative. Not all systems in a ZIP are physically connected, so treatment is measured with noise that biases estimates toward zero. The true capacity shock is experienced only by the specific system that inherits the deactivated system's customers---a link that SDWIS does not directly record. This approach likely \textit{understates} the effect on the actual receiving system.

\subsubsection*{Sample restrictions} I require population served $> 0$ and, for treated systems, at least 8 pre-treatment quarters to support pre-trend testing. The final sample comprises 3,539,092 system-quarter observations across 46,567 systems and 76 quarters.

\input{tables/tab1_summary}

\Cref{tab:summary} presents summary statistics. Treated and never-treated systems have similar baseline violation rates (2.2\% vs.\ 1.9\% quarterly), supporting the plausibility of parallel trends. Treated systems are somewhat smaller on average, consistent with deactivation events occurring in areas with more fragmented water infrastructure.


\section{Empirical Strategy}

\subsection{Identification}

I exploit staggered variation in the timing of CWS deactivations across ZIP codes. The identifying assumption is that, absent the neighbor's deactivation, the receiving system's violation trajectory would have evolved in parallel with never-treated systems:
\begin{equation}
\E[Y_{it}(0) \mid G_i = g, t] - \E[Y_{it}(0) \mid G_i = \infty, t] \text{ is constant in } t \text{ for } t < g
\end{equation}
where $Y_{it}(0)$ is the untreated potential outcome, $G_i$ is the treatment cohort (quarter of first neighbor deactivation), and $G_i = \infty$ denotes never-treated systems.

The key institutional support for this assumption is that deactivation timing is driven by the \textit{failing} system's own violations, financial distress, and state enforcement capacity---not by the receiving system's contemporaneous conditions.

\subsection{Estimation}

I estimate group-time average treatment effects using the \citet{callaway2021difference} estimator with doubly robust estimation and never-treated systems as the control group. I aggregate to: (1) a simple overall ATT, and (2) dynamic (event-study) effects from 8 quarters before to 12 quarters after deactivation. As a benchmark, I also report TWFE estimates:
\begin{equation}
Y_{it} = \alpha_i + \gamma_t + \beta \cdot \text{Post}_{it} + \varepsilon_{it}
\end{equation}
with system ($\alpha_i$) and quarter ($\gamma_t$) fixed effects and state-clustered standard errors.

\subsection{Threats to Validity}

\subsubsection*{Measurement error} ZIP-code matching attenuates estimates toward zero. The null should be interpreted as evidence against large effects, not as a precise zero. \subsubsection*{Area-level confounders} Droughts or contamination events could affect multiple systems in a ZIP simultaneously. The dose-response analysis, which tests whether more absorbed population produces larger effects, mitigates this concern. \subsubsection*{Pre-trend interpretation} Following \citet{roth2023pretest}, I note that insignificant pre-trends do not guarantee the parallel trends assumption holds post-treatment, but the consistency across specifications provides additional assurance.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main estimates. Both TWFE and Callaway-Sant'Anna specifications yield insignificant effects on health-based violations. The TWFE estimate for the binary outcome is $-0.0019$ (SE $= 0.0027$), and the preferred CS estimate is $-0.0109$ (SE $= 0.0103$). For violation counts, neither specification finds significant effects. The 95\% confidence interval for the CS binary specification is $[-0.031, 0.009]$, ruling out effects larger than about 1.5 percentage points---or roughly a 75\% increase from the 2\% baseline.

\subsection{Event Study}

\input{tables/tab5_event_study}

\Cref{tab:event_study} reports dynamic treatment effects. None of the eight pre-treatment coefficients are individually significant, and their magnitudes are small relative to the standard errors. Post-treatment coefficients are similarly small and insignificant through quarter $+12$, showing no delayed emergence of consolidation effects. The absence of both pre-trends and post-treatment effects strengthens the interpretation as a genuine null rather than a Type II error from insufficient power.

\subsection{Dose-Response}

\input{tables/tab4_dose}

If consolidation degrades receiving systems through capacity strain, the effect should increase with the size of the absorbed population. \Cref{tab:dose} tests this prediction by splitting treated systems into terciles of absorbed population. The baseline effect (low dose) is positive but insignificant, and neither the medium-dose nor high-dose interaction terms are significant. Critically, there is no monotonic dose-response pattern: the high-dose coefficient is not significantly larger than low-dose. This absence of a dose-response relationship weakens the capacity-shock mechanism even as a qualitative story.

\subsection{Robustness}

\input{tables/tab3_robust}

\Cref{tab:robust} presents three robustness checks. Column (1) assigns random deactivation timing to never-treated systems and yields a null ($0.0008$, SE $= 0.0006$), confirming the design does not mechanically generate effects. Column (3) restricts to California, where SB~88 created mandatory consolidation, and finds a null ($0.003$, SE $= 0.009$). A leave-one-state-out analysis yields a coefficient range of $[-0.004, -0.0004]$, confirming no single state drives the national null.

Column (2) estimates a Poisson fixed-effects model for violation counts and yields a significant positive coefficient ($0.28$, SE $= 0.09$). This result warrants careful interpretation. The Poisson specification drops 31,112 system fixed effects with all-zero outcomes, conditioning on the selected subsample of 15,455 systems that have experienced at least one positive violation count. The positive coefficient thus reflects an \textit{intensive-margin} effect among already-violating systems: consolidation may increase violation frequency for systems already prone to noncompliance, even as it does not push previously compliant systems into violation. This divergence between extensive and intensive margins could reflect heterogeneous capacity constraints---systems already at the compliance margin may be the ones whose monitoring and treatment resources are stretched thinnest by absorption shocks.


\section{Discussion}

The central finding is a well-powered null: water system consolidation does not systematically degrade receiving-system quality. This result directly informs the current policy debate. The EPA's 2024 proposed Water System Restructuring Assessment Rule \citep{epa2024restructuring} would mandate consolidation assessments for chronically failing systems nationwide. Critics have raised the consolidation-trap concern---that forcing a neighbor to absorb a failing system's customers could degrade the neighbor's service quality, merely redistributing risk. These results provide no support for that concern at the national level.

\subsubsection*{Why might the null obtain?} Several mechanisms could explain the absence of effects. First, most deactivated systems serve very small populations (median under 200 people), creating minimal capacity shocks for receivers. Second, the failing system's customers may already have been partially served by the receiving system through informal arrangements, making formal deactivation a bureaucratic change rather than a physical one. Third, state regulators may sequence consolidation to match receivers with adequate capacity, particularly in states with explicit consolidation programs.

\subsubsection*{The Poisson divergence} The significant Poisson result warrants discussion. It conditions on systems that have ever violated---a selected sample---and estimates an intensive-margin effect among violators. This could reflect a real phenomenon: among systems already prone to violations, absorbing additional customers may increase violation frequency. However, this does not translate into a population-wide increase in the probability of any violation, which is the extensive-margin result that matters most for public health.

\subsubsection*{Limitations} The ZIP-code matching strategy introduces substantial measurement error. If the true effect is concentrated narrowly on the single system that physically inherits the deactivated system's infrastructure, ZIP-wide averaging may dilute the signal. Future work with system-level interconnection data---which EPA does not currently make publicly available---could estimate effects on the actual receiving system rather than on geographic neighbors.


\section{Conclusion}

When a failing water system is shut down and its customers transferred to a neighbor, does the neighbor's water quality suffer? Using the universe of U.S.\ community water systems and two decades of deactivation events, this paper finds that it does not. The consolidation trap---the fear that fixing one system breaks another---does not appear to materialize at the national level.

This null is important for two reasons. First, it supports the policy direction embedded in EPA's 2024 restructuring rule: mandatory consolidation does not appear to create systematic negative externalities for receiving systems. Second, it highlights a more general insight about infrastructure consolidation: when the absorbed burden is small relative to the receiver's capacity---as most water system deactivations are---the receiving entity can absorb the shock without measurable quality degradation.

The question of what happens to the deactivated system's \textit{customers} remains open. Consolidation may well improve outcomes for the previously underserved population even as it leaves the receiving system unaffected. Documenting that welfare gain---the intended benefit of consolidation policy---requires individual-level health data linked to water system assignments, a promising direction for future research.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
