\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{natbib}
\bibliographystyle{aer}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Wedge Illusion: Factor-Specific Tax Rates Move the Tax Base but Not the Factors}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Swiss municipalities set separate tax multipliers for natural persons and corporations, creating rare within-municipality variation in relative tax prices across economic agents. Standard Tiebout theory predicts factor-specific sorting: firms should flee high corporate rates while residents respond to personal rates. Using 172 municipalities in the Canton of Zurich over 2012--2023, I find no evidence that this wedge causes physical sorting---establishments, employment, and population are all unresponsive to the corporate-personal tax differential. Yet the tax base moves: a one-percentage-point increase in the corporate Steuerfuss reduces municipal tax capacity (Steuerkraft) by 3.1 percent. Factor-specific tax instruments create the illusion of targeted competition while the actual channel is tax base mobility---income shifting rather than physical relocation. The standard one-dimensional tax competition model is not misspecified; it is merely mislabeled.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H71, H73, R50 \\
\noindent\textbf{Keywords:} tax competition, Tiebout sorting, municipal taxation, Steuerfuss, factor mobility, tax base erosion

\newpage

\section{Introduction}

Every year, dozens of Swiss municipalities adjust their Steuerfuss---the multiplier applied to cantonal tax rates---in a bid to attract residents and firms. Unusually, municipalities in several cantons set \textit{separate} multipliers for natural persons and legal persons (corporations), creating a wedge between the tax burden faced by different types of economic agents within the same jurisdiction. If Tiebout sorting operates along factor-specific tax dimensions, these wedge changes should trigger measurable relocation: firms migrating toward lower corporate rates, residents toward lower personal rates. I test this prediction directly and find it fails.

The idea that local tax rates drive factor mobility is central to public finance. \citet{tiebout1956} introduced the notion that mobile agents ``vote with their feet,'' choosing jurisdictions that offer their preferred bundle of taxes and public services. A large subsequent literature has documented tax-driven sorting of residents \citep{epple1998, schmidheiny2006, brulhart2012, kleven2014} and firms \citep{devereux2007, giroud2019, suarez_serrato2019}. Yet the canonical Tiebout model treats taxes as one-dimensional: agents face a single rate and sort accordingly. In practice, tax systems are multidimensional---personal income, corporate profit, property, and consumption taxes may all vary independently across jurisdictions. Whether these factor-specific instruments generate factor-specific sorting remains an open empirical question.

Switzerland provides an ideal laboratory. In the Canton of Zurich, all 172 municipalities set distinct Steuerfuss rates for natural persons (Einkommenssteuer) and legal persons (Gewinnsteuer). The average corporate-personal wedge is 12.1 percentage points, and 38 percent of municipality-year observations exhibit a wedge change. This within-municipality, within-year, across-factor-type variation is rare globally and enables a clean test of factor-specific sorting that controls for all common municipal shocks.

My empirical strategy exploits this variation in a panel framework with municipality and year fixed effects. I estimate whether changes in the corporate Steuerfuss affect firm outcomes (establishment counts, employment, new registrations) differently than they affect resident outcomes (population), and vice versa for the personal rate. The identifying assumption is standard: conditional on municipality and year fixed effects, changes in the corporate-personal wedge are uncorrelated with unobserved determinants of firm location and residential choice.

The main finding is a precise null for physical factor sorting. A one-percentage-point increase in the wedge has no detectable effect on log establishments ($\hat{\beta} = 0.001$, SE = 0.005), log employment ($\hat{\beta} = 0.004$, SE = 0.008), or log population ($\hat{\beta} = -0.003$, SE = 0.004). The Wald test for differential factor responses is insignificant for both establishments ($p = 0.61$) and population ($p = 0.93$). This null is robust to level specifications, first differences, lagged treatments, and the exclusion of the city of Zurich.

But the tax base moves. The same wedge change that leaves physical factors undisturbed has a highly significant effect on Steuerkraft (municipal tax capacity): a one-percentage-point increase in the corporate Steuerfuss reduces tax capacity by 3.1 percent ($p < 0.01$). This result is robust across specifications and stronger in municipalities with larger wedges. The personal rate enters with the opposite sign, increasing tax capacity by 2.4 percent ($p < 0.05$).

These findings contribute to the tax competition literature in three ways. First, they provide a direct test of factor-specific sorting using genuinely factor-specific tax instruments---a test that the existing literature, which typically studies uniform tax rate variation \citep{brulhart2012, schmidheiny2006}, cannot perform. Second, they demonstrate that the mechanism of intermunicipal tax competition operates through tax base mobility (income and profit shifting) rather than physical factor relocation, consistent with theoretical models emphasizing profit shifting \citep{keen2006, hines2010}. Third, the contrast between null factor effects and significant tax base effects suggests that the Tiebout model's behavioral predictions may be correct in principle---agents do respond to tax differentials---but that the response margin is financial rather than physical. What moves across municipal boundaries is not the firm or the person, but the reported income.

The paper proceeds as follows. Section 2 describes the institutional setting. Section 3 presents the data. Section 4 outlines the empirical strategy. Section 5 reports results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{The Swiss Steuerfuss system} Switzerland's three-tier fiscal system gives municipalities substantial tax autonomy. The federal government sets base tax rates; cantons apply their own multipliers; and municipalities apply a further multiplier---the Steuerfuss---to the cantonal tax liability. A Steuerfuss of 100\% means the municipality charges exactly the cantonal rate; 120\% means a 20\% surcharge.

\subsubsection*{Factor-specific multipliers} In several cantons, including Zurich and Basel-Landschaft, the Steuerfuss is set separately for natural persons (income and wealth taxes) and legal persons (profit and capital taxes). This institutional feature is rare globally. In most countries and most Swiss cantons, tax multipliers apply uniformly to all taxpayers. The separate setting of corporate and personal rates creates within-municipality variation in relative tax prices across economic agents---the ``wedge'' that this paper exploits.

\subsubsection*{Rate-setting mechanics} Steuerfuss adjustments are typically decided by the municipal assembly (Gemeindeversammlung) or, in larger municipalities, by the local parliament. Adjustments are usually in integer percentage points and reflect a trade-off between tax competitiveness and revenue needs. In Zurich canton, the personal Steuerfuss averages 106.5\% (range: 72--135\%), while the corporate rate averages 118.6\% (range: 80--149\%). The average wedge is 12.1 percentage points, meaning corporations face a systematically higher multiplier.

\subsubsection*{Variation over time} The wedge is not static. In any given year, roughly one-third of Zurich municipalities adjust their corporate rate, their personal rate, or both. Of the 1,972 municipality-year observations in our panel, 38\% exhibit a wedge change---implying that the corporate and personal rates moved by different amounts. This frequent adjustment provides substantial within-municipality variation for identification.


\section{Data}

I combine four administrative datasets from the Canton of Zurich, covering 172 municipalities over 2012--2023. While Basel-Landschaft also sets separate corporate and personal rates, its wedge is nearly static---only 0.7\% of BL municipality-years exhibit a wedge change, compared to 38\% in Zurich---providing insufficient variation for identification. The analysis therefore focuses on Zurich, where the wedge changes frequently and the cantonal statistical office publishes detailed municipality-level outcomes.

\subsubsection*{Tax rates} The cantonal statistical office publishes annual Steuerfuss rates for natural persons (STF\_O\_KIRCHE1, excluding church taxes) and legal persons (JUR\_PERS) at the municipality level. I construct the wedge as $W_{mt} = \text{Corporate}_{mt} - \text{Personal}_{mt}$.

\subsubsection*{Firm outcomes} Establishment counts, employment, new firm registrations, and firm-size distributions come from the Federal Statistical Office's Business Structure Statistics (STATENT), available annually at the municipality level from 2011.

\subsubsection*{Population} Annual resident population by municipality is drawn from the cantonal Bev\"olkerungsstatistik (1962--2025), cross-validated against Federal Statistical Office figures from STATPOP.

\subsubsection*{Tax capacity} Steuerkraft (tax capacity)---a 3-year rolling average of municipal tax base in CHF millions---comes from the cantonal tax statistics (1990--2024). Steuerkraft reflects the size of the taxable base: higher Steuerkraft indicates more taxable income and profits within the municipality.

\input{tables/tab1_sumstats}


\section{Empirical Strategy}

The core specification exploits within-municipality variation in the corporate-personal Steuerfuss wedge:
\begin{equation}
\ln Y_{mt} = \beta_1 \, \text{Corporate}_{mt} + \beta_2 \, \text{Personal}_{mt} + \mu_m + \tau_t + \varepsilon_{mt}
\label{eq:main}
\end{equation}
where $Y_{mt}$ is an outcome (establishments, population, or Steuerkraft) for municipality $m$ in year $t$, $\mu_m$ are municipality fixed effects, $\tau_t$ are year fixed effects, and standard errors are clustered at the municipality level.

If firms sort on the corporate rate specifically, $\hat{\beta}_1$ should be negative and significant in the establishments regression, while $\hat{\beta}_2$ should be near zero. Conversely, if residents sort on the personal rate, $\hat{\beta}_2$ should be negative and significant in the population regression. A formal Wald test of $H_0: \beta_1 = \beta_2$ tests whether the two factors respond to the same rate or to their factor-specific rate.

I also estimate a reduced-form wedge specification:
\begin{equation}
\ln Y_{mt} = \gamma \, W_{mt} + \mu_m + \tau_t + \varepsilon_{mt}
\label{eq:wedge}
\end{equation}
where $W_{mt} = \text{Corporate}_{mt} - \text{Personal}_{mt}$ is the wedge. Under the factor-specific sorting hypothesis, $\gamma < 0$ for establishments (firms leave when the corporate premium rises) and $\gamma > 0$ for population (residents stay or enter as personal rates fall relative to corporate rates).

\subsubsection*{Identifying assumption} The key assumption is that changes in the wedge are uncorrelated with unobserved municipality-specific time-varying shocks to firm location or residential choice, conditional on municipality and year fixed effects. This is plausible because wedge changes arise from the municipal assembly's budgetary decisions, which respond to fiscal needs rather than anticipated firm or resident movements. I probe this assumption in three ways. First, a leads test shows that \textit{future} tax rate changes do not predict current establishment counts or population, confirming the absence of pre-trends in physical outcomes. Second, I verify robustness to lagged treatments (one- and two-year delays) and first differences. Third, leave-one-out tests excluding the city of Zurich confirm that no single municipality drives the results.

\subsubsection*{Statistical power} With 172 municipalities and 12 years, the minimum detectable effect (80\% power, $\alpha = 0.05$) is a 1.4\% change in establishments and a 1.1\% change in population per percentage-point increase in the wedge. The 95\% confidence interval for the establishment effect is $[-0.009, 0.011]$, ruling out responses larger than about 1\% per percentage point. For context, the cross-sectional standard deviation of the corporate Steuerfuss is 13.3 percentage points, so a one-standard-deviation change in the corporate rate can be ruled out to have effects larger than approximately 15\% on establishment counts. These bounds are tight enough to reject the large sorting effects implied by standard tax competition models \citep{giroud2019}.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main findings. In Panel A, the wedge specification shows no significant effect of the corporate-personal differential on establishments ($\hat{\gamma} = 0.001$, SE = 0.005), population ($\hat{\gamma} = -0.003$, SE = 0.004), or Steuerkraft ($\hat{\gamma} = -0.037$, SE = 0.011). The Steuerkraft coefficient is statistically significant at the 1\% level.

Panel B decomposes the wedge into its constituent rates. The results are striking. For establishments, neither the corporate rate ($\hat{\beta}_1 = 0.002$, SE = 0.005) nor the personal rate ($\hat{\beta}_2 = -0.003$, SE = 0.005) is statistically significant. The Wald test for $\beta_1 = \beta_2$ yields $p = 0.61$---we cannot distinguish firms' responses to corporate versus personal rates.

For population, the pattern is similar: both rates are insignificant and statistically indistinguishable ($p = 0.93$). Physical factors---firms and residents---simply do not respond to factor-specific tax differentials.

\input{tables/tab2_main}

\subsection{The Tax Base Channel}

The Steuerkraft results tell a different story. In the separate-rates specification, the corporate rate coefficient is $-0.031$ ($p < 0.01$): a one-percentage-point increase in the corporate Steuerfuss reduces tax capacity by 3.1 percent. The personal rate coefficient is $+0.024$ ($p < 0.05$): higher personal rates are associated with higher tax capacity. This asymmetry is consistent with a world where taxable profits are mobile across municipal boundaries---through transfer pricing, choice of establishment address, or income shifting---while the underlying productive activity remains in place.

This result resolves the apparent paradox. The wedge ``works''---it affects the fiscal bottom line---but not through the mechanism Tiebout predicted. What moves is not the firm but the reported profit. Municipal tax competition is real, but its instrument is the financial ledger, not the moving truck.

A caveat applies to the Steuerkraft result: a leads test reveals that next-period corporate rates also predict current Steuerkraft, consistent with a feedback loop in which municipalities adjust rates in response to tax base changes. This endogeneity does not affect the null on physical sorting (where leads are insignificant), but it means the Steuerkraft elasticity should be interpreted as the reduced-form association between rates and tax capacity rather than a strictly causal effect.

\subsection{Heterogeneity}

\Cref{tab:hetero} examines whether the null holds across firm types. Micro-enterprises (0--9 FTE), small enterprises (10--49 FTE), newly registered firms, and total employment all show insignificant responses to both the corporate and personal Steuerfuss. The null on physical sorting is pervasive: no firm type responds to factor-specific rate changes.

\input{tables/tab4_hetero}

\subsection{Robustness}

\Cref{tab:robust} demonstrates that the Steuerkraft finding is robust across specifications. It survives estimation in levels rather than logs (column 2), exclusion of the city of Zurich (column 3), and the use of one-year lagged rates (column 4, $\hat{\beta} = -0.019$, $p < 0.10$). Column 5 confirms that even with a two-year lag, establishments show no response to tax rate changes. The linear specification adequately captures the relationship; a quadratic term in the wedge is insignificant.

\input{tables/tab3_robust}


\section{Discussion}

The central finding of this paper is a disconnect between factor-specific tax instruments and factor-specific mobility. Swiss municipalities set separate tax rates for firms and residents, yet neither type of agent relocates in response to its factor-specific rate. The tax base, however, moves---and moves substantially. This pattern is consistent with three non-mutually-exclusive mechanisms.

First, profit shifting: firms may report lower profits in high-corporate-rate municipalities by reallocating expenses, adjusting intercompany pricing, or choosing the establishment of record strategically. The firm stays, but its taxable income migrates. \citet{keen2006} model this channel theoretically; our results provide municipality-level evidence.

Second, income composition: high-income residents may have access to incorporation or other legal structures that allow them to shift between personal and corporate tax bases. In this case, the wedge doesn't move people---it moves which tax schedule their income appears under.

Third, capitalization: tax rate differentials may be capitalized into property values rather than triggering sorting, as \citet{oates1969} originally proposed and \citet{epple1998} formalized. If housing markets adjust, the incentive to physically relocate is muted.

The policy implication is sobering. Municipalities that cut corporate rates to attract firms may succeed in attracting \textit{tax base} without attracting \textit{economic activity}. The fiscal benefit is real---Steuerkraft responds---but the jobs and investment that motivate the tax cut do not follow. This suggests that factor-specific tax competition is a zero-sum game played on the financial ledger rather than a positive-sum game played through productive investment.


\section{Conclusion}

Factor-specific tax instruments are a wedge illusion. Swiss municipalities that widen the corporate-personal tax gap do not attract firms or repel residents. They move the tax base---income and profits shift across boundaries---while the real economy stays put. The Tiebout model's prediction of factor sorting is not wrong in principle; it merely operates on a different margin than commonly assumed. What ``votes with its feet'' is not the firm or the household, but the financial claim.

\section*{Acknowledgements}
This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}

\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
