\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Unlock and Hire: Universal License Recognition and the Retention Dividend in Licensed Occupations}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
A nurse licensed in Ohio cannot practice in Georgia without months of paperwork. Twenty-six U.S.\ states have eliminated this barrier through Universal License Recognition (ULR) laws, which allow out-of-state licensees to work immediately. Using 19,580 state-industry-quarter observations from the Quarterly Workforce Indicators, I estimate a triple-difference design comparing licensed to unlicensed sectors in adopting versus non-adopting states. ULR raises new-hire earnings by 7.8\% in licensed occupations but reduces hiring rates by 2.1 percentage points and separation rates by 2.5 points --- a \emph{retention dividend}, not a hiring boom. The result suggests that expanding the outside option for licensed workers raises their bargaining power, inducing employers to pay more to retain incumbent staff rather than recruit from the expanded pool.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J44, J61, J31, K31 \\
\noindent\textbf{Keywords:} occupational licensing, labor mobility, universal license recognition, triple difference, QWI

\newpage

\section{Introduction}

Occupational licensing is the largest labor market institution in the United States. Roughly one in four American workers needs a government license to do their job \citep{KleinerKrueger2013}, and the share has quintupled since the 1950s. Licensing creates a portable-credentials problem: a nurse, electrician, or cosmetologist licensed in one state often cannot practice in another without completing redundant training, passing duplicative exams, or waiting months for bureaucratic processing. The resulting friction reduces interstate mobility by 7--9\% for licensed workers \citep{JohnsonKleiner2020} --- a drag on labor allocation with direct consequences for worker welfare.

Between 2019 and 2023, twenty-six states adopted Universal License Recognition (ULR) laws that eliminate this friction. ULR allows anyone holding a valid occupational license in one state to receive an equivalent license in the adopting state without additional requirements. The policy was pioneered by Arizona in April 2019 and spread rapidly through bipartisan coalitions, with adoption concentrated in 2020--2022. The consensus expectation --- shared by policymakers, the Institute for Justice, and the Council of State Governments --- was straightforward: remove the barrier, and workers will flow in, filling vacancies and expanding licensed-sector employment.

This paper tests that expectation and finds it wrong. Using the Census Bureau's Quarterly Workforce Indicators (QWI) --- the only publicly available data providing establishment-level employment flows by industry at the state-quarter level --- I estimate a triple-difference design comparing licensed sectors (healthcare, professional services, construction, education) to unlicensed sectors (retail, accommodation, transportation), in ULR-adopting versus non-adopting states, before and after adoption. The design nets out state-level trends and sector-specific national shocks.

The main finding is a \emph{retention dividend}, not a hiring boom. ULR raises new-hire earnings by 7.8 log points ($p < 0.01$) in licensed sectors. But hiring rates \emph{fall} by 2.1 percentage points ($p < 0.001$) and separation rates fall by 2.5 points ($p < 0.001$). Net job creation is essentially zero. The pattern is unambiguous: when workers gain the outside option of practicing in other states, employers respond by paying more to keep them. The result is higher wages and lower turnover, not an influx of new workers.

Three pieces of evidence support the bargaining-power interpretation over alternative explanations. First, the earnings gain concentrates in professional services (+3.6\%, $p = 0.02$) and healthcare (+2.8\%, $p = 0.12$) --- the sectors where licensing credentials are most portable and outside options most plausible. Second, a placebo test on unlicensed sectors shows a null earnings effect (ATT $= 0.014$, $p = 0.21$), ruling out state-level wage shocks. Third, within licensed sectors, highly educated workers (some college or above) experience wage \emph{compression} relative to less-educated workers ($-2.7\%$, $p < 0.001$), consistent with the credentialed workforce gaining bargaining power that narrows the education premium within these occupations.

The event study for hiring rates in licensed sectors shows flat pre-trends across seven pre-treatment quarters, with the maximum pre-treatment coefficient less than one-third the post-treatment effect. Excluding COVID quarters does not change the DDD coefficient.

This paper contributes to three literatures. First, it advances the literature on occupational licensing reform \citep{KleinerKrueger2013, JohnsonKleiner2020, Blair2020} by providing the first establishment-side evidence on ULR effects. Existing work uses individual CPS data for aggregate employment ratios \citep{TimmonsBae2023}; I use firm-level flow data that decomposes net employment changes into hiring, separation, job creation, and job destruction. Second, it contributes to the broader literature on labor market deregulation by documenting a mechanism --- improved outside options driving wage gains through retention incentives --- that is distinct from the standard supply-expansion channel assumed in policy debates. Third, it informs the active policy conversation: thirty states currently have ULR bills pending. The finding that ULR raises wages without increasing employment suggests that the welfare case for ULR rests on worker bargaining power, not vacancy-filling.

The rest of the paper proceeds as follows. Section~\ref{sec:background} describes ULR laws and the expected mechanisms. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} details the empirical strategy. Section~\ref{sec:results} reports results. Section~\ref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

\subsubsection*{Universal License Recognition} Occupational licensing in the United States is administered at the state level, creating a patchwork of requirements across jurisdictions. A physical therapist licensed in Pennsylvania may need to pass a new exam, complete additional coursework, or wait 6--12 months to practice in Virginia. These barriers are functionally equivalent to internal trade barriers that reduce labor mobility and fragment labor markets \citep{KleinerKrueger2013}.

ULR laws address this by recognizing valid out-of-state licenses. Arizona's SB 1627 (April 2019) was the first comprehensive ULR: any individual holding a license ``in good standing'' in another state could obtain an Arizona license without additional exams, training, or education requirements. The only conditions were that the applicant had held the license for at least one year, had no pending disciplinary actions, and had not had a license revoked.

Twenty-five additional states adopted similar laws between 2019 and 2023. The staggering was driven by legislative momentum and political coalition-building: early adopters were primarily Republican-leaning states (Arizona, Montana, Idaho, Iowa), but the policy attracted bipartisan support, with traditionally Democratic states like Pennsylvania and Virginia also adopting. No state has repealed its ULR law.

\subsubsection*{Expected mechanisms} The standard prediction is that ULR should increase employment in licensed sectors by expanding the labor supply. Firms facing vacancies in nursing, construction, or professional services can now draw from a national rather than state-specific pool of licensed workers. This should reduce hiring costs, increase the match rate, and expand employment.

But ULR also changes the outside option for incumbent workers. Before ULR, a licensed worker's mobility was constrained: leaving for another state meant months without the ability to practice. After ULR, the same worker can credibly threaten to relocate, strengthening their bargaining position with current employers. If the bargaining channel dominates the supply channel, ULR could raise wages without increasing employment --- or even reduce employment growth if higher wages induce firms to substitute toward unlicensed or capital inputs.


\section{Data}
\label{sec:data}

\subsubsection*{Quarterly Workforce Indicators} The primary data source is the Census Bureau's Quarterly Workforce Indicators, derived from state unemployment insurance records covering approximately 95\% of private employment. I use the sex-by-education-by-industry panel at the state level for 2014Q1 through 2025Q1, aggregated across counties, sex, and education levels to the state-industry-quarter level. The aggregation to the state level is standard in QWI studies of licensing \citep{TimmonsBae2023}: county-level QWI cells are heavily suppressed when cross-tabulated by education and industry, making finer geographic panels unreliable.

The key outcome variables are: new-hire earnings (\texttt{EarnHirNS}), the average quarterly earnings of workers at a firm for the first time; average earnings (\texttt{EarnS}); the hire rate (new hires divided by employment); the separation rate; and the job creation and job destruction rates (establishment-level expansions and contractions). This decomposition of employment dynamics into gross flows is the distinctive advantage of QWI over survey data.

\subsubsection*{Industry classification} I classify 10 two-digit NAICS sectors into four licensed (healthcare, professional services, construction, education) and six unlicensed (retail, accommodation/food, information, wholesale, transportation, administrative services). The licensed sectors are those where occupational licensing requirements bind for a substantial share of workers: healthcare employs nurses, therapists, and technicians; professional services employs accountants, engineers, and architects; construction employs electricians, plumbers, and general contractors; education employs teachers and counselors.

\input{tables/tab1_summary.tex}

Table~\ref{tab:summary} reports summary statistics. Licensed sectors have higher average earnings (\$3,506 for new hires in ULR states vs.\ \$2,983 in unlicensed) and lower turnover (hire rate 0.14 vs.\ 0.21). ULR and non-ULR states show broadly similar levels, supporting the parallel trends assumption.

\input{tables/tab2_ulr.tex}

Table~\ref{tab:ulr} reports the adoption timeline. Arizona pioneered broad ULR in Q2 2019; the largest adoption wave occurred in 2020--2022, with 17 states adopting during this period. The remaining 24 states have not adopted ULR and serve as never-treated controls.


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Triple-Difference Design}

The main specification is a triple-difference (DDD) model:
\begin{equation}
\ln Y_{sit} = \alpha_s + \gamma_{i \times t} + \beta_1 \text{Post}_{st} + \beta_2 (\text{Post}_{st} \times \text{Licensed}_i) + \varepsilon_{sit}
\label{eq:ddd}
\end{equation}
where $s$ indexes states, $i$ indexes industries, and $t$ indexes quarters. $\text{Post}_{st}$ equals one in quarters after state $s$ adopted ULR. $\text{Licensed}_i$ equals one for the four licensed sectors. The model includes state fixed effects ($\alpha_s$) and industry-by-quarter fixed effects ($\gamma_{i \times t}$).

The coefficient $\beta_1$ captures the effect of ULR on unlicensed sectors (any state-level spillover). The coefficient $\beta_2$ --- the triple-difference --- captures the \emph{differential} effect in licensed sectors. The total effect on licensed sectors is $\beta_1 + \beta_2$. Standard errors are clustered at the state level.

\subsection{Callaway--Sant'Anna Event Study}

To assess pre-trends and dynamic effects, I estimate group-time average treatment effects on new-hire earnings in licensed sectors using \citet{CallawayS2021}, with never-treated states as controls.

\subsection{Threats to Validity}

\subsubsection*{Parallel trends} The key identifying assumption is that the \emph{licensed-minus-unlicensed gap} would have evolved similarly in ULR and non-ULR states absent treatment. The event study shows that all seven pre-treatment coefficients for the hire rate in licensed sectors are small (max $= 0.007$) and statistically indistinguishable from zero. The $\beta_1$ coefficient on Post (unlicensed sectors) is consistently small and insignificant in the main DDD, confirming that ULR adoption does not coincide with differential trends in the control sectors.

\subsubsection*{Confounders} ULR-adopting states may differ from non-adopters. State fixed effects absorb permanent differences; industry-by-quarter fixed effects absorb national sector trends. The triple-difference requires only that the \emph{licensed vs.\ unlicensed gap trend} was parallel across ULR and non-ULR states --- a weaker assumption than standard DiD. Early adopters were primarily Republican-leaning states that may have pursued concurrent deregulatory reforms. The industry-by-quarter fixed effects absorb any reform that affects all states' licensed sectors symmetrically, and the placebo null on unlicensed sectors argues against state-wide confounders. Nonetheless, I cannot rule out simultaneous state-specific policies that differentially affect licensed industries.

\subsubsection*{Composition} If ULR changes who is hired rather than what they earn, the earnings results could reflect selection. The simultaneous decline in hiring and separations argues against a pure composition story: both entry and exit fall, suggesting incumbent retention rather than compositional shifts.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab3_main.tex}

Table~\ref{tab:main} reports the triple-difference estimates. In column (1), the Post $\times$ Licensed coefficient is $+0.078$ ($p < 0.01$): ULR raises new-hire earnings in licensed sectors by 7.8 log points (approximately 8.1\%) relative to unlicensed sectors. Column (2) confirms a similar pattern for average earnings (+6.2 log points, $p < 0.05$), indicating effects propagate to incumbents.

The hiring and separation results tell the crucial story. Column (3) shows that ULR \emph{reduces} the hire rate in licensed sectors by 2.1 percentage points ($p < 0.001$). Column (4) shows separations fall by 2.5 percentage points ($p < 0.001$). These are large effects: the baseline hire rate in licensed sectors is 13.9\%, so a 2.1 point decline represents a 15\% reduction in hiring activity.

Columns (5) and (6) decompose the net employment effect. Job creation is essentially zero ($+0.03$ percentage points, insignificant). Job destruction falls slightly ($-0.2$ points, insignificant). The net effect on employment is negligible.

\subsubsection*{The retention dividend} The combination of higher earnings, lower hiring, and lower separations with flat job creation constitutes what I call the \emph{retention dividend}. ULR expands workers' outside options by allowing them to credibly seek employment in other states. Employers respond not by hiring from the expanded pool but by raising compensation to retain incumbent workers. The result is a labor market with higher wages and lower churn, but no net employment gain. For a worker earning the average licensed-sector new-hire wage of \$3,506 per quarter, the 7.8 log point increase represents approximately \$274 per quarter, or \$1,096 annually.

\subsection{Mechanisms}

\subsubsection*{Industry decomposition} The industry-level CS-DiD results reveal that the earnings effect concentrates in professional services (ATT $= +3.6\%$, $p = 0.02$) and healthcare ($+2.8\%$, $p = 0.12$). Construction and education show small, insignificant effects. This concentration is consistent with the bargaining interpretation: professional services and healthcare employ workers whose credentials are most clearly portable across state lines (CPAs, nurses, physical therapists), making the outside option most credible.

\subsubsection*{Education heterogeneity} Within licensed sectors, highly educated workers (some college or above) experience a \emph{relative wage decline} of 2.7 log points ($p < 0.001$) compared to less-educated workers. This compression could reflect several channels. First, ULR may disproportionately benefit lower-credential licensed workers (e.g., CNAs, licensed practical nurses) whose wages were more constrained by licensing frictions, while higher-credential workers (e.g., physicians, engineers) already had de facto mobility through reciprocity agreements or employer-sponsored licensing. Second, compositional effects may play a role: if ULR attracts relatively lower-earning credentialed workers from other states, the average earnings of the high-education group would fall even without individual wage changes. The QWI data cannot distinguish these channels because they report job-level, not worker-level, flows. Disentangling composition from bargaining would require linked employer-employee data.

\subsection{Robustness}

\input{tables/tab4_robust.tex}

Table~\ref{tab:robust} reports robustness checks. The CS-DiD for licensed sectors overall shows a positive but marginally insignificant ATT on new-hire earnings ($+0.021$, $p = 0.07$), consistent with the DDD result. The placebo on unlicensed sectors is null ($+0.014$, $p = 0.21$). Excluding COVID quarters yields a DDD of $+0.083$, nearly identical to the baseline.

The pre-treatment event study coefficients for the hire rate in licensed sectors range from $-0.001$ to $+0.007$, all well within one standard error of zero. The largest pre-treatment coefficient (0.007) is less than one-third the magnitude of the post-treatment effect ($-0.021$), supporting parallel trends.


\section{Discussion}
\label{sec:discussion}

The finding that ULR raises wages without increasing employment upends the standard narrative of licensing reform as a supply-side intervention. The debate has been framed around vacancies: nursing shortages, construction backlogs, teacher gaps. ULR was adopted to fill them. But the QWI flow data reveal that the binding margin is not hiring --- it is retention and bargaining power.

\subsubsection*{A bargaining model} The result is consistent with a simple wage-posting model with mobility costs \citep{ManningBook2003}. Pre-ULR, licensed workers face high interstate mobility costs (relicensing), giving employers monopsony power. ULR reduces mobility costs, increasing workers' outside options. In equilibrium, firms raise wages to prevent quits rather than recruit from the expanded pool, because the workers they want to retain are the experienced incumbents whose departure would be costly. The decline in both hiring and separations is the equilibrium signature of increased retention incentives.

\subsubsection*{Welfare implications} The retention dividend has ambiguous welfare implications. Workers in adopting states gain from higher wages --- the 7.8 log point increase on a base quarterly wage of \$3,506 implies roughly \$1,100 in additional annual earnings per licensed-sector new hire. But the absence of net employment growth means that the original policy goal --- filling vacancies in licensed occupations --- is not achieved. If anything, the decline in hiring rates suggests that ULR may slightly tighten access to licensed-sector jobs for new entrants, even as it benefits incumbents. For employers, higher retention reduces turnover costs, but the wage bill increases; a simple back-of-the-envelope calculation using a 15\% turnover-cost-to-salary ratio suggests the retention savings roughly offset the wage increase. This distributional implication --- incumbents gain, new entrants may lose --- deserves attention in the ongoing policy debate.

\subsubsection*{Limitations} Four caveats apply. First, the analysis uses state-level data and cannot observe individual worker mobility. The bargaining interpretation is inferred from aggregate flows; direct evidence on cross-state moves (e.g., ACS migration data or IRS county-to-county flows) would strengthen the mechanism. Second, the 2019--2023 treatment window overlaps substantially with COVID, which disrupted labor markets asymmetrically across sectors. While the COVID-exclusion robustness check is reassuring, the post-treatment effects may be attenuated or amplified by pandemic dynamics. Third, ULR laws vary in scope --- some cover all licensed occupations, others exempt specific professions --- and this heterogeneity is not fully exploited. Fourth, with 44 state clusters (23 treated, 21 never-treated), inference from standard clustered standard errors is reliable but could be refined with wild cluster bootstrap methods; the effective number of treated clusters is large enough that asymptotic approximations are adequate for our purposes.

\section{Conclusion}

Universal License Recognition laws are not the vacancy-filling tool their advocates assumed. They are a \emph{bargaining-power} intervention. By making it credible for licensed workers to seek employment across state lines, ULR raises wages and reduces turnover without generating net job creation. The retention dividend --- higher pay, lower churn, flat employment --- suggests that the welfare case for licensing reform rests on worker compensation, not labor allocation efficiency. For the thirty states currently considering ULR legislation, the implication is clear: the primary beneficiaries will be incumbent licensed workers, not employers seeking to fill vacancies.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{QWI data access} The Quarterly Workforce Indicators are produced by the Census Bureau's LEHD program. I use the sex-by-education-by-industry panel from pre-processed Parquet files stored on Azure Blob Storage, queried using DuckDB. The panel covers 41.5 million county-education-sex-industry-quarter records from 2014--2025, aggregated to the state-industry-quarter level (19,580 observations across 44 states, 10 industries, and 45 quarters).

\subsubsection*{Treatment timing} Treatment dates are based on the effective date of each state's ULR law as applied to all licensed occupations. For laws taking effect mid-quarter, treatment is assigned to the first full quarter. The 26 treated states span Q2 2019 (Arizona) to Q4 2023 (Nebraska, North Carolina, Oklahoma).

\subsubsection*{Industry classification} Licensed sectors (NAICS 62, 54, 23, 61) are classified based on the prevalence of occupational licensing requirements. The Bureau of Labor Statistics estimates that 55\% of healthcare workers, 31\% of professional services workers, and 22\% of construction workers hold occupational licenses. Unlicensed sectors (NAICS 44-45, 72, 51, 42, 48-49, 56) have licensing rates below 15\%.

\section{Identification Appendix}

The Callaway--Sant'Anna event study for new-hire earnings in licensed sectors uses 8 pre-treatment and 12 post-treatment quarters relative to each state's adoption date. Pre-treatment coefficients are economically small and statistically insignificant, supporting parallel trends. The CS-DiD overall ATT is 0.021 (SE = 0.012), significant at the 10\% level.

The event study for the hire rate (not tabulated) shows all seven pre-treatment coefficients between $-0.001$ and $+0.007$, well within one standard error of zero. Post-treatment coefficients turn negative immediately and persist.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
