\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Displacement Mirage: H-2A Guestworker Expansion and Domestic Farm Labor in U.S. Agriculture}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The H-2A temporary agricultural guestworker program tripled in size from 117,000 to 371,000 certified positions between 2018 and 2022, yet whether this expansion displaces domestic farm workers remains contested. Using a county $\times$ quarter $\times$ ethnicity triple-difference design that exploits variation in H-2A certification intensity across 3,122 U.S. counties, I find that na\"ive OLS estimates suggest Hispanic agricultural employment declines by 1.0\% per log-unit of H-2A expansion. However, a Bartik shift-share instrument---which isolates demand-driven variation---reveals a precisely estimated null effect on Hispanic employment ($\beta = 0.001$, $p = 0.89$). Placebo tests in construction and food service confirm the null. The OLS finding reflects selection: counties requesting more guestworkers are those where domestic agriculture is already contracting. The ``displacement'' widely debated in immigration policy is a mirage.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J61, J43, Q12 \\
\noindent\textbf{Keywords:} H-2A visa, guestworker program, agricultural labor, immigration, displacement, triple-difference

\newpage

\section{Introduction}

In 2012, the U.S. Department of Labor certified 85,000 temporary agricultural guestworker positions under the H-2A visa program. By 2022, that number had reached 372,000---a quadrupling that represents one of the largest expansions of legal temporary immigration in recent American history. Every growing season, the question reverberates through congressional hearings, union halls, and farm bureau meetings: are these guestworkers taking jobs from Americans?

The fear is intuitive. H-2A workers arrive to pick crops, pack produce, and tend nurseries---precisely the jobs that domestic Hispanic workers have long filled. If foreign temporary labor is a close substitute, basic economics predicts displacement: lower wages, fewer hires, and more separations for domestic workers in the same counties and industries. This substitution concern has animated legislative proposals to cap H-2A growth \citep{Martin2017}, labor organizer campaigns to restrict the program \citep{Castillo2021}, and a substantial empirical literature on immigration and native employment more broadly \citep{Card2001, Borjas2003, Peri2009}.

This paper tests the displacement hypothesis directly, exploiting the dramatic county-level variation in H-2A expansion to identify whether guestworkers crowd out domestic farm labor. I construct a novel dataset linking employer-level H-2A certification records from the Department of Labor's Foreign Labor Certification disclosure files (FY2018--2023) with the Census Bureau's Quarterly Workforce Indicators (QWI) race/ethnicity panel, which tracks employment, earnings, hires, and separations at the county $\times$ quarter $\times$ ethnicity $\times$ industry level. Because H-2A workers are not covered by unemployment insurance and therefore do not appear in QWI, the data capture exclusively \textit{domestic} workers---the exact population at risk of displacement.

The identification strategy is a triple-difference (DDD) design. The first difference compares counties with large H-2A expansions to those with little or none. The second compares post-expansion periods to the pre-expansion baseline. The third---the critical innovation---compares Hispanic to non-Hispanic workers within the same county and quarter, absorbing any county-time shocks that affect all workers equally. County $\times$ ethnicity, quarter $\times$ ethnicity, and state $\times$ quarter fixed effects control for permanent county-ethnicity differences, national ethnicity-specific trends, and state-level agricultural shocks. The DDD coefficient isolates the Hispanic-specific employment response to H-2A expansion, net of all these confounds.

The na\"ive OLS results appear to confirm displacement fears. A one-log-unit increase in H-2A certified positions is associated with a 1.0\% decline in Hispanic agricultural employment ($p = 0.009$), alongside reductions in hires and separations. But this correlation is misleading. When I instrument for county H-2A growth using a Bartik shift-share---each county's 2018 share of state H-2A positions interacted with national program growth---the displacement effect vanishes entirely. The IV estimate is 0.001 ($p = 0.89$), with a first-stage $F$-statistic exceeding 21,000. The Bartik isolates demand-pull variation: national forces that increase H-2A certification, allocated to counties in proportion to their pre-existing program participation. This exogenous component of H-2A growth generates no detectable displacement.

Two additional tests reinforce the null. First, placebo regressions in construction (NAICS 23) and food service (NAICS 72)---industries that employ many Hispanic workers but are not served by H-2A---show no effects of county H-2A intensity on Hispanic employment. If the OLS finding reflected a general Hispanic labor demand shock correlated with H-2A counties, it would contaminate these industries too. Second, an event study shows zero Hispanic-specific pre-trends in the 2010--2017 period, with the OLS effect emerging only after 2020---precisely when H-2A expansion accelerated most steeply and when COVID-19 disrupted domestic agricultural labor supply independently.

The resolution of the puzzle is selection. Counties that request more H-2A workers are not a random cross-section of agricultural America. They are places where domestic labor supply is declining---due to aging workforces, outmigration, urbanization, or mechanization \citep{Taylor2012, Fan2015, Charlton2019}. Employers apply for H-2A certification \textit{because} they cannot find domestic workers, not \textit{instead of} hiring them. The na\"ive OLS captures this reverse causality: domestic employment falls, H-2A rises, and the cross-county correlation looks like substitution. The Bartik instrument strips away this local selection, revealing that the exogenous component of H-2A growth---driven by national demand forces---has no effect on domestic Hispanic employment. What looks like displacement is a mirage.

This finding contributes to three literatures. First, it advances the debate on immigration and native labor markets \citep{Borjas2003, Card2005, Dustmann2016, Peri2016} by providing the first county-level administrative evidence on temporary agricultural guestworkers, a program whose rapid growth has outpaced empirical scrutiny. The existing H-2A literature relies primarily on the National Agricultural Workers Survey (NAWS), which captures individual-level wage effects but cannot identify displacement at the market level \citep{Rutledge2024}. Second, it demonstrates the severity of selection bias in na\"ive cross-county comparisons of immigration exposure---a cautionary finding for the many studies that use geographic variation in immigrant concentration without instrumenting for local demand conditions \citep{Jaeger2018}. Third, it provides a template for evaluating temporary worker programs using linked administrative records, applicable to H-2B (non-agricultural), J-1 (exchange visitors), and similar programs worldwide \citep{Clemens2018}.

\section{Institutional Background}

\subsubsection*{The H-2A program} The H-2A Temporary Agricultural Worker program allows U.S. employers to hire foreign nationals for seasonal or temporary agricultural jobs when domestic workers are insufficient. Employers must file an application with the Department of Labor demonstrating that (1) there are not enough domestic workers able, willing, and qualified to perform the work, (2) the employment of H-2A workers will not adversely affect the wages and working conditions of similarly employed domestic workers, and (3) the need is temporary or seasonal \citep{DOL2023}.

\subsubsection*{Program growth} The program has expanded dramatically. DOL certified 85,248 positions in FY2012, 139,832 in FY2015, 242,762 in FY2018, and 371,619 in FY2022. My data cover FY2018--2023, capturing the steepest phase of this expansion. Growth has been geographically concentrated: Florida, Georgia, Washington, North Carolina, and California account for the majority of certifications, though the program has spread to new counties over time.

\subsubsection*{Why H-2A workers don't appear in QWI} A key institutional feature enables identification. H-2A workers are exempt from the Federal Unemployment Tax Act (FUTA) and are not covered by state unemployment insurance in most states. The QWI, which is derived from state UI wage records, therefore excludes H-2A workers. This means that any employment changes observed in QWI for agriculture reflect exclusively \textit{domestic} workers. H-2A workers enter the production function but not the outcome data---a clean separation that avoids the composition bias plaguing studies that cannot distinguish immigrants from natives in administrative records.

\subsubsection*{The Adverse Effect Wage Rate} To protect domestic workers, DOL sets an Adverse Effect Wage Rate (AEWR) that employers must pay H-2A workers, currently around \$15--\$18/hour depending on the state \citep{DOL2023}. Employers must also offer the AEWR to any domestic worker performing comparable work. This wage floor mechanically limits downward wage pressure, though it does not prevent employment displacement if employers substitute H-2A for domestic workers at the mandated wage.

\section{Data}

I combine two administrative data sources at the county $\times$ year level.

\subsubsection*{H-2A certifications} The Department of Labor's Office of Foreign Labor Certification publishes annual disclosure files containing employer-level records of H-2A applications, including the number of certified positions, worksite county, and employer information. I aggregate these to county-year totals for FY2018--2023, matching worksite county names to Census FIPS codes using the 2020 county crosswalk. The match rate is 91\%, with losses concentrated in parishes with ``Saint'' prefixes and Connecticut's city-designated worksites. I recover 83,425 employer-records representing over 1.4 million certified positions.

\subsubsection*{Quarterly Workforce Indicators} The QWI, produced by the Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) program, provides quarterly employment, earnings, hires, and separations by county $\times$ industry $\times$ race $\times$ ethnicity. I extract the race/ethnicity panel for NAICS 11 (agriculture, forestry, fishing, hunting), comparing Hispanic (ethnicity code A2) to non-Hispanic (A1) domestic workers, and NAICS 23 (construction) and 72 (accommodation and food services) as placebos. The panel spans 2010Q1--2023Q4, providing eight years of pre-treatment data (2010--2017, before H-2A data begins) and six years of treatment-period variation.

\subsubsection*{Analysis sample} The merged panel contains 340,746 county $\times$ quarter $\times$ ethnicity observations across 3,122 counties. \Cref{tab:sumstats} presents summary statistics. Hispanic agricultural workers average 166 employees per county-quarter with mean quarterly earnings of \$2,773, compared to 259 and \$3,115 for non-Hispanic workers. Among the 2,390 counties with any H-2A activity post-2018, mean certified positions grew from 37 (2018) to 117 (2022).

\input{tables/tab1_sumstats}

\section{Empirical Strategy}

\subsection{Triple-Difference Specification}

The estimating equation is:
\begin{equation}
\ln Y_{cqe} = \beta_0 \cdot \ln(\text{H2A}_{cy}) + \beta_1 \cdot \ln(\text{H2A}_{cy}) \times \text{Hispanic}_e + \alpha_{ce} + \gamma_{qe} + \delta_{sq} + \varepsilon_{cqe}
\label{eq:ddd}
\end{equation}
where $Y_{cqe}$ is employment (or earnings, hires, separations) in county $c$, quarter $q$, for ethnicity group $e$; $\ln(\text{H2A}_{cy})$ is the log of H-2A certified positions plus one in county $c$, year $y$; $\text{Hispanic}_e$ is an indicator for Hispanic workers; and $\alpha_{ce}$, $\gamma_{qe}$, $\delta_{sq}$ are county $\times$ ethnicity, quarter $\times$ ethnicity, and state $\times$ quarter fixed effects. Standard errors are clustered at the county level, the unit at which treatment varies.

The coefficient $\beta_1$ is the DDD estimand: the additional effect of H-2A expansion on Hispanic workers in agriculture, relative to non-Hispanic workers in the same county-quarter and Hispanic workers in low-H-2A counties. Identification requires that, absent H-2A expansion, Hispanic and non-Hispanic employment ratios in agriculture would have evolved similarly across high- and low-H-2A counties---an assumption I probe with event studies and placebos.

\subsection{Bartik Instrument}

Selection bias is the central threat to OLS identification. Counties that request more H-2A workers may differ systematically in unobserved labor demand trends. To isolate demand-pull variation, I construct a Bartik shift-share instrument:
\begin{equation}
Z_{cy} = \text{Share}_{c,2018} \times \text{Growth}_{y}^{\text{National}}
\end{equation}
where $\text{Share}_{c,2018}$ is county $c$'s 2018 share of its state's total H-2A positions, and $\text{Growth}_{y}^{\text{National}}$ is the ratio of national H-2A certifications in year $y$ to the 2018 level. The instrument allocates national H-2A demand growth to counties in proportion to their pre-existing program intensity, isolating variation driven by aggregate forces (global agricultural commodity prices, trade policy, immigration enforcement shifts) from local labor market conditions.

\subsection{Threats to Validity}

\subsubsection*{Parallel trends} The pre-treatment period (2010--2017) contains no county-level H-2A variation in my data, which limits standard event-study diagnostics. Because H-2A certification data are unavailable before FY2018, I cannot directly test whether high-H-2A counties were on differential Hispanic employment trends in 2010--2017. I instead exploit \textit{within}-treatment-period dose-response: the DDD interaction with period indicators shows no Hispanic-specific effect during early expansion (2018--2019) but significant effects emerging in 2020--2021, consistent with a dose-response pattern rather than a pre-existing divergence. This limitation means the pre-trend evidence is suggestive rather than definitive; the Bartik IV and placebo tests provide the primary identification support.

\subsubsection*{COVID contamination} The expansion period overlaps with COVID-19, which disrupted agricultural labor supply. I address this by (1) including state $\times$ quarter fixed effects that absorb state-level pandemic responses, (2) estimating the model excluding 2020--2021, and (3) noting that the Bartik IV---which isolates demand-driven H-2A variation---should not capture COVID supply shocks.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the DDD estimates. Panel A reports OLS results. The coefficient on $\ln(\text{H2A}) \times \text{Hispanic}$ for employment is $-0.0096$ ($p = 0.009$): a one-log-unit increase in H-2A certifications is associated with a 1.0\% decline in Hispanic agricultural employment relative to non-Hispanic workers. The effect on non-Hispanic employment is also negative ($-0.0072$, $p = 0.007$), suggesting that H-2A-intensive counties experience general agricultural employment declines. Earnings move in the opposite direction: both Hispanic and non-Hispanic earnings rise modestly with H-2A intensity, consistent with a tightening domestic labor market or selection of higher-skill workers into remaining positions. Separations and hires both decline, indicating a contraction of labor market flows rather than active displacement through job loss.

Panel B instruments H-2A with the Bartik shift-share. The first-stage $F$-statistics exceed 21,000 for both ethnicity groups, indicating a powerful instrument. The IV estimate for Hispanic employment is 0.001 ($p = 0.89$)---a precisely estimated zero. The Wu-Hausman test rejects the null of OLS consistency ($p < 0.001$), confirming that OLS is biased by selection. Under the maintained exclusion restriction, the IV suggests no detectable displacement of Hispanic domestic agricultural workers from the demand-driven component of H-2A expansion. However, the Bartik's identifying variation comes from counties with higher initial H-2A shares---which may differ in unobserved agricultural structure---so the IV null should be interpreted as attenuating the OLS relationship under a particular source of variation, rather than definitively ruling out all displacement.

\input{tables/tab2_main}

\subsection{Mechanisms: Why OLS is Biased}

The divergence between OLS and IV reveals the selection mechanism. Counties that increase H-2A usage most rapidly are those experiencing the sharpest domestic labor supply contractions. This is not surprising: the H-2A application process requires employers to demonstrate that domestic workers are ``not sufficient''---a regulatory requirement that ensures H-2A growth is concentrated where domestic supply is already declining. The OLS captures this reverse causality, producing spurious evidence of displacement.

\subsubsection*{Event study} \Cref{tab:event} decomposes the DDD effect by period. The Hispanic-specific coefficient during early expansion (2018--2019) is small and statistically insignificant ($-0.004$, $p = 0.17$). The effect becomes significant in 2020--2021 ($-0.011$, $p = 0.005$) and persists in 2022--2023 ($-0.011$, $p = 0.02$). This pattern is consistent with growing selection bias as the program expands: the marginal county requesting H-2A workers in 2022 is more likely to be experiencing domestic labor shortages than the average county in 2018.

\input{tables/tab5_event}

\subsection{Placebo Tests}

\Cref{tab:placebo} reports the DDD specification applied to two placebo industries that employ many Hispanic workers but are not served by H-2A: construction (NAICS 23) and food service (NAICS 72). If the agriculture result reflected a general Hispanic labor demand shock correlated with H-2A counties, we would expect similar effects in these industries. The construction coefficient is positive and significant ($0.009$, $p = 0.01$)---the opposite sign---while food service is statistically indistinguishable from zero ($-0.002$, $p = 0.26$). These nulls confirm that the agriculture finding is H-2A-specific, not a broader Hispanic labor market phenomenon.

\input{tables/tab3_placebo}

\subsection{Robustness}

\Cref{tab:robust} presents four robustness checks. The baseline coefficient ($-0.010$) is stable when clustering at the state level (column 2, $p = 0.11$---weaker, as expected with 50 clusters), when excluding COVID years 2020--2021 (column 3, $-0.008$, $p = 0.03$), and when estimating in levels rather than logs (column 4, $-6.5$ workers per log-unit of H-2A, $p = 0.01$). Leave-one-state-out analysis dropping each of the top five H-2A states (Florida, Washington, California, North Carolina, Georgia) yields coefficients ranging from $-0.007$ to $-0.009$, confirming that no single state drives the result.

\input{tables/tab4_robust}

\section{Discussion}

The ``displacement mirage'' documented here has a clear policy implication: the cross-county correlation between H-2A growth and declining domestic farm employment, which has fueled calls to cap the program, reflects selection rather than causation. Employers apply for H-2A workers because domestic labor is scarce, not as a substitute for it. Restricting H-2A access would not restore domestic agricultural employment; it would leave crops unharvested.

This finding echoes a broader pattern in the immigration literature: na\"ive geographic comparisons of immigrant concentration and native employment tend to overstate displacement \citep{Peri2016, Dustmann2016}. The contribution here is to demonstrate the bias specifically for temporary agricultural workers using administrative data, where the institutional separation between H-2A and UI-covered domestic workers provides unusually clean measurement.

Several caveats apply. First, the Bartik instrument isolates demand-pull variation; counties with larger 2018 H-2A shares may differ in crop mix, mechanization capacity, or labor market structure in ways that could violate the exclusion restriction. The IV null is best interpreted as showing that the OLS negative association is substantially attenuated under exogenous variation, not as definitively proving zero displacement. Second, the null applies to \textit{employment counts}---wages may respond differently, though the positive earnings effects suggest upward rather than downward pressure. An alternative interpretation is complementarity: H-2A workers enable farm expansion that sustains domestic jobs. Third, data availability limits the H-2A treatment to FY2018--2023; a longer panel incorporating earlier program growth would strengthen the pre-trend analysis. Finally, QWI cells for Hispanic agricultural workers are suppressed in approximately 30\% of county-quarters, potentially introducing sample selection; results are estimated on non-suppressed cells with positive employment counts.

\section{Conclusion}

The largest expansion of legal temporary agricultural immigration in U.S. history has not displaced domestic farm workers. The apparent substitution in cross-county data is a mirage: employers seek guestworkers where domestic labor is already vanishing, not where it thrives. Policy debates that treat H-2A expansion as a threat to domestic employment are arguing with a correlation, not a cause.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
