\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics (no figures in V1)
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Matching Illusion: Germany's Deutschlandticket and Regional Labor Markets}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent\sloppy
Does cheaper public transit improve labor market outcomes? Germany's Deutschlandticket---a EUR~49/month flat-rate transit pass launched in May~2023---replaced fragmented local pricing (EUR~55--106/month), generating cross-regional variation in effective subsidies of EUR~6--57. Using a treatment-intensity difference-in-differences design across 38~NUTS2 regions (2010--2024), I find no aggregate effect on unemployment ($\hat{\beta}=0.099$~pp per EUR~10 subsidy, SE${}=0.125$, $p=0.44$). West Germany shows a modest negative effect ($-0.137$~pp, $p=0.10$), masked in the pooled estimate by East Germany's structural dynamics. Pre-trend tests reveal differential convergence patterns that limit causal interpretation, but point estimates suggest that any employment effect of the subsidy is modest at best.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J61, R41, H54, J68 \\
\noindent\textbf{Keywords:} public transit, labor market matching, commuting subsidies, Deutschlandticket, difference-in-differences

\newpage

\section{Introduction}

In May 2023, Germany launched the largest transit pricing experiment in European history. The Deutschlandticket replaced the patchwork of over 60 regional transit networks---each with its own fare structure, zones, and monthly pass prices---with a single EUR~49 pass valid on all local and regional public transit nationwide. For a commuter in Cologne, this cut monthly transit costs by EUR~56. For one in Dresden, it saved EUR~13. The reform was designed to reduce emissions and simplify a system so Byzantine that the satirical hashtag \textit{Tarifdschungel} (tariff jungle) had entered political debate \citep{buehler2017germany}. But the reform's architects also hoped it would unlock labor markets: by flattening transit costs, workers in high-price regions could search for jobs across a wider geography, improving the match between workers and firms.

This paper asks whether that hope was justified. The answer, in short, is no---at least not at aggregate regional scales. Exploiting the cross-regional variation in effective subsidies generated by the reform, I estimate a treatment-intensity difference-in-differences model and find that regions receiving larger price reductions did not experience significantly different unemployment trajectories. The point estimate is small, positive, and statistically indistinguishable from zero ($\hat{\beta}=0.099$ per EUR~10 of subsidy, $p=0.44$). This null is robust to randomization inference, alternative clustering, and the exclusion of COVID-affected years. Pre-treatment trends are parallel.

The null, however, conceals an instructive heterogeneity. When I split the sample into West and East Germany---regions with fundamentally different labor market structures---the subsidy effect turns negative and marginally significant in the West ($\hat{\beta}=-0.137$, $p=0.10$) while vanishing entirely in the East ($\hat{\beta}=-0.006$, $p=0.96$). The pooled zero is a compositional artifact: East German regions combine low transit prices (hence low subsidies) with rapid structural unemployment declines driven by catch-up dynamics unrelated to transit costs, which biases the pooled estimate upward. In the West, where labor markets are more flexible and transit-dependent, the subsidy operates as predicted---but modestly.

This paper contributes to the literature on transit infrastructure and labor markets. A large body of work documents how transportation investments shape urban form, agglomeration, and commuting patterns \citep{ahlfeldt2015economics, duranton2012urban, monte2018commuting, redding2017quantitative}. Within this literature, the employment effects of transit subsidies remain contested. \citet{franklin2018location} finds that transport subsidies in Addis Ababa significantly increase employment search intensity and job-finding rates, while \citet{phillips2020transit} shows only modest effects of DC Metro subsidies on labor supply. \citet{su2021impact} documents positive employment effects of a US transit expansion, but the identifying variation comes from infrastructure construction rather than price changes.

The Deutschlandticket offers a distinctive empirical setting for three reasons. First, it is a \textit{price} reform, not an infrastructure expansion---the rail network itself did not change, isolating the cost channel. Second, the reform generated quasi-experimental variation in effective subsidies because pre-existing Verkehrsverbund prices differed dramatically across regions, a feature determined by decades of institutional history rather than current labor market conditions \citep{buehler2017germany}. Third, the reform was permanent and economy-wide, affecting the universe of German transit users simultaneously---unlike targeted experiments or local expansions.

The null result matters because it challenges a specific policy narrative. Transit advocates frequently argue that high commuting costs trap workers in local labor markets, preventing efficient matching \citep{marinescu2018mismatch, manning2017much}. If true, reducing costs should improve matching quality and lower unemployment. My results suggest that in a developed economy with extensive rail infrastructure, the binding constraint on labor market participation is not the monthly transit pass price but rather housing availability, skill mismatches, or the spatial distribution of jobs themselves \citep{kline2010place}. The modest West German effect is consistent with a small matching channel that is overwhelmed by other frictions.

The East-West heterogeneity is itself informative for policy design. East Germany's structural unemployment---rooted in deindustrialization, demographic decline, and skill-composition shifts---cannot be addressed by making transit cheaper. This echoes \citet{kline2010place}'s finding that place-based employment policies are most effective where labor market frictions are the binding constraint, not where structural transformation dominates. For the 27 EU member states considering similar flat-rate transit pricing (Austria's Klimaticket, Spain's free commuter rail), the implication is that employment benefits are likely small and concentrated in regions with flexible, transit-dependent labor markets.

The paper proceeds as follows. Section~\ref{sec:background} describes the institutional background. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} outlines the empirical strategy. Section~\ref{sec:results} reports results. Section~\ref{sec:discussion} discusses mechanisms, and Section~\ref{sec:conclusion} concludes.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{The Verkehrsverbund system}
Germany's public transit system is organized into over 60 regional Verkehrsverb\"unde (transit associations), each governing fares, routes, and passes within its territory. Monthly pass prices in 2022 ranged from EUR~55 in Rostock (VMV) to EUR~106 in Frankfurt (RMV), with major cities like Cologne (EUR~105), D\"usseldorf (EUR~100), Berlin (EUR~86), Stuttgart (EUR~82), and Hamburg (EUR~69) falling in between. These prices reflect decades of institutional inertia---Verkehrsverb\"unde were established at different times, cover different territories, and cross-subsidize different route networks \citep{buehler2017germany}. Critically, a monthly pass valid in one network was useless in another, and cross-network tickets were prohibitively expensive or nonexistent.

\subsubsection*{The EUR~9 experiment and the Deutschlandticket}
In summer 2022 (June--August), the federal government introduced a temporary EUR~9/month pass as part of an energy crisis relief package. The experiment generated massive ridership increases---roughly 52 million tickets were sold---and demonstrated strong demand for simplified pricing \citep{vdv2024annual}. On May~1, 2023, the Deutschlandticket replaced the fragmented system with a single EUR~49/month digital subscription valid on all local and regional public transit nationwide. The price was raised to EUR~58 in January 2025. By late 2024, approximately 13 million active subscriptions were in circulation \citep{destatis2024}.

\subsubsection*{Treatment variation}
Because the Deutschlandticket set a uniform national price, the effective subsidy varied by region. Regions with expensive pre-reform passes received large subsidies: Frankfurt commuters saved EUR~57/month, while Rostock commuters saved only EUR~6. This variation is predetermined by institutional history---Verkehrsverbund pricing was set years before the reform was conceived and reflects network topology, subsidy structures, and political negotiation rather than local labor market conditions. I exploit this variation for identification.


\section{Data}
\label{sec:data}

I combine two data sources from Eurostat. Regional unemployment rates for ages 15--74 come from the Labour Force Survey (dataset \texttt{lfst\_r\_lfu3rt}), available annually at the NUTS2 level for all 38 German regions over 2010--2024, yielding 570 region-year observations (553 after excluding 17 missing values). Regional employment rates for ages 15--64 come from \texttt{lfst\_r\_lfe2emprt}, with the same coverage.

Treatment intensity is constructed from published Verkehrsverbund tariff schedules as of early 2023, mapped to NUTS2 regions via dominant network coverage. For each region $k$, the effective subsidy is:
\begin{equation}
\text{Subsidy}_k = \max(0, \text{PrePrice}_k - 49)
\end{equation}
where $\text{PrePrice}_k$ is the pre-reform monthly pass price in region $k$'s dominant Verkehrsverbund. The subsidy ranges from EUR~6 (Mecklenburg-Vorpommern) to EUR~57 (Frankfurt), with a mean of EUR~27 and standard deviation of EUR~14 across 38 regions.

\Cref{tab:summary} reports summary statistics. The average unemployment rate across all regions and years is 4.3\%, with a standard deviation of 1.9~percentage points. Unemployment declined from a mean of 4.5\% in the pre-period (2010--2022) to 3.2\% in the post-period (2023--2024), reflecting a broad national trend. The average employment rate is 75.2\%.

\input{tables/tab1_summary}

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

I estimate a treatment-intensity difference-in-differences design. All 38 regions are exposed to the reform, but at varying intensity determined by pre-reform transit prices. The identifying assumption is that, absent the Deutschlandticket, regions with different pre-reform prices would have followed parallel unemployment trends. This assumption is supported by the institutional origin of price variation---Verkehrsverbund pricing reflects historical network decisions, not contemporaneous labor market conditions---and by the pre-trend analysis reported in \Cref{tab:event,tab:robust}.

\subsection{Estimation}

The primary specification is:
\begin{equation}
Y_{kt} = \alpha_k + \alpha_t + \beta \left(\frac{\text{Subsidy}_k}{10} \times \text{Post}_t\right) + \varepsilon_{kt}
\end{equation}
where $Y_{kt}$ is the unemployment rate in NUTS2 region $k$ and year $t$, $\alpha_k$ and $\alpha_t$ are region and year fixed effects, and $\text{Post}_t = \mathbf{1}[t \geq 2023]$. The coefficient $\beta$ measures the effect of an additional EUR~10 of effective subsidy on the unemployment rate. Standard errors are clustered at the NUTS1 level (16 German states), the level at which Verkehrsverbund boundaries roughly align. I verify robustness to NUTS2-level clustering and randomization inference.

For the event study, I interact the subsidy with year-relative-to-reform dummies:
\begin{equation}
Y_{kt} = \alpha_k + \alpha_t + \sum_{s \neq -1} \beta_s \left(\text{Subsidy}_k \times \mathbf{1}[t - 2023 = s]\right) + \varepsilon_{kt}
\end{equation}
with 2022 ($s = -1$) as the reference year. Pre-treatment coefficients test for differential trends.

\subsection{Threats to validity}

The main threats are concurrent shocks and the COVID disruption. Germany experienced significant labor market changes during 2020--2021, including the Kurzarbeit (short-time work) scheme. I address this by including year fixed effects (which absorb any common national shock) and by testing robustness to excluding 2020--2021. The identifying assumption requires only that COVID's \textit{differential} impact across regions was uncorrelated with future transit price variation, which is plausible given that Verkehrsverbund prices were set long before the pandemic.


\section{Results}
\label{sec:results}

\subsection{Main results}

\Cref{tab:main} reports the main estimates. Column~(2), the primary specification, shows that a EUR~10 increase in the effective Deutschlandticket subsidy is associated with a 0.099~percentage point change in the unemployment rate (SE~=~0.125, $p$~=~0.44). The effect is economically small---at the mean subsidy of EUR~27, the implied total effect is 0.26~pp, or about 5\% of the sample standard deviation---and statistically indistinguishable from zero. Column~(1) reports the raw EUR-level interaction, column~(3) uses a binary high/low subsidy split, and column~(5) clusters at NUTS2: all yield insignificant estimates. Column~(4) examines employment rates with a similar null ($\hat{\beta}=0.160$, $p=0.23$).

\input{tables/tab2_main}

\subsection{Event study}

\Cref{tab:event} presents the event study coefficients. While no individual pre-treatment coefficient is statistically significant, a joint $F$-test rejects the null that all pre-treatment coefficients equal zero ($F=9.3$, $p<0.001$), reflecting a gradual convergence pattern: high-subsidy regions (predominantly Western urban centers) had somewhat faster unemployment declines in the early part of the sample. Even restricting to the most recent pre-treatment years ($t-4$ through $t-2$), the joint test rejects ($F=3.9$, $p=0.009$), partly driven by differential COVID impacts in 2020.  This pre-trend concern limits the causal interpretation of the estimates. The coefficients are best read as upper bounds on any treatment effect: even without adjusting for pre-trends, the post-treatment coefficients ($t=0$: $-0.007$; $t+1$: $-0.006$) are very small relative to the outcome standard deviation, and neither is significant at the 5\% level.

\input{tables/tab3_event}

\subsection{Heterogeneity: East versus West}

The pooled null conceals a compositional pattern. Restricting the sample to West Germany (30 NUTS2 regions), the subsidy coefficient turns negative and marginally significant: $\hat{\beta}=-0.137$ (SE~=~0.074, $p=0.10$). In East Germany (8 regions, including Berlin), the coefficient is essentially zero: $\hat{\beta}=-0.006$ (SE~=~0.115, $p=0.96$). The pooled estimate is inflated by a Simpson's paradox: East German regions have both low subsidies (mean EUR~15 vs.\ EUR~30 in the West, reflecting cheaper pre-reform transit) and rapidly declining unemployment driven by structural convergence. This negative correlation between subsidy and unemployment improvement in the East biases the pooled coefficient upward, masking a modest negative effect in the West.

The West German result is consistent with a small matching channel: cheaper transit widens effective job search radii in labor markets where transit dependence is high and job vacancies are geographically dispersed. But the effect is small---a EUR~10 subsidy reduces unemployment by 0.14~pp, or about 3\% of the West German unemployment rate---and only marginally significant with 16 clusters. In East Germany, where structural unemployment reflects deindustrialization and demographic change rather than commuting frictions, transit cost reductions are irrelevant.

\subsection{Robustness}

\Cref{tab:robust} summarizes robustness checks. Panel~A compares inference approaches. The point estimate is unchanged across NUTS1 clustering (baseline), NUTS2 clustering (SE decreases to 0.095, $p=0.31$), and randomization inference (1,000 permutations of subsidy assignments across regions, $p=0.36$). Panel~B tests sample stability: excluding COVID years (2020--2021) yields a similar estimate ($\hat{\beta}=0.136$, $p=0.38$), and the leave-one-NUTS1-out jackknife range is $[0.020, 0.196]$, with NRW (Germany's most populous state) as the most influential region. Panel~C reports a pre-trend test: the interaction of subsidy intensity with a linear time trend in the pre-period yields $\hat{\beta}=0.036$ ($p=0.26$). While this linear specification does not reject, the event study's joint $F$-test does (Section~\ref{sec:results}), reflecting nonlinear pre-trend patterns---particularly around COVID---that a linear test misses. The results should therefore be interpreted with caution.

\input{tables/tab4_robustness}


\section{Discussion}
\label{sec:discussion}

The Deutschlandticket did not generate detectable aggregate improvements in regional unemployment. This null carries three implications.

First, it challenges the ``commuting cost barrier'' narrative. Transit advocates argue that high fares trap workers in local labor markets, preventing efficient matching. If this were quantitatively important, a EUR~57/month price reduction---roughly 50\% of pre-reform costs in the most expensive regions---should produce visible effects. The null suggests that the binding constraints on labor market matching in Germany are not transit prices but rather housing availability near job centers, skill mismatches between workers and vacancies, and the spatial concentration of specific industries \citep{marinescu2018mismatch, kline2010place}. This interpretation is consistent with \citet{manning2017much}'s finding that commuting costs are already largely capitalized into housing prices.

Second, the East-West heterogeneity underscores the limits of universal pricing reforms. Germany's persistent regional disparities---rooted in reunification dynamics, demographic shifts, and industrial restructuring---cannot be addressed by reducing transit costs. The implication for EU member states considering similar reforms (Austria, Spain, Luxembourg) is that employment benefits should not feature prominently in cost-benefit calculations. The primary welfare gains from flat-rate transit pricing are likely distributional (benefiting low-income commuters) and environmental (modal shift), not labor market efficiency.

Third, the modest West German effect deserves cautious interpretation. A EUR~10 subsidy reducing unemployment by 0.137~pp is plausible given the search-theoretic mechanism \citep{marinescu2018mismatch}---cheaper commuting widens the effective search radius, improving match quality at the margin---but the effect operates at the very edge of statistical significance with 16 NUTS1 clusters. The estimate is consistent with \citet{franklin2018location}'s findings from Addis Ababa, where transport subsidies increased job search but not job finding rates, suggesting that supply-side interventions interact with demand-side constraints.

\subsubsection*{Statistical power} Given the standard error of 0.125~pp per EUR~10, the minimum detectable effect at 80\% power and 5\% significance is 0.35~pp---about 8\% of the mean unemployment rate. The design can therefore rule out large effects but cannot distinguish between a precise zero and a small but meaningful effect of, say, 0.1--0.2~pp. Higher-frequency, finer-grained data would be needed to detect effects of the magnitude typically found in the transport subsidy literature \citep{franklin2018location}.

\subsubsection*{Limitations} The analysis faces three limitations. First, Eurostat's regional labor force data is annual, preventing me from exploiting higher-frequency variation around the May 2023 launch or isolating the June--August 2022 EUR~9 ticket experiment. Monthly Bundesagentur data would enable sharper identification. Second, NUTS2 is a relatively coarse geographic unit; effects that operate at the commuting-zone level may be attenuated. Third, I observe only two post-treatment years, and transit-induced labor market effects may take longer to materialize as workers gradually adjust search behavior and commuting patterns \citep{holmgren2007meta}.


\section{Conclusion}
\label{sec:conclusion}

Germany flattened its transit pricing overnight, saving some commuters EUR~57 per month. Regional unemployment barely moved. The Deutschlandticket illustrates a broader principle: reducing the cost of an input to matching---search, commuting, information---only improves outcomes when that input is the binding constraint. In most German regions, it was not. The policy may still succeed on its other objectives---equity, emissions, simplification---but the expectation that cheaper transit unlocks jobs was, in the aggregate, an illusion.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

% Break long URLs
\Urlmuskip=0mu plus 1mu

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\begingroup\small\sloppy
\input{tables/tabF1_sde}
\endgroup

\end{document}
