\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{No Crossing: Reproductive Rights Are Not Capitalized at State Borders}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
On June 24, 2022, Missouri's trigger-law abortion ban activated overnight while neighboring Illinois maintained full protections---creating the sharpest possible reproductive policy discontinuity at the Mississippi River in the St. Louis metropolitan area. Using monthly ZIP-level Zillow Home Value Index data and a geographic difference-in-discontinuities design, I test whether this loss of reproductive rights is capitalized into property values at the border. Within 15 kilometers of the state line, the estimated effect is $1.0\%$ (SE $= 3.6\%$), ruling out effects larger than 8 percentage points. At Kansas City, where Missouri borders abortion-protecting Kansas, the point estimate reverses sign ($-1.4\%$). Over a 30-month horizon, I find no detectable capitalization of reproductive rights at either border, consistent with cross-border clinic access neutralizing the Tiebout sorting mechanism in dense metropolitan areas.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R21, R23, I18, J13 \\
\noindent\textbf{Keywords:} reproductive rights, property values, capitalization, Tiebout sorting, Dobbs, spatial discontinuity

\newpage

\section{Introduction}

At 10:10 a.m. on June 24, 2022, the Supreme Court issued its decision in \textit{Dobbs v. Jackson Women's Health Organization}. By 10:11 a.m., Missouri Attorney General Eric Schmitt had signed the paperwork activating the state's trigger law, imposing a near-total ban on abortion. Across the Mississippi River, Illinois's Reproductive Health Act of 2019 codified abortion as a fundamental right. In the span of one minute, the 370,000 women of reproductive age living in the St. Louis metropolitan area faced a binary policy divide determined entirely by which side of a river they called home.

If reproductive rights are valued as an amenity---as the Tiebout model would predict for any policy affecting quality of life---this overnight policy shock should be visible in housing prices. The logic is direct: households that value reproductive access face higher costs of staying in Missouri (travel to Illinois for care) or can relocate across the border. Either channel should depress home values on the Missouri side relative to the Illinois side, with effects concentrated near the border where cross-state substitution is cheapest. This is the Tiebout prediction in its purest form: an exogenous, overnight, binary policy change between two jurisdictions sharing a dense metropolitan border.

This paper tests whether the \textit{Dobbs} decision is capitalized into residential property values at the Missouri-Illinois border. I construct a monthly panel of 235 ZIP codes in the St. Louis metropolitan area (129 in Missouri, 106 in Illinois) from January 2020 through December 2024 using the Zillow Home Value Index, and estimate a geographic difference-in-discontinuities design \citep{grembi2016fiscal}. The running variable is signed distance from each ZIP centroid to the state border. The design differences out time-invariant cross-border gaps and common temporal trends, isolating the change in the border discontinuity attributable to \textit{Dobbs}.

The central finding is a non-result. In the narrow-bandwidth specification most appropriate for a border design---ZIP codes within 15 kilometers of the state line---the estimated effect is $0.010$ (SE $= 0.036$), with a 95\% confidence interval of $[-6.1\%, 8.0\%]$. The design rules out large effects but cannot detect economically meaningful responses of 2--5 percent. At wider bandwidths, estimates turn weakly positive but are driven by pre-existing differential appreciation between the Missouri and Illinois interiors, not by a border discontinuity. At Kansas City, where Missouri borders Kansas---which reinforced abortion protections via a decisive August 2, 2022 referendum---the basic DiD estimate is $-0.014$ (SE $= 0.007$, $p = 0.054$), opposite in sign. The sign reversal across replication borders is consistent with metro-specific housing dynamics rather than a reproductive rights channel.

This non-result contributes to three literatures. First, it speaks to the growing body of work on the economic consequences of \textit{Dobbs}. \citet{dench2025dobbs} use state-level synthetic control methods and find modest property value effects in ban states, but cannot distinguish border-level Tiebout sorting from state-level responses. My spatial design fills this gap: whatever aggregate effects exist, they do not manifest as a sharp discontinuity in property values at the metropolitan border. Second, the paper adds to the capitalization literature pioneered by \citet{oates1969effects} and extended to amenities from school quality \citep{black1999better} to environmental risk \citep{davis2011effect} to crime \citep{linden2008estimates}. In that literature, capitalization effects of 5--10 percent are typical for amenities affecting daily life \citep{chay2005does,cellini2010value}. The null here is informative because the policy shock is unusually clean---overnight, binary, and unrelated to local economic conditions---yet produces no comparable response, consistent with reproductive healthcare being an episodic rather than daily-use amenity. Third, the paper contributes to the growing evidence that reproductive policy affects migration decisions \citep{jones2022migration,kearney2023dobbs} while showing that any such responses are not large enough to move local housing markets over a 30-month window.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional setting. \Cref{sec:data} presents the data. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} reports results. \Cref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{The Dobbs Decision and Missouri's Trigger Law} The Supreme Court's decision in \textit{Dobbs v. Jackson Women's Health Organization} (June 24, 2022) overturned \textit{Roe v. Wade} (1973), returning abortion regulation to the states. Missouri's General Assembly had enacted a trigger law (HB 126, 2019) that would impose a near-total ban on abortion upon any Supreme Court decision ``restoring to the states their authority to regulate abortion.'' The law took effect within minutes of the \textit{Dobbs} ruling, making Missouri one of the first states to implement a comprehensive ban. The ban prohibits abortion at all stages of pregnancy with narrow exceptions for medical emergencies, and imposes criminal penalties on providers.

\subsubsection*{Illinois's Countervailing Protection} Illinois took the opposite path. The Reproductive Health Act (Public Act 101-0013, signed June 12, 2019) established a ``fundamental right'' to reproductive healthcare, including abortion, codified in state statute. Governor Pritzker signed additional protections in 2022 and 2023, including shielding patients and providers from out-of-state legal actions. Illinois became a destination for patients from Missouri and other ban states, with Planned Parenthood of the St. Louis Region reporting a 30-fold increase in out-of-state patients at its Fairview Heights, Illinois clinic in the months after \textit{Dobbs}.

\subsubsection*{The St. Louis Metropolitan Border} The St. Louis MSA (CBSA 41180) spans the Missouri-Illinois border along the Mississippi River. The metro area contains approximately 2.8 million residents, with major population centers on both sides: St. Louis City and County in Missouri; Madison and St. Clair counties in Illinois. The border runs through a dense, economically integrated metropolitan area where residents routinely commute, shop, and access services across state lines. This integration makes the border ideal for a geographic discontinuity design: housing markets on either side serve the same labor market, and the policy change is orthogonal to local economic conditions.

\subsubsection*{Kansas City as a Replication Border} The Kansas City MSA (CBSA 28140) provides a natural replication. Missouri's ban applies equally at this border, but Kansas voters rejected a constitutional amendment to remove abortion protections in an August 2, 2022 referendum by a 59--41 margin---one of the first direct votes on reproductive rights after \textit{Dobbs}. The Kansas referendum reinforced the policy discontinuity at the MO-KS border, providing a second testing ground with a different institutional mechanism for the protection-side policy.

\subsubsection*{The Politico Leak} On May 2, 2022---52 days before the official decision---Politico published a leaked draft of the majority opinion in \textit{Dobbs}. The leak was widely covered and signaled that \textit{Roe} would likely be overturned. I exploit this event to test for anticipatory capitalization.

\section{Data}
\label{sec:data}

\subsubsection*{Home Values} The primary outcome is the Zillow Home Value Index (ZHVI) at the ZIP code level, measured monthly. ZHVI is a smoothed, seasonally adjusted estimate of the typical home value for all homes (single-family residences and condominiums) within a ZIP code, calculated by Zillow from its proprietary valuation model applied to the universe of U.S. homes. I use data from January 2020 through December 2024, providing 30 months before and 30 months after the \textit{Dobbs} decision (excluding June 2022 as a transition month). The outcome variable is the natural logarithm of ZHVI.

\subsubsection*{Geographic Data} ZIP code centroids come from the Census Bureau's 2020 ZCTA shapefile. State boundary geometries come from the Census TIGER/Line program. I compute signed distance from each ZIP centroid to the nearest point on the state border, with positive values indicating Missouri (ban state) and negative values indicating Illinois or Kansas (protection states).

\subsubsection*{Sample Construction} I identify ZIP codes in the St. Louis MSA using the Census Bureau's ZCTA-to-county relationship file, selecting all ZCTAs with area in counties belonging to CBSA 41180. The analysis sample contains 235 ZIP codes (129 Missouri, 106 Illinois) observed over 59 months (excluding June 2022), yielding 13,514 ZIP-month observations. The Kansas City replication uses 202 ZIP codes (131 Missouri, 71 Kansas), yielding 11,841 observations.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} reports summary statistics. Missouri ZIP codes have higher average home values (\$252,710 vs. \$152,435 for Illinois), reflecting the inclusion of affluent St. Louis County suburbs on the Missouri side. This level difference is absorbed by ZIP fixed effects. The average distance from ZIP centroids to the state border is approximately 30--40 kilometers on each side, confirming adequate spatial coverage.

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

I employ a geographic difference-in-discontinuities (diff-in-disc) design \citep{grembi2016fiscal}. The estimand is the change in the border discontinuity in home values attributable to \textit{Dobbs}:
\begin{equation}
\tau = \underbrace{\left[\lim_{d \downarrow 0} \E[Y|d, \text{Post}] - \lim_{d \uparrow 0} \E[Y|d, \text{Post}]\right]}_{\text{Post-Dobbs border gap}} - \underbrace{\left[\lim_{d \downarrow 0} \E[Y|d, \text{Pre}] - \lim_{d \uparrow 0} \E[Y|d, \text{Pre}]\right]}_{\text{Pre-Dobbs border gap}}
\end{equation}
where $d$ is signed distance to the state border (positive = Missouri, negative = Illinois) and $Y$ is log ZHVI. The first differencing (across space at the border) removes smooth spatial confounders. The second differencing (pre vs. post) removes the time-invariant border discontinuity.

\subsection{Estimation}

The panel specification is:
\begin{equation}
\ln(\text{ZHVI}_{it}) = \mu_i + \lambda_t + \tau \cdot \text{MO}_i \times \text{Post}_t + f(d_i) \times \text{Post}_t + \varepsilon_{it}
\label{eq:main}
\end{equation}
where $\mu_i$ and $\lambda_t$ are ZIP and year-month fixed effects, $\text{MO}_i$ indicates Missouri, $\text{Post}_t$ indicates months after June 2022, and $f(d_i)$ is a polynomial in signed distance to the border interacted with the post indicator. The coefficient $\tau$ identifies the diff-in-disc effect. Standard errors are clustered at the ZIP level.

I estimate five specifications: (1) basic DiD without distance controls; (2) with linear $f(d_i) \times \text{Post}_t$; (3) with quadratic; (4) restricting to ZIP codes within 30 km of the border; (5) restricting to 15 km. I complement the panel approach with a cross-sectional RDD on the change in average log ZHVI from pre to post, estimated using the \texttt{rdrobust} package \citep{cattaneo2020rdrobust} with MSE-optimal bandwidth selection.

\subsection{Threats to Validity}

The key identifying assumption is that the change in the border discontinuity in home values would have been zero absent \textit{Dobbs}. Three concerns merit discussion.

\subsubsection*{Pre-existing differential trends} The event study (\Cref{sec:results}) reveals that Missouri ZIP codes were appreciating faster than Illinois ZIP codes before \textit{Dobbs}. The diff-in-disc framework addresses this through the distance$\times$post interaction, which absorbs differential spatial trends. The temporal placebo test (fake \textit{Dobbs} in January 2021) yields an insignificant coefficient of 0.016 (SE = 0.011), confirming that the design does not spuriously detect effects in the pre-period.

\subsubsection*{Anticipation} The May 2, 2022 Politico leak could trigger anticipatory capitalization. I test this directly and find a coefficient of 0.018 (SE = 0.007) at the leak date---positive, not negative, confirming it captures the pre-existing trend rather than anticipatory sorting away from Missouri.

\subsubsection*{Concurrent policies} Missouri and Illinois differ on policies beyond abortion (taxes, gun laws, marijuana legalization). ZIP fixed effects absorb time-invariant differences, and the diff-in-disc design requires only that no coincident policy change occurred precisely at the border precisely at the \textit{Dobbs} date. The Kansas City replication, where the protection-side policy has a different institutional origin (referendum rather than statute), provides a check on this concern.

\subsubsection*{ZHVI measurement} The Zillow Home Value Index is a smoothed, model-based measure, not a transaction price index. Smoothing attenuates short-run breaks and induces serial correlation, potentially reducing power to detect sharp capitalization effects. This means the design is better suited to detecting sustained shifts in valuations than transitory shocks. Standard errors are clustered at the ZIP level; results are qualitatively unchanged under two-way clustering by ZIP and year-month (not reported).

\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

\Cref{tab:main} reports the main diff-in-disc estimates. Column (1) shows the basic DiD: Missouri home values increased by 5.3 percent relative to Illinois after \textit{Dobbs} ($p < 0.001$). This large positive coefficient is misleading---it reflects the pre-existing Missouri appreciation trend, not a causal effect of \textit{Dobbs}.

Adding a linear distance$\times$post control in Column (2) reduces the estimate to 2.9 percent (SE = 1.7\%, $p = 0.086$), demonstrating that much of the naive DiD estimate reflects differential spatial trends rather than a border discontinuity. Adding a quadratic distance control in Column (3) produces a virtually identical estimate, confirming that the spatial relationship is well-captured by the linear term.

For a paper about capitalization at the border, the most informative specifications restrict the sample to ZIP codes near the state line. Column (4), using a 30 km bandwidth, yields an estimate of 3.9 percent that is statistically insignificant ($p = 0.145$). Column (5), using a 15 km bandwidth---the preferred specification, where Tiebout sorting should be most visible---produces an estimate of 1.0 percent with a standard error of 3.6 percent. The 95\% confidence interval of $[-6.1\%, 8.0\%]$ comfortably includes zero but does not rule out economically meaningful effects in the 2--5 percent range. At the narrowest bandwidth of 10 km (\Cref{tab:bandwidth}), the point estimate is $-3.1$\% (SE $= 4.5$\%), still insignificant but notable for its sign reversal relative to wider bandwidths---consistent with a spatial trend driving the positive coefficients at greater distances, not a border discontinuity.

The cross-sectional RDD on the change in home values gives a point estimate of 6.0 percent with a robust $p$-value of 0.134 and a 95\% confidence interval of $[-1.9\%, 13.9\%]$, confirming the panel findings.

\subsection{Event Study}

The event study (\Cref{eq:main} with month-specific coefficients replacing the Post dummy) reveals the temporal pattern. Pre-\textit{Dobbs} coefficients are uniformly negative and significant, indicating that Missouri was appreciating relative to Illinois in the months before the reference period (May 2022). In the months immediately following \textit{Dobbs} (July--November 2022), coefficients are briefly positive and significant (+1.0\% to +1.5\%), suggesting a temporary acceleration of the pre-trend. By December 2022 (month 7), coefficients return to zero and remain statistically insignificant through December 2024. There is no visual evidence of a sustained break in the series at the \textit{Dobbs} date.

\subsection{Kansas City Replication}

\input{tables/tab4_kc.tex}

\Cref{tab:kc} reports the Kansas City replication. The basic DiD estimate is $-1.4$\% (SE $= 0.7\%$, $p = 0.054$)---negative, not positive. Missouri home values \textit{declined} relative to Kansas after \textit{Dobbs}. With distance controls, the KC estimate is $-1.2$\% ($p = 0.335$). A caveat applies: Kansas's protection was reinforced by the August 2, 2022 referendum, not instantaneously on June 24, creating roughly six weeks of policy uncertainty on the Kansas side. Nevertheless, the sign reversal across two replication borders---one with statutory protection (Illinois), one with referendum-affirmed protection (Kansas)---is difficult to reconcile with a general reproductive-rights capitalization channel and is more consistent with metro-specific housing dynamics.

\subsection{Robustness}

\input{tables/tab3_bandwidth.tex}

\subsubsection*{Bandwidth sensitivity} \Cref{tab:bandwidth} varies the distance bandwidth from 10 to 75 km and the full MSA. The coefficient is negative at 10 km ($-3.1$\%, insignificant), near zero at 15 km, and positive at wider bandwidths. This pattern is consistent with a smooth spatial trend---the ``effect'' grows as the sample extends further from the border into the interiors of each state, where Missouri happens to be appreciating faster---not with a border discontinuity.

\subsubsection*{Placebo tests} \Cref{tab:placebo} reports temporal and spatial placebo tests. The temporal placebo (fake \textit{Dobbs} at January 2021, using only pre-period data) yields 0.016 ($p = 0.143$), confirming no false positive in the pre-period. Spatial placebos shifting the border 10--20 km into each state produce mixed results: estimates at the shifted ``borders'' are of similar magnitude to those at the true border, further indicating that the estimates capture spatial trends rather than a discontinuity.

\subsubsection*{Anticipation test} The Politico leak test (restricting to July 2021--June 2022) gives a positive coefficient of 0.018 ($p = 0.005$). A genuine anticipation effect would be negative; the positive sign confirms this captures the ongoing pre-trend.

\input{tables/tab5_placebo.tex}

\section{Discussion}
\label{sec:discussion}

The central finding is that the overnight loss of reproductive rights in Missouri did not produce detectable Tiebout capitalization at the Missouri-Illinois border over a 30-month horizon. At 15 km, the 95\% confidence interval rules out effects larger than 8 percent but does not rule out smaller responses in the 2--5 percent range that would be economically meaningful. The finding is consistent across bandwidths (the coefficient is never significantly different from zero near the border), survives temporal and spatial placebo tests, and the sign reversal at Kansas City weighs against a general capitalization channel.

What explains the null? Three candidate mechanisms deserve discussion. First, the time horizon may be too short for housing markets to fully adjust. Property values are forward-looking, but migration and sorting take time. The 30-month post-period may capture only the beginning of a longer adjustment. However, the event study shows no trending divergence even 30 months out, suggesting the effect is not building.

Second, reproductive rights may not enter housing demand in the way the Tiebout model predicts. Housing demand responds to amenities that affect daily life---school quality, crime, commute time. Reproductive healthcare is episodic and accessed infrequently; the marginal home buyer may not weigh it heavily against mortgage rates, lot size, or school districts. This is consistent with survey evidence showing that while Americans express strong preferences over reproductive policy, few rank it among their top concerns when choosing where to live.

Third, the availability of cross-border care may neutralize the sorting incentive. Even under Missouri's ban, St. Louis residents can access abortion services at clinics in Granite City and Fairview Heights, Illinois---less than 20 minutes across the river. If the policy's bite is blunted by geographic proximity to a protection state, border property values would not adjust. This interpretation is specific to border communities and does not generalize to interior Missouri.

From a policy perspective, the finding is cautiously reassuring for property tax bases in border-adjacent communities, though it may not generalize to the interior of ban states far from cross-border clinics, where the ``access blunting'' mechanism is weaker. Nor does it speak to longer-run adjustments---migration and construction responses may take years to capitalize into smoothed home value indices.

\section{Conclusion}

This paper asks whether the loss of reproductive rights is capitalized into property values at state borders. Using the sharpest possible natural experiment---an overnight trigger-law ban at a dense metropolitan border---I find no detectable border capitalization over 30 months. The result holds across bandwidths, placebos, event-study dynamics, and reverses sign at a second replication border. In a setting where cross-border clinic access is a 20-minute drive, the Tiebout mechanism appears blunted: reproductive rights may matter for individual welfare without registering as a priced amenity in local housing markets.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsubsection*{Zillow Home Value Index} The ZHVI data are downloaded from Zillow Research (\url{https://www.zillow.com/research/data/}). I use the ``ZHVI All Homes (SFR, Condo/Co-op) Time Series, Smoothed, Seasonally Adjusted'' file at the ZIP code level. The dataset contains 26,300 ZIP codes with monthly observations from January 2000 through January 2026. I restrict to the analysis window of January 2020--December 2024, excluding June 2022 as a transition month.

\subsubsection*{Geographic Boundaries} State boundary polygons and ZCTA centroids are from the U.S. Census Bureau's TIGER/Line program, accessed via the \texttt{tigris} R package \citep{walker2016tigris}. The Missouri-Illinois shared border is computed as the geometric intersection of state boundary lines. Signed distance from each ZCTA centroid to the nearest point on this border is computed using the \texttt{sf} package \citep{pebesma2018sf}. Distance is measured in kilometers, with positive values indicating Missouri and negative values indicating Illinois.

\subsubsection*{MSA Membership} ZIP codes are assigned to metropolitan statistical areas using the Census Bureau's 2020 ZCTA-to-county relationship file. St. Louis MSA (CBSA 41180) counties include St. Louis City and County, St. Charles, Jefferson, Franklin, Lincoln, and Warren (Missouri) and Madison, St. Clair, Bond, Calhoun, Clinton, Jersey, Macoupin, and Monroe (Illinois). Kansas City MSA (CBSA 28140) counties include Jackson, Cass, Clay, Platte, Lafayette, Ray, Caldwell, Bates, and Clinton (Missouri) and Johnson, Wyandotte, Leavenworth, Miami, and Jefferson (Kansas).

\section{Identification Appendix}
\label{app:identification}

\subsubsection*{Covariate Balance} ZIP fixed effects absorb all time-invariant differences between Missouri and Illinois ZIP codes, including baseline demographics, income levels, and housing stock composition. The identifying assumption requires only that the \textit{change} in these characteristics at the border is smooth through the \textit{Dobbs} date.

\subsubsection*{McCrary Test Analogue} In a geographic RDD, the manipulation concern is that units can sort across the border in response to the policy. Unlike standard RDD running variables, geographic location is fixed for structures---houses cannot relocate. New construction could respond, but the time horizon for construction responses exceeds the 30-month post-period.

\section{Robustness Appendix}

Results from bandwidth sensitivity, placebo tests, and the leak anticipation test are reported in \Cref{tab:bandwidth} and \Cref{tab:placebo} in the main text. Key robustness findings:

\begin{enumerate}
\item The coefficient is negative at the narrowest bandwidth (10 km: $-3.1\%$, SE $= 4.5\%$) and positive only at wider bandwidths, consistent with a spatial trend rather than a border discontinuity.
\item The temporal placebo (fake \textit{Dobbs} at January 2021) yields 0.016 ($p = 0.14$).
\item Spatial placebos at shifted ``borders'' produce estimates of similar magnitude to the true border, confirming that estimates reflect spatial trends.
\item The Politico leak anticipation test yields a \textit{positive} coefficient, opposite to the predicted direction, confirming it captures the pre-trend.
\end{enumerate}

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
