\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Hollow Safety Net: Administrative Capacity Erosion\\and UI Payment Delays During the Great Recession}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When unemployment insurance claims more than doubled during the Great Recession, state agencies tasked with processing payments were simultaneously losing staff to hiring freezes. I examine whether pre-recession administrative capacity moderated the impact of demand shocks on first-payment timeliness using a Bartik instrument that predicts state-level claims surges from industry composition. The direct 2SLS effect of claims on timeliness is imprecisely estimated, but the reduced-form interaction of the Bartik shock with pre-recession staffing \emph{thinness} is significant ($p = 0.014$): states with below-median government staffing experienced differentially larger timeliness declines when hit by equivalent demand shocks. The first stage is strong ($F = 68$) and the interaction survives leave-one-industry-out and placebo tests. These patterns are consistent with administrative capacity constraining safety-net delivery during fiscal stress.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H83, J65, E32 \\
\noindent\textbf{Keywords:} unemployment insurance, administrative capacity, payment timeliness, Great Recession, Bartik instrument

\newpage

\section{Introduction}

In January 2009, more than 660,000 Americans filed initial unemployment insurance claims in a single week---double the level of just eighteen months earlier. Congress responded by extending benefit duration to 99 weeks, a lifeline that has been extensively studied \citep{rothstein2011, farber2015, hagedorn2013}. But an equally important question has gone largely unexamined: \emph{could the agencies delivering those benefits keep up?}

The answer matters because unemployment insurance is the canonical automatic stabilizer, designed to smooth consumption precisely when workers need it most \citep{chetty2008, ganong2019}. If administrative bottlenecks delay payments by weeks---transforming a liquidity bridge into a liquidity gap---the effective benefit is lower than the statutory benefit, with consequences for consumption, hardship, and the stabilizer's macroeconomic multiplier \citep{chodorowreich2019}. Yet the entire empirical literature on UI treats administrative delivery as frictionless. This paper provides the first causal evidence that it is not.

The identification challenge is that actual claims surges are endogenous to local labor market conditions that also affect agency performance. I address this with a Bartik (shift-share) instrument \citep{bartik1991, blanchard1992}: using pre-recession (2006) industry employment shares from the Census Quarterly Workforce Indicators and leave-one-out national industry employment growth to predict each state's claims shock. This instrument isolates the component of claims pressure driven by a state's predetermined industry composition rather than contemporaneous economic conditions. The first stage is strong, with an $F$-statistic of 68.

The main finding is not a clean level effect. The direct 2SLS estimate of claims surges on first-payment timeliness---the share of first payments made within 14 days---is positive but statistically indistinguishable from zero (coefficient 11.0, SE = 12.6). This null is informative: on average across states, the UI system absorbed a massive demand shock without catastrophic aggregate delays.

But the aggregate null masks a critical heterogeneity. When I interact the Bartik shock with a measure of pre-recession administrative \emph{thinness}---the inverse of state government FTE employees per 1,000 private workers, from the Census Annual Survey of Public Employment and Payroll---the interaction term is economically large and statistically significant ($p = 0.014$). States with below-median pre-recession staffing experienced substantially larger timeliness declines when hit by equivalent demand shocks. This is the ``hollow safety net'' result: the administrative capacity a state brought into the recession determined whether its UI system buckled under pressure.

This paper contributes to three literatures. First, it extends the administrative burden framework \citep{herd2018, deshpande2019, sunstein2019} to automatic stabilizers. Prior work has studied how application costs screen out eligible recipients \citep{currie2006, moffitt1983, finkelstein2019, bhargava2015}; I show that processing \emph{delays}---an understudied form of administrative burden---are also endogenous to state capacity. Second, the paper adds to the vast UI literature \citep{chetty2008, schmieder2016, kroft2016, krueger2002} by documenting that benefit design and benefit delivery are complements: expanding statutory generosity during a recession may have diminishing returns if administrative capacity simultaneously contracts. Third, it contributes methodologically by applying modern shift-share diagnostics \citep{borusyak2022, goldsmithpinkham2020, adao2019} to a government capacity question.

The rest of the paper proceeds as follows. Section 2 describes the institutional setting. Section 3 presents the data. Section 4 develops the empirical strategy. Section 5 reports results. Section 6 discusses implications and limitations.

\section{Institutional Background}

\subsubsection*{The UI System} Unemployment insurance in the United States is a federal-state partnership. The federal government sets broad eligibility and benefit guidelines through the Federal Unemployment Tax Act (FUTA), while states administer claims, determine benefit amounts within federal parameters, and process payments \citep{wandner2010}. This decentralized structure means that the speed and quality of benefit delivery varies substantially across states.

\subsubsection*{The Capacity Crunch} Two forces collided during the Great Recession. On the demand side, initial UI claims surged from approximately 300,000 per week in early 2007 to over 660,000 in March 2009---a 120\% increase in under two years. On the supply side, state governments faced severe revenue shortfalls: 26 states imposed layoffs and 22 implemented furloughs in FY2010 alone \citep{gao2012}. Federal administrative funding through FUTA had declined roughly 30\% in real terms from 1999 to 2009, leaving many agencies with aging IT systems and insufficient staff even before the recession \citep{oig2012}. The result was a capacity crunch in which agencies needed \emph{more} staff precisely when they were losing them.

\subsubsection*{First Payment Timeliness} The Department of Labor tracks first payment timeliness through the Benefits Timeliness and Quality (BTQ) reporting system. States report the share of intrastate UI first payments made within 7, 14, 21, and 28+ days of the first compensable week. The federal performance target is 87\% within 14 days. National 14-day timeliness averaged roughly 90\% in 2006--2007 but fell to approximately 83\% by 2009--2010, with some states dropping below 50\% \citep{oig2012}.

\section{Data}

I construct a state-year panel covering 49 states from 2006 to 2012 (343 observations) by merging four administrative data sources.

\subsubsection*{Outcome} First payment timeliness comes from DOL BTQ Category 1 reports. The primary outcome is the share of first payments made within 14 days of the first compensable week. This variable ranges from 46.3\% to 99.3\% in the sample, with a mean of 86.3\% and standard deviation of 8.4 percentage points.

\subsubsection*{Bartik Instrument} I construct a shift-share predicted claims shock using the Census Quarterly Workforce Indicators (QWI). For each state $s$ in year $t$, the Bartik shock is:
\begin{equation}
B_{st} = \sum_{k} \omega_{sk}^{2006} \cdot g_{-s,kt}
\label{eq:bartik}
\end{equation}
where $\omega_{sk}^{2006}$ is the share of state $s$'s 2006 employment in industry $k$ (19 two-digit NAICS sectors), and $g_{-s,kt}$ is the leave-one-out national employment growth rate in industry $k$ at time $t$. The leave-one-out construction follows \citet{borusyak2022}.

\subsubsection*{Administrative Capacity} State government full-time equivalent (FTE) employment comes from the Census Annual Survey of Public Employment and Payroll (ASPEP) for 2007. I measure administrative \emph{thinness} as the inverse of government FTE per 1,000 private-sector workers. States range from 19 to 90 government employees per 1,000 private workers, with a mean of 37.3.

\subsubsection*{Claims} Annual initial claims come from DOL ETA 539 weekly reports, aggregated to the state-year level.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

\subsection{Identification}

The challenge is that actual UI claims are endogenous: states with worse economic conditions file more claims \emph{and} may simultaneously face other pressures on their agencies. I instrument actual claims with the Bartik predicted shock from Equation \eqref{eq:bartik}, which isolates the component of claims pressure attributable to a state's predetermined (2006) industry composition.

\subsubsection*{First Stage} The first stage is:
\begin{equation}
\ln(\text{Claims}_{st}) = \pi_0 + \pi_1 B_{st} + \delta_s + \theta_t + \nu_{st}
\end{equation}
where $\delta_s$ and $\theta_t$ are state and year fixed effects. The coefficient $\pi_1$ captures how well the Bartik shock predicts actual claims variation. I cluster standard errors at the state level.

\subsubsection*{Second Stage} The structural equation is:
\begin{equation}
\text{Timeliness}_{st} = \beta_0 + \beta_1 \widehat{\ln(\text{Claims}_{st})} + \delta_s + \theta_t + \varepsilon_{st}
\end{equation}
where $\widehat{\ln(\text{Claims}_{st})}$ is the predicted value from the first stage. The coefficient $\beta_1$ identifies the causal effect of a demand surge on processing timeliness.

\subsubsection*{Capacity Interaction} The key specification interacts the Bartik shock with pre-recession administrative thinness:
\begin{equation}
\text{Timeliness}_{st} = \gamma_0 + \gamma_1 B_{st} + \gamma_2 (B_{st} \times \text{Thinness}_s) + \delta_s + \theta_t + \varepsilon_{st}
\label{eq:interaction}
\end{equation}
Since $\text{Thinness}_s$ is time-invariant and absorbed by state fixed effects, $\gamma_2$ identifies how administrative capacity \emph{moderates} the effect of demand shocks on timeliness.

\subsection{Identifying Assumptions}

The Bartik instrument requires two key assumptions \citep{goldsmithpinkham2020, borusyak2022}: (1) pre-recession industry composition affects timeliness only through the claims channel; and (2) conditional on state and year fixed effects, pre-recession staffing levels are uncorrelated with unobserved determinants of processing quality.

I probe these assumptions through Rotemberg weight diagnostics (to assess which industries drive the instrument), leave-one-industry-out analysis, and a placebo test on the pre-recession period.

\subsection{Threats to Validity}

\subsubsection*{Pre-Trends} In an event study using 2008 Bartik exposure interacted with year indicators, the 2006 coefficient is borderline significant ($p = 0.05$). This may reflect that states with more recession-vulnerable industry compositions were also structurally underprepared---consistent with the capacity erosion thesis but a concern for strict exogeneity. However, the pre-recession reduced-form placebo (2006--2007 only) shows no significant relationship ($p = 0.17$), and the event study coefficients are imprecisely estimated throughout, limiting the informativeness of any single pre-period coefficient.

\subsubsection*{Federal Interventions} The American Recovery and Reinvestment Act (ARRA) provided administrative grants to some states. If these grants correlated with both the Bartik shock and timeliness improvements, the IV estimate would be biased. I note this as a limitation.

\section{Results}

\subsection{First Stage and Reduced Form}

\input{tables/tab2_firststage.tex}

Table 2 reports the first stage and reduced-form regressions. The Bartik shock is a strong predictor of actual claims: a one-standard-deviation increase in predicted employment contraction increases log initial claims by 0.57 ($F = 68$), well above conventional thresholds for instrument relevance. Construction (28\%), manufacturing (20\%), and administrative services (19\%) contribute the largest Rotemberg weights.

The reduced form (Column 1) shows a negative but imprecise relationship between the Bartik shock and timeliness ($p = 0.39$). However, the interaction with administrative thinness (Column 2) is significant at the 5\% level: the coefficient on Bartik $\times$ Thinness is 2.20 (SE = 0.86). Since thinness is the negative of staff per 1,000 workers, a state one standard deviation thinner (14.6 fewer staff per 1,000 workers) has thinness 14.6 points higher. For a one-standard-deviation Bartik contraction of $-0.022$, the differential effect between a thin and average state is roughly $2.20 \times 0.022 \times 14.6 \approx 0.7$ percentage points per year---a modest but statistically significant differential that cumulates over the multi-year recession.

\subsection{Two-Stage Least Squares}

\input{tables/tab3_main.tex}

Table 3 reports the 2SLS estimates. The full-sample IV estimate of log claims on timeliness (Column 1) is positive but far from significant (11.0, SE = 12.6). The point estimate is wrong-signed relative to the theoretical prediction that claims surges \emph{reduce} timeliness, though the wide confidence interval includes economically meaningful negative effects. This null is consistent with the possibility that, on average, states adjusted to demand surges---through overtime, federal assistance, or processing shortcuts---in ways that prevented catastrophic aggregate delays.

Columns 3 and 4 split the sample at the median of pre-recession staffing. In thin states (below-median government FTE per worker), the IV coefficient is $-1.2$ (SE = 15.7)---correctly signed but imprecise. In thick states, the coefficient is 23.2 (SE = 18.9)---positive but statistically indistinguishable from zero. The sign difference across subsamples, while not individually precise, is consistent with the reduced-form interaction: thin-staffed states experienced relative deterioration while thick-staffed states were able to absorb demand shocks, plausibly through reserve capacity, overtime authorization, or targeted receipt of federal ARRA administrative grants.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

Table 4 presents robustness checks. Panel B shows that the main 2SLS estimate is stable when dropping the highest-Rotemberg-weight industries individually. The largest movement occurs when dropping manufacturing: the coefficient shifts from 11.0 to 20.9, but remains far from significant. Panel C confirms the null extends to alternative timeliness thresholds (7-day and 21-day). Panel D shows the placebo test is clean: the pre-recession reduced form is not significant ($p = 0.17$).

\section{Discussion}

I emphasize that the direct 2SLS estimate is a null, and the central finding---the reduced-form interaction---identifies a pattern consistent with the capacity erosion mechanism but falls short of a fully identified heterogeneous causal effect. The aggregate null---that claims surges did not, on average, reduce timeliness---is itself informative. It suggests that the UI system possesses substantial absorptive capacity, consistent with federal ARRA administrative grants and state-level emergency measures documented by the DOL Inspector General \citep{oig2012}. But this absorptive capacity was not uniformly distributed. States that entered the recession with thinner administrative infrastructure experienced differentially larger timeliness declines, a pattern that holds across leave-one-industry-out specifications and is absent in the pre-recession placebo.

This finding connects to the broader administrative burden literature. \citet{herd2018} argue that administrative frictions---application complexity, documentation requirements, processing delays---function as de facto policy instruments that ration access to benefits. My results extend this framework from \emph{application} burdens to \emph{delivery} burdens: even after a claim is approved, the speed of payment depends on institutional capacity. In a state where the typical first payment arrived in 14 days rather than 7, a laid-off worker's consumption smoothing was interrupted for an additional week---precisely the period when liquidity constraints bind hardest \citep{ganong2019, chetty2008}.

The results also have implications for fiscal federalism. The UI system's federal-state structure means that administrative capacity is a state-level choice variable, but the consequences of underfunding fall on workers and on the macroeconomy \citep{chodorowreich2019}. If thin-staffed states deliver benefits more slowly during downturns, the automatic stabilizer function of UI is systematically weaker in precisely the places that may need it most.

\subsubsection*{Limitations} The annual frequency of the BTQ data prevents analysis of within-year dynamics. Monthly data would enable sharper event-study designs and more precise estimates. The pre-trend in the 2006 event-study coefficient, while borderline, suggests that pre-recession industry composition may correlate with pre-existing administrative trajectories. Future work should explore whether specific ARRA provisions differentially aided thin-staffed states, and whether the capacity crunch affected benefit \emph{amounts} or \emph{denial rates} in addition to processing speed.

\section{Conclusion}

The safety net is only as strong as the institutions that deliver it. During the Great Recession, Congress expanded UI benefits to an unprecedented 99 weeks, but the agencies tasked with writing the checks were hemorrhaging staff. I find that while the UI system on average absorbed the claims surge without catastrophic delays, states with thinner pre-recession administrative capacity experienced significantly larger timeliness declines---an interaction that identifies the binding constraint as institutional capacity, not statutory design. The hollow safety net is not a story about bad policy but about the gap between what programs promise and what agencies can deliver.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{DOL BTQ Data} First payment timeliness data come from the Department of Labor's Benefits Timeliness and Quality (BTQ) Category 1 reports, accessed via POST requests to \texttt{oui.doleta.gov/unemploy/btq/btqrpt.asp}. The data represent the annual share of intrastate UI first payments made within specified time thresholds ($\leq$7 days, 14 days, 21 days, etc.) from the first compensable week. Two states are dropped due to missing data, yielding 49 states over 2006--2012.

\subsubsection*{Census QWI} Industry employment data for the Bartik instrument come from the Census Bureau's Quarterly Workforce Indicators (QWI), accessed via the Census API. I use 19 two-digit NAICS sectors covering all private industries, aggregated from quarterly to annual frequency. Employment shares for the Bartik base year (2006) are computed as state-industry employment divided by total state employment across all 19 sectors.

\subsubsection*{Census ASPEP} State government employment data come from the 2007 Annual Survey of Public Employment and Payroll (ASPEP). I use total state government full-time equivalent (FTE) employees (function code 000) as the capacity measure.

\subsubsection*{DOL ETA 539} Weekly initial claims data come from DOL Employment and Training Administration Form 539, covering all 50 states and DC from 1986 to present. I aggregate weekly initial claims (column c3) to annual totals.

\subsubsection*{BEA State Personal Income} State personal income data from the Bureau of Economic Analysis (SAINC1 table) serve as a control variable.

\section{Identification Appendix}

\subsubsection*{Rotemberg Weights} Following \citet{goldsmithpinkham2020}, I compute approximate Rotemberg weights to assess which industries drive the Bartik instrument. Construction (27.7\%), manufacturing (20.3\%), and administrative/waste services (19.0\%) together account for 67\% of the instrument's variation. These are precisely the sectors that experienced the largest employment contractions during 2008--2010, confirming that the instrument captures recession-driven demand shocks rather than idiosyncratic industry trends.

\subsubsection*{Event Study} Using the 2008 Bartik exposure interacted with year indicators (reference year: 2007), the event study shows: 2006 ($-402$, $p = 0.05$), 2008 ($-99$, $p = 0.60$), 2009 ($65$, $p = 0.87$), 2010 ($-439$, $p = 0.32$), 2011 ($-556$, $p = 0.14$), 2012 ($-105$, $p = 0.78$). The borderline 2006 coefficient is discussed in the text.

\section{Robustness Appendix}

The leave-one-industry-out analysis (Table 4, Panel B) drops the five highest-Rotemberg-weight industries one at a time. The 2SLS coefficient varies from 6.0 (dropping mining) to 20.9 (dropping manufacturing), but none achieves statistical significance. The first-stage $F$-statistic ranges from 19.8 to 70.6, remaining above 10 in all specifications.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
