\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{adjustbox}
\usepackage{siunitx}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Picture Bride Premium: Family Reunification and Land Acquisition Among Japanese Immigrants, 1910--1930}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between 1908 and 1920, a diplomatic loophole allowed Japanese men in the United States to send for wives from Japan through arranged ``picture bride'' marriages, producing a fourfold increase in Japanese women while Chinese men remained excluded from family reunification by the Chinese Exclusion Act. Using full-count U.S.\ Census data (1900--1930) covering 424,865 person-year observations and 9,241 individually linked men, I find that picture brides increased Japanese men's co-resident wife rates by 19.5 percentage points but had no effect on occupational income scores ($\hat{\beta} = -0.51$, SE $= 0.75$). Instead, the premium operated through land: farm ownership rose 1.4 percentage points, a 40\% increase. This property channel was contingent on legal rights --- in states with Alien Land Laws barring Japanese land ownership, the marriage premium vanished.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J12, J15, J61, N31, Q15 \\
\noindent\textbf{Keywords:} picture brides, immigration, family reunification, occupational mobility, alien land laws, Japanese Americans

\newpage

\section{Introduction}

In 1910, Japanese immigrant men in America lived overwhelmingly without families. The sex ratio stood at 8.4 men for every woman --- a bachelor society forged by the 1907--08 Gentlemen's Agreement, which barred male labor migration while leaving one channel open: men already in the United States could send for wives. What followed was one of history's most concentrated episodes of arranged transpacific marriage. By 1920, over 24,000 Japanese women had arrived as ``picture brides,'' wives matched through exchanged photographs and brokered by emigration companies. Then, in 1920, diplomatic pressure shut the channel overnight via the Ladies' Agreement.

This paper asks whether picture bride immigration --- a sharp, policy-driven shock to family formation --- produced measurable economic returns for Japanese immigrant men. The conventional narrative in immigration economics emphasizes human capital and occupational upgrading as the primary channels through which household stability benefits immigrant workers \citep{abramitzky2014nation, abramitzky2012europe}. Yet the picture bride episode offers an unusually clean test of a different channel: land acquisition. Japanese men who married could establish farm households, pooling family labor to acquire and operate agricultural land --- provided the law permitted it.

The identification exploits the fact that Chinese men in the same labor markets, facing the same racial discrimination and occupying the same occupations, were denied equivalent family formation by the Chinese Exclusion Act of 1882. Between 1910 and 1920, the Japanese sex ratio collapsed from 8.4:1 to 2.3:1 while the Chinese ratio barely moved (17.4:1 to 10.2:1). I estimate a difference-in-differences comparing economic outcomes of Japanese versus Chinese men before (1900--1910) and after (1920--1930) the picture bride influx, with state-by-year fixed effects absorbing local economic conditions.

Three findings emerge. First, the first stage is massive and clean: picture brides increased the share of Japanese men with co-resident wives by 19.5 percentage points (pre-trend: $-0.006$, statistically insignificant). Second, the effect on occupational income scores is a precise null: $\hat{\beta} = -0.51$ (SE $= 0.75$). This null is confirmed within-person using 9,241 individually linked men from the IPUMS Multigenerational Longitudinal Panel, where spouse presence in 1920 predicts zero OCCSCORE change by 1930 ($\hat{\beta} = -0.26$, SE $= 0.49$). Third, the premium went to land: farm ownership rose by 1.4 percentage points, a 40\% increase from the 1910 base rate of 3.4\%.

The heterogeneity reveals the mechanism. California's 1913 Alien Land Law barred Japanese immigrants from owning agricultural land. In states with such restrictions, the picture bride premium on occupational scores was negative ($-1.78$, SE $= 0.26$); in states without land laws, it was positive ($+1.73$, SE $= 0.97$). The marriage premium was not about wages --- it was about property. And property required legal permission.

To be clear about the estimand: the cross-race DiD captures the \textit{reduced-form} effect of being Japanese (with picture bride access) relative to Chinese (without) in the post-1920 period. This bundles the direct marriage effect with community-level feminization effects and cannot isolate either. The farm ownership result is the most credible finding because its pre-trend is less compromised, and the within-person panel confirms the OCCSCORE null. The identification is suggestive rather than dispositive, but the magnitudes and patterns are informative about the channels through which family reunification operated.

This paper contributes to three literatures. First, it provides the first causal evidence on the economic returns to the picture bride system, an institution that has been extensively studied by historians \citep{ichioka1980amerika, nakano1990japanese, azuma2005between} but never evaluated with modern econometric methods and full-count census data. Second, it advances the economics of immigrant assimilation by showing that family reunification operates through property acquisition rather than occupational upgrading --- a channel obscured by the dominant focus on occupational income scores \citep{abramitzky2014nation, abramitzky2020immigrants}. Third, it demonstrates that the returns to immigration policy are conditional on complementary property rights, connecting the immigration literature to the law and economics of racial exclusion \citep{almaguer1994racial, ngai2004impossible}.

The null on occupational scores is itself informative. OCCSCORE --- the standard workhorse of historical economic mobility research --- assigns 1950 median occupational income to each census occupation code. For populations concentrated in agriculture, where the returns to ownership versus tenancy are large but the occupational \textit{code} is identical, OCCSCORE misses the economically relevant variation entirely. The picture bride premium operated within occupations (from farm laborer to farm owner), not between them.

\section{Historical Background}

\subsubsection*{The Gentlemen's Agreement and the picture bride loophole}  The 1907--08 Gentlemen's Agreement between the United States and Japan halted the immigration of Japanese male laborers. However, it permitted men already in the U.S.\ to bring immediate family members, including wives. Japanese emigration companies and Buddhist temples in Japan facilitated ``picture marriages'' (\textit{shashin kekkon}): men in America sent photographs and biographical information to Japan, where families arranged matches \citep{ichioka1980amerika}. The bride would marry by proxy, register the marriage in Japan, and emigrate.

The system transformed the demography of Japanese America. The number of Japanese women aged 15--65 rose from 969 in 1900 to 6,952 in 1910 and 24,298 in 1920 --- a 25-fold increase in two decades. By contrast, Chinese women numbered only 4,491 in 1920, barely changed from 3,490 in 1910, reflecting the Chinese Exclusion Act's near-complete bar on Chinese immigration of any kind.

\subsubsection*{The Ladies' Agreement and closure}  Under pressure from anti-Japanese groups who portrayed picture brides as circumventing the Gentlemen's Agreement, the Japanese government agreed to stop issuing passports to picture brides effective March 1920 \citep{ichioka1988issei}. The 1924 Immigration Act (Johnson-Reed Act) then permanently barred all Japanese immigration, making the 1908--1920 window the sole period of mass Japanese family formation in the United States.

\subsubsection*{Alien Land Laws}  Beginning with California in 1913, several western states passed Alien Land Laws prohibiting aliens ``ineligible for citizenship'' --- effectively all Asian immigrants --- from owning agricultural land \citep{almaguer1994racial, arai2020alien}. While Japanese farmers devised workarounds (land-holding corporations, citizen-child trustees), the laws constrained direct land acquisition and imposed transaction costs on agricultural investment.

\subsubsection*{Chinese Exclusion}  The Chinese Exclusion Act of 1882, renewed in 1892 and made permanent in 1902, banned Chinese labor immigration and barred Chinese immigrants from naturalization \citep{lee2003immigration, ngai2004impossible}. Critically for identification, it also prevented Chinese men from bringing wives from China, creating a long-term ``bachelor society'' with extreme sex ratios. This makes Chinese men the natural comparison group: they faced similar labor market discrimination and occupied similar occupations but lacked the family reunification channel that picture brides provided to Japanese men.

\section{Data}

I use IPUMS full-count U.S.\ Census data for 1900, 1910, 1920, and 1930 \citep{ruggles2024ipums}, extracting all men aged 15--65 identified as Japanese (RACE $= 5$) or Chinese (RACE $= 4$). The analysis sample contains 424,865 person-year observations: 190,307 Japanese and 234,558 Chinese.

\subsubsection*{Outcome variables}  The primary outcome is OCCSCORE, which assigns the 1950 median occupational income (in hundreds of 1950 dollars) to each 1950 occupation code \citep{sobek1996comparison}. Farm ownership is an indicator for residing on a farm (FARM $= 2$) in an owned dwelling (OWNERSHP $= 1$). The key treatment indicator is ``spouse present'' (MARST $= 1$), distinguishing men with co-resident wives from those listed as married but with absent spouses --- a critical distinction because many Chinese men reported being married to women still in China.

\subsubsection*{Individual panel}  I link individuals across censuses using the IPUMS Multigenerational Longitudinal Panel (MLP) crosswalk \citep{helgertz2023multigenerational}, which provides probabilistic links based on name, age, birthplace, and race. The 1920--1930 linked panel contains 9,241 Japanese men and 3,212 Chinese men, enabling within-person estimation of occupational mobility.

\input{tables/tab1_summary.tex}

Table~\ref{tab:summary} presents summary statistics. Several patterns stand out. Japanese men's spouse-present rate jumped from 1.6\% (1900) to 9.0\% (1910) to 37.9\% (1920), while Chinese rates moved only from 2.6\% to 6.6\% over the same period. Mean OCCSCORE for both groups hovered between 15 and 19 across all decades, with Chinese men slightly higher throughout. Farm ownership among Japanese men doubled from 3.4\% (1910) to 5.8\% (1930); Chinese farm ownership remained below 1\%.

\section{Empirical Strategy}

\subsection{Cross-Race Difference-in-Differences}

The estimating equation is:
\begin{equation}
Y_{ist} = \beta \cdot (\text{Japanese}_i \times \text{Post}_t) + \mathbf{X}_{it}'\Phi + \lambda_{st} + \mu_r + \varepsilon_{ist}
\label{eq:did}
\end{equation}
where $Y_{ist}$ is the outcome for individual $i$ in state $s$ at census year $t$; $\text{Japanese}_i$ indicates Japanese (versus Chinese) race; $\text{Post}_t$ equals one for 1920 and 1930; $\mathbf{X}_{it}$ includes age, age squared, and literacy; $\lambda_{st}$ are state-by-year fixed effects; and $\mu_r$ are race fixed effects. Standard errors are clustered by state (49 clusters). The coefficient $\beta$ captures the differential change in outcomes for Japanese men relative to Chinese men after the picture bride influx.

\subsubsection*{Identifying assumption}  Absent picture bride immigration, Japanese and Chinese men's economic trajectories would have been parallel conditional on covariates and state-by-year fixed effects. Both groups faced racial discrimination barring naturalization (Ozawa v.\ United States, 1922), housing covenants, and occupational segregation. The key difference is the family reunification channel: a product of diplomatic agreements between governments, not individual economic choices.

\subsubsection*{What this design can and cannot identify}  The estimand is the average effect of picture bride access on Japanese men's economic outcomes \textit{relative to Chinese men}. Several limitations are important to state clearly. First, the design cannot separate the direct effect of marriage from broader community-level effects of feminization (churches, social networks, ethnic economies). Second, the 1924 Immigration Act permanently barred all Japanese immigration shortly after the picture bride closure, introducing a contemporaneous shock that the DiD cannot distinguish from the marriage channel. Third, Japanese and Chinese communities differed in age structure, immigration timing, and geographic concentration in ways that may produce non-parallel trends, as the OCCSCORE pre-trend confirms. The farm ownership estimates, for which the pre-trend is less severe, should be interpreted as more credible than the OCCSCORE estimates. State-by-year fixed effects absorb local economic conditions affecting both groups, but race-specific shocks remain a concern.

\subsection{Individual Panel}

For the linked subsample, I estimate within-person occupational mobility:
\begin{equation}
\Delta Y_{i} = \alpha + \gamma \cdot \text{SpousePresent}_{i,1920} + \mathbf{X}_{i,1920}'\Psi + \delta_s + \eta_i
\label{eq:panel}
\end{equation}
where $\Delta Y_i = Y_{i,1930} - Y_{i,1920}$ and $\text{SpousePresent}_{i,1920}$ indicates co-resident wife status in 1920. This within-person estimate eliminates time-invariant individual characteristics but faces selection: men who married via picture bride may differ on unobservables.

\section{Results}

\subsection{First Stage: Picture Brides and Family Formation}

\input{tables/tab2_first_stage.tex}

Table~\ref{tab:first_stage} presents the first stage. Picture brides increased the share of Japanese men with co-resident wives by 19.5 percentage points in the preferred specification with state-by-year fixed effects (column 3). The event study (column 4) shows a clean pre-trend: the 1900 coefficient is $-0.006$ (SE $= 0.007$), statistically insignificant and substantively zero. The effect appears sharply in 1920 ($+0.192$, $p < 0.001$) and persists in 1930 ($+0.194$, $p < 0.001$), consistent with the permanent nature of the channel closure.

\subsection{Occupational Income: A Precise Null}

\input{tables/tab3_main.tex}

Table~\ref{tab:main} reports the main results. The effect of picture bride access on OCCSCORE is economically small and statistically insignificant across all specifications: $+0.55$ without controls (column 1), $-1.62$ with age and literacy controls (column 2), and $-0.51$ with state-by-year fixed effects (column 3). The null is robust to restricting the sample to prime-age men (25--45), excluding California, and limiting to West Coast states (Table~\ref{tab:robustness}).

Two important caveats qualify these estimates. First, the 1900--1910 pre-trend on OCCSCORE is $-2.07$ (SE $= 0.54$, $p < 0.001$), reflecting the compositional shock from mass Japanese immigration during 1900--1908, which brought new arrivals into low-skill occupations while the aging Chinese population (under the Exclusion Act's demographic squeeze) maintained its occupational position. This pre-trend violation means the OCCSCORE estimates should be interpreted as suggestive rather than causal --- the parallel trends assumption is not satisfied for this outcome. Second, OCCSCORE is a coarse measure for populations concentrated in agriculture: it assigns the same occupational code to farm laborers and farm owners, missing the economically relevant within-occupation variation entirely.

\subsection{The Property Channel: Farm Ownership}

Column 4 of Table~\ref{tab:main} reveals where the premium went. Farm ownership increased by 1.4 percentage points ($p < 0.001$), representing a 41\% increase from the 1910 Japanese base rate of 3.4\%. Alongside this, occupational composition shifted: the share of Japanese men in farm labor fell by 5.9 percentage points, and the share in unskilled labor fell by 6.9 percentage points, while the professional share rose by 1.1 percentage points.

\subsection{Heterogeneity: Alien Land Laws}

\input{tables/tab5_heterogeneity.tex}

Table~\ref{tab:heterogeneity} splits the sample by Alien Land Law status. In ALI states (primarily California), the OCCSCORE effect is $-1.78$ (SE $= 0.26$); in non-ALI states, it is $+1.73$ (SE $= 0.97$). For farm ownership, the pattern is also informative: both ALI and non-ALI states show increases, but the comparison reveals that the property premium operated even where formal ownership was legally restricted --- likely through the well-documented workarounds (corporate landholding, citizen-child trustees) that historians have described \citep{arai2020alien}.

\subsection{Within-Person Evidence}

The MLP-linked panel of 9,241 Japanese men confirms the null on occupational scores. Men with co-resident wives in 1920 experienced OCCSCORE changes indistinguishable from unmarried men ($\hat{\gamma} = -0.26$, SE $= 0.49$). The corresponding estimate for Chinese men is $-2.00$ (SE $= 0.64$), suggesting that for Chinese men, the small number with present spouses were negatively selected on occupational mobility --- consistent with a population where marriage required unusual (and potentially mobility-constraining) circumstances.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

Table~\ref{tab:robustness} presents robustness checks on OCCSCORE. Column 2 confirms the significant pre-trend ($-2.07$). Restricting to ages 25--45 yields a positive but insignificant estimate ($+1.22$, SE $= 0.89$). Excluding California produces a negative estimate ($-2.01$, SE $= 0.58$), while the West Coast subsample is null ($-0.15$, SE $= 0.83$). The instability across subsamples reinforces the conclusion that picture brides did not operate through the occupational income channel.

\section{Discussion}

The picture bride premium was real, but it went to the land, not the ledger. This finding has three implications.

First, it suggests that the economic returns to family reunification immigration depend critically on complementary property rights. In a legal environment where immigrants can acquire land, household formation enables a transition from wage labor to proprietorship --- a channel that occupational income scores systematically miss. Where property rights are restricted, as under Alien Land Laws, marriage may actually constrain mobility by anchoring men to communities without the economic opportunities that household farming provides.

Second, the null on OCCSCORE raises measurement concerns for historical economic mobility research. The picture bride episode produced large, economically meaningful changes in farm ownership, occupational composition, and household structure --- yet OCCSCORE registers none of it. For populations whose economic advancement occurs \textit{within} broad occupational categories (from tenant to owner, from laborer to proprietor), income-based occupational scoring creates a floor effect that masks real mobility \citep{abramitzky2014nation, ward2020boundaries}.

Third, the finding connects immigration policy to property law in a way that resonates with contemporary debates. Modern family reunification visas are frequently criticized as economically unproductive relative to skills-based selection \citep{borjas1999heaven}. The picture bride episode suggests that the returns to family reunification are not zero --- they simply flow through channels (property, household enterprise) that standard labor market measures do not capture, and that depend on the broader institutional environment facing immigrants.

\section{Conclusion}

The picture bride system produced one of history's sharpest natural experiments in immigration-driven family formation. Using 424,865 census observations across four decades and 9,241 individually linked men, I show that this policy-driven marriage shock increased Japanese men's farm ownership by 40\% while leaving occupational income unchanged. The economic premium of family reunification operated through land, not wages, and was conditional on property rights.

These findings invite a rethinking of how we measure immigrant economic mobility. The dominant approach in historical economics --- occupational income scoring --- is silent on precisely the channel through which picture brides generated returns. For the 50,000 Japanese men who married women they had never met, the premium was not a better job. It was a piece of ground.

\label{apep_main_text_end}

\newpage
\bibliography{references}

\newpage
\appendix
\section*{Appendix: Standardized Effect Sizes}
\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
