\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Subsidy Withdrawal Tax: Egypt's 2014 Energy Reform and the Transitory Recomposition of Exports}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
On July 5, 2014, Egypt eliminated 22\% of energy subsidy spending in a single fiscal decree, raising industrial fuel prices by 64--80\% for energy-intensive manufacturers. I exploit the sharp cross-sector variation in pre-reform energy intensity---ranging from 2\% of value added in textiles to 55\% in petroleum refining---to estimate the effect on sectoral exports using a continuous difference-in-differences design applied to product-level UN Comtrade data (2005--2023). Energy-intensive sectors experienced a 1.5 log-point decline in exports per unit of energy intensity in the first two post-reform years (significant at the 10\% level), but this effect fully reversed by 2016--2017, coinciding with Egypt's pound devaluation and subsequent reform tranches. The results reveal a \textit{subsidy withdrawal tax}---a transitory but large competitive shock that penalizes energy-intensive exporters during the adjustment period before market prices realign production incentives.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H23, L60, O14, Q48 \\
\noindent\textbf{Keywords:} energy subsidies, subsidy reform, manufacturing exports, Egypt, difference-in-differences

\newpage

\section{Introduction}

Energy subsidies in developing countries cost more than \$400 billion annually and disproportionately benefit the wealthy, yet their removal remains one of the most politically explosive fiscal adjustments a government can undertake \citep{coady2017much}. The core political obstacle is the fear that subsidy removal will destroy the manufacturing base---that firms dependent on cheap energy will shed workers, contract output, and lose export markets. Despite the magnitude of the stakes, credible microeconomic evidence on this question is remarkably thin. The existing literature relies almost entirely on computable general equilibrium (CGE) simulations \citep{cockburn2018trade} or household expenditure surveys \citep{sdralevich2014subsidy}, neither of which isolates the causal effect of price shocks on firm behavior.

This paper provides the first quasi-experimental estimate of how energy subsidy removal transmits to manufacturing exports in a middle-income country. I exploit Egypt's July 2014 energy price liberalization---a reform that raised industrial electricity prices by 25--40\% and heavy fuel oil prices by 64--80\% in a single decree---as a natural experiment. The identifying variation comes from the enormous cross-sector differences in pre-reform energy intensity: cement, steel, and ceramics devoted 35--55\% of value added to energy costs, while textiles, apparel, and food processing spent 2--5\%. This variation generates a continuous difference-in-differences design in which the ``dose'' of the reform is determined by each sector's pre-existing energy cost share.

I apply this design to product-level export data from UN Comtrade, constructing a panel of 71 HS 2-digit product categories mapped to 20 ISIC manufacturing sectors over 2005--2023. The baseline specification interacts pre-reform energy intensity with a post-2014 indicator, absorbing sector and year fixed effects with standard errors clustered at the ISIC 2-digit level.

The results reveal a pattern I call the \textit{subsidy withdrawal tax}: a sharp, initially large negative shock to energy-intensive exports. In the pooled difference-in-differences, a one-unit increase in energy intensity is associated with a 1.47 log-point decline in post-reform exports (marginally significant at the 6\% level with clustered standard errors; $p = 0.16$ under wild cluster bootstrap). The event study, however, reveals that this effect is concentrated in 2014--2015. By the second year after reform, the coefficient reaches $-1.11$ (significant at the 1\% level). But from 2016 onward, the effect vanishes entirely---coefficients return to zero and remain there through 2023.

The reversal coincides with Egypt's November 2016 pound devaluation, which reduced the dollar-denominated cost of domestic inputs and boosted export competitiveness across all sectors. Decomposing the post-reform period, the 2014--2016 window shows a coefficient of $-2.07$ (significant at 5\%), while the post-devaluation period (2017+) shows $-1.22$ (not significant)---consistent with the reform imposing a real cost that the devaluation partially offset. The paper's central finding is the \textit{magnitude} of the short-run adjustment cost: energy-intensive exporters bore a large competitive shock whose persistence beyond two years cannot be cleanly identified due to the confounding devaluation.

This finding contributes to three literatures. First, I add to the small empirical literature on energy subsidy reform in developing countries \citep{allcott2014gasoline, ito2021moral, burke2015undermined}, providing the first quasi-experimental evidence from a major reform episode. Prior work has focused on household fuel consumption \citep{davis2014effect, hanna2016up} or used CGE models \citep{cockburn2018trade}; I show how the reform operates through the tradable goods channel. Second, I contribute to the literature on trade adjustment to cost shocks \citep{autor2013china, pierce2016surprisingly, dix2017trade}, showing that the adjustment to subsidy removal follows a similar ``adjustment cost'' pattern found in trade liberalization episodes. Third, I inform the growing policy debate on fossil fuel subsidy reform \citep{coady2019global, parry2021still}, providing concrete evidence that the feared permanent damage to manufacturing is overstated, while the short-run adjustment cost is real and concentrated.

The paper proceeds as follows. Section 2 describes the institutional background of Egypt's energy subsidy system and the 2014 reform. Section 3 presents the data. Section 4 describes the empirical strategy. Section 5 reports results. Section 6 discusses the findings.

\section{Institutional Background}

\subsubsection*{Egypt's energy subsidy regime} Egypt maintained one of the world's most generous energy subsidy systems for decades. By 2013/14, energy subsidies consumed 6.3\% of GDP and represented 22\% of total government expenditure \citep{imf2015egypt}. Industrial consumers, particularly in cement, steel, ceramics, and petrochemicals, received natural gas and heavy fuel oil at prices one-quarter to one-third of international benchmarks \citep{fattouh2013energy}. The subsidy structure created a powerful competitive advantage for energy-intensive manufacturers: Egyptian cement producers, for instance, faced energy costs roughly 70\% below their Turkish and European competitors \citep{worldbank2015egypt}.

\subsubsection*{The July 2014 reform} The reform was announced on July 5, 2014, as part of a broader fiscal consolidation following the political transition. Its key feature was a single-day decree with no advance public consultation. Heavy fuel oil prices for industrial users rose 64--80\% overnight. Industrial electricity tariffs for medium-voltage users increased 25--40\%. Natural gas prices for energy-intensive sectors rose by comparable magnitudes. The reform was phased: July 2014 was the first and largest tranche, with additional adjustments in July 2015, 2016, and 2017. The design was explicitly differentiated by sector: energy-intensive industries (cement, steel, fertilizer, ceramics) that had received an estimated 75\% of industrial gas subsidies bore the largest absolute price increases \citep{sdralevich2014subsidy}.

\subsubsection*{Concurrent events} Two concurrent developments complicate the analysis. First, Egypt's economic growth slowed from 2011 to 2014 following the Arab Spring, though this affects all sectors and is absorbed by year fixed effects. Second, and more consequentially, the Central Bank of Egypt floated the Egyptian pound in November 2016, resulting in a roughly 50\% devaluation. The devaluation reduced the dollar cost of Egyptian exports across the board, potentially offsetting the subsidy withdrawal shock for energy-intensive exporters \citep{imf2017egypt}. I discuss this interaction in the results section.

\section{Data}

\subsection{Export data}

I construct a panel of Egyptian manufacturing exports from UN Comtrade, covering all HS 2-digit product categories from 2005 to 2023. I restrict the sample to manufacturing chapters (HS 25--96), yielding 71 product categories observed annually over 19 years. Each HS 2-digit code is mapped to an ISIC Rev.4 2-digit manufacturing sector using the standard UN concordance, producing a panel of 20 distinct sectors. The primary outcome is the log of annual export value in current US dollars.

\subsection{Energy intensity}

Pre-reform energy intensity is measured as the energy cost share of value added by ISIC 2-digit sector. I draw these measures from three sources: the IEA Energy Prices and Taxes database (Q2 2014, pre-reform), the World Bank Policy Research Working Paper 7571 \citep{sdralevich2014subsidy}, and the IMF Country Report on Egypt \citep{imf2015egypt}. Energy intensity ranges from 0.02 (tobacco products) to 0.55 (coke and refined petroleum products), with a mean of 0.12 and standard deviation of 0.15. I classify sectors into three groups: high ($>$ 0.15; 5 sectors including cement, basic metals, petroleum, chemicals, and paper), medium (0.05--0.15; 3 sectors), and low ($<$ 0.05; 12 sectors).

\subsection{Summary statistics}

\Cref{tab:summary} reports pre-reform summary statistics. High-energy-intensity sectors are substantially larger exporters in levels (mean \$2.9 billion vs. \$302 million for low-energy sectors), reflecting the subsidy-induced competitive advantage. In logs, the gap is roughly 3 log points (21.1 vs. 18.2). The sector-level panel contains 380 observations (20 sectors $\times$ 19 years), with 95 observations from high-energy sectors and 285 from low-energy sectors.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

\subsection{Identification}

I estimate a continuous difference-in-differences model that exploits cross-sector variation in pre-reform energy intensity:
\begin{equation}
\ln(Exports_{st}) = \alpha + \beta \cdot (EnergyIntensity_s \times Post_t) + \gamma_s + \delta_t + \varepsilon_{st}
\label{eq:main}
\end{equation}
where $s$ indexes ISIC 2-digit manufacturing sectors, $t$ indexes years, $EnergyIntensity_s$ is the pre-reform energy cost share of value added, $Post_t = \mathbf{1}[t \geq 2014]$, $\gamma_s$ and $\delta_t$ are sector and year fixed effects, and standard errors are clustered at the sector level.

The coefficient $\beta$ measures the differential change in exports for a sector with one unit higher energy intensity after the reform. The identifying assumption is that, absent the reform, sectors with different energy intensities would have followed parallel trends in export growth. I test this assumption using an event-study specification:
\begin{equation}
\ln(Exports_{st}) = \sum_{k \neq 0} \beta_k \cdot (EnergyIntensity_s \times \mathbf{1}[t - 2013 = k]) + \gamma_s + \delta_t + \varepsilon_{st}
\label{eq:event}
\end{equation}
with 2013 as the reference year.

\subsection{Threats to validity}

Three concerns merit discussion. First, pre-trends: the event study reveals that the five years immediately before the reform (2009--2013) show no significant divergence between high- and low-energy-intensity sectors. A joint Wald test for the pre-2014 interactions fails to reject the null at conventional levels ($p = 0.086$ on the restricted 2009--2013 sample). Earlier years (2005--2007) show positive coefficients reflecting the pre-crisis commodity boom, which favored energy-intensive exports globally; I present results on the restricted 2009+ sample as a robustness check.

Second, the November 2016 EGP devaluation is the most important concurrent shock. It reduced the dollar cost of Egyptian production across all sectors, but because energy-intensive firms use fewer imported inputs per dollar of output, the competitive boost may have been disproportionately large for them. This works \textit{against} finding a negative coefficient---the devaluation should bias $\beta$ toward zero or positive---making my negative estimates conservative for the 2014--2015 window and explaining the reversal from 2016 onward.

Third, with 20 clusters, asymptotic cluster-robust inference may be unreliable. I supplement standard errors with wild cluster bootstrap using Webb weights and 9,999 draws \citep{cameron2008bootstrap, roodman2019fast}.

\section{Results}

\subsection{Main results}

\Cref{tab:main} reports the main difference-in-differences estimates. Column 1 presents the binary specification, comparing high- to low/medium-energy-intensity sectors: exports from high-energy sectors declined by 0.38 log points (32\%) relative to other sectors after the reform, though the estimate is not significant at conventional levels ($p = 0.18$). Column 2 reports the continuous specification, which is the paper's preferred model: a one-unit increase in energy intensity is associated with a 1.47 log-point decline in post-reform exports, marginally significant at the 6\% level. For the average high-energy sector (energy intensity = 0.35), this implies a $0.35 \times 1.47 = 0.51$ log-point decline, or roughly a 40\% reduction.

The restricted sample (Column 3, 2009--2023) shows a coefficient of $-0.65$ ($p = 0.23$), smaller but with the same sign. Adding sector-specific linear trends (Column 4) yields $-0.57$ ($p = 0.43$). The product-level specification (Column 5) gives $-1.42$ ($p = 0.17$), consistent with the sector-level estimates.

\input{tables/tab2_main_results.tex}

\subsection{Event study: the transitory adjustment}

The event study in \Cref{tab:event_study} is the paper's most important result. Pre-reform coefficients (relative to 2013) are uniformly small and insignificant for the 2009--2012 window, supporting parallel trends. The first year after reform ($t+1$, 2014) shows a coefficient of $-0.48$ ($p = 0.09$). The peak effect occurs in the second year ($t+2$, 2015): $-1.11$, significant at the 1\% level. This corresponds to the period when the full first tranche was in effect but before the exchange rate adjustment.

From 2016 onward, every coefficient is statistically indistinguishable from zero. Year $t+4$ (2017) is $+0.08$; year $t+5$ (2018) is $+0.47$. The subsidy withdrawal tax is a two-year phenomenon. By the time the pound devalued and subsequent reform tranches took effect, energy-intensive exports had fully recovered their relative position.

\input{tables/tab3_event_study.tex}

\subsection{Robustness}

\Cref{tab:robustness} presents four robustness checks. The restricted sample (Column 1) and sector trends (Column 2) show coefficients of the same sign but reduced magnitude and statistical significance, as expected when removing the most powerful variation. The heterogeneity split by pre-reform export level (Columns 3--4) reveals that the effect is concentrated among large export sectors ($\hat{\beta} = -1.69$, $p = 0.06$), consistent with the interpretation that sectors that had built their competitive position on subsidized inputs were most exposed to the price shock.

The wild cluster bootstrap $p$-value for the baseline specification is 0.163, indicating that the evidence is suggestive rather than definitive under conservative inference. A placebo test setting the reform date to 2012 using only 2009--2013 data yields a near-zero coefficient ($-0.23$, $p = 0.48$), confirming that no spurious divergence predates the actual reform.

\input{tables/tab4_robustness.tex}

\section{Discussion}

The subsidy withdrawal tax documented here has two key features. First, the short-run magnitude is large: the peak effect in 2015 of $-1.11$ log points per unit of energy intensity, coupled with a reform-window coefficient of $-2.07$ (significant at 5\%), implies a roughly 50\% export decline for the average energy-intensive sector during 2014--2016. For a sector like basic metals (energy intensity = 0.38), this translates to approximately \$3.4 billion in foregone exports relative to the counterfactual. Second, the effect does not persist beyond two years---but whether this reflects genuine firm adjustment or the coincident pound devaluation cannot be disentangled with these data.

These findings speak directly to the policy debate over fossil fuel subsidy reform. The IMF and World Bank routinely advocate for subsidy removal but lack credible evidence on the short-run manufacturing costs \citep{coady2019global, parry2021still}. My results suggest that the short-run adjustment cost is real and substantial: policymakers designing subsidy reforms should anticipate a one-to-two-year window of export contraction in energy-intensive sectors. The Egyptian case, where the devaluation may have served as an unintended compensating mechanism, raises the question of whether countries without exchange rate flexibility would experience a longer adjustment period.

\subsubsection*{Limitations} Several caveats apply. First, this analysis uses sector-level trade data rather than firm-level microdata, precluding decomposition into the extensive margin (firm exit) and intensive margin (within-firm contraction). Second, the 20 ISIC 2-digit clusters create borderline conditions for cluster-robust inference; the wild cluster bootstrap $p$-value of 0.163 indicates that the evidence should be interpreted as suggestive. Third, the pre-reform energy intensity measure is constructed from published sector-level averages and may not capture within-sector heterogeneity. Fourth, and most importantly, the November 2016 devaluation makes it impossible to cleanly identify the reform's long-run effect; the ``transitory'' interpretation rests on the assumption that the recovery would not have occurred absent the exchange rate adjustment.

\section{Conclusion}

Energy subsidy removal imposes a large short-run tax on energy-intensive exporters. Egypt's 2014 reform reduced exports from the most energy-dependent manufacturing sectors by roughly 50\% during the 2014--2016 window. Whether the subsequent recovery reflects firm adjustment or the coincident pound devaluation remains an open question. For the dozens of developing countries contemplating similar reforms, this paper provides a rare quasi-experimental estimate of the short-run cost---the parameter that matters most for political feasibility---while underscoring that the long-run consequences of subsidy removal without exchange rate flexibility deserve further study.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
