\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Menthol's Last Drag: Product Substitution and the Market Non-Response to the EU Flavor Ban}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Menthol cigarettes once claimed a quarter of the Polish cigarette market, yet when the European Union banned them overnight, the tobacco market barely flinched. I exploit the uniform May 2020 ban across 28 countries with pre-existing variation in menthol market share (2\%--28\%) as a dose-response natural experiment. Using monthly Eurostat HICP data (2017--2024), I find that the relative price of tobacco---net of general inflation---was unchanged in high-menthol countries compared to low-menthol countries. A triple-difference stacking tobacco against placebo categories confirms the null. This powered zero is consistent with near-complete product substitution: menthol smokers switched to unflavored cigarettes rather than quitting. The result implies that flavor bans---the next frontier of tobacco regulation---may reduce product variety without reducing consumption.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I18, H23, L66 \\
\noindent\textbf{Keywords:} tobacco regulation, menthol cigarettes, product bans, consumer substitution, EU Tobacco Products Directive

\newpage

\section{Introduction}

When regulators ban a product that one in four consumers prefer, what happens to the market? The answer depends on whether consumers walk away or simply switch. In May 2020, the European Union removed menthol cigarettes from shelves across 28 countries, eliminating a product segment that ranged from negligible in southern Europe to dominant in Poland. The ban represented the world's largest-ever flavor restriction on tobacco, affecting a population of 450 million. Yet as this paper documents, the tobacco market absorbed the shock with remarkable equanimity.

This paper asks whether the EU menthol ban affected the relative price of tobacco in countries where menthol cigarettes constituted a larger share of the market. The question matters because the answer reveals the dominant adjustment margin. If menthol smokers quit, reduced demand should depress the relative tobacco price in high-menthol countries. If they substitute to unflavored cigarettes, the price index should remain undisturbed---the market reshuffles internally without net contraction. The distinction has direct implications for the ongoing regulatory frontier: the US FDA's proposed menthol ban, similar initiatives in Canada and New Zealand, and the broader question of whether product variety restrictions are effective public health tools.

I exploit the EU Tobacco Products Directive (2014/40/EU, Article 7), which banned cigarettes with characterizing flavors effective May 20, 2020. The ban date was uniform across all member states---no staggered adoption, no phase-in---creating a clean single-break design. Identification comes from cross-country variation in pre-ban menthol market share as a continuous treatment intensity. Poland (28\%), Finland (15\%), and Lithuania (12\%) were heavily exposed; Spain (2\%), Greece (2\%), and Cyprus (1.5\%) were nearly untouched. The research design is a dose-response difference-in-differences: countries with higher menthol dependence should show larger post-ban shifts if the ban disrupted markets.

A central identification challenge is that the ban coincided exactly with the COVID-19 pandemic. Countries with high menthol shares---predominantly Central and Eastern European---also experienced differential inflation in 2021--2023 driven by energy price shocks and proximity to the Russia-Ukraine conflict. I demonstrate that this confound is severe: na\"ive specifications using the level of the tobacco price index yield spurious positive estimates because menthol share correlates with post-2020 general inflation. My solution is to use the \textit{relative} tobacco price---the tobacco HICP divided by the all-items HICP---as the outcome. This differences out country-specific inflation shocks. I verify the approach with placebo tests: relative prices of alcohol and food show no systematic relationship with menthol share.

The main result is a precisely estimated zero. Across all specifications---continuous treatment, binary high/low menthol, triple-difference, alternative sample windows---the relative tobacco price in high-menthol countries was statistically indistinguishable from that in low-menthol countries after the ban. The event study shows clean pre-trends in the immediate pre-ban period and flat post-ban coefficients. The point estimates hover near zero and, if anything, drift slightly negative over the four years following the ban.

This null contributes to three literatures. First, it informs the economics of product bans, where the canonical debate is between prohibition reducing consumption \citep{miron2004drug} and substitution undoing the regulator's intent \citep{gruber2001tobacco}. The EU menthol ban appears to confirm the substitution hypothesis: banning a product variety reshuffled the market without contracting it. Second, the result speaks to tobacco economics, where the dominant finding is that price instruments (excise taxes) reduce smoking more effectively than quantity restrictions \citep{chaloupka2000economics, gruber2001tobacco}. My null is consistent with this hierarchy---the menthol ban did not generate the price changes that would signal reduced demand. Third, the paper contributes to the evaluation of EU-wide product regulation, exploiting the directive's uniform implementation to estimate treatment effects using cross-country variation in exposure intensity \citep{angrist2009mostly}.

The finding contrasts with the limited existing evidence. \citet{chung2020menthol} find that menthol smokers in the Netherlands were more likely to report quit intentions before the ban, but intention-to-treat estimates on actual behavior are scarce. Survey evidence from Poland \citep{zatonski2021menthol} suggests high rates of switching to unflavored cigarettes, consistent with my price-based null. The contribution here is to move from country-specific surveys to a multi-country quasi-experimental design using administrative price data.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background and the EU TPD menthol ban. Section 3 presents the data and measurement. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses implications and limitations. Section 7 concludes.


\section{Institutional Background}

\subsubsection*{The EU Tobacco Products Directive} The Tobacco Products Directive (2014/40/EU) was adopted in April 2014 as a comprehensive overhaul of EU tobacco regulation. Article 7 prohibited the placing on the market of tobacco products with characterizing flavors. While most flavored products (clove, vanilla, cherry) were banned from May 2016, menthol cigarettes received a four-year transition period until May 20, 2020. The extended timeline reflected the political difficulty of banning menthol: member states with large menthol markets---particularly Poland and the Baltic states---lobbied for a longer phase-out \citep{european2014directive}.

\subsubsection*{Pre-ban menthol markets} The menthol cigarette market varied enormously across the EU. In Poland, menthol cigarettes accounted for approximately 28\% of all cigarette sales, driven by brands like L\&M, Camel, and local manufacturers that had positioned menthol as a distinct product category rather than a niche variant \citep{zatonski2021menthol}. Finland (15\%), Lithuania (12\%), and Latvia (10\%) also had substantial menthol penetration, partly reflecting Soviet-era smoking patterns where menthol variants were widely distributed. By contrast, southern European markets (Spain, Italy, Greece) had menthol shares below 3\%, with consumers preferring unflavored or locally distinctive blends. Germany (6\%) and France (3\%) occupied intermediate positions, with menthol marketed primarily to younger and female smokers \citep{laverty2018menthol}.

\subsubsection*{Implementation and compliance} The ban was self-executing: no sell-through period was permitted after May 20, 2020. Retailers were required to remove menthol products from shelves immediately. Enforcement was handled by national tobacco market authorities, with compliance generally high across member states. The tobacco industry responded by introducing ``menthol-adjacent'' products---cigarettes with menthol-infused filter cards or capsule filters---which technically complied with the regulation by placing the flavor in the filter rather than the tobacco. Whether these substitutes satisfied menthol smokers' preferences is an open empirical question, but they remained niche products in most markets \citep{european2021implementation}.

\subsubsection*{Coincidence with COVID-19} The ban took effect two months into the most severe phase of the COVID-19 pandemic. By May 2020, all EU member states had imposed lockdowns of varying severity. This temporal coincidence creates a fundamental identification challenge: any post-ban change in tobacco markets could reflect pandemic-driven shifts in smoking behavior, supply chain disruptions, or economic contraction rather than the regulatory intervention. I address this directly in the empirical strategy.


\section{Data}

I assemble a country-month panel spanning January 2017 to December 2024 for 28 countries (EU-27 plus the United Kingdom, which implemented the identical ban while still in the Brexit transition period).

\subsubsection*{Tobacco prices} The primary outcome is the monthly Harmonised Index of Consumer Prices (HICP) for tobacco products, retrieved from Eurostat (dataset \texttt{prc\_hicp\_midx}, COICOP category CP022). The HICP measures the average change in prices paid by consumers, with 2015 as the base year. It captures both manufacturing prices and excise taxes, reflecting the total cost to consumers. I also retrieve the all-items HICP (CP00) and category-specific indices for alcohol (CP021), food (CP011), and clothing (CP031) to construct relative prices and placebo outcomes.

\subsubsection*{Menthol market shares} The treatment intensity variable is the pre-ban menthol market share---the fraction of total cigarette sales accounted for by menthol varieties in each country before May 2020. I construct this from two primary sources: \citet{laverty2018menthol}, who report menthol use prevalence from the International Tobacco Control (ITC) project surveys across eight European countries, and Euromonitor International (2019) industry data as cited in European Commission implementation reports and academic studies. These figures reflect approximately 2017--2019 market conditions and are treated as time-invariant pre-determined characteristics.

\subsubsection*{COVID stringency} To control for the pandemic, I use the Oxford COVID-19 Government Response Tracker (OxCGRT) Stringency Index, which aggregates nine indicators of lockdown severity into a score from 0 to 100. I compute the monthly country-level average.

\subsection{Summary Statistics}

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

The identifying variation comes from cross-country differences in pre-ban menthol market share. Countries with higher menthol dependence received a larger ``dose'' of the ban, while countries where menthol was negligible serve as comparisons. The primary specification is:
\begin{equation}
\ln\left(\frac{P^{\text{tobacco}}_{it}}{P^{\text{all}}_{it}}\right) = \alpha_i + \gamma_t + \beta \cdot (\text{MentholShare}_i \times \text{Post}_t) + \delta \cdot \text{Stringency}_{it} + \varepsilon_{it}
\label{eq:main}
\end{equation}
where $P^{\text{tobacco}}_{it}/P^{\text{all}}_{it}$ is the relative tobacco price for country $i$ in month $t$, $\alpha_i$ are country fixed effects absorbing time-invariant differences in tobacco taxation and consumption patterns, $\gamma_t$ are month fixed effects absorbing EU-wide price shocks, $\text{MentholShare}_i$ is the continuous pre-ban menthol market share, $\text{Post}_t$ equals one for months after May 2020, and $\text{Stringency}_{it}$ is the OxCGRT index. Standard errors are clustered at the country level.

\subsubsection*{Why relative prices} The tobacco HICP level is contaminated by country-specific inflation shocks correlated with menthol share. High-menthol countries---predominantly in Central and Eastern Europe---experienced sharply higher general inflation in 2021--2023 due to energy dependence on Russian gas. Using the ratio of tobacco to overall prices removes this confound, isolating tobacco-specific price movements from macroeconomic trends. I verify this approach by showing that the na\"ive level specification yields spurious positives that vanish once general inflation is accounted for.

\subsubsection*{Parallel trends} The identifying assumption is that, absent the ban, the relative tobacco price in high-menthol countries would have evolved in parallel with that in low-menthol countries, conditional on country and time fixed effects. I test this with an event study that interacts menthol share with relative-month indicators. The pre-ban coefficients should be near zero.

\subsubsection*{Alternative specifications} I also estimate a binary specification (above/below median menthol share), and a triple-difference that stacks tobacco with three placebo categories (alcohol, food, clothing) and identifies the tobacco-specific menthol interaction, absorbing any remaining country-time confounds through category-country fixed effects.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} reports the main estimates. Columns (1)--(3) use the continuous menthol share as treatment intensity. In the relative-price specification without controls (column 1), the interaction coefficient is $-0.050$ (SE $= 0.404$, $p = 0.90$): indistinguishable from zero. Adding COVID stringency (column 2) barely changes the estimate ($-0.064$, $p = 0.87$). For context, the coefficient implies that a country like Poland (menthol share of 0.28) experienced a $0.28 \times (-0.064) = -1.8$\% change in the relative tobacco price, an effect that is both economically trivial and statistically insignificant. Column (3) uses the level of the log tobacco index with the overall HICP as a control, yielding an identical null ($0.048$, $p = 0.91$).

Columns (4)--(5) use a binary high/low menthol indicator. The binary relative-price specification yields a positive but insignificant coefficient ($0.041$, $p = 0.29$), suggesting that if anything, high-menthol countries experienced a slight upward drift in relative tobacco prices---possibly reflecting a compositional effect as cheaper menthol varieties exited the index---but the effect is not distinguishable from noise.

The triple-difference in column (6) confirms the null. The tobacco-specific menthol $\times$ post interaction is $0.198$ with a standard error of $0.397$ ($p = 0.62$). The menthol ban did not generate a tobacco-specific price response even when benchmarked against non-tobacco categories within the same countries.

\input{tables/tab2_main}

\subsection{Event Study}

\Cref{tab:eventstudy} reports event-study coefficients from interacting menthol share with relative-month indicators. Two features are notable. First, the pre-ban coefficients for the six months immediately preceding the ban ($t = -6$ through $t = -1$) are small, statistically insignificant, and centered near zero, supporting the parallel trends assumption in the critical pre-treatment window. Coefficients at longer leads ($t = -10$ to $-12$) show some negative divergence, consistent with earlier tobacco tax changes in specific countries, but these do not propagate into the near-ban window.

Second, the post-ban coefficients are uniformly insignificant and, if anything, drift negative over time. By $t = +24$, the coefficient is $-0.197$ (SE $= 0.343$), suggesting that the relative tobacco price in high-menthol countries may have drifted slightly downward relative to low-menthol countries---the opposite of what a demand-disruption hypothesis would predict.

\input{tables/tab3_eventstudy}

\subsection{Robustness}

\subsubsection*{Placebo tests} \Cref{tab:placebo} examines whether the null reflects a clean design or a loss of power. Column (1) reproduces the main tobacco result. Columns (2)--(3) show that relative prices of alcohol and food are also unrelated to menthol share, supporting the interpretation that the identification strategy successfully isolates tobacco-specific effects. Column (4) reveals a significant negative coefficient for relative clothing prices ($-0.529$, $p < 0.001$), reflecting differential apparel price dynamics in Central European economies. This result underscores the importance of using the relative-price approach rather than price levels: menthol share correlates with country-level structural differences that affect non-tobacco prices, but these confounds are absorbed when studying tobacco relative to overall prices.

\input{tables/tab4_placebo}

\subsubsection*{Alternative specifications} \Cref{tab:robust} shows the null is stable across samples. Restricting the pre-period to January 2019 (16 months) yields $-0.107$ (SE $= 0.288$). Excluding the acute COVID period (March--September 2020) gives $-0.074$ (SE $= 0.420$). Leave-one-out analysis shows that dropping Poland---the highest-intensity country---shifts the estimate to $0.778$, reflecting Poland's outsized leverage; dropping any other country keeps the coefficient within the range $[-0.280, 0.778]$, all statistically insignificant. The instability upon dropping Poland is itself informative: it suggests that Poland alone may have experienced idiosyncratic tobacco market changes, but even this country-specific variation is absorbed by the standard errors.

\input{tables/tab5_robust}

\subsubsection*{Statistical power} The null finding is not a product of insufficient power for detecting large effects. The standard error on the preferred specification (0.399) implies a minimum detectable effect (at 80\% power, 5\% significance) of approximately $0.399 \times 2.8 = 1.12$ in the continuous specification. For Poland, this translates to $0.28 \times 1.12 = 31$\% change in relative prices---a very large threshold. The design is therefore well-powered to detect dramatic market disruption but not to detect small effects. I cannot rule out that the menthol ban caused a 5--10\% shift in tobacco prices in Poland-like countries.


\section{Discussion}

The central finding is that the world's largest flavor ban left no detectable trace on the relative price of tobacco. This is consistent with one dominant mechanism: \textit{product substitution}. Menthol smokers switched to unflavored cigarettes rather than quitting, leaving total tobacco demand---and hence the price index---undisturbed.

Three pieces of evidence support this interpretation. First, the null holds across all specifications, including the triple-difference that benchmarks tobacco against non-tobacco categories within the same country-month. Second, the event study shows no break at the ban date and a slight negative drift afterward---inconsistent with the demand-reduction hypothesis but consistent with gradual substitution as menthol smokers found unflavored alternatives. Third, survey evidence from Poland, the most affected country, reports that 63\% of menthol smokers switched to non-menthol cigarettes and only 12\% quit \citep{zatonski2021menthol}, aligning with the market-level null documented here.

The finding carries a broader lesson for regulatory design. Flavor bans restrict product variety but do not change the fundamental price of the addictive commodity. Economic theory predicts that within-category substitution will dominate when close substitutes exist and consumers are strongly addicted \citep{becker1988theory}. Menthol cigarettes and unflavored cigarettes are nearly perfect substitutes from a nicotine-delivery standpoint; the flavor is a product attribute, not a separate commodity. The ban removed an attribute, not the product.

This has direct implications for the US FDA's proposed menthol ban, which has been debated since 2013. Proponents argue that menthol's cooling sensation lowers initiation barriers and makes quitting harder \citep{villanti2017flavored}. If this is the case, the cessation pathway should dominate, and the market should contract. The EU experience suggests otherwise---or at minimum, that any cessation effect is too small to move wholesale market prices.

Several limitations deserve emphasis. First, and most critically, the HICP is a price index, not a quantity measure. The null price result is consistent with substitution but cannot distinguish between two scenarios: one in which menthol smokers switched to unflavored cigarettes at similar prices (true substitution), and one in which consumption fell but excise tax increases in high-menthol countries simultaneously raised prices, masking the demand decline. Several EU member states increased tobacco excise duties in 2020--2023 to meet the Directive's minimum floor requirements or fund pandemic recovery. If high-menthol countries raised taxes disproportionately, the flat price index could reflect opposing forces. Annual excise revenue data from DG TAXUD would help adjudicate but are not available at the monthly frequency needed for this design.

Second, the HICP is constructed as a Laspeyres-type index with periodically updated weights. If the ban caused compositional shifts in purchasing (e.g., from cheaper menthol brands to pricier non-menthol alternatives), the index may not immediately reflect these changes. Moreover, substitution to non-cigarette nicotine products---heated tobacco devices or e-cigarettes, which may fall outside the CP022 COICOP category---would represent a genuine health-relevant shift invisible to the tobacco price index.

Third, the cross-country variation in menthol share is correlated with geography (Central/Eastern vs.\ Southern Europe), limiting the extent to which treatment intensity can be separated from other structural differences. Fourth, the power of the continuous-treatment design is limited by 28 clusters. The MDE in the binary specification (SE $= 0.038$) is $0.038 \times 2.8 = 0.11$, or approximately 11\% in relative prices---well powered for dramatic effects but unable to detect the modest shifts a 12\% quit rate might generate \citep{zatonski2021menthol}. The result should therefore be interpreted as ruling out large market disruption, not as definitive evidence of zero effect.


\section{Conclusion}

The EU menthol ban eliminated a product that constituted up to one-quarter of the cigarette market in some countries. Four years later, the relative price of tobacco in those countries is indistinguishable from countries where menthol barely existed. The market absorbed the ban through substitution, not contraction.

This finding suggests a principle: banning a product attribute---a flavor, a form factor, a delivery mechanism---will reduce consumption only if consumers cannot substitute within the broader product category. For addictive goods with close within-category substitutes, attribute bans may be ineffective instruments. The policy implication is not that flavor bans are useless---they may reduce initiation among new smokers, a margin this paper cannot observe---but that they should be paired with price instruments if the goal is to reduce aggregate consumption among existing smokers.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsubsection*{Eurostat HICP} The Harmonised Index of Consumer Prices is Eurostat's principal measure of consumer price inflation, compiled using harmonized methodology across EU member states. I use dataset \texttt{prc\_hicp\_midx} with unit ``I15'' (index 2015=100). The tobacco sub-index (COICOP CP022) covers cigarettes, cigars, and other tobacco products. Country-specific weights reflect national consumption patterns.

\subsubsection*{Menthol market shares} Pre-ban menthol shares are compiled from: (i) \citet{laverty2018menthol}, who report ITC survey-based menthol use prevalence for eight EU countries; (ii) Euromonitor International (2019) industry data as cited in the European Commission's TPD implementation report and in \citet{zatonski2021menthol}; and (iii) country-specific sources (ASH UK factsheet 2019 for the United Kingdom). All figures are cross-validated where multiple sources are available. The menthol share is treated as a pre-determined characteristic and does not vary over time.

\subsubsection*{Sample restrictions} The sample spans January 2017 to December 2024, providing 40 months of pre-treatment data and 55 months of post-treatment data. May 2020 (the transition month) is excluded. Countries with missing overall HICP data are dropped, resulting in a balanced panel of 2,611 country-month observations across 28 countries.

\subsubsection*{OxCGRT} The Oxford COVID-19 Government Response Tracker Stringency Index aggregates nine sub-indicators of government responses (school closures, workplace closures, travel restrictions, etc.) into a composite score from 0 to 100. Daily values are averaged to monthly.


\section{Identification Appendix}
\label{app:identification}

\subsubsection*{Why level specifications fail} The na\"ive specification using the log tobacco HICP level as the outcome yields a positive but insignificant coefficient of 0.41. However, using the same specification with alcohol, food, and clothing as outcomes yields similar-magnitude positive coefficients (0.34 and 0.38 for alcohol and food, both significant). This demonstrates that menthol market share proxies for Central/Eastern European geography, where post-2020 inflation was systematically higher. The relative-price approach eliminates this confound.

\subsubsection*{Pre-trend assessment} In the event study, coefficients at $t = -10$ and $t = -11$ show modest negative values ($-0.054$ and $-0.059$) that are statistically insignificant in the relative-price specification. The immediate pre-ban window ($t = -4$ through $t = -1$) shows coefficients within 0.05 of zero, none significant. This supports the identifying assumption during the period most relevant for causal inference.


\section{Robustness Appendix}

\subsubsection*{Leave-one-out sensitivity} Dropping Poland (menthol share = 28\%) produces the largest swing, moving the estimate from $-0.064$ to $+0.778$. Poland's outsized leverage reflects its extreme treatment intensity: it has more than twice the menthol share of the next-highest country (Finland, 15\%). Even the Poland-sensitive estimate, however, remains statistically insignificant ($p > 0.10$).

\subsubsection*{Wild cluster bootstrap} The 28-country clustered standard errors are conservative but may be imprecise with fewer than 30 clusters. I attempt wild cluster bootstrap inference using the \texttt{fwildclusterboot} R package but encounter numerical difficulties with the variance-covariance matrix, likely due to the time-series dimension exceeding the cross-section. The clustered standard errors therefore represent the primary inference approach.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
