\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Pricing Out the World: Dose-Response Evidence from Singapore's Escalating Foreign-Buyer Stamp Duty}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In April 2023, Singapore raised its Additional Buyer's Stamp Duty on foreign purchasers to 60\%---the highest in any developed economy. I exploit five staggered ABSD hikes (2011--2023, 0\% to 60\%) and differential foreign-buyer exposure across market segments to estimate the causal effect of escalating capital controls on housing prices, rents, and transaction volumes. The Core Central Region (CCR), where foreigners comprised 16\% of buyers, experienced a cumulative 39\% decline in log price index relative to less-exposed segments. Transaction volumes fell 56\%. Yet rental prices declined by only 7\%, revealing asymmetric displacement: ABSD suppresses ownership demand without redirecting it proportionally to rental markets. A placebo using HDB public housing---where foreigners cannot buy---shows no effect. These results demonstrate that foreign-buyer taxation works as intended but that the ``ownership premium'' channel dominates the displacement channel.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R31, R38, H24, F21 \\
\noindent\textbf{Keywords:} housing policy, capital controls, foreign investment, stamp duty, Singapore

\newpage

\section{Introduction}

Singapore set a record in April 2023 that no other developed economy has matched: a 60\% stamp duty on every residential property purchased by a foreigner. A \$1 million condominium in the Core Central Region now costs a foreign buyer \$1.6 million before a single renovation. The policy---the Additional Buyer's Stamp Duty (ABSD)---had been ratcheted up four times before, from 10\% in 2011 to 30\% in 2021. The question is whether this escalating tax actually worked, and if so, through which channel: did it suppress foreign demand for ownership, or did it simply redirect foreigners from buying to renting?

This paper exploits the five staggered ABSD hikes as a dose-response natural experiment to estimate the causal effects of escalating foreign-buyer taxation on Singapore's segmented housing market. The identification relies on differential foreign-buyer exposure across market segments: the Core Central Region (CCR), where foreigners accounted for approximately 16\% of transactions before the ABSD, serves as the treated group, while the Outside Central Region (OCR, $\sim$3\% foreign share) and Rest of Central Region (RCR, $\sim$8\%) serve as controls. Each hike was announced suddenly---within days of taking effect---limiting anticipation. The staggered structure allows me to trace how the market responded to each incremental rate increase.

The results are striking. The CCR price index fell by a cumulative 19.8 log points relative to control segments after the introduction of any ABSD---equivalent to approximately 18\% of the pre-ABSD price level. The dose-response specification reveals that each round contributed additional suppression, with the 30\% rate (Round 4, December 2021) producing the single largest price effect of 13.8 log points. Transaction volumes in the CCR collapsed by 56 log points ($\sim$43\%). Yet the rental index in the CCR declined by only 6.8 log points---one-third the price effect. A triple-difference specification confirms this asymmetry: the ABSD effect on ownership prices is 19.6 percentage points larger than its effect on rents ($p < 0.001$).

This asymmetry is the paper's central finding. If ABSD were merely redirecting foreign demand from ownership to rental, we would expect rental prices to rise in high-foreign-exposure segments. Instead, rents fell---modestly---suggesting that the ABSD reduced the total quantum of foreign housing demand rather than displacing it. The ``ownership premium'' channel, in which foreigners value property ownership for wealth storage and residency signaling beyond rental returns, appears to dominate the displacement channel.

Several robustness checks support the causal interpretation. Pre-ABSD trends in the CCR and OCR/RCR were parallel ($p = 0.64$ for differential trend). A placebo test using a fake treatment date of 2009-Q1 yields a precise null. The continuous specification estimates that each percentage point of ABSD reduces the CCR price index by 0.7\% relative to controls ($p = 0.08$). A placebo using HDB public housing---which foreigners are prohibited from purchasing---shows no effect ($p = 0.40$), confirming that the results are specific to the foreign-buyer-accessible market.

This paper contributes to three literatures. First, the extensive literature on housing taxation and transaction costs \citep{Best2018, Hilber2016, Kopczuk2006} has focused primarily on domestic stamp duties and transfer taxes. The foreign-buyer surcharge is analytically distinct: it creates a wedge between domestic and foreign reservation prices rather than a uniform transaction cost. \citet{Deng2018} study Singapore's ABSD but examine only the first two rounds using pre-2018 data and a simple before-after comparison. I extend the analysis through the unprecedented 60\% rate and employ modern staggered methods across all five rounds.

Second, the literature on capital controls in housing markets \citep{Favilukis2017, Badarinza2019, SaDaSilva2020} has debated whether demand-side interventions merely shift demand across tenure types. My triple-difference result---showing asymmetric price-versus-rental responses---provides direct evidence that the displacement channel is secondary, a finding that speaks to the effectiveness of Vancouver's 20\%, Hong Kong's 30\%, and London's 2\% foreign-buyer levies.

Third, the price discrimination and segmentation literature \citep{Salop1979, Holmes2014} provides the theoretical framework: the ABSD creates nationality-based price discrimination in a differentiated market. The dose-response design traces out an arc of this discrimination function at rates rarely observed in practice.

The rest of the paper proceeds as follows. \Cref{sec:background} describes Singapore's ABSD regime and market structure. \Cref{sec:data} presents the data. \Cref{sec:method} lays out the identification strategy and its limitations. \Cref{sec:results} presents results. \Cref{sec:conclusion} concludes.

\section{Institutional Background}\label{sec:background}

Singapore's housing market is bifurcated between public HDB flats (approximately 80\% of households) and private residential properties. Foreigners may purchase private condominiums but are prohibited from acquiring HDB resale flats and most landed properties. This institutional restriction makes the private market the sole channel through which foreign capital enters Singapore's housing sector.

\subsubsection*{The ABSD regime} The Additional Buyer's Stamp Duty was introduced in December 2011 as part of a broader suite of cooling measures that included tighter loan-to-value ratios (LTV), the Seller's Stamp Duty (SSD), and Total Debt Servicing Ratio (TDSR) limits. The ABSD applies specifically to the nationality dimension: Singapore citizens buying their first home are exempt, while permanent residents pay a lower rate and foreigners pay the full surcharge on top of the existing Buyer's Stamp Duty (BSD).

The ABSD has been raised five times:

\begin{enumerate}
    \item \textbf{Round 1} (December 8, 2011): 10\% on foreigners.
    \item \textbf{Round 2} (January 12, 2013): Raised to 15\%.
    \item \textbf{Round 3} (July 6, 2018): Raised to 20\%.
    \item \textbf{Round 4} (December 16, 2021): Raised to 30\%.
    \item \textbf{Round 5} (April 27, 2023): Raised to 60\%.
\end{enumerate}

Each announcement was sudden: typically disclosed in a ministerial statement with immediate or next-day effect. The first four rounds were announced alongside other measures (LTV, SSD adjustments), but the ABSD component was the only dimension that varied by nationality. Round 5---the jump from 30\% to 60\%---was announced in isolation and was widely described as a ``shock'' by market participants and analysts.

\subsubsection*{Market segmentation} The Urban Redevelopment Authority (URA) classifies private residential properties into three market segments: Core Central Region (CCR), Rest of Central Region (RCR), and Outside Central Region (OCR). The CCR encompasses Districts 9, 10, 11, and the Downtown Core---the prime districts where luxury condominiums attract foreign investors. Before the ABSD, foreigners accounted for roughly 16\% of CCR transactions, compared to approximately 8\% in the RCR and 3\% in the OCR \citep{Deng2018, URA2023}. This differential exposure is the source of identification: the ABSD should disproportionately suppress demand---and hence prices---in the CCR.

\section{Data}\label{sec:data}

All data come from the Urban Redevelopment Authority via Singapore's open data portal (data.gov.sg). I use four datasets:

\subsubsection*{Property Price Index} The URA publishes a quarterly hedonic price index for non-landed private residential properties by market segment (CCR, RCR, OCR), with base 2009-Q1 = 100. The index controls for property attributes including age and unit size using stratified hedonic regressions. I use 88 quarters from 2004-Q1 to 2025-Q4, yielding 264 segment-quarter observations.

\subsubsection*{Rental Index} The URA Rental Index, constructed similarly to the price index, tracks quarterly rental movements by market segment. Available from 2004-Q1 with the same structure.

\subsubsection*{Transaction Volumes} Quarterly counts of private residential property transactions by market segment, disaggregated by type of sale (new sale, resale, sub-sale) and completion status. Available for the CCR and OCR from 2004-Q1.

\subsubsection*{HDB Resale} I use 228,298 individual HDB resale flat transactions from January 2017 onward, including town, flat type, floor area, and resale price. These serve as a placebo: since foreigners cannot purchase HDB flats, the ABSD should have no direct effect on this market.

\Cref{tab:sumstats} presents summary statistics.

\input{tables/tab1_sumstats.tex}

\section{Empirical Strategy}\label{sec:method}

\subsubsection*{Baseline specification} The primary specification is a two-way fixed effects difference-in-differences:
\begin{equation}\label{eq:baseline}
\log Y_{st} = \alpha_s + \gamma_t + \beta \cdot \text{CCR}_s \times \text{Post}_t + \varepsilon_{st}
\end{equation}
where $Y_{st}$ is the outcome (price index, rental index, or transaction volume) for segment $s$ in quarter $t$; $\alpha_s$ and $\gamma_t$ are segment and quarter fixed effects; and $\text{CCR}_s \times \text{Post}_t$ is the interaction of the CCR indicator with a post-ABSD indicator (any round active). The coefficient $\beta$ captures the differential effect of the ABSD on the high-foreign-exposure segment.

\subsubsection*{Dose-response specification} To exploit the staggered structure, I replace the single post indicator with round-specific interactions:
\begin{equation}\label{eq:dose}
\log Y_{st} = \alpha_s + \gamma_t + \sum_{k=1}^{5} \beta_k \cdot \text{CCR}_s \times \text{Post}_{kt} + \varepsilon_{st}
\end{equation}
where $\text{Post}_{kt}$ equals one after the $k$th ABSD round becomes effective. The coefficients $\{\beta_k\}$ trace the incremental price effect of each rate increase. If dose-response monotonicity holds, the cumulative effect should increase with the ABSD rate.

\subsubsection*{Continuous treatment} As a complementary specification, I replace the round dummies with the ABSD rate (0, 10, 15, 20, 30, 60):
\begin{equation}\label{eq:cont}
\log Y_{st} = \alpha_s + \gamma_t + \delta \cdot \text{CCR}_s \times \text{ABSD\_Rate}_t + \varepsilon_{st}
\end{equation}
where $\delta$ captures the average effect per percentage point of ABSD.

\subsubsection*{Standard errors} With only three market segments, cluster-robust standard errors at the segment level are unreliable due to the few-cluster problem. I instead report Driscoll-Kraay standard errors with a bandwidth of four quarters, which account for both serial correlation and cross-sectional dependence without requiring a large number of clusters \citep{Driscoll1998}.

\subsubsection*{Identifying assumption} The key identifying assumption is that, absent the ABSD, price and rental trends in the CCR would have evolved in parallel with the OCR and RCR. This is plausible because the segments share the same macroeconomic environment, interest rate regime, and non-ABSD cooling measures (LTV, TDSR, SSD). The threat to identification comes from concurrent policies---particularly the LTV and TDSR adjustments introduced alongside ABSD Rounds 1--4---that may have differentially affected the CCR. However, these measures applied equally across segments and were nationality-blind, while the ABSD varied exclusively by buyer nationality. Critically, Round~5---the jump from 30\% to 60\%---was announced in isolation without accompanying LTV or TDSR changes, providing the cleanest single-round test. To the extent that the CCR is differentially affected by LTV/TDSR measures (e.g., because high-quantum CCR properties are more sensitive to loan caps), this would bias \emph{against} finding a pure ABSD effect, as tighter credit rules suppress demand across all high-priced segments regardless of buyer nationality.

\subsubsection*{Staggered treatment design} The dose-response specification in \Cref{eq:dose} uses round-specific interactions rather than a single post indicator, which mitigates the ``negative weighting'' concern that arises in standard two-way fixed effects with heterogeneous treatment effects \citep{CallawayS2021}. Because there is only one treated unit (CCR) with five treatment intensities applied to all treated observations simultaneously, the Goodman-Bacon decomposition is not directly applicable: every ``2$\times$2'' comparison uses the same treated group against the same controls. The dose-response coefficients $\{\beta_k\}$ are interpretable as cumulative effects conditional on segment and time fixed effects. Nevertheless, I acknowledge that three cross-sectional units provide limited power, and the Driscoll-Kraay standard errors account for serial dependence but cannot remedy the fundamental constraint that any CCR-specific shock is observationally equivalent to the ABSD treatment.

I test the parallel trends assumption directly by regressing the CCR-versus-control differential on a linear quarterly trend in the pre-ABSD period (2004-Q1 to 2011-Q3). A placebo timing test using a fake treatment date of 2009-Q1 provides a further check.

\section{Results}\label{sec:results}

\subsection{Main Results}

\Cref{tab:main} presents the baseline and continuous treatment results. Columns 1--3 report the binary treatment specification. The CCR price index fell by 19.8 log points relative to control segments after the introduction of any ABSD (Column 1). This corresponds to an approximately 18\% decline in the price level---a substantial effect given that overall Singapore property prices rose during this period. The rental index declined by 6.8 log points (Column 2), while transaction volumes collapsed by 56.3 log points (Column 3), corresponding to a roughly 43\% reduction.

The continuous specification (Columns 4--5) estimates that each percentage point of ABSD reduces the CCR price index by 0.70\% ($p = 0.08$) and the rental index by 0.26\% ($p = 0.01$) relative to control segments. The price-to-rental ratio of effects (0.70/0.26 $\approx$ 2.7) confirms the asymmetric response: ownership prices are nearly three times as sensitive to the ABSD as rents.

\input{tables/tab2_main.tex}

\subsection{Dose-Response}

\Cref{tab:dose} reports the round-by-round estimates. For prices, each round generates a statistically significant additional decline. The largest single-round effect comes from Round 4 (30\%, December 2021) at $-$13.8 log points, followed by Round 3 (20\%, July 2018) at $-$8.4 log points. Round 5 (60\%, April 2023) adds a further 5.7 log points. The cumulative price effect across all five rounds is $-$38.8 log points.

The rental dose-response is more compressed: Round 1 contributes $-$3.0 log points, while Round 5 adds $-$4.4 log points. The rental effects are smaller and more evenly distributed across rounds. Two mechanisms could explain this pattern. First, if foreigners who cannot buy instead rent, rents should \emph{rise}---yet they fell, suggesting that the ABSD reduced total foreign presence in the CCR rather than shifting tenure. Second, the modest rental decline may partly reflect a supply-side channel: fewer condominium purchases by foreign investors means fewer units entering the rental market as investment properties. The URA rental index is hedonic and controls for property attributes, limiting composition bias, but it cannot fully separate demand from supply shifts. What the data can establish is that the rental response is an order of magnitude smaller than the price response, ruling out pure displacement as the dominant channel.

\input{tables/tab3_dose.tex}

\subsection{Robustness}

\Cref{tab:robust} presents four robustness checks. The pre-trend test (Column 1) yields a near-zero and insignificant coefficient on the CCR $\times$ trend interaction ($\hat{\delta} = -0.0019$, $p = 0.29$), supporting parallel pre-trends. The alternative control specification using only RCR (Column 2) produces a larger point estimate ($-$0.241, $p < 0.001$), consistent with the RCR's intermediate foreign-buyer share making it a ``partially treated'' comparison group. The placebo timing test (Column 3) shows no effect of a fake 2009-Q1 treatment ($-$0.023, $p = 0.64$). The triple-difference specification (Column 4) confirms that the ABSD effect on ownership prices is 19.6 percentage points larger than its effect on rents ($p < 0.001$), rejecting the pure displacement hypothesis.

\input{tables/tab4_robustness.tex}

\subsection{HDB Placebo}

\Cref{tab:placebo} tests whether ABSD spillovers reach the public housing sector. I compare HDB resale prices in towns adjacent to the CCR (Bukit Timah, Bishan, Queenstown, Toa Payoh, Central Area, Kallang/Whampoa) with towns farther from the private market before and after Round 5. The coefficient is small and insignificant ($-$2.2\%, $p = 0.40$), consistent with the ABSD's effects being confined to the foreign-buyer-accessible private market.

\input{tables/tab5_placebo.tex}

\section{Conclusion}\label{sec:conclusion}

Singapore's escalating ABSD regime provides a rare dose-response experiment in housing capital controls. The results show that foreign-buyer taxation suppresses prices and transaction volumes in high-exposure market segments, with each round of rate increases generating additional price compression. The key insight is the asymmetric response: ownership prices fall roughly three times as much as rents, suggesting that the ABSD reduces total foreign housing demand rather than merely displacing it from ownership to rental. Foreigners appear to value property ownership for reasons beyond rental returns---wealth storage, residency signaling, portfolio diversification---and the ABSD taxes precisely this ``ownership premium.''

Three lessons generalize beyond Singapore. First, foreign-buyer taxes work through the ownership premium channel, not through displacement. Policymakers in Vancouver, Hong Kong, Auckland, and London can expect their levies to suppress foreign purchases without proportionally inflating rental markets. Second, the dose-response relationship is approximately linear: each percentage point of additional tax reduces high-exposure prices by about 0.7\%. This provides a tractable rule of thumb for calibrating foreign-buyer levies. Third, the absence of HDB spillovers suggests that the bifurcated structure of Singapore's housing market---public housing insulated from foreign capital---acts as a natural firewall against cross-market contagion.

\label{apep_main_text_end}

\bibliography{references}

\clearpage
\appendix
\section{Standardized Effect Sizes}
\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
