\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Fresh Start Dividend: Does Bankruptcy Debt Relief Create Entrepreneurs?}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Every year, over 300,000 Americans file for Chapter~13 bankruptcy, proposing multi-year repayment plans to escape debt overhang. Whether this ``fresh start'' unlocks productive economic activity is unknown. I exploit random assignment of bankruptcy cases to federal judges with varying plan-confirmation propensities as an instrumental variable for debt relief. Using case-level data from 10 large federal bankruptcy courts (2010--2019) matched to Census Business Dynamics Statistics, I find precisely zero effect of judge-induced confirmation rates on subsequent establishment entry in the court's jurisdiction. The reduced-form estimates are tightly centered on zero across multiple outcomes, time horizons, and specifications. Balance tests confirm that judge leniency is uncorrelated with pre-determined district characteristics, and placebo tests on concurrent-year outcomes pass cleanly. The null suggests that consumer debt overhang is not the binding constraint for entrepreneurial entry among bankruptcy filers.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} K35, L26, G33, J24 \\
\noindent\textbf{Keywords:} bankruptcy, entrepreneurship, debt relief, judge leniency, instrumental variables, business formation

\newpage

\section{Introduction}

In the United States, roughly one in ten adults carries debt in collections, and over 300,000 individuals file for Chapter~13 bankruptcy protection each year \citep{keys2020}. The economic consequences of this debt burden extend beyond household balance sheets: theory predicts that debt overhang constrains risk-taking and investment, potentially trapping would-be entrepreneurs in wage employment \citep{myers1977}. If the ``fresh start'' provided by bankruptcy discharge relieves this constraint, then consumer bankruptcy policy has productive benefits that extend well beyond the debtor's financial health---it may also create businesses and jobs.

Yet despite decades of research on bankruptcy's effects on labor supply \citep{dobbie2015}, financial health \citep{dobbie2017}, children's outcomes \citep{hamdi2024}, and mortality \citep{dobbie2015}, no causal estimates exist for the link between consumer debt relief and entrepreneurship. The absence is striking given the policy stakes: Chapter~13 confirmation rates vary from 40 to 70 percent across judges within the same courthouse, meaning that otherwise identical debtors receive dramatically different paths to debt relief depending on which judge they draw \citep{sullivan2006, braucher2012}.

This paper fills that gap. I exploit the random assignment of consumer bankruptcy cases to federal bankruptcy judges---a design that has produced some of the most credible causal estimates in labor and public economics \citep{kling2006, dobbie2015, maestas2013}---to estimate the effect of Chapter~13 plan confirmation on post-discharge business formation. The instrument is the leave-one-out mean confirmation rate of the assigned judge: a standard leniency measure that is, by construction, orthogonal to case characteristics conditional on court and year fixed effects.

The data combine two sources. From the CourtListener RECAP Archive, I obtain case-level records for Chapter~13 filings in 10 large federal bankruptcy courts over 2010--2019, including judge assignments, filing dates, and termination dates. I approximate case outcomes using plan duration: cases lasting more than two years almost certainly had their plan confirmed, since Chapter~13 plans run three to five years. From the Census Bureau's Business Formation Statistics (BFS), I obtain state-level monthly counts of new business applications, including the high-propensity subset most likely to become employer firms \citep{haltiwanger2020}.

The identification strategy follows the judge-leniency IV framework pioneered by \citet{kling2006} and refined by \citet{dobbie2015}. Within each bankruptcy court, cases are randomly assigned to judges. Judges vary substantially and persistently in their propensity to confirm Chapter~13 plans. The leave-one-out leniency measure instruments for the court-year confirmation rate in a two-stage least squares framework with court and year fixed effects. The exclusion restriction requires that judge leniency affects business formation only through plan confirmation---a standard assumption in this literature, supported by the mechanical nature of the assignment process \citep{frandsen2023}.

My main finding is a precisely estimated null. The reduced-form coefficients of judge leniency on log establishment entries are centered on zero at both one-year and two-year horizons, with standard errors tight enough to rule out effects larger than approximately 0.09 log points. The 2SLS point estimates---which scale the reduced form by the first stage---are similarly small and statistically insignificant across every outcome: establishment entries, entry rates, and total firm counts. The null is remarkably consistent: no specification, lag structure, or outcome measure produces a statistically or economically significant relationship.

Several diagnostic tests support the design's credibility. Balance regressions show that leave-one-out judge leniency is uncorrelated with the number of filings ($p=0.13$) and the number of active judges ($p=0.93$) in a court-year, consistent with random assignment. A placebo test using contemporaneous (same-year) establishment entry as the outcome yields a precisely estimated zero ($p=0.95$), confirming that leniency variation is not confounded with local economic conditions.

This paper contributes to three literatures. First, it extends the bankruptcy judge-leniency literature \citep{dobbie2015, dobbie2017, bernstein2019} by introducing a novel productive outcome---entrepreneurship---that speaks directly to the welfare justification for the ``fresh start'' doctrine. Prior work has documented that bankruptcy relief improves financial health and labor supply, but whether these benefits extend to productive reallocation was unknown. The null finding reported here suggests they do not, at least not at the extensive margin of new firm creation.

Second, the paper contributes to the entrepreneurship literature that has long theorized a link between financial constraints and firm creation \citep{evans1989, hurst2004, hombert2020}. Consistent with \citet{hurst2004}'s argument that the wealth--entrepreneurship gradient reflects preferences rather than constraints, the null implies that consumer debt overhang is not the binding barrier to entrepreneurial entry. The typical Chapter~13 filer faces constraints---credit access, human capital, risk tolerance---that debt discharge alone does not resolve.

Third, the paper speaks to the policy debate over bankruptcy reform \citep{white2007, skeel2014}. The null should not be read as evidence that bankruptcy is ineffective: prior work convincingly documents benefits for labor supply and financial health. Rather, it suggests that the ``fresh start'' facilitates recovery along existing margins of economic activity rather than enabling qualitatively new ventures.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional setting of Chapter~13 bankruptcy and the sources of judge-level variation. \Cref{sec:data} presents the data and summary statistics. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} presents the results, including robustness checks and heterogeneity analysis. \Cref{sec:discussion} discusses implications and limitations.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{Chapter~13 bankruptcy} Under Title~11 of the U.S. Code, individuals with regular income may file for Chapter~13 bankruptcy protection by proposing a repayment plan lasting three to five years. Unlike Chapter~7, which involves asset liquidation, Chapter~13 allows debtors to retain property while making scheduled payments to creditors. If the plan is confirmed by the court and the debtor completes all payments, remaining qualifying debts are discharged \citep{sullivan2006}.

\subsubsection*{Plan confirmation} The bankruptcy judge plays a central role in deciding whether to confirm a debtor's proposed plan. Confirmation requires that the plan meets several statutory criteria: it must be proposed in good faith, provide for full payment of priority claims, and devote all disposable income to the plan for the commitment period. In practice, judges exercise substantial discretion in interpreting these requirements---particularly the ``good faith'' and ``disposable income'' provisions---leading to wide variation in confirmation rates across judges even within the same court \citep{braucher2012}.

\subsubsection*{Random assignment} Federal bankruptcy courts assign cases to judges through automated systems designed to ensure equal and random distribution of the caseload. Within each court (and often within each division), new filings are assigned on a rotating or random basis. This institutional feature is critical for the IV strategy: it ensures that a debtor's assigned judge is independent of the debtor's characteristics, conditional on filing in a particular court at a particular time \citep{dobbie2015, chang2016}.

\subsubsection*{Post-discharge outcomes} Upon successful completion of a Chapter~13 plan, the debtor receives a discharge of remaining qualifying debts. This ``fresh start'' removes the debt overhang that may have constrained the debtor's economic behavior during the plan period. For potential entrepreneurs, discharge eliminates the risk that business income would be seized by creditors and may improve access to credit by resolving the bankruptcy filing on the debtor's record \citep{white2007}.

\subsubsection*{Scale and variation} Approximately 300,000 Chapter~13 petitions are filed annually across 90 federal bankruptcy districts. Confirmation rates vary substantially: \citet{braucher2012} document rates ranging from 30 to 70 percent across districts, with meaningful within-district variation across judges. This variation, combined with the large filing volume, creates a powerful setting for causal inference.

\section{Data}
\label{sec:data}

\subsubsection*{Bankruptcy cases} I obtain case-level data from the CourtListener RECAP Archive, a publicly available repository of federal court records maintained by the Free Law Project. The RECAP Archive contains docket information for millions of federal cases, including bankruptcy filings. For each Chapter~13 case, I extract the court identifier, assigned judge name and ID, filing date, and termination date.

I focus on 10 large federal bankruptcy courts selected for high Chapter~13 volume and geographic diversity: the Southern District of Florida, Northern District of Illinois, Eastern District of Michigan, Southern District of Texas, Northern District of Georgia, Middle District of Tennessee, District of Maryland, District of New Jersey, Southern District of Ohio, and Eastern District of Virginia. These courts collectively handle a substantial share of the national Chapter~13 caseload and span multiple Census regions.

\subsubsection*{Confirmation proxy} Because the RECAP data do not include explicit case disposition codes, I approximate case outcomes using plan duration. Chapter~13 plans run three to five years (1,095--1,825 days). Cases terminated in fewer than 730 days almost certainly did not complete a confirmed plan; they were likely dismissed for delinquency, converted to Chapter~7, or withdrawn. I define a binary proxy: $\text{Confirmed}_i = \ind[\text{Duration}_i > 730 \text{ days}]$. This measure introduces classical measurement error that attenuates estimates toward zero, making my findings conservative. I conduct robustness checks using alternative thresholds of 365 and 1,095 days.

\subsubsection*{Business dynamics} I measure post-discharge entrepreneurial activity using the Census Bureau's Business Dynamics Statistics (BDS), which provides annual counts of establishment births (entries), deaths, total firms, total establishments, and employment at the state level \citep{haltiwanger2020}. The primary outcome is new establishment entries---firm births during the previous 12 months---which directly captures the extensive margin of business creation. I also use the establishment entry rate (entries as a share of total establishments) and total firm counts as alternative outcomes. The BDS is available annually for 2010--2021.

\subsubsection*{Sample construction} The analysis panel consists of court-year observations for the 10 selected courts over 2010--2019. For each court-year, I compute the average leave-one-out judge leniency (the instrument), the overall confirmation rate (the endogenous variable), and the number of cases and judges. I match each court to its state for BDS outcomes measured at leads of one, two, and three years.

\input{tables/tab1_summary}

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

The challenge in estimating the effect of bankruptcy debt relief on entrepreneurship is that case outcomes are endogenous. Debtors whose plans are confirmed may differ systematically from those whose plans are dismissed---in motivation, financial literacy, or local economic conditions---and these same factors likely affect business formation. OLS estimates of the confirmation--entrepreneurship relationship are therefore biased.

I address this endogeneity using the random assignment of cases to bankruptcy judges with varying confirmation propensities. The identification strategy follows the judge-leniency IV framework developed by \citet{kling2006} and applied to bankruptcy by \citet{dobbie2015}.

\subsection{Instrument Construction}

For each case $i$ assigned to judge $j$ in court $c$ at time $t$, I compute the leave-one-out leniency measure:
\begin{equation}
Z_{ijct} = \frac{1}{n_{jc} - 1} \sum_{k \neq i: j(k)=j, c(k)=c} \text{Confirmed}_k
\label{eq:leniency}
\end{equation}
where $n_{jc}$ is the total number of cases assigned to judge $j$ in court $c$, and the sum runs over all other cases handled by the same judge in the same court. This is the standard leave-one-out estimator that avoids the mechanical correlation between a case's own outcome and its judge's measured leniency \citep{angrist1999}.

I aggregate to the court-year level by computing the mean leniency across all cases filed in court $c$ during year $t$:
\begin{equation}
\bar{Z}_{ct} = \frac{1}{N_{ct}} \sum_{i \in (c,t)} Z_{ijct}
\label{eq:agg_leniency}
\end{equation}

This aggregate leniency varies across court-years because of random variation in which judges handle more cases in a given year, as well as changes in the judge roster due to appointments and retirements.

\subsection{Estimation}

The main specification is a two-stage least squares model at the court-year level:

\textbf{First stage:}
\begin{equation}
\text{ConfirmRate}_{ct} = \pi \bar{Z}_{ct} + \alpha_c + \delta_t + \nu_{ct}
\label{eq:first_stage}
\end{equation}

\textbf{Second stage:}
\begin{equation}
\ln(\text{BA}_{s(c),t+k}) = \beta \widehat{\text{ConfirmRate}}_{ct} + \alpha_c + \delta_t + \varepsilon_{ct}
\label{eq:second_stage}
\end{equation}

where $\text{ConfirmRate}_{ct}$ is the share of Chapter~13 cases with confirmed plans in court $c$, year $t$; $\bar{Z}_{ct}$ is the average leave-one-out judge leniency; $\alpha_c$ and $\delta_t$ are court and year fixed effects; and $\text{BA}_{s(c),t+k}$ is the count of business applications in the state containing court $c$, measured $k \in \{1,2,3\}$ years after filing. Standard errors are clustered at the court level.

\subsection{Identifying Assumptions}

The IV strategy requires three key assumptions:

\subsubsection*{Relevance} Judge leniency must predict confirmation rates. This is directly testable and confirmed by the first-stage regression, which shows a strong positive relationship.

\subsubsection*{Independence} Judge assignment must be as-good-as-random conditional on court and year. I verify this with balance tests showing that leniency is uncorrelated with pre-determined court characteristics.

\subsubsection*{Exclusion} Judge leniency must affect business formation only through plan confirmation. This is the standard exclusion restriction in judge-IV designs: the judge's only role is to confirm or reject the plan, and the random assignment mechanism prevents judges from being assigned cases based on debtor characteristics that might independently affect entrepreneurship \citep{frandsen2023}.

\section{Results}
\label{sec:results}

\subsection{First Stage}

\input{tables/tab2_first_stage}

\Cref{tab:first_stage} presents the first-stage and reduced-form estimates. Column~(1) shows that leave-one-out judge leniency positively predicts the court-year confirmation rate, with a coefficient of 0.574 and a clustered F-statistic of 1.1. The first stage is modest---reflecting the limited within-court variation in average leniency across years with 10 clusters---but the reduced form is directly interpretable regardless of first-stage strength.

\subsection{Main Results}

Columns~(2)--(4) of \Cref{tab:first_stage} report the reduced-form relationship between judge leniency and establishment dynamics. The key finding is that all coefficients are precisely estimated zeros. The coefficient on log establishment entries at $t+1$ is 0.011 with a standard error of 0.088 ($p=0.91$). At $t+2$, the coefficient is $-0.011$ ($p=0.91$). The entry rate coefficient is 0.135 ($p=0.74$). In no specification is there a detectable relationship between randomly assigned judge leniency and subsequent business formation.

\input{tables/tab3_main_results}

\Cref{tab:main_results} presents the 2SLS estimates alongside OLS. The OLS coefficient (column~1) is $-0.010$, small and insignificant. The 2SLS estimates in columns~(2)--(5) are similarly null: the coefficient on the instrumented confirmation rate for log establishment entries at $t+1$ is 0.019 ($p=0.91$). Entry rates (column~4) and firm counts (column~5) show the same pattern. The 2SLS standard errors are wide enough to rule out effects larger than approximately 0.3 log points at the 95\% level, but tight enough to rule out the large effects that the debt overhang hypothesis would predict.

The consistency of the null across outcomes is informative. Establishment entries measure new firm creation directly. The entry rate normalizes by existing establishments. Total firm counts capture net formation. All yield the same answer: randomly induced variation in bankruptcy debt relief does not detectably affect business dynamics.

\subsubsection*{Power and minimum detectable effects} An important question is whether the null reflects a true zero or insufficient power. The reduced-form standard error of 0.088 on log establishment entries implies that the 95\% confidence interval excludes effects larger than approximately $\pm 0.17$ log points. Given that a 10-percentage-point increase in the confirmation rate is a large shift, this means the design can rule out effects of roughly 1.7\% on state-level establishment entry per percentage point of confirmation. Individual-level effects on filers' own entrepreneurship could be much larger while remaining undetectable in state aggregates, since Chapter~13 filers constitute a small share of a state's entrepreneurial population. This aggregation attenuation is a fundamental limitation: the null should be interpreted as ``no detectable aggregate effect'' rather than ``no individual-level effect.''

\subsubsection*{Timing caveat} Chapter~13 plans run 3--5 years. Because outcomes are measured relative to the filing year, the $t+1$ and $t+2$ horizons capture periods when most filers are still in their repayment plans and have not yet received discharge. The null at these horizons may therefore reflect the mechanical fact that debt relief has not yet materialized, rather than a true absence of post-discharge entrepreneurial response. Future work aligning outcomes to actual discharge dates could provide sharper tests of the debt overhang hypothesis.

\subsection{Robustness}

\input{tables/tab4_robustness}

\Cref{tab:robustness} presents diagnostic tests that validate the null as a design success rather than a design failure. Panel~A reports balance regressions: judge leniency does not predict log case counts ($p=0.13$) or the number of judges ($p=0.93$), consistent with random assignment. The marginally elevated balance coefficient for case counts likely reflects sampling noise rather than systematic assignment violation.

Panel~B reports placebo tests using concurrent-year outcomes. Judge leniency in year $t$ does not predict establishment entry ($p=0.95$) or employment ($p=0.81$) in the same year, ruling out confounding from local economic conditions.

The null is robust to alternative confirmation thresholds. Using 365 days (more inclusive) or 1,095 days (more conservative) produces reduced-form coefficients of 0.021 ($p=0.81$) and 0.045 ($p=0.59$), respectively---both near zero. There is no heterogeneity by baseline entry rates: the interaction between leniency and high-entry districts is insignificant.

\section{Discussion}
\label{sec:discussion}

The null result---that randomly induced variation in bankruptcy debt relief does not increase business formation---has important implications for both theory and policy.

\subsubsection*{Why the null is informative} Three candidate mechanisms can explain why debt relief fails to promote entrepreneurship among Chapter~13 filers. First, the typical filer may lack the human capital, industry networks, or risk tolerance required for entrepreneurship, regardless of balance sheet position. Debt overhang may bind for investment by existing firms \citep{myers1977} without binding at the extensive margin of firm creation among consumer debtors. Second, the 3--5 year repayment period itself may delay or deter entrepreneurial plans: even with eventual confirmation, the years spent under court supervision may foreclose the optimal window for starting a business. Third, the credit market consequences of a bankruptcy filing---including the long-lived mark on the debtor's credit report---may offset the benefit of discharge by constraining access to startup capital \citep{keys2020}.

\subsubsection*{Relation to prior work} The null is broadly consistent with \citet{hurst2004}'s finding that the wealth--entrepreneurship gradient is driven by preferences rather than liquidity constraints. It also complements \citet{dobbie2015}'s finding that bankruptcy protection increases labor supply but does not increase self-employment income. Combined, these results paint a picture in which the fresh start facilitates recovery through wage employment---the margin most accessible to debtors---rather than through the more capital- and skill-intensive margin of entrepreneurship.

\subsubsection*{Limitations} Several caveats apply. The use of case duration as a proxy for confirmation introduces measurement error that attenuates estimates toward zero. The state-level outcome matching is imprecise for states with multiple bankruptcy districts. With only 10 court clusters, inference relies on asymptotic approximations that may be inexact. And the aggregation from individual cases to court-year panels sacrifices the case-level variation that makes judge-IV designs powerful. Future work linking individual bankruptcy records directly to state business registrations could provide sharper estimates at the individual margin.

\section{Conclusion}

The fresh start provided by Chapter~13 bankruptcy helps debtors recover---but it does not, by itself, create entrepreneurs. Using the cleanest available source of exogenous variation in consumer debt relief, this paper finds no detectable effect on business formation. The null is consistent across outcomes, time horizons, and specifications, and survives extensive diagnostic testing. The finding reframes the policy debate: the ``fresh start'' is valuable for what it demonstrably delivers---improved financial health, higher labor supply, better outcomes for children---rather than for the productive spillovers that debt overhang theory might predict. For the broader literature on financial constraints and entrepreneurship, the result underscores that the barriers to firm creation among consumer debtors lie beyond the balance sheet.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{CourtListener RECAP Archive} The RECAP Archive (\url{https://www.courtlistener.com/}) is a free, public repository of federal court documents maintained by the Free Law Project. I access case-level data through the REST API (v4), filtering for Chapter~13 bankruptcy cases (search type ``r'', query ``chapter 13'') filed between January 1, 2010 and December 31, 2019 in the 10 selected courts. For each case, I extract: court identifier, assigned judge name and ID, filing date, and termination date. I sample cases in quarterly batches (20 per court-quarter) and clean judge names by standardizing capitalization and removing system artifacts.

\subsubsection*{Census Business Dynamics Statistics} The BDS data are accessed through the Census Bureau's API (\url{https://api.census.gov/data/timeseries/bds}). I retrieve annual state-level counts of establishment entries (births), exits, total firms, total establishments, and employment for the 10 states containing the selected bankruptcy courts over 2010--2021. The BDS covers the universe of employer businesses with paid employees.

\subsubsection*{Confirmation proxy construction} I define plan confirmation using a duration threshold of 730 days. Cases with $\text{Duration} > 730$ are classified as confirmed; cases with $\text{Duration} \leq 730$ are classified as dismissed or converted. This threshold is conservative: Chapter~13 plans run 36--60 months, so most confirmed plans will have durations well above 730 days. The threshold excludes early completions (rare in Chapter~13) and includes some cases that may have been dismissed after prolonged proceedings; both sources of misclassification bias estimates toward zero.

\section{Identification Appendix}

\subsubsection*{Balance tests} I regress pre-determined court-year characteristics (log case counts, number of active judges) on leave-one-out judge leniency with court and year fixed effects. Under random assignment, leniency should be orthogonal to these variables conditional on fixed effects.

\subsubsection*{Placebo test} I use contemporaneous (year $t$) business applications as a placebo outcome. Since Chapter~13 plans run 3--5 years, filers in year $t$ are still in their repayment plans during year $t$ and have not yet received discharge. Any effect of leniency on same-year business formation would indicate confounding rather than a causal channel through plan confirmation.

\section{Robustness Appendix}

\subsubsection*{Alternative thresholds} I re-estimate the main specification using 365-day and 1,095-day thresholds for the confirmation proxy. The 365-day threshold is more inclusive (classifying more cases as confirmed), while the 1,095-day threshold is more conservative.

\subsubsection*{Duration-based leniency} As an alternative to the binary confirmation proxy, I use the leave-one-out mean case duration for each judge as a continuous leniency measure. This avoids the arbitrary threshold choice but may be noisier.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
