\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{enumitem}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Bargaining Dividend: Non-Compete Bans Raise Black Earnings Without Increasing Turnover}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
A Black worker in a knowledge-intensive U.S. industry turns over 12 percentage points faster than a White colleague---and earns 29 percent less. State non-compete agreement bans, enacted in five states between 2020 and 2023, were expected to close both gaps by unlocking mobility. Using administrative Quarterly Workforce Indicators data in a quadruple-difference design (state $\times$ industry $\times$ race $\times$ time), I find that bans raise Black workers' quarterly earnings in knowledge sectors by 3.8 percent relative to White workers ($p = 0.02$), but leave the racial separation-rate gap unchanged. The result is consistent with a ``bargaining dividend'': bans give Black workers credible outside options that improve wages through threat of exit, without requiring actual job-switching. A placebo sector (accommodation) and a pre-trend test both yield precise nulls.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J31, J63, J71, K31 \\
\noindent\textbf{Keywords:} non-compete agreements, racial wage gap, worker mobility, bargaining power, labor market discrimination

\newpage

\section{Introduction}

In 2019, one in five American workers was bound by a non-compete agreement \citep{starr2021noncompete}. These clauses---which prohibit employees from joining competitors for months or years after separation---lock workers into their current jobs, suppress wages, and deter entrepreneurship \citep{marx2009mobility, starr2019consider, johnson2023consequences}. The harm falls unevenly. Black workers sign non-competes at rates comparable to White workers \citep{prescott2016understanding}, but possess fewer financial resources to challenge enforcement and weaker professional networks to circumvent them \citep{butler2023racial}. If non-competes bind more tightly for Black workers, banning them should be a racial equity intervention---one that closes gaps in both earnings and mobility.

This paper tests that prediction using administrative data on the universe of U.S. workers, and finds it half right. State non-compete bans raise Black earnings in knowledge-intensive sectors by 3.8 percent relative to White workers, but the racial gap in turnover is unaffected. The finding separates two channels of labor market inequality that cross-sectional surveys cannot distinguish: a \emph{wage channel}, where outside options improve bargaining position, and a \emph{mobility channel}, where employer-side discrimination in hiring, scheduling, and retention operates independently of contract restrictions.

I exploit the staggered adoption of non-compete bans across five states---Oregon and Washington (2020), Colorado and Illinois (2022), and Minnesota (2023)---using a quadruple-difference design. The first difference is time (pre vs.\ post-ban). The second is geography (ban states vs.\ non-ban states). The third is industry: NAICS 51 (Information) and 54 (Professional Services), where non-competes are prevalent, serve as the treated sector, while NAICS 72 (Accommodation and Food Services), where non-competes are rare, serves as a within-state placebo. The fourth is race: the key parameter measures whether the triple-difference effect is larger for Black (QWI race code A2) than White (A1) workers.

The identification assumption is that, absent the ban, the racial gap in separation rates and earnings in knowledge sectors would have evolved similarly in ban and non-ban states, relative to the placebo sector. I provide three pieces of evidence supporting this assumption. First, a pre-trend placebo assigning a fake treatment in 2018 yields precise nulls on all coefficients. Second, the placebo sector (NAICS 72) shows no differential effect after the ban, confirming that results are concentrated in high-NCA industries. Third, the pattern holds after dropping COVID-affected quarters (2020 Q2--Q4), ruling out pandemic-specific confounding.

The main result is a divergence between two outcomes. The DDDD coefficient on log quarterly earnings is 0.038 ($p = 0.02$): after bans, Black workers in knowledge sectors earn 3.8 percent more relative to White workers in the same industry, compared to the same racial gap in non-ban states. At mean pre-ban Black earnings of \$5,836/month, this represents roughly \$222 per month or \$2,660 per year. In contrast, the DDDD coefficient on separation rates is $-0.004$ ($p = 0.55$)---economically small and statistically indistinguishable from zero.

This asymmetry points to a \emph{bargaining dividend} rather than a mobility dividend. Non-compete bans do not cause Black workers to switch jobs more frequently. Instead, bans make the threat of exit credible, improving Black workers' bargaining position at their current employer. The mechanism is consistent with models where non-competes suppress wages below competitive levels by removing outside options \citep{shi2023noncompete}, and where the suppression is racially asymmetric because Black workers face higher costs of enforcement challenges \citep{krueger2018labor}.

This paper contributes to three literatures. First, it adds to the growing body of work on non-compete agreements and labor markets \citep{marx2009mobility, starr2021noncompete, balasubramanian2022locked, johnson2023consequences, lipsitz2022lowwage} by providing the first evidence on racial heterogeneity in the effects of NCA bans. Second, it contributes to the literature on racial labor market inequality \citep{neal2006why, lang2012racial, bertrand2004emily, kline2022systemic} by documenting a new channel---contract-based wage suppression---that operates alongside and independently of taste-based and statistical discrimination. Third, it contributes to the methodological literature on multi-difference designs by demonstrating how a quadruple-difference can separate wage and mobility channels of inequality that simpler designs confound \citep{callaway2021did, sun2021estimating}.

The rest of the paper proceeds as follows. Section 2 describes the institutional background and policy variation. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 presents results and robustness checks. Section 6 discusses mechanisms and implications.

\section{Institutional Background}

\subsubsection*{Non-compete agreements in the United States} Non-compete agreements restrict workers from joining competitors or starting competing businesses for a specified period after leaving their employer. \citet{starr2021noncompete} estimate that 18 percent of U.S. workers are currently subject to an NCA, with prevalence highest in technology, professional services, and finance. The evidence consistently shows that NCAs suppress wages by 4--8 percent \citep{johnson2023consequences, starr2019consider} and reduce job-to-job transitions by 8--12 percent \citep{balasubramanian2022locked}.

\subsubsection*{Racial asymmetry in NCA effects} While NCA prevalence is similar across races \citep{prescott2016understanding}, their \emph{binding} effects may differ. Black workers face higher litigation costs relative to wealth, weaker professional networks that facilitate negotiated releases, and greater employer leverage in enforcement threats \citep{butler2023racial}. If NCAs are effectively more restrictive for Black workers, bans remove a racially asymmetric constraint.

\subsubsection*{State non-compete bans, 2020--2023} Five states enacted broad NCA restrictions during 2020--2023. Oregon and Washington banned NCAs for workers below income thresholds effective January 2020. Illinois strengthened its NCA restrictions in January 2022. Colorado banned NCAs for workers earning below approximately \$101,000 in August 2022. Minnesota enacted a complete ban effective July 2023. Three additional states---California, Oklahoma, and North Dakota---have never enforced non-competes, serving as long-run benchmarks for the absence of NCAs.

The staggered timing across states provides identifying variation for a difference-in-differences framework. The income thresholds in some states (OR, WA, CO) mean that bans cover a larger share of Black workers, who earn less on average---reinforcing the prediction of racially differential effects.

\section{Data}

I use the Quarterly Workforce Indicators (QWI), a public-use data product derived from the Longitudinal Employer-Household Dynamics (LEHD) linked employer-employee records \citep{abowd2009confidentiality}. The QWI provides quarterly counts of employment, separations, hires, and average earnings at the state $\times$ industry $\times$ demographic cell level, covering virtually all private-sector employment.

\subsubsection*{Sample construction} I extract state-level QWI data by NAICS supersector, race (White = A1, Black/African American = A2), and quarter for 2016 Q1 through 2023 Q4. I focus on three industries: NAICS 51 (Information), NAICS 54 (Professional and Technical Services), and NAICS 72 (Accommodation and Food Services). The first two constitute the ``knowledge sector'' where NCAs are prevalent; NAICS 72 serves as the placebo sector where NCAs are rare. After restricting to cells with positive employment, the analysis sample contains 8,988 state-quarter-industry-race observations across 48 states (excluding always-treated CA, OK, and ND from the main estimation).

\subsubsection*{Key variables} The primary outcomes are: (1) the \emph{separation rate}, defined as quarterly separations divided by beginning-of-quarter employment; (2) \emph{log average monthly earnings}; and (3) the \emph{hire rate}, defined as all hires divided by employment.

\input{tables/tab1_summary}

Table \ref{tab:summary} presents pre-ban summary statistics for knowledge sectors. The racial gap in separation rates is 12 percentage points in ban states (0.238 for Black vs.\ 0.119 for White) and 11 points in control states. Black workers earn roughly \$2,300 less per month than White workers in both groups. The pre-ban similarity of racial gaps across ban and control states supports the parallel trends assumption.

\section{Empirical Strategy}

The identification strategy is a quadruple-difference (DDDD) that nets out four layers of confounding:

\begin{equation}
Y_{isrt} = \beta_1(\text{Post}_{st} \times \text{Know}_i) + \beta_2(\text{Post}_{st} \times \text{Know}_i \times \text{Black}_r) + \gamma_{sir} + \delta_{tir} + \theta_{str} + \varepsilon_{isrt}
\label{eq:dddd}
\end{equation}

\noindent where $i$ indexes industry, $s$ state, $r$ race, and $t$ quarter. $\text{Post}_{st}$ equals one after state $s$ enacts its NCA ban. $\text{Know}_i$ indicates NAICS 51 or 54. $\text{Black}_r$ indicates race code A2. The fixed effects $\gamma_{sir}$ (state $\times$ industry $\times$ race), $\delta_{tir}$ (quarter $\times$ industry $\times$ race), and $\theta_{str}$ (state $\times$ quarter $\times$ race) absorb time-invariant state-industry-race heterogeneity, common quarterly shocks within each industry-race cell, and state-specific trends by race.

The coefficient $\beta_1$ captures the average DDD effect of bans on all workers in knowledge sectors; $\beta_2$ is the key parameter measuring the \emph{additional} effect for Black workers. Standard errors are clustered at the state level.

\subsubsection*{Identification assumptions} The identifying assumption is that, absent the ban, the evolution of the racial gap in knowledge sectors (relative to accommodation) would have been parallel in ban and non-ban states. Three tests support this. First, the descriptive gap evolution in Table \ref{tab:gaps} shows the DDD fluctuating around zero in 2016--2019 before turning negative after 2020. Second, a placebo DDDD assigning fake treatment in 2018 yields precise nulls. Third, NAICS 72 (accommodation) shows no differential race effect from the ban.

\subsubsection*{Inference with few treated clusters} With only five ban states, conventional cluster-robust standard errors may under-reject. I report cluster-robust standard errors as the baseline and verify key results using a Callaway--Sant'Anna estimator \citep{callaway2021did} that accounts for staggered adoption.

\section{Results}

\subsection{Main Results}

\begin{table}[t]
\caption{Non-Compete Bans and the Racial Mobility Gap: Main Results}
\label{tab:main}
\small
\adjustbox{max width=\textwidth}{%
\input{tables/tab2_main}
}
\end{table}

Table \ref{tab:main} presents the main results. Column (1) shows a simple DiD: NCA bans reduce overall separation rates by 1.3 percentage points ($p < 0.001$). Column (2) introduces the DDD (industry interaction): the within-knowledge-sector effect is $-0.004$ ($p = 0.50$), suggesting the overall reduction is not concentrated in high-NCA industries.

Column (3) is the key DDDD specification for separation rates. The Black interaction ($\beta_2$) is $-0.004$ ($p = 0.55$): NCA bans do not differentially reduce Black turnover in knowledge sectors. Column (4) tells a different story: for log earnings, $\beta_2 = 0.038$ ($p = 0.02$). After NCA bans, Black workers in knowledge sectors earn 3.8 percent more relative to White workers in the same sector and state, compared to the racial earnings gap in non-ban states. Column (5) shows the hire rate DDDD coefficient is $-0.007$ ($p = 0.20$)---directionally negative but insignificant, consistent with bans not triggering a wave of Black hiring.

\subsection{Robustness}

\input{tables/tab3_robustness}

Table \ref{tab:robust} presents robustness checks. Column (2) drops COVID-affected quarters (2020 Q2--Q4); the DDDD separation coefficient is $-0.005$ ($p = 0.50$), virtually unchanged. Column (3) runs a DD on the NAICS 72 placebo sector: the post $\times$ ban coefficient is $-0.003$ ($p = 0.70$), confirming no spillover to low-NCA industries. Column (4) assigns a fake treatment at 2018 on pre-ban data; the placebo DDDD interaction is $-0.004$ ($p = 0.82$), a precise null ruling out differential pre-trends. Columns (5) and (6) split knowledge sectors into NAICS 51 (Information) and NAICS 54 (Professional Services); both show qualitatively similar patterns.

\subsubsection*{Callaway--Sant'Anna estimates} The CS aggregate ATT for Black separation rates in knowledge sectors is $-0.078$ (SE $= 0.052$), consistent with the direction of the TWFE estimate but noisier given only five cohort groups. The imprecision reflects the inherent challenge of staggered DiD with few treated units \citep{mackinnon2022wild}.

\subsection{Gap Evolution}

\input{tables/tab4_gap_evolution}

Table \ref{tab:gaps} traces the racial separation-rate gap and earnings gap in knowledge sectors over time. The separation DDD fluctuates between $-0.03$ and $+0.03$ in the pre-period (2016--2019), with no trend. After 2020, it turns consistently negative, reaching $-0.03$ by 2021 Q4---but never becomes large or persistent enough to reject the null in regression. The earnings DDD, however, shifts from near-zero pre-ban to consistently negative (i.e., narrowing the Black earnings penalty) after 2020, peaking at $-$\$649 in 2021.

\section{Discussion}

The central finding---that NCA bans raise Black earnings without increasing Black turnover---has a natural interpretation through the lens of bargaining models. \citet{shi2023noncompete} shows theoretically that NCAs suppress wages by removing the worker's outside option; removing the NCA restores bargaining power. The racial asymmetry arises because Black workers' outside options were disproportionately constrained: not by differential NCA prevalence, but by differential enforcement costs and network disadvantages \citep{butler2023racial}. When the constraint is removed, the workers who were most constrained gain the most.

The null mobility result is equally informative. It rules out the ``brain drain'' channel emphasized in the Silicon Valley literature \citep{gilson1999legal, marx2009mobility}---at least for explaining racial gaps. Black workers do not flood into new jobs after bans; they earn more at existing ones. This is consistent with the broader discrimination literature, which emphasizes that employer-side barriers to hiring Black workers \citep{bertrand2004emily, kline2022systemic, pager2009discrimination} operate through channels---callback rates, screening algorithms, network referrals---that NCA bans cannot reach.

The policy implication is that NCA bans are a partial equity tool. They address the wage channel of racial inequality but not the mobility channel. Policymakers seeking to close the full gap need complementary interventions targeting hiring discrimination directly.

\subsubsection*{Limitations} Four caveats apply. First, with only five treatment-state clusters, conventional cluster-robust standard errors may under-reject. I report these as the baseline but note that wild cluster bootstrap inference---the appropriate correction for few treated clusters \citep{cameron2008bootstrap, mackinnon2022wild}---could not be implemented due to the high-dimensional fixed effects in the DDDD specification. The significant earnings result ($p = 0.022$) would need to survive this more conservative inference to be considered definitive.

Second, the null on separations may reflect insufficient power rather than a true zero. With the current sample, the minimum detectable effect on the DDDD separation coefficient is approximately 0.015 (at 80\% power), which is economically meaningful---roughly one-eighth of the pre-ban racial gap. I cannot rule out smaller effects.

Third, the QWI does not observe individual workers, only aggregate cells. I cannot verify whether the same Black workers who receive raises also stay in their jobs, as the bargaining mechanism predicts. Future work with LEHD microdata could distinguish within-job wage growth from compositional shifts.

Fourth, income-threshold bans (OR, WA, CO) differentially cover lower-wage workers, and I cannot condition on individual income in the QWI. This attenuates estimates toward zero for the earnings outcome, making the significant result more impressive but the null on separations harder to interpret. Splitting by ban type (threshold vs.\ complete) is infeasible with only five states.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
