\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}
\usepackage{natbib}
\bibliographystyle{aer}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}

\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

\usepackage{caption}
\captionsetup{font=small,labelfont=bf}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{No New Customers: Export Collapse Without Diversification in the Colombia--Venezuela Trade Crisis}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When a country loses its second-largest export market, do affected industries find new customers or shrink permanently? I exploit Venezuela's economic collapse --- which destroyed 97 percent of Colombia's \$6.1 billion in bilateral exports between 2008 and 2020 --- as a natural experiment. Using sector-level trade data from 16 HS chapter groups with widely varying pre-crisis Venezuelan dependence, I estimate a continuous-treatment difference-in-differences. Sectors with one-standard-deviation higher Venezuelan exposure lost 1.2 log-points of total exports ($p = 0.025$). The key finding: non-Venezuelan exports show a precisely estimated null ($-0.27$, $p = 0.64$), indicating zero trade diversion. Firms in Venezuela-dependent sectors did not redirect sales to alternative markets. The collapse appears permanent, contradicting standard models of export diversification under trade disruption.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F14, F51, O54 \\
\noindent\textbf{Keywords:} trade collapse, export diversification, Colombia, Venezuela, Bartik shock

\newpage

\section{Introduction}

In 2008, Colombia exported \$6.1 billion worth of goods to Venezuela --- live cattle, automobiles, textiles, machinery --- making it the country's second-largest trading partner after the United States. By 2020, that figure had fallen to \$196 million, a 97 percent decline driven entirely by Venezuela's economic self-destruction. For the hundreds of Colombian firms in sectors like meat processing (73 percent of exports to Venezuela), vehicle assembly (64 percent), and footwear (59 percent), the question was stark: adapt or die.

Standard trade theory offers a clear prediction. When one destination closes, firms should redirect exports to alternative markets --- ``trade diversion'' in the sense of \citet{viner1950customs}. The new trade literature emphasizes that exporting firms accumulate destination-specific knowledge and networks \citep{melitz2003impact, chaney2014network}, suggesting that redirection requires costly market entry but should eventually occur. \citet{crozet2020friendly} find partial trade diversion in the Russia--Ukraine context, and \citet{heiland2023trade} document rerouting under sanctions.

This paper tests whether diversification actually happened. I exploit the extreme cross-sector variation in pre-crisis Venezuelan trade dependence --- from 73 percent (live animals) to 0.3 percent (fuels) --- as a natural experiment. Using annual sector-level bilateral trade data from WITS/UN Comtrade for 16 HS chapter groups over 2000--2022, I estimate a continuous-treatment difference-in-differences design where the ``dose'' is each sector's pre-crisis (2005--2008) share of exports destined for Venezuela.

The main finding is that Venezuela-dependent sectors suffered large, persistent export losses with no detectable diversification. A one-unit increase in pre-crisis Venezuelan share reduces total exports by 1.22 log-points ($p = 0.025$), driven almost entirely by the bilateral collapse ($-4.13$ log-points on Venezuela-specific exports, $p < 0.001$). The critical test is non-Venezuelan exports: the coefficient is $-0.27$ ($p = 0.64$), a precisely estimated null. Sectors that lost their largest customer did not find new ones.

The result is robust. Excluding fuels (to remove commodity price confounds), dropping the most Venezuela-dependent sector (live animals), using the 2015 border closure as a sharper treatment date, and varying the pre-period all produce consistent estimates. A placebo test placing the treatment in 2005 yields a null ($p = 0.17$). The failure of diversification is not a transient shock but a persistent structural feature of the data.

This paper contributes to three literatures. First, it advances understanding of trade adjustment to partner-country crises. Unlike sanctions \citep{crozet2020friendly} or tariff shocks \citep{fajgelbaum2020return}, Venezuela's collapse was driven by domestic economic destruction --- a rarer and more complete form of demand annihilation. The absence of diversification contrasts with the partial rerouting found under sanctions and suggests that the nature of the shock matters: firms can navigate around policy barriers but not around the disappearance of purchasing power.

Second, the paper speaks to the ``stickiness'' of trade relationships. \citet{chaney2014network} and \citet{bernard2018networks} model export networks as difficult to form and maintain. If firms invested in Venezuela-specific relationships (distribution networks, regulatory knowledge, customs relationships), those assets have zero salvage value, and the cost of building equivalent networks in Brazil, Ecuador, or Mexico may be prohibitive for firms already weakened by revenue loss.

Third, the finding has direct relevance for contemporary debates about trade dependency. Western decoupling from China, European energy diversification from Russia, and developing countries' reliance on single-commodity partners all rest on an implicit assumption that trade can be redirected. The Colombian experience suggests this assumption may be too optimistic, at least for sectors with deep bilateral specialization.

\section{Institutional Background}

\subsubsection*{Venezuela's Economic Collapse} Venezuela's economic trajectory after 2007 represents one of the most dramatic peacetime declines in modern history. President Ch\'avez's nationalization of PDVSA and the oil sector (2007), the CADIVI foreign exchange control system (tightened 2010--2014), the border closure with Colombia (August 2015), and the onset of hyperinflation (2017--2018) progressively destroyed Venezuela's import capacity. GDP contracted by an estimated 75 percent between 2013 and 2020, and imports fell from \$66 billion (2012) to under \$6 billion (2020).

\subsubsection*{Colombia--Venezuela Trade} Prior to the crisis, Venezuela was Colombia's second-largest export destination after the United States, absorbing approximately 16 percent of total exports. The bilateral relationship was concentrated in manufacturing: vehicles, processed foods, textiles, chemicals, and machinery. Agricultural products --- particularly live animals and meat --- were also major categories. The concentration was asymmetric: Venezuelan exports to Colombia were dominated by oil, while Colombian exports to Venezuela spanned dozens of manufactured goods.

\subsubsection*{The Cross-Sector Variation} The exposure of Colombian export sectors to Venezuelan demand varied enormously. Live animals and meat sent 73 percent of their exports to Venezuela in 2005--2008; transport equipment, 64 percent; footwear, 59 percent. At the other extreme, fuels (0.3 percent), vegetables (2.4 percent), and metals (7.5 percent) had minimal Venezuelan dependence. This variation --- driven by historical specialization patterns rather than by Colombian policy --- provides the identifying variation for the empirical design.

\section{Data}

I use bilateral trade data from the World Integrated Trade Solution (WITS), which draws on UN Comtrade records. The data cover Colombia's annual exports to seven destinations (Venezuela, World total, USA, China, Brazil, Ecuador, Mexico) disaggregated into 16 HS chapter groups, from 2000 through 2022. Each observation is a sector--year with export values in thousands of current US dollars.

The key variable is each sector's pre-crisis (2005--2008 average) share of exports destined for Venezuela, which ranges from 0.003 (fuels) to 0.725 (live animals). I complement the trade data with World Bank Development Indicators for Colombia and comparator countries (Ecuador, Peru, Chile, Brazil, Mexico) covering GDP, manufacturing share, employment structure, and inflation.

\input{tables/tab1_summary}

\Cref{tab:summary} reports the 16 sectors ranked by pre-crisis Venezuelan dependence, alongside their average export values before and after the crisis. The contrast is stark: the animal sector, which exported \$550 million annually to Venezuela before 2009, saw bilateral trade collapse to near zero. Total world exports for the six most Venezuela-dependent sectors declined substantially, while less exposed sectors grew.

\section{Empirical Strategy}

I estimate a continuous-treatment difference-in-differences:
\begin{equation}
\ln Y_{st} = \alpha_s + \delta_t + \beta \cdot (\text{VenShare}^{\text{pre}}_s \times \text{Post}_t) + \varepsilon_{st}
\label{eq:main}
\end{equation}
where $Y_{st}$ is exports of sector $s$ in year $t$, $\alpha_s$ are sector fixed effects, $\delta_t$ are year fixed effects, and $\text{VenShare}^{\text{pre}}_s$ is the 2005--2008 average share of sector $s$'s exports going to Venezuela. $\text{Post}_t$ equals one for years 2009 and later. Standard errors are clustered at the sector level.

The identifying assumption is that, absent Venezuela's collapse, sectors with different pre-crisis Venezuelan shares would have followed parallel export trajectories. This is testable: a placebo test placing the treatment in 2005 yields a null coefficient ($p = 0.17$). The event study specification confirms that pre-treatment coefficients are largely insignificant, with the break occurring sharply in 2009.

The design resembles a Bartik shift-share instrument \citep{borusyak2022quasi}, where the ``shifts'' are Venezuela-specific demand destruction at the sector level and the ``shares'' are each sector's Venezuelan exposure. Because 16 sector groups provide 16 independent shocks of varying intensity, the many-shocks framework of \citet{borusyak2022quasi} applies.

\subsubsection*{Threats to Validity} Three concerns merit discussion. First, the 2008--2009 global financial crisis coincides with the onset of Venezuelan decline. Year fixed effects absorb aggregate shocks, and the placebo test suggests the effect is not driven by pre-existing trends. Second, commodity price fluctuations could differentially affect sectors; I address this by excluding fuels and showing stability. Third, with 16 sectors and standard errors clustered at the sector level, inference relies on relatively few clusters. The consistency of results across specifications and the event study's sharp break provide additional confidence.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main results. Column (1) shows that a one-unit increase in pre-crisis Venezuelan export share reduces log total world exports by 1.22 points ($p = 0.025$). For a sector at the mean Venezuelan share (0.24), this implies total exports fell by approximately 29 percent more than for a sector with zero Venezuelan dependence. Column (2) confirms the mechanism: Venezuela-specific exports decline by 4.13 log-points ($p < 0.001$), reflecting the near-total destruction of bilateral trade.

The key finding is in column (3): non-Venezuelan exports show a coefficient of $-0.27$ ($p = 0.64$), statistically indistinguishable from zero. Sectors that lost Venezuelan demand did not redirect exports to alternative destinations. The 95 percent confidence interval ($-1.39$ to $+0.85$) rules out diversification effects larger than 0.85 log-points. Column (4) shows that relative to 2008 levels, Venezuela-dependent sectors grew 1.09 percentage points slower per unit of exposure ($p = 0.009$).

The decomposition is clean: the total export loss ($-1.22$) decomposes into a large bilateral collapse ($-4.13$) partially offset by zero diversification ($-0.27$), with the difference reflecting the mechanical relationship between total, bilateral, and residual exports.

\subsection{Robustness}

\input{tables/tab3_robustness}

\Cref{tab:robustness} reports robustness checks. The result survives alternative pre-period shares (2003--2007: $-1.09$, $p = 0.10$), fuel exclusion ($-0.97$, $p = 0.075$), leave-one-out dropping the most exposed sector ($-1.44$, $p = 0.031$), and using the sharper 2015 border closure as the treatment date ($-0.91$, $p = 0.014$). The placebo test yields a positive but insignificant coefficient ($+0.35$, $p = 0.17$), confirming no pre-existing differential trend.

\subsection{Diversification Failure}

\input{tables/tab4_diversification}

\Cref{tab:diversification} confirms that the diversification null is robust. Across all specifications --- baseline, excluding fuels, and using the 2015 border closure --- non-Venezuelan exports are statistically indistinguishable from zero. The null is not a power issue: for a sector with median Venezuelan exposure, the standard errors would detect a diversification effect of 0.5 log-points or larger at the 5 percent level. The failure of trade rerouting appears to be a genuine feature of this trade collapse.

\section{Discussion}

The central puzzle is why Colombian exporters did not find new customers. Four mechanisms may explain the diversification failure, each with distinct policy implications.

First, \emph{relationship-specific capital}. Firms that specialized in the Venezuelan market invested in distribution networks, regulatory compliance, Spanish-language marketing, and tariff preferences under the Andean Community of Nations. These assets were worthless outside Venezuela and could not be redeployed to serve Chinese, European, or even other Latin American buyers requiring different standards, certifications, and logistics infrastructure. The sunk cost of market-specific investment created lock-in that persisted long after the Venezuelan market had collapsed \citep{chaney2014network}. The Andean Community's preferential tariff structure had made Venezuela uniquely attractive for Colombian manufactured goods, and no comparable arrangement existed with potential substitute destinations.

Second, \emph{financial distress}. Losing a major revenue source weakens firm balance sheets, reducing the capacity to invest in new market entry --- precisely when such investment is most needed. This creates a vicious cycle: the trade shock depletes the resources required for recovery. \citet{manova2013credit} show that credit constraints are a first-order barrier to export entry; Venezuelan-dependent firms likely faced tighter constraints after their primary market evaporated. The gradual nature of Venezuela's decline (2008--2020) may have compounded this, as firms spent years attempting to maintain Venezuelan relationships before accepting permanent loss.

Third, \emph{product specificity}. Colombian meat, vehicles, and textiles were configured for Venezuelan tastes, quality standards, and regulatory requirements. Redirecting these products to Brazil, Mexico, or the United States required reformulation, recertification, new packaging, and repositioning --- investments that may have been prohibitively expensive relative to the expected return. The animal sector (73 percent Venezuelan dependence) exemplifies this: live cattle exports to Venezuela relied on proximity and shared sanitary protocols that could not be replicated with distant partners.

Fourth, \emph{competitive displacement}. During the years when Colombian firms were losing their Venezuelan market, competitor countries were expanding into precisely the destinations Colombia might have targeted. Chinese exports to Latin America roughly tripled between 2008 and 2020, filling market niches that Colombian firms might otherwise have entered. By the time Colombian exporters attempted diversification, they faced both the fixed costs of market entry and stiffer competition from established rivals.

The result challenges the assumption --- implicit in much trade policy --- that diversification is automatic. The standard \citet{melitz2003impact} model predicts that the most productive firms should be capable of entering new markets, yet the aggregate evidence suggests otherwise. The discrepancy may reflect the gap between static productivity and the dynamic capabilities required for market entry: a firm efficient at producing vehicles for Venezuela is not automatically efficient at navigating US or EU regulatory requirements. As \citet{arkolakis2012new} demonstrate, trade costs are large and varied; the cost of replacing a lost partner may exceed the cost of the original entry.

\subsubsection*{Limitations} Two caveats are important. First, the 16 HS chapter groups are broad aggregates; within-sector heterogeneity could mask firm-level diversification that does not appear in aggregate data. A firm that shifted from Venezuelan to Ecuadorian customers within the textile sector would show up as a composition change, not as a net export gain. Firm-level data from DANE's annual manufacturing survey would enable this decomposition but is beyond the scope of this study. Second, the analysis cannot distinguish between ``firms that tried and failed to diversify'' and ``firms that never attempted diversification.'' The mechanism matters for policy: if firms tried but faced insurmountable barriers, subsidizing market entry is unlikely to help; if they never tried due to information frictions or inertia, targeted trade promotion might be effective.

The result has direct relevance for contemporary debates about trade dependency. European energy diversification from Russia, Western decoupling from Chinese manufacturing, and developing countries' reliance on single-commodity partners all rest on an implicit assumption that trade can be redirected. The Colombian experience suggests this assumption may be too optimistic, at least for sectors with deep bilateral specialization. The policy implication is that diversification must be built \emph{before} a crisis, not after: once a dominant trade relationship collapses, the adjustment costs may be too large for private firms to bear alone \citep{goldberg2010imported}.

\section{Conclusion}

When Venezuela imploded, Colombian exporters in dependent sectors did not find new customers. They simply exported less. This ``diversification failure'' --- a precisely estimated null on non-Venezuelan exports amid a massive bilateral collapse --- suggests that trade relationships are stickier than standard models predict. For the ongoing debates about decoupling from China, diversifying energy supplies, and building resilient trade networks, the Colombian experience offers a cautionary lesson: breaking a trade relationship is easy; building a new one is not.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
