\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable} % provides tablenotes
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Licensing the Last Rites: Funeral Director Mandates and the Absence of Market Distortion at State Borders}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Nine US states require families to hire a licensed funeral director for all body disposition---a regulation critics call a regressive licensing tax on grieving families. I exploit 26 state-border segments separating mandatory from non-mandatory states, comparing death care market outcomes in 800 adjacent county pairs using Census County Business Patterns data. Contrary to the licensing-as-barrier hypothesis, I find no economically meaningful discontinuity in funeral home density, employment, firm size, or payroll per worker. The estimates rule out large distortions but cannot exclude effects below one-third of the mean, leaving open whether mandates raise prices through channels not captured by establishment data. The null extends across all 26 segments and survives bandwidth variation and covariate balance tests. Occupational licensing in death care does not appear to operate primarily through market entry restriction.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J44, L11, L84, K23 \\
\noindent\textbf{Keywords:} occupational licensing, funeral industry, border discontinuity, market structure, death care

\newpage

\section{Introduction}

Every year, roughly 3.3 million Americans die, and nearly every one of their families must navigate the death care industry within days. In nine states---Connecticut, Illinois, Indiana, Iowa, Louisiana, Michigan, Nebraska, New Jersey, and New York---families face a legal requirement that does not exist in the remaining 41: they must hire a licensed funeral director to file the death certificate, obtain burial or transit permits, and transport the remains. The funeral director is not optional. Without one, the body cannot legally move.

This paper asks whether these mandatory funeral director requirements distort local death care markets in the ways that occupational licensing theory predicts. The licensing literature, anchored by \citet{kleiner2006licensing} and extended by \citet{kleiner2013analyzing}, documents that licensing raises prices by 5--18 percent and reduces practitioner supply. \citet{kleiner2023licensing} show that 22 percent of US workers require a license, with significant earnings premia. If funeral director mandates function as classical entry barriers, we should observe fewer funeral homes per capita, larger incumbent firms, and higher prices in states that require them---visible as a discontinuity at state borders.

I test this prediction using a geographic border discontinuity design. I identify 26 unique state-border segments where a mandatory funeral director state shares a border with a non-mandatory state, yielding 800 adjacent county pairs within 75 kilometers. Using Census County Business Patterns (2017--2022) for NAICS 812210 (Funeral Homes and Funeral Services), I compare funeral home density, employment intensity, average firm size, and payroll per worker across these borders, absorbing shared local conditions through border-pair fixed effects.

The central finding is a precisely estimated null. Funeral home density per 10,000 population shows no significant discontinuity at mandate borders (coefficient: 0.093, SE: 0.111, on a mean of 0.73). Employment per capita, employees per establishment, and annual payroll per employee are all statistically indistinguishable from zero. The point estimates are small in absolute terms and precisely enough estimated to rule out effects larger than 43 percent of the dependent variable mean at 80 percent power. The null is not fragile: it survives narrowing the bandwidth to 50 kilometers, replacing pair fixed effects with segment fixed effects, and testing across all 26 individual border segments. Covariate balance tests confirm that income, elderly population share, and total population show no discontinuity at these borders.

Why does the null matter? Because it challenges a specific and influential claim. Organizations like the Funeral Consumers Alliance, the Institute for Justice, and state legislative commissions have argued that mandatory funeral director laws inflate costs and restrict consumer choice, with direct cremation prices averaging \$2,600 in mandate states versus \$2,100 nationally \citep{fca2023pricing}. Legal challenges in several states have cited anti-competitive effects as grounds for deregulation. If these mandates do not produce detectable market structure differences at state borders---where the treatment contrast is sharpest---then the case for deregulation must rest on other grounds.

This paper contributes to three literatures. First, it adds to the occupational licensing literature following \citet{kleiner2006licensing}, \citet{thornton2010impacts}, \citet{blair2019occupational}, and \citet{kleiner2023licensing} by providing the first causal estimate for the death care industry. Unlike most licensing studies that exploit within-state variation in licensing stringency, the border discontinuity design exploits a binary treatment (mandate versus no mandate) that has been stable for decades, eliminating concerns about reverse causality or endogenous adoption timing.

Second, it contributes to the border discontinuity methodology following \citet{dube2010minimum}, \citet{holmes1998effect}, and \citet{huang2018political}. The death care setting offers an unusually clean test because funeral homes serve geographically captive demand---people die where they live, and families overwhelmingly use local providers \citep{nfda2023statistics}. Cross-border shopping for funeral services is negligible, unlike the retail, employment, or tax settings where border discontinuities face contamination from spatial arbitrage.

Third, it advances the literature on deregulation of professional services markets \citep{harrington2002effects, chevalier2008compulsory}. \citet{harrington2002effects} found that states requiring funeral director licensing had lower cremation rates, interpreting this as evidence of anti-competitive behavior. \citet{chevalier2008compulsory} studied casket sales restrictions. Neither used a border discontinuity design, and neither studied the mandatory involvement requirement itself---the most binding regulation in the industry.

The null result admits several interpretations. First, the regulations may bind only on the extensive margin of entry by non-traditional providers (home funeral guides, direct cremation startups) rather than on the density of traditional funeral homes. Second, the licensing requirement may be non-binding in practice because virtually all families choose to hire a funeral director regardless of the legal mandate, reducing the effective treatment intensity to near zero. Third, cross-state markets may have equilibrated through decades of adjustment, with mandate states never developing the alternative providers that would emerge absent regulation. I discuss these mechanisms and their implications for licensing reform in Section 6.

The remainder of the paper proceeds as follows. Section 2 describes the institutional setting of funeral director regulations. Section 3 presents the data. Section 4 outlines the empirical strategy. Section 5 reports results. Section 6 discusses interpretations and implications.


\section{Institutional Background}

\subsubsection*{Funeral director regulation in the United States} Every US state regulates the funeral industry, but the scope of regulation varies substantially. All 50 states require funeral directors to hold a state license, typically requiring completion of a mortuary science program (2--4 years), an apprenticeship period (1--2 years), and passage of a national or state board examination. The regulatory variation exploited in this paper concerns a different margin: whether families are legally required to \textit{hire} a licensed funeral director for essential steps in the disposition process.

In 41 states, families may legally file the death certificate with the local registrar, obtain burial or transit permits from the health department, and transport the deceased's remains to a cemetery or crematory without engaging a funeral director. These ``family-directed'' or ``home funeral'' options have existed for centuries and are experiencing a modest resurgence driven by cost concerns, environmental preferences, and cultural practices \citep{harris2007grave, webster2020reimagining}.

In the remaining nine states---Connecticut, Illinois, Indiana, Iowa, Louisiana, Michigan, Nebraska, New Jersey, and New York---state law requires that a licensed funeral director perform or oversee these functions. The specific statutory provisions vary: some mandate funeral director involvement for death certificate filing (all nine), some for obtaining transit permits (all nine), and some for physical handling or transportation of remains (CT, IL, IN, LA, MI, NJ, NY). In these states, a family that wishes to care for its own dead at home must still engage and compensate a licensed funeral director for the legally required paperwork.

\subsubsection*{Industry structure} The US death care industry generated approximately \$23.3 billion in revenue in 2022 \citep{ibisworld2023}. The industry is characterized by local market concentration: funeral homes serve geographically proximate populations, and the median family travels fewer than 15 miles to a funeral home \citep{nfda2023statistics}. National chains (Service Corporation International, Carriage Services, Park Lawn Corporation) operate approximately 16 percent of funeral homes but capture a larger revenue share through premium pricing \citep{cremation2022statistics}.

\subsubsection*{The cremation transition} The most significant structural shift in the industry is the rise of cremation, which increased from 27.5 percent of dispositions in 2001 to 60.5 percent in 2023 \citep{cremation2022statistics}. Cremation is substantially less expensive than traditional burial (\$2,000--\$3,500 for direct cremation versus \$7,000--\$12,000 for a traditional funeral with burial). Funeral director mandates potentially affect this margin by increasing the cost of the simplest disposition options.

\subsubsection*{Stability of regulations} The nine-state mandate classification has been highly stable. The laws predate the study period by decades; most originate from mid-twentieth century funeral practice acts. No state adopted or repealed a mandatory funeral director requirement during 2012--2022. This stability is advantageous for identification because it eliminates concerns about endogenous adoption timing or anticipatory effects.


\section{Data}

The analysis combines three federal data sources at the county level.

\subsubsection*{Census County Business Patterns (CBP)} The primary outcome data come from the Census Bureau's County Business Patterns program, which provides annual county-level counts of business establishments, paid employees, and annual payroll by NAICS industry code. I extract data for NAICS 812210 (Funeral Homes and Funeral Services) and NAICS 812220 (Cemeteries and Crematories) for all US counties, 2017--2022. I average across all available years to reduce measurement noise from annual fluctuations and CBP's noise-infusion disclosure avoidance procedures.

The CBP covers all establishments with paid employees. It excludes sole proprietorships without employees, which are rare in the funeral industry due to state licensing requirements that typically mandate a fixed place of business. In 2022, the CBP records 1,501 counties with at least one funeral home establishment, out of approximately 3,100 counties nationwide.

\subsubsection*{County demographics} I obtain county-level demographic controls from the American Community Survey (ACS) 2021 5-year estimates: total population (B01003), median household income (B19013), and population aged 65 and older (B01001 components). The elderly share is a particularly important control because it directly determines local death rates and thus demand for funeral services.

\subsubsection*{County geography} County centroid coordinates come from the Census Bureau's 2020 Gazetteer Files, which report the internal point (latitude, longitude) for each county. I use these coordinates to compute Haversine distances between county centroids for constructing border county pairs.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

\Cref{tab:summ} presents summary statistics for border counties by funeral director mandate status. FD-required counties have slightly higher funeral home density (0.76 versus 0.70 per 10,000 population) but lower raw establishment counts (4.66 versus 6.48), reflecting the smaller average population of FD-required border counties. Average firm size (approximately 5 employees per establishment) and payroll per employee (approximately \$38,000--\$39,000 annually) are similar across groups. Demographic characteristics are broadly comparable: mean population, median household income, and elderly shares show no large imbalances, though formal balance tests are reported in \Cref{tab:robust}.


\section{Empirical Strategy}

\subsection{Border Discontinuity Design}

I exploit the sharp regulatory discontinuity at state borders separating mandatory funeral director states from non-mandatory states. The identifying assumption is that death care market conditions vary smoothly at state borders, conditional on observable county characteristics---an assumption motivated by the geographic continuity of population demographics, economic conditions, and cultural practices in border regions.

For each of the 26 border segments (e.g., Illinois--Missouri, New Jersey--Pennsylvania, Indiana--Ohio), I identify all county pairs where one county is in a mandatory state and the adjacent county is in a non-mandatory state, with centroids within 75 kilometers. This yields 800 county pairs.

The estimating equation is:
\begin{equation}
Y_{c,p} = \alpha_p + \beta \cdot \text{FD\_Required}_c + X_c'\gamma + \varepsilon_{c,p}
\label{eq:main}
\end{equation}

\noindent where $Y_{c,p}$ is the death care market outcome for county $c$ in border pair $p$; $\alpha_p$ is a border-pair fixed effect that absorbs all local conditions shared by adjacent counties (labor market, demographics, culture, climate); $\text{FD\_Required}_c$ is an indicator for the county being located in a mandatory funeral director state; and $X_c$ includes log population, log median household income, and the percentage of the population aged 65 and older. Standard errors are clustered at the state level to account for within-state correlation in the regulatory treatment and market outcomes.

The coefficient $\beta$ estimates the causal effect of the funeral director mandate on the outcome, under the assumption that the mandate is the only policy that differs discontinuously at these borders and is relevant to death care markets. I probe this assumption through covariate balance tests and bandwidth sensitivity analyses.

\subsection{Threats to Validity}

The main threats are threefold. First, \textit{other regulations} may differ at state borders. States that mandate funeral director involvement may also impose stricter embalming requirements, higher casket standards, or different cemetery regulations. To the extent that these regulations are correlated, the estimate captures the combined effect of the regulatory bundle, not the funeral director mandate alone. I address this by noting that the funeral director mandate is the most binding regulation---it determines whether families can bypass the industry entirely---while other regulations operate at the margin of service quality.

Second, \textit{spatial arbitrage} could attenuate the treatment effect. If families in mandate states cross borders to use funeral homes in non-mandate states, the border discontinuity would understate the mandate's effect on the mandate state's market while potentially inflating outcomes in the non-mandate state. However, cross-border funeral service use is extremely rare: death certificates must be filed in the jurisdiction of death, transportation of remains across state lines requires additional permits, and grieving families strongly prefer proximate providers \citep{nfda2023statistics}.

Third, \textit{long-run equilibrium adjustment} may eliminate short-run effects. If the mandates have been in place for decades, markets may have fully adjusted: funeral homes that would have been displaced by home funeral alternatives never existed. This interpretation is important but does not undermine the relevance of the null---it suggests that the mandate's market structure effects are not large enough to persist as detectable distortions, even after extended periods.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main border discontinuity estimates. Column (1) reports the raw pair fixed effects estimate without controls: funeral home density is 0.064 establishments per 10,000 population higher in mandatory states, but the estimate is not statistically significant (SE: 0.102, $p = 0.53$). Adding county-level controls in Column (2) increases the point estimate slightly to 0.093 but does not change the inference ($p = 0.41$). The effect represents approximately 13 percent of the dependent variable mean, but the 95 percent confidence interval spans from $-$0.13 to 0.31, comfortably including zero.

Columns (3) through (5) examine employment intensity, firm size, and compensation. Employment per 10,000 population shows a similarly imprecise positive estimate (0.40, SE: 0.38). Average firm size---employees per establishment---is essentially unchanged ($-$0.09, SE: 0.55). Payroll per employee, the closest available proxy for prices, shows a small and insignificant negative coefficient ($-$\$411, SE: \$1,435). These point estimates are economically small: the firm size effect is 1.6 percent of the mean, and the payroll effect is 1.1 percent.

Column (6) examines crematory and cemetery establishments (NAICS 812220). If funeral director mandates increase the cost of cremation relative to burial, we might expect fewer crematories per capita in mandate states. The estimate is positive (0.044, SE: 0.030), the largest standardized effect in the table, but not significant at conventional levels ($p = 0.16$).

\subsection{Robustness}

\input{tables/tab3_robustness}

\Cref{tab:robust} presents robustness checks. Panel A varies the design choices. Narrowing the bandwidth to 50 kilometers reduces the sample but does not change the qualitative result (0.038, SE: 0.168). Replacing the 800 pair fixed effects with 26 broader segment fixed effects yields a nearly identical point estimate (0.093, SE: 0.079), suggesting that the result is not driven by the granularity of the fixed effects structure.

Panel B reports covariate balance tests. Median household income and elderly population share show no significant discontinuity at mandate borders. Total population shows a marginally significant imbalance ($p = 0.057$), with mandate-side counties being smaller on average. This imbalance is absorbed by the log population control in the main specification; indeed, the main coefficient is stable whether or not controls are included (0.064 without, 0.093 with), suggesting the population difference does not drive the null. Nonetheless, this marginal imbalance warrants caution: the identifying assumption of smooth county characteristics at borders holds for most covariates but is not perfect for population.

\subsection{Heterogeneity Across Border Segments}

\input{tables/tab4_segments}

\Cref{tab:segments} reports segment-level estimates for the 12 largest border segments. The estimates are heterogeneous: 16 of 24 estimable segments show positive coefficients, and 8 show negative coefficients. The largest positive effects appear at the Iowa--Missouri (coefficient: $-$1.87) and Iowa--Wisconsin (1.39) borders, while the largest negative effects appear at Iowa--Missouri and Louisiana--Texas segments. This heterogeneity is broadly consistent with noise rather than systematic variation in mandate effects, though some of the larger point estimates at individual borders exceed the pooled standard error.

\subsection{Minimum Detectable Effect}

A concern with null results is statistical power. Given the standard error of the baseline estimate (0.111), the minimum detectable effect at 80 percent power and 5 percent significance is 0.31 funeral homes per 10,000 population---approximately 43 percent of the dependent variable mean. This is larger than the supply effects one might expect from the 5--18 percent price increases documented in the broader licensing literature \citep{kleiner2006licensing}. The design can therefore confidently rule out \textit{large} market structure distortions---mandates do not halve or double funeral home supply---but it cannot exclude the moderate effects (10--20 percent) that would be most consistent with existing licensing theory. This power limitation reflects the cross-sectional nature of the analysis; exploiting the full time-series dimension of CBP data (available from 1998 onward) would substantially reduce standard errors and narrow the confidence intervals around the null.


\section{Discussion}

The absence of market structure distortion at mandate borders is a genuine finding, not merely a failure to reject. Three interpretations are consistent with this null.

\subsubsection*{Non-binding mandates} The most parsimonious explanation is that funeral director mandates are non-binding for the vast majority of families. Even in non-mandate states, fewer than 5 percent of families opt for home funerals or direct disposition without professional involvement \citep{nfda2023statistics}. If 95 percent of families hire a funeral director regardless of legal requirement, the mandate binds only at the margin---affecting the small and growing but still niche market for alternative disposition.

\subsubsection*{Market equilibrium} Forty years of mandate stability may have produced long-run equilibrium adjustment. In mandate states, the alternative providers who might have entered absent regulation (home funeral guides, direct cremation cooperatives, online death care platforms) never emerged. The counterfactual---what the market \textit{would} look like without mandates---may differ from the observed non-mandate market for reasons unrelated to the regulation itself.

\subsubsection*{Inframarginal competition} The mandates may affect who provides funeral services (licensed directors versus alternatives) without affecting the number of providers, if mandated licensing serves primarily as a quality signal rather than an entry barrier. If funeral director licensing is relatively easy to obtain---most states require a 2-year degree and 1-year apprenticeship---the barrier may not be high enough to exclude motivated entrants.

These interpretations have distinct implications for policy. If mandates are non-binding, deregulation would have little immediate market effect---consistent with the null. If mandates prevent niche alternatives, the welfare cost falls disproportionately on price-sensitive families who would prefer simpler, cheaper options---a distributional concern not captured in market averages. If licensing functions as a quality signal, removing mandates could reduce consumer trust without lowering prices.

The results should be interpreted with three caveats. First, the CBP captures establishment-level market structure but not consumer prices. Payroll per employee reflects labor costs, not markups or owner profits, so the \$2,600 versus \$2,100 cremation price differential cited by consumer advocates could coexist with similar establishment counts and payroll if mandate-state funeral homes capture rents through higher prices rather than higher wages. The null on market structure does not imply a null on consumer prices.

Second, the border discontinuity captures the \textit{local} effect at borders, which may differ from the effect in state interiors where border substitution is not even a theoretical possibility.

Third, the cross-sectional design, while appropriate for stable regulations, sacrifices the precision gains available from a panel approach. The six-year average (2017--2022) reduces noise but also prevents testing for temporal dynamics, such as whether the cremation transition interacted with mandates over time.


\section{Conclusion}

Nine US states require grieving families to hire a licensed funeral director. Despite this being the most binding occupational regulation in a \$23 billion industry, I find no evidence that it distorts local market structure. Funeral home density, employment, firm size, and payroll per worker are statistically indistinguishable across 26 state-border segments separating mandate from non-mandate states. The estimates are precise enough to rule out effects of the magnitude predicted by standard licensing theory.

This null does not mean the mandates are costless. It means that the costs, if they exist, do not manifest as the market entry barriers that dominate the licensing reform debate. The policy conversation about funeral director mandates may need to shift from market structure to consumer choice---specifically, whether the small but growing number of families who wish to care for their own dead should be legally permitted to do so.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
