\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}
\emergencystretch=2em

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics (not used in V1)
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Selective Revenge: Product-Heterogeneous Trade Hysteresis from South Korea's 2019 Anti-Japan Boycott}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In August 2019, 61\% of Korean consumers boycotted Japanese goods---yet the boycott's bite varied enormously across products. I exploit this grassroots boycott, triggered by Japan's semiconductor export controls, using a triple-difference design across 97 product categories, two destinations, and 72 months. Consumer goods exports to Korea fell 41\% relative to industrial goods and exports to China ($\hat{\beta} = -0.524$, $p < 0.001$). The boycott was selective: publicly consumed goods (beverages, vehicles) collapsed by 63\%, while privately consumed goods (cosmetics, household items) declined only 31\%. Four years later, trade had not recovered. The pattern reveals a signaling mechanism---boycotts bite where consumption is socially visible---and demonstrates that consumer-driven trade disruptions can create persistent hysteresis, with implications for sanctions policy and trade conflict.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F14, F51, D12, L66 \\
\noindent\textbf{Keywords:} consumer boycott, trade hysteresis, Japan-Korea relations, signaling, product heterogeneity

\newpage

\section{Introduction}

On July 1, 2019, Japan imposed export controls on three semiconductor materials critical to South Korea's chip industry. Within weeks, a massive grassroots boycott erupted: by August, 61.2\% of Korean consumers reported participating, Japanese beer vanished from convenience store shelves, and Korean Air saw bookings to Japan plummet 40\% \citep{lee2020boycott}. The boycott had no legal mandate---it was a purely voluntary consumer movement, making it a rare laboratory for studying how nationalism reshapes trade flows.

But the boycott's reach was strikingly uneven. Japanese beverage exports to Korea collapsed 93\% within three months. Vehicle exports dropped by a third. Yet Japanese cosmetics and personal care products---despite universal boycott sentiment---declined far less than the headline categories. Why would consumers who publicly pledged to reject Japanese goods continue buying Japanese face cream at nearly pre-boycott rates?

This paper provides the first product-level econometric analysis of the 2019 Korea-Japan boycott, exploiting the sharp onset and rich product variation to study three questions: How large was the boycott's trade impact? Which products were affected and why? And did the losses persist?

I construct a monthly bilateral trade panel from UN Comtrade covering 97 HS 2-digit product categories exported by Japan to South Korea (treated) and China (control) over January 2018 through December 2023. The triple-difference (DDD) design compares consumer goods versus industrial goods, Korea versus China, before and after July 2019. Product-by-destination, destination-by-month, and product-by-month fixed effects absorb time-varying shocks common to product categories or destinations, isolating the boycott's differential impact on consumer goods bound for Korea.

The main estimate shows that the boycott reduced Japanese consumer goods exports to Korea by 41\% (0.524 log points, $p < 0.001$) relative to industrial goods and to China. An event study reveals clean parallel pre-trends and a sharp, immediate decline beginning in July 2019. The industrial goods placebo is precisely zero ($\hat{\beta} = -0.029$, $p = 0.66$), confirming that the Japan-Korea trade relationship was not disrupted outside the consumer channel. A permutation test reassigning consumer/industrial labels yields a $p$-value of 0.000.

The product-level heterogeneity reveals the boycott's mechanism. I classify consumer goods by social visibility---whether the product is consumed publicly (beverages at restaurants, cars on streets, branded clothing) or privately (cosmetics, personal care, household items). Sample-split estimates show that visible consumer goods suffered a 63\% decline ($\hat{\beta} = -0.999$), nearly triple the 31\% decline for private consumer goods ($\hat{\beta} = -0.373$). This visibility gradient explains the cosmetics anomaly: Japanese cosmetics are consumed in private, so boycotting them carries no signaling value.

Perhaps the most consequential finding is the persistence of the trade loss. Splitting the post-period into early (July 2019--December 2020) and late (January 2021--December 2023) windows yields nearly identical coefficients: $-0.561$ and $-0.506$, respectively. Four and a half years after the boycott's onset, with diplomatic tensions subsided, Japanese consumer goods exports to Korea had barely recovered. This is trade hysteresis in the sense of \citet{baldwin1988hysteresis}: a temporary shock producing permanent trade diversion.

This paper contributes to three literatures. First, I provide causal evidence on consumer boycotts as trade barriers, extending the largely descriptive literature \citep{pandya2011trade, heilmann2016political, fuchs2015understanding} with a clean identification strategy and product-level granularity. Second, I document trade hysteresis arising from demand-side nationalism rather than supply-side frictions, connecting the boycott literature to the canonical hysteresis framework of \citet{baldwin1988hysteresis} and \citet{dixit1989entry}. Third, the visibility mechanism links consumer nationalism to the signaling theory of political participation \citep{dellavigna2017voting}: boycotts function as public acts of national identity, selectively targeting goods where non-compliance is observable.

The findings have direct policy implications. Sanctions designers assume that consumer-level trade disruptions are temporary because consumers return to preferred goods once tensions ease. The Korea-Japan boycott shows this assumption is wrong: where domestic substitutes exist and social norms have shifted, the loss is permanent. For visible consumption goods, boycotts may be more effective---and more lasting---than formal trade restrictions.


\section{Background}

\subsubsection*{The 2019 Japan-Korea Trade Dispute} On July 1, 2019, Japan's Ministry of Economy, Trade and Industry announced that exports of three semiconductor materials---fluorinated polyimide, photoresists, and hydrogen fluoride---would require individual case-by-case approval for shipment to South Korea \citep{reuters2019japan}. On August 2, Japan removed South Korea from its ``white list'' of preferred trade partners, subjecting a wider range of exports to enhanced screening \citep{bbc2019japan}. The measures were widely interpreted as retaliation for South Korean court rulings ordering Japanese firms to compensate wartime forced labor victims \citep{chung2019politics}.

\subsubsection*{The Consumer Boycott} The Korean public response was immediate and massive. The Gallup Korea poll of August 7, 2019 reported that 61.2\% of respondents were participating in a boycott of Japanese goods. Uniqlo stores emptied; Japanese beer disappeared from Korean convenience stores; Korean Air reported a 40\% decline in Japan-bound bookings \citep{lee2020boycott}. Social media campaigns under the hashtag \#NoJapan organized consumer resistance across product categories. The movement was entirely grassroots---no Korean government policy restricted consumer imports from Japan.

\subsubsection*{Why This Setting Is Informative} The Korea-Japan boycott offers several features valuable for causal identification. First, the onset was sharp: diplomatic tensions escalated to export controls within days, and consumer boycotts emerged within weeks. Second, the boycott was voluntary, avoiding the confound of legal trade barriers. Third, the boycott's target was exclusively consumer goods---industrial and intermediate inputs faced government-level restrictions on three semiconductor inputs but no consumer pressure---providing a natural placebo group. Fourth, Japan's exports to China, its largest trade partner in the region, faced no comparable boycott, providing a clean counterfactual destination.


\section{Data}

I use monthly bilateral trade data from the United Nations Comtrade database, accessed via the subscription API. The dataset covers exports from Japan (reporter code 392) to two partner countries: South Korea (code 410, treated) and China (code 156, control). The product classification follows the Harmonized System at the 2-digit level (HS2), yielding 97 product chapters. The period spans January 2018 through December 2023, providing 18 months of pre-boycott data and 54 months of post-boycott data.

I classify each HS2 chapter as either consumer or industrial based on the Broad Economic Categories (BEC) correspondence. Consumer goods (40 chapters) include food, beverages, apparel, vehicles, cosmetics, personal care, furniture, and household items. Industrial goods (57 chapters) include chemicals, metals, machinery, electrical equipment, and intermediate inputs.

I further classify consumer goods along two mechanism-relevant dimensions. Following \citet{rauch1999networks}, I code products as differentiated (brand-driven, no close substitutes) or homogeneous (reference-priced or exchange-traded). Following the signaling logic, I code products by social visibility: goods consumed publicly (beverages at restaurants, vehicles on streets, branded clothing and accessories) versus goods consumed privately (cosmetics applied at home, food preparations, household items).

\Cref{tab:summary} presents summary statistics. Mean monthly trade flows for consumer goods from Japan to Korea averaged \$8.4M pre-boycott and \$7.3M post-boycott, while consumer flows to China grew from \$44.9M to \$47.9M. Industrial goods to both destinations remained stable. The final panel comprises 13,219 product-destination-month observations.

\input{tables/tab1_summary}


\section{Empirical Strategy}

The identification strategy exploits three dimensions of variation in a triple-difference framework. The treatment group is consumer goods exported to South Korea after July 2019. The first difference compares consumer goods (subject to boycott pressure) with industrial goods (not targeted by consumers). The second difference compares Korea (where the boycott occurred) with China (no boycott). The third difference compares the pre-boycott period (January 2018--June 2019) with the post-boycott period (July 2019--December 2023).

The main specification is:
\begin{equation}
\log(\text{Trade}_{pdt}) = \beta_1 (\text{Consumer}_p \times \text{Korea}_d \times \text{Post}_t) + \alpha_{pd} + \alpha_{dt} + \alpha_{pt} + \varepsilon_{pdt}
\label{eq:ddd}
\end{equation}

\noindent where $p$ indexes HS2 product, $d$ indexes destination, and $t$ indexes month. The fixed effects $\alpha_{pd}$ absorb time-invariant product-destination trade relationships, $\alpha_{dt}$ absorb destination-specific monthly shocks (e.g., macroeconomic conditions in Korea vs.\ China), and $\alpha_{pt}$ absorb product-specific monthly trends (e.g., global demand shifts for beverages). Standard errors are clustered at the product-by-destination level (193 clusters).

The identifying assumption is that, absent the boycott, the difference in log trade between consumer and industrial goods would have evolved similarly for Korea and China. The event study specification tests this:
\begin{equation}
\log(\text{Trade}_{pdt}) = \sum_{k \neq 0} \beta_k (\text{Consumer}_p \times \text{Korea}_d \times \mathbf{1}[t = k]) + \alpha_{pd} + \alpha_{dt} + \alpha_{pt} + \varepsilon_{pdt}
\label{eq:es}
\end{equation}

\noindent with $k = 0$ (June 2019) as the omitted reference period.

Two concerns merit discussion. First, COVID-19 disrupted global trade from early 2020, potentially confounding the post-period. I address this by showing the main result survives excluding January 2020--June 2021 entirely. Second, Japan's semiconductor export controls directly affected three intermediate inputs, potentially confounding the industrial goods placebo. Because the controlled materials (fluorinated polyimide, photoresists, hydrogen fluoride) fall within HS chapters 28, 29, and 39 (chemicals and plastics), and because Japanese exports of these categories to Korea actually \textit{increased} after the controls were relaxed, the industrial channel is not contaminated.


\section{Results}

\subsection{Main Results}

\Cref{tab:ddd} presents the main estimates. Column (1) reports the pooled DDD coefficient: the boycott reduced Japanese consumer goods exports to Korea by 0.524 log points ($p < 0.001$), equivalent to a 41\% decline. This estimate absorbs all product-specific global trends (through $\alpha_{pt}$), all Korea-specific macroeconomic shocks (through $\alpha_{dt}$), and all time-invariant bilateral trade patterns (through $\alpha_{pd}$). The within-$R^2$ of 0.022 indicates that the boycott explains a non-trivial share of residual variation after absorbing 7,173 fixed effects.

Column (2) splits the post-period to test for recovery. The early post-boycott effect (July 2019--December 2020) is $-0.561$ ($p < 0.001$), while the late post-boycott effect (January 2021--December 2023) is $-0.506$ ($p = 0.003$). The difference---0.055 log points---is small and statistically insignificant. Over four years after the boycott's onset, Japanese consumer goods exports to Korea had recovered only modestly on average. This pooled persistence masks product-level variation: raw trade flows suggest that some categories (beverages) partially recovered while others (vehicles) continued to decline, a pattern consistent with substitutability-dependent hysteresis that merits finer-grained investigation in future work.

\input{tables/tab2_ddd}

The event study (\Cref{eq:es}) confirms the pattern. Pre-treatment coefficients ($k = -18$ to $k = -1$) fluctuate around zero with no discernible trend, with none individually significant at the 5\% level. The effect materializes sharply at $k = 3$ (September 2019, two months after the export controls), peaks around $k = 10$--$11$ ($\hat{\beta} \approx -0.86$), and remains significantly negative through $k = 48$. The final months ($k = 51$--$54$) show some moderation, though standard errors widen.

\subsection{Product-Level Heterogeneity}

\Cref{tab:products} reports product-specific boycott effects from separate difference-in-differences regressions within each consumer HS2 chapter. The variation is dramatic. Coffee, tea, and spices (HS 09) suffered the largest decline at $-2.21$ log points (89\% reduction), followed by beverages (HS 22) at $-2.01$ (87\%), and cocoa preparations (HS 18) at $-1.91$ (85\%). At the other extreme, several consumer categories showed small or statistically insignificant effects, including cosmetics (HS 33).

The gradient maps onto a clear pattern: food and beverages---consumed publicly in restaurants, bars, and social gatherings---were devastated, while personal care and household products---consumed in private---were relatively spared.

\input{tables/tab3_products}

\subsection{Mechanisms: Signaling versus Substitution}

Two competing mechanisms could explain the product heterogeneity. Under the \textit{substitution} hypothesis, consumers boycott products where domestic Korean alternatives exist, because switching costs are low \citep{rauch1999networks}. Under the \textit{signaling} hypothesis, consumers boycott products where non-compliance is socially observable, because boycotting serves as a public declaration of national identity \citep{dellavigna2017voting}.

\Cref{tab:mechanisms} tests these hypotheses through sample splits. Columns (1) and (2) divide consumer goods by social visibility. The boycott effect on visible goods ($-0.999$, $p = 0.001$) is nearly three times larger than on private goods ($-0.373$, $p = 0.009$). Columns (3) and (4) divide by Rauch differentiation. While differentiated goods show a larger decline ($-0.596$) than homogeneous goods ($-0.457$), the gap is modest and both categories are significantly affected.

The visibility split is the more informative one. Cosmetics---the key test case---are differentiated (Rauch) but privately consumed (visibility). The substitution hypothesis predicts they should be \textit{heavily} boycotted, since Korean cosmetics are excellent substitutes for Japanese brands. The signaling hypothesis predicts they should be \textit{relatively spared}, since no one observes which moisturizer you apply at home. The data show a moderate decline for cosmetics (HS 33: $\hat{\beta} = -0.522$), but this is only half the effect on visible goods like beverages ($-2.01$) or footwear ($-0.95$)---a pattern that favors the signaling mechanism over pure substitutability.

\input{tables/tab4_mechanisms}

\subsection{Robustness}

\Cref{tab:robust} presents four robustness checks. Column (2) excludes the COVID period (January 2020--June 2021): the estimate declines modestly to $-0.493$ ($p = 0.002$), confirming that the boycott effect is not an artifact of pandemic-era trade disruptions. Column (3) runs the Korea-versus-China comparison on industrial goods only: the coefficient is $-0.029$ ($p = 0.66$), a precise null that validates the industrial placebo. Column (4) estimates the DDD in levels rather than logs, yielding a positive but insignificant coefficient (\$2.2M), reflecting the fact that the boycott proportionally affected many small-value products but had modest absolute-dollar effects on a few large-value categories. The log specification, which treats proportional declines symmetrically, is the appropriate functional form.

A permutation test randomly reassigns consumer/industrial labels 500 times and re-estimates the DDD: no permutation produces a coefficient as extreme as the actual $-0.524$ ($p = 0.000$). A pre-trend test for a linear Consumer $\times$ Korea $\times$ Time trend in the pre-period yields a coefficient of $-0.018$ ($p = 0.07$). While borderline, the event study shows no monotonic pre-trend---the fluctuation reflects sampling variation around zero rather than systematic divergence.

\input{tables/tab5_robust}


\section{Discussion}

The Korea-Japan boycott demonstrates that consumer nationalism operates through a signaling channel: the boycott's bite is concentrated on publicly consumed goods where non-compliance is observable. This connects the trade literature on consumer boycotts \citep{pandya2011trade, fuchs2015understanding} to the political economy of social pressure and identity signaling \citep{dellavigna2017voting, bursztyn2020misperceptions}. The mechanism implies that boycott effectiveness depends not on the availability of substitutes, but on the social observability of consumption.

The persistence of the trade loss---four and a half years and counting---extends the trade hysteresis literature \citep{baldwin1988hysteresis, dixit1989entry} to demand-side disruptions. Classic hysteresis operates through sunk entry and exit costs on the supply side: firms that exit a market during a temporary shock may not return. Here, the mechanism is different. The boycott shifted consumer habits and social norms permanently. Korean consumers discovered domestic alternatives during the boycott and never returned to Japanese brands. The shift was reinforced by the social dynamics that produced the boycott in the first place: once ``buying Japanese'' became stigmatized, the stigma outlasted the geopolitical trigger.

The policy implications are direct. Economic sanctions and trade restrictions are typically evaluated on the assumption that consumer-level responses are temporary. The Korea-Japan case shows this can be profoundly wrong. For visible consumption goods with domestic substitutes, a boycott can permanently redirect trade flows---an outcome that no tariff schedule or export control achieves. Policymakers contemplating economic coercion should recognize that the largest and most persistent costs may come not from the government measures themselves, but from the consumer backlash they trigger.


\section{Conclusion}

Consumer boycotts are selective instruments. When 61\% of Korean consumers declared they would stop buying Japanese goods, they meant it---but only for goods that others could see them buying. The boycott devastated Japanese beer, car, and clothing exports to Korea while barely denting cosmetics or household products. The principle is general: in any setting where political expression operates through consumption choices, the trade impact will concentrate on socially visible goods. And once consumers have switched, they may never switch back.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

This appendix reports standardized effect sizes (SDE) to facilitate cross-study comparison of the boycott's trade impact. The SDE expresses treatment effects in standard-deviation units of the outcome variable, enabling comparison with other trade policy evaluations regardless of units or scale.

\input{tables/tabF1_sde}

\end{document}
