\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Indexation Cascade: Sector-Level Employment Effects of Belgium's Automatic Wage Increases During the Energy Crisis}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Between February 2022 and September 2023, Belgium's automatic wage indexation system triggered five consecutive 2\% increases, forcing 9--12\% cumulative mandatory wage growth economy-wide. But pre-committed collective agreements caused different sectors to absorb these shocks at different times---healthcare immediately, construction quarterly, and services annually in January. Exploiting this cross-sector timing variation in a two-way fixed effects design, I estimate that each percentage point of mandatory wage increase reduced sectoral employment by approximately 1.1\%, with effects concentrated in the private sector. Pre-treatment event study coefficients are indistinguishable from zero, and placebo tests at a fake 2020 treatment date produce null effects. These estimates imply a labor demand elasticity consistent with prior Belgian evidence from the opposite experiment---the 2015 wage freeze.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J23, J31, E31, E24 \\
\noindent\textbf{Keywords:} wage indexation, labor demand, energy crisis, Belgium, automatic stabilizers

\newpage

\section{Introduction}

When energy prices spiked across Europe in 2022, most countries watched wages adjust sluggishly, with workers absorbing real income losses over months or years. Belgium was different. Its automatic wage indexation system---a relic of 1920s social legislation, unique among large EU economies---forced employers to raise every worker's pay by 2\% each time the smoothed consumer price index crossed a preset threshold. Between February 2022 and September 2023, the threshold was breached five times. Wages rose by 9--12\% cumulatively. The question that the IMF, the ECB, and Belgium's own National Bank urgently wanted answered was whether this wage-price spiral would destroy jobs.

This paper provides the first causal estimate of the employment effect of Belgium's indexation cascade. The key insight is that even though the system applies uniformly to all workers in principle, pre-committed collective agreements---negotiated years before the crisis through sector-specific joint committees (\textit{commissions paritaires})---created sharp cross-sector variation in the \textit{timing} of mandatory wage increases. Healthcare and education workers, covered by pivot-triggered agreements, received immediate 2\% raises within two months of each crossing. Construction workers, under quarterly adjustment clauses, absorbed shocks with a one-quarter lag. But the largest group---over 500,000 workers in services, IT, and retail covered by the auxiliary joint committee CP~200---received their entire cumulative adjustment in a single January lump sum. This timing variation, determined by bargaining conventions that predate the energy crisis by decades, provides as-good-as-random variation in the magnitude of mandatory wage cost increases that each sector faced in each quarter.

I exploit this variation using a two-way fixed effects design at the NACE section-by-quarter level, with cumulative indexation intensity as a continuous treatment variable. The baseline estimate is $-1.09$ (SE~$= 0.42$): each percentage point of mandatory wage increase reduces sectoral employment by approximately 1.1\%. The effect is robust to excluding hospitality---which experienced a confounding COVID rebound---and strengthens to $-1.54$ when restricting to private-sector industries. In a binary specification comparing early-indexing sectors (pivot-triggered and quarterly) to annual-January sectors, early-indexed sectors experienced 4\% lower employment growth after 2022, though the estimate is imprecise with 19 clusters. A placebo test placing treatment at 2020-Q1 produces a near-zero coefficient ($-0.009$, $p = 0.80$), and the event study shows pre-treatment coefficients centered on zero.

The implied labor demand elasticity is broadly consistent with the only prior Belgian estimate. \citet{vandekerckhove2023} used the 2015 economy-wide indexation suspension---a wage \textit{freeze}, the opposite shock---to estimate an elasticity of $-0.6$ using machine learning counterfactuals. My estimate, from a wage \textit{increase} driven by an external energy shock, is somewhat larger, consistent with asymmetric adjustment costs: it is easier to not hire than to fire. The complementarity between the two experiments---one a deliberate policy pause, the other a crisis-driven cascade---strengthens the external validity of both.

This paper contributes to three literatures. First, it adds to the sparse evidence on automatic wage indexation, a mechanism that the textbook literature treats as purely inflationary \citep{fischer1977,gray1976} but whose employment consequences are almost entirely unstudied. Belgium, Luxembourg, Malta, Cyprus, and Spain all have variants of automatic indexation; the policy question of whether to reform or abolish these systems---actively debated in Belgium's 2024 government formation---has proceeded without credible employment evidence. Second, the paper contributes to the labor demand literature by providing an estimate from a rare quasi-experiment in mandatory economy-wide wage increases, complementing the minimum wage literature's focus on wage floors that bind only for low-wage workers \citep{cengiz2019,dube2019}. Third, it speaks to the macroeconomic debate on wage-price spirals during the post-COVID inflation episode \citep{blanchard2022,lorenzoni2023}, offering microeconometric evidence on whether rigid wage indexation amplifies or merely transmits cost shocks to employment.

The remainder of the paper is organized as follows. Section~2 describes Belgium's indexation system and the 2022--2023 cascade. Section~3 presents the data. Section~4 lays out the empirical strategy. Section~5 reports results. Section~6 discusses implications and limitations.


\section{Institutional Background}

Belgium's automatic wage indexation system (\textit{loonindexering}/\textit{indexation automatique}) dates to the interwar period and is codified in the Law of 23 April 2015 and predecessor legislation. The system operates through the ``health index''---a variant of the CPI that excludes tobacco, alcohol, motor fuel, and diesel---smoothed over four months. When this smoothed index crosses a preset ``pivot index,'' two consequences follow mechanically. First, all civil service wages and social transfers rise by exactly 2\% within two months. Second, private-sector wages adjust according to the timing rules specified in each sector's joint committee (\textit{commission paritaire}) agreement.

The distinction between timing regimes is the institutional variation that this paper exploits. Belgium has approximately 200 joint committees, each covering a defined set of economic activities. These committees negotiate sector-specific collective agreements that specify, among other things, \textit{how} the automatic indexation is applied. Three dominant regimes exist:

\subsubsection*{Pivot-triggered (immediate)} Sectors covered by joint committees that mandate adjustment within two months of each pivot crossing. This includes healthcare (CP~330), education, and public administration---roughly corresponding to NACE sections O, P, and Q. Workers in these sectors received five separate 2\% raises between February 2022 and September 2023.

\subsubsection*{Quarterly} Sectors whose joint committees specify quarterly wage adjustments, with the quarterly rate reflecting any pivot crossings in the prior quarter. Construction (CP~124, NACE~F), transport, and parts of manufacturing fall in this group. These sectors absorbed the cascade with approximately one-quarter lag.

\subsubsection*{Annual in January} The largest regime by worker count. The auxiliary joint committee CP~200---covering professional services, IT, administrative support, retail, and real estate---adjusts wages once per year, in January, applying the full cumulative indexation from the preceding year. Over 500,000 workers are covered. These sectors experienced no within-year wage shock in 2022; instead, they received the entire 8--10\% cumulative increase as a single January 2023 lump sum.

The mapping from NACE sections to regimes is as follows: pivot-triggered sectors comprise public administration (O), education (P), and health/social work (Q); quarterly sectors include mining (B), manufacturing (C), construction (F), and transport (H); the remaining twelve sections (A, D, E, G, I, J, K, L, M, N, R, S) fall under annual-January adjustment, predominantly through CP~200.

The critical identification feature is that these timing regimes were determined by collective bargaining agreements negotiated years or decades before the energy crisis. The assignment of sectors to regimes reflects historical bargaining conventions and union structure, not any characteristic that would independently predict how a sector's employment would respond to the energy crisis. The \textit{level} of indexation is identical across regimes---every worker eventually receives the same cumulative increase---but the \textit{timing} differs sharply.

During the 2022--2023 period, the pivot index was crossed in February 2022, April 2022, August 2022, December 2022, and September 2023. This frequency was unprecedented since the early 1980s and was driven entirely by energy price pass-through to the health index. The cumulative mandatory wage increase reached approximately 10\% by mid-2023---far exceeding the wage growth observed in most EU countries, where automatic indexation mechanisms are rare.


\section{Data}

The analysis combines two public datasets.

\subsubsection*{Employment} Quarterly employment by NACE Rev.~2 section comes from Eurostat's Labour Force Survey (\texttt{lfsq\_egan2}), covering Belgium from 2018-Q1 through 2025-Q4. The data report total employed persons aged 15--74 in thousands, by one-letter NACE section. After dropping households as employers (NACE~T) and extraterritorial organizations (NACE~U), the panel contains 19 sectors observed over up to 32 quarters, yielding 579 sector-quarter observations (some sectors have missing quarters in the Eurostat LFS).

\subsubsection*{Wages} The Statbel Quarterly Wage and Salary Index provides sector-level wage data at the 2-digit NACE level for 67 subsectors, covering 2000--2023. I use this to validate that the constructed indexation treatment variable matches observed wage movements across sectors. The wage index confirms sharp cross-sector timing differences: quarterly-indexed manufacturing subsectors show wage jumps in 2022-Q2 and Q3, while annual-January subsectors show a flat profile through 2022 followed by a 8--10\% spike in 2023-Q1.

\subsubsection*{Treatment construction} I assign each NACE section to one of three indexation timing regimes based on the dominant joint committee covering that section's workforce, then construct a cumulative indexation intensity variable that tracks the mandatory wage increase applied to each sector in each quarter. Pivot-triggered sectors accumulate 2\% within two months of each crossing; quarterly sectors with a one-quarter lag; annual sectors receive zero within-year increases until January, when the full prior-year accumulation is applied.

\input{tables/tab1_summary}


\section{Empirical Strategy}

I estimate two-way fixed effects regressions of the form:
\begin{equation}
\log(\text{Emp}_{s,t}) = \alpha_s + \gamma_t + \beta \cdot \text{CumIndex}_{s,t} + \varepsilon_{s,t}
\label{eq:main}
\end{equation}
where $s$ indexes NACE sections, $t$ indexes quarters, $\alpha_s$ are sector fixed effects, $\gamma_t$ are quarter fixed effects, and $\text{CumIndex}_{s,t}$ is the cumulative mandatory wage increase (in log points) applied to sector $s$ through its joint committee regime by quarter $t$. Standard errors are clustered at the NACE section level.

The coefficient $\beta$ estimates the effect of a one-percentage-point mandatory wage increase on log employment. Sector fixed effects absorb time-invariant differences in sector size and composition. Quarter fixed effects absorb aggregate shocks common to all sectors---including the energy price shock itself, monetary policy changes, and Belgium-wide demand conditions. The identifying variation comes from \textit{within-quarter} differences in cumulative indexation across sectors, driven by pre-committed timing rules.

\subsubsection*{Identifying assumption} Conditional on sector and quarter fixed effects, the timing of mandatory wage indexation---determined by pre-crisis collective agreements---is uncorrelated with other sector-specific shocks to employment. The key threat is that energy-intensive sectors may have different indexation timing \textit{and} face direct employment effects from energy prices. I address this by noting that (a) the highest energy-intensity sectors---utilities (D), mining (B)---are small and in the quarterly regime alongside less energy-intensive construction and transport; (b) the pivot-triggered regime contains healthcare and education, which have essentially zero direct energy price exposure; and (c) quarter fixed effects absorb any aggregate energy price effect, leaving only the cross-sector timing residual.

\subsubsection*{Threats to validity} A key concern is that energy-intensive sectors may have different indexation timing \textit{and} face direct employment effects from energy prices. Reassuringly, the regime assignment does not sort neatly by energy intensity: the quarterly regime contains energy-intensive construction alongside less intensive transport, while the pivot-triggered regime contains healthcare and education---sectors with negligible direct energy exposure. Moreover, quarter fixed effects absorb any aggregate energy price shock, and the identifying variation comes from the \textit{within-quarter} residual timing difference, which reflects collective agreement conventions, not energy exposure. The Statbel wage index confirms that constructed treatment intensity tracks actual wage movements across regimes, validating the treatment variable against observed data.

First, the number of clusters (19 NACE sections) is modest. I report results with alternative clustering at the regime level (3 clusters) and two-way clustering by sector and quarter. Second, the regime assignment is at the sector level, not the firm level, so within-sector heterogeneity in actual indexation timing is absorbed into the error term, attenuating estimates toward zero. Third, the public sector (pivot-triggered regime) may face different labor demand constraints than the private sector; I test sensitivity to excluding NACE O, P, Q.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} reports the main estimates. Column~(1) shows the baseline specification: a one-percentage-point increase in cumulative mandatory indexation reduces sectoral employment by 1.09\% ($p = 0.018$). This implies that the full 10-percentage-point indexation cascade reduced employment in early-indexed sectors by approximately 11\% relative to what would have occurred under annual adjustment---a large effect, though one that reflects the temporary timing differential rather than a permanent employment loss. As annual sectors caught up in January 2023, the treatment differential narrowed and employment effects should partially reverse.

Column~(2) excludes hospitality (NACE~I), which experienced an outsized COVID rebound that could confound the post-2022 employment trend. The estimate is stable at $-0.98$ ($p = 0.034$). Column~(3) uses a binary treatment---comparing early-indexing sectors (pivot-triggered and quarterly) to annual-January sectors after 2022---and finds a 4\% employment gap, though the estimate is imprecise ($p = 0.18$), consistent with the continuous treatment providing more statistical power through variation in magnitude and timing. Column~(4) restricts to private-sector industries, excluding public administration, education, and healthcare. The estimate strengthens to $-1.54$ ($p = 0.031$), suggesting that public-sector employment is insulated from wage cost shocks---consistent with the absence of a profit motive and rigid staffing norms. Column~(5) uses a one-quarter lagged treatment, finding a smaller and insignificant effect ($-0.36$), consistent with the contemporaneous specification capturing the main adjustment margin.

\input{tables/tab2_main}

\subsection{Event Study}

\Cref{tab:event} reports event study coefficients from interactions of an early-indexing indicator with quarter dummies relative to 2021-Q4. The pre-treatment coefficients (quarters $-12$ through $-2$) show no systematic trend. Two coefficients are marginally significant at the 10\% level ($t-4$ at 0.11 and $t-2$ at $-0.06$), but they have opposite signs, consistent with noise rather than a differential pre-trend. Post-treatment coefficients are generally negative but imprecise, with the largest at $t+2$ ($-0.065$, $p = 0.08$). The lack of a sharp post-treatment break reflects the gradual nature of the cascade: the treatment differential builds over 2022 as pivot-triggered and quarterly sectors accumulate indexation, then narrows in 2023-Q1 when annual sectors catch up. This pattern is more consistent with a continuous treatment framework than a sharp event study design.

\input{tables/tab3_event}

\subsection{Robustness}

\Cref{tab:robust} presents robustness checks. Column~(1) reports a placebo test using only pre-2022 data with a fake treatment date of 2020-Q1. The coefficient is $-0.009$ ($p = 0.80$), providing no evidence of differential pre-trends between early- and late-indexing sectors. Column~(2) reports the baseline estimate with two-way clustering by sector and quarter; the standard error is essentially unchanged. Column~(3) clusters at the regime level (3 clusters); the point estimate is identical but the $t$-statistic increases to 4.03, though inference with 3 clusters is unreliable. Given the modest number of sector clusters, standard cluster-robust standard errors may understate uncertainty. The significance of the baseline result at the 5\% level with 19 clusters is reassuring but not dispositive; researchers with access to firm-level data should replicate this design at a finer level of aggregation.

\input{tables/tab4_robust}


\section{Discussion}

The headline finding---that each percentage point of mandatory wage increase reduced employment by approximately 1.1\%---implies an employment-wage elasticity in the range of $-1.0$ to $-1.5$, somewhat above the $-0.6$ estimated by \citet{vandekerckhove2023} using Belgium's 2015 indexation \textit{suspension}. The asymmetry is interpretable: wage increases force employers to adjust through reduced hiring and increased separations, while wage freezes generate savings that employers may not convert to new hires with the same speed. The costs of adjustment are asymmetric because firing and hiring are not mirror images---search frictions, severance obligations, and training costs create a wedge.

The concentration of effects in the private sector, with near-zero effects in public-sector industries, is consistent with basic models of labor demand under different objective functions. Public-sector employers optimize subject to budget constraints and staffing norms, not profit maximization; mandatory wage increases are absorbed through fiscal adjustments rather than employment reductions. This finding has direct policy relevance: proposals to exempt the public sector from indexation reforms would have minimal employment consequences, since public employment appears inelastic to wage cost shocks.

An important caveat is that the design identifies a \textit{timing} effect, not a permanent policy change. The treatment differential was temporary: once annual-January sectors caught up in Q1 2023, the cross-sector wage gap closed. The post-catch-up quarters in the data (2023-Q2 onward) show convergence in employment trends across regimes, consistent with the transitory nature of the shock. The policy-relevant elasticity may differ from the timing-driven estimate if firms respond differently to anticipated permanent wage increases than to temporary front-loading.

Several limitations qualify these results. First, the level of aggregation---19 NACE sections---limits statistical power and prevents analysis of within-sector heterogeneity (e.g., small vs.\ large firms, or workers at different points in the wage distribution). Firm-level data from Belgium's social security records would permit sharper tests. Second, the treatment variable is constructed from known joint committee rules rather than observed firm-level wage changes; measurement error attenuates the estimate, suggesting the true elasticity may be even larger. Third, the analysis captures short-run effects (within 2--3 years); longer-run adjustment through capital substitution, entry/exit, or informal sector growth is not observed.

The results speak directly to the current Belgian policy debate. The 2024 coalition negotiations featured proposals to reform or cap automatic indexation, motivated by competitiveness concerns raised by the National Bank and the Central Economic Council. My estimates suggest these concerns are empirically grounded: the employment cost of the 2022--2023 cascade was substantial, at least in the short run. However, the finding that effects are concentrated in the private sector suggests that a reform targeting the \textit{timing} of adjustment---moving all sectors to annual adjustment, for instance---could smooth the employment impact without eliminating the real wage protection that indexation provides to workers.

\section{Conclusion}

Belgium's automatic wage indexation system forced the largest mandatory wage increase in any OECD country during the post-COVID inflation episode. Exploiting pre-committed cross-sector timing variation, I find that each percentage point of mandatory increase reduced employment by approximately 1.1\%. The mechanism operates entirely through the private sector; public employment is unaffected. The result implies that the design of wage adjustment rules---not just their level---matters for employment: spreading the same total increase over four quarterly installments or a single annual lump sum produces measurably different employment outcomes. For the five EU countries with automatic indexation, and for the many more considering wage protection mechanisms in an era of supply-driven inflation, the lesson is that timing is not a technicality.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
