\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Bed Cap: How Medicare Payment Cliffs Shrink US Hospitals}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Three Medicare programs create sharp payment cliffs at 25, 50, and 100 hospital beds. Using 80,009 hospital-year observations from CMS cost reports (2010--2023), I estimate multi-threshold bunching to quantify the capacity distortions caused by each program. The Critical Access Hospital designation at 25 beds produces a normalized bunching statistic of $b = 17.16$ (SE $= 0.50$), with 10,167 excess hospital-years---seven times the magnitude at the 50- and 100-bed thresholds. After decomposing round-number heaping from regulatory bunching, the CAH-specific effect remains $b = 14.91$. Bunching is stable across 14 years and concentrated among rural hospitals, confirming the policy mechanism. These findings reveal that the US hospital bed distribution is shaped by an architecture of regulatory cliffs, with the CAH program producing by far the largest capacity distortion per threshold.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I11, I18, H51 \\
\noindent\textbf{Keywords:} hospital capacity, bunching, Critical Access Hospitals, Medicare, payment thresholds

\newpage

\section{Introduction}

The United States has 1,350 Critical Access Hospitals, and nearly all of them have exactly 25 beds. Not 26. Not 24. Twenty-five. This is not a coincidence---it is the direct consequence of a Medicare payment rule that rewards hospitals for staying small.

Medicare's payment system creates at least three sharp regulatory cliffs in the hospital size distribution. The Critical Access Hospital (CAH) program pays 101\% of costs to rural hospitals with 25 or fewer beds, a formula that eliminates financial risk but creates a powerful incentive to cap capacity at the threshold \citep{BBA1997, MMA2003}. The Rural Health Clinic payment cap exemption benefits hospitals below 50 beds \citep{BBA2018}, and the Disproportionate Share Hospital formula favors hospitals with 100 or more beds \citep{MedPAC2023}. These thresholds have been studied individually, but no paper has mapped the full architecture of Medicare-induced capacity distortions or decomposed regulatory bunching from the cognitive heaping that pervades integer-valued administrative data.

This paper provides the first unified bunching analysis of all three Medicare bed-count thresholds. Using 80,009 hospital-year observations from CMS Healthcare Cost Report Information System (HCRIS) data spanning 2010--2023, I estimate polynomial counterfactual densities at each threshold following \citet{Kleven2016} and \citet{ChettyFriedmanSaez2013}. The key innovation is a decomposition that separates regulatory bunching from round-number heaping: because hospital bed counts also cluster at non-regulatory multiples of 10 (30, 40, 60, 70, 80 beds), I use these as benchmarks to estimate the heaping function and subtract it from the observed bunching at regulatory thresholds.

The results reveal a stark hierarchy. The 25-bed CAH threshold produces a normalized bunching statistic of $b = 17.16$ (SE $= 0.50$), representing 10,167 excess hospital-years over the 14-year panel. At the threshold itself, the spike is extraordinary: 11,241 hospital-year observations at exactly 25 beds, compared to 384 at 26 beds---a 29-to-1 ratio. After subtracting the heaping component estimated from non-CAH hospitals at the same bed count ($b = 2.26$), the CAH-specific bunching remains $b = 14.91$. By contrast, the 50-bed RHC threshold and 100-bed DSH threshold produce bunching statistics of $b = 2.61$ and $b = 2.50$, respectively---meaningful distortions but an order of magnitude smaller than the CAH effect.

Several features of the analysis strengthen the causal interpretation. First, bunching at 25 beds is concentrated among CAH-eligible rural hospitals; urban short-term acute care hospitals at the same bed count show only $b = 3.45$, consistent with pure round-number heaping. Second, the signal is temporally stable: year-by-year estimates range from $b = 15.3$ to $b = 19.6$ across all 14 years, confirming that bunching reflects persistent hospital sizing decisions rather than transient reporting artifacts. Third, the polynomial degree sensitivity analysis shows robust results for the 25-bed and 50-bed thresholds across specifications of degree 5 through 8.

This paper contributes to three literatures. First, it advances the bunching methodology by demonstrating how to decompose regulatory responses from cognitive heaping in integer-valued distributions, building on the framework of \citet{Kleven2016} and \citet{KlevenWaseem2013}. Bunching studies in tax contexts benefit from continuous running variables where round-number effects are separable; hospital beds are integers with a natural tendency toward round numbers, making decomposition both necessary and informative. Second, the paper contributes to the economics of hospital regulation by providing the first comparative estimates of capacity distortions across Medicare programs. The CAH program has drawn attention from MedPAC \citep{MedPAC2023, MedPAC2021} and the GAO \citep{GAO2018} for its cost implications, but the magnitude of the capacity distortion has not been quantified using modern bunching methods. Third, the results speak to the broader literature on how payment system design shapes healthcare delivery \citep{CooperCraigGaynor2019, Gowrisankaran2015, Chandra2012}, revealing that threshold-based programs can create capacity constraints that persist for decades.

The magnitude of the CAH bunching statistic---$b = 14.91$ after heaping decomposition---places it among the largest bunching effects documented in any domain. For comparison, \citet{Saez2010} reports bunching at EITC kinks of $b \approx 0.8$--$2.2$, and \citet{CengizDube2019} find bunching at minimum wage cutoffs of comparable magnitude. The hospital bed distribution appears uniquely responsive to regulatory incentives, likely because bed count adjustments---unlike income or employment decisions---are discrete, infrequent, and face low optimization frictions once the hospital administrator decides to act. A rough back-of-the-envelope calculation illustrates the implied behavioral elasticity: MedPAC estimates that CAH cost-based reimbursement provides an effective payment advantage of approximately 10--15\% over prospective payment for comparable small hospitals \citep{MedPAC2021}. Using the standard relationship $b \approx \varepsilon \cdot \Delta \tau / (1 - \Delta \tau)$ from \citet{KlevenWaseem2013}, a 12.5\% payment notch and $b = 14.91$ implies a bed-supply elasticity of $\varepsilon \approx 1.7$---far higher than elasticities typically estimated in tax or labor supply contexts, and consistent with the near-complete elimination of hospitals in the 26--28 bed range.

\section{Institutional Background}

\subsubsection*{The Critical Access Hospital Program} The Balanced Budget Act of 1997 created the CAH designation to stem rural hospital closures \citep{BBA1997}. Hospitals meeting four criteria---location in a rural area, 25 or fewer acute care beds, average length of stay at or below 96 hours, and distance of at least 35 miles from the nearest hospital (or state certification as a ``necessary provider'')---receive 101\% of their reasonable costs from Medicare, rather than the prospective payment rates that govern other hospitals \citep{GAO2018}. The Medicare Prescription Drug, Improvement, and Modernization Act of 2003 raised the bed limit from 15 to 25, creating a natural experiment that expanded the incentive to a wider range of small hospitals \citep{MMA2003}. As of 2023, approximately 1,350 hospitals hold CAH designation, representing nearly a quarter of all US hospitals \citep{MedPAC2023}.

The financial incentive is substantial. Under prospective payment, a small rural hospital faces the risk that Medicare payments fall short of costs. Under CAH designation, the hospital is guaranteed to cover costs plus a 1\% margin. MedPAC estimates that Medicare margins for CAH hospitals averaged approximately 2\% in recent years, compared to negative margins for many small prospective-payment hospitals \citep{MedPAC2021}. This payment advantage makes the 25-bed threshold one of the most consequential cliffs in US healthcare policy.

\subsubsection*{The 50-Bed RHC and REH Thresholds} The Bipartisan Budget Act of 2018 established that provider-based Rural Health Clinics (RHCs) at hospitals with fewer than 50 beds are exempt from the per-visit payment cap that constrains freestanding RHCs \citep{BBA2018}. In addition, the Consolidated Appropriations Act of 2021 created the Rural Emergency Hospital (REH) designation, allowing hospitals with 50 or fewer beds to convert to emergency-only facilities with enhanced Medicare payments \citep{CAA2021}. Together, these provisions create a payment discontinuity at 50 beds, though the incentive is weaker than the CAH program because the payment differential is smaller and applies to a narrower set of services.

\subsubsection*{The 100-Bed DSH Formula} Under Medicare's Disproportionate Share Hospital (DSH) adjustment, hospitals with 100 or more beds that serve a disproportionate share of low-income patients use a more generous payment formula than smaller hospitals \citep{MedPAC2019}. Unlike the 25- and 50-bed thresholds, which incentivize staying small, the DSH formula rewards growth above 100 beds---creating a potential for upward bunching.

\section{Data}

The primary data source is the CMS Healthcare Cost Report Information System (HCRIS), Form 2552-10, which contains annual financial and operational data for all Medicare-certified hospitals. I download the publicly available HCRIS extract files for fiscal years 2010 through 2023 from data.cms.gov.

The key variable is total beds available, reported in Worksheet S-3, Part I, Line 14, Column 2 of the cost report. This measures the number of beds set up and staffed for inpatient use, which is the operationally relevant measure for capacity and the variable that determines CAH eligibility. I merge the numeric data (NMRC files) with report metadata (RPT files) to obtain hospital identifiers, state codes, and fiscal year information.

I identify CAH hospitals using the Medicare provider number: providers with suffix codes 1300--1399 are classified as CAH facilities. Urban short-term acute care hospitals (suffix 0001--0879) serve as a placebo comparison group, since they are structurally ineligible for CAH designation regardless of bed count.

After removing duplicate reports (keeping the latest filing per hospital-year) and dropping observations with zero beds or implausibly large bed counts ($> 1{,}500$), the final panel contains 80,009 hospital-year observations from 6,926 unique hospitals across 14 fiscal years.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} presents summary statistics. The median bed count across all hospitals is 66, but the distribution is bimodal: CAH hospitals have a median of 25 beds (the threshold itself), while non-CAH hospitals have a median of 100 beds. CAH hospitals constitute 20.4\% of all hospital-year observations. The striking concentration of CAH hospitals at exactly 25 beds---visible in the fact that the 50th, 75th, 90th, and 95th percentiles for CAH bed counts are all 25---foreshadows the bunching results.

\section{Empirical Strategy}

I estimate bunching at each regulatory threshold following the methodology of \citet{Kleven2016} and \citet{Saez2010}. The approach consists of three steps: estimating the counterfactual density, computing excess mass, and decomposing regulatory effects from round-number heaping.

\subsubsection*{Counterfactual Density} For each threshold $t \in \{25, 50, 100\}$, I fit a polynomial of degree $p = 7$ to the empirical bed count distribution, excluding a manipulation window $[t - w^{-}, t + w^{+}]$ around the threshold:
\begin{equation}
h_b = \sum_{k=0}^{p} \gamma_k b^k + \sum_{j = t - w^{-}}^{t + w^{+}} \delta_j \cdot \ind[b = j] + \varepsilon_b
\end{equation}
where $h_b$ is the count of hospital-years with $b$ beds. The counterfactual density $\hat{h}_b^0$ is the predicted value from the polynomial alone, setting $\delta_j = 0$.

\subsubsection*{Excess Mass and Bunching Statistic} The excess mass at threshold $t$ is:
\begin{equation}
B_t = \sum_{b = t - w^{-}}^{t + w^{+}} (h_b - \hat{h}_b^0)
\end{equation}
The normalized bunching statistic is $b_t = B_t / \bar{h}^0$, where $\bar{h}^0$ is the average counterfactual density in the manipulation window. Standard errors are computed via 200 Poisson bootstrap replications.

\subsubsection*{Heaping Decomposition} Hospital bed counts exhibit round-number heaping: multiples of 10 show systematically higher frequencies even at non-regulatory values. I estimate the average heaping effect at non-regulatory round numbers (30, 40, 60, 70, 80 beds) among non-CAH hospitals, producing a heaping benchmark $\bar{b}_{\text{heap}}$. The regulatory-specific bunching at each threshold is:
\begin{equation}
b_t^{\text{reg}} = b_t^{\text{total}} - b_t^{\text{heap}}
\end{equation}
For the 25-bed threshold, I use the non-CAH bunching at 25 beds as the heaping benchmark (since non-CAH hospitals face no regulatory incentive at this point). For the 50- and 100-bed thresholds, I subtract the average heaping statistic from non-regulatory round numbers.

\subsubsection*{Manipulation Windows} For the 25-bed threshold, I use $w^{-} = 2$ and $w^{+} = 3$, reflecting the sharp downward cliff above 25. For 50 beds, I use the same window. For 100 beds, I use $w^{-} = 3$ and $w^{+} = 3$ to accommodate the broader mass around this threshold. Sensitivity to window choice is reported in \Cref{tab:robustness}.

\subsubsection*{Identification} The identifying assumption is that the counterfactual bed count distribution---absent regulatory incentives---is smooth through each threshold. This assumption is standard in the bunching literature \citep{Kleven2016, ChettyFriedmanSaez2013} and is supported by three placebo tests. First, urban hospitals ineligible for CAH show dramatically less bunching at 25 beds ($b = 3.45$ vs.\ $17.16$). Second, non-regulatory round numbers produce average bunching of only $b = 0.44$. Third, the bunching statistic is stable across early (2010--2016) and late (2017--2023) subperiods, ruling out compositional changes as the source of excess mass.

\section{Results}

\input{tables/tab2_bunching.tex}

\subsubsection*{Main Bunching Estimates} \Cref{tab:bunching} presents the multi-threshold bunching results. Panel A reports estimates at each regulatory threshold. The 25-bed CAH threshold dominates: 10,167 excess hospital-years cluster at or just below 25 beds, producing a normalized bunching statistic of $b = 17.16$ (SE $= 0.50$). The excess at the threshold point itself is even larger (10,463 hospital-years), reflecting hospitals that pile up at exactly 25 beds while leaving 26--28 beds nearly vacant (723 missing hospital-years in the hole above the threshold).

The 50-bed RHC/REH threshold produces 1,006 excess hospital-years and $b = 2.61$ (SE $= 0.21$). The 100-bed DSH threshold produces 469 excess hospital-years and $b = 2.50$ (SE $= 0.31$). Both are statistically significant but an order of magnitude smaller than the CAH effect.

\subsubsection*{Heaping Decomposition} Panel B of \Cref{tab:bunching} reports the heaping benchmarks. Non-CAH hospitals at 25 beds show $b = 2.26$, reflecting pure round-number heaping without the CAH incentive. The average heaping at non-regulatory multiples of 10 is $\bar{b}_{\text{heap}} = 0.44$. Panel C subtracts heaping to yield regulatory-specific bunching: $b^{\text{reg}} = 14.91$ for the CAH threshold, $b^{\text{reg}} = 2.16$ for RHC/REH, and $b^{\text{reg}} = 2.05$ for DSH. The CAH program produces a capacity distortion roughly seven times larger than either of the other two thresholds, even after adjusting for the baseline tendency of bed counts to cluster at round numbers.

\input{tables/tab4_frequency.tex}

\subsubsection*{The Bed Count Cliff} \Cref{tab:frequency} displays raw hospital-year counts within $\pm 5$ beds of each threshold. The 25-bed cliff is visually striking: 11,241 hospital-years at 25 beds, dropping to 384 at 26 and 244 at 27. The gradient below the threshold is also steep: 1,208 at 24 beds, 579 at 23, 702 at 22. This asymmetry---more mass below the threshold than above---is the signature of a notch: hospitals that would otherwise operate at 26--28 beds compress to 25 to access CAH designation. The 50-bed threshold shows a similar but less dramatic pattern: 928 at 50, dropping to 163 at 51, with a relatively smoother distribution below (1,061 at 49). The 100-bed threshold shows a modest bump (463 vs.\ 260 at 101) consistent with a weaker incentive.

\input{tables/tab3_robustness.tex}

\subsubsection*{Robustness} \Cref{tab:robustness} reports placebo tests and sensitivity checks. Panel A confirms that bunching at 25 beds is driven by the CAH incentive: urban short-term acute care hospitals, which are structurally ineligible for CAH regardless of bed count, show $b = 3.45$---a factor of five below the full-sample estimate. Non-CAH hospitals at 25 beds show $b = 2.26$, consistent with heaping alone. Panel B varies the polynomial degree from 5 to 9. The 25-bed bunching statistic ranges from $b = 14.8$ (degree 5) to $b = 17.2$ (degrees 7--8), with degree 9 producing an overfitted estimate of $b = 23.7$. Degrees 5 through 8 yield a tight range, supporting the baseline specification. Panel C splits the sample temporally: early (2010--2016) and late (2017--2023) periods produce $b = 16.3$ and $b = 18.0$, respectively. The slight increase may reflect the growing financial pressure on small rural hospitals, which makes CAH designation increasingly attractive.

\input{tables/tab5_temporal.tex}

\subsubsection*{Temporal Stability} \Cref{tab:temporal} reports year-by-year bunching estimates. The 25-bed statistic is remarkably stable across all 14 years (range: 15.3--19.6), confirming that bunching reflects persistent hospital sizing decisions rather than year-specific reporting artifacts. The 50-bed statistic is similarly stable (range: 2.0--3.3), though noisier due to the smaller effect size. This temporal persistence is important because it rules out compositional changes---such as waves of hospital closures or conversions---as the driver of the bunching signal.

\section{Discussion}

The results reveal a striking hierarchy of capacity distortions. The CAH program at 25 beds produces a bunching effect that is not merely larger than the effects at 50 and 100 beds---it is categorically different. A normalized bunching statistic of $b = 14.91$ means that the observed mass at the threshold is roughly 15 times the counterfactual density. In practical terms, approximately 10,000 hospital-year observations are ``misallocated'' relative to a counterfactual without the CAH incentive. This represents hospitals that either (a) downsized from above 25 beds to qualify for cost-based reimbursement, or (b) chose not to grow beyond 25 beds despite demand that would have justified expansion.

The welfare implications of this distortion depend on what would happen to these hospitals absent the CAH program. If the counterfactual is closure, then the 25-bed cap is a price worth paying for maintaining rural access---a point emphasized by \citet{Holmes2006} and \citet{Flex2020}. But if hospitals are suppressing capacity that would otherwise serve patients, the threshold creates a genuine access cost. The missing mass above 25 beds---723 hospital-years in the 26--28 bed range---suggests that a meaningful number of hospitals are operating below their unconstrained optimal size, potentially rationing services or diverting patients to more distant facilities. At an average daily census of approximately 5 patients per bed in small rural hospitals, 723 missing hospital-years in the 26--28 bed range correspond to roughly 1,000--3,600 forgone bed-years of capacity over the panel. \citet{Joynt2011} document that CAH hospitals have higher mortality for some conditions than comparable non-CAH facilities, though the direction of causality is contested.

Two important caveats apply to the welfare interpretation. First, the bunching analysis identifies distortions in the distribution of \textit{reported} beds available, which may diverge from actual staffed capacity if hospitals strategically report exactly 25 beds while maintaining surge capacity above the threshold. The HCRIS variable measures beds ``set up and staffed,'' which mitigates but does not eliminate this concern. Second, the heaping decomposition assumes that cognitive rounding at 25 beds operates similarly for CAH-eligible rural hospitals and the non-CAH benchmark. If rural administrative reporting cultures produce stronger heaping at round numbers, the regulatory-specific estimate of $b = 14.91$ may be slightly overstated. However, the urban hospital placebo ($b = 3.45$) and the average heaping at non-regulatory round numbers ($b = 0.44$) bound the heaping contribution well below the total effect, making it unlikely that heaping accounts for more than a modest fraction of the observed bunching.

The decomposition of regulatory bunching from round-number heaping is a methodological contribution with broad applicability. In any setting with integer-valued running variables---firm size thresholds, school enrollment cutoffs, emissions caps---researchers face the challenge of distinguishing policy-induced responses from cognitive or administrative rounding. The approach here---using non-regulatory round numbers as a heaping benchmark---provides a simple, transparent correction. In this application, the correction matters modestly: heaping accounts for $b = 2.26$ of the total $b = 17.16$ at 25 beds (13\%). But in settings where regulatory and round-number thresholds coincide more closely (e.g., at 50 beds), the decomposition is essential for credible inference.

One limitation is that the 100-bed DSH threshold is sensitive to polynomial specification: the estimate ranges from $b = 0.4$ (degree 5 and 8) to $b = 2.5$ (degree 6--7). This sensitivity likely reflects the broader, smoother distribution of bed counts in the 60--140 range, where the polynomial must fit a more complex shape. I interpret the 100-bed bunching as suggestive rather than definitive. Importantly, the DSH formula's incentive structure differs from the other two thresholds: it rewards hospitals \textit{above} 100 beds with a more generous reimbursement formula, whereas the CAH and RHC/REH programs reward hospitals \textit{below} their respective thresholds. The observed excess mass at 100 beds is therefore more likely to reflect hospitals growing \textit{to} 100 (or avoiding contraction below 100) rather than capping capacity---a qualitatively different behavioral response that a symmetric manipulation window may not fully capture. Future work could employ asymmetric windows or local polynomial methods to better characterize the DSH response.

\section{Conclusion}

Medicare's hospital payment system has been shaped by decades of legislative accretion. This paper shows that one consequence of that history is a hospital size distribution distorted by an architecture of regulatory cliffs. The Critical Access Hospital program, by offering 101\% cost-based reimbursement to hospitals with 25 or fewer beds, has created one of the most dramatic bunching effects in any economic domain. The question for policymakers is not whether the distortion exists---it is whether the access benefits of the CAH program justify a payment structure that discourages rural hospitals from growing, even when communities might benefit from expanded capacity.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
