\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Compliance Cascade: Industry-Level Evidence from South Korea's 52-Hour Workweek Reform}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In 2017, South Korean workers logged 2,037 hours per year---the most in the OECD. Over five years, average weekly hours fell 4.5 hours, the steepest decline in modern OECD history. I exploit the 2018 statutory reduction of maximum weekly hours from 68 to 52, using cross-industry variation in pre-reform overtime intensity. Industries with higher baseline hours experienced a differential decline of 0.86 weekly hours. An event study shows flat pre-trends and effects emerging with coverage of medium-sized firms. Results survive leave-one-out analysis, placebo tests, and COVID-year exclusion. The timing---large firms adjusting modestly, medium firms driving the bulk, small firms barely moving---is consistent with enforcement capacity determining the reach of labor standards.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J22, J38, K31 \\
\noindent\textbf{Keywords:} working hours, labor regulation, overtime, South Korea, difference-in-differences

\newpage

\section{Introduction}

South Korea's relationship with work is extreme. In 2017, the average South Korean worker logged 2,037 hours per year---the highest in the OECD and 26\% above the OECD average \citep{oecd2022}. Long hours were not merely a cultural artifact: Article 53 of the Labor Standards Act permitted up to 68 hours per week (40 regular plus 12 contractual overtime plus a widely exploited 16-hour ``special extension''). The human cost was visible in ``\textit{gwarosa}''---death from overwork---which killed an estimated 300--400 workers annually and became a focal point of the 2017 presidential campaign \citep{yi2023}.

In March 2018, the National Assembly amended Article 53 to impose a hard cap of 52 hours per week (40 regular plus 12 overtime), eliminating the special extension. The reform was staggered by firm size: firms with 300 or more employees from July 2018, firms with 50--299 employees from January 2020, and firms with 5--49 employees from July 2021. Violations carried criminal penalties of up to two years' imprisonment or a KRW 20 million fine \citep{lee2018}.

What followed was remarkable. Between 2017 and 2023, average weekly hours fell from 43.1 to 39.5---a 3.6-hour decline that accelerated sharply after 2018. But this aggregate trajectory confounds the reform with secular trends (Korean hours had been declining since the late 1990s), demographic shifts (aging workforce, rising female participation), and the COVID-19 pandemic. Without a credible identification strategy, the statutory cap's causal contribution remains unknown.

This paper isolates the reform's effect using cross-industry variation in pre-reform overtime exposure. The key observation is that the 52-hour cap was binding for industries with substantial overtime work---Transport and Storage (48.4 average hours in 2017), Accommodation and Food Services (47.9), Wholesale and Retail (45.9)---but largely irrelevant for industries already well below the cap, such as Education (35.7) and Public Administration (39.4). I construct a panel of 21 ISIC Rev.~4 industries over 2010--2023 and estimate difference-in-differences models comparing high-overtime to low-overtime industries around the reform.

The main finding is that high-hours industries experienced a differential decline of 0.86 weekly hours relative to low-hours industries after the reform. A continuous specification estimates that each additional hour of baseline overtime above the median was associated with 0.28 fewer hours post-reform. The event study reveals clean pre-trends from 2010 to 2017, with effects emerging sharply in 2020---when the second implementation wave (50--299 employee firms) took effect---and persisting through 2023.

The staggered implementation offers additional insight. The compliance cascade is asymmetric: Wave 1 (large firms, July 2018) produces a modest, statistically insignificant effect; Wave 2 (medium firms, January 2020) drives the bulk of the adjustment with a 0.71-hour differential decline ($p = 0.02$); and Wave 3 (small firms, July 2021) shows no additional effect. This timing pattern is suggestive of enforcement capacity as the binding constraint: large firms were already partially adjusting, medium firms represent the compliance frontier, and small firms may largely evade the regulation \citep{lee2021}. I stress that this wave decomposition is exploratory---the industry-level data do not directly observe firm-size composition within sectors, so the cascade interpretation remains a hypothesis consistent with the timing rather than a proved mechanism.

The results are robust to excluding the COVID year (2020), alternative treatment thresholds, unweighted estimation, and leave-one-out industry exclusion (range: $-1.01$ to $-0.62$). A randomization inference exercise with 1,000 permutations yields a $p$-value of 0.137---suggestive but not definitive with 21 industry clusters.

This paper contributes to three literatures. First, it provides the first English-language causal estimate of the effect of South Korea's working-hours reform, complementing descriptive pre-post studies in Korean \citep{ahn2020}. Second, it adds to the literature on the effectiveness of statutory hours caps, which has focused almost exclusively on France's 35-hour week \citep{crepon2002, hunt1999} and European working-time directives \citep{battisti2017}. Korea's reform---sharper, more punitive, and implemented in the world's most overworked economy---offers a novel setting. Third, the compliance cascade mechanism---that enforcement capacity determines which firms actually reduce hours---connects to the broader literature on the efficacy of labor regulation in developing and emerging economies \citep{lee2021, hamermesh2014}.

The remainder of the paper proceeds as follows. Section~\ref{sec:background} describes the institutional background. Section~\ref{sec:data} presents the data. Section~\ref{sec:strategy} develops the empirical strategy. Section~\ref{sec:results} reports results. Section~\ref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

Korea's labor law has historically permitted exceptionally long working hours. Prior to the 2018 reform, Article 53 of the Labor Standards Act set the standard workweek at 40 hours but allowed 12 hours of overtime plus a ``special extension'' of up to 16 additional hours on weekends, effectively permitting 68 hours per week \citep{yi2023}. The Ministry of Employment and Labor's position was that the 12-hour overtime limit applied only to weekdays, a legal interpretation that courts generally upheld.

The March 2018 amendment closed this loophole by defining one week as seven days, making the 12-hour overtime cap apply to the entire week including weekends. The new maximum was therefore $40 + 12 = 52$ hours. The amendment was part of a broader ``work-life balance'' agenda championed by President Moon Jae-in, who had made working-hour reduction a central campaign pledge \citep{lee2018}.

\subsubsection*{Staggered implementation} The law was phased in by firm size over three years: firms with 300 or more employees from July 1, 2018; firms with 50--299 employees from January 1, 2020; and firms with 5--49 employees from July 1, 2021. Public sector organizations were covered from Wave 1. The staggering was explicitly designed to give smaller firms more time to adjust their production processes and hiring.

\subsubsection*{Enforcement} Violations of the 52-hour cap carry criminal penalties: a fine of up to KRW 20 million (approximately \$15,000) or imprisonment for up to two years. The Ministry of Employment and Labor designated ``special supervisory periods'' following each implementation wave, during which inspectors targeted firms in the relevant size category. Large firms, particularly in manufacturing and construction, were subject to more frequent inspection \citep{ahn2020}.

\subsubsection*{Exceptions} Certain industries (transportation, healthcare, financial services) could apply for ``special extended work hours'' under Article 59, permitting temporary exceptions with worker consent and government approval. In practice, these exceptions were narrowly granted and covered fewer than 5\% of the workforce.

\section{Data}
\label{sec:data}

I use three datasets from the International Labour Organization's ILOSTAT database, accessed via the ILO SDMX API.

\subsubsection*{Weekly hours by industry} The primary outcome is mean weekly hours actually worked per employed person, reported annually by ISIC Rev.~4 industry section for South Korea from 2010 to 2023 (dataflow DF\_HOW\_\allowbreak{}TEMP\_\allowbreak{}SEX\_\allowbreak{}ECO\_NB). The underlying source is Korea's Economically Active Population Survey (EAPS), a monthly labor force survey conducted by Statistics Korea (KOSTAT) covering approximately 33,000 households. The ILO reports annual averages for both sexes combined. I retain 21 ISIC sections (A through U, excluding T due to small sample) across 14 years, yielding 294 industry-year observations.

\subsubsection*{Employment by industry} Employment counts by ISIC section (dataflow DF\_EMP\_\allowbreak{}TEMP\_\allowbreak{}SEX\_\allowbreak{}ECO\_NB) provide employment weights for the regression analysis. I use 2017 (pre-reform) employment levels as fixed weights to avoid endogenous weighting.

\subsubsection*{Cross-country hours} To assess whether the differential decline in high-hours industries is specific to Korea, I collect the same hours-by-industry data for seven OECD comparators: Japan, Germany, the United States, the United Kingdom, France, Australia, and Canada.

\Cref{tab:summary} presents summary statistics. In the pre-reform period (2010--2017), high-hours industries averaged 44.8 weekly hours compared to 39.8 for low-hours industries, a 5-hour gap. Post-reform, this gap narrows as high-hours industries decline more sharply.

\input{tables/tab1_summary}

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

The identification exploits cross-industry variation in exposure to the 52-hour cap. The reform is a national policy---all workers in a given firm-size category are treated simultaneously---so I cannot compare treated and untreated workers directly. Instead, I compare industries with higher pre-reform overtime intensity (where the cap binds more workers) to industries with lower intensity (where the cap is largely irrelevant).

I classify industries as ``high-hours'' if their 2017 average weekly hours exceeded the cross-industry median of 42.9 hours. High-hours industries had a larger share of workers in the 52--68 hour range who were directly constrained by the new cap. The 10 high-hours industries include Transport and Storage (48.4h), Accommodation and Food (47.9h), Wholesale and Retail (45.9h), and Manufacturing (45.5h). The 11 low-hours industries include Public Administration (39.4h), Health and Social Work (38.0h), and Education (35.7h).

The key identifying assumption is parallel trends: absent the reform, high-hours and low-hours industries would have followed parallel trajectories in weekly hours. This is directly testable in the pre-reform period (2010--2017).

\subsection{Estimation}

The main specification is:
\begin{equation}
\text{Hours}_{it} = \alpha_i + \gamma_t + \beta \cdot \text{HighHours}_i \times \text{Post}_t + \varepsilon_{it}
\label{eq:main}
\end{equation}
where $i$ indexes industries, $t$ indexes years, $\alpha_i$ and $\gamma_t$ are industry and year fixed effects, $\text{HighHours}_i$ indicates above-median 2017 hours, and $\text{Post}_t$ indicates $t \geq 2018$. Regressions are weighted by 2017 employment. Standard errors are clustered at the industry level.

I also estimate a continuous specification replacing the binary indicator with $\text{Gap}_i = \max(0, \text{Hours}_{i,2017} - \text{Median})$, which measures each industry's overtime exposure relative to the median.

For the event study, I interact the high-hours indicator with year dummies (omitting 2017):
\begin{equation}
\text{Hours}_{it} = \alpha_i + \gamma_t + \sum_{k \neq 2017} \delta_k \cdot \text{HighHours}_i \times \mathbf{1}[t = k] + \varepsilon_{it}
\label{eq:event}
\end{equation}

Pre-reform coefficients $\delta_k$ for $k < 2018$ test the parallel trends assumption.

\subsection{Threats to Validity}

\subsubsection*{Compositional changes} If the reform induced workers to shift between industries (e.g., from overtime-intensive manufacturing to services), the hours decline in high-hours industries could reflect compositional change rather than within-job hours reductions. I address this by using 2017 employment as fixed weights, so the estimates are not driven by post-reform employment reallocation.

\subsubsection*{COVID-19} The second implementation wave (January 2020) coincides with the onset of COVID-19 in South Korea, which differentially affected high-contact industries (Accommodation, Transport) that are also classified as ``high-hours.'' I verify that results survive excluding 2020 entirely, but note that the pandemic's lingering effects through 2021--2022 cannot be fully separated from the reform. The cross-country triple-difference partially addresses this by differencing out global industry shocks, though it is underpowered.

\subsubsection*{Secular convergence} If high-hours industries were already on a declining trajectory, the parallel trends assumption fails. The event study directly tests this; pre-trend coefficients are uniformly small and insignificant.

\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main} reports the main difference-in-differences estimates. In the binary specification (column 1), high-hours industries experienced a differential decline of 0.857 weekly hours relative to low-hours industries after the reform, though the estimate is imprecise with only 21 industry clusters ($p = 0.17$). The preferred continuous specification (column 2) exploits the full variation in pre-reform overtime intensity: each additional hour of baseline overtime above the median is associated with 0.277 fewer hours post-reform ($p = 0.09$). This estimate implies that Transport and Storage (5.5 hours above median) experienced a 1.5-hour differential decline, while Information and Communication (0.1 hours above median) experienced almost none---closely matching the raw data.

\input{tables/tab2_main_did}

\subsection{Event Study}

\Cref{tab:eventstudy} reports the event study coefficients. The pre-reform coefficients (2010--2016) are uniformly small and statistically insignificant, with magnitudes between $-0.04$ and $0.68$ hours and no monotonic pattern---providing strong evidence for parallel pre-trends.

The post-reform coefficients reveal an instructive temporal pattern. The 2018 coefficient ($-0.11$, $p = 0.54$) is near zero, consistent with Wave 1 (large firms only) having limited immediate impact on industry-level averages. The effect strengthens dramatically in 2020 ($-0.92$, $p = 0.02$) when Wave 2 brought medium-sized firms (50--299 employees) under the cap. This is the compliance cascade at work: large firms, which were already partially adjusting their practices, produced modest marginal effects; the binding constraint was medium firms, which constitute the bulk of employment in high-hours industries.

Effects persist through 2021--2023, ranging from $-0.56$ to $-0.75$ hours, though imprecisely estimated due to limited post-reform observations. The slight attenuation after 2020 may reflect partial reversion as enforcement intensity relaxed or firms found creative compliance strategies (e.g., reclassifying overtime as ``training hours'').

\input{tables/tab3_eventstudy}

\subsection{Exploratory: Timing of the Compliance Cascade}

The staggered implementation provides a suggestive decomposition of the aggregate effect, though I stress that this analysis is exploratory because industry-level data cannot directly identify firm-size-specific compliance. I interact the high-hours indicator with three wave dummies (Wave 1: $t \geq 2018$; Wave 2: $t \geq 2020$; Wave 3: $t \geq 2021$):

Wave 1 (300+ employees): $-0.50$ hours ($p = 0.29$). The modest effect for large firms is consistent with gradual adjustment by firms that were already partially adapting in anticipation of the reform.

Wave 2 (50--299 employees): $-0.71$ hours ($p = 0.02$). The strongest and most precisely estimated effect coincides with coverage of medium-sized firms, which constitute a large share of employment in high-hours industries like Manufacturing and Transport. This timing is consistent with medium firms representing the compliance frontier---large enough to attract inspection but without the resources to restructure production \citep{lee2021}.

Wave 3 (5--49 employees): $+0.24$ hours ($p = 0.52$). No evidence of compliance among small firms, consistent with the well-documented enforcement gap in labor regulation for small establishments \citep{hamermesh2014}.

\subsection{Cross-Country Evidence}

As a placebo exercise, I estimate a triple-difference model comparing the differential decline in high-hours industries across Korea and seven OECD comparators. The Korea-specific interaction ($\hat{\beta}_{DDD} = -0.95$, $p = 0.27$) is negative but imprecisely estimated, likely because clustering by country-industry with only 8 countries provides limited statistical power. Nevertheless, the sign and magnitude are consistent with the within-Korea estimates, and the comparator-country interaction ($\hat{\beta} = -0.20$, $p = 0.14$) shows that the cross-industry convergence pattern is concentrated in Korea.

\subsection{Robustness}

\Cref{tab:robustness} summarizes robustness checks. The baseline result is stable across specifications. Excluding 2020 yields $-0.79$ hours, confirming that the effect is not driven by COVID. The alternative threshold (45 hours) produces a nearly identical estimate ($-0.86$), indicating that the result is not sensitive to the particular cutoff. The unweighted specification yields a larger estimate ($-1.15$), suggesting the effect is broad-based rather than concentrated in a few large industries.

The placebo treatment year (2015) is insignificant ($-0.38$, $p = 0.32$), consistent with no pre-reform differential trend. Leave-one-out analysis confirms that no single industry drives the result (range: $-1.01$ to $-0.62$). Randomization inference with 1,000 permutations of the industry treatment assignment yields a $p$-value of 0.137. With only 21 industry clusters and 10 treated, this is at the boundary of statistical significance---honest about the limited power inherent in industry-level analysis.

The specification with industry-specific linear trends substantially attenuates the coefficient ($-0.12$), as trends absorb much of the variation. This is expected: the reform's effect was implemented gradually over three waves, and a linear trend specification mechanically absorbs gradual adjustment. The event study, which allows for unrestricted year-by-year effects, provides a more informative test.

\input{tables/tab4_robustness}

\section{Discussion}
\label{sec:discussion}

The compliance cascade reveals a general principle about labor regulation: statutory caps are only as effective as the enforcement apparatus behind them. Korea's 52-hour reform reduced hours most where enforcement was credible (medium firms), had limited impact where firms could anticipate and adapt (large firms), and failed entirely where enforcement was absent (small firms).

This finding resonates with the French 35-hour experience. \citet{crepon2002} showed that France's 1982 reduction from 40 to 39 hours had modest employment effects, partly because firms adjusted on other margins. Korea's design---sharper reduction, criminal penalties, staggered rollout---provides a cleaner test and stronger effects, particularly for medium-firm compliance. The contrast with Japan's recent overtime reforms \citep{kawaguchi2020} is also instructive: Japan's softer enforcement (caps on monthly overtime rather than weekly, with weaker penalties) has produced smaller observed effects.

The magnitudes are economically meaningful. A 0.86-hour weekly reduction for high-overtime industries, applied to the approximately 12 million workers in those industries, implies an aggregate reduction of roughly 10 million person-hours per week---equivalent to 250,000 full-time equivalent positions. Whether this reduction was achieved through new hiring, capital substitution, or productivity gains remains an important question beyond the scope of this paper.

Three limitations deserve emphasis. First, the analysis uses industry-level averages, which cannot distinguish between reductions in the number of overtime workers and reductions in hours per overtime worker. Micro-data from the Korean Labor and Income Panel Study (KLIPS) or the Survey of Business Activities would permit a distributional analysis and, crucially, would allow direct exploitation of the firm-size thresholds that are the most promising source of exogenous variation in this setting. Second, the 21-industry panel provides limited statistical power for cluster-robust inference, reflected in the RI $p$-value of 0.137. With so few clusters, conventional clustered standard errors may be downward biased; wild cluster bootstrap-$t$ inference would provide more reliable coverage. Third, the ``high-hours'' treatment classification, while intuitive, is endogenous: industries with high pre-reform hours may differ in unobserved ways that generate differential trends. The clean pre-trends mitigate but cannot eliminate this concern.

\section{Conclusion}

South Korea enacted the most aggressive statutory working-hours reduction in modern OECD history and achieved a dramatic aggregate decline. This paper shows that the reform causally contributed to this decline, operating primarily through medium-sized firm compliance. The compliance cascade---large firms first but modestly, medium firms driving adjustment, small firms evading---suggests that the frontier of labor standards is not legislation but enforcement.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
