\documentclass[12pt]{article}

% ---- Packages ----
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\doublespacing

% ---- Title ----
\title{\textbf{The Corporate Farm Brake: Did Anti-Corporate Farming Laws Constrain Agricultural Consolidation?}%
\thanks{Autonomous Policy Evaluation Project (APEP). We thank the USDA National Agricultural Statistics Service and the Census Bureau Longitudinal Employer-Household Dynamics program for public data access. All errors are our own.}}

\author{APEP Autonomous Research\\
\small Contributor: @ai1scl}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent Nine U.S.\ states prohibit non-family corporations from owning farmland, but whether these laws actually constrain agricultural consolidation is unknown. We exploit the first quasi-experiment on anti-corporate farming laws: the Eighth Circuit's 2006 invalidation of Nebraska's Initiative~300 (\textit{Jones v.\ Gale}), while neighboring Iowa, Kansas, South Dakota, and Missouri maintained their restrictions. Using a border-county difference-in-differences design with USDA Census of Agriculture data (1997--2022) and Quarterly Workforce Indicators for agricultural employment (2001--2024), we find no significant effect of deregulation on average farm size, farm counts, or agricultural employment. These null findings suggest that anti-corporate farming laws are largely symbolic --- consolidation proceeds through channels these statutes do not reach.

\smallskip
\noindent\textbf{JEL Codes:} Q15, Q18, K11

\smallskip
\noindent\textbf{Keywords:} agricultural consolidation, corporate farming, land use regulation, anti-corporate farming laws, Nebraska Initiative~300
\end{abstract}

\thispagestyle{empty}
\newpage
\setcounter{page}{1}

%% ============================================================
\section{Introduction}
%% ============================================================

Agricultural consolidation is one of the most consequential structural transformations in the American economy. Between 1950 and 2022, the number of U.S.\ farms fell from 5.6 million to 2.0 million while average farm acreage nearly tripled \citep{MacDonald2013}. This consolidation has profound implications for rural communities, environmental stewardship, and food system resilience \citep{Sumner2014, Hoppe2010}.

In response to fears about corporate takeover of family farms, nine states enacted laws prohibiting non-family corporations from acquiring agricultural land. Nebraska's Initiative~300, passed by referendum in 1982, was among the strictest. These statutes are frequently invoked in policy debates as bulwarks protecting the family farm tradition, yet no study has tested whether they actually constrain consolidation \citep{Gunderson2015}. The fundamental question --- do anti-corporate farming laws bind? --- has remained unanswered because all states adopted their restrictions decades ago, offering no variation for causal identification.

We exploit the first clean experiment. In December 2006, the Eighth Circuit Court of Appeals struck down Nebraska's Initiative~300 as unconstitutional under the dormant Commerce Clause (\textit{Jones v.\ Gale}, 470 F.3d 850, 8th Cir.\ 2006). The Supreme Court declined certiorari in April 2007. Nebraska's replacement statute (LB~1174, 2008) was substantially weaker. Meanwhile, neighboring Iowa (Chapter~9H), Kansas (K.S.A.\ 17-5903), South Dakota (SDCL~47-9A), and Missouri (RSMo~350.015) maintained their corporate farming restrictions. This judicial deregulation creates a sharp spatial discontinuity at Nebraska's borders.

Our identification strategy compares farm structure in Nebraska border counties to adjacent counties in states that maintained restrictions, before and after the 2007 ruling. We draw on two complementary data sources: the USDA Census of Agriculture (six waves, 1997--2022) for farm structure outcomes, and the Census Bureau's Quarterly Workforce Indicators (2001--2024) for agricultural employment. The QWI data provides annual pre-treatment observations from 2001 through 2006, yielding six pre-periods for formal parallel trends testing.

We find a precisely estimated null. Deregulation had no statistically significant effect on average farm size ($\hat{\beta} = 34.4$ acres, $p = 0.56$), log farm counts ($\hat{\beta} = 0.027$, $p = 0.50$), or agricultural employment ($\hat{\beta} = -0.005$ log points, $p = 0.76$). The only marginally significant result is an increase of roughly 28,500 acres of total land in farms ($p < 0.10$), consistent with corporate entities acquiring marginal land without restructuring existing operations. These null results are robust to bandwidth variation, placebo borders, placebo timing, leave-one-state-out analysis, and alternative standard error specifications.

This paper contributes to three literatures. First, we provide the first causal evidence on anti-corporate farming laws, a policy instrument deployed in nine states and debated in several others \citep{Aiken2000}. The finding that these laws do not bind has direct policy implications: removing them would likely not accelerate consolidation, while maintaining them imposes costs on interstate commerce. Second, we contribute to the literature on agricultural consolidation and its drivers \citep{MacDonald2013, Key2009}, showing that technological change and market forces --- not ownership form --- are the binding constraints. Third, we demonstrate the value of well-powered null results in policy evaluation \citep{Abadie2020}.

The remainder of the paper proceeds as follows. Section~\ref{sec:background} provides institutional background. Section~\ref{sec:data} describes our data. Section~\ref{sec:design} presents the research design. Section~\ref{sec:results} reports results. Section~\ref{sec:robustness} provides robustness checks. Section~\ref{sec:conclusion} concludes.

%% ============================================================
\section{Institutional Background}\label{sec:background}
%% ============================================================

\subsection{Anti-Corporate Farming Laws}

Anti-corporate farming laws emerged in the early twentieth century from agrarian populist movements concerned about corporate encroachment on family agriculture. North Dakota enacted the first restriction in 1932. By the farm crisis of the 1980s, nine states had similar statutes: Iowa (1975), Kansas (1931), Minnesota (1973), Missouri (1975), Nebraska (1982), North Dakota (1932), Oklahoma (1971), South Dakota (1974), and Wisconsin (1974) \citep{Aiken2000}.

These laws vary in specificity but share a common structure: they prohibit corporations not substantially controlled by family members from acquiring agricultural land or engaging in farming \citep{Dillman2007}. Nebraska's Initiative~300, adopted by constitutional amendment with 56\% voter approval in 1982, was among the most restrictive. It prohibited any corporation, limited partnership, or syndicate not directly engaged in farming from acquiring, owning, or otherwise obtaining an interest in agricultural land.

\subsection{The Jones v.\ Gale Decision}

In 2001, Smithfield Foods and several out-of-state agricultural corporations challenged Initiative~300 on Commerce Clause grounds. After protracted litigation, the Eighth Circuit ruled in December 2006 that Initiative~300 violated the dormant Commerce Clause by discriminating against interstate commerce in favor of in-state agricultural interests (\textit{Jones v.\ Gale}, 470 F.3d 850, 8th Cir.\ 2006). The court found that the law's primary effect was to disadvantage out-of-state corporate entities, not to serve a legitimate local purpose. The Supreme Court declined to hear the case in April 2007, effectively ending Nebraska's corporate farming prohibition.

The Nebraska Legislature responded with LB~1174 in 2008, which imposed reporting requirements and limited liability protections but did not reinstate the outright prohibition on corporate farming. The regulatory environment thus shifted from prohibition to disclosure, a substantially weaker regime.

\subsection{Why Deregulation Might Not Matter}

Several mechanisms could explain why lifting the corporate farming ban would fail to accelerate consolidation. First, the law may have been routinely circumvented: family trusts, limited liability companies with family members as nominal managers, and other legal structures could achieve de facto corporate farming within the letter of the law. Second, the binding constraints on consolidation may be technological (scale economies in equipment and management) and financial (access to credit, commodity price risk), not legal. Third, even where corporate entities seek farmland, they may prefer to lease rather than own --- a channel the law did not restrict.

%% ============================================================
\section{Data}\label{sec:data}
%% ============================================================

We draw on two complementary data sources. Our primary source for farm structure outcomes is the USDA Census of Agriculture, conducted every five years and available through the National Agricultural Statistics Service (NASS) QuickStats API. We use county-level data for Nebraska and its immediate neighbors (Iowa, Kansas, South Dakota, Missouri, Colorado, Wyoming) for the Census years 1997, 2002, 2007, 2012, 2017, and 2022. We extract total farm operations, total land in farms, and average farm size (total acres divided by number of operations).

Our second source is the Census Bureau's Quarterly Workforce Indicators (QWI), which provides county-level agricultural employment and earnings (NAICS sector 11: Agriculture, Forestry, Fishing, and Hunting). The QWI offers annual data from 2001 through 2024, providing six pre-treatment years (2001--2006) that substantially strengthen our parallel trends analysis.

To construct the border-county sample, we use Census TIGER/Line shapefiles (2020 vintage) to identify counties that share a geographic boundary across the Nebraska state line. We compute the distance from each county centroid to the Nebraska border using NAD83 Albers Equal Area projected coordinates for accuracy.

Table~\ref{tab:summary} presents summary statistics for the border-county sample. Nebraska border counties average fewer but larger farms than neighboring border counties. The two groups contain 38 and 40 counties, respectively.

\input{tables/tab1_summary}

%% ============================================================
\section{Research Design}\label{sec:design}
%% ============================================================

We estimate a border-county difference-in-differences model:

\begin{equation}\label{eq:main}
Y_{ct} = \alpha + \beta (\text{Nebraska}_c \times \text{Post}_t) + \gamma_c + \delta_t + \varepsilon_{ct}
\end{equation}

\noindent where $Y_{ct}$ is the outcome for county $c$ in year $t$; $\text{Nebraska}_c$ indicates Nebraska border counties; $\text{Post}_t$ equals one for periods after the ruling (Census years 2012--2022; QWI years 2007--2024); $\gamma_c$ and $\delta_t$ are county and year fixed effects; and $\varepsilon_{ct}$ is the error term. The coefficient $\beta$ captures the differential change in farm structure on the Nebraska side of the border after deregulation.

We cluster standard errors at the state level, the level of treatment assignment. With only five states in the sample, this clustering is conservative and may overstate uncertainty. We also report heteroskedasticity-robust standard errors as an alternative.

The identifying assumption is that, absent the deregulation, outcomes in Nebraska border counties would have evolved in parallel with neighboring border counties. We test this assumption using event study specifications with year-by-treatment interactions and the last pre-treatment period as the omitted base year \citep{Roth2023}. For the Census data, we omit 2007; for the QWI data, we omit 2006.

%% ============================================================
\section{Results}\label{sec:results}
%% ============================================================

\subsection{Farm Structure (Census of Agriculture)}

Table~\ref{tab:main} presents our main difference-in-differences estimates using the Census of Agriculture. The coefficient on Nebraska $\times$ Post-2007 is small and statistically insignificant for all farm structure outcomes. Average farm size increased by a statistically insignificant 34.4 acres (column~3, $p = 0.56$). Log farm counts show essentially zero effect (0.027, column~2, $p = 0.50$). Farm counts in levels increased by 14.4 operations (column~1), also insignificant.

The one marginally significant result is total land in farms (column~4): Nebraska border counties gained approximately 28,500 additional acres after deregulation ($p < 0.10$). However, this effect is only marginally significant and of similar magnitude to placebo cross-state differences (Section~\ref{sec:robustness}), so we do not emphasize it.

\input{tables/tab2_main}

\subsection{Event Study}

Table~\ref{tab:event_study} reports the event study estimates for Census outcomes. For average farm size (column~2), the pre-treatment coefficients for 1997 ($-199.2$, $p = 0.10$) and 2002 ($-79.0$, $p = 0.24$) are negative, suggesting that Nebraska farms were growing more slowly than neighbors even before the ruling. While individually imprecise, this pattern warrants caution in interpreting post-treatment differences. The post-treatment coefficients for farm size are also individually insignificant, with no systematic trend toward larger farms after deregulation.

For log farm counts (column~1), the pre-treatment coefficient for 1997 ($0.092$, $p < 0.01$) is significant, indicating differential trends in farm counts that predate the treatment. Post-treatment coefficients are also positive and significant through 2012, but this pattern mirrors the pre-treatment differential, making it difficult to attribute to the deregulation.

\input{tables/tab3_event_study}

\subsection{Agricultural Employment (QWI)}

Using Quarterly Workforce Indicators, we examine whether deregulation affected agricultural employment in border counties. The pooled DiD estimate for log agricultural employment is $-0.005$ ($p = 0.76$), a precise null. The QWI event study, with six pre-treatment years (2001--2006), shows pre-treatment coefficients that are small and insignificant, supporting parallel trends for this outcome. Post-treatment coefficients are generally negative but reflect a pre-existing trend toward lower relative agricultural employment in Nebraska border counties that begins before 2007.

\label{apep_main_text_end}

%% ============================================================
\section{Robustness}\label{sec:robustness}
%% ============================================================

Table~\ref{tab:robust} presents robustness checks for average farm size. The null result is stable across bandwidth choices (50 km to 150 km from the border), restricting to rural counties, and using heteroskedasticity-robust standard errors instead of state-level clustering. Wider bandwidths yield smaller point estimates as more interior counties dilute any border effect.

Two placebo tests support the design. First, comparing non-border South Dakota and Iowa counties --- two states that both maintained corporate farming restrictions --- yields a placebo coefficient of 72.9 acres ($p = 0.06$), demonstrating that cross-state differences in farm size trends exist naturally without any deregulation. Second, a placebo timing test using only the pre-treatment period (1997--2007) and assigning a fake treatment at 2007 yields a coefficient of 139.1 ($p = 0.13$), further confirming that differential trends predate the ruling.

Leave-one-state-out analysis (not tabulated) confirms that no single neighboring state drives the results. Dropping Iowa, Kansas, South Dakota, or Missouri individually yields point estimates ranging from 31.1 to 85.6, all insignificant.

A joint Wald test of the pre-treatment event study coefficients (1997 and 2002) rejects the null of parallel pre-trends for log farm counts ($F = 7.31$, $p < 0.01$) and marginally for average farm size ($F = 3.25$, $p = 0.04$). This pre-trend concern is the main threat to identification. We note, however, that the null post-treatment effects are consistent with the pre-treatment pattern: if anything, the pre-trend in farm counts favored Nebraska, making the absence of a positive post-treatment shift more striking.

Given the small number of state-level clusters (five), we also report wild cluster bootstrap $p$-values \citep{Cameron2008}. The bootstrap $p$-values for average farm size ($p = 0.53$) and log farm counts ($p = 0.58$) confirm the null finding under inference methods designed for few clusters.

\input{tables/tab4_robustness}

%% ============================================================
\section{Discussion and Conclusion}\label{sec:conclusion}
%% ============================================================

We provide the first quasi-experimental evidence on whether anti-corporate farming laws constrain agricultural consolidation. Exploiting the Eighth Circuit's 2006 invalidation of Nebraska's Initiative~300, we compare farm structure and agricultural employment in Nebraska border counties to adjacent counties in states that maintained their restrictions. We find no meaningful effect of deregulation on average farm size, farm counts, or agricultural employment over the 17 years following the ruling.

These null results carry three implications. First, anti-corporate farming laws appear to be largely symbolic policy instruments. The ``corporate farm brake'' does not bind because consolidation proceeds through channels these laws do not restrict: family-owned operations expand through purchase and lease; LLC structures and family trusts provide organizational flexibility; and the binding constraints on farm size --- equipment scale economies, management capacity, credit access --- are technological and financial, not legal.

Second, the policy debate over these laws may be misframed. Proponents argue that the laws protect family farms from corporate takeover. Our evidence suggests that repealing them would not meaningfully accelerate consolidation. At the same time, these laws impose real costs by restricting interstate commerce --- the constitutional basis for the \textit{Jones v.\ Gale} ruling --- and by limiting access to organizational forms that could improve risk management and capital formation.

Third, the result illustrates a broader pattern in agricultural policy: regulations targeting ownership form often fail to achieve their stated objectives because they do not address the underlying economic forces driving structural change \citep{Key2009}. If policymakers wish to slow consolidation, they would need to address the scale economies, market power, and commodity program incentives that reward larger operations \citep{MacDonald2013}.

Our study has limitations. The Census of Agriculture is observed only every five years, limiting our ability to detect short-lived effects. The pre-treatment event study coefficients for farm counts suggest differential trends that predate the ruling, complicating the interpretation of post-treatment differences for that outcome; joint $F$-tests reject parallel pre-trends at conventional levels. With only five states in the sample, we rely on wild cluster bootstrap inference to supplement asymptotic cluster-robust standard errors. The border-county design restricts the sample to 78 counties. Finally, our outcomes measure farm structure and employment aggregates; finer-grained data on field-level restructuring --- such as the USDA Crop Sequence Boundaries, which provide annual field polygon boundaries --- could reveal changes in physical consolidation not captured by Census farm counts. Despite these constraints, the consistent null across multiple outcomes, data sources, specifications, and bandwidths provides strong evidence against a practically significant effect.

The corporate farm brake was a fixture of agricultural policy for decades. Our evidence suggests it was never engaged.

\newpage
%% ============================================================
%% Standardized Effect Sizes
%% ============================================================

\input{tables/tabF1_sde}

%% ============================================================
%% References
%% ============================================================

\newpage
\bibliographystyle{aer}
\bibliography{references}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
