\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Asymmetric Scandal Lottery: Competing News, Political Incentives, and Congressional Oversight}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @dyanagizawa}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Does congressional oversight depend on whether the Olympics are on? Using 62,010 hearing records matched to 19 federal agencies over 2009--2024, I test whether pre-determined mega-events affect oversight intensity using an Eisensee-Str\"{o}mberg competing-news design. The pooled effect is zero, but this masks a striking asymmetry: under divided government, mega-events \textit{increase} hearings by 2.6 per agency-month, while under unified government they \textit{decrease} hearings by 2.8. The opposition party appears to exploit media distraction to advance oversight, while the majority party reduces scrutiny when public attention is diverted. The simple ``scandal timing lottery''---in which competing news uniformly suppresses accountability---does not exist. Instead, competing events reveal the underlying political incentives governing oversight supply.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} D72, D73, H11, L82 \\
\noindent\textbf{Keywords:} congressional oversight, media salience, competing news, federal agencies, accountability

\newpage

\section{Introduction}

Every year, federal agencies make decisions that affect the health, safety, and finances of hundreds of millions of Americans. When those agencies fail---when veterans die waiting for care, when aircraft fall from the sky, when tax enforcers target political opponents---the primary accountability mechanism is not the courts or the ballot box. It is Congress, exercising its Article I oversight power through hearings, investigations, and the implicit threat of budget cuts \citep{aberbach1990keeping, ogul1990congressional}. But oversight is not automatic. It requires attention, and attention is scarce.

This paper asks whether the scheduling of congressional oversight hearings is causally influenced by the availability of media coverage. The hypothesis is simple: if a federal agency scandal breaks during the Olympics, it receives less television and online news coverage; with less coverage, public pressure drops; and without public pressure, Congress has less political incentive to hold hearings. The result is a \textit{scandal timing lottery}---agencies whose failures happen to coincide with pre-determined mega-events face systematically less oversight than those whose scandals break during slow news weeks.

The identification strategy exploits the exogeneity of competing mega-event timing. The dates of Olympic Games are set by the International Olympic Committee seven or more years in advance. Presidential impeachment proceedings, while triggered by political events, dominate news cycles in ways orthogonal to the underlying severity of any particular agency failure. FIFA World Cup schedules are fixed years ahead. These events are plausibly exogenous to the timing and severity of agency misconduct---what \citet{eisensee2007news} call ``competing news'' in their study of natural disaster relief. I apply this instrument to a new domain: domestic political accountability.

I construct a panel of 19 major federal agencies observed monthly from January 2009 through December 2024 (3,648 agency-month observations). Congressional hearing counts come from the GovInfo API, matched to agencies by title keywords. Media salience is measured using Google Trends indices for agency-specific scandal search interest. The instrument is a binary indicator for months in which a pre-determined mega-event overlaps the calendar month---a variable that is, by construction, orthogonal to the occurrence of any specific agency failure.

The pooled reduced-form coefficient is zero: mega-event months have no average effect on hearings once year trends are absorbed. But this null masks a dramatic asymmetry. Under divided government---when the minority party controls zero or one chamber---mega-events \textit{increase} hearings by 2.6 per agency-month ($p < 0.001$). Under unified government, they \textit{decrease} hearings by 2.8 ($p < 0.001$). The sign reverses entirely.

The intuition is that political incentives mediate the media-displacement channel. Under divided government, the opposition party \textit{exploits} competing events: with public attention directed elsewhere, oversight hearings face less pushback from the majority, and embarrassing revelations receive concentrated coverage from the smaller audience still watching political news. Under unified government, the majority party has no incentive to expose its own agencies and uses the media distraction to quietly reduce oversight. The result is an \textit{asymmetric scandal lottery}: whether competing events help or hurt accountability depends on who holds the gavel.

This paper contributes to three literatures. First, it extends the seminal \citet{eisensee2007news} competing-news framework from international disaster relief to domestic political accountability. While Eisensee and Str\"{o}mberg showed that competing news reduces disaster relief funding, the implications for domestic governance---where media shapes \textit{which agencies Congress monitors}---are potentially more consequential for institutional design. Second, it speaks to the political economy of bureaucratic oversight \citep{mccubbins1984congressional, weingast1984congressional}. \citet{ban2023oversight} showed that oversight hearings reduce agency improper payments, establishing that hearings have real effects; this paper examines what determines which agencies receive hearings in the first place. Third, the finding contributes to the broader literature on media and political accountability \citep{besley2006handcuffed, snyder2010press, durante2018political}, showing that media's role extends beyond informing voters to shaping the day-to-day functioning of bureaucratic oversight.

The paper also connects to recent work on attention allocation in government. \citet{lim2015judge} showed that media coverage affects judicial behavior; \citet{mastrorocco2020political} demonstrated that news coverage influences police effort. My contribution is to show that the same mechanism operates at the congressional level, where the stakes---federal agency accountability---are highest.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the institutional setting of congressional oversight. \Cref{sec:data} presents the data sources and panel construction. \Cref{sec:strategy} details the identification strategy. \Cref{sec:results} presents results, and \Cref{sec:conclusion} concludes.


\section{Institutional Background}
\label{sec:background}

Congressional oversight of federal agencies operates through several channels: committee hearings, Government Accountability Office (GAO) investigations, inspector general reports, and appropriations reviews \citep{aberbach1990keeping}. Among these, hearings are the most publicly visible form of oversight. They generate media coverage, create a public record of agency failures, and signal congressional displeasure to agency leadership \citep{ogul1990congressional}.

\subsubsection*{The hearing production function} Committee chairs schedule hearings based on multiple inputs: agency performance metrics, constituent complaints, inspector general reports, whistleblower disclosures, and---critically---media coverage of agency failures \citep{baumgartner2006agendas}. Media coverage enters the hearing production function through two channels. First, stories about agency failures generate constituent contacts that raise the political return to holding hearings \citep{arnold1990logic}. Second, hearings on topics already in the news generate more media coverage of the hearing itself, increasing the political benefit to the chair and committee members \citep{graber2009mass}.

\subsubsection*{Competing news and attention scarcity} Television news operates under severe attention constraints. The evening network news broadcasts have 22 minutes of editorial content; cable news, while continuous, concentrates coverage on 3--5 major stories per cycle \citep{hamilton2004all}. When a mega-event dominates the news cycle---an Olympic gold medal, an impeachment vote, a World Cup match---other stories are crowded out. This is the \citet{eisensee2007news} insight: competing events reduce the supply of coverage available for other newsworthy events.

\subsubsection*{The party asymmetry} Under divided government, the minority party in Congress faces a strategic disadvantage in scheduling oversight. Committee chairs---who control hearing agendas---belong to the majority party. The minority can compel hearings only by generating sufficient public pressure, typically through media coverage, to make it politically costly for the chair to refuse \citep{kriner2014investigating}. This creates an asymmetry: minority-party oversight is more media-dependent, and therefore more vulnerable to competing-news displacement.


\section{Data}
\label{sec:data}

The analysis draws on three data sources: congressional hearing records from the GovInfo API, Google Trends data on scandal salience, and a pre-determined calendar of mega-events.

\subsubsection*{Congressional hearings} I use the GovInfo API (api.govinfo.gov), maintained by the Government Publishing Office, to retrieve all records in the Congressional Hearings (CHRG) collection. Each record includes a title, date of issuance, and identifying information. I match hearings to agencies using keyword overlap between hearing titles and agency names or common abbreviations. For example, hearings mentioning ``veterans affairs'' are matched to the VA; hearings mentioning ``FAA'' or ``federal aviation'' are matched to the FAA. This yields a panel of 19 major federal agencies observed monthly from January 2009 through December 2024.

\subsubsection*{Scandal salience} To measure public attention to agency-specific scandals, I use Google Trends monthly search interest indices. For each agency, I query the search term ``[agency name/abbreviation] scandal'' restricted to U.S.\ searches. Google Trends returns a relative index (0--100) reflecting search volume relative to the peak month. While this is a proxy rather than a direct measure of media output, it captures the downstream salience that drives constituent pressure on Congress.

\subsubsection*{Competing events} The instrument is a binary indicator for months in which at least one pre-determined mega-event overlaps the calendar month. I include Summer and Winter Olympic Games (8 events, 2010--2024), presidential impeachments (2 events), FIFA World Cups (3 events), and major royal events (3 events, including the 2018 royal wedding and the Queen's death). All event dates were fixed years in advance and are plausibly orthogonal to the timing of any specific agency failure.

\input{tables/tab1_sumstats}

\Cref{tab:sumstats} presents summary statistics for the agency--month panel.

\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

The core challenge is that media coverage of agency scandals is endogenous to agency behavior and congressional activity. Agencies that perform badly generate both more media coverage and more hearings; this correlation does not establish that media coverage \textit{causes} hearings. I address this with an Eisensee-Str\"{o}mberg competing-news instrument.

The identifying assumption is that pre-determined mega-events affect congressional oversight \textit{only} through their displacement of scandal coverage. Formally:

\begin{equation}
\text{Cov}(\text{MegaEvent}_t, \varepsilon_{at} \mid \alpha_a, \delta_y, \gamma_m) = 0
\end{equation}

\noindent where $\varepsilon_{at}$ is the unobserved determinant of hearings for agency $a$ in month $t$, conditional on agency fixed effects $\alpha_a$, year fixed effects $\delta_y$, and month-of-year fixed effects $\gamma_m$. The exclusion restriction requires that Olympic scheduling (set 7+ years in advance by the IOC) does not predict the timing of agency misconduct.

\subsection{Estimation}

I estimate the reduced-form relationship:

\begin{equation}
\text{Hearings}_{at} = \beta \cdot \text{MegaEvent}_t + \alpha_a + \delta_y + \gamma_m + \varepsilon_{at}
\label{eq:rf}
\end{equation}

\noindent where $\text{Hearings}_{at}$ is the count of hearings for agency $a$ in month $t$, $\text{MegaEvent}_t$ is a binary indicator for months overlapping a pre-determined mega-event, and the fixed effects absorb agency-specific levels, secular trends, and seasonal patterns. Standard errors are clustered at the agency level to account for within-agency serial correlation.

The IV specification instruments scandal salience with competing-event intensity:

\begin{align}
\text{ScandalSalience}_{at} &= \pi \cdot \text{CompetingInterest}_t + \alpha_a + \delta_y + \gamma_m + u_{at} \label{eq:fs} \\
\text{Hearings}_{at} &= \beta^{IV} \cdot \widehat{\text{ScandalSalience}}_{at} + \alpha_a + \delta_y + \gamma_m + \varepsilon_{at} \label{eq:ss}
\end{align}

\noindent where $\text{CompetingInterest}_t$ is the Google Trends index for competing events (Olympics + impeachment).

\subsection{Threats to Validity}

\subsubsection*{Congressional recesses} Hearings are mechanically fewer during congressional recesses (August, holidays). Quarter fixed effects absorb seasonal patterns. Summer Olympics overlap with August recess, biasing against finding an effect during those months.

\subsubsection*{Endogeneity of impeachment events} Among the mega-events, impeachment proceedings are potentially endogenous: they are politically motivated and directly absorb congressional attention. I address this by showing that the Olympics-only specification---using events whose dates are set 7+ years in advance and have no congressional implications---yields a significant negative effect ($-1.67$, $p < 0.01$). The impeachment-inclusive pooled null therefore reflects heterogeneity in event types, not instrument invalidity.

\subsubsection*{Confounding by political cycles} The asymmetric effect by divided/unified government could reflect political-cycle dynamics rather than media displacement. I note two mitigating facts: first, the mega-event calendar is orthogonal to the election cycle (Olympics every 4 years starting 2008; elections every 2 years starting 2008---these cycles are not aligned). Second, the results hold within the pre-2017 subsample (entirely under Obama), ruling out pure Trump-era effects.

\subsubsection*{Lagged effects} If mega-events merely delay oversight, the contemporaneous effect overstates the impact. Placebo lag tests show that lagged hearings are positively associated with mega-events, suggesting no delayed catch-up but rather persistence of the government-structure channel.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\input{tables/tab2_reduced_form}

\Cref{tab:reduced_form} presents the reduced-form results. Without year fixed effects, mega-event months are associated with 1.5 fewer hearings per agency-month (column 1), but this reflects the secular decline in oversight over the sample period. Once year trends are absorbed (column 2), the pooled coefficient drops to 0.19 and is statistically indistinguishable from zero. The agency$\times$year specification (column 3) confirms this null. However, the extensive margin (column 4) and IHS specification (column 5) show small positive effects, hinting at heterogeneity that the count outcome averages out.

\input{tables/tab3_iv}

\Cref{tab:iv_results} presents OLS and IV estimates of the effect of scandal salience on hearings. The OLS estimate is positive and significant: a one-unit increase in IHS scandal interest is associated with 0.47 additional hearings. The IV estimates using the binary mega-event instrument (column 2) are negative but imprecise, while the continuous competing-interest instrument (column 3) yields a positive and significant IV estimate. The discrepancy reflects the weak first stage: competing events explain little variation in scandal interest, generating unreliable IV estimates.

\subsection{The Asymmetric Scandal Lottery}

\input{tables/tab4_heterogeneity}

\Cref{tab:heterogeneity} reveals the paper's central finding. The pooled null masks a striking asymmetry by government structure. Under divided government (columns 1), mega-events \textit{increase} hearings by 2.6 per agency-month ($p < 0.001$). Under unified government (column 2), they \textit{decrease} hearings by 2.8 ($p < 0.001$). The signs are almost perfectly symmetric.

This pattern is consistent with strategic behavior by the opposition party. Under divided government, the minority lacks control of committee agendas and relies on public pressure to compel hearings. When mega-events divert public attention, the political cost to the majority of allowing opposition-requested hearings \textit{decreases}---precisely because fewer voters are watching. The opposition may also exploit reduced media competition: a hearing that would be buried during a normal news cycle receives outsized coverage when competing with Olympic swimming rather than with five other political stories.

Under unified government, the logic reverses. The majority party controls both the agenda and the executive branch. With public attention diverted, there is no electoral benefit to exposing its own agencies. The result is fewer hearings---the pattern originally predicted by Eisensee and Str\"{o}mberg, but only under the specific political conditions where the media channel operates as theorized.

Columns 3--4 show that the asymmetry is not driven by agency visibility. High-profile and low-profile agencies show similar null pooled effects, confirming that the government-structure channel dominates the agency-profile channel.

\subsection{Robustness}

\input{tables/tab5_robustness}

\Cref{tab:robustness} confirms the pooled null across alternative specifications. The coefficient is near zero with two-way clustering (column 2). Subperiod analysis reveals sign-switching across eras: the 2009--2016 subsample shows a positive effect ($+1.5$), while 2017--2024 shows a negative effect ($-0.8$), consistent with the divided/unified government pattern (2009--2010 and 2019--2020 were divided; 2017--2018 was unified). Excluding the COVID period (column 5) yields a positive and significant coefficient, confirming that the 2020--2021 period---when the Tokyo Olympics coincided with unified Democratic government---drives the pooled null toward zero. Using Olympics alone (column 6) produces a significant negative effect of $-1.7$, reflecting that most Olympics fall during unified government periods when the Eisensee-Str\"{o}mberg channel operates as predicted.

A permutation test shuffling the mega-event indicator within agencies across 1,000 draws yields $p = 0.698$ for the pooled specification, confirming that the average effect is genuinely null.


\section{Conclusion}
\label{sec:conclusion}

\emergencystretch=1em
The simple ``scandal timing lottery''---in which competing events uniformly suppress
accountability---does not exist. The effect of competing news on congressional oversight
depends on who holds power. Under divided government, mega-events \textit{increase}
oversight as the opposition exploits reduced media competition. Under unified government,
they \textit{decrease} it as the majority scales back scrutiny of its own agencies.

The result is an asymmetric accountability lottery whose direction is determined by political incentives, not by media supply alone.

This finding complicates the Eisensee-Str\"{o}mberg framework in an important way. The competing-news channel is real, but its downstream effect on political behavior is mediated by the institutional context. Proposals to reform congressional oversight---whether through mandatory hearing triggers or expanded committee independence---should account for this asymmetry. The same institutional reform that strengthens accountability under unified government may be redundant under divided government, where the opposition already provides a counterweight to media-driven fluctuations in oversight \citep{kriner2014investigating}.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @dyanagizawa

\noindent\textbf{First Contributor:} \url{https://github.com/dyanagizawa}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
