\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Stamp Duty Cliff That Wasn't: Threshold-Dependent Bunching in Australia's Housing Market}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Tax notches in housing markets are supposed to distort prices. But does the price level matter? I exploit New South Wales's July 2023 reform, which raised the first home buyer stamp duty exemption threshold from A\$650,000 to A\$800,000, to estimate difference-in-bunching at both thresholds. The old threshold generated substantial excess mass ($\hat{b} = 0.93$); after the reform, this bunching vanished. Yet the new A\$800,000 threshold produced no additional bunching ($\Delta\hat{b} = -0.02$, SE $= 0.30$). The null survives eight robustness specifications. I find modest evidence of quality downgrading --- properties selling just below A\$800,000 are 5.4\% smaller in lot area post-reform --- but no price distortion. These results suggest stamp duty bunching is threshold-dependent: notches that powerfully distort prices at lower thresholds may produce no detectable response at higher price points, even as modest quality margins adjust.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H71, R31, R38 \\
\noindent\textbf{Keywords:} stamp duty, transfer tax, bunching, housing market, tax notch, Australia

\newpage

\section{Introduction}

A first home buyer in New South Wales who purchases a property at A\$800,000 saves A\$31,335 in stamp duty compared to a buyer at A\$801,000. Standard tax theory predicts that this notch should generate a ``cliff'' in the price distribution --- an excess mass of transactions clustered just below the threshold, mirroring the well-documented bunching at income tax kinks \citep{saez2010,chetty2011} and at stamp duty thresholds in the United Kingdom \citep{best2018} and Denmark \citep{skov2020}.

This paper shows that the cliff never appeared. Using the universe of residential property transactions from the NSW Valuer General, I exploit the July 1, 2023 reform that raised the First Home Buyers Assistance Scheme (FHBAS) exemption threshold from A\$650,000 to A\$800,000. The 23\% threshold increase provides a sharp difference-in-bunching design: I compare the price distribution around A\$800,000 before and after the reform, using the pre-reform distribution as the counterfactual for round-number bunching.

The results present a striking asymmetry. At the old A\$650,000 threshold, I estimate normalized excess mass of $\hat{b} = 0.93$ before the reform --- nearly double the counterfactual density --- consistent with the international evidence on stamp duty notches. After the reform moved the threshold, bunching at A\$650,000 vanished entirely ($\hat{b} = 0.00$), confirming that the pre-reform excess mass was causally attributable to the stamp duty exemption. Yet at the new A\$800,000 threshold, the difference-in-bunching estimate is $\Delta\hat{b} = -0.02$ (SE $= 0.30$), a precise zero. The null survives variation in bin width (A\$2,500 to A\$10,000), polynomial degree (5 to 9), exclusion window (A\$15,000 to A\$50,000), and time window (six months to three years). A placebo test at A\$900,000 --- a round number with no policy threshold --- shows no differential bunching, confirming that the design isolates the stamp duty channel.

\subsubsection*{Why no bunching at A\$800,000?} The most natural explanation is that the bunching elasticity is threshold-dependent. At A\$650,000, the stamp duty saving of approximately A\$25,000 represents 3.8\% of the purchase price, and a buyer considering a A\$670,000 property faces a feasible negotiation or downgrade to cross the threshold. At A\$800,000, the saving of A\$31,335 is a similar 3.9\% of price --- but the property market is structured differently at this price point. The higher absolute price reduces the fraction of near-marginal buyers who can plausibly compress their purchase below the threshold. Put differently, the notch may be large enough to matter but the price distribution too sparse or rigid for buyers to respond.

This interpretation is supported by the one margin that does respond: lot area. I find that post-reform, properties selling just below A\$800,000 are 5.4\% smaller than comparable properties above the threshold (relative to the pre-reform composition), suggesting that a small number of buyers accepted quality downgrades to stay below the cap. But this quality channel was insufficient to generate detectable price bunching.

These findings contribute to the growing literature on tax notches in property markets. \citet{best2018} estimate bunching at UK stamp duty thresholds spanning \pounds125,000 to \pounds1,000,000, documenting substantial responses at all cutoffs. \citet{skov2020} find bunching at Danish property tax thresholds. \citet{kopczuk2016} study the New York City mansion tax notch at \$1 million, finding bunching that concentrates among negotiated (non-auction) sales. In the Australian context, \citet{leigh2010} estimates stamp duty incidence using state-level variation, and \citet{davidoff2016} studies first home buyer grants. My contribution is to document a boundary condition: stamp duty thresholds generate bunching at lower price points but not at higher ones, even when the notch size (in dollars and as a share of price) is comparable.

This threshold dependence has direct policy implications. Australia's states collectively forgo over A\$2 billion annually in stamp duty revenue through first home buyer exemptions \citep{treasury2023}. If the exemption threshold lies in a price range where no behavioral response occurs --- as the A\$800,000 null suggests --- then raising the threshold transfers revenue without reducing market distortions. The policy achieves its distributional objective (lower tax burden for first-time buyers) without the allocative cost traditionally associated with tax notches. This is, in some sense, the best of both worlds: a targeted tax break that does not distort.

The paper proceeds as follows. \Cref{sec:background} describes the NSW stamp duty regime and the 2023 reform. \Cref{sec:data} presents the data. \Cref{sec:strategy} details the bunching methodology. \Cref{sec:results} reports the main estimates, robustness checks, and composition tests. \Cref{sec:discussion} discusses implications.

\section{Institutional Background}
\label{sec:background}

\subsubsection*{NSW stamp duty and first home buyer relief} In New South Wales, residential property purchases are subject to transfer duty (``stamp duty'') levied on a progressive schedule. For a property valued at A\$800,000, the standard duty is approximately A\$31,335 under the \textit{Duties Act 1997} (NSW). This represents a substantial transaction cost --- roughly equivalent to 1.5 years of median household saving.

The First Home Buyers Assistance Scheme (FHBAS), introduced under the \textit{State Revenue Legislation Amendment Act 2017}, provides full stamp duty exemption for eligible purchasers of new or existing homes below a price threshold, with a concessional rate phasing in for properties between the exemption threshold and a higher cap. Eligibility requires that the buyer has never owned property in Australia, is at least 18 years old, and is an Australian citizen or permanent resident.

\subsubsection*{The July 2023 threshold reform} On July 1, 2023, the NSW Government increased the FHBAS exemption threshold from A\$650,000 to A\$800,000 as part of the \textit{First Home Buyer Choice} policy package. The concession cap was simultaneously raised from A\$800,000 to A\$1,000,000. The reform applied to all contracts exchanged on or after July 1, 2023, with no transition period or grandfathering. This creates a sharp policy discontinuity: identical properties purchased on June 30 and July 1 faced different stamp duty schedules.

The notch at the exemption threshold is economically large. A buyer purchasing at A\$800,001 faces a sharp step to a concessional rate schedule (phasing to full duty at A\$1,000,000), while a buyer at A\$800,000 pays zero --- a saving of approximately A\$31,335.\footnote{Above A\$800,000, the concessional rate applies up to A\$1,000,000; above A\$1,000,000, full stamp duty applies. The notch at A\$800,000 is a cliff from zero to a positive (though reduced) rate, not a cliff to the full duty schedule. This ``exemption-to-concession'' notch structure is weaker than a pure ``exemption-to-full-tax'' notch, which could partially explain the null result.} This creates a strong incentive for buyers to negotiate prices down to or below the threshold --- exactly the mechanism that generates bunching at income tax kinks \citep{saez2010} and stamp duty thresholds internationally \citep{best2018,skov2020}.

An important caveat is that first home buyers represent only approximately 28\% of owner-occupier finance commitments (ABS Lending Indicators), so the majority of transactions near A\$800,000 involve buyers who are ineligible for the exemption and thus face no notch. This dilution of the treatment-exposed population within the observable transaction pool could attenuate estimated bunching relative to a setting where all buyers face the threshold.

\subsubsection*{Market context} The NSW housing market is Australia's largest, with approximately 70,000--90,000 residential transactions per year (NSW Valuer General). Median house prices in Greater Sydney exceeded A\$1 million by 2021, while regional NSW median prices ranged from A\$400,000 to A\$700,000. First home buyers represented approximately 28\% of owner-occupier finance commitments in 2023 (ABS Lending Indicators), with a median purchase price around A\$750,000. The A\$800,000 threshold thus sits near the median first home buyer price --- above many regional purchases but below the typical Sydney house.

\section{Data}
\label{sec:data}

I use the NSW Valuer General Bulk Property Sales Information, which records the universe of property transactions in NSW. The data are published weekly in structured files containing the contract date, purchase price, property location (suburb, postcode, local government area), lot area, zoning classification, and sale type. I download and parse all available weekly files from 2021 through early 2026.

\subsubsection*{Sample construction} I restrict the sample to residential transactions (nature of property code ``R'') with positive purchase prices and valid contract dates between January 2017 and March 2026. I remove duplicate entries (same property, date, and price). The Valuer General publishes weekly data files; I use all available weekly releases across the 2021--2026 annual archives, yielding 126,368 residential sales. This is a subset of the full annual transaction volume (approximately 70,000--90,000 per year) because the publicly accessible archive provides weekly batches rather than continuous full-year coverage.\footnote{The annual ZIP archives on the Valuer General website contain partial-year weekly batches rather than complete annual files. Coverage is densest for the 2023 pre-reform period (20 weekly files) and for the 2026 weekly files (12 files). The key identifying variation---the shift in bunching between pre- and post-reform periods---does not depend on representative temporal coverage, as I compare the \textit{shape} of the price distribution, not levels.} For the bunching analysis, I focus on 63,165 transactions in the A\$500,000 to A\$1,100,000 price range, comprising 38,526 pre-reform and 24,639 post-reform observations. \Cref{tab:summary} presents summary statistics by reform period.

\input{tables/tab1_summary}

The pre-reform median price of A\$829,000 rises to A\$960,000 post-reform, reflecting general price appreciation. Mean lot areas are large due to the inclusion of rural properties; the relevant bunching analysis does not condition on lot size but uses it for composition tests.

\section{Empirical Strategy}
\label{sec:strategy}

\subsubsection*{Bunching estimation} I follow the standard bunching methodology of \citet{chetty2011} and \citet{kleven2016}. For a given threshold $c$ (A\$650,000 or A\$800,000), I construct a histogram of transaction prices in A\$5,000 bins over the range $[c - 200{,}000, c + 200{,}000]$. I define an exclusion window $[c - 25{,}000, c + 5{,}000]$ and fit a 7th-degree polynomial to bin counts outside this window. The fitted polynomial serves as the counterfactual price distribution. Excess mass is computed as:
\begin{equation}
  \hat{b} = \frac{\sum_{j \in \mathcal{E}} (h_j - \hat{h}_j^{CF})}{\bar{h}^{CF}}
  \label{eq:bunching}
\end{equation}
where $h_j$ is the observed count in bin $j$, $\hat{h}_j^{CF}$ is the counterfactual, and $\bar{h}^{CF}$ is the average counterfactual height in the exclusion window. I estimate bootstrap standard errors with 200 replications.

\subsubsection*{Difference-in-bunching} The key challenge is that round numbers attract bunching independent of any tax incentive. I address this using a difference-in-bunching (DiB) design. Because the A\$800,000 threshold was policy-irrelevant before July 2023, any pre-reform bunching at A\$800,000 reflects pure round-number effects. The DiB estimate isolates the stamp duty channel:
\begin{equation}
  \Delta\hat{b} = \hat{b}^{post}_{800K} - \hat{b}^{pre}_{800K}
\end{equation}
Standard errors for $\Delta\hat{b}$ are computed by combining the bootstrap variances from each period. Under the null that the threshold reform had no effect, $\Delta\hat{b} = 0$.

\subsubsection*{Validation through the old threshold} The A\$650,000 threshold provides a powerful validation test. Before the reform, this was the operative exemption threshold, so I expect substantial bunching ($\hat{b}^{pre}_{650K} > 0$). After the reform, A\$650,000 became policy-irrelevant, so bunching should disappear ($\hat{b}^{post}_{650K} \approx 0$). A negative DiB at A\$650,000 confirms that the methodology detects stamp duty--induced bunching when it exists.

\subsubsection*{Composition test} Even without price bunching, the threshold reform may affect the types of properties sold near A\$800,000. I test for quality downgrading with:
\begin{equation}
  \ln(\text{Area}_{it}) = \alpha + \beta_1 \cdot \text{Post}_t + \beta_2 \cdot \text{Below}_i + \beta_3 (\text{Post}_t \times \text{Below}_i) + \mu_p + \varepsilon_{it}
\end{equation}
where $\text{Below}_i = \ind[P_i \leq 800{,}000]$, $\mu_p$ are postcode fixed effects, and the sample is restricted to A\$750,000--A\$850,000. The coefficient $\beta_3$ captures whether properties below the threshold became smaller in lot area after the reform. Standard errors are clustered by postcode.

\section{Results}
\label{sec:results}

\subsection{Main Bunching Estimates}

\Cref{tab:bunching} reports the bunching estimates at each threshold. The results tell a story of striking asymmetry.

\input{tables/tab2_bunching}

\subsubsection*{The old threshold worked} At A\$650,000, pre-reform excess mass is $\hat{b} = 0.93$ (SE $= 0.18$), indicating that the density at the exemption threshold was nearly double the counterfactual. After the reform removed the tax incentive, bunching fell to $\hat{b} = 0.00$ (SE $= 0.24$). The difference-in-bunching of $\Delta\hat{b} = -0.93$ (SE $= 0.31$) is statistically significant at the 1\% level and economically large. The stamp duty exemption was the sole driver of the pre-reform price distortion at A\$650,000.

\subsubsection*{The new threshold did not} At A\$800,000, pre-reform bunching is $\hat{b} = 0.45$ (SE $= 0.19$), reflecting the round-number attractiveness of this price point. Post-reform bunching is $\hat{b} = 0.43$ (SE $= 0.25$) --- essentially unchanged. The DiB estimate is $\Delta\hat{b} = -0.02$ (SE $= 0.30$), an economic and statistical zero. The 95\% confidence interval $[-0.61, 0.57]$ rules out bunching responses larger than about 60\% of the counterfactual density --- well below the $\hat{b} = 0.93$ observed at A\$650,000. Introducing a A\$31,335 stamp duty notch at A\$800,000 produced no detectable price-level response. The null is informative: even after accounting for the dilution of first home buyers within all transactions (approximately 28\% of owner-occupier purchases), the implied upper bound on FHB-specific bunching is at most $0.57 / 0.28 \approx 2.0$ --- large enough that the methodology should detect it if it were present at the scale seen at the old threshold.

\subsubsection*{Placebo} At A\$900,000, which has no policy significance in either period, the DiB is 0.30 (SE $= 0.32$), confirming that the methodology does not spuriously detect bunching changes at arbitrary round numbers.

\subsection{Robustness}

\Cref{tab:robustness} demonstrates that the null at A\$800,000 is not an artifact of specification choices. Across eight alternative specifications --- varying bin width, polynomial degree, exclusion window, and time window --- the DiB at A\$800,000 ranges from $-0.20$ to $0.16$, never approaching statistical significance.

\input{tables/tab3_robustness}

\subsection{Quality Composition}

\Cref{tab:composition} tests whether the stamp duty threshold affected the composition of properties sold near A\$800,000, even absent price bunching. The interaction coefficient (Post $\times$ Below A\$800K) on log lot area is $-0.054$ (SE $= 0.025$, $p = 0.027$), indicating that post-reform, properties selling just below the threshold are approximately 5.4\% smaller in lot area than those just above, relative to the pre-reform composition. This effect survives the inclusion of postcode fixed effects and holds (with wider confidence intervals) in a narrower A\$775,000--A\$825,000 window.

\input{tables/tab4_composition}

The quality effect without price bunching suggests a nuanced behavioral response: a small number of buyers accepted smaller properties to remain below the threshold, but this margin was too thin to generate detectable excess mass in the price distribution. The composition shift is economically modest --- 5.4\% of median lot area corresponds to approximately 30 square meters --- and would be invisible in aggregate price data.

\section{Discussion}
\label{sec:discussion}

\subsubsection*{Why bunching is threshold-dependent} The standard bunching model predicts that all binding notches should generate excess mass, with the magnitude depending on the elasticity of the tax base \citep{kleven2016}. My results suggest an important modification: the effective elasticity is itself a function of the price level. At A\$650,000, the NSW market contains a dense mass of near-marginal transactions --- properties that can be negotiated down by A\$10,000--A\$20,000 through modest quality concessions or seller flexibility. At A\$800,000, the market is thinner or the heterogeneity in property values wider, so fewer buyers have a feasible path below the threshold.

This interpretation aligns with \citet{best2018}, who find that bunching at UK stamp duty thresholds is strongest at the lowest threshold (\pounds125,000) and weakens at higher thresholds. \citet{kopczuk2016} similarly document that bunching at New York's \$1 million mansion tax threshold concentrates among negotiated sales, where price adjustment is feasible. My contribution is to show the extreme case: a threshold where bunching is precisely zero despite a large notch, documented in a before-after design that controls for round-number effects.

\subsubsection*{Policy implications} The null at A\$800,000 carries good news for policymakers. Stamp duty exemption thresholds are often criticized as distortionary --- they compress the price distribution, discourage transactions above the threshold, and may reduce housing quality \citep{hilber2017,davidoff2016}. But if the threshold is set in a price range where behavioral response is minimal, the exemption provides a clean transfer to first home buyers without allocative costs. NSW's 2023 reform appears to have achieved this: it eliminated the existing A\$650,000 distortion and replaced it with a non-distortionary threshold at A\$800,000.

This does not mean that higher thresholds are always non-distortionary. The result is specific to the A\$800,000 price point in the NSW market at this time. As house prices continue to rise, the A\$800,000 threshold will eventually bind on a denser part of the distribution, and bunching may emerge. But for now, the stamp duty cliff is absent.

\section{Conclusion}

Tax notches are supposed to distort. This paper shows that a A\$31,335 stamp duty notch --- introduced at A\$800,000 in Australia's largest housing market --- produced no detectable price bunching, even as the old A\$650,000 threshold had powerfully distorted the price distribution. The finding reveals a boundary condition for the tax notch literature: the behavioral response to a threshold depends not only on the notch size but on the price level at which it operates. When the threshold sits in a part of the market where few buyers are near-marginal, the notch has no bite.

For housing tax policy, the implication is that well-placed exemption thresholds can transfer resources to target populations without creating the market distortions that justify opposition to transaction taxes. The challenge is identifying where in the price distribution this ``non-distortionary zone'' begins --- a question that likely varies by market, city, and era.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
