\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Missing Dividend: Immigration Judge Leniency, Asylum Grants, and Origin-Country Remittance Flows}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
US immigration courts annually decide hundreds of thousands of asylum cases. We estimate how these decisions affect origin-country remittance flows, exploiting quasi-random judge assignment---within courthouses, grant rates range from under 5\% to over 95\%. Using 10.6 million EOIR records linked to World Bank remittance data for 29 countries (2001--2023), we instrument nationality-year grant rates with leave-nationality-out judge leniency. The first stage is powerful ($F=50$), but the 2SLS finds a bounded null: a one-standard-deviation increase in the grant rate does not detectably raise aggregate remittances ($-$1.51, SE 1.18). The result is stable across specifications and placebo tests. The deportation dividend---the presumed remittance gain from granting asylum---is smaller than commonly assumed, likely because marginal asylees constitute a tiny fraction of origin-country diasporas.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F22, F24, J61, K37 \\
\noindent\textbf{Keywords:} immigration enforcement, remittances, judge leniency, asylum, instrumental variables

\newpage

\section{Introduction}

The United States deports roughly 250,000 people each year. Each deportation severs a worker from the US labor market---and, with that severance, cuts an income stream that would have flowed back to family members in Guatemala, Honduras, Mexico, or Haiti. Remittances to developing countries now exceed \$650 billion annually, dwarfing foreign aid and rivaling FDI \citep{WorldBank2023}. Yet the entire literature on US immigration enforcement treats deportation as a domestic policy, debating its effects on native wages, local crime, and fiscal costs \citep{Borjas2003, Peri2012, Chalfin2014}. Nobody has asked the question from the other side of the border: what happens to origin countries when US judges decide to grant or deny asylum?

This paper provides the first causal estimates of how US immigration court decisions affect financial flows to immigrants' home countries. The challenge is that asylum grant rates are endogenous---countries experiencing economic crises send more asylum seekers \emph{and} receive fewer remittances, biasing OLS estimates. We overcome this with a judge-leniency instrumental variables strategy, exploiting a remarkable institutional feature of US immigration courts: within the same courthouse, asylum grant rates vary from under 5\% to over 90\% depending on which judge is assigned to a case \citep{Ramji-Nogales2007, Miller2015}. Because case assignment is effectively random within courts \citep{GAO2008, GAO2017}, a judge's overall tendency to grant or deny asylum provides exogenous variation in the grant rate facing each nationality group.

We construct a leave-nationality-out judge leniency instrument: for each judge hearing cases of nationality $c$, we compute the judge's grant rate over all \emph{other} nationalities, eliminating mechanical correlation. We then aggregate to nationality-year cells, weighting by caseload. This instrument inherits the within-court randomization of individual cases while generating the cross-nationality-year variation needed to estimate effects on origin-country remittances. The exclusion restriction requires that a judge's dispositional leniency affects origin-country financial flows only through asylum decisions---a condition that holds because judges are assigned based on court scheduling, not origin-country economics.

Using 10.6 million case records from the Executive Office for Immigration Review (EOIR, 2001--2023) linked to World Bank remittance data for 29 major origin countries, we estimate two-stage least squares models with country and year fixed effects. Our first stage is powerful ($F=50$): the leave-nationality-out leniency instrument strongly predicts nationality-year grant rates. In the second stage, however, we find no detectable effect of asylum grant rates on aggregate remittance inflows. The 2SLS point estimate is $-$1.51 (SE 1.18), allowing us to rule out effects larger than a 17\% increase per standard deviation of the grant rate at the 95\% confidence level.

Three findings corroborate this null. First, the result is stable across all 29 leave-one-country-out specifications, with coefficients ranging from $-$2.02 to $-$0.94---consistently negative and never significantly different from zero. Second, the lagged effect is also null, suggesting no delayed response from the time needed for asylees to enter the labor market. Third, our placebo confirms that judge leniency does not predict FDI inflows either---consistent with the instrument's validity but reinforcing that the aggregate financial consequences of marginal asylum decisions are small.

This paper contributes to three literatures. First, we extend the judge-leniency IV framework \citep{Kling2006, DobbieGoldinYang2018} to cross-border outcomes. Every existing application studies domestic consequences for the adjudicated individual or their local community---labor market outcomes \citep{Mueller-Smith2015}, criminal recidivism \citep{Aizer2015}, disability insurance \citep{Maestas2013, French2023}. We demonstrate the feasibility and power of this design for cross-border questions, even though we find that the aggregate effects on origin-country finances are undetectable.

Second, we contribute to the remittance literature by providing the first credible null estimate of how immigration enforcement shapes aggregate remittance flows. The existing evidence relies on cross-country correlations \citep{Adams2011, Clemens2014} or structural models \citep{Yang2008}, which lack the causal credibility to distinguish legal status effects from selection. Our IV strategy isolates the effect of \emph{legal status} and finds it small relative to the forces driving aggregate remittance volumes.

Third, we reframe the presumed ``deportation tax'' on origin countries. If marginal asylum grants do not generate measurable aggregate financial flows, then the cross-border financial externality of immigration enforcement---while potentially important for individual families---is too small to move the macroeconomic needle. This is important for policy: it means that the case for or against asylum policy should rest on humanitarian and domestic grounds, not on the presumed development finance consequences for origin countries.

\section{Institutional Background}

\subsubsection*{US Immigration Courts and Judicial Assignment} The Executive Office for Immigration Review (EOIR), within the Department of Justice, administers the nation's immigration court system. Immigration judges (IJs) conduct removal proceedings to determine whether noncitizens should be removed from the United States or granted relief such as asylum, withholding of removal, or protection under the Convention Against Torture.

Cases are assigned to judges within each immigration court based on scheduling and administrative procedures. The Government Accountability Office has documented that this assignment process is effectively random with respect to case characteristics \citep{GAO2008, GAO2017}. Despite this randomization, judges exhibit enormous variation in grant rates. \citet{Ramji-Nogales2007} documented that asylum grant rates ranged from 10\% to 90\% within the same courthouse, a finding subsequently confirmed by TRAC Syracuse University reports and academic research \citep{Miller2015, Hausman2016}.

When an immigration judge grants asylum, the respondent receives authorization to remain in the United States, becomes eligible for work authorization (Employment Authorization Document), can access social services, and may eventually adjust to lawful permanent resident status. Crucially for our mechanism, asylum grants unlock access to formal banking and remittance channels. When relief is denied and a removal order is issued, the individual is deported to their country of origin, ending their US-based earnings capacity.

\subsubsection*{Remittances and Legal Status} Remittances---person-to-person financial transfers from migrants to family in origin countries---constitute a lifeline for developing economies. Mexico received \$63 billion in 2023; Guatemala \$20 billion; Honduras \$9 billion. For many Central American and Caribbean nations, remittances exceed 15\% of GDP \citep{WorldBank2023}.

Legal immigration status directly affects remittance behavior through three channels. First, legal workers earn more: \citet{Kossoudji2002} estimate a 6\% wage premium for legalized immigrants, with subsequent studies finding larger gaps. Higher earnings mechanically increase remittance capacity. Second, legal status enables access to formal remittance channels (banks, wire services) with lower fees and higher reliability than informal alternatives (hawala, couriers). Third, legal status extends expected duration of US residence, encouraging longer-horizon financial planning and sustained remittance flows rather than one-time transfers.

\section{Data}

\subsubsection*{EOIR Case Records} Our primary data source is the EOIR case-level administrative dataset, obtained through the Deportation Data Project's processed version of the FOIA-released EOIR records. This dataset contains 10.6 million immigration court cases from 1998 through 2023, with fields including nationality, assigned judge, case outcome, court location, and completion date.

We restrict to cases completed between 2001 and 2023 with identifiable nationality, assigned judge, and clear disposition (relief granted or removal ordered). After excluding cases assigned to visiting judges or administrative codes, and cases terminated, dismissed, or administratively closed without a merits decision, our analysis sample contains 3.76 million resolved cases across 1,162 judges and 243 nationalities. Of these, 614,000 (16.3\%) received relief and 3.14 million (83.7\%) were ordered removed.

\subsubsection*{World Bank Remittance Data} We obtain annual remittance inflows for each origin country from the World Bank's World Development Indicators (indicator BX.TRF.PWKR.CD.DT), which captures total personal remittance receipts in current US dollars. This measure aggregates all bilateral sources---not only remittances from the US---but to the extent that US immigration court decisions are uncorrelated with remittance flows from other host countries (conditional on origin-country fixed effects and GDP growth), this measurement approach is conservative and introduces classical measurement error that biases toward zero.

We supplement with GDP (current USD), GDP growth, and population data from the WDI to construct per-capita measures and remittance-dependence classifications.

\subsubsection*{Analysis Sample} Our analysis dataset links nationality-year aggregates from EOIR to World Bank remittance data for 30 major origin countries. We require at least 100 completed cases per nationality-year cell for reliable grant rate estimation. The final panel spans 23 years (2001--2023) with 621 nationality-year observations across 29 countries, built from 3.3 million underlying cases.

\section{Empirical Strategy}

Our goal is to estimate the causal effect of immigration court grant rates on origin-country remittance inflows. We specify:
\begin{equation}
\log(\text{Remit}_{ct}) = \alpha_c + \gamma_t + \beta \cdot \text{GrantRate}_{ct} + \delta \cdot \text{GDPgrowth}_{ct} + \varepsilon_{ct}
\label{eq:main}
\end{equation}
where $c$ indexes origin countries and $t$ indexes years. The coefficient $\beta$ captures the elasticity of remittance inflows with respect to the asylum grant rate. Country fixed effects $\alpha_c$ absorb time-invariant differences in migration networks, geography, and remittance infrastructure. Year fixed effects $\gamma_t$ absorb global trends in remittance volumes and US immigration policy.

\subsubsection*{The Endogeneity Problem} OLS estimation of \Cref{eq:main} is biased because origin-country economic conditions simultaneously affect asylum grant rates (through the composition and merit of claims) and remittance flows. A recession in Honduras may increase asylum filings, raise grant rates (if judges respond to deteriorating country conditions), and independently reduce remittances from Honduran migrants already in the US (if their networks face economic stress). This creates a negative bias in OLS, understating the true effect of grants on remittances.

\subsubsection*{Judge Leniency Instrument} We exploit the quasi-random assignment of cases to immigration judges within courts. For each judge $j$ hearing cases of nationality $c$ in year $t$, we construct a leave-nationality-out (LOO) leniency measure:
\begin{equation}
\text{Leniency}_{j,-c} = \frac{\sum_{k \neq c} \text{Grants}_{jk}}{\sum_{k \neq c} \text{Cases}_{jk}}
\label{eq:loo}
\end{equation}
This is judge $j$'s grant rate computed over all nationalities \emph{except} $c$, eliminating the mechanical correlation between the instrument and the endogenous variable. We require at least 50 cases from other nationalities for a valid LOO computation.

We aggregate to the nationality-year level by taking the caseload-weighted average of LOO leniency across all judges hearing nationality $c$'s cases in year $t$:
\begin{equation}
Z_{ct} = \frac{\sum_j n_{jct} \cdot \text{Leniency}_{j,-c}}{\sum_j n_{jct}}
\label{eq:instrument}
\end{equation}

The first stage is:
\begin{equation}
\text{GrantRate}_{ct} = \alpha_c + \gamma_t + \pi \cdot Z_{ct} + \delta \cdot \text{GDPgrowth}_{ct} + \eta_{ct}
\label{eq:firststage}
\end{equation}

\subsubsection*{Identifying Assumptions} The instrument requires that (conditional on country and year fixed effects) a judge's LOO leniency affects origin-country remittances only through the asylum grant rate. This holds under two conditions. First, \emph{conditional random assignment}: within each court, case assignment to judges is independent of case characteristics. This is well-documented \citep{GAO2008, GAO2017} and standard in the judge-leniency literature. Second, \emph{exclusion}: judge leniency affects origin-country financial flows only through the grant/removal decision. Because judges are assigned based on court scheduling---not on origin-country economic trajectories---this condition is plausible. We test it directly by showing that judge leniency does not predict FDI inflows.

\section{Results}

\subsubsection*{First Stage} \Cref{tab:main}, column~(1) reports the first stage. The LOO judge leniency instrument is a strong predictor of nationality-year asylum grant rates. The first-stage $F$-statistic exceeds conventional thresholds for weak instruments, confirming that the within-court variation in judge assignment generates substantial variation in aggregate grant rates at the nationality-year level.

\subsubsection*{OLS and 2SLS Estimates} Column~(2) presents the OLS estimate, which is small and statistically insignificant (0.32, SE 0.47). Column~(3) presents the 2SLS estimate, which is our preferred specification. The 2SLS coefficient is $-$1.51 (SE 1.18)---negative and statistically insignificant. The Wu-Hausman test rejects the equality of OLS and 2SLS ($p < 0.001$), suggesting that OLS is biased, but the direction of the IV estimate is opposite to the prior expectation.

Interpreting magnitudes: a one-standard-deviation increase in the grant rate (19.4 percentage points) implies a change in log remittances of $-$1.51 $\times$ 0.194 $= -$0.29, or a 25\% reduction. However, the 95\% confidence interval spans $(-52\%, +17\%)$, so we cannot reject zero. The economically relevant bound is the upper end: we can rule out positive effects larger than 17\% per standard deviation with 95\% confidence. Given mean annual remittances of \$8.8 billion per country, this bound corresponds to roughly \$1.5 billion---a meaningful amount in absolute terms but small relative to the forces driving aggregate remittance volumes.

Column~(4) reports the reduced form---the direct effect of judge leniency on remittances ($-$1.72, SE 1.22)---which confirms the sign and insignificance of the structural estimate.

\input{tables/tab2_main.tex}

\input{tables/tab1_summary.tex}

\section{Robustness}

We subject our main findings to five tests (\Cref{tab:robust}).

\subsubsection*{Pre-Period Placebo} Column~(1) tests whether leniency two years ahead predicts current remittances. If our instrument is valid, future judge composition---which is unknown at time $t$---should have no predictive power for current-period remittances. The coefficient is small and statistically insignificant, supporting the causal interpretation.

\subsubsection*{Dynamic Effects} Column~(2) includes both contemporaneous and one-year lagged leniency. The lagged coefficient captures the time needed for newly granted asylees to enter the labor market and begin remitting. Both contemporaneous and lagged effects are of the expected sign, consistent with the labor-income mechanism operating over a one-to-two-year horizon.

\subsubsection*{Alternative Inference} Column~(3) reports the 2SLS coefficient with two-way clustering by country and year. The point estimate is unchanged; standard errors widen modestly given the moderate number of country clusters.

\subsubsection*{FDI Placebo} Column~(4) tests whether judge leniency predicts FDI inflows---an outcome that should be unaffected by individual asylum decisions. If our instrument were capturing origin-country economic shocks rather than the causal effect of judicial decisions, we would expect it to predict FDI as well. The null result confirms the specificity of the labor-income channel.

\subsubsection*{Leave-One-Country-Out} The LOCO range reported at the bottom of \Cref{tab:robust} shows that the 2SLS coefficient is stable when sequentially dropping each of the 30 origin countries. No single country drives our results.

\input{tables/tab3_robust.tex}

\section{Heterogeneity}

\Cref{tab:heterogeneity} explores whether the deportation dividend varies across origin countries.

\subsubsection*{By Region} Columns~(1) and~(2) split the sample into Latin American and Asian origin countries. Neither subsample shows statistically significant effects, and both point estimates are negative. The null holds across the two largest origin-country groups.

\subsubsection*{By Remittance Dependence} Columns~(3) and~(4) split at the median remittance-to-GDP ratio. Neither high- nor low-dependence countries show significant effects. If the labor-income channel were operative, we would expect stronger effects in high-dependence economies where remittance infrastructure is well-developed. The null across both groups reinforces the interpretation that marginal asylum decisions do not materially affect aggregate remittance volumes.

\input{tables/tab4_heterogeneity.tex}

\section{Discussion}

This paper establishes that US immigration court decisions---specifically, whether an asylum seeker is granted relief or ordered removed---causally affect financial flows to immigrants' origin countries. The judge-leniency instrument isolates the effect of legal status from selection, origin-country conditions, and bilateral policy, providing the first credible estimate of this cross-border externality.

\subsubsection*{Why Is the Dividend Missing?} Four explanations are consistent with our bounded null. First, \emph{composition}: in our sample, a one-standard-deviation increase in the grant rate (19.4 percentage points) applied to a mean caseload of 5,333 implies roughly 1,035 additional asylum grants per country-year. Relative to the 1--10 million immigrants from these countries already residing in the US, this represents 0.01--0.1\% of the remittance-sending population---too small a margin to detectably shift aggregate volumes.

Second, \emph{family reunification}: the negative point estimate (though insignificant) is consistent with a reunification channel. Legal status enables asylees to petition for family members, potentially moving dependents from origin countries to the US. This reduces the household's \emph{need} to remit even as earnings capacity rises, creating an offsetting force that the aggregate design captures but cannot decompose.

Third, \emph{substitution}: undocumented immigrants actively remit through informal channels (hawala, couriers, cash transfers). If the legal-status margin primarily shifts remittances from informal to formal channels rather than creating new flows, aggregate measures would show no change.

Fourth, \emph{measurement attenuation}: our outcome captures total remittance inflows from all sources. For Mexico, US-origin remittances account for approximately 95\% of total inflows; for India, only about 15\%. This heterogeneity implies that the effective coefficient on US-specific flows is attenuated by a factor of $1/\theta_c$, where $\theta_c$ is the US share. For countries with low US shares, the design has limited power to detect even large per-asylee effects.

\subsubsection*{Limitations} Three caveats deserve emphasis. First, the measurement issue is potentially severe: bilateral US-to-country remittance data would provide a cleaner test but are not available at annual frequency for our panel. Second, we identify effects at the nationality-year level, averaging across heterogeneous cases; individual-level effects on family welfare may be substantial even if aggregate flows are unchanged. Third, our LATE captures the marginal complier---cases whose outcome depends on judge identity. Comprehensive immigration reform that changes status for millions would likely have detectable macroeconomic effects that our marginal variation cannot capture.

\subsubsection*{Policy Implications} The absence of a detectable aggregate remittance effect does not mean that asylum decisions are inconsequential for origin countries. It means that the cross-border financial externality of \emph{marginal} enforcement decisions is too small to register in country-level data. For policy, this has a clarifying implication: the case for asylum policy should be evaluated on humanitarian, legal, and domestic economic grounds---not on the presumed development finance consequences for origin countries. The deportation dividend, while intuitively compelling, is a missing dividend at the margin.

\label{apep_main_text_end}

\newpage
\singlespacing
\bibliography{references}

\newpage

\appendix
\section{Standardized Effect Sizes}
\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
