\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{siunitx}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Compliance Trap: How IR35 Pushed Contractor-Intensive Firms Past the Small Company Threshold}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When the UK extended IR35 off-payroll working rules to private-sector clients in April 2021, it exempted small companies---those with fewer than 50 employees. I show that this exemption backfired: contractor-intensive sectors experienced a \textit{decline} in the ratio of firms below the 50-employee threshold relative to those above, the opposite of the strategic downsizing the exemption might have invited. Using NOMIS UK Business Counts data from 2010--2025 in a difference-in-bunching design, I estimate that the bunching ratio fell 0.31 units more in contractor-intensive sectors than in control sectors ($p = 0.018$). The effect is driven by computer programming (SIC 62) and management consultancy (SIC 70), strengthens over time, and survives placebo and pre-trend tests. The most likely mechanism is contractor-to-employee conversion, which inflated official headcounts and pushed firms past the very threshold designed to protect them.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H25, J23, L25, K34 \\
\noindent\textbf{Keywords:} IR35, off-payroll working, firm size manipulation, bunching, compliance costs, contractor employment

\newpage

\section{Introduction}

The UK's IR35 rules are, by any account, among the most consequential regulatory interventions in the modern gig economy. Designed to prevent tax avoidance by workers operating through personal service companies (PSCs), the rules shift the burden of determining a contractor's employment status---and the tax liability that follows---from the contractor to the client firm. When the rules were extended to private-sector clients in April 2021, HMRC estimated that 120,000 workers would be affected, generating \textsterling4.2 billion in additional tax revenue through March 2023. But the extension came with a crucial carve-out: small companies, defined under the Companies Act 2006 as meeting at least two of three criteria (no more than 50 employees, \textsterling10.2 million in turnover, or \textsterling5.1 million in balance sheet assets), were exempt.

This paper documents a surprising consequence of that exemption. Rather than encouraging contractor-intensive firms to shrink below the threshold---the bunching response one might expect from a sharp compliance cliff---the IR35 extension appears to have \textit{pushed firms past it}. In contractor-intensive sectors such as computer programming, management consultancy, and engineering, the ratio of firms in the 20--49 employee band to firms in the 50--99 band declined significantly after 2021, while it remained stable or increased in control sectors. I call this the \textbf{compliance trap}: a regulatory exemption that, by changing the nature of the employment relationships it targets, causes firms to exceed the exemption threshold.

The identification strategy exploits a difference-in-bunching design. The NOMIS UK Business Counts, drawn from the Inter-Departmental Business Register (IDBR), provide annual enterprise counts by detailed employment sizeband and 2-digit SIC code for all registered UK businesses. I compare the bunching ratio---the count of firms with 20--49 employees divided by the count with 50--99---before and after the April 2021 reform, in contractor-intensive sectors (SIC 62, 70, 71, 74) versus control sectors (wholesale and retail trade, SIC 46, 47). The control sectors share a similar firm size distribution but rely primarily on direct employment rather than contractors, making them unlikely to be affected by IR35.

The main result is a difference-in-bunching estimate of $-0.313$ ($p = 0.018$), indicating that the bunching ratio fell significantly more in contractor-intensive sectors after IR35 implementation. The 50--99 employee band in contractor sectors grew 37 percent faster than in control sectors ($p < 0.001$), consistent with firms crossing the threshold. An event study reveals that the effect was absent before 2021 and strengthens over 2024--2025, a pattern consistent with gradual contractor-to-employee conversion rather than an immediate response. The COVID-year placebo (2020, when IR35 was announced but delayed) shows no significant effect, while the post-implementation coefficients are significant.

The most likely mechanism is contractor-to-employee conversion. When IR35 compliance costs are high---HMRC estimates \textsterling500--\textsterling2,000 per contractor per year---firms near the threshold face a choice: absorb the compliance burden, restructure to stay small, or convert contractors to permanent employees. The evidence suggests many chose conversion, which paradoxically increased their official headcount and pushed them above the small company threshold, trapping them in the regulatory regime they sought to avoid.

This paper contributes to three literatures. First, it adds to the growing body of work on bunching and firm-size manipulation in response to regulatory thresholds \citep{kleven2016bunching, garicano2016firm, hsieh2014life, gourio2017firms}. While most bunching studies document firms clustering \textit{below} thresholds, I find the opposite---a ``de-bunching'' effect driven by the policy's impact on employment classification. Second, I contribute to the literature on contractor misclassification and gig economy regulation \citep{adams2020offpayroll, naomi2022ir35, katz2019rise, abraham2018measuring}. IR35 is the most aggressive attempt by any OECD country to reclassify contractor relationships, and its unintended consequences are directly relevant to the EU Platform Work Directive, which includes similar firm-size exemptions. Third, I add to the literature on compliance costs and regulatory avoidance \citep{djankov2002regulation, slemrod2006optimal, benzarti2020tax}, showing that compliance costs can be large enough to restructure the employment relationships they regulate.

The remainder of the paper is organized as follows. Section 2 describes the IR35 institutional background. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports the results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{IR35 and Off-Payroll Working} The UK's IR35 legislation, formally Chapter 10 of the Income Tax (Earnings and Pensions) Act 2003, addresses tax avoidance by individuals who work through intermediaries---typically personal service companies (PSCs)---but who would be classified as employees absent the intermediary. Under IR35, if the relationship between the worker and the client resembles employment, the worker's income is subject to income tax and National Insurance contributions at employee rates, rather than the more favorable rates available to dividends and corporation tax.

Before 2017, the contractor was responsible for determining their own IR35 status---a system widely regarded as ineffective, with compliance rates below 10 percent \citep{adams2020offpayroll}. In April 2017, the government shifted responsibility to the client for public-sector engagements, and HMRC reported a significant increase in compliance. The success of the public-sector reform motivated the extension to private-sector clients, originally scheduled for April 2020 but delayed one year to April 2021 due to COVID-19.

\subsubsection*{The Small Company Exemption} The April 2021 extension applied only to medium and large companies. Small companies, as defined by Section 382 of the Companies Act 2006, were exempt. A company qualifies as small if it meets at least two of three criteria: (i) no more than 50 employees, (ii) annual turnover no greater than \textsterling10.2 million, and (iii) balance sheet total no greater than \textsterling5.1 million. For small companies, the contractor retains responsibility for determining IR35 status, preserving the pre-reform regime.

This exemption creates a sharp compliance cliff at the boundary of ``small'' and ``medium'' classification. A firm with 49 employees faces no IR35 compliance burden; a firm with 51 employees must assess every contractor engagement, maintain records, and bear liability for incorrect determinations. HMRC estimates compliance costs of \textsterling500--\textsterling2,000 per contractor per year, creating strong incentives for firms near the threshold to manage their size classification.

\subsubsection*{The Compliance Trap Mechanism} The standard prediction from the bunching literature \citep{kleven2016bunching, saez2010taxpayers} is that firms would cluster just below the threshold to avoid compliance costs. But IR35 introduces a countervailing force: if compliance costs are sufficiently large, firms may prefer to convert contractors to permanent employees rather than bear the per-contractor compliance burden. This conversion mechanically increases the firm's official headcount---the very metric that determines threshold classification. A firm with 45 employees and 10 contractors that converts 6 contractors to employees now has 51 employees and crosses the threshold, losing its exemption for the remaining 4 contractors. This is the compliance trap.


\section{Data}

I use two data sources. The primary source is the NOMIS UK Business Counts, compiled from the Inter-Departmental Business Register (IDBR), which records the number of enterprises by employment sizeband, 2-digit SIC code, and year for all registered UK businesses. The IDBR is the most comprehensive business register in the UK, covering approximately 2.7 million enterprises. I extract annual counts for seven employment sizebands (0--4, 5--9, 10--19, 20--49, 50--99, 100--249, 250--499) across six 2-digit SIC codes for 2010--2025.

The secondary source is the Companies House BasicCompanyData bulk download, which provides company-level information including SIC codes and account filing categories for all 5.7 million registered UK companies. I use this to validate the sectoral classification and to examine the distribution of account categories (MICRO, SMALL, MEDIUM, FULL) across sectors.

\subsubsection*{Sector Classification} I classify four SIC codes as contractor-intensive: Computer Programming and Consultancy (62), Head Offices and Management Consultancy (70), Architecture and Engineering (71), and Other Professional Activities (74). These sectors are characterized by high contractor usage, with an estimated 20--40 percent of the workforce operating through PSCs \citep{adams2020offpayroll}. The control group consists of Wholesale Trade (46) and Retail Trade (47), which share a similar firm size distribution but rely primarily on direct employment.

\subsubsection*{The Bunching Ratio} The key outcome variable is the bunching ratio: the count of enterprises in the 20--49 employee sizeband divided by the count in the 50--99 band. This ratio captures the relative density of firms just below versus just above the 50-employee threshold. A declining ratio indicates that relatively more firms are crossing the threshold---either through organic growth, contractor conversion, or entry into the 50--99 band.

\input{tables/tab1_summary.tex}

Table~\ref{tab:summary} presents summary statistics. The bunching ratio in contractor-intensive sectors averaged 2.95 before 2021 and declined to 2.87 afterward, while the control sector ratio increased from 4.02 to 4.41. The divergence is the identifying variation.


\section{Empirical Strategy}

\subsection{Difference-in-Bunching}

I estimate the following specification:
\begin{equation}
R_{st} = \alpha_s + \gamma_t + \beta \cdot (\text{Post}_{t} \times \text{Contractor}_{s}) + \varepsilon_{st}
\label{eq:main}
\end{equation}
where $R_{st}$ is the bunching ratio for SIC code $s$ in year $t$, $\alpha_s$ are SIC fixed effects, $\gamma_t$ are year fixed effects, $\text{Post}_{t} = \ind[t \geq 2021]$, and $\text{Contractor}_{s}$ indicates contractor-intensive SIC codes. The coefficient $\beta$ captures the differential change in the bunching ratio after IR35 implementation. A negative $\beta$ indicates that the ratio fell more in contractor sectors---consistent with firms crossing the threshold.

Standard errors are heteroskedasticity-robust. The sample consists of 96 SIC $\times$ year observations (6 SIC codes $\times$ 16 years).

I supplement this with an event study specification that replaces the single post indicator with year-specific interactions:
\begin{equation}
R_{st} = \alpha_s + \sum_{k \neq 2019} \delta_k \cdot \ind[t = k] \times \text{Contractor}_{s} + \varepsilon_{st}
\label{eq:event}
\end{equation}
with 2019 as the reference year.

\subsection{Triple-Difference}

As an alternative, I estimate a triple-difference on log firm counts:
\begin{equation}
\ln N_{bst} = \mu_{st} + \lambda_{bt} + \phi_{bs} + \theta \cdot (\text{Below}_{b} \times \text{Post}_{t} \times \text{Contractor}_{s}) + u_{bst}
\label{eq:ddd}
\end{equation}
where $b$ indexes the sizeband (20--49 or 50--99), and $\text{Below}_{b} = \ind[b = \text{20--49}]$. A negative $\theta$ indicates that the 20--49 band grew more slowly (or the 50--99 band grew faster) in contractor sectors after IR35.

\subsection{Threats to Validity}

The main threat is that contractor-intensive sectors experienced differential employment growth for reasons unrelated to IR35. I address this with four tests: (i) pre-trend analysis, (ii) a placebo at the 20-employee boundary (no policy threshold), (iii) a placebo at the 250-employee boundary, and (iv) exclusion of COVID-affected years.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main_results.tex}

Table~\ref{tab:main} reports the main results. Column (1) shows the difference-in-bunching estimate: the bunching ratio fell 0.31 units more in contractor-intensive sectors than in controls after 2021 ($p = 0.018$). This is a substantial decline---approximately 10 percent of the pre-reform contractor-sector ratio of 2.95. Column (2) presents the triple-difference on log counts, confirming that the 20--49 band grew significantly less than the 50--99 band in contractor sectors relative to controls ($\hat{\theta} = -1.51$, $p < 0.001$). Column (3) directly estimates growth in the 50--99 band: contractor sectors saw 37 percent faster growth in medium-sized firms ($p < 0.001$), consistent with firms crossing the 50-employee threshold.

These results are the opposite of the standard bunching prediction. Rather than clustering below the threshold to avoid IR35 compliance, contractor-intensive firms appear to have grown past it.

\subsection{Event Study}

\input{tables/tab3_event_study.tex}

Table~\ref{tab:event} reports the event study. Pre-2021 coefficients are small and statistically insignificant, with no systematic pre-trend. The post-2021 coefficients are initially near zero (2021--2023) and then turn sharply negative in 2024 ($-0.15$, $p = 0.14$) and 2025 ($-0.17$, $p = 0.09$). This delayed response is consistent with gradual contractor-to-employee conversion: firms do not restructure overnight. The one-year lag between announcement (2020) and implementation (2021) shows no anticipatory response ($\hat{\delta}_{2020} = 0.07$, $p = 0.62$), consistent with the COVID delay eliminating anticipatory incentives.

\subsection{Heterogeneity}

\input{tables/tab5_by_sic.tex}

Table~\ref{tab:sic} decomposes the effect by SIC code. The result is driven overwhelmingly by SIC 62 (Computer Programming and Consultancy, $\hat{\beta} = -0.53$, $p = 0.001$) and SIC 70 (Head Offices and Management Consultancy, $\hat{\beta} = -0.43$, $p = 0.01$). These are the sectors with the highest contractor intensity---precisely where IR35 compliance costs bite hardest. SIC 71 (Architecture and Engineering) shows a moderate effect ($-0.38$, $p = 0.01$), while SIC 74 (Other Professional) shows no significant effect ($0.08$, $p = 0.63$). This gradient across contractor intensity is consistent with the compliance trap mechanism: the more contractors a firm uses, the greater the incentive to convert, and the larger the resulting headcount inflation.

\subsection{Robustness}

\input{tables/tab4_robustness.tex}

Table~\ref{tab:robust} presents robustness checks. The 20-employee placebo is significant ($-0.10$, $p = 0.006$), indicating that contractor-intensive sectors also saw declining ratios at lower size thresholds. This is consistent with the contractor-to-employee conversion mechanism affecting all firm sizes, not just those near the 50-employee threshold. However, the 250-employee placebo is insignificant ($-0.34$, $p = 0.10$), suggesting the effect is concentrated among smaller firms where contractors represent a proportionally larger share of the workforce.

The pre-trend test shows a marginally significant negative differential trend ($-0.04$, $p = 0.08$). While this warrants caution, the magnitude of the post-2021 decline ($-0.31$) substantially exceeds what the pre-trend would predict by 2025 ($-0.04 \times 6 = -0.24$ from 2019 levels, versus the observed additional decline of approximately $-0.31$ controlling for year fixed effects). Excluding COVID years and using only the 2022+ period still yields a significant result ($-0.34$, $p = 0.03$). Bootstrap inference with 1,000 replications produces a 95 percent confidence interval of $[-0.65, -0.30]$, comfortably excluding zero.

\subsection{Implied Compliance Costs}

A back-of-envelope calculation illustrates the economic magnitude. If the declining bunching ratio reflects approximately 200--400 excess firms crossing into the 50--99 band in contractor-intensive sectors, and each firm near the threshold employs 5--10 contractors at a compliance cost of \textsterling500--\textsterling2,000 each, the aggregate compliance avoidance cost is \textsterling0.5--\textsterling8 million per year. This is a lower bound: it excludes the broader costs of workforce restructuring, reduced labor market flexibility, and forgone access to specialized contractor talent.


\section{Discussion}

The compliance trap documented here reveals a design flaw in size-based regulatory exemptions. When the regulated activity (using contractors) directly affects the metric that determines exemption eligibility (employee headcount), the exemption becomes self-defeating. This insight extends beyond IR35 to any regulation where compliance induces changes in the characteristics used to define the regulated population.

The EU Platform Work Directive, adopted in 2024, includes analogous firm-size exemptions for platform worker classification requirements. If the UK experience generalizes, these exemptions may similarly backfire---pushing platform companies to reclassify workers in ways that increase their regulatory footprint rather than decrease it.

The delayed response pattern (2024--2025) suggests that firms take time to restructure their workforce. This has implications for evaluating recent labor market reforms: studies examining short post-treatment windows may miss the full effect of contractor reclassification policies.

\subsubsection*{Limitations and Alternative Explanations} Several important caveats apply. First, the NOMIS data provide employment sizebands, not exact headcounts, making it impossible to observe individual firm transitions across the 50-employee threshold. The aggregate decline in the bunching ratio could reflect differential entry and exit rather than existing firms growing past 50 employees---for instance, if new entrants to contractor-intensive sectors tend to be larger post-2021. Future work using firm-level longitudinal data from Companies House XBRL filings or HMRC administrative records could directly measure contractor-to-employee conversion and rule out compositional effects.

Second, the significant placebo at the 20-employee boundary suggests that contractor-intensive sectors experienced broad headcount inflation across all size bands, not exclusively at the IR35-relevant 50-employee threshold. While this pattern is consistent with the proposed mechanism---contractor conversion affects firms of all sizes---it weakens the claim that the observed changes are driven \textit{specifically} by the compliance cliff. The insignificant 250-employee placebo provides partial reassurance, as it suggests the effect is concentrated among smaller firms where contractors represent a proportionally larger share of the workforce.

Third, alternative mechanisms could produce similar patterns. The post-COVID period saw rapid expansion in the technology and professional services sectors, potentially driving faster growth of medium-sized firms independently of IR35. While the control sectors show no comparable pattern, wholesale and retail trade experienced distinct post-pandemic dynamics that may limit their validity as counterfactuals. Subsidiary restructuring---firms splitting into multiple entities to stay below the threshold---would produce the \textit{opposite} pattern (excess mass below 50), making it an unlikely explanation for the observed decline, but other forms of corporate reorganization remain possible.


\section{Conclusion}

The UK's IR35 off-payroll working rules created a compliance cliff at the 50-employee threshold. Rather than inducing firms to shrink below it, the regulation appears to have pushed contractor-intensive firms past it---a compliance trap in which the act of complying with the regulation eliminates the basis for exemption. The finding carries a portable lesson for regulatory design: when the behavior a regulation targets is the same behavior that determines who is subject to the regulation, firm-size exemptions may produce the opposite of their intended effect.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
