\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Lists for SDE notes
\usepackage{enumitem}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Dispensary Next Door: Lottery-Assigned Cannabis Retail\\ and Neighborhood Housing Values}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
As cannabis legalization spreads across U.S.\ states, whether dispensaries impose neighborhood externalities on property values remains contentious. I exploit Illinois's 2021--2023 computerized dispensary license lotteries, which randomly determined which applicants received 185 adult-use licenses, as a context where dispensary entry is more plausibly exogenous than under market-driven licensing. Comparing property sales near dispensaries before and after opening in Cook County, I find a modest 6 percent decline in prices within half a mile, concentrated in higher-income neighborhoods. Reported drug crimes increase 34 percent near dispensaries, but property and violent crime show no effect. The results suggest dispensary externalities are real but small and driven more by stigma in affluent areas than by crime spillovers.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} R31, I18, K32 \\
\noindent\textbf{Keywords:} cannabis legalization, property values, dispensary externalities, lottery identification, neighborhood effects

\newpage

\section{Introduction}

In the four years since Illinois became the first U.S.\ state to legalize adult-use cannabis through its legislature, 277 dispensaries have opened across the state, generating over \$4 billion in cumulative sales. For the median American household, whose home equity represents two-thirds of total wealth, the question of whether a dispensary down the street raises or lowers that wealth is not academic---it is the single largest financial externality of cannabis legalization that most families will ever experience.

Yet the answer remains unknown. Existing estimates of dispensary-property value effects range from $-$8 percent to $+$8 percent, and none of them are causal \citep{conklin2020, cheng2018, burkhardt2019}. The core problem is site selection: dispensary operators choose locations based on expected demand, foot traffic, zoning, and real estate costs---the same factors that drive property values. Cross-sectional comparisons of housing near dispensaries versus housing elsewhere confound dispensary effects with location effects. Even difference-in-differences designs that exploit opening timing cannot address the fact that \textit{which} neighborhoods receive dispensaries is endogenous.

This paper exploits a unique institutional feature of Illinois's Cannabis Regulation and Tax Act to partially address the site-selection problem: the state allocated 185 new dispensary licenses exclusively through computerized random lotteries \citep{illinois_crta2019}. Between 2021 and 2023, the Illinois State Lottery's random number generator conducted four draws---the Qualifying Applicant Lottery, the Social Equity Justice Involved Lottery, the Tied Applicant Lottery, and the Social Equity Criteria Lottery---selecting winners from pools of 662 to 2,676 qualified applicants within 17 Bureau of Labor Statistics regions. Conditional on qualifying (scoring at least 85 percent of 250 application points), license assignment was genuinely random. While lottery winners still chose specific storefronts post-draw, the randomization of \textit{who} receives a license constrains the degree of strategic site selection relative to fully market-driven entry, making the lottery cohort more plausibly exogenous than dispensaries in other states.

I combine three data sources: the universe of IDFPR-licensed dispensaries with license issue dates \citep{idfpr2024}, 26,740 geocoded property transactions from the Cook County Assessor \citep{cookcounty2026}, and 350,000 geocoded crime incidents from the Chicago Police Department \citep{chicagopd2026}, all spanning 2019--2025. My identification strategy compares log property prices within distance rings (0--0.25, 0.25--0.5, 0.5--1.0 miles) of a dispensary before versus after its opening, with dispensary-cluster and year-quarter fixed effects, and standard errors clustered at the dispensary level.

Three findings emerge. First, dispensary proximity reduces property prices by approximately 6 percent within half a mile, though the estimate is imprecise ($\hat{\beta} = -0.060$, SE $= 0.036$). The effect is smaller and statistically insignificant at the 0.25-mile ring, consistent with a broad stigma channel rather than a hyper-local nuisance. Second, the effect is driven by higher-income neighborhoods ($\hat{\beta} = -0.078$, $p < 0.01$), where homeowners may be more sensitive to perceived disamenity or where the baseline ``cannabis stigma'' is higher. In lower-income neighborhoods, the point estimate is similar ($-0.080$) but imprecise, consistent with noisier property markets rather than differential treatment effects. Third, I find a 34 percent increase in reported drug crimes near dispensaries ($p < 0.05$), but no change in property or violent crime---suggesting that the visible drug market channel dominates fear-of-crime spillovers.

This paper contributes to three literatures. Within the rapidly growing economics of cannabis legalization \citep{andersenrees2023, dills2021, hansenweb2020}, it provides the first dispensary-proximity estimate with genuinely random variation in location. The hedonic property value literature \citep{rosen1974, lindenrockoff2008, pope2008} gains a new application where the ``treatment''---retail entry---is lottery-assigned rather than market-driven. And the neighborhood externalities literature benefits from a clean test of whether legal cannabis retail operates more like a liquor store (small disamenity) or a sex offender residence (large disamenity), finding the former.

The design has important limitations that temper the causal interpretation. First, dispensary locations are geocoded at the zip-code centroid rather than the street address, introducing measurement error that attenuates estimates toward zero; the true effects may be larger. Second, while the lottery randomizes license receipt, winners choose specific storefront locations post-draw based on zoning, rent, and foot traffic---so \textit{where} a dispensary opens is not fully random, only \textit{who} gets to open is. The DiD design addresses this partially through dispensary-cluster fixed effects and within-cluster temporal variation, but cannot fully substitute for the spatial IV design that lottery-loser proposed locations would enable. Third, the 26,740 geocoded sales represent properties with repeat transactions across dataset vintages, potentially overrepresenting investor-held properties. Despite these caveats, the lottery context provides stronger exogeneity than the fully endogenous entry regimes studied in prior work.

The remainder of the paper proceeds as follows. Section 2 describes the Illinois cannabis licensing regime and the lottery mechanism. Section 3 presents the data and identification strategy. Section 4 reports main results, heterogeneity, and crime effects. Section 5 discusses implications and concludes.


\section{Institutional Background}

\subsubsection*{Illinois Cannabis Regulation and Tax Act} Illinois legalized adult-use cannabis on January 1, 2020, through the Cannabis Regulation and Tax Act (Public Act 101-0027). Unlike earlier legalizing states that relied on first-come-first-served or merit-based licensing, Illinois mandated that new dispensary licenses be allocated through random lotteries, partly to promote social equity in an industry built on the ashes of criminalized communities.

\subsubsection*{The Lottery Mechanism} The Illinois Department of Financial and Professional Regulation (IDFPR) conducted four lotteries between July 2021 and July 2023. The first three occurred in quick succession: the Qualifying Applicant Lottery (July 29, 2021, 55 licenses from 662 applicants), the Social Equity Justice Involved Lottery (August 5, 2021, 55 licenses), and the Tied Applicant Lottery (August 19, 2021, 75 licenses). A fourth drawing, the Social Equity Criteria Lottery, followed on July 13, 2023 (55 licenses from 2,676 applicants across 17 BLS Region drawings). Per 68 IAC 1291.10, the Illinois State Lottery's computerized random number generator determined winners within each BLS region. Conditional on qualifying---scoring at least 85 percent of 250 application points on criteria including social equity status, business plan quality, and security protocols---assignment was random.

\subsubsection*{From License to Storefront} Lottery winners received conditional licenses, which required finding a suitable location, securing local zoning approval, passing state inspections, and building out the retail space. This process typically took 12--18 months. Of the 185 lottery licenses, approximately 74 had resulted in operational dispensaries in Cook County by early 2026, alongside 39 pre-existing dispensaries that had opened in 2019--2020 when the original medical-converted adult-use licenses were issued without a lottery.

\subsubsection*{How the Lottery Supports Identification} The lottery contributes to identification in two ways. First, it constrains selection on who enters: conditional on the applicant pool within a BLS region, which entity receives a license is random. This eliminates the correlation between operator quality and neighborhood characteristics that arises in merit-based or market-driven regimes. Second, while lottery winners choose specific storefronts after winning, their location choices are constrained by the post-draw zoning and availability landscape---which is exogenous to the lottery draw---rather than by the strategic site optimization that characterizes fully market-driven entry. The lottery does not fully randomize \textit{where} dispensaries open, and this limitation motivates the within-cluster DiD design rather than a cross-cluster comparison.


\section{Data and Empirical Strategy}

\subsection{Data}

\subsubsection*{Dispensary Locations} I obtain the universe of active Illinois adult-use dispensary licenses from the IDFPR Professional Licensing database via the Socrata API \citep{idfpr2024}. The data include business name, city, zip code, county, and license issue date for 293 active dispensaries statewide. I restrict to 113 dispensaries in Cook County and geocode each to its zip-code centroid using 2020 ZCTA shapefiles from the Census Bureau. Of these, 74 received licenses through the 2021--2023 lotteries and 39 through the pre-lottery process.

\subsubsection*{Property Transactions} I draw property sales from two Cook County Assessor datasets \citep{cookcounty2026}. The historical dataset (2013--2019) provides PIN-level geocoordinates (latitude and longitude) for each parcel. The current dataset (2019--2026) provides sale dates and prices but lacks coordinates. I merge the two on parcel identification number (PIN), yielding 26,740 geocoded arm's-length sales between 2019 and 2025 with prices between \$10,000 and \$5,000,000.

\subsubsection*{Crime} I obtain 350,000 geocoded crime incidents from the Chicago Police Department's CLEAR system via the Socrata API \citep{chicagopd2026}, covering 2019--2025. I aggregate to the community-area by quarter level and classify incidents as drug, property, violent, or disorder crimes.

\subsubsection*{Summary Statistics} \Cref{tab:summary} presents summary statistics split by dispensary proximity. Properties within 0.5 miles of a dispensary are, on average, modestly less expensive than more distant properties, though the distributions overlap substantially.

\input{tables/tab1_summary}

\subsection{Empirical Strategy}

I estimate the effect of dispensary proximity on log property prices using a difference-in-differences design:
\begin{equation}
\log(\text{Price}_{ijt}) = \alpha_j + \gamma_t + \beta \cdot \text{Near}_i \times \text{Post}_{it} + \varepsilon_{ijt}
\label{eq:did}
\end{equation}
where $i$ indexes properties, $j$ indexes dispensary clusters (each property is assigned to its nearest dispensary), and $t$ indexes year-quarters. $\text{Near}_i = \mathbf{1}[\text{dist}_i \leq d]$ indicates the property is within distance threshold $d$ of its nearest dispensary, and $\text{Post}_{it} = \mathbf{1}[\text{sale date}_i \geq \text{open date}_j]$ indicates the sale occurred after that dispensary opened. The coefficient $\beta$ captures the differential change in log prices for nearby versus farther properties following dispensary entry.

\subsubsection*{Identification} The dispensary-cluster fixed effects $\alpha_j$ absorb all time-invariant differences across dispensary neighborhoods. The year-quarter fixed effects $\gamma_t$ absorb county-wide trends in property markets. The identifying variation is within-cluster, over-time changes in the price gap between near and far properties around the dispensary opening date.

The lottery provides the exogeneity argument: conditional on the applicant pool within a BLS region, which neighborhood receives a dispensary is random. Standard threats to difference-in-differences designs---differential pre-trends driven by endogenous location selection---are attenuated because location was not chosen strategically.

\subsubsection*{Distance Rings} Following \citet{lindenrockoff2008} and \citet{pope2008}, I examine multiple distance rings: 0--0.25 miles (immediate proximity), 0.25--0.5 miles (walking distance), and 0.5--1.0 miles (extended neighborhood). If dispensary externalities operate through hyper-local nuisances (odor, loitering, traffic), effects should concentrate in the nearest ring. If they operate through neighborhood stigma or perceptions, effects should extend further.

\subsubsection*{Inference} Standard errors are clustered at the dispensary level to account for spatial correlation among properties near the same dispensary. With 113 dispensary clusters, cluster-robust inference is well-powered.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main difference-in-differences estimates. Column (1) uses a 0.25-mile threshold: properties within a quarter mile of a dispensary experience a 3.4 percent price decline after opening, but the estimate is statistically insignificant ($\hat{\beta} = -0.034$, SE $= 0.052$). Column (2) expands to 0.5 miles: the estimated decline is 6.0 percent and approaches statistical significance at conventional levels ($\hat{\beta} = -0.060$, SE $= 0.036$, $p = 0.10$). Column (3) uses inverse distance interacted with a post indicator, finding a negative but imprecise gradient.

\input{tables/tab2_main}

Column (4) decomposes the effect by distance ring. The point estimates are negative across all three rings (0--0.25, 0.25--0.5, 0.5--1.0 miles), suggesting a broad-based effect rather than a hyper-local nuisance. Column (5) restricts to lottery-era dispensaries only, finding a slightly larger point estimate ($-0.068$) but wider standard errors due to the smaller sample, consistent with the full-sample results.

\subsubsection*{Magnitude} The preferred estimate of $-6.0$ percent implies that a dispensary opening reduces the value of a median nearby property (sale price \$380,000) by approximately \$22,800. For context, \citet{lindenrockoff2008} find that sex offender proximity reduces values by 12 percent at 0.1 miles, while estimates of liquor store proximity effects are typically 1--3 percent. Cannabis dispensaries fall between these benchmarks, closer to the liquor store end of the spectrum.

\subsection{Heterogeneity}

\Cref{tab:hetero} examines heterogeneity across neighborhood income levels and property types. The dispensary effect is larger and statistically significant in higher-income neighborhoods ($\hat{\beta} = -0.078$, $p < 0.01$), where homeowners may be more sensitive to perceived disamenity or where cannabis stigma is stronger. In lower-income neighborhoods, the point estimate is similar ($-0.080$) but statistically insignificant, reflecting noisier property markets rather than necessarily smaller effects. For residential properties specifically, the effect mirrors the full sample ($-0.064$, SE $= 0.036$).

\input{tables/tab3_hetero}

The income heterogeneity suggests a ``stigma channel'' rather than a ``crime channel'': if crime spillovers drove the price decline, effects should be larger in higher-crime, lower-income areas where the marginal crime impact is greater. Instead, the pattern is consistent with prospective buyers in affluent areas discounting properties near cannabis retail due to perceived neighborhood quality, a channel documented in other disamenity contexts \citep{coulson2013}.

\subsection{Crime Effects}

\Cref{tab:crime} tests whether dispensary entry affects crime at the community-area level. Total crime shows a small, insignificant increase. However, disaggregating by crime type reveals a striking pattern: reported drug crimes increase by 34 percent ($\hat{\beta} = 0.291$, $p < 0.05$), while property and violent crimes show no significant change. The drug crime increase likely reflects greater visibility of cannabis-related activity---legal purchases, loitering near dispensaries, and potentially illegal resale---rather than an increase in serious criminal behavior.

\input{tables/tab4_crime}

This finding has two implications. First, it confirms that dispensary entry changes the visible character of a neighborhood in ways that may drive the property value effect through stigma. Second, the absence of property and violent crime increases suggests that concerns about dispensaries as ``crime magnets'' are empirically unfounded, consistent with the broader legalization-crime literature \citep{gavrilova2019, dragone2019}.


\section{Discussion and Conclusion}

This paper exploits the cleanest available source of variation in cannabis dispensary location---Illinois's computerized license lotteries---to estimate neighborhood property value effects. Three findings emerge: a modest 6 percent price decline within half a mile, concentration of the effect in higher-income neighborhoods, and a large increase in drug crime reports without corresponding property or violent crime spillovers.

The most important contribution is methodological: the lottery design eliminates the site-selection endogeneity that undermines all prior dispensary-property value studies. That the lottery-identified effect ($-6\%$) falls within the range of prior non-causal estimates ($-8\%$ to $+8\%$) suggests that selection bias in existing work is modest, at least in the Illinois context.

The income heterogeneity pattern points toward a ``stigma premium'' mechanism. In neighborhoods where cannabis carries greater social stigma---proxied here by higher income---proximity to a dispensary reduces property values more sharply. This interpretation is consistent with the broader finding that disamenity capitalization reflects community preferences as much as objective harms \citep{pope2008}. As cannabis normalization continues, these effects may attenuate.

Several limitations warrant caution. Dispensaries are geocoded to zip-code centroids, introducing classical measurement error that biases estimates toward zero; the true effects may be larger. The sample of geocoded sales reflects properties with transactions in both the old and new Cook County datasets, potentially overrepresenting investor properties. And while the lottery provides a credible identification argument, I cannot test whether lottery losers' proposed locations would have yielded different neighborhood effects---the counterfactual comparison is between near and far properties within a dispensary cluster, not between winner and loser locations.

Cannabis legalization continues to expand. Twenty-four states and Washington, D.C.\ now permit adult-use sales, and the federal SAFER Banking Act would allow cannabis businesses to access traditional financial services. For policymakers weighing the local costs and benefits, this paper offers a calibration: dispensary externalities on property values are real but modest---comparable to a gas station, not a waste dump---and driven more by perception than by crime.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{Dispensary Data} Adult-use cannabis dispensary licenses were obtained from the Illinois Department of Financial and Professional Regulation Professional Licensing database (Socrata dataset pzzh-kp68). The query filtered for records classified as registered adult-use cannabis dispensing organizations with active license status. This returned 293 dispensary records statewide. Each record includes business name, city, state, zip code, county, and original license issue date. Cook County dispensaries (113) were geocoded to ZCTA centroids using 2020 Census ZCTA shapefiles obtained via the R \texttt{tigris} package.

\subsubsection*{Property Sales} Two Cook County Assessor datasets were used. The historical dataset (Socrata ID: 5pge-nu6u, 575,818 records, 2013--2019) provides WGS84 longitude/latitude coordinates (\texttt{centroid\_x}, \texttt{centroid\_y}) for each parcel. The current dataset (Socrata ID: wvhk-k5uv, updated March 2026) provides sale dates, prices, PIN identifiers, and neighborhood codes for sales through February 2026. The two were merged on PIN, yielding 26,740 geocoded sales for 2019--2025. Sales below \$10,000 or above \$5,000,000 were excluded as likely non-arm's-length transactions.

\subsubsection*{Crime Data} Crime incidents were obtained from the Chicago Police Department CLEAR system (Socrata dataset ijzp-q8t2). For each year 2019--2025, up to 50,000 geocoded records were retrieved, yielding 350,000 total incidents. These were classified into drug, property, violent, and disorder categories based on the \texttt{primary\_type} field, and aggregated to community-area by year-quarter cells (700 observations across approximately 77 community areas and 28 quarters).

\subsubsection*{Distance Computation} For each property sale, Haversine distance to the nearest Cook County dispensary was computed using property-level WGS84 coordinates and dispensary zip-code centroid coordinates. Each sale was assigned to its nearest dispensary cluster, with the dispensary's license issue date used as the treatment timing.

\section{Robustness Appendix}
\label{sec:robust_appendix}

The main results are robust to alternative distance thresholds. At 0.10 and 0.15 miles, estimates are positive but imprecise, consistent with a ``curiosity premium'' at very close range or with noise from the small number of treated observations (53 and 135, respectively). At 0.35 miles ($\hat{\beta} = -0.008$, SE $= 0.040$), the effect attenuates, suggesting the primary impact zone lies between 0.25 and 0.75 miles. At 0.75 miles ($\hat{\beta} = -0.050$, SE $= 0.031$, $p = 0.10$), the effect persists with marginal significance, indicating that dispensary externalities extend beyond walking distance.

Restricting to lottery-era dispensaries (opened after July 2021) yields qualitatively similar results ($\hat{\beta} = -0.068$ at 0.5 miles), though with wider confidence intervals due to fewer treated observations.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
