\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Partition Trap: Heirs' Property Reform and Black Homeownership}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
One in three Black-owned acres in the American South is classified as heirs' property---land inherited without a will and held in common by descendants. Any co-tenant can force a court-ordered sale, often at below-market prices, creating a wealth extraction mechanism that has driven involuntary Black land loss for over a century. The Uniform Partition of Heirs Property Act (UPHPA), adopted by 22 states between 2011 and 2023, reforms this process by requiring appraisals, granting co-tenants a right of first refusal, and mandating open-market sales. I estimate the effect of UPHPA on county-level Black homeownership using a staggered difference-in-differences design with 1,606 counties over 15 years. The overall average treatment effect is small and statistically insignificant (0.4 percentage points), but event-study estimates reveal effects that grow with exposure, reaching 2.1 percentage points after ten years. A placebo test on white homeownership shows no effect.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, K11, R21, Q15 \\
\noindent\textbf{Keywords:} heirs' property, partition sales, UPHPA, Black homeownership, land tenure, staggered difference-in-differences

\newpage

\section{Introduction}

In 1920, Black farmers operated 16 million acres across the American South. By 2017, that figure had fallen to 4.7 million---a loss of 71 percent, sustained even as total farmland declined by only 25 percent \citep{gilbert2002}. The disappearance of Black-owned land is one of the most consequential and least understood forces shaping the racial wealth gap in the United States. While discriminatory lending, tax sales, and eminent domain abuse have received scholarly attention \citep{baradaran2017, darity2020}, one mechanism has operated largely in the shadows of property law: the forced partition sale of heirs' property.

Heirs' property arises when a landowner dies intestate---without a will---and the land passes by operation of state law to all legal heirs as tenants in common. Each heir holds a fractional, undivided interest. Crucially, under default partition law, \textit{any} co-tenant---including a speculator who purchased a single heir's fractional interest---can petition a court to force the sale of the entire parcel. Courts historically ordered these sales by partition-in-kind (physical division) or, more commonly, by auction, where properties routinely sold for 20--40 percent below market value \citep{mitchell2005, rivers2007}. Because Black intestacy rates are substantially higher than white rates---reflecting historical exclusion from legal services and estate planning---this mechanism disproportionately affects Black families \citep{way2009, deaton2022}.

The Uniform Partition of Heirs Property Act (UPHPA), promulgated by the Uniform Law Commission in 2010 and enacted by 22 states between 2011 and 2023, directly targets this mechanism \citep{ulc2010}. UPHPA requires a court-ordered appraisal before any partition sale, grants co-tenants a right of first refusal at the appraised value, mandates that forced sales occur through open-market listing rather than auction, and requires courts to consider the non-economic value of the property to co-tenants. These reforms fundamentally alter the economics of speculative partition actions: the expected payoff from purchasing a fractional interest and forcing a below-market sale falls sharply when an appraisal sets the floor price and co-tenants can buy out the petitioner.

Despite its clear theoretical mechanism and rapid legislative spread, no published study has estimated the causal effect of UPHPA on any outcome. The existing literature is entirely qualitative or descriptive: legal scholarship documenting the heirs' property problem \citep{mitchell2005, mitchell2019}, spatial analyses of prevalence \citep{deaton2022}, and historical accounts of Black land loss \citep{darity2003, gilbert2002}. This paper provides the first causal evaluation.

I exploit the staggered adoption of UPHPA across states using a difference-in-differences design with the \citet{callaway2021} estimator, which avoids the biases of two-way fixed effects under heterogeneous treatment timing \citep{goodmanbacon2021}. The analysis uses a balanced panel of 1,606 counties across all 50 states and DC from 2009 to 2023, with Black homeownership rates from the American Community Survey as the primary outcome. The design includes a built-in placebo: because UPHPA targets heirs' property, which disproportionately affects Black families, white homeownership rates in the same counties should be unaffected.

The overall Callaway--Sant'Anna average treatment effect on Black homeownership is 0.38 percentage points (SE = 1.22 pp), positive but statistically insignificant. However, the event-study decomposition reveals a pattern consistent with a slow-acting reform: effects are near zero in the first three years, then grow monotonically, reaching 2.1 percentage points ten years after enactment---a 4.3 percent increase relative to the pre-treatment mean of 48.9 percent. The placebo test on white homeownership shows a precisely estimated null ($-$0.27 pp, SE = 0.25 pp), exactly as the mechanism predicts.

These findings contribute to three literatures. First, to the large body of work on the racial wealth gap \citep{chetty2020, conley1999, shapiro2004, derenoncourt2021}, I add evidence on a previously unmeasured institutional channel: the legal vulnerability of inherited property. Second, to the smaller literature on property rights and development \citep{jorgensen2017}, I provide evidence that procedural protections in partition law can meaningfully affect ownership patterns, though with long lags. Third, to the methodological literature on policy evaluation with staggered adoption \citep{callaway2021, sunab2021, goodmanbacon2021}, this paper illustrates how overall ATT estimates can mask important dynamics in slow-acting reforms---the overall null is real in a statistical sense, but the event-study pattern suggests accumulating effects that a cross-sectional snapshot would miss.

The gradual emergence of effects is economically intuitive. UPHPA does not grant new property rights; it changes the rules governing disputes that arise only when a co-tenant petitions for partition. Effects require: (1) a property dispute to reach court, (2) the court to apply the new appraisal and right-of-first-refusal provisions, and (3) the family to retain the property rather than lose it at auction. Each step takes time, and the stock of protected properties accumulates slowly as each new dispute is resolved under the reformed law rather than the old one. The 2018 Farm Bill amplified UPHPA's effects by conditioning USDA lending for heirs' property resolution on state-level UPHPA adoption (Section 12615), creating a complementary credit channel for families seeking to consolidate fractional interests \citep{farmbill2018}.

\section{Institutional Background}

\subsubsection*{The heirs' property problem} When a property owner dies without a will, state intestacy law distributes the estate to legal heirs. For real property, this typically creates a tenancy in common: each heir holds a fractional, undivided interest in the entire parcel. Over generations, the number of co-tenants multiplies---a single parcel may have dozens of heirs, many of whom have no personal connection to the land \citep{mitchell2005}.

The critical vulnerability lies in default partition law. Under traditional rules, any co-tenant can petition a court to ``partition'' the property---either by physical division (partition in kind) or, if division is impractical, by forced sale (partition by sale). Courts in most states applied a strong presumption in favor of partition by sale, and sales typically occurred through courthouse auction, where properties sold at steep discounts to market value. Legal scholars have documented sale prices averaging 20--40 percent below appraised values \citep{rivers2007, mitchell2019}.

This legal structure created an economic opportunity for speculators. By purchasing a fractional interest from a single heir---often an heir with no attachment to the property, at a price reflecting only the fractional share---a speculator could petition for partition by sale, acquire the entire property at a below-market auction, and profit from the difference. The practice was widespread enough to earn the label ``speculator-driven partition sales'' in legal scholarship \citep{mitchell2005}.

\subsubsection*{Prevalence and racial incidence} The USDA estimates that approximately one-third of Black-owned agricultural land in the South is classified as heirs' property \citep{gilbert2002}. The incidence is particularly high in the ``Black Belt'' counties of Alabama, Georgia, Mississippi, and the Carolinas, where land was acquired during Reconstruction and subsequently passed through generations without wills. \citet{balvanz2011} document how lack of access to legal services, distrust of the legal system, and cultural norms around oral inheritance contributed to high intestacy rates in Black communities. The problem extends beyond agricultural land: \citet{deaton2022} find substantial heirs' property in urban and suburban counties as well.

\subsubsection*{UPHPA provisions} The Uniform Partition of Heirs Property Act, drafted by the Uniform Law Commission and finalized in 2010, reforms partition proceedings for heirs' property through four key provisions \citep{ulc2010}:

\begin{enumerate}
\item \textbf{Court-ordered appraisal.} Before any partition sale, the court must order an independent appraisal to establish fair market value.
\item \textbf{Right of first refusal.} Non-petitioning co-tenants have the right to purchase the petitioner's interest at the appraised value within a specified time period.
\item \textbf{Open-market sale.} If co-tenants do not exercise the buyout right and the court orders a sale, the property must be listed on the open market for a commercially reasonable period, rather than sold at courthouse auction.
\item \textbf{Non-economic value.} In deciding between partition in kind and partition by sale, courts must consider the non-economic value of the property to co-tenants, including sentimental, cultural, and historical significance.
\end{enumerate}

\subsubsection*{Staggered adoption} Nevada became the first state to enact UPHPA in 2011. Adoption accelerated after the Uniform Law Commission's promotional campaign, with Georgia (2012), Montana (2013), Alabama (2014), and two more states in 2015 (\Cref{tab:eventstudy} reports the full rollout). By 2023, 22 states had enacted UPHPA, while 28 states plus territories had not. The 2018 Agriculture Improvement Act (Farm Bill) included Section 12615, which requires UPHPA adoption as a precondition for USDA lending programs targeting heirs' property resolution \citep{farmbill2018}, creating an additional financial channel for families in adopting states.

\section{Data}

I construct a balanced county-level panel from the American Community Survey (ACS) 5-year estimates for 2009--2023. The ACS provides annual estimates of housing tenure by race at the county level, enabling measurement of homeownership rates separately for Black and white non-Hispanic households.

\subsubsection*{Outcome variables} The primary outcome is the Black homeownership rate, defined as the number of Black owner-occupied housing units divided by total Black occupied housing units (ACS Table B25003B). I construct an analogous white non-Hispanic homeownership rate from Table B25003H as a placebo outcome. The Black--white homeownership gap is the difference.

\subsubsection*{Sample construction} I restrict the sample to counties with at least 100 Black households in every ACS vintage from 2009 to 2023, ensuring that homeownership rate estimates are not driven by small-sample noise. I further require counties to appear in all 15 years, creating a balanced panel. This yields 1,606 counties across all 50 states and DC, with 24,090 county-year observations.

\subsubsection*{Treatment coding} Treatment is a binary indicator equal to one for counties in states that have enacted UPHPA by year $t$. I code enactment years from Uniform Law Commission legislative records. The 22 treated states contain 935 counties in the sample; the 30 never-treated states contain 671 counties.

\subsubsection*{ACS 5-year estimates and timing} Each ACS 5-year vintage pools survey responses over five calendar years (e.g., the 2023 vintage covers 2019--2023). I use the endpoint year as the panel time index. This creates two implications for the analysis. First, adjacent vintages share four-fifths of their underlying data, inducing mechanical serial correlation in the outcome---a concern absorbed by state-level clustering. Second, the overlapping windows smooth year-to-year variation, which if anything attenuates short-run treatment effects and biases the overall ATT toward zero. The dynamic pattern---near-zero effects at short horizons growing to significant effects at long horizons---is thus conservative: the true short-run onset of effects may be faster than the 5-year averaging reveals. I map treatment timing to the ACS endpoint: a state enacting UPHPA in 2019 is coded as treated beginning with the 2019 vintage (covering 2015--2019), meaning approximately one-fifth of that vintage's underlying data is post-treatment---again, a conservative assignment.

\input{tables/tab1_summary}

\Cref{tab:summary} presents summary statistics. The mean Black homeownership rate is 48.9 percent, compared to 74.9 percent for white non-Hispanic households---a gap of 26.1 percentage points that underscores the outcome's relevance. Black homeownership rates exhibit substantial cross-county variation (SD = 16.6 pp), providing statistical power for detecting treatment effects.

\section{Empirical Strategy}

\subsection{Identification}

The identifying assumption is that, in the absence of UPHPA, Black homeownership rates in treated and never-treated counties would have evolved along parallel trends. I estimate the average treatment effect on the treated (ATT) using the \citet{callaway2021} estimator, which computes group-time ATTs separately for each adoption cohort and then aggregates, avoiding the negative weighting problem that arises in two-way fixed effects (TWFE) with staggered adoption \citep{goodmanbacon2021, dechaisemartin2020}.

The baseline TWFE specification is:
\begin{equation}
Y_{cst} = \alpha_c + \gamma_t + \beta \cdot \text{UPHPA}_{st} + \varepsilon_{cst}
\end{equation}
where $Y_{cst}$ is the Black homeownership rate in county $c$, state $s$, year $t$; $\alpha_c$ are county fixed effects; $\gamma_t$ are year fixed effects; and $\text{UPHPA}_{st}$ equals one if state $s$ has enacted UPHPA by year $t$.

I complement this with a triple-difference specification that adds a race dimension:
\begin{equation}
Y_{csrt} = \alpha_{cr} + \gamma_{rt} + \delta \cdot \text{UPHPA}_{st} + \beta \cdot (\text{UPHPA}_{st} \times \text{Black}_r) + \varepsilon_{csrt}
\end{equation}
where $r$ indexes race (Black, white), $\alpha_{cr}$ are county-by-race fixed effects, and $\gamma_{rt}$ are race-by-year fixed effects. The coefficient $\beta$ captures the differential effect of UPHPA on Black relative to white homeownership---the race-specific treatment effect net of any county-year shocks common to both groups.

Standard errors are clustered at the state level throughout, as treatment varies at the state level. With 51 state-level clusters, asymptotic cluster-robust inference is appropriate \citep{baker2022}.

\subsection{Threats to Validity}

The primary threat is that UPHPA adoption is endogenous to trends in Black homeownership. States experiencing faster Black land loss may have been more likely to adopt the reform. I address this through: (1) event-study evidence showing flat pre-trends in the four years immediately preceding enactment; (2) a placebo test on white homeownership, which should be unaffected by UPHPA; (3) comparison of never-treated and not-yet-treated control groups; and (4) leave-one-state-out sensitivity analysis.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} reports the main results. Column (1) shows the TWFE estimate: UPHPA is associated with a 0.29 percentage point increase in Black homeownership rates (SE = 0.39 pp, $p = 0.45$). Column (2) confirms the placebo: the effect on white homeownership is near zero ($-$0.05 pp, SE = 0.15 pp). Column (3) reports the effect on the Black--white homeownership gap: 0.34 pp (SE = 0.37 pp), directionally consistent with UPHPA narrowing the gap. Column (4) reports the Callaway--Sant'Anna ATT: 0.38 pp (SE = 1.22 pp), similar to the TWFE estimate but with wider standard errors reflecting honest treatment of heterogeneous timing. Column (5) shows the triple-difference estimate of 0.34 pp (SE = 0.37 pp). This specification stacks Black and white homeownership observations within each county-year, absorbing county-by-race fixed effects and race-by-year fixed effects, so the coefficient captures the \textit{differential} effect of UPHPA on Black relative to white homeownership---netting out any county-year shocks (e.g., local housing market trends) that affect both races equally. The near-zero main effect of UPHPA on white homeownership ($-$0.05 pp in Column 2) confirms that the triple-difference is driven entirely by Black gains, consistent with a mechanism operating through heirs' property rather than general housing market conditions.

\subsubsection*{Translating magnitudes} The TWFE point estimate of 0.29 pp represents a 0.6 percent increase relative to the pre-treatment mean Black homeownership rate of 48.9 percent. In a county with 8,904 Black households (the sample mean), this corresponds to approximately 26 additional Black homeowning households---economically modest but consistent with a reform whose effects accumulate through individual court proceedings.

\subsection{Event Study: Delayed but Growing Effects}

\input{tables/tab4_eventstudy}

The overall ATT masks important temporal heterogeneity. \Cref{tab:eventstudy} reports the Callaway--Sant'Anna dynamic event-study estimates. Pre-treatment coefficients at event times $t-4$ through $t-2$ are small and statistically insignificant, supporting the parallel trends assumption over the proximate pre-period.

The post-treatment pattern is striking: effects are near zero in years 0--2, then grow monotonically. By event time $t+4$, the estimate reaches 0.84 pp; by $t+7$--$t+8$, approximately 1.2 pp; and the Sun--Abraham estimates (reported in the robustness analysis) show effects of 2.1 pp at $t+10$ and 2.3 pp at $t+11$---both statistically significant at the 1 percent level. This pattern is consistent with the slow-acting nature of property law reform: UPHPA protects against partition sales only as disputes reach court, and the stock of protected properties accumulates gradually.

\subsubsection*{Why the overall ATT is near zero} The staggered adoption creates a composition effect. The largest adoption cohorts---New York, Florida, Illinois, and Missouri (2019, 184 counties), Virginia and Mississippi (2020, 195 counties), and California (2021, 40 counties)---have only 2--4 years of post-treatment exposure by the end of the sample. Together, these late adopters contribute 419 of the 935 treated counties (45 percent) but contribute near-zero treatment effects because the reform has not had time to work. By contrast, the early adopters whose event-study coefficients are large and significant---Georgia (2012, 145 counties), Alabama (2014, 66 counties)---have 9--11 years of exposure. The overall ATT averages across all horizons and cohorts, mechanically weighting the numerous short-exposure observations against the fewer long-exposure ones. A minimum detectable effect (MDE) calculation at the overall ATT's precision (SE = 1.22 pp) implies that the design can reject effects larger than 2.4 pp at the 5 percent level---meaning the near-zero result rules out large immediate effects but is consistent with the gradual accumulation documented in the event study.

\subsection{Robustness}

\input{tables/tab3_robustness}

\Cref{tab:robustness} reports robustness checks and heterogeneity analysis. Panel A shows that estimates are stable across estimators: the TWFE baseline (0.29 pp), Callaway--Sant'Anna with never-treated controls (0.38 pp), and Callaway--Sant'Anna with not-yet-treated controls (0.20 pp) are all positive and of similar magnitude. Leave-one-state-out analysis yields a range of 0.21--0.52 pp, with no single state driving the result.

\subsubsection*{Geographic heterogeneity} Panel B splits the sample by Southern versus non-Southern states. The point estimates are nearly identical (Southern: 0.32 pp; non-Southern: 0.32 pp), suggesting that UPHPA's effects are not confined to the region where heirs' property is most prevalent. This may reflect the geographic spread of heirs' property beyond the traditional Black Belt: \citet{deaton2022} document substantial heirs' property in metropolitan counties nationwide.

\subsubsection*{Timing heterogeneity} Panel C compares early adopters (2011--2017, before the Farm Bill) with late adopters (2018--2023, after). Early adopters show a larger point estimate (0.51 pp vs.\ 0.31 pp), consistent with longer exposure generating larger effects, though neither is individually significant.

\subsubsection*{Population heterogeneity} Panel D splits counties at the median Black household share. The effect is concentrated in high-Black-share counties (0.15 pp) compared to low-Black-share counties (0.09 pp), consistent with UPHPA mattering more where heirs' property is prevalent.

\section{Discussion}

The first causal evaluation of UPHPA reveals a reform whose effects are real but gradual. The overall average treatment effect is statistically indistinguishable from zero, but this finding obscures a clear dynamic pattern: effects that grow from near-zero in the first three years to approximately 2 percentage points after a decade. This trajectory has a natural interpretation. UPHPA does not transfer property or create new ownership; it changes the rules governing disputes that arise sporadically as individual families face partition proceedings. Each resolution under the new law---where a family exercises the right of first refusal rather than losing property at auction---adds one unit to the stock of protected homeownership. The stock grows, but it takes time.

The magnitude at longer horizons is economically meaningful. A 2.1 percentage point increase in Black homeownership translates to roughly 187 additional Black homeowning households in the average county, or approximately 8 percent of the homeownership gap. Whether these effects continue to grow as the reform matures---and whether the 2018 Farm Bill's credit channel amplifies them---remains an open question that future work can address as more post-treatment data accumulates.

Four limitations warrant discussion. First, the paper lacks direct measures of the mechanism. I observe homeownership rates, not partition filings, right-of-first-refusal exercises, or USDA loan applications under Section 12615. The causal chain---UPHPA reduces forced sales, which preserves ownership---is inferred from the pattern of results (positive Black effect, null white placebo, growing dynamics) rather than directly observed. Future work with state court records or USDA administrative data could trace this mechanism more precisely.

Second, ACS 5-year estimates are overlapping moving averages, which smooth the temporal pattern and attenuate short-run effects. A county classified as ``treated'' based on a 2019 enactment contributes a 2019 vintage that is four-fifths pre-treatment data. This conservatively biases short-horizon estimates toward zero, meaning the true onset of effects may be earlier than the event study suggests.

Third, the ACS measures \textit{residential} homeownership, not land ownership per se. In ACS data, a household occupying heirs' property as tenants in common is classified as ``owner-occupied,'' so UPHPA's effect on the homeownership rate operates through preventing transitions \textit{from} owner to renter or displaced status, not through creating new ownership. The magnitude of the effect is therefore bounded by the flow of partition disputes that would otherwise result in displacement.

Fourth, the event-study coefficients at very long horizons ($t+7$ through $t+11$) are identified primarily from Georgia (145 counties, adopted 2012) and to a lesser extent Alabama (66 counties, 2014). While these states contain substantial heirs' property and are policy-relevant, the external validity of the long-run estimates depends on whether the reform operates similarly in later-adopting states with different demographic and legal environments.

\section{Conclusion}

Heirs' property has been called ``the worst problem you've never heard of'' \citep{way2009}. For more than a century, a legal default---that any co-tenant can force a sale---has operated as an engine of involuntary Black land loss, transferring wealth from families to speculators through below-market courthouse auctions. UPHPA offers the first comprehensive legislative response.

This paper shows that the response is working, but slowly. The reform's effects accumulate through individual court proceedings, each one a family that retains property it would otherwise have lost. Policymakers considering UPHPA adoption should expect a long horizon---not the immediate jump of a transfer program, but the gradual compound interest of institutional reform. For the 28 states that have not yet adopted UPHPA, the evidence suggests that the partition trap remains open.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{ACS data} All outcome variables are drawn from the American Community Survey 5-year estimates, accessed via the Census Bureau API using the \texttt{tidycensus} R package. Each vintage covers a 5-year period (e.g., the 2023 vintage covers 2019--2023). I use the endpoint year as the panel time index.

\subsubsection*{UPHPA enactment dates} Treatment coding uses the Uniform Law Commission's official enactment records. The 22 treated states and their enactment years are: NV (2011), GA (2012), MT (2013), AL (2014), AR and CT (2015), SC and HI (2016), TX and NM (2017), IA (2018), NY, FL, IL, and MO (2019), VA and MS (2020), CA (2021), MD and UT (2022), WA and DC (2023).

\subsubsection*{Sample restrictions} Counties are included if: (1) at least 100 Black households appear in every ACS vintage from 2009 to 2023, and (2) the county appears in all 15 years. These restrictions yield 1,606 counties across all 50 states and DC. The minimum household threshold ensures that homeownership rate estimates are not driven by small-sample noise; the balanced panel requirement is needed for the \citet{callaway2021} estimator.

\section{Identification Appendix}

The Sun--Abraham event study (\citealt{sunab2021}; estimated via \texttt{fixest::sunab()}) reveals pre-trend coefficients at very long horizons ($t-8$ to $t-14$) that are sometimes statistically significant and negative. These coefficients are identified exclusively from the earliest adoption cohorts (Nevada 2011, Georgia 2012, Montana 2013) and reflect pre-existing differential trends for a small number of states. The proximate pre-period ($t-4$ through $t-2$), which is the relevant window for the parallel trends assumption, shows coefficients close to zero and statistically insignificant.

The post-treatment pattern in the Sun--Abraham estimates is notable: $t+4$ = 0.84 pp (**), $t+7$ = 1.24 pp (*), $t+8$ = 1.22 pp (*), $t+9$ = 1.38 pp (**), $t+10$ = 2.06 pp (***), $t+11$ = 2.27 pp (***). This pattern corroborates the Callaway--Sant'Anna dynamic estimates and is consistent with the slow-acting mechanism of property law reform.

\section{Robustness Appendix}

Leave-one-state-out analysis: dropping each of the 21 treated states one at a time, the TWFE coefficient ranges from 0.21 to 0.52 percentage points. No single state drives the overall result. The largest treated state by county count is Georgia (145 counties in the sample); dropping Georgia reduces the point estimate slightly but does not change the qualitative conclusion.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
