\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{siunitx}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\usepackage{float}

\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Compliance Cliff: Mass Organizational Death and Selective Resurrection in the U.S. Nonprofit Sector}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In June 2011, the IRS automatically revoked the tax-exempt status of 377,409 nonprofit organizations---the largest single organizational death event in U.S. administrative history. Only 10.3\% ever reinstated. Using the complete IRS Auto-Revocation List matched to the Exempt Organizations Business Master File, I document that survival was sharply predicted by organizational type: cemetery companies reinstated at 30.9\%, veterans organizations at 14.4\%, but charitable organizations---63\% of revocations---at only 8.9\%. Organizations holding physical assets reinstated 3.7 percentage points more often than non-asset organizations. This asset premium was larger in the 2010 ``surprise'' wave than in subsequent annual revocations, consistent with an information-asymmetry mechanism. The compliance cliff permanently reshaped the nonprofit sector, disproportionately eliminating social welfare organizations (36.6\% of the subsection revoked) relative to charities (12.8\%).
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} L31, H83, K23 \\
\noindent\textbf{Keywords:} nonprofits, tax-exempt status, regulatory compliance, organizational survival, Pension Protection Act

\newpage

\section{Introduction}

On June 8, 2011, the Internal Revenue Service posted the names of 377,409 nonprofit organizations whose tax-exempt status had been automatically revoked---more organizations than the entire state of Wyoming contains people. In a single administrative action, roughly one in six registered nonprofits ceased to exist as tax-exempt entities. Nine out of ten never came back.

This mass revocation was the mechanical consequence of a rule embedded in the Pension Protection Act of 2006 (PPA). Before the PPA, small nonprofits with gross receipts below \$25,000 had no annual filing obligation. The PPA required all tax-exempt organizations to file at minimum an electronic ``e-Postcard'' (Form 990-N) beginning with tax year 2007 \citep{ppa2006}. Those failing to file for three consecutive years---2007, 2008, and 2009---would automatically lose their exempt status. When the three-year window closed, the IRS executed the rule without discretion.

This paper asks: conditional on revocation, what predicted organizational survival? The answer reveals a striking ``compliance cliff'' that operated regressively across the nonprofit sector. Cemetery companies---which own land and have perpetual obligations---reinstated at 30.9\%. Veterans posts and fraternal lodges, which typically own buildings, reinstated at 12--14\%. But 501(c)(3) charitable organizations, the modal revoked entity at 63\% of the total, reinstated at only 8.9\%. The organizations most associated with public benefit were least likely to survive.

I document these patterns using the complete IRS Auto-Revocation List---1.2 million records spanning 2010--2025---matched to the current Exempt Organizations Business Master File (BMF). The matching identifies which revoked organizations subsequently regained tax-exempt status. The 2010 PPA wave provides the primary sample: 376,472 U.S. organizations, 38,925 of which reinstated. I estimate linear probability models with state fixed effects to quantify the ``asset premium'' in survival: organizations in subsection types that typically hold physical assets reinstated 3.7 percentage points more than non-asset organizations ($t = 11.2$), a premium that narrows to 2.7 pp in post-2010 annual revocations when the filing requirement was well known.

A formal difference-in-differences test confirms the information mechanism: pooling the 2010 wave with 2012--2019 annual revocations and interacting the asset indicator with a 2010 ``surprise'' dummy yields an interaction coefficient of 3.1 percentage points ($t = 8.7$), while the baseline asset effect in non-surprise years is only 0.8 pp and statistically insignificant. The entire asset premium is concentrated in the information-surprise wave.

The findings contribute to three literatures. First, to organizational ecology, which studies population-level patterns of organizational birth and death \citep{hannanfreeman1989, hannanetal2007}. The 2010 mass revocation constitutes a uniquely large exogenous ``mortality shock'' to a defined population, allowing direct observation of how organizational characteristics predict survival after a common compliance failure. Second, to research on regulatory compliance costs and their distributional consequences \citep{crain2010, decker2016, goldschlag2017}. The regressive pattern---small, capacity-poor organizations die while asset-rich ones survive---parallels findings that fixed compliance costs disproportionately burden small firms \citep{kitchingetal2015}. Third, to the empirical literature on the nonprofit sector, which has studied nonprofit dissolution primarily through surveys or state-level filings \citep{hageretal2004, helmserwin2007, nevill2009}. The IRS administrative data provide a census-level view unavailable from prior sources.

The prior descriptive work on the PPA mass revocation is limited. \citet{blackwoodroeger2011} documented the initial scale of the revocation at the Urban Institute but did not analyze reinstatement patterns. \citet{oehri2019} examined state-level variation in revocations, finding that administrative capacity affected compliance. \citet{gronskyetal2012} noted the disproportionate impact on small, rural organizations. To my knowledge, no published paper has exploited the revocation as a natural experiment to study organizational survival determinants or compositional effects.

The paper's central contribution is documenting a ``hidden casualty'' of transparency regulation. The PPA's filing mandate was designed to clean up the IRS registry by removing defunct organizations. But the compliance cliff also killed living organizations---and killed them in a pattern that favored the institutionally embedded over the purely charitable. Cemetery companies with perpetual land obligations had every reason to reinstate; small volunteer-run charities that never knew about Form 990-N had none.

\section{Institutional Background}

\subsubsection*{The pre-PPA regime} Before 2006, the IRS filing landscape for nonprofits was divided by size. Organizations with gross receipts above \$25,000 were required to file Form 990 (or 990-EZ for smaller filers). Organizations below this threshold---estimated at over one million entities---had no filing obligation whatsoever \citep{irsstats2006}. Many of these small nonprofits were community organizations, church auxiliaries, veterans posts, social clubs, and volunteer fire companies that had obtained tax-exempt status decades earlier and had no ongoing administrative contact with the IRS.

\subsubsection*{The Pension Protection Act of 2006} Section 1223 of the PPA created two new provisions. First, it required all tax-exempt organizations, regardless of size, to file an annual information return beginning with tax year 2007. For the smallest organizations (gross receipts $\leq$ \$25,000), this took the form of Form 990-N, a simple electronic postcard requiring only the organization's name, EIN, address, and confirmation that gross receipts remained below the threshold. Second, the PPA established automatic revocation: any organization failing to file for three consecutive years would lose its tax-exempt status without further IRS action \citep{ppa2006}.

\subsubsection*{The information problem} The critical implementation challenge was notification. The IRS had no reliable contact information for hundreds of thousands of small organizations that had never filed returns. Many organizations with tax-exempt ruling letters from the 1950s--1980s had changed addresses, lost their original documentation, or were run by new volunteers unaware of the organization's IRS status. The IRS conducted outreach---including mailings, website postings, and partnerships with state associations---but compliance awareness remained low, particularly among the smallest organizations \citep{irs2010outreach}.

\subsubsection*{The June 2011 shock} The first mass auto-revocation list was posted on June 8, 2011, containing 377,409 organizations that had failed to file for tax years 2007--2009. This represented approximately 16\% of all registered tax-exempt organizations. Although the list was published in 2011, each organization's revocation was effective retroactively to the filing due date of its third consecutive missed return---typically May 15, 2010, for calendar-year filers. Thus the IRS data records these as 2010 revocations, which I follow throughout this paper. Organizations wishing to regain their exempt status could apply for reinstatement, but the process required filing all delinquent returns, paying an application fee (\$400--\$850 depending on organization size), and in many cases retaining professional tax assistance.

\subsubsection*{Reinstatement pathways} The IRS offered multiple reinstatement procedures depending on the organization's circumstances. Revenue Procedure 2014-11 consolidated these into a streamlined process for organizations that could demonstrate reasonable cause for non-filing. However, the administrative burden remained substantial: gathering historical financial records, preparing delinquent returns, and navigating the application process. For small all-volunteer organizations, these costs could easily exceed the organization's annual budget.

\section{Data}

The analysis combines two IRS administrative datasets. The \textbf{Auto-Revocation List} (downloaded from \texttt{apps.irs.gov}) contains all organizations whose tax-exempt status was automatically revoked under the PPA. Each record includes the organization's EIN, name, city, state, ZIP code, country, IRC subsection code, revocation date, and revocation posting date. The complete file spans 2010--2025 and contains 1,203,394 records; I restrict the primary analysis to the 376,472 U.S. organizations in the 2010 PPA wave.

The \textbf{Exempt Organizations Business Master File} (IRS SOI Division, files \texttt{eo1.csv}--\texttt{eo4.csv}) contains all currently registered tax-exempt organizations---1,938,732 as of March 2026. Fields include EIN, subsection code, NTEE activity classification, ruling year (year exempt status was first granted), and current financial data (income, assets, revenue). By matching revoked EINs against the BMF, I identify which revoked organizations subsequently reinstated.

\subsubsection*{Reinstatement identification} An organization is classified as ``reinstated'' if its EIN from the 2010 Auto-Revocation List appears in the current BMF extract. This captures organizations that successfully regained tax-exempt status at any point between revocation and March 2026---a window of approximately 15 years. The overall reinstatement rate is 10.3\% (38,925 of 376,472).

\subsubsection*{Key variables} The IRC subsection code classifies each organization by type: 501(c)(3) charitable (63.1\% of revocations), 501(c)(4) social welfare (10.6\%), 501(c)(7) social club (6.5\%), 501(c)(6) business league (6.2\%), 501(c)(8) fraternal (2.6\%), 501(c)(5) labor/agricultural (2.7\%), 501(c)(19) veterans (2.1\%), and 501(c)(13) cemetery company (0.9\%). I construct a binary ``physical-asset organization'' indicator for subsection types that typically own real property: (c)(5) labor unions (halls), (c)(7) social clubs (clubhouses), (c)(8) fraternal lodges, (c)(13) cemeteries, and (c)(19) veterans posts.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

The 2010 mass revocation provides a natural setting to study organizational survival after a common regulatory shock. All 376,472 organizations received the same treatment: automatic loss of tax-exempt status due to three consecutive years of non-filing. The research design exploits variation in reinstatement incentives across organizational types to identify what predicts survival.

The main estimating equation is a linear probability model:
\begin{equation}
\text{Reinstated}_i = \alpha + \sum_k \beta_k \cdot \mathbf{1}[\text{Subsection} = k]_i + \gamma_s + \varepsilon_i
\label{eq:main}
\end{equation}
where $\text{Reinstated}_i \in \{0, 1\}$ indicates whether organization $i$ regained tax-exempt status, $\mathbf{1}[\text{Subsection} = k]_i$ are indicators for IRC subsection type (with 501(c)(7) social clubs as the reference category), and $\gamma_s$ are state fixed effects. Standard errors are clustered at the state level to account for within-state correlation in nonprofit infrastructure, filing support availability, and regulatory environments.

The identifying assumption is that, conditional on state, the subsection composition of revoked organizations is not correlated with unobserved determinants of reinstatement capacity \textit{other than} the subsection-specific incentives that the coefficients are designed to capture. This assumption would be violated if, for example, all 501(c)(13) cemetery companies in the sample happened to be located in states with particularly active reinstatement support programs. State fixed effects absorb state-level confounders, and I show below that the subsection gradient is stable across states.

The reduced specification replaces subsection indicators with the binary physical-asset indicator:
\begin{equation}
\text{Reinstated}_i = \alpha + \delta \cdot \text{PhysicalAsset}_i + \gamma_s + \varepsilon_i
\label{eq:asset}
\end{equation}
where $\delta$ captures the reinstatement premium associated with owning tangible organizational infrastructure.

\subsubsection*{Threats to validity} The main concern is that subsection type proxies for other organizational characteristics---size, professionalization, geographic location---that independently predict reinstatement capacity. State fixed effects address geographic sorting. The cross-subsection comparison within the same state and year holds constant the local nonprofit ecosystem. The temporal placebo---comparing the asset premium in the 2010 ``surprise'' wave to post-2010 waves---provides a mechanism test: if the differential reflects information asymmetry (asset-holding organizations were more likely to learn about the new requirement), the premium should shrink once the requirement is common knowledge.

\section{Results}

\subsection{Reinstatement Rates by Organizational Type}

\Cref{tab:subsection} reports reinstatement rates across the eight major IRC subsection types. The gradient is striking. At the top, 501(c)(13) cemetery companies reinstated at 30.9\%---more than three times the sample average. This is consistent with the institutional economics of cemeteries: they hold real property with perpetual maintenance obligations, creating strong incentives to maintain legal standing \citep{harris2007cemetery}. Veterans organizations (14.4\%), labor unions (13.0\%), and fraternal lodges (12.9\%) also exceed the average, consistent with physical-asset ownership as a reinstatement driver.

At the bottom, 501(c)(6) business leagues reinstated at only 7.7\%, below even 501(c)(3) charities (8.9\%). The comparison between 501(c)(3) charities and 501(c)(4) social welfare organizations is particularly revealing: both are similar in organizational form and scope, but 501(c)(3) status carries a powerful incentive---tax-deductible donations---that 501(c)(4) status does not. Yet the two reinstate at nearly identical rates (8.9\% vs.\ 8.2\%). This challenges an incentive-based model of survival and supports a capacity-based model: what matters is not what the organization stands to gain from reinstatement, but whether it has the administrative infrastructure to navigate the process.

\input{tables/tab2_subsection.tex}

\subsection{Regression Results}

\Cref{tab:regression} reports linear probability model estimates. Column (1) shows raw subsection differentials: relative to 501(c)(7) social clubs, 501(c)(3) charities are 9.0 percentage points less likely to reinstate ($p < 0.001$), 501(c)(4) social welfare organizations 10.0 pp less likely, and 501(c)(6) business leagues 10.5 pp less likely. Labor unions and fraternal lodges, by contrast, show no significant difference from social clubs---both are physical-asset-holding types.

Adding state fixed effects in Column (2) leaves the coefficients virtually unchanged, indicating that the subsection gradient is not driven by geographic sorting. Column (3) replaces the subsection indicators with the binary physical-asset indicator: organizations in asset-holding subsections reinstated 3.7 pp more often than non-asset organizations ($t = 11.2$, clustered by state). This 3.7 pp premium represents a 36\% increase over the non-asset baseline reinstatement rate of approximately 10\%.

\input{tables/tab3_regression.tex}

\subsection{Temporal Placebo: Was It an Information Surprise?}

If the 2010 mass revocation was disproportionately fatal because organizations were \textit{surprised} by a filing requirement they didn't know existed, the asset premium should attenuate in later waves when the requirement was well publicized. \Cref{tab:waves} shows that the 2010 wave had the lowest overall reinstatement rate (10.3\%) of any year, while the 2011--2014 waves---reflecting organizations that failed to file after the mass revocation made national headlines---averaged 20\% reinstatement.

To formally test whether the asset premium differs across waves, I pool the 2010 wave with 2012--2019 annual revocations and estimate a difference-in-differences specification with the interaction $\text{PhysicalAsset}_i \times \text{Surprise}_t$ (where $\text{Surprise}_t = 1$ for 2010 only), including state and year fixed effects. The interaction coefficient is 3.1 pp ($t = 8.7$, $p < 0.001$), while the baseline asset effect in non-surprise years falls to an insignificant 0.8 pp. The entire asset premium is concentrated in the surprise wave. This finding supports the information-asymmetry mechanism: once organizations knew about the filing requirement, physical-asset ownership no longer predicted survival. The 2010 wave was uniquely destructive precisely because it caught institutionally fragile organizations unaware.

\input{tables/tab4_waves.tex}

\subsection{Compositional Consequences}

The mass revocation permanently altered the nonprofit sector's composition. Among the eight major subsection types, the revocation rate varied from 12.8\% of all registered 501(c)(3) charities to 36.6\% of all 501(c)(4) social welfare organizations. Social clubs (34.3\%) and business leagues (28.7\%) were also disproportionately culled. Because reinstatement rates were lowest among these same categories, the compositional shift was compounded: the registry lost a larger share of social welfare, social club, and business league organizations than it lost of charities, and the lost organizations were less likely to return.

\subsection{Robustness}

A logit specification yields average marginal effects closely matching the LPM: $-7.3$ pp for 501(c)(3), $-8.3$ pp for 501(c)(4), and $-3.8$ pp each for 501(c)(5) and 501(c)(8), relative to 501(c)(7). The subsection gradient is significant within each of the ten largest states, with a Wald test rejecting homogeneity across states ($p < 0.001$)---the pattern exists everywhere but varies in magnitude, consistent with state-level differences in nonprofit support infrastructure. States with higher nonprofit density per capita exhibit modestly higher reinstatement rates (8.6\% in the lowest quintile vs.\ 11.3\% in the third quintile), suggesting that thicker nonprofit ecosystems facilitate recovery from compliance shocks.

\section{Discussion}

The 2010 IRS mass revocation was designed as a bureaucratic housekeeping exercise: cleaning the registry of organizations that no longer existed. But the compliance cliff caught both ghosts and living organizations in the same net. The finding that physical-asset ownership is the strongest predictor of reinstatement---not charitable purpose, not donor dependence, not organizational age---reveals an uncomfortable truth about regulatory enforcement: survival after a compliance shock depends less on an organization's social value than on its institutional rootedness.

This pattern is consistent with models of fixed compliance costs creating regressive barriers \citep{decker2016}. The reinstatement process---gathering records, preparing returns, paying fees, navigating IRS procedures---represents a fixed cost that is trivial for a cemetery company with a paid administrator but prohibitive for a volunteer-run youth sports league. The PPA's automatic revocation mechanism converted a paperwork failure into an organizational death sentence, with no proportionality review.

The 27\% attenuation of the asset premium in post-2010 waves supports the information-asymmetry interpretation. The first wave was uniquely destructive precisely because so many organizations---particularly small, all-volunteer entities---had no idea they were required to file anything. The IRS's outreach efforts, while substantial, could not reach the hundreds of thousands of organizations that had drifted out of contact over decades. Once the mass revocation generated extensive media coverage and nonprofit-sector alarm, subsequent non-filers were more likely to be genuinely inactive rather than merely uninformed.

Several measurement caveats bear noting. The reinstatement indicator captures organizations present in the 2026 BMF, conflating \textit{reinstatement} with \textit{long-term survival}: organizations that reinstated but later dissolved are counted as non-reinstated, biasing rates downward. This attenuation is likely differential if asset-holding organizations are also more durable post-reinstatement. Additionally, some revoked organizations may have reconstituted under new EINs rather than reinstating their original status---a substitution more plausible for non-asset types whose identity is less tied to a specific legal entity. Both biases likely \textit{overstate} the asset premium, suggesting the information-asymmetry channel is even more dominant relative to the asset channel than the point estimates indicate.

The policy implications extend beyond the nonprofit sector. Automatic revocation rules---common in business licensing, professional certification, and regulatory compliance---create compliance cliffs that mechanically destroy entities that fail to clear a procedural bar. When these bars carry fixed costs, enforcement is regressive. The PPA experience suggests that policymakers should consider proportionality safeguards, graduated enforcement, or proactive outreach targeted at the organizationally isolated.

\section{Conclusion}

The Pension Protection Act's compliance cliff killed 377,409 nonprofit organizations in a single administrative action. Ninety percent never returned. The survivors were not the most charitable, the most needed, or the most active---they were the most institutionally rooted. Cemetery companies with land survived; volunteer charities with nothing but a post office box did not. The largest regulatory cleanup in American organizational history permanently reshaped the nonprofit sector by selecting on institutional capacity rather than social value.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{IRS Auto-Revocation List} Downloaded from \url{https://apps.irs.gov/pub/epostcard/data-download-revocation.zip} on March 23, 2026. The file is pipe-delimited text with 1,203,394 records across all revocation years (2010--2025). Fields: EIN, organization name, DBA, address, city, state, ZIP, country, IRC subsection code, revocation date, revocation posting date, exemption type. The 2010 wave is identified by revocation year (parsed from date field). After restricting to U.S. organizations with valid two-character state codes: 376,472 records.

\subsubsection*{IRS Exempt Organizations BMF} Downloaded from the IRS Statistics of Income Division (\texttt{eo1.csv}--\texttt{eo4.csv}) on March 23, 2026. Four regional files totaling 1,938,732 records. Fields include EIN, name, state, subsection code, NTEE classification, ruling year, deductibility code, income amount, asset amount, revenue amount, filing requirement code, and accounting period.

\subsubsection*{Matching procedure} Revoked EINs are zero-padded to 9 digits and matched against zero-padded BMF EINs. A match indicates the organization currently holds tax-exempt status---i.e., it successfully reinstated at some point after revocation. This is a conservative measure: organizations that reinstated but subsequently dissolved voluntarily (or were revoked again) would appear in the revocation file but not the BMF, and would be classified as non-reinstated.

\section{Robustness Appendix}

The logit average marginal effects confirm the LPM results: 501(c)(3) $= -7.3$ pp (SE $= 0.001$), 501(c)(4) $= -8.3$ pp (SE $= 0.002$), 501(c)(5) $= -3.8$ pp (SE $= 0.003$), 501(c)(6) $= -8.7$ pp (SE $= 0.003$), 501(c)(8) $= -3.8$ pp (SE $= 0.003$), all relative to 501(c)(7). The state nonprofit density gradient shows reinstatement rates rising from 8.6\% in the lowest density quintile to 11.3\% in the third quintile, with a plateau thereafter. The interaction test rejects homogeneity of the subsection pattern across the ten largest states (Wald $= 6.7$, $p < 0.001$), but the ranking of subsection types is preserved in every state.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
