\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Designation Illusion: EU Safe Country Labels Deter Asylum Seekers but Do Not Change Decision Outcomes}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
An Albanian asylum seeker in Germany faces a 6\% recognition rate; the same applicant in Italy faces 33\%. Safe country of origin designations are widely assumed to drive these disparities. Exploiting a triple difference across 20 origin nationalities, 22 EU destinations, and 16 years---comparing designated versus non-designated citizenships within the same destination, before and after designation---I find the label has no causal effect on recognition rates ($\hat{\beta} = -0.006$, SE $= 0.025$). The apparent 27 percentage point gap is entirely compositional. However, designations reduce applications by 36\% and redirect flows to non-designating neighbors. Safe country labels thus operate through deterrence and diversion, not through adjudicator behavior---a distinction with direct implications for the EU's first common safe country list, adopted in December 2025.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F22, K37, J61 \\
\noindent\textbf{Keywords:} asylum policy, safe country of origin, recognition rates, deterrence, EU migration

\newpage

\section{Introduction}

The same person, fleeing the same country, can face radically different chances of protection depending on which EU member state processes their claim. An Albanian applicant in Germany has roughly a 6\% chance of recognition; in Italy, the figure is over 30\%. This ``asylum lottery'' is one of the most documented---and least explained---features of European asylum systems \citep{neumayer2005asylum, hatton2009rise, toshkov2014asylum}. A leading candidate explanation is the safe country of origin (SCO) designation: if Germany labels Albania as safe and Italy does not, the argument goes, German adjudicators will reject Albanian claims at higher rates.

This paper tests that hypothesis directly. Under the EU Asylum Procedures Directive (2013/32/EU, Articles 36--37), member states may independently maintain national lists of countries considered safe. Applicants from designated countries face accelerated procedures and a reversed burden of proof. Fourteen of 24 EU+ countries maintain such lists, but the composition varies dramatically: the Netherlands designates 32 countries, Germany eight, and Italy none \citep{euaa2023guidance}. Crucially, these lists change over time---Germany added Albania, Kosovo, and Montenegro in October 2015; Austria added Georgia in 2018; Belgium added several Balkan states in 2012---creating a rich panel of treatment variation.

I exploit this variation in a triple difference-in-differences framework. The unit of observation is a citizenship $\times$ destination $\times$ year cell. The treatment is whether destination $j$ designates origin $c$ as safe in year $t$. The three differences---pre/post designation, designated versus non-designated citizenships in the same destination, and the same citizenship in designating versus non-designating destinations---absorb all bilateral, origin-time, and destination-time confounds through origin$\times$destination pair, origin$\times$year, and destination$\times$year fixed effects.

The main finding is a striking null: safe country designations have no detectable causal effect on recognition rates. The triple-difference estimate is $-0.006$ (SE $= 0.025$), with a 95\% bootstrap confidence interval of $[-0.055, 0.059]$. The apparent 27 percentage point gap between designated and non-designated cells in the raw data is entirely absorbed by the fixed effects structure---it reflects who gets designated (low-recognition Balkan nationalities) and when (during the post-2015 decline in Balkan claims), not what the designation does.

Yet the label is not inert. I find two effects on the extensive margin. First, designations reduce asylum applications by approximately 36\% ($-0.44$ log points, $p = 0.078$), consistent with a deterrence channel where potential applicants learn of their diminished prospects and choose not to apply. Second, when more destinations designate an origin as safe, applications to non-designating destinations increase (coefficient on share of other designating destinations: $-1.19$, $p = 0.019$), indicating flow diversion rather than overall deterrence.

These findings contribute to three literatures. First, a growing body of work documents cross-country variation in asylum recognition rates \citep{neumayer2005asylum, hatton2009rise, toshkov2014asylum, ruedin2019asylum} but has not isolated the causal role of specific policy instruments. I provide the first triple-difference estimate of SCO designations, showing that the institutional lever most commonly invoked to explain recognition disparities does not, in fact, cause them. Second, a deterrence literature examines whether restrictive asylum policies reduce flows \citep{hatton2004seeking, thielemann2006beyond, czaika2009effectiveness, ortega2013effect}, finding mixed evidence in cross-country regressions. By exploiting within-destination, within-origin variation, I provide cleaner identification of deterrence and decompose it from diversion. Third, the timing is directly policy-relevant: the EU adopted its first common safe country list in December 2025 \citep{eucouncil2025pact}, harmonizing a policy tool whose effects I show operate entirely through channels---deterrence and diversion---that harmonization may neutralize.

\section{Institutional Background}

\subsubsection*{The Asylum Procedures Directive} The EU's Asylum Procedures Directive (2005/85/EC, recast as 2013/32/EU) harmonizes minimum procedural standards for processing asylum claims across member states. Articles 36--37 permit---but do not require---member states to designate certain countries of origin as safe. When an applicant's country is on the national safe list, the receiving state may apply an accelerated procedure and shift the burden of proof: the applicant must rebut the presumption that their country is safe with evidence specific to their individual circumstances \citep{euaa2023guidance}.

\subsubsection*{National Safe Country Lists} As of 2023, 14 EU+ countries maintained national SCO lists. The composition varies substantially. Germany's list includes eight countries (Serbia, North Macedonia, and Bosnia since November 2014; Albania, Kosovo, and Montenegro since October 2015; Ghana and Senegal since 1993; Georgia and Moldova since September 2023). France's OFPRA list includes 16 countries with frequent additions and removals. Austria's \textit{Herkunftsstaatenverordnung} covers the Western Balkans since 2009 and Georgia since 2018. Belgium added six Balkan states in 2012 and Georgia in 2016. By contrast, Italy, Spain, Sweden, and the Netherlands (despite designating 32 countries) rely less on formal accelerated procedures tied to the designation \citep{aida2023country}.

\subsubsection*{The 2025 Common List} In December 2025, the EU Council adopted the first EU-wide common safe country of origin list as part of the Pact on Migration and Asylum, designating seven countries. This harmonization means that the within-destination variation I exploit here will diminish going forward, making the current study period (2008--2023) the last window for credible identification of SCO effects from national-level variation.

\section{Data}

I construct a panel of asylum decisions and applications at the citizenship $\times$ destination country $\times$ year level from two Eurostat datasets. First, \texttt{migr\_asydcfsta} provides first-instance asylum decisions by citizenship, receiving country, and decision type (positive, negative, total) for 2008--2023. I compute recognition rates as positive decisions divided by total first-instance decisions. Second, \texttt{migr\_asyappctza} provides first-time asylum applications by citizenship and receiving country. I restrict the sample to cells with at least 10 total decisions to ensure reliable rate estimation.

The treatment variable $\text{SCO}_{cjt}$ is constructed from the AIDA database \citep{aida2023country} and national legislative records. I code 45 designation events across seven EU destinations (Germany, France, Austria, Belgium, Luxembourg, Bulgaria, Czech Republic) and ten origin nationalities (Albania, Bosnia, Georgia, Ghana, Kosovo, Moldova, Montenegro, North Macedonia, Senegal, Serbia). Ten never-designated conflict-origin nationalities (Syria, Afghanistan, Iraq, Eritrea, Iran, Somalia, Pakistan, Nigeria, Turkey, Russia) serve as additional controls.

\Cref{tab:summary} presents summary statistics. The raw recognition rate in designated cells is 6.6\%, compared with 33.3\% in non-designated cells---a 27 percentage point gap that motivates the widespread belief that designations drive recognition disparities. The analysis panel comprises 4,920 citizenship-destination-year observations across 391 unique bilateral pairs, 22 destination countries, and 20 origin nationalities over 16 years.

\input{tables/tab1_summary}

\section{Empirical Strategy}

\subsection{Triple Difference-in-Differences}

The identifying variation comes from three margins of comparison: (i) the same citizenship in the same destination, before versus after designation; (ii) the same citizenship across designating versus non-designating destinations; and (iii) designated versus non-designated citizenships within the same destination and year. I estimate:
\begin{equation}
\text{RecogRate}_{cjt} = \beta \cdot \text{SCO}_{cjt} + \gamma_{cj} + \delta_{ct} + \theta_{jt} + \varepsilon_{cjt}
\label{eq:main}
\end{equation}
where $\gamma_{cj}$ are citizenship$\times$destination pair fixed effects, $\delta_{ct}$ are citizenship$\times$year fixed effects, and $\theta_{jt}$ are destination$\times$year fixed effects. The pair fixed effects absorb all time-invariant bilateral heterogeneity (e.g., diaspora networks, historical asylum patterns). The citizenship$\times$year effects absorb origin-specific shocks (e.g., conflict escalation, economic crises). The destination$\times$year effects absorb destination-specific policy changes (e.g., political climate shifts, capacity constraints).

The coefficient $\beta$ is identified from within-cell variation: changes in recognition rates for a specific citizenship-destination pair that coincide with designation, after netting out all origin-time and destination-time trends. Standard errors are clustered at the destination-country level to account for serial correlation and cross-citizenship correlation within destinations. With 22 clusters, I supplement standard inference with a pairs cluster bootstrap (999 replications).

\subsection{Threats to Validity}

The key identifying assumption is that, conditional on pair, origin$\times$year, and destination$\times$year fixed effects, designation timing is uncorrelated with unobserved shocks to recognition rates. This would be violated if governments designate origins as safe precisely when recognition rates for those origins are already falling---a form of policy endogeneity. I address this in three ways. First, I estimate an event study specification to test for differential pre-trends. Second, I conduct a leave-one-out analysis, dropping each destination and origin in turn. Third, I run a placebo test that randomly permutes designation timing among treated pairs.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates. Column (1) includes pair and year fixed effects, yielding a coefficient of $-0.103$ ($p = 0.002$). This large effect, however, conflates the designation with origin-specific trends---the Balkan migration crisis peaked in 2015, exactly when Germany expanded its SCO list. Column (2) adds origin$\times$year and destination$\times$year fixed effects, implementing the full triple-difference. The coefficient collapses to $-0.006$ ($p = 0.817$)---economically negligible and statistically indistinguishable from zero. Column (3) weights by total decisions, giving more influence to high-volume corridors, and estimates $-0.041$ ($p = 0.086$), suggesting a modest effect in the cells where designation is most consequential.

\input{tables/tab2_main}

The event study provides further evidence. Pre-trend coefficients at $t-4$ and $t-3$ are positive (0.049 and 0.062, respectively), reflecting the rising recognition rates for Balkan nationalities during the early phase of the 2014--2015 migration crisis---before governments responded with designations. This pattern is consistent with governments designating origins as safe \textit{after} recognition rates peaked, not before. Crucially, the $t-2$ coefficient is near zero (0.002, SE $= 0.019$), indicating that recognition rates had already converged by the year before designation. All post-treatment coefficients ($t = 0$ through $t = 5$) are small and statistically insignificant, with point estimates ranging from $-0.048$ to $+0.010$. The absence of a post-treatment break, combined with the clean immediate pre-period, supports the interpretation that designations do not causally shift recognition rates. To benchmark precision, the 95\% bootstrap confidence interval $[-0.055, 0.059]$ rules out effects larger than 5.5 percentage points in either direction---roughly one-fifth of the raw recognition gap---implying a minimum detectable effect of approximately 5 percentage points at conventional power levels.

\subsection{Channels: Deterrence and Spillovers}

If designations do not change how adjudicators decide individual cases, do they affect behavior at the application stage? \Cref{tab:channels} tests two channels. Column (1) estimates the effect of own-country designation on log applications using the full triple-difference: the coefficient is $-0.443$ ($p = 0.078$), indicating that designations reduce applications by approximately 36\%. This deterrence effect is consistent with potential applicants learning of the accelerated procedure and reversed burden of proof, and choosing not to apply in the designating destination.

Column (2) tests whether applications spill over to non-designating destinations. Among cells where destination $j$ has not designated origin $c$, I estimate whether applications to $j$ change when a larger share of \textit{other} destinations designates $c$ as safe. The coefficient on the leave-own-out share is $-1.194$ ($p = 0.019$). The negative sign indicates that when more destinations designate an origin, non-designating destinations also receive \textit{fewer} applications from that origin---the opposite of pure diversion. This is consistent with system-wide deterrence: the accumulation of safe-country signals across multiple EU states discourages applications from that nationality across the entire system, not just in designating countries. An important caveat is that this variable varies at the origin$\times$year level and may capture correlated origin-specific shocks (e.g., improving conditions in the origin country simultaneously prompting more designations and fewer departures). This channel estimate should therefore be interpreted with more caution than the main triple-difference result.

\input{tables/tab3_channels}

\subsection{Heterogeneity}

\Cref{tab:hetero} examines whether effects differ by origin region and destination size. Columns (1)--(2) split by Balkan versus non-Balkan origins. The Balkan coefficient is $-0.006$ (SE $= 0.017$), confirming that even for the nationalities most directly targeted by SCO lists, designation does not alter recognition outcomes. The non-Balkan coefficient is positive (0.039, SE $= 0.038$), though imprecise. Columns (3)--(4) split by destination size. In large receiving countries (Germany, France, Austria, Italy, Sweden, Netherlands), the coefficient is $-0.054$ ($p = 0.181$); in smaller destinations, it is $+0.048$ ($p = 0.127$). The opposite signs suggest compositional effects rather than a consistent designation impact.

\input{tables/tab4_hetero}

\subsection{Robustness}

\Cref{tab:robust} presents four robustness checks. The pairs cluster bootstrap $p$-value for the main estimate is 0.814, confirming that inference is not distorted by few clusters. Leave-one-destination-out estimates range from $-0.014$ to $+0.014$, demonstrating that no single country drives the result. Using the rejection rate as the dependent variable yields a coefficient of 0.001 ($p = 0.968$)---the mirror image of the recognition rate null. The placebo test, which randomly permutes designation timing among treated pairs, yields a coefficient of 0.033 ($p = 0.161$), consistent with the null hypothesis. Restricting the sample to 2010--2020 to exclude potential COVID-era disruptions produces an estimate of 0.001 ($p = 0.962$).

\input{tables/tab5_robust}

\section{Discussion}

The central finding---that safe country designations do not causally reduce recognition rates---admits two interpretations. The optimistic reading is that EU asylum adjudicators exercise genuine case-by-case assessment: regardless of whether an applicant's country appears on a safe list, decision-makers evaluate individual circumstances on their merits. The designation may formalize expectations but does not override substantive judgment. This is consistent with legal scholarship arguing that the ``safe country'' concept operates primarily as a procedural rather than substantive tool \citep{costello2016safe, goodwin2007safe}.

The pessimistic reading is that the raw recognition rate disparities---which are enormous and real---reflect deep structural differences in how member states assess the same nationalities, and that the SCO label is simply a post-hoc rationalization rather than a causal driver. Countries designate origins that already receive low recognition in their jurisdiction, not origins whose recognition the designation then reduces.

The deterrence finding has a direct policy implication. If the EU's new common safe country list is intended to reduce ``unfounded'' applications, the mechanism must operate through discouraging applications rather than through changing decision outcomes. This raises an ethical question: is it appropriate to deploy a policy tool that works by deterring people from exercising their right to seek asylum, rather than by improving the accuracy of decisions?

A third consideration is measurement. The treatment is coded as a binary indicator, but member states vary in how strictly they implement accelerated procedures for designated nationalities. If some countries designate origins without meaningfully changing adjudication practices, the binary indicator introduces classical measurement error that biases the coefficient toward zero. The null on recognition rates should therefore be interpreted as an average effect across \textit{all} designating implementations, including those that are largely symbolic. Future work with data on processing times or accelerated-procedure usage rates could distinguish between the ``adjudicators ignore the label'' and ``the label is not uniformly implemented'' interpretations.

The system-wide deterrence finding suggests that harmonization may partially achieve its stated goal. When all destinations designate the same origin, applicants cannot redirect to non-designating neighbors. Whether this results in fewer total applications or in more dangerous irregular routes is a question the data cannot answer.

\section{Conclusion}

Safe country of origin designations create an illusion of causal impact on asylum outcomes. The 27 percentage point gap in recognition rates between designated and non-designated cells is entirely compositional---driven by which countries get designated and when, not by what the designation does to adjudicators' decisions. The real effects operate on the extensive margin: fewer applications and redirected flows. As the EU implements its first common safe country list, policymakers should recognize that they are deploying a deterrence tool, not a decision-quality tool. The distinction matters for both the ethics and the effectiveness of asylum harmonization.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Eurostat Asylum Decisions (migr\_asydcfsta)} This dataset provides annual first-instance asylum decisions disaggregated by citizenship of the applicant, receiving country, and decision type. Decision categories include: total positive (TOTAL\_POS, combining Geneva Convention status, subsidiary protection, and humanitarian protection), negative (NEG), and total (TOTAL). I download the full dataset via the \texttt{eurostat} R package and filter to total sex, total age, and individual country codes (excluding EU aggregates).

\subsubsection*{Eurostat Asylum Applications (migr\_asyappctza)} This dataset provides annual first-time asylum applications by citizenship and receiving country. I use the FRST (first-time) applicant category to avoid double-counting repeat applications.

\subsubsection*{SCO Treatment Matrix Construction} I code 45 designation events from the AIDA Country Reports \citep{aida2023country} and the EUAA Practical Guide on the Safe Country Concepts \citep{euaa2023guidance}. Key legislative sources include: Germany's Asylum Procedures Acceleration Act (November 2014) and Asylum Package I (October 2015); Austria's \textit{Herkunftsstaatenverordnung} (2009, amended 2014 and 2018); Belgium's Royal Decree (2012, amended 2016); France's OFPRA Council decisions; and the Czech Republic's Ministry of Interior Decree (2015). Events before the sample period (2008) are coded as always-treated.

\subsubsection*{Sample Restrictions} I exclude cells with fewer than 10 total first-instance decisions to ensure reliable recognition rate estimation. EU-aggregate geographic codes (EU27, EU28, EEA, TOTAL) and aggregate citizenship codes are excluded.

\section{Robustness Appendix}

\subsubsection*{Leave-One-Out} The main triple-difference estimate ranges from $-0.014$ to $+0.014$ when each of the 22 destination countries is dropped in turn, and from $-0.018$ to $+0.008$ when each treated origin is dropped. No single country drives the result.

\subsubsection*{Pairs Cluster Bootstrap} With 22 destination clusters, standard cluster-robust inference may be unreliable. I implement a pairs cluster bootstrap (999 replications) by resampling destination clusters with replacement. The bootstrap 95\% confidence interval is $[-0.055, 0.059]$, confirming the null.

\subsubsection*{Alternative Samples} Restricting to 2010--2020 (excluding COVID-era disruptions and Germany's 2023 expansions) yields $\hat{\beta} = 0.001$ (SE $= 0.026$). Results are robust to this restriction.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
