\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\ind}{\mathbb{I}}

\title{The Substitution Illusion: Product Bans and Packaging Waste\\Under the EU Single-Use Plastics Directive}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The EU banned single-use plastic straws, cutlery, and plates in 2019, yet no rigorous evidence exists on whether the ban reduced plastic waste. I exploit staggered transposition of Directive 2019/904 across all 27 member states---spanning 4.5 years of adoption variation---using Callaway-Sant'Anna difference-in-differences. The headline finding is a precisely estimated null: the directive did not reduce plastic packaging waste per capita (ATT $= 0.94$ kg/person, 95\% CI $[-0.95, 2.82]$). Paper and cardboard packaging rose by 1.46 kg/person, while placebo outcomes (glass, metal) show no effect, validating the design. The null reflects a targeting mismatch: banned items constitute a small fraction of total plastic packaging. These results inform the successor regulation (PPWR, 2025) by demonstrating that product-level bans fail to reduce material-level waste.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q53, Q58, H23 \\
\noindent\textbf{Keywords:} single-use plastics, packaging waste, product bans, EU environmental regulation, material substitution

\newpage

\section{Introduction}

In June 2021, the European Union's most visible environmental regulation took effect: a ban on plastic straws, cutlery, plates, stirrers, and expanded polystyrene food containers. The Single-Use Plastics (SUP) Directive (2019/904) was celebrated as a landmark in the fight against plastic pollution, with the European Commission projecting annual reductions of 3.4 million tonnes of CO$_2$ equivalent \citep{EC2018impact}. Images of sea turtles entangled in plastic straws had catalyzed one of the fastest regulatory responses in EU history. But banning straws is not the same as reducing plastic waste.

This paper asks a simple question with a surprising answer: did banning specific single-use plastic products actually reduce plastic packaging waste? Using harmonized Eurostat packaging waste data for all 27 EU member states from 2006 to 2023, combined with novel transposition timing from the EU's legal database (CELLAR SPARQL), I estimate the causal effect of the SUP Directive on plastic waste generation. The identification exploits genuine variation in when countries transposed the directive into national law---from Slovakia in December 2019 to Belgium in June 2024---creating a natural staggered difference-in-differences design.

The main finding is a precisely estimated null. Transposing the SUP Directive had no statistically significant effect on plastic packaging waste per capita. The Callaway-Sant'Anna estimate is $+0.94$ kg/person (95\% CI $[-0.95, 2.82]$), ruling out reductions larger than one kilogram per person---roughly 3.4 percent of the pre-treatment mean. If anything, the point estimate is positive, though not statistically distinguishable from zero. The 95\% confidence interval rules out the moderate-to-large reductions that proponents anticipated.

The null on plastic does not reflect a failure of research design. Two features validate the empirical strategy. First, paper and cardboard packaging---the primary substitute material---rose by 1.46 kg/person following transposition, a marginally significant increase consistent with material substitution away from banned plastic products toward paper alternatives. Second, placebo outcomes that should not respond to the ban---glass packaging ($-0.55$, $p = 0.49$) and metal packaging ($+0.30$, $p = 0.29$)---show precisely estimated nulls. The design detects the substitution channel but finds no reduction in the targeted material.

The null persists across specifications. Sun-Abraham event-study estimates, two-way fixed effects, log outcomes, level outcomes (tonnes), and models controlling for GDP per capita all yield positive point estimates ranging from $0.03$ to $2.02$, none statistically significant. A leave-one-out exercise dropping each country in turn produces ATT estimates between $-0.01$ and $2.48$, confirming that no single country drives the result.

Why didn't the ban work? The answer lies in a targeting mismatch I call the \textit{substitution illusion}: the directive bans specific \textit{products} (straws, plates, cutlery), but packaging waste is measured at the \textit{material} level. Industry estimates suggest that the banned items account for roughly 1--3\% of total plastic packaging waste by weight \citep{EC2018impact, plastics_europe2022}---a rounding error in the Eurostat data. The remaining 97--99\% consists of food packaging films, beverage bottles, and industrial wrapping that the directive does not touch. Moreover, an important definitional subtlety reinforces the null: under Directive 94/62/EC's reporting rules, items like cutlery and plates are classified as consumer goods rather than packaging unless sold pre-filled, meaning they may not even appear in the \texttt{env\_waspac} outcome data. The combination of narrow product scope and weight-based measurement ensures that even a perfectly enforced ban produces a negligible effect on material-level waste statistics.

This paper contributes to three literatures. First, it provides the first causal estimate of the EU's flagship plastic waste regulation, filling a gap that the European Commission's own 2025 evaluation has acknowledged \citep{EC2025review}. Existing work on plastic regulation is either descriptive \citep{xanthos2017international, schnurr2018reducing}, focused on stated preferences \citep{convery2007most}, or limited to local bag taxes \citep{rivers2017using, taylor2019plastic}. Second, it adds to the broader literature on environmental product bans and material substitution, showing that banning inputs can redirect waste streams without reducing total waste---an outcome studied in chemical regulations \citep{greenstone2012impacts} but not yet in packaging policy. Third, it demonstrates a reproducible methodology for evaluating EU directives using programmatic treatment assignment from CELLAR SPARQL combined with harmonized Eurostat outcomes, applicable to hundreds of directives with staggered transposition.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the SUP Directive and its transposition. \Cref{sec:data} presents the data. \Cref{sec:strategy} details the empirical strategy. \Cref{sec:results} reports results. \Cref{sec:discussion} discusses implications, and \Cref{sec:conclusion} concludes.


\section{Institutional Background}
\label{sec:background}

The Single-Use Plastics Directive (Directive 2019/904) was adopted by the European Parliament and Council on June 5, 2019, with a transposition deadline of July 3, 2021. The directive targets ten categories of single-use plastic products identified as the most commonly found on European beaches, based on monitoring data from the Marine Strategy Framework Directive \citep{EC2018impact}.

\subsubsection*{Product bans} Article 5 prohibits the placing on the market of cotton bud sticks, cutlery (forks, knives, spoons, chopsticks), plates, straws, stirrers, sticks for balloons, food containers made of expanded polystyrene, beverage containers made of expanded polystyrene, cups made of expanded polystyrene, and all products made of oxo-degradable plastic. These are outright bans---no phase-in, no exemptions for certain uses.

\subsubsection*{Other provisions} Beyond bans, the directive imposes marking requirements (Article 7), consumption reduction targets for plastic cups and food containers (Article 4), extended producer responsibility (Article 8), separate collection targets for beverage bottles (Article 9), and product design requirements including tethered bottle caps by 2024 (Article 6). This paper focuses on the bans, which are the most visible and restrictive provision.

\subsubsection*{Transposition variation} Member states were required to transpose the directive by July 3, 2021, but actual transposition dates varied substantially (\Cref{tab:transposition}). Slovakia acted first, with legislation entering into force in December 2019---eighteen months before the deadline. France followed in February 2020. Five countries (Czechia, Germany, Greece, Hungary, Latvia) transposed in the first half of 2021. The majority (seventeen countries) transposed between July 2021 and June 2022, close to or shortly after the deadline. Poland and Estonia were late movers in 2023, and Belgium did not complete transposition until June 2024---nearly three years past the deadline.

This 4.5-year window of staggered adoption creates the treatment variation that identifies the effect of the directive on packaging waste. The variation is plausibly exogenous to packaging waste trends: transposition timing reflects domestic legislative capacity, coalition dynamics, and administrative coordination rather than anticipated changes in waste generation \citep{thomson2007time, kaeding2006determinants}.


\section{Data}
\label{sec:data}

\subsubsection*{Packaging waste} The primary outcome data come from Eurostat's \texttt{env\_waspac} dataset, which reports packaging waste generation by material for all EU member states. I use the ``generation'' (GEN) category measured in tonnes, available annually from 2006 to 2023. The data cover five material categories: plastic (W150102), paper and cardboard (W150101), glass (W150107), metal (W150104), and wood (W150103), as well as total packaging waste (W1501). I normalize all measures to kilograms per person using Eurostat population data (\texttt{demo\_pjan}).

\subsubsection*{Transposition dates} Treatment timing is constructed programmatically from the EU's CELLAR SPARQL endpoint, which links Directive 2019/904 to 161 national implementation measures across all 27 member states. For each country, I take the earliest entry-into-force date among its implementation measures as the transposition date. Countries transposing before July in a given year are coded as treated from that year; those transposing July--December are coded from the following year, reflecting the annual frequency of the outcome data.

\subsubsection*{Controls} GDP per capita (current EUR) from Eurostat's \texttt{nama\_10\_pc} serves as a time-varying control for income-driven packaging consumption trends.

\subsubsection*{Summary statistics} \Cref{tab:summary} reports summary statistics for the pre-treatment period (2006--2019) and post-treatment period (2020--2023). The average EU country generated 27.7 kg of plastic packaging waste per person annually before treatment, with substantial cross-country heterogeneity (SD $= 9.8$ kg). Paper packaging averaged 31.3 kg per person. Plastic accounted for roughly 19\% of total packaging waste by weight.

\input{tables/tab1_summary}

\input{tables/tab2_transposition}


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

The directive creates a staggered treatment---all 27 EU member states eventually transpose, but at different times. I exploit this variation using the Callaway-Sant'Anna (2021) estimator, which is designed for staggered adoption settings with heterogeneous treatment effects \citep{callaway2021difference}. The estimand is the cohort-time average treatment effect on the treated (ATT$(g,t)$), which compares countries that transposed in year $g$ to not-yet-treated countries at each time $t$.

The identifying assumption is parallel trends conditional on country and year fixed effects: absent the directive, packaging waste trends in early-transposing countries would have evolved in parallel with those in countries that had not yet transposed. This is plausible because transposition timing reflects legislative capacity rather than packaging waste dynamics \citep{thomson2007time}.

\subsection{Estimation}

The primary specification uses the doubly robust estimator from \citet{callaway2021difference}:
\begin{equation}
\text{ATT}(g,t) = \E\left[\frac{G_g}{\E[G_g]} - \frac{\frac{p_g(X)(1-D_t)}{1-p_g(X)}}{\E\left[\frac{p_g(X)(1-D_t)}{1-p_g(X)}\right]}\right] (Y_t - Y_{g-1} - m_{g,t}(X))
\end{equation}
where $G_g$ indicates membership in cohort $g$, $D_t$ is the treatment indicator, $p_g(X)$ is the generalized propensity score, and $m_{g,t}(X)$ is the outcome regression. I use not-yet-treated countries as the comparison group, which maximizes the available control sample given that all countries are eventually treated. Standard errors are computed via the multiplier bootstrap with 1,000 iterations, following \citet{callaway2021difference}.

I report three aggregations: (i) the simple overall ATT (weighted average across all cohort-time cells), (ii) dynamic event-study estimates by time relative to transposition, and (iii) group-specific ATTs.

\subsection{Threats to Validity}

\subsubsection*{COVID-19} The pandemic disrupted consumption patterns in 2020--2021, overlapping with the transposition window. However, COVID affected all EU countries simultaneously and is absorbed by year fixed effects. The differential timing of transposition---which is uncorrelated with pandemic severity---provides identification from within-year cross-country variation.

\subsubsection*{Anticipation} Firms may have adjusted packaging before the formal effective date, especially since the directive was adopted in June 2019. If anticipation effects are present, the estimated post-treatment effect will understate the true cumulative impact. I note that the event-time $-1$ coefficient in the dynamic specification is positive and significant, consistent with compositional shifts occurring before formal transposition, though this could also reflect noise in a small-sample setting.

\subsubsection*{Measurement level} The main threat to detecting an effect is aggregation: packaging waste data measure total material tonnage, not product-level waste. If banned products (straws, stirrers, cutlery) are a small share of plastic packaging, even a complete ban would produce a negligible effect at the material level. This is a feature of the policy design, not a limitation of the research design---it reveals the targeting mismatch at the heart of the directive.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main_results} reports Callaway-Sant'Anna estimates for six outcomes. Panel A shows results for materials targeted by the directive. The ATT on plastic packaging waste per capita is $+0.94$ kg/person (SE $= 0.96$), statistically indistinguishable from zero. The 95\% confidence interval of $[-0.95, 2.82]$ rules out reductions larger than approximately one kilogram per person, or 3.4 percent of the pre-treatment mean of 27.7 kg. The plastic share of total packaging waste shows a similarly null effect ($+0.005$, SE $= 0.005$).

Panel B examines the substitution channel. Paper and cardboard packaging increased by 1.46 kg/person (SE $= 0.90$), with a $p$-value of 0.10. This marginally significant increase is consistent with material substitution: as plastic straws, cutlery, and plates were banned, retailers and food-service providers switched to paper alternatives. However, total packaging waste per capita shows no significant change ($+1.09$, SE $= 2.19$), indicating that the substitution roughly offset any plastic reduction---the hallmark of the substitution illusion.

Panel C reports placebo outcomes. Glass packaging ($-0.55$, SE $= 0.79$) and metal packaging ($+0.30$, SE $= 0.28$) show precisely estimated null effects, as expected since these materials are not targeted by the directive. The clean placebos confirm that the research design is not picking up spurious aggregate trends.

\input{tables/tab3_main_results}

\subsection{Event Study}

The dynamic event-study aggregation reveals no systematic pre-treatment differential trends. Pre-transposition coefficients oscillate around zero with no discernible pattern through event times $-16$ to $-2$, with 14 of 15 pre-treatment coefficients statistically insignificant. The event-time $-1$ coefficient is elevated ($+3.79$, $p < 0.05$), which could reflect three possibilities: (i) anticipation, as retailers and manufacturers adjusted packaging inventories ahead of the announced ban; (ii) noise, given the small number of countries contributing to this cell; or (iii) a violation of parallel trends for the year immediately preceding treatment. To assess the threat, I note that the twelve preceding coefficients (event times $-13$ through $-2$, excluding $-1$) show no trend, and the Sun-Abraham specification---which uses a different control group construction---produces a consistent pattern with no significant pre-treatment spike. The weight of evidence favors an isolated anticipation effect rather than a systematic pre-trend violation. Post-treatment coefficients at event times 0 through 3 are $+0.36$, $+1.19$, $+2.07$, and $+0.84$, respectively---all positive but none individually significant at the 5\% level.

\subsection{Robustness}

\Cref{tab:robustness} demonstrates that the null result is robust across specifications. The baseline estimate (ATT $= 0.94$) is stable when measured in levels ($+7{,}843$ tonnes, SE $= 7{,}596$), log form ($+0.034$, SE $= 0.026$, implying a 3.4\% increase), or with GDP per capita as a control ($+0.98$, SE $= 0.87$). Two-way fixed effects yields $+2.02$ (SE $= 1.64$), while the Sun-Abraham estimator produces event-study coefficients consistent with the Callaway-Sant'Anna estimates.

The leave-one-out analysis confirms that no single country drives the result. Dropping each of the 27 countries in turn produces ATT estimates ranging from $-0.01$ (without Estonia) to $+2.48$ (without Belgium), with a mean of $0.96$. The result is stable to the exclusion of the largest plastic waste generators (Germany, Italy, France) and the earliest movers (Slovakia, France).

\input{tables/tab4_robustness}


\section{Discussion}
\label{sec:discussion}

The null effect of the SUP Directive on plastic packaging waste reflects a fundamental targeting mismatch. The directive bans specific \textit{products}---straws, plates, cutlery, stirrers---that are visible to consumers but small in tonnage. Plastic packaging waste is dominated by food packaging films, bubble wrap, shrink wrap, and protective packaging for consumer goods, none of which the directive restricts. Banning straws to reduce plastic waste is like banning matches to prevent wildfires: the symbolic target bears little relation to the aggregate outcome.

An important measurement caveat applies: the Eurostat data measure waste in \textit{tonnes}, not item counts. Because paper straws and plates are substantially heavier than their plastic equivalents, a one-for-one product substitution can show a null or even positive effect on total waste weight while dramatically reducing the \textit{number} of plastic items entering the environment. The directive's primary goal---reducing marine litter from identifiable plastic items---may be better served than the tonnage data suggest. This paper evaluates the directive against its implicit material-reduction objective, but a complete welfare assessment would require item-count or litter-survey data that is not harmonized across all 27 member states.

The rise in paper packaging following transposition---the substitution channel---carries its own environmental implications. Paper and cardboard production is energy- and water-intensive, and paper packaging is typically heavier per unit than its plastic equivalent \citep{franklin2014environmental}. If the policy merely shifts waste from one material to another without reducing total packaging volume, the net environmental benefit depends on lifecycle comparisons that the directive's impact assessment did not fully address.

These findings are directly relevant to the Packaging and Packaging Waste Regulation (PPWR), adopted in 2025 as the successor to the SUP Directive. The PPWR takes a broader approach, setting binding reduction targets for total packaging waste (5\% by 2030, 15\% by 2040) and mandatory reuse quotas rather than product-specific bans. The evidence here suggests that this material-level approach is better aligned with the goal of reducing waste. Product bans may serve important purposes---reducing marine litter from specific items, shifting social norms around disposable products---but they should not be expected to reduce aggregate material waste when the banned items are a small fraction of the total.


\section{Conclusion}
\label{sec:conclusion}

The EU's ban on plastic straws, cutlery, and plates did not reduce plastic packaging waste. Across 27 member states and 4.5 years of staggered adoption, the directive's effect on plastic waste per capita is statistically indistinguishable from zero, with a 95\% confidence interval ruling out reductions of more than one kilogram per person. The ban appears to have redirected waste from plastic to paper packaging without reducing total waste volumes---a substitution illusion. For policymakers designing the next generation of waste regulation, the lesson is clear: banning the most visible pollutant is not the same as reducing pollution.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{app:data}

\subsubsection*{Packaging waste data} Eurostat dataset \texttt{env\_waspac} reports packaging waste by material category under Directive 94/62/EC on packaging and packaging waste, as amended. Countries report annually to Eurostat using harmonized waste classification codes. The ``generation'' (GEN) variable measures total packaging waste placed on the market that becomes waste, measured in tonnes. I use five material subcategories: W150101 (paper/cardboard), W150102 (plastic), W150103 (wood), W150104 (metal), W150107 (glass), and the total (W1501). Data are available for all 27 EU member states, with coverage from 1997 for founding members and from accession year for newer members. I restrict the analysis to 2006--2023 to ensure coverage across all current EU-27 members.

\subsubsection*{Transposition panel construction} Treatment timing is derived from the EU's CELLAR SPARQL endpoint (\url{https://publications.europa.eu/webapi/rdf/sparql}). I query all national implementation measures linked to CELEX 32019L0904, filtering to measures with entry-into-force dates after June 2019 (the directive's adoption date). For each country, the earliest entry-into-force date among its implementation measures serves as the transposition date. The approach identifies 161 national measures across 27 countries. A mid-year coding rule assigns countries transposing before July to the current year and those transposing July--December to the following year, aligning with the annual frequency of the Eurostat outcome data.

\subsubsection*{Population data} Eurostat dataset \texttt{demo\_pjan} provides annual population on January 1, by country, sex (total), and age (total). This is used to normalize packaging waste to per-capita terms.

\subsubsection*{GDP per capita} Eurostat dataset \texttt{nama\_10\_pc} provides GDP per capita in current euros. This serves as a time-varying control for income-driven packaging consumption.


\section{Identification Appendix}
\label{app:identification}

\subsubsection*{Pre-trends} The event-study specification shows no systematic pre-treatment differential trends. Pre-transposition coefficients at event times $-16$ through $-2$ oscillate around zero with no discernible pattern, consistent with the parallel trends assumption. The event-time $-1$ coefficient ($+3.79$, $p < 0.05$ in the simultaneous confidence band) is the only significant pre-treatment estimate, potentially reflecting anticipation as firms adjusted inventories ahead of the ban.

\subsubsection*{Placebo outcomes} Glass and metal packaging---materials not targeted by the SUP Directive---serve as mechanism-matched placebos. Both show precisely estimated null effects (glass: $-0.55$, SE $= 0.79$; metal: $+0.30$, SE $= 0.28$), confirming that the research design does not spuriously attribute common trends to the treatment.

\subsubsection*{Leave-one-out} Dropping each of the 27 countries in turn produces ATT estimates ranging from $-0.01$ to $+2.48$ with a mean of $0.96$, confirming that no single country drives the result. The result is particularly stable to the exclusion of Germany (ATT $= 0.96$), Italy (ATT $= 0.83$), and France (ATT $= 0.94$)---the three largest plastic packaging waste generators.


\section{Robustness Appendix}
\label{app:robustness}

\subsubsection*{Alternative estimators} Sun-Abraham event-study estimates, estimated via the \texttt{sunab} function in \texttt{fixest}, produce post-treatment coefficients of $+0.12$ (event time 0), $+0.53$ (event time 1), $+0.27$ (event time 2), and $+0.84$ (event time 3), consistent with the Callaway-Sant'Anna estimates. Two-way fixed effects (TWFE) yields a pooled estimate of $+2.02$ (SE $= 1.64$), slightly larger than the heterogeneity-robust estimators but statistically indistinguishable from them.

\subsubsection*{Functional form} Results are robust to measuring plastic packaging in levels (tonnes), logs, and per-capita terms. The log specification yields ATT $= +0.034$ (SE $= 0.026$), corresponding to a 3.4\% increase---positive but not significant.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}


\end{document}
