¥begin{table}[t]
¥centering
¥caption{Standardized Effect Sizes}
¥label{tab:sde}
¥small
¥begin{adjustbox}{max width=¥textwidth}
¥begin{tabular}{lcccccl}
¥toprule
Outcome & $¥hat{¥beta}$ & SE & SD($Y$) & SDE & SE(SDE) & Classification ¥¥
¥midrule
¥multicolumn{7}{l}{¥textit{Panel A: Pooled}} ¥¥[3pt]
Cross-hazard substitution (DDD) & -0.0322 & 0.0103 & 2.326 & -0.0138 & 0.0044 & Small negative ¥¥
Respiratory conditions (DiD) & -0.0018 & 0.0011 & 0.034 & -0.0531 & 0.0328 & Moderate negative ¥¥
Total injuries (DiD) & -0.1346 & 0.0429 & 4.682 & -0.0287 & 0.0092 & Small negative ¥¥
Days away from work (DiD) & -0.8201 & 0.9564 & 84.056 & -0.0098 & 0.0114 & Small negative ¥¥
¥midrule
¥multicolumn{7}{l}{¥textit{Panel B: Heterogeneous (firm size, cross-hazard substitution)}} ¥¥[3pt]
Small firms ($<$500 emp.) & -0.0338 & 0.0104 & 2.326 & -0.0145 & 0.0045 & Small negative ¥¥
Large firms ($¥geq$500 emp.) & 0.0481 & 0.0294 & 2.326 & 0.0207 & 0.0126 & Small positive ¥¥
¥bottomrule
¥end{tabular}
¥end{adjustbox}
¥begin{tablenotes}
¥footnotesize
¥item ¥textit{Notes:} ¥textbf{Country:} United States. ¥textbf{Research question:} Does OSHA's 2018 crystalline silica standard, which targeted respiratory hazards in manufacturing, cause cross-hazard injury substitution to non-targeted hazard categories within the same establishments? ¥textbf{Policy mechanism:} The standard reduced the permissible exposure limit for respirable crystalline silica from 250 to 50 $¥mu$g/m$^3$ in general industry, requiring engineering controls, respiratory protection programs, medical surveillance, and exposure assessments---imposing substantial compliance costs on silica-intensive manufacturers. ¥textbf{Outcome definition:} Injury/illness rates per 100 full-time equivalent workers (count $¥div$ total hours worked $¥times$ 200,000), separately by hazard category (total injuries, respiratory conditions, hearing loss, skin disorders, other illnesses). ¥textbf{Treatment:} Binary: high-silica manufacturing (NAICS 327, 331, 332) versus low-silica manufacturing (all other NAICS 31--33 subsectors). ¥textbf{Data:} OSHA Injury Tracking Application Form 300A, 2016--2024, establishment-year level, 254,933 observations from 43,954 manufacturing establishments. ¥textbf{Method:} Triple-difference (high-silica $¥times$ non-targeted hazard $¥times$ post-2021) with establishment$¥times$hazard, hazard$¥times$year, and establishment$¥times$year fixed effects; standard errors clustered at establishment level. ¥textbf{Sample:} Manufacturing establishments (NAICS 31--33) present in the ITA for $¥geq$4 years with positive employment and hours; excludes construction and services. SDE $= ¥hat{¥beta} / ¥text{SD}(Y)$ where SD($Y$) is the pre-treatment standard deviation. Classification refers to magnitude, not statistical significance: Large ($|$SDE$| > 0.15$), Moderate ($0.05$--$0.15$), Small ($0.005$--$0.05$), Null ($< 0.005$).
¥end{tablenotes}
¥end{table}
