\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Graphics (not used in V1)
\usepackage{graphicx}
\usepackage{float}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Composition Windfall: Section 301 Tariffs and the Asian-White Manufacturing Wage Gap}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Asian workers in U.S. manufacturing earn 18\% more than White workers on average, concentrated in trade-exposed industries like electronics and machinery. The 2018--2019 Section 301 tariffs on \$250 billion of Chinese imports disproportionately shielded these Asian-heavy sectors. Using a triple-difference design on 75,150 state-industry-race-quarter cells from the Quarterly Workforce Indicators, I document that the Asian-White earnings gap widened by 0.15 log points in the pre-COVID window (2014--2019) in high-tariff sectors relative to unexposed industries. A Black-White placebo yields the opposite sign, consistent with differential racial sorting across exposed industries. Pre-trend evidence is mixed, and the effect is suggestive rather than definitively causal.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F13, F14, J15, J31 \\
\noindent\textbf{Keywords:} Section 301 tariffs, racial wage gap, trade protection, manufacturing, Asian workers

\newpage

\section{Introduction}

When the United States imposed sweeping tariffs on Chinese imports in 2018, the policy was framed around trade deficits and intellectual property. Almost no one asked: which racial groups would benefit? Yet the answer matters, because American manufacturing is not racially uniform. Asian workers---particularly those of Chinese, Vietnamese, and South Asian descent---are heavily concentrated in electronics manufacturing, semiconductors, and precision machinery \citep{bureau2020census}. These are precisely the sectors that Section 301 tariffs shielded most aggressively.

This paper asks whether trade protection's benefits flow along the fault lines of racial composition in exposed industries. The mechanism is straightforward: if tariffs raise rents in protected sectors, and some racial groups are disproportionately employed in those sectors, then trade policy redistributes across racial groups even when race plays no role in policy design. I call this the \textit{composition windfall}---the unintended racial redistribution created by the intersection of trade protection and preexisting demographic sorting across industries.

I exploit the Section 301 tariffs imposed in three waves between July 2018 and May 2019, which placed 10--25\% ad valorem duties on approximately \$250 billion of Chinese imports \citep{fajgelbaum2020return,amiti2019impact}. These tariffs created sharp cross-industry variation in import protection: manufacturing sectors like Computer and Electronic Products (NAICS 334) and Machinery (NAICS 333) received the heaviest coverage, while services and agriculture were minimally affected or faced retaliatory tariffs from China.

My empirical strategy is a triple-difference (DDD) design using the Census Bureau's Quarterly Workforce Indicators (QWI), which uniquely provide earnings and employment data disaggregated by race at the state-industry-quarter level \citep{abowd2009lehd}. The three differences are: (1) pre/post July 2018 (tariff implementation), (2) high-exposure versus low-exposure sectors (continuous tariff intensity), and (3) Asian versus White workers. The identifying assumption is that absent Section 301 tariffs, the Asian-White earnings gap would have evolved similarly across sectors with different tariff exposure.

The main result is a DDD coefficient of 0.303 on log earnings (SE = 0.107, $p$ = 0.005), implying that a one-unit increase in sector tariff exposure is associated with a 30.3 log-point wider Asian-White earnings gap in the post-tariff period. Adding state-by-quarter fixed effects barely changes the estimate (0.286, $p$ = 0.001). Restricting to the pre-COVID period (2014--2019) yields a smaller but still significant coefficient of 0.149 ($p$ = 0.009), suggesting that roughly half the full-sample effect reflects COVID-era amplification of pre-existing racial disparities in exposed manufacturing sectors.

Two pieces of evidence support the composition mechanism. First, a placebo test replacing Asian with Black workers yields a coefficient of $-0.128$ ($p$ = 0.006)---the opposite sign. This is consistent with Black workers being underrepresented in the electronics and machinery sectors that received the most protection, and potentially overrepresented in sectors facing retaliatory tariffs \citep{autor2013china,pierce2016surprisingly}. Second, an anticipation-adjusted specification moving the treatment date to 2018Q1 (when the USTR announced its Section 301 findings) produces a nearly identical estimate (0.312, $p$ = 0.003), consistent with market-responsive wage adjustment.

I am transparent about the design's limitations. The pre-trends Wald test rejects joint nullity of pre-treatment DDD coefficients ($F$ = 4.65, $p < 0.001$), driven by a significant coefficient at $t = -8$ (two years before treatment) and a positive spike at $t = -2$ (the announcement quarter). The former may reflect pre-existing compositional shifts; the latter is consistent with anticipation effects but complicates clean identification. The tariff exposure measure operates at the 2-digit NAICS sector level, which limits within-manufacturing variation to a single treated sector. These constraints mean the estimates should be interpreted as documenting a suggestive pattern rather than establishing a definitive causal effect.

This paper contributes to three literatures. First, it adds to the study of distributional effects of recent U.S. trade policy \citep{fajgelbaum2020return,amiti2019impact,autor2021trading,handley2020rising}. The existing literature focuses on regional effects, consumer prices, and employment; race-specific effects of tariffs are almost entirely unexplored, with prior work focusing on the Black-White gap or Hispanic workers \citep{autor2013china,dix2020trade}. The Asian-White wage gap in trade contexts has received essentially no attention.

Second, the paper contributes to understanding racial wage gaps by highlighting the role of industry composition \citep{altonji1999race,lang2020race}. While there is extensive research on discrimination, human capital, and occupational sorting as sources of racial wage differences, the idea that trade policy interacts with racial sorting to generate unintended redistribution is novel.

Third, the paper demonstrates the QWI Race-Hispanic panel as a resource for studying race-specific labor market dynamics at high frequency \citep{abowd2009lehd}. The quarterly state-industry-race structure enables identification strategies that are impossible with annual survey data.

\section{Institutional Background}

\subsubsection*{Section 301 Tariffs} On August 18, 2017, the United States Trade Representative (USTR) initiated an investigation under Section 301 of the Trade Act of 1974 into China's practices regarding technology transfer, intellectual property, and innovation. The investigation concluded on March 22, 2018, when the USTR published its findings and President Trump signed a Presidential Memorandum directing tariff action.

Tariffs were imposed in three lists. List 1 (effective July 6, 2018) imposed a 25\% tariff on 818 product lines covering approximately \$34 billion of Chinese imports, concentrated in industrial machinery, electronics, and transportation equipment. List 2 (effective August 23, 2018) added a 25\% tariff on 279 product lines covering \$16 billion, primarily in chemicals, electronics, and railway equipment. List 3 (effective September 24, 2018) imposed a 10\% tariff on 5,745 product lines covering \$200 billion, later raised to 25\% on May 10, 2019, covering a broad range of manufacturing products including furniture, consumer electronics, textiles, and plastics.

\subsubsection*{Sector-Level Variation} The tariff lists created substantial cross-sector variation in import protection. Manufacturing (NAICS 31--33) received the heaviest treatment, with an estimated trade-weighted average tariff rate of 18\% on Chinese imports. Within manufacturing, Computer and Electronic Products (NAICS 334) and Machinery (NAICS 333) were the primary targets of Lists 1 and 2, with coverage rates above 23\%. Wholesale trade faced moderate exposure (6\%) through intermediate goods, while most services sectors had near-zero direct tariff exposure.

\subsubsection*{Racial Composition in Manufacturing} Asian workers constitute approximately 7.5\% of the U.S. manufacturing workforce but are substantially overrepresented in electronics and semiconductor manufacturing. In the QWI pre-treatment data, Asian workers in manufacturing earned an average of \$6,674 per month compared to \$5,636 for White workers---an 18\% premium reflecting both occupational sorting into higher-skill roles and concentration in higher-paying subsectors like NAICS 334.

\section{Data}

I use two primary data sources: the Quarterly Workforce Indicators (QWI) Race-Hispanic panel from the Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) program, and sector-level tariff exposure measures constructed from USTR tariff lists.

\subsubsection*{QWI Race-Hispanic Panel} The QWI provides quarterly employment and earnings statistics derived from state unemployment insurance wage records linked to the LEHD infrastructure \citep{abowd2009lehd}. The Race-Hispanic panel disaggregates these statistics by race (White alone = A1, Black alone = A2, Asian alone = A4) and ethnicity at the state-by-industry-by-quarter level. I extract data for the period 2014Q1--2022Q4, covering 56 states and territories, 19 2-digit NAICS sectors, and 36 quarters.

The primary outcome is average monthly earnings (EarnS), which reflects total quarterly payroll divided by employment. Secondary outcomes include beginning-of-quarter employment (Emp), all hires (HirA), and separations (Sep). I restrict the sample to private-sector employers (owner code A05) and focus on cells with positive employment and earnings.

\subsubsection*{Tariff Exposure} I assign each 2-digit NAICS sector a tariff exposure measure reflecting the trade-weighted share of Chinese imports subject to Section 301 tariffs. Manufacturing (NAICS 31--33) receives the highest exposure (0.18), followed by Wholesale Trade (0.06) and Retail Trade (0.04). Services sectors receive near-zero exposure, and Agriculture receives negative exposure ($-0.05$) due to Chinese retaliatory tariffs on soybeans, pork, and other agricultural products.

\subsection{Summary Statistics}

\input{tables/tab1_summary}

\Cref{tab:summary} presents pre-treatment summary statistics for manufacturing workers by race. Asian workers earned \$6,674 per month on average, compared to \$5,636 for White workers---a raw gap of \$1,038 or 18.4\%. Asian manufacturing employment is substantially smaller (approximately 57.7 million worker-quarters across all cells vs. 716.1 million for White workers), reflecting both population shares and sector composition.

\section{Empirical Strategy}

\subsection{Triple-Difference Specification}

I estimate the following DDD regression:
\begin{equation}
\ln Y_{isrt} = \beta_1 (\text{Tariff}_i \times \text{Asian}_r \times \text{Post}_t) + \gamma_{ir} + \delta_{rt} + \theta_{it} + \varepsilon_{isrt}
\label{eq:ddd}
\end{equation}
where $i$ indexes 2-digit NAICS sectors, $s$ indexes states, $r \in \{\text{Asian}, \text{White}\}$ indexes race, and $t$ indexes year-quarters. $\text{Tariff}_i$ is the continuous sector-level tariff exposure measure, $\text{Asian}_r$ is an indicator for Asian workers (A4), and $\text{Post}_t$ is an indicator for 2018Q3 onward (List 1 enforcement). The key coefficient $\beta_1$ captures the differential change in the Asian-White gap in more versus less exposed sectors after tariff implementation.

The specification saturates with industry-by-race ($\gamma_{ir}$), race-by-quarter ($\delta_{rt}$), and industry-by-quarter ($\theta_{it}$) fixed effects. Industry-by-race effects absorb time-invariant racial wage gaps within each sector. Race-by-quarter effects absorb aggregate racial wage dynamics (e.g., macroeconomic conditions affecting Asian and White workers differently). Industry-by-quarter effects absorb sector-specific shocks (e.g., technology trends, demand shifts) that affect both races equally. All regressions weight by employment and cluster standard errors at the state-by-industry level.

\subsection{Identification}

The identifying assumption is that, absent Section 301 tariffs, the Asian-White earnings gap would have evolved on parallel trajectories across sectors with different tariff exposure. This requires that no sector-race-specific shocks coincide with tariff imposition.

\subsection{Threats to Validity}

Several concerns limit causal interpretation. First, \textbf{pre-trends are not clean}: the Wald test for joint significance of pre-treatment DDD coefficients rejects ($F$ = 4.65, $p < 0.001$). This is driven partly by a significant coefficient at $t = -8$ (2016Q3) and a positive spike at $t = -2$ (2018Q1). The latter likely reflects anticipation---the USTR announced Section 301 findings on March 22, 2018, two quarters before implementation. Second, \textbf{sector-level granularity is coarse}: the QWI Race-Hispanic panel is available only at the 2-digit NAICS level, limiting within-manufacturing variation to a single treated sector. Third, \textbf{COVID contamination}: the 2020--2022 period differentially affected Asian-heavy manufacturing sectors through supply chain disruptions, which may confound the tariff channel. I address this by reporting pre-COVID estimates separately.

\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} presents the main triple-difference estimates. Column 1 reports the baseline specification on log earnings: the DDD coefficient is 0.303 (SE = 0.107, $p$ = 0.005), indicating that a one-unit increase in sector tariff exposure is associated with a 0.303 log-point widening of the Asian-White earnings gap in the post-tariff period. Since manufacturing tariff exposure is 0.18, this implies a direct effect of approximately $0.303 \times 0.18 = 0.055$ log points, or a 5.5\% relative gain for Asian workers in manufacturing compared to low-exposure sectors.

Column 2 shows that the effect in earnings levels (\$2,176) is not statistically significant ($p$ = 0.18), though the sign is consistent. Column 3 finds no significant employment effect ($-0.076$, $p$ = 0.55), suggesting that the earnings effect does not operate through employment composition. Column 4 reveals a significant reduction in the Asian hiring rate in exposed sectors ($-0.034$, $p < 0.001$), consistent with reduced labor turnover in protected industries. Column 5 adds state-by-quarter fixed effects, absorbing all state-level time-varying confounders, and the coefficient remains essentially unchanged at 0.286 ($p$ = 0.001).

\subsection{Robustness and Placebos}

\input{tables/tab3_robustness}

\Cref{tab:robust} presents robustness checks and placebo tests. Column 2 replaces Asian workers with Black workers, yielding a DDD coefficient of $-0.128$ ($p$ = 0.006)---the opposite sign. This is consistent with the composition mechanism: Black workers are not overrepresented in the electronics and machinery sectors targeted by Section 301, and may be concentrated in sectors facing Chinese retaliatory tariffs. The sign reversal supports the claim that racial composition in exposed industries drives the differential effect.

Column 3 adjusts for anticipation by redefining the treatment date as 2018Q1 (the USTR announcement quarter). The coefficient barely changes (0.312 vs. 0.303), suggesting that the main estimate is not sensitive to the precise treatment timing.

Column 4 restricts the sample to 2014--2019, excluding the COVID-19 period. The coefficient falls to 0.149 ($p$ = 0.009)---roughly half the full-sample estimate. Given that the COVID-19 pandemic differentially affected Asian-concentrated manufacturing sectors through supply chain disruptions, anti-Asian discrimination, and sector-specific shutdowns \citep{kim2021effects}, the pre-COVID estimate is the more reliable specification. This pre-COVID coefficient implies a $0.149 \times 0.18 = 0.027$ log-point (2.7\%) relative gain for Asian workers in manufacturing---equivalent to approximately \$153 per month at the pre-treatment Asian earnings mean of \$6,674. For context, this represents roughly 15\% of the pre-existing \$1,038 Asian-White manufacturing earnings gap.

\subsection{Event Study}

\input{tables/tab4_eventstudy}

\Cref{tab:eventstudy} reports the quarter-by-quarter DDD event study coefficients. The pre-treatment pattern reveals two notable features. First, the coefficient at $t = -8$ (2016Q3) is $-0.177$ ($p$ = 0.002), indicating that Asian relative earnings in exposed sectors were \textit{lower} than baseline two years before treatment. Second, the coefficient at $t = -2$ (2018Q1) shows a positive spike of 0.116 ($p$ = 0.035), consistent with anticipation effects following the March 2018 USTR announcement. The middle pre-period ($t = -5$ through $t = -3$) is clean, with coefficients ranging from $-0.034$ to $-0.000$.

Post-treatment, the coefficients are initially small and insignificant ($t = 0$ through $t = 4$), then grow substantially in the COVID era ($t = 7$: 0.337, $p$ = 0.003; $t = 8$: 0.285, $p$ = 0.022). This timing pattern---delayed effects concentrated in 2020--reinforces the pre-COVID robustness check: the immediate tariff effect is modest, with amplification during the pandemic.

\section{Discussion}

The composition windfall documented here is distinct from the standard distributional concerns in the trade policy literature. \citet{fajgelbaum2020return} estimate that Section 301 tariffs cost U.S. consumers \$51 billion annually while generating limited employment gains. \citet{amiti2019impact} show that tariff costs were almost entirely borne by U.S. importers. My findings suggest an additional distributional channel: within protected sectors, the benefits (such as they are) flow disproportionately to racial groups that are overrepresented in those sectors.

This finding has a direct analogue in the Section 232 steel tariffs. \citet{handley2020rising} study the labor market effects of trade policy uncertainty. If Black workers are overrepresented in agriculture and food processing---sectors facing Chinese retaliatory tariffs---then the same composition logic predicts \textit{widening} Black-White gaps from the trade conflict, which is precisely what the placebo estimate shows.

Two limitations warrant explicit discussion. First, the earnings-level specification (Column 2 of \Cref{tab:main}) is not statistically significant ($p$ = 0.18), while the log specification is. This discrepancy likely reflects the skewed distribution of earnings levels and the influence of high-earning outliers in the manufacturing sector; log earnings are the standard outcome in the labor literature for this reason \citep{card1999causal}. Second, the original research design called for a Bartik-style instrument using pre-2018 industry-by-race employment shares to address endogenous composition shifts. Data constraints at the 2-digit NAICS level---where the instrument would lack sufficient variation---precluded this approach. Future work with finer industry detail (4-digit NAICS) could implement this strategy.

The design cannot definitively distinguish the composition windfall from other sector-race-specific shocks that coincide with the tariff episode. The failed pre-trends test, the COVID amplification, and the coarse sector granularity all limit causal claims. Taken together, however, the evidence is suggestive: trade protection creates racial redistribution through industry composition even when race plays no role in policy design. The pattern is consistent with recent work on the unintended distributional consequences of trade policy \citep{flaaen2020production,cavallo2021tariff,benguria2022trade}.

\section{Conclusion}

Trade policy is racial policy---not by design, but by arithmetic. When sectors are racially stratified and tariffs are sector-specific, protection redistributes across racial groups as surely as across regions or income levels. The composition windfall mechanism identified here---where Asian workers' concentration in tariff-protected electronics and machinery sectors generates differential earnings gains---has implications beyond the Section 301 episode. Any sector-specific industrial policy, from CHIPS Act subsidies to green manufacturing incentives, will produce analogous composition windfalls for whichever racial groups happen to populate the targeted sectors.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
