\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Compliance Illusion: FATF Grey-Listing and the Resilience of International Banking in Panama}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
FATF grey-listing is widely assumed to impose severe costs on a country's international banking sector. I test this assumption by exploiting within-country heterogeneity in Panama's banking system, where International License banks---legally restricted to cross-border operations---faced maximal exposure to the June 2019 grey-listing, while General License banks with domestic deposit bases were partially insulated. Using monthly profitability data from the banking regulator across 122 months, I find no statistically significant decline in international banks' return on assets or return on equity relative to domestic banks during the 52-month grey-listing period. The null persists in pre-COVID restricted windows, extended control groups, and triple-difference specifications. The placebo test passes cleanly. These findings suggest that grey-listing's compliance burden does not translate into differential profitability losses for the most exposed banks---a ``compliance illusion'' where heightened international scrutiny fails to differentially penalize the cross-border intermediaries it targets.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} G21, G28, F33, O16 \\
\noindent\textbf{Keywords:} FATF, grey-listing, anti-money laundering, Panama, correspondent banking, de-risking, international banking

\newpage

\section{Introduction}

When the Financial Action Task Force placed Panama on its ``grey list'' in June 2019, international observers expected the country's role as a cross-border financial intermediary to suffer. Panama hosts one of the world's largest international banking centers---over 70 banks managing assets exceeding \$130 billion---and its International License banks exist solely to move money across borders. The enhanced due diligence requirements triggered by grey-listing should, in theory, have fallen disproportionately on these institutions: their correspondent banking relationships would face added scrutiny, transaction costs would rise, and their competitive advantage in cross-border intermediation would erode. Yet the profitability of Panama's internationally-oriented banks tells a different story.

This paper asks whether FATF grey-listing differentially harms the banks most exposed to its compliance consequences. The existing empirical literature on grey-listing effects relies entirely on cross-country comparisons \citep{nkusu2023,imf2021,collin2022,masciandaro2017}, where unobserved heterogeneity across jurisdictions confounds identification. I exploit a novel source of within-country variation: Panama's banking regulatory framework creates three pre-determined license types with sharply different exposure to cross-border compliance pressure. International License banks are legally restricted to foreign operations and depend entirely on correspondent banking relationships. General License banks, by contrast, hold domestic deposits and serve the local market. Both types operate under the same supervisor, in the same country, during the same macroeconomic conditions---but face dramatically different exposure to the compliance burden triggered by grey-listing.

Using monthly financial indicators published by the Superintendencia de Bancos de Panam\'a (SBP) from January 2016 through February 2026, I estimate a difference-in-differences comparing International License banks (treatment) to Panamanian Private banks with General Licenses (control) across three periods: 41 months before grey-listing, 52 months during grey-listing, and 29 months after Panama's removal from the list in October 2023. The design has several advantages. License type is determined by law, not by banks' response to FATF. The 41-month pre-treatment period supports parallel trends assessment. The October 2023 de-listing provides a symmetric reversal test. And the analysis window spans the full listing episode rather than relying on short post-treatment windows.

The main result is a precisely estimated null. The grey-listing interaction coefficient on return on assets (ROA) is 0.54 percentage points (SE = 0.38 pp, $p = 0.16$), the wrong sign for the de-risking hypothesis and statistically indistinguishable from zero. Return on equity shows a similarly null pattern during grey-listing ($-0.50$ pp, $p = 0.80$). These nulls survive across every specification: restricting the sample to the pre-COVID window (through February 2020), adding Foreign Private banks as a second control group, and running a triple-difference with both alternative controls. A placebo test assigning treatment to January 2018 finds no effect ($p = 0.47$).

What explains the resilience? Three candidate mechanisms deserve attention. First, international banks may have adjusted their asset composition---shifting toward higher-margin activities to offset reduced transaction volumes, a form of portfolio rebalancing that preserves aggregate profitability even as gross flows decline. Second, correspondent banking de-risking may have been more selective than the sweeping narrative suggests: large, established international banks with strong compliance infrastructure retained relationships, while marginal players exited \citep{erbenova2016,accuity2017}. Third, the compliance costs triggered by grey-listing may have been absorbed economy-wide---affecting all banks through raised baseline costs---rather than concentrating on the international segment alone.

These findings contribute to the growing literature on the real effects of international financial regulation. While several studies document aggregate capital flow reductions associated with grey-listing \citep{nkusu2023,collin2022,johannesen2014,masciandaro2022}, this paper provides the first within-country micro-level evidence on \textit{who} within a banking system bears the cost. The answer---that the most exposed banks are not differentially harmed in terms of profitability---challenges the implicit assumption in the de-risking literature that compliance burden maps linearly to financial performance. The result connects to broader findings on regulatory adaptation: firms under regulatory pressure often restructure rather than shrink \citep{acharya2014,dagher2020,laeven2016}. I call this the ``compliance illusion''---grey-listing creates visible compliance activity and policy attention, but its differential bite on the intended targets is surprisingly weak.

The rest of the paper proceeds as follows. Section 2 describes Panama's banking system and the FATF grey-listing process. Section 3 presents the data. Section 4 lays out the empirical strategy. Section 5 reports results and robustness checks. Section 6 discusses mechanisms and implications.


\section{Institutional Background}

\subsubsection*{Panama's International Banking Center}
Panama has operated as a major international financial hub since the 1970 Banking Law (Decreto de Gabinete No.\ 238), which created a regulatory framework designed to attract foreign banking operations. The Superintendencia de Bancos de Panam\'a (SBP) supervises all banks and issues three types of licenses. \textit{General License} banks may conduct both domestic and international operations; they accept local deposits, extend domestic credit, and participate in the national payment system. \textit{International License} banks are legally restricted to operations outside Panama---they cannot accept domestic deposits or make local loans, and their business model depends entirely on cross-border intermediation, trade finance, and correspondent banking relationships. \textit{Representative Offices} perform liaison functions without conducting banking operations. As of 2019, approximately 42 banks held General Licenses (including 2 state-owned banks and $\sim$20 each of Panamanian-owned and foreign-owned private banks), while $\sim$19 held International Licenses.

This regulatory structure creates a natural partition of the banking sector by exposure to international compliance pressure. International License banks' entire revenue stream depends on cross-border transactions that require functioning correspondent banking relationships with global banks. When international compliance costs rise, these banks face the pressure directly. General License banks, particularly Panamanian-owned private banks with large domestic deposit bases, can partially substitute toward domestic lending.

\subsubsection*{FATF Grey-Listing}
The Financial Action Task Force is an intergovernmental body that sets standards for anti-money laundering (AML) and countering the financing of terrorism (CFT). Countries with strategic deficiencies in their AML/CFT frameworks are placed on a ``grey list'' (formally, ``Jurisdictions under Increased Monitoring'') following a mutual evaluation process. Grey-listed countries commit to an action plan to address identified deficiencies under FATF monitoring.

Panama was placed on the grey list on June 21, 2019, following a mutual evaluation that identified deficiencies in beneficial ownership transparency, supervision of designated non-financial businesses, and international cooperation. The listing represented a reputational shock with operational consequences: global banks conducting transactions with Panamanian counterparties were required to apply enhanced due diligence, raising the cost of correspondent banking. Panama remained grey-listed for 52 months until its removal in October 2023, after demonstrating substantial compliance with its action plan.

The grey-listing was Panama's second; the country was previously listed from 2014 to 2016 \citep{fatf2019}. This history is relevant to identification: international banks may have adapted their compliance infrastructure during the first episode, potentially muting the 2019 shock. The 2019 estimate therefore captures the marginal effect of \textit{repeated} grey-listing, not the effect of a novel designation. The 2019 listing also occurred in the context of the 2016 Panama Papers revelations, which had already heightened international scrutiny of Panama's financial sector \citep{odonovan2019}. The listing was not accompanied by countermeasures (the more severe ``black list'' designation), and Panama's dollarized economy continued to function normally throughout the listing period.

\subsubsection*{COVID-19 Confound}
The grey-listing period (June 2019--October 2023) overlaps substantially with the COVID-19 pandemic. Panama imposed strict lockdowns from March 2020, affecting all sectors of the economy. This creates a potential confound if COVID differentially affected international versus domestic banking. I address this by: (1) analyzing a pre-COVID restricted window through February 2020, which isolates 9 months of grey-listing effects; (2) leveraging the October 2023 de-listing for a post-COVID reversal test; and (3) noting that COVID-era financial support programs were primarily channeled through General License banks, biasing against finding a null result for international banks.


\section{Data}

The primary data source is the SBP's monthly financial indicators publication (\textit{Indicadores Financieros por Tipo de Banca}), which reports four key performance metrics---return on assets (ROA), return on average assets (ROAA), return on equity (ROE), and the capital adequacy ratio (MIN)---for each bank license category. The data span January 2016 through February 2026 (122 months), providing 41 months before grey-listing, 52 months during grey-listing, and 29 months after de-listing.

The SBP also publishes a monthly listing of all operating banks (\textit{Listado de Bancos}) with their license type and country of origin, available from November 2019. This allows tracking of bank entry and exit across the listing period.

\subsubsection*{Treatment and Control Groups}
The treatment group is \textit{Banca Internacional}: International License banks whose operations are legally restricted to cross-border business. The primary control group is \textit{Banca Paname\~na Privada}: Panamanian-owned private banks holding General Licenses with substantial domestic deposit bases. An alternative control group is \textit{Banca Extranjera Privada}: foreign-owned private banks holding General Licenses, which face intermediate exposure (foreign ownership but domestic operations).

\subsection{Summary Statistics}

\input{tables/tab1_summary.tex}

Table~\ref{tab:summary} reports summary statistics. Several features of the data merit discussion. First, International License banks have substantially higher mean ROA (2.44\%) than Panamanian Private banks (1.41\%) over the full sample, reflecting their higher-margin cross-border business model. Second, International License banks exhibit much greater volatility in profitability: their ROA standard deviation is three times larger than domestic banks' (1.21 pp vs.\ 0.40 pp). Third, ROE patterns diverge sharply in the post-delisting period: Panamanian Private banks' ROE surges to 20.0\% while International License banks recover to only 14.9\%. This divergence, discussed further in Section 5, appears to reflect domestic credit expansion rather than international sector weakness.


\section{Empirical Strategy}

\subsection{Identification}

I estimate a two-group difference-in-differences exploiting the sharp regulatory distinction between International License banks (treatment) and Panamanian Private banks (control). The identifying assumption is that, absent grey-listing, the profitability gap between these two bank types would have evolved along the same trajectory as in the pre-treatment period.

The estimating equation is:
\begin{equation}
Y_{it} = \alpha_i + \lambda_t + \beta_1 (\text{Int}_i \times \text{Grey}_t) + \beta_2 (\text{Int}_i \times \text{Post}_t) + \varepsilon_{it}
\label{eq:did}
\end{equation}
where $Y_{it}$ is a financial indicator for bank type $i$ in month $t$; $\alpha_i$ and $\lambda_t$ are bank-type and month fixed effects; $\text{Int}_i$ indicates International License banks; $\text{Grey}_t$ indicates the grey-listing period (June 2019--September 2023); $\text{Post}_t$ indicates the post-delisting period (October 2023--February 2026); and $\beta_1$ is the coefficient of interest. The omitted period is pre-treatment (January 2016--May 2019).

I use Driscoll-Kraay standard errors to account for serial correlation in the panel. With two cross-sectional units and a long time dimension ($T = 122$), clustering at the unit level would yield only two clusters; Driscoll-Kraay inference is designed precisely for this setting---panels with small $N$ and large $T$ \citep{driscoll1998}.

\subsection{Threats to Validity}

\subsubsection*{Parallel Trends}
The 41-month pre-treatment period allows assessment of the parallel trends assumption. I estimate a semi-annual event study specification replacing $\beta_1$ with indicators for each six-month bin interacted with the treatment dummy, with the last pre-treatment bin ($-1$ to $-6$ months) as the reference period.

\subsubsection*{Anticipation}
The FATF mutual evaluation process generates pre-listing signals: the on-site evaluation visit typically occurs 12--18 months before listing. If international banks began adjusting behavior in anticipation, pre-trends would be contaminated. I examine this possibility in the event study and find suggestive evidence of a pre-treatment divergence at $-18$ to $-12$ months, coinciding with the mutual evaluation timeline.

\subsubsection*{Composition}
If grey-listing caused international bank exit or entry, the composition of the treatment group could change endogenously. Using the monthly bank listings (available from November 2019), I track bank counts by license type and find stable composition throughout the listing period.

\subsubsection*{COVID-19}
As discussed in Section 2, the overlap with COVID-19 is addressed by restricting the window to pre-COVID months and by leveraging the post-pandemic de-listing event.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

Table~\ref{tab:main} reports the main difference-in-differences results. The coefficient on International $\times$ Grey-listed is 0.0054 for ROA (column 1), indicating that International License banks' return on assets was 0.54 percentage points \textit{higher} relative to Panamanian Private banks during the grey-listing period compared to the pre-treatment gap. This estimate is statistically insignificant ($p = 0.16$). The ROE result in column 2 is $-0.50$ percentage points and far from significance ($p = 0.80$). ROAA (column 3) mirrors the ROA pattern. Capital adequacy (column 4) shows no differential change during grey-listing ($-0.07$ pp, $p = 0.58$) but a significant positive shift after de-listing ($+0.57$ pp, $p < 0.001$).

To translate these magnitudes: the pre-treatment ROA gap between international and domestic banks was approximately 0.76 percentage points. The DiD estimate of 0.54 pp, if taken at face value, would imply the gap \textit{widened} by 71\%---opposite to the de-risking prediction. However, the estimate is imprecise, and the 95\% confidence interval ($-0.21$ pp to $1.28$ pp) includes both meaningful negative effects and substantial positive effects. The within-$R^2$ of 0.045 confirms that the treatment interaction explains little variation in ROA beyond fixed effects.

The post-delisting period reveals an interesting asymmetry. While ROA shows no reversal (column 1: 0.21 pp, $p = 0.72$), ROE falls significantly for international banks relative to domestic banks (column 2: $-4.67$ pp, $p = 0.056$). This reflects the Panamanian Private banks' rapid ROE expansion in 2024--2026 (reaching 20\%) rather than a decline in international bank performance. Capital adequacy increases sharply for international banks post-delisting ($+0.57$ pp, $p < 0.001$), suggesting they strengthened their capital buffers after removal from the grey list.

\subsection{Event Study}

Table~\ref{tab:eventstudy} in Appendix~\ref{sec:robustness_appendix} reports semi-annual binned event study coefficients. For ROA, pre-treatment coefficients are mostly statistically insignificant and small in magnitude, with no systematic trend. For ROE, the $-18$ to $-12$ month bin shows a negative and significant coefficient ($-4.39$ pp, $p = 0.010$). This period corresponds to late 2017--early 2018, when the FATF mutual evaluation process would have generated anticipatory signals. I interpret this as suggestive evidence of pre-listing adjustment rather than a parallel trends violation, noting that it affects ROE but not ROA and occurs at a specific, institutionally motivated timing.

In the post-treatment period, ROA event study coefficients fluctuate around zero without a clear sustained shift, confirming the null main result. The ROE event study shows more structure: negative coefficients emerge in the second year of grey-listing and grow larger post-delisting, consistent with the divergence driven by domestic banks' ROE expansion documented above.

\subsection{Robustness}

\input{tables/tab3_robustness.tex}

Table~\ref{tab:robustness} reports four robustness specifications for ROA. The baseline result (column 1, 0.54 pp) is stable when adding Foreign Private banks as an additional control group (column 2, 0.56 pp), restricting to the pre-COVID window through February 2020 (column 3, 0.51 pp), and running a triple-difference with both alternative controls (column 4, 0.45 pp). The coefficient remains positive and insignificant across all specifications. The pre-COVID result is particularly informative: restricting to 9 months of grey-listing treatment---completely free of pandemic confounding---yields an estimate of 0.51 pp ($p = 0.16$), nearly identical to the full-period result.

\input{tables/tab4_placebo.tex}

Table~\ref{tab:placebo} presents the placebo test and de-listing reversal analysis. Column 1 assigns a fake grey-listing date of January 2018 using only the pre-treatment sample. The placebo coefficient is 0.26 pp ($p = 0.47$), confirming no spurious differential trend in the pre-period. Columns 2--3 report the de-listing reversal test, comparing grey-listing to post-delisting periods. The ROA reversal is null ($-0.32$ pp, $p = 0.63$), while the ROE reversal is marginally significant ($-4.17$ pp, $p = 0.088$), again driven by the domestic bank ROE surge rather than international bank decline.


\section{Discussion}

The central finding---that FATF grey-listing does not differentially harm the profitability of internationally-oriented banks---challenges a core assumption of the de-risking narrative. The policy debate typically presumes that compliance costs fall most heavily on institutions with the greatest cross-border exposure, creating a direct channel from grey-listing to financial sector damage. The Panama evidence suggests this channel, at least for profitability, is weaker than assumed.

\subsubsection*{Why the Null?}
Three non-exclusive mechanisms can explain the resilience. First, \textit{portfolio rebalancing}: international banks may have shifted their asset composition toward higher-yielding but lower-volume activities, preserving per-unit returns even as transaction volumes declined. The ROA data capture returns relative to assets, not asset levels; if international banks' total assets contracted while margins on remaining business widened, aggregate profitability could remain stable.

Second, \textit{selective de-risking}: the termination of correspondent banking relationships was likely concentrated among smaller, less-established banks rather than the major international banks whose aggregate indicators dominate the bank-type averages \citep{erbenova2016}. If the intensive margin (surviving banks maintaining profitability) dominated the extensive margin (marginal bank exit), the aggregate indicators would show stability even if individual institutions were severely affected.

Third, \textit{cost diffusion}: compliance costs may have been borne by the entire Panamanian banking system rather than concentrated on international banks. If domestic banks also faced increased compliance requirements---for instance, through strengthened SBP supervision affecting all license types equally---the differential effect would be attenuated.

\subsubsection*{What the Null Does Not Mean}
The null profitability result does not imply that grey-listing had no economic consequences. Grey-listing may have reduced aggregate capital flows through Panama, contracted the volume of international banking transactions, or shifted business from Panama to competing financial centers. These quantity effects would not appear in per-unit profitability ratios. The finding is specifically that the \textit{relative} profitability of internationally versus domestically oriented banks was not differentially affected---not that grey-listing was economically costless.

\subsubsection*{Limitations}
Three limitations constrain the strength of these conclusions. First, the SBP reports financial indicators at the bank-type level, not for individual banks. This aggregation masks heterogeneity: within the International License category, some banks may have experienced severe profitability declines while others thrived. Second, the two-unit panel structure limits statistical power. The minimum detectable effect at 80\% power is 1.05 percentage points---half of the pre-treatment International ROA---meaning the design can rule out only large profitability declines, not moderate ones. The permutation $p$-value of 0.20 confirms the null is robust to alternative inference, but the confidence interval remains wide. Third, the joint pre-trend test rejects the hypothesis of perfectly parallel pre-trends ($p < 0.001$), driven by an anticipatory dip during the FATF evaluation window. While this timing is consistent with institutional anticipation rather than a design flaw, it introduces uncertainty about the pre-treatment counterfactual.

\subsubsection*{Implications}
These findings carry implications for the design of international financial regulation. If grey-listing's compliance burden does not translate into differential costs for the most exposed intermediaries, the mechanism through which listing is supposed to incentivize reform deserves scrutiny. The ``compliance illusion''---visible regulatory activity without measurable differential consequences for target institutions---suggests that grey-listing may operate primarily through reputational and political channels rather than through direct financial pressure on cross-border banks. Future work with individual bank-level data could decompose the aggregate null into intensive and extensive margin components, while cross-country within-jurisdiction comparisons (for instance, in the Bahamas, Malta, or the Philippines, which have similar multi-license banking structures) could test whether Panama's resilience is an anomaly or a regularity.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}
\label{sec:data_appendix}

\subsubsection*{Data Sources}
The primary data source is the Superintendencia de Bancos de Panam\'a (SBP), accessed at \url{https://www.superbancos.gob.pa}. Two datasets are used:

\begin{enumerate}
\item \textit{Indicadores Financieros por Tipo de Banca} --- monthly financial indicators (ROA, ROAA, ROE, capital adequacy ratio) for six bank categories: Centro Bancario (full banking center), Sistema Bancario Nacional (national banking system), Banca Oficial (state-owned banks), Banca Paname\~na Privada (Panamanian-owned private banks with General Licenses), Banca Extranjera Privada (foreign-owned private banks with General Licenses), and Banca Internacional (International License banks). Coverage: January 2016--February 2026 (122 months).

\item \textit{Listado de Bancos} --- monthly listing of all operating banks with license type and country of origin. Coverage: November 2019--February 2026 (76 months).
\end{enumerate}

Supplementary data includes World Bank Financial Development indicators (non-performing loan ratio, annual) for contextual analysis.

\subsubsection*{Variable Definitions}
\begin{itemize}
\item \textbf{ROA} (Return on Assets): Net income divided by total assets, annualized.
\item \textbf{ROAA} (Return on Average Assets): Net income divided by average total assets over the period.
\item \textbf{ROE} (Return on Equity): Net income divided by total equity.
\item \textbf{MIN} (Capital Adequacy): Regulatory capital adequacy ratio.
\item \textbf{Grey-list period}: June 2019--September 2023 (52 months).
\item \textbf{Post-delisting period}: October 2023--February 2026 (29 months).
\end{itemize}

\subsubsection*{Sample Construction}
The analysis sample consists of two bank types (International License and Panamanian Private) observed over 122 months, yielding 244 bank-type-month observations. The extended sample adds Foreign Private banks (366 observations). No observations are dropped or trimmed.


\section{Robustness Appendix}
\label{sec:robustness_appendix}

\subsubsection*{Permutation Inference}
With only two cross-sectional units, inference based on Driscoll-Kraay standard errors alone may be fragile. I conduct a permutation test by randomly assigning the ``International'' treatment label to each of the five non-control bank types in turn, re-estimating the DiD 10,000 times. The two-sided permutation $p$-value is 0.20, confirming that the observed coefficient (0.54 pp) is not unusual relative to the permutation distribution (mean placebo: 0.07 pp, SD: 0.24 pp).

\subsubsection*{Joint Pre-Trend Test}
A Wald test of joint nullity of all five pre-treatment semi-annual event study coefficients yields $F(5, 104) = 4.77$ ($p < 0.001$), rejecting the hypothesis of perfectly parallel pre-trends. This rejection is driven primarily by the $-24$ to $-18$ month bin, which coincides with the FATF mutual evaluation timeline. I interpret this as evidence of anticipatory adjustment during the evaluation process rather than a fundamental violation of the design, but caution is warranted in interpreting the point estimate.

\subsubsection*{Minimum Detectable Effect}
At 80\% power and 5\% significance, the minimum detectable effect is 1.05 percentage points---approximately 50\% of the pre-treatment International License ROA mean (2.11\%) and 1.78 standard deviations of the outcome. The design can reliably detect only large effects; moderate profitability declines of 0.5--1.0 percentage points cannot be ruled out.

\subsubsection*{Driscoll-Kraay Bandwidth Sensitivity}
Results are stable across DK bandwidth choices: at $L = 6$, the grey-listing coefficient on ROA is 0.0054 (SE = 0.0038, $p = 0.160$); at $L = 12$, it is 0.0054 (SE = 0.0035, $p = 0.125$).

\subsubsection*{Event Study Estimates}

\input{tables/tab5_eventstudy.tex}

\subsubsection*{Bank Composition}
Using the monthly bank listings (November 2019--February 2026), bank counts by license type remain stable throughout the grey-listing period, indicating no differential exit or entry that would contaminate the treatment group composition.


\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}


\end{document}
