\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{When the Rice Stopped Flowing: Global Price Pass-Through of India's 2023 Export Ban}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
On July 20, 2023, India banned non-basmati white rice exports overnight, removing 40\% of global rice trade. I estimate the pass-through of this supply shock to retail food prices across 59 countries using 329,030 market-commodity-month observations from WFP food price monitoring data. A within-market, across-commodity difference-in-differences design---comparing rice to non-rice staples in the same market---with continuous treatment intensity based on pre-ban Indian import dependence reveals that moving from zero to full Indian dependence raises rice prices by 13.1\% relative to control commodities ($p < 0.001$). The effect is concentrated in highly dependent countries (10.6\% increase) and absent in low-dependence markets. Pre-ban parallel trends are clean, and the estimate is stable across alternative clustering, outlier trimming, and leave-one-country-out checks.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} F14, Q17, Q18, O13 \\
\noindent\textbf{Keywords:} export bans, food prices, rice trade, supply shocks, food security

\newpage

\section{Introduction}

Three billion people eat rice as a staple. On July 20, 2023, the world's largest rice exporter removed itself from the market overnight. India's Ministry of Commerce Notification No.\ 20/2023 banned all non-basmati white rice exports---approximately 10 million tonnes annually, or 40\% of global rice trade---with no advance notice, no phase-in, and no sunset clause. For import-dependent countries in Sub-Saharan Africa, South Asia, and the Middle East, this was the equivalent of an unannounced supply amputation.

How quickly and completely do unilateral export restrictions propagate through global food markets to local retail prices? Despite the centrality of this question to food security policy, the existing evidence relies almost entirely on aggregate national price indices or small samples of wholesale markets \citep{headey2011rethinking, anderson2012trade, giordani2016food}. No study traces the pass-through from a single export ban to retail prices across thousands of individual food markets spanning dozens of countries. This paper fills that gap.

I construct a novel panel of 329,030 market-commodity-month observations covering 1,530 food markets in 59 countries from the World Food Programme's (WFP) Food Price Monitoring system (2021--2025), merged with bilateral rice trade data from the FAO Detailed Trade Matrix. The identification strategy exploits a within-market, across-commodity difference-in-differences design: comparing the price trajectory of rice to non-rice staple commodities (maize, millet, sorghum, wheat, beans, groundnuts, cassava, potatoes, lentils) sold in the \textit{same} market, before and after July 2023. Market-by-time fixed effects absorb all local shocks---currency movements, transport disruptions, seasonal demand, weather---leaving only the differential commodity-specific price change. Cross-country variation in pre-ban Indian import dependence provides continuous treatment intensity.

The preferred specification, with market$\times$year-month and country$\times$commodity fixed effects and standard errors clustered at the country level (59 clusters), yields a pass-through elasticity of 0.131 (SE = 0.028, $p < 0.001$). This means that moving from zero to full Indian rice dependence increases rice prices by 13.1\% relative to control commodities after the ban. The event study shows clean parallel trends before July 2023 and a sharp divergence immediately after. In the first two quarters post-ban, rice prices rose approximately 6.7\% relative to control commodities across all markets, with the effect concentrated entirely in countries sourcing more than 35\% of rice imports from India. In these highly dependent countries---including Nepal, Ethiopia, Sri Lanka, Kuwait, and Senegal---rice prices rose 10.6\% (SE = 2.1\%) relative to other staples. In low-dependence countries, the effect is a precise zero.

The results are robust. The coefficient ranges from 0.116 to 0.147 across leave-one-country-out iterations (59 countries), changes negligibly under two-way clustering (0.131, SE = 0.031), price trimming (0.128), or the addition of commodity-specific linear trends (0.138). A pre-ban placebo test assigning a fake treatment date of January 2022 produces an insignificant coefficient (-0.658, SE = 0.473).

This paper contributes to three literatures. First, to the trade and food security literature studying export restrictions \citep{martin2012export, anderson2014food, laborde2021covid, gotz2020russia}. While \citet{headey2011rethinking} and \citet{giordani2016food} estimate the price effects of export bans using cross-country price indices, I provide the first market-level evidence from 59 countries, showing that aggregate analyses mask enormous heterogeneity in pass-through driven by bilateral trade dependence. Second, to the cost-of-living and commodity pass-through literature \citep{atkin2015retail, nakamura2018high, cavallo2023online}. The within-market, across-commodity design identifies pass-through while absorbing confounders that plague cross-market comparisons. Third, to the literature on international trade fragmentation and food sovereignty \citep{fajgelbaum2020return, evenett2022trade}. India's ban illustrates how concentrated supply chains create fragile price transmission channels---a finding with immediate implications for food import diversification policy.

The remainder of the paper proceeds as follows. Section 2 describes the institutional background of India's export ban. Section 3 presents the data. Section 4 details the empirical strategy. Section 5 reports results. Section 6 discusses implications.


\section{Institutional Background}

\subsubsection*{India's dominance in global rice trade} India has been the world's largest rice exporter since 2012, shipping approximately 22 million tonnes in 2022 \citep{usda2023grain}. Non-basmati white rice---the staple variety consumed across Sub-Saharan Africa, South Asia, and the Middle East---constituted roughly 10 million tonnes of these exports. India's share of global rice exports exceeded 40\%, creating a market structure in which dozens of importing countries sourced the majority of their rice from a single supplier.

\subsubsection*{The July 2023 ban} On July 20, 2023, the Ministry of Commerce and Industry issued Notification No.\ 20/2023, prohibiting the export of non-basmati white rice (HS 1006.30.90) with immediate effect. The stated rationale was domestic food price stabilization ahead of state elections and the approaching festival season. Several features make this policy shock particularly well-suited for causal analysis. First, the ban was \textit{unanticipated}: no advance legislative discussion, no phased implementation schedule, and no exemption mechanism. Second, it was \textit{commodity-specific}: basmati rice and parboiled rice exports continued unrestricted, providing within-rice placebos. Third, the trade disruption was \textit{massive}: non-basmati exports fell 88\% within three months \citep{reuters2023india}.

\subsubsection*{Importing countries} The ban's impact varied sharply across importing nations. Countries in West Africa (Senegal, Benin, Togo), East Africa (Ethiopia, Kenya, Madagascar), and the Middle East (Saudi Arabia, UAE, Kuwait) sourced 50--99\% of rice imports from India. Other importers---particularly in Southeast Asia and Latin America---had diversified supply chains with substantial volumes from Thailand, Vietnam, and Pakistan. This variation in bilateral trade dependence provides the continuous treatment intensity that identifies the pass-through elasticity.

\subsubsection*{Partial reversal} In October 2024, India partially lifted the ban by introducing a Minimum Export Price (MEP) mechanism, allowing some non-basmati exports at prices above a floor. This partial reversal provides an additional falsification opportunity: if the initial price increase was caused by the ban, prices should attenuate after the lifting.


\section{Data}

The analysis combines two data sources.

\subsubsection*{WFP Food Price Monitoring} The World Food Programme's Vulnerability Analysis and Mapping (VAM) unit collects monthly retail food prices from over 3,000 markets in 98 countries, published through the Humanitarian Data Exchange (HDX). I use the global yearly files for 2021--2025, which provide standardized USD-denominated prices for staple commodities at the market-commodity-month level. The key variables are market identity, commodity type, price type (retail), and USD price. I restrict the sample to retail prices with valid USD values, markets that report both rice and at least one control commodity, and markets with at least six months of data on both sides of the July 2023 ban. The final sample contains 329,030 observations across 59 countries and 1,530 markets.

\subsubsection*{FAO Detailed Trade Matrix} I construct country-level Indian rice import dependence using the Food and Agriculture Organization's bilateral trade flow data for rice (all HS 1006 subcategories) averaged over 2020--2022. For each importing country, the treatment intensity variable is the share of total rice import value sourced from India. This measure ranges from zero (e.g., Thailand, a net exporter) to 0.997 (Bhutan, nearly completely dependent on Indian rice).

\subsubsection*{Commodity classification} I classify WFP commodities into rice (all varieties: imported, local, milled, coarse, etc.) and control staples (maize, millet, sorghum, wheat flour, beans, groundnuts, cassava, potatoes, lentils). Control commodities are chosen because their supply chains are independent of Indian rice trade: they are grown domestically or imported from different source countries. A measurement concern is that WFP's rice category includes all varieties---not only the banned non-basmati white rice but also locally produced rice and, in some markets, basmati or parboiled varieties that were not subject to the ban. This attenuates the measured treatment effect toward zero, making the estimates conservative: if anything, the true pass-through for the banned variety is larger than what I report.

\input{tables/tab1_summary}


\section{Empirical Strategy}

The central challenge is isolating the effect of India's export ban on rice prices from concurrent shocks that affect food prices generally. I address this with a within-market, across-commodity difference-in-differences design augmented by cross-country treatment intensity.

\subsubsection*{Specification} The main estimating equation is:
\begin{equation}
\log(p_{imct}) = \beta \cdot (\text{Rice}_c \times \text{Post}_t \times \text{IndiaShare}_i) + \alpha_{mt} + \gamma_{ic} + \varepsilon_{imct}
\label{eq:main}
\end{equation}
where $i$ indexes countries, $m$ markets, $c$ commodities, and $t$ year-months. $\text{Rice}_c$ is an indicator for rice commodities, $\text{Post}_t$ equals one after July 2023, and $\text{IndiaShare}_i$ is the pre-ban share of country $i$'s rice imports from India (continuous, 0--1). The coefficient $\beta$ measures the differential price change of rice vs.\ control commodities, within the same market, after the ban, scaled by Indian import dependence: it is the pass-through per unit of India share. Market$\times$time fixed effects $\alpha_{mt}$ absorb all local shocks common across commodities within a market-month (currency, transport, seasonal demand, weather, conflict), as well as any lower-order terms that vary only at the market-time level (including $\text{Post}_t$ and $\text{IndiaShare}_i \times \text{Post}_t$). Country$\times$commodity fixed effects $\gamma_{ic}$ absorb time-invariant commodity price levels that differ across countries, including the $\text{Rice}_c$ and $\text{Rice}_c \times \text{IndiaShare}_i$ interactions. What remains after these absorptions is the triple interaction: the within-market, within-country-commodity variation in rice vs.\ control prices that correlates with pre-ban Indian dependence after the ban. Standard errors are clustered at the country level (59 clusters) to account for correlated shocks within national food markets.

\subsubsection*{Identifying assumptions} The key assumption is that, absent the ban, rice and control commodity prices would have followed parallel trends within each market. This is testable: I estimate a dynamic event study replacing $\text{Post}_t$ with quarterly event-time dummies relative to July 2023 and examine whether pre-ban coefficients are jointly zero. The assumption is supported by the institutional structure: market$\times$time fixed effects absorb any shock that affects all commodities in a market equally, and the remaining variation comes only from the differential commodity-specific price change.

\subsubsection*{Threats} The main threat is a contemporaneous shock that differentially affected rice versus non-rice commodities in high-India-dependence countries. I address this through several robustness checks: (i) a pre-ban placebo test with a fake treatment date of January 2022; (ii) leave-one-country-out analysis; (iii) controlling for commodity-specific linear trends; and (iv) trimming price outliers.


\section{Results}

\subsection{Main Estimates}

\Cref{tab:main} reports the main results. Column (1) estimates a simple Rice $\times$ Post effect pooling across all countries: rice prices rose 11.0\% relative to control commodities, though this estimate is imprecise ($p = 0.36$) due to the enormous heterogeneity across importing and exporting nations. Columns (2) and (3) introduce the triple interaction with India Import Share; the sign and magnitude are sensitive to the fixed effects structure. Column (4), the preferred specification with market$\times$year-month and country$\times$commodity fixed effects, yields a pass-through coefficient of 0.131 (SE = 0.028), significant at the 1\% level. A country with 50\% Indian rice dependence experienced a 6.5\% differential rice price increase relative to a country with no dependence. The dramatic shift from Columns (2)--(3) to Column (4) reflects the importance of absorbing persistent cross-country differences in how different commodities are priced: without country$\times$commodity fixed effects, the regression conflates the ban's effect with pre-existing cross-national price level gaps between rice and control staples (e.g., rice is systematically cheaper per kilogram than beans or potatoes in many markets). Once these level differences are absorbed, the triple interaction cleanly identifies the within-country, within-commodity, post-ban differential.

\input{tables/tab2_main}

To translate this estimate: the average highly-dependent country in our sample sources 55\% of rice imports from India. For such a country, the ban increased retail rice prices by approximately 7.2\% relative to non-rice staples---equivalent to a meaningful reduction in real purchasing power for households spending 30--50\% of income on food.

\subsection{Event Study}

\Cref{tab:event} reports quarterly event-study coefficients. The pre-ban quarters ($q = -8$ through $q = -2$) show no significant differential trend between rice and control commodities, with coefficients centered near zero. The effect materializes sharply in the first post-ban quarter ($q = 0$ to $q = +1$), reaching 6.7\% by the second quarter ($p = 0.004$), and persists through $q = +8$. This pattern---flat pre-trends followed by a sharp and sustained break at the exact policy date---is consistent with a causal effect of the export ban rather than a pre-existing trend.

\input{tables/tab3_event}

\subsection{Heterogeneity}

\Cref{tab:hetero} splits the sample at the median India Import Share among importing countries (35\%). In highly dependent countries (mean India share = 55\%), rice prices rose 10.6\% relative to control commodities (SE = 2.1\%, $p < 0.001$). In low-dependence countries, the effect is a precise zero ($-0.001$, SE = 0.044). The difference of 10.7 percentage points (SE = 4.9\%) is significant at the 5\% level. This dose-response pattern---monotonically increasing in treatment intensity---provides further evidence for a causal pass-through channel. Analysis by dependence quartile (among importing countries with positive India share) confirms the gradient: Q1 and Q2 show near-zero effects, while Q3 (10.4\%) and Q4 (11.0\%) are highly significant.

\input{tables/tab5_hetero}

\subsection{Robustness}

\Cref{tab:robust} summarizes robustness checks. The baseline coefficient of 0.131 is remarkably stable across specifications. Two-way clustering by country and year-month slightly increases the standard error to 0.031 without affecting the point estimate. Trimming the top and bottom 1\% of USD prices yields 0.128. Adding commodity-specific linear time trends produces 0.138. The leave-one-country-out range is [0.116, 0.147]---no single country drives the result, though Madagascar and Gambia are the most influential (consistent with their high dependence and large market coverage in the WFP data). The pre-ban placebo test, assigning a fake treatment date of January 2022 using only pre-ban observations, produces an insignificant coefficient of $-0.658$ (SE = 0.473), confirming that the parallel trends assumption holds in the pre-period. Finally, I test for price attenuation following India's partial lifting of the ban in October 2024 via a Minimum Export Price mechanism. Adding a post-reversal interaction to the main specification yields a negative (attenuating) coefficient of $-0.144$ (SE = 0.129). While imprecise---consistent with the MEP being a partial rather than full reversal---the sign is in the expected direction, providing suggestive evidence that the price increase was driven by the supply restriction rather than a permanent structural break.

\input{tables/tab4_robustness}


\section{Discussion}

India's 2023 rice export ban offers a stark lesson in the fragility of concentrated food supply chains. The pass-through from a single exporter's policy decision to retail prices in 59 countries was swift, substantial, and concentrated---a 10.6\% price increase in highly dependent countries within months of the ban. This finding has three implications.

First, aggregate price indices understate the local impact of trade disruptions. National food price inflation figures---the standard metric in the export ban literature---average over markets and commodities, diluting the signal. Within individual markets, the rice-specific price shock was large enough to materially affect household food budgets, particularly in countries where rice accounts for 20--40\% of caloric intake.

Second, the clean dose-response relationship between bilateral trade dependence and price pass-through---effectively zero for diversified importers, over 10\% for dependent ones---suggests that import diversification is a quantitatively important insurance mechanism against unilateral trade restrictions. Countries that had shifted some procurement toward Thai, Vietnamese, or Pakistani rice before 2023 were largely insulated from the shock. This provides direct empirical support for food security strategies that emphasize supplier diversification over stockpiling \citep{gotz2020russia, laborde2021covid}.

Third, this paper cannot identify the welfare consequences beyond retail price increases. The ban likely induced substitution toward other staples, changes in household dietary composition, and adjustments in wholesale market structure that are not captured by the retail price data alone. These remain important avenues for future work.


\section{Conclusion}

When India banned non-basmati white rice exports on July 20, 2023, it did not simply restrict trade---it conducted an unplanned experiment on the price sensitivity of global food markets. Using granular retail price data from 1,530 markets across 59 countries, I show that this single policy decision raised rice prices by over 10\% relative to other staples in import-dependent countries, with the effect emerging within weeks and persisting for over a year. The pass-through was precisely proportional to bilateral trade dependence: countries that had diversified their rice imports experienced no measurable impact. In a world where export restrictions on food commodities are proliferating---from India's wheat ban of 2022 to Indonesia's palm oil restrictions---these results provide a clear empirical foundation for quantifying the costs of supply chain concentration in global food markets.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
