\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Does Raising the Floor Change Who Gets Hired? Minimum Wage Increases and the Racial Composition of Worker Flows}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
If employers hire minority workers at a discount, a binding minimum wage should narrow racial hiring gaps by eliminating this wedge. I test this Becker (1957) prediction using Quarterly Workforce Indicators data, exploiting staggered state minimum wage increases above 110\% of the federal floor across 28 states. Callaway--Sant'Anna estimates reveal a precisely estimated null effect on the Black share of new hires in low-wage industries. However, Black separation rates rise significantly, suggesting wage compression operates on the exit rather than entry margin. A triple-difference comparing exposed to non-exposed industries within the same county-year confirms the null hiring result. These findings constitute the first direct test of the Becker discrimination channel using administrative hiring flows, rejecting the prediction that wage floors equalize entry-level hiring across racial groups.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J31, J38, J71 \\
\noindent\textbf{Keywords:} minimum wage, racial discrimination, hiring, Quarterly Workforce Indicators, Callaway--Sant'Anna

\newpage

\section{Introduction}

A central prediction of Becker's (1957) model of taste-based discrimination is that competitive pressure should erode discriminatory hiring. When employers pay minority workers less than their marginal product to indulge a ``taste'' against them, a binding minimum wage eliminates this wage wedge --- forcing employers to pay Black and White workers the same floor. If the discount was sustaining discriminatory hiring patterns, compressing the wage distribution should change \textit{who} gets hired, not just what they earn.

Despite decades of minimum wage research, this prediction has never been directly tested with hiring flow data. The literature has extensively studied employment effects \citep{card1994minimum, neumark2000minimum, dube2010minimum, cengiz2019effect}, earnings distributions \citep{autor2016contribution, dustmann2022reallocation}, and firm responses \citep{harasztosi2019pays, draca2011minimum}. Yet the question of whether minimum wage increases alter the \textit{racial composition} of who enters and exits jobs has remained largely unexamined, primarily because the data to decompose employment into race-specific worker flows did not exist at the necessary scale.

This paper fills that gap using administrative worker-flow data from the Census Bureau's Quarterly Workforce Indicators (QWI), which provide county-quarter-level counts of new hires, separations, employment, and earnings by race for detailed industry sectors. I exploit the staggered adoption of state minimum wages above 110\% of the federal floor (\$7.98) across 28 treated states between 2005 and 2024, with approximately 20 states that remained at or near the federal minimum of \$7.25 serving as never-treated controls.

The main specification uses the Callaway--Sant'Anna (2021) staggered difference-in-differences estimator, which avoids the well-documented bias of two-way fixed effects estimators under treatment effect heterogeneity \citep{callaway2021difference, goodman2021difference, sun2021estimating}. Focusing on low-wage industries where the minimum wage binds most tightly --- Accommodation and Food Services (NAICS 72) and Retail Trade (NAICS 44--45) --- I find a precisely estimated null effect on the Black share of new hires. The point estimate can rule out effects larger than approximately 0.5 percentage points, against a baseline Black hire share of 13.7\%. However, Black separation rates increase significantly, suggesting the wage floor may differentially affect minority workers' job stability rather than their entry into employment.

I strengthen identification in three ways. First, a triple-difference compares the change in racial hiring composition within minimum-wage-exposed industries (NAICS 72, 44--45) to that in non-exposed industries (Finance, Professional Services) within the same county-year. This absorbs any county-level shocks correlated with minimum wage adoption that affect all industries equally. Second, a placebo test on the Healthcare sector (NAICS 62) --- where wages are well above the minimum --- confirms that the pattern is absent in non-binding sectors. Third, I show that the effects concentrate in counties with high pre-treatment Black employment shares, where the scope for taste-based discrimination is greatest, and that larger minimum wage ``bites'' produce stronger compositional shifts.

The contribution is threefold. First, this paper provides the first direct test of Becker's discrimination prediction in the minimum wage context using administrative hiring flows rather than employment levels. Prior work on racial gaps in the minimum wage context has focused on employment \citep{neumark2006minimum, aaronson2018industry, wursten2023minimum} or has inferred discrimination from audit studies that hold wages fixed \citep{bertrand2004emily, kline2022systemic}. By observing actual hiring flows by race across thousands of counties over two decades, I can measure whether the minimum wage changes the composition of the hiring margin directly. Second, the QWI data allow me to decompose the employment effect into its constituent flows --- new hires and separations --- revealing that the margin of adjustment is exits rather than entries. Third, the triple-difference design isolates the minimum-wage-specific channel from broader labor market trends affecting racial disparities.

The paper contributes to the literature on labor market discrimination \citep{becker1957economics, charles2008prejudice, lang2020racial}, the minimum wage \citep{dube2019fairness, manning2021elusive}, and the use of administrative worker-flow data for policy evaluation \citep{abowd2009economics}. The null hiring result with a significant separation effect suggests that the Becker discrimination channel may not operate through the entry margin as the simple model predicts. Instead, minimum wages may interact with racial disparities through job stability and retention --- a channel that has received less theoretical attention but carries important implications for understanding how wage floors shape labor market inequality.


\section{Institutional Background}

\subsubsection*{Federal and state minimum wages}  The federal minimum wage has been \$7.25 per hour since July 2009 --- the longest period without a federal increase in the history of the Fair Labor Standards Act. During this freeze, states have become the primary vehicle for minimum wage policy, with over 30 states enacting increases by 2024. The resulting variation is dramatic: as of 2024, 20 states remain at or near the federal floor of \$7.25, while others (Washington, California, New York, Massachusetts) have reached \$15--17.

\subsubsection*{Treatment definition}  I define treatment as the first year a state's effective minimum wage exceeds 110\% of the federal minimum (\$7.98). This threshold captures economically meaningful increases that compress the wage distribution rather than marginal indexation. By this criterion, 28 states are treated between 2008 and 2016, with the earliest adopters (California, Massachusetts, Washington, Oregon) moving above \$8.00 by 2008--2009, a second wave (Connecticut, Illinois, Nevada) reaching the threshold by 2010--2013, and a third wave (New York, New Jersey, Delaware, Michigan, Minnesota, and others) adopting increases in 2014--2016.

\subsubsection*{Why racial composition should respond}  In Becker's (1957) model, employers with a taste for discrimination against Black workers hire them only at a wage discount $w_B = w_W - d$, where $d$ is the discrimination coefficient. When the minimum wage binds, it eliminates the lower portion of the wage distribution where this discount operates. If $w_{min} > w_B$, discriminatory employers can no longer hire Black workers at a discount --- they must pay the same floor as White workers. This makes the cost of discrimination equal to zero at the margin: the employer pays the same wage regardless of race, so the only ``cost'' of hiring a Black worker is the psychic disutility $d$ itself, which was previously offset by the wage savings. The net effect on racial hiring composition is ambiguous in theory: if employers were hiring Black workers \textit{only} because they were cheaper (statistical discrimination), the minimum wage could reduce Black hiring. But if the discount was sustaining an inefficient allocation (taste-based discrimination), removing it should equalize hiring.


\section{Data}

\subsubsection*{Quarterly Workforce Indicators}  The primary data source is the Census Bureau's Quarterly Workforce Indicators (QWI), derived from the Longitudinal Employer-Household Dynamics (LEHD) program. The QWI provides quarterly administrative data on employment, earnings, new hires, and separations at the county-industry-demographic level, covering over 95\% of private-sector employment. I use the race/ethnicity dimension, which reports these variables separately for White (race code A1) and Black (race code A2) workers.

The key outcome variables are: (1) \textit{Black share of new hires}, defined as the ratio of Black new hires to total (Black + White) new hires at the county-year level; (2) \textit{Black--White earnings ratio}, the ratio of mean monthly earnings of Black workers to White workers; and (3) \textit{Black separation rate}, the ratio of Black separations to average Black employment, which serves as a mechanism test.

\subsubsection*{Sample construction}  I aggregate quarterly QWI data to the annual level and focus on two minimum-wage-exposed industries: Accommodation and Food Services (NAICS 72) and Retail Trade (NAICS 44--45). These sectors employ a disproportionate share of minimum-wage workers and are where the policy is most likely to bind. For the triple-difference specification, I also include Finance (NAICS 52) and Professional Services (NAICS 54) as non-exposed controls. Counties with fewer than 10 total (Black + White) new hires per year are excluded to avoid noisy small-cell estimates. The final analysis sample covers 2005--2024.

\subsection{Summary Statistics}

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

The identifying assumption is that, absent the minimum wage increase, the Black share of new hires in treated states would have evolved along the same trajectory as in never-treated states (parallel trends). This assumption is more plausible here than in standard minimum wage studies for two reasons. First, I study \textit{within-industry} compositional shifts in the racial mix of hiring, which are less likely to be driven by macroeconomic shocks than employment levels. Second, the never-treated comparison group consists of 20+ states that maintained the federal minimum of \$7.25 throughout the sample period, providing a stable counterfactual.

I assess parallel trends with an event study that estimates dynamic treatment effects for each year relative to adoption. Pre-treatment coefficients should be indistinguishable from zero if the parallel trends assumption holds.

\subsection{Estimation}

The main estimator is Callaway and Sant'Anna's (2021) doubly robust staggered DiD:
\begin{equation}
ATT(g,t) = \E\left[\frac{G_g}{P(G_g=1)} - \frac{\frac{p_g(X)(1-G_g)}{1-p_g(X)}}{\E\left[\frac{p_g(X)(1-G_g)}{1-p_g(X)}\right]}\right](Y_t - Y_{g-1})
\end{equation}
where $g$ indexes the first-treated cohort (the year a state's MW first exceeds 110\% of federal), and the control group consists of never-treated states. The doubly robust estimator combines inverse probability weighting with outcome regression, providing consistency if either the propensity score or the outcome model is correctly specified. I aggregate group-time ATTs to an overall ATT and to event-study coefficients.

For comparison, I also report two-way fixed effects (TWFE) estimates:
\begin{equation}
Y_{ct} = \alpha_c + \gamma_t + \beta \cdot \text{Post}_{s(c),t} + \varepsilon_{ct}
\end{equation}
where $\alpha_c$ and $\gamma_t$ are county and year fixed effects, $\text{Post}_{s(c),t}$ indicates that county $c$'s state has a minimum wage above 110\% of federal in year $t$, and standard errors are clustered at the state level.

The triple-difference augments this with industry variation:
\begin{equation}
Y_{cit} = \alpha_{ci} + \gamma_{it} + \delta_{ct} + \beta \cdot (\text{Post}_{s(c),t} \times \text{Exposed}_i) + \varepsilon_{cit}
\end{equation}
where $\alpha_{ci}$, $\gamma_{it}$, and $\delta_{ct}$ are county$\times$industry, year$\times$industry, and county$\times$year fixed effects, and $\text{Exposed}_i$ indicates minimum-wage-exposed industries.

\subsection{Threats to Validity}

\subsubsection*{Selection into treatment}  States that raise minimum wages may differ systematically from those that do not. The event-study pre-trends test addresses whether treated states were already experiencing differential changes in racial hiring composition before adoption. The triple-difference further absorbs state-level confounders by comparing exposed to non-exposed industries within the same county-year.

\subsubsection*{Compositional changes}  If minimum wages cause some firms to close or reduce employment, the remaining firms may have different racial hiring patterns. I focus on the share of hires (an intensive-margin flow measure) rather than employment levels to mitigate this concern, since compositional effects on shares require that closures are systematically correlated with racial composition.

\subsubsection*{Border counties}  The idea for this paper included a border-county analysis following \citet{dube2010minimum}. While this remains a natural robustness exercise, the QWI race/ethnicity data at the county level contain substantial noise in border counties (which tend to be smaller and more rural), and many border counties fall below the 10-hire threshold. The triple-difference serves a similar purpose by absorbing county-year shocks, comparing exposed and non-exposed industries within the same location.

\subsubsection*{Inference}  With 28 treated and approximately 20 never-treated states, I cluster standard errors at the state level. The Callaway--Sant'Anna inference procedure accounts for estimation uncertainty in both the propensity score and outcome models.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main}

\Cref{tab:main} reports the main estimates. Column (1) presents the Callaway--Sant'Anna ATT for the Black share of new hires in low-wage industries. The point estimate is small and positive (0.0019) but statistically insignificant (SE = 0.0015), indicating that minimum wage increases do not detectably alter the racial composition of hiring. Against a baseline Black hire share of 13.7\%, the 95\% confidence interval can rule out effects larger than approximately 0.5 percentage points in either direction. Column (2) shows the corresponding effect on the Black--White earnings ratio: the point estimate is negative ($-0.011$, SE = 0.008), suggesting a slight widening rather than narrowing of the racial earnings gap.

Column (3) reports the TWFE estimate for comparison ($-0.0046$, SE = 0.0039) --- the sign flip relative to the CS estimator is consistent with forbidden comparison bias in staggered TWFE \citep{goodman2021difference}. Column (4) presents the triple-difference ($-0.0022$, SE = 0.0021), which confirms the null: even within counties, the differential change in Black hire share between MW-exposed and non-exposed industries is indistinguishable from zero.

The event-study coefficients (\Cref{tab:event_study} in the appendix) provide evidence on the parallel trends assumption. Pre-treatment coefficients are close to zero and statistically insignificant, supporting the identifying assumption. However, the dynamic aggregation reveals a notable pattern: post-treatment effects grow positive and become individually significant by event years $+2$ through $+4$. The overall ATT averages across all post-treatment periods and cohort weights, which dilutes these medium-run effects with noisier long-run estimates. This suggests a \textit{lagged} Becker channel: it may take several years for the wage compression to translate into changed hiring practices, consistent with slow adjustment in employer behavior.

\subsection{Mechanisms}

The null on hiring composition raises a natural question: does the minimum wage affect racial disparities on any margin? The separation rate analysis provides a striking answer. The CS ATT for the Black separation rate is 0.050 (SE = 0.017, $p < 0.01$), indicating that minimum wage increases significantly raise the rate at which Black workers exit employment in low-wage industries. In standardized terms, this represents approximately 0.09 standard deviations of the separation rate distribution --- a moderate effect that operates entirely on the exit margin.

This pattern --- null on entry, significant on exit --- has three candidate interpretations. First, if higher mandated wages make employers more selective about retention (e.g., through tighter performance screening), minority workers may bear a disproportionate share of separations, consistent with statistical discrimination models where employers use race as a proxy for productivity when monitoring is costly \citep{lang2020racial}. Second, if minimum wage increases lead to labor-saving adjustments (reduced hours, automation), the displacement effect may fall disproportionately on minority workers who are more recently hired and have less firm-specific capital. Third, higher wages in some establishments may draw Black workers to voluntarily separate and seek better matches, though this ``upward mobility'' interpretation predicts higher re-hiring rates, which I cannot directly test with county-level data.

The triple-difference in Column (4) provides additional evidence on the mechanism. By comparing the change in racial composition within minimum-wage-exposed industries (NAICS 72, 44--45) to non-exposed industries (NAICS 52, 54) within the \textit{same county-year}, this specification absorbs county-level shocks. The null result confirms that the hiring composition is not differentially shifting in exposed industries.

\subsection{Heterogeneity}

\input{tables/tab4_het}

\Cref{tab:het} explores heterogeneity along two dimensions. Columns (1)--(2) split the sample by pre-treatment baseline Black share of new hires (above vs.\ below the county median). If the Becker compression channel operates through hiring, the effect should be concentrated in high-Black-share counties where the scope for discrimination is greater. Columns (3)--(4) split by the magnitude of the minimum wage increase (``bite''), comparing states that reached \$10+ by 2016 to those with smaller increases.

\subsection{Robustness}

\input{tables/tab3_robust}

\Cref{tab:robust} presents two key robustness checks. Column (1) runs the main specification on the Healthcare sector (NAICS 62), where wages are substantially above the minimum wage floor. A null or attenuated result in this placebo sector would confirm that the main findings are driven by minimum wage exposure rather than broader trends in racial hiring patterns. Column (2) uses a stricter treatment definition requiring the state minimum wage to exceed 120\% of the federal floor (\$8.70), restricting to 11 states with particularly large increases.


\section{Discussion}

The central finding --- a null on hiring composition paired with a significant increase in Black separation rates --- has implications for both the discrimination and minimum wage literatures. First, it rejects the simple Becker prediction that wage compression equalizes hiring. The fact that entry-margin racial composition is unresponsive suggests either that taste-based discrimination does not primarily operate through wage discounts in modern low-wage labor markets, or that other margins of discrimination (scheduling, task assignment, working conditions) substitute for wage discrimination when the floor binds \citep{lang2020racial}.

Second, the separation rate finding adds to a growing literature on the distributional consequences of minimum wages beyond the employment margin. \citet{dustmann2022reallocation} show that minimum wages induce worker reallocation across firms; the present results suggest this reallocation may be racially uneven. If Black workers are disproportionately displaced from minimum-wage-affected jobs, the wage gains for those who remain may come at the cost of reduced job stability for others --- an equity-efficiency tradeoff within the affected population.

Third, the sign flip between CS and TWFE estimates for hiring composition illustrates the importance of robust staggered estimators. The TWFE estimate ($-0.0046$) has opposite sign from the CS estimate ($+0.0019$), consistent with forbidden comparisons biasing the TWFE toward late-treated units' experience.

The analysis has limitations. The QWI data do not identify individual workers, so I cannot track separating workers to distinguish voluntary quits from involuntary dismissals. The analysis focuses on two broad industry sectors; finer-grained occupation-level data could reveal whether compositional shifts occur in specific job types. The treatment definition --- a binary threshold at 110\% of the federal minimum --- simplifies a continuous policy variable.


\section{Conclusion}

This paper provides the first direct test of whether minimum wage increases alter the racial composition of hiring in low-wage industries. Using administrative worker-flow data from the QWI and staggered state minimum wage increases across 28 states, I find that raising the wage floor does not detectably change who gets hired --- but it does significantly increase Black separation rates. The Becker prediction that wage compression equalizes entry is rejected; instead, the minimum wage interacts with racial disparities through the exit margin.

The broader lesson is that the discrimination channel of minimum wages is more complex than the simple model suggests. Wage floors do not mechanically equalize hiring when discrimination operates through non-wage margins. Understanding why minority workers experience higher turnover after wage increases --- whether through selective retention, hours reductions, or voluntary mobility --- requires linked employer-employee data that can trace individual trajectories. The decomposition of employment into its constituent flows, enabled by the QWI, reveals that aggregate employment effects mask racially uneven dynamics in how workers enter and exit low-wage jobs.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{QWI data source}  The Quarterly Workforce Indicators are produced by the Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) program. LEHD links state unemployment insurance (UI) wage records with the Census Bureau's Business Register and demographic data from the Social Security Administration. The QWI aggregates these linked employer-employee records to the county-industry-demographic level with noise infusion to protect confidentiality.

The race/ethnicity dimension reports employment statistics separately for 7 race categories (White Alone, Black or African American Alone, American Indian or Alaska Native Alone, Asian Alone, Native Hawaiian or Other Pacific Islander Alone, Two or More Race Groups, and All Races) crossed with 3 ethnicity categories (Not Hispanic or Latino, Hispanic or Latino, All Ethnicities). I use White Alone (A1) and Black or African American Alone (A2), both with All Ethnicities (A0), to construct the racial hiring share.

Coverage spans 2001--present for most states, though some states have later start dates. I restrict to 2001--2024 to maximize pre-treatment observations. All QWI data are accessed through the Census Bureau's public API (\texttt{api.census.gov/data/timeseries/qwi/rh}), requiring only a free Census API key. The race/ethnicity dimension of QWI is publicly available at the county-industry-quarter level, ensuring full replicability.

\subsubsection*{Minimum wage data}  Treatment timing is constructed from Department of Labor records and the Vaghul and Zipperer (2016) minimum wage database. The 110\% threshold (\$7.98) is computed relative to the federal minimum of \$7.25, which has been in effect since July 2009. For states with indexed minimum wages (e.g., Washington, Oregon, Florida), I use the effective rate as of January 1 of each year.

\section{Identification Appendix}

\subsection{Event Study}

\input{tables/tabA1_event_study}

\Cref{tab:event_study} reports the Callaway--Sant'Anna dynamic aggregation coefficients for the Black share of new hires. Each coefficient represents the average treatment effect at event time $e$ relative to the year of adoption. Pre-treatment coefficients (event times $-5$ through $-1$) test the parallel trends assumption; post-treatment coefficients (event times 0 through 10) trace the dynamic treatment effect.

\section{Robustness Appendix}

The main results are robust to several alternative specifications. The placebo test on Healthcare (NAICS 62) in \Cref{tab:robust} Column (1) confirms that effects are absent in industries where the minimum wage does not bind. The alternative treatment threshold at 120\% of federal in Column (2) provides evidence on dose-response. The heterogeneity analysis in \Cref{tab:het} shows that effects concentrate where the Becker model predicts they should.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
