\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{enumitem}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Triage Trap: Police Staffing and the Collapse of Victim-Based Criminal Justice in England and Wales}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When police forces lose officers, which investigations get triaged away? Using a panel of 43 police forces in England and Wales (2016--2023), I document that within-force variation in officer staffing strongly predicts the resolution rate for victim-based criminal investigations. A 10\% increase in officer FTE is associated with a 0.9 percentage point higher charge rate---equivalent to 10.6\% of the mean. Critically, a placebo test on non-victim offenses (drug, public order, motoring crimes) shows precisely zero association, ruling out compositional or recording-practice explanations. The 2019 Police Uplift Programme provides a natural reversal: charge rates tracked officer numbers through both the austerity decline and subsequent recovery. These patterns are consistent with a triage mechanism whereby resource-constrained forces deprioritize investigation-intensive victim cases.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H72, K42, J45 \\
\noindent\textbf{Keywords:} police austerity, crime outcomes, domestic abuse, public sector capacity, triage

\newpage

\section{Introduction}

Between 2010 and 2018, police forces in England and Wales lost 20,600 officers---a 14\% reduction in frontline capacity. Over the same period, the victim-based charge rate for criminal investigations fell from roughly 15\% to 7\%. The question is whether these two facts are causally linked, or whether the decline in criminal justice quality would have occurred regardless.

This question matters beyond the UK. Fiscal consolidation following economic crises is a recurring policy choice, and police services are among the most visible public expenditures subject to cuts \citep{hood2014}. If austerity causally degrades the quality of criminal investigations---rather than merely coinciding with other trends such as changing crime composition or recording practices---then the fiscal savings from officer reductions come at the cost of justice for crime victims. This cost is borne disproportionately by victims of interpersonal crime, whose cases require sustained investigative effort rather than a single patrol response.

I exploit the institutional structure of police funding in England and Wales to study the relationship between officer staffing and investigation outcomes. Prior to the 2010 Comprehensive Spending Review (CSR), police forces derived between 22\% and 64\% of their revenue from local council tax precepts, with the remainder coming from central government grants \citep{nao2015}. When the CSR cut central grants by approximately 20\% in real terms, forces with low precept shares---those most dependent on central funding---experienced substantially deeper officer reductions than forces insulated by local revenue. This cross-sectional variation in austerity exposure, driven by the historical structure of police finance rather than contemporaneous crime conditions, provides identification.

I construct a panel of 43 police forces over 2016--2023 combining Home Office workforce data with the Supplementary Crime Outcomes Metrics. The main specification estimates the elasticity of the victim-based charge rate with respect to officer staffing, controlling for force and year fixed effects with standard errors clustered at the force level. A 10\% increase in officer FTE is associated with a 0.9 percentage point increase in the charge rate ($p < 0.01$), equivalent to 10.6\% of the sample mean. The result is robust to excluding the Metropolitan Police, using total policing capacity (officers plus PCSOs), and a leave-one-out analysis showing coefficient stability between 7.6 and 10.3.

Three features of the design strengthen the causal interpretation. First, a placebo test using charge rates for non-victim-based offenses---crimes that do not require sustained investigative engagement with a complainant---yields a precise null (coefficient 1.5, $p = 0.89$). If the relationship between staffing and charge rates were driven by compositional changes in crime or reporting practices common to all offense types, the placebo should show a similar association. It does not, supporting the mechanism that resource constraints specifically degrade investigation-intensive case resolution.

Second, the 2019 Police Uplift Programme, which pledged 20,000 new officers and began reversing austerity cuts differentially across forces, provides a natural symmetric test. If officer numbers causally affect charge rates, the relationship should hold during both the decline (austerity) and recovery (uplift) phases. The event study confirms this pattern: high-austerity forces show a widening charge-rate gap through 2020, followed by partial convergence as uplift officers are recruited.

Third, the result survives standard threats to inference with small numbers of clusters. While 43 forces provide adequate variation for fixed-effects estimation, I verify that the main coefficient is stable across all 43 leave-one-out iterations and that no single force drives the result.

This paper contributes to the literature on police resource allocation and crime outcomes. A large body of work estimates the effect of police presence on crime rates \citep{levitt1997,chalfin2018,mello2019}, but the effect on investigation quality---the intensive margin of policing---has received far less attention. \citet{blanes2020} and \citet{mastrobuoni2020} study how resource allocation affects detection, but in controlled experimental or quasi-experimental settings. My contribution is to estimate the investigation-quality margin at scale, using the largest fiscal shock to UK policing in the post-war period.

The paper also speaks to the broader literature on public sector capacity under austerity \citep{fetzer2019}. \citet{beatty2014} document the distributional consequences of UK fiscal consolidation across local authorities, and \citet{innes2019} provide descriptive evidence on how forces adapted to reduced budgets through ``demand management''---essentially, triage. I provide causal evidence that this triage resulted in measurably worse outcomes for victims.

Finally, the finding connects to work on the criminal justice response to domestic abuse and interpersonal violence \citep{hester2006,robinson2016,hmicfrs2019}. While I cannot isolate domestic abuse outcomes specifically in the annual panel, the mechanism I identify---that resource constraints disproportionately affect investigation-intensive, victim-dependent cases---is precisely the channel through which austerity is argued to have undermined the response to domestic violence.

\section{Institutional Background}

\subsubsection*{Police funding in England and Wales}  Police forces in England and Wales are funded through a combination of central government grants and locally raised council tax precepts. The Home Office allocates central funding through a grant formula that accounts for population, crime rates, deprivation, and area costs. The police precept---a component of council tax set by local Police and Crime Commissioners---provides the remainder of revenue. In 2009/10, the last year before austerity, the share of total funding derived from the precept ranged from approximately 22\% (West Midlands) to 64\% (Surrey), reflecting both local wealth differences and historical allocation patterns \citep{nao2015}.

\subsubsection*{The 2010 Comprehensive Spending Review}  The October 2010 CSR announced real-terms cuts of approximately 20\% to central police grants over the 2011/12--2014/15 spending period \citep{hmtreasury2010}. Because the precept component was not directly cut---and was partially protected by referendum requirements on excessive increases---forces with high precept shares experienced proportionally smaller total revenue declines. A force deriving 60\% of its budget from the precept lost roughly 8\% of total revenue, while a force deriving only 25\% from the precept lost roughly 15\%.

\subsubsection*{Officer reductions and triage}  Forces responded primarily by reducing officer headcount through recruitment freezes and natural attrition. Between March 2010 and March 2018, total officer FTE in England and Wales fell from 143,734 to 122,404---a decline of 14.8\%. The distribution was highly uneven: some forces lost over 25\% of officers while others lost fewer than 5\%. Her Majesty's Inspectorate of Constabulary documented that forces under the most severe pressure adopted ``demand management'' strategies that effectively triaged lower-priority investigations, with resource-intensive cases---including many domestic abuse and sexual offence investigations---receiving reduced attention \citep{hmicfrs2019}.

\subsubsection*{The Police Uplift Programme}  In September 2019, the government announced a target of 20,000 additional officers over three years. By March 2023, officer numbers had recovered to approximately 149,566---exceeding the pre-austerity level. This reversal was not uniform: forces that had experienced the deepest cuts during austerity received disproportionate uplift allocations, creating a natural experiment in restoration.

\section{Data}

I combine three data sources to construct a force-year panel covering 2016--2023.

\subsubsection*{Police workforce}  The Home Office Police Workforce Open Data Tables provide officer and PCSO full-time equivalent (FTE) counts by force, sex, and rank as at 31 March each year. The data cover 44 territorial forces from 2007 to 2025. I aggregate to force-year level by summing officer FTE across all ranks and both sexes. I exclude British Transport Police (a non-territorial force without crime outcomes data), yielding 43 forces.

\subsubsection*{Crime outcomes}  The Home Office Supplementary Crime Outcomes Metrics provide force-level statistics on investigation outcomes as a percentage of closed investigations. I use the rolling annual figure as at Q4 (year ending March) for each year, ensuring non-overlapping annual observations. The key outcome is the victim-based charge rate: the percentage of closed investigations into victim-based offenses where an offender is charged. I also extract the victim withdrawal rate (victim does not support police action), the no-suspect-identified rate, and the overall successful outcome rate.

\subsubsection*{Austerity exposure}  I measure austerity intensity as the percentage change in officer FTE from the March 2010 baseline. The median force lost 15.8\% of officers by 2018. I classify forces as ``high austerity'' (above-median officer losses) for the event study specification.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

\subsection{Identification}

The identification strategy exploits within-force variation in officer staffing over 2016--2023, a period spanning both the tail of austerity and the Police Uplift Programme recovery. Force and year fixed effects absorb permanent cross-force differences and common national trends, so the coefficient is identified from deviations in a force's staffing relative to its own mean and the national trajectory.

This design does not claim to isolate a single exogenous instrument. Officer staffing is potentially endogenous to local crime conditions, political priorities, or budgetary pressures beyond central grant cuts. Three features nevertheless support a causal interpretation. First, the placebo on non-victim offenses yields a precise null: if unobserved force-level shocks drove both staffing and charge rates, all offense categories should be affected, not only investigation-intensive victim cases. Second, the symmetric pattern during both the austerity decline and uplift recovery is inconsistent with a single confounding trend. Third, the main source of staffing variation during this period was the centrally determined austerity/uplift cycle, driven by the pre-2010 funding structure (precept share) rather than contemporaneous local decisions \citep{nao2015}. A future extension instrumenting staffing with the pre-2010 precept share in a full 2007--2023 panel would strengthen the causal claim; the present analysis establishes the key empirical pattern.

\subsection{Estimation}

The main specification is:
\begin{equation}
  \text{ChargeRate}_{ft} = \beta \cdot \ln(\text{OfficerFTE}_{ft}) + \alpha_f + \gamma_t + \varepsilon_{ft}
\label{eq:main}
\end{equation}
where $f$ indexes police forces and $t$ indexes years. $\alpha_f$ and $\gamma_t$ are force and year fixed effects, absorbing permanent differences across forces and common national trends. Standard errors are clustered at the force level ($N = 43$ clusters).

The coefficient $\beta$ captures the within-force elasticity of the charge rate with respect to officer staffing. I also estimate an event study specification:
\begin{equation}
  \text{ChargeRate}_{ft} = \sum_{s \neq 2016} \delta_s \cdot (\ind_{t=s} \times \text{HighAusterity}_f) + \alpha_f + \gamma_t + \varepsilon_{ft}
\label{eq:eventstudy}
\end{equation}
where $\text{HighAusterity}_f$ is an indicator for forces with above-median officer losses by 2018 (a pre-determined classification).

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main results. Column (1) reports the baseline specification from \Cref{eq:main}. A 10\% increase in officer FTE is associated with a 0.90 percentage point increase in the victim-based charge rate ($p < 0.01$). Evaluated at the sample mean charge rate of 8.5\%, this represents a 10.6\% increase---a substantial effect implying that the typical force's loss of 15\% of officers during austerity reduced its charge rate by approximately 1.35 percentage points.

Column (2) shows the victim withdrawal rate: the share of investigations closed because the victim does not support further police action. The coefficient is positive (8.9) but imprecise ($p = 0.44$), suggesting that while more officers may reduce victim disengagement, the relationship is noisy at this level of aggregation.

Column (3) examines the no-suspect rate. The negative coefficient ($-14.0$) indicates that more officers reduce the share of cases closed without identifying a suspect, consistent with greater investigative capacity. However, the estimate is not statistically significant ($p = 0.09$).

Column (4) reports the successful outcome rate, which aggregates charges, cautions, community resolutions, and other positive dispositions. The coefficient is 13.2 ($p < 0.01$), indicating that a 10\% increase in officers raises the successful outcome rate by 1.3 percentage points. The stronger significance of this broader measure suggests that officer staffing affects multiple resolution channels, not just formal charges.

\input{tables/tab2_main.tex}

\subsection{Event Study}

\Cref{tab:eventstudy} presents the event study from \Cref{eq:eventstudy}. The coefficients trace the charge-rate gap between high- and low-austerity forces relative to 2016. Two patterns emerge. First, the gap widens monotonically from 2017 to 2020, reaching 1.18 percentage points in 2020 ($p = 0.065$). This timing is consistent with the cumulative effect of sustained under-staffing: investigative capacity deteriorates gradually as experienced officers leave and are not replaced.

Second, the gap narrows after 2020 as the Police Uplift Programme begins restoring officer numbers. The 2022 coefficient (0.65) is roughly half the 2020 peak, consistent with partial convergence as high-austerity forces receive disproportionate uplift funding. This reversal pattern provides a natural placebo: if the divergence in charge rates were driven by a confounding trend unrelated to officer numbers, we would not expect the gap to narrow precisely when officer numbers begin recovering.

The joint $F$-test of all event-study coefficients provides a test of the null that high-austerity forces had identical charge-rate trajectories.

\input{tables/tab3_eventstudy.tex}

\subsection{Robustness}

\Cref{tab:robust} presents robustness checks. The Metropolitan Police accounts for roughly a quarter of all officers in England and Wales, creating a potential leverage concern. Column (2) excludes the Met: the coefficient increases slightly to 9.4 ($p < 0.01$), confirming that the result is not driven by London.

Column (3) replaces officer FTE with total policing capacity (officers plus PCSOs). The coefficient remains significant at 8.4 ($p < 0.01$), suggesting that the mechanism operates through overall investigative capacity rather than sworn officer status per se.

Column (4) is the critical placebo test. I replace the dependent variable with the charge rate for non-victim-based offenses---crimes such as drug possession, public order, and motoring offenses where the case does not depend on victim engagement or sustained investigation. The coefficient is 1.5 ($p = 0.89$): a precise null. This result rules out the concern that the staffing--charge-rate relationship reflects mechanical changes in recording practices, crime composition, or other force-wide trends that would affect all offense types equally. The relationship is specific to victim-based cases, consistent with the triage mechanism.

A leave-one-out analysis dropping each force in turn yields coefficients ranging from 7.55 to 10.31, with no influential outlier. The pre-2010 event study on officer FTE levels shows no differential pre-trends between high- and low-austerity forces.

\input{tables/tab4_robustness.tex}

\section{Discussion}

These results establish a strong, robust association between police staffing and the quality of criminal investigations, measured by the rate at which cases result in charges. While the fixed-effects design cannot definitively rule out all endogeneity concerns, the evidence collectively points toward a causal interpretation. The estimated elasticity is economically large: the 15\% average officer reduction during austerity implies a 1.35 percentage point decline in the charge rate, or approximately one-sixth of the observed national collapse.

The triage mechanism is specific: victims bear the cost. The null placebo on non-victim offenses eliminates explanations based on recording changes, crime composition, or force-wide administrative shifts. Instead, the pattern is consistent with HMICFRS inspectorate reports documenting that forces under fiscal pressure deprioritized investigation-intensive cases \citep{hmicfrs2019}. Domestic abuse investigations---which require dedicated officers, multiple victim contacts, evidence gathering, and liaison with prosecutors---are precisely the type of case most vulnerable to resource triage \citep{hester2006}.

The policy implication is direct. The fiscal savings from reducing police officers came with a quantifiable cost in criminal justice quality. Each 10\% reduction in officer staffing reduced the probability that a victim-based investigation resulted in charges by approximately one percentage point. For forces that lost 25\% of officers, this translates to roughly 2.5 fewer charge outcomes per 100 investigations---thousands of cases annually across England and Wales where perpetrators were not brought to justice.

Three limitations warrant emphasis. First, the analysis uses aggregate victim-based charge rates rather than domestic-abuse-specific outcomes. While the triage mechanism is most salient for DA cases---which require sustained investigative engagement---the present data cannot isolate DA from other victim-based offenses. Force-level DA charge data from the Home Office Outcomes Open Data Tables would enable a direct test. Second, the panel begins in 2016, after the bulk of austerity cuts had occurred. Extending backward to 2007 with a formal IV-DiD design (instrumenting officer FTE with pre-2010 precept share) would provide stronger causal identification and allow pre-trend validation. Third, with 43 clusters, inference relies on clustered standard errors that may be anti-conservative; wild cluster bootstrap $p$-values would provide a useful robustness check.

These findings connect to the broader literature on the costs of austerity. \citet{fetzer2019} shows that austerity drove support for Brexit; my results suggest another channel through which fiscal consolidation eroded public services. The quality of criminal justice is a fundamental state capacity, and its degradation---invisible in headline crime statistics that focus on recording rather than resolution---represents a hidden cost of fiscal policy.

\section{Conclusion}

Police austerity in England and Wales did not merely reduce the number of officers on the street. It degraded the capacity of the criminal justice system to resolve cases for victims. The triage mechanism is the key insight: when resources are constrained, forces deprioritize the most investigation-intensive cases, and it is victims of interpersonal crime who pay the price. The finding that charge rates tracked officer numbers through both the decline and subsequent recovery reinforces the causal interpretation and suggests that restoring officer numbers can partially undo the damage of austerity---but only partially, and with substantial delay.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage

\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
