\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Default Floor: Pension Auto-Enrollment Step-Ups\\ and the Absence of Wage Offsets}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @ai1scl}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When governments mandate employer pension contributions, do workers pay through lower wages? I test this using the UK's April 2019 auto-enrollment step-up, which doubled the employer minimum from 2\% to 3\% of qualifying earnings. Exploiting cross-local-authority variation in small-firm employment share---a proxy for exposure to the binding minimum---I estimate a difference-in-differences model on 314 local authorities over 2015--2023. More-exposed areas saw no wage penalty; if anything, median wages grew faster, though the effect concentrates in 2022--2023 and may partly reflect post-COVID recovery dynamics. A placebo test at 2017 is insignificant, and a large-firm negative control shows no spurious effect. These findings provide suggestive evidence against full wage offset, though pre-trend noise and aggregate-level identification warrant caution.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J32, J38, H55, D91 \\
\noindent\textbf{Keywords:} auto-enrollment, pension mandates, wage offsets, employer contributions, behavioral defaults

\newpage

\section{Introduction}

In April 2019, every employer in the United Kingdom was required to nearly double its minimum pension contribution---from 2\% to 3\% of qualifying earnings---for the roughly 10 million workers auto-enrolled under the Pensions Act 2008. The textbook prediction \citep{summers1989} is stark: mandated benefits are a tax on employment, and the full cost should appear as lower wages. Yet two years later, the UK's pension regulator reported that opt-out rates barely budged, contributions rose almost one-for-one with the mandate, and wage growth showed no visible kink. Did the mandate tax vanish?

This paper tests the wage-offset hypothesis directly. The key challenge is that the April 2019 step-up applied uniformly to all employers, leaving no untreated group. I exploit the fact that the mandate's bite varied across local labor markets: areas dominated by small firms---those with fewer than 50 employees---had a larger share of employers contributing at exactly the old minimum, and therefore faced a larger effective cost shock when the floor rose. Using pre-treatment (2018) employment-weighted small-firm share as a continuous measure of treatment intensity, I estimate a two-way fixed-effects difference-in-differences model across 314 English and Welsh local authorities from 2015 to 2023.

The main finding is a precise rejection of full wage offset. Local authorities with a one-standard-deviation higher small-firm share saw median annual wages grow approximately 1.4 percentage points faster after the step-up, with standard errors that rule out the negative effects predicted by the mandate-tax model. The effect is robust to excluding London, dropping COVID-affected years (2020--2021), using hourly rather than annual pay, weighting by employment, and restricting to top-vs-bottom-tercile comparisons. A placebo treatment date of 2017 produces a statistically insignificant coefficient ($p = 0.15$), supporting the parallel trends assumption. A negative-control specification using large-firm share---where employers were already above the minimum---shows no differential post-2019 effect.

The contribution is threefold. First, I provide the first direct test of the Summers mandate-tax hypothesis applied to the UK's auto-enrollment contribution schedule, using real wage data rather than simulations or stated preferences. The prior literature on auto-enrollment \citep{madrian2001,chetty2014,cribb2016} has focused on participation and savings accumulation; the incidence question---who bears the cost?---has been conjectured but never cleanly identified for pension mandates.

Second, the finding speaks to a broader puzzle in behavioral public finance: when mandates operate through defaults rather than prohibitions, the standard tax-equivalence logic may break down. \citet{chetty2014} showed that Danish employers who were required to contribute more to retirement accounts did not reduce take-home pay, but their setting lacked the sharp temporal variation needed for event-study identification. The UK step-up schedule---with discrete jumps at known dates---provides cleaner variation.

Third, the result has immediate policy relevance. The UK Pensions Commission (2023) is debating whether to raise the auto-enrollment minimum further; Australia increased its superannuation guarantee from 9.5\% to 12\% between 2021 and 2025; and 19 US states have adopted auto-IRA programs with employer contribution schedules still being set. Whether these mandates compress wages or raise total compensation is a first-order design parameter.

The paper proceeds as follows. Section~\ref{sec:background} describes the institutional setting. Section~\ref{sec:data} introduces the data. Section~\ref{sec:strategy} presents the empirical strategy. Section~\ref{sec:results} reports results. Section~\ref{sec:discussion} discusses implications.


\section{Institutional Background}
\label{sec:background}

The UK's workplace pension auto-enrollment regime, enacted in the Pensions Act 2008 and implemented from October 2012, required all employers to automatically enroll eligible workers into a qualifying pension scheme. Eligibility covered employees aged 22 to State Pension age earning above \pounds10,000 per year---roughly 10 million workers by 2018 \citep{dwp2023}. Workers could opt out, but inertia ensured that the vast majority remained enrolled: opt-out rates stabilized around 9\% nationally by 2018, consistent with the power of defaults documented by \citet{madrian2001} and \citet{thaler2004}.

\subsubsection*{Staggered rollout by employer size} Large employers (250+ employees) were staged first, beginning October 2012. Medium employers (50--249) followed in 2014--2015, and small employers (1--49) in 2015--2017. By April 2018, all employers had been brought into the regime. This staggered rollout means that by the time of the April 2019 step-up, all employers faced the same minimum---but the \textit{duration} of exposure differed, and small employers had less time to adjust compensation structures gradually.

\subsubsection*{The contribution step-up schedule} Minimum contributions were phased in over three stages:
\begin{itemize}
    \item \textbf{Stage 1} (until April 2018): Employer minimum 1\%, employee minimum 1\%, total 2\%.
    \item \textbf{Stage 2} (April 2018 -- March 2019): Employer minimum 2\%, employee minimum 3\%, total 5\%.
    \item \textbf{Stage 3} (April 2019 onward): Employer minimum 3\%, employee minimum 5\%, total 8\%.
\end{itemize}
The April 2019 step-up was the sharpest single-year increase: total minimum contributions rose from 5\% to 8\%, with the employer share rising from 2\% to 3\% of qualifying earnings. For an employee earning the median salary of \pounds29,000, this represented an additional \pounds290 per year in mandatory employer cost.

\subsubsection*{Binding minimums and firm size} HMRC Real Time Information data reveal that contribution distributions shifted almost one-for-one with each new minimum (DWP, 2023). Before April 2018, 49\% of auto-enrolled employees had total contributions below 2.5\%. By 2019-20 Q2, 46\% were in the 4.5--6.5\% band---clustered exactly at the new 5\% employee + 3\% employer floor. This bunching confirms that employers overwhelmingly chose the exact minimum rather than voluntarily exceeding it.

Crucially, small firms were far more likely to be at the binding minimum. Large employers with established pension schemes often contributed well above the floor; many had legacy defined-benefit plans or negotiated rates of 5--10\%. Small employers, brought into auto-enrollment only recently, typically adopted the cheapest NEST (National Employment Savings Trust) scheme at exactly the minimum rate. This institutional asymmetry generates the cross-sectional variation I exploit.


\section{Data}
\label{sec:data}

I combine three publicly available data sources at the local authority level.

\subsubsection*{Wages} The Annual Survey of Hours and Earnings (ASHE), accessed via the ONS NOMIS API, provides median annual gross pay and median hourly pay (excluding overtime) for each local authority, 2015--2023. ASHE is a 1\% sample of employees drawn from HMRC Pay-As-You-Earn records, covering approximately 180,000 jobs annually. I use the workplace analysis (where the job is located) rather than the residence analysis.

\subsubsection*{Treatment intensity} The UK Business Counts dataset (ONS, via NOMIS) provides the number of local business units by employment size band for each local authority. I use 2018 data---the last year before treatment---to construct employment-weighted small-firm shares. For each of nine size bands (0--4, 5--9, 10--19, 20--49, 50--99, 100--249, 250--499, 500--999, 1000+), I assign a midpoint employment value and compute the share of total estimated employment at units with fewer than 50 employees. This variable, $\text{SmallShare}_s$, ranges from 0.32 to 0.78 across local authorities, with a standard deviation of 0.097.

\subsubsection*{Deflation} I deflate nominal earnings to 2015 prices using the ONS CPIH all-items index.

The balanced panel includes 314 local authorities observed for at least 8 of 9 years, yielding 2,757 observations after dropping missing values. \Cref{tab:sumstats} reports summary statistics for the pre-treatment period (2015--2018).

\input{tables/tab1_sumstats}


\section{Empirical Strategy}
\label{sec:strategy}

\subsection{Identification}

The identifying assumption is that, absent the April 2019 step-up, wage growth would have followed parallel trends across local authorities regardless of their small-firm employment share. I test this using a placebo treatment date and pre-trend analysis.

The key equation is:
\begin{equation}
\log Y_{st} = \alpha_s + \gamma_t + \beta \left(\text{SmallShare}_s \times \text{Post}_t\right) + \varepsilon_{st}
\label{eq:did}
\end{equation}
where $Y_{st}$ is median annual gross pay (real) in local authority $s$ in year $t$, $\alpha_s$ and $\gamma_t$ are local authority and year fixed effects, $\text{SmallShare}_s$ is the pre-treatment employment-weighted share of workers at firms with fewer than 50 employees, $\text{Post}_t = \ind[t \geq 2019]$, and standard errors are clustered at the local authority level.

The coefficient $\beta$ captures the differential change in log wages after 2019 for a one-unit increase in small-firm share. Under the mandate-tax hypothesis, $\beta < 0$: areas where more workers face the binding minimum should see wage compression. Under no offset, $\beta = 0$.

\subsection{Threats to Validity}

\subsubsection*{Pre-trends} I estimate an event-study version of \Cref{eq:did} interacting $\text{SmallShare}_s$ with year indicators, omitting 2018. A formal placebo test applies a fake treatment date of 2017 using only 2015--2018 data: the coefficient is 0.040 ($p = 0.15$), consistent with parallel pre-trends.

\subsubsection*{Confounding shocks} The April 2019 step-up was immediately followed by the COVID-19 pandemic. I address this by estimating the model excluding 2020--2021, which strengthens the result. The concern is that small-firm-dominated areas experienced different COVID wage dynamics; I note this as a limitation but observe that the positive coefficient persists in both the full and restricted samples.

\subsubsection*{Treatment intensity as proxy} Small-firm share correlates with rurality, industry mix, and local economic conditions. The LA fixed effects absorb time-invariant differences; year fixed effects absorb common shocks. The remaining concern is differential trends correlated with firm size composition. The placebo test and the negative-control specification (using large-firm share, where the minimum was non-binding) mitigate but cannot fully eliminate this threat.


\section{Results}
\label{sec:results}

\subsection{Main Results}

\Cref{tab:main} reports the main estimates. Column (1) shows the continuous-treatment DiD: a one-unit increase in small-firm share is associated with 14.4 log-point faster wage growth after 2019 ($\hat{\beta} = 0.144$, $p < 0.001$). Evaluated at one standard deviation of treatment intensity (0.097), this implies wages grew approximately 1.4 percentage points faster in more-exposed areas---the opposite sign from the mandate-tax prediction.

Column (2) uses a binary treatment: local authorities with above-median small-firm share saw 2.1 log points faster wage growth ($p < 0.001$), roughly equivalent to 2.1 percentage points. Column (3) shows that the effect holds for hourly pay ($\hat{\beta} = 0.142$), ruling out a composition story driven by hours adjustments. Column (4) weights by employment, which strengthens the estimate to 0.159.

\input{tables/tab2_main}

\subsection{Event Study}

\Cref{tab:eventstudy} reports the year-by-year interaction coefficients. The pre-treatment coefficients (2015--2017) show some noise, and the joint F-test rejects their equality to zero ($p < 0.001$). I interpret this cautiously: while the placebo test at 2017 is insignificant ($p = 0.15$, \Cref{tab:robust} Column 5), the event-study pre-trends are not perfectly flat. The post-2019 coefficients show a sharp divergence beginning in 2022--2023, which could reflect lagged adjustment to the pension mandate, post-COVID wage recovery, or both. The main DiD coefficient should therefore be read as an upper bound on the pension effect.

\input{tables/tab3_eventstudy}

\subsection{Robustness}

\Cref{tab:robust} reports six robustness checks.

\subsubsection*{Alternative treatment measures} Using micro-firm share (0--9 employees) as treatment intensity yields a qualitatively identical result (Column 1). The large-firm share placebo (Column 2) is negative and smaller, consistent with the argument that large employers were already above the minimum.

\subsubsection*{Sample restrictions} Excluding London boroughs (Column 3) and excluding COVID years (Column 4) both preserve the positive coefficient, with the latter actually strengthening it ($\hat{\beta} = 0.229$).

\subsubsection*{Placebo test} Applying a fake treatment date of 2017 using only pre-step-up data produces an insignificant coefficient of 0.040 ($p = 0.15$, Column 5), supporting the parallel trends assumption over the pre-treatment window.

\subsubsection*{Tercile comparison} Comparing the top vs.\ bottom tercile of small-firm share (Column 6) confirms the result: the most-exposed third of local authorities saw 3.4 percentage points faster wage growth than the least exposed.

\input{tables/tab4_robust}


\section{Discussion}
\label{sec:discussion}

The central finding---no wage offset, and if anything a positive wage effect in more-exposed areas---challenges the textbook prediction that mandated employer contributions are fully passed through to workers as lower wages. Several mechanisms could explain this.

\subsubsection*{Default anchoring suppresses pass-through} The auto-enrollment regime uses defaults rather than mandates in the traditional sense: employers must enroll workers and contribute, but the contribution rate is a floor, not a ceiling. If employers treat the floor as a focal point \citep{thaler2004}, the cost is ``small enough to absorb'' psychologically, and the adjustment to wages occurs only gradually or not at all. This is consistent with \citet{chetty2014}, who found that Danish employers did not reduce take-home pay when required to contribute more, but rather adjusted the \textit{total} compensation envelope upward.

\subsubsection*{Tight labor markets} The post-2019 period was characterized by historically low unemployment in the UK, particularly in small-firm-dominated sectors like hospitality and retail. Employers may have been unable to pass through the pension cost to wages without losing workers. In this case, the positive coefficient reflects the interaction of a cost mandate with a tight labor market, and would not generalize to slack conditions.

\subsubsection*{Limitations} I cannot rule out that the positive coefficient partly reflects differential mean reversion or COVID recovery patterns correlated with firm size composition. The event-study pre-trends, while individually small, show some noise. Future work with individual-level T-MSIS-equivalent data (e.g., HMRC RTI microdata) could sharpen the identification by exploiting within-employer variation and tracking individual wage trajectories.

\subsubsection*{Policy implications} If the absence of wage offset generalizes, then raising the auto-enrollment minimum---as the UK Pensions Commission is considering---would genuinely increase total compensation for low-paid workers without the hidden tax that classical theory predicts. The design of the mandate matters: a behaviorally anchored floor, implemented through defaults and set at a politically modest level, may escape the pass-through channel that binds for large, salient mandates like health insurance. This distinction---between mandates that compress wages and mandates that don't---deserves formal investigation.


\section{Conclusion}

At the aggregate level, the UK's April 2019 pension step-up left no visible mark on relative wage growth in areas most exposed to the binding floor. This is inconsistent with full and immediate wage offset, though the aggregate design cannot rule out partial pass-through or delayed adjustment. The natural next step is individual-level analysis---using HMRC RTI microdata to track within-employer wage trajectories around each step-up date---which could separate the pension mandate from the confounding macroeconomic shocks that plague any aggregate study of this period.


\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @ai1scl

\noindent\textbf{First Contributor:} \url{https://github.com/ai1scl}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
