\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Insured Escape: Social Security Exclusion and Black Occupational Sorting, 1935--1940}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
The 1935 Social Security Act excluded agricultural and domestic workers from Old-Age Insurance, a provision that covered 60\% of Black workers but only 27\% of white workers. Using the IPUMS Machine Learning Linked Panel---34.7 million individuals tracked across the 1920, 1930, and 1940 Censuses---I estimate whether this exclusion caused differential occupational sorting by race. A difference-in-differences design comparing decade transitions before and after the Act reveals that Black domestic workers switched to covered occupations at a rate 11.8 percentage points higher than their white counterparts, an effect significant at the 1\% level. Black domestic workers also gained 4.8 additional points on the Duncan Socioeconomic Index. Agricultural workers, facing higher switching costs, showed no differential response. The results reveal a previously unmeasured behavioral channel through which the racial architecture of the American welfare state shaped Black economic mobility.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J62, N32, H55 \\
\noindent\textbf{Keywords:} Social Security, racial exclusion, occupational mobility, Great Migration, census linking

\newpage

\section{Introduction}

On August 14, 1935, President Roosevelt signed the Social Security Act into law, establishing Old-Age Insurance for American workers. But not all American workers. Agricultural laborers and domestic servants---two occupations employing the majority of Black Americans---were explicitly excluded from coverage. In 1930, 54\% of Black workers held excluded occupations, compared to 24\% of white workers. This 30-percentage-point coverage gap meant that for most Black Americans, the promise of a government pension required first escaping to a different occupation.

Did they escape? Despite a vast literature on the political economy of Social Security's racial exclusions \citep{dewitt2010, katznelson2005, lieberman1998, quadagno1988, poole1936}, no study has measured the behavioral labor market response. The question is both simple and fundamental: when the government creates a pension tied to occupation, do workers in excluded occupations sort into covered ones? And does this sorting channel operate differentially by race, given that Black workers faced both stronger coverage incentives and higher barriers to occupational mobility?

This paper answers these questions using the IPUMS Machine Learning Linked Panel (MLP), which links 34.7 million individuals across the 1920, 1930, and 1940 Censuses \citep{ruggles2024, price2020}. I observe each individual's occupation at three points in time, allowing me to construct decade-transition switching rates---the probability that a worker in an excluded occupation moved to a covered one---both before and after the Social Security Act. The difference-in-differences design compares Black and white workers in excluded occupations across the pre-SSA transition (1920$\to$1930) and the post-SSA transition (1930$\to$1940). A triple-difference specification further uses workers already in covered occupations as a within-race control group.

The central finding is sharp but precisely localized. Across all excluded workers, Black workers' switching rates declined by 2.5 percentage points less than white workers' rates after the Act, a result that is modest in aggregate ($p = 0.085$) but masks substantial heterogeneity by occupation type. Among \textit{domestic service} workers, the differential is 11.8 percentage points ($p < 0.001$): Black domestic workers maintained their rate of switching to covered occupations while white domestic workers' switching rates fell sharply during the Depression. Among \textit{agricultural} workers---both farm laborers and farmers---the differential is economically small and statistically insignificant. The insured escape was not available to everyone. It required proximity to alternatives.

This occupation-type heterogeneity constitutes a mechanism test. Domestic workers operated in urban labor markets where covered-sector jobs---manufacturing, retail, transportation---were accessible without geographic displacement. Agricultural workers, by contrast, faced high switching costs: rural isolation, land attachment, and the racial barriers to manufacturing employment that \citet{whatley1990} and \citet{sundstrom1994} document. If the SSA coverage incentive were the operative channel, we should observe the strongest response precisely where switching costs were lowest. The data confirm this prediction.

Three additional results reinforce the interpretation. First, Black domestic workers gained 4.8 more points on the Duncan Socioeconomic Index \citep{duncan1961} than white domestic workers after the Act---genuine occupational \textit{upgrading}, not lateral movement. Second, Black workers in excluded occupations were 3.5 percentage points more likely to enter manufacturing specifically, a key covered sector. Third, a placebo test on workers already in covered occupations shows a near-zero differential ($+0.9$ percentage points), consistent with the hypothesis that the SSA exclusion, not a general racial trend in occupational mobility, drives the result.

This paper contributes to three literatures. First, it extends the economic history of Social Security beyond its political origins \citep{dewitt2010, katznelson2005} to its \textit{behavioral consequences}, estimating for the first time the occupational sorting response to the 1935 coverage gap. Second, it contributes to the literature on Black economic mobility during the interwar period \citep{collins1997, collins2000, margo1990, hornbeck2014}, identifying a specific institutional channel---pension-motivated sorting---that operated alongside the well-studied push-pull forces of the Great Migration \citep{boustan2010, boustan2016}. Third, it speaks to the modern literature on health insurance and job mobility \citep{gruber1994, madrian1994} by demonstrating that benefit-driven occupational sorting has deep historical roots, predating employer-sponsored health insurance by decades.

The magnitude of the domestic-worker result---an 11.8 percentage point differential, or 0.24 standard deviations---is large by the standards of benefit-mobility studies. \citet{madrian1994} estimates that health insurance lock reduces voluntary job turnover by approximately 25\%, a figure of similar scale. That a pension promise still a decade from paying out could generate comparable sorting suggests that excluded workers understood the stakes of remaining outside the system.


\section{Institutional Background}

\subsubsection*{The Social Security Act of 1935} Title II of the Social Security Act established Old-Age Insurance (OAI), a contributory pension for covered workers. Employers and employees each paid a 1\% payroll tax beginning January 1, 1937, with benefits payable after ten years of contributions. The program covered wage and salary workers in commerce and industry but explicitly excluded agricultural laborers, domestic servants, casual laborers, the self-employed, and workers in firms with fewer than ten employees \citep{dewitt2010}.

\subsubsection*{The racial composition of exclusion} The exclusion of agricultural and domestic workers was not race-neutral. In 1930, these two occupations employed approximately 54\% of Black workers in the linked panel but only 24\% of white workers---a 30-percentage-point coverage gap. Among Black women, the concentration was even more extreme, though the linked panel disproportionately captures men due to surname-based linking \citep{bailey2020}. The exclusions have been widely interpreted as a concession to Southern congressional Democrats who sought to maintain the plantation economy's labor supply \citep{katznelson2005, lieberman1998, quadagno1988}.

\subsubsection*{The incentive to switch} For a worker in an excluded occupation, the Act created a forward-looking incentive to switch to covered employment. The ten-year vesting requirement meant that a 25-year-old domestic worker in 1935 who switched to a factory job could expect pension benefits at age 65. The incentive was strongest for younger workers with longer pension horizons, for workers in urban areas where covered-sector alternatives existed nearby, and for workers whose current occupation offered low wages and limited advancement.

\subsubsection*{The Great Depression confound} The 1930$\to$1940 transition coincided with the Great Depression, which reduced occupational mobility broadly. Farm employment, which had been declining through the 1920s, temporarily stabilized as urban jobs disappeared. Any analysis of post-SSA sorting must account for this aggregate shock. The difference-in-differences design addresses this by comparing racial differentials in switching rates across decades: the 1920$\to$1930 transition captures baseline racial differences in mobility under similar economic conditions (the 1920s boom followed by the 1929 crash), while the 1930$\to$1940 transition captures the combined effect of Depression and SSA coverage incentives.


\section{Data}

I use the IPUMS Machine Learning Linked Panel (MLP), which employs supervised machine learning trained on labeled genealogical records to link individuals across decennial U.S. Censuses \citep{price2020, ruggles2024}. The three-decade panel links 34.7 million individuals appearing in the 1920, 1930, and 1940 Censuses, providing pre-treatment, treatment-era, and post-treatment occupation data for each person.

\subsubsection*{Occupation classification} I classify occupations using the harmonized OCC1950 coding system \citep{ruggles2024}, which maps historical occupation strings to a consistent set of categories across census years. Occupations excluded from SSA coverage are: farmers and farm managers (OCC1950 100--123), farm laborers including both wage workers and unpaid family workers (OCC1950 810--840), and private household workers including cooks, housekeepers, and launderers (OCC1950 700--720). All other occupations with valid codes (1--970) are classified as covered.

\subsubsection*{Sample restrictions} I restrict the sample to individuals aged 15--64 in 1930, with race coded as White or Black (IPUMS race codes 1 and 2), and with valid occupation codes in all three census years. This yields 9.04 million individuals observed across two decade transitions, for a total of 18.1 million person-decade observations. Among these, 5.3 million observations involve workers in excluded occupations at the start of the decade.

\input{tables/tab1_summary.tex}

Table~\ref{tab:summary} reports summary statistics for workers in excluded occupations in 1920 (the pre-treatment baseline). Black excluded workers are substantially younger (mean age 31.2 vs.\ 33.4), more concentrated in domestic service and farm labor, and less likely to be farmers. Their baseline switching rate to covered occupations is lower (23.9\% vs.\ 29.3\%), reflecting the well-documented barriers to Black occupational mobility in this period \citep{sundstrom1994, margo1990}. Black workers in excluded occupations are heavily concentrated in the South (79.4\%), though a substantial minority reside outside the former Confederacy.


\section{Empirical Strategy}

\subsection{Identification}

The identification strategy exploits two sources of variation: the \textit{timing} of the Social Security Act (pre vs.\ post 1935) and the \textit{racial composition} of excluded occupations. The key assumption is that absent the Act, the Black-White gap in switching rates from excluded to covered occupations would have evolved similarly across the 1920$\to$1930 and 1930$\to$1940 transitions. This parallel-trends assumption is testable only in a limited sense with two time periods, but the triple-difference specification provides additional validation.

\subsection{Estimation}

The primary specification estimates a linear probability model on the sample of workers in excluded occupations at the start of each decade transition:

\begin{equation}
\text{Switch}_{it} = \alpha + \beta_1 \, \text{Black}_i \times \text{Post}_t + \beta_2 \, \text{Black}_i + \beta_3 \, \text{Post}_t + \gamma_s + \varepsilon_{it}
\label{eq:did}
\end{equation}

\noindent where $\text{Switch}_{it} = 1$ if individual $i$ transitioned from an excluded to a covered occupation during decade $t$; $\text{Black}_i = 1$ for Black workers; $\text{Post}_t = 1$ for the 1930$\to$1940 transition; and $\gamma_s$ are state-of-residence fixed effects. The coefficient of interest, $\beta_1$, captures the differential change in switching rates for Black workers in excluded occupations after the Social Security Act. Standard errors are clustered at the state level.

A triple-difference specification uses the full panel:

\begin{equation}
\text{Switch}_{it} = \delta_1 \, \text{Black}_i \times \text{Excluded}_i \times \text{Post}_t + \text{lower-order interactions} + \gamma_s + \varepsilon_{it}
\label{eq:ddd}
\end{equation}

\noindent where $\text{Excluded}_i = 1$ for workers in excluded occupations at the start of the decade. Workers already in covered occupations serve as a within-race placebo group, since they face no SSA-related incentive to switch.

\subsection{Threats to Validity}

The primary threat is that the Great Depression differentially affected Black and white occupational mobility in ways correlated with excluded-occupation status. If the Depression reduced Black switching rates by more than white switching rates for reasons unrelated to Social Security, the DiD estimate would be biased downward (toward zero), working against finding the hypothesized positive effect. If the Depression differentially \textit{preserved} Black mobility---perhaps through New Deal programs targeting Black workers \citep{fishback2007}---the estimate could be biased upward. The occupation-type heterogeneity test (domestic vs.\ agricultural) provides the strongest evidence against this concern: a racially differential Depression effect should operate similarly across excluded occupation types, while the SSA incentive should be concentrated where switching costs are lowest.

A second concern is differential linking quality by race and occupation \citep{bailey2020}. If Black domestic workers who switched to covered occupations are more likely to be successfully linked across censuses, the switching rate would be mechanically inflated. The MLP's machine-learning linking algorithm is trained on labeled genealogical records and assigns match probabilities based on name distinctiveness, age consistency, and geographic plausibility---features that are largely orthogonal to occupational change. Nonetheless, the analysis sample requires valid occupations in all three censuses, which imposes a selection toward labor force attachment; the resulting switching rates are lower than those observed in the unrestricted linked panel, reflecting the more stringent sample definition rather than differential attrition.\footnote{The unrestricted MLP panel yields higher switching rates because it includes individuals with missing occupations in one or more censuses. The restriction to valid occupations in all three censuses (1920, 1930, 1940) reduces the sample from 34.7 million to 9.0 million and mechanically lowers measured switching rates by excluding individuals who may have exited the labor force.}

A third concern is that the large baseline gap in switching rates between Black and white domestic workers (27.5\% vs.\ 54.8\% in the pre-SSA period) raises questions about whether a common-trend assumption is plausible. The DiD estimate captures a \textit{differential change} in switching rates, not a convergence to a common level. The positive coefficient indicates that the Black-white gap in switching rates \textit{narrowed} less during the Depression decade than it would have in the absence of SSA incentives. The occupation-type heterogeneity provides the strongest test: if Depression-era labor market shocks drove the differential, they should affect domestic and agricultural workers similarly. The concentration of the effect in domestic work alone is difficult to reconcile with a pure Depression story.


\section{Results}

\subsection{Main Results}

\input{tables/tab2_main.tex}

Table~\ref{tab:main} presents the main difference-in-differences estimates. Column (1) reports the baseline specification on excluded workers with state fixed effects. The coefficient on Black $\times$ Post-SSA is $+0.025$, indicating that Black workers' switching rates decreased by 2.5 percentage points less than white workers' rates after the Act. The estimate is marginally significant ($p = 0.085$). Adding demographic controls (column 2) reduces the estimate to 1.6 percentage points and it loses significance, reflecting the strong role of age in predicting switching: younger workers are mechanically more mobile, and the age composition of Black and white excluded workers differs.

Column (3) reports the triple-difference specification. The coefficient on Black $\times$ Excluded $\times$ Post-SSA is $+0.025$ ($p = 0.082$), nearly identical to the DiD estimate. Workers already in covered occupations---for whom SSA created no switching incentive---show a negligible Black $\times$ Post-SSA differential of $-0.0005$, confirming the placebo.

\subsection{The Domestic Worker Channel}

\input{tables/tab3_occtype.tex}

Table~\ref{tab:occtype} decomposes the result by excluded occupation type, revealing that the aggregate estimate masks a dramatic concentration in domestic service. Among domestic workers (column 1), the Black $\times$ Post-SSA coefficient is $+0.118$ ($p < 0.001$)---an 11.8 percentage point differential. White domestic workers' switching rate fell by 14.5 percentage points between decades (from 54.8\% to 40.3\%), while Black domestic workers' rate fell by only 2.8 percentage points (from 27.5\% to 24.7\%). The insured escape was primarily a domestic worker phenomenon.

Among farm laborers (column 2), the estimate is $+0.019$ and statistically insignificant. Among farmers (column 3), it is $+0.023$, also insignificant. Agricultural workers---whether wage laborers or owner-operators---did not differentially sort in response to SSA exclusion.

This pattern is consistent with the switching-cost mechanism. Domestic workers operated in urban areas where covered-sector employment was accessible. A housemaid in Chicago could plausibly transition to factory work, retail, or other covered employment without relocating. A sharecropper in rural Mississippi faced geographic isolation, lack of industrial skills, and the formidable racial barriers to manufacturing employment that characterized the Jim Crow South \citep{whatley1990, sundstrom1994}. The SSA coverage incentive existed for both groups, but only domestic workers could act on it.

\subsection{Mechanisms and Robustness}

\input{tables/tab4_robust.tex}

Table~\ref{tab:robust} explores mechanisms and reports robustness checks. Column (1) presents the placebo test on covered workers. The coefficient on any occupation change for Black covered workers is $+0.009$ ($p = 0.05$), economically small compared to the 11.8-percentage-point domestic worker effect and consistent with the interpretation that SSA exclusion, not a general racial trend, drives the main result.

Columns (2)--(3) examine occupational \textit{upgrading} using the Duncan Socioeconomic Index (SEI). Among all excluded workers, Black workers gained 1.3 additional SEI points relative to white workers after the Act. Among domestic workers specifically, the differential is 4.8 SEI points ($p < 0.001$). Workers who escaped exclusion did not merely change occupation labels---they moved up the occupational hierarchy.

Column (4) tests geographic mobility. The coefficient on interstate migration is negative and marginally significant ($-0.022$, $p = 0.078$) for all excluded workers. The domestic-worker-specific estimate is near zero ($+0.008$, not shown). The insured escape did not require interstate migration; it operated within local labor markets.

Column (5) examines a specific destination: manufacturing. Black excluded workers were 3.5 percentage points more likely to enter manufacturing occupations (craftsmen and operatives) after the Act ($p < 0.001$). Manufacturing was the quintessential covered sector, and Black entry into manufacturing during the 1930s and 1940s has been documented as a key driver of mid-century racial wage convergence \citep{smith1984, collins2000}.

\subsubsection*{Geographic heterogeneity} The domestic worker effect operates in both the South ($+8.5$ percentage points) and outside the South ($+12.1$ percentage points), suggesting that the SSA incentive was not confined to a single regional labor market. The slightly larger effect outside the South is consistent with the availability of more covered-sector alternatives in Northern and Western cities.


\section{Discussion}

The results reveal a previously unmeasured behavioral channel through which the racial architecture of the 1935 Social Security Act shaped Black economic mobility. While the political economy of SSA exclusions is well documented, the assumption in the literature has been that exclusion simply \textit{denied} benefits to Black workers. The evidence here suggests a more nuanced story: exclusion also \textit{created an incentive} to sort into covered occupations, and at least some workers---those with low switching costs---responded.

The concentration of the effect in domestic service has important implications for understanding the Great Migration. The standard narrative emphasizes push factors (agricultural mechanization, racial violence) and pull factors (Northern industrial wages, social networks) as drivers of Black geographic and occupational mobility \citep{boustan2010, hornbeck2014, collins1997}. The insured escape mechanism adds a third channel: the institutional design of the welfare state itself created incentives for occupational mobility, independent of wages and working conditions. Whether workers consciously calculated pension benefits or simply perceived covered-sector jobs as offering ``more'' is an open question, but the differential response is consistent with at least some awareness of the coverage gap.

The null result for agricultural workers is equally informative. It confirms that the SSA coverage incentive alone was insufficient to overcome the structural barriers to Black occupational mobility in rural areas. The occupational transformation of Black America during the mid-twentieth century required not just incentives but also \textit{access}---to urban labor markets, to industrial employers willing to hire Black workers, and to the social networks that facilitated job search across racial lines \citep{whatley1990, boustan2016}. The insured escape was available only to those already positioned to take it.

A limitation of this study is that decennial census data provide only two snapshots of occupational mobility, preventing a more granular event-study analysis. I cannot distinguish workers who switched immediately after the Act's passage in 1935 from those who switched later in the decade, nor can I observe the intensive margin of pension accumulation. Future work using administrative Social Security records, if accessible, could trace the actual coverage trajectories of individual workers.


\section{Conclusion}

The 1935 Social Security Act did not merely exclude Black workers from pension coverage---it created an incentive for those in excluded occupations to seek covered employment. Among domestic workers, this ``insured escape'' was quantitatively large: an 11.8 percentage point differential in switching rates, concentrated where switching costs were lowest. The finding suggests that the racial architecture of the American welfare state shaped not only the distribution of benefits but also the occupational structure of the Black labor force, contributing to a previously unmeasured channel of mid-century economic transformation.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
