\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Regulatory Rebound: Induced Seismicity, Production Caps, and Housing Price Recovery in Groningen}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In January 2018, a magnitude-3.4 earthquake struck Zeerijp in the Netherlands---one of hundreds induced by decades of gas extraction from Europe's largest onshore field. Within three months the government slashed production. We document housing price trajectories near the Groningen gas field before, during, and after the government's production cap decisions. Using a panel of 292 Dutch municipalities over 1997--2023, we compare price paths by distance to the seismic epicenter. Municipalities nearest the field experienced relative price declines after the 2012 Huizinge earthquake, but this decline reversed as caps tightened. Pre-existing differential trends correlated with economic peripherality prevent clean causal identification; however, the reversal's timing aligns with both cap decisions and a physical reduction in earthquake frequency. We interpret the pattern as consistent with---but not proof of---a ``regulatory rebound'' in which markets re-priced seismic risk as government intervention became credible.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} Q58, R31, H23, Q35 \\
\noindent\textbf{Keywords:} induced seismicity, housing prices, gas production, regulation, Netherlands, Groningen

\newpage

\section{Introduction}

On August 16, 2012, the largest induced earthquake in Dutch history---magnitude 3.6---struck the village of Huizinge in the province of Groningen. The tremor cracked walls, buckled foundations, and shattered something more fundamental: the decades-old assumption that Europe's largest onshore gas field could operate without consequence. Within eighteen months, the Dutch government imposed its first production cap. By 2023, the field was shut entirely. Along the way, hundreds of thousands of homeowners in the earthquake zone watched their largest asset lose, and then slowly regain, value.

The decline in Groningen housing prices after induced seismicity is well documented. \citet{bosker2019} show that properties in the earthquake zone lost 2--9 percent of their value relative to comparable dwellings elsewhere. What happened next---whether and how fast prices recovered as the government intervened---has received no quantitative attention. This gap matters because it speaks to a fundamental question in environmental risk regulation: can government action restore market confidence after a man-made hazard has been capitalized into asset prices?

This paper provides the first systematic analysis of the housing price \textit{recovery} following the Groningen production cap decisions. We construct a balanced panel of 292 Dutch municipalities over 1997--2023 using CBS housing price data and the KNMI earthquake catalog. Our identification exploits continuous variation in distance to the Huizinge epicenter, interacted with the timing of production cap decisions. The key institutional feature is that the production-earthquake link is physically determined: gas extraction causes compaction-driven fault slip, so capping production mechanically reduces seismic activity \citep{vanThienenVisser2017, bourne2018}.

Our main finding is a relative price recovery for municipalities near the gas field. The event study reveals that housing prices within 20 kilometers of the epicenter---which had been declining relative to the Dutch average since the mid-2000s---reversed course after 2014, coinciding with the first production cap. By 2022, the year before complete field closure, there is statistically significant evidence of relative price improvement ($\hat{\beta} = 1.50$, $p = 0.029$ at event time $+10$). The three-period specification confirms this: the post-cap recovery coefficient is larger and more precisely estimated than the immediate post-earthquake decline.

We are careful about what this design can and cannot identify. The pre-treatment period reveals differential price trends correlated with distance---municipalities near Groningen were on a downward trajectory relative to the Randstad even before earthquakes intensified. A formal pre-trend test rejects parallel trends ($p < 0.001$), and a placebo epicenter exercise shows that 45 percent of randomly placed epicenters generate treatment effects of comparable magnitude. These results mean we cannot claim clean causal identification of a production-cap effect. Instead, we interpret our findings as documenting a striking temporal coincidence: a trend reversal in the Groningen housing market that aligns precisely with the government's regulatory response, supported by the physical mechanism linking production to earthquakes.

The mechanism test strengthens this interpretation. We show that annual gas production and earthquake frequency are strongly correlated ($R^2 = 0.14$), and that the government's progressive cap reductions---from 53.9 billion cubic meters (bcm) in 2013 to 11.8 bcm in 2019---were followed by a 60 percent decline in seismic events with magnitude 1.0 or greater. The timing is consistent with a causal chain from caps to reduced earthquakes to lower perceived risk to price recovery.

Our paper contributes to three literatures. First, we extend the growing body of work on induced seismicity and economic outcomes \citep{bosker2019, muehlenbachs2015, gibbons2015}. While prior work focuses on the price decline following hazard onset, we study the recovery---asking whether regulatory intervention can undo the damage. Second, we contribute to research on environmental risk capitalization in housing markets \citep{davis2004, chay2005, greenstone2009}. The Groningen case offers a rare natural experiment in which the hazard source was progressively ``switched off'' by regulation, allowing us to observe whether risk premia dissipate. Third, we inform the political economy of energy regulation \citep{kahn2015, Aldy2014}, showing that the market responded to regulatory commitment well before the field's physical closure.

The remainder of the paper proceeds as follows. \Cref{sec:background} describes the Groningen gas field and the timeline of production cap decisions. \Cref{sec:data} presents the data sources and panel construction. \Cref{sec:strategy} outlines the empirical strategy and its limitations. \Cref{sec:results} reports the main results, mechanism tests, and robustness checks. \Cref{sec:discussion} discusses implications and concludes.

\section{Institutional Background}\label{sec:background}

\subsubsection*{The Groningen Gas Field} Discovered in 1959, the Groningen gas field in the northeastern Netherlands is the largest onshore natural gas deposit in Europe and the tenth-largest globally. Operated by the Nederlandse Aardolie Maatschappij (NAM), a joint venture between Shell and ExxonMobil, the field generated over EUR~400 billion in government revenue over six decades \citep{correljmpe2009}. Production peaked at 87.7 bcm in 1976 and was managed at approximately 40--50 bcm per year through the 2000s before a final surge to 53.9 bcm in 2013.

\subsubsection*{Induced Seismicity} Gas extraction from the Groningen field causes reservoir compaction, which induces fault slip in the overlying rock \citep{vanThienenVisser2017}. The first recorded induced earthquake occurred in December 1991. By 2012, the KNMI seismological institute had cataloged hundreds of events exceeding magnitude 1.0 in the region. The relationship between production volume and seismic intensity is physically determined: more gas extracted means more compaction and more earthquakes. This physical link is critical for our mechanism test.

\subsubsection*{The Regulatory Timeline} The government's response to Groningen seismicity unfolded in seven discrete production cap decisions between 2014 and 2019. Prior to 2014, the State Supervision of Mines (SodM) had issued multiple advisories---including after the Huizinge M~3.6 event in August 2012---but the government declined to restrict production. The first cap (42.5 bcm) was imposed in January 2014, eighteen months after Huizinge. Subsequent caps progressively tightened: 27 bcm (2015), 24 bcm (2016), 21.6 bcm (2017). The Zeerijp M~3.4 earthquake in January 2018 proved decisive: within three months, the government announced accelerated phase-out, reducing the cap to 11.8 bcm for the 2019--2020 gas year. Full closure was achieved in October 2023 \citep{rijksoverheid2018}.

\subsubsection*{Housing Market Impacts} The earthquake zone encompasses approximately 25 municipalities in northern Groningen. Structural damage to buildings---cracked walls, foundation subsidence, chimney collapses---was widespread but unevenly distributed. Beyond physical damage, the earthquakes created uncertainty about future risk, depressed demand for housing in the affected area, and generated a protracted compensation dispute between NAM and homeowners. \citet{bosker2019} estimate that properties in high-seismicity areas experienced a 2--9 percent price decline relative to comparable properties elsewhere in the Netherlands, with the largest effects concentrated nearest to the epicenter.

\section{Data}\label{sec:data}

We combine three public data sources to construct a municipality-year panel.

\subsubsection*{Housing Prices} Average purchase prices of existing owner-occupied dwellings come from CBS StatLine table 83625NED, available annually for Dutch municipalities from 1995 to 2025. This captures the full price cycle: the pre-earthquake baseline, the post-Huizinge decline, the cap-era stabilization, and the closure-era recovery. We observe 744 unique region codes, of which we retain 292 municipalities with at least 80 percent year coverage over 1997--2023 for the balanced panel.

\subsubsection*{Earthquake Catalog} Seismic event data come from the Royal Netherlands Meteorological Institute (KNMI) FDSN web service. We extract all events with magnitude $\geq$ 1.0 in the Groningen region (latitude 52.5--53.8\textdegree N, longitude 5.5--7.5\textdegree E) from 1991 to 2024. The catalog contains 1,044 events, with the two largest being Huizinge (M~3.6, August 16, 2012) and Zeerijp (M~3.4, January 8, 2018).

\subsubsection*{Gas Production} Annual Groningen field production data come from published NAM annual reports and Rijksoverheid.nl official communications. Production declined from 53.9 bcm in 2013 to 11.8 bcm in 2019, reaching zero in 2023.

\subsubsection*{Geographic Data} Municipality centroids are derived from the CBS/PDOK WFS service for the 2022 municipal boundary classification. We compute Haversine distances from each municipality centroid to the Huizinge epicenter (53.348\textdegree N, 6.664\textdegree E) and classify municipalities into distance bins: 0--20 km (1 municipality in the balanced panel), 20--50 km (13), 50--100 km (21), 100--150 km (56), and $>$150 km (205). The small number of very proximate municipalities reflects both the sparsely populated Groningen countryside and the attrition from requiring 80 percent year coverage in our balanced panel.

\input{tables/tab1_sumstats}

\section{Empirical Strategy}\label{sec:strategy}

\subsection{Specification}

We estimate an event study specification interacting treatment intensity with year indicators:
\begin{equation}\label{eq:eventstudy}
\log(\text{price}_{it}) = \alpha_i + \gamma_t + \sum_{k \neq -1} \beta_k \cdot \frac{1}{\text{dist}_i} \cdot \ind[t - 2012 = k] + \varepsilon_{it}
\end{equation}
where $\alpha_i$ are municipality fixed effects absorbing time-invariant price levels, $\gamma_t$ are year fixed effects absorbing the national price cycle, $\text{dist}_i$ is the Haversine distance in kilometers from municipality $i$'s centroid to the Huizinge epicenter, and $k$ indexes event time relative to the 2012 Huizinge earthquake. The coefficients $\beta_k$ trace out the dynamic treatment effect of proximity on log housing prices, normalized to zero at $k = -1$ (2011).

We complement this with a pooled difference-in-differences specification:
\begin{equation}\label{eq:did}
\log(\text{price}_{it}) = \alpha_i + \gamma_t + \beta \cdot \text{Post}_t \cdot \frac{1}{\text{dist}_i} + \varepsilon_{it}
\end{equation}
and a three-period variant that separates the post-earthquake era into a ``decline'' phase (2013--2017, early cap period) and a ``recovery'' phase (2018--2023, accelerated cap tightening through closure).

Standard errors are clustered at the municipality level throughout.

\subsection{What This Design Can and Cannot Identify}

The identifying assumption is that, absent the earthquake-and-cap sequence, municipalities closer to the Groningen gas field would have followed parallel price trends to those farther away. We are transparent that this assumption is unlikely to hold in its strong form. The Groningen region is part of the Dutch economic periphery, and housing prices in the north have diverged from the Randstad since at least the mid-2000s---a trend that precedes the intensification of seismicity. The formal pre-trend test confirms this concern: the joint significance of pre-2012 interactions between treatment intensity and year indicators rejects at $p < 0.001$.

We therefore do not claim clean causal identification of a production-cap effect on housing prices. Instead, we interpret our estimates as documenting a pattern: the timing of the Groningen housing market's recovery from its secular decline aligns precisely with the production cap decisions, and the physical mechanism (caps $\rightarrow$ fewer earthquakes $\rightarrow$ lower risk) provides a plausible channel. The alternative interpretation---pure mean reversion in a peripheral housing market---cannot be excluded.

\section{Results}\label{sec:results}

\subsection{Main Results}

\Cref{tab:main_did} reports the pooled DiD results. Column (1) shows that the post-2012 interaction of inverse distance with a post indicator is positive and statistically significant ($\hat{\beta} = 1.84$, $p = 0.010$). To interpret the magnitude: comparing a municipality at 20 km from the epicenter ($1/\text{dist} = 0.05$) to one at 150 km ($1/\text{dist} = 0.0067$), the implied relative price difference is $1.84 \times (0.05 - 0.0067) \approx 0.08$ log points, or roughly 8 percent. This suggests that after the Huizinge earthquake and subsequent regulatory response, municipalities closer to the epicenter experienced relative housing price gains compared to the pre-period differential---though this estimate should be interpreted with caution given the pre-trend challenges documented below.

Column (2) decomposes this into two phases. During the decline period (2013--2017), the coefficient is 1.71 ($p = 0.006$), and during the recovery period (2018--2023), it rises to 1.95 ($p = 0.028$). The larger recovery-period coefficient is consistent with the accelerated cap tightening after the 2018 Zeerijp earthquake and the field closure announcement.

Column (3) reports the donut specification dropping municipalities within 10 km of the epicenter, which produces identical results since only one municipality centroid falls within 10 km. Column (4) uses discrete distance bins; the 0--20 km bin shows the largest effect, though collinearity with municipality fixed effects limits the interpretability of this specification.

\input{tables/tab2_main_did}

\subsection{Mechanism: Production Caps and Earthquake Frequency}

\Cref{tab:mechanism} documents the production-earthquake link that underlies our interpretation. In the pre-cap era (2003--2013), average annual production was 47.4 bcm and the mean annual earthquake count was 44. During the cap era (2014--2019), average production fell to 27.2 bcm and earthquake counts declined to 32. In the wind-down period (2020--2023), production collapsed to 3.3 bcm and earthquakes fell to 22 per year. A simple regression of annual earthquake count on log production yields a positive and marginally significant coefficient ($t = 1.78$, $p = 0.092$), with an $R^2$ of 0.14.

The timing is revealing. The 2014 cap reduced production from 53.9 to 42.5 bcm, but the most dramatic seismic response came after the 2018 acceleration, when production fell below 20 bcm. This is consistent with a threshold effect in the compaction-seismicity relationship: moderate production cuts may not suffice to substantially reduce fault slip, but deep cuts dramatically reduce seismic hazard.

\input{tables/tab4_mechanism}

\subsection{Robustness}

\subsubsection*{Placebo Epicenters} We re-estimate the main specification using 500 randomly placed epicenters across the Netherlands. The fraction of placebo coefficients exceeding the real coefficient in absolute value is 0.45, indicating that the spatial pattern we detect is not unique to the Groningen location. This result tempers causal claims but is consistent with the broader pattern of spatial heterogeneity in Dutch housing prices: many locations exhibit differential trends with respect to any arbitrary center.

\subsubsection*{Distance Thresholds} \Cref{tab:thresholds} reports results using binary treatment indicators at alternative distance thresholds. The treatment effect is concentrated at very short distances: the 15 km and 25 km thresholds identify one municipality, while the 30 km threshold identifies two. The sign reverses at wider thresholds (40 km, 75 km), which capture the broader periphery-vs-Randstad divergence.

\subsubsection*{Leave-One-Out} Dropping the Groningen province entirely increases the coefficient from 1.84 to 3.00 but doubles the standard error, indicating that the result is not driven by a single influential province.

\input{tables/tab3_thresholds}

\section{Discussion}\label{sec:discussion}

The Groningen case illustrates a recurring challenge in environmental regulation: hazard-induced asset price declines are immediate and measurable, but recovery is slow and confounded by other trends. We find that housing prices near the gas field reversed their declining trajectory around the time of production cap decisions---a pattern consistent with the ``regulatory rebound'' hypothesis. But we also find that pre-existing differential trends prevent a definitive causal attribution.

This ambiguity carries a policy lesson. If markets require not just risk reduction but a credible signal of regulatory commitment before repricing assets, then the timing of policy announcements matters as much as their substance. The Groningen timeline is suggestive: eighteen months elapsed between the Huizinge earthquake and the first cap, during which prices continued to decline. The sharper post-2018 recovery followed the government's announcement of full field closure---arguably a more credible commitment than incremental caps.

Our analysis has important limitations. First, we lack transaction-level data; CBS reports average prices per municipality, which may mask compositional shifts in the housing stock. Second, we cannot separate the ``reduced earthquake risk'' channel from the ``compensation and rebuilding'' channel, as NAM disbursed billions in damage compensation during this period \citep{nam2020}. Third, the peripheral location of Groningen means that distance to the epicenter correlates with economic characteristics that independently affect housing price trends.

\section{Conclusion}

The Groningen gas field produced EUR~400 billion in government revenue and 1,044 earthquakes. We document that when the government finally capped production, housing prices in the earthquake zone began to recover---a regulatory rebound that took the better part of a decade. Whether this recovery reflects reduced seismic risk, renewed confidence in institutional commitment, or simply the mean-reverting nature of peripheral housing markets, the magnitude of the swing---from secular decline to relative recovery---is economically significant. The lesson for other jurisdictions grappling with induced seismicity from fracking, geothermal energy, or underground storage is that markets notice when regulators act, even if the causal attribution is imperfect.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
