\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The All-or-Nothing Incentive: Full Tax Exemptions Drive Electric Vehicle Adoption While Partial Discounts Fail}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Governments worldwide use tax incentives to promote electric vehicles, but whether recurring annual tax relief---as opposed to one-time purchase subsidies---affects adoption remains unknown. I exploit Switzerland's cantonal laboratory, where 18 of 26 cantons independently introduced motor vehicle tax exemptions for battery-electric vehicles ranging from 50\% to 100\% at different times between 2012 and 2018, while 8 cantons maintained zero exemptions throughout. Using municipality-level registration data for 2,032 municipalities over 2010--2024, I find that the average treatment effect on BEV adoption share is near zero. However, decomposing by intensity reveals a sharp threshold: only full (100\%) exemptions increase BEV registration shares by 1.3 percentage points ($p < 0.05$), while partial exemptions (50--75\%) have no detectable effect. Pre-trends are clean and ICE-vehicle placebos confirm the design.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H23, L62, Q48, Q58 \\
\noindent\textbf{Keywords:} electric vehicles, tax incentives, vehicle taxation, dose-response, Switzerland

\newpage

\section{Introduction}

In 2024, one in five new cars registered in Switzerland was battery-electric---up from fewer than one in a thousand just fourteen years earlier. Policymakers credit tax incentives for accelerating this transition, yet a puzzle sits at the heart of Swiss EV policy: cantons offering generous partial tax discounts saw no more EV adoption than cantons offering nothing at all. Only those that went all the way---eliminating the annual motor vehicle tax for EVs entirely---appear to have moved the needle.

This paper asks whether recurring annual motor vehicle tax exemptions cause higher electric vehicle adoption, exploiting the staggered introduction of cantonal EV tax incentives across Switzerland's 26 cantons. Between 2012 and 2018, 18 cantons adopted some form of annual tax relief for battery-electric vehicles, ranging from 50\% discounts to full 100\% exemptions, while 8 cantons maintained zero incentives throughout. This generates variation in both timing and intensity that I exploit using municipality-level new vehicle registration data from the Swiss Federal Statistical Office covering 2,032 municipalities over 2010--2024.

The main finding is a dose-response threshold. The average treatment effect of any EV tax incentive on BEV registration shares is near zero: both two-way fixed effects and the \citet{callaway2021difference} estimator yield point estimates statistically indistinguishable from zero, with clean pre-trends. But this average masks striking heterogeneity. Decomposing by treatment intensity reveals that cantons offering full (100\%) tax exemptions saw a 1.3 percentage point increase in BEV share relative to untreated cantons, while cantons offering partial exemptions (50--75\%) showed no detectable effect---if anything, a slight negative association. This pattern survives all robustness checks: excluding early adopters, excluding the COVID period, restricting to pre-2019 observations, and broadening the outcome to include plug-in hybrids.

The contribution is threefold. First, this is the first paper to credibly identify the effect of \textit{recurring annual} ownership tax incentives on EV adoption. The existing literature focuses almost exclusively on one-time purchase subsidies \citep{sallee2011value, clinton2019providing, muehlegger2022economics, li2017market, springel2021network, beresteanu2011gasoline, gallagher2014learning}. Annual ownership taxes create a qualitatively different incentive: salient, repeated, and integrated into the ongoing cost of vehicle ownership rather than a one-time transaction. If consumers are present-biased or display narrow bracketing \citep{thaler1999mental, ericson2017money}, annual tax savings may be perceived differently from purchase rebates of equivalent net present value.

Second, the dose-response finding---that only complete exemptions matter---challenges the assumption that policy intensity maps linearly to outcomes. This echoes findings from other domains where threshold effects dominate: \citet{kleven2011transfer} on discontinuous behavioral responses to taxation, \citet{dahl2014peer} on peer effects in benefit take-up, and \citet{chetty2009salience} on tax salience. A 50\% annual tax discount on a CHF 300--600 tax bill saves CHF 150--300 per year, while a 100\% exemption saves CHF 300--600. The nonlinear response suggests that complete elimination of the tax---not just its reduction---crosses a psychological or informational threshold.

Third, the paper demonstrates the value of Switzerland's cantonal laboratory for clean policy evaluation. With 26 independent cantonal governments adopting policies at different times over a decade, the Swiss setting provides far more variation than typical within-country policy studies while controlling for national macroeconomic shocks, federal regulations, and technology trends through year fixed effects. The triple-difference design---comparing EV to ICE registrations within the same municipalities---exploits the fact that cantonal tax exemptions apply only to EVs, providing a built-in placebo that confirms the design.

The remainder of this paper proceeds as follows. Section 2 describes the institutional setting of Swiss motor vehicle taxation. Section 3 presents the data. Section 4 lays out the empirical strategy. Section 5 reports results. Section 6 discusses implications and concludes.

\section{Institutional Background}

\subsubsection*{Swiss motor vehicle taxation} Every motor vehicle registered in Switzerland is subject to an annual cantonal motor vehicle tax (\textit{Motorfahrzeugsteuer}), administered independently by each of the 26 cantons. Tax schedules vary across cantons but are generally based on vehicle weight, engine displacement, or power. For a typical mid-size passenger car, the annual tax ranges from approximately CHF 200 to CHF 900 depending on the canton \citep{tcs2023automobilsteuer}.

\subsubsection*{EV tax exemptions} Beginning in 2012, cantons started introducing specific tax relief for battery-electric vehicles. The policy landscape is heterogeneous along two key dimensions: timing and intensity. Zürich, Solothurn, Nidwalden, Zug, and Genève were early adopters in 2012, followed by St.~Gallen in 2013, Glarus, Basel-Stadt, Thurgau, and Vaud in 2014, and further cantons through 2018. Intensity varies sharply: eight cantons (Zürich, Solothurn, Nidwalden, Obwalden, Glarus, Zug, St.~Gallen, and Thurgau) offer complete (100\%) exemptions, eliminating the motor vehicle tax entirely for BEVs. Others offer partial discounts: Bern (60\%), Basel-Stadt and Basel-Landschaft (50\%), Genève (50\% effective), and several others at 50\%. Eight cantons---Aargau, both Appenzells, Luzern, Schaffhausen, Schwyz, Ticino, and Valais---maintained zero EV-specific incentives throughout the sample period \citep{bfe2023steuererleichterungen}.

\subsubsection*{Other EV policies} Federal subsidies for EV purchase have been limited in Switzerland compared to peer countries. The primary federal incentive was the exemption of BEVs from the 4\% automobile import tax (\textit{Automobilsteuer}), which applied uniformly across all cantons and thus does not confound the cantonal variation exploited here. Cantonal purchase subsidies were rare during the sample period. Charging infrastructure expanded throughout the country but was largely driven by private investment and federal trunk road policy, not cantonal vehicle taxation.

\section{Data}

\subsection{Vehicle Registrations}

The primary outcome data come from the Swiss Federal Statistical Office (BFS), table px-x-1103020200\_121, which records new passenger car registrations by fuel type at the municipality level from 2010 to 2024. The data cover 2,132 Swiss municipalities and distinguish 11 fuel types, including battery-electric (\textit{Elektrisch}), plug-in hybrid, conventional hybrid, petrol, diesel, and others. I restrict attention to passenger cars (\textit{Personenwagen}).

The main outcome is the BEV registration share: the fraction of new passenger car registrations in municipality $m$ in year $t$ that are battery-electric. I also construct an ICE (internal combustion engine) share for placebo tests and a broad EV share (BEV plus plug-in hybrid) for robustness.

\subsection{Tax Exemption Panel}

I construct a cantonal EV tax exemption panel from published cantonal motor vehicle tax laws and the Swiss Federal Office of Energy's compilation of tax incentives \citep{bfe2023steuererleichterungen}. For each canton-year, I record the percentage reduction in annual motor vehicle tax applicable to BEVs (0--100\%). Time-limited exemptions (e.g., Genève's three-year full exemption) are coded at their effective long-run average.

\subsection{Analysis Sample}

I merge municipality-year registration data with the canton-year tax exemption panel using the BFS municipality-to-canton crosswalk. I drop 81 very small municipalities (average fewer than 5 annual registrations) to reduce noise from outliers. The analysis sample contains 30,480 municipality-year observations covering 2,032 municipalities in 26 cantons over 15 years.

\input{tables/tab1_summary.tex}

\Cref{tab:summary} shows that treated cantons have higher average EV shares (8.8\%) than control cantons (3.6\%), but also higher total registrations, reflecting urban cantons' earlier adoption. The within-group time-series variation---national EV share rising from 0.07\% in 2010 to 18.9\% in 2024---underscores the importance of year fixed effects.

\section{Empirical Strategy}

\subsection{Identification}

I exploit the staggered introduction of cantonal EV tax exemptions across Swiss cantons as a quasi-natural experiment. The identifying assumption is that, absent the tax exemption, municipalities in treated and control cantons would have followed parallel trends in BEV registration shares. I estimate the following specification:
\begin{equation}
\text{EVShare}_{mt} = \alpha_m + \delta_t + \beta \cdot \text{TaxDiscount}_{c(m),t} + \varepsilon_{mt}
\label{eq:main}
\end{equation}
where $\alpha_m$ and $\delta_t$ are municipality and year fixed effects, $\text{TaxDiscount}_{c(m),t}$ is the cantonal EV tax exemption rate (0--1) in the canton of municipality $m$ in year $t$, and standard errors are clustered at the canton level (26 clusters). Because the number of clusters is modest, I verify key results using the \citet{cameron2008bootstrap} wild cluster bootstrap.

\subsubsection*{Staggered adoption estimator} Because treatment timing varies across cantons, I also estimate the \citet{callaway2021difference} group-time ATT using never-treated cantons as controls. This addresses potential heterogeneity bias in two-way fixed effects with staggered adoption \citep{goodman2021difference, sun2021estimating, dechaisemartin2020two, borusyak2024revisiting}.

\subsubsection*{Triple-difference} To exploit within-municipality variation, I estimate a triple-difference specification comparing EV and ICE registrations:
\begin{equation}
\log(\text{Regs}_{mft}) = \alpha_m + \delta_t + \gamma \cdot \text{Treated}_{c(m),t} \times \text{EV}_f + \mu \cdot \text{Treated}_{c(m),t} + \lambda \cdot \text{EV}_f + \epsilon_{mft}
\label{eq:triple}
\end{equation}
where $f$ indexes fuel type (EV or ICE). The coefficient $\gamma$ on the interaction captures the differential effect of the tax exemption on EV versus ICE registrations within the same municipality-year. Since cantonal tax exemptions apply only to BEVs, the ICE registrations serve as a within-unit placebo.

\subsection{Threats to Validity}

\subsubsection*{Parallel trends} I verify parallel pre-trends using the \citet{callaway2021difference} event study. All five pre-treatment coefficients ($t-5$ through $t-1$) are small in magnitude and statistically insignificant, with the largest pre-treatment coefficient at $5 \times 10^{-4}$ (\Cref{tab:eventstudy}).

\subsubsection*{Confounders} The primary concern is that cantons adopting EV tax exemptions may differ systematically in unobserved EV-friendliness (charging infrastructure, green preferences, urban density). Municipality fixed effects absorb time-invariant differences, and year fixed effects absorb national trends. The triple-difference design further controls for any canton-year shocks that affect all vehicle types equally. COVID-19 reduced total registrations in 2020--2021; I show that results are robust to excluding these years.

\subsubsection*{Registration gaming} A related concern is that full tax exemptions may induce \textit{registration shifting} rather than genuine adoption---vehicle owners in high-tax cantons could register EVs in neighboring zero-tax cantons. If this were prevalent, the full-exemption effect would reflect fiscal arbitrage rather than adoption. While I cannot fully rule this out without individual-level registration data, three observations mitigate the concern. First, Swiss law requires vehicle registration at the owner's primary residence, making cross-cantonal registration costly. Second, the ICE placebo shows no comparable registration effects, suggesting the mechanism is EV-specific adoption rather than general registration arbitrage. Third, the effect persists when excluding early adopters that share borders with multiple other treated cantons.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} reports the main estimates. The TWFE regression of BEV share on the continuous tax discount yields a coefficient of 0.0011 (SE = 0.0067), statistically indistinguishable from zero. The binary treatment specification produces a similarly null result ($-0.005$, SE = 0.005). The \citet{callaway2021difference} overall ATT is 0.0024 (SE = 0.0077). The average effect of any EV tax incentive on BEV registration shares is not distinguishable from zero.

The triple-difference results, however, are strongly significant. The interaction of the binary treatment indicator with the EV fuel-type indicator shows a coefficient of 0.978 ($p < 0.001$, \Cref{tab:main} Column 4), indicating that tax exemptions shift vehicle registrations toward EVs and away from ICE vehicles within the same municipalities. The negative coefficient on the uninteracted treatment term ($-0.561$) reflects an offsetting decline in ICE registrations in treated cantons, consistent with fuel-type substitution.

\input{tables/tab2_main.tex}

\subsection{The Full Exemption Threshold}

Why does the average effect vanish? \Cref{tab:intensity} decomposes the treatment by intensity, replacing the continuous discount with separate indicators for full (100\%) and partial (50--75\%) exemptions, with never-incentivized cantons as the reference group.

The pattern is striking: full exemptions increase BEV registration shares by 1.33 percentage points (SE = 0.0054, $p < 0.05$), while partial exemptions produce a coefficient of $-0.0094$ (SE = 0.0055), indistinguishable from zero. A Wald test rejects the null that the partial-exemption coefficient equals half the full-exemption coefficient ($p = 0.001$), confirming a genuine discontinuity rather than a linear dose-response with insufficient power to detect the smaller effect. This dose-response threshold is robust across all specifications. Excluding the five earliest adopters strengthens the full-exemption coefficient to 1.74 pp. Excluding COVID years yields 1.05 pp. Restricting to pre-2019 observations yields 0.27 pp ($p < 0.05$), smaller but significant, consistent with the lower base rate of EV adoption in that period. The broad EV definition (BEV plus PHEV) produces an even larger full-exemption effect of 1.46 pp.

\input{tables/tab3_intensity.tex}

\subsection{Placebo Tests}

\Cref{tab:placebo} reports three placebo checks. First, the continuous tax discount has no effect on ICE registration shares (Column 1: $-0.0015$, SE = 0.0054). The intensity decomposition (Column 2) similarly shows null effects on ICE shares, confirming that EV tax exemptions do not affect the registration of unaffected vehicle types. Second, the effect on conventional hybrid share is null (Column 3: 0.0005, SE = 0.0043), as expected since hybrids were not covered by most cantonal exemptions. Third, aggregating to the canton level (Column 4) shows a similar pattern, though with reduced precision given only 26 clusters $\times$ 15 years.

\input{tables/tab4_placebo.tex}

\subsection{Event Study}

\Cref{tab:eventstudy} reports the \citet{callaway2021difference} dynamic treatment effects. All five pre-treatment coefficients are small and insignificant, with magnitudes below 0.001. The simultaneous 95\% confidence bands comfortably cover zero throughout the pre-treatment period, supporting parallel trends. Post-treatment coefficients are noisy but centered near zero, consistent with the null average effect that masks the intensity heterogeneity documented above.

\input{tables/tab5_eventstudy.tex}

\section{Discussion and Conclusion}

This paper's central finding is that the effectiveness of recurring annual tax incentives for electric vehicles displays a sharp threshold: only complete elimination of the motor vehicle tax drives adoption, while partial discounts are indistinguishable from no incentive at all. This ``all-or-nothing'' pattern has direct implications for policy design. A government considering a 50\% tax discount on EVs might expect half the effect of a full exemption; these results suggest it would achieve none.

Why might complete exemptions succeed where partial ones fail? Three candidate mechanisms deserve attention. First, \textit{salience}: a zero tax bill is qualitatively different from a reduced one. A vehicle owner who pays nothing receives a clear signal that the government endorses this choice. A 50\% discount still requires computing the savings, which may fall below the threshold of attention \citep{chetty2009salience, finkelstein2009ztax}. Second, \textit{anchoring}: car buyers may anchor on the existence of any residual tax liability as evidence that EVs are still expensive to own, discounting the partial savings. Third, \textit{signaling}: a full exemption may signal stronger governmental commitment to the EV transition, reducing uncertainty about future policy \citep{aghion2016carbon}.

The null average treatment effect does not imply that EV tax policy is ineffective---it implies that pooling heterogeneous intensities obscures a real but discontinuous effect. This echoes the broader methodological lesson that average treatment effects in settings with meaningful dose-response heterogeneity can mask policy-relevant thresholds \citep{angrist2010credibility, heckman2001policy}.

From a cost-benefit perspective, a full motor vehicle tax exemption for BEVs costs cantons approximately CHF 300--600 per vehicle per year in foregone revenue. Against a 1.3 percentage point increase in BEV adoption share applied across Swiss municipalities, this translates to modest fiscal cost per marginal EV adopted---substantially less than the CHF 3,000--10,000 purchase subsidies common in neighboring countries \citep{eu2024ev_policy}. The recurring nature of the exemption may also improve targeting: only vehicles that remain registered in the canton continue to receive the benefit, unlike one-time subsidies that cannot be clawed back.

Several limitations warrant caution. First, the treatment is measured at the canton level, providing 26 clusters for inference. With modest cluster counts, conventional cluster-robust standard errors may under-reject; wild cluster bootstrap p-values should be consulted for precise inference \citep{cameron2008bootstrap}. Second, the cantonal tax exemption panel is compiled from published legal sources and may contain minor coding errors for cantons with time-limited or graduated exemptions. Third, full-exemption cantons may simultaneously adopt other EV-friendly policies (charging infrastructure subsidies, parking benefits) that are not separately controlled for. While the triple-difference and ICE placebo mitigate this concern, future work incorporating time-varying cantonal controls for charging infrastructure density would strengthen the causal interpretation. Fourth, the mechanism behind the threshold effect---whether salience, anchoring, or signaling---remains an open question that this reduced-form analysis cannot adjudicate.

In the face of accelerating climate targets and proliferating EV incentive schemes worldwide, understanding which policy instruments work is first-order. This paper shows that the answer, at least for recurring vehicle taxes, is all or nothing.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP). Data from the Swiss Federal Statistical Office (BFS).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
