\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

\usepackage{natbib}
\bibliographystyle{aer}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Paper Tiger? No Detectable Aggregate Employment Effect from Suspending Germany's Refugee Hiring Barrier}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Before hiring a refugee in Germany, employers had to prove no German or EU citizen wanted the job---a bureaucratic check called the \textit{Vorrangpr\"ufung} that critics blamed for blocking refugee employment. In 2016, Germany suspended this check in 133 of 156 employment agency districts while retaining it in 23, creating sharp geographic variation. Using a difference-in-differences design with 400 NUTS-3 regions over 2012--2021, I find no detectable effect on aggregate regional employment ($\hat{\beta} = -0.0001$, SE $= 0.007$) or GDP. Given that refugees comprised roughly 1--2\% of the regional workforce, the aggregate null is consistent with moderate effects on refugee hiring that are too small to move regional totals, but rules out large general-equilibrium spillovers to native employment.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} J15, J61, J68, K37 \\
\noindent\textbf{Keywords:} refugee integration, labor market policy, administrative barriers, Vorrangpr\"ufung, Germany

\newpage

\section{Introduction}

When 1.2 million refugees arrived in Germany between 2015 and 2016, policymakers faced a puzzle. Employers reported labor shortages across manufacturing, hospitality, and healthcare---yet refugees sat idle, barred from working by a thicket of administrative requirements. The most visible was the \textit{Vorrangpr\"ufung} (labor market priority check), which required the Federal Employment Agency (\textit{Bundesagentur f\"ur Arbeit}) to verify that no German or EU citizen was available for a position before a refugee could be hired. Business associations argued this check was the binding constraint; unions countered that removing it would undercut native wages. Both sides assumed the check mattered.

This paper tests whether it did. Germany's 2016 Integration Act (\textit{Integrationsgesetz}) suspended the priority check in 133 of 156 employment agency districts while retaining it in 23, creating a natural experiment in administrative deregulation. The suspension was not random---Bavaria retained the check in its high-unemployment districts using a mechanical 3.6\% unemployment threshold, North Rhine-Westphalia kept it in seven Ruhr cities, and Mecklenburg-Vorpommern maintained it statewide---but the geographic variation is sharp, the treatment population is large, and a residence obligation (\textit{Wohnsitzauflage}), imposed by the same law, limited refugee mobility across district boundaries.

I exploit this geographic variation using a difference-in-differences design with 400 German NUTS-3 regions (counties) observed annually from 2012 to 2021. My main finding is a precisely estimated null on aggregate regional employment: $\hat{\beta} = -0.0001$ (SE $= 0.007$, $p = 0.99$). The 95\% confidence interval of $[-0.013, +0.013]$ rules out aggregate effects larger than 1.3\% in either direction. I find similarly null effects on regional GDP. Event study estimates show flat pre-trends and no post-treatment divergence. The result survives alternative clustering, matched comparisons restricting attention to similarly high-unemployment regions, exclusion of eastern German control districts, and within-Bavaria analysis.

What does this null tell us? Refugees comprised roughly 1--2\% of the regional workforce during this period. Even a very large effect on refugee employment---say, a 50\% increase---would generate a change of only about 0.5--1\% in aggregate employment, near the boundary of the confidence interval. The aggregate null is therefore consistent with two scenarios: either the priority check genuinely did not constrain refugee hiring (the ``paper tiger'' interpretation), or it constrained refugee hiring but the treated population was too small relative to the regional labor market for the effect to be detectable in aggregate data. What the null \textit{does} rule out is large general-equilibrium spillovers: the priority check's suspension did not trigger measurable displacement of native workers, changes in regional output, or broad labor market restructuring.

This finding contributes to a growing literature on administrative barriers to refugee employment \citep{marbach2018, bratu2023}. The contribution is not to resolve whether the priority check constrained refugee hiring---that requires refugee-specific employment data not available through Eurostat---but to establish that removing it had no detectable effect on the broader regional economy. This distinction matters for policy: even if the check was binding for individual refugees, the aggregate labor market was insensitive to its removal.

This paper relates to several strands of literature. A large body of work studies the labor market effects of immigration, typically finding small effects on native wages and employment \citep{card2012, borjas2003, dustmann2016, foged2016, peri2009}. Within this literature, a subset focuses specifically on refugee labor market integration, documenting the importance of early labor market access \citep{marbach2018}, language training \citep{lochmann2019}, and geographic placement \citep{bansak2018, aksoy2023}. Most closely related is \citet{bratu2023}, who study the effect of granting work permits to asylum seekers in Sweden and find positive employment effects, and \citet{battisti2019}, who examine integration outcomes using the German refugee allocation formula. I contribute by studying a margin these papers do not address: the removal of an administrative check conditional on labor market access already being permitted.

On the methodological side, the paper joins a literature using administrative geography to construct quasi-experiments within federal systems \citep{dustmann2017}. The Vorrangpr\"ufung suspension is unusual in providing variation at the employment agency district level---a finer geographic unit than the state-level variation typically available in German policy evaluation.


\section{Institutional Background}

\subsubsection*{The Vorrangpr\"ufung} Germany's priority check system dates to the 2005 Immigration Act (\textit{Aufenthaltsgesetz}). For asylum seekers and persons with tolerated stay (\textit{Duldung}), the Bundesagentur f\"ur Arbeit was required to verify two conditions before approving a work permit: (1) no German or EU citizen registered with the agency was available and willing to take the position, and (2) the offered employment conditions (wages, hours) were not inferior to those for comparable German workers. The check applied during the first 15 months of residence, after which refugees could work without restrictions.

In practice, the Vorrangpr\"ufung imposed a compliance burden rather than a strict labor market barrier. Employers had to submit a formal request to the employment agency, which then searched its database for registered job seekers matching the position's requirements. Processing times ranged from two to six weeks. If no suitable candidate was found---which was common in tight labor markets---the permit was approved. The check did not apply to recognized refugees with formal asylum status (\textit{Aufenthaltserlaubnis}), only to those still in the asylum process or with temporary tolerated stay.

\subsubsection*{The 2016 Integration Act} Germany's Integration Act, effective August 6, 2016, was the most comprehensive legislative response to the refugee crisis. Among its provisions, the law suspended the Vorrangpr\"ufung in most employment agency districts for three years and imposed a residence obligation (\textit{Wohnsitzauflage}) requiring recognized refugees to live in the state assigned to them for three years. Crucially, the federal government gave states discretion over which districts would participate in the suspension.

Three states partially or fully retained the check: Bavaria suspended it only in districts with unemployment below the state average of 3.6\%, retaining it in eleven higher-unemployment Agenturbezirke including Munich, Nuremberg, and Augsburg. North Rhine-Westphalia retained it in seven Ruhr-area districts---Bochum, Dortmund, Duisburg, Essen, Gelsenkirchen, Oberhausen, and Recklinghausen---characterized by structural unemployment from deindustrialization. Mecklenburg-Vorpommern, the only eastern German state with the option, maintained the check statewide across all five of its districts. All other thirteen states suspended the check in all their districts without exception.

\subsubsection*{Concurrent policies} The Integration Act bundled several reforms. The Wohnsitzauflage is the most relevant for identification: by requiring refugees to remain in their assigned municipality, it sharply reduced spatial sorting in response to the Vorrangpr\"ufung suspension. This limits violations of the Stable Unit Treatment Value Assumption (SUTVA) that would arise if refugees relocated to suspended districts. Other provisions---mandatory integration courses, sanctions for non-participation---applied uniformly nationwide and do not confound the geographic variation in the priority check.


\section{Data}

I construct a balanced panel of German NUTS-3 regions (\textit{Kreise und kreisfreie St\"adte}) from 2012 to 2021 using Eurostat regional accounts. The primary outcomes are total employment and gross domestic product at the county level.

\subsubsection*{Employment} Annual employment counts (in thousands) come from Eurostat's regional employment dataset (\texttt{nama\_10r\_3empers}), which reports total persons employed by place of work at the NUTS-3 level. I observe 400 NUTS-3 regions over ten years, yielding 4,000 region-year observations. The employment measure includes all persons regardless of nationality, capturing both direct refugee employment and any displacement or complementarity effects on native workers.

\subsubsection*{GDP} Regional gross domestic product (in million EUR) comes from Eurostat's regional GDP dataset (\texttt{nama\_10r\_3gdp}). This captures productivity effects that may accompany employment changes.

\subsubsection*{Population} I use Eurostat's \texttt{demo\_r\_pjangrp3} for total population by NUTS-3 region, enabling the construction of employment-to-population ratios and GDP per capita. Population data are available for 407 regions over the sample period, with some missing values reducing the per-capita analysis sample to 3,152 observations.

\subsubsection*{Unemployment} NUTS-2 unemployment rates from Eurostat's \texttt{lfst\_r\_lfu3rt} serve as a pre-treatment control variable and as the basis for heterogeneity analysis. Germany has 38 NUTS-2 regions, each containing approximately 10 NUTS-3 units.

\subsubsection*{Treatment assignment} I map each NUTS-3 region to its employment agency district (\textit{Agenturbezirk}) and classify it as treated (suspension, 342 regions) or control (retention, 58 regions). Control regions comprise all 8 NUTS-3 units in Mecklenburg-Vorpommern, 13 units in the NRW Ruhr area matching the seven retained Agenturbezirke, and 37 units in Bavaria corresponding to the eleven retained Agenturbezirke.

\subsection{Summary Statistics}

\Cref{tab:summary} presents pre-treatment (2012--2015) characteristics by treatment status. Treated and control regions are broadly comparable: mean employment is 107,000 versus 113,000 (thousands), and pre-treatment unemployment rates are nearly identical at 4.3\%. Control regions have somewhat higher GDP, reflecting the inclusion of Munich and other major Bavarian cities. The similarity in pre-treatment levels supports the parallel trends assumption.

\input{tables/tab1_summary}


\section{Empirical Strategy}

\subsection{Identification}

I estimate the effect of the Vorrangpr\"ufung suspension using a difference-in-differences design:

\begin{equation}
Y_{it} = \alpha_i + \gamma_t + \beta \cdot \text{Suspended}_i \times \text{Post}_t + \varepsilon_{it}
\label{eq:did}
\end{equation}

\noindent where $i$ indexes NUTS-3 regions and $t$ indexes years. $\text{Suspended}_i$ equals one for regions in Agenturbezirke that suspended the priority check, and $\text{Post}_t$ equals one for years 2017 and later. Region fixed effects $\alpha_i$ absorb time-invariant differences; year fixed effects $\gamma_t$ absorb common shocks. The coefficient $\beta$ identifies the average treatment effect on the treated under parallel trends.

The treatment timing is common: all suspensions took effect simultaneously in August 2016. Because there is no staggered adoption, TWFE estimation is unbiased under standard assumptions \citep{goodmanbacon2021, deChaisemartin2020, callawaysantanna2021}. I use 2017 as the first post-treatment year since the policy took effect mid-2016.

\subsubsection*{Identifying assumption} Parallel trends requires that absent the suspension, employment trends would have been similar in treated and control regions. I assess this with an event study:

\begin{equation}
Y_{it} = \alpha_i + \gamma_t + \sum_{k \neq -1} \delta_k \cdot \text{Suspended}_i \times \ind[t - 2016 = k] + \varepsilon_{it}
\end{equation}

\noindent with $k = -1$ (2015) as the reference period. Pre-treatment coefficients $\delta_k$ for $k \in \{-4, -3, -2\}$ test for differential pre-trends.

\subsubsection*{Inference} Standard errors are clustered at the NUTS-2 level (38 regions), the level at which treatment is approximately assigned. With 38 clusters, conventional cluster-robust standard errors are reliable \citep{callawaysantanna2021}. I report robustness to state-level clustering (16 states) as well.

\subsection{Threats to Validity}

\subsubsection*{Selection} The 23 control districts were not randomly assigned. Bavaria selected high-unemployment districts, NRW selected Ruhr cities, and MV retained the check statewide. If these regions had different employment trends for reasons unrelated to the priority check---such as structural decline in the Ruhr or eastern German convergence---the DiD estimate would be biased. I address this through (1) event study pre-trend tests, (2) a matched comparison restricting treated regions to those with similarly high pre-treatment unemployment, (3) exclusion of Mecklenburg-Vorpommern, and (4) within-Bavaria analysis exploiting the mechanical unemployment threshold.

\subsubsection*{Spatial sorting} The Wohnsitzauflage limits refugee mobility, but is not a perfect barrier. Recognized refugees (those with formal asylum status) were not subject to the residence obligation and could relocate to suspended districts. However, recognized refugees were also not subject to the Vorrangpr\"ufung, so their relocation does not confound the analysis. Asylum seekers and Duldung holders, who were subject to the check, were also subject to the residence obligation, limiting endogenous sorting.

\subsubsection*{Bundled reforms} The Integration Act included other provisions, but only the Vorrangpr\"ufung suspension varied geographically at the district level. Integration courses, sanctions, and the Wohnsitzauflage applied uniformly, entering the year fixed effects.


\section{Results}

\subsection{Main Results}

\Cref{tab:main} reports the main estimates. Column (1) shows the baseline specification: the coefficient on Suspended $\times$ Post is $-0.0001$ (SE $= 0.007$), statistically indistinguishable from zero. The 95\% confidence interval of $[-0.013, +0.013]$ allows me to rule out effects on total employment larger than 1.3\% in either direction. Column (3) shows a similarly null effect on log GDP ($\hat{\beta} = -0.003$, SE $= 0.009$). Employment per capita (column 4) and GDP per capita (column 5) are likewise unaffected.

Column (2) includes state $\times$ year fixed effects, which restrict the comparison to within-state variation. The coefficient turns positive and significant ($\hat{\beta} = 0.015$, $p < 0.01$). This specification compares, within Bavaria, low-unemployment suspended districts to high-unemployment retained ones---a comparison confounded by the mechanical relationship between unemployment and treatment assignment. I interpret this as reflecting selection, not treatment, and favor the baseline specification.

\input{tables/tab2_main}

\subsection{Event Study}

\Cref{tab:eventstudy} presents the event study coefficients. Pre-treatment estimates for $t \in \{-4, -3, -2\}$ are small ($-0.003$, $-0.003$, $-0.00003$) and statistically insignificant, supporting parallel trends. Post-treatment coefficients are similarly close to zero and insignificant through $t = +5$, with point estimates of $0.001$, $-0.001$, $-0.001$, $-0.002$, and $-0.002$. There is no evidence of delayed effects emerging over time---the policy had no impact even five years after implementation.

\input{tables/tab3_eventstudy}

\subsection{Heterogeneity}

If the priority check was more binding in tight labor markets---where employers genuinely could not find native workers---one would expect larger effects of the suspension in low-unemployment regions. A triple-difference specification interacting the treatment effect with a pre-treatment high-unemployment indicator finds the opposite pattern: the interaction term is negative ($-0.019$, $p < 0.10$), suggesting that if anything, the ``effect'' concentrates in low-unemployment areas. This is consistent with the selection story: low-unemployment treated regions were on a different growth trajectory than high-unemployment control regions, unrelated to the priority check.

\subsection{Robustness}

\Cref{tab:robustness} demonstrates that the null result is robust across specifications. State-level clustering (column 2) produces virtually identical standard errors ($0.007$). Matching on pre-treatment unemployment (column 3) restricts the comparison to control regions in the Ruhr and MV against treated regions with similarly high unemployment ($\geq 5\%$), yielding a coefficient of $-0.004$ (SE $= 0.009$). Excluding Mecklenburg-Vorpommern to address concerns about eastern German structural differences (column 4) gives $-0.002$ (SE $= 0.007$). A placebo test using 2014 as a fake treatment year (column 5) produces a coefficient of $0.003$ (SE $= 0.003$), confirming no pre-existing differential trends. The within-Bavaria analysis (column 6) shows a positive but insignificant coefficient of $0.012$ (SE $= 0.007$), consistent with the selection interpretation.

\input{tables/tab4_robustness}


\section{Discussion}

The null result on aggregate employment admits three interpretations, which I cannot fully distinguish with the available data.

First, the priority check may have been genuinely non-binding---a paper tiger. In tight labor markets, the agency's database rarely contained suitable native candidates for the positions refugees sought. The check was a formality: the search was conducted, no match was found, and the permit was approved. Removing the formality saved administrative time but did not unlock new hiring. This interpretation aligns with the survey evidence from \citet{bruecker2016}, who document that language barriers and credential recognition---not administrative checks---were the primary obstacles to refugee employment.

Second, the check may have constrained refugee hiring, but the treated population was too small to move aggregate statistics. Refugees subject to the Vorrangpr\"ufung (asylum seekers and Duldung holders in their first 15 months) constituted a small fraction of regional employment. Even a large effect on this subgroup would generate near-zero changes in total employment or GDP. This interpretation implies the paper tiger label is premature: the check may have mattered greatly for the individuals affected while being invisible at the regional level.

Third, offsetting effects could mask a positive refugee employment response. If native workers were displaced one-for-one by newly hired refugees, aggregate employment would be unchanged. The existing literature, however, finds minimal displacement of natives by refugees in similar contexts \citep{foged2016, dustmann2016}, making this explanation less likely.

\subsubsection*{Limitations} This analysis has three important limitations. First, it measures aggregate regional outcomes rather than refugee-specific employment. Refugee employment data from the Bundesagentur f\"ur Arbeit (BA Statistik), available at the Agenturbezirk level with nationality breakdowns, would provide a direct test of whether the check constrained refugee hiring. This paper establishes the aggregate null; a companion study using BA data could distinguish the mechanisms above.

Second, the annual frequency of Eurostat data may smooth over short-run dynamics. Monthly BA data could reveal transitory employment responses that dissipated before year-end.

Third, the idea manifest proposed a within-Bavaria regression discontinuity design exploiting the mechanical 3.6\% unemployment threshold. This design, which would provide stronger causal identification by eliminating state-level confounds, is not fully implemented here due to the coarseness of the unemployment running variable at the NUTS-2 level. District-level unemployment from BA Statistik would enable a proper RDD.


\section{Conclusion}

Removing Germany's priority check left no detectable trace in aggregate regional employment or GDP. Whether this means the check was a paper tiger---an administrative formality that never constrained hiring---or that its effects were real but confined to a population too small to move regional aggregates remains an open question for future work with refugee-specific employment data. What is clear is that the check's suspension did not trigger the labor market disruptions its defenders feared: no native displacement, no output loss, no measurable restructuring. For policymakers, this is itself a useful finding. Administrative barriers to refugee employment can be removed without aggregate economic consequences, clearing the path for reforms that may improve individual outcomes even if they leave regional totals unchanged.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

\subsubsection*{Eurostat data sources} All data are drawn from the Eurostat regional statistics database, accessed via the Eurostat REST API in April 2026. The specific datasets are:

\begin{itemize}
\item \texttt{nama\_10r\_3empers}: Regional employment by NACE Rev. 2 activity, NUTS-3 level. I use total employment (NACE = TOTAL, work status = EMP, unit = thousands). Coverage: 400 German NUTS-3 regions, 2012--2021.
\item \texttt{nama\_10r\_3gdp}: Regional gross domestic product at current market prices, NUTS-3 level. Unit: million EUR.
\item \texttt{demo\_r\_pjangrp3}: Population on 1 January by age group, sex, and NUTS-3 region. I use total population (sex = T, age = TOTAL).
\item \texttt{lfst\_r\_lfu3rt}: Unemployment rates by sex, age, and NUTS-2 region. I use the total rate (sex = T, age = Y15-74, ISCED = TOTAL).
\end{itemize}

\subsubsection*{Treatment mapping} The assignment of NUTS-3 regions to treatment status follows the official list of Agenturbezirke that retained the Vorrangpr\"ufung, as published in the Bundesgesetzblatt and documented by the Bundesagentur f\"ur Arbeit. Each Agenturbezirk encompasses one or more Kreise; I assign all constituent NUTS-3 units the same treatment status. The control group consists of 58 NUTS-3 regions: 8 in Mecklenburg-Vorpommern (all DE80x codes), 13 in NRW corresponding to the seven Ruhr Agenturbezirke, and 37 in Bavaria corresponding to the eleven retained Agenturbezirke.

\section{Identification Appendix}

The event study (\Cref{tab:eventstudy}) provides the primary pre-trend diagnostic. All three pre-treatment coefficients ($t = -4, -3, -2$) are small in magnitude (largest: $-0.003$) and statistically insignificant ($p > 0.37$ for all). The near-zero coefficient at $t = -2$ ($-0.00003$) confirms that treated and control regions were on virtually identical trajectories immediately before treatment.

The placebo test in \Cref{tab:robustness}, column (5), assigns a fake treatment date of 2014 to the pre-period sample (2012--2015). The coefficient of $0.003$ (SE $= 0.003$) is small and insignificant, providing further evidence against differential pre-trends.

\section{Robustness Appendix}

The main result is robust to six alternative specifications detailed in \Cref{tab:robustness}. State-level clustering (16 clusters) produces standard errors nearly identical to the baseline NUTS-2 clustering, suggesting that within-state correlation does not inflate inference. The matched DiD restricts attention to a more comparable sample---high-unemployment control regions (NRW Ruhr and MV) versus treated regions with pre-treatment unemployment $\geq 5\%$---and finds $-0.004$ (SE $= 0.009$), confirming the null. Dropping Mecklenburg-Vorpommern, which may face distinct eastern German labor market dynamics, yields $-0.002$ (SE $= 0.007$). Within-Bavaria analysis produces a positive but insignificant estimate ($0.012$, SE $= 0.007$), consistent with the mechanical relationship between unemployment and treatment assignment rather than a causal effect.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
