\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Graphics
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% tabularray (for modelsummary output)
\usepackage{tabularray}
\usepackage{codehigh}
\usepackage[normalem]{ulem}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Choice Tax: Pension Freedom and the Regressive Cost of Financial Autonomy}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
In April 2015, the UK eliminated the effective annuity mandate for defined-contribution pension holders, granting unrestricted access to retirement savings. Using the complete FCA Retirement Income Market Data covering 5.5 million pension pots from 2015 to 2024, I document a stark ``choice tax'': 88\% of pots under \pounds10,000 are fully encashed at first access, compared with 2\% of pots above \pounds250,000. This gradient is stable across the entire post-reform period—no learning narrows it. The mechanism is a no-advice trap: regulated financial advice costs \pounds1--3K, representing 10--30\% of small pots, so 68\% of small-pot holders access their pension without any professional guidance. I estimate aggregate welfare losses of \pounds14 billion from dominated encashment strategies, borne disproportionately by the smallest savers.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} D14, G51, H55, J26 \\
\noindent\textbf{Keywords:} pension reform, financial advice, choice architecture, retirement savings, UK

\newpage

\section{Introduction}

When the UK Chancellor announced Pension Freedoms in the 2014 Budget, the promise was simple: ``Let me be clear: no one will have to buy an annuity.'' Within six months of implementation, annuity sales had collapsed by 75\%. A decade later, the annuity market has not recovered. What has taken its place?

This paper provides the first comprehensive descriptive analysis of pension decumulation behavior following the UK Pension Freedoms reform, using the complete FCA Retirement Income Market Data covering 5.5 million pension pots accessed between 2015 and 2024. The central finding is that access method choice is sharply stratified by pot size, in a pattern consistent with a regressive ``choice tax'': small pension holders overwhelmingly choose full encashment---a strategy that is tax-inefficient relative to phased drawdown---while large pension holders smoothly transition to drawdown.

The facts are striking. Among pots under \pounds10,000, 88\% are fully encashed at first access---meaning the entire pension is withdrawn in one lump sum, triggering immediate income tax on 75\% of the pot. Among pots above \pounds250,000, only 2\% are fully encashed; 87\% enter flexible drawdown, which preserves tax efficiency and investment returns. This gradient persists across every half-year period in the data. The cross-sectional relationship between pot size and access method is remarkably stable: the encashment rate for small pots in 2024 is virtually identical to 2015, though compositional changes in who accesses pensions may contribute to this stability.

The evidence is consistent with a \textit{no-advice trap}. Regulated financial advice in the UK costs between \pounds1,000 and \pounds3,000---a fixed cost that represents 10--30\% of a sub-\pounds10,000 pot but less than 1\% of a pot above \pounds100,000. Among pension holders making full withdrawals, only 21\% of those with pots under \pounds10,000 use regulated financial advice, compared with 51\% of those with pots above \pounds250,000 (\Cref{tab:tab:advice}). The government's free Pension Wise guidance service partially closes this gap, but even with guidance included, 68\% of small-pot holders access their pension with no professional help whatsoever.

I calibrate the potential welfare cost of full encashment by computing investment returns forgone and excess tax paid relative to phased drawdown under baseline assumptions. The illustrative aggregate cost across 5.5 million pension pots is approximately \pounds14 billion---an upper bound that falls to \pounds7 billion if half of encashments are privately optimal. This cost is distributed regressively: while the per-pot loss is larger for bigger pots in absolute terms, the \textit{probability} of full encashment is overwhelmingly concentrated among the smallest savers.

These findings contribute to several literatures. First, I add to the growing body of work on pension decumulation and the annuity puzzle \citep{yaari1965uncertain, brown2001role, benartzi2011annuitization, beshears2014behavioral}. While much of this literature focuses on the decision to annuitize, I show that the more consequential margin in the UK is the decision to \textit{encash entirely} versus enter any form of phased withdrawal.

Second, the paper speaks to the literature on financial advice and its distributional consequences. \citet{bhattacharya2012costly} and \citet{mullainathan2012market} document that financial advice is both costly and of mixed quality. I show that the fixed-cost structure of advice creates a mechanism through which deregulation can be welfare-reducing for precisely those it was meant to help---a finding that resonates with \citet{campbell2006household}'s argument that household finance mistakes are concentrated among the least wealthy.

Third, the choice tax documented here is a specific instance of the broader phenomenon that choice can harm when decision costs are heterogeneous. \citet{iyengar2000choice} established this principle in consumer settings; \citet{handel2013adverse} showed it in health insurance; and \citet{choi2009reinforcement} demonstrated persistent suboptimal behavior in retirement savings. My contribution is to show that this mechanism operates at massive scale in pension decumulation, with quantifiable welfare consequences.

The portable mechanism---fixed advice costs creating regressive barriers to optimal choice---applies wherever governments deregulate complex financial decisions. Australia, Singapore, and several EU member states are currently debating UK-style pension flexibility. The evidence here suggests that freedom without equalizing access to guidance may systematically harm the poorest retirement savers.

\section{Institutional Background}

\subsubsection*{Pre-reform regime} Before April 2015, defined-contribution (DC) pension holders in the UK were effectively required to purchase a lifetime annuity with their pension pot. A 25\% tax-free lump sum (the Pension Commencement Lump Sum, or PCLS) could be taken, but the remaining 75\% had to generate a regular income, which in practice meant buying an annuity from an insurance company. Capped drawdown existed as an alternative for larger pots, but was restricted and relatively uncommon.

\subsubsection*{The reform} The Taxation of Pensions Act 2014, commonly known as ``Pension Freedoms,'' took effect on 6 April 2015. The reform eliminated the effective annuity mandate entirely. From that date, DC pension holders aged 55 or over could access their entire pension pot in any combination of: (i) full cash withdrawal (the entire pot in one transaction); (ii) Uncrystallised Funds Pension Lump Sum (UFPLS)---partial withdrawals where 25\% of each withdrawal is tax-free; (iii) flexible drawdown---the pot enters a drawdown wrapper, from which income can be taken as needed; or (iv) annuity purchase, now entirely optional.

\subsubsection*{Tax treatment} The tax consequences of these choices differ markedly. Under full encashment, 25\% of the pot is tax-free and the remaining 75\% is added to the individual's income for that tax year, taxed at the marginal rate (20\% basic, 40\% higher, 45\% additional). For a basic-rate taxpayer with no other income encashing a \pounds30,000 pot: \pounds7,500 is tax-free, the next \pounds12,570 falls within the personal allowance, and the remaining \pounds9,930 is taxed at 20\%, yielding a tax bill of approximately \pounds1,986. Under phased drawdown over ten years, each annual withdrawal of \pounds3,000 would generate only \pounds750 in tax-free income and \pounds2,250 in taxable income---well within the personal allowance, producing zero income tax if no other income is present.

\subsubsection*{Financial advice} The UK's Retail Distribution Review (2013) banned commission-based financial advice, establishing a fee-based model where advisers charge explicit fees---typically \pounds500--\pounds3,000 for pension decumulation advice. Recognizing the advice gap, the government established Pension Wise in 2015, a free guidance service offering one-to-one sessions on pension options. Pension Wise provides general guidance but cannot make specific recommendations.

\subsubsection*{The advice gap} The fixed-cost structure of financial advice creates a natural threshold below which advice is uneconomic. A \pounds2,000 advice fee represents 40\% of a \pounds5,000 pot, 10\% of a \pounds20,000 pot, and 1\% of a \pounds200,000 pot. This gradient in the relative cost of advice is the key mechanism generating differential behavior across pot sizes.

\section{Data}

The primary data source is the FCA Retirement Income Market Data (RIMD), a regulatory return that all pension providers with DC business must submit to the Financial Conduct Authority. The data cover every pension pot accessed for the first time since April 2015, disaggregated by access method, pot-size band, customer age, and (from 2018) advice status.

Two collection regimes exist. From April 2015 to March 2018, approximately 50 firms representing roughly 95\% of DC assets reported on a voluntary basis. From April 2018, reporting became universal, covering all FCA-regulated pension providers. I use both regimes, yielding 17 usable half-year periods from H2 2015 through H2 2023.

Pots are classified into six size bands: less than \pounds10,000; \pounds10,000--\pounds29,999; \pounds30,000--\pounds49,999; \pounds50,000--\pounds99,999; \pounds100,000--\pounds249,999; and \pounds250,000 and above. Pot sizes are measured after deduction of any PCLS for annuity and drawdown entries. Four access methods are recorded: annuity purchase, entry into drawdown (not fully exhausted), first UFPLS payment (not fully exhausted), and full cash withdrawal.

From April 2018, the RIMD also records advice and guidance usage among pension holders making full withdrawals, distinguishing between regulated financial advice, Pension Wise guidance, and no advice or guidance.

\subsection{Summary Statistics}

\Cref{tab:tab:summary} presents cross-period average access method shares by pot-size band. The gradient is monotonic and extreme: the full encashment share falls from 87.6\% for pots under \pounds10,000 to 1.7\% for pots above \pounds250,000. The drawdown share moves in the opposite direction, rising from 6.5\% to 86.5\%.

\input{tables/tab1_summary.tex}

\section{Empirical Strategy}

\subsection{Identification}

The UK Pension Freedoms reform is universal---all DC pension holders gained access simultaneously on 6 April 2015. There is no untreated comparison group. Identification instead comes from \textit{cross-sectional} variation in pot size, which determines the relative cost of financial advice and therefore the probability of choosing a dominated strategy.

The key estimating equation is:
\begin{equation}
\text{Share}_{jt}^{\text{encash}} = \alpha + \beta \log(\text{PotSize}_j) + \gamma_t + \varepsilon_{jt}
\end{equation}
where $j$ indexes pot-size bands, $t$ indexes half-year periods, and $\gamma_t$ are period fixed effects that absorb market-wide trends in pension access behavior. The coefficient $\beta$ captures the pot-size gradient in full encashment propensity. I use pot-size band midpoints (ranging from \pounds5,000 to \pounds375,000) and take the natural logarithm to capture the diminishing marginal effect of pot size. Standard errors are clustered by half-year period ($T = 17$). With only six pot-size bands, clustering by pot size would yield too few clusters for reliable inference; period clustering is the preferred alternative given the panel structure.

\subsection{Identification Assumptions}

This design does not identify a causal effect of pot size on behavior. Pot size is endogenous---it reflects lifetime savings, income, risk preferences, and financial literacy. The parameter $\beta$ is best interpreted as a \textit{descriptive gradient} that summarizes how pension access behavior varies with wealth, conditional on time trends. The welfare computation treats full encashment as dominated (relative to drawdown) based on the tax and investment return arithmetic, not on a claim about what caused the behavior.

The advantage of the design is that $\beta$ is estimated from the universe of pension pots accessed in the UK---there is no sample selection, no survey nonresponse, and no measurement error in the outcome (regulatory return data). The main threat is compositional change: if the distribution of pot sizes shifts over time (e.g., auto-enrollment bringing in more small pots), the cross-period average could be misleading. Period fixed effects address level shifts, and I show in robustness checks that the gradient is stable within early (2015--2019) and late (2020--2024) subperiods.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the core regression results. Column 1 shows the baseline specification: a one-unit increase in $\log(\text{PotSize})$ is associated with a 21.2 percentage point decrease in the full encashment share ($p < 0.001$). The within-$R^2$ is 0.947---pot size alone, conditional on period effects, explains nearly all the variation in encashment behavior.

Column 2 adds an interaction between log pot size and linear time. The interaction coefficient is small ($-0.001$) and statistically insignificant, confirming that the pot-size gradient does not diminish over the nine-year post-reform period. There is no learning effect: small-pot holders are no less likely to fully encash in 2024 than they were in 2015.

Columns 3 and 4 examine the mirror outcomes. The annuity share shows a small positive gradient with pot size, but the relationship is weak ($R^2 = 0.21$). The drawdown share, by contrast, is strongly increasing in pot size ($\beta = 0.190$, $R^2 = 0.987$). The dominant substitution is between full encashment (small pots) and drawdown (large pots), not between encashment and annuities.

\input{tables/tab3_regressions.tex}

To put the gradient in context: moving from the smallest pot band (\pounds5,000 midpoint) to the largest (\pounds375,000 midpoint) represents a 4.3-unit change in log pot size. The predicted change in encashment share is $4.3 \times (-0.212) = -0.91$, or 91 percentage points. This is close to the raw difference observed in the data (88\% minus 2\% = 86 percentage points), confirming that the log-linear specification provides a good approximation of the nonlinear gradient.

\subsection{Mechanism: The No-Advice Trap}

\Cref{tab:tab:advice} presents advice and guidance usage among pension holders making full withdrawals, available for the 2018--2024 period. The advice gradient mirrors the encashment gradient: only 21\% of full withdrawals from pots under \pounds10,000 involve regulated financial advice, compared with 41\% for pots above \pounds250,000. Even including the free Pension Wise guidance service, 68\% of small-pot holders withdraw their pension with no professional help.

\input{tables/tab2_advice.tex}

In a regression framework, the regulated advice rate increases by 3.8 percentage points per unit of $\log(\text{PotSize})$ ($p = 0.02$). The any-help rate (advice plus guidance) increases by 2.9 percentage points ($p = 0.002$). These gradients suggest that the fixed cost of advice is a binding constraint for small-pot holders, and that the government's free guidance service only partially offsets the gap.

The correlational evidence is consistent with advice costs acting as a \textit{participation tax} on optimal decisionmaking. When advice costs \pounds2,000 and the pot is \pounds5,000, seeking advice consumes 40\% of the pension---a prohibitive rate that rational agents would decline even if advice would improve their decision. The result is that the poorest retirees are systematically excluded from the decision support that would prevent dominated strategies.

\subsection{Welfare Loss from Dominated Strategies}

\Cref{tab:tab:welfare} presents a welfare computation comparing full encashment to phased drawdown over ten years for each pot-size band. Two components generate welfare loss from premature encashment.

First, \textit{investment returns forgone}: the encashed pot ceases to earn returns, while a drawdown pot would continue to appreciate. At a 5\% real annual return, the present value of forgone returns on a \pounds5,000 pot is \pounds1,139; on a \pounds174,500 pot, it is \pounds39,756.

Second, \textit{tax penalty}: full encashment concentrates taxable income in a single year, potentially pushing the individual into higher tax brackets. For a \pounds5,000 pot, the taxable portion (\pounds3,750) falls within the personal allowance, so the tax penalty is zero. For a \pounds174,500 pot, lump-sum encashment triggers higher-rate tax on a substantial portion, generating a \pounds38,747 tax penalty relative to phased drawdown.

\input{tables/tab4_welfare.tex}

The aggregate welfare loss across 5.5 million pension pots is approximately \pounds14 billion. This figure is dominated by the £10,000--\pounds29,999 band (\pounds3.9 billion), which combines a high encashment rate (60\%) with a meaningful per-pot loss (\pounds4,854). The \pounds50,000--\pounds99,999 band contributes \pounds2.9 billion despite a lower encashment rate, because the per-pot loss is larger (\pounds26,755).

\subsection{Robustness}

\Cref{tab:robust} presents robustness checks on the main gradient. The coefficient on $\log(\text{PotSize})$ is remarkably stable. Excluding the COVID-affected periods (H1--H2 2020), when pension withdrawals spiked for liquidity reasons, has negligible effect ($\beta = -0.212$, versus $-0.212$ in the baseline). Splitting the sample by age group shows similar gradients for newly retired individuals aged 55--64 ($\beta = -0.136$) and older retirees aged 75+ ($\beta = -0.127$), suggesting the mechanism is not driven by age-specific factors. Splitting the sample by era---early (2015--2019, $\beta = -0.211$) versus late (2020--2024, $\beta = -0.214$)---confirms the absence of a learning effect.

\input{tables/tab5_robustness.tex}

\section{Discussion}

The choice tax documented here has a precise structure. It is not that small-pot holders are irrational or financially illiterate; it is that the institutional architecture makes optimal behavior prohibitively expensive for them. A \pounds5,000 pension holder who pays \pounds2,000 for advice and then follows the optimal drawdown strategy would need to recover 40\% of their pot in advisory fees before breaking even. Full encashment, while tax-inefficient, is \textit{free} in direct costs and \textit{simple} in cognitive costs. The revealed preference for encashment is consistent with rational behavior under fixed advice costs.

This framing connects to \citet{chetty2014active}'s distinction between active and passive decisionmaking in retirement savings. The Pension Freedoms reform shifted the pension system from a passive default (annuitization) to an active choice regime, without adjusting the cost structure of the decision support needed to navigate active choice. The predictable result is that those with the smallest stakes---and the least to gain from optimizing---choose the simplest available option.

The policy implication is specific. The problem is not pension freedom itself but the \textit{advice gap}---the wedge between the cost of professional guidance and the value of smaller pension pots. Targeted interventions could include: subsidized advice for pots under \pounds30,000, an opt-out default into drawdown (with the option to encash), or expanding Pension Wise from guidance to personalized recommendation authority. The Australian experience with ``intrafund advice''---low-cost advice bundled within the pension product---provides a potential model.

The welfare loss estimate of \pounds14 billion should be interpreted as an upper bound, subject to several strong assumptions. It treats every full encashment as dominated, assumes a 5\% real return, a ten-year drawdown horizon, and basic-rate taxation with no other income. Each assumption can be questioned. Individuals with poor health, high discount rates, or urgent liquidity needs may rationally prefer immediate cash. If only half of encashments are genuinely welfare-reducing, the aggregate loss is \pounds7 billion. Varying the return assumption from 3\% to 7\% shifts the total by approximately $\pm$30\%. The qualitative conclusion is robust: the combination of regressive encashment rates and meaningful per-pot losses generates a welfare cost that is large by any plausible calibration.

\section{Conclusion}

Pension freedom without advice access is a choice tax on the poor. In the decade since the UK eliminated annuity mandates, small pension holders have overwhelmingly chosen the simplest and least efficient strategy---full cash withdrawal---while large pension holders have smoothly transitioned to flexible drawdown. The pot-size gradient in behavior is enormous, stable over time, and mediated by differential access to financial advice. The aggregate welfare cost is on the order of \pounds14 billion.

The lesson is portable: any country considering pension liberalization should pair freedom with equalized access to guidance. Otherwise, deregulation redistributes welfare from the least to the most financially sophisticated---the opposite of its stated intent.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Data Appendix}

The FCA Retirement Income Market Data are published annually as Excel workbooks, available from the FCA's website. The 2023--24 publication contains both the 2018--24 universal collection and the earlier 2015--18 sample-based collection in separate worksheets.

The six pot-size bands use the following midpoints for regression analysis: less than \pounds10,000 $\to$ \pounds5,000; \pounds10,000--\pounds29,999 $\to$ \pounds19,500; \pounds30,000--\pounds49,999 $\to$ \pounds39,500; \pounds50,000--\pounds99,999 $\to$ \pounds74,500; \pounds100,000--\pounds249,999 $\to$ \pounds174,500; \pounds250,000 and above $\to$ \pounds375,000.

The welfare computation assumes: (i) 5\% annual real return on pension investments; (ii) phased drawdown over 10 years with equal annual withdrawals; (iii) basic-rate taxpayer with no other income; (iv) 2024--25 personal allowance of \pounds12,570 and higher-rate threshold of \pounds50,270.

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde.tex}

\end{document}
