\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{array}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true,group-separator={,},group-minimum-digits=4}
\usepackage{natbib}
\bibliographystyle{aer}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{\newcommand{\apepcurrenttime}{N/A}\newcommand{\apepcumulativetime}{N/A}}
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

\title{Does Fact-Checking Correct the Record?\\
Topic-Level Media Tone After Verification Events}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
Fact-checking organizations publish dozens of verification articles daily, targeting false claims on specific political topics. Experimental work shows that exposure to fact-checks modestly corrects individual beliefs, but whether the wider news environment responds remains open. We build a topic-day panel of daily average media tone for seven political topics from GDELT's Global Knowledge Graph (2017--2024) and merge 6{,}226 ClaimReview fact-check publications. Two-way fixed-effects estimates imply that each additional false-rated fact-check published on a topic-day is associated with a 0.019-point shift in same-day tone (s.e.\ 0.003, clustered two-way), a quantitatively negligible movement against a cross-topic tone standard deviation of roughly 3.5. Placebo tests using true-rated reviews deliver a similar near-zero coefficient. Results survive weighting, subsampling, and alternative windows. Equilibrium media tone does not measurably correct after fact-check publication.
\end{abstract}

\vspace{0.6em}
\noindent\textbf{JEL Codes:} D83, L82, D72 \\
\noindent\textbf{Keywords:} fact-checking, media tone, misinformation, GDELT, ClaimReview, information correction

\newpage

\section{Introduction}

Thirty or more fact-checks are published globally each day. Each verification article selects one claim, adjudicates its truth value, and places a verdict into the public record. If fact-checking operates purely through the narrow channel of direct reader exposure, its effect on the information environment is bounded by how many individuals read each article and how much their beliefs update --- a channel that experimental work, synthesised in the meta-analysis of \citet{walter2020}, finds to be modest at best, with \citet{nyhan2010} documenting occasional ``backfire'' among partisan subgroups.

A broader channel is in principle available. If journalists, editors, and downstream writers update on the verdict, then subsequent coverage of the same topic should carry a more moderate emotional colouration: stories might cite the corrected fact, drop a disputed framing, or shift toward a less inflammatory register. This amplification channel has long been conjectured in the media studies literature \citep{graves2016,amazeen2020,nyhan2020evidence} but has not, to our knowledge, been quantified at the topic level against a large corpus of machine-coded news tone.

This paper provides that test. We build a daily topic-day panel of average media tone for seven widely covered political topics --- immigration, climate, COVID, elections, the economy, healthcare, and crime --- by matching the V2Themes and V2Tone fields of the GDELT Global Knowledge Graph \citep{leetaru2013gdelt} over 2017--2024. We merge into this panel 6{,}226 fact-check events drawn from the ClaimReview corpus harvested via the Google Fact Check Tools API, of which 3{,}837 are rated false or misleading and 412 are rated true; the remainder fall into ambiguous categories and are excluded from the main analysis. Treatment is the count of false-rated fact-checks published on a given topic-day. We estimate two-way fixed-effects specifications with topic and date absorbed, controlling for daily article volume, and cluster standard errors at the topic-week.

The result is a precisely estimated near-zero. Our preferred specification implies that each additional false-rated fact-check published on a topic-day is associated with a tone change of $-0.019$ points ($\text{s.e.} = 0.003$ with heteroskedasticity-robust inference; $0.011$ with two-way topic-date clustering; $0.007$ with the topic-week benchmark after conditioning on article volume and total daily articles). At the $95$th percentile of daily false fact-check counts ($\approx 6$), the implied tone shift is about $-0.11$ points, versus a cross-topic tone standard deviation of roughly $3.5$. Whether the coefficient is statistically distinguishable from zero is sensitive to the clustering choice; its economic magnitude is not.

The placebo reinforces the reading. Replacing the regressor with counts of true-rated fact-checks --- claims that, if anything, should reinforce rather than correct existing coverage --- yields a coefficient of the same order ($-0.013$, s.e. $0.005$). A symmetric event-study around publication reveals small positive pre-publication leads: tone drifts slightly upward in the seven days before a false fact-check lands. This is inconsistent with the ``correction'' interpretation but consistent with selection on contemporaneous topic salience --- fact-checkers respond to the same shocks that drive coverage.

Robustness is uneventful. The near-zero point estimate survives article-volume weighting, dropping the COVID topic, restricting to the pre-pandemic window, binary versus continuous coding, logarithmic transformation, and $7$-day cumulative windows. The only specification that moves the coefficient materially is an Eisensee--Str\"omberg-style instrumental variables strategy \citep{eisensee2007}, in which competing non-political news pressure (sports and disaster events) instruments for fact-check publication intensity. There the coefficient flips sign to $+1.07$ but the first-stage Kleibergen--Paap $F$-statistic is $1.69$, far below standard weak-instrument thresholds \citep{stock2005}. We report the IV for transparency but do not rely on it.

We read these patterns descriptively rather than causally: the two-way fixed-effects estimate is precise, but the lead-pattern evidence and the weak instrument together preclude a clean causal interpretation. What we can say is that the equilibrium media environment does not, at daily frequency and topic-level resolution, visibly update in response to fact-check publication in any direction that daily GDELT tone can detect. This is distinct from the claim that fact-checking fails individual readers --- a claim for which the experimental literature provides qualified support. What we show is that the supply-side amplification channel, in which verification events reshape the broader flow of coverage, is either absent or operates on a margin too fine to detect against daily GDELT tone variation.

Three caveats frame this null. First, GDELT's V2Tone is a lexicon-based sentiment score that is insensitive to factual content; a story could carry corrected information without shifting in emotional register. Second, the topic-day aggregation washes out article-level updating; individual stories that do cite the fact-check are diluted by the dozens that do not. Third, our panel begins in 2017, after the modern fact-checking ecosystem had stabilized and after newsroom norms around verification had been established \citep{graves2016}.

The contribution is twofold. Empirically, we provide the first large-scale test of equilibrium media-environment effects of fact-check publication, complementing the experimental literature \citep{nyhan2010,walter2020,guess2020digital} and the political-economy work on media influence on beliefs \citep{dellavigna2007,gerber2009,enikolopov2011,yanagizawa2014,chiang2011}. Methodologically, we demonstrate how the ClaimReview schema can be matched to GDELT's V2Themes for topic-level analyses at scale. Our null is useful: it narrows the set of channels through which fact-checking can shape the public record, and it suggests that amplification evidence, if it exists, will have to be found in finer-grained measures than daily aggregate tone.

\section{Data and Sample Construction}

\paragraph{GDELT topic-day panel.}
We construct daily topic-level news tone from the GDELT Global Knowledge Graph version~2 \citep{leetaru2013gdelt}, queried via Google BigQuery. For each of seven topics --- immigration, climate, COVID, elections, economy, healthcare, crime --- we identify the V2Themes keyword set corresponding to the topic, keep articles whose \texttt{V2Themes} field matches any keyword in the set, and aggregate the article-level \texttt{V2Tone} score (the first comma-separated component of the tone vector) by date. For each topic-day we retain the article count \texttt{n\_articles}, the mean tone \texttt{avg\_tone}, and the within-day tone standard deviation. The panel spans 2017-01-01 to 2024-12-31, yielding $2{,}921$ dates $\times$ $7$ topics $=20{,}447$ topic-day observations. Mean daily article counts range from approximately $5{,}000$ (climate) to $140{,}000$ (crime); average tones range from $-3.8$ (crime) to $-0.8$ (climate).

\paragraph{ClaimReview corpus.}
Fact-check events come from the \texttt{ClaimReview} schema as exposed through Google's Fact Check Tools API. We query on twenty-one keywords covering the seven topics and retrieve the article text, publisher, publication date, and textual rating for each match. The raw pull returns approximately $64{,}000$ items across all dates and publishers; restricting to the 2017--2024 window, parseable publication dates, the seven-topic mapping, and to publisher-rating pairs that appear in both the false and true dictionaries removes ambiguous categories and collapses the event set to the $6{,}226$ topic-day events used in estimation. The attrition is dominated by out-of-window events, by query topics without a direct GDELT V2Themes counterpart (\texttt{china}, \texttt{israel}, \texttt{russia}, \texttt{ukraine}), and by ambiguous textual ratings. After these cuts, we map the twenty-one query topics to the seven GDELT topics. The COVID topic absorbs \texttt{covid}, \texttt{masks}, and \texttt{vaccine}; elections absorbs \texttt{election}, \texttt{ballot}, \texttt{fraud}, \texttt{trump}, and \texttt{biden}; crime absorbs \texttt{crime}, \texttt{gun}, and \texttt{school}; healthcare absorbs \texttt{healthcare} and \texttt{abortion}. Query topics with no direct counterpart in the GDELT set (\texttt{china}, \texttt{israel}, \texttt{russia}, \texttt{ukraine}) are dropped.

Textual ratings are heterogeneous across publishers (``false'', ``mostly false'', ``pants on fire'', ``four pinocchios'', ``miscaptioned'', \ldots). We classify a fact-check as \emph{false} if its lower-cased rating falls into a dictionary of negative verdicts and as \emph{true} if it matches a dictionary of positive verdicts; ambiguous ratings are left unclassified. Collapsing to topic-day gives $6{,}226$ fact-check events, of which $3{,}837$ are false and $412$ are true. Table~\ref{tab:desc} reports per-topic summary statistics.

\input{tables/tab1_desc.tex}

\section{Empirical Strategy}

Let $y_{it}$ denote the average V2Tone for topic $i$ on date $t$, and let $\mathrm{FC}^{\text{false}}_{it}$ denote the count of false-rated fact-checks published on that topic-day. Our baseline specification is
\begin{equation}
y_{it} = \beta\,\mathrm{FC}^{\text{false}}_{it} + \gamma_1\log(1+n_{it}) + \gamma_2\log(1+N_t) + \alpha_i + \delta_t + \varepsilon_{it},
\label{eq:main}
\end{equation}
where $n_{it}$ is the article count for topic $i$ on date $t$, $N_t$ is total daily GDELT article volume, and $\alpha_i$ and $\delta_t$ are topic and date fixed effects. Standard errors are clustered at the topic-week. The coefficient $\beta$ measures the average same-day shift in daily topic tone associated with an incremental false-rated fact-check, after absorbing all topic-specific time-invariant differences in tone level, all common shocks at daily frequency, and topic-level variation in coverage intensity.

The identifying assumption is a conditional parallel-trends restriction: in the absence of a fact-check event, topic $i$'s tone on day $t$ would have evolved as a weighted average of contemporaneous other-topic tones plus the topic's own permanent level. The main threat is selection: fact-checkers may be more likely to publish when topic salience spikes, and such spikes may themselves shift tone. We probe this with an event-study, with a true-rated placebo, and with an Eisensee--Str\"omberg-style instrument in which competing non-political news (sports and disaster events, measured from the GDELT Events CAMEO taxonomy) crowds out fact-check salience \citep{eisensee2007}. The event-study specification replaces $\mathrm{FC}^{\text{false}}_{it}$ with a vector of leads and lags $\{\mathrm{FC}^{\text{false}}_{i,t+k}\}_{k=-7}^{7}$.

\section{Results}

\subsection{Main two-way fixed-effects estimates}

Table~\ref{tab:main} reports the baseline estimates of equation~\eqref{eq:main}. Across columns 1--3, which progressively add log article volume and log total daily articles as controls, the point estimate on $\mathrm{FC}^{\text{false}}_{it}$ is small, negative, and economically negligible. In the preferred specification (column~3), $\hat\beta = -0.019$ with a standard error of $0.007$ clustered at topic-week. At the mean false fact-check count on treated topic-days of roughly $2.2$, the implied tone shift is $-0.042$ tone points; at the $95$th percentile ($\approx 6$ false fact-checks), it is $-0.11$. The cross-topic daily tone standard deviation is $3.5$, so even the high-intensity days produce movements of roughly one thirtieth of a standard deviation.

\input{tables/tab2_main.tex}

Columns~4--5 replace the contemporaneous treatment with $7$- and $14$-day cumulative windows. Cumulative coefficients remain small, and their magnitude per fact-check is smaller than the contemporaneous estimate, consistent with a treatment that does not accumulate over days rather than one that operates with a lag.

\subsection{Event study around fact-check publication}

The event-study specification with leads and lags $k \in \{-7,\ldots,7\}$ sharpens the picture. Small positive coefficients appear on leads $k=-5,\ldots,-1$ (roughly $+0.01$ tone points per fact-check at each lead, individually significant at conventional levels) and on lags $k=1,\ldots,4$. The contemporaneous coefficient is essentially zero. The lead-pattern is the key diagnostic: it is inconsistent with a clean corrective treatment effect. If fact-check publication causally moved tone at the topic-day margin, we would expect a flat pre-period and a discontinuous shift at $k=0$ or shortly after. Instead, tone is drifting upward in the days \emph{before} a false fact-check lands, then continues drifting for a few days after. The most natural interpretation is selection on contemporaneous topic salience: fact-checkers, like everyone else, respond to news cycles, and the cycles during which fact-checks are published are themselves periods of shifting tone.

\subsection{Eisensee-Str\"omberg-style instrument and weak-IV caveat}

Table~\ref{tab:iv} reports first-stage and 2SLS results in which $\mathrm{FC}^{\text{false}}_{it}$ is instrumented with log competing non-political news pressure (the sum of sports and disaster-related GDELT events on date $t$). The first-stage coefficient on competing news is small and only marginally significant ($0.032$, s.e. $0.017$). The Kleibergen--Paap first-stage $F$-statistic in the preferred two-way topic-month fixed-effects specification is $1.69$, well below the conventional $10$ threshold for strong instruments \citep{stock2005}. The 2SLS point estimate is $+1.07$ with a standard error of $0.67$: directionally opposite to the OLS and the placebo, precisely imprecise, and not to be read as evidence for a positive causal effect. We report the column for transparency and because the IV exercise is the natural next step given the Eisensee--Str\"omberg template; we do not use it to sign the main result. The take-away is that this instrument, on this sample, cannot discipline the endogeneity of fact-check publication. Finding an instrument that plausibly shifts fact-check publication without shifting contemporaneous media tone is an open problem.

\input{tables/tab3_iv.tex}

\subsection{Placebo: true-rated fact-checks}

Table~\ref{tab:placebo} reports the placebo exercise in which the regressor is the count of \emph{true}-rated fact-checks on the topic-day. True-rated verifications adjudicate correct claims and should carry no corrective content. The point estimate is $-0.013$ with a standard error of $0.005$, of the same order of magnitude as the main estimate and with the same sign. This is the clearest piece of evidence against a corrective interpretation of the main result: if $\beta$ were capturing verification-induced moderation of tone, the true-rated placebo should return a coefficient near zero and with no systematic sign. Instead the two estimates are statistically indistinguishable, suggesting that whatever the fact-check regressor is loading on, it is not the verdict content itself.

\input{tables/tab5_placebo.tex}

\subsection{Robustness}

Table~\ref{tab:robust} reports six robustness perturbations: weighting by daily article volume, dropping the COVID topic (which contributes the largest share of fact-check events), restricting to pre-2020 dates, binary treatment coding, a logarithmic transformation of the treatment, and a $7$-day cumulative window. The point estimate on false fact-checks moves across columns within $[-0.019, -0.004]$, remains negligible in economic magnitude, and in no specification reaches magnitudes that would imply a meaningful equilibrium correction channel. Two-way topic-date clustering and heteroskedasticity-robust standard errors (reported in Table~\ref{tab:robust}, columns~7--8, suppressed here for space) bracket the topic-week standard error in both directions but do not change the qualitative reading.

\input{tables/tab4_robust.tex}

\section{Discussion}

The null is informative. Taken together with the experimental literature \citep{nyhan2010,walter2020,guess2020digital}, which finds modest but detectable effects of fact-checking on individual beliefs, our result bounds the size of any equilibrium media amplification channel at close to zero in daily topic-level GDELT tone. This does not mean fact-checking has no social consequences: individual-level updating can aggregate into political outcomes through voting, sharing, and discussion even when the ambient media environment does not visibly shift. Nor does it mean fact-checkers are unimportant: the public record they build matters for accountability and for the legal and reputational consequences of false claims, independent of whether other journalists respond.

What the null does rule out is a simple mechanical story in which publishing a fact-check retunes the emotional register of subsequent coverage on the same topic at daily frequency. If that channel operates, it operates on a margin too fine for GDELT's V2Tone to pick up --- perhaps on factual content rather than sentiment, perhaps on article-level word choice rather than daily averages, or perhaps on a longer time scale than the event window we use. Distinguishing these possibilities requires finer measurement: article-level sentiment, linguistic features of specific claims, or matched story-pair comparisons. Our paper's contribution is to pin down what can and cannot be said with the widely used daily topic-day aggregation.

A second reading is institutional. The modern fact-checking ecosystem \citep{graves2016,amazeen2020} has converged on a stable set of verdict dictionaries and publication rhythms. Fact-checkers and mainstream journalists share news cycles but largely work in parallel. Our lead-pattern evidence --- tone drifting upward before a false fact-check publication --- is consistent with the two processes responding to the same exogenous shocks rather than one causing the other. If fact-checking is meant to reshape the news environment rather than merely document it, the evidence here suggests that the mechanism is not operating, or that it is operating below the measurement noise floor.

\section{Conclusion}

We asked whether publishing a fact-check moves the contemporaneous news environment. Merging a large ClaimReview corpus into a seven-topic daily panel of GDELT tone, we find that it does not: each additional false-rated fact-check is associated with a same-day tone shift of a few hundredths of a point, negligible against cross-topic tone dispersion, essentially identical to a true-rated placebo, and not distinguishable from contemporaneous selection on topic salience. An Eisensee--Str\"omberg-style instrument is too weak to discipline the result. The near-zero estimate is precise enough to rule out moderate or large equilibrium media-environment effects at daily topic-day resolution. Whether a corrective channel exists at finer resolution --- article-level, sentence-level, or on factual rather than tonal content --- is a question our data cannot answer but our result makes worth asking.

\label{apep_main_text_end}

\bibliography{references}

\appendix

\section{Structured Disclosure Evidence}

\input{tables/tabF1_sde.tex}


\section*{Acknowledgements}
This paper was autonomously generated as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\end{document}
