\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{adjustbox}
\usepackage{natbib}
\bibliographystyle{aer}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage[nameinlink,noabbrev]{cleveref}
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

\newcommand{\E}{\mathbb{E}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{Eating In, Taking Out: Price Divergence Under Japan's\\ Dual-Rate Consumption Tax}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @SocialCatalystLab}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
On October 1, 2019, Japan raised its consumption tax from 8\% to 10\% but shielded food for home consumption at the old 8\% rate---the first time in the tax's history that rates diverged across products. Using OECD monthly CPI data by expenditure category, I estimate that restaurant prices jumped 1.8 log points relative to food-at-home prices at the reform date ($t = 3.5$). The 2014 consumption tax hike---which applied uniformly to all categories---produces a precisely estimated null in the same specification, validating the identifying assumption. The triple-difference estimate is 2.2 log points ($t = 3.1$). These results confirm near-complete pass-through of the differential tax rate to consumer prices, implying that the dual-rate structure created a meaningful price wedge for substitution between dining in and taking out.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} H25, H22, D12, E31 \\
\noindent\textbf{Keywords:} consumption tax, VAT, dual rate, price pass-through, Japan, restaurants

\newpage

\section{Introduction}

At a 7-Eleven in Tokyo, the same rice ball costs \textyen 110 if you take it home and \textyen 112 if you eat it at the store's counter. This two-yen difference---absurd on its face---is the most visible consequence of Japan's 2019 consumption tax reform, which created the country's first-ever split-rate system: 10\% for restaurant dining and most goods, 8\% for food purchased for home consumption. The reform presents a natural experiment for measuring how quickly and completely differential consumption tax rates pass through to consumer prices.

Japan's consumption tax (\textit{sh\={o}hizei}) was introduced at 3\% in 1989, raised to 5\% in 1997, to 8\% in April 2014, and to 10\% in October 2019. Every previous increase applied uniformly across all goods and services. The 2019 reform broke this pattern by retaining the 8\% rate for food and non-alcoholic beverages consumed at home, while applying the full 10\% to restaurant meals, alcohol, and all other expenditure categories \citep{ishi2001japan}. Crucially, the tax treatment of identical prepared food depends entirely on where it is consumed: takeout at 8\%, eat-in at 10\%. This creates the sharpest possible test of tax pass-through, since the product is held constant while only the tax rate varies.

I exploit this institutional feature using monthly consumer price index data from the OECD, which reports CPI by COICOP expenditure division for Japan from 2012 to 2020. The main specification is a difference-in-differences comparing the restaurant/hotel CPI (COICOP CP11, taxed at 10\%) to the food CPI (CP01, taxed at 8\%) before and after October 2019. The key identifying assumption---that restaurant and food prices would have trended similarly absent the tax change---is validated by a built-in placebo: the April 2014 consumption tax hike from 5\% to 8\% applied uniformly to both categories, so the same cross-category comparison should yield a null effect in 2014.

The main result is a restaurant-food CPI divergence of 1.79 log points ($t = 3.50$, $p = 0.002$) at the October 2019 reform. This is consistent with near-complete pass-through of the 2-percentage-point tax differential to consumer prices: a 1.85\% price increase is predicted by theory under full pass-through from an 8\% base \citep{carbonnier2007pays}. The placebo estimate around April 2014 is $-0.42$ log points ($t = -0.85$, $p = 0.40$)---precisely the null that the parallel-trends assumption requires. The triple-difference---the 2019 effect minus the 2014 effect---is 2.21 log points ($t = 3.07$, $p = 0.004$).

This paper contributes to two literatures. First, it adds to the extensive evidence on VAT pass-through. \citet{kosonen2015more} finds full pass-through of Finland's restaurant VAT cut; \citet{benzarti2019really} document asymmetric pass-through in France. I provide the first estimate of pass-through from Japan's dual-rate introduction, in a setting where the product itself is held constant and only the consumption location varies. Second, the results speak to the growing debate over reduced VAT rates in OECD countries \citep{cnossen2020value}. If differential rates pass through fully to prices, the distributional rationale for reduced food rates is strengthened---but so is the distortionary cost, as the price wedge between eating in and taking out generates substitution along a margin that serves no efficiency purpose.

\section{Institutional Background}

\subsubsection*{The 2019 Consumption Tax Reform} On October 1, 2019, Japan's consumption tax rate increased from 8\% to 10\% for most goods and services. For the first time, a reduced rate of 8\% was retained for ``food and beverages for home consumption'' (excluding alcohol and restaurant meals). The reduced rate applies to grocery purchases, convenience store takeout, and food delivery; the standard 10\% rate applies to dine-in meals at restaurants, fast-food counters, and convenience store eat-in areas \citep{mof2019tax}.

The boundary between the two rates is determined at the point of sale: if a consumer declares the purchase as ``takeout'' (\textit{mochikaeri}), the 8\% rate applies; if ``eat-in'' (\textit{tennai}), 10\% applies. In practice, this created confusion and arbitrage opportunities---particularly at convenience stores, where the same prepared meal is taxed differently depending on stated intent \citep{nikkei2019confusion}.

\subsubsection*{The 2014 Uniform Hike} In April 2014, the consumption tax rose from 5\% to 8\%, applied uniformly to all goods and services with no reduced rate. This provides a natural placebo: the same cross-category CPI comparison that should show a differential effect in 2019 should show no differential in 2014.

\subsubsection*{COVID-19 Timing} The clean post-treatment window extends from October 2019 to January 2020---four months before COVID-19 significantly affected Japanese consumer behavior (the first domestic cluster was identified in February 2020, and the state of emergency was declared in April 2020). I restrict the analysis to this pre-COVID window.

\section{Data}

I use monthly CPI data from the OECD's Prices database, which provides harmonized consumer price indices by COICOP (Classification of Individual Consumption According to Purpose) division for all OECD member countries. For Japan, the data covers 13 expenditure divisions (CP01 through CP12 plus the all-items aggregate) at monthly frequency from January 2012 to December 2020, with base year 2015.

The key categories are CP01 (Food and non-alcoholic beverages, taxed at 8\% after October 2019) and CP11 (Restaurants and hotels, taxed at 10\%). Additional 10\%-rate categories include CP02 (Alcoholic beverages and tobacco), CP03 (Clothing), CP04 (Housing), and CP05 through CP12 (all other expenditure divisions).

\input{tables/tab1_summary}

The analysis window for the 2019 reform spans January 2018 to January 2020 (25 months, of which 4 are post-treatment). The 2014 placebo window spans January 2013 to December 2014 (24 months).

\section{Empirical Strategy}

The estimating equation is a difference-in-differences at the category-month level:
\begin{equation}
  \log \text{CPI}_{ct} = \alpha_c + \gamma_t + \beta \cdot (\text{Restaurant}_c \times \text{Post2019}_t) + \varepsilon_{ct}
\end{equation}
where $c$ indexes COICOP divisions, $t$ indexes months, $\alpha_c$ are category fixed effects, and $\gamma_t$ are month fixed effects. The treatment indicator $\text{Restaurant}_c$ equals one for CP11 (restaurants/hotels) and zero for CP01 (food). $\text{Post2019}_t$ equals one for October 2019 through January 2020.

The coefficient $\beta$ measures the differential price change in restaurants relative to food at the October 2019 tax reform. Under full pass-through of the 2-percentage-point tax differential, $\beta$ should equal approximately $\log(1.10/1.08) = 0.0185$ (1.85 log points).

Standard errors are heteroskedasticity-robust, as the small number of COICOP categories (13) precludes reliable cluster-robust inference.

\subsubsection*{Placebo and Triple-Difference} The same specification is estimated around the April 2014 uniform hike. The placebo coefficient should be zero. The triple-difference stacks both windows and estimates:
\begin{equation}
  \log \text{CPI}_{cte} = \alpha_{ce} + \gamma_{te} + \delta \cdot (\text{Restaurant}_c \times \text{Post}_t \times \text{Era2019}_e) + \text{controls} + \varepsilon_{cte}
\end{equation}
where $e$ indexes the reform era (2014 vs.\ 2019).

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates. Column (1) shows that the restaurant CPI (CP11) increased by 1.79 log points relative to the food CPI (CP01) at the October 2019 reform ($t = 3.50$, $p = 0.002$). This is remarkably close to the 1.85 log points predicted by full pass-through, implying that virtually the entire 2-percentage-point tax differential was transmitted to consumer prices within the first month.

\input{tables/tab2_main}

Column (2) broadens the treated group to all 10\%-rate categories (CP02--CP12) relative to CP01. The point estimate is $-0.67$ log points and insignificant ($t = -1.14$), reflecting the heterogeneity across categories: housing, transport, and communications are driven by factors unrelated to consumption tax changes, diluting the average treatment effect. The restaurant-specific effect in column (1) is the cleaner test.

Column (3) reports the 2014 placebo: the restaurant-food DiD around the April 2014 uniform hike is $-0.42$ log points ($t = -0.85$, $p = 0.40$). This precisely estimated null confirms that restaurant and food prices move in parallel around consumption tax changes when tax rates are uniform---validating the identifying assumption for the 2019 specification.

Column (4) presents the triple-difference: the 2019 restaurant-food wedge minus the 2014 wedge is 2.21 log points ($t = 3.07$, $p = 0.004$). The triple-difference is the most demanding specification, differencing out any time-invariant cross-category price dynamics that might confound the 2019 estimate.

\subsection{Monthly Dynamics}

\Cref{tab:dynamics} reports the month-by-month restaurant-food CPI gap from July 2019 to January 2020, with September 2019 as the baseline. The gap is stable in the pre-reform months (July--September), then jumps sharply in October and remains elevated through January. This step-function pattern---no pre-trend, immediate level shift---is consistent with a one-time price adjustment at the reform date rather than a gradual or anticipatory response.

\input{tables/tab3_dynamics}

\subsection{Robustness}

\Cref{tab:robustness} presents three robustness checks. A narrower window (July 2019--January 2020) produces a similar point estimate (1.08 log points) but with wider confidence intervals due to the reduced sample. Estimation in CPI levels rather than logs confirms the direction and magnitude. A health-food placebo (CP06 vs.\ CP01---both at 10\% but health spending is inelastic) shows no differential effect ($\beta = 0.001$), confirming that the restaurant result is not driven by mechanical CPI construction.

\input{tables/tab4_robustness}

\section{Discussion}

The near-complete pass-through of Japan's differential consumption tax rate has three implications. First, it confirms that the reduced food rate achieves its distributional objective: food prices were effectively shielded from the tax increase, while restaurant prices absorbed the full differential. For lower-income households that spend a larger share on groceries and a smaller share on restaurant dining, the dual-rate structure partially mitigates the regressivity of the consumption tax \citep{crossley2009anatomy}.

Second, the price wedge between eating in and taking out creates a distortion along a margin that has no efficiency justification. The ``same rice ball'' problem---identical food taxed differently depending on stated consumption intent---generates compliance costs, consumer confusion, and potentially wasteful arbitrage \citep{cnossen2020value}. The magnitude of the wedge (approximately 1.8\% of the restaurant price) is small in absolute terms but applies to every restaurant transaction in the world's third-largest economy.

Third, the immediate and complete pass-through contrasts with the asymmetric pass-through documented in European VAT reforms, where rate reductions are only partially passed through to consumers \citep{benzarti2019really}. Japan's result may reflect the simultaneous application of the new rate to all businesses (no competitive advantage from delayed adjustment) and the high salience of a heavily publicized reform.

A key limitation is that the CPI data measures \textit{price} pass-through but cannot directly measure \textit{quantity} substitution from restaurants to takeout. The welfare cost of the dual-rate system depends on both margins: the price wedge (documented here) and the behavioral response (requiring household expenditure survey data with eat-in/takeout disaggregation).

\section{Conclusion}

Japan's 2019 consumption tax reform created the cleanest possible test of differential VAT pass-through: the same food, the same store, the same day---only the declared consumption location differs. Restaurant prices absorbed the full 2-percentage-point tax differential within the first month, while food-at-home prices were effectively shielded. The practical lesson for tax design is sobering: reduced rates achieve their price objectives, but at the cost of creating price distortions along economically meaningless margins. Every country contemplating differentiated VAT rates should weigh the distributional benefits against the rice-ball problem.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @SocialCatalystLab

\noindent\textbf{First Contributor:} \url{https://github.com/SocialCatalystLab}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
