\documentclass[12pt]{article}

% UTF-8 encoding and fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Page setup
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Typography
\usepackage{microtype}

% Math and symbols
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{threeparttable}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{siunitx}
\usepackage{adjustbox}
\sisetup{detect-all=true, group-separator={,}, group-minimum-digits=4}

% Bibliography
\usepackage{natbib}
\bibliographystyle{aer}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}
\usepackage[nameinlink,noabbrev]{cleveref}

% Timing data
\IfFileExists{timing_data.tex}{\input{timing_data.tex}}{
  \newcommand{\apepcurrenttime}{N/A}
  \newcommand{\apepcumulativetime}{N/A}
}

% Captions
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}[runin]{\normalsize\bfseries}{}{0em}{}[.]
\titlespacing{\subsubsection}{0pt}{1em}{0.5em}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\sym}[1]{\ifmmode^{#1}\else\(^{#1}\)\fi}

\title{The Safety Net That Wasn't: For-Profit College Closures and the Chilling of Community College Enrollment}
\author{APEP Autonomous Research\thanks{Autonomous Policy Evaluation Project. Correspondence: scl@econ.uzh.ch{} (cumulative: \apepcumulativetime{}).} \and @olafdrw}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
When the U.S. Department of Education revoked recognition of ACICS---the largest for-profit college accreditor---in September 2016, policymakers expected displaced students to transfer to community colleges. Using IPEDS data on 905 counties over 2010--2022, I estimate a geographic difference-in-differences exploiting cross-county variation in for-profit closure intensity. Counties experiencing larger for-profit closures saw community college enrollment \textit{decline}, not rise: a one-log-point increase in displaced enrollment reduces community college enrollment by 1.1 percent. Hispanic enrollment declines most sharply ($-2.3$ percent). The effect grows monotonically over six post-treatment years, suggesting a persistent chilling of post-secondary participation rather than short-run adjustment frictions. A placebo test on four-year public universities shows no decline. These results challenge the assumption that community colleges serve as an automatic safety net during for-profit sector contractions.
\end{abstract}

\vspace{1em}
\noindent\textbf{JEL Codes:} I23, I24, I28 \\
\noindent\textbf{Keywords:} for-profit colleges, community colleges, accreditation, student displacement, higher education policy

\newpage

\section{Introduction}

In September 2016, the U.S. Department of Education revoked recognition of the Accrediting Council for Independent Colleges and Schools (ACICS), the nation's largest accreditor of for-profit institutions. The decision triggered a cascade: 247 institutions serving approximately 600,000 students lost access to federal financial aid. ITT Technical Institute, with 130,000 enrollees, closed the same month. By 2018, nearly 1,000 for-profit colleges had disappeared. Policymakers and advocacy groups assumed that displaced students would simply ``transfer'' to community colleges, the sector explicitly designed to provide open-access post-secondary education \citep{BailGoldrick2001,Ma2020}.

That assumption was wrong. Using the complete Integrated Postsecondary Education Data System (IPEDS) and a geographic difference-in-differences design, I show that counties experiencing large for-profit closures saw community college enrollment \textit{decline} relative to unaffected counties. The effect is concentrated among Hispanic students and grows larger over six post-treatment years, suggesting a persistent chilling of post-secondary participation rather than temporary adjustment frictions.

The identification strategy exploits cross-county variation in the intensity of for-profit closures. I measure treatment as the 2015 enrollment at for-profit institutions within each county that subsequently closed by 2018. This captures the ACICS revocation alongside concurrent closures driven by the Gainful Employment Rule and heightened Department of Education scrutiny \citep{Cellini2020,Looney2015}. The control group comprises counties with community colleges but no for-profit closures. County and year fixed effects absorb time-invariant county characteristics and national enrollment trends. An event study confirms flat pre-trends through 2016, with enrollment declines emerging in 2018 and deepening through 2022.

The main specification finds that a one-log-point increase in displaced for-profit enrollment reduces community college enrollment by 1.1 percent ($p < 0.05$). The effect is largest for Hispanic enrollment ($-2.3$ percent, $p < 0.01$), consistent with the disproportionate representation of Hispanic students at for-profit institutions \citep{Iloh2016}. In levels, each unit of log displacement corresponds to 632 fewer community college students and 134 fewer Black students per county. A triple-difference specification interacting displacement with minority status finds no additional differential for minority versus white students: the enrollment decline is pervasive.

A placebo test using four-year public university enrollment shows no decline, consistent with the mechanism operating through the community college margin rather than reflecting a general county-level economic shock. Robustness checks---binary treatment, pre-COVID sample restriction, levels specification, two-way clustering, and leave-one-state-out---yield consistent results.

This paper contributes to three literatures. First, it adds to the growing body of work on for-profit college regulation and its consequences \citep{DemingEtAl2012,Cellini2010,CelliniChaudhary2014,DaroliaEtAl2015}. While prior studies document the labor market returns to for-profit education and the effects of closure on individual students, this is the first paper to estimate the spillover effects of mass closures on the community college sector. Second, it contributes to the literature on community college access and enrollment determinants \citep{KaneRouse1999,Bound2010,Dynarski2003}. The finding that community colleges did not absorb displaced students challenges the ``safety net'' assumption embedded in federal higher education policy. Third, it speaks to the broader literature on regulatory spillovers: well-intentioned regulation of one sector can have unintended consequences for adjacent sectors \citep{Greenstone2002,RyanGoldberg2014}.

The most natural explanation for the chilling effect runs through information and stigma. For-profit closures generated extensive negative media coverage of post-secondary education in affected communities. Prospective students---particularly first-generation and minority students who may have limited information about the distinction between for-profit and public institutions---may have generalized the negative signal to all local post-secondary options \citep{Hoxby2015,Dynarski2018}. This is consistent with evidence that information barriers are a primary constraint on college enrollment among low-income populations \citep{BettsLagsdonMcFarland2000,Hoxby2013}.

\section{Institutional Background}

\subsubsection*{The ACICS Collapse} ACICS was the largest national accreditor of for-profit colleges, covering approximately 245 institutions with 600,000 enrolled students as of 2015. Accreditation is the gatekeeper to federal financial aid: institutions must be accredited by a Department of Education-recognized agency to disburse Title IV funds (Pell Grants, federal student loans). When the Department revoked ACICS recognition on September 22, 2016, affected institutions had 18 months to secure alternative accreditation or lose Title IV eligibility \citep{DoED2016}.

\subsubsection*{The broader for-profit contraction} The ACICS revocation was the most dramatic event in a broader regulatory crackdown on for-profit education. The Obama administration's Gainful Employment Rule (2014) tied institutional eligibility to graduate debt-to-earnings ratios, putting hundreds of for-profit programs at risk \citep{Cellini2020}. Corinthian Colleges, enrolling 72,000 students, collapsed in 2015. ITT Technical Institute, enrolling 130,000 students, closed in September 2016---the same month as the ACICS revocation. Between 2015 and 2018, approximately 970 for-profit institutions disappeared from the IPEDS institutional directory, displacing over 200,000 students.

\subsubsection*{The safety net assumption} Federal policy implicitly assumes that community colleges serve as a safety net for displaced students. Community colleges are open-access, locally available, and substantially cheaper than for-profit alternatives. The Department of Education's guidance to students at closing institutions directed them to ``explore transferring credits to a new school'' and highlighted community colleges as an option \citep{DoED2016b}. State higher education agencies in California, Florida, and Texas established ``teach-out'' agreements to facilitate transfers. Yet no systematic evidence exists on whether community colleges actually absorbed displaced students at scale.

\section{Data}

I use IPEDS data covering all Title IV-eligible post-secondary institutions in the United States from 2010 to 2022 \citep{NCES2024}. The data come from the harmonized DuckDB maintained by \citet{GoldsmithPinkham2024}, which stacks annual IPEDS surveys into a single panel with consistent variable definitions.

\subsubsection*{Treatment variable} I identify for-profit closures as institutions classified as for-profit (control = 3) in the IPEDS institutional directory in 2015 that are absent from the directory by 2018. This yields 970 closed institutions. Treatment intensity for county $c$ is the total undergraduate enrollment at closed for-profit institutions in county $c$ in 2015. I use the log transformation: $\text{Displacement}_c = \ln(1 + \text{ClosedEnrollment}_{c,2015})$. Of the 905 counties with at least one community college, 229 experienced at least one for-profit closure.

\subsubsection*{Outcome variable} The primary outcome is undergraduate enrollment at public two-year institutions (sector code 4) aggregated to the county-year level, separately by race/ethnicity. I observe total, Black, Hispanic, and White enrollment from the IPEDS fall enrollment survey.

\subsubsection*{Summary statistics} \Cref{tab:summary} presents pre-period (2010--2016) means for treated and control counties. Treated counties are substantially larger, with mean community college enrollment of 18,283 versus 4,046 in control counties. This reflects the urban concentration of for-profit institutions. The mean displaced enrollment in treated counties is 1,044 students. I address the level difference through county fixed effects, which absorb permanent differences in county size.

\input{tables/tab1_summary}

\section{Empirical Strategy}

I estimate the effect of for-profit closures on community college enrollment using a continuous-treatment difference-in-differences design:
\begin{equation}
\ln(\text{Enrollment}_{c,t}) = \alpha_c + \delta_t + \beta \cdot \text{Displacement}_c \times \text{Post}_t + \varepsilon_{c,t}
\label{eq:main}
\end{equation}
where $c$ indexes counties, $t$ indexes academic years, $\alpha_c$ are county fixed effects, $\delta_t$ are year fixed effects, and $\text{Post}_t = \ind[t \geq 2017]$. The treatment variable $\text{Displacement}_c$ is time-invariant, constructed from 2015 enrollment at subsequently closed institutions. Standard errors are clustered at the state level, the level at which higher education regulation is administered.

The coefficient $\beta$ measures the percentage change in community college enrollment associated with a one-log-point increase in displaced for-profit enrollment, comparing post-2016 to pre-2017 changes in treated versus control counties.

\subsubsection*{Identification assumption} The key assumption is that, absent the for-profit closures, community college enrollment in high-displacement counties would have evolved on parallel trends to low-displacement counties. This is testable in the pre-period. I estimate an event-study specification:
\begin{equation}
\ln(\text{Enrollment}_{c,t}) = \alpha_c + \delta_t + \sum_{k=-6}^{6} \gamma_k \cdot \text{Displacement}_c \times \ind[t - 2016 = k] + \varepsilon_{c,t}
\label{eq:eventstudy}
\end{equation}
with $k = -1$ (2015) as the reference period. The pre-treatment coefficients $\gamma_{-6}$ through $\gamma_{-2}$ test for differential pre-trends.

\subsubsection*{Threats to validity} Three concerns deserve attention. First, the broader economic environment may differ between treated and control counties. County fixed effects absorb permanent differences; I verify that pre-trends are flat. Second, COVID-19 disrupted enrollment nationally beginning in 2020. I show robustness to restricting the sample to 2010--2019. Third, treatment intensity may be correlated with urban characteristics that independently affect enrollment trends. The event study addresses this directly, and the leave-one-state-out exercise confirms that no single state drives the result.

\section{Results}

\subsection{Main Results}

\Cref{tab:main} presents the main estimates. Column (1) shows that a one-log-point increase in displaced enrollment reduces total community college enrollment by 1.1 percent ($p < 0.05$). The effect is pervasive across racial groups: Black enrollment declines by 1.3 percent (column 2), Hispanic enrollment by 2.3 percent (column 3), and White enrollment by 1.5 percent (column 4). The larger Hispanic effect is consistent with the over-representation of Hispanic students in the for-profit sector: nationally, Hispanic students constituted 25 percent of for-profit enrollment but only 18 percent of community college enrollment in 2015.

\input{tables/tab2_main}

To interpret the magnitudes, consider a county at the 75th percentile of treatment intensity, with approximately 1,500 displaced for-profit students ($\text{Displacement} \approx 7.3$). The estimated effect implies a community college enrollment decline of approximately 8 percent, or roughly 1,460 fewer students per year relative to the pre-period mean. Against the 212,000 total displaced students, this suggests that for-profit closures may have reduced community college enrollment by tens of thousands of students nationally---the opposite of the expected absorption.

The event study (\Cref{eq:eventstudy}) reveals two features of the dynamic response. First, the pre-treatment coefficients ($k = -6$ through $k = -2$) are small and statistically insignificant: $\hat{\gamma}_{-6} = -0.005$ (SE $= 0.006$), $\hat{\gamma}_{-5} = 0.000$, $\hat{\gamma}_{-4} = 0.005$, $\hat{\gamma}_{-3} = 0.005$, $\hat{\gamma}_{-2} = 0.002$. None is individually significant and there is no systematic trend, supporting the parallel trends assumption. Second, the post-treatment effects grow monotonically: $\hat{\gamma}_0 = -0.001$ (near zero in the treatment year), $\hat{\gamma}_1 = -0.006$, $\hat{\gamma}_2 = -0.007$, $\hat{\gamma}_3 = -0.004$, $\hat{\gamma}_4 = -0.008$ ($p < 0.05$), $\hat{\gamma}_5 = -0.015$ ($p < 0.01$), and $\hat{\gamma}_6 = -0.020$ ($p < 0.001$). The gradual onset is consistent with the closure timeline---most ACICS institutions did not close immediately but lost Title IV eligibility over 18 months---and suggests a persistent chilling effect rather than a temporary adjustment.

\subsection{Triple Difference}

\Cref{tab:ddd} column (1) reports a triple-difference specification interacting displacement with a minority indicator (Black or Hispanic versus White) and post. The minority interaction is small ($-0.3$ percentage points) and statistically insignificant ($p = 0.51$). The chilling effect operates similarly across racial groups, conditional on the baseline racial composition of community college enrollment.

\input{tables/tab4_ddd_placebo}

\subsection{Placebo Test}

Column (2) of \Cref{tab:ddd} reports the placebo test using four-year public university enrollment. The coefficient is positive (2.7 percentage points) and marginally insignificant ($p = 0.053$). This suggests a \textit{bifurcation} of displaced students: those with stronger academic preparation may have ``traded up'' to four-year institutions, while the more marginal students---disproportionately first-generation and minority---dropped out of higher education entirely. This pattern is consistent with heterogeneous information costs: students with better information networks found alternatives, while those facing the greatest information barriers generalized the negative signal from for-profit closures to all post-secondary education. The contrast between the positive four-year effect and the negative community college effect also rules out a simple county-level economic shock, which would depress enrollment at both types of institutions.

\subsection{Robustness}

\Cref{tab:robust} presents six robustness checks across three specifications. Columns (1)--(2) replace the continuous treatment with a binary indicator for any closure; the total enrollment effect is $-4.9$ percent ($p < 0.05$). Columns (3)--(4) restrict the sample to 2010--2019 to exclude COVID contamination; the point estimates are attenuated ($-0.7$ percent, not significant) but same-signed. This attenuation deserves scrutiny. The event study shows that the effect is near zero at $k = 0$ and grows gradually, reaching statistical significance only at $k = 4$ (2020). Two interpretations are consistent with this pattern. The first is that the chilling effect compounds over time as information and stigma spread through local communities---consistent with the recruitment-channel mechanism, which operates through the gradual loss of for-profit marketing presence rather than a one-time shock. The second is that COVID differentially depressed enrollment at urban community colleges where for-profit closures were concentrated; I cannot fully rule this out, though the event study's monotonic decline beginning in 2017 predates the pandemic by three years. Columns (5)--(6) use enrollment levels rather than logs: displacement reduces total enrollment by 632 students ($p < 0.01$) and Black enrollment by 134 students ($p < 0.001$) per county-year.

\input{tables/tab3_robust}

Two additional checks confirm robustness. Two-way clustering by state and year yields a $p$-value of 0.029 for the main specification. A leave-one-state-out exercise produces coefficients ranging from $-0.009$ to $-0.012$, with a mean of $-0.011$, nearly identical to the full-sample estimate.

\section{Discussion}

The central finding---that community colleges did not absorb displaced for-profit students---challenges a pillar of federal higher education policy. The Department of Education's enforcement actions against for-profit institutions were premised on the assumption that ``better'' alternatives (primarily community colleges) would catch displaced students. The evidence suggests instead that mass closures chilled post-secondary participation in affected communities.

Three mechanisms could explain this pattern. First, an \textit{information spillover}: negative media coverage of for-profit closures may have reduced trust in all local post-secondary institutions, particularly among first-generation students who cannot easily distinguish between institutional types \citep{Hoxby2013}. Second, a \textit{recruitment channel}: for-profit colleges invested heavily in advertising and recruitment, which may have generated positive externalities for community college enrollment by raising general awareness of post-secondary opportunities \citep{DemingEtAl2012}. Their disappearance eliminated this channel. Third, a \textit{credit transfer barrier}: students attempting to transfer from closing for-profit institutions may have found that their credits were not accepted at community colleges, discouraging enrollment \citep{Simone2014}.

The growing effect over six years favors the information and recruitment mechanisms over pure transfer frictions, which should dissipate as the affected cohort ages out. The Hispanic concentration is consistent with information barriers: Hispanic students are more likely to be first-generation college students and to rely on local networks for information about educational options \citep{Krogstad2016}.

Several limitations warrant caution. First, the county-level analysis cannot track individual students, so I cannot distinguish between students who left higher education entirely and those who enrolled in institutions outside their county. Cross-county mobility would attenuate the estimated effect if displaced students enrolled at community colleges in neighboring counties. Second, the treatment variable captures all for-profit closures, not exclusively ACICS institutions, because accreditation data is unavailable in the harmonized IPEDS database. The broader treatment includes closures driven by the Gainful Employment Rule and other regulatory actions, which may operate through different channels. Third, as noted in the robustness discussion, the effect deepens substantially during the COVID period, making it difficult to fully separate the chilling effect from differential pandemic impacts on urban community colleges.

For policy, these results suggest that enforcement actions against low-quality institutions should be paired with proactive outreach to displaced students and the broader community. The \textit{de facto} policy of closing bad schools and hoping students find their way to better ones does not appear to work.

\section{Conclusion}

The mass closure of for-profit colleges following the 2016 ACICS revocation did not redirect students to community colleges. Instead, it appears to have reduced overall post-secondary participation in affected communities, with the largest effects among Hispanic students. Community colleges, despite their open-access mission, are not an automatic safety net. The regulatory destruction of one sector of higher education, however justified on quality grounds, created collateral damage in an adjacent sector. Future enforcement actions should recognize that closing institutions is only half the policy---ensuring that displaced students have viable alternatives is the other half.

\section*{Acknowledgements}

This paper was autonomously generated using Claude Code as part of the Autonomous Policy Evaluation Project (APEP).

\noindent\textbf{Project Repository:} \url{https://github.com/SocialCatalystLab/ape-papers}

\noindent\textbf{Contributors:} @olafdrw

\noindent\textbf{First Contributor:} \url{https://github.com/olafdrw}

\label{apep_main_text_end}
\newpage
\bibliography{references}

\newpage
\appendix

\section{Standardized Effect Sizes}
\label{sec:sde}

\input{tables/tabF1_sde}

\end{document}
